# Constructs vs. Variables {#sec-construct-vs-variable}

Almost everything marketing scientists care about is invisible. Brand equity,
customer satisfaction, perceived quality, purchase intention, trust, involvement,
loyalty—none of these can be read off an instrument the way a thermometer reads
temperature. They are *constructs*: abstractions the research community invents,
names, and agrees to treat as if they were real, precisely because doing so
organizes a sprawl of otherwise disconnected observations. What we actually
record—a click, a seven-point rating, a repeat purchase, a star count—are
*variables*: observed quantities standing in for the construct we cannot see.
The entire empirical enterprise of marketing rests on the slender bridge between
the two, and most of what goes wrong in applied research goes wrong on that
bridge.

This chapter is about that bridge. It develops, formally and with reproducible
code, the distinction between a construct and the variables that measure it; the
act of *operationalization* that connects them; and the two properties—*validity*
and *reliability*—that determine whether the connection is any good. These ideas
are foundational in the literal sense: every later chapter that estimates a brand's
worth (@sec-branding), models a response function, or runs a regression is
implicitly trusting that its left- and right-hand-side variables measure the
constructs the theory is about. When that trust is misplaced, no amount of
econometric sophistication downstream can repair it. Measurement error is not a
nuisance to be cleaned up after the interesting modeling; for latent constructs it
is the modeling.

By the end of the chapter the reader should be able to: state what a latent
construct is and write down its measurement model; distinguish *reflective* from
*formative* operationalizations and explain why the choice changes which validity
tests apply; decompose an observed score into true score and error and derive
reliability from that decomposition; estimate reliability (coefficient alpha, the
intraclass correlation, Cohen's $\kappa$) and the components of validity
(convergent, discriminant, nomological) from data; and recognize the
*construct-to-measure gap*—the slippage between what we mean and what we
record—in its many disguises. We lead with intuition and follow immediately with
the formalism, because both audiences for this book need both.

## Constructs, Variables, and the Gap Between Them

A **variable** is anything that varies across the units of analysis and can be
recorded: it has a name, a measurement procedure, and a set of admissible values.
A respondent's age, the price paid, whether a coupon was redeemed, the number on a
Likert item—all are variables. Variables are *operational*: their meaning is
exhausted by the procedure that produces them. There is no ambiguity about what "the
integer the respondent circled on item 3" means once the questionnaire is fixed.

A **construct** is a different kind of object. It is a conceptual term,
deliberately constructed by researchers, that denotes a property of the unit which
is *not* directly observable and whose meaning is given by a theoretical
definition rather than a measurement procedure. "Customer satisfaction" is not
identical to any single rating; it is the postulated psychological state that, the
theory says, *causes* people to give high ratings, recommend the product, and buy
again. The construct lives in the theory; the variables live in the dataset.

::: {.callout-note}
The vocabulary is not standardized across fields. Psychometrics says *latent
variable*; econometrics says *unobservable* or *latent factor*; structural-equation
modeling says *factor* or *construct*; classical test theory says *true score*. We
use **construct** for the theoretical concept, **latent variable** $\eta$ for its
formal representation inside a model, and **indicator**, **measure**, or **manifest
variable** $x$ for the observed quantity that proxies it. The relationships are the
same regardless of dialect.
:::

The distinction matters because constructs and variables fail to coincide in
predictable, consequential ways. Call the slippage the **construct-to-measure
gap**: the difference between the construct as theoretically intended and the
construct as actually captured by the chosen measure. The gap has at least three
faces.

First, **deficiency**: the measure omits part of the construct. A satisfaction
scale that asks only about product quality misses satisfaction with price, service,
and delivery; it is *construct-deficient*. Second, **contamination**: the measure
captures things outside the construct. A self-report of "how healthy is your diet"
is contaminated by social-desirability bias—respondents report the construct *plus*
their desire to look good. Third, **distortion**: the measure relates to the
construct nonlinearly or non-monotonically, so that equal differences in the
measure do not correspond to equal differences in the construct. @fig-cm-gap
diagrams the three failure modes against the ideal of a measure that maps cleanly
onto its construct.

```{mermaid}
%%| label: fig-cm-gap
%%| fig-cap: "The construct-to-measure gap. The intended construct (left) and the recorded measure (right) overlap only partially. Region D is construct deficiency (intended but unmeasured); region C is contamination (measured but unintended); the intersection V is valid variance. Distortion warps the mapping even within V."
flowchart LR
  subgraph Construct["Intended construct"]
    D["D: deficiency<br/>(meant, not measured)"]
    V1["V: valid<br/>overlap"]
  end
  subgraph Measure["Recorded measure"]
    V2["V: valid<br/>overlap"]
    C["C: contamination<br/>(measured, not meant)"]
  end
  V1 --- V2
```

No measure is perfectly free of all three faults; the practical question is whether
the valid overlap $V$ is large enough, and the contamination $C$ small and
*random* enough, for the measure to support the inferences the study wants to draw.
Making that question precise is the work of measurement theory.

> "A construct is some postulated attribute of people, assumed to be reflected in
> test performance." A construct is admitted into a science not because it can be
> observed but because a network of theoretical relations—a *nomological
> network*—gives it empirical consequences that can be checked.[^cronbachmeehl]

[^cronbachmeehl]: This is the central idea of construct validation in the
classical psychometric tradition: a construct earns its keep through the lawful
relations it enters into, not through any single observation.
<!-- CITE-NEEDED: Cronbach & Meehl (1955), "Construct validity in psychological tests," Psychological Bulletin -->

## Latent Constructs and the Measurement Model

To reason formally we need to represent a construct inside a model. Let $\eta$
(eta) denote the **latent variable** standing for the construct for a generic unit,
and let $\mathbf{x} = (x_1,\dots,x_K)^\top$ be the vector of $K$ observed
indicators intended to measure it. A **measurement model** is a statement about how
$\mathbf{x}$ and $\eta$ are related. Two such statements dominate marketing, and
choosing between them is the single most consequential measurement decision a
researcher makes [@coltman2008; @diamantopoulos2013quantitative].

### Reflective Measurement

In a **reflective** (effect-indicator) model the construct is a *common cause* of
its indicators: each indicator is a noisy manifestation of the same underlying
$\eta$. The classic linear, unidimensional form is

$$
x_k = \tau_k + \lambda_k\,\eta + \varepsilon_k, \qquad k = 1,\dots,K,
$$ {#eq-reflective}

where $\lambda_k$ is the **loading** (how strongly indicator $k$ reflects the
construct), $\tau_k$ an intercept, and $\varepsilon_k$ a measurement error
satisfying $\mathbb{E}[\varepsilon_k]=0$, $\mathrm{Cov}(\eta,\varepsilon_k)=0$, and
$\mathrm{Cov}(\varepsilon_j,\varepsilon_k)=0$ for $j\neq k$ (uncorrelated errors,
the *local-independence* assumption). The arrows in a path diagram point *from*
$\eta$ *to* each $x_k$, as in @fig-measurement-models.

Three properties follow immediately and are diagnostic of the reflective view.
Because all indicators share the single common cause $\eta$, they should be
positively **intercorrelated**; because they are interchangeable expressions of the
same thing, dropping one should not change the *meaning* of the construct, only the
precision with which it is measured; and because each carries information about the
same $\eta$, **internal consistency** (do the items hang together?) is the relevant
reliability question. The price premium, perceived quality, and loyalty items of a
brand-equity scale are typically modeled this way [@yoo2001; @keller1993].

### Formative Measurement

In a **formative** (causal-indicator or composite) model the arrows reverse: the
indicators are *causes* or *components* that together compose the construct,

$$
\eta = \sum_{k=1}^{K} w_k\,x_k + \zeta,
$$ {#eq-formative}

with weights $w_k$ and a disturbance $\zeta$ capturing the part of $\eta$ not
accounted for by the chosen indicators. Here the indicators need *not* be
correlated—they describe different facets—and they are emphatically *not*
interchangeable: drop one and the construct itself changes. "Socioeconomic status"
formed from income, education, and occupation is the textbook case; in marketing,
many "marketing-mix exposure," "deprivation," or driver-based indices are formative,
and whether brand equity is reflective or formative is a live and unsettled dispute
(@sec-branding) [@henseler2017; @coltman2008].

::: {.callout-warning}
Misspecifying the direction is not a stylistic slip—it biases structural estimates
and invalidates the usual reliability and validity diagnostics. Applying
internal-consistency tests (which *expect* correlated indicators) to a formative
index will reject perfectly good causal indicators for the "crime" of being
uncorrelated; conversely, applying formative weighting to reflective items throws
away the information that their shared variance is the construct
[@diamantopoulos2013quantitative; @bendle2018misuse].
:::

The two models are not nested and imply different psychometrics, summarized in
@tbl-reflective-formative. The decision is theoretical, not empirical: it is
settled by asking whether the construct *produces* its indicators (reflective) or
is *defined by* them (formative), a question data alone cannot answer
[@coltman2008].

| Property | Reflective (@eq-reflective) | Formative (@eq-formative) |
|---|---|---|
| Causal direction | construct $\rightarrow$ indicators | indicators $\rightarrow$ construct |
| Indicator intercorrelation | high, expected | not required |
| Interchangeability | items interchangeable | items not interchangeable |
| Effect of dropping an item | precision only | construct meaning changes |
| Reliability logic | internal consistency ($\alpha$, $\omega$, CR) | n/a; check indicator weights & VIF |
| Validity logic | convergent + discriminant + nomological | nomological + indicator validity |
| Error attaches to | each indicator ($\varepsilon_k$) | the construct ($\zeta$) |

: Reflective vs. formative measurement models and their consequences for assessment. {#tbl-reflective-formative}

```{mermaid}
%%| label: fig-measurement-models
%%| fig-cap: "Reflective vs. formative measurement. In the reflective model the latent construct η is a common cause of its indicators (arrows out, errors on each indicator). In the formative model the indicators compose the construct (arrows in, error on the construct)."
flowchart LR
  subgraph R["Reflective"]
    direction TB
    eta1(("η")) --> xr1["x₁"]
    eta1 --> xr2["x₂"]
    eta1 --> xr3["x₃"]
    e1>"ε₁"] --> xr1
    e2>"ε₂"] --> xr2
    e3>"ε₃"] --> xr3
  end
  subgraph F["Formative"]
    direction TB
    xf1["x₁"] --> eta2(("η"))
    xf2["x₂"] --> eta2
    xf3["x₃"] --> eta2
    z>"ζ"] --> eta2
  end
```

## Operationalization: From Concept to Instrument

**Operationalization** is the act of specifying the procedure that turns a
construct into a variable—the rule that says "to measure perceived quality, ask
these four questions on these scales and average them." It is where theory makes
contact with data, and it is irreversible in the sense that downstream analysis can
never recover construct meaning that the operationalization failed to capture.

The classical paradigm for developing a reflective scale proceeds through a
disciplined sequence: (1) **specify the domain** of the construct from theory; (2)
**generate an item pool** broad enough to cover the domain (guarding against
deficiency); (3) **purify** the items by collecting data and discarding those that
correlate poorly with the rest; (4) **assess reliability** on a fresh sample; and
(5) **assess validity** against external criteria and related constructs.[^churchill]
The logic is iterative because operationalization is fallible: the first item pool
is almost always both deficient and contaminated, and only data exposes which items
do which.

[^churchill]: This sequence is the backbone of scale development in marketing; the
purification step in particular—iteratively pruning items by item-to-total
correlation and re-estimating reliability—originates in the marketing measurement
literature of the late 1970s.
<!-- CITE-NEEDED: Churchill (1979), "A paradigm for developing better measures of marketing constructs," JMR -->

A single construct can be operationalized in radically different ways, and the
choice carries assumptions. Customer satisfaction can be measured by a
*self-reported* rating (cheap, but contaminated by mood and response styles), by
*revealed behavior* such as repurchase (objective, but deficient—repurchase also
reflects switching costs and availability), or by a *physiological* or *textual*
signal mined from reviews (scalable, but distortion-prone). Each operationalization
trades the three faces of the gap differently; none is uniquely correct, and
reporting which one was used is part of reporting the result. A recurring error in
applied work is to treat a convenient managerial *metric* (e.g., a single
top-box satisfaction percentage, or a Net Promoter score) as if it were the
construct itself, when it is one deficient operationalization among many
[@bendle2018; @bendle2018misuse].

## Classical Test Theory and the Anatomy of a Score

To define reliability we need a model of what an observed score *is*. **Classical
test theory** (CTT) supplies the simplest one. It decomposes an observed score $X$
for a single indicator into a **true score** $T$ and an **error** $E$:

$$
X = T + E, \qquad T \equiv \mathbb{E}[X], \qquad \mathbb{E}[E] = 0, \qquad
\mathrm{Cov}(T, E) = 0.
$$ {#eq-ctt}

The true score is *defined* as the expected value of the observed score over
hypothetical independent replications of the measurement on the same unit—not as
the Platonic construct value, a subtlety we return to under validity. Error is
*random*: by construction it is mean-zero and uncorrelated with the true score.
Because $T$ and $E$ are uncorrelated, the observed variance partitions cleanly,

$$
\sigma_X^2 = \sigma_T^2 + \sigma_E^2 .
$$ {#eq-var-decomp}

**Reliability** $\rho_{XX'}$ is then defined as the share of observed variance that
is true-score variance,

$$
\rho_{XX'} \;=\; \frac{\sigma_T^2}{\sigma_X^2} \;=\; \frac{\sigma_T^2}{\sigma_T^2 + \sigma_E^2}
\;=\; \mathrm{Corr}(X, X')^2\ \text{-free form}\ = \mathrm{Corr}(X, X'),
$$ {#eq-reliability}

where $X'$ is a *parallel* measurement (same $T$, same $\sigma_E^2$, independent
error). The last equality is the operational content of the definition: reliability
equals the correlation between two parallel measurements of the same thing, which
is why test–retest and split-half correlations estimate it. Reliability ranges from
0 (all noise) to 1 (no error); it has no units and is a property of *scores in a
population*, not of an instrument in the abstract.

Two immediate consequences shape everything downstream. First, **attenuation**:
unreliability in a predictor biases its estimated effect *toward zero*. If a true
relationship has correlation $\rho_{TY}$ but $X$ measures $T$ with reliability
$\rho_{XX'}$, the observable correlation is

$$
\rho_{XY} = \rho_{TY}\,\sqrt{\rho_{XX'}\,\rho_{YY'}},
$$ {#eq-attenuation}

so noisy measurement systematically *understates* effects—a fact that makes
"we found no effect" uninterpretable without a reliability figure. Second,
**aggregation buys reliability**: averaging $K$ parallel indicators leaves the
true score intact but shrinks error variance by $1/K$, so the reliability of a
$K$-item sum rises with $K$ via the Spearman–Brown relation
$\rho_K = K\rho_1 / [1 + (K-1)\rho_1]$. This is the formal reason multi-item scales
beat single items, and why @eq-reflective is worth the trouble.

## Reliability in Practice

Reliability is estimated, not known. The estimator depends on which replication the
data approximate.

### Internal Consistency: Coefficient Alpha

The most reported estimate is **coefficient alpha** (Cronbach's $\alpha$), which
treats the $K$ items of one administration as approximately parallel replications.
With $\sigma_i^2$ the variance of item $i$ and $\sigma_X^2$ the variance of the
total score $X=\sum_i x_i$,

$$
\alpha = \frac{K}{K-1}\left(1 - \frac{\sum_{i=1}^{K}\sigma_i^2}{\sigma_X^2}\right).
$$ {#eq-alpha}

Alpha is a *lower bound* on reliability under the assumption of (essentially)
$\tau$-equivalent items—equal loadings, @eq-reflective with all $\lambda_k$ equal.
When loadings differ, $\alpha$ underestimates reliability, and McDonald's $\omega$
(which uses the estimated loadings directly) is preferable. Two cautions are worth
stating plainly: $\alpha$ rises mechanically with $K$, so a high $\alpha$ can simply
mean "many items," not "good items"; and $\alpha$ presupposes a *reflective,
unidimensional* construct—it is meaningless for a formative index
(@tbl-reflective-formative).[^alpha]

[^alpha]: A common abuse is to report $\alpha$ for a deliberately multidimensional
battery and conclude the battery is "reliable." Alpha conflates dimensionality with
consistency; a factor analysis must establish unidimensionality first.

```{r}
#| message: false
#| warning: false
#| label: alpha-demo
#| fig-cap: "Simulated four-item reflective scale: true score plus item-specific error."
set.seed(2718)

n  <- 500                       # respondents
K  <- 4                         # reflective indicators
lambda <- c(0.9, 0.8, 0.7, 0.6) # item loadings

eta   <- rnorm(n)                                   # latent construct (standardized)
err   <- sapply(sqrt(1 - lambda^2),                 # error sd from loadings
                function(s) rnorm(n, sd = s))
X     <- sweep(err, 2, lambda, `+`) +               # x_k = lambda_k * eta + e_k
         outer(eta, lambda)

# Coefficient alpha from equation (eq-alpha), implemented directly
cronbach_alpha <- function(items) {
  k <- ncol(items)
  item_var  <- sum(apply(items, 2, var))
  total_var <- var(rowSums(items))
  (k / (k - 1)) * (1 - item_var / total_var)
}

alpha_hat <- cronbach_alpha(X)
round(alpha_hat, 3)
```

### Stability and Agreement: ICC and Cohen's $\kappa$

When the replication is across *time* (test–retest) or across *raters*, reliability
is an **intraclass correlation** (ICC): the share of total variance attributable to
stable between-unit differences rather than to occasion- or rater-specific noise.
For a one-way model with unit effect $u_i$ and residual $e_{ij}$, the ICC is
$\sigma_u^2 / (\sigma_u^2 + \sigma_e^2)$, the same true-over-total ratio as
@eq-reliability, now estimated from a variance-components decomposition. The
forms—single vs. average rater, consistency vs. absolute agreement—must be matched
to the design, a distinction made canonical for the field by @Shrout_1979.

For *categorical* judgments (does this review express complaint? is this ad
emotional?), correlation is inappropriate and the relevant statistic is **Cohen's
$\kappa$**, which corrects observed agreement $p_o$ for the agreement
$p_e$ expected by chance:

$$
\kappa = \frac{p_o - p_e}{1 - p_e}.
$$ {#eq-kappa}

A $\kappa$ of 0 means raters agree only as often as random labeling would predict;
1 means perfect agreement [@Cohen_1960]. The widely used verbal benchmarks—slight,
fair, moderate, substantial, almost perfect—come from @Landis_1977, though they are
conventions, not laws. The example below estimates both an ICC for ratings and a
$\kappa$ for binary codings on simulated data.

```{r}
#| message: false
#| warning: false
#| label: icc-kappa

set.seed(1414)

## ---- Intraclass correlation (two raters, continuous ratings) -------------
n  <- 200
true_q <- rnorm(n, mean = 0, sd = 1)        # stable latent quality of object i
r1 <- true_q + rnorm(n, sd = 0.6)            # rater 1
r2 <- true_q + rnorm(n, sd = 0.6)            # rater 2

# One-way ICC via variance components (between-object / total)
ms_between <- 2 * var(rowMeans(cbind(r1, r2)))           # 2 raters per object
ms_within  <- mean((r1 - r2)^2) / 2
icc <- (ms_between - ms_within) / (ms_between + ms_within)
round(icc, 3)

## ---- Cohen's kappa (two coders, binary labels) ---------------------------
latent  <- rnorm(n)
code1 <- as.integer(latent + rnorm(n, sd = 0.7) > 0)   # coder 1
code2 <- as.integer(latent + rnorm(n, sd = 0.7) > 0)   # coder 2

tab <- table(code1, code2)
p_o <- sum(diag(tab)) / sum(tab)                        # observed agreement
p_e <- sum(rowSums(tab) * colSums(tab)) / sum(tab)^2    # chance agreement
kappa <- (p_o - p_e) / (1 - p_e)                        # equation (eq-kappa)
round(c(p_o = p_o, p_e = p_e, kappa = kappa), 3)
```

High reliability is necessary but not sufficient. A bathroom scale that reads three
kilograms heavy is perfectly reliable and perfectly wrong: it returns the same
number every time (no error variance) while measuring the wrong true score. The
missing property is validity.

## Validity: Are We Measuring the Right Thing?

**Validity** asks whether a measure captures the construct it claims to capture—
whether the true score $T$ in @eq-ctt actually equals (or tracks) the construct
$\eta$, rather than some neighboring or contaminating quantity. Modern measurement
theory treats validity not as a property a measure simply has, but as a body of
*evidence* accumulated for a particular interpretation and use. Several
complementary strands of evidence are standard.

**Content validity** is the (largely judgmental) question of whether the item pool
covers the construct's domain without straying outside it—directly the
deficiency/contamination trade-off of @fig-cm-gap, assessed by expert review before
any data are collected.

**Convergent validity** asks whether indicators that *should* measure the same
construct actually correlate. In a reflective model it is read off the loadings:
high, significant $\lambda_k$ mean the indicators share the common cause. A standard
summary is the **average variance extracted** (AVE), the mean squared standardized
loading,

$$
\mathrm{AVE} = \frac{1}{K}\sum_{k=1}^{K}\lambda_k^2 ,
$$ {#eq-ave}

interpreted as the share of indicator variance the construct explains; $\mathrm{AVE}
\geq 0.5$ (the construct explains more than half the indicator variance) is the
conventional threshold [@fornell1981a].

**Discriminant validity** asks the reverse: whether a construct is empirically
*distinct* from other constructs it is not supposed to be identical to. The classic
@fornell1981a criterion requires a construct's AVE to exceed its squared correlation
with every other construct—each construct shares more variance with its own
indicators than with any other construct. (More recent work argues the
heterotrait–monotrait ratio of correlations detects discriminant-validity problems
the Fornell–Larcker criterion misses, especially when loadings are similar across
constructs.)[^htmt]

[^htmt]: The HTMT criterion compares average cross-construct item correlations to
average within-construct item correlations; values near 1 signal that two
"different" constructs are empirically the same.
<!-- CITE-NEEDED: Henseler, Ringle & Sarstedt (2015), "A new criterion for assessing discriminant validity," JAMS -->

**Nomological (construct) validity** is the deepest and the one that ties back to
the definition of a construct: does the measure relate to *other* constructs the
way theory predicts? A valid satisfaction measure should predict repurchase and
word of mouth, correlate with perceived quality, and be largely unrelated to,
say, respondent handedness. The web of predicted relations is the **nomological
network**, and a measure earns construct validity by behaving as the network says
it should. This is also where validity and reliability finally separate cleanly:
reliability concerns the *consistency* of $X$ around its true score $T$; validity
concerns whether $T$ is the *right* $T$. @fig-validity-types lays out the strands
and the questions they answer.

```{mermaid}
%%| label: fig-validity-types
%%| fig-cap: "Strands of measurement validity and the question each answers. Reliability (left) is logically prior—an unreliable measure cannot be valid—but high reliability does not imply validity."
flowchart TB
  M["Observed measure x"]
  M --> REL["Reliability:<br/>is x consistent<br/>around its true score?"]
  M --> VAL["Validity:<br/>is the true score<br/>the right construct?"]
  VAL --> CON["Content:<br/>does it cover<br/>the domain?"]
  VAL --> CONV["Convergent:<br/>do same-construct<br/>items agree? (AVE)"]
  VAL --> DISC["Discriminant:<br/>is it distinct from<br/>other constructs?"]
  VAL --> NOMO["Nomological:<br/>does it relate to<br/>other constructs<br/>as theory predicts?"]
```

### A Worked Convergent/Discriminant Check

The reflective machinery is concrete. The example below simulates two correlated
latent constructs, each with three reflective indicators, then recovers the loadings
by factor analysis, computes each construct's AVE from @eq-ave, and applies the
Fornell–Larcker discriminant test: AVE must exceed the squared inter-construct
correlation.

```{r}
#| message: false
#| warning: false
#| label: validity-demo

set.seed(3142)
n <- 600

# Two latent constructs with a true correlation of 0.45
Sigma <- matrix(c(1, 0.45, 0.45, 1), 2)
L     <- chol(Sigma)
F     <- matrix(rnorm(n * 2), n, 2) %*% L
colnames(F) <- c("eta_A", "eta_B")

lam_A <- c(0.85, 0.80, 0.75)   # loadings, construct A
lam_B <- c(0.80, 0.78, 0.70)   # loadings, construct B

make_block <- function(eta, lam) {
  sapply(lam, function(l) l * eta + rnorm(length(eta), sd = sqrt(1 - l^2)))
}
A <- make_block(F[, "eta_A"], lam_A)
B <- make_block(F[, "eta_B"], lam_B)
dat <- data.frame(A, B)
names(dat) <- c(paste0("a", 1:3), paste0("b", 1:3))

# Recover loadings with a 2-factor maximum-likelihood EFA
fa <- factanal(dat, factors = 2, rotation = "promax")
load <- fa$loadings[, 1:2]

# Average variance extracted per construct (eq-ave), using the dominant factor
ave_A <- mean(load[1:3, which.max(colSums(abs(load[1:3, ])))]^2)
ave_B <- mean(load[4:6, which.max(colSums(abs(load[4:6, ])))]^2)

# Inter-construct correlation from estimated factor scores
fs <- factanal(dat, factors = 2, rotation = "promax", scores = "regression")$scores
phi2 <- cor(fs[, 1], fs[, 2])^2

cat("AVE construct A:", round(ave_A, 3), "\n")
cat("AVE construct B:", round(ave_B, 3), "\n")
cat("Squared factor correlation:", round(phi2, 3), "\n")
cat("Fornell-Larcker passes (both AVE > shared var)? ",
    ave_A > phi2 && ave_B > phi2, "\n")
```

Both AVEs comfortably exceed 0.5 (convergent validity) and exceed the squared
inter-construct correlation (discriminant validity): the indicators measure their
own construct better than they measure the neighbor. Had we forced the two
constructs to correlate near 1, the discriminant test would fail—the data would be
telling us we had relabeled one construct as two.

## Identification: What Pins a Latent Model Down

A latent variable has no inherent scale—nothing in @eq-reflective fixes the units
or origin of $\eta$, since multiplying $\eta$ by a constant and dividing every
$\lambda_k$ by the same constant leaves all observables unchanged. A measurement
model is **identified** only after this indeterminacy is removed and enough
constraints are imposed that the parameters are recoverable from the covariance
matrix of the indicators. The standard fixes are to *standardize* the latent
variable ($\mathrm{Var}(\eta)=1$) or to fix one loading (a *marker* indicator,
$\lambda_1 = 1$), which sets the scale; the construct's mean is fixed similarly by
an intercept constraint.

Beyond scaling, identification requires *enough indicators*. With a single latent
factor and uncorrelated errors, the covariance structure has $K(K+1)/2$ knowns and
$2K$ unknowns (loadings and error variances) plus one variance; the rule of thumb is
that **at least three indicators** per reflective factor are needed for a single
factor to be just-identified, and two suffice only when the factor is embedded in a
larger model that supplies extra constraints. What *breaks* identification is then
predictable: too few indicators (the model has more parameters than the covariance
matrix can determine); error terms that are in fact correlated (violating local
independence, so shared method variance masquerades as the construct); and—most
relevant to the reflective/formative debate—a **formative** construct that is not
emitted into the model with at least two outgoing paths to other constructs or
indicators, without which @eq-formative has no scale and the weights $w_k$ are not
separately identified [@diamantopoulos2013quantitative; @henseler2017]. The lesson
generalizes: identification failures in latent-variable models are usually a
symptom that the measurement model and the theory disagree about how many things are
being measured.

## Common-Method Variance and Other Contamination

The CTT assumption that errors are *random and uncorrelated* (@eq-ctt,
@eq-reflective) is the one most often violated in practice, and the violation has a
name. **Common-method variance** (CMV) is systematic error shared across indicators
because they were collected by the same *method*—the same respondent, on the same
survey, at the same moment, using the same response scale. CMV induces correlation
among the $\varepsilon_k$, which the reflective model attributes to the construct,
inflating loadings, reliability, and—most damagingly—the correlations *between*
constructs measured by the same method. A satisfaction–loyalty correlation
estimated from a single self-report survey is partly a measurement of the
respondent's consistent way of answering surveys.

CMV is contamination in the precise sense of @fig-cm-gap: variance that is measured
but not meant. It cannot be detected by reliability statistics—shared method
variance *raises* $\alpha$—which is exactly why a high $\alpha$ is no defense.
Procedural remedies (separating the measurement of predictor and outcome in time or
source, varying scale formats, guaranteeing anonymity to reduce
social-desirability pressure) attack the cause; statistical remedies (modeling a
latent method factor that loads on all indicators, or using marker variables
theoretically unrelated to the constructs) attempt to partition method variance from
trait variance after the fact.[^cmv] The deeper point for this chapter is that the
construct-to-measure gap is not a one-time design choice but a property that travels
with the data into every downstream model; an analyst who inherits a single-source
survey inherits its contamination whether or not the original authors named it.

[^cmv]: No statistical fix fully substitutes for designing the contamination out.
The latent-method-factor and marker-variable approaches make assumptions (about
which variance is method variance) that are themselves untestable from the
contaminated data.
<!-- CITE-NEEDED: Podsakoff, MacKenzie, Lee & Podsakoff (2003), "Common method biases in behavioral research," Journal of Applied Psychology -->

## From Measurement Error to Decision Error

Constructs and variables ultimately serve *decisions*: launch or not, target or not,
attribute the lift to the campaign or not. It is worth closing the loop between the
measurement error studied above and the decision error it produces, because the two
are routinely confused. Suppose a measure is used to classify—high-value vs.
low-value customer, satisfied vs. dissatisfied, treatment worked vs. did not. The
construct has a true state; the measure yields a verdict; and the four combinations
form the confusion matrix in @tbl-error-matrix.

| | Construct true | Construct false |
|---|---|---|
| **Measure says true** | correct (true positive) | **Type I error** (false positive) |
| **Measure says false** | **Type II error** (false negative) | correct (true negative) |

: Decision errors induced by an imperfect measure. Reliability and validity shortfalls translate into Type I and Type II error rates; which error dominates depends on where the decision threshold sits relative to the construct distribution. {#tbl-error-matrix}

The connection to the earlier formalism is direct. A measure with low *validity*
shifts the whole verdict distribution off the construct, trading one error type for
the other systematically (a biased scale, like the three-kilo bathroom scale). A
measure with low *reliability* fattens the spread of verdicts around the truth,
raising *both* error rates at once (a noisy scale). And attenuation (@eq-attenuation)
guarantees that unreliable predictors make real effects look smaller than they are,
so the most common decision error in marketing analytics—concluding "no effect"
from a noisy measure—is a measurement failure wearing the costume of a null result.
The discipline of distinguishing construct from variable is, in the end, the
discipline of knowing how much of a number to believe.

## Key Takeaways

- A **construct** is a theoretical concept defined by its role in a nomological
  network; a **variable** is an observed quantity defined by its measurement
  procedure. The **construct-to-measure gap**—deficiency, contamination,
  distortion—is where applied research most often fails (@fig-cm-gap).
- The **measurement model** must be specified as **reflective** (@eq-reflective,
  construct causes indicators) or **formative** (@eq-formative, indicators compose
  construct); the choice is theoretical, not empirical, and it dictates which
  reliability and validity tests even apply (@tbl-reflective-formative)
  [@coltman2008; @diamantopoulos2013quantitative].
- **Reliability** is the true-score share of observed variance (@eq-reliability);
  estimate it by coefficient $\alpha$ (@eq-alpha) for internal consistency, the ICC
  [@Shrout_1979] for raters/occasions, and Cohen's $\kappa$ (@eq-kappa)
  [@Cohen_1960; @Landis_1977] for categorical codings. Unreliability **attenuates**
  estimated effects toward zero (@eq-attenuation).
- **Validity** asks whether the true score is the *right* construct: content,
  convergent (AVE, @eq-ave), discriminant (Fornell–Larcker), and nomological
  evidence together build the case [@fornell1981a]. High reliability does *not*
  imply validity.
- Latent models must be **identified** (scale set, $\geq 3$ reflective indicators,
  local independence); **common-method variance** violates the random-error
  assumption and is invisible to reliability statistics, so it must be designed out,
  not tested away.

## Further Reading

The reflective/formative distinction and its consequences for marketing measurement
are developed in @coltman2008, @diamantopoulos2013quantitative, and @henseler2017;
the misuse of metrics that confuse a managerial number for a construct is dissected
in @bendle2018 and @bendle2018misuse. The classical convergent/discriminant
apparatus traces to @fornell1981a, agreement and rater reliability to @Shrout_1979,
@Cohen_1960, and @Landis_1977. The branding chapter (@sec-branding) shows these
abstractions doing concrete work, where whether brand equity is reflective or
formative remains an open and consequential question.
