# Trust, Commitment, and Loyalty {#sec-relational}

If satisfaction (@sec-satisfaction) is the evaluative residue of a single
consumption episode, the *relational* constructs are what accumulate across
many. Marketing's "relational turn" of the late 1980s reframed exchange not as a
sequence of independent transactions but as an ongoing relationship with a past
and an expected future, and it imported a family of constructs to describe the
glue that holds such relationships together: **trust**, the willingness to rely
on a partner one is confident in; **commitment**, the enduring desire to
maintain a valued relationship; and **loyalty**, the resulting disposition to
keep choosing the same partner despite situational pulls to defect. These three
are the most theorized intangibles in relationship marketing, and they form a
short causal chain—trust and commitment are the psychological mediators, loyalty
the behavioral-attitudinal outcome—whose links have been estimated and
re-estimated for three decades.

This chapter treats them as a system rather than three isolated scales. We open
with the relational turn that made the constructs necessary, then define and
formalize each in turn: trust as a two-dimensional belief, commitment as a
multi-component bond, and loyalty as a two-by-two of attitude and behavior layered
over a four-phase developmental sequence. We then assemble them into the
**commitment–trust theory** of relationship marketing—the field's canonical
mediation model—and show, with reproducible code, how it is identified and
estimated as a structural model of latent constructs. The chapter closes with the
meta-analytic evidence on what drives and what follows from these constructs, and
with the disciplining findings that puncture the naive view that more
relationship is always better. Throughout, the measurement apparatus of
@sec-construct-vs-variable is doing silent work: trust, commitment, and loyalty
are latent constructs measured by reflective indicators, and every estimate below
inherits the reliability and validity problems developed there.

## The Relational Turn {#sec-rel-turn}

The intellectual pivot is @dwyer1987, who argued that marketing had over-invested
in the *discrete transaction*—a clean, anonymous, price-mediated exchange with no
past and no future—and under-theorized *relational exchange*, in which the
parties' history and anticipated future govern present behavior. They mapped a
developmental sequence (awareness, exploration, expansion, commitment,
dissolution) through which a buyer–seller dyad passes, and in doing so they
relocated the object of study from the *act* of purchase to the *relationship*
that produces a stream of purchases. The move had a measurement consequence that
this chapter inherits: once the relationship is the unit of theory, the constructs
worth measuring are the ones that characterize relationships—how much the parties
trust each other, how committed they are, how loyal the buyer is—rather than the
attributes of any single transaction.

The relational turn drew its theoretical spine from **social exchange theory**:
relationships persist when each party's outcomes exceed both their internal
comparison level (what they feel they deserve) and the comparison level of
alternatives (what they could get elsewhere). Trust and commitment are, in this
reading, the psychological residues of a history of exchanges whose outcomes have
repeatedly cleared those bars. The constructs are therefore not free-floating
attitudes but *relationship-specific* states, which is why they are typically
measured with reference to a named partner ("this supplier," "this brand") rather
than in the abstract.

## Trust {#sec-rel-trust}

**Trust** is the willingness of a party to rely on an exchange partner in whom it
has confidence. The most durable conceptualization, from @moorman1992trust and
@moorman1993trust, decomposes it into two beliefs that are routinely
distinguished both theoretically and empirically:

- **Credibility (reliability):** the belief that the partner has the expertise and
  dependability to perform effectively and as promised—a competence judgment.
- **Benevolence:** the belief that the partner is genuinely interested in the
  truster's welfare and will not act opportunistically when new conditions arise
  for which no commitment was made—a motive judgment.

The distinction matters because the two beliefs have different antecedents and
different failure modes. A supplier can be highly credible (it always delivers)
yet low in benevolence (it will exploit any contractual gap), and a partner that
is benevolent but unreliable inspires affection without dependence. @moorman1993trust
define trust operationally as a *willingness to rely* on a partner—deliberately
separating the **belief** (confidence in the partner) from the **behavioral
intention** (willingness to act on that confidence), because a party may trust and
yet decline to rely when the stakes are too high. This separation foreshadows the
identification problem in the commitment–trust model: trust is a latent belief, and
the behaviors we observe (continued reliance, information sharing) are its
consequences, not the construct itself.

### Formalizing reliance under uncertainty {#sec-rel-trust-formal}

Trust does its work precisely when outcomes are uncertain and the partner could
defect. A minimal decision-theoretic statement makes the role of the two beliefs
explicit. Let a truster choose whether to rely ($r = 1$) or not ($r = 0$) on a
partner who may honor the relationship (probability $p$) or exploit it
(probability $1 - p$). Reliance yields gain $G$ if honored and loss $L$ if
exploited; non-reliance yields a safe outcome normalized to $0$. The truster
relies when expected value is positive,

$$
p\,G - (1 - p)\,L > 0 \quad\Longleftrightarrow\quad p > \frac{L}{G + L}.
$$ {#eq-rel-trust}

The two belief dimensions enter through different terms. **Credibility** governs
$p$, the perceived probability the partner is *able* to honor the relationship;
**benevolence** governs the truster's assessment that the partner is *willing* to
forgo the exploitation payoff even when $1 - p$ situations arise. Equation
@eq-rel-trust also exposes why trust substitutes for contracts: raising perceived
$p$ (through reputation, history, or relational norms) lowers the threshold
performance a formal safeguard would otherwise have to guarantee, which is the
economic content of the claim that trust reduces transaction costs. It also shows
trust is *stake-dependent*: as the loss $L$ from betrayal grows relative to the
gain $G$, the required confidence rises, so the same belief supports reliance on
small matters but not large ones—exactly the belief-versus-behavior gap
@moorman1993trust insist on.

## Commitment {#sec-rel-commitment}

**Commitment** is an enduring desire to maintain a valued relationship—a pledge of
relational continuity that, unlike a single satisfied transaction, persists across
episodes and resists short-term temptation. @gundlach1995 give the construct its
standard internal structure, distinguishing three components that are now treated
as conceptually (if not always empirically) separable:

- **Affective (attitudinal) commitment:** an emotional attachment and identification
  with the partner—staying because one *wants* to.
- **Calculative (instrumental) commitment:** a cost-based bond reflecting switching
  costs and the scarcity of alternatives—staying because leaving is *expensive*.
- **Temporal/behavioral commitment:** consistency of the bond over time, evidenced by
  durable inputs and pledges rather than momentary intentions.

@gundlach1995 emphasize that genuine commitment requires *inputs*—credible,
often irreversible investments that signal intent and would be costly to
abandon—so commitment is revealed by what a party stakes, not merely by what it
says. This is the logic of @andersonweitz1992 on **pledges**: idiosyncratic
investments and other self-binding actions build and sustain commitment in
distribution channels precisely because they are hard to reverse, making the
pledging party's continuation credible to the other side. The asymmetry between
the two motives is consequential downstream: affective commitment predicts
advocacy and willingness to sacrifice, whereas purely calculative commitment can
coexist with resentment and evaporates the moment a cheaper alternative appears—a
distinction that reappears in @sec-rel-loyalty as the difference between *true* and
*spurious* loyalty.

## The Commitment–Trust Theory {#sec-rel-kmv}

The synthesis that organizes the field is the **commitment–trust theory** of
@morganhunt1994, universally abbreviated KMV (for the "key mediating variables"
model). Its central claim is a mediation hypothesis: relationship marketing
succeeds not because relational inputs act directly on outcomes, but because they
operate *through* commitment and trust, which are the **key mediating variables**
standing between a set of relational antecedents and a set of relational outcomes.
Trust additionally is a *cause* of commitment—one commits to a partner one trusts—
so the two mediators are themselves ordered.

@morganhunt1994 specify five antecedents (relationship termination costs, relationship
benefits, shared values, communication, and opportunistic behavior) and five
outcomes (acquiescence, propensity to leave, cooperation, functional conflict, and
decision-making uncertainty), with commitment and trust mediating between them.
@fig-rel-kmv renders the core mediation skeleton. The model's enduring influence
comes from this architecture rather than from any single coefficient: it
established that relational constructs are *mediators*, which means that a study
which regresses an outcome directly on relationship investments—omitting trust and
commitment—is mis-specified, and its direct effect is a reduced-form blend of the
true structural paths.

```{mermaid}
%%| label: fig-rel-kmv
%%| fig-cap: "The commitment–trust (KMV) mediation skeleton. Relational antecedents act on relational outcomes only through the key mediating variables, trust and commitment; trust is also an antecedent of commitment. A model that omits the mediators mis-specifies the mechanism."
flowchart LR
  SV["Shared values"] --> T["Trust"]
  COMM["Communication"] --> T
  OPP["Opportunistic<br/>behavior"] -->|–| T
  BEN["Relationship<br/>benefits"] --> C["Commitment"]
  TC["Termination<br/>costs"] --> C
  SV --> C
  T --> C
  C --> ACQ["Acquiescence"]
  C --> LEAVE["Propensity<br/>to leave (–)"]
  T --> COOP["Cooperation"]
  T --> UNC["Decision<br/>uncertainty (–)"]
```

### Estimating the mediation model {#sec-rel-kmv-est}

The KMV model is a structural-equation model in the sense of
@sec-cvv-econometric-bridge: each construct is latent, measured by reflective
indicators, and the structural paths connect the latent variables. The chunk below
simulates a compact version—two antecedents (shared values, communication) feeding
trust, trust plus relationship benefits feeding commitment, and commitment driving
acquiescence—each construct measured by three reflective items, then fits it in
`lavaan` [@rosseel2012lavaan] and recovers the structural paths. The point is the
one @morganhunt1994 make conceptually: the antecedents' effect on the outcome is
*indirect*, routed through the mediators, and a structural model recovers that
routing while a naive regression of the outcome on the antecedents would not.

```{r}
#| message: false
#| warning: false
#| label: rel-kmv-sem
set.seed(914)

have_lavaan <- requireNamespace("lavaan", quietly = TRUE)

n <- 800
# --- latent constructs (structural model) ---
shared <- rnorm(n)                              # shared values
comm   <- rnorm(n)                              # communication
trust  <- 0.45 * shared + 0.35 * comm + rnorm(n, sd = 0.7)
benef  <- rnorm(n)                              # relationship benefits
commit <- 0.50 * trust + 0.30 * benef + rnorm(n, sd = 0.7)
acquie <- 0.55 * commit + rnorm(n, sd = 0.8)    # acquiescence (outcome)

# --- reflective measurement: 3 indicators per construct ---
ind <- function(eta, lambdas) {
  sapply(lambdas, function(l) l * eta + rnorm(length(eta), sd = sqrt(1 - l^2)))
}
mk <- function(eta) ind(scale(eta), c(0.85, 0.78, 0.72))
dat <- data.frame(
  setNames(as.data.frame(mk(shared)), paste0("sv", 1:3)),
  setNames(as.data.frame(mk(comm)),   paste0("cm", 1:3)),
  setNames(as.data.frame(mk(trust)),  paste0("tr", 1:3)),
  setNames(as.data.frame(mk(benef)),  paste0("bn", 1:3)),
  setNames(as.data.frame(mk(commit)), paste0("co", 1:3)),
  setNames(as.data.frame(mk(acquie)), paste0("aq", 1:3))
)

if (have_lavaan) {
  kmv <- "
    SV =~ sv1 + sv2 + sv3
    CM =~ cm1 + cm2 + cm3
    TR =~ tr1 + tr2 + tr3
    BN =~ bn1 + bn2 + bn3
    CO =~ co1 + co2 + co3
    AQ =~ aq1 + aq2 + aq3
    TR ~ SV + CM
    CO ~ TR + BN
    AQ ~ CO
  "
  fit <- lavaan::sem(kmv, data = dat, std.lv = TRUE)
  pe  <- lavaan::parameterEstimates(fit, standardized = TRUE)
  print(pe[pe$op == "~", c("lhs", "rhs", "std.all", "pvalue")], row.names = FALSE)
  # Indirect effect of shared values on acquiescence (SV -> TR -> CO -> AQ)
  b <- function(l, r) pe$std.all[pe$op == "~" & pe$lhs == l & pe$rhs == r]
  cat(sprintf("\nIndirect SV -> AQ via TR, CO: %.3f\n",
              b("TR","SV") * b("CO","TR") * b("AQ","CO")))
} else {
  cat("lavaan unavailable; skipping KMV SEM demo.\n")
}
```

The recovered structural coefficients track the population paths, and the
indirect-effect calculation makes the mediation explicit: shared values reach
acquiescence only by first raising trust, which raises commitment, which raises
acquiescence. Omitting the mediators would collapse this chain into a single
attenuated direct effect and discard the mechanism the theory is *about*.

## Loyalty {#sec-rel-loyalty}

**Loyalty** is the downstream construct the relationship is built to produce, and it
is the one most often mismeasured—because the behavior (repeat buying) and the
attitude (commitment to the brand) can come apart. @dickbasu1994 give the field its
organizing scheme by crossing **relative attitude** (how strongly, and how
differentiated from alternatives, the consumer evaluates the brand) with **repeat
patronage** (observed behavioral frequency), yielding four cells (@tbl-rel-loyalty):

| | High repeat patronage | Low repeat patronage |
|---|---|---|
| **High relative attitude** | **Loyalty** (true) | **Latent loyalty** |
| **Low relative attitude** | **Spurious loyalty** | **No loyalty** |

: The Dick & Basu loyalty typology: loyalty as the conjunction of favorable relative attitude and repeat patronage [@dickbasu1994]. {#tbl-rel-loyalty}

The typology's payoff is diagnostic. **Spurious loyalty**—repeat buying without
favorable attitude—looks identical to true loyalty in transaction data but is
fragile, sustained by inertia, switching costs, or lack of alternatives rather than
preference, and it collapses when a competitor removes the friction. **Latent
loyalty**—favorable attitude without repeat buying—signals that situational or
market barriers (stockouts, distribution gaps, budget constraints) suppress
behavior the attitude would otherwise produce. The managerial prescriptions differ
sharply across cells, which is the whole point: a firm reading only behavioral
loyalty cannot tell a defensible franchise from a captive one, and will
mis-allocate retention spending accordingly. This is the construct-to-measure gap
of @fig-cm-gap in operational dress—behavioral repeat purchase is a *deficient and
contaminated* measure of the loyalty construct.

### Loyalty as a developmental sequence {#sec-rel-loyalty-phases}

@oliver1999loyalty deepens the static typology into a *temporal* one, arguing that
loyalty develops through four phases, each vulnerable to a different competitive
threat:

1. **Cognitive loyalty:** preference grounded in beliefs about the brand's
   attributes (information-based, hence vulnerable to better competitor information).
2. **Affective loyalty:** a liking or attitude toward the brand built from
   cumulative satisfaction (vulnerable to competitor-induced dissatisfaction and to
   the affect's own decay).
3. **Conative loyalty:** a behavioral *intention*, a commitment to rebuy (vulnerable
   to counter-persuasion).
4. **Action loyalty:** intention converted into action accompanied by a willingness
   to overcome obstacles—the only phase that is behaviorally inertial and hard for
   competitors to disrupt.

The sequence matters because it locates *where* a loyalty program or recovery effort
must intervene: defending a merely cognitively loyal customer requires different
levers than defending an action-loyal one, and the deepest loyalty is reached only
when satisfaction has been converted, stage by stage, into a self-sustaining
behavioral disposition. Oliver's account also clarifies why satisfaction is
*necessary but not sufficient* for loyalty—satisfaction feeds affective loyalty but
must still be carried through the conative and action phases—anticipating the
asymmetric, nonlinear satisfaction–loyalty link discussed below.

### Visualizing the typology {#sec-rel-loyalty-viz}

The chunk below simulates consumers on the two Dick–Basu axes and classifies them
into the four cells, illustrating how attitude and behavior dissociate and why a
behavior-only measure conflates true with spurious loyalty.

```{r}
#| message: false
#| warning: false
#| label: rel-loyalty-typology
#| fig-cap: "Simulated consumers in the Dick & Basu loyalty space. Repeat behavior alone (the horizontal axis) cannot separate true loyalty from spurious loyalty; relative attitude (the vertical axis) is what distinguishes them."
set.seed(2718)
n <- 600

# Attitude and behavior are correlated but far from identical (rho ~ 0.45):
# the gap between them is where spurious and latent loyalty live.
rel_att  <- rnorm(n)
repeat_b <- 0.45 * rel_att + rnorm(n, sd = sqrt(1 - 0.45^2))

cell <- ifelse(rel_att > 0 & repeat_b > 0, "True loyalty",
        ifelse(rel_att > 0 & repeat_b <= 0, "Latent loyalty",
        ifelse(rel_att <= 0 & repeat_b > 0, "Spurious loyalty", "No loyalty")))

cat("Cell counts:\n"); print(table(cell))
cat(sprintf("\nShare of repeat buyers (behavior > 0) that are SPURIOUS: %.1f%%\n",
            100 * mean(cell[repeat_b > 0] == "Spurious loyalty")))

plot(repeat_b, rel_att, col = factor(cell), pch = 19, cex = 0.5,
     xlab = "Repeat patronage (behavior)", ylab = "Relative attitude",
     main = "Dick & Basu loyalty space")
abline(h = 0, v = 0, lty = 2)
legend("topleft", legend = levels(factor(cell)),
       col = 1:4, pch = 19, cex = 0.7, bty = "n")
```

A sizeable share of the repeat buyers turn out to be *spurious* loyals—high behavior,
low attitude—who would be invisible to a churn model trained on transactions alone.
That is the empirical cost of treating a convenient behavioral metric as if it were
the construct.

## Relationship Quality: A Higher-Order Construct {#sec-rel-quality}

Trust, commitment, and satisfaction co-vary so strongly in relational settings that
a literature treats them as facets of a single higher-order construct,
**relationship quality**—the overall strength of a relationship, of which the three
are reflective sub-dimensions [@hennigthurau2002; @garbarino1999]. @garbarino1999
show the facets are not interchangeable across customer types: for *low-relational*
customers, overall satisfaction drives future intentions, whereas for
*high-relational* customers, trust and commitment carry the predictive load. This
is a moderation result with a measurement moral—the same battery of items maps onto
behavior differently depending on relationship stage, so a model that imposes one
structure on a mixed customer base will fit neither segment well.

The second-order structure is itself a testable measurement model. The chunk fits a
second-order CFA in which a single relationship-quality factor explains the
covariation among first-order trust, commitment, and satisfaction factors, each with
its own reflective indicators—the construct-validity machinery of @sec-construct-vs-variable
applied to a relational battery.

```{r}
#| message: false
#| warning: false
#| label: rel-quality-cfa
set.seed(4040)
n <- 700

# One second-order factor RQ drives three first-order factors,
# each measured by three reflective items.
RQ   <- rnorm(n)
trust  <- 0.80 * RQ + rnorm(n, sd = 0.6)
commit <- 0.75 * RQ + rnorm(n, sd = 0.6)
satis  <- 0.70 * RQ + rnorm(n, sd = 0.6)

mk <- function(eta, lam = c(0.85, 0.80, 0.74))
  sapply(lam, function(l) l * scale(eta) + rnorm(length(eta), sd = sqrt(1 - l^2)))
rq_dat <- data.frame(
  setNames(as.data.frame(mk(trust)),  paste0("tr", 1:3)),
  setNames(as.data.frame(mk(commit)), paste0("co", 1:3)),
  setNames(as.data.frame(mk(satis)),  paste0("sa", 1:3))
)

if (have_lavaan) {
  rq_mod <- "
    TR =~ tr1 + tr2 + tr3
    CO =~ co1 + co2 + co3
    SA =~ sa1 + sa2 + sa3
    RQ =~ TR + CO + SA          # second-order relationship-quality factor
  "
  fit <- lavaan::cfa(rq_mod, data = rq_dat, std.lv = TRUE)
  pe  <- lavaan::parameterEstimates(fit, standardized = TRUE)
  cat("Second-order loadings (RQ -> facets):\n")
  print(pe[pe$op == "=~" & pe$lhs == "RQ", c("rhs", "std.all")], row.names = FALSE)
  fm <- lavaan::fitMeasures(fit, c("cfi", "rmsea", "srmr"))
  cat(sprintf("\nFit: CFI=%.3f  RMSEA=%.3f  SRMR=%.3f\n", fm[1], fm[2], fm[3]))
} else {
  cat("lavaan unavailable; skipping relationship-quality CFA.\n")
}
```

When the second-order loadings are high and the model fits, relationship quality is
defensible as a single construct; when a facet loads weakly, the data are saying that
facet carries relationship-specific information the umbrella construct discards—the
same reflective/formative judgment call of @tbl-reflective-formative, now at the level
of whether three constructs are one.

## Consequences and Antecedents: The Meta-Analytic Evidence {#sec-rel-consequences}

Because the relational constructs have been measured in hundreds of studies, the most
authoritative statements about them are meta-analytic. @palmatier2006meta synthesize
the relationship-marketing literature and deliver several findings that discipline the
individual-study noise:

- **Commitment and trust are the central mediators**, but they do not mediate
  equally: relational antecedents that build trust and commitment translate into
  performance, and *objective* relationship outcomes (e.g., exchange performance) are
  predicted better by some constructs than the seller's financial performance is.
- **Relationship investments and relationship benefits** are among the strongest
  antecedents, consistent with the social-exchange logic of @sec-rel-turn.
- The constructs' effects are **stronger in service than in product contexts and in
  channel/B2B than in consumer settings**, because relational governance matters most
  where exchange is ongoing, customized, and hard to contract on completely.

@palmatier2007comparative push further by pitting theoretical perspectives against
one another longitudinally, and @palmatier2013 add a *dynamic* construct—**relationship
velocity**, the rate and direction of change in relationship quality—showing that the
*trajectory* of a relationship predicts performance beyond its current level, so two
relationships of equal present quality diverge in value depending on whether they are
improving or decaying. On the outcome side, @verhoef2002relational link relational
constructs to **customer referrals and cross-buying**, and @watson2015loyalty review
how loyalty is built, measured, and—crucially—*monetized*, integrating the
attitudinal/behavioral distinction with the customer-lifetime-value machinery of
@sec-clv. @chaudhuri2001 close the loop to brand outcomes: brand trust and brand
affect drive both purchase and attitudinal loyalty, and through them market-share and
relative-price outcomes, tying the relational constructs of this chapter to the brand
constructs of @sec-branding.

## Disciplining Findings {#sec-rel-disciplining}

As with satisfaction (@sec-satisfaction), the relational constructs invite a naive
maximization that the evidence does not support. Three findings discipline it.

First, **loyalty is not automatically profitable.** The presumption that loyal
customers cost less to serve, pay more, and spread favorable word of mouth holds only
conditionally; @watson2015loyalty and the broader loyalty literature document that
behavioral loyalty unaccompanied by attitudinal loyalty (the *spurious* cell of
@tbl-rel-loyalty) often reflects captivity rather than preference and carries little
of loyalty's promised margin. Managing the behavioral metric without the attitude can
buy retention that does not pay.

Second, **the satisfaction–loyalty link is asymmetric and nonlinear.** Oliver's phase
model (@sec-rel-loyalty-phases) and the delight research of @oliver1997delight imply
that satisfaction must be carried through successive phases to yield action loyalty,
and that merely *meeting* expectations secures far less loyalty than *exceeding* them
into delight—while dissatisfaction destroys loyalty faster than satisfaction builds
it, echoing the loss-aversion asymmetry priced by markets in @sec-satisfaction.

Third, **relationships have a dark side.** The same idiosyncratic investments and
pledges that build commitment [@andersonweitz1992; @gundlach1995] also create
lock-in, complacency, and vulnerability to opportunism; @aaker2004 show that strong
consumer–brand relationships make transgressions *more* damaging, not less, because a
trusted partner's betrayal violates a stronger expectation. Commitment, in other
words, raises the stakes of @eq-rel-trust: the deeper the relationship, the larger the
loss $L$ a betrayal inflicts, and the more a single failure can cost. Relationship
strength is an asset that is also a hostage.

## Connections {#sec-rel-connections}

The relational constructs sit at the center of the book's construct map. Their
measurement is the reflective-construct theory of @sec-construct-vs-variable; their
upstream driver is the satisfaction of @sec-satisfaction; their monetization is the
customer-lifetime-value and loyalty-program material of @sec-clv; and their
brand-level expression—brand trust, brand attachment, brand love—is developed among
the brand-and-self constructs and in @sec-branding. The KMV mediation logic also
recurs whenever a marketing input is theorized to work *through* a psychological
state rather than on behavior directly, which is most of the time.

## Key Takeaways {#sec-rel-takeaways}

- The relational turn [@dwyer1987] relocated marketing's object of study from the
  discrete transaction to the ongoing relationship, making **trust**, **commitment**,
  and **loyalty** the central intangibles—latent constructs measured reflectively,
  inheriting every reliability and validity concern of @sec-construct-vs-variable.
- **Trust** is a two-dimensional belief—credibility (competence) and benevolence
  (motive)—and a *willingness to rely* distinct from the belief itself
  [@moorman1992trust; @moorman1993trust]; it supports reliance only when confidence
  clears a stake-dependent threshold (@eq-rel-trust).
- **Commitment** is multi-component—affective, calculative, temporal
  [@gundlach1995]—and is revealed by credible, costly **pledges** rather than stated
  intent [@andersonweitz1992]; affective and calculative commitment have sharply
  different downstream behavior.
- The **commitment–trust (KMV) theory** [@morganhunt1994] establishes trust and
  commitment as the **key mediating variables** between relational antecedents and
  outcomes; the model is a latent-variable SEM (@sec-rel-kmv-est), and omitting the
  mediators mis-specifies the mechanism.
- **Loyalty** is the conjunction of favorable relative attitude and repeat patronage
  [@dickbasu1994], developing through cognitive → affective → conative → action phases
  [@oliver1999loyalty]; behavior-only measures cannot separate **true** from
  **spurious** loyalty (@tbl-rel-loyalty).
- Meta-analytic evidence makes commitment and trust the central mediators, strongest
  in service and B2B settings [@palmatier2006meta], with relationship *velocity*
  adding a dynamic dimension [@palmatier2013] and loyalty monetized through CLV
  [@watson2015loyalty; @verhoef2002relational; @chaudhuri2001].
- Disciplining findings: loyalty is profitable only conditionally, the
  satisfaction–loyalty link is asymmetric and nonlinear [@oliver1997delight], and
  strong relationships amplify the damage of betrayal [@aaker2004]—commitment raises
  the stakes of trust, not just its rewards.

## Further Reading {#sec-rel-further}

The foundational relational-turn statement is @dwyer1987; the trust conceptualization
is @moorman1992trust and @moorman1993trust, and the commitment structure is
@gundlach1995 with @andersonweitz1992 on pledges. The canonical synthesis is the
commitment–trust theory of @morganhunt1994, extended longitudinally by
@palmatier2007comparative and dynamically by @palmatier2013. On loyalty, read
@dickbasu1994 for the typology and @oliver1999loyalty for the phase model, with
@watson2015loyalty for the modern measurement-and-monetization integration. The
meta-analytic anchor is @palmatier2006meta; the customer-type moderation is
@garbarino1999; the relational-benefits and relationship-quality integration is
@hennigthurau2002; and the consumer–brand-relationship and dark-side strands run
through @fournier1998 and @aaker2004 into @sec-branding.
