# Engagement, Involvement, and Word of Mouth {#sec-engagement}

The constructs so far have described the consumer as an evaluator (@sec-evaluative),
a relationship partner (@sec-relational), and a brand-identifier (@sec-brand-self).
This chapter turns to the consumer as a *participant*—someone whose contribution to a
firm extends past the purchase to include attention, effort, advocacy, content, and
referrals. Three constructs organize this participatory turn. **Involvement** is the
motivational state—the personal relevance of an object that determines how much
cognitive effort a consumer will expend on it. **Engagement** is the set of behavioral
and psychological manifestations of an active relationship that go *beyond* transactions.
And **word of mouth** (WOM) is the social output—the informal communication about
products and brands that consumers direct at other consumers, and that has become, in
the digital era, the most consequential and most measurable of the three.

These constructs have moved from the periphery of marketing to its center as firms have
recognized that a customer's value is not exhausted by what they buy: a customer who
writes reviews, refers friends, and defends the brand creates value through channels
that transaction data alone cannot see. We develop involvement first, as the
motivational antecedent and a moderator that reshapes how all persuasion works; then
customer engagement, including its decomposition into a total *engagement value*; then
word of mouth as a construct distinct from the *mechanics* of diffusion treated in
@sec-virality. As throughout this part, the measurement subtext is discriminant
validity: engagement, involvement, participation, and WOM-intention are easy to define
apart and hard to keep apart in data.

## Involvement {#sec-eng-involvement}

**Involvement** is a person's perceived relevance of an object based on inherent needs,
values, and interests [@zaichkowsky1985]. It is a *motivational* construct: high
involvement means the object matters to the consumer, which mobilizes attention,
information search, and elaboration; low involvement means it does not, so decisions are
made with minimal cognitive effort. @zaichkowsky1985 develops the Personal Involvement
Inventory to measure it reflectively, and the construct is standardly partitioned along
two axes—**enduring** involvement (a stable interest in a category) versus **situational**
involvement (temporarily heightened by a purchase occasion or risk), and by object
(*product*, *advertising*, and *purchase-decision* involvement).

Involvement's theoretical importance is less as an outcome than as a **moderator** of how
persuasion operates. The Elaboration Likelihood Model [@petty1983] holds that persuasion
travels two routes: a **central route**, in which highly involved consumers scrutinize the
quality of an argument, and a **peripheral route**, in which uninvolved consumers respond
to surface cues (source attractiveness, number of arguments, music) without evaluating
substance. Involvement determines which route dominates, so the *same* advertisement
persuades through different mechanisms depending on the audience's involvement—a result
that reshapes advertising strategy (@sec-advertising) and that makes involvement a
prerequisite variable for almost any study of message effects.
@fig-eng-elm-routes sketches the two routes and involvement's gatekeeping role between them.

```{mermaid}
%%| label: fig-eng-elm-routes
%%| fig-cap: "The two routes of the Elaboration Likelihood Model. The same message persuades through scrutiny of argument quality when involvement is high and through surface cues when it is low—involvement routes persuasion."
flowchart LR
  MSG["Persuasive message:<br/>the same advertisement"] --> CEN["Central route:<br/>scrutiny of argument quality"]
  MSG --> PER["Peripheral route: surface cues—<br/>source attractiveness,<br/>number of arguments, music"]
  INV["Involvement:<br/>perceived personal relevance"] -.->|"high"| CEN
  INV -.->|"low"| PER
  CEN --> ATT["Attitude"]
  PER --> ATT
```

### Involvement as a moderator of message effects {#sec-eng-elm}

The ELM's central prediction is an *interaction*: argument quality matters more for
high-involvement consumers, while peripheral cues matter more for low-involvement
consumers. The chunk simulates this crossover and recovers the two interactions,
illustrating why a study that ignores involvement will estimate an average message effect
that describes no actual consumer.

```{r}
#| message: false
#| warning: false
#| label: eng-elm-moderation
set.seed(1983)
n <- 4000

involvement <- rbinom(n, 1, 0.5)            # 1 = high, 0 = low
arg_quality <- rnorm(n)                     # strength of the message's arguments
periph_cue  <- rnorm(n)                     # attractiveness/number of cues

# High-involvement: persuasion via argument quality (central route).
# Low-involvement:  persuasion via peripheral cues.
attitude <- 0.10 +
  0.55 * involvement * arg_quality +        # central route (only when involved)
  0.10 * (1 - involvement) * arg_quality +
  0.50 * (1 - involvement) * periph_cue +   # peripheral route (only when uninvolved)
  0.10 * involvement * periph_cue +
  rnorm(n, sd = 0.8)

fit <- lm(attitude ~ arg_quality * involvement + periph_cue * involvement)
co  <- round(summary(fit)$coefficients[, c("Estimate", "Pr(>|t|)")], 3)
# Show the main effects and their involvement interactions (robust to term naming).
keep <- grepl("arg_quality|periph_cue", rownames(co))
print(co[keep, ])
```

The positive `arg_quality:involvement` interaction and the negative
`periph_cue:involvement` interaction are the ELM crossover: involvement routes persuasion.
An analyst who omitted involvement and regressed attitude on argument quality alone would
recover a muddy average, understating the central route for the involved and missing the
peripheral route for the uninvolved entirely.

## Customer Engagement {#sec-eng-engagement}

**Customer engagement** generalizes the relationship beyond the transaction. @vandoorn2010
define **customer engagement behaviors** (CEBs) as a customer's behavioral manifestations
toward a firm, beyond purchase, that result from motivational drivers—writing reviews,
making recommendations, helping other customers, blogging, and otherwise contributing.
@brodie2011 give the construct its conceptual domain, framing engagement as an
**interactive, co-creative** experience with a focal object that comprises cognitive,
emotional, and behavioral dimensions and that arises within an iterative process of
relationship development. @hollebeek2011 develops the brand-specific version. The common
thread is that engagement captures value-creating customer activity that purchase metrics
miss, which is why it rose to prominence exactly as social media made such activity
visible and abundant.

@pansari2017 supply the nomological network: engagement is built on a foundation of
satisfaction *and* emotion (a relationship must be both satisfying and emotionally
connected to become engaged), and it produces value through **direct** contributions
(purchases) and **indirect** contributions (referrals, influence, feedback). The
direct/indirect split is the conceptual bridge to a measurement program that tries to put
a number on the whole of a customer's value.
@fig-eng-engagement-net assembles this nomological network.

```{mermaid}
%%| label: fig-eng-engagement-net
%%| fig-cap: "The nomological network of customer engagement. A relationship that is both satisfying and emotionally connected becomes engaged; engagement spans cognitive, emotional, and behavioral dimensions, manifests in behaviors beyond purchase, and produces direct and indirect value."
flowchart LR
  SAT["Satisfaction"] --> ENG["Customer engagement:<br/>cognitive, emotional, and<br/>behavioral dimensions"]
  EMO["Emotional connection"] --> ENG
  ENG --- CEB["Engagement behaviors: writing reviews,<br/>recommending, helping other<br/>customers, blogging"]
  ENG --> DIR["Direct contributions:<br/>purchases"]
  ENG --> IND["Indirect contributions:<br/>referrals, influence, feedback"]
```

### Total customer engagement value {#sec-eng-cev}

@kumar2010cev decompose a customer's total worth into four additive components—the move
that lets engagement be *valued*, not just described:

- **Customer Lifetime Value (CLV):** the discounted margin from the customer's own
  purchases (the construct developed in @sec-clv).
- **Customer Referral Value (CRV):** value from customers acquired through this customer's
  formal referrals.
- **Customer Influencer Value (CIV):** value from this customer's influence on other
  customers and prospects through word of mouth and social spread.
- **Customer Knowledge Value (CKV):** value from the feedback and ideas the customer
  contributes to the firm.

@fig-eng-cev-formative depicts the decomposition: four distinct value sources summed into
one formative index.

```{mermaid}
%%| label: fig-eng-cev-formative
%%| fig-cap: "Total customer engagement value as a formative index. The four components are distinct, summed value sources—not interchangeable reflections of a latent engagement—so the arrows point from components to construct."
flowchart LR
  CLV["CLV: discounted margin from<br/>the customer's own purchases"] --> CEV["Total customer<br/>engagement value"]
  CRV["CRV: customers acquired through<br/>this customer's formal referrals"] --> CEV
  CIV["CIV: influence on customers and<br/>prospects via word of mouth<br/>and social spread"] --> CEV
  CKV["CKV: feedback and ideas<br/>contributed to the firm"] --> CEV
```

A customer who buys little but refers and influences many can be worth more than a heavy
buyer who does neither, which is the entire point: optimizing CLV alone (@sec-clv)
*undervalues* engaged customers and can lead a firm to fire its most valuable advocates.
The decomposition is **formative**—the four components are distinct value sources that are
summed, not interchangeable reflections of a latent engagement (@tbl-reflective-formative)—so
the appropriate reliability question is not internal consistency but whether the components
are correctly weighted and non-overlapping. The chunk simulates a customer base and shows
how rank-ordering customers by total engagement value reshuffles the priority list relative
to CLV alone.

```{r}
#| message: false
#| warning: false
#| label: eng-cev-decomp
set.seed(2010)
n <- 5000

# Four value components; deliberately NOT perfectly correlated, so that
# CLV alone mis-ranks customers relative to total engagement value.
clv <- rgamma(n, shape = 2, scale = 120)                 # own-purchase value
crv <- rgamma(n, shape = 1.2, scale = 60) * (runif(n) < .5)  # referral value
civ <- rgamma(n, shape = 1.1, scale = 90) * (runif(n) < .4)  # influencer value
ckv <- rgamma(n, shape = 1.0, scale = 25) * (runif(n) < .3)  # knowledge value

cev <- clv + crv + civ + ckv                              # total engagement value

top_clv <- order(clv, decreasing = TRUE)[1:500]           # top 500 by CLV
top_cev <- order(cev, decreasing = TRUE)[1:500]           # top 500 by CEV
overlap <- length(intersect(top_clv, top_cev))

cat(sprintf("Correlation CLV vs total engagement value: %.2f\n", cor(clv, cev)))
cat(sprintf("Of the top-500 customers by CLV, only %d are also top-500 by CEV.\n", overlap))
cat(sprintf("=> %d high-engagement customers are INVISIBLE to a CLV-only ranking.\n",
            500 - overlap))
```

A non-trivial share of the most valuable customers by total engagement value are *not* in
the CLV top tier, so a firm that targets and rewards on CLV alone systematically neglects
its referral and influence engine. Engagement, made measurable through the CEV
decomposition, is the construct that corrects that blind spot.

## Word of Mouth {#sec-eng-wom}

The most-studied engagement behavior is **word of mouth**—informal, consumer-to-consumer
communication about products and brands. It is essential to distinguish WOM-the-*construct*
from the *diffusion mechanics* of @sec-virality: this chapter treats WOM as a property of
consumers (their propensity to talk, the valence and volume of what they say, and its
antecedents), while the virality chapter treats the network dynamics by which messages
spread. As a construct, WOM is characterized by **valence** (positive vs. negative),
**volume** (how much), and the gap between **intention** and **behavior**.

Its antecedents are well mapped. @anderson1998wom establishes that the satisfaction–WOM
relationship is **asymmetric and U-shaped**: both highly satisfied *and* highly
dissatisfied customers generate more word of mouth than indifferent ones, and the
dissatisfied generate disproportionately much—the negativity that recurs throughout this
book (the loss-aversion theme of @sec-satisfaction). @brown2005wom identify the drivers of
*positive* WOM intentions and behaviors, finding commitment and identification (the
constructs of @sec-relational and @sec-brand-self) central. @berger2014wom synthesizes the
psychology of *why* people talk, organizing the drivers into five functions—impression
management, emotion regulation, information acquisition, social bonding, and persuasion—and
the meta-analysis of @dematos2008 quantifies the antecedents across studies, confirming
satisfaction, loyalty, commitment, perceived quality, and trust as robust drivers. WOM is
thus the social *output* of nearly every other construct in this part.
@fig-eng-wom-net gathers these antecedents into a single map.

```{mermaid}
%%| label: fig-eng-wom-net
%%| fig-cap: "The antecedents of word of mouth. Satisfaction drives WOM from both tails, with the dissatisfied loudest; commitment and identification drive positive WOM; five psychological functions explain why people talk at all."
flowchart LR
  SU["Satisfaction: U-shaped and<br/>asymmetric—both tails talk,<br/>the dissatisfied loudest"] --> WOM["Word of mouth:<br/>valence, volume,<br/>intention vs behavior"]
  CI["Commitment and identification:<br/>drivers of positive WOM"] --> WOM
  META["Meta-analytic drivers: satisfaction,<br/>loyalty, commitment,<br/>perceived quality, trust"] --> WOM
  FUN["Five functions of talking:<br/>impression management, emotion regulation,<br/>information acquisition, social bonding,<br/>persuasion"] --> WOM
```

### The U-shaped satisfaction–WOM relationship {#sec-eng-wom-ushape}

The @anderson1998wom finding is a clean nonlinearity worth making explicit, because a
linear model would conclude—wrongly—that satisfaction monotonically increases word of
mouth. The chunk simulates WOM volume as a U-shaped function of satisfaction with a
steeper negative arm (the dissatisfied talk more), and contrasts the misleading linear fit
with the correct quadratic one.

```{r}
#| message: false
#| warning: false
#| label: eng-wom-ushape
#| fig-cap: "Word-of-mouth volume as a U-shaped function of satisfaction (Anderson 1998): the very satisfied and the very dissatisfied talk most, with a steeper dissatisfaction arm. A linear fit (dashed) misses the mechanism entirely."
set.seed(1998)
n <- 1500
sat <- runif(n, 1, 7)                          # satisfaction, 1..7
center <- 4
# U-shape with a steeper negative arm; minimum WOM near the middle.
wom <- 1.2 * (sat - center)^2 +
       0.9 * pmax(center - sat, 0) +           # extra volume on the dissatisfied side
       rnorm(n, sd = 1.5)
wom <- pmax(wom, 0)

lin  <- lm(wom ~ sat)
quad <- lm(wom ~ poly(sat, 2))
cat(sprintf("Linear fit R^2    : %.3f\n", summary(lin)$r.squared))
cat(sprintf("Quadratic fit R^2 : %.3f\n", summary(quad)$r.squared))

plot(sat, wom, pch = 19, cex = 0.3, col = "grey60",
     xlab = "Satisfaction", ylab = "WOM volume", main = "Satisfaction and word of mouth")
ord <- order(sat)
abline(lin, lty = 2, lwd = 2)
lines(sat[ord], predict(quad)[ord], col = "firebrick", lwd = 2)
legend("top", c("Linear (misleading)", "Quadratic (correct)"),
       lty = c(2, 1), col = c("black", "firebrick"), lwd = 2, bty = "n")
```

The quadratic fit captures the U-shape the linear model flattens into near-nothing: word
of mouth is generated at *both* tails of satisfaction, and the dissatisfied tail is louder.
A firm reading a near-zero linear coefficient would wrongly conclude satisfaction does not
drive WOM, when in fact it drives it nonlinearly and asymmetrically—the same lesson, in a
new construct, that the satisfaction chapter drew about asymmetric market pricing.

## Measurement and the Construct Boundary {#sec-eng-measurement}

Engagement is the construct most at risk of conceptual inflation. Because it is defined
expansively—any value-creating activity beyond purchase—it threatens to absorb involvement
(its motivational antecedent), participation, co-creation, loyalty (@sec-relational), and
word of mouth (one of its behaviors) into an undifferentiated whole. Two disciplines keep
the boundary sharp. First, distinguishing the **psychological state** of engagement (a
motivational, cognitive-affective disposition) from **engagement behaviors** (the observable
CEBs), and measuring each with its own indicators rather than conflating cause and
manifestation. Second, recognizing that *engagement value* (CEV) is **formative** while the
*engagement state* is **reflective**—applying internal-consistency reliability to the former
or summative weighting to the latter is the model-misspecification error of
@sec-construct-vs-variable. Word of mouth, similarly, must keep **intention** and **behavior**
separate (the gap is large and predictable) and **valence** explicit (positive and negative
WOM have different antecedents and opposite effects), or its measures collapse distinctions the
theory depends on.

## Key Takeaways {#sec-eng-takeaways}

- **Involvement** is perceived personal relevance [@zaichkowsky1985]—enduring vs.
  situational, by object—and matters chiefly as a **moderator**: it routes persuasion
  between the central and peripheral paths of the ELM [@petty1983], so message effects are
  involvement-conditional (@sec-eng-elm).
- **Customer engagement** captures value-creating activity beyond purchase—engagement
  *behaviors* [@vandoorn2010] and an interactive, co-creative *state* [@brodie2011;
  @hollebeek2011]—built on satisfaction plus emotion and producing direct and indirect value
  [@pansari2017].
- **Total customer engagement value** decomposes a customer into CLV + referral + influencer
  + knowledge value [@kumar2010cev]; it is **formative**, and CLV-only targeting (@sec-clv)
  systematically undervalues engaged advocates (@sec-eng-cev).
- **Word of mouth** is a construct of valence, volume, and intention-vs-behavior, distinct
  from the diffusion mechanics of @sec-virality; the satisfaction–WOM link is **U-shaped and
  asymmetric** [@anderson1998wom] (@sec-eng-wom-ushape), positive WOM is driven by commitment
  and identification [@brown2005wom], and the psychology of talking spans five functions
  [@berger2014wom], with robust meta-analytic antecedents [@dematos2008].
- The family's measurement discipline is to separate **state from behavior** and **reflective
  from formative**, and to keep WOM **intention/behavior** and **valence** explicit, on pain
  of the misspecification errors of @sec-construct-vs-variable.

## Further Reading {#sec-eng-further}

Involvement begins with @zaichkowsky1985 and its moderating role with the ELM of @petty1983.
For customer engagement, read @vandoorn2010 for the behavioral foundation, @brodie2011 for
the conceptual domain, and @pansari2017 for the antecedents-and-consequences synthesis, with
@kumar2010cev for the engagement-value decomposition (and @sec-clv for CLV proper). On word
of mouth, @anderson1998wom establishes the U-shape, @brown2005wom the positive-WOM
antecedents, @berger2014wom the psychology of transmission, and @dematos2008 the meta-analytic
evidence; the diffusion side is @sec-virality. The discriminant- and specification-validity
cautions are, as ever, the apparatus of @sec-construct-vs-variable.
