# Frontier Constructs {#sec-frontier}

The construct map of marketing is not closed. As markets, technologies, and the
texture of consumer life change, new constructs enter the nomological network—some
because new phenomena demand them (consumers now react to *algorithms*, share rather
than own, and trade data for personalization), some because old intuitions finally
acquire measurement. This chapter surveys six constructs at the current frontier:
**customer experience**, **psychological ownership**, **brand authenticity**,
**algorithm aversion** (and the broader family of consumer responses to AI),
**privacy concern**, and **access-based consumption**. Each has accumulated enough
lawful relations to meet the Cronbach–Meehl admission test that opens this part, and
each illustrates a distinct frontier—experiential, possessive, identity-authenticity,
technological, informational, and ownership-model.

A frontier construct carries a special hazard worth stating before we begin. The
newest constructs are, almost by definition, the *least validated*: their scales are
young, their discriminant validity against neighboring constructs is often unestablished,
and the literature has had little time to expose them to the **jingle–jangle fallacies**
(calling two different things by one name; calling one thing by two names). The
measurement discipline of @sec-construct-vs-variable therefore weighs *more* heavily at
the frontier, not less. The excitement of a new construct is not evidence that it is
distinct from the constructs it joins, and the burden of proof for a frontier construct
is the full convergent/discriminant/nomological battery, applied with extra skepticism.

## Customer Experience {#sec-fr-experience}

The most integrative frontier construct is **customer experience** (CX). @lemon2016
define it as the customer's *multidimensional response*—cognitive, emotional, sensorial,
social, and behavioral—to a firm's offerings across the entire **customer journey**, from
pre-purchase through purchase to post-purchase, accumulated over many **touchpoints** the
firm only partly controls (brand-owned, partner-owned, customer-owned, and
social/external). CX subsumes and reorganizes several constructs in this part: satisfaction
(@sec-satisfaction) becomes one episodic, post-purchase slice of experience; perceived
quality and value (@sec-evaluative) are cognitive components; engagement (@sec-engagement)
is a behavioral one. The construct's frontier status comes from its measurement difficulty
rather than novelty of idea: experience is *dynamic* (it evolves along the journey),
*distributed* (across touchpoints, many beyond the firm's control), and *multidimensional*,
so a single cross-sectional survey scale is structurally inadequate to it. Measuring CX well
requires journey-level, multi-touchpoint, multi-moment data—exactly the kind of behavioral
and unstructured data the methodology and multimodal parts of this book develop—and the gap
between the expansive construct and the thin measures usually used for it is the
construct-to-measure gap of @fig-cm-gap at its widest.

## Psychological Ownership {#sec-fr-ownership}

**Psychological ownership** is the feeling that a target—a product, a brand, an idea—is
"MINE," in the absence of (or in addition to) legal ownership. The construct matters because
that *feeling* of ownership, not the legal fact, drives valuation: it is the psychological
engine behind the **endowment effect**, the tendency to value a good more once it feels
one's own. @peck2009 show that *mere touch* raises perceived ownership and, through it,
valuation and willingness to pay—physical contact, or even vivid imagined contact, induces
the feeling of mine-ness. @fuchs2010 demonstrate that **empowerment** strategies (letting
customers participate in product selection or design) raise psychological ownership of the
resulting product, which in turn raises demand—a "I-designed-it-myself" effect that converts
participation into preference. Psychological ownership has clear antecedents (control over
the target, intimate knowledge of it, investment of the self into it) and clear consequences
(higher valuation, stewardship, willingness to pay), making it a well-formed construct with a
genuine nomological network.

### Ownership and valuation: an endowment simulation {#sec-fr-ownership-sim}

The core nomological claim—psychological ownership raises valuation—is directly simulable as
an endowment/touch manipulation. The chunk assigns consumers to an ownership-inducing
condition (e.g., touch or self-design) versus a control, and shows the induced ownership
shifts willingness to pay upward, the @peck2009 and @fuchs2010 result in miniature.

```{r}
#| message: false
#| warning: false
#| label: fr-ownership-endowment
set.seed(2009)
n <- 2000

owned <- rbinom(n, 1, 0.5)                      # 1 = touch/self-design condition
base_value <- rnorm(n, mean = 20, sd = 5)       # intrinsic valuation of the good

# Psychological ownership is induced by the condition, with individual variation,
# and ownership raises willingness to pay (the endowment channel).
psych_own <- 0.9 * owned + rnorm(n, sd = 0.4)   # latent felt ownership
wtp <- base_value + 6.0 * psych_own + rnorm(n, sd = 3)

cat(sprintf("Mean WTP, control            : $%.2f\n", mean(wtp[owned == 0])))
cat(sprintf("Mean WTP, ownership-induced  : $%.2f\n", mean(wtp[owned == 1])))
# Mediation: does felt ownership carry the condition's effect on WTP?
tot <- coef(lm(wtp ~ owned))["owned"]
dir <- coef(lm(wtp ~ owned + psych_own))["owned"]
cat(sprintf("Total effect of condition on WTP   : $%.2f\n", tot))
cat(sprintf("Direct effect controlling ownership: $%.2f  (mostly mediated)\n", dir))
```

The ownership condition lifts willingness to pay, and the lift is largely *mediated* by felt
ownership—the direct effect shrinks once psychological ownership is controlled—confirming the
construct's role as the mechanism rather than the manipulation being a mere framing trick.
This is the endowment effect re-described as a measurable consumer construct with a touchable
antecedent.

## Brand Authenticity {#sec-fr-authenticity}

**Brand authenticity** is the consumer's perception that a brand is *genuine, true to itself,
and not merely a commercial construction*. @morhart2015 give the construct an integrative
framework and a measurement scale with four dimensions—**continuity** (the brand's timelessness
and historical consistency), **credibility** (its ability and willingness to deliver what it
promises), **integrity** (its moral purity and care), and **symbolism** (its serving as a
resource for consumers' identity). The construct is a frontier one because the "authenticity
economy" has made perceived genuineness a competitive battleground, and because it is, like
all the constructs in this book, *perceived*: authenticity is not a property a brand objectively
has but a judgment consumers render, contaminated by marketing communications that try to
manufacture the very impression of un-manufactured-ness. That tension—authenticity claimed
through inauthentic means—makes its measurement and its nomological links to trust (@sec-relational),
brand love (@sec-brand-self), and self-connection an active and subtle research area.

## Consumer Responses to AI: Algorithm Aversion and Beyond {#sec-fr-ai}

A fast-growing construct family concerns how consumers respond to **algorithms and artificial
intelligence** as decision agents. @dietvorst2015 document **algorithm aversion**: people
erroneously avoid algorithms after seeing them err, abandoning a statistically superior
algorithm for an inferior human judge because they penalize algorithmic mistakes more harshly
than identical human mistakes. @longoni2019 sharpen the mechanism in the medical domain—**resistance
to medical AI**—and trace it to **uniqueness neglect**: consumers resist AI care because they
believe an algorithm cannot account for their personal uniqueness the way a human provider can,
even when the AI is more accurate. @longoni2022wordofmachine add the **word-of-machine effect**:
consumers prefer AI recommendations for *utilitarian* attributes (where accuracy dominates) but
human recommendations for *hedonic/experiential* attributes (where they doubt the machine can
appreciate subjective experience). These are distinct constructs—an aversion, a resistance, a
domain-contingent preference—but they form a coherent family describing the consumer's
*trust in and acceptance of* non-human agents, a family that did not exist fifteen years ago and
that is now central to marketing as AI mediates more of the customer journey (@sec-ai-ml).

### Why a superior algorithm gets abandoned {#sec-fr-ai-sim}

The @dietvorst2015 finding hinges on an *asymmetry*: an identical-sized error damages trust more
when an algorithm makes it than when a human does. The chunk simulates consumers choosing between
a more-accurate algorithm and a less-accurate human after observing each err, and shows that the
heavier trust penalty on algorithmic error drives adoption of the *inferior* human—algorithm
aversion as an emergent consequence of asymmetric error-attribution.

```{r}
#| message: false
#| warning: false
#| label: fr-algorithm-aversion
set.seed(2015)
n <- 3000

# The algorithm is genuinely more accurate (smaller errors) than the human.
err_algo  <- abs(rnorm(n, sd = 0.6))
err_human <- abs(rnorm(n, sd = 1.0))

# But algorithmic error is penalized ~2x as heavily (the aversion asymmetry).
penalty_algo  <- 2.0
penalty_human <- 1.0
trust_algo  <- 10 - penalty_algo  * err_algo
trust_human <- 10 - penalty_human * err_human

choose_algo <- trust_algo > trust_human
cat(sprintf("Algorithm is more accurate (mean error %.2f vs %.2f).\n",
            mean(err_algo), mean(err_human)))
cat(sprintf("Yet only %.1f%% of consumers choose the (superior) algorithm,\n",
            100 * mean(choose_algo)))
cat("because algorithmic errors are penalized twice as heavily as human ones.\n")
# Remove the asymmetry: penalize both equally -> adoption flips.
fair <- (10 - err_algo) > (10 - err_human)
cat(sprintf("Under symmetric error penalties, %.1f%% choose the algorithm.\n",
            100 * mean(fair)))
```

Under the asymmetric penalty the superior algorithm is rejected by a majority; remove the
asymmetry and adoption flips to favor it. Algorithm aversion is thus not (only) about accuracy
beliefs but about a *differential trust response to machine error*—a construct with direct
design implications: deploying AI requires managing the error-attribution asymmetry, not merely
proving the AI is better.

## Privacy Concern {#sec-fr-privacy}

As marketing runs on personal data, **privacy concern**—the consumer's apprehension about the
collection and use of their personal information—has become a first-order construct. @martinmurphy2017
provide an integrative framework, organizing privacy as a construct with antecedents (data
practices, control, transparency), states (concern, perceived risk, perceived intrusion), and
consequences (trust, disclosure, patronage). The defining tension is the **personalization–privacy
paradox**: the same data that lets a firm personalize—raising perceived value (@sec-evaluative) and
relationship quality (@sec-relational)—simultaneously raises privacy concern, so personalization
both attracts and repels, and the net effect is contingent on trust and control. The construct
ties directly to the firm-performance evidence on data practices and to the deeper treatment of
privacy regulation and its marketing consequences in @sec-privacy. Its measurement inherits a
sharp version of the social-desirability contamination of @sec-construct-vs-variable: stated
privacy concern routinely *overstates* the behavior it predicts (the "privacy paradox" in which
people who report high concern nonetheless disclose freely), a convergent-validity gap between the
attitudinal measure and revealed behavior that any serious use of the construct must confront.

## Access-Based Consumption {#sec-fr-access}

Finally, the rise of the sharing economy has produced a construct that reframes ownership itself.
@bardhi2012 conceptualize **access-based consumption**—consumption in which no transfer of
ownership occurs (car sharing, streaming, rental, subscription)—and show it is *not* simply
diluted ownership but a distinct mode with its own psychology: consumers relate to accessed
objects with less identification and less stewardship, treating them instrumentally rather than
as extensions of the self. This inverts the psychological-ownership construct of @sec-fr-ownership:
where ownership induces mine-ness and care, access deliberately withholds them, and the firm
managing an access model must build value without the identification that ownership supplies. The
construct matters because the ownership-to-access shift undercuts the self–brand connection and
psychological-ownership mechanisms that much of this part relies on, demanding new accounts of how
value, loyalty, and brand meaning form when the consumer never owns anything.

## The Frontier-Construct Discipline {#sec-fr-discipline}

These six constructs share a methodological situation that is the chapter's closing lesson.
Frontier constructs are proposed faster than they are validated, and the field's incentive to
name a "new" construct outruns its incentive to prove the construct is distinct from its
neighbors. Customer experience risks absorbing satisfaction, quality, value, and engagement
without demonstrating it adds variance beyond their sum; brand authenticity overlaps trust and
self-connection; the AI-response constructs proliferate (aversion, resistance, word-of-machine,
and more) faster than their discriminant validity is established. The remedy is not to distrust
new constructs but to hold them to the *same* standard as old ones—the convergent, discriminant,
and nomological evidence of @sec-construct-vs-variable—with extra vigilance for the jingle–jangle
fallacies that the rush to name invites. Increasingly, frontier construct measurement is also
*computational*: experience, authenticity, and privacy concern are scored from text, behavior, and
multimodal traces rather than surveys, which relocates but does not repeal the validity problem,
exactly as @sec-cvv-frontiers argues. The frontier is where the constructs of tomorrow are formed,
and it is precisely there that the measurement theory of this part does its most important work.

## Key Takeaways {#sec-fr-takeaways}

- The construct map is open; **frontier constructs** earn admission by the same Cronbach–Meehl
  test as established ones, and—being youngest—carry the *highest* discriminant-validity and
  jingle–jangle risk, so the discipline of @sec-construct-vs-variable weighs more, not less.
- **Customer experience** [@lemon2016] is a multidimensional response across the whole journey
  and touchpoints; it subsumes satisfaction/quality/value/engagement and is hard to measure
  because it is dynamic, distributed, and multidimensional.
- **Psychological ownership**—felt mine-ness without legal title—drives valuation via the
  endowment effect, induced by mere touch [@peck2009] and empowerment/self-design [@fuchs2010]
  (@sec-fr-ownership-sim).
- **Brand authenticity** [@morhart2015] is a perceived four-dimensional judgment (continuity,
  credibility, integrity, symbolism), contested because it is often manufactured through
  inauthentic means.
- Consumer **responses to AI** form a construct family—algorithm aversion [@dietvorst2015],
  resistance to medical AI via uniqueness neglect [@longoni2019], and the word-of-machine effect
  [@longoni2022wordofmachine]—unified by *asymmetric trust in machine error/judgment*
  (@sec-fr-ai-sim).
- **Privacy concern** [@martinmurphy2017] centers the personalization–privacy paradox and a
  stated-vs-revealed convergent-validity gap; it links to @sec-privacy. **Access-based consumption**
  [@bardhi2012] inverts psychological ownership, withholding identification and stewardship and
  forcing new accounts of value and loyalty.
- Frontier measurement is increasingly **computational**, relocating but not repealing the
  validity problem (@sec-cvv-frontiers).

## Further Reading {#sec-fr-further}

Customer experience is @lemon2016. Psychological ownership runs through @peck2009 and @fuchs2010;
brand authenticity through @morhart2015. The consumer-AI family is @dietvorst2015, @longoni2019,
and @longoni2022wordofmachine, developed further in @sec-ai-ml. Privacy concern is
@martinmurphy2017 with the firm-performance evidence in @martin2017data and the regulatory
treatment in @sec-privacy; access-based consumption is @bardhi2012. The overarching methodological
frame—that frontier and computational constructs are governed by, not exempt from, classical
measurement theory—is laid out in @sec-construct-vs-variable and @sec-cvv-frontiers.
