# Innovation {#sec-innovation}

Innovation is the engine of long-run growth in marketing, and it is also one of
the field's hardest objects to measure and to manage. A new product is a bet:
the firm commits resources today against an uncertain stream of future demand,
and the value of that bet is realized—or destroyed—through the joint behavior of
consumers (who must adopt), competitors (who respond), and capital markets (who
price the firm's prospects long before the cash arrives). This chapter treats
innovation as the connective tissue between three literatures that are often read
separately: the **diffusion** of new technologies through a population of
adopters, the **organizational and strategic** determinants of who innovates and
how much, and the **financial** valuation of innovation as an intangible asset.

The chapter is organized around the life of an innovation. It begins with
*diffusion*—the canonical models of how an installed base accumulates over
time—because diffusion supplies the formal backbone on which forecasting,
substitution, and life-cycle tracking are built. It then turns to *measurement*:
how researchers quantify a construct that has no natural unit, leaning heavily on
the patent-citation machinery that the empirical literature has standardized. With
constructs and measures in hand, it examines the *new-product-development* (NPD)
process inside the firm—teams, capabilities, channels, co-creation, and
governance—and the *organizational* forces (acquisitions, franchising, going
public) that expand or erode a firm's capacity to innovate. It closes with the
*marketing–finance* linkage, where innovation is priced by investors and the
construct is finally expressed in returns. Throughout, intuition leads and the
formalism follows in full: a reader should leave able to specify a diffusion
model, defend a patent-based innovation measure against its known biases, and
reason about why markets reward some innovations seven times more than others.

By the end of the chapter the reader will be able to (i) write down and estimate
the Bass diffusion model and its successive-generation extension, (ii) construct
and critique patent-based innovation measures, and (iii) connect new-product
activity to firm value through the event-study logic developed at length in
@sec-marketing-finance.

## Theoretical Foundations {#sec-innovation-theory}

Before the formal diffusion machinery, it is worth naming the theories that give
that machinery its behavioral content. Four strands govern how an innovation spreads
and why some spread faster than others.

**Diffusion of innovations**, in Rogers's classic synthesis, describes adoption as a
process unfolding through communication channels over time across adopter categories:
innovators, early adopters, early majority, late majority, and laggards. The
**Bass model** [@Bass_1969] is the formal, estimable expression of exactly this idea:
its coefficient of innovation $p$ captures Rogers's external-influence (media,
advertising) channel and its coefficient of imitation $q$ captures the internal,
word-of-mouth contagion among prior adopters, as developed in @eq-04-bass-hazard.
Rogers supplies the sociology; Bass supplies the differential equation.

**The technology acceptance model** (TAM), Davis's adaptation of the theory of
reasoned action to information technology, holds that adoption intention is driven by
two beliefs: perceived usefulness and perceived ease of use [@davis1989tam]. TAM
explains adoption at the level of the *individual* decision that the Bass model
aggregates, and it is the microfoundation most often invoked when the innovation is
a digital product or platform.

::: callout-note
## Controversy: an entire journal issue on whether to retire one model

TAM is among the most-cited models in the social sciences, and information systems
devoted a special issue of the *Journal of the Association for Information Systems*
to arguing about whether it had become an obstacle. Marketing should care: TAM is
routinely used here as an off-the-shelf adoption model, and the objections are not
about IS specifically.

**1989 — the model.** @davis1989tam introduces perceived usefulness and perceived
ease of use as the determinants of acceptance. It is parsimonious, easy to
operationalize, and travels well, and it is applied to essentially every technology
over the following two decades.

**2007 — the special issue.** @hirschheim2007quovadis introduces nine pieces
running from page 203 to page 286 under the title *Quo Vadis TAM*. The framing
paper, @benbasat2007quovadis, argues that TAM's very success has become the
problem: a model this easy to apply crowds out theory development, generates
endless small variations, and diverts attention from the design characteristics
that actually determine whether a system is useful.

**The issue does not agree with itself.** @goodhue2007comment replies directly,
eight pages later, that the diagnosis overstates the damage. @bagozzi2007legacy—a
marketing scholar—argues the deeper defect is conceptual, in how the model links
intention to behavior, and proposes a paradigm shift rather than a repair. And the
closing piece is by TAM's own originator with co-authors, @venkatesh2007deadoralive,
asking *Dead or Alive?* about the research programme they built.

**The transferable lesson.** This is what happens when a model becomes
infrastructure: the reasons for its dominance—simplicity, easy measurement,
comparability across studies—are largely independent of whether it is the best
available explanation, and they compound. The same dynamic is documented in this
book for a service-quality instrument (@sec-eval-servqual) and a customer metric
(@sec-satisfaction). If you reach for TAM because it is standard, that is a reason
to check the 2007 issue first, not a reason to skip it.
:::

**Rogers's perceived attributes** explain the *rate* of diffusion, that is, why the
Bass $p$ and $q$ differ so sharply across categories. Five attributes predict adoption
speed: **relative advantage** (the improvement over what it replaces),
**compatibility** (fit with existing values and practices), **complexity** (inverse
ease of understanding and use), **trialability** (the degree to which it can be
experimented with before commitment), and **observability** (the visibility of its
benefits to others). High relative advantage, compatibility, trialability, and
observability, and low complexity, accelerate diffusion. These attributes are the
behavioral content behind an otherwise atheoretical $q/p$ ratio.

**Network effects** modify the diffusion logic for products whose value to each
adopter rises with the size of the installed base (communication platforms,
standards, marketplaces). Under direct or indirect network externalities the
imitation channel is not merely informational but *value-creating*: each new adopter
raises the relative advantage of adoption for everyone else, which steepens the
S-curve, can produce tipping and winner-take-most outcomes, and connects diffusion to
the platform dynamics treated elsewhere in the book.

## Diffusion of Innovations

Diffusion is the process by which an innovation spreads through a population of
potential adopters over time. The central empirical regularity is that cumulative
adoption traces an **S-curve**: slow at first, accelerating as word spreads, then
saturating as the pool of non-adopters is exhausted. The modeling task is to give
this curve a behavioral micro-foundation so that its parameters carry meaning and
can be used for forecasting.

### The Bass Model

The workhorse is the Bass model [@Bass_1969]. Its premise is that the propensity
to adopt at time $t$, conditional on not having adopted yet, is a linear function
of how many others have already adopted. Let $F(t)$ be the cumulative fraction of
the market that has adopted by time $t$, and let $h(t)$ denote the **hazard**—the
conditional rate of first adoption among those who have not yet adopted. The Bass
specification is

$$
h(t) \;=\; \frac{f(t)}{1 - F(t)} \;=\; p + q\,F(t),
$$ {#eq-04-bass-hazard}

where $f(t) = F'(t)$ is the adoption density. The two parameters carry distinct
behavioral content. The **coefficient of innovation** $p$ captures the tendency to
adopt independently of social influence—external pressure from advertising, media,
or intrinsic need—and governs the *level* of the curve at $t=0$. The
**coefficient of imitation** $q$ captures adoption driven by social contagion—word
of mouth, observation, network effects—and governs the *acceleration* once early
adopters seed the population. A scale parameter $m$, the ultimate **market
potential**, converts the fraction $F(t)$ into a count of adopters $N(t) = m\,F(t)$.

> The Bass model partitions adopters into *innovators*, who adopt under external
> influence captured by $p$, and *imitators*, whose adoption probability rises
> linearly with the cumulative number of previous adopters, captured by $q$
> [@Bass_1969].

Substituting $f(t) = h(t)\bigl(1 - F(t)\bigr)$ into @eq-04-bass-hazard yields the
ordinary differential equation $f(t) = \bigl(p + qF(t)\bigr)\bigl(1 - F(t)\bigr)$,
whose solution under $F(0)=0$ is the closed form

$$
F(t) \;=\; \frac{1 - e^{-(p+q)t}}{1 + \tfrac{q}{p}\,e^{-(p+q)t}}.
$$ {#eq-bass-cumulative}

The non-cumulative adoption rate $n(t) = m\,f(t)$ is hump-shaped whenever $q > p$,
peaking at $t^{\*} = \tfrac{1}{p+q}\ln(q/p)$; when $q \le p$ adoption declines
monotonically from launch and the S-curve degenerates. The interior peak is what
gives durable-goods sales their characteristic rise-and-fall, and the location of
that peak is the quantity managers most want to forecast.

**Estimation.** The original approach regresses discrete-time adoptions on the
installed base. Writing $n_t = m f(t)$ and $N_{t-1} = m F(t-1)$, the discretized
analogue of @eq-04-bass-hazard is the quadratic

$$
n_t \;=\; p\,m \;+\; (q - p)\,N_{t-1} \;-\; \frac{q}{m}\,N_{t-1}^{2} \;+\; u_t
\;=\; \beta_0 + \beta_1 N_{t-1} + \beta_2 N_{t-1}^{2} + u_t,
$$ {#eq-04-bass-ols}

which is linear in the composite parameters $\beta_0, \beta_1, \beta_2$ and can be
estimated by ordinary least squares (OLS). The structural parameters are then
recovered by inversion: $m$ solves $\beta_2 m^2 + \beta_1 m + \beta_0 = 0$ (taking
the positive root), $p = \beta_0/m$, and $q = -\beta_2 m$. OLS on
@eq-04-bass-ols is consistent only under strong assumptions, and three failures of
identification recur in practice. First, $N_{t-1}$ is mechanically correlated with
the error $u_t$ when adoption shocks are serially correlated, biasing the
coefficients—maximum likelihood or nonlinear least squares on @eq-bass-cumulative
is preferred. Second, the parameters are weakly identified before the sales peak:
without observations on both sides of $t^{\*}$, $m$ and $q$ trade off and estimates
are unstable, which is why early-life forecasts are notoriously fragile. Third, the
basic model assumes a *fixed* market potential and *constant* $p, q$; in reality
marketing-mix variables, price declines, and successive product generations all
shift these quantities, motivating the extensions below. @fig-bass-curve plots the
cumulative S-curve alongside the hump-shaped adoption rate it implies.

```{r bass-curve, message=FALSE, warning=FALSE}
#| label: fig-bass-curve
#| fig-cap: "Bass diffusion: cumulative adoption F(t) and the hump-shaped adoption
#|   rate n(t) for p = 0.03, q = 0.38, m = 1. The adoption rate peaks at the
#|   inflection point of the S-curve."
set.seed(1969)
p <- 0.03; q <- 0.38; m <- 1
t <- seq(0, 25, by = 0.1)

F_t <- (1 - exp(-(p + q) * t)) / (1 + (q / p) * exp(-(p + q) * t))
f_t <- (p + q * F_t) * (1 - F_t)   # adoption density n(t)/m

t_peak <- log(q / p) / (p + q)

op <- par(mar = c(4, 4, 2, 4))
plot(t, F_t, type = "l", lwd = 2, ylim = c(0, 1),
     xlab = "Time since launch", ylab = "Cumulative fraction F(t)")
par(new = TRUE)
plot(t, f_t, type = "l", lwd = 2, lty = 2, col = "grey40",
     axes = FALSE, xlab = "", ylab = "")
axis(4); mtext("Adoption rate n(t)/m", side = 4, line = 2.5)
abline(v = t_peak, col = "red", lty = 3)
legend("topleft", legend = c("Cumulative F(t)", "Rate n(t)/m", "Peak"),
       lty = c(1, 2, 3), lwd = c(2, 2, 1),
       col = c("black", "grey40", "red"), bty = "n")
par(op)
```

::: callout-note
## Controversy: can you estimate a Bass model before the peak?

The model above is the most successful forecasting tool marketing has produced, and
the most consequential thing to know about it is a limitation established twenty-eight
years after it appeared. It bears directly on the only use managers actually have for
it, which is forecasting a launch that has not peaked yet.

**1969 — the model.** @Bass_1969 publishes the two-parameter formulation in
*Management Science*. It fits historical diffusion curves remarkably well across
categories, and that fit is the basis of its adoption.

**1997 — the identification problem.** @vandenbulte1997bias show that the parameter
estimates are biased and, worse, *systematically change* as more data accumulate.
Estimates of $p$, $q$ and the market size $m$ obtained before the adoption peak are
not merely imprecise—they drift in predictable directions as the series lengthens.
The mechanism is that $p$, $q$ and $m$ are jointly weakly identified from the
pre-peak portion of an S-curve: many combinations trace nearly the same early path
and diverge only afterwards.

**1998 — an erratum follows** [@vandenbulte1998erratum], which is itself worth
noticing in a chapter about diffusion: the correction is part of the public record.

**2004 — the author looks back.** @bass2004comments revisits the model thirty-five
years on in a retrospective for *Management Science*, an unusual document in which
the originator assesses what the model did and did not establish.

**The practical consequence, stated bluntly.** Excellent in-sample fit to a completed
diffusion curve is close to uninformative about forecasting ability for an incomplete
one, because the fit is achieved by a parameter combination that the early data
cannot pin down. A Bass forecast made before the peak should be reported with the
range of $(p, q, m)$ combinations consistent with the data so far, not as a point
estimate. @sec-strategic-dynamic-models develops the identification issue formally,
and @sec-cred-marketing-debates places this among the field's standing methodological
disagreements.
:::

### Successive Generations and Substitution

Most durable categories do not diffuse once; they diffuse repeatedly as new
technological generations replace old ones (mainframes to minicomputers to PCs;
3G to 4G to 5G). A model of a *single* generation misattributes the decline of an
incumbent technology to saturation when it is in fact substitution. @Chandrasekaran_2020
develop a successive-generations framework that separates two rates the basic Bass
model conflates: the rate at which adopters **disengage** from an older technology
and the rate at which they **adopt** the newer one. Allowing these to differ is
what lets the model fit the empirically observed overlap, in which an older
generation is still gaining late adopters even as the newer one accelerates—an
overlap a single substitution rate cannot reproduce.

The practical payoff is life-cycle tracking. @meade2004 review the forecasting
toolkit for the technology-adoption life cycle and stress that model choice should
follow the data regime: pre-peak data demand methods robust to the weak
identification noted above, whereas post-peak data permit richer multi-generation
specifications. The diffusion literature thus supplies not a single model but a
ladder of models indexed by how much of the life cycle the analyst can observe.
@fig-successive-generations sketches how each generation carries its own adoption
process while a distinct disengagement rate governs substitution.

```{mermaid}
%%| label: fig-successive-generations
%%| fig-cap: "Successive-technology diffusion. Each generation has its own adoption
%%|   process; substitution is governed by a disengagement rate from the incumbent
%%|   that is distinct from the adoption rate of the successor."
flowchart LR
    NP["Non-adopters<br/>(market potential m)"] -->|"adopt Gen 1<br/>(p₁, q₁)"| G1["Gen 1<br/>installed base"]
    G1 -->|"disengage"| SW["Switching pool"]
    NP -->|"adopt Gen 2<br/>(p₂, q₂)"| G2["Gen 2<br/>installed base"]
    SW -->|"upgrade"| G2
    G2 -->|"disengage"| G3["Gen 3 …"]
```

## Growth Curves Beyond Bass: Trend Extrapolation for a New Channel {#sec-growth-curves}

The Bass model is one member of a larger family of **growth curves**, and the family
matters whenever the task is not to explain adoption but to *project a young,
saturating series forward*. A new product's cumulative sales, a technology's installed
base, and—as @kaiser2026chatgpt show—a new marketing *channel's* conversion rate all
trace an S: near-flat at launch, a steepening middle, a plateau as the addressable
ceiling binds. The forecasting difficulty is that early data show only the bottom of
the S, and the functional form the analyst imposes on the trend, not the data, then
decides where the curve is headed. Kaiser and Schulze make the point concrete by
projecting organic-LLM referral outcomes under four trend specifications and reporting
that the choice "substantially impacts predictions." Those four specifications are
worth stating precisely, because they span the space of sensible answers.

The two **linear** forms are the null against which the curves earn their keep. A
plain linear trend and its *centered* variant,
$$
y_t = \beta_0 + \beta_1 t
\qquad\text{and}\qquad
y_t = \tilde\beta_0 + \beta_1 (t - \bar t),
$$ {#eq-04-growth-linear}
fit the identical line—centering shifts only the intercept's meaning (from the
extrapolated value at $t=0$ to the level at the sample midpoint) and decorrelates
intercept from slope, which stabilizes the starting values the nonlinear fits below
require. Their shared defect is fatal for a saturating quantity: a straight line has
no ceiling. Extrapolated a year past a nascent channel's data, a linear conversion
rate marches past every plausible bound, and a share past 100%. Centering cures none
of this; it is a conditioning device, not a model of saturation.

The **logistic** (sigmoid) curve is the symmetric saturating alternative. It is the
solution of the autocatalytic law $\dot y = \tfrac{k}{L}\,y\,(L - y)$—growth
proportional to both the level attained and the headroom remaining—giving
$$
y_t = \frac{L}{1 + \exp\!\big(-k\,(t - t_0)\big)},
$$ {#eq-04-growth-logistic}
with a **hard upper limit** $L$ (the ceiling, imposed a priori or estimated), growth
rate $k$, and midpoint $t_0$ at which $y = L/2$. Its inflection sits exactly halfway
up, so a logistic rises and levels off as mirror images. The **Gompertz** curve
relaxes that symmetry. It solves $\dot y = c\,y\,\ln(a/y)$—growth proportional to the
*log* headroom—so the brake engages later and the approach to the asymptote is slower
than the take-off:
$$
y_t = a\,\exp\!\big(-b\,\exp(-c\,t)\big),
$$ {#eq-04-growth-gompertz}
with asymptote $a$, and inflection at $t = (\ln b)/c$ where $y = a/e \approx 0.368\,a$,
below the logistic's half-way point—the signature of an **asymmetric S-curve**
[@gompertz1825nature]. A channel that matures fast then crawls toward its ceiling is
Gompertz-shaped; one that accelerates and decelerates evenly is logistic. The data
rarely say which until the plateau is in view, and that is the whole problem: before
the inflection, market potential and growth rate trade off just as $m$ and $q$ do in
the Bass model (@eq-bass-cumulative), so competing curves fit the observed rise almost
equally well while disagreeing sharply about the ceiling. The honest output of an
early-life projection is therefore an *envelope* of forms, not a single line.

**Estimation** is ordinary least squares for the linear forms and nonlinear least
squares for the curves—the same `nls` machinery used for the cumulative Bass model,
with the asymptote and rate given explicit starting values (or supplied by R's
self-start models `SSlogis` and `SSgompertz`). @fig-growth-curves fits all four to a
simulated year of a maturing channel and projects each a year beyond the data.

```{r growth-curves, message=FALSE, warning=FALSE}
#| label: fig-growth-curves
#| fig-cap: "Four trend specifications fit to twelve months of a maturing channel's
#|   monthly conversion rate (points), then projected twelve months past the end of the
#|   data (dotted vertical line); the dashed horizontal line is the true 3% ceiling. In
#|   sample the fits nearly coincide; out of sample the linear trend climbs past any
#|   plausible ceiling while the logistic and Gompertz curves saturate—at different
#|   levels, which is the weak identification of a pre-inflection series made visible."
set.seed(2024)

# Twelve months of a new channel's monthly conversion rate, truly approaching a
# ceiling of 3% along a logistic path, observed with noise.
months  <- 1:12
L_true  <- 0.030; k_true <- 0.55; t0_true <- 7
cr_true <- L_true / (1 + exp(-k_true * (months - t0_true)))
cr_obs  <- pmax(cr_true + rnorm(12, 0, 0.0012), 1e-4)
d <- data.frame(t = months, cr = cr_obs)

ctrl <- nls.control(maxiter = 200, warnOnly = TRUE)   # return a fit rather than abort
fit_lin <- lm(cr ~ t, data = d)
fit_cen <- lm(cr ~ I(t - mean(t)), data = d)          # identical fit, centered trend
fit_log <- nls(cr ~ L / (1 + exp(-k * (t - t0))),
               start = list(L = 0.03, k = 0.4, t0 = 7), data = d, control = ctrl)
fit_gom <- nls(cr ~ a * exp(-b * exp(-c * t)),
               start = list(a = 0.03, b = 5, c = 0.4), data = d, control = ctrl)

newt <- data.frame(t = 1:24)
pr <- sapply(list(Linear = fit_lin, Logistic = fit_log, Gompertz = fit_gom),
             function(m) predict(m, newdata = newt))

matplot(newt$t, pr, type = "l", lwd = 2, lty = 1,
        col = c("firebrick", "steelblue", "darkgreen"),
        xlab = "Month since launch", ylab = "Monthly conversion rate",
        ylim = range(0, pr, cr_obs))
points(d$t, d$cr, pch = 19)
abline(v = 12, lty = 3, col = "grey50")
abline(h = L_true, lty = 2, col = "grey60")
legend("topleft", bty = "n", lwd = 2,
       col = c("firebrick", "steelblue", "darkgreen", "black", "grey60"),
       lty = c(1, 1, 1, NA, 2), pch = c(NA, NA, NA, 19, NA),
       legend = c("Linear", "Logistic", "Gompertz", "Observed", "True ceiling"))

cat(sprintf("Projected CR at month 24 — linear %.3f, logistic %.3f, gompertz %.3f\n",
            pr[24, "Linear"], pr[24, "Logistic"], pr[24, "Gompertz"]))
cat(sprintf("Centered-linear reproduces the linear fit exactly: %s\n",
            isTRUE(all.equal(unname(fitted(fit_lin)), unname(fitted(fit_cen))))))
```

The fits are nearly indistinguishable over the twelve observed months yet part company
violently once extrapolated: the linear projection sails through the true ceiling while
the two S-curves flatten toward different asymptotes. That spread *is* the forecast
uncertainty a single fitted line hides, and it is why fitting the family—and reporting
where the members agree and where they diverge—is the disciplined move for any young,
saturating series.

::: {.callout-note}
## The trend-extrapolation discipline

A saturating quantity demands a saturating model. Never linear-extrapolate a rate, a
share, or a penetration figure: the linear form has no ceiling and will always, given
enough horizon, predict the impossible. Fit at least one symmetric (logistic) and one
asymmetric (Gompertz) curve, treat the gap between their projected asymptotes as the
honest forecast interval, and distrust any point forecast made before the series has
shown its inflection—the region where, exactly as in the Bass model, the ceiling and
the growth rate are only weakly separated by the data.
:::

These curves are the reduced-form cousins of the Bass model of @sec-innovation-theory.
Where Bass decomposes the S into interpretable innovation and imitation rates
(@eq-04-bass-hazard), the logistic and Gompertz fit the shape directly, which is the
pragmatic choice when only cumulative levels are observed and no adopter-level story is
needed. @kaiser2026chatgpt turn exactly this apparatus on a marketing *channel* rather
than a product—projecting how fast LLM referral converges toward established channels—
and pair it with the overdispersed rate model of @sec-quasibinomial to keep the
inference honest on sparse data. Diffusion logic, it turns out, describes the maturation
of marketing infrastructure as readily as the adoption of the products that flow
through it.

## Earned Media and the Local Diffusion of a Green Innovation {#sec-innovation-news-diffusion}

Bass and its descendants put the diffusion mechanism inside the consumer population:
imitation is contagion among adopters. For a great many products that is only half the
story, because the consumer cannot adopt until a local business decides to *carry* the
thing. Plant-based meat is the clean case. A household's willingness to try an
Impossible burger is irrelevant in a county where no restaurant puts one on the menu,
so the object to be explained is a **supply-side adoption decision** by thousands of
small, independent merchants—each of whom faces a fixed cost of menu redesign,
supplier onboarding, and staff training, and each of whom is guessing at local demand.

@xu2026food study exactly this margin for plant-based ("impossible") meat over
2015–2019, and their first contribution is a measurement one. There is no registry of
which restaurants and stores carry the product, so they construct a location-specific
adoption metric from the merchants' *own social-media announcements*—the post that says
the burger is now on the menu—which turns an unobservable rollout into a
county-quarter panel. The substantive question is then whether **local news coverage**
of the product causes these merchants to adopt it.

The obstacle is that coverage is not assigned at random. Editors write about what their
audience already cares about, so a county with latent enthusiasm for green food
generates both more column inches and more adoption; and once a few restaurants adopt,
that itself becomes local news. Coverage and adoption are jointly determined by an
unobserved local taste, and reverse causality runs alongside it. The authors' answer is
to instrument coverage with **quasi-random variation in county-quarter news production
across topics**. Local news space is a rivalrous resource: an outlet has a fixed number
of slots per quarter, so a flood of unrelated newsworthy events crowds the focal topic
out, while the producer's own national news flow pushes it in. This is the
news-crowd-out logic of @eisensee2007news, refined to the county-quarter-topic cell and
resting on the local-press-supply tradition of @snyder2010press. The topic decomposition
is what makes the exclusion restriction credible: the identifying variation comes
predominantly from news about the **producer's financials**—funding rounds, valuations,
earnings—which is generated in corporate boardrooms and financial markets, not by
editorial interest in the county's food scene, and which has no plausible direct channel
to a local restaurateur's menu except through the coverage itself.

```{mermaid}
%%| label: fig-news-adoption-id
%%| fig-cap: "Identifying the effect of local news on local adoption. The unobserved local taste for green food (dashed) confounds coverage and adoption in both directions. The two news-supply shifters enter only through coverage: the producer's national financial news flow, scaled by the county's press capacity, and crowd-out from rival topics competing for the same finite news hole."
flowchart LR
    F["Producer's national\nfinancial news flow\n(funding, valuations, earnings)"]
    O["County press capacity\n(outlets, newsroom size)"]
    X["Rival-topic news pressure\n(crowd-out of the news hole)"]
    N["Local coverage of\nthe green product"]
    A["Local merchant adoption\n(menu / shelf announcements)"]
    T["Unobserved local\ntaste for green food"]
    L["County ideology\n(liberal vote share)"]
    F --> N
    O --> N
    X -- "crowds out" --> N
    N --> A
    T -. "confound" .-> N
    T -. "confound" .-> A
    A -. "reverse causality" .-> N
    L -- "moderates" --> A
```

Three results carry beyond the setting. First, local coverage does raise adoption by
restaurants and stores, and the magnitude is of the same order as what packaged-goods
firms buy with television advertising—except that it is *earned*, which reframes public
relations from a reputational function into a distribution-building one. Second, the
effect is markedly stronger in **liberal-leaning counties**. Green-product diffusion is
therefore spatially sorted by ideology and not only by the usual income and density
gradients, echoing the finding that political identity conditions the appeal of
environmental attributes even when the private economics are held fixed
[@gromet2013ideology]. A national average effect is a weighted mixture over a
politically partitioned map, and a rollout plan built on that average will
systematically over-invest in one half of the country and under-invest in the other.
Third—and this is the methodological payoff—**the topic composition of news is
information, not nuisance**. Decomposing coverage by topic told the authors which slice
of the variation was doing the identifying work, and that slice is the one whose
exclusion restriction can be argued from institutional detail.

The replication below builds the confound deliberately, then removes it. A panel of
counties over twenty quarters carries an unobserved green-taste shock that raises both
coverage and adoption; coverage is additionally moved by two supply-side shifters that
have no business in the adoption equation. Ordinary least squares with county and
quarter fixed effects cannot rescue the estimate—the confound varies *within* county
over time—while two-stage least squares on the news-supply instruments recovers the
truth. The last block interacts the endogenous regressor with county ideology, using
interacted instruments, to recover the political moderation.

```{r innovation-news-iv, message=FALSE, warning=FALSE}
set.seed(404)
library(fixest)

# --- A county-quarter panel of local-business adoption ------------------------
n_cty <- 400; n_qtr <- 20
cty <- rep(seq_len(n_cty), each = n_qtr)
qtr <- rep(seq_len(n_qtr), times = n_cty)
N   <- n_cty * n_qtr

# County traits, both time-invariant: liberal vote share and local press capacity.
lib     <- rep(plogis(rnorm(n_cty, 0, 0.8)), each = n_qtr)
outlets <- rep(exp(rnorm(n_cty, 0, 0.5)), each = n_qtr)

# Two exogenous shifters of local news SUPPLY, neither of which belongs in the
# adoption equation:
#  (1) the producer's national financial news flow (funding rounds, earnings),
#      which reaches a county in proportion to its press capacity;
#  (2) crowd-out -- unrelated topics competing for the same county-quarter news hole.
fin_national <- rep(rnorm(n_qtr, 0, 1), times = n_cty)
z_fin   <- scale(log(outlets) * fin_national)[, 1]
z_crowd <- scale(rgamma(N, shape = 2, scale = 1))[, 1]

# The confound: unobserved local taste for green food, which moves BOTH the
# editor's willingness to cover the product AND the merchant's willingness to stock it.
taste <- 0.6 * scale(lib)[, 1] + rnorm(N, 0, 1)

# Observed local coverage of the product: the endogenous regressor.
news <- 0.75 * z_fin - 0.35 * z_crowd + 0.50 * taste + rnorm(N, 0, 0.5)

# Adoption: local restaurants and stores announcing the product this quarter.
beta_main <- 0.25   # true average effect of coverage
beta_lib  <- 0.60   # true extra effect per unit of (demeaned) liberal share
alpha_c   <- rep(rnorm(n_cty, 0, 0.7), each = n_qtr)
tau_t     <- rep(rnorm(n_qtr, 0, 0.4), times = n_cty)
lib_c     <- lib - mean(lib)
adopt <- (beta_main + beta_lib * lib_c) * news +
         alpha_c + tau_t + 0.8 * taste + rnorm(N, 0, 0.8)

d <- data.frame(adopt, news, lib_c, z_fin, z_crowd, cty, qtr)

# --- (1) OLS with county and quarter fixed effects: still confounded ----------
m_ols <- feols(adopt ~ news | cty + qtr, data = d)

# --- (2) 2SLS: instrument coverage with the two news-supply shifters ----------
m_iv <- feols(adopt ~ 1 | cty + qtr | news ~ z_fin + z_crowd, data = d)

# --- (3) Which instrument does the identifying work? --------------------------
Fstat   <- function(m) fitstat(m, "wf")$wf$stat
f_fin   <- feols(news ~ z_fin   | cty + qtr, data = d)
f_crowd <- feols(news ~ z_crowd | cty + qtr, data = d)

# --- (4) Ideology heterogeneity: two endogenous terms, two interacted instruments
m_het <- feols(adopt ~ 1 | cty + qtr |
                 news + news:lib_c ~ z_fin + z_crowd + z_fin:lib_c + z_crowd:lib_c,
               data = d)

cat(sprintf("True average effect of coverage        : %.3f\n", beta_main))
cat(sprintf("OLS, county + quarter FE               : %.3f\n", coef(m_ols)["news"]))
cat(sprintf("2SLS, news-supply instruments          : %.3f  (SE %.3f)\n",
            coef(m_iv)["fit_news"], se(m_iv)["fit_news"]))
cat(sprintf("First-stage F -- producer financials   : %.0f\n", Fstat(f_fin)))
cat(sprintf("First-stage F -- rival-topic crowd-out : %.0f\n", Fstat(f_crowd)))
cat(sprintf("True ideology interaction              : %.3f\n", beta_lib))
cat(sprintf("2SLS ideology interaction              : %.3f  (SE %.3f)\n",
            coef(m_het)["fit_news:lib_c"], se(m_het)["fit_news:lib_c"]))
```

Fixed effects absorb the level of a county's enthusiasm and the national trajectory of
the category, and they still leave the OLS coefficient at roughly twice the truth,
because the taste shock that drives coverage moves quarter by quarter inside the county.
Two-stage least squares on the two news-supply shifters lands on the true parameter, and
the first-stage diagnostics reproduce the paper's finding that producer-financial news
carries the identification: it is by far the stronger instrument, and it is also the one
whose exclusion restriction survives an institutional argument. The interacted
specification then recovers the ideology moderation, which is the practical output—it
converts a single national elasticity into a map.

Two cautions travel with the design. The instruments are only as good as the claim that
national corporate news reaches a county solely through the local news hole; if the same
funding round moves the *supplier's* wholesale terms or the parent chain's national menu
policy, the exclusion restriction fails and no first-stage F will reveal it. And because
the estimate is a local average treatment effect, it is identified off merchants whose
adoption decision is responsive to coverage—plausibly the undecided middle, not the
committed early adopters that a Bass model's innovation parameter is meant to capture.
The two frameworks are complements: @sec-innovation-theory characterizes the shape of
the curve, this design identifies what moves it.

## Measuring Innovation {#sec-innovation-measure}

Innovation has no natural unit, so every empirical claim about it rests on a
constructed measure, and the credibility of the claim is only as strong as the
measure's construct validity. Three families dominate. **Input** measures (R&D
expenditure) capture effort but not outcome and are reported inconsistently across
firms. **Output** measures based on new-product introductions—drawn from
commercial databases such as product-launch analytics [@wies2023innovation]—capture
commercialized innovation but miss the upstream invention that precedes it.
**Patent-based** measures sit between input and market outcome and have become the
field's default because patents are dated, classified, linked to assignees, and
connected to one another through citations.

### Patent Counts and Their Truncation Problems

The canonical pipeline, exemplified by @chang2021board, draws granted-patent
records (historically from the NBER patent database covering U.S. Patent and
Trademark Office grants) and proxies a firm's annual innovation by two quantities:
the **count** of patents the firm applied for in a year and ultimately had granted,
and the **citations** those patents subsequently receive. Two measurement
decisions are load-bearing. First, patents are dated by *application* year rather
than grant year, because application date is closer to the moment of inventive
activity. Second, both measures are logged to tame the heavy right tail of the
patent distribution.

Two truncation biases must be corrected or the estimates are not interpretable.
The **application–grant lag** truncates recent application years (patents applied
for late in the sample have not yet been granted), so the sample is typically
ended a few years before the data cutoff. The **citation-accumulation lag**
truncates forward citations for recent patents (they have had less time to be
cited), which is addressed with technology-class-by-year fixed effects that net out
the systematic component of citation timing. These are not cosmetic adjustments:
without them, a spurious downward trend in innovation appears at the end of every
sample.

### Backward Citations: Novelty

A patent cites *prior art*. The **backward** citations a patent makes—the patents
it builds on—index how much existing knowledge it recombines, and thus inversely
index its novelty.

> Citing propensity is defined as the average number of backward citations
> excluding citations to public research per patent; it reflects a firm's overall
> level of citing [@roach2013lens, p. 512].

@chandy2006invention use the same logic at the patent level: novelty is measured
by the number of backward citations a patent makes, with *higher* backward
citations implying *lower* novelty, on the reasoning that an idea drawing heavily
on existing patents is by construction less of a departure from what came before.
Backward counts are available at application time, which makes them attractive as
an *ex ante* measure—unlike the forward citations discussed next.

### Forward Citations: Importance

The **forward** citations a patent receives—later patents that cite it—index its
downstream influence, and the literature treats them as the best available proxy
for the *importance* or *quality* of an invention rather than its mere volume.
@tellis2009radical and @cao2023innovation both weight patents by forward citations
on precisely this ground, and @ahuja2001entrepreneurship operationalize
**breakthrough** inventions as the top 1% of patents by citation weight within
their application year. The external validity of the measure is anchored by
@trajtenberg1990penny, who shows that the more citations a CT-scanner patent
receives, the higher its correlation with the technology's realized improvement
surplus—a rare direct link between a citation count and an independent welfare
measure [@lampe2016patent]. @chandy2006invention summarize idea *importance* as the
total forward-citation count for a patent.

Forward citations have well-understood virtues and a sharp limitation, which the
following table organizes.

| Property | Forward citations | Backward citations |
|---|---|---|
| Construct measured | Importance / quality / influence | Novelty (inverse) |
| Timing | *Ex post* (accrue after grant) | *Ex ante* (fixed at application) |
| Manipulability by firm | Low—examiners verify prior art; over-citing narrows own scope | Higher—applicants control what they cite |
| Principal bias | Right-truncation: older patents accrue more citations | Less time-sensitive |
| Standard fix | Long window (≈17+ yrs) or class-year fixed effects | — |

: Forward versus backward patent citations as innovation measures {#tbl-citations}

As @tbl-citations records, the chief threat to forward-citation measures is
**temporal**: because citations accrue over time, older patents mechanically look
more important. The standard remedy is a citation window long enough—on the order
of seventeen years or more—that the bias from differential accumulation is
negligible, supplemented by the class-year fixed effects noted above. The
*ex post* nature of the measure remains a genuine caveat: forward citations cannot
recover the importance an inventor or investor could have perceived *at the time*
of patenting, which is exactly the quantity a forward-looking valuation would want.
@galasso2011ceo exploit this measurement apparatus to study how managerial risk
exposure shapes the quantity and the citation-weighted quality of innovation.

### Measuring Creativity with Embeddings {#sec-innovation-creativity-embeddings}

Patent measures capture *firm* invention, but most consumer creativity leaves no patent
and no citation trail: a recipe, a playlist, a graphic-design template, an outfit.
Representation learning fills the gap. @sozuer2026recipe learn an **ingredient
embedding**—each ingredient's vector inferred from the company it keeps across more than
57,000 recipes, the same representation-learning idea as the product embeddings of
@sec-bundle-rec applied to a dish's components—and from that geometry read two properties
of a recipe: its **fit** (how coherently its ingredients belong together) and its
**novelty** (how unusual its ingredients are for its category). The measure needs no
expert taster and no citation graph; it recovers creativity from the statistics of
recombination, exactly as the backward citations above recover a patent's novelty from
the prior art it recombines.

The payoff is a result that complicates the folklore that creativity is uniformly
rewarded. High ingredient fit lifts *both* trial and ratings, but **unique** ingredients
*cut* trial while *raising* ratings among those who do try—novelty carries an adoption
cost and a quality dividend at once. The recipe that wins is therefore not the most novel
but the most *balanced*: novel in its parts, coherent as a whole. This is the
innovation-side echo of the atypicality-and-engagement result in social content
(@sec-effectiveness-drivers)—distinctiveness pays, but only when it stays legible against
a familiar ground—and it turns the embedding from a descriptive gauge into a design tool:
the authors build a generator that proposes adding, dropping, or swapping an ingredient
to move a recipe toward that balanced-creativity frontier.

The measurement move here—define a construct, operationalize it as a geometric property
of a learned embedding, validate it against outcomes it was not fit to, and end in a tool
rather than a coefficient—recurs across the creator economy. Its pricing counterpart is
@sikdar2026artwork, who fuse a listing's structured attributes, photographs, and text to
tell a novice artist what to charge and how long the work will take to sell
(@sec-mmf-artpricing). Both study creators without institutional credentials, both let a
learned representation stand in for the expert judgment such creators cannot buy, and
both are careful to remain associational: the creator chooses the artifact and its
description jointly, so the representation is entangled with the unobserved skill that
also drives success. The general statement of that template, and the conditions under
which reading a representation is measurement rather than the interpretability error,
is in @sec-aiml-representation-measurement.

## New Product Development

A diffusion curve and a patent count describe innovation from the outside. Inside
the firm, innovation is produced by **new-product-development** processes whose
design—who is on the team, what information they have, how decisions are made—shapes
whether the resulting product is innovative enough to survive. The recurring
finding is that NPD outcomes are *contingent*: capabilities and team structures
have value not in the abstract but conditional on the informational and
competitive environment.

### Capabilities, Teams, and the Information Environment

@moorman1999contingency advance a contingency view of NPD capability. Rather than
asking whether technological and marketing capabilities raise performance on
average, they ask *when*: the informational content of the external environment
modulates how firms deploy their capabilities, affecting both the magnitude and the
pace of product development. The value of a capability, on this account, lies in
its adaptive deployment against environmental cues, not in its mere existence—a
theme that recurs across the NPD literature.

Team composition is the next lever. @sethi2001cross study cross-functional NPD
teams and locate the primary cause of new-product failure in a *lack of
innovativeness*, defined as the delivery of uniquely meaningful benefits.
Innovativeness rises with a strong superordinate team identity, encouragement to
take risks, customer influence, and active senior-management monitoring—but it
*falls* with high social cohesion, because cohesive teams suppress the dissent that
generates novel options. The interactions are as important as the main effects: the
benefit of a superordinate identity is amplified by risk encouragement and damped
by social cohesion, while functional diversity per se does not move innovativeness.
The emphasis on risk encouragement echoes a robust finding from the creativity
literature, that the *expectation of evaluation* lowers creative output
[@amabile1990]: teams that anticipate being judged retreat to safe, conventional
options, which is precisely the suppression of novelty that undermines
innovativeness.
@slotegraaf2011product add a temporal dimension, showing that the relationship
between team **stability** and the comprehensiveness of decision making is
*curvilinear*—too little stability fragments knowledge, too much breeds
complacency—and that decision comprehensiveness improves new-product advantage,
especially at its upper range.

A structural caution runs through this literature. @grewal2013environments show
that the often-mixed findings on how market orientation interacts with the
environment can reflect **unobserved heterogeneity**: estimating finite-mixture
(latent-class) regressions reveals multiple regimes with different effect signs, so
that pooled estimates suffer aggregation bias. The methodological lesson
generalizes—latent heterogeneity in firms' innovation responses can masquerade as a
null or unstable average effect, and disaggregated estimation is the corrective.

### Channels, Co-creation, and Forecasting

NPD does not end at the factory gate. @luo2007new examine new-product development
under the constraint of **channel acceptance**, recognizing that a product that
retailers will not stock cannot diffuse regardless of consumer appeal.
@ho2020using study **continuous-improvement products** (CIPs)—products engineered
for post-purchase improvement rather than full replacement—and show that sharing
*development-progress* information raises adoption of the current version, mediated
by consumers' perception of the developer's commitment to ongoing improvement and
moderated by product familiarity. Disclosure, in other words, is itself a marketing
instrument.

Co-creation has a demand-side counterpart that the firm-side literature sometimes
skips: whether consumers actually value the output. @schreier2006valueincrement
measures the willingness-to-pay increment for mass-customized over standard products
and finds it substantial but heterogeneous, which is what makes customization a
segmentation decision rather than a universal upgrade. Where the firm solicits ideas
rather than configurations, the design of the solicitation dominates the volume of
it: @toubia2006ideageneratio shows that incentive schemes rewarding *others'* use of
an idea, rather than the idea itself, raise both the quantity and the creativity of
what is submitted.

Two comparative studies bound how far any of this generalizes.
@parry1997crossnational compare NPD practices across nations and find the
practice–performance links are not invariant, and @nijssen2006exploringproduct show
that product and service innovation differ enough in their antecedents that a single
NPD framework applied to both will mis-specify one of them. On the organizational
antecedent, @kim1998marketorientati ask whether the market orientation–performance
link is mediated by innovation rather than direct — the answer largely determining
whether a firm should invest in listening or in building.

On the demand-sensing side, @kim2016brand challenge the static view of consumer
co-creation. Their concept of **brand-embedded interaction**—dynamic, personalized
engagement that steers consumers toward ideas satisfying both their own needs and
the brand's aspirations—yields more *constructive* NPD ideas in two Twitter-based
quasi-field experiments than passive solicitation does. Co-creation quality, that
is, is endogenous to how the firm structures the interaction.

Forecasting a durable good *before* launch is the hardest demand-sensing problem,
because there is no sales history to extrapolate. @harz2021 deploy **virtual
reality** (VR) for prelaunch sales forecasting of durables, arguing that VR's
*visualization* capability (simulation scope, similarity to reality, immersion) and
*automated-tracking* capability (interactivity) induce behavioral consistency
between participants' information search, stated preferences, and purchase behavior.
The forecasting backbone is the **macro-flow model** (@fig-macro-flow), which projects new-product
sales over time by specifying a set of behavioral states (awareness, preference,
purchase), the flows between them, and the determinants of those flows
[@urban1990]. @harz2021 extend this model with VR-derived virtual purchases and
third-party awareness inputs and find improved accuracy—after adjusting for
advertising—at both aggregate and individual levels, with *presence* and *vividness*
as the operative psychological mechanisms.

```{mermaid}
%%| label: fig-macro-flow
%%| fig-cap: "The macro-flow model underlying prelaunch forecasting: consumers flow
%%|   through behavioral states, and the model estimates the determinants of each
%%|   transition. Virtual-reality simulations supply the flow inputs before launch
%%|   [@urban1990; @harz2021]."
flowchart LR
    A["Aware"] -->|"info search"| K["Knowledgeable /<br/>Preference formed"]
    K -->|"intend"| I["Purchase intent"]
    I -->|"trial / virtual purchase"| P["Purchase"]
    P -->|"satisfaction"| R["Repeat / WOM"]
    R -.->|"contagion"| A
```

### Governance and Organizational Structure

Innovation is governed from the top of the firm. @srinivasan2018corporate study
**board interlock centrality**—how connected a firm's directors are to other
boards—and find it raises incremental new-product introductions by channeling
market intelligence into the firm; the effect strengthens with internal and
marketing leadership on the board and with a CEO of marketing background, and
weakens under pronounced intra-industry external leadership. @you2020 push the
upper-echelons logic further, showing that CEO and CMO characteristics—personality,
demographics, experience, values—shape both innovation and the stock returns it
generates. At the front end of strategy, @molner2018 show that managers' **market-
scoping mindset** conditions which markets an innovation reaches: ambiguity
*avoidance* pulls managers toward a downstream, end-user orientation that courts
commercialization failure, whereas ambiguity *acceptance* fosters an upstream
orientation that uncovers viable market spaces.

Structure also has to absorb distance. @anderson2018distributed analyze qualitative
data from 70 distributed product-development projects spanning 14 countries and
involving team members who speak 10 different languages, and find that the
integration strategies usually prescribed for such projects, modular product
architectures and colocating team members, are by themselves insufficient to
coordinate the work. What their field interviews point to instead is investment in
*design information systems* built with features that carry design intent across time
zones and languages, together with project managers empowered to act on them. The
finding is a caution for any account of new-product development that treats
modularity as a complete answer to coordination. Modularity partitions the work, but
it does not tell a team in one time zone what a team in another meant, and the
residual has to be carried by an information system and by authority.


### Generative AI in the Front End: Effort Saved Is Not Creativity Gained {#sec-innovation-genai-ideation}

The front end of the funnel above is where generative AI has been adopted fastest,
on the reasoning that turning a vague idea into a developed concept is laborious and
the labor can be offloaded. @gordetzki2026ideation test that reasoning and find it
backwards in an instructive way.

In an online experiment, 276 participants refined ideas for an innovation challenge
either alone, with AI-generated *text*, or with AI-generated *images*. Image-based
support did what it was supposed to do on the effort margin: it significantly reduced
effort relative to both working alone and text-based support. It also produced ideas
that were **18% less creative** than those from the more effortful text-based
support. The explanation is about what each format leaves undone. An image specifies
every detail of a concept, and a fully specified concept leaves the human nothing to
complete. Text leaves gaps, and filling gaps is effortful precisely because it is
generative. The effect is moderated by how far the idea has already been developed:
textual input helps most when the concept is already mature, because there is enough
structure for the human's completions to attach to.

The result should be read as a claim about **agency configuration** rather than about
model quality—the question is how much of the concretization work the tool takes over,
not how good the tool is. That reframing matters for NPD practice because it makes the
design variable something a manager controls. It also supplies a caution for the
measurement discussion in @sec-innovation-measure: an ideation tool evaluated on
throughput or on time-to-concept will score the image condition best, and an
evaluation restricted to those metrics would recommend exactly the configuration that
degrades the output the process exists to produce.

### Quality Without Variety: The Diversity Cost of LLM Ideation {#sec-innovation-llm-diversity}

The companion question is what happens when the model is asked to *generate* the ideas
rather than help concretize them, and there the accounting runs the other way: the
per-idea numbers look excellent and the portfolio numbers do not.
@terwiesch2026creativity run eight studies on one tightly bounded ideation task—a new
physical product for the U.S. college-student market retailing under \$50—chosen because
it is the challenge format long used in product-design courses, so a large pool of
pre-LLM human ideas exists for comparison. Idea quality is measured as this literature
has long measured it: average **purchase intent** elicited from target-market consumers,
the best available single-number predictor of an idea's eventual value.

On that metric the LLM wins outright. LLM-generated ideas have higher average quality
than human ideas and are **seven times more likely** to land in the pooled top decile,
and a second study rules out the obvious deflationary explanation—that the model is
merely writing better sales copy around equally good ideas. The weakness appears only
when the unit of analysis moves from the idea to the *set*: AI ideas are less novel
individually and, more consequentially, less **diverse** collectively. A fifth study
re-analyzes prior LLM creativity experiments and finds the same diversity shortfall in
every one, which upgrades the result from an artifact of one prompt to a property of the
generator.

That ordering—quality up, diversity down—matters more than either finding alone because
of what an idea pool is *for*. A tournament does not consume the mean of its pool; it
consumes the maximum, and the expected maximum of a sample rises with dispersion as well
as with location. A generator that shifts the quality distribution rightward while
compressing its spread, and while concentrating the draws in one region of the idea
space, can dominate on every per-idea metric and still yield a worse best idea, because
the search has been narrowed. The measurement lesson is the one
@sec-innovation-creativity-embeddings makes with patent text: novelty and diversity are
*set-level* quantities that no average of item-level scores recovers, so an evaluation
reporting only mean purchase intent is structurally blind to the cost being incurred.

The remaining studies buy the variance back, and in doing so establish that the
shortfall is a configuration problem rather than a hard ceiling. Newer model versions
generate more diverse sets than older ones but still fall short of human sets. Three
interventions close most of what remains: **pooling** ideas across vendors, so that one
model's training idiosyncrasies do not define the search; **prompt engineering**—
chain-of-thought, and injecting heterogeneous personas or constraints, each of which
pushes the sampler off its modal completion; and **creative agents** instructed to
explore the solution landscape deliberately. A final study exploits the economics
directly: because the marginal cost of one more AI idea is near zero, simply generating
far more of them steadily improves coverage of the idea space and approaches human-level
coverage. For an innovation manager the synthesis is that the human pool is less a
diversity benchmark to be matched by a cleverer prompt than a coverage target that can
be purchased with scale, heterogeneity, or both.

Read alongside @gordetzki2026ideation, the two results point at the same design variable
from opposite ends of the front end. Concretization support degrades creativity by
leaving the human nothing to complete; generation support degrades diversity by leaving
the *search* nothing to explore. In both cases the metric that would have caught the
problem—effort in one, set-level coverage in the other—is the first one a
throughput-oriented evaluation drops.

## Organizational Determinants of Innovation

Beyond the project and the boardroom lies the question of which *organizational*
arrangements expand or erode a firm's stock of innovative capacity—what several
authors call its **innovation base**, the accumulated inventions, ideas, and
discoveries that propel future innovation.

### Acquisitions and the Innovation Base

The evidence on whether acquisitions help or harm innovation is genuinely mixed,
and the resolution is conditional. @hitt1991 document a *negative* average effect:
acquisition activity reduces R&D inputs and outputs, including patent counts, as
managerial attention and capital are absorbed by integration. @mishra2013building,
analyzing 352 firms across five industries over seventeen years, qualify this:
firms that acquire build a *more* robust innovation base than those that do not,
but the benefit depends on the **type** of acquisition—vertical (across supply-chain
stages) versus horizontal (same stage, similar chains)—and, critically, on the
**breadth of knowledge** in the acquiring firm, which determines whether either
acquisition type actually enriches the base. @king2008performance reconcile the
two by emphasizing **resource interactions**: acquisitions raise performance only
when target and acquirer resources are *complementary*. Marketing–technology
complementarity (strong-marketing acquirer, strong-technology target) creates
value, whereas technology–technology overlap is substitutive and can destroy it.
The apparent contradiction in the literature thus dissolves into a moderation
structure—average effects are uninformative when complementarity varies across
deals.

### Firm Boundaries: Franchising and External Knowledge

Firms also reshape their innovation capacity by reconfiguring their boundaries.
@mallapragada2017innovativeness study **franchising** as such a lever, using panel
data on 38 U.S. restaurant chains (1992–2005) and a nonlinear seemingly-unrelated-
regression model. They find a positive relationship between franchising emphasis and
*product* innovativeness that intensifies under high financial leverage and weakens
under high slack resources, with a parallel but distinctly moderated pattern for
*process* innovativeness. Franchising, on this evidence, is a conditional substitute
for alliances and joint ventures in shaping innovation outcomes.

Knowledge also enters the firm through less formal channels. @baruffaldi2024like
show that firms learn at **scientific conferences**, especially when they engage
intensely through both sponsorship and scientific contribution. Using direct
flights as an instrumental variable for attendance—an instrument that shifts the
cost of attending without plausibly affecting innovation through other channels—they
find that co-attendance with other scientists raises the likelihood of leveraging
external knowledge and of future collaboration, with learning skewed toward
prominent firms and scientists. The instrument is what makes the learning claim
*causal* rather than merely correlational: firms that attend more conferences differ
in unobservable ways from those that do not, and direct flights break that
confound.

### Who You Know in the Lab: Networks and Absorptive Capacity {#sec-innovation-academic-networks}

Conferences are a scheduled, purchasable channel. @arts2026harnessing ask about the
unscheduled one, and open with a puzzle the chapter's measurement section has already
half-stated: firms conduct less scientific research internally than they once did, yet
their patents lean *more* heavily on science. If the science is not being made in
house, how does it get in? The public-good answer is that it does not need a channel—
published findings are there for anyone to read, and a firm's problem is only whether
it is looking. The authors' answer is that reading is not the binding constraint.

They construct a social network covering nearly the entire population of life-science
academic scientists and corporate inventors listed on U.S. patents, then confront the
selection problem that has dogged every prior attempt at this question. A firm that
cites a professor's paper may do so because it knows the professor, or because the
paper was the best and most commercializable science available—and the two are
correlated, because good scientists attract both citations and connections. The design
that breaks the tie is **paper twins**: pairs of scientific papers reporting nearly
identical findings, published at nearly the same time, by independent academic teams.
Within a twin pair the knowledge itself is held fixed, along with its timing and its
commercial potential; what differs is which team the firm's inventors happen to be
connected to. Comparing which twin a corporate patent cites is therefore a comparison
of social distance with content controlled, the same maneuver as a matched-pairs
experiment and the reason the estimate can be read as transmission rather than quality
(@sec-graph-identification).

Connections win. Academic science travels into corporate patents along *close*
interpersonal ties—within two degrees of separation—so the twin whose authors sit a
coauthor, or a coauthor's coauthor, away from the inventor is the one that gets cited.
Two degrees is a short reach in a network of that size, and the shortness is the
substantive result: this is not diffuse network embeddedness paying off, it is knowing
someone, or knowing someone who does. Science may be a public good in the sense that
no one is excluded from reading it, but absorbing it is rival, and the rationing device
is the social graph.

The moderators are where the paper meets this chapter's recurring theme. The value of a
connection depends on the corporate inventor's own active engagement in scientific
research, and on the alignment between that inventor's scientific expertise and the
academic knowledge being integrated. A tie into a laboratory is worth little to an
inventor who has stopped doing science, or whose expertise lies elsewhere. That is
**absorptive capacity** recovered at the level of the individual inventor rather than
the firm, and it is the same conditional structure as the acquisition results above and
the open-source results below: external knowledge is not an average effect but a
complementarity, and internal research capability is the organ that digests it. It also
sharpens how the science-linkage measures of @sec-innovation-measure should be read. A
firm's count of patent citations to scientific papers is usually taken as a measure of
how science-intensive its technology is; on this evidence it is partly a measure of
whose calls its inventors can get returned. (INFORMS and SSRN were both inaccessible
when this was written; the account above rests on the published abstract.)

The same conditional structure appears when the network is inside the firm.
@breschi2026ethnic ask how ethnic diversity among a firm's inventors is converted
into valuable innovation and argue that composition and collaborative structure
are different things. From patent and co-inventor data on 890 publicly traded
U.S. firms between 1990 and 2015 they build a network index of ethnic integration,
the ratio of cross-ethnic to within-ethnic co-invention ties, and link it to the
stock-market reaction around patent grants, the forward-looking value measure
that @tbl-citations notes forward citations cannot supply. Integration raises
patent value, most in firms that are both ethnically diverse and working on
technologically complex innovation, and the estimates survive an
instrumental-variables strategy for the endogeneity of collaboration structure.
Diversity without integration returns little. (Wiley's site was inaccessible when
this was written, so the instrument itself could not be inspected; the account
rests on the published abstract.)

### Opening the Boundary: Open-Source Product Development

The franchising and conference channels move knowledge *toward* a firm that still owns
the development process. **Open-source product development** dissolves that ownership:
the firm builds its product in public on a social coding platform such as GitHub, and
outside volunteers commit code alongside its own employees. Whether that openness pays
is a question about the firm's income statement, not about community goodwill, and
@singh2026opensource answer it with a panel of 977 U.S. high-tech firms observed from
2001 to 2025. Firms that adopt open-source product development realize, on average, a
**4–5% increase in gross margin**, an estimate that survives staggered
difference-in-differences, generalized synthetic control, instrumental-variable, and
dynamic-panel specifications (@sec-causal-inference)—four designs with different
identifying assumptions converging on the same magnitude, which is what makes the
result hard to dismiss as an artifact of any one of them.

The interesting content is in the mechanism and its limits. Framing open source as
**distributed knowledge integration** in the knowledge-based view of the firm, the
authors decompose more than 323,000 project-level contributions across over 44,000
repositories into *internal employee* and *external volunteer* sources, and show that
**labor productivity** partially mediates the margin effect: outside expertise enlarges
the pool of specialized knowledge applied to the product without a commensurate rise in
the wage bill. But the mediated path is not always positive. The indirect effect turns
discernibly positive only past a **threshold of roughly 35% external volunteer
contribution** (@sec-causal-mediation-firm-panel reads the moderated mediation as a
design)—below it, the firm is paying the coordination and review costs of
openness without importing enough outside knowledge to cover them. A configurational
analysis then adds a second condition: R&D intensity is present in *every*
high-profitability configuration, the absorptive capacity required to metabolize
externally sourced knowledge (the method is set out in @sec-qual-fsqca, and the
three-stage architecture of the paper as a whole in @sec-sw-emb). That is the same
conditional logic as the acquisitions literature above, in a different guise—openness,
like acquisition, is not an average effect but a complementarity, and a firm without
internal R&D has no organ with which to digest what the crowd contributes.

Generative AI changes the economics of that crowd. @song2026copilot ask what an AI
pair programmer does to open-source development, where participation is voluntary,
membership is fluid, and the integration of contributions runs through peer discussion
rather than through a manager. GitHub supplied proprietary project-level data on the
share of contributors using Copilot, which the authors joined to the public activity
records of 9,244 active GitHub projects observed monthly from January 2021, when
Copilot entered technical preview, to December 2022, stopping before ChatGPT could
contaminate the panel. Treated projects are those in which Copilot was both supported
by the contributors' coding environments and actually used; controls are projects
whose environments did not support it. Because adoption is staggered and treated
projects outnumber controls, the authors estimate a generalized synthetic control
(@sec-causal-inference) on the 7,637 projects with at least one contribution, and back
it with within-environment comparisons, a matched difference-in-differences, and an
instrument built from contributors' exposure to Copilot through their collaborators on
other projects.

The result is a tradeoff rather than a productivity headline. Copilot use raises
merged pull requests per project by 5.9%, and the gain decomposes along both margins
that the cost–benefit account of open-source participation predicts: 3.4% more
developers with merged code and 2.1% more merged code per developer. But the average
time to merge a pull request rises by 8%, and the mechanism is more talk, not less
review. Comments per merged request rise 6.5%, distinct commenters 1.7%, and comments
per commenter 5.9%, with more topics per discussion and more revision cycles, while
core developers complete *more* reviews after adoption, not fewer. The two forces net
out positive, with 3.5% to 5.1% more requests merged within one, three, or ten days,
and bug and issue counts per merged request are flat, so the extra code is not worse
code.

The distributional result is the one that matters for the complementarity argument
above. The core developers' share of merged code rises by about 5 percentage points
(6.5% on a base of 0.83), and the relative time to merge a peripheral contributor's
code rises by 5.4%, with peripheral contributions attracting disproportionately more
discussion and more requests for context. This is not a skill story: Copilot usage is
if anything higher among peripheral developers and their language expertise is
comparable. It is a role story. Core developers know the architecture well enough to
turn AI-generated suggestions into changes the project can absorb and capture the
reputational return on them; peripheral developers, especially new ones, get a lower
cost of contributing but not a higher benefit, and their cheaper code costs the
project more to integrate. The authors read the pattern as a warning that AI pair
programmers may concentrate open-source activity in the core. Read against
@singh2026opensource, it is also a warning for firms: if the volunteer share of
contribution is what makes openness pay, a tool that tilts contribution toward
insiders thins exactly the channel through which external knowledge arrives, and the
binding constraint shifts from writing code to coordinating it.

### Inventor Job Security and the Direction of Invention

Organizational arrangements shape not only how much a firm invents but *what kind*
of invention it gets, and the mechanism can run through a single employment term.
The intuitive prediction is that protecting inventors from dismissal buys
exploration: an inventor who cannot be fired for a failed project has the safety
net that risky, distant search requires. @wehrheim2026paradox test that prediction
against the staggered adoption of wrongful-discharge laws across U.S. states
between 1980 and 1999, using matched employer–employee data built from patent
records and a difference-in-differences design on the cross-state timing. The
prediction fails, and the reason is instructive. Job security does not primarily
change an inventor's tolerance for failure; it changes the expected length of the
relationship, and a longer expected tenure raises the return to investing in
*firm-specific* technological knowledge over generally portable knowledge. Under
stronger protection, inventors produced patents drawing more heavily on their own
firm's technologies, less novel by the backward-citation logic of
@sec-innovation-measure, and of higher private value to the employer.

The result is a clean case of the private–social wedge that runs through this
chapter's measurement discussion. Every outcome improved from the employer's
ledger; the novelty of the technological frontier did not. It also sharpens what
"organizational determinant" means. The wrongful-discharge laws were labor policy,
written about dismissal and evaluated on employment, and their measurable
consequence landed on the composition of the patent stock—an untargeted margin of
exactly the kind @sec-policy-spillovers is about. A firm that wants exploration
from a protected workforce has to buy it with something other than protection,
because protection is an incentive to specialize.

### Where Inventors Live: Residential Proximity Inside the Firm

The organizational arrangements above are all things a firm chooses.
@pennington2026residential study one it does not: where its inventors go home at
night. The agglomeration literature has long held that geographic proximity between
*firms* raises collaboration and knowledge spillovers, and the workplace literature
that proximity between *desks* does the same. The authors ask whether the mechanisms
behind the second claim—chance encounters, repeated observation, the lowered search
and social costs of asking a familiar face for help—operate outside the building, so
that two employees of the same facility who live near one another are more likely to
end up on the same patent.

The setting does much of the identifying work. The orthopedic-implant cluster around
Warsaw, Indiana held the research facilities of three of the five largest orthopedic
firms in the world (DePuy, Zimmer, and Biomet) within a short drive of one another,
with no university, no related industry, and no major city nearby to sort inventors'
residences by knowledge type or prior affiliation. Because the firms required
inventors to list home addresses in patent-application correspondence, the authors
could hand-code residential street addresses from the USPTO Image File Wrapper
documents rather than rely on the regional-level addresses in standard patent
bibliographies. In the dissertation version of the paper
[@pennington2021dissertation], the sample is 1,913 applications filed by the three
firms between 2001 and 2015, 702 disambiguated inventors at 752 addresses, and a
conditional logit in which every pair of inventors active at the assignee in a
five-year window around the filing is a potential coinventor pair on that application
(about 122 inventors, hence about 7,400 pairs, per choice set). Each kilometer of
distance between two inventors' homes lowers the odds that they appear together on a
patent by about 0.3%. A firm whose average pair lived half a standard deviation
farther apart would see roughly 5% fewer collaborations per application, and the one
firm with the most dispersed inventors was about 17% less collaborative than the
other two.

The value of the paper is in what it rules out. Homophily is the obvious rival:
people who are alike may both live near one another and work together. A shared
prior employer or alma mater raises a pair's collaboration odds by about 50% but does
not predict how close they live, and it does not attenuate the distance effect when
interacted with it; newcomers to the region settle *farther* from people they already
know, not closer; differences in block-group household income between the two homes
matter on their own but barely dent the proximity coefficient; and the pair's average
distance to the town center, where all the R&D facilities sit, is insignificant once
pairwise distance is in the model, so it is proximity to each other rather than to
work that predicts collaboration. A within-pair test on inventors who moved during
the sample period—the same two people, before and after their residential distance
changed—recovers an effect of the same sign at about 1% per kilometer with every
fixed pair characteristic differenced away. The published version adds that the
effect holds for pairs in different product divisions, so residential proximity
relaxes an organizational constraint on who collaborates with whom.

The one dimension along which proximity does *not* relax a constraint is gender.
Same-gender pairs collaborate about 28% more often than mixed-gender pairs, and when
distance is interacted with a same-gender indicator the entire proximity effect loads
onto same-gender pairs; for mixed-gender pairs, living nearby does nothing. The
authors read this as social cost: encounters at the grocery store or the youth sports
field escalate into work conversations more readily between two men, or two women,
than across genders, so the collaborations that proximity creates are the unplanned
ones, and unplanned collaborations inherit the homophily of unplanned social life
(@sec-graph-identification). The chapter's recurring lesson about organizational
determinants applies with a twist. A geography the firm never designed shapes its
patent network, and the same mechanism that adds collaborations subtracts diversity
from them. The result is also a caution about *measurement scale* for anyone using
location as a variable (@sec-geospatial): a within-commuting-region indicator, the
proximity measure most patent studies can afford, would have missed the whole effect,
which lives at the resolution of a five-minute versus a ten-minute drive to the same
store.

The limiting case of proximity without organizational affiliation is the coworking
space, where people with different employers, occupations, and goals rent desks in
the same room. @hymer2026coworking ask what relationships form there and what value
they carry. Applying grounded-theory techniques to observational, archival, and
interview data from 80 workers across 21 coworking spaces (@sec-qualitative-research),
they arrive at a construct they call the **work companion**: someone who works
alongside you with a shared focus on work but without the obligations of family,
friendship, or a common employer. What makes the relationship valuable, on their
account, is exactly that absence. Because neither party owes the other anything,
members can choose moment to moment between socially engaging and working in
parallel, and they extract both personal and professional value from that interaction
flexibility; where a family, friendship, or employment tie *is* present, obligations
rigidify the interaction and the value falls, which the authors treat as the
construct's boundary condition. Space managers matter as curators of the conditions
under which companions form. The paper is qualitative and builds theory rather than
testing it, but it sits naturally beside the inventor results above. Pennington and
Shaver show that proximity outside the firm produces collaboration inside it; Hymer
and coauthors describe the micro-relationship through which proximity among
strangers does its work, and why a firm cannot simply manufacture it by seating its
own employees together. (The published abstract was the only part of the paper
accessible when this was written.)

Proximity can also be priced. @catalini2020travel move the question from where
collaborators live to what it costs them to meet. They build a simple framework in
which travel costs shape both whether scientists collaborate at all and which kinds
of project get developed locally rather than over distance, then test it against the
entry of new low-cost airline routes, a quasi-experiment that lowers the cost of a
face-to-face meeting between two cities without changing anything about the science
in either. After a low-cost carrier enters, collaborations between the newly
connected locations rise by between 0.3 and 1.1 times, a result that survives
multiple falsification tests. The gain accrues disproportionately to high-quality
scientists otherwise embedded in weaker local environments, which is to say the
people for whom the local collaboration pool was the binding constraint. Lower travel
costs also change the *kind* of project undertaken, and that is the part with
consequences for @sec-innovation-measure: a friction does not simply scale
collaboration up or down, it selects which collaborations happen, so a count of
coauthored patents is a function of the cost of meeting as much as of the willingness
to work together. Read beside the residential result, the two papers bracket the
geography of collaboration at both ends, kilometers between homes in one town and
airfares between cities, and both find that the physical cost of an encounter shapes
an output the firm books as intellectual.

### Dominance, Radicalness, and the Competition–Innovation Curve

Who produces *radical* innovation—and who profits from it? @sorescu2003sources find
that **dominant** firms introduce *fewer* radical innovations than non-dominant
firms, that the financial rewards to radical innovation vary enormously across firms
and are tied to the resource base, and that firms with deeper marketing and
technology support and broader product portfolios extract more value from the
radical innovations they do produce. The relationship between competition and
innovation is itself non-monotonic: @aghion2005 establish an **inverted-U**, in
which innovation is highest at intermediate product-market competition—too little
competition removes the spur to innovate, too much removes the rents that reward it.

The contrast between dominant and entrant innovators is sharpened by what happens to
a firm's innovation *after* it goes public. @Wies_2015 show that following an
initial public offering (IPO), firms raise their innovation *levels*—they innovate
more—but the innovations become *less risky*, with fewer breakthroughs, as the
discipline of public markets tilts the portfolio toward incremental safety. This
post-IPO retreat from radicalness is the empirical hinge connecting the
organizational and financial halves of the chapter.

## Innovation and Firm Value

Innovation is ultimately an intangible asset, and capital markets price it. The
marketing–finance literature—developed in full in @sec-marketing-finance—asks how
investors value innovation activity and treats the question with event-study and
asset-pricing machinery rather than survey scales.

### How Markets Price New-Product Activity

@srinivasan2009a provide the foundational empirical link between product innovation,
marketing investment, and stock returns. Their headline magnitudes are striking and
worth stating precisely: investors react far more favorably to **pioneering**
innovations than to minor updates—on the order of *seven times* greater—and the
advertising support behind pioneering innovations is roughly *nine times* more
effective in moving returns. Perceived product quality raises returns, while price
promotions *lower* them, plausibly because a promotion signals weak underlying
demand. The asymmetry between pioneering and incremental innovation is the financial
counterpart to the radical-versus-incremental distinction in @sorescu2003sources.

The effect of innovation on sales is itself conditioned by brand assets.
@slotegraaf2008 show that the permanent and cumulative sales effects of product
introductions are larger for brands with higher equity and more frequent releases,
while lower-equity brands benefit *more* from each individual introduction—a
substitution between brand stock and product flow. Collaboration timing also moves
markets: @wu2015sleeping analyze 831 announcements of **horizontal collaborations**
(partnering with competitors for NPD) over twelve years and find that markets react
*positively* to collaboration at the *initiation* phase but *negatively* during
development and commercialization, with the new product's innovativeness and the
partner's relative strengths providing asymmetric moderation.

### Innovation Signaling at the IPO

Innovation is informative precisely when information is scarcest—at the IPO, when
outside investors know least about firm quality. @cao2022express show that firms
deploy their **innovation potential** as a credible quality signal at IPO:
innovation potential is positively associated with the IPO's initial value and
first-day returns and negatively associated with the extent of insider share sales.
The components differ in what they signal—patents weigh most heavily on insider
selling, while preannouncements move first-day returns most—consistent with a
signaling logic in which costlier, harder-to-fake signals (granted patents) carry
the most weight where credibility matters most.

### Sustaining Innovation After the IPO

The post-IPO innovation decline documented by @Wies_2015 raises the question of what
can arrest it. @wu2024innovation examine the role of **AI analytics** in sustaining
post-IPO innovation, analyzing patent data from over 1,000 publicly traded firms.
Firms that acquire AI-analytics capabilities after their IPO suffer a *smaller*
decline in innovation quality, with the effect concentrated in machine-learning
capabilities specifically. The mechanism is **recombination**: sustained innovation
is driven by the ability to recombine existing technologies into new ones—exactly
the backward-citation logic of @sec-innovation-measure—and AI analytics supports
that recombination. It mitigates two of the three classic barriers to post-IPO
innovation—short-term financial pressure and disclosure requirements (with the
disclosure benefit largest for firms with long product cycles)—while doing little
for managerial-incentive problems. The broader lesson is that *technology* is itself
an input into the production of innovation, and that firms and investors can
partially offset the predictable post-IPO innovation slump by investing in it.

## Key Takeaways

- Diffusion is the formal backbone of innovation forecasting. The **Bass model**
  (@eq-04-bass-hazard, @eq-bass-cumulative) decomposes adoption into innovation
  ($p$) and imitation ($q$); its parameters are weakly identified before the sales
  peak, and successive-generation extensions [@Chandrasekaran_2020] are needed when
  substitution, not saturation, drives an incumbent's decline.
- Diffusion often turns on a **supply-side** adoption decision by local merchants,
  not only on consumer contagion. @xu2026food show that local news coverage causes
  restaurants and stores to adopt plant-based meat, identifying the effect from
  quasi-random county-quarter news production across topics
  (@sec-innovation-news-diffusion); the effect is larger in liberal-leaning counties,
  so a national average masks a politically sorted map.
- Patent **citations** are the field's default innovation measure: backward
  citations proxy (inverse) novelty and are available *ex ante*, while forward
  citations proxy importance but accrue *ex post* and require truncation
  corrections (@tbl-citations).
- NPD outcomes are **contingent**—on the information environment
  [@moorman1999contingency], team structure [@sethi2001cross; @slotegraaf2011product],
  and latent firm heterogeneity [@grewal2013environments]—so average effects mislead.
- Acquisitions help innovation only under resource **complementarity**
  [@king2008performance; @mishra2013building], reconciling the otherwise-negative
  average effect [@hitt1991]; competition and innovation trace an inverted-U
  [@aghion2005].
- External **scientific** knowledge is not free for the taking. Holding the science
  itself fixed with *paper twins*, academic findings enter corporate patents along
  interpersonal ties within two degrees of separation, and only for inventors who are
  still active in research and whose expertise matches what they are absorbing—
  absorptive capacity at the individual level [@arts2026harnessing]. Firms also learn
  at conferences, identified with direct flights as an instrument for attendance
  [@baruffaldi2024like].
- Some organizational determinants are not chosen. Where a firm's inventors *live*
  predicts who patents together: each kilometer between two coworkers' homes lowers
  their odds of coinventing by about 0.3%, the effect survives homophily controls and
  a within-pair mover test, and it operates only within same-gender pairs
  [@pennington2026residential].
- Generative AI reshapes the open-source crowd itself. GitHub Copilot raises merged
  contributions per project by 5.9% but lengthens merge time by 8% through more code
  discussion, and tilts contribution toward core developers and away from the
  peripheral volunteers whose knowledge is what makes openness pay
  [@song2026copilot; @singh2026opensource].
- Markets price innovation asymmetrically—pioneering innovations earn roughly seven
  times the reaction of incremental ones [@srinivasan2009a]—and the post-IPO retreat
  from radical innovation [@Wies_2015] can be partly offset by technology investment
  [@wu2024innovation]. The valuation machinery is developed in @sec-marketing-finance.
- **Generative AI is not one lever at the front end.** Used to concretize a concept,
  visual support cuts ideator effort but yields ideas 18% less creative than the more
  effortful textual support [@gordetzki2026ideation]; used to generate concepts, an LLM
  raises average purchase intent and is seven times likelier to produce a top-decile idea
  while producing **less novel and less diverse sets**, a shortfall that recurs across
  prior LLM creativity studies and is recovered by pooling vendors, chain-of-thought and
  persona prompting, creative agents, or sheer scale [@terwiesch2026creativity]. Because
  a tournament consumes the maximum of a pool rather than its mean, set-level coverage is
  the metric to report (@sec-innovation-llm-diversity).
