# Entrepreneurship and New Ventures {#sec-entrepreneurship}

The preceding chapter treated entry as something *firms* do: an established
organization decides when to bring a product into a category and how to defend
the position it takes. But a large share of entry is performed by organizations
that did not exist before the entry decision was made. The founding of a firm is
simultaneously an entry decision, a resource-assembly decision, and a
positioning decision, and it is made by people whose knowledge was accumulated
somewhere else—in a university laboratory, in an incumbent's product division,
or in the course of using a product and finding it inadequate.

This chapter develops the resulting research program. Its organizing claim is
that the *origin* of a founder's knowledge is a strategically consequential
variable, and that it predicts not merely how well a venture performs but
**what kind of economic value the venture is able to capture at all**. A venture
founded on scientific knowledge and a venture founded on industry knowledge are
not two draws from the same distribution with different means; they are firms
solving different problems, transacting in different markets, and failing in
different ways. Treating founder origin as a control variable in a performance
regression discards exactly the variation that matters.

We proceed in four movements. @sec-ent-origins establishes the knowledge-origins
taxonomy and the evidence that founders carry more than human capital out of
their prior organization. @sec-ent-value-capture defines modes of value capture
and develops the chapter's central result: the correspondence between where
knowledge came from and which market a venture monetizes in.
@sec-ent-model formalizes that correspondence and shows why the interesting
founders are the ones who fit neither category. @sec-ent-teams and
@sec-ent-resources then turn to founding-team composition and to the evaluation
problem ventures face when they seek resources from investors, regulators, and
employees. @sec-ent-marketing closes with what marketing has to contribute to a
literature that has so far been written almost entirely without customers in it.

A methodological thread runs underneath. Much of the best work here is
*descriptive*: it maps a pattern rather than estimating a treatment effect. That
is a deliberate design choice with its own standards of rigor, and it is treated
as a craft problem in @sec-sw-descriptive.

## Where Ventures Come From: The Knowledge-Origins View {#sec-ent-origins}

The foundational move in this literature is to stop treating "entrepreneur" as a
homogeneous category. @agarwalshah2014 organize firm formation around three
distinct knowledge sources, each associated with a different pre-founding
context.

**Academic entrepreneurs** commercialize knowledge generated in scientific
research, typically at universities or public laboratories. Their knowledge is
codified, novel, and often protected by patents assigned to or licensed from the
parent institution. It is deep on the technical dimension and thin on everything
downstream: who the customer is, what the distribution channel looks like, how a
regulator will react.

**Employee entrepreneurs** (also called spinouts) leave an incumbent firm in the
same or an adjacent industry. Their knowledge is the mirror image: rich in
customer requirements, supplier relationships, pricing norms, and the tacit
operating detail of an industry, but bounded by the technological trajectory the
parent was already on.

**User entrepreneurs** found firms on knowledge generated by *using* a product
and encountering an unmet need. Their knowledge is neither scientific nor
producer-side; it is need knowledge, and it explains why user-founded firms
cluster in categories where the need is legible to those who experience it and
invisible to those who do not.

The taxonomy matters because it predicts different founding rates, different
entry timing relative to industry emergence, and different survival profiles.
The evidence that founders carry organizational content, and not merely skill,
out of the parent is by now substantial.

@zucker2002commercializing supply the original demonstration on the academic
side. Studying biotechnology, they show that firm performance is tied to the
*capture* of university science by identifiable star scientists working with the
firm, not to proximity to universities in the abstract. Knowledge moved with
people, and the firms that got the people got the performance.

@feldman2019heritage extend the claim from knowledge to *practice*. Using survey
and registrar data on spawns and their parents, they find that organizational
practices transfer from parent to spinout: the overlap in practices between a
spawn and its parent is roughly ten percent greater than between the spawn and
other established firms, and about seventy percent of comparisons show start-ups
to be less similar to unrelated organizations than parents are to their spawns.
Crucially, transfer is selective. The practices that travel are the ones that fit
a start-up's requirements *and* are clearly defined and causally unambiguous.
Heritage is not wholesale replication; it is a filtered inheritance.

@ahn2025cultural push the same logic into culture. Linking Crunchbase start-ups
to natural-language measures of culture derived from Glassdoor reviews, they show
that founders transmit cultural elements from the parent organization, and that
transmission is stronger when the founder's tenure at the parent was longer, when
the parent's culture was internally coherent, and when it was atypical relative
to other organizations. The contingencies are the contribution: an incoherent
parent culture supplies no transmissible toolkit, and a generic one supplies
nothing distinguishable.

Finally, the pathway *into* founding is itself structured by employment.
@law2025joiners follow more than eight thousand individuals through the Venture
For America program and find that working as a start-up employee raises the
likelihood of later founding—and that this joiner-to-founder effect is markedly
stronger for Black women than for other demographic groups. Their qualitative
evidence points to self-reflection prompted by the start-up employment
experience as a mechanism. The result is a reminder that the supply of founders
is not exogenous to the organizations that employ them.

@fig-ent-origins summarizes the pathway from prior context to venture.

```{mermaid}
%%| label: fig-ent-origins
%%| fig-cap: "The knowledge-origins view of firm formation. The founder's prior employment context determines the composition of the knowledge profile carried into the venture, and that profile — not the founder's credential — is what conditions downstream strategy. Dashed arrows mark inheritance of organizational content beyond individual human capital."
flowchart LR
    U["University /<br/>public laboratory"]
    F["Incumbent firm<br/>in the industry"]
    C["Use context<br/>(unmet need)"]
    S["Scientific knowledge<br/>codified, novel, patentable"]
    K["Commercial knowledge<br/>customers, channels, regulators"]
    N["Need knowledge<br/>problem definition"]
    V["Venture knowledge profile"]
    H["Inherited practices<br/>and culture"]
    U --> S
    F --> K
    C --> N
    S --> V
    K --> V
    N --> V
    F -.-> H
    H -.-> V
```

### What makes founding salient {#sec-ent-salience}

Origin explains what a founder brings. It does not explain *when* the decision to
found gets made, and the trigger turns out to be partly a marketing artifact.
@huang2026salience study entry onto a large Chinese e-commerce marketplace and
find that individuals are more likely to open an online store after observing the
salient success of a nearby one. The source of variation is the platform's
own store rating system: when a local store's rating is upgraded, new store
entries in the vicinity rise significantly. The effect decays with physical
distance and strengthens with the salience of the upgrade—the two gradients a
salience account predicts, and the reason the result is not simply local demand
information diffusing.

The second half of the paper is what makes it worth teaching. The entries induced
by a local upgrade event are *worse*: they underperform in sales and are more
likely to exit, and the excess exit concentrates in downturns in overall market
conditions. That asymmetry is the discriminating test. If a neighbor's rating
upgrade revealed genuine unexploited local demand, entrants responding to it
should do at least as well as other entrants; that they systematically do worse
is consistent instead with salience-based decision making, in which highly visible
local information is overweighted relative to its actual diagnosticity
[@bordalo2013salience]—the same primitive that a salience nudge perturbs
(@sec-thn-foundations).

Two implications follow for marketing. First, on the supply side of entry: the
entry-order literature in @sec-market-entry takes the set of entrants as given and
asks who moves when, but the composition of that set is being set in part by which
local successes happen to be visible, and visibility is manufactured by a rating
system. Second, on platform design: a ratings mechanism built to solve a buyer-side
information problem also functions as an entry-inducement device on the seller
side, and it induces the marginal entrant least equipped to survive a downturn.
That is an externality of the design, not of the market, and it belongs with the
other realized externalities of platform structure in
@sec-platform-entry-externality.

## Modes of Economic Value Capture {#sec-ent-value-capture}

The knowledge-origins taxonomy becomes strategically interesting only when it is
connected to an outcome that is *categorical* rather than scalar. That outcome is
the mode of value capture.

::: {.callout-note}
## Definition: modes of economic value capture

A venture founded on a technological opportunity may realize economic value in
one of two markets. In the **market for technology** it monetizes the technology
itself—licensing, selling patents, being acquired for its intellectual property,
or entering a research collaboration in which a partner commercializes the
output. In the **market for product** it monetizes an artifact sold to end
users, which requires assembling the complementary assets that production,
distribution, regulatory clearance, and after-sales service demand.

A third possibility, **no value capture**, is not a residual category but a
substantive outcome: many ventures founded on genuine technological
opportunities never transact in either market. Any study that conditions on
having captured value selects on the dependent variable.
:::

The distinction descends from a long line of work on the appropriability of
innovation, but its use as a *dependent variable in its own right* is more
recent. @moeen2017incubation study the incubation stage of an industry—the
interval between a technological breakthrough and the first commercialization of
it—in agricultural biotechnology, and deliberately report what they call
*stylized findings* rather than tests of hypotheses. Among them: knowledge
evolution precedes product evolution in an industry's life cycle; the firms
investing during incubation are heterogeneous in type; and different types
converge on different modes of value capture. Incumbents of the obsolescing
industry tend to become acquisition targets, science-based start-ups tend toward
alliances and acquisitions, and diversifying firms tend toward product
commercialization. Their managerial takeaway is the methodological one restated:
success and failure must be measured against multiple yardsticks, not against
product commercialization alone.

@moeen2020building extend the incubation program by asking how firms reduce
uncertainty across industry milestones, and @agarwal2025industries synthesize
the resulting body of work into an account of industry creation as the joint
product of heterogeneous actors. Together these establish the setting in which
the chapter's centerpiece operates.

### Founder knowledge and the mode of capture

@moeen2026valuecapture asks the question the incubation literature had set up
but not answered at the level of the individual venture: given that value can be
captured in either market, which founders end up in which?

The design separates founders by prior employment context—academic scientists
versus former employees of within-industry firms—and observes the mode of value
capture in a population of U.S. medical device start-ups. The setting is chosen
well. Medical devices leave an institutional paper trail on both sides of the
distinction: intellectual-property transactions on the technology side, and FDA
clearance and marketing authorization on the product side. The dependent
variable is therefore classified by regulators and registries rather than by the
researcher.

Three findings follow, and their structure repays study.

1. **Academic start-ups tend to capture value in markets for technology.** The
   scientific knowledge that founded them is precisely the asset a technology
   market prices, and it is not sufficient for the complementary-asset assembly
   that product markets demand.
2. **Within-industry employee start-ups tend to capture value in markets for
   product.** Their commercial knowledge is what product commercialization
   requires, and it is not the novel scientific content a technology market
   pays for.
3. **Founders whose history bridges the two dimensions shift toward pioneering
   products.** Academics with industry experience, and employees with scientific
   training, do not simply split the difference between the two modes. They move
   to a third outcome that neither pure profile reaches.

The third finding is what converts the study from a stereotype into a mechanism
claim, and the next section explains why.

## A Formal Sketch: Why Bridging Founders Pioneer {#sec-ent-model}

The three findings above have a compact formal representation that makes the
role of the bridging case transparent. Let a founding team be summarized by a
knowledge profile
$$
\mathbf{k} \;=\; (\sigma, \kappa), \qquad \sigma \ge 0,\ \kappa \ge 0,
$$
where $\sigma$ is depth of *scientific* knowledge and $\kappa$ is depth of
*commercial* knowledge about the focal industry. Origin enters only through the
distribution from which $\mathbf{k}$ is drawn: academic founders draw high
$\sigma$ and low $\kappa$, within-industry employee founders the reverse.

Suppose the venture chooses among three modes. Licensing into the market for
technology yields a payoff increasing in scientific depth alone; selling a
conventional product into an established category yields a payoff increasing in
commercial depth alone; and pioneering a product—one that is both technically
novel and successfully commercialized—requires the two knowledge types as
*complements*:

$$
\begin{aligned}
V_T(\mathbf{k}) &= \alpha\,\sigma - f_T + \varepsilon_T, \\
V_P(\mathbf{k}) &= \beta\,\kappa - f_P + \varepsilon_P, \\
V_N(\mathbf{k}) &= \lambda\,\sigma\kappa - f_N + \varepsilon_N,
\end{aligned}
$$ {#eq-ent-payoffs}

with $f_T < f_P < f_N$ the fixed costs of operating in each mode and
$\varepsilon_m$ an idiosyncratic venture-specific shock capturing everything the
knowledge profile does not determine. The mode is chosen as
$$
m^{*}(\mathbf{k}) \;=\; \arg\max_{m \in \{T,\,P,\,N,\,\varnothing\}}\ V_m(\mathbf{k}),
$$ {#eq-ent-mode}
where $\varnothing$ denotes no value capture and yields zero. Because the shocks
are non-degenerate, the model predicts *tendencies* rather than deterministic
sorting—which is the form the empirical claim takes as well.

The structural content of @eq-ent-payoffs is the contrast between *additive* and
*multiplicative* aggregation in the systematic part of the payoff. $V_T$ and
$V_P$ are each linear in one coordinate and flat in the other, so each is
maximized along an axis of the knowledge plane. $V_N$ is multiplicative, so it
vanishes whenever either coordinate is near zero and can only dominate in the
*interior*. Three consequences follow immediately, and they are the paper's three
findings.

- A profile with high $\sigma$ and $\kappa \approx 0$ has $V_N \approx 0$, so the
  technology market wins. Academic founding maps to markets for technology.
- A profile with high $\kappa$ and $\sigma \approx 0$ likewise has $V_N \approx
  0$, so the product market wins. Employee founding maps to markets for product.
- A profile bounded away from both axes can have $\lambda\sigma\kappa$ exceed
  both linear terms even when neither coordinate is individually extreme.
  Bridging founders pioneer.

The pioneering region is therefore not a blend of the other two; it is a region
of the knowledge plane that neither pure profile can reach, because reaching it
requires a product of coordinates rather than a sum. This is the formal sense in
which the bridging cell identifies the mechanism: if mode of capture were driven
by *credential*—university versus firm—the bridgers would follow their
institution. Because they instead follow their knowledge profile into a third
mode, the operative variable is the profile.

The following simulation makes the partition concrete. It draws founders from
two origin distributions, applies @eq-ent-mode, and cross-tabulates origin
against realized mode.

```{r ent-mode-partition, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({library(dplyr); library(ggplot2)})
set.seed(20260811)

# Payoff parameters: complementarity lambda, mode-specific fixed costs,
# and the scale of the idiosyncratic shock.
alpha <- 1.00; beta <- 1.00; lambda <- 1.20
f_T   <- 0.25; f_P  <- 0.45; f_N    <- 0.70
s_eps <- 0.50

n <- 3000
draw <- function(origin, n) {
  if (origin == "Academic") {
    # deep science, thin commercial knowledge (right tail = industry stints)
    sigma <- rgamma(n, shape = 6.0, rate = 3.0)
    kappa <- rgamma(n, shape = 1.6, rate = 3.0)
  } else {
    # deep commercial knowledge, thin science (right tail = scientific training)
    sigma <- rgamma(n, shape = 1.6, rate = 3.0)
    kappa <- rgamma(n, shape = 6.0, rate = 3.0)
  }
  data.frame(origin = origin, sigma = sigma, kappa = kappa)
}

founders <- rbind(draw("Academic", n), draw("Employee", n)) |>
  mutate(
    V_T  = alpha  * sigma         - f_T + rnorm(dplyr::n(), 0, s_eps),
    V_P  = beta   * kappa         - f_P + rnorm(dplyr::n(), 0, s_eps),
    V_N  = lambda * sigma * kappa - f_N + rnorm(dplyr::n(), 0, s_eps),
    # argmax over the three modes and the outside option (no capture)
    mode = c("None", "Technology", "Product", "Pioneer")[
      max.col(cbind(0, V_T, V_P, V_N), ties.method = "first")
    ],
    # "bridging" = neither coordinate near zero, i.e. the interior of the plane
    bridging = sigma > 1 & kappa > 1
  )

# Mode shares by founder origin
round(prop.table(table(founders$origin, founders$mode), margin = 1), 3)
```

The cross-tabulation reproduces the first two findings: academic founders
concentrate in the technology mode and employee founders in the product mode.
The third finding is a statement about the interior, and is best seen by
conditioning on it.

```{r ent-bridging, message=FALSE, warning=FALSE}
# Among bridging founders only, the pioneering mode becomes the modal outcome
round(prop.table(table(founders$origin[founders$bridging],
                       founders$mode[founders$bridging]), margin = 1), 3)

cat(sprintf("\nPioneering share, non-bridging founders: %.3f\n",
            mean(founders$mode[!founders$bridging] == "Pioneer")))
cat(sprintf("Pioneering share, bridging founders    : %.3f\n",
            mean(founders$mode[founders$bridging]  == "Pioneer")))
```

Conditioning on the interior flips the modal outcome for *both* origins: among
bridging founders, pioneering becomes the majority mode regardless of whether
the founder came from a university or a firm. That is the point. The shift
toward pioneering is a property of the knowledge profile, not of the founder's
institutional label, and the label is what a credential-based explanation would
have to run on. @fig-ent-partition plots the resulting partition of the
knowledge plane, with the pioneering region visible as a wedge that touches
neither axis.

```{r fig-ent-partition, message=FALSE, warning=FALSE}
#| fig-cap: "Partition of the knowledge plane by the mode with the highest *expected* payoff under @eq-ent-payoffs. The technology and product regions hug their respective axes; the pioneering region is bounded away from both, because its payoff is multiplicative in the two knowledge types. Points are simulated founders, marked by origin; idiosyncratic shocks mean individual ventures need not choose the region they fall in."
grid <- expand.grid(sigma = seq(0, 3.5, length.out = 260),
                    kappa = seq(0, 3.5, length.out = 260))
grid$mode <- c("None", "Technology", "Product", "Pioneer")[
  max.col(cbind(0,
                alpha  * grid$sigma,
                beta   * grid$kappa,
                lambda * grid$sigma * grid$kappa) -
          cbind(0, f_T, f_P, f_N)[rep(1, nrow(grid)), ],
          ties.method = "first")
]

ggplot() +
  geom_raster(data = grid, aes(sigma, kappa, fill = mode), alpha = 0.55) +
  geom_point(data = dplyr::slice_sample(founders, n = 700),
             aes(sigma, kappa, shape = origin), size = 0.9, alpha = 0.7) +
  scale_fill_manual(values = c(None       = "grey88",
                               Technology = "#9ecae1",
                               Product    = "#a1d99b",
                               Pioneer    = "#fdae6b")) +
  coord_cartesian(xlim = c(0, 3.5), ylim = c(0, 3.5), expand = FALSE) +
  labs(x = expression(paste("Scientific knowledge  ", sigma)),
       y = expression(paste("Commercial knowledge  ", kappa)),
       fill = "Optimal mode", shape = "Founder origin") +
  theme_minimal(base_size = 11)
```

Two cautions about what this exercise does and does not establish. It is a
*consistency* demonstration: it shows that a complementarity in the value-capture
technology is sufficient to generate the observed correspondence, not that it is
the only mechanism that could. And the empirical claim in @moeen2026valuecapture
is a correspondence between origin and mode in observational data, not a
treatment effect of origin on mode; founders select into universities and firms
for reasons that may independently predict which market they can transact in.
The contribution is the map, and the map is what disciplines the causal designs
that should follow.

### Positioning as a regulatory act

The product-market pathway carries an obstacle the technology-market pathway
does not: in regulated categories, clearance precedes revenue.
@pahnke2026positioning study 239 new medical device firms' 510(k) clearances
between 2001 and 2019 and show that how a venture positions its device
*relative to already-cleared products* materially changes time to clearance.
For a firm's first product, clearance is accelerated by citing fewer reference
products while positioning against same-category products and category
exemplars; for the second product, by referencing the firm's own cleared device.

This is a positioning problem in the strict marketing sense—choosing a reference
set and a claimed similarity—with a regulator rather than a consumer as the
audience, and with the optimal strategy changing as the firm accumulates its own
precedent. It connects the venture literature directly to the categorization and
positioning material developed in @sec-ms-positioning, and it supplies a concrete
reason why the product mode is costlier than the technology mode in
@eq-ent-payoffs.

## Founding Teams and the Distribution of Outcomes {#sec-ent-teams}

Founder knowledge is carried by people, and most ventures are founded by more
than one. Four recent studies sharpen how team composition maps to outcomes,
and the first of them changes what "outcome" should mean.

@agarwal2026startup examine how pre-entry knowledge and post-entry learning
shape start-up performance, and report the result as a statement about the
*distribution* rather than the mean. Pre-entry knowledge does not simply shift
average performance; it reshapes the whole distribution, with effects on the
upper and lower tails that differ from each other and from the effect on the
mean. A literature that reports only average returns to founder experience
therefore understates what founder knowledge does.

This is a general point about resource assembly, and it recurs wherever
performance depends on combining resources whose fit is uncertain: the
interesting action is in the second moment, because good fit and bad fit are
both amplified by the act of combination. The same logic governs acquisition
performance, where it is developed at length in @sec-ms-corpdev.

The following simulation isolates why a mean-only summary is misleading when a
covariate shifts both location and spread.

```{r ent-distributional, message=FALSE, warning=FALSE}
set.seed(915)
N <- 20000
know <- runif(N)                              # pre-entry knowledge, standardized to [0,1]

# Knowledge raises the mean modestly, but raises the SPREAD sharply:
# it amplifies both good and bad realizations of fit.
mu <- 0.30 * know
sd <- 0.40 + 1.30 * know
perf <- rnorm(N, mean = mu, sd = sd)

grp <- cut(know, breaks = quantile(know, seq(0, 1, 0.2)),
           include.lowest = TRUE, labels = paste0("Q", 1:5))

tab <- t(sapply(split(perf, grp), function(x)
  c(mean = mean(x), p10 = quantile(x, .10), p50 = quantile(x, .50),
    p90 = quantile(x, .90))))
round(tab, 3)

# The mean effect understates the tail effects in OPPOSITE directions.
cat(sprintf("\nMean   Q5 - Q1: %+.3f\n",
            mean(perf[grp == "Q5"]) - mean(perf[grp == "Q1"])))
cat(sprintf("p90    Q5 - Q1: %+.3f  <- upside widens\n",
            quantile(perf[grp == "Q5"], .90) - quantile(perf[grp == "Q1"], .90)))
cat(sprintf("p10    Q5 - Q1: %+.3f  <- downside deepens\n",
            quantile(perf[grp == "Q5"], .10) - quantile(perf[grp == "Q1"], .10)))
```

The mean difference is positive and modest. The ninetieth-percentile difference
is much larger and the tenth-percentile difference is *negative*: high-knowledge
founders both fly higher and crash harder. Reporting only the mean would license
the conclusion that knowledge is mildly good, when the accurate statement is that
knowledge raises variance far more than it raises location.

@dacunto2026teams change the unit of measurement in a way that matters. Rather
than reading skills off founders' credentials, they take the career histories of
a start-up's *initial employees*—for the full cohort of U.S. start-ups, linked
through Census employer–employee records—and treat each person's prior industry
as a proxy for the skillset carried in. A team's skill profile is then the
distribution of those prior industries, and the variable of interest is its
diversity.

Start-ups whose teams span more prior industries grow faster than same-industry
local peers: a one-standard-deviation increase in diversity is associated with
roughly 16% higher five-year employment growth and 10% higher sales growth, net
of demographic diversity, pre-start-up wages, and the usual observables. The
association concentrates in innovative industries and in teams whose coordination
costs are low—the two conditions under which combining distinct skillsets is
worth the friction of combining them.

The identification problem is the familiar one: teams choose their own
composition, and whatever makes a venture promising may also let it recruit
across industries. The authors' answer is to use a shock to composition rather
than a choice about it—episodes in which a team abruptly loses a member whose
skills no teammate duplicates. The contrast between losing an unduplicated member
and losing a redundant one is what isolates the role of the diversity.

The phrase the paper settles on is worth keeping: teams do well by combining
individual specialization with team-level diversification, so that they are
"jacks of all trades and masters of each." That is a different claim from the
classic jack-of-all-trades argument, which locates breadth inside a single
founder. Here breadth is a property of the team and depth a property of the
person, and the two have to travel together—breadth without individual depth buys
nothing.

```{r ent-team-diversity}
set.seed(2026)
K <- 6      # prior industries
M <- 4      # members per founding team
N <- 6000   # start-ups
B <- 8      # human capital per member -- held FIXED, only its allocation varies
tau <- 1    # an industry pays off only once team capability clears this floor

# `focus` is the share of a member's budget in their own prior industry:
# 1/K is a pure generalist, 1 a pure specialist. Home industries are drawn
# independently of focus, so team diversity is not mechanically tied to it.
draw_team <- function(focus) {
  home <- sample.int(K, M, replace = TRUE)
  d <- matrix(B * (1 - focus) / (K - 1), nrow = M, ncol = K)
  d[cbind(seq_len(M), home)] <- B * focus
  list(depth = d, home = home)
}

# "Masters of each": the team's capability in an industry is its DEEPEST member
# there, and shallow coverage below tau contributes nothing at all.
growth <- function(tm) sum(sqrt(pmax(0, apply(tm$depth, 2, max) - tau)))

focus <- runif(N, 1 / K, 1)
teams <- lapply(focus, draw_team)
div   <- vapply(teams, function(tm) length(unique(tm$home)), integer(1))
g     <- vapply(teams, growth, numeric(1))

# (1) The headline association, with total human capital constant by construction.
#     The magnitude is not calibrated to theirs; only the sign and mechanism are.
b <- coef(lm(g ~ div + focus))["div"]
cat(sprintf("+1 SD diversity: %+.1f%% growth from the mean\n",
            100 * b * sd(div) / mean(g)))

# (2) ...but only where members are individually deep. Split on specialization.
spec <- ifelse(focus > 0.75, "specialists", "generalists")
round(tapply(g, list(spec, div), mean), 2)
```

```{r ent-team-shock}
# (3) The identification strategy: a team abruptly loses one member. Removing an
# UNDUPLICATED member and removing a redundant one delete identical human
# capital, so any gap between them is what the diversity was contributing.
uniq <- dup <- numeric(0)
for (tm in teams[focus > 0.75]) {
  for (i in seq_len(M)) {
    d <- growth(list(depth = tm$depth[-i, , drop = FALSE])) - growth(tm)
    if (sum(tm$home == tm$home[i]) == 1) uniq <- c(uniq, d) else dup <- c(dup, d)
  }
}
cat(sprintf("Lose an unduplicated member: %+.2f\n", mean(uniq)))
cat(sprintf("Lose a duplicated member   : %+.2f\n", mean(dup)))
```

The split table makes the interaction visible. In the generalist rows every member
already spans every industry at shallow depth, so adding another prior industry to
the team changes almost nothing; it is only when members are deep in one thing each
that spanning more things gives the team anything to combine. The loss shock then
reproduces the asymmetry the empirical design leans on: departures that remove
identical amounts of human capital have very different consequences depending on
whether the leaver's industry was covered by anyone else. The redundant departure
costs exactly nothing here only because the stylization is extreme---team
capability is its single deepest member---but the asymmetry it exaggerates is the
one the design exploits.

Because the measure is built from employees rather than founders, this paper is
already most of the way to the argument of @sec-ent-ehc below—that the venture's
relevant human capital does not stop at the founding team.

Two studies then address who founds together. @howell2026solo take up the
assumption that founding teams dominate solo founders—more human capital, more
diverse networks—and note that solo founding nonetheless persists at high rates.
Examining its viability, they find the lone-genius-versus-lonely-fool dichotomy
too coarse: solo founding is viable under identifiable conditions rather than
uniformly inferior.

@kovacs2026political examine team assembly on a dimension previously
unexamined. Using Crunchbase and voter-file data on 1,125 U.S. start-ups, they
document strong political homophily in founding-team formation—politically
similar teams form more often than diverse ones, net of gender, age, location,
and industry—and find that politically heterogeneous founding teams are more
likely to shut down. The authors are careful about the evidentiary status of the
secondary outcomes: associations with funding, headcount, and platform rankings
are directionally consistent but vary in robustness, and they say so. That
calibration is itself a model for how to report a descriptive result whose
components differ in strength.

### Beyond the founders: the entrepreneurial content of the payroll {#sec-ent-ehc}

The three studies above stop at the founding team. @braunerhjelm2026staying push
the same variable one layer down, into the people the venture hires. Working with
Swedish longitudinal register data from 1997 to 2016—which lets them identify
every new venture founded over two decades and link it to the full prior careers
of its employees—they define a venture's *entrepreneurial human capital* (EHC) as
the share of its employees who have previously started and managed a firm of their
own. After controlling for an extensive set of confounders, a higher EHC share is
associated with a higher probability of venture survival.

The paper's contribution is less the sign of that association than the three
mechanisms it separates. First, an enlarged resource base, magnified by learning:
the association strengthens with the length of an employee's prior spell in
entrepreneurship, which is what a learning account predicts and a pure
selection-on-type account does not. Second, an organizational fit: the hiring
venture must be structured to absorb and use what these employees carry. Third, a
human capital fit, in which the competencies acquired in running a firm match the
competence requirements of an early-stage venture rather than being generically
valuable. The three are stated as conditions under which the association
holds rather than as a single flat effect, which is why the headline number
understates what happens in the ventures where the fit is good.

Two cautions belong with the result. The estimand is associational, as the authors
state; register data buy an unusually rich confounder set but do not shut down
sorting, and the most obvious alternative story—that ventures with better
prospects both attract former founders and survive longer—is exactly the one the
mechanism tests are working to constrain. And survival is a coarse outcome: it is
institutionally observable and comparable across the whole population, which is
why it is the right dependent variable at this scale, but it is silent about the
distributional questions @agarwal2026startup raises for founder knowledge.

Read alongside @law2025joiners, the two papers close a loop. Law and coauthors
trace the joiner-to-founder channel, in which start-up employment raises the
likelihood of later founding; @braunerhjelm2026staying trace the same channel in
reverse, with former founders returning to wage employment and carrying their
entrepreneurial capital onto someone else's payroll. Neither the supply of
founders nor the supply of employees is exogenous to the population of ventures.
For marketing this matters because most of a young venture's customer-facing
capability—pricing, positioning, channel, the first sales hires—sits with
employees rather than founders, and this literature's usual practice of treating
the founder as the venture's entire stock of human capital measures the wrong
person.
## Getting Resources: Evaluation, Signaling, and Persuasion {#sec-ent-resources}

A venture must persuade before it can transact. Its earliest audiences are not
customers but investors, regulators, and prospective employees, and the
persuasion problem it faces with them is structurally the same one marketing
studies with consumers: an audience forms an evaluation under uncertainty from
signals the venture partly controls.

@gius2025disagreement reports a result that inverts the usual reading of
evaluator consensus. Across 67 venture competitions, the *more* judges disagree
about a start-up's quality, the more likely it is to succeed—particularly when
its proposition is unique. The argument is that common opinion cannot itself be a
source of competitive advantage, so value accrues disproportionately to founders
with unconventional ideas, which are exactly the ideas that provoke
disagreement. The paper also examines which evaluators disagree more (former
entrepreneurs) and which aspects of a venture are most polarizing (the business
model). For anyone who aggregates expert ratings into a quality score, the
implication is direct: the variance of the ratings carries signal that the mean
discards.

@conti2025opensource study signaling through observable activity rather than
pitch. Linking 160,065 U.S. start-ups in Crunchbase to their GitHub activity,
they find in a matched difference-in-differences design that engaging with open
source *communities* raises the likelihood of subsequent funding, while using
GitHub for internal development only does so much more weakly. The effect is
stronger for firms developing novel technologies and weaker in highly
competitive environments—a visible trade-off between community engagement and
appropriability. Machine-learning classification of technology use cases and
data on product launches support access to external knowledge as an important
channel.

@sanchezruiz2025narcissism take the founder's self-presentation as the signal.
Separating narcissistic *admiration* from narcissistic *rivalry*, and using
Shark Tank pitch data, they find the two associate with opposite funding
patterns through opposite movements in investor sentiment: admiration raises
sentiment and funding, rivalry lowers both. The methodological lesson
generalizes past this setting—a construct that behaves inconsistently across
studies may be two constructs that have been summed.

@kanze2026empathy inverts the direction of evaluation. Venture funding is a
two-sided matching process, yet the literature has attended almost exclusively to
investors evaluating founders. Across an archival study of 677 entrepreneurs
evaluating 408 investors and a preregistered experiment on 481 entrepreneurs,
investors with prior entrepreneurial experience are perceived as more
empathetic, and perceived empathy mediates the relationship between shared
experience and how favorably founders rate them. Linguistic analysis separates
cognitive from affective perspective-taking as components. The paper also
documents that women are less likely to move from entrepreneurship into
investing, which constrains their access to these relational returns.

@fig-ent-resources places these audiences in one frame.

```{mermaid}
%%| label: fig-ent-resources
%%| fig-cap: "The venture's persuasion problem. Before it reaches a customer, a new venture must be evaluated by investors, regulators, and prospective employees, each reading a different signal and applying a different decision rule. The dashed path marks the two-sided nature of investor matching: founders evaluate investors as well."
flowchart LR
    V["New venture"]
    P["Pitch and<br/>self-presentation"]
    A["Observable<br/>activity"]
    R["Positioning against<br/>a reference set"]
    I["Investors"]
    G["Regulator"]
    E["Prospective<br/>employees"]
    O["Resources:<br/>capital, clearance, talent"]
    V --> P
    V --> A
    V --> R
    P --> I
    A --> I
    R --> G
    P --> E
    I --> O
    G --> O
    E --> O
    I -.->|"founders evaluate investors"| V
```

## Where Ventures Operate: The Legal Environment {#sec-ent-legal}

The sections above treat the venture's environment as a set of audiences to be
persuaded. There is a second environment, less visible and rarely modeled, that
sets the expected cost of operating at all: the courts that will hear the
venture's disputes.

@canayaz2026judges supply the first systematic evidence that small businesses
respond to it. Their finding is that shifts toward more liberal—that is, less
probusiness—federal courts predict reductions in local small business activity.
A one-standard-deviation increase in liberal ideology predicts a 1% reduction in
firm count *relative to the county it borders in a different federal circuit*,
and the effect has more than doubled since 2000. The reduction is
concentrated in circuit-years with elevated small-business-related federal cases
or heightened public attention to the legal environment, which is what one
expects if the channel is the *salience* of expected litigation costs rather
than a mechanical legal constraint.

Two features of the result matter more than the headline magnitude. First, the
adjustment runs through **exit of young firms rather than relocation**. Firms do
not pick up and move across the circuit line; marginal young firms fail to
survive on the less favorable side. This is an important discipline on how the
finding should be read: judicial ideology is not shifting the map of where
existing businesses sit, it is shifting the survival threshold for businesses
that have not yet established themselves. Second, the decline coincides with a
**compositional shift away from litigation-sensitive industries**. The local
economy does not merely shrink; its industry mix tilts toward activities with
less legal exposure. A count of firms conceals both mechanisms, and reporting
only the count would have made the result look like a simple level effect.

### The Design: A Boundary, Not a Cutoff Score

The identification strategy is worth isolating, because it is the version of
regression discontinuity that geographic marketing data most often admits.
Federal appellate jurisdictions partition the United States into circuits, and
those circuit boundaries run *through* otherwise continuous economic geography.
Counties on either side of such a boundary share labor markets, weather, input
prices, and regional demand shocks, but fall under courts whose ideological
composition moves independently. Comparing a county to the county it physically
borders, when the two sit in different circuits, differences out the local
economic conditions that would otherwise confound any cross-sectional
comparison of business activity and legal climate.

This is a **boundary discontinuity** design, developed at greater length in
@sec-causal-rd-boundary. The running variable is not a score a firm can
accumulate but a location, and the identifying assumption correspondingly
changes: what must hold is not that firms cannot manipulate a test score but
that nothing *else* discrete changes at the same line. That assumption is
substantive rather than statistical, and here it bites hard, because federal
circuits are built from whole states—so every circuit boundary is *also* a state
boundary, carrying tax codes, licensing regimes, minimum wages, and
incorporation law, each of which independently affects small business formation.

What rescues the design is that the treatment has a time dimension. Circuit
ideology turns over as judges are appointed and replaced, for reasons having
nothing to do with any particular county's economy, so identification can come
from *changes* within a fixed pair of bordering counties rather than from the
permanent difference between them. The permanent state-policy gap is then
absorbed rather than assumed away, and what remains to be ruled out is the much
narrower possibility that state policy shifts in step with judicial turnover.
The general lesson travels: a boundary design whose treatment is static is
usually estimating the effect of a bundle, while one whose treatment moves along
a fixed boundary over time is estimating something closer to what it claims.

### Why This Belongs in a Marketing Reader

The paper is a finance paper, and it is not obvious on its face why a marketing
doctoral student should read it. Three reasons.

It supplies a **cost shifter that entry models routinely omit**. The formal
sketch in @sec-ent-model treats the fixed cost of the product mode, $f_P$, as a
parameter. Expected litigation cost is part of that parameter, it varies across
locations for reasons unrelated to demand, and it is therefore a candidate
instrument as well as a candidate covariate—an exclusion restriction that market
conditions cannot supply. The entry-timing questions of @sec-market-entry inherit
the same opportunity.

It makes the **exit margin measurable**. Marketing's customer-side models are
overwhelmingly models of acquisition and retention on the demand side; the
supply side of small business survival is usually left to industrial
organization. Here is a design in which the firms that disappear are identified
and located, which is exactly the variation needed to ask whether the ventures
that fail under legal pressure are the ones with weaker customer franchises.

It is a **public-policy result about a policy no one legislated**. The welfare
apparatus of @sec-welfare-calibration presumes a policy instrument with a lever
attached. Judicial ideology has no lever; it drifts with appointments made for
unrelated reasons, and its consequences for small business fall on communities
that never voted on it. That combination—a real distributional effect, no
policy-maker who chose it—is a specific and underexploited genre for the
policy-and-welfare paper described in @sec-policy-papers-marketing.

The caution to carry with it is the one the paper itself observes. The verb
throughout is *predict*, and the mechanism—expected litigation costs—is inferred
from the pattern of heterogeneity rather than measured directly. Nobody observes
a founder's estimate of what a lawsuit would cost. The heterogeneity results
(salience, litigation-sensitive industries) are what make the interpretation
credible, and they are doing more work than the point estimate is.

## What Marketing Brings, and What Is Missing {#sec-ent-marketing}

Read as a body, this literature has a conspicuous absence: the customer. Value
capture is measured in licences, clearances, funding rounds, and survival.
Positioning is studied with a regulator as the audience. Signaling is studied
with investors as the audience. The demand side enters, when it enters at all,
as an unmodeled residual.

Four openings follow directly, and each is a place where marketing has
constructs and data that the strategy literature does not.

*Mode of value capture has a demand-side determinant that is never measured.*
@eq-ent-payoffs treats $f_P$, the fixed cost of the product mode, as a
parameter. In reality much of it is the cost of building the customer-facing
assets—brand, channel access, salesforce—whose measurement is the subject of
@sec-branding and @sec-sales. A venture with a technically weak but
commercially legible proposition may enter the product market precisely because
its founders can build those assets cheaply, and nothing in the current
literature measures that.

*Pioneering products are studied without pioneer advantage.* The bridging
founders in @moeen2026valuecapture arrive at pioneering products, which is the
exact object of @sec-market-entry: whether being first confers durable share,
and through which mechanism. The two literatures describe the same firms and do
not cite each other. Whether founder knowledge profile predicts the *durability*
of a pioneer's advantage, rather than merely the act of pioneering, is an open
and answerable question.

*Founder heritage should have a brand signature.* If practices
[@feldman2019heritage] and culture [@ahn2025cultural] transfer from parent to
spawn, positioning and brand-identity elements plausibly do too, and these are
directly measurable from the venture's own communications. A spawn whose
positioning is inherited from its parent is competing in a category the parent
defined, which has implications for differentiation that neither literature has
drawn.

*Evaluator disagreement is a measurement problem marketing already solves.* The
result in @gius2025disagreement—that rating variance predicts success—is a claim
about the information content of dispersion in a panel of judges. Marketing has
an extensive apparatus for exactly this in concept testing and expert panels
(@sec-preference-measurement), including models that separate genuine
disagreement from scale-use heterogeneity. Applying it would establish whether
the disagreement result survives the correction.

## Reading This Literature: The Descriptive Design {#sec-ent-descriptive}

A reader arriving from empirical marketing will notice that several of the
studies above make no causal claim. @moeen2017incubation reports "stylized
findings." @moeen2026valuecapture reports a correspondence between founder
origin and mode of capture. @kovacs2026political reports associations and grades
them by robustness. None of these is a design failure; each is a deliberate
choice to map a phenomenon whose shape was unknown, on the reasoning that a
causal estimate of the wrong quantity is worse than an honest description of the
right one.

The genre has explicit methodological warrant. @helfat2007stylized argues that
stylized facts are a legitimate and necessary input to theory development in
management, and @hambrick2007devotion argues that the field's insistence on
theory in every paper suppresses exactly the descriptive work that theory needs
to feed on. @choudhury2026abduction give the positive account: abduction, the
mode of inference that formulates explanations for unexplained phenomena, is a
distinct and teachable research design rather than a deduction that failed.

How to *write* such a paper—what makes a descriptive contribution defensible,
where the design typically fails, and what the norms are in marketing as opposed
to management—is treated as a craft problem in @sec-sw-descriptive. When the
paper goes one step further and argues for a particular explanation of the
pattern it documents, the reporting structure for that argument is the PEEBI
scheme of @pillai2026peebi, developed in @sec-sw-peebi.

## Key Takeaways

- **Founder origin is a knowledge variable, not a demographic control.**
  Academic, employee, and user founders carry systematically different knowledge
  profiles out of systematically different prior contexts [@agarwalshah2014],
  and what transfers includes practices [@feldman2019heritage] and culture
  [@ahn2025cultural], not only individual skill.
- **The timing of founding is partly a marketing artifact.** Entry onto an
  e-commerce platform rises when a nearby store's rating is upgraded, decaying
  with distance and rising with the salience of the upgrade; the entrants so
  induced underperform and exit more, especially in downturns
  [@huang2026salience]. A rating system built for buyers doubles as an
  entry-inducement device for sellers (@sec-ent-salience).
- **The right dependent variable is often a category, not a magnitude.** Modes
  of value capture—technology market, product market, no capture—are
  institutionally observable and carry the strategic content that a performance
  scalar averages away [@moeen2017incubation; @moeen2026valuecapture].
- **Origin predicts mode, and the exception identifies the mechanism.**
  Academic founding maps to markets for technology and employee founding to
  markets for product; founders bridging both dimensions shift to pioneering
  products [@moeen2026valuecapture]. Formally, the pioneering mode is
  multiplicative in the two knowledge types and so is reachable only from the
  interior of the knowledge plane (@eq-ent-payoffs, @fig-ent-partition).
- **Report distributions, not means.** Pre-entry knowledge widens the
  performance distribution far more than it shifts its location
  [@agarwal2026startup]; a mean-only summary reverses the sign of the tail
  effects.
- **Entrepreneurial human capital is not only the founder's.** The share of a
  venture's employees who have previously run a firm is associated with survival,
  and the association operates through learning from longer prior spells,
  organizational absorptive fit, and competence fit rather than as a single flat
  effect [@braunerhjelm2026staying]. The joiner-to-founder channel
  [@law2025joiners] and the founder-to-joiner channel run in both directions
  through the same labor market (@sec-ent-ehc).
- **The first audience is not the customer.** Ventures are evaluated by
  investors, regulators, and employees, each reading a different signal
  [@gius2025disagreement; @conti2025opensource; @sanchezruiz2025narcissism;
  @pahnke2026positioning], and investor matching runs in both directions
  [@kanze2026empathy].
- **The operating environment is a cost shifter, not a control.** Judicial
  ideology—an institution no founder chose and no legislature set—predicts local
  small business counts, and it does so through the exit of young firms rather
  than relocation, alongside a compositional shift away from litigation-sensitive
  industries [@canayaz2026judges]. The contiguous-county design that identifies it
  is the geographic form of regression discontinuity (@sec-causal-rd-boundary).
- **Descriptive is a design, not a deficiency**—when the outcome categories are
  institutionally classified, the typology is declared in advance, and the
  paper states plainly what it is not claiming (@sec-ent-descriptive,
  @sec-sw-descriptive).

## Further Reading

Begin with @agarwalshah2014 for the knowledge-sources taxonomy that organizes
the field, then @agarwal2025industries for the synthesis of how heterogeneous
actors create industries. The value-capture program is best read in sequence:
@moeen2017incubation for the stylized findings and the multiple-yardsticks
argument, @moeen2020building for uncertainty reduction across industry
milestones, and @moeen2026valuecapture for the founder-level correspondence that
is this chapter's centerpiece. On heritage, @zucker2002commercializing remains
the cleanest demonstration that knowledge moves with people;
@feldman2019heritage and @ahn2025cultural extend it to practices and culture.
For teams and outcomes, pair @agarwal2026startup with @howell2026solo and
@kovacs2026political. On evaluation and resource acquisition,
@gius2025disagreement, @conti2025opensource, @sanchezruiz2025narcissism, and
@kanze2026empathy cover pitch, activity, self-presentation, and two-sided
matching respectively; @pahnke2026positioning covers the regulator.
On the legal environment that sets the cost of operating, @canayaz2026judges is
the entry point, and it doubles as a worked example of the boundary-discontinuity
design treated in @sec-causal-rd-boundary. @law2025joiners is the entry point on how employment structures the supply of
founders, and @braunerhjelm2026staying is its mirror image, tracing former
founders back onto the payrolls of new ventures. On what triggers the
founding decision in the first place, @huang2026salience is the entry point, and
it reads well against the salience machinery in @sec-thn-foundations. On the descriptive genre itself, read @helfat2007stylized and
@hambrick2007devotion alongside @suttonstaw1995, and then
@choudhury2026abduction for the constructive account. The entry-order material
this chapter presumes is in @sec-market-entry, and the resource-assembly logic it
mirrors is in @sec-ms-corpdev.
