# Communication

## Information Theory

### Entropy

Most of the examples are taken from `philentropy` package in R, which can be accessed via this
[link](https://cran.r-project.org/web/packages/philentropy/vignettes/Information_Theory.html)

#### Discrete Entropy

If $\mathbf{p}_X = (p_{x1},p_{x2},...,p_{xn})$ is the probability distribution of the discrete random variable X with
$P(X = x) = p_{xi}$. Then, the **entropy** of the distribution is [@Shannon_1948]

$$
H(\mathbf{p}_X) = -\sum_{x_i} p_{xi}\log_2 p_{xi}
$$

where

-   entropy unit is **bits**, and $H(p) \in [0,1]$ and max at $p = 0.5$
-   or the unit is **nats** ($log_e$)
-   or the unit is **dits** ($log_{10}$)

In general, if you have different bases, you can convert between units. Given bases a and b

$$
\log_a k = \log_a b \times log_b k 
$$

where $\log_a b$ is a constant.

Generalize

$$
H_b(X) = log_b a \times H_a(X)
$$

Equivalently,

$$
H(X) = E(\frac{1}{\log p(x)})
$$

Entropy is the measure of how much information gained from learning the value of X [@Zwillinger_1995, pp.262]. Entropy
depends on the distribution of X.

\
Example:

If X has 2 values, then $\mathbf{p}= (p,1-p)$ and

$$
H(\mathbf{p}_X ) = H(p,1-p) = -p \log_2 p - (1-p) \log_2 (1-p)
$$

which is known as binary entropy function

Plot of p vs. H(p) where max of $H(\mathbf{p}_X) = \log_2 n$ when $X \sim U$

```{r}
hb <- function(gamma) {
    -gamma * log2(gamma) - (1 - gamma) * log2(1 - gamma) # output in bits
}

xs <- seq(0, 1, len = 100)
plot(xs,
     hb(xs),
     type = 'l',
     xlab = "p",
     ylab = "H(p)")
```

```{r}
library("philentropy")
Prob <- 1:10/sum(1:10) # probabilities P(X)
H(Prob) # Shannon's Entropy
```

#### Mutual Information

If X and Y are two discrete random variables and $\mathbf{p}_{X \times Y}$ is the their joint distribution, the **mutual
information** (i.e., learning of a value of X gives information about the value of Y) of X and Y is

$$
I(X,Y) = I(Y,X) = H(\mathbf{p}_X)+ H(\mathbf{p}_Y)- H(\mathbf{p}_{X \times Y})
$$

where $I(X,Y) \ge 0$ and $I(X,Y) = 0$ iff X and Y are independent (information about X gives no information about Y).

Equivalently,

$$
MI(X,Y) = \sum_{i=1}^n \sum_{j=1}^m P(x_i, y_j) \times \log_b (\frac{P(x_i,y_j)}{P(x_i) \times P(y_j)})
$$

Properties:

-   $I(X;X) = H(X)$ known as self-information
-   $I(X;Y) \ge 0$ with equality iff $X \perp Y$

```{r}
P_x <- 1:10/sum(1:10) # distribution P(X)
P_y <- 20:29/sum(20:29) # distribution P(Y)
P_xy <- 1:10/sum(1:10) # joint distribution P(X,Y)
MI(P_x, P_y, P_xy) # Mutual Information
```

#### Relative Entropy

The measure of the distance between distribution p and q is called **relative entropy**, denoted by $D(p ||q)$

Equivalently, it measures the inefficiency of assuming distribution q when the true distribution is p.

The relative entropy between p(x) and q(x) is

$$
D(p||q) = \sum_x p(x) \log \frac{p(x)}{q(X)}
$$

where

-   $D(p||q) \ge 0$ and equality iff $p = q$

#### Joint-Entropy

$$
H(X,Y) = - \sum_{i=1}^n \sum_{j = 1}^m P(x_i, y_j) \times \log_b(P(x_i,y_j))
$$

```{r}
P_xy <- 1:10/sum(1:10) # joint distribution P(X,Y)
JE(P_xy) # Joint-Entropy
```

<br>

#### Conditional Entropy

$$
H(Y|X) = - \sum_{i=1}^n \sum_{j = 1}^m P(x_i, y_j) \times \log_b(\frac{P(x_i)}{P(x_i,y_j)})
$$

```{r}
P_x <- 1:10/sum(1:10) # distribution P(X)
P_y <- 1:10/sum(1:10) # distribution P(Y)
CE(P_x, P_y) # Joint-Entropy
```

Relation among entropy measures

$$
H(X,Y) = H(X) + H(Y|X) = H(Y) + H(X|Y)
$$

Properties:

-   Chain rule: $H(X_1, ..., X_n) = \sum_{i=1}^n H(X_i | X_{i-1},...,X_1)$
-   $0 \le H(Y|X) \le H(Y)$ Information on X does not increase the entropy measure of Y
-   If $X \perp Y$, then $H(Y|X) = H(Y); H(X,Y) = H(X) + H(Y)$
-   $H(X) \le log |n_X|$ where $n_X$ = number of elements in X
-   $H(X) = \log (n_X)$ iff $X \sim U$
-   $H(X_1,...,X_n) \le \sum_i^n H(X_i)$
-   $H(X_1,...,X_n) = \sum_i^n H(X_i)$ iff $X_i$'s are mutually independent.

#### Continuous Entropy

If X is a continuous variable, its entropy is

$$
h(\mathbf{X}) = - \int_{R^d} p(\mathbf{x})\log p(\mathbf{x}) dx
$$

where d is the dimension of X.

If X and Y are continuous random variables with density functions $p(\mathbf{x})$ and $q(\mathbf{y})$, the relative
entropy is

$$
H(X,Y) = \int_{R^d} p(x) \log \frac{p(x)}{q(x)} dx
$$

More advance analysis can be access [here](https://elife-asu.github.io/rinform/#4_time_series_measures)

### Divergence

Three metrics for divergence can be found
[here](https://cran.r-project.org/web/packages/philentropy/vignettes/Information_Theory.html):

-   Kullback-Leibler
-   Jensen-Shannon
-   Generalized Jensen-Shannon

<br>

### Channel Capacity

The information channel capacity of a discrete memoryless channel is

$$
C = \max_{p(x)} I(X;Y)
$$

which is the highest rate use at which information can be sent without much error.\
More information can be found [here](http://www.di.fc.ul.pt/~jpn/r/inf_theory/informationtheory.html)
