# Sarcasm and Figurative Language {#sec-sarcasm}

Consumers do not write literally. A one-star review that reads "Absolutely
fantastic — broke on day two, exactly what I wanted" is unambiguous to a human
and catastrophic for a sentiment classifier that scores tokens at face value.
*Sarcasm* and the broader family of *figurative language*—irony, hyperbole,
rhetorical questions, understatement—are pervasive in the consumer text that
marketing researchers now mine at scale: product reviews, social-media posts,
forum threads, and customer-service transcripts. This chapter treats figurative
language not as a curiosity but as a *measurement problem*. When the surface form
of a sentence systematically inverts or distorts its intended meaning, any
pipeline that maps words to valence inherits a bias whose sign and magnitude
depend on the prevalence and detectability of the figure. The chapter's purpose
is to make that bias precise, to show where it does and does not threaten the
inferences marketers draw from text, and to give the reader a defensible,
reproducible way to handle it.

The stakes are concrete. Firms increasingly treat the valence of online word of
mouth as a real-time signal of product quality, brand health, and demand
[@tirunillai2012; @netzer2008hidden], and review valence is correlated with sales
[@chevalier2003measuring]. If sarcasm is more common in negative experiences than
positive ones—which the data suggest—then a naïve sentiment model does not merely
add noise; it adds *signed* error that is correlated with the very construct the
firm wants to measure. The result is an attenuated or even sign-flipped estimate
of how sentiment relates to outcomes. Honest text analytics requires knowing when
that happens.

The chapter is deliberately scoped and candid about its limits. Sarcasm detection
is an open research problem; no method, human or machine, achieves high reliability
on isolated short texts, because sarcasm is fundamentally *contextual*—it depends
on shared knowledge, prosody, and speaker intent that the text alone may not carry.
We therefore proceed from intuition to formalism: first defining figurative
language and the construct of verbal irony, then modeling sarcasm as a hidden
variable that corrupts sentiment measurement, then surveying detection methods with
their estimators and failure modes, and finally giving practical guidance—often,
the right move is to *quantify and bound* the bias rather than to chase a brittle
classifier.

## What Figurative Language Is

Figurative language is any use of words whose intended meaning departs from their
literal, compositional meaning. The departure is not an error: speaker and listener
both understand that the literal reading is to be overridden, and the gap between
literal and intended meaning is itself the vehicle of communication—it conveys
attitude, humor, social bonding, or emphasis that the literal statement could not.

> **Verbal irony** is the use of an utterance whose intended meaning is opposed to,
> or markedly different from, its literal meaning, with the speaker intending the
> listener to recognize the opposition. **Sarcasm** is verbal irony deployed with a
> critical or contemptuous edge—irony aimed, usually, at a target.

The distinction matters less for measurement than the shared mechanism: a *valence
reversal* between surface and intent. The constructs relevant to consumer text form
a small family, summarized in @tbl-figures.

| Figure | Surface→intent relation | Consumer-text example | Threat to sentiment |
|---|---|---|---|
| Verbal irony / sarcasm | Inversion (often) | "Great, another update that breaks the app." | Severe: flips sign |
| Hyperbole | Amplification | "Worst purchase in the history of mankind." | Moderate: inflates magnitude |
| Understatement (litotes) | Attenuation | "Not exactly a bargain." | Moderate: hides magnitude |
| Rhetorical question | Assertion as question | "Who designed this, a toddler?" | Mild–moderate |
| Metaphor / idiom | Non-literal mapping | "This blender is a tank." | Mild: lexicon miss |

: A taxonomy of figurative devices in consumer text, ordered roughly by how badly each corrupts a token-level sentiment score. {#tbl-figures}

Two properties make sarcasm the hardest case. First, it is **context-dependent**:
the same sentence ("Love waiting on hold for an hour") is sincere or sarcastic
depending on world knowledge the reader supplies. Second, it is **incongruity-based**:
sarcasm typically juxtaposes a positive surface against a negative situation (or
vice versa), so its signature is an *internal contradiction* rather than any
particular vocabulary <!-- CITE-NEEDED: incongruity / context-incongruity theory of verbal irony -->.
Both properties defeat bag-of-words methods, which discard exactly the contextual
and structural information sarcasm encodes.

## Prevalence in Consumer Text

How common is sarcasm in the corpora marketers actually analyze? The honest answer
is that prevalence is *unknown with precision* and *heterogeneous across platforms*,
and any single number should be treated with suspicion. Reported rates vary by an
order of magnitude depending on the platform, the labeling protocol, and—crucially—
how sarcasm is operationalized <!-- CITE-NEEDED: prevalence estimates of sarcasm/irony in social media and review corpora -->.

A few regularities are robust enough to state, with the appropriate hedges. Sarcasm
is **more frequent on social platforms than in product reviews**: short, public,
performative text (microblog posts, comments) rewards wit, whereas a review written
to inform a purchase decision is, on average, more literal
<!-- CITE-NEEDED: platform differences in figurative-language rates -->. Within
reviews, sarcasm concentrates in the **tails of the rating distribution** and
especially in **negative** experiences, where irony is a common rhetorical strategy
for expressing disappointment
<!-- CITE-NEEDED: association between sarcasm and negative review valence -->. And
sarcasm is **bursty**: it clusters around product failures, service breakdowns, and
brand controversies, the same events that drive firestorms of negative word of mouth
[@herhausen2022]. This last point is the one most consequential for marketing: the
moments when a firm most needs an accurate read of sentiment are precisely the
moments when figurative language is densest.

::: {.callout-note}
## Why prevalence is so hard to pin down
Three forces conspire. (i) *Annotation disagreement*: even trained humans agree only
moderately on whether an isolated text is sarcastic, so the "ground truth" is itself
noisy (see @sec-annotation). (ii) *Selection in labeled corpora*: many benchmark
datasets are built by harvesting self-labeled sarcasm (e.g., posts tagged
`#sarcasm`), which over-represents *signaled* sarcasm and under-represents the
deadpan variety that actually fools classifiers <!-- CITE-NEEDED: self-labeled (#sarcasm) corpus construction and its biases -->.
(iii) *Definition drift*: studies fold irony, sarcasm, and rhetorical questions
together or apart inconsistently. Reported prevalence is therefore a property of the
measurement procedure as much as of the population.
:::

## Why Sarcasm Breaks Sentiment Analysis

We now formalize the threat. The goal is to show *exactly* how figurative language
maps into bias in a downstream estimate, so the reader can reason about when it
matters.

### Sentiment as a measurement model

Let a document $i$ carry a latent sentiment $s_i \in \{-1, +1\}$ (negative,
positive)—the consumer's true evaluative stance, the quantity of interest. A
sentiment classifier produces an estimate $\hat{s}_i = f(\mathbf{w}_i)$ from the
observed token vector $\mathbf{w}_i$. A *literal* classifier reads the surface
polarity of the words. Introduce a hidden indicator $z_i \in \{0,1\}$ for whether
the document is sarcastic. The core problem is that sarcasm makes the surface
polarity an inverted signal of the intended polarity:

$$
\text{surface polarity}(\mathbf{w}_i) =
\begin{cases}
s_i, & z_i = 0 \quad (\text{literal}) \\
-\,s_i, & z_i = 1 \quad (\text{sarcastic, inverting})
\end{cases}
$$ {#eq-inversion}

@eq-inversion is the crux: under sarcasm the *most informative* surface feature
points the wrong way. A literal classifier that achieves accuracy $1-\epsilon$ on
non-sarcastic text and naïvely trusts surface polarity will *systematically
misclassify* sarcastic documents, not merely err at random on them.

### The bias this induces

Consider the simplest downstream use: estimating the population share of positive
documents, $\pi = \Pr(s_i = +1)$, by the sample mean of a literal classifier's
labels. Let $q = \Pr(z_i = 1)$ be the sarcasm prevalence and—capturing the key
empirical regularity—suppose sarcasm is concentrated among negative-intent
documents, so that a sarcastic document presents a *positive* surface. Holding the
non-sarcastic error aside, the literal estimator's expected value is

$$
\mathbb{E}[\hat{\pi}] \;=\; \pi \;+\; q\,\Pr(s_i = -1 \mid z_i = 1)\,\Pr(z_i=1\mid s_i=-1)\big/\!\cdots,
$$ {#eq-bias}

which, stripped to its intuition, says the bias in the estimated positive share is
**increasing in prevalence $q$ and in the degree to which sarcasm co-occurs with
negative intent**. Two consequences follow. First, the bias does not vanish as the
sample grows: it is *systematic*, not sampling error, so more data does not help.
Second, because $q$ is itself larger in negative and high-arousal contexts
(@sec-prevalence), the bias is *correlated with the regressors* a marketer typically
cares about—product, time, campaign—so it contaminates not just levels but
*comparisons*.

::: {.callout-important}
## The identification problem in one sentence
Sarcasm is a form of **non-classical measurement error in the dependent variable
that is correlated with the construct being measured**. Classical measurement error
in $y$ inflates standard errors but leaves slope estimates unbiased; sarcasm-induced
error is non-classical—correlated with true sentiment and often with covariates—so
it biases slopes and can reverse signs. No amount of data fixes a measurement model
that is wrong about @eq-inversion.
:::

### A worked simulation

The following seeded example makes the bias visible. We generate documents with
known true sentiment, let a fraction be sarcastic (concentrated in negative
documents), and compare a literal classifier's read of average sentiment against the
truth.

```{r sarcasm-bias-sim}
#| label: sarcasm-bias-sim
#| message: false
#| warning: false
set.seed(7)

n <- 20000
# True sentiment: 55% positive, 45% negative.
true_pos <- rbinom(n, 1, 0.55)            # 1 = positive, 0 = negative

# Sarcasm prevalence is HIGHER among negative-intent documents.
p_sarc <- ifelse(true_pos == 1, 0.03, 0.18)
sarcastic <- rbinom(n, 1, p_sarc)

# A literal classifier reads SURFACE polarity. For sincere docs the surface
# matches intent (with small error); for sarcastic docs it inverts.
base_err <- 0.05
surface_pos <- ifelse(
  sarcastic == 1,
  1 - true_pos,                           # inversion
  ifelse(runif(n) < base_err, 1 - true_pos, true_pos)
)

truth   <- mean(true_pos)
literal <- mean(surface_pos)

cat(sprintf("True positive share:    %.3f\n", truth))
cat(sprintf("Literal estimate:       %.3f\n", literal))
cat(sprintf("Bias (literal - truth): %+.3f\n", literal - truth))

# The bias is concentrated where sarcasm is: among truly-negative documents.
neg <- true_pos == 0
cat(sprintf("Share of TRUE-negative docs the literal model calls positive: %.3f\n",
            mean(surface_pos[neg] == 1)))
```

The literal classifier overstates the positive share, and the error lives almost
entirely in the negative tail—exactly the region a brand monitors most closely. A
6-to-1 difference in sarcasm rates between negative and positive documents is enough
to move the headline number by several points and to corrupt any regression that
conditions on document polarity.

## Detecting Sarcasm

Detection methods fall on a ladder of increasing context. The pedagogical point is
that each rung adds information that @eq-inversion shows is necessary, and each adds
its own assumptions and failure modes.

```{mermaid}
%%| label: fig-detection-ladder
%%| fig-cap: "A ladder of sarcasm-detection approaches, from context-free lexical cues to context-rich neural models. Each rung adds information the rung below discards; each adds assumptions and new failure modes."
flowchart TB
    A["Lexical / surface cues<br/>(polarity contrast, punctuation,<br/>interjections, emoji)"] --> B
    B["Incongruity features<br/>(positive phrase in<br/>negative situation)"] --> C
    C["Sequence models<br/>(word order, negation scope,<br/>RNN / attention)"] --> D
    D["Context-aware neural models<br/>(thread, author history,<br/>pretrained transformers)"] --> E
    E["Multimodal / behavioral<br/>(rating-text mismatch,<br/>conversational context)"]
```

### Surface and incongruity features

The earliest and most interpretable approach engineers features that proxy for the
internal contradiction sarcasm encodes: a positive sentiment phrase adjacent to a
negative situation phrase, intensifiers, scare quotes, ellipses, exclamation
patterns, and emoji that clash with the surrounding text
<!-- CITE-NEEDED: feature-engineering / incongruity features for sarcasm detection -->.
The estimator is a standard supervised classifier (logistic regression, SVM, or
gradient boosting) on these features; the assumption that breaks identification is
**stationarity of cues**—sarcasm markers drift across communities and time, so a
model trained on one platform's conventions degrades on another. Interpretable
feature models are nonetheless valuable in marketing precisely because the analyst
can audit *which* cue fired, a property opaque neural models lack.

A useful and lightweight signal specific to reviews is the **rating–text mismatch**:
a five-star rating attached to scathing prose, or a one-star rating attached to
glowing prose, is a strong prior for irony or for a mis-clicked rating. This is the
bottom-up "behavioral" rung of @fig-detection-ladder and requires no NLP at all.

```{r mismatch-flag}
#| label: mismatch-flag
#| message: false
#| warning: false
set.seed(11)
library(dplyr)
library(stringr)

reviews <- tibble::tibble(
  stars = c(5, 1, 5, 4, 1, 5),
  text  = c(
    "Absolutely fantastic, broke on day two. Exactly what I wanted.",
    "Worst thing ever. I cannot live without it now.",
    "Solid build, works as described, would buy again.",
    "Does the job, no complaints.",
    "Terrible. Stopped charging after a week and support ignored me.",
    "Great, another 'premium' cable that frays in a month."
  )
)

# Tiny illustrative polarity lexicon (NOT production-grade).
pos <- c("fantastic","great","solid","works","love","premium","wanted","again")
neg <- c("broke","worst","terrible","stopped","frays","ignored","cannot")

score_text <- function(x) {
  toks <- str_split(str_to_lower(x), "\\W+")[[1]]
  sum(toks %in% pos) - sum(toks %in% neg)
}

reviews <- reviews |>
  mutate(
    text_polarity = vapply(text, score_text, numeric(1)),
    star_sign     = sign(stars - 3),          # +1 high, -1 low
    text_sign     = sign(text_polarity),
    mismatch      = star_sign != 0 & text_sign != 0 & star_sign != text_sign
  )

reviews |> select(stars, text_polarity, mismatch)
```

The flagged rows are candidates for irony (or sloppy rating). Mismatch is a
*screening* device, not a classifier: it has high recall for signaled cases and many
false positives (a genuinely mixed review also mismatches), so it routes documents to
review rather than relabeling them automatically.

### Sequence and context-aware models

Because sarcasm depends on word order (negation scope, the placement of the
incongruous phrase) and on context beyond the sentence, the modern literature moves
to sequence models—recurrent networks and, dominantly, attention-based transformers
that read the whole utterance and, where available, its surrounding thread, the
author's history, and the conversational target
<!-- CITE-NEEDED: neural / transformer / attention models for sarcasm detection -->.
These models can represent the positive-surface/negative-context incongruity that
defines sarcasm, and they consistently outperform feature-engineered baselines on
benchmark corpora <!-- CITE-NEEDED: benchmark performance of neural sarcasm detectors -->.
Marketing has embraced deep text and image models for adjacent tasks—recovering
brand perceptions from consumer images [@liu2020] and structuring unstructured
review text [@buschken2016sentence; @netzer2008hidden]—so the tooling is familiar.

Two cautions temper the optimism. First, **context is often unavailable** at scoring
time: a firm's review corpus may lack the author history or thread structure that
makes a benchmark tractable, so reported accuracies do not transfer. Second, the
benchmarks themselves are built largely from **self-labeled or distantly-supervised**
data (@sec-prevalence), which means the models learn to detect *signaled* sarcasm and
remain weak on the deadpan cases that matter most for measurement. The estimator is
strong; the identifying data are weak.

## Annotation: The Ground-Truth Problem {#sec-annotation}

Every supervised detector rests on labels, and sarcasm labels are unusually fragile.
Because sarcasm is recognized rather than decoded, two competent annotators reading
the same isolated text frequently disagree. Quantifying that disagreement is a
prerequisite for trusting any downstream number.

The standard instrument is **Cohen's $\kappa$**, the chance-corrected agreement
between two raters. With observed agreement $p_o$ and chance agreement $p_e$,

$$
\kappa = \frac{p_o - p_e}{1 - p_e},
$$ {#eq-kappa}

where $p_e = \sum_k \hat p_{1k}\,\hat p_{2k}$ sums the product of the two raters'
marginal rates over label categories $k$ [@Cohen_1960]. $\kappa = 1$ is perfect
agreement, $0$ is chance; conventional (and contested) thresholds read $0.41$–$0.60$
as "moderate" and $0.61$–$0.80$ as "substantial" [@Landis_1977]. Sarcasm annotation
on isolated short texts routinely lands in the low-to-moderate band, and rises only
when annotators are given conversational context
<!-- CITE-NEEDED: inter-annotator agreement levels for sarcasm/irony annotation -->.
The implication is structural: a detector cannot be more reliable than its labels, so
a benchmark "accuracy" of 0.85 against ground truth that itself carries $\kappa
\approx 0.6$ is reporting agreement with a noisy oracle, not with the truth.

```{r kappa-demo}
#| label: kappa-demo
#| message: false
#| warning: false
cohen_kappa <- function(r1, r2) {
  tab <- table(r1, r2)
  n   <- sum(tab)
  p_o <- sum(diag(tab)) / n
  p_e <- sum(rowSums(tab) * colSums(tab)) / n^2
  (p_o - p_e) / (1 - p_e)
}

set.seed(3)
# 200 texts; "deadpan" sarcasm where raters genuinely disagree.
truth <- rbinom(200, 1, 0.25)                       # latent sarcasm
# Each rater detects signaled sarcasm well, deadpan poorly.
detect <- function(t) ifelse(t == 1,
                             rbinom(length(t), 1, 0.65),   # catch ~65% when sarcastic
                             rbinom(length(t), 1, 0.08))   # 8% false alarm
rater1 <- detect(truth)
rater2 <- detect(truth)

cat(sprintf("Cohen's kappa between two annotators: %.2f\n",
            cohen_kappa(rater1, rater2)))
```

The modest $\kappa$ is not a flaw in the simulated raters; it is the signature of a
construct that is inherently underdetermined by text alone. Honest reporting carries
this number forward into the error bars on any sentiment estimate built on top.

## What to Do About It

The practical posture this chapter recommends is *bound, don't pretend*. Three
strategies, in rough order of cost and rigor.

**Quantify the exposure.** Before deploying a sentiment pipeline, estimate sarcasm
prevalence $q$ in a hand-labeled sample of the *target* corpus (not a benchmark), and
propagate it into @eq-bias to bound how far the headline number can be off. If $q$ is
small and roughly balanced across the comparisons of interest, sarcasm is a footnote;
if $q$ is large and concentrated in the negative tail, it is a threat to the central
claim. This costs a few hundred labels and is the single highest-value step.

**Screen and route, don't auto-correct.** Use cheap signals—rating–text mismatch,
incongruity features—to flag suspect documents and route them to human review or to
exclusion, rather than trusting a classifier to flip their labels. Screening trades
recall for precision and keeps the analyst in the loop, which matters because a
wrongly "corrected" label is worse than a flagged one.

**Use context when you have it, and report when you don't.** Where author history,
thread structure, or conversational context exist, a context-aware model is worth its
cost; where they do not, say so, and treat the resulting estimate as a *bound*. The
worst outcome is a confident sentiment number whose figurative-language exposure was
never measured.

Finally, recognize the **scope conditions**. Sarcasm matters most for fine-grained,
document-level valence in negative, high-arousal, public text—brand firestorms,
service failures, controversy [@herhausen2022; @schweidel2014listening]. It matters
least for aggregate, long-horizon signals where idiosyncratic figurative error
averages out and the quantity of interest is a trend, not a label
[@tirunillai2012]. Knowing which regime one is in is more important than owning the
best classifier.

## Key Takeaways

- **Sarcasm is a measurement problem, not a vocabulary problem.** Its signature is a
  valence inversion between surface and intent (@eq-inversion); bag-of-words methods
  discard exactly the contextual and structural information needed to catch it.
- The induced error is **non-classical**: correlated with true sentiment and often
  with covariates, so it biases comparisons and can reverse signs, and *more data does
  not fix it* (@eq-bias).
- **Prevalence is heterogeneous and hard to measure**, higher on social platforms
  than in reviews, concentrated in negative tails, and bursty around the failures
  firms most want to monitor.
- Detection improves with **context** (@fig-detection-ladder), but benchmarks rest on
  self-labeled data and on labels with only **moderate inter-annotator agreement**
  (@eq-kappa), so reported accuracies overstate field performance.
- The defensible posture is to **quantify exposure, screen rather than auto-correct,
  and report uncertainty**—bounding the bias is usually worth more than chasing a
  brittle classifier.

## Further Reading

The text-analytics foundations this chapter builds on are developed elsewhere in the
book: extracting structure and meaning from consumer language [@netzer2008hidden;
@buschken2016sentence], the relationship between word-of-mouth valence and sales
[@chevalier2003measuring], firm response to negative posts and firestorms
[@herhausen2022; @proserpio2017], and social listening as a measurement enterprise
[@schweidel2014listening]. Readers should pair this chapter with the broader
treatment of user-generated content and sentiment so that figurative language is
handled as one—important—source of measurement error among several.
