# Branding {#sec-branding}

A *brand* is the set of associations that a name, symbol, or design evokes in the
people who encounter it. Those associations are an economic asset: they shift
demand, command price premiums, and—when they are strong—surface on the balance
sheet and in how equity markets value the firm. This chapter treats branding as
two tightly coupled objects: a **behavioral** construct (what a brand means to
consumers) and a **financial** one (what a brand is worth to the firm). A serious
account of branding must connect the two, and the modern empirical literature
increasingly does.

The chapter proceeds from the inside out. We begin with the raw materials of a
brand—its name, logo, and slogan. We then build to *brand equity*, the central
construct, and confront its measurement controversies head-on. From there we move
to the richer psychological structures brands inhabit—associations, personality,
relationships, love, authenticity—and to the strategic decisions managers actually
make: portfolio architecture, extensions, co-branding, crisis recovery, and global
strategy. We close with **brand valuation**, where the behavioral construct is
finally expressed in dollars, and supply reproducible code for the leading
valuation methods.

## Conceptual Foundations

Classical theory rationalizes brands as **quality signals**. When product quality
is unobservable before purchase (an experience or credence good), a brand name
that is expensive to build and easy to destroy can credibly signal quality: a firm
selling low quality cannot profitably imitate the investment because repeat
purchase will not materialize. On this view a brand economizes on search and
substitutes for information in information-poor markets.

Hyperconnectivity strains this account. @swaminathan2020 define the hyperconnected
world as the continuous, location-independent access among people, devices, and
organizations, and argue it changes branding along two axes. First, it **blurs
ownership**: when any stakeholder can broadcast, brand meaning is co-created rather
than dictated, moving brands from single to shared ownership. Second, it
**broadens boundaries**: more entities are branded (people, places, ideas,
organizations [@thomson2006; @zenker2010place]) and commercial brands increasingly
carry social missions. The signaling role weakens precisely where information is
now abundant—online reviews and search deliver quality information directly—so the
brand's value migrates from *signaling quality* toward other functions: a cue that
economizes on attention in information-rich environments, an instrument of identity
expression, a container of socially constructed meaning, and an architect of value
within networks.

Two complementary lenses organize the field. The **consumer** lens splits into an
economic view (brands as market signals) and a psychological view (brands as
knowledge structures in memory). The **firm** lens splits into a strategic view
(branding as a set of controllable actions) and a financial view (branding actions
priced in firm value). The psychological view rests on @belk1988's *extended
self*: consumers consume to construct and signal identity, so a brand becomes part
of who the consumer is. That premise—brands as identity infrastructure—recurs
throughout the chapter and explains why ostensibly "soft" constructs (personality,
love, authenticity) have hard consequences for demand. Reflecting the migration of
brand activity online, @muntinga2011 classify consumers' online brand-related
activities (COBRAs) along a ladder of engagement—*consuming* (passive), *contributing*
(rating, commenting), and *creating* (producing brand content)—and brands
increasingly cultivate owned channels and apps rather than renting attention from
marketplaces [@wichmann2021]. @fig-brand-lenses maps the two lenses and the four
views they contain.

```{mermaid}
%%| label: fig-brand-lenses
%%| fig-cap: "Two lenses on branding. The consumer lens treats the brand as a behavioral construct and splits into an economic view (brands as market signals) and a psychological view (brands as knowledge structures in memory); the firm lens treats the brand as a financial object and splits into a strategic view (controllable actions) and a financial view (actions priced in firm value)."
flowchart TB
  BR["Branding"]
  CL["Consumer lens<br/>what the brand means<br/>to consumers"]
  FL["Firm lens<br/>what the brand is worth<br/>to the firm"]
  EC["Economic view<br/>brands as market signals"]
  PS["Psychological view<br/>brands as knowledge<br/>structures in memory"]
  ST["Strategic view<br/>branding as a set of<br/>controllable actions"]
  FI["Financial view<br/>branding actions priced<br/>in firm value"]
  BR --> CL
  BR --> FL
  CL --> EC
  CL --> PS
  FL --> ST
  FL --> FI
```

## Brand Elements

The smallest controllable units of a brand are its **name**, **logo**, and
**slogan**. Each is a lever on perception before any product experience occurs.

**Names** carry semantic and phonetic content. Linguistically feminine brand names
raise perceived warmth, which in turn lifts attitudes and choice share, both
hypothetically and in consequential choice; the effect attenuates for male subjects
and for utilitarian products [@pogacar2021]. For technological products,
alphanumeric names (e.g., model-number conventions) are more acceptable than for
nontechnical products [@pavia1993]. Names are also sticky assets: in rebranding,
most of the sales gain is attributable to the pre- and post-rebranding brand
*identities* rather than to the transition itself [@tsai2015].

**Logos** trade off recognition against image. High-recognition logos tend to be
natural, harmonious, and moderately elaborate; complex, elaborate logos sustain
interest and favor more effectively [@henderson1998]. Symbol-based logos deliver
stronger self-identification benefits than wordmark-only logos [@park2013]. Logo
complexity also interacts with exposure: @meyvis2001brandlogo find that repetition
and consumer motivation jointly determine whether a complex logo is eventually
processed as rich or merely as hard, so a design that tests badly on first exposure
may still be the right long-run choice. @gorn2015yourcompany extends the argument to
descriptive versus non-descriptive marks and shows the fit between logo type and
brand meaning matters more than the aesthetic quality of either. Even the mark.s position on the package carries meaning: @noseworthy2014placelogo show that placing a logo high or low activates conceptual metaphors of power and quality, so a layout decision usually delegated to design is a positioning decision.
**Slogans** for strong brands are better liked and more recognizable than slogans
for weak brands, independent of respondents' ability to attribute them correctly
[@dahlén2005]—evidence that brand strength spills onto the evaluation of the brand's
own elements.

Whether a slogan is *memorable* is a separate question from whether it is liked, and
@aka2026slogans make it a predictive one. They assemble the memorability factors the
literature has proposed via a systematic review, operationalize them three ways—NLP
features, human ratings, and constructs derived from text embeddings—and score the
resulting models against two distinct outcomes: **recognition** (has this slogan been
seen before?) and **cued brand recall** (which brand does it belong to?). Splitting the
outcome that way matters, because a slogan can be highly recognizable and still fail the
job it was written for, which is to retrieve the brand; the two are separate dependent
variables, not two indicators of one construct.

The paper's most transferable move is what it does with the embedding model. Raw text
embeddings are used not as an alternative feature set but as a *benchmark* for
completeness: whatever predictive accuracy the embeddings achieve beyond the
literature-derived factors is an estimate of how much of slogan memorability the
published theory has failed to name. That inverts the usual role of a black-box
predictor. Instead of the interpretable model being the goal and the embedding being the
cheat, the embedding becomes an upper bound against which a body of theory is audited—
the same logic @sec-tad-applications applies to text measures generally, and a template
worth borrowing wherever a literature has accumulated a list of proposed drivers and
never checked whether the list is exhaustive. The applied payoff is a selection tool:
the models predict the memorability of *candidate* slogans, including for new brands
with no exposure history, which turns a copy decision usually settled by managerial
taste into a scored comparison.

## Brand Equity {#sec-brand-equity}

Brand equity is the value a brand name adds beyond a product's functional utility.
It is the central construct of the chapter and one of the few in marketing that has
generated two largely separate measurement literatures: a *customer-based* tradition
that reads equity from the consumer's mind (surveys, scales, choices) and a
*financial/market-based* tradition that reads it from prices, profits, and stock
returns. The two traditions rarely cite one another, yet they are measuring the same
underlying asset from opposite ends. The most consequential recent development is not
a new definition but a new *instrument set*: always-on computational measurement from
social text, images, co-purchase graphs, and, most recently, large language models.
This section leads with that computational frontier, then grounds it in the classical
customer-based and financial traditions, and closes by bridging explicitly to the
marketing-finance machinery of @sec-marketing-finance.

Two reference frameworks organize everything that follows. The first is Keller's
[@keller1993] definition of *customer-based brand equity* (CBBE) as the differential
effect of brand knowledge on consumer response to the brand's marketing: the same
product at the same price elicits a more favorable response under a strong brand. The
second is the *brand value chain* of @kellerlehmann2006, which traces equity through
four stages, marketing investment, customer mindset (awareness, associations,
attitudes), market performance (price premium, share, loyalty), and finally
shareholder value. The value chain is the spine that connects the customer-based and
financial traditions: customer-based measures sit at the mindset and market-
performance stages, financial measures at the shareholder-value stage, and a credible
account of brand equity must explain how value propagates from one to the next.
@srinivasan2009 (Srinivasan and Hanssens) supply the
empirical-methods companion to that conceptual chain, cataloguing the metrics (Tobin's
$q$, abnormal returns, event-study CARs) and methods (persistence models, vector
autoregressions, value-relevance regressions) that link marketing assets, brand equity
among them, to firm value. Bridging the two traditions matters because each is
incomplete alone: mindset measures can move without ever reaching the cash register,
and financial measures can move for reasons that have nothing to do with the brand.
@fig-brand-value-chain draws this spine and marks where each measurement tradition
reads it.

```{mermaid}
%%| label: fig-brand-value-chain
%%| fig-cap: "The brand value chain as the spine connecting the two measurement traditions. Equity propagates from marketing investment through customer mindset and market performance to shareholder value; customer-based measures read the mindset and market-performance stages, financial and market-based measures the shareholder-value stage."
flowchart LR
  MI["Marketing<br/>investment"] --> CM["Customer mindset<br/>awareness, associations,<br/>attitudes"]
  CM --> MP["Market performance<br/>price premium, share,<br/>loyalty"]
  MP --> SV["Shareholder<br/>value"]
  CB["Customer-based<br/>measures"] -.-> CM
  CB -.-> MP
  FM["Financial and<br/>market-based measures"] -.-> SV
```

A transparent operationalization at the market-performance stage is the price premium
a brand commands over an otherwise-equivalent generic,
$$
\text{BE}_b = p_b - p_{\text{generic}}
$$ {#eq-brand-premium}
which is conceptually distinct from *value added*, $p - c$ (price over cost of
production). A generic can carry value added (e.g., convenience) without carrying
brand equity; equity is specifically the increment attributable to the *name*
[@Aaker_1996]. A complementary distinction separates the **functional**, emotional,
and self-expressive benefits a brand's value proposition delivers [@Aaker_1996]:
functional benefits derive from product attributes; emotional benefits from feelings
during private use; self-expressive benefits from public signaling of identity.
Because mass production has commoditized quality, quality alone no longer confers
status [@Holt_1998], and the self-expressive component carries an increasing share of
equity.

@fig-brand-equity-routes maps the measurement routes developed in the remainder of
this section: the two classical traditions and the computational instruments that
now feed them.

```{mermaid}
%%| label: fig-brand-equity-routes
%%| fig-cap: "Routes to measuring brand equity. The customer-based tradition reads equity from the consumer mindset and from market behavior (mindset scales, revenue and price premiums, conjoint and choice models); the financial tradition reads it from prices, profits, and stock returns. The computational frontier supplies always-on instruments that stand in for periodic surveys and lead firm-value outcomes."
flowchart TB
  BE["Brand-equity<br/>measurement"]
  subgraph CBT["Customer-based tradition"]
    MS["Mindset scales<br/>loyalty, perceived quality,<br/>awareness, associations"]
    RP["Revenue and price premium<br/>brand vs private label<br/>in scanner data"]
    CC["Conjoint and choice models<br/>equalization price,<br/>brand-specific residual"]
  end
  subgraph FIN["Financial and market-based tradition"]
    TQ["Intangible-value decomposition<br/>Tobin's q logic"]
    SR["Stock returns and risk<br/>strong-brand portfolios,<br/>value relevance"]
  end
  CF["Computational frontier<br/>social text, follower networks,<br/>consumer images, co-purchase baskets,<br/>LLM synthetic respondents"]
  BE --> MS
  BE --> RP
  BE --> CC
  BE --> TQ
  BE --> SR
  CF -.->|"stands in for<br/>periodic surveys"| MS
  CF -.->|"leads returns<br/>and firm value"| SR
```

### The Computational Frontier {#sec-be-frontier}

The classical instruments for measuring equity (surveys, scanner regressions, choice
experiments) are periodic, expensive, and backward-looking. The defining shift of the
past decade is the migration of measurement onto continuous digital traces: review
text, social posts, follower graphs, consumer-posted images, co-purchase baskets, and
now synthetic respondents generated by language models. The methodological
through-line across this work is *representation plus validation*: every method pairs
a learned representation (topics, embeddings, image features, model judgments) with an
external validation against a traditional brand metric (a survey perceptual map, a
mindset tracker, a stock return). The right pedagogical emphasis is that pairing, not
any single algorithm.

The reference architecture for always-on tracking is @rust2021, who build a
continuously updating brand-reputation index from social-media streams and validate it
against established survey brand metrics. The lesson is that a well-constructed
real-time index can stand in for a quarterly survey, collapsing the measurement lag
from months to hours. Perception can also be recovered from network structure rather
than content: @culotta2016 infer brand perceptual attributes (eco-friendly, luxury,
nutritious) from *who follows whom* on social platforms, showing that the company a
brand keeps is itself a measurement instrument. From raw text, two complementary
geometries emerge. @netzer2012 recover competitive market structure from forum
co-mentions, turning unstructured chatter into a brand map; @tirunillai2014a (cited as
the LDA "chatter" study) extract latent quality dimensions, their valence, and their
dynamics from hundreds of thousands of reviews. The second wave replaces topic counts
with *embeddings*: @timoshenko2019 use word embeddings to extract non-redundant
customer needs from user-generated content, and @gabel2019map introduce P2V-MAP, a
product2vec model that learns interpretable brand and product positions from
basket co-occurrence at retail scale. @berger2020text codify the resulting toolkit
(dictionaries, topics, embeddings, classifiers) and, importantly for a measurement
chapter, its validation requirements.

Brand image is increasingly read from pixels rather than words. @liu2020 train a
deep convolutional network (BrandImageNet) that recovers perceptual brand attributes
directly from consumer-posted images, "visual listening in" as a complement to text
listening, and @hartmann2021power show that the *type* of consumer image (a "selfie"
with the brand versus a "packshot" of it) carries distinct brand meaning and drives
engagement differently. The social signal also propagates to firm value:
@tirunillai2012 establish that the volume of online chatter leads abnormal stock
returns, @luo2013 show social-media metrics are leading indicators of firm equity
value, and @colicev2018b route owned versus earned social
media through mindset metrics (awareness, satisfaction) to shareholder value, closing
the value chain in the social era.

The newest stream treats the language model itself as the measurement device.
@licastelo2024 test whether an LLM can generate brand perceptual maps and attribute
scores that match human survey perceptions, validating with a triplet-agreement
criterion (do model and humans agree which of two brands is closer to a third?). The
broader "silicon sampling" idea, simulating human respondents with a language model,
traces to @argyle2023. The stream is the most active and least settled: LLMs can scale
and pre-test brand measures cheaply, but current evidence is that they require a human
anchor for validity rather than replacing human respondents outright. We return to a
worked, API-free version of the perceptual-map idea in @sec-be-code.

### Customer-Based Measurement {#sec-be-cbbe}

The customer-based tradition measures equity from the consumer mindset. Its
managerial origin is Aaker's five-asset framework (awareness, perceived quality,
associations, loyalty, and other proprietary assets) in *Managing Brand Equity*
(Aaker 1991) and the "Brand Equity Ten" measurement battery in *Building Strong
Brands* (Aaker 1996, @Aaker_1996); its psychological grounding is Keller's CBBE
[@keller1993]. The dominant survey instruments operationalize these constructs as
multi-item scales. @yoo2001 give a parsimonious multidimensional consumer-based
brand-equity scale (loyalty, perceived quality, awareness/associations) plus a
four-item overall-equity scale; @netemeyer2004 develop and validate facet measures
(perceived quality, perceived value, uniqueness, and willingness to pay a price
premium) and show the price-premium facet is the proximal driver of purchase intent,
a result that motivates reading equity off market behavior directly.

#### Is Equity Reflective or Formative? {#sec-be-reflective-formative}

A consequential and unsettled question is whether CBBE is a reflective or a formative
construct, because the answer dictates which validity tests apply and how the
construct may legitimately be aggregated (@sec-measurement-scales). Let $\eta$ denote
latent brand equity and let $x_1,\dots,x_K$ be observed dimensions (perceived quality,
associations, awareness, loyalty). The reflective specification treats the dimensions
as manifestations of equity,
$$ x_k = \lambda_k \eta + \varepsilon_k, \qquad k = 1,\dots,K, $$
so the indicators should be internally consistent and interchangeable; this is the
view of @yoo2001. The formative specification reverses the arrows, treating the
dimensions as causes that compose equity,
$$ \eta = \sum_{k=1}^{K} w_k x_k + \zeta, $$
so the indicators need not correlate and are not interchangeable; this is the view of
@henseler2017. The two are not nested and imply different psychometrics: reflective
models are evaluated by internal-consistency and convergent/discriminant validity,
whereas formative models are evaluated by indicator weights and multicollinearity
diagnostics. A pragmatic third position treats equity as an outcome measured directly
(e.g., the premium in @eq-brand-premium) rather than scaled from its drivers; this is
often easier to defend to reviewers but is not the field's consensus.

#### Revenue and Price Premium {#sec-be-premium}

Reading equity off market behavior rather than surveys yields outcome measures that
are forward-looking and hard to game. @ailawadi2003 define the *revenue premium*, the
difference in revenue (price times volume) between a brand and a comparable
private-label or generic, net of cost, as an outcome measure of brand equity
computable directly from scanner data. The price-premium family (Aaker's "Brand Equity
Ten" premium item, @Aaker_1996) isolates the per-unit price a brand commands at equal
share. Revealed-preference decompositions in the same spirit recover equity from
single-source panels: @kamakura1993 split brand value into a tangible
(attribute-driven) and an intangible (name-driven) component within a choice model.
The revenue-premium computation, including a side estimate of own-price elasticity, is
demonstrated on simulated scanner data in @sec-be-code.

#### Conjoint and Choice-Based Equity {#sec-be-choice}

A third route recovers equity from trade-offs in choice. @parksrinivasan1994 decompose
brand equity in a survey-based conjoint into an attribute-based component (the part
explained by measured attributes) and a non-attribute, brand-specific residual, and
link the residual to extendibility. @swait1993equalization define the *equalization
price*, the price reduction that makes a no-name option as attractive as the branded
one in a discrete-choice model, giving equity a clean monetary interpretation as the
value of the brand's intangible utility. @erdemswait1998 supply the microfoundation:
brands act as *signals* that reduce perceived risk and information costs, so equity is
the credibility and clarity of that signal, mapping directly onto choice-model
utility. The equalization-price calculation is worked end-to-end on simulated choice
data in @sec-be-code.

### Financial and Market-Based Valuation {#sec-be-financial}

The financial tradition measures equity from the firm's market value. The foundational
method is @simon1993 (Simon and Sullivan), who estimate brand equity from financial-
market value: total intangible value is the excess of market value over the
replacement cost of tangible assets (a Tobin's $q$ logic), and brand equity is the
share of that intangible attributable to brand-related drivers (advertising share,
order of entry, the advertising-to-sales ratio). @barth1998brand show that published
brand-value estimates are *value-relevant*: they explain share prices and returns
incrementally over book equity and earnings, legitimizing brand value as an
accounting-relevant intangible. The Interbrand "Best Global Brands"
discounted-economic-profit method (implemented in the worked InterBrand demo later in
this chapter) is the leading practitioner instrument, and its academic legitimacy
rests on Barth et al. A Simon-Sullivan-style Tobin's $q$ decomposition on simulated
firm financials is given in @sec-be-code, and it is the explicit computational bridge
to @sec-mf-q.

#### Stock-Return and Risk Evidence {#sec-be-stock}

Brand equity shows up in stock returns and risk, not merely in accounting value.
@madden2006 (Madden, Fehle, and Fournier) form portfolios of strong-brand firms from
the Interbrand list and show they earn higher returns with lower risk than the market
and matched firms, a branding "alpha" after Fama-French and Carhart adjustment.
@mizik2008 (cited in the project as Mizik and Jacobson) show perceptual brand
attributes from the Young & Rubicam BrandAsset Valuator predict future stock returns,
implying the market does not fully or immediately impound brand-perception changes.
@aakerjacobson2001 establish the same leading relationship for brand attitude in
high-technology markets, and @johansson2012 show strong brands lost less value in the
2008 financial crisis, evidence that equity buffers downside risk. The estimation
problem of recovering the total financial impact of equity from short time series is
treated by @mizik2014. A simulated brand-portfolio long-short return spread, mirroring
@madden2006, appears in @sec-be-code.

### Bridge to the Marketing-Finance Interface {#sec-be-bridge}

The customer-based and financial traditions meet at firm value, and the meeting point
is the Tobin's $q$ valuation regression developed in @sec-mf-q. There, $q_{it}$, the
ratio of a firm's market value to the replacement cost of its assets, is regressed on
marketing assets with firm and year fixed effects,
$$ q_{it} = \beta\,\text{BrandIndex}_{it} + \mathbf{x}_{it}^{\top}\boldsymbol{\gamma}
   + \alpha_i + \delta_t + u_{it}, $$ {#eq-be-q}
so that a customer-based brand index (an embedding-derived equity score, a survey
mindset metric, a revenue premium) enters the same regression that prices any other
marketing asset. This is the literal computational bridge between the two traditions:
the customer-based literature supplies $\text{BrandIndex}_{it}$, and the
marketing-finance literature supplies the identification machinery and the firm-value
outcome. @srinivasan2009 is the methodological map for that regression, and the
caveats of @sec-mf-aatq (reverse causality, omitted intangibles, accounting-proxy
bias in $q$) apply in full—and so does the deeper one developed in
@sec-mf-marginal-vs-average: the $q_{it}$ on the left of @eq-be-q is an *average*
ratio, whereas the quantity a brand-investment argument needs is the *marginal*
value of a dollar of brand capital. Because brand assets are simultaneously an
intangible missing from the denominator and a source of the rents that inflate the
numerator, the wedge between the two loads directly onto
$\text{BrandIndex}_{it}$, so $\beta$ should be read as an upper bound
[@gala2026marginalq; @crouzet2023]. A structural counterpart is now available:
@belo2022 estimate brand capital's share of firm market value at 6–25% without
using any brand survey, which is the external benchmark against which an estimate
of $\beta$ can be sanity-checked.

The financial relevance of customer-based equity is by now well established: it is
positively associated with stock returns, firm value, and
credit ratings and negatively associated with idiosyncratic risk and earnings
volatility [@bharadwaj2011a; @larkin2013; @madden2006; @mizik2008; @rego2009], and the
asset is even priced in the managerial labor market, where executives accept lower pay
to be associated with strong brands, an effect concentrated among CEOs and younger
executives [@tavassoli2014employee]. We revisit valuation quantitatively at the end of
this chapter and formalize the event-study machinery in @sec-marketing-finance.

The bridge also carries a strategic question that brand equity alone cannot settle: how
much of the firm's marketing effort should go to building the brand rather than managing
individual customer relationships. That debate stalled for years on the absence of a
firm-level measure of emphasis. @scholdra2026focus supply one by scoring more than 95,000
earnings-call transcripts from over 1,800 firms across 22 years with separate
textbook-derived dictionaries for brand and customer management [@han2021capturing], and
report three results that should temper any general claim about brand-versus-customer
priority: the two foci are essentially uncorrelated, so neither a systematic trade-off nor
a systematic complementarity holds; they act on different performance outcomes, revenue-
versus cost-related; and a dual emphasis on both performs *worse* than committing to
either one. @sec-mf-focus develops the measurement and the monitoring pipeline.

### Worked Computations {#sec-be-code}

The five demonstrations below run on simulated data, span both measurement traditions,
and carry the marketing-finance bridge. They use base R plus `dplyr`, `ggplot2`, and
`MASS`, all already standard in this book.

A first demonstration computes the *revenue premium* of @ailawadi2003 on simulated
weekly scanner data for a national brand and a comparable private label across stores,
and recovers each brand's own-price elasticity as a side metric.

```{r be-revenue-premium, message=FALSE, warning=FALSE}
library(dplyr)
set.seed(42)

n_stores <- 30; n_weeks <- 52
grid <- expand.grid(store = 1:n_stores, week = 1:n_weeks)

# National brand: higher baseline price and stronger baseline demand (equity);
# private label: cheaper, weaker baseline. Both face log-log demand in price.
sim_brand <- function(base_logdemand, base_price, elasticity, n, sd_price) {
  price <- pmax(0.5, base_price + rnorm(n, 0, sd_price))
  units <- exp(base_logdemand + elasticity * log(price) + rnorm(n, 0, 0.20))
  data.frame(price = price, units = units)
}

nb <- sim_brand(base_logdemand = 7.0, base_price = 3.20, elasticity = -1.8,
                n = nrow(grid), sd_price = 0.25)
pl <- sim_brand(base_logdemand = 6.2, base_price = 2.10, elasticity = -2.4,
                n = nrow(grid), sd_price = 0.20)

scan_dat <- bind_rows(
  cbind(grid, brand = "National", nb),
  cbind(grid, brand = "PrivateLabel", pl)
)

# Revenue premium = (revenue of national brand) - (revenue of private label),
# aggregated to the store-week and averaged (Ailawadi, Lehmann & Neslin 2003).
rev_long <- scan_dat |>
  mutate(revenue = price * units) |>
  group_by(store, week, brand) |>
  summarise(revenue = sum(revenue), .groups = "drop")

rev_nb <- rev_long$revenue[rev_long$brand == "National"]
rev_pl <- rev_long$revenue[rev_long$brand == "PrivateLabel"]
revenue_premium <- mean(rev_nb - rev_pl)

# Own-price elasticity per brand from a log-log regression.
elas <- scan_dat |>
  group_by(brand) |>
  summarise(elasticity = coef(lm(log(units) ~ log(price)))[["log(price)"]],
            .groups = "drop")

cat("Mean revenue premium per store-week: $",
    format(round(revenue_premium), big.mark = ","), "\n", sep = "")
print(elas)
```

The second demonstration is the Simon-Sullivan financial decomposition. Across a
cross-section of simulated firms it computes Tobin's $q$, isolates intangible value as
market value net of tangible replacement cost, regresses that intangible on
brand-related drivers, and reports the fitted brand-equity share. This is the same
$q$ object used in @sec-mf-q.

```{r be-simon-sullivan, message=FALSE, warning=FALSE}
set.seed(7)
N <- 400
firm <- data.frame(
  tangible_repl = runif(N, 50, 500),                 # replacement cost of tangibles
  ad_share      = rbeta(N, 2, 5),                     # advertising share of sales
  order_entry   = sample(1:6, N, replace = TRUE),     # 1 = pioneer (more equity)
  rd_intensity  = rbeta(N, 2, 8)                      # other intangible (control)
)

# True data-generating process: intangible value rises in advertising share and
# pioneering (low order-of-entry number), plus non-brand intangibles (R&D) + noise.
brand_intangible <- 120 * firm$ad_share + 15 * (7 - firm$order_entry)
other_intangible <- 200 * firm$rd_intensity
firm$market_value <- firm$tangible_repl + brand_intangible + other_intangible +
  rnorm(N, 0, 25)

firm$q          <- firm$market_value / firm$tangible_repl
firm$intangible <- firm$market_value - firm$tangible_repl

# Regress intangible value on brand drivers (+ control); fitted brand component
# is the brand-equity share of intangible value (Simon & Sullivan 1993 logic).
fit <- lm(intangible ~ ad_share + I(7 - order_entry) + rd_intensity, data = firm)
firm$brand_equity_hat <- coef(fit)["ad_share"] * firm$ad_share +
  coef(fit)["I(7 - order_entry)"] * (7 - firm$order_entry)

brand_share <- mean(firm$brand_equity_hat) / mean(firm$intangible)

cat("Mean Tobin's q:                 ", round(mean(firm$q), 3), "\n")
cat("Brand share of intangible value:", round(100 * brand_share, 1), "%\n")
```

The third demonstration mirrors @madden2006: it simulates monthly returns for a
strong-brand and a weak-brand portfolio (the strong portfolio carries a small positive
alpha and lower idiosyncratic volatility), recovers alpha and beta from a CAPM-style
time-series regression on the strong portfolio, and reports the long-short
(strong minus weak) spread with a $t$-statistic.

```{r be-portfolio-spread, message=FALSE, warning=FALSE}
set.seed(2024)
Tm  <- 240                                   # 20 years of monthly returns
mkt <- rnorm(Tm, mean = 0.006, sd = 0.045)   # market excess return
rf  <- 0.0
alpha_strong <- 0.0025                        # ~30 bps/yr branding alpha
strong <- alpha_strong + 0.95 * mkt + rnorm(Tm, 0, 0.020)  # lower idio vol
weak   <- 0.0000      + 1.10 * mkt + rnorm(Tm, 0, 0.035)   # higher idio vol

capm   <- lm(I(strong - rf) ~ I(mkt - rf))
spread <- strong - weak
tt     <- t.test(spread)

cat("CAPM alpha (monthly):", round(coef(capm)[1], 5),
    " annualized:", round(((1 + coef(capm)[1])^12 - 1) * 100, 2), "%\n")
cat("CAPM beta:           ", round(coef(capm)[2], 3), "\n")
cat("Long-short mean spread (monthly):", round(mean(spread), 5),
    " t =", round(tt$statistic, 2), "\n")
```

The fourth demonstration computes the equalization-price equity of
@swait1993equalization on a simulated choice experiment. Three brands and a no-name
outside option vary in price and quality; a multinomial logit is fit by maximum
likelihood (hand-coded, no external choice package), and the equalization price is the
brand-specific intercept divided by the absolute price coefficient, that is, the price
cut that would make the no-name option as attractive as the branded one.

```{r be-equalization-price, message=FALSE, warning=FALSE}
set.seed(11)
n_resp <- 1500
J <- 4                                  # 3 brands + 1 no-name outside option

# True utilities: brand-specific intercepts (ASCs), price (neg), quality (pos).
asc   <- c(BrandA = 1.6, BrandB = 1.0, BrandC = 0.4, NoName = 0.0)
b_pr  <- -0.9
b_qu  <-  0.7

# Each respondent sees one choice set with random price/quality per alternative.
make_set <- function() {
  price   <- runif(J, 1, 5)
  quality <- runif(J, 0, 3)
  V <- asc + b_pr * price + b_qu * quality
  p <- exp(V) / sum(exp(V))
  choice <- sample(1:J, 1, prob = p)
  data.frame(alt = 1:J, price = price, quality = quality,
             chosen = as.integer(1:J == choice))
}
dat <- do.call(rbind, lapply(1:n_resp, function(i)
  cbind(set = i, make_set())))

# Design matrix: 3 brand dummies (NoName is the reference) + price + quality.
X <- cbind(
  BrandA  = as.integer(dat$alt == 1),
  BrandB  = as.integer(dat$alt == 2),
  BrandC  = as.integer(dat$alt == 3),
  price   = dat$price,
  quality = dat$quality
)
negll <- function(par) {
  V <- as.vector(X %*% par)
  df <- data.frame(set = dat$set, V = V, chosen = dat$chosen)
  denom <- tapply(exp(df$V), df$set, sum)
  ll <- sum(df$V[df$chosen == 1]) - sum(log(denom))
  -ll
}
est <- optim(rep(0, 5), negll, method = "BFGS")$par
names(est) <- colnames(X)

eq_price <- est[c("BrandA", "BrandB", "BrandC")] / abs(est["price"])
cat("Estimated coefficients:\n"); print(round(est, 3))
cat("\nEqualization price (value of brand over no-name), $/unit:\n")
print(round(eq_price, 3))
```

The fifth demonstration builds a brand *perceptual map* from text, in the spirit of
@netzer2012, @timoshenko2019, and @gabel2019map, without any external corpus or API.
Each brand is given a profile over latent association words; a term-by-brand matrix is
embedded with a low-rank singular value decomposition (a stand-in for word2vec or LLM
embeddings); brands are then projected to two dimensions by classical multidimensional
scaling and plotted with `ggplot2`. The same pipeline accepts real embeddings (from
`text2vec` or an `ellmer` call to a language model) in place of the simulated matrix.

```{r be-perceptual-map, message=FALSE, warning=FALSE, fig.width=6, fig.height=5}
library(ggplot2)
set.seed(99)

brands <- c("Summit", "Vela", "Norden", "Aero", "Lumen", "Cobalt")
assoc  <- c("rugged", "premium", "playful", "eco", "techy", "trusted",
            "affordable", "youthful")

# Latent positioning: each brand emphasizes a few associations.
profile <- matrix(rgamma(length(brands) * length(assoc), shape = 0.4, rate = 1),
                  nrow = length(assoc), dimnames = list(assoc, brands))
profile["rugged",  "Summit"] <- 6; profile["eco",     "Summit"]  <- 4
profile["premium", "Vela"]   <- 6; profile["trusted", "Vela"]    <- 4
profile["trusted", "Norden"] <- 6; profile["eco",     "Norden"]  <- 5
profile["techy",   "Aero"]   <- 6; profile["youthful","Aero"]    <- 4
profile["playful", "Lumen"]  <- 6; profile["youthful","Lumen"]   <- 5
profile["premium", "Cobalt"] <- 5; profile["techy",   "Cobalt"]  <- 5

# TF-IDF-style weighting, then a low-rank SVD embedding of the term-brand matrix.
tf  <- t(t(profile) / colSums(profile))
idf <- log(ncol(profile) / (rowSums(profile > 0.5) + 1))
M   <- tf * idf
sv  <- svd(M)
brand_emb <- diag(sv$d[1:4]) %*% t(sv$v[, 1:4])   # 4-dim brand embeddings
colnames(brand_emb) <- brands

# Classical MDS (base R) from cosine distances between brand embeddings.
cos_sim <- function(A) {
  A <- scale(A, center = FALSE, scale = sqrt(colSums(A^2)))
  t(A) %*% A
}
D   <- as.dist(1 - cos_sim(brand_emb))
mds <- cmdscale(D, k = 2)
map_df <- data.frame(brand = brands, Dim1 = mds[, 1], Dim2 = mds[, 2])

ggplot(map_df, aes(Dim1, Dim2, label = brand)) +
  geom_point(color = "steelblue", size = 3) +
  geom_text(vjust = -0.9) +
  labs(title = "Simulated brand perceptual map (embedding + MDS)",
       x = "Dimension 1", y = "Dimension 2") +
  theme_minimal() +
  coord_cartesian(clip = "off")
```

### Brand Loyalty, Awareness, and Prominence

**Loyalty** has two components that must not be conflated: *attitudinal* loyalty (a
favorable disposition) and *behavioral* loyalty (repeat purchase), a distinction
pioneered by Jacoby and Chestnut and reviewed by @berkowitz1978. The two diverge
under competitive promotion and can be separately diagnostic of equity.
**Awareness** (brand familiarity) correlates with market performance, but the link
is moderated by market characteristics (product homogeneity, technological
volatility) and buyer characteristics (buying-center heterogeneity, time pressure)
[@homburg2010].

**Prominence**—the conspicuousness of a brand's mark on a product—operates through
a status-signaling logic [@han2010]. Status accrues to evidence of wealth, i.e.,
to conspicuous consumption, and luxury goods confer status simply by being shown.
Crucially, *quiet* versus *loud* branding are different signals decipherable by
different audiences. @han2010 organize consumers on two dimensions—wealth and
need for status—into four types: **patricians** (high wealth, low need) pay a
premium for subtle signals only other patricians can read; **parvenus** (high
wealth, high need) buy loud signals to separate from have-nots; **poseurs** (low
wealth, high need) imitate parvenus, often with counterfeits; and **proletarians**
(low need) do not signal. Two empirical regularities follow and are observed in
market data: inconspicuously branded luxury goods can cost *more* than
conspicuously branded ones, and counterfeiters copy the louder, lower-priced items
that poseurs demand. Status-consistent communication norms reinforce this: reduced
emotionality in messaging signals high status by evoking high-status reference
groups [@lee2021], and prominence raises perceived quality and emotional value,
which feed purchase intention [@butcher2016].

Who actually buys the fakes is now visible at scale. @chen2026counterfeits analyze
millions of cross-border counterfeit orders by U.S. consumers and document a
*U-shaped* income gradient: relative to the middle-income cohort, both low- and
high-income consumers buy counterfeits more, and more repeatedly. The two tails differ
in composition in a way the signaling logic anticipates—low-income demand loads on
lower-tier brands, whereas high-income demand tilts toward ultraluxury and toward the
higher-priced listings and classic series (the Hermès Birkin, the Chanel Classic Flap),
as if paying for counterfeit *quality*. The poseur of @han2010 is real, but is only one
tail of a counterfeit market that reaches well into high-income households.

### Perceived Quality and Quality Signaling

Perceived quality is itself an asset: brand quality raises shareholder wealth
[@aaker1994]. Its dynamics are attribute-specific. As a warranty nears expiration,
satisfaction with *resolvable* attributes declines faster yet weighs more heavily on
overall quality perceptions, whereas *irresolvable* attributes decline more slowly
and weigh less over time [@slotegraaf2004longitudinal].

When quality is unobservable, firms can signal it through **specialization**.
@kalra2008signaling model a firm choosing to offer one service or two. Specializing
forgoes profit in the second category; that forgone profit is a **signaling cost**
that a high-quality firm can bear more cheaply than a low-quality imitator, so
specialization can support a separating equilibrium in which only high-quality
firms specialize. In homogeneous markets specialization is the primary quality
signal; in heterogeneous markets pricing can signal quality directly, but
specialization persists as an effective secondary signal because its signaling cost
is lower, and its use rises with competitive intensity. The general lesson—that
credible signals are costly differentially across types—recurs in the prominence and
authenticity literatures. Whether consumers actually *use* brands as signals in the
economic sense, rather than merely as heuristics, has been tested across countries:
@valenzuela2006brandssignals find the signaling account holds in markets that differ
substantially in brand infrastructure, which is the cross-country validation the
theory needed. Corporate-level associations layer on top of product-level ones, and
they are not simply additive — a corporate association can improve or damage response
to a new product depending on whether it is read as ability or as social
responsibility [@vanbruggen2005corporateassociati].

### Associations, Meaning, Image, Personality, and Coolness

Brand *associations* are the full network of brand-linked nodes in memory;
different audiences can hold contradictory associations with the same stimulus.
@batra2019 define **brand meaning** as the complete network of associations
produced by consumer interactions with the brand and its communications, sourced
from visual cues (advertising, packaging), sensory cues (music), and human cues
(endorsers, who act as conduits of cultural-meaning transfer). **Brand image**
governs transferability: extensions succeed when parent and target share attributes
*and* image (personality) similarity [@batra2004], with prestige- versus
popularity-based images interacting with the visual distance between product and
consumer to shape willingness to pay [@chu2021]. Corporate social responsibility
raises product evaluations even for attributes consumers can directly experience,
an effect dampened when the firm's motives are read as self-interested rather than
benevolent [@chernev2015]. Image can now be measured at scale: @liu2020 train a
multi-label convolutional network (BrandImageNet) to detect perceptual brand
attributes in consumer-created images, recovering survey-consistent perceptions in
near real time.

**Brand personality**—the human characteristics associated with a brand—loads on
five dimensions in @aaker1997: sincerity, excitement, competence, sophistication,
and ruggedness, with an accompanying scale; @grohmann2009 adds masculine and
feminine dimensions and shows spokespeople shape them and that
personality–self-congruence drives affective, attitudinal, and behavioral response.
@batra2010 supply a Bayesian factor model separating brand- and category-level
random effects to quantify a brand's *fit* and *atypicality*, the levers of
extension success. **Brand coolness** [@warren2014; @warren2019] is a distinct,
dynamic, socially constructed positive trait attributed to *appropriately
autonomous* objects: autonomy (diverging from norms) raises coolness only when the
divergence is appropriate, which depends on whether the violated norm is descriptive
or injunctive, the norm's perceived legitimacy, and how much the audience values
autonomy. Cool brands are first niche (subcultural, rebellious, authentic,
original) and acquire iconic and popular attributes only as the mass market adopts
them.

Association networks also operate below deliberate awareness, which complicates every
self-report measure in this section. @chartrand2008automaticbrand show that
subliminal brand exposure activates the goal a brand represents — Apple raising
creativity, a discount retailer raising thrift — and changes behavior without
recognition of the prime. The individual difference that governs how much brands do
this work at all is captured by @sprott2009importancgeneral, whose
brand-engagement-in-self-concept measure predicts which consumers incorporate brands
into identity in the first place, and therefore for whom the personality and
self-congruence findings above should be expected to hold. Which relational role the
brand is cast in matters too: @kramer2015materialiprefer find that materialistic
consumers prefer a brand positioned as servant rather than as partner, reversing the
usual prescription. And identity threat activates brand use in predictable ways —
socially excluded consumers shift preference toward brands that signal belonging or
distinctiveness depending on how they construe the exclusion [@wan2016socialexclusion].
Product placement is where these implicit mechanisms are deliberately engineered, and
@russell2002investigaeffective shows the effect depends on the fit between the
placement's modality and its plot connection rather than on exposure alone.

## Brand Authenticity

Whether **authenticity** is reflective or formative is, as with equity, unresolved,
and the two leading scales take opposite stances. @morhart2015 propose a
*reflective* perceived-brand-authenticity (PBA) scale—the extent to which consumers
perceive a brand as faithful to itself and its consumers and as supporting
consumers' own authenticity—built from four reflective facets (continuity,
integrity, credibility, symbolism) and arising from the interplay of objective
facts (indexical authenticity), subjective associations (iconic authenticity), and
existential motives (existential authenticity); its drivers predict emotional
attachment and positive word of mouth. @nunes2021 instead propose a *formative*
account in which authenticity is composed of six component judgments—accuracy,
connectedness, integrity, legitimacy, originality, and proficiency—whose weights
shift with consumption context. The unresolved measurement model is not academic
bookkeeping: it determines whether a brand should manage authenticity by
strengthening a coherent latent disposition (reflective) or by intervening on
specific, possibly uncorrelated, component judgments (formative).

Underneath both scales sits a lay theory the measurement literature tends to skip
past. @newman2014authenticcontagiou show that consumers reason about authenticity as
if brand essence were physically transmissible: an object touched by the original
creator is valued more than a physically identical copy, and the premium behaves like
contagion rather than like an inference about quality. That intuition is what makes
retro branding work — a revived brand trades on a continuity consumers believe in
rather than one the firm can document [@sherry2003teachingbrands] — and it is also
why emotional branding is unstable: the same narrative that grants a brand cultural
authority invites consumers to police it, so the brand becomes accountable to a story
it no longer controls [@thompson2006emotionalbranding]. The counterfeit market is the
clean test of what authenticity is actually doing: @wilcox2009counterfeluxury find
that consumers buy counterfeits when the brand serves a social-adjustive function and
not when it serves a value-expressive one, which means the fake substitutes for the
signal but not for the meaning.

## Brand Relationships

Consumers form relationships with brands that are valid at the level of lived
experience [@fournier1998]. Relationships require *interdependence*—partners affect
and redefine the relationship—and brands are animated through possession by a
significant other, anthropomorphization, or acting as an active partner.
@fournier1998 operationalizes **brand relationship quality** (BRQ) along six
facets: love/passion, self-connection, commitment, interdependence, intimacy, and
brand-partner quality. BRQ is richer than loyalty, with which it overlaps, and
mechanistically connected to brand personality.

Relationship *norms* govern evaluation. @aggarwal2004 distinguishes **exchange**
relationships (quid pro quo: expected, prompt repayment) from **communal**
relationships (benefits given to show concern, without expected return); when
consumers relate to a brand communally, applying exchange norms violates
expectations and damages evaluations. People judge brands as they judge people—on
**warmth** and **competence** [@aaker2010]: for-profits read as competent but cold,
nonprofits as warm but less competent, and credibility cues that raise perceived
competence recover willingness to buy from nonprofits. Anthropomorphism has a dark
side: humanized brands suffer *larger* evaluation losses under negative publicity
because consumers attribute misbehavior to stable traits, with the penalty
concentrated among entity theorists (who believe traits are fixed), for whom
compensation—not apology or denial—is the only effective response [@puzakova2013;
@macinnis2017]. Self-construal moderates which connection matters: self-concept
connection dominates under independent self-construal, country-of-origin connection
under interdependent self-construal [@swaminathan2007], and avoidance/anxiety
attachment styles predict preferences for exciting versus sincere brands
[@swaminathan2009].

### Brand Love

**Brand love** sits at the level of brand equity and subsumes brand affect.
@batra2012 establish it as a higher-order reflective construct distinct from the
transient *emotion* of love—long-lasting and spanning affective, cognitive, and
behavioral experiences. @bagozzi2016 distill a parsimonious scale with three
reflective higher-order factors—self–brand integration, positive emotional
connection, and passion-driven behavior—validated against method bias with a
multitrait–multimethod design.

## Reputation and Evaluation

**Reputation** is a global evaluation of an organization accumulated over time
[@fombrun1990; @aaker2010], carrying both competence and warmth dimensions. It is
broader than brand equity in one telling sense: equity, a marketing construct,
typically denotes the *positive* part of the firm, whereas reputation, a management
construct, spans both positive and negative. Reputation is actively managed in
review platforms. @proserpio2017 show hotels that begin responding to reviews gain
about 0.12 stars on average, but receive *fewer, longer* negative reviews
thereafter—dissatisfied customers self-censor short, unjustified complaints when
they anticipate scrutiny—so managers face a trade-off between the quantity and the
depth of negative feedback. Because responding is a non-random managerial choice,
the paper is a useful template for identification under endogenous treatment.
Evaluation is also socially contingent: the mere revealed presence of virtual
supporters shifts a target consumer's brand evaluation and purchase intention,
moderated by supporter composition and competitor salience [@naylor2012], and
social-media-based favorability measures—debiased for poster selection via a
probabilistic graphical model—track traditional survey measures [@zhang2021].

## Brand Architecture and Portfolio

A firm's **brand architecture** ranges from a *branded house* (one master brand
across products) to a *house of brands* (independent brands). The branded house
yields efficiency and lower customization and cannibalization but concentrates risk;
@Rao_2004 find corporate branding (branded house) associated with higher Tobin's
$q$ and mixed strategies with lower values. @hsu2015 extend this to four
strategies—house of brands, branded house, **sub-branding** (e.g., Intel
Pentium: master plus product brand, high risk/high return), and **endorsed
branding** (a graphically subordinate parent, reduced risk)—plus hybrids, and
decompose idiosyncratic risk into reputation, dilution, cannibalization, and
stretch components: sub-branding creates the most firm value but carries the most
risk. Portfolio profitability can rise from pruning lower-tier variants, moderated
by the upper- to lower-tier ratio [@aribarg2008], and portfolios are characterized
by the number of brands and segments, internal competition, and perceived
quality/price [@morgan2009]. @fig-brand-architecture arrays the four strategies
along the architecture spectrum.

```{mermaid}
%%| label: fig-brand-architecture
%%| fig-cap: "The brand-architecture spectrum. Strategies range from a branded house (one master brand across products) through sub-branding (master plus product brand) and endorsed branding (a graphically subordinate parent) to a house of brands (independent brands), with hybrids in between; sub-branding creates the most firm value but carries the most risk, while endorsement reduces risk."
flowchart LR
  BH["Branded house<br/>one master brand<br/>across products<br/>efficiency, concentrated risk"]
  SB["Sub-branding<br/>master plus product brand<br/>highest value, highest risk"]
  EB["Endorsed branding<br/>graphically subordinate<br/>parent, reduced risk"]
  HB["House of brands<br/>independent brands"]
  BH --- SB
  SB --- EB
  EB --- HB
```

### Co-branding and Ingredient Branding

In **brand alliances**, partners' reputations spill over onto one another
[@rao1999; @simonin1998]. Stock markets react to co-branded product introductions
as signals of innovativeness and quality, rewarding consistency between partners and
exclusive partnerships [@cao2013], with the brand-value effect moderated by the
value differential between partners and prior exploitation of the target brand
[@cao2017]. Alliance reputations are fragile: preventable, controllable, purposeful
crises damage the culpable ally, and even non-culpable partners suffer indirectly
through negative post-crisis sentiment toward the partnership [@singh2020a]. In
**ingredient branding**, co-branded trial spills over behaviorally onto both host
and ingredient brands, more so among non-loyal past consumers and when perceived
fit is high [@swaminathan2011]. Followership data can surface alliance
opportunities: @malhotra2022 introduce **brand transcendence**—the overlap between
a brand's followers and those of brands in a new category—from co-followership
patterns on social media.


@malhotra2026alliances build that observation into a full discovery pipeline, which
is a different kind of contribution from the effect estimates above. The managerial
problem they target is combinatoric: the number of candidate partnerships in a market
is quadratic in the number of brands, so the binding constraint on alliance strategy
is *screening*, not evaluation. Their framework—**brand alliance network exploitation**
(BANE)—combines offline brand attributes with online co-mention signals from digital
user traces, learns network embeddings over the resulting information network, and
predicts which pairs will ally.

Three features make it more than a ranking model. It is validated on a *temporal
holdout*, so the test is whether pre-announcement signals predict alliances that had
not yet happened, rather than whether the embedding reconstructs alliances it was
fitted on. It is validated against *people*: consumer and brand-manager surveys
confirm that BANE-identified opportunities are judged more promising than controls,
which is the step that separates a plausible-looking network artifact from a
commercially meaningful one. And it treats the predicted-but-not-realized cases as
the output of interest rather than as prediction error—the BANE matrix qualifies
those, and the framework then generates marketing-mix execution guidance for them.
The last move is the one most worth imitating: a model built to predict an observed
behavior is repurposed so that its *false positives*, properly qualified, become the
recommendation set.

### Acquisitions and Disposals

Brand-asset transactions are valued by markets through an **event study**. For firm
$i$ on day $t$, the abnormal return is the realized return minus its expectation
from a market model estimated over a pre-event window,
$$ \mathrm{AR}_{it} = R_{it} - \mathbb{E}[R_{it} \mid \text{market model}], \qquad
   \mathrm{CAR}_i = \sum_{t \in W} \mathrm{AR}_{it}, $$
and the cumulative abnormal return $\mathrm{CAR}_i$ over a short window $W$ is
regressed on transaction characteristics, after screening confounds (earnings
announcements, splits, executive changes, buybacks, dividend changes) within the
window [@mcwilliams1997a; @srinivasan2004event]. @wiles2012 apply this to brand
acquisitions and disposals and find effects that are *not* symmetric: abnormal
returns depend on the acquirer's marketing capability, its channel relationships,
and the relative price/quality positioning of the brands involved—investors reward
acquirers who realize cost synergies but penalize those touting revenue synergies.
Marketing capability is itself estimated as a **stochastic-frontier** efficiency
score: with intangible value (Tobin's $q$, residualized for technology and
management quality) as output and marketing resources (advertising, SG&A,
trademarks) as inputs, capability is the firm's distance from the efficient frontier
[@dutta1999]. Complementary findings show acquired-brand value rising in the
target's marketing capability and portfolio diversification [@bahadir2008], and that
negative reactions to acquisitions trace to a perceived loss of the brand's unique
values, conditional on brand age, leadership continuity, and value alignment
[@biraglia2022express].

## Brand Extensions

A **brand extension** attaches an established name to a product in a new category.
The classic determinant of success is perceived **fit** between parent and
extension [@aaker1990], though fit can be overridden by key brand associations that
themselves create the basis of fit [@broniarczyk1994], and fit's influence fades
once consumers receive richer attribute information about the extension
[@klink2001threats].

A choice-model formalization clarifies the mechanism. Let a consumer's utility for a
candidate extension be
$$ U = \beta_{\text{brand}} + \boldsymbol{\beta}_{\text{attr}}^{\top}\mathbf{x}
        + \beta_{\text{int}}\,(\text{brand} \times \mathbf{x}) + \epsilon, $$
where the interaction term captures brand–attribute synergy in the parent category.
@rangaswamy1993brand show that a brand deriving utility largely from such an
interaction is *less* extendable than an equally preferred brand whose utility is
free-standing; extendibility is therefore maximized by investing in
non-product-specific values (quality, style, durability, reputation). Extensions also
feed back on the parent. @swaminathan2001 model parent-brand choice with a **binary
logit**—chosen over the multinomial form because it isolates the incremental loyalty
coefficient—and find positive reciprocal effects of extension trial on parent choice
(strongest for prior non-users) and evidence of negative spillover from extension
failure (strongest for loyal users); experience with the parent drives extension
*trial* but not *repeat*. Strong brands command extension **price premiums** that
rise with fit and fall with the extension's financial and social risk
[@delvecchio2005brand], and quality reach is asymmetric: high-quality brands extend
further than low-quality brands [@heath2011]. For context-dependent consumers,
benefit-based information can rescue *low-fit* extensions while attribute-based
information depresses them; context-independent consumers judge on fit regardless
[@mathur2022].

## Strategic Branding Decisions

### Brand Crisis and Recovery

Brand crises—product-harm events, firestorms, negative word of mouth—spill across a
firm's segments and brands, with consequences sharpest within a country
[@borah2016]. Consumers interpret a firm's crisis response through their prior
expectations of the firm [@dawar2000], crises contaminate comparable brands in the
same category [@dahlén2006], and effects in foreign markets follow an inverted-U in
psychological distance that strong marketing capabilities attenuate [@dinner2018].
On social media, response *style* matters: active listening and empathy evoke
gratitude in high-arousal customers even before the failure is resolved, an effect
@herhausen2022 identify across 472,995 negative posts in 89 S&P 500 brand
communities, instrumenting response with text-derived arousal, tie strength, and
linguistic-style matching while controlling for firm-, community-, and post-level
covariates.

### Global Brand Strategy

Perceived **brand globalness** raises purchase likelihood through perceived quality
and prestige [@emsteenkamp2002; @davvetas2015], moderated by consumer ethnocentrism
[@shimp1987]; a more globally recognized brand also gains in trust, affect, and
association with global citizenship [@steenkamp2019; @torelli2012]. An alternative
route is to become an icon of the *local* culture: @xie2015a formalize **perceived
brand localness** and a **brand identity expressiveness** construct—the brand's
capacity to construct and signal self- and social identity—through which the quality
and prestige paths of the globalness model are mediated to null. In developing
markets, nonlocal-origin brands are preferred for the social status they confer,
especially among the status-susceptible and for socially visible, unfamiliar
categories [@batra2000]. Preferences are also path-dependent: consumers carry local
brand preferences to new locations, so early share advantages persist for decades
[@bronnenberg2009; @bronnenberg2010].

### Competitive Dynamics

Competitive context reshapes brand response. Publicly **complimenting a competitor**
("brand-to-brand praise") can raise sales and reputation by signaling warmth, with
the largest gains for skeptical audiences and for-profit complimenters when the
praise is authentic [@zhou2021]. Consumer support for small brands rises under
competitive threat from large brands [@paharia2014], and advertising's primary
function is often category expansion rather than share stealing, with the
complementary (category-building) function stronger in larger markets
[@dubé2005]—a theme developed in @sec-advertising. Positioning also determines who
captures a promotion's spillover: @wathieu1996asymmetripromotion show that promotion
effects run asymmetrically across the quality tier, so an identical discount transfers
share very differently depending on which side of the positioning gap the promoting
brand sits — the brand-level counterpart to the price-tier asymmetry in
@sec-pricing.

## Brand–Consumer Interaction

**Brand experience**—sensory, affective, cognitive, and behavioral responses
evoked by brand stimuli—predicts purchase and loyalty [@brakus2009], with online
experience driven by credibility and utility [@morgan-thomas2013] and the relative
value of experiential versus functional communication differing across developed
and developing markets [@zarantonello2013]. **Co-creation** is double-edged: it
boosts empowerment and brand identity in general [@fuchs2010] but can fail to lift
liking in luxury categories [@fuchs2013] and among high-power-distance, conservative
consumers [@paharia2019]. **Brand communities** sustain engagement through
oppositional loyalty—devotion to one brand defined partly against another
[@muniz2001]—and can take on quasi-religious intensity [@muñizjr.2005].

What a community actually *does* for the firm has been specified more precisely
since. @arnould2009brandcommunity decompose community into twelve value-creating
practices—welcoming, staking, badging, documenting, and the rest—which reframes the
community from a sentiment to a set of repeatable behaviors a firm can support or
disrupt. The effects are not uniformly benign: @herrmann2005socialinfluence find in a
European car-club panel that community integration raises loyalty to the brand while
simultaneously raising members' *resistance to the firm's own marketing*, so the same
mechanism that locks customers in reduces the firm's ability to steer them.
@arsel2004starbucksbrandscap traces the sharper version of this in the Starbucks
case, where the brandscape itself became material for consumers constructing
anticorporate identities. Consumers can also resist for reasons unrelated to
community: @lam2010resistancbrand shows switching resistance persisting even when a
radically new brand is objectively superior, because the incumbent brand is
load-bearing for identity rather than merely preferred.

The social-media layer adds a distinct set of regularities. Brand-related content
produced by consumers differs systematically from firm-produced content in what it
emphasizes and how it is received [@fischer2012brandrelated]; the popularity of a
brand post on a fan page is predictable from its characteristics rather than from the
brand's stature [@gensler2012popularitbrand]; and consumer-to-consumer communication
inside a branded online community shifts brand outcomes independently of anything the
firm says [@noble2009influencecommunica]. @liuthompkins2013managingbrands is the
synthesis of what this environment changes for brand management, and the essential
point is a loss of control rather than a gain in reach.

## Brand Valuation

Brand value travels under many names: *brand equity*, *brand value*, and *brand
valuation* in marketing; *brand capital* and *intangible capital* in
finance and accounting. Firm value derives from tangible assets and intangibles, the
latter including knowledge capital and brand capital. The valuation question is how
much of firm value the brand explains, and firms capture only part of the potential:
@fischer2024accessing estimate that firms realize on average just **35%** of their
financial brand-leverage potential, a "brand capabilities gap" of roughly
**\$2.1 billion per brand** (≈4.3% of firm value), closeable through identifiable
market-strategy levers.

Valuation methods divide into **output-based** measures (which read value from
realized outcomes and are forward-looking) and **input-based** measures (which
accumulate spending and are backward-looking). The evidence favors output-based
measures: an equal-weighted portfolio of top brands earns ≈3% annual abnormal
returns—concentrated in firms that build brands *internally*—while the traditional
input measure (advertising expense) earns none, and analysts systematically
underreact to brand strength, leaving excess returns after earnings announcements
[@boustanifar2025brand]. Trademark registrations tell the same story: they predict
profitability and returns and are undervalued by investors, identified using the
Federal Trademark Dilution Act as an exogenous shock to trademark protection
[@hsu2022valuation].

### Output-Based: The InterBrand Discounted-Earnings Method

The InterBrand approach values a brand in three steps: compute **economic profit**,
isolate the share attributable to the brand via a **Role of Brand Index** (RBI), and
discount projected **brand earnings** by a rate reflecting **brand strength**.
Economic profit is after-tax operating profit net of a capital charge,
$$ \text{Economic Profit}_t = \text{After-Tax Operating Profit}_t - \text{Cost of Capital}_t, $$
brand earnings apply the RBI,
$$ \text{Brand Earnings}_t = \text{Economic Profit}_t \times \text{RBI}, $$
and brand value is the present value of projected brand earnings plus a terminal
value,
$$ \text{Brand Value} = \sum_{t=1}^{T} \frac{\text{Brand Earnings}_t}{(1+r)^{t}}
   + \frac{\text{Terminal Value}}{(1+r)^{T}}, \qquad
   \text{Terminal Value} = \frac{\text{Brand Earnings}_T}{r - g}, $$
where $r$ is the discount rate and $g$ the perpetual growth rate (the Gordon-growth
terminal value, valid only for $g < r$). The following reproducible example projects
five years of history forward ten years.

```{r interbrand-valuation}
# Historical data (years 1-5)
after_tax_profit <- c(500000, 520000, 540000, 560000, 580000)
cost_of_capital  <- c(100000, 105000, 110000, 115000, 120000)
RBI              <- 0.7    # Role of Brand Index
discount_rate    <- 0.08   # annual discount rate r
future_years     <- 6:15   # 10-year projection horizon

# Step 1: historical economic profit and its average growth rate
economic_profit <- after_tax_profit - cost_of_capital
growth_rate     <- mean(diff(economic_profit) / head(economic_profit, -1))

# Step 2: project economic profit and brand earnings
projected_economic_profit <- economic_profit[length(economic_profit)] *
  (1 + growth_rate)^(seq_along(future_years))
projected_brand_earnings  <- projected_economic_profit * RBI

# Step 3: discount to present value (as of year 6)
discounted_future_earnings <- projected_brand_earnings /
  (1 + discount_rate)^(seq_along(future_years))

# Terminal value (Gordon growth, g < r)
terminal_growth_rate     <- 0.02
terminal_value           <- projected_brand_earnings[length(projected_brand_earnings)] /
  (discount_rate - terminal_growth_rate)
discounted_terminal_value <- terminal_value /
  (1 + discount_rate)^length(future_years)

total_brand_value <- sum(discounted_future_earnings) + discounted_terminal_value

cat("Average growth rate:        ", round(growth_rate, 4), "\n")
cat("Total brand value (year 6): $", format(round(total_brand_value), big.mark = ","), "\n")
```

### Output-Based: A Text Measure from SEC Filings

@boustanifar2025brand construct a text-based measure as the frequency of the words
*brand(s)* in a filing, normalized by document length. The measure is forward-looking
and covers firms that disclose no advertising expense (e.g., Tesla, Uber,
Salesforce).

```{r brand-text-measure}
# Base-R implementation (no external NLP dependencies) of the brand-intensity
# measure: count of "brand"/"brands" tokens divided by total word count.
brand_text_measure <- function(texts) {
  clean <- gsub("[^a-z ]", " ", tolower(texts))          # keep letters and spaces
  total_words <- vapply(strsplit(trimws(clean), "\\s+"),
                        function(w) sum(nzchar(w)), numeric(1))
  brand_count <- vapply(gregexpr("\\bbrands?\\b", clean),
                        function(m) if (m[1] == -1L) 0 else length(m), numeric(1))
  data.frame(
    filing_id       = seq_along(texts),
    brand_count     = brand_count,
    total_words     = total_words,
    brand_intensity = ifelse(total_words > 0, brand_count / total_words, 0)
  )
}

sec_filings <- c(
  "Our brand is our most valuable asset; we invest in the brand every year.",
  "This filing discusses logistics and supply chains with no brand emphasis.",
  "Brands, brand equity, and brand loyalty drive our pricing power."
)
brand_text_measure(sec_filings)
```

### Input-Based: The Perpetual-Inventory Method

The input view treats brand capital as a stock accumulated from advertising and
depreciated each period. With depreciation rate $\delta$ and advertising $A_t$,
$$ K_t = (1-\delta)\,K_{t-1} + A_t, \qquad K_0 = \frac{A_0}{g + \delta}, $$
where the initial stock $K_0$ assumes a balanced-growth path with advertising growth
$g$ [@vitorino2014understanding; @belo2014brand]. The method is standard but
limited: roughly 70% of firms—and many top brands—do not report advertising
expense, the input reflects spending rather than its quality or effectiveness, it is
backward-looking, and some valuable brands advertise minimally [@doyle1990building].
In the cross-section, firms with *lower* brand-capital investment rates and *higher*
brand-capital intensity earn higher average returns (≈5% per year), consistent with
a neoclassical investment model treating brand capital as a production factor subject
to adjustment costs [@belo2014brand].

```{r perpetual-inventory}
initial_advertising <- 1000   # A_0
growth_rate         <- 0.10   # g
depreciation_rate   <- 0.20   # delta
num_years           <- 10

initial_stock <- initial_advertising / (growth_rate + depreciation_rate)

advertising <- initial_advertising * (1 + growth_rate)^(0:(num_years - 1))
brand_capital <- numeric(num_years)
brand_capital[1] <- initial_stock
for (t in 2:num_years) {
  brand_capital[t] <- (1 - depreciation_rate) * brand_capital[t - 1] + advertising[t]
}

results <- data.frame(Year = 1:num_years,
                      Advertising = round(advertising),
                      Brand_Capital = round(brand_capital))
results

plot(results$Year, results$Brand_Capital, type = "o",
     xlab = "Year", ylab = "Value", main = "Perpetual Inventory Method",
     ylim = range(c(results$Brand_Capital, results$Advertising)))
lines(results$Year, results$Advertising, type = "o", col = "blue")
legend("topleft", legend = c("Brand Capital", "Advertising"),
       col = c("black", "blue"), lty = 1, bty = "n")
```

## Key Takeaways

- **Brand elements are now scorable before launch.** Slogan memorability splits into two
  outcomes—recognition and cued brand recall—and models built from literature-derived
  factors plus text embeddings predict both for candidate slogans, including for brands
  with no exposure history [@aka2026slogans]. The embedding's margin over the
  theory-derived features doubles as an estimate of what the slogan literature has not
  yet named.
- Brand equity is the demand advantage attributable to the *name*—formally the price
  premium in @eq-brand-premium—and is distinct from value added.
- Whether equity (and authenticity) is **reflective** or **formative** is unsettled
  and dictates measurement and management; do not assume one without argument
  (@sec-measurement-scales).
- Many brand phenomena—prominence, specialization, authenticity—are **signaling**
  problems whose logic is that credible signals are differentially costly across
  types.
- Brand actions are priced by markets; event studies and stochastic-frontier
  capability estimates connect branding to firm value
  (@sec-marketing-finance).
- **Output-based** valuations dominate **input-based** ones empirically; advertising
  expense is a weak proxy for brand strength.
