# Advertising {#sec-advertising}

*Advertising* is paid, mediated communication from an identified sponsor designed to
shift what consumers know, feel, and ultimately do. It is among the largest
discretionary line items a firm controls and one of the oldest objects of study in
marketing, yet two facts about it remain in productive tension. The first is that
advertising plainly *works*: a long tradition of meta-analysis and market-response
modeling documents that it moves consumer mindsets, behavior, and—through those
channels—firm value [@wang2008; @hewett2016]. The second is that, measured honestly,
its average effect is small, noisy, and frequently indistinguishable from zero. The
most rigorous modern field evidence finds that the bulk of brands earn statistically
insignificant or negative advertising elasticities, that observational estimates are
swamped by selection, and that a majority of observed advertising schedules carry
negative marginal returns [@Shapiro_2018; @shapiro2021tv; @Gordon_2017; @Lewis_2015].
Reconciling "advertising works" with "most advertising doesn't pay" is the central
problem of the field, and it is fundamentally a problem of *measurement*.

This chapter is organized around three lenses that the literature applies to that
problem, and the reader who finishes it should be able to move fluently among them.
The **behavioral** lens asks how an exposure is processed inside a single mind: which
psychological stages a message must clear to change a belief, when affect substitutes
for cognition, and how involvement moderates the route to persuasion. The
**theoretical** lens asks how rational agents—consumers, advertisers, and
intermediaries—interact when advertising is informative, intrusive, and costly. The
**econometric** lens asks the hardest question of all: given non-experimental data in
which firms advertise *because* they expect demand, how does one recover the causal
effect of advertising rather than the spurious correlation between spending and the
sales that motivated it? We treat the behavioral models as the source of testable
mechanisms, the theoretical models as the source of equilibrium discipline, and the
econometric models as the arbiter of what is actually true. Throughout, the unifying
thread is that advertising's effect is best understood not as a level but as an
**increment above a baseline**—and that the entire empirical enterprise turns on how
credibly that baseline is constructed [@Gordon_2019].

A note on scope and ethics frames everything that follows. Audiences dislike
intrusive advertising, especially when they are vulnerable [@kellogg_2005], and the
welfare consequences of targeting, deception, and repetition recur as substantive
questions rather than asides. Advertising that persuades by exploiting vulnerability
is both a behavioral phenomenon and a regulatory one, and the chapter treats it as
both.

## The Behavioral Approach

The behavioral tradition opens the consumer's head and asks what an advertisement has
to accomplish there. Its oldest and most durable idea is that persuasion is a
*sequence of conditional stages*, each of which can fail.

### Persuasion as a Stochastic Information-Processing Chain

@mcguire1983 models the effect of a persuasive communication as a chain of
information-processing steps, each of which the receiver must clear before the next
becomes possible. Because the steps are sequential and each is probabilistic, the
probability that an exposure ultimately changes behavior is the *product* of the
stage-wise conditional probabilities. Writing the stages as presentation (the message
reaches the receiver), attention, comprehension, yielding (attitude change),
retention (persistence of the new belief), and behavior, the persuasion probability is

$$
P(\text{behavior}) =
P(\text{presentation}) \cdot
P(\text{attention} \mid \text{presentation}) \cdot
P(\text{comprehension} \mid \text{attention}) \cdot
P(\text{yielding} \mid \text{comprehension}) \cdot
P(\text{retention} \mid \text{yielding}) \cdot
P(\text{behavior} \mid \text{retention}).
$$ {#eq-mcguire-chain}

The multiplicative form in @eq-mcguire-chain is the source of the model's most useful
implication: because the stages compound, a campaign is only as strong as its weakest
link, and very high performance on one stage cannot rescue a near-zero probability on
another. Each stage maps onto a distinct managerial lever and a distinct measurement
instrument—presentation to reach and media weight, attention to recognition tests,
comprehension to recall and semantic-differential profiles, yielding to measured
attitude change, retention to delayed attitude measurement, and behavior to observed
choice. The independent variables that move these probabilities are organized in
@mcguire1983 as a communication matrix crossing *source*, *message*, *channel*,
*receiver*, and *destination*. Advertising appeals operate on the yielding stage
through valence: positive (gain-framed) appeals reduce anxiety, negative
(loss-framed) appeals raise it. The model's chief limitation—and the seam along which
later work pulled it apart—is its assumption of a *single fixed hierarchy* in which
cognition always precedes affect which always precedes behavior.

::: {.callout-note}
The "hierarchy of effects" tradition that @eq-mcguire-chain formalizes traces to
practitioner think-/feel-/do- models of the 1960s. Its scientific value is not the
specific ordering, which is often wrong, but the discipline of decomposing a single
aggregate effect into separately measurable conditional stages. Most modern
diagnostics—funnels, brand-lift studies, attention metrics—are descendants of this
decomposition.
:::

### From Funnel to Journey

The managerial counterpart to the processing chain is the **purchase funnel**:
awareness, familiarity, consideration, purchase, and loyalty (with loyalty split into
active and passive forms). The funnel is a *firm's-eye* abstraction—a narrowing
pipeline the marketer pushes prospects down. The influential reframing associated with
McKinsey's *consumer decision journey* replaces the narrowing funnel with a cyclical
process organized around the consumer's perspective: an initial consideration set, a
phase of active evaluation in which the set can *expand*, a moment of purchase, and a
post-purchase experience that feeds loyalty and future consideration. The two views
are complementary rather than competing, as @fig-funnel-journey contrasts; a useful
way to hold them together is that the journey adds a dimension the funnel suppresses,
turning a one-directional narrowing into a recurring loop. @batra2016 push this
further into a *dynamic,
expanded* customer decision journey that integrates paid, owned, and earned media
across traditional and digital touchpoints and asks how communications interact
across them.

```{mermaid}
%%| label: fig-funnel-journey
%%| fig-cap: "Two complementary representations of how advertising acts on demand. The firm-centric funnel (left) is a narrowing pipeline; the consumer decision journey (right) is a recurring loop in which the consideration set can expand during active evaluation and post-purchase experience feeds back into future consideration."
flowchart TB
  subgraph Funnel["Purchase Funnel (firm's view)"]
    direction TB
    A1[Awareness] --> A2[Familiarity] --> A3[Consideration] --> A4[Purchase] --> A5[Loyalty]
  end
  subgraph Journey["Consumer Decision Journey (consumer's view)"]
    direction TB
    B1[Initial consideration set] --> B2[Active evaluation]
    B2 --> B3[Moment of purchase]
    B3 --> B4[Post-purchase experience]
    B4 -->|loyalty loop| B1
    B2 -->|set expands| B1
  end
```

### Antecedents, Processing, and Consequences

The decisive theoretical advance over the fixed hierarchy is to treat advertising
processing as a *contingent* system whose route depends on the consumer's needs,
motivation, ability, and opportunity. @maclnnis1989 supply the canonical synthesis,
organizing the field's prior strands—the hierarchy-of-effects models, the
multiattribute attitude and cognitive-response models of the 1970s—into a single
antecedents → processing → consequences architecture.

On the antecedent side, needs are partitioned into **utilitarian** needs (functional
problem-solving) and **expressive** needs, the latter splitting into *socially
expressive* needs (to project or reflect a self-image) and *experiential* needs (to
satisfy sensory or cognitive appetites). The pivotal antecedent is **motivation to
process brand information in the ad**—what the consumer-behavior literature calls
*involvement*—which may be situational (tied to the moment) or enduring (tied to the
person). Motivation governs **attention** and **processing capacity**: higher
motivation buys greater attention and greater capacity for analyzing the ad, but
capacity is finite, so capacity spent on the ad is unavailable for competing tasks.
Utilitarian needs direct attention to brand attributes; expressive needs direct it to
symbolic and experiential value. Processing then proceeds through escalating levels—
feature analysis, basic categorization, meaning analysis, information integration,
role-taking, and constructive processing—moderated by **ability** and **opportunity**.
The consequences are cognitive responses (thoughts) and emotional responses
(feelings), and these can be partitioned by referent into *message-related*
(brand-relevant), *execution-related* (brand-irrelevant), and *viewing-context-related*
(environmental) responses. The architecture's value is integrative: it subsumes the
elaboration-likelihood model of persuasion [@petty1986], Mitchell's brand-processing
model, and Lutz's typology of ad responses as special cases. A central testable claim
is that attitude toward the ad ($A_{\text{Ad}}$) mediates attitude toward the brand
($A_{\text{B}}$) specifically at the *higher* processing levels—meaning analysis,
information integration, and role-taking—implying a managerial trade-off between
investing an ad's finite real estate in brand/message content versus execution.

The complementary review by @vakratsas1999 abstracts across this entire literature
with a single reduced-form schematic—advertising input passes through *filters* (e.g.,
motivation, ability) into the consumer's cognition, affect, and experience, which in
turn drive behavior—and then taxonomizes the field's models by which of these
mediators they admit and in what order. @tbl-adv-hierarchies organizes the principal
families.

| Family | Mediating sequence | Core mechanism | Representative theory |
|---|---|---|---|
| Cognitive (C) | Cognition only | Information changes beliefs; price vs. non-price ads move price sensitivity asymmetrically | Information-economics models |
| Pure affect (A) | Affect only | Mere exposure; wear-in then wear-out yields an inverted-U | Two-factor / optimal-arousal theories |
| Persuasive hierarchy (CA) | Cognition → affect → behavior | Elaboration drives attitude; involvement moderates | ELM [@petty1986]; @maclnnis1989 |
| Low-involvement (CEA) | Cognition → experience → affect | Behavior precedes firm attitudes | Low-involvement learning |
| Integrative (CAE) | Cognition → affect → experience | Expectations then trial confirm/disconfirm | Two-stage trial models |
| Hierarchy-free (NH) | No fixed order | Route is contingent on context | @vakratsas1999 |

: Families of advertising-response models, classified by which mediators they admit and in what order. Adapted from the synthesis in @vakratsas1999. {#tbl-adv-hierarchies}

Two empirical generalizations survive this taxonomy and recur below. First,
experience, affect, and cognition all serve as mediators of advertising effects, so a
model that omits any one is misspecified for some product class. Second, **short-term
advertising elasticities are small and decline over the product life cycle**—a
behavioral fact that anticipates the econometric findings of @sec-econometric-approach.
At the aggregate level, the venerable finding that roughly three exposures per
purchase cycle is "optimal" reflects the same affective dynamics of wear-in and
wear-out [@vakratsas1999]. The broader program of cataloguing such regularities is
synthesized by @wind2009, who compile a set of empirical generalizations about
advertising while flagging the field's persistent gaps in boundary conditions,
advertising properties, and measurement.

### Cognitive and Affective Mediators

The cognitive route formalizes persuasion as the algebra of the thoughts an ad
provokes. @wright1973 distinguishes three modes of spontaneous cognitive response—
*support arguments* that reinforce the message, *counterarguments* that resist it, and
*source derogations* that attack the communicator—and models acceptance as their
weighted net balance,

$$
\text{Acceptance} =
w_{\text{SA}} \sum_{i} \text{SA}_i
- w_{\text{CA}} \sum_{j} \text{CA}_j
- w_{\text{SD}} \sum_{k} \text{SD}_k,
$$ {#eq-cognitive-response}

where $\text{SA}$, $\text{CA}$, and $\text{SD}$ count support arguments,
counterarguments, and source derogations, and the weights $w$ scale their persuasive
impact. @eq-cognitive-response makes the cognitive-response prediction precise:
anything that suppresses counterargument production—distraction, low involvement, or
the message-relevant situational factors of content-processing involvement and
message modality emphasized by @wright1973—mechanically raises acceptance. This is the
engine behind @batra1986's analysis of *repetition*: repeated exposure raises brand
attitude and purchase intention while support and counterargument production are low,
but once such production rises, attitudes level off and then decline. Repetition's
benefit therefore wears out, and the wear-out can be pushed back by varying ad
execution; the appropriate dosing interval is the purchase cycle, with roughly three
exposures per cycle a common optimum. Ability, motivation, and opportunity are the
antecedents that govern when this cognitive processing engages at all.

The affective route insists that evaluation does not reduce to cognition. The
landmark statement is @zajonc1980's, which deserves quotation because it reorients the
entire downstream literature.

> Preferences need no inferences: affective reactions to stimuli can be acquired
> first, can occur without extensive perceptual and cognitive encoding, and can
> influence subsequent judgments independently of and prior to cognition.

On this view affect is a *system-1* response that can precede and bias the slower,
deliberate *system-2* cognition rather than following from it. The mechanisms are
classical (Pavlovian) conditioning—an automatic physiological reaction transferred to
a previously neutral stimulus—and **evaluative conditioning**, the direct transfer of
affect from one stimulus to another, which @gibson2008 shows (using implicit-measure
designs) reshapes *explicit* attitudes mainly when no strong prior preference exists.
A useful distinction within affect separates diffuse, long-lasting *moods* from
discrete, source-specific *emotions*. The empirical payoff of taking affect seriously
is large and well replicated: ad-evoked feelings raise brand attitudes both directly
and indirectly through attitude toward the ad, across involvement levels and product
types, with the effect strongest for hedonic rather than utilitarian products
[@pham2013]; emotional reactions mediate advertising's effect on attitudes toward both
ad and brand [@holbrook1987]; and attitude toward the ad itself shifts brand
attitudes, with execution cues and source likeability mattering most in low-involvement
settings [@batra1986a; @mackenzie1986; @mitchell1981]. Validated scales separate the
**hedonic** and **utilitarian** dimensions of consumer attitude that this literature
relies on [@batra1991; @voss2003].

Two further results sharpen the picture. Counterintuitively, momentary *interruptions*
can promote persuasion by amplifying arousal and the need for completion, an effect
concentrated among consumers low in need for cognition [@kupor2015]. And when creative
strategy is decomposed into composite content metrics, the components rank by their
effect on advertising elasticity in a clear order—**experiential** content first, then
**cognitive**, then **affective** [@dallolio2022]—a finding that gives the
cognition/affect/experience trichotomy of @tbl-adv-hierarchies direct managerial teeth.

A distinct memory mechanism complicates all of the above: ads do not arrive in
isolation. @burke1988competitive document **competitive interference**, in which one
ad degrades memory for another. Memory for a brand's ad is impaired *retroactively* by
later ads—both from competitors in the same class and from other products in the same
manufacturer's line—and *proactively* by prior ads, and the memory benefit of
repetition is itself blunted by the presence of competing advertising. Interference is
why elasticities estimated in clutter-free laboratories overstate field effects, a
theme @sec-econometric-approach returns to.

### Involvement and the Two Routes to Persuasion

The construct that organizes the behavioral literature's contingencies is
**involvement**—loosely, how deeply a consumer wants to think about a product. Two
frameworks dominate: the **elaboration-likelihood model** (ELM) [@petty1979] and the
**heuristic-systematic model** [@chaiken1980]. The ELM's core claim is that
persuasion travels one of two routes. The **central route** engages effortful
elaboration of message arguments; attitudes formed this way are durable and predictive
of behavior. The **peripheral route** relies on cues outside the argument—source
attractiveness, mere repetition, mood—and yields attitudes that are temporary and
weakly predictive. @petty1983 supply the canonical demonstration via a crossover
interaction: argument *quality* moves attitudes more under **high** involvement, while
a peripheral cue (a celebrity endorser versus an ordinary one) moves attitudes more
under **low** involvement. This single experiment is the field's cleanest evidence
that message content and executional cues are *substitutes* whose relative potency is
governed by involvement.

::: {.callout-note}
## A terminological caution
@petty1986 frame the central/peripheral distinction in terms of *motivation* rather
than *involvement*, and there is a methodological reason to follow them: the
involvement construct is notoriously fragmented across the literature, so framing
results in terms of motivation invites less definitional dispute. The marketing
literature further unbundles the ELM's "ability" into separate **ability** and
**opportunity** factors, yielding the motivation–ability–opportunity (MAO) triad as
the joint antecedent of elaboration [@petty1983]. For review and operationalization,
@muehling1993 is the preferable entry point, with @andrews1990 a complement and
@zaichkowsky1985a the canonical scale.
:::

The most rigorous theoretical treatment of involvement is @greenwald1984's, which
recovers the construct from a cognitive-capacity foundation rather than a managerial
one. They distinguish **attentional capacity** (the limited resource Kahneman calls
*effort*, demanded in increasing amounts as task complexity rises) from **attentional
arousal** (a state of wakefulness that facilitates well-learned responses), and array
processing into four levels of increasing depth: *preattention* (little capacity),
*focal attention* (modest capacity to decipher the message), *comprehension* (capacity
to analyze it), and *elaboration* (capacity to integrate it into existing knowledge).
Their **principle of higher-level dominance** states that when the effects of
different processing levels oppose one another, the highest engaged level dominates the
net result—deeper, more deliberate judgment wins. They also draw the conceptual line,
since muddied, between **involvement** (the audience/observer's depth of processing)
and what is better termed **engagement** (the actor/participant's behavior). That
distinction has acquired fresh empirical content online, where @schivinski2016, building
on @muntinga2011, scale consumer engagement along three escalating tiers—*consumption*
(viewing, reading), *contribution* (liking, sharing, commenting), and *creation*
(posting, producing content).

Involvement also moderates the affective route. @batra1990 show that positive mood
*reduces* cognitive elaboration and biases the perceived quality of arguments,
allowing mood to influence brand attitudes peripherally. @macinnis2002 add a maturity
condition: because affective processing requires no motivation while the persuasive
use of endorsers does, affectively based executional cues most reliably lift sales for
*mature, familiar* brands whose consumers already possess the relevant product
knowledge. A standing caution from the meta-analytic literature is that laboratory
advertising studies diverge from the field precisely on the dimensions that involvement
implicates: forced exposure, unmeasured choice, and the absence of competitive clutter,
decay, repeated exposure, and brand maturity.

### Visual Cues

Finally, the *visual* construal of an advertisement interacts with the brand's
positioning to determine willingness to pay. @chu2021 show that the effect of the
*visual distance* between the depicted product and the consumer flips with brand
image: for **prestige** brands associated with status and luxury, attitudes and
willingness to pay a premium *rise* as that distance grows (distance connotes
exclusivity), whereas for **popular** brands associated with broad appeal and social
connectedness, *closeness* is what lifts attitudes and willingness to pay. Visual
strategy is therefore not separable from positioning—the same image is an asset for one
brand archetype and a liability for the other.

## The Theoretical Approach {#sec-theoretical-approach}

The behavioral models describe a single mind; the theoretical approach asks what
happens when consumers, advertisers, and intermediaries are *strategic*. The exemplar
here is the economics of privacy and targeting. @choi2023consumer build an analytical
model with three classes of actor—consumers, an advertiser, and an ad network—in which
advertising is genuinely valuable (it conveys product information that guides the
purchase journey) but also costly to the consumer along two margins: a distaste for
online tracking and **ad wear-out** from repetition. Tracking lets the advertiser infer
a consumer's journey stage and target accordingly, so the consumer's opt-in decision
trades the benefit of more relevant information against the cost of wear-out.

The model's value is that it generates non-obvious comparative statics that no purely
behavioral account would predict. Consumers opt in to tracking when ad effectiveness is
*intermediate* or when their sensitivity to wear-out is low; and—counterintuitively—
opting in can *reduce* the number of repeat ads a consumer receives when effectiveness
is intermediate, because targeting substitutes relevance for repetition. Most
strikingly, *higher* ad effectiveness, though good for the advertiser per exposure, can
*lower* the ad network's profit by reducing the number of repeated ads it can sell.
These results carry directly into policy: by deriving *why* a rational consumer would or
would not consent to tracking, the model gives regulators contemplating consent
requirements a structural account of who opts in and what it does to surplus on all
sides of the market. The general lesson is that targeting and repetition are
equilibrium objects—choosing them well requires modeling the consumer's reaction, not
just the firm's.

## The Econometric Approach {#sec-econometric-approach}

The behavioral and theoretical lenses generate mechanisms; the econometric lens asks
which of them survive contact with real, non-experimental data. Its defining problem is
**endogeneity**. Firms do not assign advertising at random—they spend *because* they
forecast demand, target their best markets and best-timed weeks, and cut spending when
they expect to sell poorly anyway. Let $y_{it}$ be a sales outcome for unit $i$ in
period $t$ and $a_{it}$ advertising. The naïve regression

$$
y_{it} = \beta\, a_{it} + \mathbf{x}_{it}^{\top}\boldsymbol{\gamma} + \varepsilon_{it}
$$ {#eq-naive-adv}

recovers the causal advertising effect $\beta$ only if $\mathbb{E}[\varepsilon_{it}\,a_{it}]=0$.
That condition almost never holds: any demand shock the firm anticipates and responds
to (a holiday, a competitor's exit, favorable weather) enters $\varepsilon_{it}$ and is
correlated with $a_{it}$, biasing $\hat\beta$. Because firms target *favorable*
conditions, the bias is typically *positive*, so naïve estimates *overstate*
effectiveness—and the bias can be large relative to the true effect.

The benchmark that disciplines this literature is the randomized experiment, and where
experiments exist they are sobering. @Lewis_2015 and @Gordon_2017 demonstrate that
advertising effects are so small relative to the variance of sales that observational
methods routinely fail to recover them—producing estimates that may be too large or too
small by orders of magnitude—and that even large field experiments require enormous
samples to detect plausible effects. The correct conceptual target, following
@Gordon_2019, is the **incremental** effect: the lift in behavior *above the baseline*
the consumer would have exhibited absent the ad. When randomization is infeasible, the
practical remedy is to *construct* a credible baseline—for instance, panel data across
geographic markets that lets one estimate each market's no-advertising counterfactual
and read the advertising effect as the deviation from it. @tbl-adv-identification
summarizes the principal strategies and what breaks each.

| Strategy | Identifying assumption | What breaks it |
|---|---|---|
| Naïve regression (@eq-naive-adv) | Advertising uncorrelated with demand shocks | Targeting/timing of spend (almost always violated) |
| Randomized experiment | Exposure assigned at random | Infeasibility, scale, non-compliance, spillovers |
| Panel / fixed effects + geo baseline | Unobservables constant within unit; cross-market counterfactual valid | Time-varying shocks correlated with spend |
| Instrumental variables | Instrument moves ads but not demand directly | Weak or invalid (non-excludable) instruments |
| Quasi-experiment / regulatory shock | Treatment timing as-good-as-random | Confounding co-events (e.g., algorithm changes) |
| Border / discontinuity designs | Outcomes continuous across an ad-market boundary | Sorting or other discontinuities at the border |

: Identification strategies for advertising effects, with the assumption each relies on and the violation that defeats it. {#tbl-adv-identification}

### A Simulation of the Endogeneity Problem

The bias in @eq-naive-adv is concrete enough to reproduce. The following simulation
constructs markets in which a manager sets advertising as an increasing function of a
demand shock she observes but the analyst does not. The naïve regression then recovers a
coefficient far above the true effect, while a specification that conditions on the
shock recovers the truth—illustrating why a credible baseline is the whole game.

```{r adv-endogeneity, message=FALSE, warning=FALSE}
set.seed(10)

n_markets <- 2000
true_beta <- 0.5                       # the causal advertising effect we want back

# Demand shock the MANAGER sees but the ANALYST does not (the confounder)
demand_shock <- rnorm(n_markets, 0, 1)

# Manager targets favorable markets: spend rises with the anticipated shock
advertising <- 2 + 1.2 * demand_shock + rnorm(n_markets, 0, 0.5)

# Sales depend on advertising (true_beta) AND the same hidden shock
sales <- 5 + true_beta * advertising + 2.0 * demand_shock + rnorm(n_markets, 0, 1)

naive   <- coef(lm(sales ~ advertising))["advertising"]
adjusted <- coef(lm(sales ~ advertising + demand_shock))["advertising"]

cat("True effect          :", true_beta, "\n")
cat("Naive estimate       :", round(naive, 3), "  (biased upward)\n")
cat("Baseline-adjusted     :", round(adjusted, 3), "  (recovers the truth)\n")
```

The naïve estimate is biased upward by exactly the mechanism @Gordon_2017 warn about:
the manager's targeting loads the omitted shock onto the advertising coefficient. Only
when the analyst controls the demand baseline—the empirical analogue of a randomized
control group or a cross-market counterfactual—does the estimate return to the truth.

### Estimating Advertising Effects and Elasticities

Once identification is taken seriously, the estimated magnitudes shrink. @shapiro2021tv
estimate the distribution of TV advertising elasticities and returns across a large set
of product categories and find elasticities *below* prior benchmarks, a large share of
estimates statistically insignificant or negative, and—on the return side—over 80% of
brands facing *negative* marginal returns, with only about a third earning a positive
return from their observed schedule. The pattern is robust to functional form and is not
an artifact of low power or measurement error, echoing the prior-distribution evidence
in @Shapiro_2018 that a substantial share of consumer-packaged-goods elasticities are
insignificant or negative. The collected experimental and quasi-experimental program—
@shapiro2016, @blake2014, @lewis2011, @guitart2021impact—reinforces the same conclusion:
credibly identified advertising effects are small, heterogeneous, and frequently
unprofitable at the margin.

These small *short-run* elasticities are consistent with the behavioral prediction of
@vakratsas1999 that elasticities decline over the life cycle, and they motivate the
search for effects on outcomes other than immediate own-brand sales—word of mouth,
website traffic, firm value—where advertising's influence may be larger or more
reliably detected.

### Advertising's Effects on Word of Mouth, Traffic, and Firm Value

A productive response to small sales elasticities is to widen the dependent variable.
Advertising shapes price sensitivity through the **consideration set** and the relative
strength of preference [@mitra1995], and informative advertising reduces the
probability that a horizontally differentiated consumer fails to choose her best
alternative by roughly ten percent [@anand2011]. A large body of work then traces
advertising into *online* outcomes. TV advertising measurably moves online word of mouth
(WOM) about both the brand and the program—the phenomenon of **social TV**, the joint
consumption of a program and social-media activity about it [@fossen2017television]—and
ads aired in programs with *higher* social-TV activity drive more online shopping, with
affective (funny, emotional) ads most effective [@fossen2019social]. Crucially, the
most-discussed programs are not always the best vehicles for brand engagement, and
online program engagement predicts a *larger ad audience* because engaged viewers are
less likely to change channels during breaks [@fossen2021online]. Product placement
follows the same logic: prominent, especially verbal, placements raise online
conversation and web traffic, with decreasing returns at high prominence and little
enhancement from nearby TV advertising [@fossen2019measuring].

When advertising and consumer-generated media are modeled jointly, content
classification becomes decisive. For movies, pre-release blog volume and advertising
drive opening-day box office while post-release blog *valence* and advertising drive
later performance, with effects heterogeneous across markets in ways tied to
demographics and identifiable through instrumental variables [@gopinath2013blogs]. Over
a product's life, only **recommendation**-oriented WOM valence directly moves sales—not
sheer WOM volume—and its influence *grows* over time while both attribute- and
emotion-based advertising *wear out*, with rational attribute messages wearing out faster
than emotional ones [@gopinath2014investigating]. The same architecture lets firms be
classified as consumer-driven (WOM dominant) or firm-driven (advertising dominant),
which dictates where the marketing dollar should go. The macro view—@hewett2016—finds
that advertising does not directly move traditional media, online WOM, or consumer
sentiment, yet *does* directly affect firm outcomes, underscoring that the channel
through which advertising pays varies by context.

Advertising's reach extends to capital markets and even to public welfare.
@fich2024target show that *target-firm* advertising raises takeover premiums and
negotiating power while lowering acquirer announcement returns—consistent with targets
using advertising to communicate value to investors and acquirers—so that even failed
bids leave a permanent value gain. Ownership structure feeds back onto advertising:
under **common ownership**, competing firms with shared institutional blockholders cut
advertising, an effect identified using mutual-fund mergers as an exogenous shock and
concentrated in competitive, advertising-intensive industries [@lu2022frenemies]. The
spillovers of a rival's *bankruptcy* depend on a firm's own marketing investments,
moderated by industry growth (for advertising) and concentration (for R&D)
[@jindal2023effects]. And advertising can move *societal* outcomes: during the COVID-19
pandemic, TV ads carrying COVID narratives raised social-distancing behavior—eleven
times more in counties without government mandates—identified through a border strategy
exploiting discontinuities across TV markets, evidence that brand-sponsored advertising
can substitute for absent public policy [@ghosh2023societal].

### Spillovers and Attribution

Two econometric problems deserve separate treatment because they recur across digital
channels. The first is **spillover**: an ad's effect is not confined to its target.
@fossen2021impact show that political TV ads exert positive *ad-to-ad* spillovers—the ad
*following* a political ad suffers an 89% smaller audience decline and gains positive
online chatter—filling a gap in research that had largely ignored how one ad conditions
response to the next. A rich literature on search and display spillovers
[@kitts2014; @joo2014; @sahni2016; @yang2010; @ghose2016; @Rutz_2011] establishes that
estimating an ad's effect while ignoring its interaction with adjacent advertising
biases the estimate, often badly.

The second is **attribution**: assigning credit for a conversion across the many
touchpoints that preceded it [@simonson2001; @lambrecht2013; @li2014; @blake2015;
@zantedeschi2017]. Attribution is hard for the same reason @eq-naive-adv is biased—the
touchpoints a consumer encounters are *selected*, not assigned—so last-click and similar
heuristics confound correlation with cause. The methodological lesson is uniform across
spillover and attribution: the unit of analysis is rarely a single isolated exposure,
and treating it as one reintroduces the endogeneity the field works so hard to remove.

### Advertising Content and Consumer Demand for Ads

Two further econometric streams round out the picture. A content literature
[@xu2014a; @teixeira2014; @tucker2015; @liaukonyte2015; @rao2015; @sudhir2016] measures
how *what an ad says and shows*—its emotional arc, its information content, its
persuasive devices—maps to attention, sharing, and purchase, connecting the behavioral
creative-strategy results of @dallolio2022 to field outcomes. And a demand-for-ads
literature [@goldstein2014; @wilbur2016; @tuchman2017] treats advertising as something
consumers themselves have preferences over—tolerating, avoiding, or even seeking it—
which closes the loop back to the theoretical model of @choi2023consumer: if consumers
choose their exposure, advertising volume and targeting are equilibrium outcomes, not
exogenous treatments.

### Deceptive Advertising

The welfare and regulatory edge of advertising is sharpest where messages are
*deceptive*. Regulatory reports of deceptive advertising depress stock-market
performance, an effect that strong brand reputation partially buffers [@wiles2010].
@rao2022 study a particularly modern form—**fake-news advertising**, promotional content
that mimics the format of surrounding news without disclosure—and ask what happens when
the Federal Trade Commission shuts the channel down. The accounting framework decomposes
a merchant's demand into organic visits, fake-news-ad referrals, and regular-ad
referrals, allowing direct effects and cross-channel spillovers to be separated:

$$
\begin{aligned}
Q_{\text{merchant}} &= Q_{\text{org}} + Q_{\text{fn}} + Q_{\text{reg}}, \\
Q_{\text{org}} &= \alpha_{\text{org}}
   + \gamma_1^{\text{fn}} Ad_{\text{fn}} + \gamma_1^{\text{reg}} Ad_{\text{reg}}, \\
Q_{\text{fn}} &= \alpha_{\text{fn}} Ad_{\text{fn}}
   + \gamma_2^{\text{reg}} Ad_{\text{reg}}\, \mathbb{I}(Ad_{\text{fn}} > 0), \\
Q_{\text{reg}} &= \alpha_{\text{reg}} Ad_{\text{reg}}
   + \gamma_3^{\text{fn}} Ad_{\text{fn}}\, \mathbb{I}(Ad_{\text{reg}} > 0),
\end{aligned}
$$ {#eq-fakenews}

where the $\alpha$ terms are direct effects and the $\gamma$ terms cross-channel
spillovers, with $fn$ denoting fake news and $reg$ regular advertising. Two effects are
of interest: a **treatment effect** (fake news directly drives $Q_{\text{fn}}$ and spills
over to organic and regular channels through $\gamma_1^{\text{fn}}$ and
$\gamma_3^{\text{fn}}$) and a **selection effect** (in the channel's absence, some
consumers substitute to other channels while others stop searching entirely). After the
shutdown, regular-ad referrals rise—fake-news and regular ads are substitutes—and the
probability that a merchant draws a complaint falls by about 8%, though somewhat
mechanically. The selection effect is small: substitution to legitimate channels is
dominated by the decline in organic demand and fake-news referrals. The estimated decline
is best read as an *upper bound* on the regulation's true impact, because a contemporaneous
search-algorithm change cannot be ruled out—a textbook illustration of the
quasi-experimental confound flagged in @tbl-adv-identification. With outcomes that are
95% zeros, the design sits near the boundary where rare-events corrections to logistic
regression become relevant [@king2001logistic].

### A Worked Estimate: Geo-Experimental Lift

To make the recommended remedy concrete, the following example implements the
geo-baseline logic of @Gordon_2019: estimate each market's counterfactual from a holdout
of untreated markets, then read the advertising effect as the lift above that baseline.
The simulation embeds a true incremental effect of a 6% sales lift and recovers it via a
difference-in-differences contrast between treated and control geographies.

```{r adv-geo-lift, message=FALSE, warning=FALSE}
set.seed(2024)

n_geo  <- 200
true_lift <- 0.06                                  # true incremental effect: +6%

# Baseline sales differ across markets and across pre/post periods
geo_base <- rnorm(n_geo, 100, 15)
period_shock <- 4                                  # common seasonal lift, both arms

treated <- rep(c(TRUE, FALSE), each = n_geo / 2)   # randomized at the GEO level

pre  <- geo_base + rnorm(n_geo, 0, 3)
post <- geo_base + period_shock +
        ifelse(treated, true_lift * geo_base, 0) + rnorm(n_geo, 0, 3)

did <- data.frame(
  geo = 1:n_geo, treated = treated,
  delta = post - pre                               # within-geo change
)

# Difference-in-differences: treated change minus control change
est_lift_pct <- (mean(did$delta[did$treated]) -
                 mean(did$delta[!did$treated])) / mean(geo_base)

cat("True incremental lift :", paste0(round(100 * true_lift, 1), "%"), "\n")
cat("Estimated lift (DiD)  :", paste0(round(100 * est_lift_pct, 1), "%"), "\n")
```

By differencing each market against its own pre-period and then against untreated
markets, the design strips out both fixed cross-market heterogeneity and the common
seasonal shock, isolating the incremental effect the naïve regression could not. This is
the empirical embodiment of the chapter's central claim: advertising's effect is a
credibly constructed *increment above a baseline*, and the quality of that baseline,
not the sophistication of the regression, is what makes the estimate believable.

## Key Takeaways

- Advertising's effect is best defined as an **increment above a baseline**, not a raw
  correlation between spending and sales; the entire empirical enterprise turns on how
  credibly that baseline is built [@Gordon_2019].
- The behavioral tradition decomposes persuasion into conditional, separately measurable
  stages (@eq-mcguire-chain) and shows the **route**—central versus peripheral, cognitive
  versus affective—is contingent on involvement, motivation, ability, and opportunity
  [@petty1983; @maclnnis1989; @vakratsas1999].
- **Affect can precede and bias cognition** rather than follow it [@zajonc1980], and
  ad-evoked feelings move brand attitudes robustly, especially for hedonic products
  [@pham2013].
- Targeting and repetition are **equilibrium objects**: modeling consumers' strategic
  response to tracking and wear-out yields non-obvious comparative statics with direct
  policy content [@choi2023consumer].
- Credibly identified advertising effects are **small, heterogeneous, and often
  unprofitable** at the margin; naïve regressions overstate them because firms target
  favorable conditions [@shapiro2021tv; @Gordon_2017; @Lewis_2015].
- Widening the dependent variable—to word of mouth, traffic, firm value, even public
  health—and respecting **spillover** and **attribution** often reveals effects that
  immediate own-brand sales elasticities miss [@hewett2016; @fossen2021impact; @ghosh2023societal].
