# Information Theory {#sec-information-theory}

Every marketing act is, at bottom, an act of communication: a firm encodes an
intention—about quality, identity, price, or fit—into a signal, sends that signal
through a noisy channel, and hopes a consumer decodes something close to what was
sent. The discipline spends enormous sums on the encoding (creative, copy, design)
and on buying the channel (media), yet the quantity that ultimately matters is how
much *uncertainty about the consumer's eventual action* the message resolves. That
quantity—information, measured in bits—is not a metaphor. It is a precise,
additive, model-free statistic with a century of theory behind it, and it gives
marketing a common currency for problems that otherwise look unrelated: how
surprising a brand name is, how much a segmentation variable tells us about
response, how redundant two ad exposures are, and how much a recommendation system
can possibly learn about a user.

This chapter develops that currency. We begin with @Shannon_1948's model of
communication and the single construct on which everything else rests—**entropy**,
the average uncertainty in a random outcome. We then build the relational
quantities marketing actually uses: **mutual information** (how much one variable
tells us about another), **relative entropy** or Kullback–Leibler divergence (how
costly it is to act on the wrong belief about a market), and the **channel
capacity** that bounds how much any message can convey. Throughout, intuition
leads and the formalism follows immediately and in full, with reproducible R for
every estimator. The information-theoretic view recurs later in the book—in the
surprise-driven account of virality (@sec-virality), in the attention economics
of advertising (@sec-advertising), and in the entropy-based diagnostics used for
text and image data (@sec-image-processing)—so it earns a careful treatment here.

::: {.callout-note}
## Scope and software
The worked examples use base R together with the `philentropy` package, which
implements the standard distance and information measures over discrete
distributions. We treat probabilities as *known* in the formal development and
return to the harder problem—*estimating* these quantities from finite marketing
data, where naive plug-in estimators are badly biased—in the section on
estimation.
:::

## The Mathematical Theory of Communication

@Shannon_1948 framed communication as a chain of five components: an information
**source** that produces a message, a **transmitter** that encodes it into a
signal, a **channel** that carries the signal and injects **noise**, a
**receiver** that decodes, and a **destination**. @weaver1963mathematical, in the
expository companion that popularized the theory, organized the problems of
communication into three levels—the *technical* problem (how accurately can
symbols be transmitted?), the *semantic* problem (how precisely do the
transmitted symbols convey the intended meaning?), and the *effectiveness*
problem (how effectively does the received meaning change conduct?). Marketing
communication research lives mostly at the second and third levels, but Shannon's
theory, which solves the first, supplies the measurement scaffolding for all
three [@severin1997communication].

```{mermaid}
%%| label: fig-shannon-channel
%%| fig-cap: "Shannon's model of communication applied to marketing. The firm (source) encodes an intended position into a message; media (the channel) transmit it subject to competitive and contextual noise; the consumer (receiver) decodes a perception that drives action. Mutual information quantifies what survives the channel."
flowchart LR
  S["Source<br/>(firm intent)"] --> E["Encoder<br/>(creative, copy)"]
  E --> C["Channel<br/>(media)"]
  N["Noise<br/>(clutter,<br/>competition)"] --> C
  C --> D["Decoder<br/>(consumer<br/>perception)"]
  D --> R["Destination<br/>(action: choice,<br/>WOM, search)"]
```

The crucial move is to treat the message as a **random variable**. Before the
consumer is exposed, her eventual perception or action is uncertain; the message
is informative exactly insofar as it reduces that uncertainty. Shannon's
contribution was to show that the average reduction in uncertainty has a unique
measure—up to choice of units—satisfying a short list of intuitive axioms
(continuity, monotonicity in the number of equally likely outcomes, and additivity
over independent sub-choices). That measure is entropy.

> "The fundamental problem of communication is that of reproducing at one point
> either exactly or approximately a message selected at another point."
>
> --- @Shannon_1948

## Entropy: The Measure of Uncertainty

Let $X$ be a discrete random variable taking values in a finite alphabet
$\{x_1,\dots,x_n\}$ with probability mass function
$\mathbf{p}_X = (p_1,\dots,p_n)$, where $p_i = \Pr(X = x_i)$. The **entropy** of
$X$ is the expected value of the *surprise* of its outcomes, where the surprise of
an outcome of probability $p$ is $\log(1/p)$—rare outcomes are surprising, certain
outcomes are not. Formally,

$$
H(\mathbf{p}_X) \;=\; -\sum_{i=1}^{n} p_i \log_2 p_i
\;=\; \mathbb{E}\!\left[\log_2 \frac{1}{p(X)}\right].
$$ {#eq-entropy}

Entropy is the average number of yes/no questions needed to pin down the value of
$X$ under an optimal questioning strategy, and equivalently the minimum expected
number of **bits** needed to encode $X$ without loss [@Shannon_1948;
@Zwillinger_1995, pp. 262]. The base of the logarithm fixes only the unit: base 2
gives *bits*, base $e$ gives *nats*, and base 10 gives *dits* (or *hartleys*).
Units convert by a constant factor, because for any bases $a$ and $b$,
$\log_a k = \log_a b \cdot \log_b k$, so

$$
H_b(X) \;=\; (\log_b a)\, H_a(X).
$$ {#eq-entropy-base}

Two properties make entropy the right primitive. First, it depends only on the
*distribution* of $X$, not on the labels of the outcomes: relabeling brands does
not change how hard the market is to predict. Second, it is maximized by the
uniform distribution and minimized by a degenerate one:

$$
0 \;\le\; H(\mathbf{p}_X) \;\le\; \log_2 n, \qquad
H(\mathbf{p}_X) = \log_2 n \iff X \sim \mathrm{Uniform}.
$$

A market in which all $n$ brands hold equal share is maximally uncertain
($\log_2 n$ bits); a monopolized market is perfectly predictable (0 bits). This is
why entropy and its monotone transforms appear as **concentration** and
**diversity** indices—the exponential of entropy, $e^{H}$, is the *effective
number* of equally likely outcomes, a model-free analogue of the number of
meaningful competitors.

### The Binary Case

The simplest instance fixes intuition. If $X$ has two outcomes with
$\mathbf{p}_X = (p, 1-p)$, entropy reduces to the **binary entropy function**

$$
H_b(p) \;=\; -p\log_2 p - (1-p)\log_2(1-p),
$$ {#eq-binary-entropy}

which is symmetric about $p = \tfrac{1}{2}$, equals $0$ at $p \in \{0,1\}$ (a
sure outcome carries no information), and peaks at exactly $1$ bit when
$p = \tfrac{1}{2}$ (a fair coin is maximally uncertain). A churn flag, a
click/no-click outcome, or a buy/no-buy decision is a Bernoulli draw whose
predictive difficulty is summarized by this curve: a campaign that pushes a 50/50
audience toward 90/10 has removed $1 - H_b(0.1) \approx 0.53$ bits of uncertainty
about who will convert.

```{r binary-entropy, fig.cap="The binary entropy function $H_b(p)$ in bits. Uncertainty is maximal (1 bit) at $p=0.5$ and vanishes at the certain outcomes $p \\in \\{0,1\\}$."}
H_binary <- function(p) {
  # 0 log 0 := 0 by continuity; guard the endpoints
  ifelse(p == 0 | p == 1, 0, -p * log2(p) - (1 - p) * log2(1 - p))
}

p_grid <- seq(0, 1, length.out = 200)
plot(p_grid, H_binary(p_grid), type = "l", lwd = 2,
     xlab = "p = Pr(X = 1)", ylab = "H(p)  [bits]")
abline(v = 0.5, lty = 3); abline(h = 1, lty = 3)
```

For a general (non-binary) distribution, the `philentropy` package computes
@eq-entropy directly. The example below uses a skewed ten-outcome distribution,
whose entropy lies strictly below the uniform maximum of $\log_2 10 \approx 3.32$
bits.

```{r shannon-entropy, message=FALSE, warning=FALSE}
library(philentropy)
Prob <- (1:10) / sum(1:10)   # a skewed 10-outcome distribution
H(Prob)                      # Shannon entropy, in bits
log2(10)                     # uniform maximum for comparison
```

## Mutual Information: How Much One Variable Tells Us About Another

Marketing rarely cares about a single variable in isolation; it cares about
*association*—does knowing a consumer's segment tell us anything about her
response? The information-theoretic answer is **mutual information**, the reduction
in uncertainty about $Y$ from learning $X$ (equivalently, about $X$ from learning
$Y$; the quantity is symmetric). Let $X$ and $Y$ have joint distribution
$\mathbf{p}_{X\times Y}$ with marginals $\mathbf{p}_X$ and $\mathbf{p}_Y$. Then

$$
I(X;Y) \;=\; H(\mathbf{p}_X) + H(\mathbf{p}_Y) - H(\mathbf{p}_{X\times Y})
\;=\; \sum_{i=1}^{n}\sum_{j=1}^{m}
   p(x_i,y_j)\,\log_2 \frac{p(x_i,y_j)}{p(x_i)\,p(y_j)}.
$$ {#eq-mutual-information}

The first form is an *inclusion–exclusion* identity: the information shared by $X$
and $Y$ is what their separate uncertainties double-count once they are observed
jointly. The second form reveals what mutual information *is*—the relative entropy
(defined below) between the true joint distribution and the product of marginals
that would obtain under independence. It therefore measures how far $X$ and $Y$ are
from independent, and it has the defining properties

$$
I(X;Y) \;\ge\; 0, \qquad I(X;Y) = 0 \iff X \perp Y, \qquad I(X;X) = H(X).
$$

The last identity—self-information equals entropy—says a variable tells us
everything about itself, namely its own entropy. Unlike Pearson correlation,
mutual information detects *any* statistical dependence, linear or not, which makes
it a natural screening statistic for nonlinear feature selection, for quantifying
how diagnostic a targeting variable is, and for measuring the redundancy between
two media exposures.[^pmi]

[^pmi]: The pointwise summand $\log_2\!\big(p(x,y)/[p(x)p(y)]\big)$ is the
**pointwise mutual information** (PMI) of the pair $(x,y)$. PMI is the workhorse
of computational linguistics, where it scores how much more often two words
co-occur than chance would predict; it underlies collocation extraction and the
word-association measures used in marketing text analytics (@sec-image-processing).
Mutual information is simply the expectation of PMI under the joint distribution.

```{r mutual-information, message=FALSE, warning=FALSE}
P_x  <- (1:10)  / sum(1:10)     # marginal P(X)
P_y  <- (20:29) / sum(20:29)    # marginal P(Y)
P_xy <- (1:10)  / sum(1:10)     # joint P(X, Y)
MI(P_x, P_y, P_xy)              # mutual information, in bits
```

## Relative Entropy: The Cost of the Wrong Belief

Suppose a manager believes the market follows distribution $q$ when it truly
follows $p$—she has mis-segmented, mis-forecast demand, or trained a model on a
stale population. How costly is that error, measured in the only currency we have,
bits? The answer is the **relative entropy**, or **Kullback–Leibler (KL)
divergence**, of $p$ from $q$:

$$
D_{\mathrm{KL}}(p \,\|\, q) \;=\; \sum_{x} p(x)\,\log_2 \frac{p(x)}{q(x)},
$$ {#eq-kl}

with the conventions $0\log\tfrac{0}{q}=0$ and $p\log\tfrac{p}{0}=\infty$.
$D_{\mathrm{KL}}(p\,\|\,q)$ is the *expected number of extra bits* incurred by
coding outcomes drawn from $p$ using a code optimized for $q$—the avoidable
inefficiency of acting on the wrong model. It satisfies **Gibbs' inequality**,

$$
D_{\mathrm{KL}}(p \,\|\, q) \;\ge\; 0, \qquad
D_{\mathrm{KL}}(p \,\|\, q) = 0 \iff p = q,
$$

so the truth is always (weakly) the most efficient belief to hold. Two cautions
matter in practice. First, KL divergence is **not a metric**: it is asymmetric,
$D_{\mathrm{KL}}(p\,\|\,q) \neq D_{\mathrm{KL}}(q\,\|\,p)$ in general, and violates
the triangle inequality. Which direction to use is a modeling decision—the
"forward" divergence $D_{\mathrm{KL}}(p\,\|\,q)$ penalizes a belief $q$ that places
little mass where $p$ is large (missing real demand), whereas the "reverse"
divergence penalizes $q$ that spreads mass where $p$ is zero (chasing phantom
demand). Second, it is unbounded and undefined when $q$ assigns zero probability to
an outcome that actually occurs, which is precisely why symmetrized, bounded
alternatives are often preferred for comparing market structures.

::: {.callout-tip}
## Connection to mutual information and likelihood
Mutual information is exactly the KL divergence between the joint distribution and
the independence benchmark:
$I(X;Y) = D_{\mathrm{KL}}\!\big(p_{X\times Y} \,\|\, p_X\, p_Y\big)$. And KL
divergence is the population analogue of the **log-likelihood ratio**: maximizing a
log-likelihood is, asymptotically, minimizing the KL divergence between the data's
true distribution and the fitted model. Information theory and maximum-likelihood
estimation are two views of the same object.
:::

### Symmetrized and Bounded Divergences

When the task is to *compare* two empirical distributions—two creative variants'
click profiles, two cities' category-share vectors, this month's versus last
month's search-term mix—analysts usually want a symmetric, finite measure. The
**Jensen–Shannon (JS) divergence** averages each distribution against their
mixture $m = \tfrac{1}{2}(p+q)$:

$$
\mathrm{JSD}(p \,\|\, q) \;=\;
\tfrac{1}{2} D_{\mathrm{KL}}(p \,\|\, m) + \tfrac{1}{2} D_{\mathrm{KL}}(q \,\|\, m),
\qquad m = \tfrac{1}{2}(p + q).
$$ {#eq-jsd}

Because $m$ has support wherever either $p$ or $q$ does, JS divergence is always
finite; it is symmetric by construction; it is bounded in $[0,1]$ bits; and its
square root is a true metric on the space of distributions. The three measures sit
in a clear hierarchy, summarized in @tbl-divergences.

| Measure | Formula | Symmetric? | Bounded? | Metric? | Typical marketing use |
|---|---|:--:|:--:|:--:|---|
| KL divergence | $\sum_x p\log\frac{p}{q}$ | No | No | No | Coding/inference cost of a wrong model; basis of likelihood |
| Jensen–Shannon | $\tfrac12 D_{\mathrm{KL}}(p\|m)+\tfrac12 D_{\mathrm{KL}}(q\|m)$ | Yes | $[0,1]$ bit | $\sqrt{\cdot}$ is | Comparing two empirical distributions (segments, periods) |
| Generalized JS | weighted $\sum_k \pi_k D_{\mathrm{KL}}(p_k\|\bar p)$ | Yes | Yes | --- | Dispersion across $\ge 3$ distributions (e.g., regional mixes) |

: Divergence measures for comparing marketing distributions. KL is the
inferential primitive but ill-behaved as a distance; Jensen–Shannon is the
preferred symmetric, bounded comparison; the generalized form extends it to a
weighted ensemble of distributions. {#tbl-divergences}

```{r divergences, message=FALSE, warning=FALSE}
# Two campaign click-profiles over five creative slots
p <- c(0.30, 0.25, 0.20, 0.15, 0.10)
q <- c(0.10, 0.15, 0.20, 0.25, 0.30)

# Asymmetry of KL: the two directions differ
KL_pq <- suppressMessages(KL(rbind(p, q), unit = "log2"))
KL_qp <- suppressMessages(KL(rbind(q, p), unit = "log2"))

# Jensen-Shannon: symmetric and bounded in [0, 1] bits
JSD_pq <- suppressMessages(JSD(rbind(p, q), unit = "log2"))

c(KL_p_to_q = as.numeric(KL_pq),
  KL_q_to_p = as.numeric(KL_qp),
  JSD       = as.numeric(JSD_pq))
```

## Joint and Conditional Entropy

To reason about messages and responses together we need the entropy of *pairs*.
The **joint entropy** of $(X,Y)$ is the uncertainty in the pair,

$$
H(X,Y) \;=\; -\sum_{i=1}^{n}\sum_{j=1}^{m}
   p(x_i,y_j)\,\log_2 p(x_i,y_j),
$$ {#eq-joint-entropy}

and the **conditional entropy** $H(Y\mid X)$ is the uncertainty that *remains* in
$Y$ once $X$ is known,

$$
H(Y\mid X) \;=\; -\sum_{i=1}^{n}\sum_{j=1}^{m}
   p(x_i,y_j)\,\log_2 \frac{p(x_i,y_j)}{p(x_i)}.
$$ {#eq-conditional-entropy}

These are bound together by the **chain rule** and its multivariate extension,

$$
H(X,Y) = H(X) + H(Y\mid X) = H(Y) + H(X\mid Y), \qquad
H(X_1,\dots,X_n) = \sum_{i=1}^{n} H(X_i \mid X_{i-1},\dots,X_1),
$$ {#eq-chain-rule}

which simply says the total uncertainty of a sequence is the running sum of each
variable's *new* uncertainty given everything before it. The remaining structural
facts follow directly and have clean marketing readings:

- **Conditioning cannot increase entropy:** $0 \le H(Y\mid X) \le H(Y)$. Observing
  a covariate $X$ never makes the response $Y$ *more* uncertain on average—at worst
  it is uninformative.
- **Independence is the equality case:** if $X \perp Y$ then
  $H(Y\mid X) = H(Y)$ and $H(X,Y) = H(X) + H(Y)$; a segmentation variable
  independent of response buys no predictive power.
- **Subadditivity:** $H(X_1,\dots,X_n) \le \sum_i H(X_i)$, with equality iff the
  variables are mutually independent. Correlated touchpoints carry *redundant*
  information; their joint entropy falls short of the sum of their marginals by
  exactly the total mutual information among them.

The first two of these connect the entropy quantities to mutual information through
the identity $I(X;Y) = H(Y) - H(Y\mid X) = H(X) - H(X\mid Y)$: mutual information
is the drop in entropy from conditioning, which is why it is non-negative and why
it vanishes precisely under independence.

```{r joint-conditional-entropy, message=FALSE, warning=FALSE}
P_xy <- (1:10) / sum(1:10)     # a joint distribution P(X, Y)
JE(P_xy)                       # joint entropy  H(X, Y)

P_x <- (1:10) / sum(1:10)
P_y <- (1:10) / sum(1:10)
CE(P_x, P_y)                   # conditional entropy  H(X | Y)
```

::: {.callout-note collapse="true"}
## Continuous (differential) entropy
For a continuous random vector $\mathbf{X} \in \mathbb{R}^d$ with density
$p(\mathbf{x})$, the analogue of @eq-entropy is the **differential entropy**
$h(\mathbf{X}) = -\int_{\mathbb{R}^d} p(\mathbf{x})\log p(\mathbf{x})\,d\mathbf{x}$.
Differential entropy retains the chain rule and the form of relative entropy—
$D_{\mathrm{KL}}(p\,\|\,q) = \int p(\mathbf{x})\log\frac{p(\mathbf{x})}{q(\mathbf{x})}\,d\mathbf{x}$—
but loses two properties of the discrete case: it can be negative, and it is *not*
invariant to a change of variables (rescaling $\mathbf{X}$ shifts $h$ by the
log-Jacobian). Mutual information, being a difference of entropies, *is* invariant
and non-negative even in the continuous case, which is why it, rather than raw
differential entropy, is the quantity to report when variables are continuous.
:::

## Channel Capacity: The Ceiling on Any Message

How much can a message *possibly* convey? Model the medium as a **discrete
memoryless channel**: an input symbol $X$ enters, and a corrupted output $Y$
emerges according to a fixed transition law $p(y\mid x)$ that captures the channel's
noise—ad clutter, competitive interference, imperfect attention, mis-decoded
positioning. The receiver's information about the input is $I(X;Y)$, which depends
both on the channel and on how the *sender chooses to use it*, i.e., on the input
distribution $p(x)$. The **channel capacity** is the best the sender can do,
optimizing over input distributions:

$$
C \;=\; \max_{p(x)} \; I(X;Y).
$$ {#eq-capacity}

Capacity is the maximum rate, in bits per use of the channel, at which information
can be transmitted with arbitrarily small error—a hard ceiling no creative
execution can beat. The reading for marketing is structural rather than literal: a
cluttered, low-attention medium has small capacity, so even a perfect message
delivers little; a high-fidelity, high-attention context (a chosen search query, an
owned app) has large capacity, so the same message lands. Capacity makes precise
the intuition that *channel choice bounds message effectiveness*, and it frames the
@weaver1963mathematical effectiveness problem as one of allocating scarce
informational bandwidth to the messages that most reduce uncertainty about
profitable action.

## Estimating Information Quantities From Data

The formulas above assume the probabilities are known. In practice we estimate them
from a finite sample, and the obvious *plug-in* estimator—compute empirical
frequencies $\hat p_i = n_i/N$ and substitute into @eq-entropy—is **systematically
biased downward**: with finite data we under-observe rare outcomes, the empirical
distribution looks more concentrated than the truth, and estimated entropy falls
short of $H$. The bias is of order $(K-1)/(2N\ln 2)$ bits for an alphabet of $K$
outcomes and $N$ observations, so it is worst exactly when it bites hardest in
marketing—high-cardinality variables (SKUs, search terms, ZIP codes) observed in
modest samples. Mutual information inherits the problem and is biased *upward*,
making independent variables appear spuriously associated; this is why a raw MI
screen will rank a high-cardinality noise variable above a genuine low-cardinality
predictor unless corrected.

Three practical defenses follow, and a careful analyst applies them before
reporting any information statistic:

1. **Bias correction.** Apply the Miller–Madow correction
   $\hat H_{\text{MM}} = \hat H_{\text{plug-in}} + (\hat K - 1)/(2N\ln 2)$, where
   $\hat K$ is the number of observed outcomes, or use a Bayesian/shrinkage
   estimator that regularizes the cell probabilities toward uniform.
2. **Permutation calibration.** To test whether an estimated $I(X;Y)$ exceeds what
   finite-sample bias alone would produce, recompute it on many random permutations
   of $Y$ (which destroys any true dependence) and compare the observed value to
   that null distribution.
3. **Binning discipline.** For continuous variables, the estimate depends on how
   values are discretized; report sensitivity to bin width, or use a
   discretization-free estimator (e.g., nearest-neighbor methods), rather than
   treating one arbitrary histogram as ground truth.

The simulation below makes the plug-in bias visible: two *independent* variables
have true mutual information $I(X;Y)=0$, yet the plug-in estimate is reliably
positive, and grows with the alphabet size.

```{r mi-bias, message=FALSE, warning=FALSE}
set.seed(11)

# Plug-in mutual information (in bits) for two independent categorical variables
plugin_mi <- function(n, K) {
  x <- sample(K, n, replace = TRUE)        # X and Y are independent by construction
  y <- sample(K, n, replace = TRUE)
  joint  <- table(x, y) / n
  px <- rowSums(joint); py <- colSums(joint)
  outer_pxpy <- outer(px, py)
  nz <- joint > 0
  sum(joint[nz] * log2(joint[nz] / outer_pxpy[nz]))   # true value is 0
}

# Average plug-in estimate across replications, for two alphabet sizes
sims <- expand.grid(K = c(5, 20), rep = 1:200)
sims$mi <- mapply(function(K) plugin_mi(n = 200, K = K), sims$K)
aggregate(mi ~ K, data = sims, FUN = mean)   # both should be 0; both are > 0
```

The estimate is positive for both alphabets and larger for the high-cardinality
case, exactly as the $(K-1)/(2N\ln 2)$ bias term predicts. Any information-based
feature ranking that ignores this will over-reward granular variables.

## Why Information Theory Earns Its Place in Marketing

Information theory supplies marketing with a small set of model-free quantities
that recur across otherwise unrelated problems. **Entropy** measures uncertainty
and, through its exponential, the effective number of competitors or
choices—turning concentration and diversity into one statistic. **Mutual
information** measures association of any functional form, giving a principled
target-selection and redundancy criterion that Pearson correlation cannot match.
**Relative entropy** prices the cost of holding the wrong model of a market and
links directly to maximum-likelihood estimation, while its symmetric
Jensen–Shannon variant gives a well-behaved distance between empirical
distributions. **Channel capacity** bounds what any message can achieve and
reframes media choice as bandwidth allocation. The unifying idea—that the value of
a communication is the uncertainty it resolves about profitable action—travels far
beyond this chapter, surfacing wherever marketing must quantify surprise,
diagnosticity, or redundancy in data.

## Key Takeaways

- **Communication is uncertainty reduction.** @Shannon_1948's model and
  @weaver1963mathematical's three-level taxonomy frame marketing messages as
  signals whose worth is the information they convey about consumer response.
- **Entropy is the primitive.** $H(\mathbf{p}_X) = -\sum_i p_i\log_2 p_i$
  (@eq-entropy) is bounded by $\log_2 n$, maximized under uniformity, and its
  exponential is the effective number of outcomes.
- **Mutual information measures any dependence.** $I(X;Y)\ge 0$ with equality iff
  $X \perp Y$ (@eq-mutual-information); it equals the KL divergence from
  independence and detects nonlinear association invisible to correlation.
- **KL divergence prices wrong beliefs but is not a distance**—asymmetric and
  unbounded; use Jensen–Shannon (@eq-jsd) to *compare* empirical distributions.
- **Channel capacity bounds message effectiveness** (@eq-capacity): channel choice
  sets a ceiling no creative can exceed.
- **Estimate with care.** Plug-in entropy is biased low and plug-in mutual
  information biased high; correct the bias and calibrate against a permutation
  null before trusting any information statistic.
