# Sales {#sec-sales}

Selling is where a firm's marketing promises are tested against a buyer's
willingness to pay. This chapter treats the sales function as two distinct but
connected measurement problems. The first is **inference about demand from
observed sales**: firms and researchers rarely see the full demand curve, but
they do see rankings, units, and revenue, and a recurring task is to reconstruct
latent demand from these partial signals. The second is **management of the sales
force**: the salespeople who convert demand into revenue are themselves an asset
whose compensation, productivity, retention, and future value can be modeled,
measured, and optimized.

The two problems share a methodological spine. In both, the object of interest
is unobserved—true demand in the first case, a salesperson's forward-looking
profit contribution in the second—and must be recovered from a noisy,
selection-prone proxy. A reader who finishes this chapter should be able to
estimate a demand curve from a sales-rank list, reason about when that estimate
breaks down, and evaluate a sales-force compensation or retention policy using
the same identification discipline applied throughout the book.

We proceed from the outside in. We begin with the economics of compensation and
its hidden health costs, because incentives are the lever managers pull most
often. We then turn to **sales from rank**, the empirical problem of inferring
demand from ordinal best-seller lists, and develop the Pareto rank–demand model
in full. We close with the **salesperson as an asset**: turnover and its
contagion through peer networks, and the forward-looking valuation of a sales
force.

## Compensation and Its Hidden Costs

The canonical justification for putting variable pay—commissions, bonuses,
quota-based incentives—into a salesperson's compensation contract is an
agency-theoretic one. The firm (principal) cannot perfectly observe the agent's
effort, only the noisy output (sales) that effort produces. Tying pay to output
realigns the agent's incentives with the firm's, and the standard prediction is
that a steeper pay–performance slope elicits more effort and higher sales. This
prediction is robust and well documented.

What the classical account omits is that effort is not free of physiological
cost. @habel2021 confirm the performance benefit of variable compensation but
show that the same incentive intensity that lifts sales also operates as a
*health hazard*: it raises stress, which manifests as emotional exhaustion and a
higher incidence of sick days. The effect is not uniform across the sales force.
It is **buffered** for salespeople with higher personal ability and richer
social resources—the people best equipped to absorb the additional psychological
load—and sharpest for those without such buffers.

The managerial implication is that the pay–performance slope has a cost the
standard model prices at zero. Let $w_i = \alpha + \beta\, s_i$ denote a linear
contract for salesperson $i$, with fixed component $\alpha$, variable rate
$\beta$, and realized sales $s_i$. The firm's familiar problem is to choose
$\beta$ to maximize expected profit net of wage cost. @habel2021 effectively add
a term: the agent's stress, and hence absenteeism and exhaustion, is increasing
in $\beta$, so the firm faces a more accurate objective in which $\beta$ carries
a shadow cost through reduced effective capacity and higher turnover risk. The
optimal incentive intensity that internalizes this health externality is lower
than the one that ignores it, and it varies with the ability and social
resources of the individual—an argument for incentive schemes that are
differentiated rather than uniform.

::: {.callout-note}
The health cost of incentives is a *moderated* effect, not a main-effect
caveat. Reporting only the average performance gain from steeper incentives, as
much of the older literature does, overstates the net benefit for the segment of
the sales force least able to bear the stress.
:::

## Sales from Rank {#sec-sales-from-rank}

Many of the most visible signals of commercial success are **ordinal**, not
cardinal. Amazon publishes a sales rank for every book; Apple's App Store
publishes ranked lists rather than download counts; the music, film, and
publishing industries all live by charts. A rank is informative—rank 1 outsells
rank 100—but it discards the magnitudes a firm actually needs: how *many* more
units does rank 1 move than rank 100? Recovering demand magnitudes from rank is
the central inference problem of this section.

The difficulty is that the map from demand to rank is many-to-one and unknown.
Two products can swap ranks after a small change in either one's sales, and the
same rank can correspond to wildly different unit volumes in a thick versus a
thin market. The literature solves this in three stages of increasing
ambition, summarized in @tbl-rank-approaches: experimentally perturb rank and
read off the demand response; supplement rank with auxiliary demand data; or
impose a parametric demand–rank law and estimate it from rank data alone.

| Study | Setting | Demand data required | Core idea |
|---|---|---|---|
| @chevalier2003measuring | Amazon books | Known/very low for chosen titles | Buy units of low-selling books; watch rank move; trace the rank–demand map |
| @brynjolfsson2003consumer | Amazon books | Publisher demand data | Calibrate the rank–demand relationship directly |
| @garg2013inferring | Apple App Store | None | Impose a Pareto law; identify it from paid and grossing ranks |
| @he2020sales | Amazon (general) | None | Generalize @chevalier2003measuring beyond books |

: Approaches to inferring demand from sales rank {#tbl-rank-approaches}

### The Experimental Approach

The cleanest identification comes from intervention. @chevalier2003measuring run
a field experiment on Amazon: they select books whose demand is either known or
known to be very small, purchase large quantities relative to that baseline
demand, and observe how each title's sales rank responds to the induced demand
shock. Because the experimenter controls the size of the shock, the otherwise
unobservable mapping between a change in units and a change in rank is revealed.
Repeating this across the rank distribution traces out the rank–demand curve.

The method's strength—a controlled demand shock—is also its limitation. It is
practical only for **low-selling** titles, where a feasible experimental purchase
is large relative to organic demand. At the head of the distribution, the
purchase required to move rank by a perceptible amount is prohibitive, so the
high-volume region of the curve cannot be identified experimentally. The
generalization of this revealed-demand logic to Amazon products beyond books is
developed in @he2020sales.

### The Auxiliary-Data Approach

When the experimenter cannot perturb demand, the next-best instrument is
external demand data. @brynjolfsson2003consumer obtain demand figures from a book
publisher and pair them with observed ranks, which pins down the rank–demand
relationship by direct calibration rather than by experiment. This dispenses with
the need to intervene but substitutes a different scarce resource—proprietary
unit-sales data—that most analysts cannot obtain.

### The Structural Approach: A Pareto Rank–Demand Law

The most general method dispenses with both intervention and demand data,
recovering magnitudes from rank alone by imposing structure on the demand
distribution. @garg2013inferring develop this approach on Apple's App Store,
which is an unusually favorable laboratory because it publishes three distinct
ranked lists that, jointly, over-identify the model:

1. **Top-free apps** — ranked by download volume among zero-price apps.
2. **Top-paid apps** — ranked by download volume among positive-price apps.
3. **Top-grossing apps** — ranked by *revenue*, i.e., price times downloads.

The key modeling assumption is that downloads follow a **Pareto** (power-law)
distribution in rank, an empirical regularity in best-seller and long-tail
markets [@garg2013inferring]. For the top-paid list, let $d_{r_p}$ denote
downloads at paid rank $r_p$. The Pareto law states

$$
d_{r_p} = b_p\, r_p^{-a_p}, \qquad 1 \le r_p \le 200,
$$ {#eq-pareto-paid}

where $a_p > 0$ is the **shape** parameter (the steepness of the long tail; a
larger $a_p$ means demand falls off faster with rank) and $b_p > 0$ is a
**scale** factor that depends on total market size for the platform (iPhone or
iPad). The truncation at 200 reflects that the published list is finite. Taking
logs of @eq-pareto-paid linearizes it, $\log d_{r_p} = \log b_p - a_p \log r_p$,
so a single list with download data would yield $(a_p, b_p)$ from an ordinary
least squares (OLS) regression of log-downloads on log-rank. The problem is that
download magnitudes are not published; only ranks are. The grossing list
supplies the missing leverage.

#### Identification from the grossing list

Assume the same app obeys a Pareto law in the grossing (revenue) list, with its
own shape $a_g$ and scale $b_g$. For an app at grossing rank $r_g$ selling at
price $p$, revenue equals price times the downloads it earns at its paid rank:

$$
p\, d_{r_p} = b_g\, r_g^{-a_g}.
$$ {#eq-pareto-grossing}

Substituting the paid-list law @eq-pareto-paid for $d_{r_p}$ and taking logs
gives, after rearranging to put grossing rank on the left,

$$
\log r_g
= \frac{1}{a_g}\log\!\left(\frac{b_g}{b_p}\right)
+ \frac{a_p}{a_g}\log r_p
- \frac{1}{a_g}\log p .
$$ {#eq-rank-mapping}

@eq-rank-mapping is linear in observables. It maps an app's paid rank $r_p$ and
price $p$ into its grossing rank $r_g$, and every quantity on the right except
the structural parameters is in the data. It is therefore estimable as the
truncated OLS regression

$$
\log r_g = \beta_0 + \beta_1 \log r_p + \beta_2 \log p + \varepsilon,
$$ {#eq-rank-ols}

where the regression coefficients are nonlinear functions of the structural
parameters. Matching coefficients between @eq-rank-mapping and @eq-rank-ols
recovers the shapes directly:

$$
a_g = -\frac{1}{\beta_2}, \qquad
a_p = -\frac{\beta_1}{\beta_2}, \qquad
\frac{b_g}{b_p} = \exp\!\left(-\frac{\beta_1}{\beta_2}\right).
$$ {#eq-structural-recovery}

The regression identifies the two shape parameters and the *ratio* of scales, but
not the scales themselves, because rank data are invariant to a common rescaling
of all download volumes. To pin down the levels, @garg2013inferring add one
aggregate moment. Summing the paid-list law over all ranked apps in a day equates
the model-implied total to observed aggregate downloads $D_t$:

$$
D_t = \sum_{r_p=1}^{N} d_{r_p} = b_p \sum_{r_p=1}^{N} r_p^{-a_p}.
$$ {#eq-aggregate-downloads}

Because $a_p$ is already known from @eq-structural-recovery, the sum on the right
is a computable constant, so @eq-aggregate-downloads solves for the scale levels:

$$
b_p = \frac{\sum_{r_p=1}^{N} d_{r_p}}{\sum_{r_p=1}^{N} r_p^{-a_p}}, \qquad
b_g = \exp\!\left(-\frac{\beta_0}{\beta_1}\right)
      \cdot \frac{\sum_{r_p=1}^{N} d_{r_p}}{\sum_{r_p=1}^{N} r_p^{-a_p}}.
$$ {#eq-scale-recovery}

With $(a_p, b_p)$ in hand, @eq-pareto-paid returns a download magnitude for every
paid rank, completing the recovery of cardinal demand from ordinal lists.

#### What breaks identification

The estimator rests on assumptions that are explicit and falsifiable, and each
is a potential failure point.

- **Revenue comes only from upfront price.** The grossing equation
  @eq-pareto-grossing treats revenue as price times downloads, ignoring in-app
  purchases. @garg2013inferring argue this is reasonable for *paid* apps, which
  earn most of their money at the point of sale, while in-app monetization is
  concentrated in *free* apps, which the grossing analysis excludes
  [@garg2013inferring, p. 1256]. Where paid apps monetize heavily inside the app,
  the price term in @eq-rank-ols is misspecified and the recovered shapes are
  biased.
- **Rank and price are cross-sectionally independent.** Estimation treats each
  app-day observation as independent even when an app appears across multiple
  days, ignoring the correlation that would arise from app-level unobservables.
  Failing this, the OLS standard errors understate sampling uncertainty.
- **A single Pareto law fits the whole list.** A mixture of populations
  (e.g., games versus utilities) with different tail behavior would violate the
  single-shape assumption, and the linearized regression would average over
  heterogeneous slopes.

@garg2013inferring estimate the model on roughly 200 paid apps, 200 grossing
apps, and their prices over April–May 2011, drawing on Apple's public lists and
third-party trackers of the period.[^datasources] The pipeline from the three
published lists to recovered demand is summarized in @fig-rank-pipeline.

[^datasources]: The original data were assembled from Apple's lists and
contemporaneous trackers including Appshopper (shut down in 2021) and AppAnnie
(now data.ai), with later coverage from providers such as Mobilewalla. The
specific vendors are incidental to the method; any source of ranked lists plus a
single aggregate-download moment suffices.

```{mermaid}
%%| label: fig-rank-pipeline
%%| fig-cap: "Recovering cardinal demand from ordinal App Store lists. Ranked lists and prices identify the Pareto shapes and the scale ratio via regression; one aggregate-download moment pins down the scale levels; the paid-list law then returns downloads at every rank."
flowchart TD
    A[Top-paid ranks r_p] --> D[Regress log r_g on log r_p and log p]
    B[Top-grossing ranks r_g] --> D
    C[Prices p] --> D
    D --> E["Shapes a_p, a_g and scale ratio b_g / b_p"]
    F[Aggregate downloads D_t] --> G[Recover scale levels b_p, b_g]
    E --> G
    G --> H["Pareto law d = b_p r_p^(-a_p): downloads at every rank"]
```

#### A reproducible illustration

The following simulation generates app-level ranks and prices from a known
data-generating process, then recovers the structural parameters through
@eq-rank-ols and @eq-structural-recovery. Because the truth is known, the example
doubles as a check that the estimator is unbiased when its assumptions hold.

```{r rank-demand-sim, message=FALSE, warning=FALSE}
set.seed(12)

# True structural parameters
a_p_true <- 0.80   # paid-list shape
a_g_true <- 0.95   # grossing-list shape
b_p_true <- 5e5    # paid-list scale
b_g_true <- 2e6    # grossing-list scale

N <- 200
r_p   <- 1:N                                  # paid ranks
price <- round(runif(N, 0.99, 9.99), 2)       # app prices

# Downloads from the paid-list Pareto law (eq-pareto-paid)
downloads <- b_p_true * r_p^(-a_p_true)

# Grossing rank implied by the structural mapping (eq-rank-mapping),
# with mild noise standing in for measurement error
log_rg <- (1 / a_g_true) * log(b_g_true / b_p_true) +
          (a_p_true / a_g_true) * log(r_p) -
          (1 / a_g_true) * log(price) +
          rnorm(N, 0, 0.05)
r_g <- exp(log_rg)

# Estimate the truncated OLS regression (eq-rank-ols)
fit <- lm(log(r_g) ~ log(r_p) + log(price))
b   <- coef(fit)

# Recover structural parameters (eq-structural-recovery)
a_g_hat     <- -1 / b[["log(price)"]]
a_p_hat     <- -b[["log(r_p)"]] / b[["log(price)"]]
b_ratio_hat <- exp(-b[["log(r_p)"]] / b[["log(price)"]])

# Recover scale levels from the aggregate-download moment (eq-scale-recovery)
D_t       <- sum(downloads)
b_p_hat   <- D_t / sum(r_p^(-a_p_hat))

data.frame(
  parameter = c("a_p", "a_g", "b_g/b_p", "b_p"),
  truth     = c(a_p_true, a_g_true, b_g_true / b_p_true, b_p_true),
  estimate  = c(a_p_hat, a_g_hat, b_ratio_hat, b_p_hat)
)
```

The recovered shapes and scale track their true values, illustrating that the
ordinal lists, plus one cardinal anchor, suffice to reconstruct the full demand
schedule.

## The Salesperson as an Asset {#sec-salesperson-asset}

The second measurement problem shifts from the product to the person. A sales
force is a portfolio of human assets, and two questions dominate its management:
who leaves, and what is each person worth going forward. Both resist naive
measurement—turnover is contagious in ways that confound individual attribution,
and a salesperson's value is forward-looking while the convenient metrics are
backward-looking.

### Turnover: Own and Peer Effects {#sec-turnover}

Salesperson turnover is expensive: it destroys accumulated account relationships
and product knowledge and imposes replacement costs. Most prior work studied the
*consequences* of voluntary turnover; @sunder2017salespeople instead model its
*antecedents*, and in particular separate two channels that the data tend to
confound.

The first channel is the salesperson's **own** standing. Drawing on identity
theory, a salesperson's role identity is reinforced or threatened by personal
performance, so individual achievement shapes the propensity to stay. The second
channel is **peer** influence. Drawing on social identity theory, a salesperson's
attachment is shaped by the group, so the behavior of peers—above all, **peer
turnover**—moves the individual's own exit decision. The empirical contribution
is to estimate both channels jointly on a panel of 6,727 salespeople observed
over two years, and the central finding is that **peer effects dominate own
effects**: a colleague's departure is a stronger predictor of a salesperson's
exit than that salesperson's own performance.

::: {.callout-note}
## Identification: distinguishing contagion from common shocks
The headline that "peers drive turnover" is a causal claim, and peer effects are
notoriously hard to identify. The **reflection problem** (a peer's outcome
reflects the same group-level shocks that drive ego's outcome) and **homophily**
(similar people self-select into the same teams) both generate peer correlations
with no underlying contagion. Reading a peer-turnover coefficient as a causal
contagion effect requires that team membership and the timing of peer exits be
plausibly exogenous to ego's latent propensity to leave—conditioning on common
shocks, not merely correlating outcomes.
:::

A reduced-form hazard specification makes the two channels concrete. Let the
hazard that salesperson $i$ on team $j$ exits in period $t$ be

$$
h_{ijt} = h_0(t)\,
\exp\!\big(\gamma\, \text{Perf}_{ijt} + \delta\, \text{PeerTurnover}_{jt}
+ \boldsymbol{\theta}^{\top}\mathbf{x}_{ijt}\big),
$$ {#eq-turnover-hazard}

where $h_0(t)$ is a baseline hazard, $\text{Perf}_{ijt}$ captures own performance
(the identity channel), $\text{PeerTurnover}_{jt}$ is recent turnover among $i$'s
peers (the social-identity channel), and $\mathbf{x}_{ijt}$ collects controls.
The @sunder2017salespeople finding is that the estimated peer coefficient
$\hat\delta$ is large relative to the own-performance coefficient $\hat\gamma$.

### Valuing a Salesperson's Future Contribution {#sec-salesperson-value}

Traditional sales-force evaluation is **retrospective**: it ranks salespeople on
realized volume, the metric that is easiest to observe but least informative about
what the firm will earn from each person next year. As marketing has shifted
toward a customer-centric, lifetime-value orientation, the natural analogue for
the sales force is a **forward-looking, profit-based** valuation of each
salesperson—an internal parallel to customer lifetime value.

@kumar2014measuring propose exactly such a metric and embed it in a **latent
class** model that recognizes the sales force is not homogeneous. The intuition
for latent classes is that salespeople fall into a small number of unobserved
**segments** that respond differently to managerial levers, so a single average
response masks divergent effects. Formally, with $C$ latent segments, the
expected future value of salesperson $i$ is the mixture

$$
\widehat{\text{Value}}_i
= \sum_{c=1}^{C} \pi_{ic}\, \mathbb{E}\!\left[V_i \mid \text{segment } c\right],
$$ {#eq-salesperson-value}

where $\pi_{ic}$ is the posterior probability that $i$ belongs to segment $c$
and $\mathbb{E}[V_i \mid \text{segment } c]$ is the discounted future profit
contribution conditional on that segment's response parameters. Both the
segment memberships $\pi_{ic}$ and the segment-specific response parameters are
estimated jointly, typically by maximum likelihood via the
expectation–maximization (EM) algorithm.

The substantive payoff is a precise statement of when "one size fits all" fails.
@kumar2014measuring find that segments respond **differently** to **training**
and to **incentives**, so a uniform program is inefficient: the same training
budget yields different returns across segments, and the ranking of training
versus incentives can even reverse depending on the **time horizon**—an
intervention that looks inferior in the short run may dominate over a longer one.
The forward-looking metric thus reframes sales-force management from rewarding
*past* volume to allocating development resources toward the salespeople and
interventions with the highest *future* return.

@fig-salesperson-value contrasts the retrospective and forward-looking views of
sales-force evaluation.

```{mermaid}
%%| label: fig-salesperson-value
%%| fig-cap: "From retrospective to forward-looking sales-force evaluation. Realized volume looks backward; a latent-class future-value metric segments the sales force and allocates training and incentives to the highest forward return."
flowchart LR
    A[Realized sales volume] --> B[Retrospective ranking]
    C[Salesperson behavior and responses] --> D[Latent-class segmentation]
    D --> E[Segment-specific future value]
    E --> F[Allocate training and incentives by horizon]
    B -. backward-looking .-> G((Evaluation))
    F -. forward-looking .-> G
```

#### A reproducible illustration of latent-class value

The following example simulates two latent segments of salespeople with different
responses to training, fits a two-component mixture by EM, and recovers the
segment-specific effects. It is a deliberately minimal stand-in for the richer
specification in @kumar2014measuring, intended to make the mixture logic of
@eq-salesperson-value concrete.

```{r latent-class-sim, message=FALSE, warning=FALSE}
set.seed(34)

n <- 600
# Two latent segments with different training -> future-value slopes
seg     <- rbinom(n, 1, 0.4)            # segment indicator (unobserved in practice)
train   <- runif(n, 0, 10)             # training hours
slope   <- ifelse(seg == 1, 1.8, 0.3)  # high vs. low responders
value   <- 20 + slope * train + rnorm(n, 0, 4)
dat     <- data.frame(value, train)

# Fit a 2-component mixture of regressions by EM (base R)
em_mixreg <- function(y, x, K = 2, iters = 100) {
  n <- length(y)
  pi_k  <- rep(1 / K, K)
  beta0 <- quantile(y, c(0.3, 0.7)); beta1 <- c(0.3, 1.5); sig <- rep(sd(y), K)
  for (it in seq_len(iters)) {
    # E-step: responsibilities
    dens <- sapply(1:K, function(k)
      pi_k[k] * dnorm(y, beta0[k] + beta1[k] * x, sig[k]))
    r <- dens / rowSums(dens)
    # M-step: weighted regressions
    for (k in 1:K) {
      w  <- r[, k]
      fk <- lm(y ~ x, weights = w)
      beta0[k] <- coef(fk)[1]; beta1[k] <- coef(fk)[2]
      sig[k]   <- sqrt(sum(w * residuals(fk)^2) / sum(w))
      pi_k[k]  <- mean(w)
    }
  }
  list(pi = pi_k, slope = beta1, intercept = beta0)
}

out <- em_mixreg(dat$value, dat$train, K = 2)
data.frame(
  segment        = c(1, 2),
  mix_weight     = round(out$pi, 3),
  training_slope = round(out$slope, 3)
)
```

The estimator separates the high-responding segment (training slope near 1.8)
from the low-responding one (near 0.3), reproducing the qualitative lesson of
@kumar2014measuring: returns to a managerial lever are segment-specific, and the
forward-looking value in @eq-salesperson-value averages over those segments
weighted by membership.

## Key Takeaways

- Variable compensation lifts sales but carries a **health cost**—stress,
  exhaustion, absenteeism—that is buffered by individual ability and social
  resources, so the profit-maximizing incentive slope is lower and more
  differentiated than the standard agency model implies [@habel2021].
- **Sales rank is ordinal**; recovering cardinal demand requires either an
  experimental demand shock [@chevalier2003measuring; @he2020sales], auxiliary
  demand data [@brynjolfsson2003consumer], or a parametric demand law identified
  from multiple ranked lists [@garg2013inferring].
- The Pareto rank–demand model (@eq-pareto-paid–@eq-scale-recovery) identifies
  tail **shapes** and a **scale ratio** from rank regressions, but needs one
  aggregate-download moment to fix scale **levels**; its identification breaks if
  revenue is not upfront, if cross-sectional independence fails, or if a single
  power law does not fit the list.
- Salesperson **turnover is contagious**: peer turnover predicts individual exit
  more strongly than own performance, but reading this as causal contagion
  demands care with the reflection problem and homophily
  [@sunder2017salespeople].
- A **forward-looking, latent-class** valuation of the sales force
  (@eq-salesperson-value) shows that returns to training and incentives are
  segment- and horizon-specific, so uniform development programs are inefficient
  [@kumar2014measuring].
