# Sales {#sec-sales}

Selling is where a firm's marketing promises are tested against a buyer's
willingness to pay. This chapter treats the sales function as two distinct but
connected measurement problems. The first is **inference about demand from
observed sales**: firms and researchers rarely see the full demand curve, but
they do see rankings, units, and revenue, and a recurring task is to reconstruct
latent demand from these partial signals. The second is **management of the sales
force**: the salespeople who convert demand into revenue are themselves an asset
whose compensation, productivity, retention, and future value can be modeled,
measured, and optimized.

The two problems share a methodological spine. In both, the object of interest
is unobserved—true demand in the first case, a salesperson's forward-looking
profit contribution in the second—and must be recovered from a noisy,
selection-prone proxy. A reader who finishes this chapter should be able to
estimate a demand curve from a sales-rank list, reason about when that estimate
breaks down, and evaluate a sales-force compensation or retention policy using
the same identification discipline applied throughout the book.

We proceed from the outside in. We begin with the economics of compensation and
its hidden health costs, because incentives are the lever managers pull most
often. We then turn to **sales from rank**, the empirical problem of inferring
demand from ordinal best-seller lists, and develop the Pareto rank–demand model
in full. We continue with the **salesperson as an asset**: turnover and its
contagion through peer networks, and the forward-looking valuation of a sales
force. We close by carrying the same logic into a setting with no price on either
side of the exchange—**nonprofit fundraising**—where allocating scarce
solicitation labor across methods is the sales-portfolio problem stripped to its
essentials.

One boundary condition frames all of it. The occupation this chapter models is
being restructured by digital tools that automate qualification, prospecting, and
follow-up, which changes both what a salesperson does and which parts of the job the
firm can observe [@mullins2019salesprofessio]. Technology adoption at the individual
level has measurable performance consequences, though they run through mediating
mechanisms — adaptive selling and effort — rather than directly
[@ahearne2008hightouch]. Where customer orientation itself becomes counterproductive
is an empirical question with an interior answer: @homburg2011whenshould find an
inverted-U, so the optimum is a calibrated level rather than a maximum.

## Theoretical Foundations {#sec-sales-theory}

The sales function is governed by a small set of theories that explain both the
buyer-seller relationship and the firm's control of its own sales force. Four
strands organize the chapter.

**Social exchange theory** frames selling as a relationship rather than a sequence of
isolated transactions. Parties continue an exchange when its rewards exceed its costs
relative to available alternatives, and recurring exchange builds norms of
reciprocity and mutual obligation. This is the lens that justifies treating the
salesperson-customer dyad (and, internally, the salesperson-firm dyad) as an evolving
relationship whose quality, not merely its current price, determines future revenue.

**The commitment-trust theory of relationship marketing** specifies which mediating
variables make that exchange succeed. Morgan and Hunt argue that relationship
*commitment* and *trust* are the central mediators between the conditions of an
exchange (shared values, communication, the absence of opportunism) and cooperative
outcomes such as loyalty and acquiescence [@morganhunt1994]. For sales, this explains
why relationship quality predicts retention and why opportunistic behavior, on either
side, is so corrosive to long-run value.

**Agency theory** governs salesforce control and underlies the compensation analysis
of @sec-sales, "Compensation and Its Hidden Costs." Because the firm (principal)
cannot perfectly observe the salesperson's (agent's) effort and the two have
divergent goals, it must choose between *behavior-based* control (salary, monitoring,
supervision) and *outcome-based* control (commission tied to noisy sales). The linear
contract $w_i = \alpha + \beta s_i$ formalizes exactly this trade-off: a steeper
$\beta$ aligns incentives but shifts risk onto a risk-averse agent, which, together
with the health externality documented in this chapter, is why the profit-maximizing
incentive slope is interior, not maximal. The governance-cost logic connects to the
broader transaction-cost account of when firms integrate versus outsource selling
[@williamson1981tce].

**Adaptive selling** describes the behavior the contract is meant to elicit. It is
the practice of altering sales behavior across and within customer encounters in
response to information about the selling situation, the antithesis of a fixed,
canned pitch. Adaptive selling theory predicts that the value of discretion (and
therefore of training and information systems that support it) rises with the
heterogeneity of customers, which is why the latent-class, segment-specific returns
to training in @sec-salesperson-value are exactly what the theory anticipates. What
adaptation looks like in practice has been unpacked in two directions: the influence
tactics a salesperson selects, and how well they read the situation that should
govern the selection [@mcfarland2006influencetactics]. Perceived listening is the
under-modeled half of this — customers infer effort and competence from whether they
believe they were heard, largely independent of what was subsequently proposed
[@sohi1997listeningyour] — and the buyer's inference about *motive* can undo an
otherwise well-adapted message, since a pitch that triggers suspicion of ulterior
motive is discounted rather than merely disbelieved [@decarlo2005salesmessage].

The synthesis that pulled these strands into one agenda is
@weitz1999personalselling, which reframed personal selling from a transaction skill
into the firm's principal relationship-management instrument; the practitioner-facing
framework in @zoltners2008salesforce maps the same territory onto the decisions a
sales organization actually makes about sizing, structure, and deployment.

## The Sales Conversation: Talk as the Instrument {#sec-sales-conversation}

Adaptive selling names the behavior but does not say what the behavior is made of. The
unit adaptation actually operates on is the *speaking turn*, and until conversations
could be transcribed and acoustically analyzed at scale the turn was invisible to
measurement. @balducci2026leads open it up for the prospecting call—the stage of B2B
selling at which a lead either becomes a live opportunity or is lost.

Their framework, built from symbolic interactionism and a grounded analysis of the calls
themselves, decomposes salesperson speech into two channels. **Linguistic acts** are what
is said: *legitimizing* (establishing the standing of the caller and the call), *piquing*
(creating interest), and *nudging* (moving the lead toward commitment). **Paralinguistic
acts** are how it is said—vocal intensity, speech rate, and pitch, the same acoustic
features @sec-audio-prosody treats as measures of a construct rather than as bookkeeping
quantities. Between the acts and the outcome sits **lead receptivity**, a turn-level
indicator of the lead's openness to the message. That construct is the paper's
methodological move as much as its theoretical one: it turns a conversation into a panel
instead of a single observation.

The estimates come from 783 lead-calling conversations comprising 11,566 speaking turns.
Receptivity earns its place in the chain—a one-standard-deviation increase raises
conversion likelihood by **13.44%**—but the more useful results are the interactions with
position in the call, which show that communicative acts are not uniformly effective but
*positionally* so. A more specific nudge raises receptivity by **11.82%** early in a
conversation and lowers it by **81.40%** later in the same conversation. Speaking faster
is roughly neutral early and costs **8.26%** late.

The sign flip, and the order-of-magnitude asymmetry around it, is the finding worth
carrying away. A conventional analysis regressing conversion on the *rate* of specific
nudges per call would average an early-call gain against a late-call collapse many times
larger, recover something near zero, and conclude that nudging does not matter. The
lesson generalizes well past sales: when a treatment is delivered repeatedly inside an
episode and its effect depends on where in the episode it lands, the episode-level
aggregate is not a summary of the turn-level effect—it is a different estimand, and
usually an uninterpretable one. This is the within-episode timing problem that
@mcfarland2006influencetactics posed for influence tactics and that the frontline
literature meets in @sec-frontline; what has changed is that transcription and acoustic
feature extraction (@sec-audio-voice, @sec-tad-applications) make the turn observable, so
the timing can be estimated rather than assumed.

The managerial reading is uncomfortably specific. "Be concrete" and "keep the energy up"
are not wrong as coaching, they are *unconditioned*—instructions about content where the
evidence says the binding constraint is timing. A script that deploys them uniformly
across a call collects the early gain and then pays the late penalty.

## Compensation and Its Hidden Costs

The canonical justification for putting variable pay—commissions, bonuses,
quota-based incentives—into a salesperson's compensation contract is an
agency-theoretic one. The firm (principal) cannot perfectly observe the agent's
effort, only the noisy output (sales) that effort produces. Tying pay to output
realigns the agent's incentives with the firm's, and the standard prediction is
that a steeper pay–performance slope elicits more effort and higher sales. This
prediction is robust and well documented.

Compensation is only one of the control instruments a sales organization holds,
and the others are exercised by supervisors rather than by contracts.
@challagalla1996dimensiontypes separate supervisory control into activity, capability,
and outcome varieties and show they act on performance and satisfaction through
different routes — a distinction the single behavior-versus-outcome dichotomy
obscures. Leadership style is a further margin: servant leadership raises
salesperson performance through the trust and commitment it builds rather than
through effort extraction [@chonko2009examiningimpact]. And the incentive climate
interacts with disposition. Trait-competitive salespeople respond to intra-firm
competition in ways their less competitive colleagues do not
[@cron1998traitcompetiti], which is one reason a single contest design produces such
heterogeneous responses across a sales force. Discretionary effort outside the formal
role — the in-role and extra-role behaviors no contract specifies — is what much of
this control apparatus is ultimately trying to reach
[@ahearne1998somepossible].

What the classical account omits is that effort is not free of physiological
cost. @habel2021 confirm the performance benefit of variable compensation but
show that the same incentive intensity that lifts sales also operates as a
*health hazard*: it raises stress, which manifests as emotional exhaustion and a
higher incidence of sick days. The effect is not uniform across the sales force.
It is **buffered** for salespeople with higher personal ability and richer
social resources—the people best equipped to absorb the additional psychological
load—and sharpest for those without such buffers.

The managerial implication is that the pay–performance slope has a cost the
standard model prices at zero. Let $w_i = \alpha + \beta\, s_i$ denote a linear
contract for salesperson $i$, with fixed component $\alpha$, variable rate
$\beta$, and realized sales $s_i$. The firm's familiar problem is to choose
$\beta$ to maximize expected profit net of wage cost. @habel2021 effectively add
a term: the agent's stress, and hence absenteeism and exhaustion, is increasing
in $\beta$, so the firm faces a more accurate objective in which $\beta$ carries
a shadow cost through reduced effective capacity and higher turnover risk. The
optimal incentive intensity that internalizes this health externality is lower
than the one that ignores it, and it varies with the ability and social
resources of the individual—an argument for incentive schemes that are
differentiated rather than uniform.

::: {.callout-note}
The health cost of incentives is a *moderated* effect, not a main-effect
caveat. Reporting only the average performance gain from steeper incentives, as
much of the older literature does, overstates the net benefit for the segment of
the sales force least able to bear the stress.
:::

The hidden cost need not run through health, and it need not require a contract at
all. Firms also reward workers *discretionarily*—an unexpected bonus, a wage set
above what the market requires—on a logic that comes not from agency theory but
from its behavioral rival, the **gift-exchange** (or efficiency-wage) hypothesis:
treat the employment relation as a reciprocal exchange of gifts, pay the worker
more than you must, and the worker returns the kindness with effort the contract
cannot compel. Much of the case for discretionary sales bonuses and "we look after
our people" pay rests, implicitly, on this reciprocity. Yet when
@bogliacino2026gifts put the logic to a clean test in a natural field experiment,
the gift went unreturned. Workers hired for a data-entry task, paired and observed
over an initial round, then had a discretionary bonus paid to one or both of them,
assigned on one of three grounds—relative productivity, economic need, or an
avowedly arbitrary draw. Contrary to the gift-exchange prediction, post-bonus
output *fell*: a statistically significant decline on the order of 8 to 15 percent
across their two experiments, and largest when both members of a pair were paid.
The bonus, it seems, was read as a *signal* rather than a gift—receiving it told
workers the employer was already content, which both lowered the perceived risk of
dismissal for slacking and dissolved the sense of obligation the gift was meant to
create. Reciprocity ran backward.

The lesson reaches past data entry to any discretionary bonus, the salesforce's
included. A payment meant as goodwill is not a costless lever: whether it buys
effort or licenses coasting depends on what the recipient infers from receiving
it—that they are valued and should reciprocate, or that their position is secure
and they can ease off. The agency model prices this signaling channel at zero, but
the field evidence says it can be large enough to reverse the intended effect. The
implication is not that bonuses fail, but that a discretionary payment
communicates, and what it communicates can as easily be *you may relax* as *do more
of this*.

## Sales from Rank {#sec-sales-from-rank}

Many of the most visible signals of commercial success are **ordinal**, not
cardinal. Amazon publishes a sales rank for every book; Apple's App Store
publishes ranked lists rather than download counts; the music, film, and
publishing industries all live by charts. A rank is informative—rank 1 outsells
rank 100—but it discards the magnitudes a firm actually needs: how *many* more
units does rank 1 move than rank 100? Recovering demand magnitudes from rank is
the central inference problem of this section.

The difficulty is that the map from demand to rank is many-to-one and unknown.
Two products can swap ranks after a small change in either one's sales, and the
same rank can correspond to wildly different unit volumes in a thick versus a
thin market. The literature solves this in three stages of increasing
ambition, summarized in @tbl-rank-approaches: experimentally perturb rank and
read off the demand response; supplement rank with auxiliary demand data; or
impose a parametric demand–rank law and estimate it from rank data alone.

| Study | Setting | Demand data required | Core idea |
|---|---|---|---|
| @chevalier2003measuring | Amazon books | Known/very low for chosen titles | Buy units of low-selling books; watch rank move; trace the rank–demand map |
| @brynjolfsson2003consumer | Amazon books | Publisher demand data | Calibrate the rank–demand relationship directly |
| @garg2013inferring | Apple App Store | None | Impose a Pareto law; identify it from paid and grossing ranks |
| @he2020sales | Amazon (general) | None | Generalize @chevalier2003measuring beyond books |

: Approaches to inferring demand from sales rank {#tbl-rank-approaches}

### The Experimental Approach

The cleanest identification comes from intervention. @chevalier2003measuring run
a field experiment on Amazon: they select books whose demand is either known or
known to be very small, purchase large quantities relative to that baseline
demand, and observe how each title's sales rank responds to the induced demand
shock. Because the experimenter controls the size of the shock, the otherwise
unobservable mapping between a change in units and a change in rank is revealed.
Repeating this across the rank distribution traces out the rank–demand curve.

The method's strength—a controlled demand shock—is also its limitation. It is
practical only for **low-selling** titles, where a feasible experimental purchase
is large relative to organic demand. At the head of the distribution, the
purchase required to move rank by a perceptible amount is prohibitive, so the
high-volume region of the curve cannot be identified experimentally. The
generalization of this revealed-demand logic to Amazon products beyond books is
developed in @he2020sales.

### The Auxiliary-Data Approach

When the experimenter cannot perturb demand, the next-best instrument is
external demand data. @brynjolfsson2003consumer obtain demand figures from a book
publisher and pair them with observed ranks, which pins down the rank–demand
relationship by direct calibration rather than by experiment. This dispenses with
the need to intervene but substitutes a different scarce resource—proprietary
unit-sales data—that most analysts cannot obtain.

### The Structural Approach: A Pareto Rank–Demand Law

The most general method dispenses with both intervention and demand data,
recovering magnitudes from rank alone by imposing structure on the demand
distribution. @garg2013inferring develop this approach on Apple's App Store,
which is an unusually favorable laboratory because it publishes three distinct
ranked lists that, jointly, over-identify the model:

1. **Top-free apps** — ranked by download volume among zero-price apps.
2. **Top-paid apps** — ranked by download volume among positive-price apps.
3. **Top-grossing apps** — ranked by *revenue*, i.e., price times downloads.

The key modeling assumption is that downloads follow a **Pareto** (power-law)
distribution in rank, an empirical regularity in best-seller and long-tail
markets [@garg2013inferring]. For the top-paid list, let $d_{r_p}$ denote
downloads at paid rank $r_p$. The Pareto law states

$$
d_{r_p} = b_p\, r_p^{-a_p}, \qquad 1 \le r_p \le 200,
$$ {#eq-pareto-paid}

where $a_p > 0$ is the **shape** parameter (the steepness of the long tail; a
larger $a_p$ means demand falls off faster with rank) and $b_p > 0$ is a
**scale** factor that depends on total market size for the platform (iPhone or
iPad). The truncation at 200 reflects that the published list is finite. Taking
logs of @eq-pareto-paid linearizes it, $\log d_{r_p} = \log b_p - a_p \log r_p$,
so a single list with download data would yield $(a_p, b_p)$ from an ordinary
least squares (OLS) regression of log-downloads on log-rank. The problem is that
download magnitudes are not published; only ranks are. The grossing list
supplies the missing leverage.

#### Identification from the grossing list

Assume the same app obeys a Pareto law in the grossing (revenue) list, with its
own shape $a_g$ and scale $b_g$. For an app at grossing rank $r_g$ selling at
price $p$, revenue equals price times the downloads it earns at its paid rank:

$$
p\, d_{r_p} = b_g\, r_g^{-a_g}.
$$ {#eq-pareto-grossing}

Substituting the paid-list law @eq-pareto-paid for $d_{r_p}$ and taking logs
gives, after rearranging to put grossing rank on the left,

$$
\log r_g
= \frac{1}{a_g}\log\!\left(\frac{b_g}{b_p}\right)
+ \frac{a_p}{a_g}\log r_p
- \frac{1}{a_g}\log p .
$$ {#eq-rank-mapping}

@eq-rank-mapping is linear in observables. It maps an app's paid rank $r_p$ and
price $p$ into its grossing rank $r_g$, and every quantity on the right except
the structural parameters is in the data. It is therefore estimable as the
truncated OLS regression

$$
\log r_g = \beta_0 + \beta_1 \log r_p + \beta_2 \log p + \varepsilon,
$$ {#eq-rank-ols}

where the regression coefficients are nonlinear functions of the structural
parameters. Matching coefficients between @eq-rank-mapping and @eq-rank-ols
recovers the shapes directly:

$$
a_g = -\frac{1}{\beta_2}, \qquad
a_p = -\frac{\beta_1}{\beta_2}, \qquad
\frac{b_g}{b_p} = \exp\!\left(-\frac{\beta_1}{\beta_2}\right).
$$ {#eq-structural-recovery}

The regression identifies the two shape parameters and the *ratio* of scales, but
not the scales themselves, because rank data are invariant to a common rescaling
of all download volumes. To pin down the levels, @garg2013inferring add one
aggregate moment. Summing the paid-list law over all ranked apps in a day equates
the model-implied total to observed aggregate downloads $D_t$:

$$
D_t = \sum_{r_p=1}^{N} d_{r_p} = b_p \sum_{r_p=1}^{N} r_p^{-a_p}.
$$ {#eq-aggregate-downloads}

Because $a_p$ is already known from @eq-structural-recovery, the sum on the right
is a computable constant, so @eq-aggregate-downloads solves for the scale levels:

$$
b_p = \frac{\sum_{r_p=1}^{N} d_{r_p}}{\sum_{r_p=1}^{N} r_p^{-a_p}}, \qquad
b_g = \exp\!\left(-\frac{\beta_0}{\beta_1}\right)
      \cdot \frac{\sum_{r_p=1}^{N} d_{r_p}}{\sum_{r_p=1}^{N} r_p^{-a_p}}.
$$ {#eq-scale-recovery}

With $(a_p, b_p)$ in hand, @eq-pareto-paid returns a download magnitude for every
paid rank, completing the recovery of cardinal demand from ordinal lists.

#### What breaks identification

The estimator rests on assumptions that are explicit and falsifiable, and each
is a potential failure point.

- **Revenue comes only from upfront price.** The grossing equation
  @eq-pareto-grossing treats revenue as price times downloads, ignoring in-app
  purchases. @garg2013inferring argue this is reasonable for *paid* apps, which
  earn most of their money at the point of sale, while in-app monetization is
  concentrated in *free* apps, which the grossing analysis excludes
  [@garg2013inferring, p. 1256]. Where paid apps monetize heavily inside the app,
  the price term in @eq-rank-ols is misspecified and the recovered shapes are
  biased.
- **Rank and price are cross-sectionally independent.** Estimation treats each
  app-day observation as independent even when an app appears across multiple
  days, ignoring the correlation that would arise from app-level unobservables.
  Failing this, the OLS standard errors understate sampling uncertainty.
- **A single Pareto law fits the whole list.** A mixture of populations
  (e.g., games versus utilities) with different tail behavior would violate the
  single-shape assumption, and the linearized regression would average over
  heterogeneous slopes.

@garg2013inferring estimate the model on roughly 200 paid apps, 200 grossing
apps, and their prices over April–May 2011, drawing on Apple's public lists and
third-party trackers of the period.[^datasources] The pipeline from the three
published lists to recovered demand is summarized in @fig-rank-pipeline.

[^datasources]: The original data were assembled from Apple's lists and
contemporaneous trackers including Appshopper (shut down in 2021) and AppAnnie
(now data.ai), with later coverage from providers such as Mobilewalla. The
specific vendors are incidental to the method; any source of ranked lists plus a
single aggregate-download moment suffices.

```{mermaid}
%%| label: fig-rank-pipeline
%%| fig-cap: "Recovering cardinal demand from ordinal App Store lists. Ranked lists and prices identify the Pareto shapes and the scale ratio via regression; one aggregate-download moment pins down the scale levels; the paid-list law then returns downloads at every rank."
flowchart TD
    A[Top-paid ranks r_p] --> D[Regress log r_g on log r_p and log p]
    B[Top-grossing ranks r_g] --> D
    C[Prices p] --> D
    D --> E["Shapes a_p, a_g and scale ratio b_g / b_p"]
    F[Aggregate downloads D_t] --> G[Recover scale levels b_p, b_g]
    E --> G
    G --> H["Pareto law d = b_p r_p^(-a_p): downloads at every rank"]
```

#### A reproducible illustration

The following simulation generates app-level ranks and prices from a known
data-generating process, then recovers the structural parameters through
@eq-rank-ols and @eq-structural-recovery. Because the truth is known, the example
doubles as a check that the estimator is unbiased when its assumptions hold.

```{r rank-demand-sim, message=FALSE, warning=FALSE}
set.seed(12)

# True structural parameters
a_p_true <- 0.80   # paid-list shape
a_g_true <- 0.95   # grossing-list shape
b_p_true <- 5e5    # paid-list scale
b_g_true <- 2e6    # grossing-list scale

N <- 200
r_p   <- 1:N                                  # paid ranks
price <- round(runif(N, 0.99, 9.99), 2)       # app prices

# Downloads from the paid-list Pareto law (eq-pareto-paid)
downloads <- b_p_true * r_p^(-a_p_true)

# Grossing rank implied by the structural mapping (eq-rank-mapping),
# with mild noise standing in for measurement error
log_rg <- (1 / a_g_true) * log(b_g_true / b_p_true) +
          (a_p_true / a_g_true) * log(r_p) -
          (1 / a_g_true) * log(price) +
          rnorm(N, 0, 0.05)
r_g <- exp(log_rg)

# Estimate the truncated OLS regression (eq-rank-ols)
fit <- lm(log(r_g) ~ log(r_p) + log(price))
b   <- coef(fit)

# Recover structural parameters (eq-structural-recovery)
a_g_hat     <- -1 / b[["log(price)"]]
a_p_hat     <- -b[["log(r_p)"]] / b[["log(price)"]]
b_ratio_hat <- exp(-b[["log(r_p)"]] / b[["log(price)"]])

# Recover scale levels from the aggregate-download moment (eq-scale-recovery)
D_t       <- sum(downloads)
b_p_hat   <- D_t / sum(r_p^(-a_p_hat))

data.frame(
  parameter = c("a_p", "a_g", "b_g/b_p", "b_p"),
  truth     = c(a_p_true, a_g_true, b_g_true / b_p_true, b_p_true),
  estimate  = c(a_p_hat, a_g_hat, b_ratio_hat, b_p_hat)
)
```

The recovered shapes and scale track their true values, illustrating that the
ordinal lists, plus one cardinal anchor, suffice to reconstruct the full demand
schedule.

## The Salesperson as an Asset {#sec-salesperson-asset}

The second measurement problem shifts from the product to the person. A sales
force is a portfolio of human assets, and two questions dominate its management:
who leaves, and what is each person worth going forward. Both resist naive
measurement—turnover is contagious in ways that confound individual attribution,
and a salesperson's value is forward-looking while the convenient metrics are
backward-looking.

### Turnover: Own and Peer Effects {#sec-turnover}

Salesperson turnover is expensive: it destroys accumulated account relationships
and product knowledge and imposes replacement costs. Most prior work studied the
*consequences* of voluntary turnover; @sunder2017salespeople instead model its
*antecedents*, and in particular separate two channels that the data tend to
confound.

The first channel is the salesperson's **own** standing. Drawing on identity
theory, a salesperson's role identity is reinforced or threatened by personal
performance, so individual achievement shapes the propensity to stay. The second
channel is **peer** influence. Drawing on social identity theory, a salesperson's
attachment is shaped by the group, so the behavior of peers—above all, **peer
turnover**—moves the individual's own exit decision. That peer structure is itself
measurable rather than assumed: mapping the intra-organizational network shows which
positions confer performance advantages and how those positions are acquired, which
supplies the microfoundation the contagion story needs
[@bolander2015socialnetworks]. The same logic explains how a market orientation
propagates through a firm at all — not by proclamation but by social learning across
exactly these ties [@lam2010diffusionmarket]. The empirical contribution
is to estimate both channels jointly on a panel of 6,727 salespeople observed
over two years, and the central finding is that **peer effects dominate own
effects**: a colleague's departure is a stronger predictor of a salesperson's
exit than that salesperson's own performance.

::: {.callout-note}
## Identification: distinguishing contagion from common shocks
The headline that "peers drive turnover" is a causal claim, and peer effects are
notoriously hard to identify. The **reflection problem** (a peer's outcome
reflects the same group-level shocks that drive ego's outcome) and **homophily**
(similar people self-select into the same teams) both generate peer correlations
with no underlying contagion. Reading a peer-turnover coefficient as a causal
contagion effect requires that team membership and the timing of peer exits be
plausibly exogenous to ego's latent propensity to leave—conditioning on common
shocks, not merely correlating outcomes.
:::

A reduced-form hazard specification makes the two channels concrete. Let the
hazard that salesperson $i$ on team $j$ exits in period $t$ be

$$
h_{ijt} = h_0(t)\,
\exp\!\big(\gamma\, \text{Perf}_{ijt} + \delta\, \text{PeerTurnover}_{jt}
+ \boldsymbol{\theta}^{\top}\mathbf{x}_{ijt}\big),
$$ {#eq-turnover-hazard}

where $h_0(t)$ is a baseline hazard, $\text{Perf}_{ijt}$ captures own performance
(the identity channel), $\text{PeerTurnover}_{jt}$ is recent turnover among $i$'s
peers (the social-identity channel), and $\mathbf{x}_{ijt}$ collects controls.
The @sunder2017salespeople finding is that the estimated peer coefficient
$\hat\delta$ is large relative to the own-performance coefficient $\hat\gamma$.

### Valuing a Salesperson's Future Contribution {#sec-salesperson-value}

Traditional sales-force evaluation is **retrospective**: it ranks salespeople on
realized volume, the metric that is easiest to observe but least informative about
what the firm will earn from each person next year. As marketing has shifted
toward a customer-centric, lifetime-value orientation, the natural analogue for
the sales force is a **forward-looking, profit-based** valuation of each
salesperson—an internal parallel to customer lifetime value.

@kumar2014measuring propose exactly such a metric and embed it in a **latent
class** model that recognizes the sales force is not homogeneous. The intuition
for latent classes is that salespeople fall into a small number of unobserved
**segments** that respond differently to managerial levers, so a single average
response masks divergent effects. Formally, with $C$ latent segments, the
expected future value of salesperson $i$ is the mixture

$$
\widehat{\text{Value}}_i
= \sum_{c=1}^{C} \pi_{ic}\, \mathbb{E}\!\left[V_i \mid \text{segment } c\right],
$$ {#eq-salesperson-value}

where $\pi_{ic}$ is the posterior probability that $i$ belongs to segment $c$
and $\mathbb{E}[V_i \mid \text{segment } c]$ is the discounted future profit
contribution conditional on that segment's response parameters. Both the
segment memberships $\pi_{ic}$ and the segment-specific response parameters are
estimated jointly, typically by maximum likelihood via the
expectation–maximization (EM) algorithm.

The substantive payoff is a precise statement of when "one size fits all" fails.
@kumar2014measuring find that segments respond **differently** to **training**
and to **incentives**, so a uniform program is inefficient: the same training
budget yields different returns across segments, and the ranking of training
versus incentives can even reverse depending on the **time horizon**—an
intervention that looks inferior in the short run may dominate over a longer one.
The forward-looking metric thus reframes sales-force management from rewarding
*past* volume to allocating development resources toward the salespeople and
interventions with the highest *future* return.

@fig-salesperson-value contrasts the retrospective and forward-looking views of
sales-force evaluation.

```{mermaid}
%%| label: fig-salesperson-value
%%| fig-cap: "From retrospective to forward-looking sales-force evaluation. Realized volume looks backward; a latent-class future-value metric segments the sales force and allocates training and incentives to the highest forward return."
flowchart LR
    A[Realized sales volume] --> B[Retrospective ranking]
    C[Salesperson behavior and responses] --> D[Latent-class segmentation]
    D --> E[Segment-specific future value]
    E --> F[Allocate training and incentives by horizon]
    B -. backward-looking .-> G((Evaluation))
    F -. forward-looking .-> G
```

#### A reproducible illustration of latent-class value

The following example simulates two latent segments of salespeople with different
responses to training, fits a two-component mixture by EM, and recovers the
segment-specific effects. It is a deliberately minimal stand-in for the richer
specification in @kumar2014measuring, intended to make the mixture logic of
@eq-salesperson-value concrete.

```{r latent-class-sim, message=FALSE, warning=FALSE}
set.seed(34)

n <- 600
# Two latent segments with different training -> future-value slopes
seg     <- rbinom(n, 1, 0.4)            # segment indicator (unobserved in practice)
train   <- runif(n, 0, 10)             # training hours
slope   <- ifelse(seg == 1, 1.8, 0.3)  # high vs. low responders
value   <- 20 + slope * train + rnorm(n, 0, 4)
dat     <- data.frame(value, train)

# Fit a 2-component mixture of regressions by EM (base R)
em_mixreg <- function(y, x, K = 2, iters = 100) {
  n <- length(y)
  pi_k  <- rep(1 / K, K)
  beta0 <- quantile(y, c(0.3, 0.7)); beta1 <- c(0.3, 1.5); sig <- rep(sd(y), K)
  for (it in seq_len(iters)) {
    # E-step: responsibilities
    dens <- sapply(1:K, function(k)
      pi_k[k] * dnorm(y, beta0[k] + beta1[k] * x, sig[k]))
    r <- dens / rowSums(dens)
    # M-step: weighted regressions
    for (k in 1:K) {
      w  <- r[, k]
      fk <- lm(y ~ x, weights = w)
      beta0[k] <- coef(fk)[1]; beta1[k] <- coef(fk)[2]
      sig[k]   <- sqrt(sum(w * residuals(fk)^2) / sum(w))
      pi_k[k]  <- mean(w)
    }
  }
  list(pi = pi_k, slope = beta1, intercept = beta0)
}

out <- em_mixreg(dat$value, dat$train, K = 2)
data.frame(
  segment        = c(1, 2),
  mix_weight     = round(out$pi, 3),
  training_slope = round(out$slope, 3)
)
```

The estimator separates the high-responding segment (training slope near 1.8)
from the low-responding one (near 0.3), reproducing the qualitative lesson of
@kumar2014measuring: returns to a managerial lever are segment-specific, and the
forward-looking value in @eq-salesperson-value averages over those segments
weighted by membership.

## Selling Without a Price: Fundraising Productivity {#sec-fundraising-productivity}

Not every sales force sells. Nonprofit fundraisers solicit gifts rather than close
transactions, and the exchange they manage has no price on either side: the donor
receives no product, and the organization posts no margin. Yet the managerial
problem is the sales problem in all its essentials. A scarce pool of skilled labor
must be allocated across solicitation methods that differ in cost, cycle length,
and relational intensity; performance must be measured against that labor
investment rather than against the raw dollars it produced; and the allocation
decision is a portfolio choice whose payoff depends on the organization's stage of
development. This section takes the nonprofit case as the cleanest available
laboratory for a question the commercial literature usually confounds with pricing
and product: **how concentrated should a revenue-generating function be across the
methods available to it?**

### Productivity Versus Effectiveness {#sec-fundraising-denominator}

The fundraising literature has long evaluated performance by the numerator alone.
Total dollars raised measures **effectiveness**; the cost-to-raise-a-dollar ratio
measures financial efficiency but treats labor as an expense line rather than as
the capacity constraint it actually is. Neither answers the question a development
director faces, which is what the next fundraiser is worth. **Fundraising
productivity** puts labor in the denominator directly,

$$
\text{Productivity}_{it} \;=\; \frac{\text{Fundraising revenue}_{it}}{\text{Fundraising FTE}_{it}},
$$ {#eq-fundraising-productivity}

for organization $i$ in year $t$, where FTE is full-time-equivalent headcount
dedicated to fundraising. The choice of denominator is not cosmetic. A strategy
that raises total revenue by hiring is effective but need not be productive, and
the two rankings of the same organizations can differ sharply. @han2026fundraising
make the return on labor investment the object of study for exactly this reason:
in a labor-intensive function, revenue per FTE is the quantity a
capacity-constrained manager can actually act on.

### The Concentration Decision {#sec-fundraising-concentration}

Fundraising revenue arrives through distinct **solicitation methods**: direct mail,
special events, grant writing, corporate and foundation appeals, online giving,
planned giving, and face-to-face personal solicitation. Let $s_{ijt}$ be the share
of organization $i$'s fundraising revenue in year $t$ arriving through method $j$.
The degree of concentration is the Herfindahl form of @sec-diversity applied to
the organization's own revenue portfolio rather than to an industry,

$$
H_{it} \;=\; \sum_j s_{ijt}^2, \qquad \tfrac{1}{J} \le H_{it} \le 1,
$$ {#eq-fundraising-hhi}

with $H_{it} = 1$ when everything comes through a single method and $1/J$ when the
$J$ methods contribute equally. Diversification is $1 - H_{it}$, or an entropy
analogue when the researcher prefers to weight small shares more heavily.

The literature's prior on the sign of $H$ was genuinely divided. The
portfolio-theoretic argument favors diversification: independent revenue streams
smooth shocks, and a nonprofit that depends on one channel is hostage to it. The
capability argument favors concentration: solicitation methods have steep learning
curves and method-specific relational capital, so spreading a small team thin
forfeits the depth that closes large gifts. Because the two arguments live at
different levels, risk versus capability, the empirical literature reported
conflicting signs without resolving them, and offered little guidance conditional
on organizational context.

### What the Food-Bank Evidence Shows {#sec-fundraising-evidence}

@han2026fundraising adjudicate this with a three-year panel of 105 U.S. food
banks, a setting where solicitation methods, dedicated fundraising headcount, and
revenue by method are all observed for organizations pursuing an identical
mission. Four results are worth carrying forward.

First, **concentration raises productivity**. Greater revenue concentration is
associated with higher fundraising revenue per FTE, favoring the capability
argument over the portfolio argument at the level of labor returns.

Second, the effect is **amplified by relational intensity**. Organizations that
prioritize highly personalized solicitation methods get more out of concentrating,
and the authors' post hoc decomposition traces the amplification to grant writing
and personal solicitation specifically. This is the learning-curve mechanism made
visible: it is precisely the methods whose returns depend on accumulated
relationships and craft that reward depth over breadth.

Third, **organizational age moderates the effect negatively**. Splitting the
sample by age, concentration significantly improves productivity for the youngest
food banks and has no statistically significant effect for mature ones. A young
organization has no channel in which it is yet good; committing its small team to
one is how it becomes good at something. A mature organization has already climbed
the learning curves it is going to climb, and further concentration buys it little.

Fourth, **labor mix conditions the return**. A higher proportion of full-time
fundraisers amplifies the benefit of concentration, continuity of relationship
being the input that method-specific capital is built from. This qualifies rather
than contradicts the finding of @sharma2024parttime that part-time staffing can
*increase* effectiveness in food-bank SNAP outreach: flexible labor suits outreach
tasks that are episodic and geographically dispersed, whereas concentrated,
relationship-intensive solicitation rewards continuity. The operative question is
not whether part-time labor is good, but which task the labor is asked to perform.

### The Tension the Result Exposes {#sec-fundraising-tradeoff}

The finding does not survive a change of dependent variable unscathed, and this is
the most useful part of the paper for a methods reader. Concentration is also
positive for **effectiveness** (total fundraising revenue) and for **revenue
growth** (year-over-year rate of increase), but *negative* for **financial
stability**, measured as the deviation of realized fundraising revenue from its
predicted value. Concentration raises the level and the trend of the revenue
stream while raising its volatility, which is the portfolio argument reappearing
where it belongs: in the second moment rather than the first.

::: {.callout-note}
This is the same structure as the marketing-finance value chain of
@sec-marketing-finance, where market-based assets are argued to *augment the level
of cash flows* and *reduce their volatility* as separate levers. The food-bank
evidence is a reminder that a single strategic choice can move the two levers in
opposite directions, so the phrase "improves performance" is empty until the
performance construct is named. It is also the nonprofit analogue of customer
concentration in B2B selling: a portfolio anchored on a few large accounts is
efficient per salesperson and fragile per shock.
:::

### A Reproducible Illustration of the Tradeoff {#sec-fundraising-sim}

The simulation below builds a synthetic panel with the qualitative structure just
described, concentration raising the *level* of revenue per FTE while also raising
its *dispersion* around the organization's own trend, and shows that a
within-organization fixed-effects regression recovers both halves of the tradeoff.
The numbers are the book's own construction, not estimates from
@han2026fundraising; the point is that the two effects are separately estimable
from the same panel and that reporting only the first would be a partial account.

```{r fundraising-concentration-sim, message=FALSE, warning=FALSE}
set.seed(2026)

n_org  <- 105   # organizations, matching the scale of the food-bank panel
n_yr   <- 3     # years
n_meth <- 6     # solicitation methods

# --- Revenue shares by method, then the Herfindahl concentration index --------
# A Dirichlet draw with a small concentration parameter yields lumpy portfolios;
# alpha varies by organization so that H has genuine between-org variation.
rdirichlet <- function(alpha) {
  g <- rgamma(length(alpha), shape = alpha)
  g / sum(g)
}

org    <- rep(seq_len(n_org), each = n_yr)
year   <- rep(seq_len(n_yr), times = n_org)
alpha0 <- rep(runif(n_org, 0.2, 2.0), each = n_yr)   # low alpha => concentrated

H <- vapply(alpha0, function(a) sum(rdirichlet(rep(a, n_meth))^2), numeric(1))

# --- Outcome: log revenue per FTE --------------------------------------------
# Level effect: concentration raises productivity (beta_H > 0).
# Volatility effect: concentration raises the SD of the idiosyncratic shock, so
# the same organization is less predictable year to year.
beta_H   <- 0.90
org_fe   <- rep(rnorm(n_org, mean = 11.5, sd = 0.35), each = n_yr)  # log $ per FTE
yr_fe    <- c(0, 0.04, 0.09)[year]
sd_shock <- 0.10 + 0.45 * H                                         # heteroskedastic in H
eps      <- rnorm(length(H), mean = 0, sd = sd_shock)

log_prod <- org_fe + yr_fe + beta_H * H + eps

d <- data.frame(org = factor(org), year = factor(year), H = H, log_prod = log_prod)

# --- First moment: does concentration raise productivity? --------------------
fit_level <- lm(log_prod ~ H + org + year, data = d)
round(coef(summary(fit_level))["H", ], 4)

# --- Second moment: does it raise instability? -------------------------------
# "Instability" mirrors the paper's construct: absolute deviation of realized
# productivity from its predicted value, here from a model that deliberately
# excludes H so the residual is not mechanically purged of it.
fit_pred <- lm(log_prod ~ org + year, data = d)
d$instab <- abs(residuals(fit_pred))
fit_stab <- lm(instab ~ H, data = d)
round(coef(summary(fit_stab))["H", ], 4)
```

```{r}
#| label: fig-fundraising-tradeoff
#| fig-cap: "Concentration raises the level of fundraising productivity (left) and its instability (right). Both panels are simulated from the data-generating process above, not from published estimates."
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1))

plot(d$H, d$log_prod, pch = 16, col = "#4477aa66",
     xlab = "Revenue concentration H", ylab = "log(revenue per FTE)",
     main = "Level")
abline(lm(log_prod ~ H, data = d), lwd = 2, col = "#bb5566")

plot(d$H, d$instab, pch = 16, col = "#22883366",
     xlab = "Revenue concentration H", ylab = "|deviation from predicted|",
     main = "Instability")
abline(fit_stab, lwd = 2, col = "#bb5566")

par(op)
```

Two cautions belong with any reading of this evidence, simulated or real. The
concentration index is **chosen, not assigned**: an organization that concentrates
may be responding to the same donor-market conditions that drive its productivity,
so the panel's fixed effects absorb time-invariant capability but not a
time-varying shock that moves strategy and outcome together. And the level and
volatility results come from the same panel, so a manager cannot treat them as
independent findings to be traded off at will; they are two projections of one
strategic posture. The identification discipline of @sec-causal-inference applies
here with no discount for the setting being a nonprofit one.

## Key Takeaways

- **Adaptive selling resolves to the speaking turn.** Across 783 prospecting calls and
  11,566 turns, salesperson speech splits into linguistic acts (legitimizing, piquing,
  nudging) and paralinguistic acts (intensity, rate, pitch) that work through **lead
  receptivity**, a turn-level openness measure whose one-SD increase lifts conversion
  13.44% [@balducci2026leads]. The acts are *positionally* effective, not uniformly so:
  a specific nudge gains 11.82% early and loses 81.40% late. Any call-level aggregate
  averages those into near-zero and reports that nothing matters
  (@sec-sales-conversation).
- Variable compensation lifts sales but carries a **health cost**—stress,
  exhaustion, absenteeism—that is buffered by individual ability and social
  resources, so the profit-maximizing incentive slope is lower and more
  differentiated than the standard agency model implies [@habel2021].
- **Discretionary bonuses are not a free lever.** A natural field experiment finds
  that surprise bonuses *reduced* subsequent output—reciprocity ran backward once
  the bonus was read as a signal of employer contentment rather than a gift to be
  repaid—so gift-exchange justifications for goodwill pay should not be assumed
  [@bogliacino2026gifts].
- **Sales rank is ordinal**; recovering cardinal demand requires either an
  experimental demand shock [@chevalier2003measuring; @he2020sales], auxiliary
  demand data [@brynjolfsson2003consumer], or a parametric demand law identified
  from multiple ranked lists [@garg2013inferring].
- The Pareto rank–demand model (@eq-pareto-paid–@eq-scale-recovery) identifies
  tail **shapes** and a **scale ratio** from rank regressions, but needs one
  aggregate-download moment to fix scale **levels**; its identification breaks if
  revenue is not upfront, if cross-sectional independence fails, or if a single
  power law does not fit the list.
- Salesperson **turnover is contagious**: peer turnover predicts individual exit
  more strongly than own performance, but reading this as causal contagion
  demands care with the reflection problem and homophily
  [@sunder2017salespeople].
- A **forward-looking, latent-class** valuation of the sales force
  (@eq-salesperson-value) shows that returns to training and incentives are
  segment- and horizon-specific, so uniform development programs are inefficient
  [@kumar2014measuring].
- **Fundraising is selling without a price**, and its productivity metric puts
  labor in the denominator (@eq-fundraising-productivity). In a three-year panel of
  105 food banks, **concentrating** revenue across fewer solicitation methods
  raised revenue per FTE—amplified by personalized methods such as grant writing
  and personal solicitation and by a larger full-time share, and concentrated among
  the youngest organizations—while *reducing* financial stability
  [@han2026fundraising]. A strategy can move the level and the volatility of a
  revenue stream in opposite directions, so "improves performance" is empty until
  the performance construct is named.
