# Customer Lifetime Value {#sec-clv}

*Customer lifetime value* (CLV)—often called *lifetime value* (LTV) in practice—is
the present value of the future cash flows a firm expects to earn from a customer
relationship over its remaining duration. It reframes the customer from a
transaction into an asset: the question is no longer "what did this person buy
today?" but "what is this relationship worth, net of the cost of sustaining it, for
as long as it lasts?" That shift in unit of analysis is the conceptual core of
customer-centric marketing. Once customers are assets, acquisition becomes a capital
budgeting decision, retention becomes asset maintenance, and the firm's total
customer base aggregates into *customer equity*—a balance-sheet-style summary of the
demand side of the business [@rust2002customer; @gupta2006modeling].

CLV matters for two reasons that map onto the book's two audiences. Scientifically,
it forces an explicit stochastic model of three primitives—how long a customer stays
(retention or survival), how often and how much they buy (transaction flow and
spend), and how those translate into margin net of marketing cost—each of which is a
latent process observed only through noisy transaction data. Commercially, CLV is
the metric that allocates acquisition budget, prioritizes service, sizes loyalty
programs, and, when aggregated, has been shown to track shareholder value
[@gupta2006modeling; @kumar2010]. A century-old practitioner heuristic captures the
stakes: a single year's projected business typically recovers only about half of a
customer's full lifetime value, a ratio that varies widely by industry
[@kumar_2007]. Truncating the horizon at one year therefore discards roughly half
the asset.

This chapter builds CLV from its primitives. It first defines the construct
formally and states the assumptions under which the familiar closed-form formulas
hold. It then develops the two dominant modeling traditions—the *managerial
discounted-margin* formulation and the *probabilistic* (buy-till-you-die and
shifted-beta-geometric) tradition—and shows, with reproducible code, how to estimate
a retention curve from sparse data and roll it forward into a value forecast. It
closes by extending the asset view beyond the focal customer's own purchases to the
value a customer creates by *referring* others (customer referral value) and by
surveying the identification problems that make CLV easy to compute and hard to
compute *correctly*.

## The Construct and Its Assumptions

Fix a customer $i$ and let periods be indexed by $j$ (annual or semiannual). Three
latent quantities drive value: the probability $p(\text{Buy}_{ij}=1)$ that the
customer is active and transacts in period $j$; the contribution margin
$\text{CM}_{ij}$ the customer generates conditional on transacting; and the
marketing cost $\text{MC}_{ij}$ the firm directs at the customer to keep the
relationship alive. Future cash flows are discounted at a per-period rate $r$.
Writing $T$ for the end of the observation (calibration) window and $N$ for the
number of periods projected forward, the discounted-margin definition of CLV is

$$
\text{CLV}_i \;=\; \sum_{j=T+1}^{T+N}
   \frac{p(\text{Buy}_{ij}=1)\,\widehat{\text{CM}}_{ij}}{(1+r)^{\,j-T}}
   \;-\; \sum_{j=T+1}^{T+N} \frac{\widehat{\text{MC}}_{ij}}{(1+r)^{\,j-T}} .
$$ {#eq-clv-general}

@eq-clv-general is the formulation popularized in the customer-equity program of
Kumar and colleagues [@Kumar_2008; @kumar2010]. Several features deserve emphasis
because they are routinely elided in practice. The summation starts at $j=T+1$: CLV
is a *forward-looking* quantity, the value of the relationship from today onward, not
the historical revenue already booked.[^historic] The purchase probability,
contribution margin, and marketing cost are all *predicted* ($\widehat{\,\cdot\,}$),
so CLV inherits the sampling error of three separate forecasting models. And the
discount factor must be expressed at the period frequency of the data: a 15% annual
rate corresponds to a semiannual factor of $r \approx 0.07238$, since
$(1.07238)^2 \approx 1.15$ [@Kumar_2008].

[^historic]: A common conflation is between CLV and *historic customer value* (the
realized margin already earned). The two answer different questions: historic value
scores past contribution; CLV is the basis for forward-looking decisions such as how
much to spend acquiring a *similar* prospect. Only the latter belongs in a capital
budgeting calculation.

> **Customer lifetime value.** The present value of all future cash flows
> attributable to the relationship with a customer—margin from the customer's own
> purchases, net of the marketing cost of serving and retaining them—over the
> expected remaining lifetime of that relationship.

A useful special case exposes the economics. Suppose margin per active period is a
constant $m = \text{CM} - \text{MC}$ and the customer is retained from one period to
the next with constant probability $\rho$ (the *retention rate*). Then the expected
discounted value of an infinite-horizon relationship collapses to a geometric series,

$$
\text{CLV} \;=\; m \sum_{t=0}^{\infty} \left(\frac{\rho}{1+r}\right)^{t}
            \;=\; m \,\frac{1+r}{\,1+r-\rho\,},
$$ {#eq-clv-geometric}

valid whenever $\rho < 1+r$. The factor $(1+r)/(1+r-\rho)$ is the *margin
multiple*: at $\rho = 0.8$ and $r = 0.1$ it equals $1.1/0.3 \approx 3.67$, so a
loyal customer is worth roughly three and two-thirds years of current margin. Two
modeling decisions are buried in @eq-clv-geometric and matter enormously: the
constancy of $\rho$ (real retention curves are not memoryless, as the next section
shows) and the choice of horizon (the infinite sum versus a finite $N$). Both are
where naive CLV estimates go wrong.

### Two Traditions

The literature splits into two complementary approaches to operationalizing
@eq-clv-general, summarized in @tbl-clv-approaches. The *managerial* tradition
predicts the components—purchase incidence, margin, marketing cost—with separate
regression or machine-learning models, typically on a contractual or
panel-with-covariates dataset, and is the workhorse of the customer-equity
literature [@Kumar_2008; @kumar2010; @gupta2006modeling]. The *probabilistic*
tradition instead posits a small generative model for the latent purchase and
dropout processes and derives expected value in closed form; its canonical members
are the Pareto/NBD and BG/NBD "buy-till-you-die" models for noncontractual settings
[@schmittlein1982] and the shifted-beta-geometric (sBG) model for discrete-time
contractual settings, developed below. A third, structural strand models the
customer's underlying utility and budget constraint and derives CLV as an
implication of choices [@sunder2016measuring], trading tractability for behavioral
fidelity.

| Dimension | Managerial / component models | Probabilistic models |
|---|---|---|
| What is modeled | Each cash-flow component separately | Latent purchase + dropout processes |
| Typical data | Contractual or covariate-rich panel | Recency–frequency transaction logs |
| Covariates | Rich (demographics, marketing) | Few or none |
| Output | Per-customer point forecast | Posterior expected value |
| Strengths | Targeting, what-if on marketing | Parsimony, honest uncertainty |
| Weakness | Many models to maintain and validate | Hard to attach drivers/covariates |

: Two traditions for operationalizing customer lifetime value. {#tbl-clv-approaches}

The remainder of this chapter develops the probabilistic tradition in depth, because
it most cleanly exposes the latent-process assumptions on which *every* CLV number
depends, and because it estimates cleanly from the sparse retention data firms
actually hold. @fig-clv-pipeline shows how the pieces compose.

```{mermaid}
%%| label: fig-clv-pipeline
%%| fig-cap: "From transaction data to customer lifetime value. A latent retention or survival process is estimated, projected forward, combined with a margin process, discounted, and aggregated across customers into customer equity."
flowchart LR
  A[Transaction /<br/>retention data] --> B[Estimate latent<br/>survival process]
  B --> C[Project retention<br/>over horizon N]
  M[Margin & cost<br/>process] --> D
  C --> D[Discount expected<br/>per-period cash flows]
  D --> E[CLV per customer]
  E --> F[Aggregate:<br/>customer equity]
```

## Retention, Survival, and the Shifted-Beta-Geometric Model

The single most consequential primitive in @eq-clv-general is the path of the
purchase probability over time. In a *contractual* setting—subscriptions,
memberships, insurance—the firm observes exactly when each customer churns, and the
object of interest is the **survivor function** $S(t)=\Pr(\text{still active at }t)$.
The naive approach assumes a constant retention rate $\rho$ and sets
$S(t)=\rho^{t}$, the memoryless geometric curve underlying @eq-clv-geometric. This
is almost always wrong in a specific and predictable direction: *observed aggregate
retention rates rise with tenure*. Customers who have stayed many periods are
disproportionately the intrinsically loyal ones; the churn-prone have already left.
Aggregate retention climbs not because any individual becomes more loyal but because
the surviving population is increasingly self-selected toward loyalty—a pure
*sorting* (heterogeneity) effect, not a duration-dependence effect.

The shifted-beta-geometric model formalizes exactly this intuition and is the
discrete-time benchmark for contractual CLV. Each customer has a fixed but
unobserved per-period churn probability $\theta$. Conditional on $\theta$, the
period of churn follows a (shifted) geometric distribution: the customer survives
each period with probability $1-\theta$ and churns with probability $\theta$, so

$$
P(T = t \mid \theta) = \theta\,(1-\theta)^{\,t-1}, \qquad t = 1, 2, \dots
$$ {#eq-sbg-geometric}

Heterogeneity across customers is captured by mixing $\theta$ over the population
with a Beta prior, $\theta \sim \text{Beta}(\alpha,\beta)$, whose density is
$f(\theta)=\theta^{\alpha-1}(1-\theta)^{\beta-1}/B(\alpha,\beta)$. Integrating the
individual churn probability over this prior yields closed-form aggregate churn and
survivor functions that depend only on the two parameters $(\alpha,\beta)$:

$$
P(T = t) = \frac{B(\alpha+1,\;\beta+t-1)}{B(\alpha,\beta)}, \qquad
S(t) = \frac{B(\alpha,\;\beta+t)}{B(\alpha,\beta)} .
$$ {#eq-sbg-survivor}

These satisfy the convenient forward recursions
$P(1)=\alpha/(\alpha+\beta)$, $P(t)=P(t-1)\,(\beta+t-2)/(\alpha+\beta+t-1)$, and
$S(t)=S(t-1)-P(t)$, which is exactly how the estimation code below evaluates them.
The model's key qualitative property—rising aggregate retention from fixed
individual propensities—emerges purely from the Beta mixture; no individual's
$\theta$ ever changes.

**Estimation.** With cohort data giving the number of customers $n_t$ surviving at
each tenure $t$ (equivalently, the number churning, $d_t = n_{t-1}-n_t$), the
parameters are estimated by maximum likelihood. The log-likelihood for a cohort
observed through period $\tau$ sums the log-probability of each observed churn over
its churn period plus the log-survivor probability for those still active at the end
of the window:

$$
\ell(\alpha,\beta) = \sum_{t=1}^{\tau} d_t \,\log P(T=t)
   \;+\; n_{\tau}\,\log S(\tau).
$$ {#eq-sbg-loglik}

The last term is the right-censoring contribution: customers still active at $\tau$
contribute the probability of surviving *at least* that long. Identification is weak
when the window is short relative to the heterogeneity—$\alpha$ and $\beta$ are
separately identified only once the curvature of the survival curve is visible—which
is why the projections below stabilize as more months of history are used.

The following example reconstructs the analysis of a subscription business: four
cohorts ("cases") with markedly different retention dynamics, each observed for eight
months. The goal is to fit the sBG model and project retention forward. We first
visualize the raw curves.

```{r clv-setup}
#| message: false
#| warning: false
#| label: fig-retention-observed
#| fig-cap: "Observed monthly retention curves for four subscription cohorts. All decline, but at different rates and with different curvature."
library(tidyverse)

set.seed(13)

# Observed retention rate by months since signup (month 0 = 100%)
df_ret <- tibble(
  month_lt = 0:7,
  case01 = c(1, .531, .452, .423, .394, .375, .356, .346),
  case02 = c(1, .869, .743, .653, .593, .551, .517, .491),
  case03 = c(1, .677, .562, .486, .412, .359, .332, .310),
  case04 = c(1, .631, .468, .382, .326, .289, .262, .241)
) |>
  pivot_longer(-month_lt, names_to = "example", values_to = "retention_rate")

ggplot(df_ret, aes(month_lt, retention_rate, colour = example)) +
  geom_line() + geom_point() +
  facet_wrap(~ example) +
  scale_colour_manual(values = c("#4e79a7", "#f28e2b", "#e15759", "#76b7b2")) +
  labs(x = "Months since signup", y = "Retention rate") +
  theme_minimal() +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))
```

We next implement the sBG churn, survivor, and (negative) log-likelihood functions
following @eq-sbg-survivor and @eq-sbg-loglik, then optimize.

```{r sbg-functions}
# Aggregate churn probability P(T = t) via the forward recursion implied by
# the Beta-geometric mixture (equation @eq-sbg-survivor).
churn_bg <- Vectorize(function(alpha, beta, period) {
  p <- alpha / (alpha + beta)
  if (period > 1)
    p <- churn_bg(alpha, beta, period - 1) *
         (beta + period - 2) / (alpha + beta + period - 1)
  p
}, vectorize.args = "period")

# Survivor function S(t) = S(t-1) - P(t), with S(1) = 1 - P(1).
survival_bg <- Vectorize(function(alpha, beta, period) {
  s <- 1 - churn_bg(alpha, beta, 1)
  if (period > 1)
    s <- survival_bg(alpha, beta, period - 1) - churn_bg(alpha, beta, period)
  s
}, vectorize.args = "period")

# Negative log-likelihood (equation @eq-sbg-loglik): churners contribute
# log P(T = t); the still-active at the end contribute log S(tau).
neg_loglik <- function(par, active, lost) {
  alpha <- par[1]; beta <- par[2]
  tau <- length(active)
  -( sum(lost * log(churn_bg(alpha, beta, 1:tau))) +
       active[tau] * log(survival_bg(alpha, beta, tau)) )
}

# Convert retention rates into active/lost counts for a notional 1,000-customer
# cohort (the likelihood depends only on the shape, not the cohort size).
df_ret <- df_ret |>
  group_by(example) |>
  mutate(active = 1000 * retention_rate,
         lost   = replace_na(lag(active) - active, 0)) |>
  ungroup()
```

The instructive exercise is to ask how the *projected* retention curve evolves as
the firm accumulates more months of actual data. With only one or two months
observed, identification is weak and the projection is unreliable; as the window
lengthens, the fitted curve locks onto the truth. We refit the model for each cohort
using the first $i$ months only, for $i = 1, \dots, 7$, and overlay every projection
on the observed curve.

```{r sbg-projections}
#| label: fig-retention-projected
#| fig-cap: "Observed retention (bold) with sBG projections fit on the first i months only, for i = 1..7 (faint lines). Projections stabilize toward the observed curve as the calibration window lengthens."
project_cohort <- function(d) {
  map_dfr(1:7, function(i) {
    fit <- d |> filter(month_lt >= 1, month_lt <= i)
    opt <- optim(c(1, 1), neg_loglik,
                 active = fit$active, lost = fit$lost)
    tibble(month_lt   = 0:7,
           fact_months = i,
           retention_pred = round(
             c(1, survival_bg(opt$par[1], opt$par[2], 1:7)), 3))
  })
}

ret_preds <- df_ret |>
  group_by(example) |>
  group_modify(~ project_cohort(.x)) |>
  ungroup()

df_plot <- df_ret |>
  select(month_lt, example, retention_rate) |>
  left_join(ret_preds, by = c("month_lt", "example"))

ggplot(df_plot, aes(month_lt, retention_rate, colour = example)) +
  geom_line(aes(y = retention_pred, group = fact_months), alpha = 0.4) +
  geom_line(linewidth = 1.2) + geom_point(size = 1.2) +
  facet_wrap(~ example) +
  scale_colour_manual(values = c("#4e79a7", "#f28e2b", "#e15759", "#76b7b2")) +
  labs(x = "Months since signup", y = "Retention rate") +
  theme_minimal() +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))
```

The faint projection lines fan out wildly when fit on one month and converge onto
the observed curve as the calibration window grows—a visual diagnostic of the
identification problem in @eq-sbg-loglik. In practice this argues against committing
to a CLV figure from a single early cohort snapshot.

### From Retention to Value

With a fitted survivor function, lifetime value follows directly. For a contractual
product with constant per-period margin $m$ (here a \$1 subscription) and discount
factor $r$, expected discounted value over a horizon of $N$ periods is

$$
\widehat{\text{CLV}} \;=\; m \sum_{t=0}^{N} \frac{\widehat{S}(t)}{(1+r)^{\,t}},
$$ {#eq-clv-from-survivor}

with $\widehat{S}(t)$ taken from observed data where available and from the sBG
projection beyond the calibration window. Setting $r=0$ for transparency (the firm
wants undiscounted expected revenue per signup), we compute average LTV for cohort 3
from only two months of history, projected over a 24-month horizon.

```{r clv-forecast}
fit3 <- df_ret |> filter(example == "case03", month_lt >= 1, month_lt <= 2)
opt3 <- optim(c(1, 1), neg_loglik, active = fit3$active, lost = fit3$lost)

projected <- tibble(
  month_lt       = 3:24,
  retention_pred = round(survival_bg(opt3$par[1], opt3$par[2], 3:24), 3))

ltv3 <- df_ret |>
  filter(example == "case03", month_lt <= 2) |>
  select(month_lt, retention_rate) |>
  bind_rows(projected) |>
  mutate(retention = coalesce(retention_rate, retention_pred),
         ltv_monthly = retention * 1,          # $1 subscription price
         ltv_cum     = round(cumsum(ltv_monthly), 2))

tail(ltv3, 3)
cat("Average 24-month LTV for cohort 3: $", tail(ltv3$ltv_cum, 1), "\n", sep = "")
```

The cumulative figure—roughly \$9.33 per subscriber over 24 months—blends two months
of *observed* retention with twenty-two months of *projected* retention from a
two-parameter model estimated on those same two months. That this is even possible is
the practical appeal of the probabilistic approach; that it rests on extrapolating a
heterogeneity model far beyond its calibration window is its central caveat, and
@fig-retention-projected is the warning label.

## Customer Referral Value

CLV as defined so far captures only the margin a customer generates through their
*own* purchases. A customer also creates value by bringing in others—through
word of mouth and formal referral programs—and this *customer referral value* (CRV)
is empirically distinct from CLV: high-CLV customers are frequently *not* the best
referrers, so the two must be managed separately [@kumar_2007]. The managerial
implication is a two-by-two **customer value matrix** crossing CLV (low/high) with
CRV (low/high): "champions" score high on both, "affluents" buy heavily but refer
little, "advocates" refer well but buy little, and the remainder warrant minimal
investment. The diagnostic point is that targeting on CLV alone leaves the referral
asset unmanaged.

Measuring referral value requires separating two kinds of referred customer
[@kumar_2007]. A **type-one** referral is incremental—the referred customer would
*not* have joined absent the referral—so their entire margin is attributable to the
referrer. A **type-two** referral *would* have joined anyway through other channels;
the referrer's contribution is not the customer's margin but the *acquisition cost
the firm saved* by not having to recruit them through paid channels. Empirically the
split is roughly even between the two types [@kumar_2007]. Letting $n_1$ index the
incremental (type-one) referrals and $n_1{+}1,\dots,n_2$ index the would-have-joined
(type-two) referrals,

$$
\text{CRV}_i \;=\; \sum_{t=1}^{T}\sum_{y=1}^{n_1}
   \frac{A_{ty} - a_{ty} - M_{ty} + \text{ACQ1}_{ty}}{(1+r)^{t}}
   \;+\; \sum_{t=1}^{T}\sum_{y=n_1+1}^{n_2}
   \frac{\text{ACQ2}_{ty}}{(1+r)^{t}},
$$ {#eq-crv}

where $A_{ty}$ is the gross margin from incremental customer $y$, $a_{ty}$ the cost
of the referral incentive, $M_{ty}$ the marketing cost to retain the referred
customer, $\text{ACQ1}_{ty}$ the acquisition cost saved on incremental customers,
$\text{ACQ2}_{ty}$ the acquisition cost saved on would-have-joined customers, and $r$
the per-period discount factor (again $\approx 0.07238$ semiannually for a 15% annual
rate [@Kumar_2008]). The first sum books full net margin for incremental customers;
the second books *only* avoided acquisition cost for the rest. A practical detail
governs the attribution window: referrals attributable to an incentive campaign
accrue for about a year after it runs, so CRV is typically accumulated over a
one-year horizon and referrals beyond it are treated as organic
[@kumar_2007].

## Identification and What Breaks It

CLV is arithmetically simple and inferentially treacherous. Every term in
@eq-clv-general is a forecast, and the headline number compounds their errors. The
recurring failure modes are worth stating as assumptions whose violation is
diagnostic.

*Constant-retention bias.* Assuming a fixed $\rho$ (@eq-clv-geometric) when the true
process is heterogeneous understates the value of long-tenured customers and
overstates churn for survivors, because it misreads population sorting as individual
duration dependence. The sBG model exists precisely to correct this; the symptom is a
geometric model that systematically underpredicts late-tenure survival.

*Horizon and discounting.* Because roughly half of lifetime value sits beyond the
first year [@kumar_2007], the choice of $N$ and $r$ is not a rounding decision. The
infinite-horizon shortcut in @eq-clv-geometric requires $\rho < 1+r$ and a genuinely
stationary margin; the Gordon-style terminal value used in valuation work is valid
only when the growth rate is strictly below the discount rate. A short horizon
truncates the asset; an over-long one extrapolates a calibration-window model past
its evidence (@fig-retention-projected).

*Endogenous marketing.* The marketing cost $\text{MC}_{ij}$ is not exogenous—firms
direct more spend at customers they already believe are valuable. Treating observed
$\text{MC}$ and observed retention as if the latter were the causal response to the
former conflates targeting with treatment effect, biasing the implied return on
retention spending. Clean estimates of how marketing *moves* CLV require experimental
or quasi-experimental variation, not the historical correlation between spend and
value.

*Aggregation and the noncontractual problem.* In noncontractual settings (retail,
e-commerce) churn is never observed: a customer who has not purchased recently may
have defected or may simply be between trips. Disentangling "dead" from "dormant"
requires a latent-attrition model such as the Pareto/NBD [@schmittlein1982], whose
recency–frequency sufficient statistics carry exactly the information needed to infer
the probability a customer is still alive. Ignoring latent attrition and counting all
lapsed customers as retained inflates customer equity.

*Heterogeneous response and structural drivers.* Reduced-form component models give
per-customer forecasts but say little about *why* value differs or how it responds to
prices and competition. Structural approaches embed CLV in a model of consumer
utility: @sunder2016measuring derive brand- and category-level CLV in consumer
packaged goods by accounting for multiple-discreteness (consumers buy several brands),
brand-switching, and a latent budget constraint inferred from transaction data alone
via Bayesian estimation, and show that ignoring the budget constraint—as simpler
heuristics do—mismeasures value and the response to asymmetric price changes. The
gain in behavioral fidelity is bought with strong functional-form assumptions; the
trade-off between the two traditions in @tbl-clv-approaches has no free lunch.

The financial relevance of getting these right is direct: aggregated CLV is *customer
equity*, and the demonstrated link from customer equity to firm value
[@gupta2006modeling; @kumar2010] is the bridge from this chapter to the
marketing–finance interface developed in @sec-marketing-finance.

## Key Takeaways

- CLV (@eq-clv-general) treats a customer as a forward-looking asset: the discounted
  present value of future margin from their purchases net of the marketing cost of
  retaining them. Aggregated across customers it is *customer equity*.
- The constant-retention shortcut (@eq-clv-geometric) is biased whenever customers
  are heterogeneous; observed aggregate retention rises with tenure because of
  *sorting*, not increasing individual loyalty. The shifted-beta-geometric model
  (@eq-sbg-survivor) captures this with a two-parameter Beta mixture and estimates
  cleanly from sparse cohort data.
- Projections extrapolated from a short calibration window are unstable
  (@fig-retention-projected); commit to a CLV figure only once the survival curve's
  curvature is identified.
- Customer referral value (@eq-crv) is distinct from CLV and must separate
  *incremental* referrals (book full margin) from *would-have-joined* referrals (book
  only saved acquisition cost). High-CLV customers are not automatically high-CRV
  customers.
- The hard part is identification, not arithmetic: constant-retention bias, horizon
  and discounting choices, endogenous marketing spend, and latent attrition in
  noncontractual settings each bias the headline number in a predictable direction.
