# Customer Lifetime Value {#sec-clv}

*Customer lifetime value* (CLV), often called *lifetime value* (LTV) in practice, is
the present value of the future cash flows a firm expects to earn from a customer
relationship over its remaining duration. It reframes the customer from a
transaction into an asset: the question is no longer "what did this person buy
today?" but "what is this relationship worth, net of the cost of sustaining it, for
as long as it lasts?" That shift in unit of analysis is the conceptual core of
customer-centric marketing. Once customers are assets, acquisition becomes a capital
budgeting decision, retention becomes asset maintenance, and the firm's total
customer base aggregates into *customer equity*, a balance-sheet-style summary of the
demand side of the business [@rustLemonZeithaml2004; @gupta2006modeling].

CLV matters for two reasons that map onto the book's two audiences. Scientifically,
it forces an explicit stochastic model of three primitives (how long a customer stays
(retention or survival), how often and how much they buy (transaction flow and
spend), and how those translate into margin net of marketing cost) each of which is a
latent process observed only through noisy transaction data. Commercially, CLV is
the metric that allocates acquisition budget, prioritizes service, sizes loyalty
programs, and, when aggregated, has been shown to track shareholder value
[@gupta2006modeling; @kumar2010; @kumarshah2009]. A practitioner heuristic captures the
stakes: a single year's projected business typically recovers only about half of a
customer's full lifetime value, a ratio that varies widely by industry. Truncating
the horizon at one year therefore discards roughly half the asset.

This chapter builds CLV from both ends. It leads with the way value is computed in
production (the unit-economics formulas, the company-type playbooks, and the pipeline
that turns a data warehouse into a per-customer number) and grounds that practice in
the classical models that explain when the production shortcuts are safe and when they
are biased. It defines the construct formally, lays out the contractual versus
non-contractual taxonomy that decides which model class is even admissible, and then
develops, with reproducible code, the four modeling traditions in turn: the
probabilistic "buy-till-you-die" family, the econometric and hierarchical-Bayes
extensions, the machine-learning and deep-learning frontier, and the
causal/decision-focused layer that asks not what CLV *is* but how to *act* on it. It
covers the full set of adjacent constructs (customer equity, CAC, the LTV:CAC ratio
and payback, net and gross revenue retention, RFM, churn, share of wallet, customer
referral value, customer-based corporate valuation, and negative CLV) and closes on
the identification problems that make CLV easy to compute and hard to compute
*correctly*.

## The Construct and Its Assumptions {#sec-clv-construct}

Fix a customer $i$ and let periods be indexed by $j$ (annual or semiannual). Three
latent quantities drive value: the probability $p(\text{Buy}_{ij}=1)$ that the
customer is active and transacts in period $j$; the contribution margin
$\text{CM}_{ij}$ the customer generates conditional on transacting; and the
marketing cost $\text{MC}_{ij}$ the firm directs at the customer to keep the
relationship alive. Future cash flows are discounted at a per-period rate $r$.
Writing $T$ for the end of the observation (calibration) window and $N$ for the
number of periods projected forward, the discounted-margin definition of CLV is

$$
\text{CLV}_i \;=\; \sum_{j=T+1}^{T+N}
   \frac{p(\text{Buy}_{ij}=1)\,\widehat{\text{CM}}_{ij}}{(1+r)^{\,j-T}}
   \;-\; \sum_{j=T+1}^{T+N} \frac{\widehat{\text{MC}}_{ij}}{(1+r)^{\,j-T}} .
$$ {#eq-clv-general}

@eq-clv-general is the formulation popularized in the customer-equity program of
Kumar and colleagues [@Kumar_2008; @kumar2010]. Several features deserve emphasis
because they are routinely elided in practice. The summation starts at $j=T+1$: CLV
is a *forward-looking* quantity, the value of the relationship from today onward, not
the historical revenue already booked.[^historic] The purchase probability,
contribution margin, and marketing cost are all *predicted* ($\widehat{\,\cdot\,}$),
so CLV inherits the sampling error of three separate forecasting models. And the
discount factor must be expressed at the period frequency of the data: a 15% annual
rate corresponds to a semiannual factor of $r \approx 0.07238$, since
$(1.07238)^2 \approx 1.15$ [@Kumar_2008].

[^historic]: A common conflation is between CLV and *historic customer value* (the
realized margin already earned). The two answer different questions: historic value
scores past contribution; CLV is the basis for forward-looking decisions such as how
much to spend acquiring a *similar* prospect. Only the latter belongs in a capital
budgeting calculation.

> **Customer lifetime value.** The present value of all future cash flows
> attributable to the relationship with a customer (margin from the customer's own
> purchases, net of the marketing cost of serving and retaining them) over the
> expected remaining lifetime of that relationship.

A useful special case exposes the economics. Suppose margin per active period is a
constant $m = \text{CM} - \text{MC}$ and the customer is retained from one period to
the next with constant probability $\rho$ (the *retention rate*). Then the expected
discounted value of an infinite-horizon relationship collapses to a geometric series,

$$
\text{CLV} \;=\; m \sum_{t=0}^{\infty} \left(\frac{\rho}{1+r}\right)^{t}
            \;=\; m \,\frac{1+r}{\,1+r-\rho\,},
$$ {#eq-clv-geometric}

valid whenever $\rho < 1+r$. The factor $(1+r)/(1+r-\rho)$ is the *margin
multiple*: at $\rho = 0.8$ and $r = 0.1$ it equals $1.1/0.3 \approx 3.67$, so a
loyal customer is worth roughly three and two-thirds years of current margin. Two
modeling decisions are buried in @eq-clv-geometric and matter enormously: the
constancy of $\rho$ (real retention curves are not memoryless, as later sections
show) and the choice of horizon (the infinite sum versus a finite $N$). Both are
where naive CLV estimates go wrong. The closed-form CLV expressions of the managerial
tradition, of which @eq-clv-geometric is the simplest, descend from
@bergernasr1998.

### Two Traditions and a Taxonomy {#sec-clv-traditions}

The literature splits into complementary approaches to operationalizing
@eq-clv-general, summarized in @tbl-clv-approaches. The *managerial* tradition
predicts the components (purchase incidence, margin, marketing cost) with separate
regression or machine-learning models, typically on a contractual or
panel-with-covariates dataset, and is the workhorse of the customer-equity
literature [@Kumar_2008; @kumar2010; @gupta2006modeling; @venkatesankumar2004]. The
*probabilistic* tradition instead posits a small generative model for the latent
purchase and dropout processes and derives expected value in closed form; its
canonical members are the Pareto/NBD and BG/NBD "buy-till-you-die" models for
non-contractual settings [@paretonbd1987; @fader2005bgnbd] and the
shifted-beta-geometric (sBG) model for discrete-time contractual settings
[@faderhardie2007sbg]. A third, structural strand models the customer's underlying
utility and budget constraint and derives CLV as an implication of choices
[@sunder2016measuring], trading tractability for behavioral fidelity.

| Dimension | Managerial / component models | Probabilistic models |
|---|---|---|
| What is modeled | Each cash-flow component separately | Latent purchase + dropout processes |
| Typical data | Contractual or covariate-rich panel | Recency-frequency transaction logs |
| Covariates | Rich (demographics, marketing) | Few or none |
| Output | Per-customer point forecast | Posterior expected value |
| Strengths | Targeting, what-if on marketing | Parsimony, honest uncertainty |
| Weakness | Many models to maintain and validate | Hard to attach drivers/covariates |

: Two traditions for operationalizing customer lifetime value. {#tbl-clv-approaches}

Which tradition is even *admissible* is decided by a single two-by-two taxonomy,
shown in @tbl-clv-taxonomy, that organizes the entire field. The first axis is
whether the setting is *contractual* (the firm observes exactly when a customer
leaves: subscriptions, insurance, B2B SaaS) or *non-contractual* (the customer can
lapse silently and churn is never directly observed: retail, e-commerce, direct-to-
consumer). The second axis is whether purchase opportunities arrive in *discrete*
periods or in *continuous* time [@faderhardie2009tutorial]. The chapter's sBG model
occupies the contractual-discrete cell; the buy-till-you-die models occupy the
non-contractual cells.

| | Discrete time | Continuous time |
|---|---|---|
| Contractual (churn observed) | shifted-beta-geometric (sBG) for retention; BG/BB for transactions | survival models (exponential, Weibull, Cox) on time-to-churn |
| Non-contractual (churn latent) | BG/BB | Pareto/NBD, BG/NBD, Gamma-Gamma |

: The organizing taxonomy of customer-base analysis. The cell decides which model class is admissible. {#tbl-clv-taxonomy}

@fig-clv-model-choice recasts the same taxonomy as the sequence of questions an
analyst answers before any estimation begins.

```{mermaid}
%%| label: fig-clv-model-choice
%%| fig-cap: "Choosing a customer-base model. The contractual versus non-contractual distinction (is churn observed?) and the discrete versus continuous timing of purchase opportunities jointly decide which model class is admissible."
flowchart TD
  Q1{"Is churn observed?<br/>(contractual setting)"}
  Q1 -->|"yes: subscriptions,<br/>insurance, B2B SaaS"| Q2{"Purchase opportunities<br/>in discrete periods?"}
  Q1 -->|"no: retail, e-commerce,<br/>direct-to-consumer"| Q3{"Purchase opportunities<br/>in discrete periods?"}
  Q2 -->|"discrete"| M1["sBG for retention;<br/>BG/BB for transactions"]
  Q2 -->|"continuous"| M2["Survival models on time-to-churn<br/>(exponential, Weibull, Cox)"]
  Q3 -->|"discrete"| M3["BG/BB"]
  Q3 -->|"continuous"| M4["Pareto/NBD, BG/NBD,<br/>Gamma-Gamma"]
```

@fig-clv-pipeline shows how the pieces compose, from raw transactions through a
latent survival process to a discounted, aggregated customer-equity figure.

```{mermaid}
%%| label: fig-clv-pipeline
%%| fig-cap: "From transaction data to customer lifetime value. A latent retention or survival process is estimated, projected forward, combined with a margin process, discounted, and aggregated across customers into customer equity."
flowchart LR
  A[Transaction /<br/>retention data] --> B[Estimate latent<br/>survival process]
  B --> C[Project retention<br/>over horizon N]
  M[Margin & cost<br/>process] --> D
  C --> D[Discount expected<br/>per-period cash flows]
  D --> E[CLV per customer]
  E --> F[Aggregate:<br/>customer equity]
```

## CLV in Industry Practice {#sec-clv-production}

Academic CLV and operational LTV diverge sharply. In production, the formula a company
uses is dictated by its business model and its data, and most LTV is computed with far
cruder tools than the journals describe, which is exactly why the production pitfalls
catalogued below recur. We lead with practice because it motivates everything that
follows: each classical model in the later sections exists to repair a specific
failure of the naive production formula.

### The Formulas Practitioners Actually Use {#sec-clv-formulas}

The single most widely used LTV formula, ubiquitous in SaaS and subscription
businesses, is

$$
\text{LTV} \;=\; \frac{\text{ARPA} \times \text{gross margin \%}}{\text{churn rate}},
$$ {#eq-clv-naive}

where ARPA is average revenue per account per period and $1/\text{churn}$ is the
implied (geometric) expected lifetime. @eq-clv-naive is exactly the chapter's
@eq-clv-geometric with $r=0$ and constant retention $\rho = 1 - \text{churn}$. Its
appeal is that it needs only two numbers off a dashboard. Its dangers are four. First,
it assumes a single constant churn rate, but real retention curves flatten with tenure
(the sorting effect formalized by the sBG model in @sec-clv-sbg), so it *underestimates*
the value of survivors. Second, it divides by churn, so as churn approaches zero the
estimate explodes: a firm that improves monthly churn from 2% to 1% sees its headline
LTV double, an artifact of the functional form rather than a real doubling of value.
Third, it ignores discounting. Fourth, it ignores expansion revenue, so for any
business with net revenue retention above 100% (defined in @sec-clv-saas-metrics) it
understates value badly.

The correct version is the discounted, cohort-based form, identical to the
@eq-clv-from-survivor derived later: sum the per-period margin weighted by the survivor
function $S(t)$ and discounted at the period rate. Most companies should use this and
do not.

Above the customer level sit the board-room *unit-economics ratios* that govern
whether a growth model is solvent:

- *CAC* (customer acquisition cost) is total sales-and-marketing spend in a period
  divided by new customers acquired.
- The *LTV:CAC ratio* compares the value of an acquired customer to the cost of
  acquiring them. A common venture heuristic treats $\geq 3{:}1$ as healthy, below
  $1{:}1$ as losing money on every customer, and much above $5{:}1$ as a sign of
  *under*-investment in growth.
- The *CAC payback period* is CAC divided by monthly margin per customer, the number
  of months to recoup acquisition cost; under twelve months is a frequent SaaS target.

These ratios carry no academic DOI; they are practitioner and venture-capital
heuristics, but they are how acquisition budgets are actually defended. @sec-clv-unit-econ
computes them from simulated cohorts.

### SaaS Retention Metrics: Logo versus Dollar, Gross versus Net {#sec-clv-saas-metrics}

Subscription businesses report retention at the cohort level along two crossed
distinctions. The first is *logo* (customer count) retention versus *dollar* retention:
a firm can lose many small logos yet grow dollars if its surviving accounts expand, or
the reverse. The second is *gross* versus *net*. Gross revenue retention (GRR) is

$$
\text{GRR} \;=\; \frac{\text{start MRR} - \text{churn} - \text{contraction}}{\text{start MRR}},
$$ {#eq-grr}

which measures pure leakage and is capped at 100%. Net revenue retention (NRR, also
net dollar retention) adds expansion (upsell, seat growth, cross-sell) back in:

$$
\text{NRR} \;=\; \frac{\text{start MRR} - \text{churn} - \text{contraction} + \text{expansion}}{\text{start MRR}},
$$ {#eq-nrr}

and can exceed 100%. NRR above 100% means the existing customer base grows *without any
new logos*, which is why it is the single most-watched metric in subscription software.
A firm with NRR of 120% needs no new customers to grow 20% a year; a firm with NRR of
85% is on a treadmill, replacing a churning base just to stand still. @fig-clv-nrr
renders the expansion-versus-churn waterfall that produces an NRR above 100%.

### A Company-Type Playbook {#sec-clv-playbook}

The admissible model and the operating metric both depend on the business model.
@tbl-clv-playbook is the cross-industry map.

| Company type | Contract? | Data | How CLV is computed in production | Typical tooling |
|---|---|---|---|---|
| SaaS / B2B subscription | Contractual | Billing/CRM, MRR by cohort | Cohort retention curves plus NRR/GRR; LTV:CAC and payback; sometimes sBG or survival on logo churn. Boards see unit economics, not per-customer CLV. | dbt/SQL cohorts; `lifelines`; spreadsheets |
| E-commerce / DTC | Non-contractual | Order logs (id, date, amount) | BTYD: BG/NBD for transactions plus Gamma-Gamma for spend yields per-customer pLTV, feeding segmentation and paid-acquisition bids. | R `CLVTools`/`BTYD`; Python `lifetimes`; two-stage ML |
| Retail / CPG | Non-contractual | Loyalty-card panels, basket data | Panel or loyalty BTYD, or structural CLV [@sunder2016measuring]; share of wallet; category CLV. Hard without identified customers. | Panel econometrics; hierarchical Bayes; warehouse |
| Telecom / insurance / banking | Contractual | Tenure, usage, claims | Survival or churn models (Cox, gradient boosting, survival forests) yield a retention curve, then discounted CLV; uplift for save-desk targeting. | `survival`/`lifelines`; XGBoost; SAS |
| Mobile gaming / apps | Non-contractual, freemium | Event telemetry | Early-LTV/pLTV with ML from the first hours or days of play; whale identification; ZILN or two-stage heads; ad-LTV for user-acquisition bidding. | TensorFlow; gradient boosting; feature stores |
| Media / streaming | Contractual subs plus ad tiers | Viewing plus subscription | Cohort retention with content-engagement covariates; survival on subscription; ad-supported tiers blend ARPU models. | Warehouse cohorts; survival; ML |

: How CLV is computed across company types. {#tbl-clv-playbook}

### The Production Stack and Its Recurring Pitfalls {#sec-clv-pitfalls}

The standard pipeline runs event and transaction data in a warehouse (BigQuery,
Snowflake, Redshift) into feature engineering in dbt or SQL, then into a CLV model
(R `CLVTools`/`BTYD`, or Python `lifetimes`/`lifelines` plus a gradient-boosting or
deep model), and writes predictions back for segmentation, value-based paid-media
bidding (Google and Meta both consume per-user pLTV), CRM suppression and targeting,
and finance (customer-based corporate valuation, @sec-clv-adjacent). The pitfalls that
recur in essentially every such system are worth stating as named failure modes:

1. *The naive constant-churn formula* (@eq-clv-naive) biases LTV and explodes as churn
   approaches zero.
2. *Survivorship bias.* Fitting a retention curve only on customers who survived to the
   present inflates retention; the censored and churned customers must be included.
3. *Look-ahead, or target, leakage.* Using information from inside the prediction window
   to build features inflates offline metrics and collapses in production. Strict
   temporal train/validation splits are mandatory.
4. *Ignoring discounting.* Undiscounted multi-year LTV overstates present value; pick
   $r$ at the data's period frequency.
5. *Cohort blending.* Mixing acquisition cohorts of different channels and vintages
   hides that paid cohorts usually churn faster than organic ones; always analyze by
   cohort.
6. *Treating predicted CLV or churn as a targeting policy.* The level of predicted churn
   is not the same as the *persuadability* of a customer (@sec-clv-causal).
7. *Extrapolating a short calibration window*, the warning made visual in
   @fig-retention-projected.
8. *Endogenous marketing spend.* Observed retention is partly a response to targeted
   spend, so the spend-retention correlation is not a causal return.

Several of these are not fixable by better data hygiene alone; they require the
classical models the rest of the chapter develops. We turn to those, starting with the
contractual-discrete benchmark.

## Retention, Survival, and the Shifted-Beta-Geometric Model {#sec-clv-sbg}

The single most consequential primitive in @eq-clv-general is the path of the
purchase probability over time. In a *contractual* setting (subscriptions,
memberships, insurance) the firm observes exactly when each customer churns, and the
object of interest is the survivor function $S(t)=\Pr(\text{still active at }t)$.
The naive approach assumes a constant retention rate $\rho$ and sets
$S(t)=\rho^{t}$, the memoryless geometric curve underlying @eq-clv-geometric and the
production formula @eq-clv-naive. This is almost always wrong in a specific and
predictable direction: observed aggregate retention rates rise with tenure.
Customers who have stayed many periods are disproportionately the intrinsically loyal
ones; the churn-prone have already left. Aggregate retention climbs not because any
individual becomes more loyal but because the surviving population is increasingly
self-selected toward loyalty, a pure *sorting* (heterogeneity) effect, not a
duration-dependence effect.

The shifted-beta-geometric model formalizes exactly this intuition and is the
discrete-time benchmark for contractual CLV [@faderhardie2007sbg]. Each customer has
a fixed but unobserved per-period churn probability $\theta$. Conditional on
$\theta$, the period of churn follows a (shifted) geometric distribution: the
customer survives each period with probability $1-\theta$ and churns with probability
$\theta$, so

$$
P(T = t \mid \theta) = \theta\,(1-\theta)^{\,t-1}, \qquad t = 1, 2, \dots
$$ {#eq-sbg-geometric}

Heterogeneity across customers is captured by mixing $\theta$ over the population
with a Beta prior, $\theta \sim \text{Beta}(\alpha,\beta)$, whose density is
$f(\theta)=\theta^{\alpha-1}(1-\theta)^{\beta-1}/B(\alpha,\beta)$. Integrating the
individual churn probability over this prior yields closed-form aggregate churn and
survivor functions that depend only on the two parameters $(\alpha,\beta)$:

$$
P(T = t) = \frac{B(\alpha+1,\;\beta+t-1)}{B(\alpha,\beta)}, \qquad
S(t) = \frac{B(\alpha,\;\beta+t)}{B(\alpha,\beta)} .
$$ {#eq-sbg-survivor}

These satisfy the convenient forward recursions
$P(1)=\alpha/(\alpha+\beta)$, $P(t)=P(t-1)\,(\beta+t-2)/(\alpha+\beta+t-1)$, and
$S(t)=S(t-1)-P(t)$, which is exactly how the estimation code below evaluates them.
The model's key qualitative property, rising aggregate retention from fixed
individual propensities, emerges purely from the Beta mixture; no individual's
$\theta$ ever changes.

**Estimation.** With cohort data giving the number of customers $n_t$ surviving at
each tenure $t$ (equivalently, the number churning, $d_t = n_{t-1}-n_t$), the
parameters are estimated by maximum likelihood. The log-likelihood for a cohort
observed through period $\tau$ sums the log-probability of each observed churn over
its churn period plus the log-survivor probability for those still active at the end
of the window:

$$
\ell(\alpha,\beta) = \sum_{t=1}^{\tau} d_t \,\log P(T=t)
   \;+\; n_{\tau}\,\log S(\tau).
$$ {#eq-sbg-loglik}

The last term is the right-censoring contribution: customers still active at $\tau$
contribute the probability of surviving *at least* that long. Identification is weak
when the window is short relative to the heterogeneity ($\alpha$ and $\beta$ are
separately identified only once the curvature of the survival curve is visible),
which is why the projections below stabilize as more months of history are used.

The following example reconstructs the analysis of a subscription business: four
cohorts ("cases") with markedly different retention dynamics, each observed for eight
months. The goal is to fit the sBG model and project retention forward. We first
visualize the raw curves in @fig-retention-observed.

```{r clv-setup}
#| message: false
#| warning: false
#| label: fig-retention-observed
#| fig-cap: "Observed monthly retention curves for four subscription cohorts. All decline, but at different rates and with different curvature."
library(tidyverse)

set.seed(13)

# Observed retention rate by months since signup (month 0 = 100%)
df_ret <- tibble(
  month_lt = 0:7,
  case01 = c(1, .531, .452, .423, .394, .375, .356, .346),
  case02 = c(1, .869, .743, .653, .593, .551, .517, .491),
  case03 = c(1, .677, .562, .486, .412, .359, .332, .310),
  case04 = c(1, .631, .468, .382, .326, .289, .262, .241)
) |>
  pivot_longer(-month_lt, names_to = "example", values_to = "retention_rate")

ggplot(df_ret, aes(month_lt, retention_rate, colour = example)) +
  geom_line() + geom_point() +
  facet_wrap(~ example) +
  scale_colour_manual(values = c("#4e79a7", "#f28e2b", "#e15759", "#76b7b2")) +
  labs(x = "Months since signup", y = "Retention rate") +
  theme_minimal() +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))
```

We next implement the sBG churn, survivor, and (negative) log-likelihood functions
following @eq-sbg-survivor and @eq-sbg-loglik, then optimize.

```{r sbg-functions}
# Aggregate churn probability P(T = t) via the forward recursion implied by
# the Beta-geometric mixture (equation @eq-sbg-survivor).
churn_bg <- Vectorize(function(alpha, beta, period) {
  p <- alpha / (alpha + beta)
  if (period > 1)
    p <- churn_bg(alpha, beta, period - 1) *
         (beta + period - 2) / (alpha + beta + period - 1)
  p
}, vectorize.args = "period")

# Survivor function S(t) = S(t-1) - P(t), with S(1) = 1 - P(1).
survival_bg <- Vectorize(function(alpha, beta, period) {
  s <- 1 - churn_bg(alpha, beta, 1)
  if (period > 1)
    s <- survival_bg(alpha, beta, period - 1) - churn_bg(alpha, beta, period)
  s
}, vectorize.args = "period")

# Negative log-likelihood (equation @eq-sbg-loglik): churners contribute
# log P(T = t); the still-active at the end contribute log S(tau).
neg_loglik <- function(par, active, lost) {
  alpha <- par[1]; beta <- par[2]
  tau <- length(active)
  -( sum(lost * log(churn_bg(alpha, beta, 1:tau))) +
       active[tau] * log(survival_bg(alpha, beta, tau)) )
}

# Convert retention rates into active/lost counts for a notional 1,000-customer
# cohort (the likelihood depends only on the shape, not the cohort size).
df_ret <- df_ret |>
  group_by(example) |>
  mutate(active = 1000 * retention_rate,
         lost   = replace_na(lag(active) - active, 0)) |>
  ungroup()
```

The instructive exercise is to ask how the *projected* retention curve evolves as
the firm accumulates more months of actual data. With only one or two months
observed, identification is weak and the projection is unreliable; as the window
lengthens, the fitted curve locks onto the truth. We refit the model for each cohort
using the first $i$ months only, for $i = 1, \dots, 7$, and overlay every projection
on the observed curve.

```{r sbg-projections}
#| label: fig-retention-projected
#| fig-cap: "Observed retention (bold) with sBG projections fit on the first i months only, for i = 1..7 (faint lines). Projections stabilize toward the observed curve as the calibration window lengthens."
project_cohort <- function(d) {
  map_dfr(1:7, function(i) {
    fit <- d |> filter(month_lt >= 1, month_lt <= i)
    opt <- optim(c(1, 1), neg_loglik,
                 active = fit$active, lost = fit$lost)
    tibble(month_lt   = 0:7,
           fact_months = i,
           retention_pred = round(
             c(1, survival_bg(opt$par[1], opt$par[2], 1:7)), 3))
  })
}

ret_preds <- df_ret |>
  group_by(example) |>
  group_modify(~ project_cohort(.x)) |>
  ungroup()

df_plot <- df_ret |>
  select(month_lt, example, retention_rate) |>
  left_join(ret_preds, by = c("month_lt", "example"))

ggplot(df_plot, aes(month_lt, retention_rate, colour = example)) +
  geom_line(aes(y = retention_pred, group = fact_months), alpha = 0.4) +
  geom_line(linewidth = 1.2) + geom_point(size = 1.2) +
  facet_wrap(~ example) +
  scale_colour_manual(values = c("#4e79a7", "#f28e2b", "#e15759", "#76b7b2")) +
  labs(x = "Months since signup", y = "Retention rate") +
  theme_minimal() +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))
```

The faint projection lines fan out wildly when fit on one month and converge onto
the observed curve as the calibration window grows, a visual diagnostic of the
identification problem in @eq-sbg-loglik. In practice this argues against committing
to a CLV figure from a single early cohort snapshot. The same identification problem,
addressed by borrowing strength across cohorts with limited information, is the subject
of @schweidel2008retention.

### From Retention to Value {#sec-clv-retention-to-value}

With a fitted survivor function, lifetime value follows directly. For a contractual
product with constant per-period margin $m$ (here a \$1 subscription) and discount
factor $r$, expected discounted value over a horizon of $N$ periods is

$$
\widehat{\text{CLV}} \;=\; m \sum_{t=0}^{N} \frac{\widehat{S}(t)}{(1+r)^{\,t}},
$$ {#eq-clv-from-survivor}

with $\widehat{S}(t)$ taken from observed data where available and from the sBG
projection beyond the calibration window. Setting $r=0$ for transparency (the firm
wants undiscounted expected revenue per signup), we compute average LTV for cohort 3
from only two months of history, projected over a 24-month horizon.

```{r clv-forecast}
fit3 <- df_ret |> filter(example == "case03", month_lt >= 1, month_lt <= 2)
opt3 <- optim(c(1, 1), neg_loglik, active = fit3$active, lost = fit3$lost)

projected <- tibble(
  month_lt       = 3:24,
  retention_pred = round(survival_bg(opt3$par[1], opt3$par[2], 3:24), 3))

ltv3 <- df_ret |>
  filter(example == "case03", month_lt <= 2) |>
  select(month_lt, retention_rate) |>
  bind_rows(projected) |>
  mutate(retention = coalesce(retention_rate, retention_pred),
         ltv_monthly = retention * 1,          # $1 subscription price
         ltv_cum     = round(cumsum(ltv_monthly), 2))

tail(ltv3, 3)
cat("Average 24-month LTV for cohort 3: $", tail(ltv3$ltv_cum, 1), "\n", sep = "")
```

The cumulative figure, roughly \$9.33 per subscriber over 24 months, blends two months
of *observed* retention with twenty-two months of *projected* retention from a
two-parameter model estimated on those same two months. That this is even possible is
the practical appeal of the probabilistic approach; that it rests on extrapolating a
heterogeneity model far beyond its calibration window is its central caveat, and
@fig-retention-projected is the warning label.

## Probabilistic Models for Customer-Base Analysis {#sec-clv-btyd}

In *non-contractual* settings, churn is never observed: a customer who has not bought
recently may have defected or may simply be between trips. The buy-till-you-die (BTYD)
family solves this by positing two latent processes, a *purchase* process active while
the customer is "alive" and a *dropout* (death) process that ends the relationship at
an unobserved time, and deriving expected future transactions and value in closed form
from each customer's *recency* (time of last purchase) and *frequency* (number of
repeat purchases). These statistics carry exactly the information needed to infer the
probability a customer is still alive.

### Pareto/NBD and BG/NBD {#sec-clv-pnbd}

The origin model is the *Pareto/NBD* [@paretonbd1987]. While alive, customer $i$
purchases according to a Poisson process with rate $\lambda_i$, and the rate varies
across the population as $\lambda \sim \text{Gamma}(r,\alpha)$ (the negative-binomial,
or NBD, half). The customer's lifetime is exponential with dropout rate $\mu_i$, and
$\mu \sim \text{Gamma}(s,\beta)$ across customers (the Pareto half). The expected number
of transactions in a future interval, and the probability that a customer is still
alive given recency and frequency, both have closed forms. The model's one
inconvenience is computational: the likelihood involves Gaussian hypergeometric
functions and can be fragile to optimize.

The *BG/NBD* model, "counting your customers the easy way," replaces the
analytically awkward Pareto death process with a beta-geometric one
[@fader2005bgnbd]. After each purchase, the customer flips a coin and, with
probability $p$, becomes inactive; $p$ varies across customers as $\text{Beta}(a,b)$.
This swap makes estimation far easier with almost identical fit, and BG/NBD is now the
single most-used BTYD model. The *Gamma-Gamma* monetary model [@fader2005rfmclv]
supplies the spend dimension: average transaction value varies within a customer
around a latent mean, and that mean is Gamma-distributed across the population, with
spend assumed independent of transaction frequency. Multiplying expected future
transactions by expected spend (and a margin rate) gives a monetary pLTV.

A unifying observation closes the loop with practice: RFM scoring and the BTYD models
are two views of the same object. @fader2005rfmclv show this with *iso-value curves*,
contours in the recency-frequency plane that group customers with different histories
but equal predicted future value.

### A Worked Non-Contractual Pipeline {#sec-clv-btyd-code}

We simulate a transaction log from the Pareto/NBD data-generating process, then fit
BG/NBD plus Gamma-Gamma with the maintained `CLVTools` package and cross-check the
transaction model against the original `BTYD` package. The simulation draws each
customer's purchase rate and lifetime from Gamma priors, runs a Poisson process up to
the (possibly censored) lifetime, and attaches Gamma-distributed spend.

```{r btyd-sim}
#| message: false
#| warning: false
set.seed(42)

n_cust  <- 1500
T_total <- 78                       # weeks of potential observation

lambda    <- rgamma(n_cust, shape = 0.55, rate = 10)  # weekly purchase rate
mu        <- rgamma(n_cust, shape = 0.40, rate = 30)  # dropout rate
cust_mean <- rgamma(n_cust, shape = 4,   rate = 0.08) # latent mean spend (~ $50)

sim_one <- function(i) {
  end <- min(rexp(1, mu[i]), T_total)   # exponential lifetime, censored at T_total
  ts <- 0; t <- 0                       # initial purchase at t = 0
  repeat {
    t <- t + rexp(1, lambda[i])         # Poisson interarrival while alive
    if (t > end) break
    ts <- c(ts, t)
  }
  tibble(cust = i, t = ts)
}

log_df <- map_dfr(seq_len(n_cust), sim_one) |>
  mutate(amount = rgamma(n(), shape = 6, rate = 6 / cust_mean[cust]),
         date   = as.Date("2020-01-01") + round(t * 7)) |>
  transmute(cust_id = paste0("C", cust), date, amount = round(amount, 2))

cat("rows:", nrow(log_df),
    " customers:", n_distinct(log_df$cust_id),
    " span:", as.character(min(log_df$date)), "to",
    as.character(max(log_df$date)), "\n")
```

We hand the log to `CLVTools`, fit the Pareto/NBD transaction model and the
Gamma-Gamma spend model on the full history, and predict a discounted per-customer CLV
52 weeks forward. We use Pareto/NBD here because `CLVTools` derives a closed-form
discounted expected residual transaction count (DERT), and hence a discounted CLV,
directly from it.

```{r btyd-clvtools}
#| message: false
#| warning: false
library(CLVTools)

clv_dat <- clvdata(log_df, date.format = "ymd", time.unit = "week",
                   name.id = "cust_id", name.date = "date", name.price = "amount")

fit_pnbd <- pnbd(clv_dat, verbose = FALSE)   # Pareto/NBD transaction model
fit_gg   <- gg(clv_dat,   verbose = FALSE)   # Gamma-Gamma spend model

round(coef(fit_pnbd), 3)

# Predict 52 weeks ahead; continuous.discount.factor is the per-week rate.
pred <- predict(fit_pnbd, predict.spending = fit_gg,
                prediction.end = 52,
                continuous.discount.factor = 0.0029,  # ~ 15% per year
                verbose = FALSE) |>
  as.data.frame()

pred |>
  transmute(Id, PAlive = round(PAlive, 3), CET = round(CET, 2),
            spend = round(predicted.mean.spending, 1),
            DERT  = round(DERT, 2),
            pLTV  = round(predicted.CLV, 2)) |>
  head(6)

cat("Mean predicted 1-year CLV: $", round(mean(pred$predicted.CLV), 2), "\n", sep = "")
```

`PAlive` is the inferred probability a customer is still active, `CET` the conditional
expected transactions over the horizon, `DERT` the discounted expected residual
transactions, and `predicted.CLV` the monetary lifetime value. For independent
confirmation, we re-estimate the BG/NBD transaction model with the original `BTYD`
package on the same log and compare the implied expected transactions.

```{r btyd-crosscheck}
#| message: false
#| warning: false
#| results: hide
library(BTYD)

end_cal <- as.Date("2020-01-01") + 39 * 7   # split calibration / holdout

elog <- log_df |>
  transmute(cust = cust_id, date = as.Date(date), sales = amount) |>
  as.data.frame()

elog_m  <- dc.MergeTransactionsOnSameDate(elog)
cbs_cbt <- dc.ElogToCbsCbt(elog_m, per = "week", T.cal = end_cal)
cal_cbs <- cbs_cbt$cal$cbs

params_bg <- bgnbd.EstimateParameters(cal_cbs)
cet_btyd  <- bgnbd.ConditionalExpectedTransactions(
  params_bg, T.star = 13,
  x = cal_cbs[, "x"], t.x = cal_cbs[, "t.x"], T.cal = cal_cbs[, "T.cal"])
```

```{r btyd-crosscheck-out}
cat("BTYD BG/NBD params (r, alpha, a, b):", round(params_bg, 3), "\n")
cat("Mean 13-week expected transactions (BTYD BG/NBD):",
    round(mean(cet_btyd), 3), "\n")
```

The two packages, two model variants, and two estimation routines deliver coherent
parameters and forecasts from the same simulated log, which is the trust-building step
before any of these numbers reaches a paid-media bid. We visualize the BTYD intuition
directly in @fig-clv-palive: the probability a customer is alive as a function of how
many times they have bought and how long ago they last appeared.

```{r palive-heatmap}
#| label: fig-clv-palive
#| fig-cap: "Probability a customer is still alive under the fitted BG/NBD model, over the recency-frequency grid. Many purchases with a recent last visit imply near-certain life; a long silence after few purchases implies likely dropout."
#| message: false
#| warning: false
grid <- expand_grid(x = 0:10, t.x = seq(0, 39, by = 3)) |>
  mutate(palive = bgnbd.PAlive(params_bg, x = x, t.x = t.x, T.cal = 39))

ggplot(grid, aes(t.x, x, fill = palive)) +
  geom_tile() +
  scale_fill_viridis_c(name = "P(alive)", limits = c(0, 1)) +
  labs(x = "Recency: time of last purchase (weeks)",
       y = "Frequency: number of repeat purchases") +
  theme_minimal()
```

The contours in @fig-clv-palive are the iso-value intuition of @fader2005rfmclv made
concrete: customers with very different recency-frequency histories can share the same
P(alive), and therefore the same future value.

## Econometric and Hierarchical-Bayes CLV {#sec-clv-econometric}

Where BTYD models are deliberately covariate-light, the econometric tradition predicts
the *components* of @eq-clv-general (incidence, frequency, spend, churn) with
regression and hierarchical-Bayes machinery and rich covariates such as demographics,
marketing touches, channel, and macro conditions. Its strength is targeting and
what-if analysis on marketing levers; its weakness is the number of models to maintain
and validate.

The bridge between the two traditions is @abe2009, which recasts the Pareto/NBD as a
hierarchical Bayes model. This yields individual-level posteriors, honest uncertainty,
and, crucially, the ability to let covariates shift the purchase and dropout rates,
something the closed-form Pareto/NBD cannot do. On the pure component-modeling side,
@venkatesankumar2004 build a covariate-rich CLV used to rank and select customers and
to allocate contact across channels, and @reinartzkumar2003 model *profitable* lifetime
duration with a hazard specification, the econometric face of retention.

Two themes from this literature carry practical weight. First, acquisition and retention
are not independent: @schweidel2008bivariate model jointly *when* a customer is acquired
and *how long* they then stay, and show that ignoring the correlation biases CLV.
Second, elaboration does not always pay. @donkers2007clv run a model tournament in
insurance and find that simple status-quo and Markov-chain models often match or beat
far more elaborate ones out of sample, an empirical-humility check on the impulse to
add structure.

The Markov-chain formulation [@pfeifercarraway2000] is worth stating in full because it
links CLV to the decision-theoretic view. Define a small set of states (for example,
active, lapsed, lost), a transition matrix $P$, and a per-period reward vector $\mathbf{r}$
giving the margin earned in each state. With discount factor $\gamma = 1/(1+r)$, the
vector of state-conditional lifetime values solves the linear system

$$
\mathbf{V} \;=\; (\mathbf{I} - \gamma P)^{-1}\,\mathbf{r},
$$ {#eq-clv-markov}

the discounted expected reward of a customer currently in each state. @tbl-clv-markov
computes it for a three-state migration model; @fig-clv-states draws the transition
structure the code encodes.

```{mermaid}
%%| label: fig-clv-states
%%| fig-cap: "The three-state customer-migration process behind the Markov-chain CLV. Only the active state earns margin; the lapsed state can be reactivated but leaks heavily into the absorbing lost state."
flowchart LR
  ACT["Active<br/>(reward 100)"]
  LAP["Lapsed<br/>(reward 0)"]
  LST["Lost<br/>(reward 0, absorbing)"]
  ACT -->|"0.70"| ACT
  ACT -->|"0.20"| LAP
  ACT -->|"0.10"| LST
  LAP -->|"0.30"| ACT
  LAP -->|"0.40"| LAP
  LAP -->|"0.30"| LST
  LST -->|"1.00"| LST
```

```{r markov-clv}
#| message: false
states <- c("active", "lapsed", "lost")
P <- matrix(c(0.70, 0.20, 0.10,
              0.30, 0.40, 0.30,
              0.00, 0.00, 1.00),
            nrow = 3, byrow = TRUE, dimnames = list(states, states))
reward <- c(active = 100, lapsed = 0, lost = 0)
gamma  <- 1 / 1.10                                   # 10% per-period discount
V <- solve(diag(3) - gamma * P) %*% reward           # equation @eq-clv-markov
```

```{r markov-clv-tbl}
#| echo: false
#| label: tbl-clv-markov
#| tbl-cap: "Markov-chain migration CLV by current state. An active customer is worth far more than a lapsed one because the lapsed state carries a high probability of transitioning to the absorbing lost state."
knitr::kable(
  tibble(`Current state` = states, `CLV ($)` = round(V[, 1], 2))
)
```

@tbl-clv-markov shows the absorbing structure at work: an active customer is worth
\$350 because they earn margin now and are likely to keep earning, a lapsed customer
only \$150 because the lapsed state leaks heavily into the absorbing lost state, and a
lost customer nothing. This is the same discounting logic as @eq-clv-from-survivor,
expressed as a controllable Markov reward process, and it is the conceptual link to the
decision models of @sec-clv-causal.

## The Machine-Learning and Deep-Learning Frontier {#sec-clv-ml}

The production frontier has moved from closed-form probability models to supervised
machine learning and deep learning on rich feature sets and raw event sequences.
Several recurring design patterns organize the space.

The dominant production pattern is the *two-stage churn-times-spend* (or
propensity-times-value) decomposition. A first-stage classifier (logistic regression,
gradient boosting, random forest) predicts the probability a customer is active or
repeat-purchasing over the horizon; a second-stage regressor predicts conditional
spend; their product, times margin and discounted, is pLTV. This mirrors the BTYD
frequency-times-monetary split but adds covariates and nonlinearity. The benchmark for
*whether the modeling effort pays off* is the churn-model tournament of
@neslin2006churn, whose central finding ("methods matter") is that modest gains in
predictive accuracy translate into six-figure swings in campaign profit.

We demonstrate the two-stage pattern on simulated customer features, using a random
forest for the active/inactive classifier and an elastic-net (`glmnet`) regression for
log-spend among actives, then combine the two heads.

```{r ml-two-stage}
#| message: false
#| warning: false
library(glmnet)
library(randomForest)

set.seed(11)
N <- 5000
feat <- tibble(
  recency   = runif(N, 0, 52),
  frequency = rpois(N, 3),
  monetary  = rgamma(N, 4, 0.08),
  tenure    = runif(N, 1, 104),
  promo     = rbinom(N, 1, 0.4))

eta      <- -1 + 0.03 * feat$frequency - 0.02 * feat$recency + 0.4 * feat$promo
active   <- rbinom(N, 1, plogis(eta))
spend    <- active * rgamma(N, shape = 3, rate = 3 / (feat$monetary * 0.5))
df_ml    <- feat |> mutate(active = factor(active), spend = spend)

tr <- sample(N, 0.7 * N); te <- setdiff(seq_len(N), tr)

# Stage 1: random-forest active/inactive classifier.
rf <- randomForest(active ~ recency + frequency + monetary + tenure + promo,
                   data = df_ml[tr, ], ntree = 200)
p_active_hat <- predict(rf, df_ml[te, ], type = "prob")[, "1"]

# Stage 2: elastic-net log-spend regression among training actives.
act_tr <- tr[df_ml$active[tr] == "1"]
xtr <- model.matrix(~ recency + frequency + monetary + tenure + promo,
                    df_ml[act_tr, ])[, -1]
ytr <- log(df_ml$spend[act_tr])
cvfit <- cv.glmnet(xtr, ytr, alpha = 0.5)

xte <- model.matrix(~ recency + frequency + monetary + tenure + promo,
                    df_ml[te, ])[, -1]
resid_var <- as.numeric(var(ytr - predict(cvfit, xtr, s = "lambda.min")))
spend_hat <- exp(as.numeric(predict(cvfit, xte, s = "lambda.min")) + 0.5 * resid_var)

pltv_ml <- p_active_hat * spend_hat * 0.6   # 60% margin
cat("Mean two-stage pLTV: $", round(mean(pltv_ml), 2), "\n", sep = "")
cat("Correlation with realized margin:",
    round(cor(pltv_ml, df_ml$spend[te] * 0.6), 3), "\n")
```

The lognormal back-transform includes the variance correction $\exp(0.5\sigma^2)$,
without which expected spend on the raw scale is systematically understated, a common
production bug. The positive correlation between predicted pLTV and realized margin on
the held-out customers confirms the two heads compose sensibly.

A more integrated approach uses a single probabilistic head that models the spike at
zero (non-buyers) and the heavy right tail (a few high spenders) jointly. The
*zero-inflated lognormal* (ZILN) loss, popularized by a Google deep-learning model
[@wang2019ziln], does exactly this: one model outputs a churn logit, a lognormal mean,
and a lognormal scale, and is trained on the combined likelihood. We fit a small ZILN
model by maximum likelihood (no neural network required to make the point) and recover
both the structure and a calibrated expected value. @fig-clv-ml-heads contrasts the
joint ZILN head with the two-stage decomposition.

```{mermaid}
%%| label: fig-clv-ml-heads
%%| fig-cap: "Two production architectures for machine-learning pLTV. The two-stage pattern multiplies a first-stage activity classifier by a second-stage conditional-spend regressor; the zero-inflated lognormal alternative trains one model on the combined likelihood of the zero spike and the heavy right tail."
flowchart TB
  FT["Customer features and event history"]
  subgraph TS["Two-stage: churn-times-spend"]
    C1["Stage 1 classifier:<br/>probability active over the horizon"]
    R1["Stage 2 regressor:<br/>conditional spend among actives"]
    PD["Product, times margin,<br/>discounted: pLTV"]
    C1 --> PD
    R1 --> PD
  end
  subgraph ZS["Joint head: zero-inflated lognormal"]
    ZH["One model outputs a churn logit,<br/>a lognormal mean, and a<br/>lognormal scale"]
    ZE["Trained on the combined likelihood:<br/>spike at zero plus heavy right tail"]
    ZH --> ZE
  end
  FT --> C1
  FT --> R1
  FT --> ZH
```

```{r ziln}
#| message: false
#| warning: false
set.seed(21)
M <- 6000
x      <- runif(M, -2, 2)
p_buy  <- plogis(-0.5 + 0.9 * x)
buy    <- rbinom(M, 1, p_buy)
spend2 <- buy * rlnorm(M, meanlog = 3 + 0.5 * x, sdlog = 0.6)

# ZILN negative log-likelihood: a churn (zero-inflation) head and a lognormal
# value head, each linear in x here for transparency.
ziln_nll <- function(par) {
  p  <- plogis(par[1] + par[2] * x)         # P(buy)
  ml <- par[3] + par[4] * x                 # lognormal mean (log scale)
  s  <- exp(par[5])                         # lognormal scale
  ll <- ifelse(spend2 == 0,
               log1p(-p),
               log(p) + dlnorm(pmax(spend2, 1e-8), ml, s, log = TRUE))
  -sum(ll)
}

fit_ziln <- optim(c(0, 0, 3, 0, log(0.6)), ziln_nll, method = "BFGS")
p_hat  <- plogis(fit_ziln$par[1] + fit_ziln$par[2] * x)
E_buy  <- p_hat * exp((fit_ziln$par[3] + fit_ziln$par[4] * x) +
                        exp(fit_ziln$par[5])^2 / 2)
cat("ZILN params (b0, b1, m0, m1, log-sigma):",
    round(fit_ziln$par, 3), "\n")
cat("Mean predicted E[spend]: $", round(mean(E_buy), 2),
    "  vs actual: $", round(mean(spend2), 2), "\n", sep = "")
```

The fitted ZILN expected value matches the simulated mean closely because the model
respects the data-generating structure: it does not try to force a single distribution
onto a population that is half non-buyers and half a right-skewed spending tail.

Beyond these patterns, the frontier includes *sequence models* (RNN, LSTM, seq2seq,
transformers) that consume the raw chronological event stream and learn temporal
patterns hand-built RFM features miss [@bauerjannach2021seq2seq]; *customer embeddings*
that learn dense vector representations from product and event co-occurrence and
transfer information across sparse customers, as in the ASOS production system
[@chamberlain2017embeddings]; graph neural networks for marketplace and advertising
LTV; and *early-LTV* models that predict a mobile game player's lifetime value (and
identify "whales") from the first hours of play [@sifa2015games]. A caveat is in order:
most of this work lives in KDD, RecSys, AAAI, and arXiv rather than the top marketing
journals, and the marketing-journal contribution is concentrated on the *decision* side,
to which we now turn.

## Causal and Decision-Focused CLV {#sec-clv-causal}

Predicting CLV is not the same as acting on it, and the top-journal frontier here is
about *incremental* value and resource allocation rather than point prediction. This is
where the "endogenous marketing" pitfall of @sec-clv-pitfalls becomes a research program.

The landmark result is *retention futility* [@ascarza2018futility]: the
highest-churn-risk customers are often not the most *persuadable*, so targeting
retention spend on predicted churn (or predicted CLV) wastes money on customers who
would either leave regardless or stay regardless. The correct target is the *uplift*,
the individual treatment effect of the intervention, validated by Ascarza in two field
experiments. We make the point concrete by simulating a retention treatment whose effect
is uncorrelated with churn risk and comparing a risk-targeting policy against an
uplift-targeting policy.

```{r uplift}
#| message: false
set.seed(3)
M2     <- 8000
risk   <- runif(M2)                       # churn risk
uplift <- rnorm(M2, 0.04, 0.03)           # treatment effect, ~independent of risk
margin <- 200

top_quintile <- function(v) v >= quantile(v, 0.8)
gain_risk   <- mean(uplift[top_quintile(risk)])   * margin
gain_uplift <- mean(uplift[top_quintile(uplift)]) * margin

cat(sprintf("cor(risk, uplift) = %.3f\n", cor(risk, uplift)))
cat(sprintf("Per-target retained margin: risk-targeting $%.2f vs uplift-targeting $%.2f\n",
            gain_risk, gain_uplift))
```

With risk and uplift essentially uncorrelated, the uplift-targeting policy roughly
doubles the retained margin per treated customer relative to the intuitive
target-the-high-risk policy, which is Ascarza's point in miniature. The complementary
methodological advance is off-policy evaluation: @simester2020policy show how to assign
customers to *actions* and then evaluate *any* targeting policy, including a
CLV-maximizing one, without re-running an experiment for each candidate policy. At the
budget level, @reinartzthomaskumar2005 jointly optimize acquisition and retention spend
against a CLV objective, the canonical resource-allocation-to-CLV result. The throughline
is that the decision-relevant quantity is the *derivative* of CLV with respect to a
marketing action, not its level, and recovering that derivative requires experimental or
quasi-experimental variation rather than the historical correlation between spend and
value.

## Adjacent and Related Constructs {#sec-clv-adjacent}

CLV sits at the center of a constellation of constructs, summarized in
@tbl-clv-constructs, that extend, aggregate, or operationalize it.

| Construct | One-line definition | Anchor reference |
|---|---|---|
| Customer equity (CE) | Sum of all customers' CLV; the demand-side asset on a quasi-balance-sheet | @rustLemonZeithaml2004 |
| Valuing customers to value firms | CE drives firm value; lets one value high-growth, negative-earnings firms | @guptaLehmannStuart2004 |
| CLV modeling synthesis | The review defining and surveying implementable CLV | @gupta2006modeling |
| CE to market capitalization | Aggregated CLV tracks shareholder value | @kumarshah2009 |
| CBCV, contractual | Value a public subscription firm from disclosed customer metrics | @mccarthy2017 |
| CBCV, non-contractual | Same, for non-subscription firms, from a BTYD model on disclosed data | @mccarthyfader2018 |
| Customer referral value (CRV) | Value a customer creates by referring others | @kumarCRV2010 |
| Customer engagement value (CEV) | CLV plus CRV plus influencer and knowledge value | @kumar2010cev |
| Negative CLV | Loyal is not the same as profitable; some long-tenured customers destroy value | @reinartzkumar2002hbr |
| RFM | The classic scoring heuristic, equivalent to a BTYD view | @fader2005rfmclv |
| Churn / retention prediction | The predictive substrate of CLV | @neslin2006churn |
| Structural CLV (CPG) | CLV from a utility and budget model with brand switching | @sunder2016measuring |

: The constructs adjacent to customer lifetime value. {#tbl-clv-constructs}

*Customer equity* is the headline aggregate: the sum of every customer's CLV, treated
as the demand-side asset of the firm. @rustLemonZeithaml2004 build it into a "return on
marketing" framework that traces marketing actions through value, brand, and relationship
equity to customer equity. The link to firm value is the bridge to the
marketing-finance interface: @guptaLehmannStuart2004 show that valuing customers lets one
value the firm itself, which is decisive for high-growth, negative-earnings companies whose
worth lives entirely in their customer base, and @kumarshah2009 demonstrate that aggregated
CLV tracks market capitalization. The most direct realization of this idea is
*customer-based corporate valuation* (CBCV), which values a public company from disclosed
customer metrics: for contractual subscription firms [@mccarthy2017] and, using a BTYD
model fit to disclosed transaction data, for non-contractual firms [@mccarthyfader2018].

Two constructs widen the definition of customer value beyond the focal customer's own
purchases. *Customer referral value* extends it to the customers a person brings in,
developed in detail in @sec-clv-crv. *Customer engagement value* widens it further to
include a customer's influencer value (social-media and word-of-mouth reach) and
knowledge value (feedback and co-creation), so that the "total" engagement asset is CLV
plus CRV plus influencer plus knowledge value [@kumar2010cev].

A construct that runs the other way is *negative CLV*. Loyalty is not the same as
profitability: some long-tenured customers consume more service, discount, and support
than they generate in margin, and @reinartzkumar2002hbr document this "mismanagement of
customer loyalty" directly. The practical corollary is that a firm should be willing to
*demarket* to, or fire, its unprofitable customers, a recommendation that follows from
@eq-clv-general the moment the marketing-cost term exceeds the margin term over the
horizon. Finally, *share of wallet*, the fraction of a customer's category spend the firm
captures, complements CLV by indicating headroom: a high-CLV customer with low share of
wallet is an expansion opportunity, the customer-level analogue of the net-revenue-retention
expansion in @sec-clv-saas-metrics.

## Customer Referral Value {#sec-clv-crv}

CLV as defined so far captures only the margin a customer generates through their
*own* purchases. A customer also creates value by bringing in others, through
word of mouth and formal referral programs, and this *customer referral value* (CRV)
is empirically distinct from CLV: high-CLV customers are frequently not the best
referrers, so the two must be managed separately [@kumarCRV2010]. The managerial
implication is a two-by-two *customer value matrix* crossing CLV (low/high) with
CRV (low/high): "champions" score high on both, "affluents" buy heavily but refer
little, "advocates" refer well but buy little, and the remainder warrant minimal
investment. The diagnostic point is that targeting on CLV alone leaves the referral
asset unmanaged.

Measuring referral value requires separating two kinds of referred customer
[@kumarCRV2010]. A *type-one* referral is incremental: the referred customer would
not have joined absent the referral, so their entire margin is attributable to the
referrer. A *type-two* referral would have joined anyway through other channels, so
the referrer's contribution is not the customer's margin but the *acquisition cost
the firm saved* by not having to recruit them through paid channels. Empirically the
split is roughly even between the two types. Letting $n_1$ index the
incremental (type-one) referrals and $n_1{+}1,\dots,n_2$ index the would-have-joined
(type-two) referrals,

$$
\text{CRV}_i \;=\; \sum_{t=1}^{T}\sum_{y=1}^{n_1}
   \frac{A_{ty} - a_{ty} - M_{ty} + \text{ACQ1}_{ty}}{(1+r)^{t}}
   \;+\; \sum_{t=1}^{T}\sum_{y=n_1+1}^{n_2}
   \frac{\text{ACQ2}_{ty}}{(1+r)^{t}},
$$ {#eq-crv}

where $A_{ty}$ is the gross margin from incremental customer $y$, $a_{ty}$ the cost
of the referral incentive, $M_{ty}$ the marketing cost to retain the referred
customer, $\text{ACQ1}_{ty}$ the acquisition cost saved on incremental customers,
$\text{ACQ2}_{ty}$ the acquisition cost saved on would-have-joined customers, and $r$
the per-period discount factor (again $\approx 0.07238$ semiannually for a 15% annual
rate [@Kumar_2008]). The first sum books full net margin for incremental customers;
the second books *only* avoided acquisition cost for the rest. A practical detail
governs the attribution window: referrals attributable to an incentive campaign
accrue for about a year after it runs, so CRV is typically accumulated over a
one-year horizon and referrals beyond it are treated as organic [@kumarCRV2010].

## Unit Economics and the Discounting of a CLV Stream {#sec-clv-unit-econ}

We close the production thread of @sec-clv-production with three quantitative
demonstrations: the LTV:CAC and payback dashboard, the net-revenue-retention waterfall,
and a discounting and horizon sensitivity analysis that quantifies the chapter's opening
claim that roughly half the asset sits beyond the first year.

@tbl-clv-unit-econ computes the board-level ratios per acquisition channel from simulated
channel economics, using the geometric LTV of @eq-clv-naive (here with $r=0$ for
transparency).

```{r unit-econ}
#| message: false
channels <- tribble(
  ~channel,       ~cac, ~m_month, ~churn,
  "Organic",        20,       18,   0.03,
  "Paid Search",    95,       20,   0.06,
  "Social",        140,       16,   0.09,
  "Affiliate",      60,       14,   0.05)

unit <- channels |>
  mutate(ltv        = m_month / churn,             # geometric LTV, r = 0
         ltv_cac    = round(ltv / cac, 2),
         payback_mo = round(cac / m_month, 1))
```

```{r unit-econ-tbl}
#| echo: false
#| label: tbl-clv-unit-econ
#| tbl-cap: "Unit economics by acquisition channel. Organic acquisition is wildly efficient; the Social channel, with an LTV:CAC near 1, barely recovers its acquisition cost and would be a candidate to cut or rework."
knitr::kable(
  unit |> transmute(Channel = channel, CAC = cac, `Margin/mo` = m_month,
                    Churn = churn, LTV = round(ltv, 0),
                    `LTV:CAC` = ltv_cac, `Payback (mo)` = payback_mo)
)
```

@tbl-clv-unit-econ makes the resource-allocation logic immediate: the Social channel,
with an LTV:CAC ratio near $1{:}1$ and an payback of nearly nine months, barely earns
back its acquisition cost and is the first candidate for rework, while Organic at $30{:}1$
is so efficient it likely signals under-investment in scaling it.

Next we render the gross- and net-revenue-retention decomposition of @sec-clv-saas-metrics
as a waterfall in @fig-clv-nrr, showing how expansion revenue pushes net retention above
100% even as gross retention leaks below it.

```{r nrr-waterfall}
#| label: fig-clv-nrr
#| fig-cap: "Net revenue retention waterfall. Starting MRR loses churn and contraction (GRR falls below 100%) but gains expansion, lifting NRR above 100%: the base grows with no new logos."
#| message: false
start_mrr   <- 100
churn       <- -5
contraction <- -3
expansion   <- 15
grr <- (start_mrr + churn + contraction) / start_mrr
nrr <- (start_mrr + churn + contraction + expansion) / start_mrr

wf <- tibble(
  step  = factor(c("Start MRR", "Churn", "Contraction", "Expansion", "End MRR"),
                 levels = c("Start MRR", "Churn", "Contraction",
                            "Expansion", "End MRR")),
  delta = c(start_mrr, churn, contraction, expansion, NA)) |>
  mutate(end   = c(start_mrr, cumsum(delta[1:4])[-1], NA),
         end   = ifelse(step == "End MRR", cumsum(delta[1:4]), end),
         start = lag(end, default = 0),
         start = ifelse(step %in% c("Start MRR", "End MRR"), 0, start),
         end   = ifelse(step %in% c("Start MRR", "End MRR"),
                        ifelse(step == "Start MRR", start_mrr,
                               start_mrr + churn + contraction + expansion), end),
         fill  = case_when(step %in% c("Start MRR", "End MRR") ~ "total",
                           delta >= 0 ~ "gain", TRUE ~ "loss"))

ggplot(wf, aes(step, fill = fill)) +
  geom_rect(aes(xmin = as.numeric(step) - 0.4, xmax = as.numeric(step) + 0.4,
                ymin = pmin(start, end), ymax = pmax(start, end))) +
  scale_fill_manual(values = c(total = "#4e79a7", gain = "#59a14f",
                               loss = "#e15759"), guide = "none") +
  labs(x = NULL, y = "MRR ($000s)",
       subtitle = sprintf("GRR = %.0f%%   |   NRR = %.0f%%",
                          100 * grr, 100 * nrr)) +
  theme_minimal()
```

@fig-clv-nrr shows GRR at 92% (pure leakage from churn and contraction) but NRR at 107%,
because expansion of \$15k outpaces the \$8k of losses. A subscription business with this
profile grows 7% a year on its existing base alone.

Finally, @tbl-clv-dcf tabulates discounted CLV across discount rates and horizons for a
fixed retention curve, quantifying how much value lies beyond the first year and how much
discounting and horizon choice move the number.

```{r dcf-sensitivity}
#| message: false
S_curve <- 0.85^(0:60)   # monthly retention curve, 15% monthly churn
m       <- 100           # monthly margin

grid <- expand_grid(r_annual = c(0, 0.05, 0.10, 0.15),
                    N = c(12, 24, 60)) |>
  rowwise() |>
  mutate(clv = round(sum(m * S_curve[1:(N + 1)] /
                           (1 + r_annual / 12)^(0:N)), 0)) |>
  ungroup()

frac_beyond_y1 <- 1 - sum(m * S_curve[1:13]) / sum(m * S_curve)
```

```{r dcf-tbl}
#| echo: false
#| label: tbl-clv-dcf
#| tbl-cap: "Discounted CLV ($) by annual discount rate and horizon for a fixed monthly retention curve. The horizon matters more than the discount rate at these parameters; lengthening from 12 to 60 months adds more value than any plausible change in r removes."
knitr::kable(
  grid |>
    pivot_wider(names_from = N, values_from = clv, names_prefix = "N=") |>
    rename(`r (annual)` = r_annual)
)
```

```{r dcf-frac}
cat("Fraction of (undiscounted) lifetime value beyond month 12: ",
    round(frac_beyond_y1, 3), "\n", sep = "")
```

@tbl-clv-dcf separates the two levers. Moving the horizon from 12 to 60 months adds far
more value than any plausible discount rate removes, which is the quantitative form of the
warning against truncating the horizon at one year. At a steeper retention curve the
fraction of value beyond year one would be larger still; the "roughly half the asset"
heuristic in the chapter's opening corresponds to a flatter, more loyal customer base than
this illustrative 15%-monthly-churn curve.

## Identification and What Breaks It {#sec-clv-identification}

CLV is arithmetically simple and inferentially treacherous. Every term in
@eq-clv-general is a forecast, and the headline number compounds their errors. The
recurring failure modes are worth stating as assumptions whose violation is
diagnostic, and each maps onto a production pitfall from @sec-clv-pitfalls.

*Constant-retention bias.* Assuming a fixed $\rho$ (@eq-clv-geometric, @eq-clv-naive)
when the true process is heterogeneous understates the value of long-tenured customers
and overstates churn for survivors, because it misreads population sorting as individual
duration dependence. The sBG model (@sec-clv-sbg) exists precisely to correct this; the
symptom is a geometric model that systematically underpredicts late-tenure survival.

*Horizon and discounting.* Because a large share of lifetime value sits beyond the
first year (@tbl-clv-dcf), the choice of $N$ and $r$ is not a rounding decision. The
infinite-horizon shortcut in @eq-clv-geometric requires $\rho < 1+r$ and a genuinely
stationary margin; the Gordon-style terminal value used in valuation work is valid
only when the growth rate is strictly below the discount rate. A short horizon
truncates the asset; an over-long one extrapolates a calibration-window model past
its evidence (@fig-retention-projected).

*Endogenous marketing.* The marketing cost $\text{MC}_{ij}$ is not exogenous: firms
direct more spend at customers they already believe are valuable. Treating observed
$\text{MC}$ and observed retention as if the latter were the causal response to the
former conflates targeting with treatment effect, biasing the implied return on
retention spending. Clean estimates of how marketing *moves* CLV require experimental
or quasi-experimental variation, not the historical correlation between spend and
value, which is the program of @sec-clv-causal.

*Aggregation and the non-contractual problem.* In non-contractual settings (retail,
e-commerce) churn is never observed: a customer who has not purchased recently may
have defected or may simply be between trips. Disentangling "dead" from "dormant"
requires a latent-attrition model such as the Pareto/NBD [@paretonbd1987], whose
recency-frequency sufficient statistics carry exactly the information needed to infer
the probability a customer is still alive (@fig-clv-palive). Ignoring latent attrition
and counting all lapsed customers as retained inflates customer equity.

*Survivorship and leakage.* Two data-engineering errors mimic good news. Fitting a
retention curve only on customers who survived to the present inflates retention
(survivorship bias); building features from inside the prediction window inflates
offline accuracy that then collapses in production (look-ahead leakage). Both are
defeated only by strict temporal discipline in constructing the training data.

*Heterogeneous response and structural drivers.* Reduced-form component models give
per-customer forecasts but say little about *why* value differs or how it responds to
prices and competition. Structural approaches embed CLV in a model of consumer
utility: @sunder2016measuring derive brand- and category-level CLV in consumer
packaged goods by accounting for multiple-discreteness (consumers buy several brands),
brand-switching, and a latent budget constraint inferred from transaction data alone
via Bayesian estimation, and show that ignoring the budget constraint, as simpler
heuristics do, mismeasures value and the response to asymmetric price changes. The
gain in behavioral fidelity is bought with strong functional-form assumptions; the
trade-off between the two traditions in @tbl-clv-approaches has no free lunch.

The financial relevance of getting these right is direct: aggregated CLV is *customer
equity*, and the demonstrated link from customer equity to firm value
[@gupta2006modeling; @kumar2010; @kumarshah2009] is the bridge from this chapter to the
marketing-finance interface developed in @sec-marketing-finance.

## Key Takeaways {#sec-clv-takeaways}

- CLV (@eq-clv-general) treats a customer as a forward-looking asset: the discounted
  present value of future margin from their purchases net of the marketing cost of
  retaining them. Aggregated across customers it is *customer equity*, which tracks firm
  value.
- The contractual versus non-contractual and discrete versus continuous taxonomy
  (@tbl-clv-taxonomy) decides which model class is admissible before any estimation begins.
- The naive production formula (@eq-clv-naive), ARPA times margin over churn, is the
  industry default and is biased in predictable ways: it understates survivor value,
  explodes as churn approaches zero, and ignores discounting and expansion revenue.
- The constant-retention shortcut is biased whenever customers are heterogeneous;
  observed aggregate retention rises with tenure because of *sorting*, not increasing
  individual loyalty. The shifted-beta-geometric model (@eq-sbg-survivor) captures this
  with a two-parameter Beta mixture and estimates cleanly from sparse cohort data.
- In non-contractual settings, buy-till-you-die models (Pareto/NBD, BG/NBD) plus the
  Gamma-Gamma spend model infer a probability-alive and a monetary pLTV from
  recency-frequency-monetary data alone (@fig-clv-palive), and agree across the
  `CLVTools` and `BTYD` implementations.
- The machine-learning frontier decomposes pLTV into a churn head and a spend head
  (two-stage, or a joint ZILN loss), trading the parsimony of probability models for
  covariates and nonlinearity; "methods matter" only insofar as accuracy gains move
  campaign profit.
- Acting on CLV requires *uplift*, not level: targeting the highest-risk or highest-value
  customers can be futile (@ascarza2018futility) when risk and persuadability are
  uncorrelated.
- The hard part is identification, not arithmetic: constant-retention bias, horizon and
  discounting choices, endogenous marketing spend, survivorship and leakage, and latent
  attrition in non-contractual settings each bias the headline number in a predictable
  direction.
