# Celebrity Endorsement {#sec-celebrity-endorsement}

A *celebrity endorser* is an individual who enjoys public recognition and lends
that recognition to a consumer good by appearing with it in advertising
[@mccracken1989]. Firms spend heavily on this practice because a famous face
appears to do something a product claim cannot: it imports ready-made meaning—
status, competence, glamour, rebelliousness—into a brand that would otherwise
have to manufacture those associations slowly and at great cost. The endorsement
contract is, in effect, the rental of an accumulated reputation. The central
empirical questions are therefore the same ones we ask of any intangible asset:
*through what mechanism does the asset create value, how is that value measured,
and what is the risk that it evaporates?*

This chapter treats celebrity endorsement as a problem with three faces. The
first is **psychological**: why does an endorser change what a consumer thinks
and does? Two classical models—source credibility and source attractiveness—
locate the answer in properties of the *source* (expertise, trust, likability,
similarity). A third, the **meaning-transfer** model, relocates it in *culture*:
the endorser is a conduit that moves symbolic meaning from the culturally
constituted world into the brand and finally into the consumer's own identity
project [@mccracken1986; @mccracken1989]. The second face is **strategic**: which
celebrity should endorse which brand? The *match-up hypothesis* answers that
congruence between endorser and product governs effectiveness, and we give it a
formal, testable structure. The third face is **financial**: endorsement
contracts are firm-level investments whose announcement moves stock prices, whose
returns can be estimated by event study, and whose downside—**endorser risk**—is
the possibility that a scandal contaminates every brand the celebrity touches.

By the end of the chapter the reader should be able to state each model's
construct definitions formally, specify and estimate a match-up interaction,
design an event study that isolates the firm-value effect of an endorsement (and
of an endorser's disgrace), and reason quantitatively about whether an
endorsement is worth its price. Throughout, intuition leads and formalism
follows immediately, in the house manner. Celebrity endorsement is best understood
as a special case of the broader machinery developed in @sec-branding (brands as
networks of meaning and as priced firm assets) and @sec-marketing-finance
(event-study valuation of marketing actions); we draw on both.

## Conceptual Foundations

The persuasion literature offers two distinct reasons a message recipient might
be moved by *who* delivers a message rather than *what* the message says. The
distinction matters because the two routes have different boundary conditions and
different vulnerabilities, and because the dominant modern model—meaning
transfer—rejects both as incomplete.

### The Source Credibility Model

The **source credibility** model holds that a communicator's persuasiveness rises
with the audience's perception that the source both *knows* the truth and is
*willing to tell it*. Following the formalization that @mccracken1989 inherits
from the Hovland tradition, credibility has two components. **Expertise** (or
expertness) is the perceived ability of the source to make valid assertions—the
degree to which the endorser is seen as a legitimate source of knowledge about
the product class. **Trustworthiness** is the perceived willingness of the source
to make valid assertions—the audience's confidence that the endorser is reporting
honestly rather than shilling. A message is most effective when the source scores
high on both: a credentialed dermatologist endorsing a skin cream is persuasive
to the extent the audience believes she knows skin *and* is not merely paid to
say so.

> An endorser is effective to the extent the audience perceives the source as
> able to make valid assertions (expertise) and willing to make them
> (trustworthiness); credibility is the conjunction of the two.
> — after the source-credibility tradition synthesized in @mccracken1989

The model's empirical workhorse is the **Ohanian source scale**, which
operationalizes expertise, trustworthiness, and attractiveness as reflective
multi-item semantic-differential batteries.[^ohanian] Its appeal is that it
yields a clean mediation story: endorser characteristics raise perceived
credibility, which raises attitude toward the ad ($A_{ad}$), which raises
attitude toward the brand ($A_b$) and purchase intention. Its limitation, which
motivated the meaning-transfer turn, is that credibility ratings poorly predict
why a *non-expert* celebrity—an actor with no plausible product knowledge—can
nonetheless sell cars, watches, and soft drinks.

[^ohanian]: The 15-item credibility scale (attractiveness, trustworthiness,
expertise) is the field's most-used instrument; we describe its structure rather
than reproduce it, since the items are reflective indicators of three latent
factors and the construct logic is what matters here.  The source-credibility scale that operationalizes expertise, trustworthiness, and attractiveness as separable dimensions is @ohanian1990constructvalidatio.

### The Source Attractiveness Model

The **source attractiveness** model locates persuasion not in epistemic
authority but in the audience's *affinity* for the source. It decomposes
attractiveness into three perceived properties [@mccracken1989]. **Familiarity**
is knowledge of the source through exposure—the sheer accumulated awareness that
makes a face feel known. **Likability** is affection for the source arising from
physical appearance and behavior. **Similarity** is a supposed resemblance
between source and receiver. The model predicts that liked, familiar, and
similar sources persuade more, operating through identification: the receiver
accepts influence because doing so sustains a satisfying self-defining
relationship with the source.

The attractiveness model rationalizes the use of physically beautiful endorsers
and explains a robust laboratory regularity—attractive sources lift ad and brand
attitudes—but it shares the credibility model's defect. Treating attractiveness
as a *generic* positive trait predicts that the most beautiful celebrity should
sell everything equally well, which is false: an Olympic sprinter sells running
shoes better than perfume, and a supermodel the reverse. The data demand a model
in which the *content* of the celebrity's image, not merely its valence,
determines what the celebrity can sell. That is the contribution of meaning
transfer.

### The Meaning-Transfer Model

@mccracken1989 argues that both classical models are too thin: they reduce a
culturally rich figure to a few scalar traits and cannot explain
*image-specific* effectiveness. The **meaning-transfer** model replaces source
*properties* with the *symbolic meanings* a celebrity carries—status, class,
gender, age, personality, and lifestyle—accumulated across the roles, contexts,
and performances in which the public has encountered the person. The celebrity is
valuable precisely because these meanings are sharply drawn and already
established in the audience's mind.

Meaning moves in three stages, sketched in @fig-meaning-transfer. In the first
stage, cultural meaning resides in the **culturally constituted world** and
becomes attached to celebrities through their public lives. In the second,
**advertising** transfers a selected subset of that meaning from celebrity to
**product**: the ad's creative task is to make the audience see the same meanings
in the brand that they already see in the endorser. In the third, **consumption
rituals** transfer meaning from the product to the **consumer**, who acquires and
displays the good to construct and signal an identity. This third stage rests on
the same premise as the rest of @sec-branding—that consumers consume to build the
*extended self* [@belk1988]—and is enacted through the possession, exchange,
grooming, and divestment rituals through which the broader cultural system moves
meaning into goods [@mccracken1986].

```{mermaid}
%%| label: fig-meaning-transfer
%%| fig-cap: "The meaning-transfer model. Cultural meaning flows from the constituted world to the celebrity, from the celebrity to the brand via advertising, and from the brand to the consumer via consumption rituals. Each arrow is a transfer the marketer must actively engineer; meaning does not move on its own."
flowchart LR
  A["Culturally<br/>constituted world<br/><i>(status, class, gender,<br/>lifestyle)</i>"] -->|public roles<br/>& performances| B["Celebrity<br/>endorser"]
  B -->|advertising| C["Brand /<br/>product"]
  C -->|consumption<br/>rituals| D["Consumer's<br/>extended self"]
  classDef world fill:#eef,stroke:#557;
  classDef person fill:#efe,stroke:#575;
  class A world;
  class D person;
```

Four **modes of endorsement** govern how explicitly the celebrity is bound to the
product within the ad, and thus how directly meaning is invited to transfer
[@mccracken1989]. In the **explicit** mode the endorser states "I endorse this
product"; in the **implicit** mode, "I use this product"; in the **imperative**
mode, "you should use this product"; and in the **copresent** mode the celebrity
merely appears alongside the product, asserting nothing. The copresent mode is the
purest meaning-transfer device—no claim is made, yet the juxtaposition licenses
the audience to read the celebrity's meanings into the brand—and it is precisely
the mode the credibility and attractiveness models cannot explain, because no
assertion is made whose validity expertise or trust could underwrite.

The three models are best read as nested in explanatory ambition rather than as
rivals, as @tbl-three-models summarizes. Credibility and attractiveness identify
*which source traits* matter; meaning transfer explains *why a given trait matters
for a given brand* and supplies the bridge to the match-up hypothesis below.

```{r}
#| label: tbl-three-models
#| tbl-cap: "Three models of endorser influence. Each row states the locus of persuasion, the primitive constructs, the mechanism, and the empirical anomaly that the next model down was built to resolve."
#| echo: false
#| message: false
#| warning: false
library(knitr)
m <- data.frame(
  Model = c("Source credibility", "Source attractiveness", "Meaning transfer"),
  `Locus of persuasion` = c("Epistemic authority of the source",
                            "Audience affinity for the source",
                            "Cultural meanings the source carries"),
  `Primitive constructs` = c("Expertise; trustworthiness",
                            "Familiarity; likability; similarity",
                            "Status, class, gender, age, lifestyle"),
  `Unresolved anomaly` = c("Non-expert celebrities still sell",
                         "Beauty does not sell all categories equally",
                         "Requires congruence logic (match-up)"),
  check.names = FALSE
)
kable(m)
```

## The Match-Up Hypothesis

If a celebrity's value is the *specific* meaning she carries, then effectiveness
must depend on the *fit* between that meaning and the brand. The **match-up
hypothesis** states exactly this: an endorser is more effective the more
congruent the endorser's image is with the endorsed product. The hypothesis is
the strategic operationalization of meaning transfer—it tells the manager not
merely to hire a credible or attractive celebrity but to hire one whose
accumulated meanings *match* the meanings the brand wishes to own.

### Intuition and Formalization

The intuition is a congruence-as-interaction claim: endorser quality and brand
relevance are complements, not substitutes. A high-meaning endorser adds little
to a brand the meaning does not fit, and a perfectly matched but low-profile
endorser has little meaning to transfer. Let $y$ denote a response outcome
(attitude toward the brand, purchase intention, or choice). Let $E$ index a
property of the endorser (e.g., perceived attractiveness or expertise, mean-
centered) and let $M$ denote the *match*—the congruence between the endorser's
image and the product category, also mean-centered. The match-up hypothesis is
the claim that the two interact:

$$
y = \beta_0 + \beta_1 E + \beta_2 M + \beta_3\,(E \times M) + \varepsilon,
\qquad H_{\text{match-up}}:\ \beta_3 > 0.
$$ {#eq-matchup}

The substantive content is entirely in $\beta_3$. A positive interaction means
the marginal effect of endorser quality, $\partial y / \partial E = \beta_1 +
\beta_3 M$, *grows* with match: the same attractive celebrity moves the outcome
more for a beauty product than for a power tool. The main effects $\beta_1$ and
$\beta_2$ are, on the match-up logic, of secondary interest; a model that omits
the product term and reports only that "attractive endorsers work" has not tested
the hypothesis at all. This is the same interaction-versus-main-effect distinction
that governs brand-extension fit in @sec-branding, and it carries the same warning:
congruence is a *relational* property and must be modeled relationally.

### A Reproducible Test

Identification of $\beta_3$ in an experiment is clean because $E$ and $M$ are
*assigned*, not measured: a $2 \times 2$ design crosses a high- versus low-
attractiveness endorser with a high- versus low-match product, and randomization
guarantees the regressors are orthogonal to $\varepsilon$, so least squares
recovers the interaction without confounding. The estimator is ordinary least
squares (OLS); the identifying assumption is random assignment of the endorser–
product pairing (which makes $\mathbb{E}[\varepsilon \mid E, M] = 0$); and what
breaks identification in *observational* endorsement data is that firms *choose*
matched endorsers, so $M$ correlates with unobserved brand quality and $\beta_3$
absorbs selection. The simulation below generates data under @eq-matchup with a
known positive interaction and recovers it.

```{r}
#| label: matchup-sim
#| echo: true
#| message: false
#| warning: false
set.seed(1989)  # year of McCracken's meaning-transfer paper

n_per_cell <- 60
design <- expand.grid(
  endorser = c(-0.5, 0.5),   # low / high attractiveness (effect-coded)
  match    = c(-0.5, 0.5),   # low / high product match
  rep      = seq_len(n_per_cell)
)

# True data-generating process: positive match-up interaction (beta_3 = 1.4)
b0 <- 3.0; b1 <- 0.4; b2 <- 0.6; b3 <- 1.4
design$y <- with(design,
  b0 + b1 * endorser + b2 * match + b3 * (endorser * match) +
    rnorm(nrow(design), sd = 1))

fit <- lm(y ~ endorser * match, data = design)
round(summary(fit)$coefficients, 3)

# Cell means make the interaction visible: the endorser "lift" is large only
# when the product matches.
cell <- aggregate(y ~ endorser + match, data = design, FUN = mean)
lift_match    <- with(cell, y[endorser == 0.5 & match == 0.5] -
                              y[endorser == -0.5 & match == 0.5])
lift_mismatch <- with(cell, y[endorser == 0.5 & match == -0.5] -
                              y[endorser == -0.5 & match == -0.5])
cat("Endorser lift when matched:    ", round(lift_match, 2), "\n")
cat("Endorser lift when mismatched: ", round(lift_mismatch, 2), "\n")
```

The estimated `endorser:match` coefficient recovers the planted $\beta_3$, and the
two "lift" quantities make the substantive claim concrete: switching to the
attractive endorser helps far more when the product matches than when it does not.
A non-significant interaction—not a significant main effect—is what would falsify
the hypothesis.

### Scope, Mechanism, and Boundaries

The match-up effect is not unconditional, and the modern literature has mapped its
boundaries. The processing route matters: under the elaboration-likelihood logic
of @petty1983 and @petty1986, source cues such as attractiveness operate as
*peripheral* cues when involvement is low and the audience is not scrutinizing
arguments, but as substantive evidence (or as a distraction) when involvement is
high. Congruence therefore interacts with involvement, and an endorser who helps a
low-involvement impulse purchase may be irrelevant or counterproductive for a
high-involvement considered one. The mechanism also runs partly through *transfer
of affect*: positive feelings conditioned to the celebrity can become attached to
the brand through co-occurrence, a classical-conditioning account that complements
meaning transfer and that, like it, predicts congruence-dependent effects rather
than uniform lift. Finally, the **multiple-product / multiple-endorser** problem
complicates the simple match-up story: a celebrity who endorses many unrelated
brands dilutes the distinctiveness of the meaning available for transfer to any
one of them, weakening congruence even when each pairing is individually
well matched.  @tripp1994multipleproduct show the dilution directly: as the number of products an endorser fronts rises, both endorser credibility and attitude toward the ad decline.

Two cross-cultural moderators deserve emphasis because they bound external
validity. Power-distance belief—the extent to which a consumer accepts hierarchy—
moderates the celebrity effect: power-distance beliefs *strengthen* the positive
effect of celebrity (versus non-celebrity) endorsers on advertising evaluations,
because celebrities are high-status sources whose influence is more readily
accepted by those who endorse hierarchical social arrangements [@Winterich_2018].
This is a meaning-transfer result in disguise—status is one of the cultural
meanings a celebrity carries, and its persuasive force depends on whether the
audience values status. Self-construal further conditions which connection
matters, paralleling the brand-attachment findings of @swaminathan2007.

## Influencers as Endorsers

The economics of endorsement have been reshaped by social-media **influencers**—
individuals who accumulate audiences on platforms and monetize that attention
through paid brand content. Influencers occupy an intermediate position between
the classical celebrity and the ordinary peer: they are more *similar* to their
followers than a film star (raising the attractiveness model's similarity lever)
yet command genuine reach, and their persuasiveness depends sharply on perceived
authenticity. @sokolova2020 show that parasocial interaction and the perceived
credibility of the influencer drive purchase intention, with engagement operating
through identification rather than expertise. Disclosure of the commercial
relationship is the new trust frontier: sponsorship transparency interacts with
follower trust in ways the original trustworthiness construct anticipated but
could not have specified. Influencer marketing is now an industry of its own, with
intermediaries matching brands to creators [@nistor2024influencers], and even
*mega*-influencer and celebrity-creator effects have been documented in
firm-outcome data [@tian2024mega; @leung2022]. The constructs are old; the
measurement environment—observable, high-frequency, and platform-mediated—is new.

## Endorser Risk

The asset a firm rents in an endorsement is a *person*, and persons misbehave. The
defining hazard of celebrity endorsement is therefore **endorser risk**: the brand
inherits not only the celebrity's favorable meanings but also her liabilities, and
a scandal can transfer *negative* meaning into the brand through exactly the
channel of @fig-meaning-transfer that was supposed to transfer positive meaning. A
serious treatment of endorsement must price this downside, not merely the upside.

Several features make endorser risk distinctive. It is *idiosyncratic to the
person* and largely outside the firm's control, so it cannot be diversified within
a single endorsement. It is *transferable*: meaning transfer is symmetric, and the
same congruence that amplifies positive transfer amplifies the contamination when
the endorser's image sours. And it is *correlated across a celebrity's
portfolio*—one scandal simultaneously damages every brand the celebrity endorses,
which is why endorsement contracts contain morality clauses permitting termination
on disrepute. The negative information need not even be deserved: under the
attribution logic that governs brand crises in @sec-branding, audiences who view
character as stable (entity theorists) generalize a transgression to a global
judgment of the person, and that judgment flows to the brand.

The timing of harm is subtle. A *sleeper effect* can arise when a discounting
cue—"this source is untrustworthy"—decays faster than the message itself, so
persuasion *rises* over time as the audience forgets why it discounted the
source; symmetrically, the damage from a tainted endorser can dissipate as the
association weakens, so the net effect of a scandal is dynamic rather than a
one-time level shift [@wu2015sleeping]. Managers who read a scandal's immediate
stock-price reaction as the whole story may therefore mis-estimate the long-run
cost, in either direction.

## Endorsement and Firm Value

Endorsement contracts are firm-level investments, and capital markets price them.
The natural identification strategy is the **event study** developed in
@sec-marketing-finance: an endorsement announcement (or an endorser scandal) is a
discrete, dated event whose firm-value consequence is read from the abnormal
stock return in a tight window around the announcement, after the normal return is
removed.

### The Event-Study Estimator

For firm $i$ on trading day $t$, the **abnormal return** is the realized return
minus its expectation under a market model estimated on a clean pre-event window:

$$
\mathrm{AR}_{it} = R_{it} - \big(\hat\alpha_i + \hat\beta_i R_{mt}\big),
\qquad
\mathrm{CAR}_i(W) = \sum_{t \in W} \mathrm{AR}_{it},
$$ {#eq-car}

where $R_{mt}$ is the market return, $(\hat\alpha_i, \hat\beta_i)$ are OLS
estimates from the estimation window, and $\mathrm{CAR}_i(W)$ is the cumulative
abnormal return over a short event window $W$ (often $[-1, +1]$ days). The method
is the standard apparatus of @mcwilliams1997a and @srinivasan2004event and is
applied to brand-asset transactions in @sec-branding.

Three assumptions carry the inference, and each is a place identification can
break. **Market efficiency**: prices impound the news quickly, so a short window
captures the full effect; if the news leaks, the window must widen and power
falls. **No confounding**: the window must be scrubbed of contemporaneous
firm-specific news—earnings, splits, executive changes, buybacks, dividend
changes—because the event study cannot separate two events sharing a window
[@srinivasan2004event]. **Correct normal-return model**: a misspecified benchmark
($\hat\alpha, \hat\beta$) contaminates every $\mathrm{AR}$; thin trading and
event-induced variance further bias the standard errors. The estimator answers a
sharp question—*by how much did this announcement change shareholder wealth?*—only
to the extent these assumptions hold.

### A Reproducible Event Study

The following example simulates daily returns for a firm, injects a positive
endorsement-announcement effect on the event day, estimates the market model on
the pre-event window, and aggregates abnormal returns over a three-day event
window. The structure mirrors exactly what one would run on CRSP data.

```{r}
#| label: event-study-sim
#| echo: true
#| message: false
#| warning: false
set.seed(2024)

est_days   <- 120          # estimation window length
event_win  <- -1:1         # 3-day event window around the announcement
n_days     <- est_days + length(event_win)

# True market model: R_it = alpha + beta * R_mt + e_it, with a +1.8% jump on day 0
alpha <- 0.0002; beta <- 1.1; sigma <- 0.012
r_m   <- rnorm(n_days, mean = 0.0004, sd = 0.009)
shock <- rep(0, n_days); shock[est_days + which(event_win == 0)] <- 0.018
r_i   <- alpha + beta * r_m + shock + rnorm(n_days, sd = sigma)

# Estimate the market model on the clean pre-event window only
est_idx <- seq_len(est_days)
mm  <- lm(r_i[est_idx] ~ r_m[est_idx])
ab  <- coef(mm)[1]; bt <- coef(mm)[2]

# Abnormal returns over the event window
ev_idx <- (est_days + 1):n_days
AR  <- r_i[ev_idx] - (ab + bt * r_m[ev_idx])
CAR <- sum(AR)

# Test statistic: CAR scaled by the estimation-window residual SD
sd_ar  <- sd(residuals(mm))
t_stat <- CAR / (sd_ar * sqrt(length(ev_idx)))

cat("Estimated alpha, beta:", round(ab, 5), round(bt, 3), "\n")
cat("Daily abnormal returns:", paste(round(AR, 4), collapse = ", "), "\n")
cat("CAR over [-1,+1]:      ", round(CAR, 4),
    sprintf("(%.2f%%)", 100 * CAR), "\n")
cat("t-statistic:           ", round(t_stat, 2), "\n")
```

The recovered $\mathrm{CAR}$ concentrates on the event day and is statistically
distinguishable from zero, reproducing the planted announcement effect. An
endorser-scandal study is the same machinery with the sign reversed and the event
redefined as the disclosure date.

### What the Evidence Shows

Applied work using this design finds that endorsement is, on average, value
creating but heavily conditional. Announcements of celebrity-endorsement contracts
are associated with positive abnormal returns and with gains in sales, and the
effect is larger when the endorser is well matched and athletically or
reputationally distinguished [@agrawal2000; @elberse2003]. Athlete endorsements in
particular raise the sponsoring firm's sales and stock returns, with the lift
tracking the athlete's on-field success and visibility [@elberse2003]. The downside
is equally real and is priced: negative information about an endorser—arrest,
scandal, doping—transfers to the brand and to firm value, and the contamination is
stronger where the celebrity–brand link is tighter, the mirror image of the
match-up amplification [@chuluun2017firm]. Endorsement, like co-branding in
@sec-branding, is thus a *reputation-coupling* decision: the firm gains access to
the celebrity's reputational capital but assumes a share of its variance.

@tbl-endorser-risk frames the trade-off as a simple expected-value problem and
foreshadows the quantitative reasoning a manager should bring to a contract.

```{r}
#| label: tbl-endorser-risk
#| tbl-cap: "Endorsement as a risky investment. The expected firm-value effect nets the probability-weighted upside of a well-received campaign against the probability-weighted downside of an endorser scandal, minus the contract's cost. A high-match endorser raises both the upside and the contamination severity."
#| echo: false
#| message: false
#| warning: false
library(knitr)
risk <- data.frame(
  Component = c("Upside (successful campaign)",
                "Downside (endorser scandal)",
                "Contract cost",
                "Net expected value"),
  `Driver` = c("Match x reach x meaning fit",
             "Scandal probability x contamination severity",
             "Negotiated fee + activation spend",
             "E[upside] - E[downside] - cost"),
  `Match-up role` = c("Raises upside (amplifies positive transfer)",
                    "Raises severity (amplifies negative transfer)",
                    "Higher for higher-meaning endorsers",
                    "Match-up enters with opposite signs"),
  check.names = FALSE
)
kable(risk)
```

### A Back-of-the-Envelope Valuation

The table's logic is worth making numerical, because it disciplines the romance of
a marquee signing. Treat the endorsement as a one-period lottery: with probability
$1-q$ the campaign succeeds and adds value $V^{+}$; with probability $q$ a scandal
strikes and subtracts $V^{-}$; the contract costs $C$. The endorsement is
ex-ante value creating only if

$$
(1-q)\,V^{+} \;-\; q\,V^{-} \;-\; C \;>\; 0.
$$ {#eq-endorser-ev}

Match-up enters @eq-endorser-ev with *opposite signs* on $V^{+}$ and $V^{-}$: a
tighter celebrity–brand coupling raises the success payoff $V^{+}$ but also raises
the contamination loss $V^{-}$, so the optimal match is interior, not maximal,
once risk is priced. The snippet below evaluates the criterion and the breakeven
scandal probability.

```{r}
#| label: endorser-ev
#| echo: true
#| message: false
#| warning: false
V_up   <- 40e6     # firm-value gain if the campaign succeeds
V_down <- 120e6    # firm-value loss if the endorser is disgraced
C      <- 15e6     # contract + activation cost
q      <- 0.05     # annual scandal probability

ev <- (1 - q) * V_up - q * V_down - C
q_breakeven <- (V_up - C) / (V_up + V_down)

cat("Expected value of the endorsement: $",
    format(round(ev), big.mark = ","), "\n", sep = "")
cat("Breakeven scandal probability:     ",
    sprintf("%.1f%%", 100 * q_breakeven), "\n")
```

The breakeven probability—the scandal rate at which the endorsement is exactly
worth its cost—falls as the contamination loss $V^{-}$ rises, which is why brands
with the most to lose (those whose match is tightest and whose equity is largest)
are the most sensitive to endorser conduct and the most aggressive about morality
clauses. The model is deliberately spare, but it captures the governing tension:
the very congruence the match-up hypothesis prescribes is also what makes a
scandal expensive.

## Pitfalls and Identification

The recurring threat in endorsement research is **selection**. Firms do not pair
celebrities with brands at random; they hire matched, attractive, credible
endorsers precisely for valuable brands and well-funded campaigns. In
observational data the match variable $M$ in @eq-matchup is therefore correlated
with unobserved brand quality and campaign budget, so a naive regression overstates
the causal effect of congruence and the event-study $\mathrm{CAR}$ in @eq-car
blends the endorsement's effect with the information the announcement reveals about
the firm's prospects. Experiments solve this by assigning the pairing; event
studies mitigate it by differencing against the market and scrubbing confounds, but
neither fully removes the firm's *choice* of when and whom to announce.

A second pitfall is **construct conflation**. Credibility, attractiveness, and
meaning are distinct, and a study that measures only generic "endorser
favorability" cannot adjudicate among the three models or test the match-up
interaction, which requires a *product-relative* match measure. A third is
**dynamics**: the sleeper effect [@wu2015sleeping] and the gradual decay of
meaning mean that a single post-event snapshot can mislead; effects should be
traced over a horizon. A fourth, specific to influencers, is **disclosure
endogeneity**: sponsored posts are selectively disclosed and selectively
produced, so the apparent effect of an influencer confounds the creator's choice
of which products to promote with the persuasion itself [@nistor2024influencers].
Each pitfall is an instance of the general principle that runs through this book—
state the assignment mechanism before interpreting the estimate.

## Key Takeaways

- A celebrity endorser rents an accumulated reputation; the three classical
  models locate its persuasive force in *credibility* (expertise, trust),
  *attractiveness* (familiarity, likability, similarity), and—most completely—
  *meaning transfer* of specific cultural meanings from celebrity to brand to
  consumer [@mccracken1986; @mccracken1989].
- The **match-up hypothesis** is the testable, interaction-based ($\beta_3 > 0$
  in @eq-matchup) operationalization of meaning transfer; a significant main
  effect without a significant interaction does not test it.
- Endorser influence is moderated by processing route [@petty1986], cultural
  values such as power distance [@Winterich_2018], and—for influencers—perceived
  authenticity and disclosure [@sokolova2020].
- **Endorser risk** is the symmetric, transferable, portfolio-correlated downside
  of meaning transfer; tighter match amplifies *both* the upside and the
  contamination, so the value-maximizing match is interior once risk is priced
  (@eq-endorser-ev).
- Endorsement contracts and scandals are firm-level events whose value effect is
  estimated by **event study** (@eq-car); the evidence shows positive average
  returns to well-matched, high-reputation endorsements and significant losses to
  endorser scandals [@agrawal2000; @elberse2003; @chuluun2017firm], subject
  always to the selection caveat that firms choose whom, and when, to announce.
