# Influencer Marketing {#sec-influencer-marketing}

*Influencer marketing* is the practice of leveraging an individual's accumulated
audience, credibility, and content to carry a firm's marketing message. Formally,
@leung2022 describe online influencer marketing as the deployment of an
influencer's resources—follower networks, personal positioning, communication
content, and the trust followers place in them—to enhance the effectiveness of a
firm's marketing communication. The defining feature is that persuasion is rented
from a third party whose authority was earned outside any single brand relationship,
and whose continued authority depends on not appearing captured by the brands that
pay them. That tension—between the sponsor's desire for reach and control and the
influencer's need to preserve credibility—organizes nearly everything in this
chapter.

Influencer marketing matters commercially because attention has migrated to social
platforms and because a small set of accounts concentrates a disproportionate share
of persuasive power. Industry audits routinely find that a handful of recommenders
account for a large fraction of category recommendations—on the order of five
percent of recommenders generating close to half of socially influenced
purchases in apparel and footwear—so the selection of *whom* to hire dominates the
economics of a campaign. It matters scientifically because influencers are a clean
laboratory for classical questions in persuasion: source credibility, the
endorsement of meaning, two-sided signaling, and the measurement of engagement and
its translation (or failure to translate) into sales.

This chapter develops the topic from construct to model to measurement. We first
locate the influencer within a communication-theoretic frame—sender, message,
receiver—and define the constructs (the influencer as a *person brand*, credibility,
engagement) formally. We then build the econometric workhorse of the empirical
literature: a selection-corrected count model of engagement, with its estimator,
assumptions, and identification requirements stated in full. We turn next to the
economics of influencer *type*—why micro-influencers and celebrities play different
equilibrium roles—and to the frontier problem that engagement is not sales,
including deep-learning measures of product-focused attention. We close with the
practical problem managers actually face: valuing an account or a single post, and a
reproducible toolkit for doing so.

## The Influencer as a Communication System

The cleanest way to organize the determinants of influencer effectiveness is the
mathematical theory of communication [@weaver1963mathematical; @Shannon_1948], which
decomposes any act of communication into a **sender**, a **message**, and a
**receiver**. @leung2022 map the three components onto the influencer setting and
attach a measurable construct to each: the sender (the influencer) contributes
**source trustworthiness** [@wilson1993]; the message (the post) contributes
**content value** [@ducoffe1996advertising]; and the receiver (the follower)
contributes **involvement** with the message. Effectiveness is multiplicative rather
than additive in these components—a trustworthy source delivering low-value content
to an uninvolved audience persuades no one—which is why single-metric rankings of
influencers (by follower count alone, say) predict outcomes so poorly.

```{mermaid}
%%| label: fig-comm-system
%%| fig-cap: "The influencer as a communication system. Effectiveness is jointly determined by source trustworthiness, message content value, and receiver involvement; the firm intervenes through selection (whom to hire) and the brief (what to post), and observes only downstream engagement."
flowchart LR
  F["Firm<br/>(selection + brief)"] --> S
  subgraph channel["Communication channel"]
    S["Sender<br/>influencer<br/><i>trustworthiness</i>"] --> M["Message<br/>post<br/><i>content value</i>"]
    M --> R["Receiver<br/>follower<br/><i>involvement</i>"]
  end
  R --> E["Engagement<br/>(likes, comments,<br/>shares)"]
  E -.->|noisy proxy| Sales["Sales lift"]
  R -.->|reputation feedback| S
```

### The Influencer as a Person Brand

An influencer is best understood as a **person brand**: a human being who functions
as an economic asset because their name evokes a stable, ownable set of associations
[@thomson2006; @fournier2019; @kerrigan2011]. This framing inherits the full
apparatus of @sec-branding—brand associations, personality, and relationship
quality—but with one decisive difference from a traditional celebrity. Classical
celebrity endorsement works through **meaning transfer**: the celebrity, credentialed
by an institution (a studio, a league, an awards body), carries culturally
constituted meanings that flow through the advertisement into the brand and onward to
the consumer [@mccracken1989]. Social-media influencers are *not* institutionally
credentialed in this way [@mcquarrie2013]; their authority is conferred bottom-up by
an audience and is therefore both cheaper to acquire and easier to lose. The
influencer's persuasive capital is endogenous to their own posting behavior in a way
the movie star's is not, which is the root of the reputation dynamics developed in
@sec-reputation-burning.

> **Influencer (person brand).** An individual whose accumulated audience, perceived
> expertise, and earned trust constitute a transferable marketing asset, and whose
> endorsement derives credibility from an audience relationship rather than from
> institutional credentialing.

### Engagement, Formally

The dependent variable throughout the empirical literature is **engagement**.
Following @hollebeek2011, engagement is "a customer's cognitive, emotional, and
behavioral activities" with respect to brand-related interactions—a deliberately
multidimensional construct. In practice it is operationalized through observable
behavioral traces: comments, likes, shares, saves, and watch time. These traces are
*counts*, bounded below at zero and right-skewed, a fact that dictates the estimator
(@sec-engagement-model).

A useful summary of effectiveness is the **engagement elasticity**, the percentage
change in consumer engagement produced by a one-percent change in influencer
marketing spend. @leung2022 estimate this elasticity at roughly $0.46$: a one-percent
increase in spend raises engagement by about $0.457$ percent. Letting $g$ denote
engagement and $s$ denote spend, the elasticity is

$$
\varepsilon_{g,s} \;=\; \frac{\partial \log g}{\partial \log s}
\;=\; \frac{\partial g}{\partial s}\,\frac{s}{g},
$$ {#eq-engagement-elasticity}

a unit-free quantity comparable across influencers, platforms, and campaigns in a way
that raw engagement counts are not. Because @eq-engagement-elasticity is below one,
engagement returns to spend are *inelastic*—doubling the budget less than doubles
engagement—an empirical regularity consistent with diminishing returns to reach.

## Drivers of Influencer Effectiveness {#sec-effectiveness-drivers}

What makes an influencer campaign work? The literature converges on a set of drivers
attached to the three communication components, several of which act non-monotonically
or interact with campaign intent and platform. We organize them by component and then
treat the most important interactions explicitly.

**Sender (influencer) characteristics.** Three sender attributes raise effectiveness:
**originality** (distinctive content that does not read as a template),
**follower size** (reach), and the moderating role of **sponsor salience**
[@leung2022]. Source **trustworthiness** is the deeper construct underneath these
[@wilson1993]: influencers persuade to the extent that their endorsement is read as a
sincere recommendation rather than a paid placement. A counterintuitive corollary is
that signals of independence can *raise* perceived influence—an influencer who follows
relatively few accounts while commanding many followers is read as autonomous and
therefore more credible [@sokolova2020]. Several drivers, however, are
**inverted-U** shaped: influencer **activity** (posting frequency),
**follower–brand fit**, and content **positivity** each help up to a point and then
hurt, as over-posting reads as spam, perfect fit reads as paid, and relentless
positivity reads as inauthentic [@leung2022].

**Message (content) characteristics.** Content **value** is the message-level analogue
of source credibility [@ducoffe1996advertising]. A practically important
content distinction is between **product-launch** posts and ordinary endorsement
posts: product-launch content *decreases* effectiveness [@leung2022], plausibly
because launch posts foreground the commercial transaction and suppress the
audience-relationship cues that make influencer content persuasive. Identifying which
posts are launch- or promotion-related at scale is itself a measurement task; the
literature uses dictionary-based and supervised text classification on post text to
code launch versus non-launch and product versus service [@lovett2013].

**Receiver (audience) characteristics.** Effectiveness ultimately depends on follower
**involvement**—how cognitively and affectively engaged the audience is with the
message. Involvement is not directly controllable by the firm, but it is shaped by the
match between content and audience and by platform affordances, which is why the same
post performs differently across platforms.

### Sponsored Blogging: Expertise, Intent, and Platform {#sec-sponsored-blogging}

A foundational empirical study of these drivers is @hughes2019, who study sponsored
blogging—content that is simultaneously *paid* (the firm compensates the blogger) and
*earned* (it appears as the blogger's own editorial voice), placing it at the
intersection of paid and earned media [@colicev2018b; @colicev2019]. Grounding their
hypotheses in the **elaboration likelihood model**, which distinguishes central
(argument-driven) from peripheral (cue-driven) routes to persuasion, the authors show
that the *same* blogger characteristic can help or hurt depending on the campaign's
position in the purchase funnel and the platform on which the content appears.

Two moderators are central. **Campaign intent** is whether the goal is to raise
*awareness* (top of funnel) or to encourage *trial* (lower funnel). **Platform
involvement** contrasts a high-involvement environment (a blog, where readers
elaborate) with a low-involvement environment (a Facebook feed, where they skim).
@tbl-blogging-effects summarizes the key contingencies.

| Driver | Awareness campaign | Trial campaign | Platform contingency |
|---|---|---|---|
| Blogger **expertise** | More effective | Less effective | On Facebook (low involvement) source expertise has *no* effect on engagement |
| **Hedonic** content | Less effective | More effective | On Facebook, hedonic content lifts engagement, especially for trial |
| Campaign **freebies** | Mixed | Mixed | Giveaways can backfire, cannibalizing blog engagement in favor of the Facebook channel |

: How blogger and content characteristics interact with campaign intent and platform. Source: @hughes2019. {#tbl-blogging-effects}

The headline interaction is between expertise and intent: when the goal is to build
*awareness*, high blogger expertise is more effective, because an expert source lends
credibility to an unfamiliar product; when the goal is to drive *trial*, expertise
matters less and experiential, hedonic content matters more, because the audience
already knows the product and needs a reason to act. Critically, this expertise effect
is a high-involvement phenomenon: on Facebook, where readers do not elaborate, source
expertise washes out entirely. The managerial implication is that influencer selection
and the creative brief must be chosen *jointly* with the funnel objective and the
platform, not optimized in isolation.

## Modeling Engagement with Selection {#sec-engagement-model}

We now formalize the workhorse model behind studies such as @hughes2019 and
@leung2022. Two econometric facts dominate the specification. First, engagement is a
**count** with **overdispersion**—the conditional variance exceeds the conditional
mean—so ordinary least squares and even the Poisson model are inappropriate. Second,
the influencers and posts in any dataset are **not randomly assigned**: brands choose
which influencers to hire and influencers choose which campaigns to accept, so the
observed sample is selected on unobservables that may also drive engagement. Ignoring
this yields a *selection bias* that contaminates every coefficient.

### The Negative Binomial Outcome Model

Let $y_i$ be the engagement count for post $i$ (e.g., number of comments or likes),
and let $\mathbf{x}_i$ collect influencer, content, campaign, and platform covariates.
The **negative binomial** model specifies a log-linear conditional mean,

$$
\mathbb{E}[y_i \mid \mathbf{x}_i] = \mu_i = \exp(\mathbf{x}_i^\top \boldsymbol{\beta}),
$$ {#eq-nb-mean}

with the variance inflated relative to the mean by an overdispersion parameter
$\alpha > 0$,

$$
\operatorname{Var}(y_i \mid \mathbf{x}_i) = \mu_i + \alpha\,\mu_i^2 .
$$ {#eq-nb-variance}

The model arises as a Poisson–gamma mixture: conditional on an unobserved
multiplicative heterogeneity term $\nu_i \sim \text{Gamma}(1/\alpha, 1/\alpha)$,
$y_i \mid \nu_i \sim \text{Poisson}(\mu_i \nu_i)$, and integrating out $\nu_i$ yields
the negative binomial likelihood [@cameron2005microeconometrics]. The parameter
$\alpha$ measures the residual variance left by influencer- and post-level factors the
covariates do not capture; as $\alpha \to 0$ the model collapses to Poisson. The
coefficients $\boldsymbol{\beta}$ are estimated by maximum likelihood and interpreted
as semi-elasticities: a unit change in covariate $k$ multiplies expected engagement by
$\exp(\beta_k)$, holding the rest fixed.

### Correcting for Endogenous Selection

The threat to identification is that the set of (influencer, campaign) pairs we
observe is the *outcome of a choice*. Brands pursue influencers they expect to perform
well; influencers accept campaigns that fit their persona. If the unobservables
driving selection are correlated with the unobservables driving engagement, the
estimated $\boldsymbol{\beta}$ in @eq-nb-mean confounds the causal effect of a covariate
with its correlation with the selection process.

The standard remedy is a **Heckman-type selection model** [@heckman2007econometric;
@greene2003econometric]. Two equations are specified. A **selection equation** models
whether an influencer is chosen for a campaign,

$$
d_i = \mathbf{1}\!\left\{ \mathbf{z}_i^\top \boldsymbol{\gamma} + u_i > 0 \right\},
$$ {#eq-selection}

and an **outcome equation** models engagement conditional on selection
(@eq-nb-mean). Identification hinges on an **exclusion restriction**: the selection
covariates $\mathbf{z}_i$ must contain at least one variable that affects *selection*
but does not directly affect *engagement* once selection is accounted for. Concretely,
@hughes2019 construct a blogger **psychographic index**—travel/foodie, persona,
lifestyle, and values factors extracted by varimax-rotated factor analysis of profile
descriptions—as instruments: these traits plausibly drive which bloggers a brand
recruits without directly moving the engagement a post generates. @leung2022 use an
analogous strategy, instrumenting selection with the characteristics of the
*second-most-similar* influencer the focal brand could have chosen, following the
matching logic of @hughes2019.

The exclusion restriction is precisely where identification can break. If the
purportedly excluded variable *does* influence engagement directly—if, say, a
blogger's "lifestyle" loading also makes their posts more engaging—the instrument is
invalid, the correction is contaminated, and the corrected estimates can be *more*
biased than the naive ones. There is no purely statistical test of the restriction; it
must be defended on substantive grounds. A credible influencer study therefore states
its exclusion restriction explicitly and argues why the excluded variables touch
selection but not the outcome.

### A Reproducible Selection-Corrected Count Model

The following simulation builds a dataset with genuine selection on unobservables,
shows the bias from naively fitting the outcome model on the selected sample, and
recovers the truth with a control-function correction. It depends only on base R.

```{r nb-selection, message=FALSE, warning=FALSE}
set.seed(15)

n <- 4000

# --- Latent traits -------------------------------------------------------
# Exclusion-restriction variable: psychographic "lifestyle" index z.
# It drives SELECTION but NOT engagement directly.
z          <- rnorm(n)
expertise  <- rnorm(n)               # influencer expertise (in both equations)
hedonic    <- rnorm(n)               # content hedonic value (outcome only)

# Correlated unobservables linking selection and engagement.
rho   <- 0.6
e_sel <- rnorm(n)
e_out <- rho * e_sel + sqrt(1 - rho^2) * rnorm(n)

# --- Selection equation (who gets hired) ---------------------------------
sel_index <- -0.2 + 0.9 * z + 0.5 * expertise + e_sel
selected  <- as.integer(sel_index > 0)

# --- Outcome: engagement counts (negative binomial) ----------------------
beta0 <- 1.0; b_exp <- 0.40; b_hed <- 0.30
log_mu <- beta0 + b_exp * expertise + b_hed * hedonic + 0.8 * e_out
mu     <- exp(log_mu)
theta  <- 2                                  # NB size (overdispersion)
y      <- rnbinom(n, size = theta, mu = mu)

dat <- data.frame(y, z, expertise, hedonic, selected, e_out)
obs <- dat[dat$selected == 1, ]              # we only observe hired influencers

# --- Naive NB on the selected sample (biased) ----------------------------
naive <- MASS::glm.nb(y ~ expertise + hedonic, data = obs)

# --- Control-function correction -----------------------------------------
# Stage 1: probit selection on z + expertise; form the inverse Mills ratio.
sel_fit <- glm(selected ~ z + expertise, family = binomial("probit"), data = dat)
lp      <- predict(sel_fit, newdata = obs)
imr     <- dnorm(lp) / pnorm(lp)             # inverse Mills ratio for selected

# Stage 2: NB outcome including the IMR as a control function.
corr <- MASS::glm.nb(y ~ expertise + hedonic + imr, data = obs)

round(rbind(
  truth     = c(expertise = b_exp, hedonic = b_hed),
  naive     = coef(naive)[c("expertise", "hedonic")],
  corrected = coef(corr)[c("expertise", "hedonic")]
), 3)
```

The naive estimates inflate the expertise coefficient because hired influencers are
positively selected on the very unobservables that also drive engagement; adding the
inverse Mills ratio as a control function absorbs that correlation and returns the
expertise coefficient close to its true value of $0.40$. The hedonic coefficient,
which enters only the outcome and is uncorrelated with selection, is recovered well by
both methods—an illustration that selection bias contaminates *selected-on* covariates
in particular.

## Influencer Type and the Economics of Trust

Not all influencers play the same economic role, and the difference is an equilibrium
phenomenon rather than a matter of scale alone. @nistor2024influencers model a market
with **celebrity** influencers (very large reach) and **micro-influencers** (smaller,
more engaged audiences) who differ in how endorsement affects their future standing.
The model yields a sharp separation. Celebrity influencers, whose value to followers
is reach and awareness rather than curation, endorse *all* products to maximize the
number of brand deals; their endorsements drive visibility but convey little reliable
information about quality. Micro-influencers instead endorse *only high-quality
products*, because their value proposition to followers is curation—acting as a
**quality screener** so followers need not sift through comments to assess a product.
In equilibrium, micro-influencers raise the *information quality* available to their
audience while celebrities raise *awareness*; the two are complements, not substitutes,
and a brand's optimal influencer mix depends on whether it needs reach or credibility.

This screening role rationalizes a pricing puzzle: per follower, micro-influencers
often command higher rates than their reach alone would justify, because part of what
the brand buys is the influencer's *refusal* to endorse low-quality products, which is
exactly what makes the endorsement informative.

### Selecting Influencers Under a Reach–Cost Trade-off

The practical selection problem is to trade reach against cost. @tian2024mega formalize
this with the **follower elasticity of impressions** (FEI)—the percentage change in
impressions delivered per percentage change in follower count—and use it to quantify
the marginal value of a larger audience against its marginal price. Because impressions
scale sub-proportionally with followers (mega-influencers reach a smaller *share* of
their audience), FEI typically falls with size, and the cost-effective hire is often a
mid-tier influencer rather than the largest one available. The general lesson echoes
@eq-engagement-elasticity: the right object is an *elasticity*, normalizing the outcome
by the input, not the raw level of reach.

## Reputation Burning: The Cost of Sponsorship to the Influencer {#sec-reputation-burning}

Because an influencer's authority is conferred by their audience rather than by an
institution (@sec-effectiveness-drivers), sponsorship is not free to the influencer:
visibly accepting payment can erode the trust that made them valuable. @cheng2024reputation
provide direct empirical evidence of this **reputation-burning effect**. Posting a
sponsored video reduces an influencer's reputation—measured by subscriber count—by
about $0.19\%$ relative to comparable organic videos. The effect is larger for
influencers with bigger audiences, and the gap in audience response (likes and
comments) between sponsored and organic content widens with following size, consistent
with larger audiences holding their influencers to a higher standard of editorial
independence.

The loss is not a fixed tax; it is *contingent*, and the contingencies are managerially
actionable. Reputation damage is mitigated when the sponsored content **aligns** with
the influencer's typical style—so the endorsement does not read as a break in
character—and when the **promoted brand is less well known**, because endorsing an
obscure brand looks more like discovery than like cashing in. This is the
micro-influencer's screening logic (@nistor2024influencers) viewed from the supply
side: the influencer's optimal acceptance policy weighs the per-deal fee against the
content-fit-dependent reputation cost.

Identifying this effect requires comparing sponsored and organic videos that are
otherwise alike, which they are not by default—influencers choose which videos to
monetize. @cheng2024reputation construct matched treatment (sponsored) and control
(organic) groups at the influencer–video level using **DiNardo–Fortin–Lemieux
reweighting**, which reweights the control distribution of observed covariates to match
the treated distribution, so that any residual difference is attributable to
sponsorship rather than to composition. The design is a useful template for any setting
where treatment is a non-random managerial choice.

## Engagement Is Not Sales: Product-Focused Attention {#sec-engagement-not-sales}

The single most consequential critique of influencer marketing is that **engagement
need not translate into sales**. A video can be wildly engaging while directing none of
that attention to the product—the viewer remembers the creator, not the brand. Two
strands of recent work confront this gap by opening up the unstructured content of the
video itself.

@rajaram2020video build an *interpretable deep-learning* framework that decomposes a
YouTube influencer video into its textual ("what is said"), audio, and visual streams
and relates each to viewer engagement. Their central finding is that **textual content
matters more than audio or images** for engagement, and they separate drivers of
*shallow* from *deep* engagement using a dual-process view of viewer cognition. The
contribution is methodological as much as substantive: the framework predicts
engagement *and* attributes it to interpretable content features, which is what makes
it useful to a manager writing a brief.

@yang2021first attack the engagement-versus-sales problem head-on by constructing a
**Product Engagement score** (PE-score) that measures how much of a video's engaging
moments are actually *about the product*. The construction proceeds in three stages.
First, a deep three-dimensional convolutional neural network is trained to produce an
**engagement heatmap** assigning each pixel of each frame a weight reflecting its
contribution to viewer engagement; critically, the network is trained on engagement
*residuals* after partialling out influencer fixed effects, product fixed effects,
acoustic features, and transcript embeddings, so it learns visual drivers of
engagement rather than confounds. Second, an object-detection pass (SIFT-based) yields
a **product heatmap** locating the product in each frame. Third, the PE-score combines
the two: it is high when the engaging regions of the video coincide with the product's
location.

Formally, let $E_{ft}(p)$ be the engagement weight and $P_{ft}(p)$ the product
indicator for pixel $p$ in frame $f$ of video $t$. The frame-level product engagement
is the overlap

$$
\text{PE}_{ft} \;=\; \sum_{p} E_{ft}(p)\, P_{ft}(p),
$$ {#eq-pe-score}

aggregated over frames to a video-level PE-score. The measure requires only the ad and
a product image, making it usable for pre-release testing.

Validated against TikTok influencer video ads matched to Taobao product sales
(May–November 2019, 2,685 ads with linked sales), the findings are striking. Traditional
engagement metrics (likes, comments, shares) show **no significant correlation with
sales**, confirming the central critique; the PE-score, by contrast, **significantly
predicts sales lift**, identified through a difference-in-differences design around
video-posting time that uses a product search index to address reverse causality.
Neither product-placement *intensity* (mere presence) nor influencer *popularity* nor
price predicts sales—what matters is whether engagement is *aimed at the product*. An
additional dataset shows influencers' PE-scores are higher when promoting their *own*
products, evidence that incentive alignment shapes how product-focused the content is.
Reassuringly for the method's validity, the neural saliency maps correlate highly with
actual gaze maps from eye-tracking.

These results sit alongside a broader move to mine engagement and its drivers from
unstructured social data [@yang2024engagement; @lovett2013], and they carry a blunt
practical lesson: a brand paying for engagement should contract on
*product-focused* engagement, because general engagement is, empirically, uncorrelated
with the outcome the brand actually wants.

## Valuing Influencer Accounts and Posts {#sec-valuation}

The recurring managerial decision—what to pay an influencer, or what a single sponsored
post is worth—is a valuation problem, and it is productive to import the discipline of
financial valuation while respecting where social-media assets differ. The parallels
are close enough to borrow frameworks and loose enough to demand caution: an
influencer's "cash flows" are sponsorship and affiliate revenues that are sporadic,
algorithm-dependent, and contingent on an audience that can defect overnight.

@tbl-valuation-analogy maps the three canonical financial approaches onto their
social-media analogues.

| Finance method | Traditional use | Social-media analogue | Principal challenge |
|---|---|---|---|
| Discounted cash flow (DCF) | Discount projected firm cash flows to present value | Project and discount an account's future sponsorship/affiliate income | Revenue is sporadic and algorithm-dependent; the discount rate must reflect platform volatility |
| Comparable multiples | Value via peers' EBITDA/earnings multiples | Benchmark price-per-follower or price-per-engagement against peers in the niche | Metrics are not standardized across niches and platforms |
| Precedent transactions | Price from recent comparable acquisitions | Infer buyout value from comparable account/page sales | Private-account transaction data are scarce |

: Financial valuation frameworks and their influencer-marketing analogues. {#tbl-valuation-analogy}

### Post-Level Multiples

At the **post** level, valuation reduces to a chosen unit price multiplied by an
expected quantity. The common units are cost-per-thousand-impressions (CPM),
cost-per-engagement (CPE), cost-per-view (CPV), and cost-per-click (CPC):

$$
\text{Value}_{\text{CPM}} = \frac{\text{Impressions}}{1000}\times \text{CPM},
\qquad
\text{Value}_{\text{CPE}} = \text{Engagements}\times \text{CPE}.
$$ {#eq-post-multiples}

A closely related construct, **earned media value** (EMV), monetizes *organic* reach by
pricing it as if it had been bought, scaled by an adjustment factor for sentiment and
fit, $\text{EMV} = \text{Impressions}\times \text{CPM}\times \phi$, where $\phi$
captures qualitative quality. EMV's well-known weakness is its sensitivity to the
analyst's choice of $\phi$, which is why it should be reported alongside the raw
inputs rather than as a single headline figure. Because no single multiple captures
both reach (CPM) and interaction (CPE), practitioners often blend them and apply a
qualitative adjustment for brand fit and sentiment, exactly as @yang2021first would
counsel by emphasizing *product-focused* rather than gross engagement.

```{r post-valuation}
followers       <- 200000
engagement_rate <- 0.03                       # 3% of followers engage
impressions     <- 0.75 * followers           # reach < followers
engagements     <- engagement_rate * followers

cpe <- 0.12; cpm <- 10                         # market unit prices
val_cpe <- engagements * cpe
val_cpm <- (impressions / 1000) * cpm

# Blend (weight CPE 40%, CPM 60%) then apply qualitative adjustment.
blend  <- 0.40 * val_cpe + 0.60 * val_cpm
final  <- blend * 1.20                          # +10% brand fit, +10% sentiment

data.frame(
  metric = c("CPE value", "CPM value", "Blended", "Final (adj.)"),
  value  = round(c(val_cpe, val_cpm, blend, final))
)
```

### Account-Level Discounted Cash Flow

At the **account** level, the appropriate object is the present value of the
influencer's future brand-deal income, net of production and agency costs, plus a
terminal value capturing the residual worth if the influencer pivots platforms or
winds down. With net annual brand income $C_t$ growing at rate $g$, discount rate $r$,
horizon $T$, and a terminal multiple or Gordon-growth tail, account value is

$$
V_0 \;=\; \sum_{t=1}^{T} \frac{C_t}{(1+r)^t}
        \;+\; \frac{\text{TV}}{(1+r)^T},
\qquad C_t = C_1 (1+g)^{t-1}.
$$ {#eq-account-dcf}

The discount rate $r$ should *exceed* a typical corporate weighted-average cost of
capital, because an influencer's income is more exposed to platform-algorithm changes,
audience fickleness, and competitive entry than a diversified firm's. The following
example values an account generating \$120{,}000 in year-one net income, growing
$5\%$ annually, discounted at $15\%$, with a terminal value set to twice final-year
income.

```{r account-dcf}
C1 <- 120000; g <- 0.05; r <- 0.15; T <- 5
t   <- 1:T
C   <- C1 * (1 + g)^(t - 1)                    # projected net income
pv  <- C / (1 + r)^t                            # discounted income
TV  <- 2 * C[T]                                 # terminal value (2x final year)
pv_TV <- TV / (1 + r)^T

account_value <- sum(pv) + pv_TV
data.frame(
  year = t, income = round(C), pv_income = round(pv)
)
cat("PV of terminal value: $", format(round(pv_TV), big.mark = ","), "\n", sep = "")
cat("Account value:        $", format(round(account_value), big.mark = ","), "\n", sep = "")
```

These valuations are estimates resting on assumptions about growth, persistence, and
discounting, and they should be screened for the pathologies specific to social data:
fake followers and bot engagement that inflate the inputs, market saturation that
compresses multiples as supply grows, and abrupt algorithm changes that invalidate the
projection. A defensible valuation states its assumptions, blends multiple methods
(@tbl-valuation-analogy), and is refreshed as realized campaign outcomes arrive—much
as the marketing–finance event-study machinery of @sec-marketing-finance updates the
market's valuation of a firm's brand assets as news arrives.

## Key Takeaways

- Influencer marketing is persuasion *rented* from a third party; its central tension
  is that the sponsorship which generates the firm's reach also erodes the influencer's
  audience-conferred credibility (@sec-reputation-burning).
- Effectiveness is jointly determined by **source trustworthiness**, **content value**,
  and **receiver involvement**; several drivers (activity, fit, positivity) are
  **inverted-U** shaped, and key effects (expertise, hedonic content) interact with
  campaign intent and platform involvement (@sec-sponsored-blogging).
- The empirical workhorse is a **selection-corrected overdispersed count model**:
  negative binomial for the engagement outcome (@eq-nb-mean), Heckman-type correction
  for non-random influencer selection, identified by an **exclusion restriction** that
  must be defended on substance, not statistics (@sec-engagement-model).
- Influencer **type** is an equilibrium role, not a size class: micro-influencers act
  as **quality screeners** and raise information quality, while celebrities raise
  awareness (@nistor2024influencers); selection should trade reach against cost via an
  elasticity such as FEI (@tian2024mega).
- **Engagement is not sales.** Traditional engagement metrics do not predict sales lift;
  *product-focused* engagement, measured by the PE-score (@eq-pe-score), does
  [@yang2021first].
- Account and post **valuation** borrows DCF, multiples, and precedent-transaction logic
  from finance, adjusted for platform volatility and screened for inflated metrics
  (@sec-valuation).
