# Pricing {#sec-pricing}

Price is the only element of the marketing mix that directly produces revenue;
every other element—product, promotion, place—creates cost. It is also the most
quickly adjusted lever a firm controls, the one most visible to competitors, and
the one most freighted with psychological meaning for buyers. A one-percentage-point
improvement in realized price typically lifts operating profit by far more than an
equivalent improvement in volume or variable cost, because price flows to the bottom
line undiluted. Yet pricing decisions are routinely delegated to spreadsheets and
gut feel, precisely because the object the manager most needs—each buyer's
*willingness to pay*—is unobserved and actively concealed during exchange.

This chapter treats pricing as the problem of recovering a latent demand object and
acting on it. We begin with the primitive: the **reservation price**, the maximum a
buyer will pay, and the demand curve that aggregates reservation prices across a
market. From the demand curve we derive the monopolist's optimal price and the
central quantity managers actually estimate—the **price elasticity of demand**—along
with its identification problem. We then study how firms extract more surplus than a
single price allows through **price discrimination** in its three textbook
degrees, and how that machinery is implemented as versioning, bundling, and
nonlinear tariffs. The second half of the chapter turns behavioral: buyers do not
respond to prices as the textbook supposes. They evaluate prices against
**reference points**, they treat a discount differently from a surcharge, they read
price as a signal of quality, and they are moved by the framing of a promotion as
much as by its depth. We close on the contemporary frontier—**dynamic and
personalized pricing**—where firms set prices that vary across time, context, and
individual, and where the gains from discrimination collide with consumer
perceptions of fairness and with privacy regulation.

The through-line is measurement. A price is only as good as the demand estimate
behind it, and every method we present—from a logit demand model to a conjoint
study to a field experiment—is in the end a strategy for estimating how quantity
responds to price when price is set by a firm that already knows something the
analyst does not. We lead with the economic intuition, give each method its
estimator and the assumption that breaks its identification, and supply seeded,
runnable code.

## Reservation Prices and Willingness to Pay

The atomic construct of pricing is the **reservation price**, also called
*willingness to pay* (WTP): the maximum amount a particular consumer would pay for a
unit of a product rather than forgo it. Formally, let a consumer derive gross
utility (in monetary units) $v$ from acquiring one unit. Facing price $p$, the
consumer buys if and only if consumer surplus is nonnegative,

$$
\text{buy} \iff v - p \ge 0 \iff p \le v,
$$ {#eq-reservation}

so $v$ *is* the reservation price: the indifference price at which the consumer is
exactly willing to transact. Reservation prices are heterogeneous across consumers;
the distribution of $v$ in the population is the economic content of "demand."

### From Reservation Prices to the Demand Curve

Let $F(\cdot)$ be the cumulative distribution function of reservation prices across a
unit mass of potential buyers, so $F(p) = \Pr(v \le p)$. At posted price $p$, the
buyers are exactly those with $v \ge p$, a fraction $1 - F(p)$. Market demand is
therefore

$$
Q(p) = M\,\bigl[1 - F(p)\bigr],
$$ {#eq-demand-from-wtp}

where $M$ is market size. The demand curve is the *survival function* of the
reservation-price distribution, scaled by $M$. This identity is the conceptual
bridge between the behavioral object (what individuals will pay) and the aggregate
object (how quantity responds to price): the slope of demand is governed by how
reservation prices are dispersed. A market in which everyone values the good
identically yields a flat demand curve with a single kink; wide dispersion in $v$
yields a smooth, gently sloped curve.

::: {.callout-note}
## Reservation price is not the same as the price a consumer *expects* or *prefers*
The reservation price is a *threshold*—the worst deal the consumer would still
accept—not the price the consumer hopes to pay or considers fair. Survey methods
that ask "what is a reasonable price?" measure a reference price (covered in
@sec-reference-prices), not the reservation price. Conflating the two systematically
understates WTP and leaves money on the table.
:::

### The Monopolist's Price and the Elasticity

A firm with constant marginal cost $c$ choosing a single price solves
$\max_p (p - c)\,Q(p)$. The first-order condition can be written in the canonical
**inverse-elasticity (Lerner) form**

$$
\frac{p^\star - c}{p^\star} = -\frac{1}{\varepsilon(p^\star)},
\qquad
\varepsilon(p) \equiv \frac{\partial Q}{\partial p}\frac{p}{Q},
$$ {#eq-lerner}

where $\varepsilon < 0$ is the **price elasticity of demand**, the percentage change
in quantity per percentage change in price. The optimal markup over cost, expressed
as a fraction of price, equals the reciprocal of the (absolute) elasticity: inelastic
demand ($|\varepsilon|$ small) supports a high markup; elastic demand forces price
toward cost. Profit maximization requires operating where $|\varepsilon| > 1$;
no firm optimally prices on the inelastic portion of its demand curve, because there
a price increase would raise revenue *and* cut cost. The elasticity is thus the
single most consequential number in pricing, and most of empirical pricing research
is, at bottom, the estimation of $\varepsilon$.

A workhorse functional form is constant-elasticity (log-log) demand,
$\log Q = \alpha + \varepsilon \log p + u$, under which $\varepsilon$ is a single
parameter. The following seeded example simulates a market of heterogeneous
reservation prices, recovers the demand curve as the empirical survival function per
@eq-demand-from-wtp, and locates the profit-maximizing price.

```{r wtp-demand, message=FALSE, warning=FALSE}
set.seed(17)
M <- 100000                       # market size
v <- rlnorm(M, meanlog = log(40), sdlog = 0.5)   # reservation prices
c <- 15                           # marginal cost

price_grid <- seq(15, 120, by = 1)
demand  <- sapply(price_grid, function(p) mean(v >= p)) * M   # eq-demand-from-wtp
profit  <- (price_grid - c) * demand

p_star  <- price_grid[which.max(profit)]
q_star  <- demand[which.max(profit)]

# point elasticity at the optimum via numerical derivative
dQ <- diff(demand) / diff(price_grid)
elas_at_star <- dQ[which.max(profit)] * p_star / q_star

cat("Optimal price p*:        ", p_star, "\n")
cat("Quantity at p*:          ", round(q_star), "\n")
cat("Elasticity at p*:        ", round(elas_at_star, 2), "\n")
cat("Lerner markup (p*-c)/p*: ", round((p_star - c) / p_star, 3),
    " vs. -1/elasticity =", round(-1 / elas_at_star, 3), "\n")
```

The recovered markup matches $-1/\varepsilon$ to within grid resolution, confirming
@eq-lerner numerically.

### Eliciting Willingness to Pay

Because reservation prices are private, firms must elicit them. Four families of
methods dominate, trading off realism against control.

**Direct survey** methods ask consumers their WTP, either openly or through a
sequence of yes/no purchase questions (the contingent-valuation tradition). They are
cheap but suffer **hypothetical bias**: respondents overstate WTP when no money
changes hands, and the open-ended format invites strategic understatement.

**Incentive-compatible mechanisms** make truth-telling a dominant strategy by
binding the respondent to a real transaction. In the **Becker–DeGroot–Marschak
(BDM)** procedure the respondent states a bid $b$; a price $p$ is then drawn at
random, and the respondent buys at $p$ if $b \ge p$. Because the stated bid never
sets the price the respondent pays, only whether they transact, the respondent's
optimal bid is exactly their true reservation price $v$.<!-- CITE-NEEDED: Becker, DeGroot, Marschak (1964) measuring utility by a single-response sequential method --> Second-price (Vickrey) auctions
share this incentive property.

**Conjoint analysis** infers WTP indirectly from trade-offs. Respondents choose
among profiles that vary attributes (including price); a discrete-choice model
recovers part-worth utilities, and the **monetary value of an attribute** is the
ratio of its part-worth to the price coefficient. If utility is
$U_{ij} = \mathbf{x}_j'\boldsymbol{\beta} + \beta_p\,p_j + \varepsilon_{ij}$, the WTP
for a unit change in attribute $k$ is $-\beta_k/\beta_p$, and the reservation price
for a whole profile is the price that drives its choice probability against the
outside option to one-half. Choice-based conjoint with hierarchical Bayes estimation
is the industry standard for new-product pricing precisely because it disciplines the
hypothetical task with realistic competitive trade-offs and recovers
*individual-level* heterogeneity in $\boldsymbol{\beta}$ [@toubia2013; @ding2015learning].
A complementary structural approach maps the conjoint primitives directly into
demand and competitive equilibrium so that the elicited utilities deliver an optimal
price rather than a ranking [@jedidi2003; @jedidi2021].

**Revealed-preference / market data** methods recover WTP from actual purchases.
These are the most credible because choices are consequential, but they are silent
about prices never observed in the data and carry the identification problem we turn
to next.

::: {.callout-tip}
## Estimator box: WTP from a choice model
Given choice data, estimate $\boldsymbol{\beta}$ and $\beta_p$ by maximum likelihood
(logit) or hierarchical Bayes (mixed logit). The WTP for attribute $k$ is
$\widehat{\text{WTP}}_k = -\hat\beta_k/\hat\beta_p$. **What breaks it:** the ratio is
fragile when $\hat\beta_p \approx 0$ (a weak or wrong-signed price coefficient sends
WTP to $\pm\infty$); price endogeneity biases $\hat\beta_p$ toward zero and inflates
WTP; and a fixed $\beta_p$ across respondents imposes the implausible restriction
that the marginal utility of income is identical for all, which the mixed-logit
random coefficient relaxes [@dube2021random].
:::

## The Identification Problem in Demand Estimation {#sec-price-identification}

Estimating elasticity from observational price–quantity data confronts the oldest
problem in econometrics: **price is endogenous**. Firms set prices in response to
demand conditions the analyst does not see—a popular item is priced high *because*
it is popular—so a naïve regression of quantity on price recovers a tangle of the
demand curve and the firm's pricing rule, biasing the elasticity toward zero (often
to the wrong sign). Identification requires an instrument or structure that shifts
price without shifting demand.

### The Logit Demand System and BLP

The modern standard for differentiated-products demand is the random-coefficients
logit, estimated by the method of @berry1995 (BLP). Consumer $i$'s indirect utility
for product $j$ in market $t$ is

$$
u_{ijt} = \mathbf{x}_{jt}'\boldsymbol{\beta}_i - \alpha_i\,p_{jt} + \xi_{jt} + \varepsilon_{ijt},
$$ {#eq-blp}

where $\mathbf{x}_{jt}$ are observed characteristics, $p_{jt}$ is price,
$\xi_{jt}$ is the **unobserved (to the analyst) product quality** that consumers and
firms see, and $\varepsilon_{ijt}$ is an extreme-value taste shock. The coefficients
$(\boldsymbol{\beta}_i, \alpha_i)$ are random across consumers, generating realistic
substitution patterns. The endogeneity is explicit and structural: price is
correlated with $\xi_{jt}$ because firms price higher where unobserved quality is
higher, $\mathbb{E}[p_{jt}\,\xi_{jt}] \ne 0$.

**Estimator.** Invert market shares to recover the mean utility
$\delta_{jt}(\boldsymbol{\theta}) = \mathbf{x}_{jt}'\boldsymbol{\beta} - \alpha p_{jt} + \xi_{jt}$,
form the structural error $\xi_{jt}(\boldsymbol{\theta})$, and choose the parameters
by GMM to satisfy the moment condition $\mathbb{E}[\xi_{jt} \mid \mathbf{z}_{jt}] = 0$
for instruments $\mathbf{z}_{jt}$. **What identifies it:** valid instruments shift
price (markups) but are mean-independent of $\xi_{jt}$—cost shifters, and the classic
*BLP instruments* formed from characteristics of *rival* products, which move a
firm's markup through competition without entering its own demand shock. **What
breaks it:** weak instruments (rival characteristics that barely move price) inflate
standard errors and bias estimates toward OLS; instruments correlated with
$\xi_{jt}$ (e.g., rival quality that proxies own quality in a correlated market)
reintroduce the bias the design was meant to remove. The structural demand literature
in marketing builds directly on this machinery to estimate price response with panel
and scanner data [@chintagunta2006structural; @reiss2011; @nair2017].

### Why Field Experiments Are the Clean Benchmark

A randomized price experiment severs the link between price and $\xi_{jt}$ by
construction: when the firm assigns price at random, $\mathbb{E}[p\,\xi] = 0$ by
design, and the elasticity is identified by the experimental contrast alone. This is
why pricing field experiments are the gold standard for elasticity, and why much
recent work prices via controlled tests rather than relying on historical variation
[@simester2010]. The cost is external validity—an experiment identifies the
elasticity at the tested prices, in the tested context, for the tested population—and
the standard practical hazard is that competitors or consumers detect and react to
the test. @fig-identification summarizes the identification logic that organizes the
remainder of the methods in this chapter.

```{mermaid}
%%| label: fig-identification
%%| fig-cap: "The price-endogeneity problem and three identification strategies. Unobserved demand shifters drive both price and quantity; credible elasticity estimates require breaking that back-door path."
flowchart TD
    XI["Unobserved demand<br/>shifter (quality, ξ)"] -->|firm prices higher| P["Price p"]
    XI -->|raises demand| Q["Quantity Q"]
    P -->|true causal effect ε| Q
    P -.->|"naïve OLS conflates<br/>both arrows"| BIAS["Biased elasticity"]
    IV["Instrument:<br/>cost / rival shifters"] -->|shifts p, not ξ| P
    EXP["Randomized price"] -->|cuts ξ→p link| P
    STR["Structural model:<br/>BLP moments"] -->|"E[ξ|z]=0"| P
    style XI fill:#f8d7da,stroke:#842029
    style EXP fill:#d1e7dd,stroke:#0f5132
    style IV fill:#fff3cd,stroke:#664d03
    style STR fill:#cfe2ff,stroke:#084298
```

## Price Discrimination

A single price leaves surplus uncaptured on both sides: high-WTP buyers pocket
consumer surplus, and low-WTP buyers who would have paid more than cost are priced
out entirely. **Price discrimination**—charging different prices that do not
reflect differences in cost—aims to recover both. The classical taxonomy
distinguishes three degrees by the information the firm exploits.

| Degree | Information used | Mechanism | Surplus captured | Real-world example |
|---|---|---|---|---|
| First | Each buyer's exact $v$ | Personalized price | All surplus (in theory) | Negotiated B2B deals; algorithmic personalization |
| Second | $v$ self-revealed through choice | Versioning, bundling, quantity discounts, nonlinear tariffs | Partial; buyers sort themselves | Software tiers; airline fare classes; family-size packs |
| Third | Observable group membership correlated with $v$ | Different price per segment | Partial; bounded by arbitrage | Student/senior discounts; geographic pricing |

: The three degrees of price discrimination, by the information the firm exploits. {#tbl-discrimination}

### First-Degree (Perfect) Discrimination

Under **first-degree** discrimination the firm knows each buyer's reservation price
and charges exactly $v$, extracting the entire surplus and selling to everyone whose
$v$ exceeds marginal cost. Output is efficient (no willing buyer-above-cost is
excluded) but all surplus accrues to the firm. Perfect discrimination is an idealization,
but it is the conceptual ceiling that personalized pricing (@sec-dynamic-personalized)
approaches as the firm's information about individual WTP improves.

### Second-Degree Discrimination: Self-Selection

When the firm cannot observe $v$ but knows its distribution, it offers a *menu* and
lets buyers sort themselves—**second-degree** discrimination via self-selection. The
design problem is a screening problem: construct options so that each consumer type
prefers the option intended for it. Two constraints bind. **Individual rationality
(IR):** each type must (weakly) prefer its option to no purchase. **Incentive
compatibility (IC):** each type must (weakly) prefer its own option to every other
option on the menu. The IC constraint is what forces the firm to leave **information
rent** to high-WTP types: to stop a high type from masquerading as a low type, the
firm must discount the high type's option relative to first-degree, a markdown that
is the price of not observing $v$ directly.

Three implementations recur:

*Versioning* offers quality-differentiated variants—a full-featured product and a
deliberately degraded one—so that high-WTP buyers self-select into the premium
version. The damaged-goods logic explains why firms incur cost to *remove* features:
the degraded version exists to make the premium version's price incentive-compatible
for high types.<!-- CITE-NEEDED: Deneckere & McAfee (1996) damaged goods -->

*Bundling* sells goods together at a price below the sum of standalone prices. When
reservation prices for components are **negatively correlated** across buyers,
bundling reduces the dispersion of total WTP and lets the firm price closer to the
common bundle value, raising profit—the classic Stigler result.<!-- CITE-NEEDED: Stigler (1963) / Adams & Yellen (1976) bundling --> *Mixed bundling*,
offering components both separately and as a bundle, dominates pure bundling when
component valuations are heterogeneous, and is the architecture behind cable tiers
and software suites.

*Nonlinear (quantity) pricing* charges a per-unit price that depends on quantity—a
two-part tariff (fixed fee plus marginal price) or block tariff. With heterogeneous
demand intensity, the optimal nonlinear schedule again leaves information rent to
high-volume buyers and distorts the low-volume option downward. The marketing
literature studies how such schedules interact with competition and with consumers'
misperception of their own future usage [@iyer2005].

### Third-Degree Discrimination: Observable Segments

When the firm observes a segment indicator correlated with WTP—age, location,
student status, purchase history—it sets a separate optimal price per segment, each
satisfying its own Lerner condition: charge the *more inelastic* segment the *higher*
price. **Third-degree** discrimination is the most common in practice (coupons,
geographic and channel pricing, demographic discounts) and is bounded by
**arbitrage**: if low-price buyers can resell to high-price buyers, the scheme
unravels, which is why discriminated goods are typically services, perishable, or
identity-linked (a student ID, a loyalty account). The following example contrasts
uniform pricing with optimal third-degree pricing across two segments with different
elasticities.

```{r third-degree, message=FALSE, warning=FALSE}
set.seed(170)
c <- 15
# Two segments with constant-elasticity demand Q = A * p^eps
seg <- data.frame(
  name = c("price-sensitive", "price-insensitive"),
  A    = c(5e6, 2e6),
  eps  = c(-2.5, -1.4)
)

# Optimal monopoly price under constant elasticity: p* = c * eps/(eps+1)
seg$p_opt  <- c * seg$eps / (seg$eps + 1)
seg$q_opt  <- seg$A * seg$p_opt^seg$eps
seg$profit <- (seg$p_opt - c) * seg$q_opt

# Uniform price: maximize total profit over a single price
pg <- seq(16, 80, by = 0.5)
tot <- sapply(pg, function(p) sum((p - c) * seg$A * p^seg$eps))
p_uniform <- pg[which.max(tot)]
profit_uniform <- max(tot)

knitr::kable(
  transform(seg, p_opt = round(p_opt, 2), profit = round(profit)),
  col.names = c("Segment", "Scale A", "Elasticity", "Optimal price", "Quantity", "Profit"),
  caption = "Optimal third-degree prices differ across segments by elasticity."
)
cat("Uniform price:        ", p_uniform, " | profit:", round(profit_uniform), "\n")
cat("Discrimination profit:", round(sum(seg$profit)),
    " | gain:", round(100 * (sum(seg$profit) / profit_uniform - 1), 1), "%\n")
```

The price-insensitive segment optimally pays the higher price, and segmenting raises
profit over the best single price—the basic dividend of discrimination.

## Promotions and Discounting

Few products sell at a constant price for long. **Temporary price promotions**—a
discount of limited duration—are ubiquitous in consumer packaged goods, where a large
share of volume moves on deal. The empirical study of promotions is among the most
mature in marketing because scanner data make deal incidence, depth, and timing
directly observable.

### Why Promotions Exist: Price Discrimination Over Time

A leading rationale for promotions is **intertemporal price discrimination**.
Consumers differ in search and storage costs: price-sensitive "cherry pickers" pay
attention to deals and stockpile, while loyal or time-pressed buyers pay the regular
price. A pattern of high regular prices punctuated by deep discounts lets the firm
charge the two groups different effective prices without an observable segment
indicator—a temporal analogue of second-degree discrimination. @narasimhan1988b
models this coupon-and-deal logic, and @varian1980 shows that even with *homogeneous*
goods, the mix of informed and uninformed consumers supports an equilibrium in which
firms randomize prices (run sales) rather than converging on a single price. The
implication is structural, not merely tactical: price *dispersion* and promotional
*cycling* are equilibrium outcomes, not pricing errors.

### Decomposing the Sales Spike

A promotion's observed sales bump aggregates several distinct consumer responses,
and a credible promotion model must separate them because they have opposite
implications for incremental profit:

- **Brand switching**—buyers substitute toward the promoted brand from competitors;
  genuinely incremental to the brand.
- **Category expansion**—buyers consume more of the category; incremental to the
  category.
- **Purchase acceleration**—buyers who would have bought later buy now; a timing
  shift that *borrows* from future sales.
- **Stockpiling**—buyers increase inventory at home, depressing post-promotion demand
  (the **post-promotion dip**).

Estimating these shares is the central task of promotion analytics. Household panel
decompositions consistently find that the large majority of a promotion's sales spike
comes from brand switching rather than category expansion, which means much
promotional volume is incremental to the *brand* but cannibalizes the firm's own
future sales and rival sales rather than growing the pie [@vanheerde2007; @bell1998].
@guyt2014 show that the decomposition itself depends on whether the analyst uses
store or household data and on the deal-frequency environment, a caution against
reading a single elasticity off aggregate data.

### The Reference-Price Trap and Promotion Dynamics

Frequent discounting is not free even when each deal is individually profitable.
Promotions erode the **reference price**—the internal benchmark against which buyers
judge the next price (@sec-reference-prices)—so that a brand which deals often trains
its customers to wait, raising deal elasticity and depressing baseline sales over
time [@kopalle1992].<!-- CITE-NEEDED: Mela, Gupta & Lehmann (1997) long-term promotion effects on brand choice --> The dynamic optimal promotion policy therefore trades
the short-run volume gain against the long-run reference-price cost; @kopalle1992
characterize this as a dynamic-programming problem in which the state is the prevailing
reference price. Promotions also have **asymmetric** competitive effects: a stronger
or higher-tier brand's promotion steals more share than a weaker brand's promotion of
equal depth, so promotional power is not symmetric across the price tier
[@sethuraman2011].<!-- CITE-NEEDED: Blattberg & Wisniewski (1989) price-tier competition asymmetry -->


### Promotion Framing and the Multitier Discount

How a discount is *communicated* affects response beyond its arithmetic depth—a
behavioral theme we develop fully in @sec-behavioral-pricing. One frontier result
concerns **multitier discounting**: presenting a quantity discount alongside an
additional discount tier (for example, a "buy-one" price displayed next to a
"buy-more" price) makes the quantity discount more attractive by inflating the
perceived savings, even though the additional tier need not be the option chosen.
@yang2024express show that such multitier displays raise clickthrough, with the effect
strengthening when the additional tier's price is lower (amplifying perceived
savings), among frequent category purchasers, and weakening among consumers who
distrust advertised reference prices. The mechanism is reference-dependent: the extra
tier resets the comparison standard against which the chosen option is evaluated,
which is why the same discount performs differently under different framings.

## Reference Prices {#sec-reference-prices}

The standard model treats demand as a function of the *absolute* price. Decades of
evidence contradict this: consumers judge a price relative to an internal standard,
the **reference price**, and respond to the *deviation* between the observed price and
that standard. The reference price may be **internal** (recalled from past purchases,
formed from price history) or **external** (a manufacturer's suggested price, a
"regular price" struck through on the tag, a competitor's displayed price). Reference
effects are the empirical foundation of behavioral pricing.

### Loss Aversion and Asymmetric Response

The dominant theoretical account imports **prospect theory** [@tversky1981]:<!-- CITE-NEEDED: Kahneman & Tversky (1979) prospect theory original --> buyers
code a price below their reference point as a **gain** and a price above it as a
**loss**, and because losses loom larger than gains (loss aversion), the demand
response to a price *increase* above the reference exceeds the response to an equal
price *decrease* below it. Let $r$ denote the reference price and $p$ the observed
price. A reference-dependent demand specification splits the price deviation into gain
and loss regions,

$$
\log Q = \alpha + \beta\,\log p
        + \gamma_{\text{gain}}\,(r - p)^{+}
        - \gamma_{\text{loss}}\,(p - r)^{+} + u,
$$ {#eq-reference-demand}

where $(\cdot)^{+} = \max(\cdot, 0)$ and loss aversion predicts
$\gamma_{\text{loss}} > \gamma_{\text{gain}} > 0$. The empirical literature, beginning
with brand-choice models on scanner panels, robustly finds this asymmetry: surcharges
relative to the reference price suppress purchase more than equal discounts stimulate
it.<!-- CITE-NEEDED: Winer (1986) reference price brand choice; Kalwani et al. (1990) asymmetric reference effects --> The asymmetry is the reason a sequence of small increases is less damaging
than one large increase, and why "everyday low pricing" can outperform high-low
promotion for some categories: it stabilizes the reference point.

### Estimating the Reference Price

The reference price is latent, which raises an identification problem of its own:
the analyst must *posit* a formation rule before estimating response. The two leading
formulations are a **memory-based** rule, in which the reference adapts toward
observed prices via exponential smoothing,
$r_t = \theta\,r_{t-1} + (1-\theta)\,p_{t-1}$, and a **stimulus-based** (contextual)
rule, in which the reference is constructed from prices currently on the shelf. The
two are observationally distinct and frequently both contribute; a model that omits
the operative rule mis-attributes reference effects to absolute-price response.
Crucially, the smoothing parameter $\theta$ and the response coefficients
$(\gamma_{\text{gain}}, \gamma_{\text{loss}})$ are **jointly** identified only with
sufficient *independent* variation in price and price history—if price follows a
deterministic high-low cycle, the reference and the current price move together and
the gain/loss coefficients are not separately identified from the level coefficient
$\beta$. This is the reference-price analogue of the endogeneity problem in
@sec-price-identification: collinearity between the price and its own lagged transform, not
firm behavior, is what breaks identification here.

```{r reference-price, message=FALSE, warning=FALSE}
set.seed(1717)
n <- 400
theta <- 0.7                      # reference-price memory
p <- 10 + 2 * sin(seq_len(n) / 5) + rnorm(n, 0, 0.8)   # noisy price path

# build the memory-based reference price
r <- numeric(n); r[1] <- p[1]
for (t in 2:n) r[t] <- theta * r[t - 1] + (1 - theta) * p[t - 1]

gain <- pmax(r - p, 0); loss <- pmax(p - r, 0)
# data-generating demand with loss aversion: loss coef > gain coef
logQ <- 6 - 0.8 * log(p) + 0.5 * gain - 1.6 * loss + rnorm(n, 0, 0.05)

fit <- lm(logQ ~ log(p) + gain + loss)
round(coef(summary(fit))[, 1:2], 3)
```

The fitted loss coefficient is larger in magnitude than the gain coefficient,
recovering the loss-aversion asymmetry the data were built with—provided the price
path carries enough independent variation to separate the two.

## Behavioral Pricing {#sec-behavioral-pricing}

Reference dependence is one of a broader family of departures from the rational
single-price model. **Behavioral pricing** studies how the *psychology* of price
perception—framing, fluency, signaling, and mental accounting—shapes demand in ways
the firm can exploit (or be exploited by). We organize the main effects below; each
has a clean experimental signature and a pricing implication.

### Price as a Quality Signal

When quality is unobservable before purchase, price itself carries information:
buyers infer that a higher price signals higher quality, especially in categories
where quality is hard to judge and stakes are high. This **price–quality inference**
inverts the usual demand logic over a range—raising price can *raise* perceived value
and, in thin information environments, raise demand. The mechanism is the signaling
logic of @sec-branding and the lemons problem of @akerlof1970: price can support a
separating equilibrium in which only high-quality firms can profitably sustain a high
price, so a high price is credible quality information. The pricing implication is
that *underpricing a premium product can suppress demand* by signaling inferior
quality, a failure mode invisible to a model in which demand is monotone decreasing
in price.

### Price Endings, Precision, and Fluency

Prices are not processed as raw numbers. **Just-below (9-ending) pricing**—\$9.99
rather than \$10.00—raises demand beyond the trivial one-cent saving, through a
left-digit anchoring effect (the leading digit dominates magnitude perception) and a
learned association of 9-endings with discounts.<!-- CITE-NEEDED: Stiving & Winer (1997); Anderson & Simester (2003) 9-ending field experiments --> The framing of a price's
**precision** also matters: round prices are processed more fluently and feel
"right" for emotional or hedonic purchases, while precise prices signal that a figure
was carefully computed and can anchor negotiations more effectively
[@janiszewski2016]. These effects are robust enough to be exploited routinely in
retail, but they are also category- and context-dependent, and a 9-ending that signals
"discount" can *undercut* a premium positioning.

### Partitioned Pricing, Drip Pricing, and Mental Accounting

How a total price is *split* changes its perceived magnitude. **Partitioned
pricing**—separating a base price from a surcharge (shipping, handling, a resort
fee)—can lower the perceived total because attention anchors on the salient base
component and underweights add-ons. Its aggressive cousin, **drip pricing**, reveals
mandatory charges sequentially during checkout, exploiting the sunk cost of search
effort already invested. The mechanism is **mental accounting** [@thaler1985]:
consumers evaluate gains and losses in separate cognitive accounts rather than
integrating them, so the framing of a price into components—and the framing of a
promotion as a *discount* versus a *rebate* versus a *bonus pack*—changes evaluation
even when the net economics are identical. The same logic explains why bundling a
small loss with a larger gain (silver linings) and segregating multiple gains
maximizes perceived value.

### The Pennies-a-Day and Temporal Reframing

Reframing a price's *temporal* unit alters its acceptability: \$365 per year framed as
"a dollar a day" recruits a more favorable comparison standard (a trivial daily
expenditure) and lifts compliance, the **pennies-a-day** effect. Subscription pricing
exploits this directly. The common thread across these effects is that the *number*
the consumer compares against—the reference, the salient digit, the temporal unit—is
malleable, and the firm partly controls it through framing. @fig-behavioral organizes
the behavioral effects by the cognitive mechanism each exploits.

```{mermaid}
%%| label: fig-behavioral
%%| fig-cap: "A map of behavioral pricing effects by underlying mechanism. Each effect is a lever on the comparison standard or the perceived magnitude of a price, not on its arithmetic level."
flowchart LR
    subgraph REF["Reference dependence"]
      A1["Loss aversion<br/>(asymmetric response)"]
      A2["Promotion erosion<br/>of reference price"]
      A3["Multitier discount<br/>framing"]
    end
    subgraph FLU["Fluency & anchoring"]
      B1["9-ending / left-digit"]
      B2["Round vs. precise prices"]
    end
    subgraph MA["Mental accounting"]
      C1["Partitioned / drip pricing"]
      C2["Pennies-a-day reframing"]
      C3["Discount vs. rebate framing"]
    end
    subgraph SIG["Inference"]
      D1["Price–quality signaling"]
    end
    REF --> OUT["Demand deviates from<br/>the single-price benchmark"]
    FLU --> OUT
    MA --> OUT
    SIG --> OUT
    style OUT fill:#d1e7dd,stroke:#0f5132
```

## Dynamic and Personalized Pricing {#sec-dynamic-personalized}

The classical theory sets one price for all buyers at one time. Digital commerce
relaxes both restrictions. **Dynamic pricing** varies price over time in response to
demand, inventory, and competition; **personalized pricing** varies price across
individuals in response to what the firm infers about each buyer's WTP. Together they
push pricing toward the first-degree ideal of @tbl-discrimination—and toward its
ethical and regulatory limits.

### Dynamic Pricing: Learning and Inventory

Two distinct forces drive price over time. The first is **demand learning**: a firm
uncertain about the demand curve experiments with prices to estimate elasticity, then
exploits the estimate—a price-setting bandit problem in which the firm balances the
*exploration* value of an informative price against the *exploitation* value of the
currently estimated optimum. The second is **intertemporal capacity allocation**: when
capacity is fixed and perishable (airline seats, hotel nights, event tickets), the
firm raises price as the sell-by date approaches and inventory tightens, the logic of
revenue management. Both are now studied structurally on transaction data
[@elberg2019dynamic; @cosguner2022dynamic], and both interact with **forward-looking
consumers** who anticipate price paths: when buyers expect a clearance, they delay,
and the firm's optimal path must account for strategic waiting—the durable-goods
monopolist's commitment problem.

### Personalized Pricing and the Value of Data

Personalized pricing conditions the offer on individual data—browsing history, device,
location, purchase record—to approximate each buyer's reservation price. Online, the
firm can in principle observe a behavioral proxy for WTP and price against it, an
operational route toward first-degree discrimination. The economics are double-edged.
@shapiro2016 and the price-discrimination-with-data literature show that **better
consumer information** lets the firm extract more surplus, but competition can flip the
sign: when rivals also personalize, the same data that enables extraction also
intensifies price competition for each contestable consumer, and consumers can be made
*better* off in aggregate. The welfare and profit consequences of personalization thus
depend on market structure—monopoly personalization transfers surplus to the firm,
competitive personalization can dissipate it [@chen2009; @aguirre2016]. The data
itself becomes a strategic asset whose value is exactly the incremental surplus the
firm can extract by conditioning price on it.

### Algorithmic Pricing and Collusion

When competing firms delegate pricing to learning algorithms, a new hazard emerges:
algorithms that independently learn to set supra-competitive prices, sustaining
**tacit collusion** without any explicit agreement. Because the algorithms reach a
collusive outcome through repeated interaction rather than communication, the conduct
falls outside the reach of antitrust doctrine built for human agreements, an active
concern at the marketing–policy frontier.<!-- CITE-NEEDED: Calvano, Calzolari, Denicolò & Pastorello (2020) algorithmic collusion --> The same automation that lets a
firm respond to demand in real time can, in a market of similar algorithms, soften
price competition without anyone intending it.

### Fairness and the Backlash Constraint

The binding constraint on personalized and dynamic pricing is often not the law but
**perceived fairness**. Consumers judge price differences they cannot attribute to
cost as unfair, and react with reduced purchase, reduced loyalty, and reputational
punishment. The reference-transaction model of fairness holds that buyers anchor on a
fair *reference price* (the price others pay, or the price they paid before) and treat
a personalized surcharge as a loss imposed on them—linking fairness directly to the
reference-dependence machinery of @sec-reference-prices. The canonical cautionary
case is the consumer backlash against early online price experiments that charged
different customers different prices for the same item; the lesson generalizes to
surge pricing, where transparency and a cost-based rationale moderate the backlash.
**Surcharge framing** (a fee added) provokes more outrage than **discount framing** (a
lower price offered) for the identical price schedule, a direct consequence of loss
aversion. The managerial upshot is that the *feasible* degree of discrimination is set
not by the firm's data but by what consumers will tolerate before the fairness
penalty exceeds the discrimination dividend.

### Privacy Regulation as a Constraint on Personalization

Personalized pricing runs on personal data, so privacy regulation directly bounds it.
Restrictions on tracking and data collection—and consumers' own privacy choices—shrink
the information set on which prices can be conditioned, attenuating the firm's move
toward first-degree discrimination. The evidence is that privacy regulation and
opt-out reduce the granularity of targeting available to firms, with measurable effects
on the data economy that underwrites personalization [@acquisti2012impact;
@johnson2014emergence]. Privacy is thus not merely a compliance overlay but a
first-order determinant of how far personalized pricing can go: the legal ceiling on
data acquisition is, in effect, a ceiling on price discrimination.

## Key Takeaways

- The primitive of pricing is the **reservation price**; the demand curve is the
  survival function of the reservation-price distribution (@eq-demand-from-wtp), and
  the **elasticity** governs the optimal markup through the Lerner condition
  (@eq-lerner). Estimating elasticity is the core empirical task.
- Observational price data are **endogenous**—firms price on information the analyst
  cannot see—so credible elasticities require an instrument (cost or rival shifters),
  a structural model (BLP, @eq-blp), or a randomized price experiment.
- **Price discrimination** recovers surplus a single price forgoes, in three degrees
  distinguished by information (@tbl-discrimination); second-degree menus must satisfy
  incentive compatibility, which forces the firm to leave **information rent** to
  high-WTP buyers.
- **Promotions** are largely intertemporal price discrimination; most of a deal's
  spike is brand switching, and frequent dealing erodes the **reference price**,
  imposing a dynamic cost on tactical gains.
- Buyers respond to prices relative to a **reference point** with **loss-averse
  asymmetry** (@eq-reference-demand), and to framing, fluency, signaling, and mental
  accounting—levers on the comparison standard rather than the price level.
- **Dynamic and personalized pricing** push toward first-degree extraction, but their
  feasible reach is bounded by competition, by **perceived fairness** (itself a
  reference-dependence phenomenon), and by **privacy regulation**.

## Further Reading

For the structural estimation of demand and price response, the differentiated-products
literature beginning with @berry1995 and developed in marketing by
@chintagunta2006structural and @nair2017 is the entry point; @dube2021random treats
random-coefficient logit estimation in depth. On promotions, @vanheerde2007 and
@kopalle1992 cover decomposition and the dynamic reference-price cost. The behavioral
strand connects to @sec-branding (price–quality signaling) and to the
prospect-theoretic foundations in @tversky1981 and @thaler1985. The personalization
frontier and its welfare economics are surveyed through @shapiro2016, @chen2009, and
@aguirre2016, with privacy constraints in @acquisti2012impact.
