# Healthcare and Pharmaceutical Marketing {#sec-health}

Healthcare is the largest sector of most advanced economies, and the marketing of
its products differs from consumer goods in a way that organizes this entire
chapter: the person who *consumes* a prescription drug is usually not the person
who *chooses* it. A physician writes the prescription; a patient swallows the
pill; a payer reimburses the cost; a regulator decides what may be claimed about
the molecule at all. Marketing therefore acts on a chain of agents with divergent
information and incentives, and the central empirical problems—who responds to
detailing, whether direct-to-consumer advertising expands the category or merely
reallocates it, how a new therapy diffuses through a professional network, what
prescribers do when the evidence itself disagrees—are problems of *separating
persuasion from selection* in exactly this multi-agent setting.

This chapter treats pharmaceutical and healthcare marketing as a set of measurement
and identification problems rather than a catalog of tactics. We begin with the
institutional structure that constrains every instrument (the prescribing agency
relationship, regulation, and the data that researchers actually observe). We then
take the four canonical levers in turn: **physician detailing** (personal selling
to prescribers), **direct-to-consumer advertising** (DTCA), the **diffusion of
medical innovations** through professional networks, and physician response to
**conflicting information**. For each, we lead with the economic intuition, state a
formal model with its estimator, and—critically—name what breaks identification,
because the recurring threat in this literature is that firms target their effort
exactly where response would have been high anyway. We close with the welfare and
policy questions that make the sector distinctive, and with reproducible code for a
detailing-response model and a Bass diffusion fit.

By the end the reader should be able to specify a prescription-response model that
separates a detailing effect from targeting endogeneity, set up a DTCA design that
distinguishes category expansion from business stealing, fit and interpret a
diffusion model on adoption data, and reason about how a Bayesian prescriber should
update beliefs when two trials contradict each other.

## The Prescribing Agency Relationship

The defining feature of the prescription market is a **double agency**. The patient
delegates the choice of therapy to a physician who is better informed, and the
physician in turn prescribes against a budget that a third party (an insurer, a
government program, or a pharmacy-benefit manager) ultimately pays. The physician is
thus an *expert agent* who neither bears the full cost nor consumes the product.
This structure has three consequences that recur throughout the chapter.

First, **the buyer's price sensitivity is muted and indirect**. Because the
prescriber does not pay and the patient pays only a copay, demand responds less to
list price than in ordinary markets and more to clinical attributes, habit, and
information—the very things marketing supplies. Pharmaceutical demand is, to a
first approximation, an information problem.

Second, **firms market to two audiences with different instruments**.
"Push" marketing—**detailing** (sales-representative visits), free samples,
sponsored education, and journal advertising—targets the prescriber. "Pull"
marketing—**direct-to-consumer advertising**—targets the patient, who then asks the
prescriber for the drug. The two are complements as often as substitutes, and a
recurring empirical question is how they interact [@narayanan2009].

> A *prescription* drug is a credence good chosen by an expert agent on behalf of a
> principal who consumes it, and paid for by a third party. Marketing in this market
> is the management of information flows to the agent and, where permitted, to the
> principal—under a regulatory ceiling on what may be claimed.

Third, **the product is a credence good under heavy regulation**. Patients cannot
verify efficacy even after consumption (was it the drug or the disease course?), so
claims are policed. In the United States the Food and Drug Administration (FDA)
regulates promotional claims, requires "fair balance" between benefit and risk, and
since its 1997 guidance has permitted broadcast DTCA with abbreviated risk
disclosure—a regulatory regime that makes the U.S. and New Zealand the only major
markets allowing product-claim DTCA.<!-- CITE-NEEDED: FDA 1997 DTCA guidance and cross-country comparison --> Most other
jurisdictions ban it, which is why much of the credible DTCA evidence exploits
this regulatory variation. Throughout, the reader should keep the regulatory
boundary in view: it is both a constraint on firms and a source of identifying
variation for researchers. @fig-agency sketches the agency chain and the two
marketing routes, push and pull, that act on it.

```{mermaid}
%%| label: fig-agency
%%| fig-cap: "The prescription agency chain and the two marketing routes. Push instruments act on the prescriber; pull instruments act on the patient, who petitions the prescriber. A regulator constrains claims on both routes."
flowchart LR
  F[Pharma firm] -->|detailing, samples,<br/>journal ads| MD[Physician<br/>expert agent]
  F -->|DTC advertising| PT[Patient<br/>principal/consumer]
  PT -->|requests drug| MD
  MD -->|writes Rx| PH[Pharmacy]
  PH --> PT
  PAY[Payer / PBM] -.->|reimburses| PH
  REG[FDA / regulator] -.->|constrains claims| F
  classDef firm fill:#e8eef7,stroke:#33415c;
  classDef reg fill:#f7ece8,stroke:#5c3a33;
  class F firm
  class REG,PAY reg
```

### What the Researcher Observes

The empirical literature is built largely on a small number of data structures, and
the available data dictate the feasible identification strategy. Three are central.
*Physician-level panels* (e.g., audits that record, for a sample of prescribers,
their monthly new prescriptions and the detailing calls and samples they received)
support individual-level response models and are the workhorse for detailing
research [@manchanda1999; @Manchanda_2004]. *Market-level time series* (national or
regional sales against aggregate advertising and detailing spend) support
category-versus-share questions and aggregate diffusion fits but cannot, on their
own, separate persuasion from heterogeneity. *Adoption records*—the dates at which
each physician first prescribes a new molecule, sometimes with the social ties among
them—support diffusion and contagion models [@Van_den_Bulte_2001; @iyengar2003].
The selection problem differs by structure: in physician panels the threat is that
calls are *targeted*; in market series it is that advertising is *timed* to demand;
in adoption data it is that *connected* physicians are also *similar*.

## Theoretical Foundations {#sec-health-theory}

The agency structure above explains the *institutions* of the prescription market;
a parallel set of behavioral theories explains the *individual* decisions (of
patients seeking care, of prescribers changing practice) that marketing tries to
move. Five strands recur throughout the chapter.

**The health belief model** holds that protective health action follows from a
person's perceived susceptibility to a condition, the perceived severity of its
consequences, the perceived benefits of acting, and the perceived barriers to doing
so, often triggered by a cue to action. It frames the patient side of direct-to-consumer
advertising (@sec-health, "Direct-to-Consumer Advertising"): an advertisement that
raises perceived susceptibility or lowers a perceived barrier is precisely a cue
that converts latent need into a request to prescribe.

**The theory of reasoned action and its extension, the theory of planned behavior**,
hold that behavior follows intention, which is in turn determined by attitudes,
subjective norms, and, in the planned-behavior extension, perceived behavioral control
[@ajzen1991planned; @sheppard1988]. This intention-to-behavior chain is the dominant
account of health compliance and adoption, and it underwrites the survey-based
predictive models that the prescription literature deploys.

**Protection-motivation theory** sharpens the fear-appeal logic by separating two
appraisals: a threat appraisal (severity and vulnerability) and a coping appraisal
(response efficacy and self-efficacy). Protection motivation rises only when both are
favorable, which explains why risk-disclosure messaging that frightens without
supplying an efficacious response can backfire, a caution directly relevant to the
fair-balance regime constraining DTCA.

**The transtheoretical (stages-of-change) model** describes behavior change as
movement through ordered stages (precontemplation, contemplation, preparation,
action, and maintenance) rather than a single switch. It implies that interventions
must be *stage-matched*: a message effective for someone contemplating treatment is
wasted on someone not yet aware of the problem, a segmentation logic that mirrors the
heterogeneous-response findings for detailing (@sec-health, "Physician Detailing")
and conflicting information.

**Behavioral-economics nudges in health** complete the set. Because prescribers and
patients are boundedly rational, choice architecture (defaults, framing, and salience
of the easy option) shifts health behavior without changing the underlying clinical
incentives. The same reference-dependence and present-bias mechanisms developed for
nudges and pricing recur here: framing a screening as a loss avoided rather than a
gain forgone, or defaulting a formulary toward the preferred therapy, moves behavior
at the margin where the health belief model locates its barriers. Framing interacts
with the regulatory-focus orientation of the recipient: @keller2006regulatorfocus
shows that a gain-framed message persuades promotion-focused readers while the
loss-framed version of the same message persuades prevention-focused ones, so a
single "optimal" frame is a category error.

Two further findings should temper any confidence that health marketing moves
behavior in the intended direction. First, marketing a remedy can undercut the
behavior the remedy is meant to complement: @bolton2008doesdrug find that exposure to
drug and supplement advertising reduces intentions to pursue diet and exercise, an
externality the fair-balance regime does not contemplate. Second, the food side of
the health market is governed as much by perception as by nutrition. Consumers judge
package size by a single salient dimension rather than by volume, so a change that
supersizes in one dimension and downsizes in three reads as larger while delivering
less [@chandon2009supersizedimension], and when value and health conflict on the same
package, value usually wins for the consumers whose health outcomes are worst
[@haws2013whenvalue].

## Physician Detailing

**Detailing** is personal selling to physicians: a sales representative ("detail
rep") visits a prescriber, presents clinical information about a drug, leaves
samples, and builds a relationship. It is the largest line in most pharmaceutical
promotional budgets. The marketing-science question is deceptively simple—*what is
the incremental prescription return to a detailing call?*—and the answer is hard for
one reason: **calls are not randomly assigned**. Firms send representatives to
high-volume, high-potential physicians, so a naive regression of prescriptions on
detailing recovers the firm's *targeting rule* compounded with any true response.

### A Response Model and the Targeting Problem

Let $y_{it}$ denote the number of new prescriptions physician $i$ writes for the
focal drug in period $t$, and let $d_{it}$ denote detailing calls received. A
natural count model writes prescriptions as Poisson with a log-mean that responds to
a stock of detailing goodwill,

$$
y_{it} \sim \text{Poisson}(\lambda_{it}), \qquad
\log \lambda_{it} = \alpha_i + \beta\, g(D_{it}) + \mathbf{x}_{it}^{\top}\boldsymbol{\gamma},
$$ {#eq-detail-poisson}

where $\alpha_i$ is a physician-specific intercept (baseline prescribing
propensity), $\mathbf{x}_{it}$ collects controls, and $g(D_{it})$ is a
**concave**, saturating transform of the detailing stock

$$
D_{it} = (1-\delta)\,D_{i,t-1} + d_{it},
$$ {#eq-detail-stock}

with carryover $1-\delta$. Concavity—e.g., $g(D)=\log(1+D)$ or a power
$D^{\kappa}$ with $\kappa<1$—encodes the well-documented diminishing return to
repeated calls and is what lets the model speak to the *optimal* call frequency
rather than merely its sign [@manchanda1999]. The object of interest is $\beta>0$,
the responsiveness of prescribing to accumulated detailing.

The identification threat is **endogenous targeting**: $d_{it}$ is chosen by the
firm as a function of expected prescribing, so $\mathbb{E}[d_{it}\,\varepsilon_{it}]
\neq 0$ even conditional on $\mathbf{x}_{it}$, and ordinary estimation conflates
response with selection. There are three lines of defense, in rough order of
strength.

1. **Rich physician fixed effects.** The intercept $\alpha_i$ absorbs *time-invariant*
   potential, so $\beta$ is identified from within-physician variation in calls.
   This removes the cross-sectional component of targeting (reps go to big
   prescribers) but not the dynamic component (reps go to physicians who are
   *trending up*).
2. **A joint model of the firm's allocation.** @manchanda1999 estimate the response
   model *together with* the firm's detailing-allocation policy, treating calls as
   the outcome of an optimization that depends on the physician's responsiveness.
   Modeling the selection mechanism explicitly is what purges the bias; a notable
   finding from this literature is that firms do *not* allocate calls optimally,
   leaving return on the table.
3. **Latent-trait segmentation.** @Manchanda_2004 let responsiveness itself be
   heterogeneous and partly unobserved, recovering segments of physicians who differ
   in their detailing elasticity. The managerial payoff is a targeting rule keyed to
   *responsiveness* rather than to *volume*—two physicians with identical baseline
   prescribing can have very different marginal returns to a call.

A complementary mechanism operates through **samples and learning**. Free samples
are simultaneously a price instrument (they lower the patient's first-dose cost) and
an information instrument (they let the physician observe the drug's effect on her
own patients). @dong2009 model prescribing as Bayesian learning in which detailing
supplies a *signal* and samples generate *experience*, so the two instruments have
different decay and different long-run effects: persuasion fades while learned
quality persists. Distinguishing an *informative* from a *persuasive*
interpretation of detailing matters for welfare—informative detailing can be
socially valuable even when it raises a single firm's sales.

### Heterogeneity, Saturation, and the Managerial Question

Two regularities are robust enough to plan around. First, **response is concave and
saturates**: beyond a physician-specific threshold, additional calls add little, so
the optimal policy spreads a fixed sales force across more physicians at lower
frequency rather than pounding a few [@manchanda1999]. Second, **elasticities are
heterogeneous**, and the heterogeneity is not well predicted by prescribing volume
alone [@Manchanda_2004]. The practical implication is that the targeting problem and
the response problem are the *same* problem: a firm cannot set an efficient call
plan without an unbiased estimate of $\beta$ for each segment, and it cannot obtain
that estimate without modeling how it currently targets.

## Direct-to-Consumer Advertising

**Direct-to-consumer advertising (DTCA)** addresses the patient rather than the
prescriber. The patient, prompted by a television or print advertisement, raises the
drug with a physician, who may or may not prescribe it. Because product-claim DTCA is
legal in only a handful of countries, it is both economically large in the United
States and a natural laboratory for studying advertising in a market where the buyer
cannot transact directly.

The first-order policy question is the same one advertising raises everywhere
(@sec-advertising), sharpened by the agency structure: does DTCA **expand the
category** (bring undiagnosed or untreated patients into therapy) or merely **steal
business** (reallocate share among existing patients)? The welfare reading depends
entirely on the answer. Category expansion that pulls undertreated patients into
care can be health-improving; pure business stealing is a within-category transfer
that consumes real resources. The empirical literature finds DTCA's *primary* effect
is typically on **category/primary demand**—the decision to seek treatment at
all—rather than on **brand choice**, which detailing influences more
[@narayanan2009; @chintagunta2010effects]. DTCA and detailing are therefore not
redundant: they act at different stages of the prescription funnel, and their
returns interact.

### Decomposing the Two Effects

A transparent way to separate the two effects is a nested-demand decomposition.
Let category sales in market $m$, period $t$ be $Q_{mt}$ and let brand $j$'s share be
$s_{jmt}$, so brand sales are $q_{jmt}=s_{jmt}\,Q_{mt}$. In logs,

$$
\log q_{jmt} = \underbrace{\log Q_{mt}}_{\text{category (primary) demand}}
             + \underbrace{\log s_{jmt}}_{\text{brand (selective) demand}} .
$$ {#eq-dtca-decomp}

Regressing $\log Q_{mt}$ on the *category's* total DTCA recovers the
**expansion** effect; regressing $\log s_{jmt}$ on brand $j$'s *relative* DTCA and
detailing recovers the **business-stealing** effect. The empirical pattern is that
DTCA loads on the first equation and detailing on the second [@narayanan2009].
@narayanan2015 push this further by recognizing that the marketing instruments are
also *informative*: early in a drug's life, advertising and detailing reduce
consumer and physician uncertainty, so their measured effect changes over the life
cycle—an *informative* role that dominates at launch and a *persuasive/reminder*
role later. Treating marketing as carrying information, not just shifting tastes, is
the through-line connecting the detailing, DTCA, and diffusion sections.

### What Breaks Identification

Three threats recur, and a credible DTCA study must address each.

- **Advertising endogeneity (timing).** Firms raise DTCA when they expect demand to
  rise (a new indication, a seasonal peak), so $\mathbb{E}[\text{adv}\cdot
  \varepsilon]\neq 0$ and OLS overstates the effect. Defenses include border/advertising
  discontinuities (television markets that spill across regions with different
  prescribing), instruments such as the *cost* of advertising or competitors'
  spending, and within-market dynamic controls.
- **Simultaneity of the two instruments.** DTCA and detailing are set jointly and
  respond to the same demand shocks; estimating either in isolation attributes the
  common component to whichever is included. The fix is to model them jointly, as in
  @eq-dtca-decomp with both instruments entering the share equation
  [@narayanan2009].
- **Aggregation.** Market-level data cannot, by construction, distinguish "more
  patients each buying once" from "the same patients buying more," which is precisely
  the expansion-versus-stealing question. Where physician- or patient-level panels
  exist, the decomposition is far more credible.

A useful comparison of the two push/pull instruments appears in @tbl-push-pull.

| Dimension | Detailing (push) | DTCA (pull) |
|---|---|---|
| Target | Physician (expert agent) | Patient (principal) |
| Primary margin | Brand choice / selective demand | Category / primary demand |
| Mechanism | Information + persuasion + samples | Awareness + request to prescriber |
| Decay | Fast persuasion, slower if it teaches | Reminder-like, fast decay |
| Key identification threat | Endogenous *targeting* of calls | Endogenous *timing* of spend |
| Regulatory exposure | Moderate (fair balance, gifts) | High (broadcast risk disclosure) |

: Push versus pull instruments in prescription markets. The two act at different
stages of the funnel, which is why their estimated effects load on different
equations of @eq-dtca-decomp. {#tbl-push-pull}

## Diffusion of Medical Innovations

A new drug, device, or procedure does not reach all prescribers at once; it
*diffuses*. The study of how innovations spread is one of marketing's foundational
contributions, and medicine is its archetypal setting—indeed the canonical early
diffusion study tracked the adoption of the antibiotic tetracycline among physicians,
seeding decades of work on social contagion in adoption. The tetracycline study of @coleman1957diffusioninnovatio is the origin point: physicians who were socially integrated with colleagues adopted earlier than isolated ones, which is the first credible demonstration that adoption travels through ties rather than through advertising alone.
The general diffusion machinery is developed in @sec-diffusion; here we specialize it
to the professional adoption setting and emphasize what is distinctive about medicine:
adopters are experts embedded in dense referral and collegial networks, so
*word-of-mouth among prescribers* is unusually consequential.

### The Bass Model

The **Bass model** [@Bass_1969] is the standard parametric description of
first-time adoption. Let $F(t)$ be the cumulative fraction of the eventual market
that has adopted by time $t$ and $f(t)=F'(t)$ the adoption rate. Bass posits that the
*hazard* of adoption among those who have not yet adopted is linear in the fraction
who already have,

$$
\frac{f(t)}{1-F(t)} = p + q\,F(t),
$$ {#eq-19-bass-hazard}

where $p$ is the **coefficient of innovation** (the propensity to adopt independently
of others—external influence such as advertising and detailing) and $q$ is the
**coefficient of imitation** (the propensity to adopt because peers have—internal,
word-of-mouth influence). For a market of eventual size $M$, instantaneous adoptions
$n(t)=M f(t)$ solve

$$
n(t) = M\left[ p + \frac{q}{M}\,N(t) \right]\bigl(M - N(t)\bigr),
$$ {#eq-19-bass-sales}

with $N(t)$ cumulative adoptions. Equation @eq-19-bass-hazard makes the medical
interpretation transparent: $p$ captures the firm's *push and pull* marketing (the
external channels of the previous two sections), while $q$ captures *contagion among
prescribers*. A high $q/p$ ratio signals a market where collegial influence
dominates marketing—exactly the regime in which seeding opinion leaders pays off.

**Estimation and what breaks it.** The discrete-time analogue of @eq-19-bass-sales is
the classic @Bass_1969 regression of period adoptions on cumulative adoptions and its
square,

$$
n_t = \beta_0 + \beta_1 N_{t-1} + \beta_2 N_{t-1}^2 + \varepsilon_t,
$$ {#eq-19-bass-ols}

from which $(p,q,M)$ are recovered as nonlinear functions of $(\beta_0,\beta_1,
\beta_2)$. OLS on @eq-19-bass-ols is convenient but flawed: the regressors are
*generated* from the dependent variable (so they correlate with $\varepsilon_t$ by
construction), the estimates of $M$ are unstable until the peak is observed, and the
additive error contradicts the multiplicative structure of @eq-19-bass-sales.
Nonlinear least squares on @eq-19-bass-sales, or maximum likelihood, is preferred, and
the parameters are poorly identified from data that do not yet span the inflection
point [@Bass_1969; @chandrasekaran2008]. The broader empirical-generalizations
literature documents systematic patterns in $p$, $q$, and the time-to-takeoff across
categories and countries [@chandrasekaran2007; @sultan2009], and warns that takeoff
itself is a distinct, hard-to-predict event [@golder1998]. @fig-diffusion separates
the external, marketing-led inflow from the internal, contagion-led inflow that
together generate the S-shaped curve.

```{mermaid}
%%| label: fig-diffusion
%%| fig-cap: "Adopter inflows in the Bass model. The innovation coefficient p drives an external, marketing-led inflow that is largest early; the imitation coefficient q drives an internal, contagion-led inflow that peaks once enough peers have adopted. Their sum produces the familiar S-shaped cumulative curve."
flowchart LR
  M[Not-yet-adopted<br/>M - N&#40;t&#41;] -->|p : external<br/>marketing/detailing| A[Adopters N&#40;t&#41;]
  M -->|q/M &middot; N&#40;t&#41; : internal<br/>peer contagion| A
  A -.->|word of mouth<br/>raises hazard| M
```

### Social Contagion Versus Homophily

The deepest identification problem in medical diffusion is distinguishing genuine
**social contagion** (physician $i$ adopts *because* connected physician $j$ did)
from **homophily** and **common shocks** (connected physicians adopt at similar
times because they are similar—same specialty, same hospital, same patients—or
respond to the same detailing). Naively, the prescribing of one's network neighbors
predicts one's own prescribing; but neighbors are chosen, not assigned. The marketing
literature has confronted this directly. @Van_den_Bulte_2001 re-examine a celebrated
medical-diffusion finding and show that what looked like contagion is substantially
confounded by **marketing effort** that was correlated with network position—when
detailing is controlled, the apparent peer effect shrinks. @iyengar2003 use
sociometric data on physician advice and discussion ties to identify contagion in new-drug
adoption *net* of marketing, finding that opinion leaders adopt earlier and that
contagion flows along professional ties; later work distinguishes self-reported from
sociometric leadership and shows leaders are both more *innovative* and more
*in-the-loop* with detailing [@vandenbulte2007]. The methodological lesson
generalizes: any contagion estimate is credible only to the extent it controls for
the firm's effort and for the selection that makes neighbors similar.

A simple linear-in-means specification makes the threat explicit. Let $a_{it}=1$ when
physician $i$ has adopted by $t$, and let $\bar a_{-i,t-1}$ be the lagged adoption
share of $i$'s network neighbors:

$$
a_{it} = \phi\,\bar a_{-i,t-1} + \mathbf{x}_{it}^{\top}\boldsymbol{\gamma}
        + \theta\,\text{detailing}_{it} + \alpha_i + \varepsilon_{it}.
$$ {#eq-contagion}

The contagion coefficient $\phi$ is identified only if (i) the omitted detailing term
$\theta\,\text{detailing}_{it}$ is included and uncorrelated with the network after
controls, and (ii) the unobserved similarity that drives both link formation and
adoption is captured by $\alpha_i$ and $\mathbf{x}_{it}$. Failing either, $\phi$
absorbs homophily and is biased upward—the classic reflection-and-selection problem.
This is why the credible designs above either control detailing explicitly
[@Van_den_Bulte_2001] or exploit directed, measured ties [@iyengar2003].

## Conflicting Information

The agency relationship makes the prescriber a *decision-maker under uncertainty*,
and that uncertainty is rarely resolved cleanly. A drug launches on the strength of
its trials; later studies, meta-analyses, and post-marketing surveillance arrive
with mixed or contradictory verdicts; marketing communications and news coverage
selectively amplify favorable or unfavorable signals. How a prescriber updates beliefs
when the evidence itself disagrees is the subject of this final section, and the
running example is the diabetes drug **rosiglitazone (Avandia)**, whose
cardiovascular safety became the subject of openly conflicting reports
[@kalra2011understanding].

### Bayesian Updating with Biased Signals

Cast the physician as a Bayesian learner about an unknown attribute $\mu$—say the
true cardiovascular risk of a drug. Before new evidence she holds a Normal prior,
$\mu \sim \mathcal{N}(m_0, \sigma_0^2)$, with precision (inverse variance)
$\tau_0 = 1/\sigma_0^2$. A new study delivers a signal $z$ with precision
$\tau_z = 1/\sigma_z^2$. The posterior mean is the precision-weighted average

$$
m_1 = \frac{\tau_0\,m_0 + \tau_z\,z}{\tau_0 + \tau_z}, \qquad
\tau_1 = \tau_0 + \tau_z,
$$ {#eq-bayes-update}

so the *weight* a physician places on the new study is $\tau_z/(\tau_0+\tau_z)$. Three
features of @eq-bayes-update organize the empirical findings.

First, **a more confident prior moves less**. A physician with strong prior beliefs
(large $\tau_0$)—often the high-volume specialist who has prescribed the drug for
years—down-weights a discordant new study, while a less committed prescriber updates
more. Experience, expertise, and self-efficacy therefore predict *resistance* to new,
conflicting information [@kalra2011understanding].

Second, **conflicting signals net out**. When two studies of comparable precision
point opposite ways, their weighted contribution to $m_1$ partly cancels and the
posterior is dragged toward the prior. Mixed evidence thus produces *less* belief
revision than a single clear signal of the same total precision—which is exactly why
contradictory reports can be more favorable to an incumbent drug than a single bad
study would be.

Third, **signals are not neutral**: news releases and marketing communications are
*selected* and may misrepresent precision, so a sophisticated learner should
discount $\tau_z$ for the source's bias. Modeling this discount is what separates
naive from rational updating in the data.

### Heterogeneous Response, Sources, and Spillover

The Avandia episode yields three empirical regularities that map onto the model and
that any analysis of conflicting information should anticipate
[@kalra2011understanding]. **Physician segments update differently**: specialists
versus primary-care physicians, and high- versus low-volume prescribers, revise their
prescribing along distinct trajectories, consistent with heterogeneous priors
$\tau_0$ and heterogeneous trust in sources. **Information sources differ by segment**:
groups rely on different channels (peer-reviewed studies, detailing, news), so the
*same* event reaches them with different effective precision. And there is **negative
spillover across the category**: safety concerns about Avandia depressed perceptions of
its close competitor Actos as well—evidence that prescribers update about a *class* or
*category*, not only a brand, and that bad news contaminates comparable products much
as it does comparable brands in consumer markets (@sec-branding).

These dynamics connect back to the chapter's spine. Conflicting information is a
problem of *signal precision under selection*: detailing and DTCA are themselves
biased signals whose informative content @narayanan2015 emphasize; diffusion through
peers is partly the transmission of these signals along the network of
@iyengar2003; and the welfare verdict on any marketing instrument turns on whether it
raises or lowers the precision of the beliefs the prescribing agent acts on.

## Diagnostic Technology as an Upstream Marketing Variable {#sec-health-diagnostic-bias}

The chapter so far has treated the physician as an agent who receives persuasive
information—detailing, advertising, peer signals—and converts it into a
prescription. There is a prior input that the marketing literature rarely models:
the **measuring instrument** whose reading the physician treats as fact. When that
instrument is biased, and biased differentially across patient groups, every
downstream decision inherits the error, and no amount of unbiased persuasion
corrects it.

Pulse oximetry is the clean case. The device infers blood oxygen saturation from
light absorption, and it systematically overestimates saturation in patients with
darker skin pigmentation [@sjoding2020racial]. @zhang2026pulse follow the
consequences through the operational chain in an ICU cohort: under
peripheral-oximetry monitoring, Black patients are 5.2 percentage points less
likely than White patients to receive supplemental oxygen; oxygen delivery lowers
the probability of unplanned ICU readmission by 4.8 percentage points, a 37%
relative reduction; and readmissions carry a large downstream capacity penalty in
remaining length of stay. When an accurate arterial reading is available, the
therapy disparity attenuates to statistically indistinguishable from zero.

Three lessons transfer directly to marketing settings. First, a **measurement
defect upstream of a decision is a marketing problem**, not only an engineering
one, because it determines what the decision maker believes about the customer or
patient in front of them—the same structure as a biased attribution model steering
budget away from the channels it mismeasures. Second, **disparities can be produced
by instruments rather than by intentions**, which changes the remedy from
persuasion to process: confirmatory measurement, not retraining. Third, the design
that establishes the pathway is worth studying on its own terms; it is read as a
methods exemplar in @sec-causal-mediation-oximetry, where the availability of the
accurate reading functions as a moderator that switches the proposed mechanism off.

## Information Latency as an Upstream Decision Variable {#sec-health-velocity}

The previous section made measurement *bias* upstream of a decision a marketing
problem. There is a second variable of the same kind, and it is even easier to
overlook because it changes nothing about what the instrument says: **latency**—how
long it takes the decision maker to see what the instrument already knows.

@espinosa2026velocity isolate it. Their setting is a hospital emergency department
that introduced a technology which visually displayed *when* laboratory results
became available. No new test, no new variable, no change in the content or the
precision of any signal; only the delay between a result existing and a physician
seeing it. The design covers 64,152 decisions by 387 physicians, which is enough to
estimate heterogeneity rather than only an average. Faster access produced, at once,
fewer and more targeted tests—a 25% reduction in charges—a 13% reduction in patient
length of stay, roughly 75 minutes, lower hospitalization rates, and higher patient
satisfaction. The gains were largest under high workload and in less common cases.

### Why Speed Changes the *Composition* of Information, Not Only Its Timing

The interesting result is not that decisions got faster. It is that they got faster
while using *less* information and producing *better* outcomes, which the naive
reading of the value of information forbids. @eq-bayes-update explains why, once one
notices what it leaves out. That equation prices a signal by its precision $\tau_z$
and is silent about *when* the signal arrives, because it treats the physician's
information set as chosen in one shot. Under high latency that is literally true: if
results take hours to surface, the physician must commit to a *batch* of tests before
any of them reports, and the optimal batch is deliberately redundant—it buys
precision against several hypotheses at once because there will be no chance to
condition on the first result before choosing the second.

Cut the latency and the same problem becomes *sequential*. The physician sees signal
one, updates by @eq-bayes-update, and buys signal two only if the posterior is still
too diffuse to act on. Sequential acquisition weakly dominates batch acquisition: it
never buys a signal the earlier one made unnecessary. The observable consequence is
exactly what the paper reports—fewer tests, better targeted, with an earlier
resolution of uncertainty and no loss of decision quality. Data velocity is therefore
not a scaled-up version of data volume; it changes the *structure* of the information
acquisition problem.

The same lens explains the heterogeneity. In less common cases the prior is diffuse,
$\tau_0$ is small, and @eq-bayes-update gives the early signal a large weight
$\tau_z/(\tau_0+\tau_z)$, so seeing it early has more to move. Under high workload
attention is the binding constraint, and the value of releasing it early—closing one
case to start another—rises with the queue.

### What Transfers to Marketing

Three lessons carry over, and the third is a research-design lesson rather than a
managerial one.

**Latency is a decision variable, not an infrastructure constraint.** The analytics
default in marketing is to acquire more data: more touchpoints, more granularity,
more history. This study prices the other margin. An investment that shortens
time-to-insight without adding a single variable improved efficiency and quality
simultaneously, which makes the latency of a dashboard, an attribution pipeline, or a
brand-tracking study a first-class object of study rather than an IT detail
(@sec-metrics).

**Speed and quality need not trade off.** Frontline service models usually treat
throughput and quality as points on a frontier (@sec-frontline). A latency reduction
does not move the firm along that frontier; it shifts the frontier, because it
changes how much information a given unit of attention can condition on. That is a
different managerial claim from "work faster," and it has a different empirical
signature.

**The design is the contribution.** Marketing's analytics literature almost always
confounds *more* data with *faster* data, because systems that deliver one usually
deliver the other. The identifying move here is a pure timing intervention with the
information content held fixed. Any evaluation of a real-time personalization engine,
a live-bidding system, or an executive dashboard should be held to that standard:
what varied, the information or only its arrival time? Absent that separation, an
estimated "analytics effect" is a composite of the two, and the composite gives no
guidance about which investment to make.

The caveats are the usual ones for a single-site operational study. The environment
is protocolized, decisions are high-frequency, and the signal—a laboratory result—has
an unusually well-defined arrival time. The marketing analogue is rarely so clean, and
identifying what plays the role of "the test result" is the first question a transfer
of this design must answer.

## Peer Support Communities and Emotional Spillover {#sec-health-communities}

The chapter so far has treated the prescriber as the decision-maker and the firm as
the sender. A large share of health information now moves through neither: online
communities in which patients managing chronic illness, mental health conditions, and
stigmatized diagnoses support one another. The marketing-relevant object there is not
persuasion but **participation**, because a peer support community only supplies
support while its members keep showing up.

@zhou2026emotion study what sustains that. Analyzing 65,740 reply interactions in an
online HIV community, they trace emotional exchange as a dynamic process rather than
a property of individual messages. Users who *receive* more positive sentiment are
more likely to remain active in writing replies and to express positive sentiment
themselves later; receiving negative sentiment is associated with lower participation
and less positive expression. The reverse direction holds too—a user's own emotional
expression shapes how others respond to them, with positive expressions drawing more
replies and more positive ones, and negative expressions drawing fewer and less
positive ones. The effects propagate across users, threads, topics, and local social
clusters, and are strongest for feeling-oriented rather than information-oriented
messages.

The structural point is that these two directions compose into a feedback loop, which
is why the authors describe the dynamics as self-reinforcing and why the managerial
recommendation is stated at the community level. A moderator watching individual
conversations sees only messages; the quantity that determines whether the community
survives is the loop gain, and a negative episode matters less for the offense it
gives than for the withdrawal it triggers in the recipient, which removes their
future positive replies from everyone else's inbox.

Two limits are worth stating plainly. The design is observational, and the
same-direction correlation between receiving and later expressing positive sentiment
is consistent with homophily in sociability as well as with contagion; the
cross-cluster propagation evidence constrains that story but does not close it. And
a single-community study of a stigmatized condition is a hard case to generalize
from—the stakes of participation there are unlike those in a product community,
which is precisely why the effects are large enough to detect.

## Welfare, Policy, and Identification Recap

The sector's distinctive welfare question is whether marketing makes the agency
relationship *better* or *worse* informed. Informative detailing and launch-phase
DTCA can raise the precision of prescriber and patient beliefs and pull undertreated
patients into care, a social good [@narayanan2015; @chintagunta2010effects];
persuasive, reminder-style promotion and selectively amplified evidence can degrade
it. Because the same instrument can do either, the empirical decomposition is not an
academic nicety—it is the policy lever. The recurring identification discipline of the
chapter is the practical expression of this: estimate detailing response only while
modeling targeting, estimate DTCA effects only while separating timing and
aggregation, estimate contagion only while controlling marketing and homophily, and
model belief revision only while discounting biased signals.

## Worked Example: Detailing Response and Diffusion

The following reproducible examples implement two of the chapter's core models on
simulated data. The first fits the concave detailing-response model of
@eq-detail-poisson; the second fits the Bass diffusion model of @eq-19-bass-sales by
nonlinear least squares and plots the adoption curve. Both are seeded.

```{r setup-health}
#| message: false
#| warning: false
set.seed(20240617)
```

```{r detailing-response}
#| message: false
#| warning: false
#| fig-cap: "Estimated detailing-response model. Prescriptions follow a Poisson count model whose log-mean is concave in the accumulated detailing stock, with physician random intercepts."
# --- Simulate a physician panel with concave detailing response ----------------
n_md   <- 200L          # physicians
n_t    <- 24L           # months
delta  <- 0.30          # 1 - carryover
beta   <- 0.45          # true detailing-stock elasticity (on log(1 + D))
md     <- rep(1:n_md, each = n_t)
alpha  <- rnorm(n_md, mean = 0.2, sd = 0.6)[md]   # physician baseline propensity

# Detailing is TARGETED at high-baseline physicians -> endogeneity if ignored
call_rate <- exp(-0.5 + 0.8 * alpha)
d <- rpois(n_md * n_t, lambda = pmin(call_rate, 4))

# Build the detailing stock D_it within each physician
D <- numeric(length(d))
for (i in seq_along(d)) {
  D[i] <- if (i %% n_t == 1L) d[i] else (1 - delta) * D[i - 1L] + d[i]
}

lambda <- exp(alpha + beta * log1p(D))
y      <- rpois(length(lambda), lambda)
dat    <- data.frame(md = factor(md), y = y, logD = log1p(D), d = d)

# Naive Poisson (ignores targeting) vs. fixed-effects Poisson (absorbs baseline)
naive <- glm(y ~ logD, family = poisson, data = dat)
fe    <- glm(y ~ logD + md, family = poisson, data = dat)

cat("True beta:            ", beta, "\n")
cat("Naive estimate:       ", round(coef(naive)["logD"], 3), "\n")
cat("Fixed-effects estimate:", round(coef(fe)["logD"], 3), "\n")
```

The naive estimate is biased because detailing is targeted at high-baseline
physicians (the simulated `call_rate` rises with `alpha`); the physician
fixed effects absorb that time-invariant targeting and recover an estimate much
closer to the true $\beta$. This is the within-physician identification of
@eq-detail-poisson in miniature—and a reminder that it still does not address
*dynamic* targeting, which would require the joint allocation model of
@manchanda1999.

```{r bass-diffusion}
#| message: false
#| warning: false
#| fig-cap: "Bass-model fit to simulated new-drug adoptions. Points are observed monthly adoptions; the line is the nonlinear least-squares fit recovering the innovation coefficient p and imitation coefficient q."
# --- Simulate adoptions from a Bass process ------------------------------------
M_true <- 10000      # eventual adopters
p_true <- 0.015      # innovation (external: marketing/detailing)
q_true <- 0.42       # imitation  (internal: peer contagion)
Tmax   <- 36L

N <- numeric(Tmax); n <- numeric(Tmax); Nprev <- 0
for (t in 1:Tmax) {
  rate   <- (p_true + (q_true / M_true) * Nprev) * (M_true - Nprev)
  n[t]   <- max(rpois(1, lambda = rate), 0)
  Nprev  <- Nprev + n[t]
  N[t]   <- Nprev
}
period <- 1:Tmax
Ncum_lag <- c(0, head(N, -1))

# --- Fit the Bass model by nonlinear least squares -----------------------------
bass_fit <- nls(
  n ~ (p + (q / M) * Ncum_lag) * (M - Ncum_lag),
  start = list(p = 0.01, q = 0.3, M = 12000),
  control = nls.control(maxiter = 500, warnOnly = TRUE)
)
est <- coef(bass_fit)
cat(sprintf("Estimated p = %.4f (true %.3f)\n", est["p"], p_true))
cat(sprintf("Estimated q = %.4f (true %.3f)\n", est["q"], q_true))
cat(sprintf("Estimated M = %.0f (true %d)\n",   est["M"], M_true))

plot(period, n, pch = 19, col = "grey40",
     xlab = "Month", ylab = "New adoptions",
     main = "Bass diffusion of a new drug")
lines(period, predict(bass_fit), col = "firebrick", lwd = 2)
legend("topright", legend = c("Observed", "Bass fit"),
       pch = c(19, NA), lty = c(NA, 1), col = c("grey40", "firebrick"), bty = "n")
```

The recovered $q/p$ ratio is large, the signature of a contagion-driven market in
which seeding influential prescribers and supporting peer word-of-mouth—rather than
brute-force detailing—offers the higher return. Whether that recovered $q$ is *real*
contagion or disguised homophily is, as @sec-health notes, the identification
question that @eq-contagion forces the analyst to confront before acting on it.

## Key Takeaways

- Prescription marketing acts on a **double-agency** chain (patient–physician–payer)
  under a regulatory ceiling on claims; the buyer's price sensitivity is muted, so
  marketing is fundamentally an **information** problem.
- **Detailing** response (@eq-detail-poisson) is concave and heterogeneous, but
  estimates are confounded by **endogenous targeting**; credible estimates control
  rich physician effects or, better, model the firm's allocation jointly
  [@manchanda1999; @Manchanda_2004].
- **DTCA** acts mainly on **category (primary) demand** while detailing acts on
  **brand (selective) demand** (@eq-dtca-decomp); separating expansion from business
  stealing is both the welfare question and the identification challenge
  [@narayanan2009; @narayanan2015].
- The **Bass model** (@eq-19-bass-hazard) decomposes adoption into external
  (marketing-led, $p$) and internal (contagion-led, $q$) channels; in medicine the
  central threat is **contagion versus homophily**, resolved only by controlling
  marketing effort and measured ties [@Van_den_Bulte_2001; @iyengar2003].
- Physicians facing **conflicting information** update as **Bayesian learners with
  biased signals** (@eq-bayes-update): confident priors resist revision, mixed
  evidence nets out, and bad news spills across the therapeutic class
  [@kalra2011understanding].
- A **biased instrument upstream of a decision** propagates through every downstream
  choice: differential pulse-oximetry error is associated with less supplemental
  oxygen for Black ICU patients, higher unplanned readmission, and a large
  length-of-stay penalty, and the therapy gap closes when an accurate arterial
  reading is available [@sjoding2020racial; @zhang2026pulse]. The remedy is
  confirmatory measurement, a process change, rather than persuasion
  (@sec-health-diagnostic-bias).
- **Latency is the second upstream variable.** Displaying *when* lab results became
  available—no new information, only faster access—cut test charges 25%, shortened
  length of stay 13% (about 75 minutes), lowered hospitalization, and raised patient
  satisfaction, with the largest gains under high workload and in uncommon cases
  [@espinosa2026velocity]. Speed changed the **composition** of information gathered,
  not just its timing: low latency converts a batch acquisition problem into a
  sequential one, so fewer and better-targeted signals suffice
  (@sec-health-velocity).

## Further Reading

The diffusion machinery specialized here is developed in full in @sec-diffusion; the
general theory of advertising's category-versus-share effects in @sec-advertising;
and the contamination of comparable products by negative information in
@sec-branding. For the foundations of marketing as Bayesian learning under
uncertainty, see the consumer-learning models of @erdem1996 and @erdem1998 and the
quality-signaling logic of @kalra2008signaling. On the speed rather than the volume
of information, @espinosa2026velocity is the cleanest available design—a timing
intervention with information content held fixed—and is worth reading alongside the
service-operations material in @sec-service and the dashboard discussion in
@sec-metrics.
