# Games and Gamification {#sec-gaming}

A *game* is a structured system of voluntary activity governed by rules that
produce quantifiable outcomes and, crucially, *engagement*: players keep playing
because the system makes playing intrinsically rewarding. Video games have grown
into the largest single category of paid entertainment, larger than film and
recorded music combined, and the economics of the industry rest less on selling a
finished good than on sustaining a relationship in which a player returns,
progresses, and—often—pays in small increments over months or years. That shift,
from the boxed product to the *live service*, makes gaming a marketing problem of
the first order: the central managerial decisions are about engagement design,
monetization architecture, and the regulatory limits on both.

This chapter treats games as a marketing object along three axes. The first is the
**monetization architecture**: how revenue is extracted from a population of
players whose willingness to pay is enormously heterogeneous, and why the
free-to-play model with in-game purchases dominates. The second is the
**engagement mechanics**: the rule systems—progression, scarcity, social
comparison, variable rewards—that convert attention into habit, and the construct
of *consumer engagement* that the marketing literature uses to formalize them.
The third is **regulation and welfare**: because the same mechanics that drive
engagement can drive overconsumption, gaming has become one of the most heavily
regulated consumer categories, and the evaluation of those regulations is a clean
causal-inference problem that we develop in detail.

A fourth theme cuts across all of marketing rather than belonging to games alone:
**gamification**, the transfer of game design elements—points, levels, badges,
leaderboards, streaks—into non-game settings such as loyalty programs, fitness
apps, and sales-force incentives. Gamification borrows the engagement machinery of
games to serve a marketing goal, and inherits both its power and its ethical
hazards. By the end of the chapter the reader should be able to formalize a
free-to-play monetization model, define and measure consumer engagement, specify
and identify the effect of a usage-restriction regulation, and reason about when a
gamified mechanic creates value versus when it merely manufactures compulsion.

## The Economics of In-Game Purchases

### From Premium to Free-to-Play

Historically a game was a *premium* good: the player paid a fixed price once and
received the complete experience. The dominant modern model inverts this. Under
**free-to-play** (F2P), access is free and revenue comes from **in-game
purchases**—small transactions, typically a few dollars, for virtual items,
currency, cosmetic customization, time savings, or access to additional content.
These transactions are commonly called **microtransactions**, and the business
model is sometimes labeled *freemium* when a free tier coexists with paid upgrades.

The logic is a price-discrimination argument. Let consumers be indexed by a
willingness-to-pay type $\theta$ with distribution $F(\theta)$ that is highly
right-skewed: most players value the game modestly, a few value it intensely. A
single premium price $p$ excludes everyone with $\theta < p$ and leaves all
surplus from high-$\theta$ players above $p$ on the table. Free access
($p = 0$) admits the entire population, and a menu of in-game purchases then
extracts revenue increasing in $\theta$, approximating first-degree price
discrimination through self-selection. Formally, if a player of type $\theta$
chooses purchase intensity $q$ to maximize utility $u(q;\theta) - P(q)$ under a
nonlinear price schedule $P(q)$, expected revenue per installed player is

$$
R = \int_{\underline{\theta}}^{\overline{\theta}} P\big(q^\*(\theta)\big)\, dF(\theta),
$$ {#eq-f2p-revenue}

where $q^\*(\theta)$ is the player's optimal purchase intensity. Two features of
@eq-f2p-revenue drive industry practice. First, because $F$ is right-skewed,
$R$ is dominated by its upper tail: a small fraction of players—the so-called
*whales*—account for the majority of revenue, so the schedule $P(q)$ is engineered
to remain unbounded above (there is always something more to buy) rather than to
clear a median consumer.[^whales] Second, since admission is free, the marginal
admitted player has $\theta$ near zero and contributes through scale and network
effects (see @sec-platforms) rather than direct payment; the model trades certain
small per-unit revenue for a far larger base.

[^whales]: The colloquial taxonomy of "minnows," "dolphins," and "whales" maps
onto quantiles of $F(\theta)$. The concentration is extreme: in many titles fewer
than 5% of paying players, and well under 1% of all players, generate the majority
of in-game-purchase revenue. This concentration is the source of both the model's
profitability and its regulatory scrutiny, because the heaviest spenders are
disproportionately likely to be the vulnerable consumers (@sec-regulation).

The transition from a non-paying to a paying relationship is itself a measurable
event. @cao2023free study the monetization of a non-advertising-based app and show
that the timing and design of the conversion from free to paid materially shapes
both retention and revenue: converting too aggressively sheds the low-$\theta$ base
that sustains network value, while converting too late forgoes extractable surplus
from high-$\theta$ users. The managerial object is therefore not a price but a
*conversion policy*—when, and on what, to begin charging a player who arrived for
free.

### A Taxonomy of In-Game Purchases

Not all microtransactions are alike, and the distinction matters for both welfare
and regulation. @tbl-iap-types organizes the major types by what they sell and the
externality they impose on other players.

| Type | What is sold | Effect on other players | Regulatory salience |
|---|---|---|---|
| Cosmetic | Appearance, skins, emotes | None (purely expressive) | Low |
| Convenience / time-savers | Faster progression, energy refills | Mild (pacing) | Moderate |
| Pay-to-win | Competitive advantage | Strong negative | High |
| Content / expansion | New levels, characters | None | Low |
| Loot boxes / gacha | Randomized item bundles | Varies | Very high |

: A taxonomy of in-game purchases by object sold and externality. {#tbl-iap-types}

The economically and ethically distinctive category is the **loot box** (also
called *gacha* after the Japanese capsule-toy machine): a sealed bundle whose
contents are randomized, purchased before the buyer knows what it contains. A loot
box is a lottery. If a box costs $c$ and yields item $j$ with probability $\pi_j$
and subjective value $v_j$, the player buys when expected value exceeds price,

$$
\sum_j \pi_j\, v_j \;>\; c,
$$ {#eq-lootbox}

but the structural similarity to gambling is exact: a stochastic payout purchased
for a fixed stake, with the operator setting $\{\pi_j\}$ to its advantage. The
resemblance is the reason loot boxes have attracted the heaviest regulatory
attention of any monetization mechanic, a point we return to in @sec-regulation.
The behavioral force is not the expected value in @eq-lootbox but the *variable
schedule* of reinforcement the randomization creates, which we formalize next.

## Engagement Mechanics

### Consumer Engagement as a Construct

Marketing formalizes the "stickiness" of games through the construct of
**consumer engagement**. Following the engagement literature, engagement is a
consumer's investment of operant (behavioral) and operand (cognitive, emotional)
resources in interactions with a focal object beyond the act of purchase
[@hollebeek2011; @harmeling2016]. It is explicitly a multidimensional construct
with cognitive, emotional, and behavioral facets, and it is *cumulative*: today's
engagement raises the probability of tomorrow's, which is precisely the dynamic a
game's reward system is designed to exploit. Engagement sits inside the broader
**customer experience** and **customer journey** framework, the end-to-end set of
touchpoints through which a consumer moves [@lemon2016], and it is the proximal
driver of the retention and monetization outcomes in @eq-f2p-revenue.

> Customer engagement is the intensity of an individual's participation in and
> connection with an organization's offerings and activities, reflecting investment
> of resources beyond what is required for the transaction itself.

The reason engagement is the right construct—rather than, say, satisfaction or
loyalty—is that it captures *behavioral investment that is not yet a purchase*. A
player who logs in daily, completes a streak, or climbs a leaderboard has invested
without paying; that accumulated investment is what later converts to spending and
what a player is loath to abandon (a sunk-cost dynamic). Engagement is thus the
mediating state between admission and monetization.

### The Reward-Schedule Model

The engineering core of an engagement mechanic is its **reward schedule**: the
rule mapping player actions to rewards. The foundational result, from the
psychology of reinforcement, is that *variable-ratio* schedules—rewards delivered
after an unpredictable number of actions—produce the highest and most persistent
response rates, far exceeding fixed schedules.<!-- CITE-NEEDED: Skinner / variable-ratio reinforcement schedules and response persistence -->
Loot boxes, random drops, and gacha pulls are variable-ratio schedules
instantiated as commerce.

A tractable formalization treats play as a sequence of actions, each of which may
return a reward. Let the per-action reward be a random variable $r_t$ with mean
$\mu$ and variance $\sigma^2$, and let the player continue playing as long as the
*recent* experienced reward keeps an internal motivation state $m_t$ above a
quit threshold $\bar m$. A simple updating rule is

$$
m_t = \alpha\, m_{t-1} + (1-\alpha)\, r_t, \qquad 0 < \alpha < 1,
$$ {#eq-motivation}

where $\alpha$ is the persistence of motivation. Under a *fixed* reward ($r_t = \mu$
deterministically) the state converges monotonically and the player habituates: the
reward becomes predictable and $m_t$ drifts toward a constant the designer cannot
push above $\bar m$ without raising $\mu$ (costly). Under a *variable* reward with
the same mean $\mu$ but variance $\sigma^2 > 0$, occasional large draws repeatedly
spike $m_t$ and reset the approach to the threshold, sustaining play at *no
additional expected reward cost*. The designer's lever is therefore $\sigma^2$, not
$\mu$: variance is free engagement. This is the formal statement of why randomized
rewards dominate, and why they are simultaneously the most effective and the most
hazardous mechanic. The same logic explains why **streaks** and **daily login
bonuses** work—they convert the quit decision into a recurring loss frame, so that
not playing forfeits accumulated progress.

The following simulation makes the contrast concrete, comparing session length
under fixed versus variable rewards with identical means.

```{r reward-schedule-sim}
#| label: fig-reward-schedule
#| fig-cap: "Simulated player drop-out under fixed versus variable reward schedules
#|   with identical mean reward. Variable rewards sustain motivation above the quit
#|   threshold for longer by repeatedly spiking the motivation state."
#| message: false
#| warning: false
set.seed(20)

simulate_play <- function(n = 200, mu = 1, sigma = 0, alpha = 0.8,
                          quit = 0.6, m0 = 1) {
  m <- m0
  for (t in 1:n) {
    r <- if (sigma == 0) mu else rgamma(1, shape = mu^2 / sigma^2,
                                        scale = sigma^2 / mu)
    m <- alpha * m + (1 - alpha) * r
    if (m < quit) return(t)          # player quits this session
  }
  n
}

reps <- 2000
fixed_len    <- replicate(reps, simulate_play(sigma = 0))
variable_len <- replicate(reps, simulate_play(sigma = 1.5))

cat("Median session length (fixed reward):   ", median(fixed_len), "\n")
cat("Median session length (variable reward):", median(variable_len), "\n")

hist(variable_len, breaks = 30, col = rgb(0, 0, 1, 0.4),
     xlab = "Session length (actions before quit)", main = "",
     xlim = range(c(fixed_len, variable_len)))
hist(fixed_len, breaks = 30, col = rgb(1, 0, 0, 0.4), add = TRUE)
legend("topright", legend = c("Variable reward", "Fixed reward"),
       fill = c(rgb(0, 0, 1, 0.4), rgb(1, 0, 0, 0.4)), bty = "n")
```

### The Engagement-to-Revenue Loop

Engagement mechanics, monetization, and acquisition form a closed loop, shown in
@fig-engagement-loop. Acquisition delivers players into a *core loop* of action
and reward governed by @eq-motivation; sustained engagement produces the behavioral
investment that both converts to in-game purchases (@eq-f2p-revenue) and, through
social and streaming visibility, feeds acquisition of new players. Live streaming
is an empirically important arm of this loop: @huang2025promotional, using
high-frequency Twitch data and within-day variation in top streamers' broadcast
hours for identification, estimate a positive but modest and short-lived elasticity
of concurrent players to live-stream viewership (about $0.027$), with the strongest
effects for lesser-known titles where streams resolve quality uncertainty.
Influencer content cuts both ways, however: @li2024opposing, studying YouTube
influencers in the video-game industry, document *opposing* purchase and usage
effects—exposure can raise purchase intent while substituting for the consumer's
own play time—so the net effect on the engagement loop is not signed a priori.

```{mermaid}
%%| label: fig-engagement-loop
%%| fig-cap: "The engagement-to-revenue loop in a free-to-play game. The core
%%|   action-reward loop sustains engagement, which both monetizes through in-game
%%|   purchases and feeds acquisition through social and streaming visibility."
flowchart LR
    A[Acquisition<br/>free admission] --> B[Core loop<br/>action and reward]
    B --> C{Motivation<br/>state m_t}
    C -->|above threshold| B
    C -->|sustained| D[Engagement<br/>behavioral investment]
    D --> E[In-game purchases<br/>monetization]
    D --> F[Social and<br/>streaming visibility]
    F --> A
    E --> G[Revenue]
    C -->|below threshold| H[Churn]
```

## Gamification in Marketing

**Gamification** is the application of game-design elements—points, levels,
badges, leaderboards, challenges, streaks, and randomized rewards—to contexts
whose primary purpose is not play. In marketing the canonical applications are
loyalty programs (status tiers, points, progress bars), mobile engagement apps
(streaks, daily goals), referral programs (unlockable rewards), and internal
sales-force or service incentives (leaderboards, badges). The premise is that the
engagement machinery of @eq-motivation can be ported to serve a marketing
objective—repeat purchase, app usage, word of mouth—without the underlying activity
being a game.

The mechanism that gamification exploits is the conversion of an instrumental task
into one with intrinsic and social reward structure. Three behavioral levers
recur. First, **goal gradient and endowed progress**: progress toward a salient
goal accelerates effort as the goal nears, and artificially endowing early
progress (e.g., a loyalty card pre-stamped twice) increases completion.<!-- CITE-NEEDED: goal-gradient / endowed-progress effect in loyalty programs -->
Second, **operational transparency and effort visibility**: making the effort or
progress visible increases valuation and persistence. Third, **social comparison**:
leaderboards and visible status convert a private task into a positional one. The
emotional facet matters here—the satisfaction of completing a streak or holding a
status tier is an affective reward, and the literature on how time and activity are
experienced as fun or productive bears directly on gamified design [@etkin2015;
@tonietto2020].

Gamification, however, inherits the ethical hazard of its source. The same
mechanics that increase desirable engagement can manufacture compulsion, and
several documented effects warn against naive deployment. Public, social
gamification can backfire through *slacktivist* dynamics, in which a token public
act of engagement *reduces* subsequent substantive effort—an effect of
"private versus public" token support that managers of badge- and share-based
programs should anticipate [@kristofferson2013]. And because gamified metrics are
visible and incentivized, they invite gaming-of-the-metric: participants optimize
the tracked behavior rather than the underlying objective, a generic failure of
incentive design.<!-- CITE-NEEDED: gaming-the-metric / Goodhart's law in incentive design -->
The design question is therefore whether a gamified mechanic *aligns* the tracked
behavior with genuine value (a well-designed loyalty tier rewarding real repeat
custom) or merely *manufactures* a behavior that looks like value (a streak that
compels logins of no benefit to the consumer). @tbl-gamification contrasts
value-creating with value-extracting gamification along this axis.

| Dimension | Value-creating gamification | Value-extracting gamification |
|---|---|---|
| Objective alignment | Tracked behavior $\approx$ genuine value | Tracked behavior decoupled from value |
| Consumer welfare | Consumer surplus rises | Surplus transferred to firm or destroyed |
| Reward structure | Predictable, earned | Variable-ratio, engineered for compulsion |
| Transparency | Odds and rules disclosed | Odds opaque (e.g., undisclosed drop rates) |
| Exit | Low switching/quit cost | High sunk-cost lock-in |

: Value-creating versus value-extracting gamification. The same mechanics serve
either end; the distinction is whether the tracked behavior is aligned with
consumer value and whether the design preserves the consumer's capacity to exit.
{#tbl-gamification}

## Regulation of Gaming and Welfare {#sec-regulation}

### Why Regulate

The welfare case for regulating games rests on the gap between the engagement
mechanics of @eq-motivation and consumer sovereignty. When a variable-reward
schedule sustains play and spending past the point a fully-informed, self-controlled
consumer would choose, the consumer's revealed behavior no longer reveals
preference; the standard welfare theorems fail. Three regulatory concerns follow,
in rough order of consensus. The first is **minors and spending**: in-game
purchases by children, who lack both the means and the judgment to consent to
financial commitments, made through frictionless payment. The second is **gambling
mechanics**: loot boxes (@eq-lootbox) replicate the structure of gambling, and
several jurisdictions have moved to classify them as such, with consequences for
age-gating and disclosure. The third is **time use and overconsumption**: excessive
play time, particularly among adolescents, framed as a public-health concern. These
motivate three corresponding instruments—spending caps and parental controls,
loot-box disclosure or prohibition, and **usage-restriction laws** that cap or
curfew play time. The remainder of this section develops the evaluation of the
third, which has produced the cleanest natural experiment.

### A Natural Experiment: Usage-Restriction Laws

The sharpest evidence on usage restriction comes from South Korea's *Shutdown Law*
(the "Cinderella law"), which barred players under 16 from accessing online games
between midnight and 6 a.m. @jo2020protecting evaluate this policy on
individual-level usage and spending data and reach a set of conclusions that should
temper enthusiasm for blunt restriction. Regulation reduced overall game usage, but
the effect was sharply heterogeneous in *past behavior*: it bound on average
players, had a smaller effect on heavy players, and ran in the *opposite* direction
for the very heaviest—the consumers the law was most meant to protect. And the
revenue impact was negligible: restricted players reallocated rather than reduced
spending. The policy lesson is that a curfew deters light, marginal players while
the heaviest—plausibly the at-risk population—are unmoved or even intensify, so a
blunt usage cap is poorly targeted at the welfare problem it names.

### Identifying the Effect of a Usage Restriction

The Korean setting is valuable precisely because it permits credible identification,
and @jo2020protecting combine three complementary designs. We state each as an
estimator with its identifying assumption and what breaks it, because the
combination—not any single design—is what makes the conclusion robust. The cross-cutting
machinery here connects to the broader treatment of causal inference in
@sec-causal-inference.

**Difference-in-differences (DiD).** With an age-based cutoff defining treated
(under-16) and control (just over the cutoff) players observed before and after the
law, the canonical two-way fixed-effects estimator for outcome $y_{it}$ (usage or
spending) is

$$
y_{it} = \alpha_i + \gamma_t + \tau\, D_{it} + \varepsilon_{it},
$$ {#eq-did}

where $\alpha_i$ and $\gamma_t$ are individual and time fixed effects, $D_{it}=1$
for treated players in the post-period, and $\tau$ is the average treatment effect.
Identification requires the **parallel-trends** assumption: absent the law, treated
and control outcomes would have moved in parallel,
$\mathbb{E}[y_{it}(0) - y_{i,t-1}(0)\mid D]$ is independent of treatment status.
This breaks if under-16 and over-16 players were already on diverging usage
trajectories (developmental differences in gaming intensity by age make this a live
threat), or if the law triggered compositional change—e.g., minors migrating to
unregulated platforms or borrowing adult credentials, which contaminates both the
treated outcome and the control group.

**Regression discontinuity (RD).** The age threshold furnishes a sharp
discontinuity: a player just under 16 is restricted, one just over is not. Comparing
outcomes in a narrow bandwidth $h$ around the cutoff $a_0 = 16$,

$$
\tau_{\text{RD}} = \lim_{a \downarrow a_0} \mathbb{E}[y \mid a] -
                   \lim_{a \uparrow a_0} \mathbb{E}[y \mid a],
$$ {#eq-rd}

identifies the *local* effect at the cutoff under the assumption that all other
determinants of $y$ are continuous in age at $a_0$—players cannot precisely
manipulate their recorded age, and nothing else changes discontinuously at exactly
16. Manipulation (misreported birthdates to evade the curfew) is the canonical
threat and would show as a density discontinuity in the running variable; the
estimate is also only local to age 16 and need not generalize to younger children.

**Propensity-score matching (PSM).** To compare restricted and unrestricted players
who are otherwise observationally similar, each treated player $i$ is matched to
controls with a similar propensity score $e(\mathbf{x}_i) = \Pr(D_i = 1 \mid
\mathbf{x}_i)$ estimated from pre-treatment covariates $\mathbf{x}_i$ (past usage,
spending, tenure). The matched estimator is

$$
\hat\tau_{\text{PSM}} = \frac{1}{N_1}\sum_{i:\,D_i=1}
   \Big( y_i - \textstyle\sum_{j} w_{ij}\, y_j \Big),
$$ {#eq-psm}

where $w_{ij}$ weights controls $j$ by closeness in $e(\cdot)$. PSM identifies the
effect on the treated under **selection on observables** (conditional
unconfoundedness): treatment is as-good-as-random given $\mathbf{x}$, and there is
common support. It breaks under selection on *unobservables*—if an unmeasured trait
(say, a player's underlying self-control or propensity to seek out gaming) drives
both treatment-relevant behavior and outcomes, matching on observed covariates
leaves bias. PSM cannot, by construction, fix this; it is why @jo2020protecting
triangulate with RD, whose identification does not lean on observables.

The discipline of the design is that the three estimators lean on *different*
assumptions—parallel trends, continuity at the cutoff, selection on observables—so
their agreement is more convincing than any one alone. Where they would disagree
tells the analyst which assumption is binding. @tbl-identification summarizes the
comparison.

| Design | Estimand | Key assumption | What breaks it |
|---|---|---|---|
| DiD (@eq-did) | ATT, average over time | Parallel trends | Diverging age trends; migration to unregulated play |
| RD (@eq-rd) | Local effect at age 16 | Continuity at cutoff | Age manipulation; non-generalizable locally |
| PSM (@eq-psm) | ATT on matched sample | Selection on observables | Selection on unobservables (e.g., self-control) |

: Three identification strategies for a usage-restriction law and the assumption
each rests on. Their combination is robust because the assumptions are
non-overlapping. {#tbl-identification}

### A Reproducible Difference-in-Differences

The following simulation reproduces the *structure* of the usage-restriction
evaluation—not the Korean data—to make the DiD estimator and its heterogeneity
concrete. We generate treated (minor) and control players with a treatment effect
that varies by baseline intensity, mirroring the central finding of
@jo2020protecting that restriction binds on average players but not the heaviest.

```{r did-simulation}
#| label: tbl-did-results
#| tbl-cap: "Difference-in-differences estimates of a simulated usage-restriction
#|   law, overall and by baseline-intensity tercile. The overall effect masks
#|   heterogeneity: the restriction reduces usage for light and average players but
#|   not for the heaviest, mirroring the qualitative finding of Jo et al. (2020)."
#| message: false
#| warning: false
set.seed(20)
library(dplyr)

n <- 4000
players <- tibble(
  id        = 1:n,
  treated   = rbinom(n, 1, 0.5),                 # under-16 (restricted) vs. control
  baseline  = rgamma(n, shape = 2, scale = 5)    # pre-period usage (hours/week)
) |>
  mutate(intensity = ntile(baseline, 3))         # 1 = light, 3 = heaviest

# Treatment effect: -30% for light, -15% for average, ~0 for heaviest players.
te <- c(`1` = -0.30, `2` = -0.15, `3` = 0.02)

panel <- players |>
  tidyr::expand_grid(post = c(0, 1)) |>
  mutate(
    effect = treated * post * te[as.character(intensity)],
    usage  = baseline * (1 + effect) +
             rnorm(n(), 0, 1) +                   # noise
             post * 0.5,                          # common time trend (seasonality)
    D      = treated * post
  )

# Two-way fixed-effects DiD via the interaction term.
did_overall <- lm(usage ~ treated + post + D, data = panel)
overall_tau <- coef(did_overall)["D"]

# Heterogeneity: estimate tau within each baseline-intensity tercile.
by_tercile <- panel |>
  group_by(intensity) |>
  group_modify(~ broom::tidy(lm(usage ~ treated + post + D, data = .x))) |>
  filter(term == "D") |>
  transmute(intensity, tau = estimate, se = std.error)

knitr::kable(
  bind_rows(
    by_tercile,
    tibble(intensity = NA, tau = overall_tau,
           se = summary(did_overall)$coefficients["D", "Std. Error"])
  ),
  digits = 3,
  col.names = c("Intensity tercile", "DiD estimate (hours/week)", "Std. error"),
  caption = NULL
)
```

The estimated $\hat\tau$ is sharply negative for the lightest tercile, attenuated
for the middle, and indistinguishable from zero for the heaviest—the pattern that
makes a blunt curfew a poorly targeted instrument. A regulator reading only the
pooled coefficient would conclude the law "works"; the heterogeneity reveals that it
works on exactly the players who were not the problem.

### Loot Boxes, Disclosure, and the Limits of Restriction

The loot-box debate sharpens the regulatory dilemma. Because @eq-lootbox makes a
loot box a lottery, the live policy options are (i) **prohibition**, classifying
loot boxes as gambling and banning their sale to minors (the route taken in some
jurisdictions); (ii) **disclosure**, mandating publication of the drop
probabilities $\{\pi_j\}$ so that @eq-lootbox is at least computable by the
consumer; and (iii) **spending caps and friction**, limiting the rate at which a
player can purchase. Disclosure is the lightest-touch instrument and the most
defensible on consumer-sovereignty grounds—it restores the information the variance
lever of @eq-motivation conceals—but it does not address the behavioral force,
since the appeal of randomized rewards survives full knowledge of the odds.
Prohibition addresses the mechanic directly but at the cost of the surplus
legitimate players derive from cosmetic randomization, and invites substitution
toward unregulated grey-market trading. The general lesson of @jo2020protecting—that
restrictions bind on marginal, not at-risk, consumers and induce reallocation
rather than reduction—travels to loot boxes and counsels against assuming any single
instrument is sufficient.<!-- CITE-NEEDED: empirical evaluation of loot-box disclosure mandates and gambling reclassification -->

## Key Takeaways

- The dominant **free-to-play** model is a price-discrimination scheme
  (@eq-f2p-revenue): free admission maximizes the base, and a menu of in-game
  purchases extracts surplus from a right-skewed willingness-to-pay distribution
  dominated by its upper tail. The managerial object is a *conversion policy*, not
  a price [@cao2023free].
- **Engagement** is the construct that mediates admission and monetization
  [@hollebeek2011; @harmeling2016; @lemon2016]; its engineering core is the
  **reward schedule**, and the decisive design lever is reward *variance*, not mean
  (@eq-motivation)—variance is free engagement, which is why randomized rewards are
  both the most effective and the most hazardous mechanic.
- **Loot boxes** are structurally lotteries (@eq-lootbox); their behavioral force
  is the variable-ratio schedule, not the expected value, which is why disclosure of
  odds is necessary but not sufficient.
- **Gamification** ports game mechanics into marketing and inherits both their power
  and their ethical hazard; the test is whether the tracked behavior is *aligned*
  with consumer value or merely *manufactured* (@tbl-gamification), and social
  gamification can backfire [@kristofferson2013].
- **Usage-restriction laws** are evaluable as a natural experiment, and the cleanest
  evidence triangulates DiD, RD, and PSM on non-overlapping assumptions
  (@tbl-identification). The central finding [@jo2020protecting] is that restriction
  binds on light and average players but not the heaviest, with negligible revenue
  effect—so blunt restriction is poorly targeted at the overconsumption it names
  (@sec-causal-inference).
