# Gambling, Sports Betting, and Prediction Markets {#sec-gambling}

The previous chapter treated games as products whose engagement mechanics can be
engineered, and closed on loot boxes—the point at which a reward schedule starts to
look like a wager. This chapter crosses that line deliberately. Gambling is a
consumer market with a marketing mix like any other: a price (the odds), a product
line (bet types), a promotion calendar (free bets and deposit matches), an
advertising budget, and a customer-relationship function that is unusually good at
identifying and retaining its heaviest users. It is also a market in which the
product is designed to be consumed past the point the consumer intended, which makes
it the sharpest available test of whether the field's constructs and methods can
speak to consumer welfare rather than only to firm performance.

Marketing has studied gambling for decades, but sporadically, and the literature sat
in an awkward place: interesting behavior, weak identification. That changed in 2018.
When the U.S. Supreme Court struck down the Professional and Amateur Sports
Protection Act in *Murphy v. NCAA*, it handed researchers a staggered, state-level,
two-channel policy experiment of a kind the field rarely gets, and the papers built
on it began landing in the top journals in 2026. This chapter is organized around
that experiment, works outward to the marketing machinery that operates inside it,
and ends with prediction markets—the same instrument turned to a different purpose,
and the place where a wager becomes a forecast.

**Learning objectives.** By the end of this chapter you should be able to (i)
describe the PASPA rollout as a research design and say precisely what its two
treatments identify, (ii) summarize what the three flagship studies found and why
their agreement matters more than any one of them, (iii) connect the operator's
customer-base economics to the harm literature and quantify the tension between them,
and (iv) distinguish a prediction market from a sportsbook on grounds other than
regulatory labeling.

## The PASPA Rollout as a Research Design {#sec-gb-paspa}

Before May 2018, only Nevada offered legal single-game sports betting. *Murphy*
returned the decision to the states, and within a month Delaware and New Jersey had
launched retail sportsbooks; by 2025, 39 states and the District of Columbia had
legalized some form of sports betting, and roughly $300 billion had been wagered
through the newly legal markets between 2018 and 2023 [@hollenbeck2026financial].

What makes this more than a large market is its internal structure. Three features do
the identifying work, and a study that ignores any of them will mismeasure the effect.

**Staggered timing.** States legalized at different dates for reasons largely
unrelated to the outcomes researchers care about—budget cycles, legislative
calendars, tribal-compact negotiations. That supports a difference-in-differences
design with not-yet-treated states as controls, which is why every paper in this
literature reaches for the staggered estimators of @sec-causal-week06 rather than a
two-way fixed-effects specification.

**Two distinct treatments.** Retail sportsbooks and online or mobile betting are
separate legalization events, and many states adopted them at different times.
This is the crucial design feature. "Sports betting became legal" is a confounded
treatment; *channel* is the variable that actually moves the cost of placing a bet
from a drive to a thumb-tap. A design that pools them estimates a weighted average of
two effects that turn out to be very different sizes.

**A pre-period with real gambling in it.** Roughly 75–80% of Americans reported some
gambling in a year before 2018, with about 10% gambling twice weekly or more
[@hollenbeck2026financial]. The counterfactual is not abstinence; it is lotteries,
casinos, and offshore books. Any welfare claim has to net out substitution from those
channels, which is a spillover problem of exactly the shape @sec-policy-spillovers
formalizes.

```{mermaid}
%%| label: fig-gb-design
%%| fig-cap: "The PASPA rollout as a two-treatment staggered design. Channel, not legality, is the variable that moves the cost of placing a bet, and the outcome families below have been measured on four independent data sources by four independent teams."
flowchart TB
  P["Murphy v. NCAA (2018)<br/>PASPA struck down"] --> R["Retail sportsbook<br/>legalization (staggered)"]
  P --> O["Online / mobile<br/>legalization (staggered)"]
  R --> N["Little detectable change<br/>on most outcomes"]
  O --> W["Well-being<br/>(hotline calls, suicide)"]
  O --> F["Household finance<br/>(credit score, bankruptcy)"]
  O --> S["Savings and portfolios<br/>(brokerage deposits)"]
  O --> H["Health behaviors<br/>(substance use)"]
  W --> C["Convergent finding:<br/>the channel is the treatment"]
  F --> C
  S --> C
  H --> C
```

## What Legalization Did {#sec-gb-consequences}

Four teams, working with four unrelated data sources, have now estimated the
consequences of this rollout. Their agreement is the substantive headline, and their
disagreement about magnitudes is a useful lesson in what different outcome measures
can see.

### Well-being: help-seeking and mortality

@ozer2026letitride ask whether legalization shows up in the two outcomes the clinical
literature has long associated with problem gambling: help-seeking and suicide. They
assemble state-level counts of calls to the National Problem Gambling Hotline
alongside state suicide counts, and exploit the combination of legalization states
occupy—online only, retail only, both, or neither.

The channel result is stark. Hotline calls are essentially uncorrelated with the
legalization of physical sportsbooks, and strongly correlated with the legalization
of online betting, where calls rise by roughly 22.8%. The authors further report that
online legalization is associated with an increase in suicides, with effects
concentrated among young, unmarried, and relatively less educated men. (The published
article is paywalled; the magnitudes here come from its abstract and the authors'
university news release, and the suicide result in particular should be read as the
correlational statement the authors make it.)

Two things are worth saying about the outcome measures. Hotline calls are a
*help-seeking* measure, not a prevalence measure, and help-seeking responds to
salience: a state that legalizes online betting also floods itself with advertising
that carries a helpline number, which mechanically raises calls. That is a real
confound and it cuts toward overstatement. Against it, the retail-versus-online
contrast is doing exactly the work a placebo should—retail legalization also comes
with advertising, and calls do not move.

### Household finance: credit and insolvency

@hollenbeck2026financial ask the same question of a financial outcome that is
measured continuously, for nearly everyone, without any self-report. They use the
University of California Consumer Credit Panel, a nationally representative 2% sample
of U.S. adults with a credit file, and after restricting to non-movers with active
accounts, work with roughly 4.4 million individuals and about 90 million quarterly
observations from March 2016 to June 2023. Treatment effects are estimated with the
@callaway2021did estimator, again with two definitions—any legal access, and online
access specifically.

The pattern repeats. Average credit scores fall by roughly 0.8 points in states with
any legal sports betting, and by about 2.75 points—roughly three times as much—where
online betting is available. Online access raises the likelihood of bankruptcy by
about 10% and debt sent to collections by about 8%, with effects emerging roughly two
years after legalization. Crucially, the damage is not spread evenly: it is
"overwhelmingly concentrated among the subprime category," people who were already
financially precarious before legal betting arrived. (These figures are from the
authors' open April 2025 working paper; the published version may differ.)

A 2.75-point average credit score decline sounds trivial, and read as an average it
is. That is the point. A small average over a large population that is concentrated
in a vulnerable tail is exactly the distributional structure @sec-ms-corpdev warned
about in a different context: the mean is the wrong summary statistic when the
mechanism operates on a subpopulation.

### Savings: what the money stopped doing

@baker2026gambling ask where the wagered money came from, using household transaction
data rather than credit files. Their answer reframes the whole debate. Legalization
did not primarily crowd out other gambling or general consumption. It crowded out
*saving*: net investment into brokerage accounts falls by about 20% following
legalization, the heaviest bettors cut their investment deposits by more than half,
and roughly 20 cents of every dollar sent to a betting app comes out of money that
would otherwise have been invested. Households with above-median savings barely
change their investing; low-savings households cut deposits substantially.

This is the most economically consequential of the three findings and the least
intuitive. The harm is not only that some people lose money betting—that is the
product working as designed—but that the losses are financed disproportionately out
of long-horizon savings by precisely the households with the least of it. The welfare
cost therefore compounds at the market rate of return over a working life, which no
single-period accounting of gambling losses will capture.

### Health behaviors

@dasgupta2026substance extend the outcome set to substance use, finding early
evidence that sports betting laws move it as well. Read alongside
@wang2025aggression on aggression, the emerging picture is that the effects are not
confined to the household balance sheet.

### Why the convergence is the argument

Any one of these studies is contestable. Hotline calls confound help-seeking with
prevalence. Credit scores are coarse and slow. Transaction panels cover a
self-selected set of account-linkers. What is hard to explain away is that four teams
using four unrelated measurement systems—a public-health hotline, a credit bureau, a
household transaction aggregator, and health surveys—all find the same *channel*
asymmetry: retail legalization does little, online legalization does a lot. Rival
explanations have to account for the pattern in all four, and the obvious ones do not.
This is triangulation in the sense of @sec-complementary-evidence, and it is the
strongest form of argument available when no single design is decisive.

Balance requires noting the dissent. @walker2026badresearch argue that a good deal of
sports-betting research is methodologically weak in ways that lead to bad regulation,
and the critique deserves engagement rather than dismissal: several widely cited
findings in this area rest on aggregate correlations, on outcome measures that
confound salience with incidence, or on treatment definitions that pool channels. The
three studies above are largely not the target—they are the ones that got the design
right—but a doctoral reader entering this literature should expect the median paper
in it to be considerably weaker than the modal paper they will hear cited.

## Why the Channel Is the Treatment {#sec-gb-channel}

The consistent retail-versus-online asymmetry demands a mechanism, and the marketing
literature already supplies most of it. Moving a sportsbook from a building to a phone
changes at least five things at once.

*Friction.* A retail bet requires a trip. An online bet requires a thumb. Every
result in @sec-nudges about the power of default and friction predicts that this
single change dominates the others.

*Continuous availability.* A casino visit is an episode with a beginning and an end.
A phone is available during the game, at work, and at 2 a.m., which converts a
discrete purchase occasion into a continuous one.

*In-play betting.* The most consequential product innovation is the ability to bet on
the next play rather than the final score, which compresses the wager-outcome cycle
from hours to seconds. The reward-schedule machinery of @sec-gaming applies directly:
a variable-ratio schedule with a short inter-trial interval is the most
reinforcement-dense arrangement known, and in-play betting is precisely that.

*Push notification.* The operator can initiate the purchase occasion. No retail
sportsbook can text a lapsed customer an offer during halftime.

*Instant, frictionless funding.* Linked bank accounts and card deposits remove the
cash constraint that a physical venue imposes, which is the most plausible proximate
explanation for the @baker2026gambling savings result.

@ma2014onlinegambling had established well before legalization that online gambling
behavior responds to cumulative and recent outcomes and to prior use—that is, that the
channel supports precisely the outcome-contingent escalation the harm literature
worries about—and @cotte2009blackjack had shown ethnographically that online and
casino gambling are experientially different activities rather than the same activity
in two locations. The 2026 causal literature can be read as the population-scale
confirmation of what those two papers described.

## The Marketing Machinery {#sec-gb-marketing}

Gambling is a marketing problem in a strict sense: every element of the mix is a
deliberate design choice, and the choices are unusually legible.

### Odds are a price

The operator's margin—the *hold* or *vig*—is embedded in the odds as a
deviation from fair pricing, which makes betting markets a natural laboratory for
price-response research where the true value is eventually revealed. The oldest
finding in that laboratory is the **favorite-longshot bias**: longshots are
systematically overpriced relative to their win probability and favorites
underpriced. @thaler1988parimutuel put it on the anomalies map, and
@ottaviani2010favorite show it can arise from noisy information and rational
inference rather than from a taste for skewness, which matters because the two
explanations imply different remedies. For a marketing reader the useful reframing is
that the bias is a *price discrimination* result: the operator earns a higher margin
on the bets that the least informed customers prefer.

### Promotions manufacture a purchase occasion

Free bets, deposit matches, and odds boosts are the industry's dominant acquisition
instrument, and they are structurally identical to the promotional lotteries
@dhar1995lottery analyzed: an uncertain reward whose expected value consumers
systematically misjudge. @rossi2021freebet document the resulting advertising volume
and content on social media, and find substantial regulatory non-compliance—a finding
that belongs to the public-policy tradition of @sec-policy-papers-marketing and
supplies exactly the kind of "instrument became cheaper on a date" shock that
chapter's template asks for.

### The customer base is the policy problem {#sec-gb-customer-base}

The sharpest connection to the rest of this book runs through customer-base analysis.
@lam2009nbd show that gambling purchases follow the NBD and NBD–Dirichlet processes
that @sec-cba develops for packaged goods, and
@narayanan2011casino estimate individual-level casino gambling behavior with the
heterogeneity machinery of @sec-clv. Those two facts together have an uncomfortable
implication that neither the harm literature nor the CLV literature usually states.

A gamma-mixed Poisson purchase process is heavy-tailed. Under it, a small fraction of
customers accounts for a large fraction of transactions—the standard result the CLV
chapters treat as a targeting opportunity. In gambling, that same tail is the set of
customers a clinical framework would classify as at risk. The operator's revenue
concentration and the regulator's harm concentration are therefore not two facts but
one fact, and any responsible-gambling instrument that binds on the tail is a direct
tax on the revenue base. The simulation below makes the exchange rate explicit.

```{r gb-concentration, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({library(ggplot2)})
set.seed(2018)

# A synthetic bettor population whose annual bet counts follow an NBD
# (gamma-mixed Poisson), the process Lam & Mizerski (2009) fit to gambling
# purchases. Stakes are lognormal and independent of frequency.
n      <- 50000
lambda <- rgamma(n, shape = 0.40, rate = 0.10)   # latent monthly bet propensity
bets   <- rpois(n, lambda * 12)                  # annual bets
stake  <- rlnorm(n, meanlog = 3.0, sdlog = 1.2)  # typical stake per bet ($)
handle <- bets * stake                           # total amount wagered
hold   <- 0.075                                  # operator margin on handle
rev    <- hold * handle                          # expected operator revenue = player loss

active <- rev[rev > 0]
sorted <- sort(active, decreasing = TRUE)
share_top <- function(p) sum(sorted[1:ceiling(p * length(sorted))]) / sum(sorted)

cat("Active bettors:          ", length(active), "of", n, "\n")
cat("Median annual loss:      $", round(median(active)), "\n")
cat("Mean annual loss:        $", round(mean(active)), "\n")
cat("Revenue share, top  1% : ", round(100 * share_top(0.01), 1), "%\n")
cat("Revenue share, top  5% : ", round(100 * share_top(0.05), 1), "%\n")
cat("Revenue share, top 10% : ", round(100 * share_top(0.10), 1), "%\n")

# A responsible-gambling instrument: cap each player's annual net loss at L.
# Revenue retained is sum(min(rev_i, L)); players affected are those above L.
# Caps are set at quantiles of the realized loss distribution so that the
# incidence comparison does not depend on the calibration above.
qs   <- c(0.90, 0.95, 0.99, 0.995, 0.999)
caps <- as.numeric(quantile(active, qs))
inc  <- data.frame(
  cap_pctile = 100 * qs,
  cap_usd    = round(caps),
  affected   = sapply(caps, function(L) 100 * mean(active > L)),
  retained   = sapply(caps, function(L) 100 * sum(pmin(active, L)) / sum(active))
)
inc$forgone_per_pp <- with(inc, (100 - retained) / affected)
print(inc, row.names = FALSE, digits = 3)
```

The numbers state the tension quantitatively, and the asymmetry is severe. The
heaviest 1% of active bettors supply roughly a quarter of revenue and the heaviest 5%
supply about half. Consequently a loss cap set at the 99th percentile of the loss
distribution—binding on one player in a hundred—costs the operator close to 12% of
revenue, and a cap at the 95th percentile, binding on one player in twenty, costs
about 30%. The final column is the *incidence* of the policy: forgone revenue per
percentage point of players affected, which rises monotonically as the cap tightens
onto the tail, from about 4 at the 90th percentile to about 24 at the 99.9th.

That column is the number a welfare analysis in the sense of
@sec-welfare-calibration needs, and it is almost never computed on a common footing.
Industry submissions to consultations report revenue at risk; public-health
submissions report players protected. They are the numerator and denominator of the
same ratio, and quoting either alone is a rhetorical choice rather than an analytical
one. Note also what the exercise does *not* settle: it prices the policy, it does not
evaluate it. Whether trading 30% of revenue for protection of the heaviest 5% is
worthwhile depends on how much of that tail's spending is harm rather than
willingly-purchased entertainment—which is a measurement question the field has not
answered and the agenda in @sec-gb-agenda returns to.

```{r gb-lorenz, message=FALSE, warning=FALSE}
#| fig-cap: "Revenue concentration among simulated bettors. The diagonal is the equal-contribution benchmark; the curve is the cumulative revenue share by player percentile, ranked from heaviest to lightest. The shaded region marks the top 5% of players, whose losses the operator books as revenue and a clinical framework books as harm."
q   <- seq(0.001, 1, by = 0.001)
cum <- sapply(q, function(p) sum(sorted[1:ceiling(p * length(sorted))]) / sum(sorted))
df  <- data.frame(pct = 100 * q, cum = 100 * cum)

ggplot(df, aes(pct, cum)) +
  annotate("rect", xmin = 0, xmax = 5, ymin = 0, ymax = 100,
           fill = "grey80", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.9) +
  labs(x = "Player percentile, heaviest first (%)",
       y = "Cumulative share of operator revenue (%)") +
  theme_minimal(base_size = 11)
```

The mechanism is worth stating plainly because it generalizes past gambling. Whenever
a firm's revenue follows a heavy-tailed consumption process *and* excess consumption
of the product is itself the harm, customer-base concentration and harm concentration
coincide by construction. That is true of gambling, and it is at least arguably true
of free-to-play games (@sec-gaming), short-form video, and alcohol. The CLV
machinery is neutral about this; it identifies the tail either way. What changes is
what the firm is supposed to do having found it.

## The Consumer Behavior of Betting {#sec-gb-behavior}

The behavioral literature explains why the demand curve for this product has such a
strange shape.

**Why people gamble at all** is not obvious under expected utility, since the
expected return is negative by construction. @barberis2012casino gives the most
useful modern answer: under cumulative prospect theory, an agent who overweights the
tails of a probability distribution finds a sequence of casino bets attractive *ex
ante*, and—this is the part that matters—will plan to stop at a threshold and then
predictably fail to, because the plan is time-inconsistent. The model derives
overstaying from the same primitives that derive entry, which is exactly what a good
theory of a self-control good should do.

@andrade2009planned confirm the prediction experimentally in the field's own journal:
what bettors *plan* to wager in sequential gambles and what they *actually* wager
diverge systematically, with prior outcomes driving the gap. That divergence is the
empirical signature of the normative-departure taxonomy in @sec-normative-benchmarks:
the consumer is not merely choosing a risky option, they are failing to execute their
own plan, which is the specific condition under which restricting choice can raise
welfare.

The mechanisms behind the divergence are well documented. @thaler1990housemoney show
that prior gains raise risk-taking ("house money") while prior losses raise it too
when a break-even option is available—two effects that between them cover most of a
losing session. @clotfelter1993gamblers document the gambler's fallacy in lottery
play, and @sundali2012nearmiss show that near-misses in casino roulette increase
subsequent betting, which is the design principle behind every slot reel that stops
one symbol short.

Finally, the market exists inside a culture that had to be built. @humphreys2010semiotic
traces how casino gambling was legitimated as a consumption practice, and
@mascarenhas1990ethical—published thirty-six years ago in *JAMS*—supplies a formal
methodology for the ethical assessment of exactly this kind of marketing phenomenon.
That marketing had the tools to evaluate this market before the market existed, and
largely did not use them, is worth sitting with.

## Prediction Markets {#sec-gb-prediction}

A prediction market is structurally a betting market whose purpose is to generate a
forecast rather than to entertain. The price of a contract paying \$1 if an event
occurs is read as the market's probability of that event, and the case for the
instrument is that it aggregates dispersed private information with a financial
incentive for honesty.

**Do they work?** @spann2008sportsforecast run the direct comparison in sports, where
outcomes resolve quickly and often, and find prediction markets competitive with
betting odds and superior to tipsters. @berg2009google used markets to forecast
Google's market capitalization before its IPO, which is the harder test—a one-shot
event with no base rate.

**Can they be manipulated?** This is the standing objection, and the answer is
encouraging with a caveat. @camerer1998manipulated ran a field experiment placing and
cancelling large racetrack bets and found the market largely undisturbed, evidence
that manipulation is expensive and self-defeating. @jian2012manipulation show
theoretically and experimentally that the answer depends on the trading mechanism and
on how information is distributed—so manipulability is a property of market *design*,
not of prediction markets in general. @slamka2013liquidity take up the design question
directly, comparing automated market makers on performance and liquidity, which is the
practical problem for any firm running an internal market too thin to clear.

**Why marketing should care.** @qiu2017predictionmarket ran a field experiment using a
social-media-based prediction market to elicit voluntary knowledge contribution,
which points at the applied use: a firm that wants an honest internal forecast of a
launch date, a demand level, or a project risk has, in a prediction market, an
elicitation mechanism that rewards accuracy rather than optimism. That makes it a
preference- and belief-measurement instrument, and it belongs beside the methods of
@sec-preference-measurement rather than in a separate literature.

**Where the boundary now sits.** The regulatory line between a prediction market and a
sportsbook has become genuinely unclear, as event-contract platforms list markets on
sporting outcomes under financial rather than gaming regulation. The distinction that
survives scrutiny is not the asset and not the regulator but the *purpose the price
serves*: a prediction market's output is consumed by third parties as information,
and its social value comes from that externality. A sportsbook's prices are consumed
by the bettor as a product. A venue can be both, and when it is, the informational
defense does not automatically transfer to the consumption side. That is an open
research question rather than a settled taxonomy, and it is the most promising place
in this chapter for a marketing paper.

## A Research Agenda {#sec-gb-agenda}

The causal literature has established the first-order facts. What is missing is
mostly marketing.

1. **Advertising is unmeasured.** The three flagship studies identify off
   *legalization*, not off exposure. Nobody has estimated the effect of sports-betting
   advertising itself, despite the fact that this is the field's core competency and
   that ad exposure varies within legalized states for reasons unrelated to
   legalization timing (national broadcast buys, team sponsorships, DMA boundaries).
2. **Promotions are unmeasured.** Free bets and deposit matches are the acquisition
   instrument. Their effect on subsequent betting intensity—not on acquisition, which
   the operator knows, but on the durable consumption path—is unestimated in public
   research, and the promotional-lottery framing of @dhar1995lottery is the natural
   theoretical starting point.
3. **The substitution question is open.** @baker2026gambling establish crowd-out of
   savings rather than of other gambling. Whether legal betting substitutes for
   illegal betting—the industry's central policy claim—remains poorly identified, and
   it is the quantity that decides whether legalization was welfare-improving on the
   industry's own terms.
4. **Responsible-gambling instruments need incidence estimates.** Deposit caps,
   cooling-off periods, and self-exclusion are evaluated, when at all, on
   participation. The simulation in @sec-gb-customer-base shows the missing number is
   the revenue incidence, and computing it requires operator data that a research
   partnership could obtain.
5. **The welfare calculation has not been done properly.** A full accounting needs the
   consumer surplus of the roughly 98% of bettors for whom this is inexpensive
   entertainment, netted against the concentrated losses of the tail. Every
   ingredient of @sec-welfare-calibration is available; nobody has assembled them, and
   the resulting bound would be the most useful single number in the policy debate.

Each of these is a five-move paper in the sense of @sec-pwm-template, and the shock,
the untargeted market, and the exposure variation are all in place.

## Key Takeaways

- The post-*Murphy* rollout is a **two-treatment staggered design**, and the treatment
  that matters is the **channel**. Retail legalization moves almost nothing; online
  legalization moves everything measured so far. A study that pools the two estimates a
  weighted average of a null and a large effect (@sec-gb-paspa).
- The three flagship results are convergent and channel-specific: problem-gambling
  hotline calls rise about 22.8% with online legalization [@ozer2026letitride]; average
  credit scores fall about 2.75 points with bankruptcy up roughly 10%, concentrated in
  the subprime tail [@hollenbeck2026financial]; and net brokerage investment falls about
  20%, with roughly 20 cents of every wagered dollar coming out of savings
  [@baker2026gambling]. Four independent measurement systems agreeing on the channel
  asymmetry is the argument (@sec-complementary-evidence).
- **Small averages concentrated in a vulnerable tail are the signature of this
  literature**, and reporting the mean is the characteristic error. The distributional
  question is the policy question.
- Gambling purchases follow the same **NBD/Dirichlet** processes the customer-base
  chapters develop [@lam2009nbd; @narayanan2011casino], which means operator revenue
  concentration and clinical harm concentration are one fact rather than two. Any
  binding responsible-gambling instrument is therefore a tax on the revenue base, and
  its **incidence** is computable (@sec-gb-customer-base).
- Betting demand is generated by **time-inconsistent plans**, not merely by risk
  preference: prospect-theoretic agents enter rationally and overstay predictably
  [@barberis2012casino], and bettors' planned and actual wagers diverge with prior
  outcomes [@andrade2009planned; @thaler1990housemoney]. That divergence is the
  specific condition under which choice restriction can raise welfare
  (@sec-normative-benchmarks).
- **Prediction markets are the same instrument aimed at information.** They forecast
  competitively [@spann2008sportsforecast; @berg2009google], and their manipulability
  is a property of market design rather than of the form [@camerer1998manipulated;
  @jian2012manipulation; @slamka2013liquidity]. The defensible distinction from a
  sportsbook is what the price is *for*, not who regulates it.
- The open work is marketing's own: **advertising and promotion effects are
  unestimated**, substitution from illegal betting is unidentified, and no one has
  assembled the welfare bound (@sec-gb-agenda).

## Further Reading

For the causal literature, read @hollenbeck2026financial first for the cleanest design
statement, then @baker2026gambling for the substitution result that reframes it, then
@ozer2026letitride for the outcomes that matter most and are hardest to measure.
@walker2026badresearch is the necessary corrective on research quality in the wider
field. For behavior, @barberis2012casino is the theoretical anchor and
@andrade2009planned the marketing counterpart. For prediction markets, start with
@spann2008sportsforecast for whether they work and @jian2012manipulation for when they
break. @eadington1999casino and @thaler1988parimutuel remain the best short
introductions to the economics, and @humphreys2010semiotic to how the market became
culturally acceptable in the first place.
