# Marketing–Finance Interface {#sec-marketing-finance}

Marketing spends real money—on advertising, salesforces, distribution, brands,
and customer relationships—and the people who allocate capital eventually ask
what those expenditures are worth. The *marketing–finance interface* is the body
of theory and empirical method that answers this question by linking marketing
actions to firm value: to cash flows, to risk, and ultimately to the price an
equity market sets on the firm. Its central claim is deceptively simple. The
things marketing builds—brand equity, a satisfied and loyal customer base,
channel and partner relationships—are *assets*, and like any asset they should be
visible in the value of the firm even though accounting rules largely keep them
off the balance sheet [@Srivastava_1998; @srinivasan2009].

Two intellectual currents meet here. From finance comes the discipline of
discounted cash flow and the apparatus of asset pricing: a firm is worth the
present value of its future cash flows, and a stock's return reflects the
revision of investors' expectations about those flows. From marketing comes a
theory of *where* those cash flows originate—in customers and the relationships,
knowledge, and reputation that produce repeat purchase, price premiums, and
lower acquisition costs. The interface insists that the two be made consistent:
actions that create value in the product market should, under an efficient
capital market, create value for shareholders [@srinivasan2009]. When they do
not, either the marketing action destroyed value or the market has not yet priced
it—both empirically interesting outcomes.

By the end of this chapter the reader should be able to state the market-based
assets framework formally, write down and estimate the workhorse models that
connect marketing to firm value (stock-return response models, persistence/VAR
models, and Tobin's $q$ regressions), reason carefully about what identifies a
causal effect in each, and recognize the measurement traps—most acutely in
accounting-based proxies for $q$—that have tripped up the literature. We treat
the event-study machinery for discrete marketing events (acquisitions, alliances,
product introductions) as a running application, since mergers and acquisitions
and strategic alliances are where the interface has produced its deepest body of
evidence. The branding chapter (@sec-branding) develops the event study and
stochastic-frontier capability estimators in the specific context of brand
assets; here we build the general scaffolding.

## Market Efficiency and the Logic of Valuation

Everything downstream rests on an assumption about how stock prices incorporate
information, so it is worth stating precisely. The *efficient market hypothesis*
classifies markets by the information set already impounded in prices. Under
**weak-form** efficiency prices reflect the history of prices themselves; under
**semi-strong** efficiency they reflect all public information in real time; under
**strong-form** efficiency they reflect public *and* private information. The
working consensus in the marketing–finance literature is that real markets sit
between weak and semi-strong form: public information about marketing—an earnings
release that reveals advertising's payoff, a product-recall announcement, an
alliance—moves prices, and it does so promptly though not always completely
[@srinivasan2009].

This has a sharp empirical implication. If markets are at least near-semi-strong
efficient, then good news about marketing raises the stock price and bad news
lowers it, and—because the price is forward-looking—the *change* in price at the
moment news arrives is an unbiased, risk-adjusted estimate of the news's value to
shareholders. Stock-market valuation is then in sync with product-market
valuation: the same actions that build value with customers build value for
investors. This equivalence is what licenses the use of stock returns to value
marketing, and it is the premise the event study exploits.

Formally, let $V_{it}$ be the market value of firm $i$'s equity at the end of
period $t$. The dividend-discount identity writes it as the expected present
value of future cash flows to equity,

$$
V_{it} = \sum_{T=t}^{\infty}
  \left(\frac{1}{1+r_{it}}\right)^{T-t}\,
  \mathbb{E}_t\!\left[\mathrm{CF}_{iT}\right],
$$ {#eq-mf-ddm}

where $r_{it}$ is the firm's cost of equity capital and $\mathbb{E}_t[\cdot]$ is
the expectation conditional on information available at $t$. Marketing enters
through the cash-flow expectations. @Srivastava_1998 organize marketing's value
contribution into four channels visible in @eq-mf-ddm: marketing can
**accelerate** cash flows (bring them forward in time), **augment** their
**level**, **reduce their risk** (volatility and vulnerability, which lowers
$r_{it}$), and raise their **residual value** (the terminal value beyond the
explicit forecast horizon). The rest of the framework is an elaboration of how
marketing assets operate through these four levers.

## Market-Based Assets

### Definition and the Resource-Based Test

The conceptual core of the interface is the **market-based asset**. The following
landmark definition frames everything that follows.

> Market-based assets "arise from the commingling of the firm with entities in
> its external environment," where an asset is "any physical, organizational, or
> human attribute that enables the firm to generate and implement strategies that
> improve its efficiency and effectiveness in the marketplace"
> [@Srivastava_1998, pp. 2–3].

The defining feature is *externality of locus*: unlike a factory or a patent, a
market-based asset lives in a relationship or in knowledge that spans the firm's
boundary. @Srivastava_1998 distinguish two kinds. **Relational** market-based
assets are outcomes of the firm's relationships with external stakeholders—
distributors, retailers, end customers, strategic partners. Brand equity is the
relational asset between a firm and its customers; channel equity is the
relational asset between a firm and its intermediaries. **Intellectual**
market-based assets are the firm's knowledge about its environment: the emerging
and potential states of market conditions and of the entities (competitors,
customers, channels, suppliers, regulators) within them.

For a resource to create *sustained* value it must pass the resource-based view's
four tests: it must be **convertible** (the firm can actually exploit it), **rare**,
**imperfectly imitable**, and have **no perfect substitutes** [@Srivastava_1998].
Knowledge and relationships pass all four, which is precisely why they can support
a durable competitive advantage where tangible assets—easily bought, copied, or
substituted—cannot. Each asset has a **stock** dimension (the amount of brand
equity or customer knowledge the firm possesses at a point in time) and a **flow**
dimension (the rate at which that stock is being augmented or is decaying)
[@Srivastava_1998, p. 5]. The stock–flow distinction matters for estimation:
marketing expenditures are flows that accumulate, with decay, into the stock that
produces cash flow—a structure we exploit in the persistence and perpetual-
inventory models below.

### From Assets to Shareholder Value

The framework's payoff is a causal chain from marketing actions, through
market-based assets and intermediate product-market outcomes, to the four
shareholder-value levers of @eq-mf-ddm. @fig-mba-chain renders the chain;
it is our own redrawing of the logic in @Srivastava_1998 and the "chain of
marketing productivity" in @Rust_2004, neither reproduced from the originals.

```{mermaid}
%%| label: fig-mba-chain
%%| fig-cap: "The marketing–finance value chain. Marketing actions build market-based assets, which generate product-market outcomes, which map onto the four shareholder-value levers of the discounted-cash-flow identity. Synthesized from @Srivastava_1998 and @Rust_2004."
flowchart LR
  A["Marketing actions<br/>(advertising, innovation,<br/>relationship investment)"]
    --> B["Market-based assets<br/>Relational: brands, customers,<br/>channels, partners<br/>Intellectual: market knowledge"]
  B --> C["Product-market outcomes<br/>faster penetration · price premium<br/>share premium · extensions<br/>lower selling/service cost<br/>loyalty / retention"]
  C --> D1["Accelerate cash flows"]
  C --> D2["Augment level of cash flows"]
  C --> D3["Reduce volatility &<br/>vulnerability of cash flows"]
  C --> D4["Enhance residual value"]
  D1 --> E["Shareholder value"]
  D2 --> E
  D3 --> E
  D4 --> E
```

The mechanisms behind each arrow are worth making explicit, because they convert
soft marketing language into the four hard financial levers.

**Accelerating cash flows.** Greater brand awareness speeds consumer response to
advertising and shortens the path to trial [@keller1993], and a large installed
base together with partnership networks compresses the product life cycle by
speeding the adoption process [@robertson1993reduce]. Time-to-market acceptance
is a cash-flow timing variable, and—because near cash is worth more than distant
cash—accelerating it raises value even when the *level* of cumulative cash flow is
unchanged.

**Augmenting the level of cash flows.** Market-based assets raise revenues (brand
extensions open new categories), lower costs (extensions and the relationship's
advertising efficiencies cut unit selling cost), and reduce working- and
fixed-capital requirements (relationship marketing and cooperative ventures—co-
branding and co-marketing alliances—let partners leverage each other's networks
and customer bases rather than build their own).

**Reducing risk.** The *vulnerability* of cash flows falls with customer
satisfaction, loyalty, and retention, and stability rises as information sharing,
automatic replenishment, and lower inventories smooth operations. Higher switching
costs and operational integration raise the net present value of the cash-flow
stream by lowering its discount rate $r_{it}$.

**Enhancing residual value.** A larger, more loyal, higher-quality customer base
buys more and erects barriers to competitive entry, raising the terminal value
that dominates @eq-mf-ddm for long-lived firms.

@Srivastava_1998 summarize the logic in three propositions: the more market-based
assets external entities help develop, the more eager and valuable those entities
become to the firm; the more such assets pass the resource-based tests, the more
external entities benefit; and when a firm leverages market-based assets to boost
cash flows, shareholder value is created. The empirical program of the interface
is, in effect, the testing of this third proposition.

### Why Not Just Use Accounting Numbers?

A natural objection is that firms already report performance, so why build a
parallel valuation apparatus. The answer is that the obvious accounting measures
are poor proxies for value creation. Book value records assets at historical cost
and ignores intangibles entirely; replacement value is conceptually better but
practically unmeasurable for relationships and knowledge; and price–earnings
multiples inherit the well-known pathologies of accrual accounting—earnings are
backward-looking, unadjusted for risk, and manipulable through accruals timing.
The *shareholder-value* approach—discounted future cash flow, as in
@eq-mf-ddm—dominates these because it is forward-looking, risk-adjusted, and
harder to game [@Srivastava_1998]. This is also the principal reason cash flow,
rather than earnings, is the preferred firm-performance variable in interface
research, and it motivates the use of $q$-type and stock-return measures in place
of profitability ratios.

## Methods for Connecting Marketing to Firm Value

Three broad estimation strategies dominate, distinguished by the nature of the
marketing event and the time structure of the data. @tbl-mf-methods contrasts
them; the subsections that follow develop the two most important.

| Approach | What it measures | Core model | Identifying assumption | Breaks when |
|---|---|---|---|---|
| **Event study / stock-return response** | Value of a discrete, dated event | Market-model abnormal returns $\mathrm{AR}_{it}$, cumulated to $\mathrm{CAR}_i$ | Market is semi-strong efficient; event date is exogenous and unconfounded within the window | Information leaks before the date; confounding events share the window; the event is anticipated |
| **Persistence modeling (VAR)** | Long-run cumulative impact of a marketing shock, separating temporary from permanent effects | Vector autoregression with impulse-response and unit-root analysis | Shocks are correctly identified (ordering/IRF); the system is stable or has a well-defined permanent component | Wrong variable ordering; structural breaks; omitted feedback channels |
| **Tobin's $q$ / valuation regression** | Cross-sectional association of marketing assets with firm value | $q_{it} = \mathbf{x}_{it}^{\top}\boldsymbol{\beta}+\alpha_i+\delta_t+u_{it}$ | Marketing regressor is exogenous conditional on controls and fixed effects | Reverse causality; omitted intangibles; accounting proxy bias (see @sec-mf-aatq) |

: Three workhorse approaches to the marketing–finance interface, with their identifying assumptions. {#tbl-mf-methods}

A fourth, the single-equation **error-correction model** (ECM), is used when the
data are nonstationary and one wants to separate a long-run equilibrium relation
from short-run dynamics; it recognizes the random-walk character of stock prices
and is a staple of the persistence tradition [@srinivasan2009a].

### Stock-Return Response Modeling

The cleanest way to value a discrete marketing event is to read the value
directly off the stock price, exploiting market efficiency. Begin again from the
market-value identity and difference it across one period. With $\mathrm{Eret}_{it}$
the expected (required) return on the equity and $r_{it}$ the discount rate,

$$
V_{it} = (1+\mathrm{Eret}_{it})\,V_{it-1}
  + \sum_{T=t}^{\infty}
    \left(\frac{1}{1+r_{it}}\right)^{T-t}
    \Delta\,\mathbb{E}_t\!\left[\mathrm{CF}_{iT}\right],
$$ {#eq-mf-valdiff}

where $\Delta\,\mathbb{E}_t[\mathrm{CF}_{iT}] \equiv
\mathbb{E}_t[\mathrm{CF}_{iT}] - \mathbb{E}_{t-1}[\mathrm{CF}_{iT}]$ is the
*revision* in the expected cash flow for date $T$ that arrives during period $t$.
The first term is the value the firm would have had absent any news—it grows at the
required return—and the second term is the capitalized value of the news itself.
Rearranging gives the realized stock return as an expected component plus an
abnormal component driven entirely by cash-flow-expectation revisions:

$$
\mathrm{StockReturn}_{it}
  = \frac{V_{it}-V_{it-1}}{V_{it-1}}
  = \mathrm{Eret}_{it}
  + \underbrace{\frac{1}{V_{it-1}}
      \sum_{T=t}^{\infty}\left(\frac{1}{1+r_{it}}\right)^{T-t}
      \Delta\,\mathbb{E}_t\!\left[\mathrm{CF}_{iT}\right]}_{\text{abnormal return }\;\mathrm{AR}_{it}}.
$$ {#eq-mf-return}

@eq-mf-return is the theoretical justification for the event study. The
**abnormal return** $\mathrm{AR}_{it}$ is the realized return net of its
expectation, and under market efficiency it equals the present value of the
cash-flow news, scaled by beginning-of-period value. To take it to data one needs
a model for the expected return $\mathrm{Eret}_{it}$. The standard choice is the
*market model*, a single-factor regression of the firm's return on the market
return estimated over a clean pre-event window $[t_0, t_1]$:

$$
R_{it} = \alpha_i + \beta_i R_{mt} + \varepsilon_{it},
\qquad
\widehat{\mathrm{AR}}_{it} = R_{it} - \hat\alpha_i - \hat\beta_i R_{mt}.
$$ {#eq-mf-marketmodel}

Cumulating over the event window $W$ gives the cumulative abnormal return
$\mathrm{CAR}_i = \sum_{t \in W}\widehat{\mathrm{AR}}_{it}$, which is then regressed
cross-sectionally on event characteristics to learn *what* drives value. The
branding chapter (@sec-branding) gives the full estimator and the standard
confound screens; the key identifying requirements are that the event date be
exogenous and that no other value-relevant news (earnings, splits, executive
turnover, buybacks, dividend changes) contaminate the window. The following code
simulates the procedure end to end so the reader can see exactly where each
quantity comes from.

```{r mf-event-study, message=FALSE, warning=FALSE}
set.seed(21)

# --- simulate a market and one firm with a known event-day jump --------------
n_est  <- 250          # estimation-window trading days
n_evt  <- 5            # event-window days (e.g., -2..+2)
alpha  <- 0.0002; beta <- 1.10
mkt_est <- rnorm(n_est, 0.0004, 0.009)
firm_est <- alpha + beta * mkt_est + rnorm(n_est, 0, 0.012)

# --- estimate the market model on the clean pre-event window -----------------
fit <- lm(firm_est ~ mkt_est)

# --- event window: true cash-flow news adds +3% on the announcement day ------
mkt_evt   <- rnorm(n_evt, 0.0004, 0.009)
news      <- c(0, 0, 0.03, 0, 0)            # day 0 carries the marketing news
firm_evt  <- coef(fit)[1] + coef(fit)[2] * mkt_evt +
             rnorm(n_evt, 0, 0.012) + news

# --- abnormal and cumulative abnormal returns --------------------------------
expected_evt <- predict(fit, newdata = data.frame(mkt_est = mkt_evt))
AR  <- firm_evt - expected_evt
CAR <- cumsum(AR)

data.frame(day = -2:2, AR = round(AR, 4), CAR = round(CAR, 4))
```

The recovered cumulative abnormal return concentrates on day 0 and is close to the
3% shock injected, illustrating why—when the efficiency and no-confound assumptions
hold—the event study delivers a clean, risk-adjusted dollar value for a marketing
event.

### Persistence Modeling

Many marketing effects are not one-day jumps but dynamic responses that build and
decay over months: an advertising pulse raises awareness, which raises sales, which
feeds back into the budget. *Persistence modeling* uses a **vector autoregression**
(VAR) to trace the full dynamic response of firm value (or sales, or cash flow) to
a marketing shock and, crucially, to separate the *temporary* component (which
dies out) from the *permanent* component (which is impounded forever)
[@srinivasan2009]. For a vector $\mathbf{y}_t$ stacking, say, log firm value, a
marketing variable, and a control,

$$
\mathbf{y}_t = \mathbf{c} + \sum_{\ell=1}^{p}
  \mathbf{\Phi}_\ell\,\mathbf{y}_{t-\ell} + \boldsymbol{\epsilon}_t,
\qquad \boldsymbol{\epsilon}_t \sim (\mathbf{0}, \mathbf{\Sigma}),
$$ {#eq-mf-var}

the **impulse-response function** $\partial \mathbf{y}_{t+h}/\partial
\boldsymbol{\epsilon}_t$ gives the effect of a unit shock $h$ periods out, and its
long-run sum measures total persistent impact. Identification of structural shocks
requires an ordering (or sign/long-run restrictions); the standard recursive
ordering assumes a Cholesky causal chain that the researcher must defend. When the
series are nonstationary—stock prices follow a near random walk—the analyst works
with the unit-root/cointegration structure directly, typically via an
error-correction representation that separates the long-run equilibrium from
short-run adjustment [@srinivasan2009a]. The two pitfalls to flag are that the
permanent/temporary decomposition is only as good as the unit-root inference behind
it, and that an omitted feedback channel (e.g., competitor response) biases the
impulse responses.

## Tobin's $q$ and Valuation Regressions {#sec-mf-q}

The most common cross-sectional measure of firm value in interface research is
**Tobin's $q$**, the ratio of a firm's market value to the replacement cost of its
assets. Intuitively, $q>1$ means the market values the firm above what it would
cost to rebuild its assets, and the gap is attributed to intangibles and growth
options—exactly the territory marketing claims. Its appeal is that it is
forward-looking, risk-adjusted, and less manipulable than accounting profitability,
which is why it is the dependent variable of choice in studies of how marketing
assets relate to firm value [@Grewal_2010; @McAlister_2016; @Morgan_2009;
@Wies_2019]. When probing the drivers of $q$, the literature recommends controlling
at minimum for **financial leverage** and **cash flow**, since both shift $q$ for
reasons unrelated to marketing [@Wies_2019].

The traditional empirical proxy assembles $q$ from financial statements
[@Rao_2004]:

$$
q = \frac{\mathrm{MVE} + \mathrm{PS} + \mathrm{DEBT}}{\mathrm{TA}},
$$ {#eq-mf-q-classic}

where $\mathrm{MVE}$ is the market value of equity (share price times shares
outstanding), $\mathrm{PS}$ the liquidating value of preferred stock,
$\mathrm{DEBT}$ is short-term liabilities net of short-term assets plus the book
value of long-term debt, and $\mathrm{TA}$ the book value of total assets. Because
both numerator and denominator scale with firm size, $q$ is scale-independent and
thus comparable as a measure of *relative* market performance across firms of
different sizes.

A typical valuation regression then takes the form

$$
q_{it} = \mathbf{x}_{it}^{\top}\boldsymbol{\beta}
  + \alpha_i + \delta_t + u_{it},
$$ {#eq-mf-qreg}

with $\mathbf{x}_{it}$ the marketing variables of interest plus leverage and cash
flow, $\alpha_i$ firm fixed effects absorbing time-invalid heterogeneity, and
$\delta_t$ time effects absorbing common shocks. Identification of a *causal*
marketing effect requires that, conditional on the controls and fixed effects, the
marketing regressor be uncorrelated with $u_{it}$. This is demanding: reverse
causality (valuable firms spend more on marketing), omitted intangibles (R&D or
organizational capital correlated with both), and measurement error in the
intangible-laden denominator all threaten it. The next two subsections address the
denominator problem, which is the deepest and least appreciated.

### Total $q$: Putting Intangibles in the Denominator

Equation @eq-mf-q-classic divides market value by *physical* assets at book value,
ignoring the intangible capital—knowledge and organizational/brand capital—that
the firm has built. @peters2017 argue that the neoclassical theory of investment
applies to intangibles just as it does to plant and equipment, and propose a
**total $q$** that puts intangible capital into the denominator at replacement
cost:

$$
\begin{aligned}
q^{\text{tot}}_{it}
  &= \frac{V_{it}}{K^{\text{phy}}_{it} + K^{\text{int}}_{it}} \\[2pt]
  &= \frac{\text{prcc\_f}\times\text{csho} + \text{dltt} + \text{dlc} - \text{act}}
          {\text{ppegt} + K^{\text{int}}_{it}},
\end{aligned}
$$ {#eq-mf-totalq}

where the market value $V_{it}$ is equity (price `prcc_f` times shares `csho`) plus
the book value of debt (`dltt` + `dlc`) minus current assets (`act`),
$K^{\text{phy}}_{it}$ is the replacement cost of physical capital (gross
property, plant, and equipment, `ppegt`), and $K^{\text{int}}_{it}$ is the
replacement cost of intangible capital. Earlier $q$ proxies omitted the intangible
term entirely [@fazzari1987; @erickson2011a], biasing $q$ upward for
intangible-intensive firms.

Intangible capital is itself accumulated by the **perpetual-inventory method**.
For internally created knowledge capital, the end-of-period stock evolves as

$$
G_{it} = (1-\delta_{\mathrm{RD}})\,G_{i,t-1} + \mathrm{RD}_{it},
$$ {#eq-mf-knowcap}

with $G_{i0}=0$, real R&D expenditure $\mathrm{RD}_{it}$ (treated as zero when
missing [@lev]), and an industry-specific depreciation rate $\delta_{\mathrm{RD}}$
(the BEA R&D rates, defaulting to 15% when unavailable [@li2018]). Organizational
and brand capital are built analogously from a fraction of past selling, general,
and administrative (SG&A) spending—which bundles advertising (brand), employee
training (human capital), and distribution systems. Intangible capital is costlier
to adjust than physical capital, a friction the neoclassical model takes seriously.
@fig-perpetual-inventory simulates the knowledge-capital recursion to make the
stock–flow logic concrete.

```{r mf-perpetual-inventory, message=FALSE, warning=FALSE}
# Knowledge-capital accumulation by perpetual inventory (eq. mf-knowcap).
set.seed(21)
years      <- 1:15
rd_flow    <- 100 * (1.08)^(years - 1) * exp(rnorm(length(years), 0, 0.05))
delta_rd   <- 0.15                       # BEA-style default depreciation
G <- numeric(length(years)); G[1] <- rd_flow[1]
for (t in 2:length(years)) {
  G[t] <- (1 - delta_rd) * G[t - 1] + rd_flow[t]
}

plot(years, G, type = "o", pch = 16,
     xlab = "Year", ylab = "Replacement cost",
     main = "Knowledge capital as a depreciated stock of R&D flows",
     ylim = range(c(G, rd_flow)))
lines(years, rd_flow, type = "o", col = "steelblue", pch = 1)
legend("topleft", bty = "n",
       legend = c("Knowledge-capital stock G_t", "Annual R&D flow"),
       col = c("black", "steelblue"), pch = c(16, 1), lty = 1)
```

The stock $G_t$ rises well above any single year's flow because past investments
persist (depreciating at $\delta_{\mathrm{RD}}$); this is exactly the wedge that
@eq-mf-totalq restores to the denominator and that a naive physical-capital $q$
ignores.

### Why Accounting Proxies for $q$ Mislead {#sec-mf-aatq}

A blunt warning is in order, because the convenience of @eq-mf-q-classic has
seduced a generation of marketing papers. @bendle2018misuse label these
financial-statement constructions **accounting-based approximations of Tobin's $q$**
(AATQ) and argue they depart from $q$'s original meaning in ways that
systematically *favor* the marketing variables researchers most want to study.
Their critique has several edges. AATQ are not comparable across industries because
the off-balance-sheet wedge varies by sector; they do not in fact isolate tangible
assets in the denominator; and there is no theoretical reason for AATQ to
equilibrate at one, contrary to a common assumption inherited from the original
$q$. Worse, when AATQ exceeds one—the empirically typical case—it can *rise even in
response to wasted investment*, so an AATQ increase is consistent with both genuine
value creation and value-neutral or value-destroying strategies. The net effect is
a bias toward overstating the effectiveness of investments in market-based assets,
precisely because those assets are the unrecorded items (brand equity, customer
satisfaction) that the AATQ denominator omits. The methodological lesson is to
prefer a $q$ that books intangibles into the denominator (@eq-mf-totalq), to lean
on within-firm variation through fixed effects, and to treat any AATQ-based effect
as suggestive until corroborated.

## Discrete Events: M&A, Alliances, and Innovation

The interface's richest evidence comes from discrete, dated events whose value the
event study can isolate. Mergers and acquisitions (M&A) are the canonical case, and
the marketing question is sharp: *what kind of combination creates value?* Two
hypotheses compete. The **similarity** (relatedness) hypothesis holds that value
comes from combining firms with overlapping products, markets, or technologies,
through scale and consolidation; the **complementarity** hypothesis holds that
value comes from combining firms whose resources differ and fill each other's gaps.
@tbl-mf-mna organizes the core findings; note that the studies disagree because the
*motive* for the merger determines which logic applies.

| Study | DV | Key IV | Finding |
|---|---|---|---|
| @singh1987 | Cumulative portfolio abnormal returns | Acquirer–target relatedness (technological / product-market) | Related firms create more value |
| @shelton1988 | Merger dollar gains | Product-market fit | Related firms create more value |
| @harrison1991 | ROA | Differences in R&D, capital, administrative intensity | R&D *complementarity* boosts unrelated acquisitions |
| @datta1992 | Wealth effects | Bids, financing, acquisition type; served-market overlap | Similar firms create more value |
| @ramaswamy1997 | ROA | Distance in coverage, efficiency, marketing, client mix, risk | Differences *impede* horizontal bank mergers |
| @hitt1998 | ROA | Product-market relatedness and resource complementarity | Resource complementarities aid success |
| @larsson1999 | Synergy realization | Combination potential, integration, low employee resistance | Complementary operations + integration raise synergy |
| @swaminathan2008 | CAR (event study) | Strategic-emphasis alignment; resource similarity/complementarity | Both similarity and complementarity create value, under different motives |

: Selected M&A studies at the marketing–finance interface, illustrating the similarity-versus-complementarity tension. {#tbl-mf-mna}

The financing and timing of deals carries information of its own. Cash deals are
received more favorably than stock-financed deals, because paying with equity
signals that managers believe their stock is overvalued; this adverse reaction to
equity financing is *permanent*, with negative acquirer returns persisting five
years post-takeover [@agrawal1992post; @loughran1997long]. Excess cash is no
panacea—announcement returns fall with an acquirer's cash holdings, consistent
with managers shielded from external capital-market discipline making poorer
investment choices [@harford1999corporate]. There is broad evidence of post-merger
underperformance on average [@agrawal2000], yet the cross-section is informative:
small acquirers earn favorable long-run returns [@mitchell2000managerial;
@moeller2004firm]; undervalued, high book-to-market "value" acquirers outperform
overvalued "glamour" acquirers, whose overconfident managers face weaker scrutiny
[@sudarsanam2003glamour]; bull-market acquisitions underperform bear-market ones as
hubris inflates synergy estimates in upswings [@bouwman2009market]; deals for
privately held or subsidiary targets generate buyer gains through illiquidity
discounts and concentrated post-deal monitoring [@fuller2002returns;
@conn2005impact]; serial acquirers with high valuations destroy value once organic
growth stalls [@moeller2005wealth]; CEO ownership lifts long-run returns
[@cosh2006board]; business similarity and the disposal of non-core assets through
divestitures and spin-offs both raise shareholder value [@megginson2004]; and
cross-border deals favor developed-market acquirers entering emerging markets,
especially in R&D- and brand-intensive businesses where intellectual assets matter
[@chari2009].

@swaminathan2008 bring marketing's distinctive variable—**strategic emphasis**—to
this literature. Following @mizik2003, they operationalize a firm's strategic
emphasis as advertising minus R&D expenditure, scaled by total assets in the
pre-merger year, and define **strategic-emphasis alignment** as the absolute
difference between acquirer and target emphasis. Their event study of 206 deals
shows that when merging firms are *poorly* aligned, diversity (complementarity)
improves value, whereas when they are *well* aligned, value is enhanced by a
consolidation motive—reconciling the similarity/complementarity debate by making the
merger's *motive* the moderator. Marketing assets thus shape M&A value creation
directly.

The interface also documents how M&A interacts with the customer franchise.
@umashankar2021 find that acquisitions tend to *reduce* customer satisfaction, as
executives' attention shifts from customers to financial integration; the resulting
dissatisfaction can cannibalize the very synergies the deal was meant to capture,
though the presence of marketing expertise in the firm's upper echelons mitigates
the damage. Even the market for the dealmakers themselves responds to organizational
structure: high-performing M&A bankers, especially early in their careers, migrate
from bulge-bracket banks toward focused boutiques as cross-subsidization of
underperforming divisions pushes talent out, which in turn shapes deal outcomes
[@gao2024big].

**Strategic alliances** extend the same logic to looser combinations.
@swaminathan2009a show that alliance announcements create value for the
announcing firm, and that *network* position governs how much: a firm's abnormal
returns are most favorable when network **efficiency** (access to non-redundant
partner capabilities) and **density** (interconnection among partners) are
*moderate*, while network reputation and centrality have no effect; a firm's
**marketing-alliance capability**—its accumulated skill at managing prior
alliances—positively drives value creation. Networks amplify alliance benefits,
facilitate compliance, and signal partner and alliance quality. Construction of the
network variables uses a window of prior years (five in @swaminathan2009a, seven in
related work [@gulati1999; @schilling2007]), and because partnerships form through
referrals rather than at random, the authors estimate a selection model alongside
the value-creation model [@verbeek1992a]. Their controls follow the
intangible-asset tradition: installed base, relationship investment, marketing and
advertising expenditures, and R&D, each entered through a Koyck (geometric-lag)
function and combined via principal components to tame multicollinearity
[@dutta1999], plus alliance experience, partner size, intra- versus inter-industry
scope [@rindfleisch2001b], and repeat partnering.

Innovation events round out the picture: a firm's aggregate investment decisions
shape not only its own value but the **growth and volatility of its entire
industry**, as @king2011industry document across 377 industries over sixteen years,
where investments in value creation and value appropriation interact intricately
with the industry environment. Related work on the value of innovation in
acquisitions is treated alongside R&D in @king2008performance.

## Marketing, Information, and the Stock Market

Beyond discrete deals, a growing literature studies how marketing assets shape the
*information environment* of the stock and shows that marketing's financial payoff
runs partly through investors, not only customers. @cheong2021 find that
advertising lowers **stock-price synchronicity**—the degree to which a firm's
return moves in lockstep with its industry—implying that advertising conveys
firm-specific information to investors and is valued for that, over and above its
demand effects. Firm-generated social content has measurable, security-level
consequences: @lacka2021 define **price impact** as the effect on the variance of a
stock's price and estimate the permanent and temporary price impacts of S&P 500 IT
firms' tweets, finding that tweets carrying *both* valence and subject matter about
consumer or competitor orientation produce *permanent* price impact, whereas tweets
carrying only one attribute move prices only temporarily—and negative-valence tweets
about competitors generate the largest permanent impacts. The broader case that
marketing creates measurable firm value, and how to assess its effectiveness and
efficiency, is made by @hanssens2016, and the meta-analytic synthesis of marketing's
firm-value effects across studies is provided by @edeling2016. The general framing
of performance outcomes in marketing—and the multiple, sometimes conflicting metrics
managers must reconcile—is laid out by @katsikeas2016, and the broad correspondence
between marketing and finance assets-to-value logic traces back to @commenta2009 and
@Srivastava_1998.

A complementary result concerns the *human-capital* side of the interface.
@anderson2018 show that improving both marketing and finance skills raises profits,
but through different pathways: marketing/sales skill lifts profit via a
**growth** focus (higher sales, more product and employee investment), while
finance/accounting skill lifts it via an **efficiency/cost** focus. The managerial
implication is a fit argument—marketing skill is the better fit for startups chasing
growth, finance skill for mature firms optimizing cost.

## Adjacent Constructs: Agility and Complexity

Two firm-level constructs increasingly appear as moderators or controls in
interface studies, and both are measured from mandatory disclosures, which makes
them attractive instruments.

**Corporate agility**—a firm's ability to adapt to environmental change—is
hypothesized to raise survival rates [@lehn2021]. Because firms must report
accurately to the SEC, the measure built from disclosures is reliable; @lehn2021
operationalize agility as the sensitivity of a firm's competitive responses to
rivals' threats—specifically, the sensitivity of its product similarity (or
dissimilarity) to rivals' products to the rivals' own similarity to the firm—and
distinguish agility from mere flexibility.

**Firm complexity** is captured by **accounting reporting complexity** (ARC), "the
difficulty to understand, prepare, audit, and analyze the financial reports,"
operationalized as the count of accounting items disclosed in XBRL 10-K filings
[@hoitash2017, p. 262]. ARC is inversely associated with financial-reporting
quality and positively associated with audit delay and audit fees, and—because it is
built from the focal firm's own line items rather than from text that may reference
*other* firms—it captures firm complexity more cleanly than dictionary-based
linguistic measures [@hoitash2022; @loughran2020; @loughran2020a]. Complementary
complexity proxies include operating-segment counts, foreign operations, and the
linguistic readability of filings via the Fog Index [@gunning1952technique] and
10-K length. Complexity has downstream consequences: as it rises, analysts struggle
to forecast accurately [@hoitash2021], and corporate social responsibility is
positively correlated with greater ARC [@garcia2020].

## Frontier Topics in Corporate Finance for Marketing

The interface continues to expand into the financing events that bracket a firm's
public life, where marketing assets act as quality signals to capital providers.

**Venture capital.** Early intellectual-property signals matter for seed funding:
@rieger2024express show, on a Crunchbase–USPTO dataset of 5,370 ventures, that
ventures filing trademark applications are more likely to secure VC funding, with the
effect strongest in the first 100 days, in low-technological-uncertainty industries,
and in non-clustered locations—evidence that trademarks are early credible signals of
venture quality.

**Initial public offerings.** Firms can use **innovation potential** as a credible
quality signal at the IPO: @cao2022express find it positively associated with the
IPO's initial value and first-day returns and negatively associated with insider
selling, with patents weighing most on insider sales and pre-announcements weighing
most on first-day returns. The modern IPO landscape has also shifted—firms emphasize
growth metrics (such as user counts) over traditional financials and go public
roughly twice as old as in the 1980s, amid a larger retail-investor base, which has
prompted proposals for triggered disclosure.

**Customer-based corporate valuation.** When a firm's value derives transparently
from its subscriber base, the customer franchise can be valued directly:
@mccarthy2017 use DISH Network and Sirius XM data to estimate firm value from
customer-level acquisition, retention, and spend dynamics—bringing the market-based-
asset logic full circle by valuing the firm *from its customers up*.

**Initial coin offerings.** A nascent literature studies token sales as a financing
mechanism, supported by emerging data infrastructure [@czaja2021; @lyandres2022;
@hsieh2021; @momtaz2020; @belitski2021; @campino2022].

## Key Takeaways

- The marketing–finance interface treats brands, customers, channels, and market
  knowledge as **market-based assets** that pass the resource-based tests and create
  shareholder value through four levers of the discounted-cash-flow identity:
  accelerating, augmenting, de-risking, and raising the residual value of cash flows
  [@Srivastava_1998].
- Under near-semi-strong market efficiency, the *change* in stock price at the
  moment marketing news arrives is a risk-adjusted estimate of its value, which is
  what the **event study** and stock-return response model exploit
  (@eq-mf-return).
- **Persistence/VAR** models separate temporary from permanent effects of marketing
  shocks; **Tobin's $q$ regressions** measure the cross-sectional value of marketing
  assets but require careful identification.
- Accounting-based proxies for $q$ (**AATQ**) are biased toward overstating
  market-based-asset effects; prefer a **total $q$** that books intangible capital
  into the denominator at replacement cost [@peters2017; @bendle2018misuse].
- In M&A and alliances, *both* resource similarity and complementarity create
  value—the merger's **motive**, captured by strategic-emphasis alignment, decides
  which logic applies [@swaminathan2008; @swaminathan2009a].

## Further Reading

The foundational statement of the market-based-assets framework is
@Srivastava_1998, with the productivity-chain elaboration in @Rust_2004; the
methodological review of stock-market approaches to marketing is @srinivasan2009
and the meta-analytic synthesis is @edeling2016. For Tobin's-$q$ measurement,
read @peters2017 alongside the cautionary @bendle2018misuse. The M&A and alliance
applications are best entered through @swaminathan2008 and @swaminathan2009a.
