# Marketing–Finance Interface {#sec-marketing-finance}

Marketing spends real money—on advertising, salesforces, distribution, brands,
and customer relationships—and the people who allocate capital eventually ask
what those expenditures are worth. The *marketing–finance interface* is the body
of theory and empirical method that answers this question by linking marketing
actions to firm value: to cash flows, to risk, and ultimately to the price an
equity market sets on the firm. Its central claim is deceptively simple. The
things marketing builds—brand equity, a satisfied and loyal customer base,
channel and partner relationships—are *assets*, and like any asset they should be
visible in the value of the firm even though accounting rules largely keep them
off the balance sheet [@Srivastava_1998; @srinivasan2009].

Two intellectual currents meet here. From finance comes the discipline of
discounted cash flow and the apparatus of asset pricing: a firm is worth the
present value of its future cash flows, and a stock's return reflects the
revision of investors' expectations about those flows. From marketing comes a
theory of *where* those cash flows originate—in customers and the relationships,
knowledge, and reputation that produce repeat purchase, price premiums, and
lower acquisition costs. The interface insists that the two be made consistent:
actions that create value in the product market should, under an efficient
capital market, create value for shareholders [@srinivasan2009]. When they do
not, either the marketing action destroyed value or the market has not yet priced
it—both empirically interesting outcomes.

By the end of this chapter the reader should be able to state the market-based
assets framework formally, write down and estimate the workhorse models that
connect marketing to firm value (stock-return response models, persistence/VAR
models, and Tobin's $q$ regressions), reason carefully about what identifies a
causal effect in each, and recognize the measurement traps—most acutely in
accounting-based proxies for $q$—that have tripped up the literature. We treat
the event-study machinery for discrete marketing events (acquisitions, alliances,
product introductions) as a running application, since mergers and acquisitions
and strategic alliances are where the interface has produced its deepest body of
evidence, and a recent voluntary-commitment study (@sec-mf-carbon) shows that the
same machinery still carries top-journal contributions. The branding chapter
(@sec-branding) develops the event study and
stochastic-frontier capability estimators in the specific context of brand
assets; here we build the general scaffolding.

## Market Efficiency and the Logic of Valuation

Everything downstream rests on an assumption about how stock prices incorporate
information, so it is worth stating precisely. The *efficient market hypothesis*
classifies markets by the information set already impounded in prices. Under
**weak-form** efficiency prices reflect the history of prices themselves; under
**semi-strong** efficiency they reflect all public information in real time; under
**strong-form** efficiency they reflect public *and* private information. The
working consensus in the marketing–finance literature is that real markets sit
between weak and semi-strong form: public information about marketing—an earnings
release that reveals advertising's payoff, a product-recall announcement, an
alliance—moves prices, and it does so promptly though not always completely
[@srinivasan2009].

This has a sharp empirical implication. If markets are at least near-semi-strong
efficient, then good news about marketing raises the stock price and bad news
lowers it, and—because the price is forward-looking—the *change* in price at the
moment news arrives is an unbiased, risk-adjusted estimate of the news's value to
shareholders. Stock-market valuation is then in sync with product-market
valuation: the same actions that build value with customers build value for
investors. This equivalence is what licenses the use of stock returns to value
marketing, and it is the premise the event study exploits.

Formally, let $V_{it}$ be the market value of firm $i$'s equity at the end of
period $t$. The dividend-discount identity writes it as the expected present
value of future cash flows to equity,

$$
V_{it} = \sum_{T=t}^{\infty}
  \left(\frac{1}{1+r_{it}}\right)^{T-t}\,
  \mathbb{E}_t\!\left[\mathrm{CF}_{iT}\right],
$$ {#eq-mf-ddm}

where $r_{it}$ is the firm's cost of equity capital and $\mathbb{E}_t[\cdot]$ is
the expectation conditional on information available at $t$. Marketing enters
through the cash-flow expectations. @Srivastava_1998 organize marketing's value
contribution into four channels visible in @eq-mf-ddm: marketing can
**accelerate** cash flows (bring them forward in time), **augment** their
**level**, **reduce their risk** (volatility and vulnerability, which lowers
$r_{it}$), and raise their **residual value** (the terminal value beyond the
explicit forecast horizon). The rest of the framework is an elaboration of how
marketing assets operate through these four levers.

## Market-Based Assets

### Definition and the Resource-Based Test

The conceptual core of the interface is the **market-based asset**. The following
landmark definition frames everything that follows.

> Market-based assets "arise from the commingling of the firm with entities in
> its external environment," where an asset is "any physical, organizational, or
> human attribute that enables the firm to generate and implement strategies that
> improve its efficiency and effectiveness in the marketplace"
> [@Srivastava_1998, pp. 2–3].

The defining feature is *externality of locus*: unlike a factory or a patent, a
market-based asset lives in a relationship or in knowledge that spans the firm's
boundary. @Srivastava_1998 distinguish two kinds. **Relational** market-based
assets are outcomes of the firm's relationships with external stakeholders—
distributors, retailers, end customers, strategic partners. Brand equity is the
relational asset between a firm and its customers; channel equity is the
relational asset between a firm and its intermediaries. **Intellectual**
market-based assets are the firm's knowledge about its environment: the emerging
and potential states of market conditions and of the entities (competitors,
customers, channels, suppliers, regulators) within them.

For a resource to create *sustained* value it must pass the resource-based view's
four tests: it must be **convertible** (the firm can actually exploit it), **rare**,
**imperfectly imitable**, and have **no perfect substitutes** [@Srivastava_1998].
Knowledge and relationships pass all four, which is precisely why they can support
a durable competitive advantage where tangible assets—easily bought, copied, or
substituted—cannot. Each asset has a **stock** dimension (the amount of brand
equity or customer knowledge the firm possesses at a point in time) and a **flow**
dimension (the rate at which that stock is being augmented or is decaying)
[@Srivastava_1998, p. 5]. The stock–flow distinction matters for estimation:
marketing expenditures are flows that accumulate, with decay, into the stock that
produces cash flow—a structure we exploit in the persistence and perpetual-
inventory models below.

### From Assets to Shareholder Value

The framework's payoff is a causal chain from marketing actions, through
market-based assets and intermediate product-market outcomes, to the four
shareholder-value levers of @eq-mf-ddm. @fig-mba-chain renders the chain;
it is our own redrawing of the logic in @Srivastava_1998 and the "chain of
marketing productivity" in @Rust_2004, neither reproduced from the originals.

```{mermaid}
%%| label: fig-mba-chain
%%| fig-cap: "The marketing–finance value chain. Marketing actions build market-based assets, which generate product-market outcomes, which map onto the four shareholder-value levers of the discounted-cash-flow identity. Synthesized from @Srivastava_1998 and @Rust_2004."
flowchart LR
  A["Marketing actions<br/>(advertising, innovation,<br/>relationship investment)"]
    --> B["Market-based assets<br/>Relational: brands, customers,<br/>channels, partners<br/>Intellectual: market knowledge"]
  B --> C["Product-market outcomes<br/>faster penetration · price premium<br/>share premium · extensions<br/>lower selling/service cost<br/>loyalty / retention"]
  C --> D1["Accelerate cash flows"]
  C --> D2["Augment level of cash flows"]
  C --> D3["Reduce volatility &<br/>vulnerability of cash flows"]
  C --> D4["Enhance residual value"]
  D1 --> E["Shareholder value"]
  D2 --> E
  D3 --> E
  D4 --> E
```

The mechanisms behind each arrow are worth making explicit, because they convert
soft marketing language into the four hard financial levers.

**Accelerating cash flows.** Greater brand awareness speeds consumer response to
advertising and shortens the path to trial [@keller1993], and a large installed
base together with partnership networks compresses the product life cycle by
speeding the adoption process [@robertson1993reduce]. Time-to-market acceptance
is a cash-flow timing variable, and—because near cash is worth more than distant
cash—accelerating it raises value even when the *level* of cumulative cash flow is
unchanged.

**Augmenting the level of cash flows.** Market-based assets raise revenues (brand
extensions open new categories), lower costs (extensions and the relationship's
advertising efficiencies cut unit selling cost), and reduce working- and
fixed-capital requirements (relationship marketing and cooperative ventures—co-
branding and co-marketing alliances—let partners leverage each other's networks
and customer bases rather than build their own).

**Reducing risk.** The *vulnerability* of cash flows falls with customer
satisfaction, loyalty, and retention, and stability rises as information sharing,
automatic replenishment, and lower inventories smooth operations. Higher switching
costs and operational integration raise the net present value of the cash-flow
stream by lowering its discount rate $r_{it}$.

**Enhancing residual value.** A larger, more loyal, higher-quality customer base
buys more and erects barriers to competitive entry, raising the terminal value
that dominates @eq-mf-ddm for long-lived firms.

The four levers are separable in the framework, but a single strategic choice need
not move them in the same direction, and the framework is most useful where it
forces that conflict into the open. @han2026fundraising supply a clean case from
outside the corporate setting: concentrating a nonprofit's revenue across fewer
solicitation methods raised both the level and the growth rate of fundraising
revenue while *lowering* its stability, measured as the deviation of realized
revenue from its predicted value. In the vocabulary of @fig-mba-chain, the same
decision augmented the level of cash flows and raised their volatility. A valuation
model that discounts the augmented stream at an unchanged $r_{it}$ will therefore
overstate the gain, and the direction of the net effect on value is an empirical
question rather than a corollary of the level result. @sec-fundraising-tradeoff
develops the case in full.

@Srivastava_1998 summarize the logic in three propositions: the more market-based
assets external entities help develop, the more eager and valuable those entities
become to the firm; the more such assets pass the resource-based tests, the more
external entities benefit; and when a firm leverages market-based assets to boost
cash flows, shareholder value is created. The empirical program of the interface
is, in effect, the testing of this third proposition.

### Why Not Just Use Accounting Numbers?

A natural objection is that firms already report performance, so why build a
parallel valuation apparatus. The answer is that the obvious accounting measures
are poor proxies for value creation. Book value records assets at historical cost
and ignores intangibles entirely; replacement value is conceptually better but
practically unmeasurable for relationships and knowledge; and price–earnings
multiples inherit the well-known pathologies of accrual accounting—earnings are
backward-looking, unadjusted for risk, and manipulable through accruals timing.
The *shareholder-value* approach—discounted future cash flow, as in
@eq-mf-ddm—dominates these because it is forward-looking, risk-adjusted, and
harder to game [@Srivastava_1998]. This is also the principal reason cash flow,
rather than earnings, is the preferred firm-performance variable in interface
research, and it motivates the use of $q$-type and stock-return measures in place
of profitability ratios.

## Measuring Strategic Emphasis: Brand Focus versus Customer Focus {#sec-mf-focus}

The two relational market-based assets of @Srivastava_1998—brand equity and the customer
base—correspond to two managerial programs that the field has argued about for thirty
years. **Brand management** builds and defends a position in the minds of a market;
**customer management** builds and defends individual relationships and allocates
resources across them. The strategy literature has claimed that one is displacing the
other, that they are complements, and that they are substitutes competing for the same
budget. The reason the argument has stayed unsettled is measurement: firms do not report
which program they emphasize, and the usual proxies (advertising intensity, CRM spend,
the presence of a chief customer officer) are noisy, sporadically disclosed, and
confounded with industry.

@scholdra2026focus solve the measurement problem with text the firms produce anyway.
Earnings conference calls are quarterly, near-universal among listed firms, and contain
management's own account of what the business is doing; the transcripts stretch back two
decades. The authors run automated content analysis on more than 95,000 transcripts from
over 1,800 firms across 22 years, complemented by manual coding and secondary data, and
derive two separate indices: a **brand-management focus** and a **customer-management
focus**. The measurement approach follows the retail-sector precursor of @han2021capturing,
who validated the same construction on 853 retailer earnings calls: rather than
hand-writing a keyword list, the dictionaries are induced from a training library of brand
management and customer management textbooks, yielding *bigram* dictionaries whose matches
against a transcript produce a focus index for each dimension.

The substantive findings are worth stating precisely, because each has a different
implication for practice.

- **The two foci are essentially uncorrelated.** Firms that talk more about brands do not
  systematically talk less about customers, or more. There is neither a general trade-off
  nor a general complementarity—which is why prior work, sampling different industries and
  eras, kept reaching opposite conclusions.
- **The patterns are stable across industries and business types.** The dimensions
  therefore behave like properties of a firm's strategy rather than artifacts of sector
  vocabulary, which is what licenses cross-industry benchmarking.
- **The two foci load on different parts of the P&L.** They have distinct effects on
  revenue-related and cost-related performance outcomes, so a single "marketing emphasis"
  score would average away the mechanism.
- **Dual emphasis underperforms.** Emphasizing both is *less* attractive than committing to
  one—an attention-and-capability result, not an accounting one, and the finding that most
  directly contradicts the "do both, they reinforce each other" prior.
- **Environmental moderators matter.** The effectiveness of each focus depends on
  conditions the firm does not choose, so the correct question is not which focus is better
  but which is better *here*.

### Building and Monitoring the Indices {#sec-mf-focus-code}

The published study uses commercial transcript data (licensed from financial-data vendors)
that cannot be redistributed, and we located no public replication package, so the code
below reproduces the *method* rather than the authors' estimates: it builds bigram
dictionaries, scores a corpus, benchmarks each firm against its industry–year peers, and
tests the dual-emphasis claim. Any firm with an archive of its own calls can run the same
pipeline on real text by substituting the corpus and expanding the dictionaries.

```{r mf-focus-score, message=FALSE, warning=FALSE}
set.seed(21)

# --- Step 1: bigram dictionaries induced from a training library ---------------
# In the paper these come from brand- and customer-management textbooks; here we
# use a short illustrative subset. Dictionaries are BIGRAMS, not single words,
# because "customer value" and "brand value" share a token but not a construct.
dict_brand <- c("brand equity", "brand awareness", "brand positioning",
                "brand portfolio", "brand image", "advertising campaign",
                "brand loyalty", "brand extension")
dict_cust  <- c("customer retention", "customer lifetime", "loyalty program",
                "customer segment", "churn rate", "customer acquisition",
                "cross selling", "customer satisfaction")

# --- Step 2: a synthetic corpus with KNOWN latent foci -------------------------
# Each firm-year has latent brand focus b and customer focus c, drawn
# independently (the paper's finding: the two dimensions are uncorrelated).
n_firm <- 120; n_year <- 12
panel <- expand.grid(firm = 1:n_firm, year = 1:n_year)
firm_b <- rnorm(n_firm, 0, 1); firm_c <- rnorm(n_firm, 0, 1)
panel$b <- firm_b[panel$firm] + rnorm(nrow(panel), 0, 0.4)
panel$c <- firm_c[panel$firm] + rnorm(nrow(panel), 0, 0.4)

make_call <- function(b, c, n_bigrams = 600) {
  # Probability a given bigram slot is a brand (customer) phrase rises with the
  # latent focus; everything else is filler, as in a real transcript. Dictionary
  # hits are sparse (about 20 per call), which is what makes the index noisy.
  p_b <- plogis(-3.6 + 0.6 * b); p_c <- plogis(-3.6 + 0.6 * c)
  kind <- sample(c("b", "c", "filler"), n_bigrams, replace = TRUE,
                 prob = c(p_b, p_c, max(1e-6, 1 - p_b - p_c)))
  c(sample(dict_brand, sum(kind == "b"), replace = TRUE),
    sample(dict_cust,  sum(kind == "c"), replace = TRUE),
    rep("filler phrase", sum(kind == "filler")))
}

# --- Step 3: score each call -> focus index per 1,000 bigrams ------------------
score_call <- function(bigrams) {
  c(brand    = 1000 * sum(bigrams %in% dict_brand) / length(bigrams),
    customer = 1000 * sum(bigrams %in% dict_cust)  / length(bigrams))
}

scores <- t(mapply(function(b, c) score_call(make_call(b, c)),
                   panel$b, panel$c))
panel$bf <- scores[, "brand"]
panel$cf <- scores[, "customer"]

cat(sprintf("Recovery: cor(latent brand, brand index)    = %.2f\n",
            cor(panel$b, panel$bf)))
cat(sprintf("           cor(latent customer, cust. index) = %.2f\n",
            cor(panel$c, panel$cf)))
cat(sprintf("Measured foci are uncorrelated with each other: r = %.2f\n",
            cor(panel$bf, panel$cf)))
```

Normalization matters more than it looks. Dividing by transcript length is not cosmetic:
calls got longer over the sample period, so an unnormalized count would manufacture a
secular rise in both foci. For *monitoring*, the useful object is not the raw index but
the firm's position relative to its industry–year peers, which nets out both the
vocabulary of the sector and any year in which everyone talked more about everything.

```{r mf-focus-monitor, message=FALSE, warning=FALSE}
# Peer-relative focus: demean within year (in practice, within industry x year).
panel$bf_rel <- panel$bf - ave(panel$bf, panel$year)
panel$cf_rel <- panel$cf - ave(panel$cf, panel$year)

# A monitoring view for one firm: where does it sit, and is it drifting?
focal <- subset(panel, firm == 7)
round(head(focal[, c("year", "bf", "cf", "bf_rel", "cf_rel")], 4), 2)
```

```{r}
#| label: fig-mf-focus-map
#| fig-cap: "A strategic-emphasis map. Each point is a firm-year positioned by its peer-relative brand focus and customer focus; the tracked firm's path (dark) shows drift over time. The two dimensions are near-orthogonal, so a firm's position is not pinned down by a single 'marketing emphasis' score."
#| message: false
#| warning: false
plot(panel$bf_rel, panel$cf_rel, pch = 16, col = "grey80", cex = 0.6,
     xlab = "Brand focus (peer-relative)", ylab = "Customer focus (peer-relative)",
     main = "Strategic-emphasis map")
abline(h = 0, v = 0, col = "grey50", lty = 3)
lines(focal$bf_rel, focal$cf_rel, type = "o", pch = 16, col = "black", cex = 0.8)
```

### Testing the Dual-Emphasis Claim, and Why It Is Hard {#sec-mf-focus-dual}

The claim that doing both is worse than committing to one is an *interaction* claim: with
performance $y_{jt}$ for firm $j$ in year $t$,
$$
y_{jt} = \alpha_j + \lambda_t + \beta_B \, \text{BF}_{jt} + \beta_C \, \text{CF}_{jt}
+ \beta_{BC} \, \text{BF}_{jt}\!\times\!\text{CF}_{jt} + \varepsilon_{jt},
$$ {#eq-mf-dual-emphasis}

and the dual-emphasis penalty is $\beta_{BC} < 0$. Interactions are the most fragile
coefficient in the model to measurement error, and a dictionary index is a noisy measure
by construction. The following simulation makes the fragility quantitative: it generates
performance from the *latent* foci with a real penalty, then estimates
@eq-mf-dual-emphasis using the latent values and using the measured indices.

```{r mf-focus-dual, message=FALSE, warning=FALSE}
# True DGP: both foci help; doing both is penalized (beta_BC < 0).
panel$y <- 0.40 * panel$b + 0.35 * panel$c - 0.25 * panel$b * panel$c +
           rnorm(nrow(panel), 0, 0.8)

# Standardize BOTH the latent constructs and the measured indices, so the two
# regressions are on the same scale and the only difference is measurement error.
z <- function(x) as.numeric(scale(x))
panel$b_z  <- z(panel$b);  panel$c_z  <- z(panel$c)
panel$bf_z <- z(panel$bf); panel$cf_z <- z(panel$cf)

fit_latent   <- lm(y ~ b_z  * c_z  + factor(year), data = panel)
fit_measured <- lm(y ~ bf_z * cf_z + factor(year), data = panel)

comparison <- data.frame(
  term      = c("brand focus", "customer focus", "interaction"),
  latent    = round(coef(fit_latent)[c("b_z", "c_z", "b_z:c_z")], 3),
  measured  = round(coef(fit_measured)[c("bf_z", "cf_z", "bf_z:cf_z")], 3),
  row.names = NULL)
comparison$attenuation <- round(1 - comparison$measured / comparison$latent, 2)
comparison
```

The main effects survive the transition from latent construct to dictionary index with
attenuation of roughly a sixth to a quarter; the interaction loses about 40% of its
magnitude, because the product of two noisy measures carries roughly the product of their
reliabilities. Two practical implications follow.
A study that fails to find a dual-emphasis penalty with a short dictionary has not
established its absence—it may have measured it away, which is why the published design
invests in a textbook-derived dictionary and a corpus of 95,000 transcripts rather than a
convenience sample. And a firm monitoring its own emphasis should expect the *level* of
each focus to be more reliably estimated than any interaction between them.

Three validity questions belong on any use of this measure, and they generalize to every
"strategy from corporate text" design. First, **talk versus action**: an earnings call is
a disclosure to investors, so the index measures emphasis as management chooses to present
it. This is a genuine limitation and also part of the construct—what a CEO commits to
publicly is costly to reverse. Second, **who is speaking**: prepared remarks and the
analyst Q&A are different data-generating processes, and analyst questions can inject
vocabulary management did not choose. Third, **dictionary drift**: over 22 years the
language of both disciplines changed, so a fixed dictionary risks measuring era rather
than emphasis; validation against manually coded calls from different periods is the check.
@sec-text-as-data develops the general machinery, and @sec-tad-pitfalls the general
warnings; the point here is that the marketing–finance interface now has a firm-level,
longitudinal, comparable measure of *what the firm is actually trying to do*, which is the
input that regressions of value on strategy have always lacked.

## Methods for Connecting Marketing to Firm Value

Three broad estimation strategies dominate, distinguished by the nature of the
marketing event and the time structure of the data. @tbl-mf-methods contrasts
them; the subsections that follow develop the two most important.

| Approach | What it measures | Core model | Identifying assumption | Breaks when |
|---|---|---|---|---|
| **Event study / stock-return response** | Value of a discrete, dated event | Market-model abnormal returns $\mathrm{AR}_{it}$, cumulated to $\mathrm{CAR}_i$ | Market is semi-strong efficient; event date is exogenous and unconfounded within the window | Information leaks before the date; confounding events share the window; the event is anticipated |
| **Persistence modeling (VAR)** | Long-run cumulative impact of a marketing shock, separating temporary from permanent effects | Vector autoregression with impulse-response and unit-root analysis | Shocks are correctly identified (ordering/IRF); the system is stable or has a well-defined permanent component | Wrong variable ordering; structural breaks; omitted feedback channels |
| **Tobin's $q$ / valuation regression** | Cross-sectional association of marketing assets with firm value | $q_{it} = \mathbf{x}_{it}^{\top}\boldsymbol{\beta}+\alpha_i+\delta_t+u_{it}$ | Marketing regressor is exogenous conditional on controls and fixed effects | Reverse causality; omitted intangibles; accounting proxy bias (see @sec-mf-aatq); and, more fundamentally, average $Q$ is not the marginal $q$ the theory names (see @sec-mf-marginal-vs-average) |

: Three workhorse approaches to the marketing–finance interface, with their identifying assumptions. {#tbl-mf-methods}

A fourth, the single-equation **error-correction model** (ECM), is used when the
data are nonstationary and one wants to separate a long-run equilibrium relation
from short-run dynamics; it recognizes the random-walk character of stock prices
and is a staple of the persistence tradition [@srinivasan2009a].

### Stock-Return Response Modeling

The cleanest way to value a discrete marketing event is to read the value
directly off the stock price, exploiting market efficiency. Begin again from the
market-value identity and difference it across one period. With $\mathrm{Eret}_{it}$
the expected (required) return on the equity and $r_{it}$ the discount rate,

$$
V_{it} = (1+\mathrm{Eret}_{it})\,V_{it-1}
  + \sum_{T=t}^{\infty}
    \left(\frac{1}{1+r_{it}}\right)^{T-t}
    \Delta\,\mathbb{E}_t\!\left[\mathrm{CF}_{iT}\right],
$$ {#eq-mf-valdiff}

where $\Delta\,\mathbb{E}_t[\mathrm{CF}_{iT}] \equiv
\mathbb{E}_t[\mathrm{CF}_{iT}] - \mathbb{E}_{t-1}[\mathrm{CF}_{iT}]$ is the
*revision* in the expected cash flow for date $T$ that arrives during period $t$.
The first term is the value the firm would have had absent any news—it grows at the
required return—and the second term is the capitalized value of the news itself.
Rearranging gives the realized stock return as an expected component plus an
abnormal component driven entirely by cash-flow-expectation revisions:

$$
\mathrm{StockReturn}_{it}
  = \frac{V_{it}-V_{it-1}}{V_{it-1}}
  = \mathrm{Eret}_{it}
  + \underbrace{\frac{1}{V_{it-1}}
      \sum_{T=t}^{\infty}\left(\frac{1}{1+r_{it}}\right)^{T-t}
      \Delta\,\mathbb{E}_t\!\left[\mathrm{CF}_{iT}\right]}_{\text{abnormal return }\;\mathrm{AR}_{it}}.
$$ {#eq-mf-return}

@eq-mf-return is the theoretical justification for the event study. The
**abnormal return** $\mathrm{AR}_{it}$ is the realized return net of its
expectation, and under market efficiency it equals the present value of the
cash-flow news, scaled by beginning-of-period value. To take it to data one needs
a model for the expected return $\mathrm{Eret}_{it}$. The standard choice is the
*market model*, a single-factor regression of the firm's return on the market
return estimated over a clean pre-event window $[t_0, t_1]$:

$$
R_{it} = \alpha_i + \beta_i R_{mt} + \varepsilon_{it},
\qquad
\widehat{\mathrm{AR}}_{it} = R_{it} - \hat\alpha_i - \hat\beta_i R_{mt}.
$$ {#eq-mf-marketmodel}

Cumulating over the event window $W$ gives the cumulative abnormal return
$\mathrm{CAR}_i = \sum_{t \in W}\widehat{\mathrm{AR}}_{it}$, which is then regressed
cross-sectionally on event characteristics to learn *what* drives value. The
branding chapter (@sec-branding) gives the full estimator and the standard
confound screens; the key identifying requirements are that the event date be
exogenous and that no other value-relevant news (earnings, splits, executive
turnover, buybacks, dividend changes) contaminate the window. The following code
simulates the procedure end to end so the reader can see exactly where each
quantity comes from.

```{r mf-event-study, message=FALSE, warning=FALSE}
set.seed(21)

# --- simulate a market and one firm with a known event-day jump --------------
n_est  <- 250          # estimation-window trading days
n_evt  <- 5            # event-window days (e.g., -2..+2)
alpha  <- 0.0002; beta <- 1.10
mkt_est <- rnorm(n_est, 0.0004, 0.009)
firm_est <- alpha + beta * mkt_est + rnorm(n_est, 0, 0.012)

# --- estimate the market model on the clean pre-event window -----------------
fit <- lm(firm_est ~ mkt_est)

# --- event window: true cash-flow news adds +3% on the announcement day ------
mkt_evt   <- rnorm(n_evt, 0.0004, 0.009)
news      <- c(0, 0, 0.03, 0, 0)            # day 0 carries the marketing news
firm_evt  <- coef(fit)[1] + coef(fit)[2] * mkt_evt +
             rnorm(n_evt, 0, 0.012) + news

# --- abnormal and cumulative abnormal returns --------------------------------
expected_evt <- predict(fit, newdata = data.frame(mkt_est = mkt_evt))
AR  <- firm_evt - expected_evt
CAR <- cumsum(AR)

data.frame(day = -2:2, AR = round(AR, 4), CAR = round(CAR, 4))
```

The recovered cumulative abnormal return concentrates on day 0 and is close to the
3% shock injected, illustrating why—when the efficiency and no-confound assumptions
hold—the event study delivers a clean, risk-adjusted dollar value for a marketing
event.

### Persistence Modeling

Many marketing effects are not one-day jumps but dynamic responses that build and
decay over months: an advertising pulse raises awareness, which raises sales, which
feeds back into the budget. *Persistence modeling* uses a **vector autoregression**
(VAR) to trace the full dynamic response of firm value (or sales, or cash flow) to
a marketing shock and, crucially, to separate the *temporary* component (which
dies out) from the *permanent* component (which is impounded forever)
[@srinivasan2009]. For a vector $\mathbf{y}_t$ stacking, say, log firm value, a
marketing variable, and a control,

$$
\mathbf{y}_t = \mathbf{c} + \sum_{\ell=1}^{p}
  \mathbf{\Phi}_\ell\,\mathbf{y}_{t-\ell} + \boldsymbol{\epsilon}_t,
\qquad \boldsymbol{\epsilon}_t \sim (\mathbf{0}, \mathbf{\Sigma}),
$$ {#eq-mf-var}

the **impulse-response function** $\partial \mathbf{y}_{t+h}/\partial
\boldsymbol{\epsilon}_t$ gives the effect of a unit shock $h$ periods out, and its
long-run sum measures total persistent impact. Identification of structural shocks
requires an ordering (or sign/long-run restrictions); the standard recursive
ordering assumes a Cholesky causal chain that the researcher must defend. When the
series are nonstationary—stock prices follow a near random walk—the analyst works
with the unit-root/cointegration structure directly, typically via an
error-correction representation that separates the long-run equilibrium from
short-run adjustment [@srinivasan2009a]. The two pitfalls to flag are that the
permanent/temporary decomposition is only as good as the unit-root inference behind
it, and that an omitted feedback channel (e.g., competitor response) biases the
impulse responses.

## Tobin's $q$ and Valuation Regressions {#sec-mf-q}

The most common cross-sectional measure of firm value in interface research is
**Tobin's $q$**, the ratio of a firm's market value to the replacement cost of its
assets. Intuitively, $q>1$ means the market values the firm above what it would
cost to rebuild its assets, and the gap is attributed to intangibles and growth
options—exactly the territory marketing claims. Its appeal is that it is
forward-looking, risk-adjusted, and less manipulable than accounting profitability,
which is why it is the dependent variable of choice in studies of how marketing
assets relate to firm value [@Grewal_2010; @McAlister_2016; @Morgan_2009;
@Wies_2019]. When probing the drivers of $q$, the literature recommends controlling
at minimum for **financial leverage** and **cash flow**, since both shift $q$ for
reasons unrelated to marketing [@Wies_2019].

The traditional empirical proxy assembles $q$ from financial statements
[@Rao_2004]:

$$
q = \frac{\mathrm{MVE} + \mathrm{PS} + \mathrm{DEBT}}{\mathrm{TA}},
$$ {#eq-mf-q-classic}

where $\mathrm{MVE}$ is the market value of equity (share price times shares
outstanding), $\mathrm{PS}$ the liquidating value of preferred stock,
$\mathrm{DEBT}$ is short-term liabilities net of short-term assets plus the book
value of long-term debt, and $\mathrm{TA}$ the book value of total assets. Because
both numerator and denominator scale with firm size, $q$ is scale-independent and
thus comparable as a measure of *relative* market performance across firms of
different sizes.

A typical valuation regression then takes the form

$$
q_{it} = \mathbf{x}_{it}^{\top}\boldsymbol{\beta}
  + \alpha_i + \delta_t + u_{it},
$$ {#eq-mf-qreg}

with $\mathbf{x}_{it}$ the marketing variables of interest plus leverage and cash
flow, $\alpha_i$ firm fixed effects absorbing time-invalid heterogeneity, and
$\delta_t$ time effects absorbing common shocks. Identification of a *causal*
marketing effect requires that, conditional on the controls and fixed effects, the
marketing regressor be uncorrelated with $u_{it}$. This is demanding: reverse
causality (valuable firms spend more on marketing), omitted intangibles (R&D or
organizational capital correlated with both), and measurement error in the
intangible-laden denominator all threaten it. But there is a deeper problem than
any of these, one the marketing literature has largely not absorbed: the left-hand
side of @eq-mf-qreg is not the quantity investment theory names. The subsections
that follow work through it in order—what $q$ was supposed to be, what belongs in
the denominator, what measurement error does to the slope, how the theoretical
object can now be estimated directly, and what all of it implies for a marketing
valuation regression.

### What $q$ Was Supposed to Be: Marginal versus Average {#sec-mf-marginal-vs-average}

Before repairing the denominator it is worth being precise about what
@eq-mf-q-classic is a proxy *for*, because it is not the object investment theory
cares about. The theory's $q$ is **marginal**: the shadow value of one additional
unit of installed capital,

$$
q^{\mathrm{mar}}_{it} \;=\; \frac{\partial V_{it}}{\partial K_{it}},
$$ {#eq-mf-marginalq}

which is a sufficient statistic for investment—under convex adjustment costs the
firm installs capital until the marginal cost of installation equals
$q^{\mathrm{mar}}$. What the researcher computes is **average** $Q$: total market
value divided by total replacement cost. @hayashi1982 established exactly when the
two coincide. The firm must be a price taker, its production and adjustment-cost
technologies must be linearly homogeneous in capital, and capital must be a single
homogeneous good installed without fixed costs. Under those conditions the value
function is linear in capital, so the average ratio equals the derivative and the
observable proxy *is* the theoretical object.

Almost every development in the $q$ literature since is a relaxation of one
Hayashi condition, and reading the literature that way turns a confusing pile of
competing "improved $q$ measures" into an ordered list (@fig-mf-q-map).

- **More than one capital good.** Firms install intangible as well as physical
  capital, and the two adjust at different costs. The fix is to put intangibles
  into the denominator at replacement cost: total $q$ [@peters2017].
- **Mismeasurement of the intended ratio.** Even the ratio one means to compute is
  observed with error—from replacement-cost and debt approximations, from noise in
  the market-value numerator, and from the capitalization parameters used to build
  intangible stocks [@lewellen1997; @erickson2000; @erickson2011a; @ewens2024].
- **Market power.** A firm earning rents collects inframarginal profit on capital
  it has already installed, so its value exceeds the marginal value of new capital
  times the stock, and average $Q$ overstates $q^{\mathrm{mar}}$ [@crouzet2023].
- **Non-convexities and idle capital.** When managers anticipate excess capacity,
  the marginal unit will be underutilized, so average $Q$ is a biased estimator of
  marginal $q$ even with no rents at all [@grullon2025].

```{mermaid}
%%| label: fig-mf-q-map
%%| fig-cap: "The $q$ literature as a set of relaxations of the @hayashi1982 equivalence conditions. Average $Q$ is a valid proxy for marginal $q$ only under all four conditions; each modern measure repairs the failure of one of them, and marketing assets are implicated in two at once."
flowchart TB
  H["Hayashi (1982) conditions<br/>price taking · linear homogeneity<br/>single capital good · no fixed costs"]
    --> EQ["Average Q = marginal q<br/>(the proxy is the object)"]
  H --> R1["Relaxed: many capital goods<br/>intangibles adjust differently"]
  H --> R2["Relaxed: ratio observed with error"]
  H --> R3["Relaxed: market power<br/>rents on installed capital"]
  H --> R4["Relaxed: non-convexity<br/>anticipated excess capacity"]
  R1 --> M1["Total q<br/>Peters & Taylor (2017)<br/>exit-price parameters:<br/>Ewens et al. (2024)"]
  R2 --> M2["Cleaned inputs: Lewellen & Badrinath (1997)<br/>Higher-moment GMM: Erickson & Whited<br/>Fitted Q: Gala et al. (2026)"]
  R3 --> M3["Q+ decomposition<br/>Crouzet & Eberly (2023)"]
  R4 --> M4["Capacity-adjusted q<br/>Grullon & Ikenberry (2025)"]
  M1 --> MQ["Marginal q estimated directly<br/>Gala, Gomes & Liu (2026)"]
  M2 --> MQ
  M3 --> MQ
  M4 --> MQ
  MQ --> USE["What a marketing valuation<br/>regression should be run on"]
  EQ --> USE
```

Marketing's stake in this taxonomy is unusually direct. Brand and customer assets
are simultaneously an *omitted intangible* and a *source of rents*, so they load on
two of the four wedges at once, and they do so in the same direction. That is the
structural reason the accounting critique in @sec-mf-aatq bites hardest exactly
where marketing most wants to make its case.

### Total $q$: Putting Intangibles in the Denominator

Equation @eq-mf-q-classic divides market value by *physical* assets at book value,
ignoring the intangible capital—knowledge and organizational/brand capital—that
the firm has built. @peters2017 argue that the neoclassical theory of investment
applies to intangibles just as it does to plant and equipment, and propose a
**total $q$** that puts intangible capital into the denominator at replacement
cost:

$$
\begin{aligned}
q^{\text{tot}}_{it}
  &= \frac{V_{it}}{K^{\text{phy}}_{it} + K^{\text{int}}_{it}} \\[2pt]
  &= \frac{\text{prcc\_f}\times\text{csho} + \text{dltt} + \text{dlc} - \text{act}}
          {\text{ppegt} + K^{\text{int}}_{it}},
\end{aligned}
$$ {#eq-mf-totalq}

where the market value $V_{it}$ is equity (price `prcc_f` times shares `csho`) plus
the book value of debt (`dltt` + `dlc`) minus current assets (`act`),
$K^{\text{phy}}_{it}$ is the replacement cost of physical capital (gross
property, plant, and equipment, `ppegt`), and $K^{\text{int}}_{it}$ is the
replacement cost of intangible capital. Earlier $q$ proxies omitted the intangible
term entirely [@fazzari1987; @erickson2011a], biasing $q$ upward for
intangible-intensive firms.

Intangible capital is itself accumulated by the **perpetual-inventory method**.
For internally created knowledge capital, the end-of-period stock evolves as

$$
G_{it} = (1-\delta_{\mathrm{RD}})\,G_{i,t-1} + \mathrm{RD}_{it},
$$ {#eq-mf-knowcap}

with $G_{i0}=0$, real R&D expenditure $\mathrm{RD}_{it}$ (treated as zero when
missing [@lev]), and an industry-specific depreciation rate $\delta_{\mathrm{RD}}$
(the BEA R&D rates, defaulting to 15% when unavailable [@li2018]). Organizational
and brand capital are built analogously from a fraction of past selling, general,
and administrative (SG&A) spending—which bundles advertising (brand), employee
training (human capital), and distribution systems. Intangible capital is costlier
to adjust than physical capital, a friction the neoclassical model takes seriously.
@fig-perpetual-inventory simulates the knowledge-capital recursion to make the
stock–flow logic concrete.

```{r}
#| label: fig-perpetual-inventory
#| fig-cap: "Knowledge capital as a depreciated stock of past R&D flows, simulated by the perpetual-inventory recursion."
#| message: false
#| warning: false
# Knowledge-capital accumulation by perpetual inventory (eq. mf-knowcap).
set.seed(21)
years      <- 1:15
rd_flow    <- 100 * (1.08)^(years - 1) * exp(rnorm(length(years), 0, 0.05))
delta_rd   <- 0.15                       # BEA-style default depreciation
G <- numeric(length(years)); G[1] <- rd_flow[1]
for (t in 2:length(years)) {
  G[t] <- (1 - delta_rd) * G[t - 1] + rd_flow[t]
}

plot(years, G, type = "o", pch = 16,
     xlab = "Year", ylab = "Replacement cost",
     main = "Knowledge capital as a depreciated stock of R&D flows",
     ylim = range(c(G, rd_flow)))
lines(years, rd_flow, type = "o", col = "steelblue", pch = 1)
legend("topleft", bty = "n",
       legend = c("Knowledge-capital stock G_t", "Annual R&D flow"),
       col = c("black", "steelblue"), pch = c(16, 1), lty = 1)
```

The stock $G_t$ rises well above any single year's flow because past investments
persist (depreciating at $\delta_{\mathrm{RD}}$); this is exactly the wedge that
@eq-mf-totalq restores to the denominator and that a naive physical-capital $q$
ignores.

The capitalization parameters are themselves estimates, and the choice of parameter
moves the answer. @ewens2024 recover the R&D depreciation rate and the long-lived
share of SG&A from the prices paid for firms at exit rather than assuming them, and
obtain intangible stocks about 15% smaller on average than the conventional
parameters imply, with considerably more variation across industries. Their stocks
outperform the status-quo measures in explaining market enterprise values and
markedly attenuate the known biases in market-to-book and return on equity. The
lesson for interface research is uncomfortable but simple: a *brand-capital* stock
built from a default SG&A fraction is a parameterized guess, and the sensitivity of
the headline valuation coefficient to that guess belongs in the robustness table,
not in a footnote.

### Measurement Error in $q$, and What It Does to Slopes {#sec-mf-q-measerror}

Suppose the denominator were right in principle. The ratio is still observed with
error, $q^{\mathrm{obs}} = q^{*} + \varepsilon$, and in the classical case the OLS
slope on $q^{\mathrm{obs}}$ is attenuated by the reliability ratio

$$
\lambda \;=\; \frac{\sigma^{2}_{q^{*}}}{\sigma^{2}_{q^{*}} + \sigma^{2}_{\varepsilon}},
$$ {#eq-mf-reliability}

while the *correctly measured* regressors—cash flow, leverage, and the marketing
variables of interest—absorb what the attenuated coefficient leaves behind. This is
why the investment literature's durable "cash flow matters even controlling for
$q$" result is contested: cash flow may be proxying for the part of $q^{*}$ that
$q^{\mathrm{obs}}$ fails to capture rather than for a binding financing constraint.
The same arithmetic applies, with the same force, when the correctly measured
regressor is advertising intensity or a customer-satisfaction score.

Two families of repairs exist. The first cleans the inputs: @lewellen1997 show that
the replacement-cost and debt-value approximations in common use materially change
firm rankings, so measured $q$ is partly a construction artifact. The second treats
the problem statistically. @erickson2000 and @erickson2011a identify the
attenuation from higher-order moments of the joint distribution of the regressors,
delivering consistent slopes without an instrument, provided the measurement error
is classical and the true regressor is non-normal. Marketing papers reporting a $q$
regression with a market-based asset on the right-hand side almost never do either,
which means the reported marketing coefficient silently carries whatever share of
$q^{*}$'s variation the proxy missed.

### Estimating Marginal $q$ Directly {#sec-mf-marginal-q}

@gala2026marginalq close the loop by estimating @eq-mf-marginalq itself instead of
proxying for it, and their method is deliberately light on structure. Under weak
regularity conditions the firm's market value is a differentiable function of a set
of observable state variables that includes the capital stock. So: **first**,
project observed market values onto those state variables, which yields a fitted
value function whose variation is driven only by fundamentals—the resulting ratio,
which they call **Fitted Q**, is a direct estimate of average $Q$ purged of
classical measurement error; **second**, differentiate the fitted value function
with respect to capital to obtain marginal $q$. No functional form is imposed on
the stochastic discount factor or on the adjustment-cost technology, which is what
makes the estimate robust to the misspecification that structural Euler-equation
approaches inherit, and what lets adjustment-cost parameters be recovered
afterwards without simulation-based indirect inference.

Estimating both objects inside one framework separates two failures the literature
routinely confounds. The step from Tobin's $Q$ to Fitted Q removes *classical
measurement error*; the step from Fitted Q to marginal $q$ removes *model
misspecification*—the Hayashi wedge. The paper's central empirical finding is that
the second step is the one that matters:

- Marginal $q$ averages **below one** and is far less dispersed than either average
  measure: roughly $0.8$, against about $1.3$ for Fitted Q and near $2.9$ for raw
  Tobin's $Q$, with a standard deviation more than an order of magnitude below
  Tobin's $Q$'s.
- Using Tobin's $Q$ as the proxy inflates implied capital adjustment costs by a
  factor of about $3.5$.
- It also **understates the sensitivity of investment to fundamentals by roughly
  70–80%**, depending on the adjustment-cost specification.
- Investment regressions run on marginal $q$ or Fitted Q deliver within-firm fit an
  order of magnitude above the Tobin's-$Q$ benchmark (within-group $R^{2}$ up to
  $0.49$, roughly fifteen times larger), and **the cash-flow coefficient loses its
  significance**, which undercuts the financial-constraints reading of that entire
  literature.
- The gap between marginal $q$ and Tobin's $Q$ has **widened over time**, and the
  widening is attributed primarily to market power and intangible capital.

Read against @hayashi1982, the result is not that neoclassical investment theory
failed; it is that the theory was being tested with the wrong observable. The
last bullet is the one that should stop a marketing reader: the two forces driving
the wedge are precisely the two that market-based assets create.

### Decomposing the Average–Marginal Gap {#sec-mf-q-gap}

Knowing a wedge exists is not the same as knowing what is inside it. @crouzet2023
supply the accounting. In their **$Q+$** framework the gap between average $Q$ and
marginal $q$ decomposes into three terms: the value of *installed intangible
capital*, the *rents earned on physical capital*, and an interaction term capturing
the *rents earned on intangibles*. The intangible-related terms contribute
substantially to the gap, especially in fast-growing sectors, which is a direct
caution against the fashionable reading that high valuations alongside weak
investment simply demonstrate rising market power. For marketing the decomposition
reads almost as a translation table: the first term is the market-based asset
itself, and the third is the pricing power that asset confers.

@belo2022 approach the same object from the valuation side, fitting a neoclassical
model with four inputs—physical capital, installed labor, knowledge capital, and
**brand capital**—to observed firm market values. The estimated model attributes on
the order of 30–40% of market value to physical capital, 20–43% to knowledge
capital, 6–25% to brand capital, and 14–22% to the installed labor force, with
physical capital's share falling and knowledge capital's rising across the sample
period. The paper holds dual citizenship: it is a finance paper whose estimated
structural parameter *is* the quantity brand-valuation research has been trying to
recover (@sec-branding), obtained with no brand-survey input at all, and it is
therefore the natural external benchmark against which a marketing brand-equity
estimate should be validated.

### Why Accounting Proxies for $q$ Mislead {#sec-mf-aatq}

A blunt warning is in order, because the convenience of @eq-mf-q-classic has
seduced a generation of marketing papers. @bendle2018misuse label these
financial-statement constructions **accounting-based approximations of Tobin's $q$**
(AATQ) and argue they depart from $q$'s original meaning in ways that
systematically *favor* the marketing variables researchers most want to study.
Their critique has several edges. AATQ are not comparable across industries because
the off-balance-sheet wedge varies by sector; they do not in fact isolate tangible
assets in the denominator; and there is no theoretical reason for AATQ to
equilibrate at one, contrary to a common assumption inherited from the original
$q$. Worse, when AATQ exceeds one—the empirically typical case—it can *rise even in
response to wasted investment*, so an AATQ increase is consistent with both genuine
value creation and value-neutral or value-destroying strategies. The net effect is
a bias toward overstating the effectiveness of investments in market-based assets,
precisely because those assets are the unrecorded items (brand equity, customer
satisfaction) that the AATQ denominator omits. The methodological lesson is to
prefer a $q$ that books intangibles into the denominator (@eq-mf-totalq), to lean
on within-firm variation through fixed effects, and to treat any AATQ-based effect
as suggestive until corroborated.

That last claim—that an AATQ can rise in response to wasted investment—is easy to
see in two lines of arithmetic, and @fig-mf-q-neutral runs it. Take a firm whose
market value is $V_0$ and let it make a **value-neutral**
marketing investment: it spends $I$ on brand building and the market marks its
value up by exactly $I$, so the net present value of the investment is zero and the
marginal $q$ of the increment is exactly one. Because advertising and most brand
spending are *expensed*, the AATQ denominator does not move, and AATQ rises from
$V_0/K^{\mathrm{phy}}$ to $(V_0+I)/K^{\mathrm{phy}}$—the accounting proxy scores a
zero-NPV project as value creation. Total $q$, which books the new intangible into
the denominator at cost, goes to $(V_0+I)/(K^{\mathrm{phy}}+K^{\mathrm{int}}+I)$
and therefore *falls* whenever $q>1$, correctly reporting that nothing was created
beyond the cost of what was bought.

```{r}
#| label: fig-mf-q-neutral
#| fig-cap: "A value-neutral marketing investment under three measures. The firm spends $I$ and the market marks value up by exactly $I$, so marginal $q$ of the increment is one by construction. The accounting proxy (AATQ) rises because expensed brand spending never enters its denominator; total $q$ falls, correctly reporting that no value was created beyond cost."
#| message: false
#| warning: false
# A zero-NPV brand investment: value rises by exactly what was spent.
V0   <- 300   # market value before the investment
Kphy <- 100   # replacement cost of physical capital (the AATQ denominator)
Kint <- 100   # replacement cost of installed intangible capital
I    <- seq(0, 100, by = 5)

aatq <- (V0 + I) / Kphy                 # expensed: denominator never moves
totq <- (V0 + I) / (Kphy + Kint + I)    # intangible booked at replacement cost
mq   <- rep(1, length(I))               # marginal q of the increment: 1 by design

plot(I, aatq, type = "l", lwd = 2, ylim = range(c(aatq, totq, mq)),
     xlab = "Value-neutral brand investment I",
     ylab = "Measure", main = "What a zero-NPV brand investment does to each q")
lines(I, totq, lwd = 2, col = "steelblue")
lines(I, mq,   lwd = 2, col = "grey50", lty = 2)
legend("left", bty = "n", lwd = 2, lty = c(1, 1, 2),
       col = c("black", "steelblue", "grey50"),
       legend = c("AATQ = (V0 + I) / Kphy",
                  "Total q = (V0 + I) / (Kphy + Kint + I)",
                  "Marginal q of the increment = 1"))

cat("AATQ:    ", round(aatq[1], 3), "->", round(aatq[length(I)], 3), "\n")
cat("Total q: ", round(totq[1], 3), "->", round(totq[length(I)], 3), "\n")
```

The three lines diverge for a project that, by construction, created nothing. Any
design that reads value creation off the *level* or the *change* in an accounting
$q$ is reading that divergence.

### Choosing a $q$ for a Marketing Regression {#sec-mf-q-ladder}

@tbl-mf-q-ladder places the variants on one ladder, from the proxy most marketing
papers still use to the object investment theory actually names.

| Measure | What it estimates | Hayashi condition repaired | Data required | Use in interface research |
|---|---|---|---|---|
| **AATQ** (@eq-mf-q-classic) | Average value per dollar of *booked* assets | none | Compustat only | Report for comparability with the prior literature; never as the sole outcome [@bendle2018misuse] |
| **Total $q$** [@peters2017] | Average value per dollar of physical *plus* intangible capital | many capital goods | Compustat + perpetual-inventory stocks (@eq-mf-knowcap) | Default baseline outcome for a valuation regression |
| **Exit-price total $q$** [@ewens2024] | The same, with capitalization parameters estimated rather than assumed | many capital goods, better calibrated | Compustat + exit-price parameters | Robustness check on any brand- or knowledge-capital stock |
| **Higher-moment corrected slopes** [@erickson2000; @erickson2011a] | Consistent coefficients, not a corrected $q$ | classical measurement error | Compustat + non-normal regressors | When the marketing coefficient, not the level of $q$, is the claim |
| **Fitted Q** [@gala2026marginalq] | Average $Q$ purged of classical measurement error | classical measurement error | Market values + firm state variables | When market mispricing or noise is the worry |
| **Marginal $q$** [@gala2026marginalq] | $\partial V/\partial K$: the shadow value of new capital | all four | Market values + firm state variables + a projection step | When the claim is about *investment behavior* rather than about the level of value |
| **$Q+$ decomposition** [@crouzet2023] | The composition of the average–marginal gap | price taking | Total $q$ inputs + markup estimates | When the question is *why* the firm is valued above replacement cost |
| **Capacity-adjusted $q$** [@grullon2025] | Marginal $q$ net of anticipated underutilization | non-convexities | Compustat + capacity-utilization data | Capital-intensive settings; declining-investment puzzles |

: A ladder of $q$ measures, ordered by how much of the @hayashi1982 gap each one closes, with the data each demands and the marketing question each answers. {#tbl-mf-q-ladder}

Four rules follow for practice, and none of them is expensive.

**Report total $q$ as the baseline and AATQ only for comparability.** The
perpetual-inventory stocks in @eq-mf-knowcap are buildable from Compustat
alone, so there is no longer a data excuse for the accounting proxy; keep it in the
table if the literature you are speaking to uses it, but do not let it carry the
claim.

**Treat a marketing coefficient estimated on an average $q$ as an upper bound.**
The wedge between average and marginal $q$ is driven by intangible capital and
rents [@gala2026marginalq; @crouzet2023], and market-based assets *are* intangible
capital that *generates* rents. The bias is therefore not a random nuisance—its
sign is predictable and it points toward the finding the researcher hoped for.

**Firm fixed effects do not solve it.** The wedge is firm-specific *and*
time-varying: it grows as the firm's intangible share and pricing power grow, which
is exactly the variation a within-firm design is exploiting. Fixed effects absorb
the level of the wedge, not its correlation with the marketing regressor.

**Match the measure to the claim.** If the question is "do markets value brand
equity?", an average measure with intangibles in the denominator is the right
outcome. If the question is "does the firm invest more in brand capital when brand
capital is worth more?"—an investment question, and increasingly the interesting
one—then average $Q$ is the wrong regressor and a marginal-$q$ construction is the
right one. The same caution applies retrospectively: the $q$-residual approach to
brand equity in @simon1993, still widely cited, inherits every wedge in
@tbl-mf-q-ladder, and its estimates should be read as an era's best available
approximation rather than as a validated measure.

## Discrete Events: M&A, Alliances, and Innovation

The interface's richest evidence comes from discrete, dated events whose value the
event study can isolate. Mergers and acquisitions (M&A) are the canonical case, and
the marketing question is sharp: *what kind of combination creates value?* Two
hypotheses compete. The **similarity** (relatedness) hypothesis holds that value
comes from combining firms with overlapping products, markets, or technologies,
through scale and consolidation; the **complementarity** hypothesis holds that
value comes from combining firms whose resources differ and fill each other's gaps.
@tbl-mf-mna organizes the core findings; note that the studies disagree because the
*motive* for the merger determines which logic applies.

| Study | DV | Key IV | Finding |
|---|---|---|---|
| @singh1987 | Cumulative portfolio abnormal returns | Acquirer–target relatedness (technological / product-market) | Related firms create more value |
| @shelton1988 | Merger dollar gains | Product-market fit | Related firms create more value |
| @harrison1991 | ROA | Differences in R&D, capital, administrative intensity | R&D *complementarity* boosts unrelated acquisitions |
| @datta1992 | Wealth effects | Bids, financing, acquisition type; served-market overlap | Similar firms create more value |
| @ramaswamy1997 | ROA | Distance in coverage, efficiency, marketing, client mix, risk | Differences *impede* horizontal bank mergers |
| @hitt1998 | ROA | Product-market relatedness and resource complementarity | Resource complementarities aid success |
| @larsson1999 | Synergy realization | Combination potential, integration, low employee resistance | Complementary operations + integration raise synergy |
| @swaminathan2008 | CAR (event study) | Strategic-emphasis alignment; resource similarity/complementarity | Both similarity and complementarity create value, under different motives |

: Selected M&A studies at the marketing–finance interface, illustrating the similarity-versus-complementarity tension. {#tbl-mf-mna}

The financing and timing of deals carries information of its own. Cash deals are
received more favorably than stock-financed deals, because paying with equity
signals that managers believe their stock is overvalued; this adverse reaction to
equity financing is *permanent*, with negative acquirer returns persisting five
years post-takeover [@agrawal1992post; @loughran1997long]. Excess cash is no
panacea—announcement returns fall with an acquirer's cash holdings, consistent
with managers shielded from external capital-market discipline making poorer
investment choices [@harford1999corporate]. There is broad evidence of post-merger
underperformance on average [@agrawal2000], yet the cross-section is informative:
small acquirers earn favorable long-run returns [@mitchell2000managerial;
@moeller2004firm]; undervalued, high book-to-market "value" acquirers outperform
overvalued "glamour" acquirers, whose overconfident managers face weaker scrutiny
[@sudarsanam2003glamour]; bull-market acquisitions underperform bear-market ones as
hubris inflates synergy estimates in upswings [@bouwman2009market]; deals for
privately held or subsidiary targets generate buyer gains through illiquidity
discounts and concentrated post-deal monitoring [@fuller2002returns;
@conn2005impact]; serial acquirers with high valuations destroy value once organic
growth stalls [@moeller2005wealth]; CEO ownership lifts long-run returns
[@cosh2006board]; business similarity and the disposal of non-core assets through
divestitures and spin-offs both raise shareholder value [@megginson2004]; and
cross-border deals favor developed-market acquirers entering emerging markets,
especially in R&D- and brand-intensive businesses where intellectual assets matter
[@chari2009].

@swaminathan2008 bring marketing's distinctive variable—**strategic emphasis**—to
this literature. Following @mizik2003, they operationalize a firm's strategic
emphasis as advertising minus R&D expenditure, scaled by total assets in the
pre-merger year, and define **strategic-emphasis alignment** as the absolute
difference between acquirer and target emphasis. Their event study of 206 deals
shows that when merging firms are *poorly* aligned, diversity (complementarity)
improves value, whereas when they are *well* aligned, value is enhanced by a
consolidation motive—reconciling the similarity/complementarity debate by making the
merger's *motive* the moderator. Marketing assets thus shape M&A value creation
directly.

The interface also documents how M&A interacts with the customer franchise.
@umashankar2021 find that acquisitions tend to *reduce* customer satisfaction, as
executives' attention shifts from customers to financial integration; the resulting
dissatisfaction can cannibalize the very synergies the deal was meant to capture,
though the presence of marketing expertise in the firm's upper echelons mitigates
the damage. Even the market for the dealmakers themselves responds to organizational
structure: high-performing M&A bankers, especially early in their careers, migrate
from bulge-bracket banks toward focused boutiques as cross-subsidization of
underperforming divisions pushes talent out, which in turn shapes deal outcomes
[@gao2024big].

**Strategic alliances** extend the same logic to looser combinations.
@swaminathan2009a show that alliance announcements create value for the
announcing firm, and that *network* position governs how much: a firm's abnormal
returns are most favorable when network **efficiency** (access to non-redundant
partner capabilities) and **density** (interconnection among partners) are
*moderate*, while network reputation and centrality have no effect; a firm's
**marketing-alliance capability**—its accumulated skill at managing prior
alliances—positively drives value creation. Networks amplify alliance benefits,
facilitate compliance, and signal partner and alliance quality. Construction of the
network variables uses a window of prior years (five in @swaminathan2009a, seven in
related work [@gulati1999; @schilling2007]), and because partnerships form through
referrals rather than at random, the authors estimate a selection model alongside
the value-creation model [@verbeek1992a]. Their controls follow the
intangible-asset tradition: installed base, relationship investment, marketing and
advertising expenditures, and R&D, each entered through a Koyck (geometric-lag)
function and combined via principal components to tame multicollinearity
[@dutta1999], plus alliance experience, partner size, intra- versus inter-industry
scope [@rindfleisch2001b], and repeat partnering.

Innovation events round out the picture: a firm's aggregate investment decisions
shape not only its own value but the **growth and volatility of its entire
industry**, as @king2011industry document across 377 industries over sixteen years,
where investments in value creation and value appropriation interact intricately
with the industry environment. Related work on the value of innovation in
acquisitions is treated alongside R&D in @king2008performance.

## Voluntary Commitments and the Modern Event Study {#sec-mf-carbon}

The event study is sometimes dismissed as a method whose best years were the 1980s
and 1990s. That is wrong as a matter of publication record: @schnabel2026carbon
place a conventional short-window CAR design in *Management Science* in 2026. What
has changed is not the estimator but **where the contribution sits**. The
first-stage average CAR is no longer the finding; it is the setup. The finding lives
in the second stage—in what the analyst can measure about the event that predicts
the cross-section of abnormal returns—and in the translation of a percentage return
into a unit that a manager or a regulator can act on.

### The Design and Its Headline Number

@schnabel2026carbon study **voluntary carbon goal announcements**: unilateral,
uncompelled public commitments by firms to reduce or eliminate greenhouse-gas
emissions in specified scopes by a stated date. The sample is 188 announcements by
U.S. publicly traded firms across a variety of industries between 2018 and 2024,
and the main event window is $(-2, +2)$ trading days. Three numbers summarize the
first stage:

- an average CAR of $+0.65\%$, so the market rewards the commitment rather than
  penalizing it as a costly distraction;
- a translation of that percentage into about \$490 million of value at the average
  market capitalization in the sample;
- a further normalization to roughly \$75 of abnormal return per current ton of
  CO$_2$-equivalent emissions—a modest price, the authors note, set against the
  substantial financial commitment that carbon neutrality implies.

The voluntariness of the event is what makes the sign interesting. A mandated
disclosure is news about regulation; a voluntary goal is news about the firm's own
beliefs, and the market must decide whether the commitment signals capability—the
firm knows how to abate cheaply, or expects abatement to pay—or signals expense, in
which case the firm is about to spend shareholder money on a non-market objective.
The positive CAR says the capability reading dominates on average, which is the
market-based-assets logic of @Srivastava_1998 applied to an environmental
commitment: the announcement revises expected cash flows upward, presumably through
some combination of demand, cost, regulatory-risk, and cost-of-capital channels that
the CAR alone cannot separate.

### Normalizing a Return into a Physical Unit

The third number is the one worth stealing. A CAR reported in percent leaves the
reader with no way to judge whether the market's reaction is large, small, or
absurd. Dividing the dollar value by the **physical quantity being committed**—here,
tons of CO$_2$-equivalent—produces a price per unit that can be compared against
external benchmarks: the social cost of carbon, prevailing carbon-market prices, and
the firm's own expected abatement cost per ton. If the market pays \$75 per ton for
a commitment that costs \$150 per ton to honor, either the market is wrong or the
commitment is not expected to be honored in full; if it pays \$75 for abatement that
costs \$20, the announcement is revealing a profitable project. The normalization
converts an unfalsifiable "the market liked it" into a quantity that other evidence
can contradict.

That normalization is also an arithmetic identity, which means a reader can audit
it. The chunk below recovers the two sample quantities implied by the three
published numbers—a habit worth applying to any paper that reports a translated
effect size.

```{r mf-carbon-implied, message=FALSE, warning=FALSE}
# Reported in Schnabel, Heidari, Hock, and Raithel (2026, Management Science):
car_pct     <- 0.0065      # average CAR over the (-2,+2) event window
value_usd   <- 490e6       # implied value change at average market cap
per_ton_usd <- 75          # abnormal return per current ton of CO2e

# The two translations are an identity, so the sample quantities behind them
# can be backed out and checked for plausibility.
implied_mktcap <- value_usd / car_pct       # average market capitalization
implied_tons   <- value_usd / per_ton_usd   # avg. current emissions per firm

data.frame(
  quantity = c("Implied average market cap ($B)",
               "Implied avg. current emissions per firm (Mt CO2e)"),
  value    = c(round(implied_mktcap / 1e9, 1),
               round(implied_tons   / 1e6, 2))
)
```

The implied average market capitalization lands in the tens of billions, consistent
with a sample of large U.S. public firms, and the implied emissions figure is a few
million tons per firm—both plausible, which is the point of the exercise. When the
implied quantities are *not* plausible, the translation is doing rhetorical work the
estimate cannot support.

### Where the Contribution Lives: The Second Stage

The cross-sectional regression of $\mathrm{CAR}_i$ on event characteristics is where
@schnabel2026carbon make their argument, and their two moderators illustrate the two
distinct ways a modern event study earns a contribution.

**A moderator built from text.** The authors apply **topic modeling** to the media
coverage surrounding each announcement and find that investors respond more
positively when coverage emphasizes the *expected business impact* of the carbon
goal rather than treating it as a purely ethical or aspirational gesture. This is
the text-as-data machinery of @sec-text-as-data feeding the second stage of a
finance design: an unstructured corpus becomes a per-event, interpretable covariate.
The causal reading requires care—media framing is not randomly assigned, and a firm
with a more credible goal will both generate business-impact coverage and earn a
higher CAR—so the estimate is best read as characterizing *which announcements the
market rewards*, not as an experiment in framing. Stated that way it is still
actionable, because it identifies what the informative content of an announcement
is.

**A moderator built from a physical quantity.** Firms with **higher current
emissions in the scopes they commit to eliminate** are rewarded more. This is the
volume argument: a ton pledged by a heavy emitter is a real ton, whereas a
carbon-neutrality claim by a firm with negligible emissions in the committed scopes
is close to free. It also answers the standard cynicism about ESG announcements—if
the market were pricing cheap talk, the reaction would not scale with the magnitude
of what is actually being given up.

**Both effects decay over the sample period.** The premium for emphasizing business
impact and for committing large current emissions diminishes across the 2018–2024
window. The economics are the order-of-entry economics of @sec-innovation: the first
credible commitment in an industry is news, the twentieth is table stakes. This is
the result with the sharpest managerial edge, and methodologically it is the reason
the sample period matters—an event study run on 2018–2020 alone and one run on
2022–2024 alone would report different moderator coefficients, and an analyst who
pooled them would have no way to see why.

The chunk below simulates the second stage so that the estimating equation and the
decay test are explicit. The data are simulated rather than the authors', but the
specification is the one their design implies. Centering the two moderators is what
makes the output readable: each main effect is then the moderator's effect at the
*start* of the sample window, and each interaction with time is the amount by which
that effect decays across the window.

```{r mf-carbon-second-stage, message=FALSE, warning=FALSE}
set.seed(7)
n <- 188                                  # announcements, as in the paper

# --- event-level covariates --------------------------------------------------
t_norm    <- sort(runif(n))               # announcement time, scaled to [0,1]
biz_share <- pmin(pmax(rnorm(n, 0.35, 0.15), 0), 1)  # topic share: business impact
log_emis  <- log(pmax(rlnorm(n, 1.6, 1.1), 0.05))    # log committed-scope emissions
size      <- rnorm(n)                     # log market cap, a standard control

biz_c  <- biz_share - mean(biz_share)     # centered: effect at window start
emis_c <- log_emis  - mean(log_emis)

# --- DGP: both moderator effects decay to zero by the end of the window ------
CAR <- 0.0065 +
  0.080  * biz_c  * (1 - t_norm) +
  0.008  * emis_c * (1 - t_norm) +
  0.0004 * size +
  rnorm(n, 0, 0.022)          # abnormal returns are mostly noise by construction

d      <- data.frame(CAR, biz_c, emis_c, t_norm, size)
m_main <- lm(CAR ~ biz_c + emis_c + size, data = d)
m_time <- lm(CAR ~ biz_c * t_norm + emis_c * t_norm + size, data = d)

round(summary(m_main)$coefficients[, c(1, 4)], 4)
round(summary(m_time)$coefficients[, c(1, 4)], 4)
```

Three features of the output carry the lesson. First, because the moderators are
centered, the intercept of `m_main` recovers the average CAR—about $0.6\%$, the
quantity the first stage was built to estimate—and both moderators enter positively,
which is the headline moderation result. Second, in `m_time` each interaction with
announcement timing is negative and roughly equal in magnitude to its own main
effect, which is the signature of an effect that starts large and is gone by the end
of the window; a specification without the interaction averages two different
regimes and understates how much value was available to an early mover. Third, note
how little of the variance in $\mathrm{CAR}$ these regressions explain—here about
$12\%$ and $22\%$. Abnormal returns are mostly noise by construction, and a second
stage that explained most of their variance would be evidence of a leak, not of a
good model.

That low explanatory power has a consequence worth confronting directly: **the decay
interaction is the fragile part of the design.** Re-running the same
data-generating process under fresh noise shows how often a sample of this size
actually recovers a decay that is genuinely there.

```{r mf-carbon-power, message=FALSE, warning=FALSE}
# Detection rate for each decay interaction, holding the DGP above fixed.
detect <- function(seed) {
  set.seed(seed)
  t_norm <- sort(runif(188))
  biz_c  <- pmin(pmax(rnorm(188, 0.35, 0.15), 0), 1)
  emis_c <- log(pmax(rlnorm(188, 1.6, 1.1), 0.05))
  biz_c  <- biz_c  - mean(biz_c)
  emis_c <- emis_c - mean(emis_c)
  CAR <- 0.0065 + 0.080 * biz_c  * (1 - t_norm) +
                  0.008 * emis_c * (1 - t_norm) + rnorm(188, 0, 0.022)
  ct <- summary(lm(CAR ~ biz_c * t_norm + emis_c * t_norm))$coefficients
  c(business_impact = ct["biz_c:t_norm",  1] < 0 && ct["biz_c:t_norm",  4] < .05,
    emissions       = ct["t_norm:emis_c", 1] < 0 && ct["t_norm:emis_c", 4] < .05)
}
round(rowMeans(sapply(1:400, detect)), 2)
```

Roughly half the time for the stronger moderator and a third of the time for the
weaker one—even though the decay is real by construction and the main effects are
detected almost always. Interactions in a CAR second stage are estimated on the
residual of an outcome that is already nearly all noise, so they cost far more
sample than main effects do. Three practical implications follow. Design the sample
period *long* enough that announcement order actually varies, since the decay is
identified only by that spread. Do not read a null interaction as evidence of no
decay. And treat a decay result that survives at this sample size as informative
precisely because it is hard to find—which is the right way to read the finding in
@schnabel2026carbon rather than as a routine robustness check.

### Design Decisions and How They Are Defended

@tbl-mf-carbon-design collects the choices any announcement event study must make,
the threat each choice addresses, and the defense reviewers expect. It is written to
be used as a checklist when designing or refereeing one.

| Decision | Threat it addresses | Standard defense |
|---|---|---|
| Short window, e.g. $(-2,+2)$ | Long windows accumulate unrelated news and lose power | Report a window grid ($(-1,+1)$, $(-2,+2)$, $(-5,+5)$) and show the effect concentrates near day 0 |
| Precise event date | Leakage before, or slow diffusion after, the announcement | Hand-code the first public dissemination; verify pre-window abnormal returns are flat |
| Confound screen | Earnings, splits, M&A, executive turnover sharing the window | Drop or flag contaminated events; report results with and without them |
| Benchmark model | The market model may misprice risk in a period of factor turbulence | Re-estimate with Fama–French/Carhart factors and with market-adjusted returns |
| Cross-sectional standard errors | Announcements cluster in calendar time (COP meetings, reporting seasons), correlating abnormal returns across firms | Cluster by event date or industry, or use calendar-time portfolios |
| Self-selection into announcing | Firms choose whether and when to announce at all | Model the announcement decision; match on observables; interpret estimates as conditional on announcing |
| Sample-period heterogeneity | Effects that decay make a pooled estimate period-specific | Interact the key moderators with announcement time, as above |

: Design decisions in a modern announcement event study, with the threat each addresses and the defense reviewers expect. {#tbl-mf-carbon-design}

### What Generalizes to Marketing Events

Nothing in this template is specific to carbon. The transferable structure is a
**voluntary, dated, unilateral commitment** whose content can be measured on a scale
the firm cannot costlessly inflate, announced into a media environment whose framing
can be quantified from text, in a population where announcement order matters.
Marketing supplies many such events: a public commitment to drop a controversial
ingredient or supplier practice, a privacy pledge that forgoes a data source
(@sec-privacy), a service guarantee that transfers risk from customer to firm
(@sec-service), a repositioning that abandons a profitable product line. Each has a
physical or financial quantity that scales the promise, each is covered by media
whose framing is measurable, and each is subject to the same first-mover decay. The
design in @schnabel2026carbon is a reusable template, and the reason to teach it
here is that the estimator is the least interesting part of it.
## Marketing, Information, and the Stock Market

Beyond discrete deals, a growing literature studies how marketing assets shape the
*information environment* of the stock and shows that marketing's financial payoff
runs partly through investors, not only customers. @cheong2021 find that
advertising lowers **stock-price synchronicity**—the degree to which a firm's
return moves in lockstep with its industry—implying that advertising conveys
firm-specific information to investors and is valued for that, over and above its
demand effects. Firm-generated social content has measurable, security-level
consequences: @lacka2021 define **price impact** as the effect on the variance of a
stock's price and estimate the permanent and temporary price impacts of S&P 500 IT
firms' tweets, finding that tweets carrying *both* valence and subject matter about
consumer or competitor orientation produce *permanent* price impact, whereas tweets
carrying only one attribute move prices only temporarily—and negative-valence tweets
about competitors generate the largest permanent impacts. The broader case that
marketing creates measurable firm value, and how to assess its effectiveness and
efficiency, is made by @hanssens2016, and the meta-analytic synthesis of marketing's
firm-value effects across studies is provided by @edeling2016. The general framing
of performance outcomes in marketing—and the multiple, sometimes conflicting metrics
managers must reconcile—is laid out by @katsikeas2016, and the broad correspondence
between marketing and finance assets-to-value logic traces back to @commenta2009 and
@Srivastava_1998.

A complementary result concerns the *human-capital* side of the interface.
@anderson2018 show that improving both marketing and finance skills raises profits,
but through different pathways: marketing/sales skill lifts profit via a
**growth** focus (higher sales, more product and employee investment), while
finance/accounting skill lifts it via an **efficiency/cost** focus. The managerial
implication is a fit argument—marketing skill is the better fit for startups chasing
growth, finance skill for mature firms optimizing cost.

## Adjacent Constructs: Agility and Complexity

Two firm-level constructs increasingly appear as moderators or controls in
interface studies, and both are measured from mandatory disclosures, which makes
them attractive instruments.

**Corporate agility**—a firm's ability to adapt to environmental change—is
hypothesized to raise survival rates [@lehn2021]. Because firms must report
accurately to the SEC, the measure built from disclosures is reliable; @lehn2021
operationalize agility as the sensitivity of a firm's competitive responses to
rivals' threats—specifically, the sensitivity of its product similarity (or
dissimilarity) to rivals' products to the rivals' own similarity to the firm—and
distinguish agility from mere flexibility.

**Firm complexity** is captured by **accounting reporting complexity** (ARC), "the
difficulty to understand, prepare, audit, and analyze the financial reports,"
operationalized as the count of accounting items disclosed in XBRL 10-K filings
[@hoitash2017, p. 262]. ARC is inversely associated with financial-reporting
quality and positively associated with audit delay and audit fees, and—because it is
built from the focal firm's own line items rather than from text that may reference
*other* firms—it captures firm complexity more cleanly than dictionary-based
linguistic measures [@hoitash2022; @loughran2020; @loughran2020a]. Complementary
complexity proxies include operating-segment counts, foreign operations, and the
linguistic readability of filings via the Fog Index [@gunning1952technique] and
10-K length. Complexity has downstream consequences: as it rises, analysts struggle
to forecast accurately [@hoitash2021], and corporate social responsibility is
positively correlated with greater ARC [@garcia2020].

## Frontier Topics in Corporate Finance for Marketing

The interface continues to expand into the financing events that bracket a firm's
public life, where marketing assets act as quality signals to capital providers.

**Venture capital.** Early intellectual-property signals matter for seed funding:
@rieger2024express show, on a Crunchbase–USPTO dataset of 5,370 ventures, that
ventures filing trademark applications are more likely to secure VC funding, with the
effect strongest in the first 100 days, in low-technological-uncertainty industries,
and in non-clustered locations—evidence that trademarks are early credible signals of
venture quality.

**Initial public offerings.** Firms can use **innovation potential** as a credible
quality signal at the IPO: @cao2022express find it positively associated with the
IPO's initial value and first-day returns and negatively associated with insider
selling, with patents weighing most on insider sales and pre-announcements weighing
most on first-day returns. The modern IPO landscape has also shifted—firms emphasize
growth metrics (such as user counts) over traditional financials and go public
roughly twice as old as in the 1980s, amid a larger retail-investor base, which has
prompted proposals for triggered disclosure.

**Customer-based corporate valuation.** When a firm's value derives transparently
from its subscriber base, the customer franchise can be valued directly:
@mccarthy2017 use DISH Network and Sirius XM data to estimate firm value from
customer-level acquisition, retention, and spend dynamics—bringing the market-based-
asset logic full circle by valuing the firm *from its customers up*.

**Initial coin offerings.** A nascent literature studies token sales as a financing
mechanism, supported by emerging data infrastructure [@czaja2021; @lyandres2022;
@hsieh2021; @momtaz2020; @belitski2021; @campino2022].

## Key Takeaways

- The marketing–finance interface treats brands, customers, channels, and market
  knowledge as **market-based assets** that pass the resource-based tests and create
  shareholder value through four levers of the discounted-cash-flow identity:
  accelerating, augmenting, de-risking, and raising the residual value of cash flows
  [@Srivastava_1998].
- Under near-semi-strong market efficiency, the *change* in stock price at the
  moment marketing news arrives is a risk-adjusted estimate of its value, which is
  what the **event study** and stock-return response model exploit
  (@eq-mf-return).
- **Persistence/VAR** models separate temporary from permanent effects of marketing
  shocks; **Tobin's $q$ regressions** measure the cross-sectional value of marketing
  assets but require careful identification.
- Accounting-based proxies for $q$ (**AATQ**) are biased toward overstating
  market-based-asset effects; prefer a **total $q$** that books intangible capital
  into the denominator at replacement cost [@peters2017; @bendle2018misuse], and
  estimate the capitalization parameters rather than assuming them [@ewens2024].
- The deeper problem is that **average $Q$ is not marginal $q$**. The two coincide
  only under the @hayashi1982 conditions—price taking, linear homogeneity, a single
  capital good, no fixed costs—and the modern literature is a set of relaxations of
  those conditions (@fig-mf-q-map).
- @gala2026marginalq estimate marginal $q$ directly by projecting market values on
  firm state variables and differentiating with respect to capital. Marginal $q$
  averages below one, Tobin's $Q$ near three; using $Q$ as the proxy inflates
  adjustment costs about $3.5\times$, understates investment's sensitivity to
  fundamentals by 70–80%, and manufactures the cash-flow effect that the
  financial-constraints literature reads as evidence.
- The average–marginal gap decomposes into installed intangibles, rents on physical
  capital, and rents on intangibles [@crouzet2023]—and market-based assets sit on
  two of those three terms, which makes the bias in a marketing $q$ regression
  **directional, not random**. Treat such a coefficient as an upper bound; firm
  fixed effects do not absorb a wedge that grows with the regressor.
- In M&A and alliances, *both* resource similarity and complementarity create
  value—the merger's **motive**, captured by strategic-emphasis alignment, decides
  which logic applies [@swaminathan2008; @swaminathan2009a].
- The event study is a **living method**, not a historical one: @schnabel2026carbon
  report a $+0.65\%$ CAR around voluntary carbon goal announcements. Its modern
  contribution sits in the **second stage**—moderators built from text (topic-modeled
  media framing) and from physical quantities (committed emissions)—and in the
  normalization of the return into an auditable per-unit price (@sec-mf-carbon).
- Announcement effects **decay with order of entry**, so an event study's moderator
  estimates are period-specific unless the design interacts them with announcement
  time.

## Further Reading

The foundational statement of the market-based-assets framework is
@Srivastava_1998, with the productivity-chain elaboration in @Rust_2004; the
methodological review of stock-market approaches to marketing is @srinivasan2009
and the meta-analytic synthesis is @edeling2016. For Tobin's-$q$ measurement the
reading has moved on, and it is best taken in order: @hayashi1982 for the
conditions under which average $Q$ is marginal $q$ at all; @peters2017 for the
intangible denominator, with @ewens2024 on where the capitalization parameters come
from; @erickson2000 and @erickson2011a on what measurement error does to the slope;
@gala2026marginalq for the direct estimate of marginal $q$ and the finding that
misspecification rather than noise is the binding problem; and @crouzet2023 with
@belo2022 for what the average–marginal gap is actually made of—the latter
delivering a structural estimate of the share of firm value attributable to brand
capital. @bendle2018misuse remains the right cautionary entry point for a marketing
audience.

The M&A and alliance applications are best entered through @swaminathan2008 and
@swaminathan2009a. For
a current, well-executed announcement event study whose contribution lives in the
second stage rather than in the average CAR—and a model of how to translate an
abnormal return into a per-unit price a manager can argue about—read
@schnabel2026carbon.
