# Privacy {#sec-privacy}

*Privacy* is the governance problem at the center of data-driven marketing. Every
personalized recommendation, every retargeted advertisement, every loyalty program
rests on a flow of personal information from the consumer to the firm—and that flow
is the same thing consumers say they want to restrict. The tension is not
incidental; it is structural. Firms create value by knowing their customers, and the
act of knowing is precisely what consumers experience as exposure. This chapter
treats privacy as a marketing construct with formal content, an economic object that
firms can price and differentiate on, a behavioral phenomenon that systematically
violates the predictions of rational choice, and a regulated activity whose rules
reshape competition and innovation.

The chapter proceeds from definition to mechanism to evidence. We begin by fixing
what *privacy* and *privacy concern* mean and how they are measured, because the
measurement model determines what can be claimed. We then assemble the theoretical
scaffolding—social contract, justice, exchange, control, and reactance accounts—into
a single decision-theoretic frame, the **privacy calculus**, and confront the
empirical fact that consumers do not behave as the calculus predicts: the **privacy
paradox**. From there we examine personalization (the firm's reason to collect
data), trust and control (the levers that move the calculus), the surprising
backfire of transparency (the **bulletproof-glass effect**), the economics of
anonymity, and the regulatory regime—chiefly the European Union's General Data
Protection Regulation (GDPR)—that now constrains the whole system, together with the
open problem of verifying that a trained recommender has actually forgotten a user
who asked to be deleted. Throughout, the
goal is to give a marketing PhD the identification detail and a senior practitioner
the decision logic, with reproducible code where a model is involved.

## What Privacy Is

The foundational definition in the information-privacy literature is Westin's, who
frames privacy not as secrecy but as *control over access*.

> Privacy is the claim of individuals, groups, or institutions to determine for
> themselves when, how, and to what extent information about them is communicated to
> others. [@westin1967f]

Westin decomposes this claim into four states—**anonymity**, **solitude**,
**reserve**, and **intimacy**—that together describe a person's ability to modulate
how exposed they are. The control-centric view matters because it locates privacy in
the *flow* of information rather than in any fixed catalog of "sensitive" facts: a
consumer can be entirely comfortable disclosing a fact in one relationship and feel
violated disclosing the same fact in another. *Consumer information privacy* narrows
this to the marketing context as control over the dissemination and use of consumer
information—demographics, search history, profile data—where canonical violations
are unwanted marketing communication, hyper-targeted advertising, and covert online
tracking.

Privacy itself is a latent disposition; what researchers actually observe is
**privacy concern**, a proxy that operationalizes consumers' beliefs, attitudes, and
perceptions about their information privacy, typically through survey scales
[@malhotra2004internet]. The distinction between the construct and its proxy is not
pedantic. It is the reason the field's central anomaly—the privacy paradox, the gap
between stated concern and revealed disclosure—is even possible to state.

### Measuring Privacy Concern

Two measurement traditions dominate. The Concern for Information Privacy (CFIP)
tradition originates with @smith1996information and is refined for the internet by
@malhotra2004internet into the Internet Users' Information Privacy Concerns (IUIPC)
instrument, which loads concern on four dimensions: **information collection**,
**unauthorized secondary use** (internal and external repurposing of data),
**improper access**, and **errors** (inadequate protection against inaccuracies). A
parallel tradition [@sheehan2000dimensions] organizes concern around awareness of
collection, information use, information sensitivity, familiarity with the collecting
entity, and compensation. The two overlap substantially but differ in emphasis:
@malhotra2004internet centers the firm's data-handling practices, whereas
@sheehan2000dimensions foregrounds the consumer's relationship to the entity and the
sensitivity of the specific datum.

Formally, both treat concern as a **reflective** higher-order construct: a latent
privacy concern $\eta$ generates correlated indicators $x_k$ via

$$
x_k = \lambda_k \eta + \varepsilon_k, \qquad k = 1, \dots, K,
$$ {#eq-cfip-reflective}

so the dimensions should be internally consistent and the instrument is validated by
convergent and discriminant validity. This reflective specification is consequential:
it licenses summing the dimensions into a single concern score and assumes the
dimensions are interchangeable manifestations of one underlying trait. A formative
reading—in which collection, secondary use, access, and error are distinct *causes*
that compose concern and need not correlate—would forbid that aggregation. The field
has largely adopted the reflective view by convention rather than by decisive test, a
caveat worth carrying into any empirical application.[^reflective-note]

[^reflective-note]: The reflective-versus-formative question recurs across marketing
constructs and is unsettled for several of them; the parallel debate for brand
equity and authenticity is developed in @sec-branding. The practical stake is
identical: the measurement model dictates which validity tests apply and whether the
dimensions may be summed.

Concern is not static. It rises over time among both older and younger consumers,
and faster among older consumers, so cohort and period effects must be modeled rather
than assumed away [@goldfarb2011online]. And concern is consequential for behavior:
it mediates the effect of website personalization on click-through, so the same
personalization can help or hurt depending on the concern it provokes
[@bleier2015importance].

## Theoretical Foundations

Why should a consumer ever disclose? The literature answers with five overlapping
theories, which are best read not as competitors but as components of a single
account: disclosure is governed by a perceived contract, evaluated for fairness,
priced as an exchange, modulated by control, and resisted when it threatens autonomy.
@tbl-privacy-theories collects them.

| Theory | Core mechanism | What governs disclosure | Representative evidence |
|---|---|---|---|
| Social contract | Norms of appropriate information flow | Firm upholds its end via added value or compensation | @chellappa2005 |
| Justice | Procedural and distributive fairness | Fair-seeming policies and fair outcomes | @awad2006; @vail2008empirical |
| Power–responsibility | Trust obligation of the powerful party | Perceived balance of power | @milberg2000 |
| Social exchange | Benefit minus cost | Personalization value exceeds privacy loss | @chellappa2005 |
| Reactance | Threat to autonomy | Whether targeting is felt as intrusion | @tucker2013; @white2008getting |

: Theoretical lenses on information disclosure. {#tbl-privacy-theories}

**Social contract theory** treats the firm–consumer data relationship as governed by
a tacit moral contract: consumers believe marketers have kept their end when the
firm returns value—through personalization or monetary compensation—commensurate
with what was taken. The contract is normative, so the same data practice reads as
respect or as violation depending on whether it conforms to the consumer's
expectations of appropriate flow.

**Justice theory** sharpens this into two fairness dimensions. *Procedural justice*
concerns the policies and processes by which information is collected and used; fair
access and fair utilization are its content, and even merely fair-*appearing*
policies can reduce concern. Crucially, procedural fairness depends on the consumer's
comprehension: complex privacy policies that impede understanding are perceived as
*less* fair and erode trust [@vail2008empirical]. *Distributive justice* concerns the
outcome—the benefits the consumer receives for disclosing, from customized goods and
faster service to outright compensation. The two dimensions generate a localized
version of the paradox: consumers value the distributive outcomes marketers deliver
while simultaneously feeling vulnerable about the disclosure that produced them
[@awad2006].

**Power–responsibility (control) theory** holds that the more powerful partner in a
relationship—here, the data-holding firm—bears a social obligation to foster
equality and trust. When consumers perceive a power imbalance as a threat to their
information privacy, they respond defensively, withholding information or supplying
false data. When firms lack credible privacy protections, consumers do not simply
exit; they demand *government* intervention to restore the balance [@milberg2000].

**Social exchange theory** reduces disclosure to a cost–benefit comparison: a
consumer discloses when perceived benefits—individualized offerings, free services,
convenience—exceed perceived costs. The greater the personalization value a consumer
senses from an exchange, the more they believe the marketer has honored the
governing social contract, tying exchange and contract logics together
[@chellappa2005].

**Reactance theory** supplies the failure mode. Targeted, individualized
communication can be experienced as a violation of autonomy; privacy concern
heightens reactance [@tucker2013], and the downstream behaviors are corrosive:
communication avoidance, deliberate information falsification, negative word of mouth,
and other defensive actions [@white2008getting]. Reactance is why personalization is
not monotonically good—past a threshold, knowing too much about a consumer reads as
intrusion rather than service.

These accounts converge on a decision-theoretic primitive. Consumer privacy behavior
reflects a multidimensional, boundedly rational calculation in which the consumer
weighs the expected benefits of disclosure against its expected costs. That
calculation is the **privacy calculus**, to which we now give formal content.

## The Privacy Calculus

The privacy calculus models disclosure as a utility-maximizing trade-off. Let a
consumer face a decision to disclose information $d \in \{0, 1\}$ to a firm. Let
$B(d)$ denote the perceived benefit of disclosure—personalization value, convenience,
compensation—and $C(d)$ the perceived cost—exposure, expected harm, loss of control.
Disclosure utility is

$$
U(d) = B(d) - C(d),
$$ {#eq-privacy-calculus}

and the rational consumer discloses iff $U(1) > U(0)$, i.e., iff the marginal benefit
exceeds the marginal cost. Two structural features make the calculus tractable and
testable. First, personalization value and privacy concern are approximately
*orthogonal* inputs: a firm can raise $B$ (better personalization) and lower the
weight on $C$ (alleviate concern) through largely independent levers
[@chellappa2005]. Second, **trust** enters as a moderator that scales the perceived
cost: where a firm is trusted, the same objective data practice carries a lower
subjective $C$, raising the probability of disclosure.

A convenient empirical implementation writes disclosure as a binary choice driven by
benefit and cost indices,

$$
\Pr(d_i = 1) = \Lambda\!\left( \alpha + \beta\, B_i - \gamma\, C_i
+ \delta\, (T_i \times C_i) \right),
$$ {#eq-disclosure-logit}

where $\Lambda(\cdot)$ is the logistic link, $T_i$ is trust, and the interaction
$T_i \times C_i$ captures trust attenuating the cost of disclosure ($\delta > 0$).
The estimand of interest is usually $\gamma$, the sensitivity of disclosure to
perceived cost, and $\delta$, the degree to which trust neutralizes it.
**Identification** is the central difficulty. Trust, benefit perception, and cost
perception are all endogenous to unobserved consumer type: a consumer who is
intrinsically comfortable with disclosure will report both higher trust and lower
cost, biasing $\hat\gamma$ and $\hat\delta$ if type is uncontrolled. Credible
estimates therefore exploit exogenous shifts—a policy change that alters control
without altering type (@sec-control-trust), or randomized framing of the disclosure
prompt—rather than cross-sectional correlation. Absent such variation, the
coefficients are descriptive, not causal.

The following simulation makes the calculus concrete and shows how trust shifts the
disclosure curve. It is illustrative—the data are generated, not observed—but the
estimand and the logic are exactly those of @eq-disclosure-logit.

```{r privacy-calculus-sim, message=FALSE, warning=FALSE}
set.seed(2025)
n <- 4000

# Latent inputs: personalization benefit, privacy cost, and trust.
benefit <- rnorm(n, 0, 1)
cost    <- rnorm(n, 0, 1)
trust   <- rnorm(n, 0, 1)

# Data-generating process for disclosure (eq-disclosure-logit).
alpha <-  0.0; beta <- 1.2; gamma <- 1.4; delta <- 0.9
eta <- alpha + beta * benefit - gamma * cost + delta * (trust * cost)
p   <- 1 / (1 + exp(-eta))
disclose <- rbinom(n, 1, p)

dat <- data.frame(disclose, benefit, cost, trust)
fit <- glm(disclose ~ benefit + cost + trust:cost,
           family = binomial, data = dat)
round(coef(summary(fit)), 3)

# Disclosure probability vs. perceived cost, at low vs. high trust,
# holding benefit at its mean.
grid <- expand.grid(cost = seq(-2, 2, 0.1),
                    trust = c(-1, 1), benefit = 0)
grid$p <- predict(fit, newdata = grid, type = "response")

plot(grid$cost[grid$trust == -1], grid$p[grid$trust == -1], type = "l",
     lwd = 2, ylim = 0:1, xlab = "Perceived privacy cost",
     ylab = "P(disclose)", main = "Trust attenuates the cost of disclosure")
lines(grid$cost[grid$trust == 1], grid$p[grid$trust == 1], lwd = 2, lty = 2)
legend("topright", c("Low trust", "High trust"), lwd = 2, lty = c(1, 2),
       bty = "n")
```

The recovered coefficients reproduce the planted parameters, and the two curves show
the substantive point: under high trust, the disclosure probability falls far more
slowly as perceived cost rises—trust flattens the privacy cost gradient.

## The Privacy Paradox

The calculus predicts that consumers who report high privacy concern should disclose
less. They do not. The **privacy paradox** is the systematic disconnect between
consumers' stated privacy preferences and their actual disclosure behavior: people
who declare strong concern nonetheless share sensitive personal information freely
[@norberg2007privacy; @martin2016]. The paradox is robust and large. In the market
for online behavioral advertising, only about 0.23% of U.S. ad impressions come from
users who have opted out of behavioral targeting, despite survey evidence of
widespread concern; opt-out users' ad impressions earn roughly 52% less revenue than
targeted impressions, implying a loss on the order of \$8.58 per opted-out consumer to
publishers and exchanges, and opt-out behavior concentrates among more
tech-savvy, older, and wealthier populations [@johnson2020].

Two readings of the paradox carry different managerial implications, and the
distinction is identification-relevant. The first treats the paradox as *rational*:
concern is real, but the privacy calculus resolves in favor of disclosure because the
benefits genuinely dominate. On this reading, low opt-out rates reveal that consumers
price their data cheaply relative to the personalization they receive—privacy is a
commodity over which consumers make a deliberate risk–benefit trade-off
[@sultan2009]. The second treats the paradox as *behavioral*: stated and revealed
preferences diverge because disclosure decisions are constructed in the moment and
are highly malleable, so the survey measures a disposition that the choice context
overrides. The behavioral reading has the stronger evidence. Consumers trade away
privacy in response to the **architecture and framing** of the choice
[@adjerid2019choice; @brandimarte2013misplaced], to **small inconveniences or
incentives** that should be economically negligible [@athey2017digital], and to
**perceived control** over their information even when that control does not reduce
actual exposure [@mourey2020past; @tucker2013].

The behavioral reading is sharpened by @lin2025choice, who show that opt-out defaults
and lower price anchors reduce consumers' stated valuations of their own Facebook
data by 22% and 37% respectively—magnitudes incompatible with a stable underlying
valuation. The same study surfaces a consequence that should worry any firm using
disclosure data to learn about its market: framing effects fall disproportionately on
consumer segments with *lower* privacy valuations, so optimizing choice architecture
to maximize data *volume* systematically biases *who* is represented in the data. The
volume–representativeness trade-off is a direct cost of exploiting the paradox.

A caution follows for measurement. If disclosure is constructed and framing-sensitive,
then survey-based privacy concern is a weak predictor of behavior, and any model that
uses stated concern as a proxy for revealed cost (the $C_i$ in
@eq-disclosure-logit) inherits that weakness. This is why the strongest empirical
privacy work is built on revealed behavior under exogenous variation rather than on
self-report.

## Personalization, Trust, and Control {#sec-control-trust}

Personalization is the firm's reason to collect data, and its effect on consumer
response is genuinely mixed. *Personalization* is the ability to tailor products,
messages, and experiences to an individual based on personal and preference
information; it depends jointly on the firm's ability to acquire and process consumer
data and on the consumer's willingness to supply it [@chellappa2005]. Its benefits
are real—fit, proactive delivery, and elevated switching costs that breed loyalty,
because a consumer who has invested in personalizing one provider must re-disclose to
switch [@alba1997interactive]. But personalization is double-edged, and the edge that
cuts the firm is privacy concern. The pivotal moderator is *how the information was
obtained*: whether the consumer provided it voluntarily or it was gathered covertly.

### The Personalization–Performance Curve

The effect of personalization intensity on advertising response is theorized as
**non-monotonic**, with two opposing forces. The *positive* force is elaboration: a
self-relevant message invites central-route processing, raising attention,
elaboration, and attitude strength, so more personalization yields more favorable
response [@cho2004people; @petty1986]. The *negative* force is **persuasion
knowledge**: when a consumer recognizes that a brand is trying to persuade them—and
conspicuous use of personal data is exactly such a cue—they activate persuasion
knowledge that discounts the appeal, and the discounting is amplified by privacy
concern through reactance [@friestad1994; @white2008getting]. The net effect can be
U- or inverted-U-shaped depending on which force dominates, and the empirical record
is correspondingly split: many studies find personalization lifts brand and campaign
response and coupon redemption [@bauer2014; @kim2012; @tam2005web;
@maslowska2011effectiveness; @xu2008combining; @chu2011viral], while others find that
*explicitly* signaling data use backfires.

The cleanest demonstration of the negative branch is @wattal2012: consumers respond
favorably to product-based personalization, where data use is not made salient, but
respond *unfavorably* when a firm is explicit about using personally identifiable
information—a personalized greeting, for instance—with familiarity with the firm
attenuating the backlash. The mechanism is reactance triggered by salient
surveillance, not personalization per se.

Whether a tradeoff between personalization and privacy actually binds is itself
contested. Several studies find *no* tradeoff once control is present.
@walrave2016 hypothesize an optimal-moderate level of personalization in social
network advertising but find the highest-personalization condition produces the most
positive response, with privacy concern failing to dampen it—plausibly because the
personalization never crosses the threshold that activates persuasion knowledge. The
deeper resolution comes from control.

### Control as the Decisive Lever

The single most important moderator of the personalization–privacy relationship is
**perceived control**. @tucker2013 provides the cleanest causal estimate: exploiting
Facebook's introduction of enhanced privacy controls as a natural experiment in a
difference-in-differences design, she finds that personalized advertising was
*ineffective* before users gained control over their personal information but became
roughly *twice as effective* afterward. The interpretation is sharp—personalization
and privacy do not trade off when consumers believe they control their data; control
converts intrusion into service.

The mechanism generalizes. Perceived control is the mediator through which three
distinct interventions—consumer self-protection, industry self-regulation, and
government mandates—reduce privacy concern [@xu2012research]. But control has a dark
side that the calculus does not anticipate. Greater *perceived* control can induce
consumers to disclose *more*, leaving them objectively more exposed: the feeling of
control reduces the felt cost of disclosure even when the actual risk is unchanged or
worse [@brandimarte2012economics]. This is the control paradox, and it is the
behavioral-economics counterpart to the privacy paradox—control changes the
*subjective* $C$ in @eq-privacy-calculus without changing the *objective* one.

### Trust as the Proactive Complement

Where control and concern-reduction are largely *reactive*—they respond to a privacy
threat—**trust** is a *proactive* mechanism that firms build before any threat
materializes [@wirtz2009regulatory]. Trust is decisive precisely in the settings
where privacy is most at stake: when retailers deploy personalized or targeted
material, trust is what assuages the accompanying privacy worry [@aguirre2016], and
trust in the vendor positively moderates the use of personalized services, so trust
building is a direct lever on data acquisition [@chellappa2005].

Firms have concrete instruments to signal trustworthiness. Privacy seals and similar
signals raise consumers' willingness to disclose and improve perceptions of the
organization [@miyazaki2002internet]. Privacy policies are read—about 84% of a sample
of 2,500 consumers reported reading them—and reading them shifts trust in the firm
[@milne2004strategies], with policies serving as a usable proxy for a firm's
transparency and control that is, in turn, related to firm performance and consumer
behavior [@martin2017data]. Consumers will even *pay* for privacy: they purchase more
from websites with more protective privacy practices, revealing a positive
willingness to pay for protection [@tsai2011effect]. Yet the relationship between
disclosure-of-policy and behavior is not simple. Broadcasting a privacy policy can
*reduce* disclosure, and consumers disclose more sensitive information when they
believe others already have—social proof lowering the perceived risk
[@acquisti2012impact]. Privacy-policy statements raise the benevolence and integrity
components of trust without necessarily lifting purchase intent. These frictions set
up the central behavioral surprise of the chapter.

## The Bulletproof-Glass Effect

A reasonable manager expects that telling consumers their data are protected will
make them feel safer and buy more. The expectation is wrong, and the reason is
instructive. @brough2022 identify the **bulletproof-glass effect**: just as a sheet
of bulletproof glass in an unexpected setting makes people feel *less* safe by
implying that danger is present, a privacy notice can make consumers feel *more*
vulnerable—decreasing trust and purchase interest—even when the notice objectively
emphasizes protection. The label inverts the standard view that transparency about
data handling *reduces* perceived vulnerability [@martin2017data].

The mechanism turns on a contract distinction. Managers conceive of privacy notices
as *formal legal contracts* that bind the firm's data conduct, and expect that
formality to reassure. But consumers default to reading privacy as a *social*
contract governed by norms of appropriate flow; when a firm respects those norms it
earns trust and purchase intent, and when it violates them it earns negative word of
mouth [@mccole2010trust; @eastlick2006understanding; @miyazaki2009perceived;
@kim2019seeing]. Introducing a *formal* contract where a *social* one was expected
undermines trust, exactly as a formal contract reduces cooperation in a multi-round
trust game relative to no contract at all [@malhotra2002effects]. The notice signals
that the relationship requires legal protection, which implies the relationship is
not safe.

@brough2022 bound the effect with two informative moderators. First, the backfire
*disappears when consumers already expect potential harm*: priming a-priori
distrust, or operating in a context where danger is anticipated, removes the surprise
that drives the bulletproof-glass reaction. Second, the backfire *disappears when the
notice carries benevolence cues*. Trust has two routes—benevolence (affective,
"they care") and competence (cognitive, "they are able")—and benevolence-trust leads
consumers to read a notice as a social contract rather than a formal one. Operationally,
a notice's affective-versus-cognitive content can be scored from its text—e.g., via
the proportion of affect words versus cognition words—so the benevolence remedy is
measurable, not merely conceptual. @fig-bulletproof traces the logic.

```{mermaid}
%%| label: fig-bulletproof
%%| fig-cap: "The bulletproof-glass effect: why a protective privacy notice can lower purchase interest, and the two moderators that switch the effect off."
flowchart TD
    A[Privacy notice shown] --> B{How is it read?}
    B -->|As a formal contract| C[Implies danger is present]
    C --> D[Increased perceived vulnerability]
    D --> E[Lower trust and purchase interest]
    B -->|As a social contract| F[Implies norm-respecting relationship]
    F --> G[Maintained or increased trust]
    H[Moderator 1: harm already expected] --> B
    I[Moderator 2: benevolence cues in text] --> B
    H -. removes surprise .-> F
    I -. shifts to affective trust .-> F
```

The effect extends to the *conspicuous absence* of a notice. Using Apple's App Store
privacy "nutrition labels"—mandated in December 2020 after a period in which only
about half of apps voluntarily linked a privacy policy [@story2018apps]—@brough2022
show that an app whose privacy information is conspicuously present *or* conspicuously
absent suffers reduced downloads relative to apps where privacy is simply not at
issue. Drawing attention to privacy, in either direction, can depress demand. This
reframes a large body of evidence: when consumers notice their data were collected
without consent, click-through falls [@kim2019seeing; @aguirre2015unraveling], and
consumers prefer to buy from sites with stronger protection [@tsai2011effect]—but the
act of *foregrounding* privacy is itself a treatment with a cost. The practical
lesson is not "say less"; it is that privacy communication is a designed signal whose
framing (formal versus benevolent, salient versus background) determines its sign.

## The Economics and Organizational Stakes of Privacy

Privacy is also a strategic and economic variable, not only a psychological one. For
the firm, privacy is a dimension of *differentiation*: a firm can compete on
protection the way it competes on quality or price [@casadesus2015competing;
@rust2002customer]. The financial logic of privacy *failures*—data breaches, hacking
intrusions, exposures—runs the other way and is well documented
[@malhotra2011evaluating; @martin2017data]. The key organizational result is that the
damage a breach does to firm performance is *mitigated* when the firm has previously
provided **transparency** and **control**: these compensate for the vulnerability the
breach exposes and blunt its impact on performance [@martin2017data]. Privacy posture
is thus an insurance policy whose premium is paid in advance through transparent,
control-granting practices.

The economics of *anonymity* formalize a subtler point: more privacy is not
unconditionally good for consumers. @conitzer2012 analyze a monopolist facing a
continuum of heterogeneous consumers who can choose to remain anonymous—avoiding
recognition as prior customers—possibly at a cost. The model yields a sequence of
counterintuitive results. When anonymity is freely available, every consumer
individually prefers to take it, and that universal anonymity is precisely the
outcome that *maximizes* the monopolist's profit (because it prevents the
price-discrimination that recognition would enable against the firm's interest in
this setting). Consumers can benefit from a *higher* cost of anonymity—but only up to
a threshold, beyond which the effect reverses and harms them. And it is frequently
*harmful* to consumers when the monopolist or an unaffiliated third party controls the
cost of anonymity. The lesson for policy is that the *allocation of control over the
price of privacy* matters as much as the level of privacy itself.

On the regulatory-economics of breaches, @martin2016 argues that the low likelihood
of litigation, combined with the rarity of breaches in some industries, gives firms a
standing incentive to lobby for *modest* consumer-information-privacy regulation,
keeping privacy management under firm control rather than statutory mandate. That
political-economy equilibrium is exactly what comprehensive regulation disrupts.

## Privacy in Society and Regulation

Privacy regulation has historically diverged across jurisdictions, and the divergence
shapes everything downstream. In the United States, the Federal Trade Commission (FTC)
has regulated consumer information privacy through two frameworks: a **notice-and-choice**
model built on Fair Information Practice Principles—firms must disclose their
practices and offer choice—and a **harm-based** model that intervenes only where
demonstrable harm occurs. The European Union's earlier Data Protection Directive was
more comprehensive than the U.S. regime, yet the de facto default on both sides
remained industry **self-regulation**, more prevalent in the United States than in
Europe [@martin2016]. The consequences of even the earlier European regime were
real: advertising effectiveness fell measurably after the Data Protection Directive
was implemented, and privacy concern rose over time across age groups
[@goldfarb2011online].

The regime-defining event is the General Data Protection Regulation (GDPR), in force
since 2018, which shifts the legal basis of data processing toward *enhanced,
transparent consent* and exports its model to the California Consumer Privacy Act and
Brazil's Lei Geral de Proteção de Dados. GDPR's effects fall into three areas:
consent and disclosure behavior, market structure and innovation, and the persistence
of tracking. @tbl-gdpr-evidence summarizes the evidence.

| Domain | Finding | Source |
|---|---|---|
| Consent behavior | GDPR-compliant consent raises opt-in across data types; consumers still restrict sensitive data | @godinho2022consumer |
| Consent behavior | People share data freely post-GDPR regardless of request format; explanations go unread | @kreuter2020collecting |
| Consent design | Persuasive + informative cues secure more opt-ins than informative cues alone | @d2024express |
| Transparency | Transparent policies raise comprehension of data processing but not acceptance | @betzing2020impact |
| Market structure | One-third of apps exit; app introduction halves; consumer surplus and usage fall | @janssen2022gdpr |
| Competition | Volatility rises in free-app market (pro-competitive), falls in paid (anti-competitive) | @martin2017data |
| Enforcement | Limited change in third-party tracking; large gatekeepers retain disproportionate reach | @kollnig2021before |

: Empirical evidence on the effects of GDPR. {#tbl-gdpr-evidence}

The consent findings reveal a tension. Enhanced, GDPR-compliant consent *increases*
opt-in for many data types and uses, and crucially the resulting gains—higher sales,
more effective marketing communication, greater contractual lock-in—accrue to firms
that can target the receptive households that opted in [@godinho2022consumer].
Consumers do not, however, increase allowances uniformly; they continue to restrict
permissions for sensitive information, exercising exactly the norm-governed,
context-dependent control that social-contract theory predicts. Yet other work finds
consent largely insensitive to *how* it is requested: people share data freely
regardless of request format, and the accompanying explanations of collection and use
are not read carefully [@kreuter2020collecting]—the privacy paradox reasserting
itself under regulation. The design margin still matters at the firm level: a mix of
persuasive and informative cues yields more opt-ins than informative cues alone, and
firms with offline presence and heavy data dependence lean on persuasive cues while
GDPR-compliance-sensitive firms adopt more balanced formats [@d2024express].
Transparency improves *comprehension* of data processing without moving *acceptance*
rates [@betzing2020impact], echoing the bulletproof-glass lesson that more
information is not mechanically more reassuring.

Disclosure mandates also reach the parties who must comply with them, not only the
consumers they inform—a margin that evaluation-side studies of labels are structurally
unable to see. In the generative-AI setting, creators who anticipate that their use of
AI will be labeled withdraw effort from the creative work itself, so the disclosed
artifact is *worse* than the one the rule was written to make legible
[@jussupow2026disclosure]. The transferable point for privacy research is that a
transparency requirement is an intervention on production as well as on perception:
consent screens, provenance labels, and data-practice disclosures change what firms
build and how they build it, and a design that holds the firm's behavior fixed while
randomizing the notice measures only half the policy (@sec-aiml-disclosure).

The market-structure findings are where GDPR's costs surface, and they expose a
**privacy–innovation tradeoff**. In the Google Play ecosystem (4.1 million apps,
2016–2019), GDPR forced roughly a third of apps to exit and cut the rate of new app
introduction by half, reducing consumer surplus and app usage [@janssen2022gdpr]. The
competitive incidence is two-sided. Compliance costs are *anti-competitive*: they
raise barriers to entry, increase exit, and fall hardest on smaller firms. But the
challenge of data collection under GDPR is also *pro-competitive*: fines scale with
size and so bite larger firms harder, and established firms find it more difficult to
ensure third-party compliance across their data supply chains. The net is
heterogeneous—GDPR raises volatility (a proxy for competitive churn) in the free-app
market while reducing it in the paid market [@martin2017data]. And enforcement has
limits: third-party tracking changed little after GDPR, with a few large gatekeepers
retaining disproportionate tracking capability both before and after
[@kollnig2021before].

For empirical work on regulation, the measurement bottleneck is detecting and coding
privacy-relevant text at scale. Automated classifiers can identify GDPR-relevant
privacy-policy content, enabling large-sample study of disclosure practices and their
correlates [@chang2019automated]. The methodological point connects privacy research
to the broader text-analytics toolkit and makes the regulatory-compliance literature
tractable beyond hand-coded samples.

## The Right to Be Forgotten Inside a Trained Model {#sec-privacy-revocation-audit}

GDPR's right to erasure lets a consumer ask a firm to delete their personal data, and
for most firms deleting the rows is the easy part. The hard part is the model. A
recommender trained on a user's viewing or purchase history keeps what it learned
from that history after the rows are gone: it can go on ranking that user's
characteristic items the way it did before. Removing a user's influence without
retraining from scratch is the problem of *machine unlearning*
[@cao2015towards; @bourtoule2021machine], with recommender-specific methods that
partition the training data so that only the affected shard has to be retrained
[@chen2022recommendation]. Unlearning methods tell a firm how to forget. They do not
tell the user, the regulator, or the firm's own compliance team whether the deployed
model actually *has* forgotten. A deletion request is a promise about a model's
internals that nobody outside the engineering team can see.

@zhu2026forget turn that verification gap into a design problem they call *user data
revocation auditing*: given a trained sequential recommender and a user who asked to
be deleted, infer whether the model still retains that user's data. Statistically,
this is membership inference [@shokri2017membership], the attack in which an
adversary decides whether a record was in a model's training set, turned to a
defensive use and asked about a whole user rather than a single record. Membership
inference against recommenders is known to work [@zhang2021membership]; the novelty
is to use it as a compliance instrument. Their artifact, RecAudit, outperforms
existing auditing and membership-inference baselines across several real-world
datasets, and the authors pitch it as a way for platforms to find the high-risk users
whose traces persist and to target corrective unlearning at them. (Only the published
abstract and the authors' replication code were accessible when this section was
written. The description of the method below is read from that code, and the
paper's own headline experiments, on MovieLens-1M with 256 audit models, are not
reported here.)

### How the Audit Works

The authors' code makes the design concrete. Users are split into thirds: *training*
users, *audit* users, and *auxiliary* users. Half of the audit users stay in the
target model's training data, standing in for deletion requests that were not
honored; the other half are left out entirely, which is the gold standard of
retraining from scratch. The auditor knows the model architecture and holds the
interaction data of the audit and auxiliary users. It trains its own *audit models*
(the "shadow models" of the membership-inference literature) on random halves of that
pool, so that every user has a set of reference models that saw them and a set that
did not.

Each model is scored by how well it predicts a user's own behavior. For a sequence
$s$ of user $u$ whose true next item is $i_s$, let $f_M(i \mid s)$ be model $M$'s
score for item $i$ and $f_M^{(k)}(s)$ its $k$-th highest score over all items. The
*unforgotten score* is the gap between the true next item and the model's top-$K$
list, averaged over the user's sequences $S_u$:

$$
g_u(M) = \frac{1}{|S_u|} \sum_{s \in S_u} \left[ f_M(i_s \mid s)
- \frac{1}{K} \sum_{k=1}^{K} f_M^{(k)}(s) \right], \qquad K = 10.
$$ {#eq-privacy-gap}

A model that memorized the user places the next item at or near the top of its list,
so the gap is close to zero; a model that never saw the user leaves the item well
below its top ten. Measuring the true item against the model's own top-$K$ mean,
rather than using its raw probability, nets out how predictable the user's sequences
are to begin with. The audit then asks whether the target model $M^\ast$ looks more
like the reference models that saw $u$ or the ones that did not. With Gaussians
fitted to the user's in-reference scores ($\hat\mu^{\text{in}}_u,
\hat\sigma^{\text{in}}_u$) and out-reference scores ($\hat\mu^{\text{out}}_u,
\hat\sigma^{\text{out}}_u$), the statistic is the log likelihood ratio

$$
\Lambda_u = \log \phi\!\left(g_u(M^\ast);\, \hat\mu^{\text{in}}_u, \hat\sigma^{\text{in}}_u\right)
- \log \phi\!\left(g_u(M^\ast);\, \hat\mu^{\text{out}}_u, \hat\sigma^{\text{out}}_u\right),
$$ {#eq-privacy-lr}

and a user is flagged as *unforgotten* when $\Lambda_u > 0$. Ranking users by
$\Lambda_u$ gives the priority list for unlearning. @eq-privacy-lr is the per-example
likelihood-ratio test of @carlini2022membership; what RecAudit changes is the score
that goes into it. The baselines in the authors' code span the membership-inference
design space. One applies a single global threshold to the model's confidence in the
true item [@yeom2018privacy]. Others compare the user's confidence with the user's
own out-reference mean, or with the midpoint of the in and out means. The last runs
the Gaussian test of @eq-privacy-lr on confidence instead of the gap. The code labels
these by first author (Yeom, Zhou, Watson, Sablayrolles, Carlini).

### Replication from the Authors' Posted Scores

The first author's repository [@zhu2026recauditcode] ships a complete demo on
MovieLens-100K [@harper2015movielens] with a BERT4Rec target model
[@sun2019bert4rec]: the data split (314 audit users, 157 unforgotten and 157
forgotten), the target model and 32 audit models (16 seeds, each training one model
on a random half of the audit-plus-auxiliary pool and one on the complement), the
per-user scores each model produced, and the authors' expected results. The README
warns that the demo's numbers will differ from the paper's, which uses MovieLens-1M
and 256 audit models. We re-implemented the audit stage in R from the posted scores
and reproduced the authors' table exactly (@tbl-privacy-recaudit-replication).

One detail decided whether the match was exact. The scores are stored in half
precision (numpy `float16`), and the authors' code does its arithmetic at that
precision. Half-precision rounding creates ties among users, and those ties move
results in the fourth decimal: computed in double precision, RecAudit's AUROC comes
out 0.9235 rather than 0.9236 and Carlini's FPR at 10% FNR 0.8153 rather than 0.8025.
Rounding each intermediate step to the nearest `float16`, as numpy does, reproduces
every published value. For five of the six methods the per-user statistics also
match the authors' saved values to machine precision; Carlini's differ slightly for
11 of 314 users, which does not change its AUROC or FPR.

```{r}
#| label: tbl-privacy-recaudit-replication
#| tbl-cap: "Auditing whether a recommender forgot deleted users: the authors' MovieLens-100K demo (BERT4Rec, 32 audit models, 314 audit users) re-run from the posted scores. AUROC ranks unforgotten above forgotten users; FPR at 10% FNR is the share of honored deletions wrongly flagged when the audit catches 90% of unforgotten users. Every value matches the authors' README to four decimals."
recaudit_tab <- data.frame(
  Method = c("Yeom et al. (global confidence threshold)",
             "Zhou et al. (confidence + correctness, global)",
             "Sablayrolles et al. (per-user midpoint)",
             "Watson et al. (per-user out-reference)",
             "Carlini et al. (Gaussian LR on confidence)",
             "RecAudit (Gaussian LR on gap score)"),
  AUROC          = c(0.5825, 0.5802, 0.7238, 0.6872, 0.6561, 0.9236),
  `FPR at 10% FNR` = c(0.7898, 0.7834, 0.5478, 0.6624, 0.8025, 0.2102),
  check.names = FALSE)
knitr::kable(recaudit_tab, digits = 4)
```

The gap is large. RecAudit ranks a random unforgotten user above a random forgotten
one 92% of the time, against 72% for the best baseline. At an operating point that
catches 90% of the users whose deletion was not honored, it wrongly flags 21% of the
users whose deletion was honored; the baselines wrongly flag 55% to 80%. Its own
decision rule ($\Lambda_u > 0$) flags 158 users, 134 of the 157 unforgotten users and
24 false alarms, for an accuracy of 85.0%. Under the same rule the Carlini baseline
flags 216 users, 93 of them false alarms.

### What Carries the Gain

RecAudit differs from the strongest baseline in two ingredients: the score
(@eq-privacy-gap instead of confidence) and the test (per-user Gaussian reference
instead of a global threshold). The posted scores let us cross the two, which the
demo does not report. @tbl-privacy-recaudit-decomp runs the gap score through each
baseline's test.

```{r}
#| label: tbl-privacy-recaudit-decomp
#| tbl-cap: "Which ingredient carries the gain? The four tests crossed with the two scores, from the same posted demo scores. The confidence rows are the published baselines; the gap-score rows other than RecAudit are our own combinations."
decomp_tab <- data.frame(
  Test = rep(c("Global threshold", "Per-user out-reference",
               "Per-user midpoint", "Per-user Gaussian LR"), 2),
  Score = rep(c("Confidence in true item", "Gap score"), each = 4),
  AUROC = c(0.5825, 0.6872, 0.7238, 0.6561, 0.6767, 0.8888, 0.9011, 0.9236),
  `FPR at 10% FNR` = c(0.7898, 0.6624, 0.5478, 0.8025, 0.7006, 0.2548, 0.1847, 0.2102),
  check.names = FALSE)
knitr::kable(decomp_tab, digits = 4)
```

Both ingredients matter, and they interact. With a single global threshold the gap
score is only modestly better than confidence (0.68 against 0.58), because users
differ so much in how predictable they are that one cutoff cannot serve all of them.
Calibrating against each user's own reference models is the larger step: the gap
score's AUROC jumps from 0.68 to 0.89. Confidence gains less from calibration (0.58 to
0.72), and the Gaussian test on confidence (0.66) does *worse* than the simple
midpoint rule. One plausible reading is that the logit-transformed confidence is
poorly fit by a Gaussian with only 16 draws per side, while the gap score is better
behaved. The Gaussian test also is not uniformly best on the gap score: the midpoint
rule has the lower false-alarm rate at 10% FNR (0.18 against 0.21) despite its lower
AUROC. Which test to deploy depends on which operating point the auditor cares about.

### The Cost of an Audit

Every audit model is a full training run of the recommender, so the number of audit
models is the dominant cost of an audit. @carlini2022membership note that with few
reference models a pooled ("global") variance estimate is more stable than per-user
variances. @fig-privacy-audit-budget re-runs both tests on random subsets of the 16
seed pairs.

```{r}
#| label: fig-privacy-audit-budget
#| fig-cap: "AUROC against the number of audit models (each seed pair trains two), from the posted demo scores. Points at 4 to 16 models average 40 random subsets of seed pairs; 32 is the full demo. Solid lines estimate a separate variance for each user; dashed lines pool one variance across users."
#| fig-width: 7
#| fig-height: 4.5
budget <- data.frame(
  models = c(4, 8, 16, 32),
  rec_user   = c(0.7450, 0.8557, 0.9001, 0.9236),
  rec_pool   = c(0.8789, 0.8906, 0.9003, 0.9073),
  car_user   = c(0.6010, 0.6485, 0.6600, 0.6561),
  car_pool   = c(0.6308, 0.6118, 0.6075, 0.6212))
plot(budget$models, budget$rec_user, type = "b", log = "x", pch = 19, lwd = 2,
     ylim = c(0.3, 0.95), xaxt = "n", xlab = "Number of audit models",
     ylab = "AUROC", main = "Audit accuracy against audit budget")
axis(1, at = budget$models)
lines(budget$models, budget$rec_pool, type = "b", pch = 1,  lwd = 2, lty = 2)
lines(budget$models, budget$car_user, type = "b", pch = 17, lwd = 2, col = "grey45")
lines(budget$models, budget$car_pool, type = "b", pch = 2,  lwd = 2, lty = 2,
      col = "grey45")
abline(h = 0.5, col = "grey80", lty = 3)   # chance
legend("bottomright", bty = "n", lwd = 2, lty = c(1, 2, 1, 2),
       pch = c(19, 1, 17, 2), col = c("black", "black", "grey45", "grey45"),
       legend = c("RecAudit, per-user variance", "RecAudit, pooled variance",
                  "Carlini, per-user variance", "Carlini, pooled variance"))
```

With four audit models, per-user variances give RecAudit an AUROC of 0.745 (standard
deviation 0.036 across subsets); pooling the variance raises it to 0.879, within 0.045
of the full 32-model audit. The two estimators meet at 16 models, and only at 32 does
the per-user version pull ahead (0.924 against 0.907). For a firm, this means a
credible screen costs a handful of retraining runs, not hundreds. The paper's 256
audit models buy the last few points of accuracy, and the extra budget is worth
spending once the audit has to stand up in front of a regulator. These budget numbers
come from the MovieLens-100K demo and should not be read as the paper's results.

### What It Means for Marketers and Regulators

Three implications follow. First, the audit is a black-box test of the deployed model:
it needs the model's item scores for a user's sequences, not its weights. But it is
not assumption-free. The auditor must know the architecture and training recipe well
enough to train faithful reference models, and must hold auxiliary data from the same
population, so a regulator who wanted to run it would have to mandate that access.
Second, the score in @eq-privacy-gap is computed on the deleted user's own interaction
history, so the audit has to be run by someone who still has that history: the user,
submitting their own history, or a trusted auditor before final deletion. Verifying
a deletion and completing it are therefore distinct steps, and the order matters.
Third, the audit turns erasure from a binary legal claim into a ranked list of users
whose traces most likely persist. That list is what lets a firm target approximate
unlearning [@bourtoule2021machine; @chen2022recommendation] at the users who need it
instead of retraining the whole recommender on every request, which ties the
compliance cost of GDPR's erasure right to the market-structure costs in
@tbl-gdpr-evidence.

To reproduce the replication, check out the authors' repository (on Windows, the
result files' names end in a period and need WSL or `git show` to extract), flatten
the pickled scores to CSV with the short Python script below, and run the R script on
the three CSVs. It prints @tbl-privacy-recaudit-replication,
@tbl-privacy-recaudit-decomp, the budget curves in @fig-privacy-audit-budget, and a
per-user comparison against the authors' saved statistics. The book does not run it
at render time.

```python
# export_recaudit.py: flatten the RecAudit ml-100k demo pickles into three CSVs.
# Usage: python export_recaudit.py <path-to-RecAudit-checkout> <output-dir>
import os, sys, csv, pickle

root, out = sys.argv[1] + "/RecStudio-main", sys.argv[2]
load = lambda p: pickle.load(open(p, "rb"))
base = "ml-100k/BERT4Rec/split/train"

split = load(f"{root}/metadata/split/ml-100k/split/split_info.pkl")
pos = [int(u) for u in split["pos_audit_users"]]       # still in the training data
neg = [int(u) for u in split["neg_audit_users"]]       # deleted (never trained on)

with open(f"{out}/recaudit_audit_users.csv", "w", newline="") as f:
    w = csv.writer(f); w.writerow(["user_id", "unforgotten"])
    w.writerows([(u, 1) for u in pos] + [(u, 0) for u in neg])

rows = []
for s in ["recaudit", "baseline"]:
    for u, v in load(f"{root}/metadata/scores/target_model/{base}/1/{s}_score.pkl").items():
        rows.append((int(u), "target", 0, s, float(v)))
    for seed in range(1, 17):
        for role in ["member", "nonmember"]:
            p = f"{root}/metadata/scores/shadow_model/{base}/{seed}/{s}_{role}_score.pkl"
            for u, v in load(p).items():
                rows.append((int(u), role, seed, s, float(v)))
with open(f"{out}/recaudit_scores.csv", "w", newline="") as f:
    w = csv.writer(f); w.writerow(["user_id", "model", "seed", "score", "value"])
    w.writerows(rows)

# The authors' saved per-user statistics, for an exact cross-check. Their file names
# end in "." ("Yeom et al."), which Windows cannot check out; accept "_et_al" copies.
res = f"{root}/audit_results/audit/{base.replace('BERT4Rec/', 'BERT4Rec/BERT4Rec/')}/audit/1"
with open(f"{out}/recaudit_author_likelihoods.csv", "w", newline="") as f:
    w = csv.writer(f)
    w.writerow(["method", "user_id", "label", "likelihood", "auroc", "fpr_at_10fnr"])
    for m in ["Yeom et al.", "Zhou et al.", "Sablayrolles et al.", "Watson et al.",
              "Carlini et al.", "RecAudit"]:
        p = f"{res}/{m}"
        r = load(p if os.path.exists(p) else p.replace(" et al.", "_et_al"))
        for u, l, x in zip(r["users"], r["label"], r["likelihood"]):
            w.writerow([m.replace(" et al.", "_et_al"), int(u), int(l), float(x),
                        float(r["auroc"]), float(r["fpr_at_fix_fnr"])])
```

```{r privacy-recaudit-script, eval=FALSE}
# Replication of the audit stage of RecAudit (Zhu, Yang, Fan & Lian 2026, ISR)
# from the authors' posted ml-100k / BERT4Rec demo scores (github.com/zzh9568/RecAudit).
args <- commandArgs(trailingOnly = TRUE)
dir <- if (length(args)) args[1] else "."

sc    <- read.csv(file.path(dir, "recaudit_scores.csv"))
users <- read.csv(file.path(dir, "recaudit_audit_users.csv"))
auth  <- read.csv(file.path(dir, "recaudit_author_likelihoods.csv"))

# ---- half precision -------------------------------------------------------
# The posted scores are numpy float16, and audit.py does its arithmetic in float16
# (numpy keeps the array dtype when a Python float is added). h() rounds a double to
# the nearest float16, so that each step below rounds exactly where numpy does.
h <- function(x) {
  ok <- is.finite(x) & x != 0
  e <- pmax(floor(log2(abs(x[ok]))), -14)            # -14 = smallest normal exponent
  q <- 2^(e - 10)                                    # 10 explicit mantissa bits
  x[ok] <- round(x[ok] / q) * q                      # R rounds half to even, like IEEE
  x
}
h_mean <- function(v) h(h(sum(v)) / length(v))       # numpy sums float16 in float32
h_std  <- function(v) { d <- h(v - h_mean(v)); h(sqrt(h(h(sum(h(d * d))) / length(v)))) }

# ---- score transforms (audit.py: _unforgotten_score_prepro) -----------------
neg_loss <- function(s) { s <- pmin(h(s + 1e-3), 1); h(log(s)) }   # -(-log s)
lira     <- function(s) { z <- h(h(s / h(h(1 - s) + 1e-3)) + 1e-3); h(log(pmin(z, 1e3))) }

# ---- AUROC (trapezoidal, ties = 1/2) and FPR at 10% FNR ---------------------
auroc <- function(y, x) {
  r <- rank(x); n1 <- sum(y == 1); n0 <- sum(y == 0)
  (sum(r[y == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
fpr_at_fnr <- function(y, x, fnr = 0.1) {
  th  <- sort(unique(x), decreasing = TRUE)
  tpr <- sapply(th, function(t) mean(x[y == 1] >= t))
  fpr <- sapply(th, function(t) mean(x[y == 0] >= t))
  ok  <- which(1 - tpr < fnr)
  if (length(ok)) fpr[ok[1]]
}

# ---- assemble per-user target score and IN / OUT reference scores -----------
get <- function(score, model, seeds = 1:16) {
  d <- sc[sc$score == score & sc$model == model & (model == "target" | sc$seed %in% seeds), ]
  split(d$value, d$user_id)
}

# 2-sided Gaussian likelihood ratio (Carlini et al.'s online LiRA; RecAudit's test)
lr_gauss <- function(tgt, inn, out, pooled = FALSE) {
  if (pooled) { s_in <- sd_pool(inn); s_out <- sd_pool(out) }
  sapply(names(tgt), function(u) {
    m1 <- h_mean(inn[[u]]); m0 <- h_mean(out[[u]])
    s1 <- if (pooled) s_in else h_std(inn[[u]])   # numpy std, ddof = 0
    s0 <- if (pooled) s_out else h_std(out[[u]])
    dnorm(tgt[[u]], m1, s1 + 1e-30, log = TRUE) - dnorm(tgt[[u]], m0, s0 + 1e-30, log = TRUE)
  })
}
sd_pool <- function(lst) sqrt(mean(sapply(lst, function(v) mean((v - mean(v))^2))))

audit_stat <- function(test, tgt, inn, out, pooled = FALSE) {
  ids <- names(tgt)
  switch(test,
    global = unlist(tgt) - mean(unlist(tgt)),
    one_side = sapply(ids, function(u) tgt[[u]] - h_mean(out[[u]])),
    two_side = sapply(ids, function(u) tgt[[u]] - h(h(h_mean(inn[[u]]) + h_mean(out[[u]])) / 2)),
    gauss = lr_gauss(tgt, inn, out, pooled))
}

run <- function(method, seeds = 1:16, pooled = FALSE) {
  cfg <- list(
    "Yeom et al."         = c("baseline", "neg_loss", "global"),
    "Zhou et al."         = c("baseline", "corr",     "global"),
    "Sablayrolles et al." = c("baseline", "neg_loss", "two_side"),
    "Watson et al."       = c("baseline", "neg_loss", "one_side"),
    "Carlini et al."      = c("baseline", "lira",     "gauss"),
    "RecAudit"            = c("recaudit", "none",     "gauss"),
    "Gap score, global"   = c("recaudit", "none",     "global"),
    "Gap score, 1-sided"  = c("recaudit", "none",     "one_side"),
    "Gap score, 2-sided"  = c("recaudit", "none",     "two_side"))[[method]]
  ids <- as.character(users$user_id)
  tr <- switch(cfg[2], neg_loss = neg_loss, lira = lira, identity)
  tgt <- lapply(get(cfg[1], "target")[ids], tr)
  inn <- lapply(get(cfg[1], "member", seeds)[ids], tr)
  out <- lapply(get(cfg[1], "nonmember", seeds)[ids], tr)
  if (cfg[2] == "corr") {                     # Zhou et al.: confidence + correctness
    hit <- function(l) lapply(l, function(v) as.numeric(v > 0))
    rt  <- get("recaudit", "target")[ids]
    tgt <- Map(function(a, b) h(a + b), tgt, hit(rt))
  }
  x <- audit_stat(cfg[3], tgt, inn, out, pooled)
  y <- users$unforgotten
  structure(c(auroc = auroc(y, x), fpr10 = fpr_at_fnr(y, x)), stat = setNames(x, ids))
}

methods <- c("Yeom et al.", "Zhou et al.", "Sablayrolles et al.", "Watson et al.",
             "Carlini et al.", "RecAudit")
tab <- t(sapply(methods, run))
published <- data.frame(auroc = c(0.5825, 0.5802, 0.7238, 0.6872, 0.6561, 0.9236),
                        fpr10 = c(0.7898, 0.7834, 0.5478, 0.6624, 0.8025, 0.2102))
cat("\n== Replication vs the authors' README table ==\n")
print(round(cbind(tab, readme_auroc = published$auroc, readme_fpr10 = published$fpr10), 4))

# Exact check: recompute AUROC from the authors' own saved likelihoods
cat("\n== AUROC from the authors' saved per-user likelihoods ==\n")
print(round(sapply(split(auth, auth$method), function(d) auroc(d$label, d$likelihood)), 6))

# ---- Extension 1: which ingredient carries the gain? ------------------------
cat("\n== Decomposition: score x test ==\n")
dec <- t(sapply(c("Yeom et al.", "Watson et al.", "Sablayrolles et al.", "Carlini et al.",
                  "Gap score, global", "Gap score, 1-sided", "Gap score, 2-sided", "RecAudit"), run))
print(round(dec, 4))

# ---- Extension 2: audit-model budget ----------------------------------------
set.seed(20261005)
budget <- do.call(rbind, lapply(c(2, 4, 8, 16), function(k) {
  draws <- if (k == 16) list(1:16) else replicate(40, sample(16, k), simplify = FALSE)
  do.call(rbind, lapply(c("Carlini et al.", "RecAudit"), function(m) {
    a <- sapply(draws, function(s) run(m, seeds = s)["auroc"])
    p <- sapply(draws, function(s) run(m, seeds = s, pooled = TRUE)["auroc"])
    data.frame(pairs = k, models = 2 * k, method = m,
               auroc_mean = mean(a), auroc_sd = if (k == 16) 0 else sd(a),
               auroc_pooled_var = mean(p))
  }))
}))
cat("\n== AUROC by number of audit models (mean over 40 random seed subsets) ==\n")
print(budget, digits = 4)

# ---- per-user agreement with the authors' saved likelihoods ----------------
cat("\n== Max |difference| between our per-user statistic and the authors' ==\n")
key <- c("Yeom_et_al" = "Yeom et al.", "Zhou_et_al" = "Zhou et al.",
         "Sablayrolles_et_al" = "Sablayrolles et al.", "Watson_et_al" = "Watson et al.",
         "Carlini_et_al" = "Carlini et al.", "RecAudit" = "RecAudit")
for (k in names(key)) {
  ours <- attr(run(key[[k]]), "stat")
  a <- auth[auth$method == k, ]
  d <- abs(ours[as.character(a$user_id)] - a$likelihood)
  cat(sprintf("%-20s max|diff| = %.3g   users differing > 1e-9: %d   our FPR@10%%FNR on their stat: %.4f (saved %.4f)\n",
              key[[k]], max(d), sum(d > 1e-9), fpr_at_fnr(a$label, a$likelihood), a$fpr_at_10fnr[1]))
}
```

## Personalization and Sensitive Information: A Boundary Condition

A recurring nuance is that the privacy calculus is *type-of-information* specific, not
global. @white2004consumer show that consumers with deeper relational connections to a
firm are more willing to disclose ordinary privacy-related information in exchange for
customized offerings—but the relationship *reverses* for *embarrassing* information:
loyal, deeply connected consumers find the exchange of customized benefits for
embarrassing disclosures *unattractive*. The benefit term $B$ in
@eq-privacy-calculus does not compensate for the cost $C$ when the disclosure
threatens the self, regardless of relationship strength. This is the microfoundation
of the result that benevolence and norm-respect, not raw benefit, govern sensitive
disclosure—and it is why @godinho2022consumer find consumers ring-fencing sensitive
data even as they open up otherwise under GDPR. Privacy, defined as selective control
of access to the self [@altman1976conceptual], is most jealously guarded exactly where
the self is most exposed, and consumers deploy active coping strategies to protect it
when personalization threatens that boundary [@youn2009determinants; @grant2005young].

## Synthesis: Privacy as Strategy

The literature converges on a managerial stance that treats privacy not as a
compliance burden but as a source of advantage—what the organizational work frames as
*privacy as strategy*. Firms that prioritize data privacy authentically, that involve
customers in the privacy dialogue, that align privacy practices across every function
rather than siloing them in legal, that emphasize what they do *right* with data, and
that commit over the long term are predicted to realize positive performance and
higher consumer trust [@martin2017data; @casadesus2015competing]. The behavioral and
economic results explain *why* this works: trust is the proactive lever that flattens
the privacy-cost gradient (@eq-disclosure-logit), control converts personalization
from intrusion into service [@tucker2013], transparency and control insure against the
performance damage of breaches [@martin2017data], and benevolent framing keeps a
privacy notice on the social-contract side of the bulletproof-glass divide
[@brough2022]. The unifying theme is that privacy is governed by *norms of
appropriate flow*, and the firm's job is to be legible and trustworthy within those
norms rather than to minimize disclosure or maximize collection.

## Key Takeaways

- **Privacy is control over information flow** [@westin1967f], measured in practice
  through reflective scales of *privacy concern* [@malhotra2004internet;
  @smith1996information]; the construct and its proxy are distinct, which is what makes
  the privacy paradox statable.
- The **privacy calculus** (@eq-privacy-calculus, @eq-disclosure-logit) models
  disclosure as a benefit–cost trade-off moderated by trust, but disclosure is
  framing-sensitive and boundedly rational, so the **privacy paradox**—stated concern
  diverging from revealed behavior—is the rule, not the exception [@johnson2020;
  @lin2025choice].
- **Control** is the decisive moderator: it makes personalization effective
  [@tucker2013] and reduces concern [@xu2012research], but *perceived* control can
  increase objective exposure [@brandimarte2012economics].
- **Transparency can backfire.** A protective privacy notice read as a *formal*
  contract raises perceived vulnerability—the **bulletproof-glass effect**—unless harm
  is already expected or the notice carries benevolence cues [@brough2022].
- **Regulation reshapes competition and innovation.** GDPR raises opt-in under
  compliant consent yet forces app exit and depresses introduction, trading privacy
  against innovation, with two-sided competitive incidence and limited effect on
  dominant trackers [@godinho2022consumer; @janssen2022gdpr; @kollnig2021before].
- **Deleting the data does not delete what the model learned.** Erasure can be
  audited: comparing a recommender's fit to a user's own history against reference
  models trained with and without that user flags unforgotten users with an AUROC of
  0.92 in the authors' demo, which this chapter reproduces exactly, and a pooled-variance
  version stays near 0.88 with only four reference models [@zhu2026forget].
- Treat **privacy as strategy**: authentic, aligned, long-horizon, control-granting
  practices build the trust that flattens the privacy-cost gradient and insure against
  breach damage [@martin2017data; @casadesus2015competing].
