# Metrics {#sec-metrics}

Marketing research lives and dies by measurement. A construct that cannot be
operationalized into a reproducible number cannot enter a regression, cannot be
priced by equity markets, and cannot be defended to a sceptical reviewer or a
sceptical chief financial officer. This chapter is a working catalogue of the
metrics that quantitative marketing scholars actually compute—the financial
indicators that connect marketing actions to firm value, the firm-level
controls drawn from accounting data, and the increasingly text- and
image-based marketing constructs that machine learning has made measurable at
scale. It is also a methods chapter: for the harder constructs—marketing
capability above all—we give the estimator, the identifying assumptions, and
runnable code that reconstructs the measure from raw Compustat and patent data.

The organizing distinction is between *outcomes* and *inputs to outcomes*. A
metric such as return on marketing investment is an outcome the firm wants to
maximize; metrics such as firm size, leverage, and profitability are covariates
that confound or moderate the marketing–performance relationship and must be
controlled. A third class—capability metrics—are neither directly observed nor
simple ratios of accounting items; they are *latent efficiencies* recovered by
econometric frontier estimation, and they demand the most care. We treat the
three classes in turn, leading each with the intuition for what the number is
supposed to capture and following immediately with its formal definition,
its data source, and the assumptions under which it is identified.

Throughout, the practical substrate is the Wharton Research Data Services (WRDS)
ecosystem: the Center for Research in Security Prices (CRSP) for market data,
Compustat for accounting fundamentals, Thomson Financial 13F filings for
institutional ownership, and the WRDS U.S. patents database for innovation
output. The reader who works through the code will end with a firm-year panel
on which any of the marketing–finance models in @sec-marketing-finance can be
estimated.

## Financial Metrics {#sec-metrics-finance}

Financial metrics translate a marketing intervention into the language that
capital markets and corporate boards speak. They fall into three families:
*return* measures that scale profit by the capital deployed to earn it,
*value-creation* measures that ask whether the firm earned more than its cost of
capital, and *covariate* measures that characterize the firm's size, risk, and
profitability for use as controls.

### Return on Investment and Return on Marketing Investment {#sec-roi}

The most elementary return metric scales net profit by the capital that produced
it. *Return on investment* (ROI) is

$$
\text{ROI} = \frac{\text{Net Profit}}{\text{Investment}},
$$ {#eq-roi}

a unitless ratio interpretable as the profit earned per dollar committed. ROI is
attractive precisely because it is dimensionless and therefore comparable across
projects of different scale, but that virtue is also its central weakness: it
discards the magnitude of the investment, so a tiny project with a spectacular
percentage return can dominate a large project that creates far more absolute
value. ROI is a ranking device, not a value-maximization objective.

For marketing specifically, the analogue isolates the *incremental* effect of the
marketing dollar. *Return on marketing investment* (ROMI), sometimes written
ROIM, is

$$
\text{ROMI} = \frac{\text{IRAM} - \text{CM}}{\text{MS}},
$$ {#eq-romi}

where IRAM is the incremental revenue attributable to marketing, CM is the
contribution cost of the marketing investment, and MS is marketing spending. A
positive ROMI signals that the marketing program returned more than it cost on
the margin. The deceptively simple numerator hides the entire identification
problem of the field: *incremental* revenue is a counterfactual quantity—the
revenue that would not have occurred absent the marketing—and recovering it
requires either an experiment or a credible model of the no-marketing baseline.
Naively crediting marketing with all post-campaign revenue (the endogeneity of
spend with demand) inflates ROMI without bound, which is why the construct is
better understood as a target for causal estimation than as an accounting ratio.

::: {.callout-note}
ROI and ROMI are *summary* statistics of a causal effect, not the effect itself.
Their managerial appeal—a single comparable percentage—is exactly what makes them
easy to game by manipulating the denominator (under-counting the true cost base)
or by attributing organic demand to the campaign. Report them alongside the
identification strategy that produced the incremental numerator.
:::

### Economic Value Added {#sec-eva}

ROI compares profit to capital but is silent on whether that profit cleared the
*cost* of capital. *Economic value added* (EVA), also called economic profit, fills
that gap. It measures the residual wealth a firm generates after deducting a charge
for all the capital it employs, equity as well as debt, on an after-tax basis:

$$
\text{EVA} = \text{NOPAT} - (\text{Invested Capital} \times \text{WACC}).
$$ {#eq-eva}

Here *net operating profit after taxes* (NOPAT) equals operating profit times one
minus the tax rate; *invested capital* is the sum of debt, capital leases, and
shareholders' equity (equivalently, equity plus long-term debt measured at the
start of the period); and WACC is the *weighted average cost of capital*, the
blended return the firm must pay its providers of capital. The product
$\text{WACC} \times \text{Invested Capital}$ is the **finance charge**: the
opportunity cost of the funds tied up in the business. EVA is positive only when
operating performance exceeds that charge, which is the precise sense in which
positive EVA means the firm created value rather than merely earned a profit.

The cost of capital itself is a weighted average over the firm's capital
structure,

$$
\text{WACC} = \frac{K_e \, E}{E + D} + \frac{K_d \,(1 - t)\, D}{E + D},
$$ {#eq-wacc}

where $E$ and $D$ are the market values of equity and debt, $K_e$ is the required
return on equity, $K_d(1-t)$ is the after-tax return on debt, and $t$ is the
marginal tax rate. The after-tax adjustment on debt reflects the tax
deductibility of interest.

Because invested capital can equivalently be written as total assets net of
current liabilities, a common balance-sheet implementation of @eq-eva is

$$
\text{EVA} = \text{NOPAT} - (\text{total assets} - \text{current liabilities}) \times \text{WACC}.
$$ {#eq-eva-bs}

EVA rests heavily on the book value of invested capital, which makes it most
informative for *asset-rich* firms whose balance sheets capture the bulk of the
resources they deploy. For firms whose value resides in intangibles—software,
brands, data, organizational capital—book invested capital understates the true
capital base, and EVA correspondingly overstates value creation. The intangibles
problem is not a minor caveat; it is the reason marketing scholars increasingly
favor market-based metrics over EVA for technology and consumer-brand firms.

### Market Value Added {#sec-mva}

Where EVA is a flow concept measured each period, *market value added* (MVA) is the
corresponding stock: the cumulative wealth the firm has created for its capital
providers since inception. It is the gap between what the market says the firm's
claims are worth and what investors originally contributed:

$$
\text{MVA} = \text{market value of shares (or enterprise value)} - \text{book value of shareholders' equity}.
$$ {#eq-mva}

A firm with persistently positive EVA accumulates positive MVA; under the residual-
income identity, MVA equals the present value of all expected future EVA. MVA thus
sidesteps the period-by-period WACC estimation that EVA requires, at the cost of
inheriting all the noise of market expectations: it tells you what the market
believes the firm has created, not what it has demonstrably created.

### Firm-Level Covariates from Accounting Data {#sec-covariates}

Most marketing–finance studies are observational, so the credibility of any
estimated marketing effect rests on controlling for the firm characteristics that
jointly drive marketing decisions and financial outcomes. The literature has
converged on a standard battery of accounting-based controls, each with a
canonical operationalization. We collect them here with their definitions, their
empirical pedigree, and the source items in Compustat.

**Profitability** is conventionally measured as operating income before
depreciation scaled by total assets, a return-on-assets variant that strips out
the financing and depreciation choices that contaminate net income
[@Grewal_2008; @McAlister_2016]:

$$
\text{profitability} = \frac{\text{operating income before depreciation}}{\text{total assets}}.
$$ {#eq-profitability}

**Firm size** enters as the natural logarithm of total assets, the log
transformation taming the extreme right skew of the firm-size distribution and
rendering coefficients interpretable as elasticities [@Grewal_2010; @McAlister_2016; @Nezami_2018]:

$$
\text{firm size} = \log(\text{total assets}).
$$ {#eq-firm-size}

**Sales growth**, the percentage change in gross sales, proxies for the firm's
demand-side momentum and growth opportunities [@Grewal_2010; @Nezami_2018; @Rao_2004].
**Cash flow**, measured as the log of operating cash flow in millions, captures
internal financing capacity and is a standard control in studies linking marketing
to firm risk [@Chakravarty_2011; @Malshe_2015].

**Financial leverage** scales long-term debt by the book value of assets, indexing
the firm's reliance on debt financing and its exposure to financial distress
[@Kashmiri_2017; @Chakravarty_2011]:

$$
\text{financial leverage} = \frac{\text{long-term debt}}{\text{book value of assets}}.
$$ {#eq-leverage}

**Abnormal stock return** is frequently dichotomized: a dummy equal to one when a
firm's stock return exceeds its industry-averaged return marks out-performers in a
way robust to the heavy tails of raw returns [@Markovitch_2005; @Chakravarty_2011; @Chakravarty_2016].

Beyond these workhorses, specialized studies build bespoke financial constructs.
*Unexpected size-adjusted advertising investment*—the residual from a model of
expected advertising given firm size—isolates the surprise component of marketing
spend that markets have not already priced [@Chakravarty_2016; @Kim_2011; @Liu_2017].
*Shareholder complaints*, drawn from the RiskMetrics governance database, proxy for
investor dissatisfaction and governance friction [@Wies_2019].

### Book Equity {#sec-book-equity}

Book equity (BE) is a deceptively intricate construct because the "right" measure
depends on data availability, and the field's conventions trace to the Fama–French
data library. The canonical definition is worth quoting in full:

> "BE is the book value of stockholders' equity, plus balance sheet deferred taxes
> and investment tax credit (if available), minus the book value of preferred stock.
> Depending on availability, we use the redemption, liquidation, or par value (in
> that order) to estimate the book value of preferred stock. Stockholders' equity is
> the value reported by Moody's or Compustat, if it is available. If not, we measure
> stockholders' equity as the book value of common equity plus the par value of
> preferred stock, or the book value of assets minus total liabilities (in that
> order)." [@davis2000]

The hierarchy of fallbacks is the point: preferred stock is valued by redemption,
then liquidation, then par; stockholders' equity is taken from the cleanest
available source and reconstructed only when necessary. The Compustat
implementation coalesces the preferred-stock items in order of preference and adds
deferred taxes:

```{r}
#| eval: false
library(tidyverse)
# Preferred-stock book value: redemption (pstkrv), then liquidation (pstkl),
# then par (pstk); add deferred taxes and investment tax credit (txdtc).
book_equity <- seq - coalesce(pstkrv, pstkl, pstk, 0) + coalesce(txdtc, 0)
```

### Net Contribution {#sec-net-contribution}

The bridge from marketing spend to profit is the *net contribution* function,
which underlies the advertising-budgeting and response-modeling literature. Net
contribution is gross margin times sales revenue, less the cost of the marketing
that produced those sales:

$$
\text{NC} = m \times S(a) - k a,
$$ {#eq-net-contribution}

where $m$ is gross margin, $S(a)$ is the sales-response function mapping marketing
effort $a$ to sales, and $k$ is the unit cost of effort. The first-order condition
$m \, S'(a^\*) = k$ characterizes the optimal effort $a^\*$: spend until the
marginal gross profit from an additional unit of effort equals its marginal cost.
The shape of $S(\cdot)$—concave, S-shaped, or saturating—governs whether an
interior optimum exists, which is the recurring substantive question of the
advertising-response literature.

### A Reference Map of Compustat Items {#sec-compustat-map}

The metrics above all reduce to combinations of a relatively small set of WRDS
data items. Because reproducing any one of them requires knowing the exact item
mnemonic and its source file, @tbl-compustat-core assembles the core mapping. All
items without a CRSP annotation come from Compustat Fundamentals Annual; the
mnemonics follow the WRDS data-items reference.[^wrds-ref]

[^wrds-ref]: The item definitions follow the WRDS data-items documentation
(`wrds_data_items`). Items prefixed in CRSP/Compustat Merged (e.g., `PRCC_C`,
`CSHO`) are drawn from the merged fundamentals files; market-price and
shares-outstanding items used for market capitalization come from CRSP.

```{r}
#| label: tbl-compustat-core
#| tbl-cap: "Core firm-level metrics and their Compustat/CRSP data items."
#| echo: false
#| message: false
#| warning: false
library(knitr)
core_items <- tibble::tribble(
  ~Metric,                          ~`Data item`,                 ~`Source file`,
  "Book value of equity",           "PRCC_C x CSHO",              "CRSP/Compustat Merged",
  "Capital intensity",              "CAPX / AT",                  "Cash flow; balance sheet",
  "Cash flow",                      "(IBC + DP) / AT",            "Cash flow; income statement",
  "Cash holdings",                  "CHE / AT",                   "Balance sheet",
  "Cost of capital (proxy)",        "XINT / DLC",                 "Income statement; balance sheet",
  "Earnings per share",             "NI / CSHO",                  "Income statement; misc.",
  "Firm size",                      "log(AT)",                    "Balance sheet",
  "Leverage",                       "(DLTT + DLC) / SEQ",         "Balance sheet",
  "Market-to-book ratio",           "MKVALT / BKVLPS",            "Supplemental; balance sheet",
  "Market value",                   "MKVALT or CSHO x PRCC_F",    "Supplemental; misc.",
  "Payout ratio",                   "(DVP + DVC + PRSTKC) / IB",  "Income statement; cash flow",
  "R&D intensity",                  "XRD / AT",                   "Income statement; balance sheet",
  "Return on assets (ROA)",         "NI / AT",                    "Income statement; balance sheet",
  "Return on equity (ROE)",         "NI / (CSHO x PRCC_F)",       "Income statement; supplemental",
  "Return on investment (ROI)",     "NI / ICAPT",                 "Income statement; balance sheet",
  "Tangibility",                    "PPENT / AT",                 "Balance sheet",
  "Tobin's Q",                      "(AT + CSHO x PRCC_F - CEQ) / AT", "Balance sheet; supplemental",
  "Total equity",                   "PSTKC + CSHO",               "Balance sheet; misc."
)
kable(core_items, booktabs = TRUE)
```

A second, finance-oriented battery—used in studies of corporate investment,
financing, and payout policy—follows the conventions catalogued by @kahle2017.
@tbl-kahle reproduces the construction rules, which differ from
@tbl-compustat-core chiefly in their use of *lagged* assets in denominators (to
avoid mechanical contemporaneous correlation) and in the explicit treatment of
missing R&D as zero.

```{r}
#| label: tbl-kahle
#| tbl-cap: "Valuation, investment, financing, and payout metrics following @kahle2017."
#| echo: false
kahle_items <- tibble::tribble(
  ~Category,       ~Metric,                       ~Construction,
  "Valuation",     "Tobin's Q",                   "(AT + CSHO x PRCC_F - CEQ) / AT",
  "Valuation",     "Market cap",                  "prc x shrout (CRSP)",
  "Valuation",     "Revenue Herfindahl",          "revt_i^2 / sum(revt) within 3-digit NAICS x year",
  "Investment",    "CapEx / assets",              "capx / lag(at)",
  "Investment",    "R&D / assets",                "xrd / lag(at); missing R&D set to 0",
  "Investment",    "Fixed assets / assets",       "ppent / at",
  "Investment",    "Cash / assets",               "che / at",
  "Profitability", "Operating cash flow / assets","(oibdp - xint - txt) / lag(at)",
  "Profitability", "ROA",                         "ib / at",
  "Financing",     "Book leverage",               "(dltt + dlc) / at",
  "Financing",     "Market leverage",             "(dltt + dlc) / (at - ceq + che x prcc_f)",
  "Financing",     "Net leverage",                "(dltt + dlc - che) / at",
  "Financing",     "Net equity issuance",         "(sstk - prstkc) / lag(at)",
  "Ownership",     "Institutional ownership",     "% shares held by institutions (13F)",
  "Ownership",     "Blockholder",                 "institution holding >= 10% of shares (13F)",
  "Payout",        "Dividends / assets",          "dvc / lag(at)",
  "Payout",        "Repurchase / assets",         "(prstkc - pstk) / lag(at)",
  "Payout",        "Total payout / assets",       "(dvc + prstkc) / lag(at)"
)
kable(kahle_items, booktabs = TRUE)
```

Institutional-ownership and blockholder variables come from Thomson Financial 13F
filings; the macro denominator for the market-cap/GDP ratio is series GDPA from the
U.S. Bureau of Economic Analysis.

### Industry Concentration and Diversity {#sec-diversity}

Several of the metrics above require an industry-concentration index, and the
literature has settled on the Herfindahl form. For revenue shares $s_i$ within an
industry, the Herfindahl–Hirschman index is $\sum_i s_i^2$; the
revenue-Herfindahl in @tbl-kahle computes this within each 3-digit NAICS industry
and year. The same quantity, applied to sector market shares, is the **Simpson
diversity index** familiar from ecology—the two are algebraically identical, both
equal to the probability that two randomly drawn units belong to the same
category. High Herfindahl (low Simpson diversity) signals a concentrated industry;
the index thus does double duty as a competition control and as a measure of how
diversified a firm's or market's activity is across sectors.

## Marketing Metrics {#sec-metrics-marketing}

The financial metrics above are computed from structured accounting data. The
marketing constructs in this section are different in kind: trust, sentiment,
willingness to pay, purchase intention, and brand reputation are *psychological*
states that classically required surveys to measure but are now increasingly
recovered from unstructured text and images by machine learning. The
methodological frontier here is the use of pre-trained language and vision models
to convert social-media content into validated marketing measures at a scale
surveys cannot reach.

### Trust {#sec-trust}

Trust is the willingness to rely on an exchange partner in whom one has confidence,
and it has long been measured by multi-item attitudinal scales. The frontier
substitutes behavioral and relational signals in social-media data for
self-report: @roy2017 develop an algorithmic measure of brand trust from the
structure and content of consumers' social-media interactions, demonstrating that
trust leaves a computational trace that can be extracted without surveying anyone.

### Sentiment {#sec-23-sentiment}

Sentiment—the valence of expressed affect—is core to human communication and is the
single most demanded text-analytic measure in marketing, applied to social media,
news, customer feedback, and corporate communications. The central practical
question is *which* method to use, because the menu ranges from simple lexicons
that map words to polarity scores to transfer-learning language models that are
far more accurate but far more demanding.

@hartmann2023more resolve this question empirically with a meta-analysis spanning
272 datasets and roughly twelve million sentiment-labeled documents. Their headline
finding is that transfer-learning models—pre-trained transformers fine-tuned on
sentiment—deliver the best performance, outperforming lexicons by more than twenty
percent in accuracy on average. The advantage is not uniform: it widens with the
number of sentiment classes and is moderated by text length, and the
leaderboard-topping benchmark model is not always the best choice for a given
research setting. Crucially for reproducibility, the authors supply a pre-trained
model (SiEBERT) and open-source scripts, lowering the barrier to applying
state-of-the-art sentiment analysis. The practical lesson is that method choice
should be made deliberately against the research question, the data, and the
available computational resources—not by reflexively reaching for a lexicon because
it is convenient.

### Willingness to Pay {#sec-wtp}

Willingness to pay (WTP)—the maximum price a consumer will accept for a good—is the
demand-side primitive that underlies pricing and welfare analysis. Recovering it
from naturally occurring text rather than from elicitation experiments is an active
frontier; @he2023express represents recent work in this direction, extracting WTP
signals from expressed consumer language.

### Purchase Intention {#sec-purchase-intention}

Purchase intention—the self-reported likelihood of buying—is a leading indicator of
behavior and a workhorse dependent variable. @hartmann2021power show that it can be
inferred from *images*, not just text, and in doing so overturn a common assumption
about social-media metrics. Smartphones have made it trivial for consumers to share
branded imagery, and the authors classify that imagery into three types using
convolutional neural networks: **packshots** (the product alone), **consumer
selfies** (a consumer's face shown with the brand), and **brand selfies** (the
product held from the consumer's own visual perspective, with no consumer face
visible). Applying language models to social-media responses across more than
250,000 brand-image posts from 185 brands on Twitter and Instagram, they find a
revealing dissociation: consumer selfies generate more likes and comments, but
brand selfies induce higher *purchase intentions*. Engagement metrics and purchase
intent diverge.

The dissociation has managerial bite. In a display-advertising field test, brand
selfies earned higher click-through rates than consumer selfies, and a laboratory
experiment traced the mechanism to self-reference: the first-person perspective of
the brand selfie invites the viewer to imagine holding the product themselves. The
broader methodological point is that machine learning can decode marketing-relevant
constructs from multimedia content, and that traditional engagement counts (likes,
comments) may *mislead* about the constructs managers actually care about. The
purchase-intention classifier is released as a fine-tuned RoBERTa model.[^purchase-model]

[^purchase-model]: The fine-tuned purchase-intention model is distributed publicly
as a large RoBERTa checkpoint, allowing researchers to score new English-language
text for expressed purchase intent without retraining.

### Brand Reputation {#sec-brand-reputation}

Brand reputation—the aggregate esteem in which a brand is held—has migrated from
survey trackers to real-time social listening. @rust2021 measure brand reputation
directly from Twitter data, demonstrating that a construct historically captured by
periodic, expensive surveys can be estimated continuously from public conversation,
with the attendant gains in timeliness and the attendant risks of platform-specific
selection.

## Marketing Capability {#sec-capability}

Capability is the most demanding metric in this chapter and the one that most
rewards careful estimation, because it is not observed at all—it is a latent
*efficiency* that must be inferred from the gap between what a firm achieves and
what the best firms achieve with comparable inputs. The construct originates with
@dutta1999, who define a firm's capability as "its ability to deploy the resources
(inputs) available to it to achieve the desired objective(s) (output)." The higher
a firm's functional capability, the more efficiently it converts its inputs into
the relevant functional output; equivalently, the lower its functional
*inefficiency*, the higher its capability. This input–output framing is what makes
stochastic frontier analysis the natural estimator.

### The Substantive Argument {#sec-capability-theory}

The motivating insight of @dutta1999 is that in high-technology markets, raw
technological prowess is not enough. A firm can possess formidable research and
development (R&D) capability—generating a stream of high-quality innovations—yet
fail commercially because it lacks the *marketing* capability to translate those
innovations into products consumers value and buy. Marketing capability has its
largest effect on quality-adjusted innovation output precisely for firms with a
strong technological base: the firms that benefit most from great marketing
capability are those that already have a strong R&D foundation, because only they
have innovations worth commercializing. The interaction of marketing and R&D
capabilities is therefore the single most important determinant of firm
performance—high-technology firms must be able both to generate innovation
continuously and to commercialize it.

This argument dictates the estimation strategy. Three capabilities are estimated
jointly because they feed one another: **marketing capability** drives sales from a
firm's technological base, advertising stock, marketing stock, customer
relationships, and installed base; **R&D capability** drives quality-adjusted
technological output; and **operations capability** drives the cost of production.
The functional relationships @dutta1999 posit are, schematically,

$$
\text{Sales} = f(\text{technological base},\ \text{advertising stock},\ \text{marketing stock},\ \text{customer relationships},\ \text{installed base}),
$$

$$
\text{Quality-adjusted output} = f(\text{technological base},\ \text{cumulative R\&D},\ \text{marketing capability}),
$$

$$
\text{Cost of production} = f(\text{output},\ \text{cost of capital},\ \text{labor cost},\ \text{technological base},\ \text{marketing capability}).
$$

The appearance of marketing capability inside the R&D and operations equations is
the formal expression of the substantive claim that the capabilities are
interdependent, not separable.

@fig-capability-flow makes the recursive structure explicit.

```{mermaid}
%%| label: fig-capability-flow
%%| fig-cap: "The interdependent capability system of @dutta1999. Stocks of marketing, advertising, and R&D investment feed three stochastic frontiers; estimated marketing efficiency enters the R&D and operations frontiers, capturing the interdependence of capabilities."
flowchart TD
    A[Advertising stock] --> M[Marketing frontier:<br/>log sales]
    MK[Marketing stock] --> M
    TB[Technological base] --> M
    REC[Receivables / CRM] --> M
    IB[Installed base] --> M
    M --> ME[Marketing efficiency]
    ME --> R[R&D frontier:<br/>log tech output]
    RD[R&D stock] --> R
    TB --> R
    R --> RE[R&D efficiency]
    ME --> O[Operations frontier:<br/>log COGS]
    LC[Labor cost] --> O
    CC[Cost of capital] --> O
    TB --> O
    O --> OE[Operations efficiency]
```

### Stock Variables and the Koyck Transformation {#sec-koyck}

Marketing and innovation investments do not affect sales only in the year they are
incurred; their effect decays geometrically over time. The standard device for
turning a flow of expenditure into a *stock* is the Koyck geometric-lag
transformation [@koyck1954distributed]. For an expenditure flow $x_t$ and a
retention rate $\lambda \in (0,1)$, the stock is

$$
\text{Stock}_t = \sum_{j=0}^{t-1} \lambda^{j} \, x_{t-j},
$$ {#eq-koyck}

so that each past dollar contributes $\lambda^j$ of its original force after $j$
years. The advertising-stock literature anchors $\lambda$ empirically: weights of
0.4 [@peles1971] and 0.5 [@wang2017] are standard for advertising stock, and
@dutta2005 use a weight of 0.5 for marketing expenditure and 0.4 for R&D
expenditure (their p. 281). The choice of $\lambda$ is consequential—too high a
retention rate over-credits ancient spending—and should be defended against the
estimated carryover in the relevant category rather than imposed by habit.

### Measuring R&D Output: Innovativeness and Width {#sec-rd-output}

A raw patent count is a poor measure of technological output because patents differ
enormously in quality; @dutta1999 deliberately avoid it in favor of two
citation-based, quality-adjusted measures. The first, **innovativeness**, follows
the citation-weighting tradition of @trajtenberg1990 and @trajtenberg1990a: a patent
that is cited often is more valuable, so patents are weighted by how far their
citation count exceeds the industry norm. The second, **width of applicability**,
follows @jaffe1993: a patent cited by firms in *other* industries has broader
applicability, so patents are weighted by the share of their citations that come
from outside the focal industry.

Concretely, the **innovativeness-adjusted** output is built in three steps. First,
compute the average number of citations received by all sample patents within an
industry—defined at one-, two-, three-, and four-digit SIC granularity—where the
original study used a single mean across all firms and years. Second, weight each
firm's patent by its citation count divided by that industry-sample average. Third,
sum the citation-weighted patents within a firm-year. The **width-adjusted** output
parallels this: for each patent compute the proportion of its citations originating
outside the focal SIC code; weight the patent by that proportion divided by the
industry-average proportion; and sum the weighted patents within a firm-year.

### The Stochastic Frontier Estimator {#sec-sfa}

The three capabilities are recovered by *stochastic frontier analysis* (SFA), which
is the right tool precisely because capability is defined as efficiency relative to
a best-practice frontier. For firm $i$ in year $t$ with output $y_{it}$ and inputs
$\mathbf{x}_{it}$, the production frontier is

$$
\log y_{it} = \mathbf{x}_{it}' \boldsymbol{\beta} + v_{it} - u_{it},
$$ {#eq-sfa}

where the composed error decomposes into a symmetric noise term
$v_{it} \sim \mathcal{N}(0, \sigma_v^2)$ and a one-sided inefficiency term
$u_{it} \ge 0$. The frontier $\mathbf{x}_{it}'\boldsymbol{\beta} + v_{it}$ is the
maximum attainable output; the firm falls short of it by $u_{it}$. Capability is
then the technical efficiency

$$
\text{TE}_{it} = \exp(-u_{it}) \in (0, 1],
$$ {#eq-efficiency}

recovered as the conditional expectation $\mathbb{E}[\exp(-u_{it}) \mid v_{it} -
u_{it}]$ following Jondrow and others. For a *production* frontier, inefficiency
*decreases* output, so the sign on $u_{it}$ is negative (`ineffDecrease = TRUE` in
the code below); for a *cost* frontier—used for the operations capability, where the
output is the cost of goods sold to be *minimized*—inefficiency *increases* cost and
the sign reverses (`ineffDecrease = FALSE`).

The identifying assumptions are exactly those that make SFA both powerful and
fragile. First, the inefficiency term must be one-sided and distributionally
specified (half-normal, truncated-normal, or exponential); the efficiency estimates
are not robust to gross misspecification of this distribution. Second, the inputs
$\mathbf{x}_{it}$ must be exogenous to $u_{it}$—if firms with high latent capability
systematically invest more, the frontier is biased and the recovered efficiencies
absorb the endogeneity. Third, the log specification requires strictly positive
inputs and outputs, which is why the data pipeline below replaces zeros and missing
stocks with small positive constants. None of these assumptions is innocuous, and
the interpretation of the recovered efficiencies as "capability" is only as good as
the frontier is correctly specified.

### Data Construction {#sec-capability-data}

We now build the firm-year panel from WRDS. The pipeline connects to the WRDS
PostgreSQL server, pulls Total Q (a refined Tobin's Q), the Compustat fundamentals
needed for the input stocks, and the patent data needed for the R&D-output measures,
and assembles them into a single panel. The replication is constrained by the WRDS
U.S. patents coverage (2011–2019 for the citation files), so the post-2010 window is
used for estimation; the original 1985–1994 study period would require building a
name-matching algorithm against raw USPTO bulk data.

```{r}
#| eval: false
library(RPostgres)
library(tidyverse)

# WRDS connection. Supply your own credentials via environment variables;
# never hard-code passwords into a reproducible script.
wrds <- dbConnect(
  Postgres(),
  host = "wrds-pgdata.wharton.upenn.edu",
  port = 9737,
  dbname = "wrds",
  sslmode = "require",
  user = Sys.getenv("wrds_user"),
  pass = Sys.getenv("wrds_pass")
)
```

Total Q provides the valuation outcome used in some replications below.

```{r}
#| eval: false
#| collapse: true
res <- dbSendQuery(wrds, "SELECT DISTINCT gvkey, fyear, q_tot
                          FROM totalq_all.total_q
                          WHERE fyear >= 2000 AND
                                gvkey IS NOT NULL AND
                                q_tot IS NOT NULL")
totalq <- dbFetch(res, n = -1)
dbClearResult(res)
totalq |> write_rds(file.path("data", "totalq.rds"))
```

```{r}
#| message: false
library(tidyverse)
totalq <- read_rds(file.path("data", "totalq.rds")) |>
  rename(year = fyear)
```

The fundamentals pull retrieves the income-statement and balance-sheet items needed
to build the input stocks, joins industry classifications, and interpolates missing
firm-years. Spline interpolation is used in preference to linear interpolation
because the underlying series (advertising, R&D, sales) are smooth and trending;
linear fill would introduce kinks that the frontier would misread as inefficiency.

```{r}
#| eval: false
#| collapse: true
res <- dbSendQuery(
  wrds,
  "SELECT DISTINCT gvkey, fyear, conm, curcd, cogs, rect, revt, sale,
                   xad, xrd, xsga, ppegt, emp, act, xopr, xint, dlc,
                   xlr, uxintd, invfg
   FROM comp_na_daily_all.funda
   WHERE fyear >= 2000 AND gvkey IS NOT NULL AND
         sale IS NOT NULL AND sale > 0 AND revt IS NOT NULL"
)
capability <- dbFetch(res, n = -1)
dbClearResult(res)

# Industry classifications.
res <- dbSendQuery(wrds,
  "SELECT gvkey, gind, gsubind, naics, sic
   FROM comp_na_daily_all.names")
ind <- dbFetch(res, n = -1)
dbClearResult(res)

capability <- capability |>
  left_join(ind, by = join_by(gvkey)) |>
  rename(year = fyear) |>
  unique() |>
  arrange(gvkey, year) |>
  group_by(gvkey, year) |>
  slice(1) |>           # one record per firm-year
  ungroup()

# Spline interpolation across the non-missing span of each series.
spline_interpolate <- function(x) {
  nz <- which(!is.na(x))
  if (length(nz) == 0) return(x)
  first_non_na <- nz[1]
  last_non_na  <- nz[length(nz)]
  x[first_non_na:last_non_na] <-
    zoo::na.spline(x[first_non_na:last_non_na], na.rm = TRUE)
  x
}
library(zoo)

capability <- capability |>
  group_by(gvkey) |>
  arrange(year, .by_group = TRUE) |>
  fill(conm, curcd, gind, gsubind, naics, sic, .direction = "downup") |>
  ungroup() |>
  group_by(gvkey) |>
  complete(year = min(year):max(year)) |>     # fill gaps within firm
  arrange(year, .by_group = TRUE) |>
  fill(conm, curcd, gind, gsubind, naics, sic, .direction = "downup") |>
  mutate(across(
    c(xrd, xsga, xad, emp, ppegt, xint, act, invfg, xlr,
      cogs, rect, revt, sale, xopr, dlc),
    spline_interpolate
  )) |>
  ungroup() |>
  arrange(gvkey, year)

capability |> write_rds(file.path("data", "capability", "capability.rds"))
```

```{r}
capability <- read_rds(file.path("data", "capability", "capability.rds"))
```

The patent–firm linkage assigns patents to `gvkey`s, keeping for each patent the
match with the highest WRDS confidence score and breaking ties deterministically.
Annual patent counts per firm follow.

```{r}
#| eval: false
#| collapse: true
res <- dbSendQuery(wrds,
  "SELECT gvkey, link_bdate, patnum, wrds_score
   FROM wrdsapps_patents.uspatents_gvkey_linking
   WHERE gvkey IS NOT NULL")
wrdsapps_patents_link <- dbFetch(res, n = -1) |>
  group_by(patnum) |> slice_max(order_by = wrds_score, n = 1) |> ungroup() |>
  group_by(patnum) |> slice(1) |> ungroup()
dbClearResult(res)

res <- dbSendQuery(wrds,
  "SELECT DISTINCT patnum, grantdate, cited_patnum, cited_pat_gdate, cite_type
   FROM wrdsapps_patents.uspatents_citations")
wrdsapps_patents_citations <- dbFetch(res, n = -1)
dbClearResult(res)

patent_output <- wrdsapps_patents_link |>
  mutate(year = year(link_bdate)) |>
  group_by(gvkey, year) |>
  summarise(pat_count = n(), .groups = "drop") |>
  complete(gvkey, year = range(year)) |>
  mutate(across(-c(gvkey, year), ~ replace_na(., 0)))

patent_output |> write_rds(file.path("data", "capability", "patent_output.rds"))
```

```{r}
patent_output <- read_rds(file.path("data", "capability", "patent_output.rds"))
```

#### Innovativeness-adjusted output {#sec-tech-innv}

Implementing the three-step innovativeness measure requires the industry citation
averages at each SIC granularity. The following computes citations per patent per
year, the industry averages at one- through four-digit SIC, the per-patent weights,
and finally the firm-year sums.

```{r}
#| eval: false
# Citations received per patent per year, with firm and industry attached.
df_cite_patent_year <- wrdsapps_patents_citations |>
  group_by(year = year(grantdate), patnum = cited_patnum) |>
  summarise(cite_count = n(), .groups = "drop") |>
  inner_join(wrdsapps_patents_link, by = join_by(patnum)) |>
  filter(year(link_bdate) <= year) |>     # patent predates the citing year
  inner_join(ind |> select(gvkey, sic) |> na.omit() |> unique(),
             by = "gvkey") |>
  select(-c(link_bdate, wrds_score)) |>
  unique()

# Industry-year average citations at each SIC granularity.
ind_avg_patent_sic <- df_cite_patent_year |>
  group_by(year, sic) |>
  summarize(ind_avg_year_patent_sic = mean(cite_count, na.rm = TRUE),
            .groups = "drop")
ind_avg_patent_sic3 <- df_cite_patent_year |>
  mutate(sic3 = substr(sic, 1, 3)) |>
  group_by(year, sic3) |>
  summarize(ind_avg_year_patent_sic3 = mean(cite_count, na.rm = TRUE),
            .groups = "drop")
ind_avg_patent_sic2 <- df_cite_patent_year |>
  mutate(sic2 = substr(sic, 1, 2)) |>
  group_by(year, sic2) |>
  summarize(ind_avg_year_patent_sic2 = mean(cite_count, na.rm = TRUE),
            .groups = "drop")
ind_avg_patent_sic1 <- df_cite_patent_year |>
  mutate(sic1 = substr(sic, 1, 1)) |>
  group_by(year, sic1) |>
  summarize(ind_avg_year_patent_sic1 = mean(cite_count, na.rm = TRUE),
            .groups = "drop")

# Per-patent weights = citations / industry average; sum within firm-year.
tech_innv <- df_cite_patent_year |>
  mutate(sic1 = substr(sic, 1, 1),
         sic2 = substr(sic, 1, 2),
         sic3 = substr(sic, 1, 3)) |>
  inner_join(ind_avg_patent_sic,  by = join_by(year, sic)) |>
  inner_join(ind_avg_patent_sic3, by = join_by(year, sic3)) |>
  inner_join(ind_avg_patent_sic2, by = join_by(year, sic2)) |>
  inner_join(ind_avg_patent_sic1, by = join_by(year, sic1)) |>
  mutate(
    weight_year_patent_sic  = cite_count / ind_avg_year_patent_sic,
    weight_year_patent_sic3 = cite_count / ind_avg_year_patent_sic3,
    weight_year_patent_sic2 = cite_count / ind_avg_year_patent_sic2,
    weight_year_patent_sic1 = cite_count / ind_avg_year_patent_sic1
  ) |>
  group_by(gvkey, year) |>
  summarise(
    tech_inv_patent      = sum(weight_year_patent_sic,  na.rm = TRUE),
    tech_inv_patent_sic3 = sum(weight_year_patent_sic3, na.rm = TRUE),
    tech_inv_patent_sic2 = sum(weight_year_patent_sic2, na.rm = TRUE),
    tech_inv_patent_sic1 = sum(weight_year_patent_sic1, na.rm = TRUE),
    .groups = "drop"
  )

tech_innv |> write_rds(file.path("data", "capability", "tech_innv.rds"))
```

```{r}
tech_innv <- read_rds(file.path("data", "capability", "tech_innv.rds"))
# Firms with no patents in a year get a small positive value so that
# the log frontier is defined.
tech_innv <- tech_innv |>
  complete(gvkey, year = min(tech_innv$year):max(tech_innv$year)) |>
  mutate(across(-c(gvkey, year), ~ replace_na(., 0.1)))
```

#### Width-of-applicability output {#sec-tech-width}

The width measure counts, for each patent, the share of its citations that come
from firms in a *different* SIC code, then weights by that share relative to the
industry-average share.

```{r}
#| eval: false
tech_width_all <- wrdsapps_patents_citations |>
  inner_join(
    wrdsapps_patents_link |> select(-wrds_score) |> rename(gvkey_cite = gvkey),
    by = join_by(patnum)) |>
  select(-link_bdate) |>
  inner_join(
    wrdsapps_patents_link |> select(-wrds_score) |> rename(gvkey_cited = gvkey),
    by = join_by(cited_patnum == patnum)) |>
  filter(year(cited_pat_gdate) >= year(link_bdate)) |>
  select(-link_bdate) |>
  inner_join(ind |> select(gvkey, sic) |> rename(sic_cite = sic),
             by = join_by(gvkey_cite == gvkey)) |>
  inner_join(ind |> select(gvkey, sic) |> rename(sic_cited = sic),
             by = join_by(gvkey_cited == gvkey)) |>
  mutate(outside = if_else(sic_cite == sic_cited, 0, 1)) |>
  unique()

df_cite_patent_year_outside <- tech_width_all |>
  filter(outside == 1) |>
  group_by(patnum = cited_patnum, year = year(grantdate),
           gvkey = gvkey_cited, sic = sic_cited) |>
  summarise(outside_cite_count = n(), .groups = "drop") |>
  unique()

tech_width <- df_cite_patent_year_outside |>
  inner_join(df_cite_patent_year, by = join_by(patnum, year, gvkey, sic)) |>
  unique() |>
  mutate(prop = outside_cite_count / cite_count) |>
  mutate(mean_prop = mean(prop)) |>
  mutate(weight = prop / mean_prop, weighted_patent = weight * cite_count) |>
  group_by(gvkey, year) |>
  summarize(tech_width = sum(weighted_patent), .groups = "drop")

tech_width |> write_rds(file.path("data", "capability", "tech_width.rds"))
```

```{r}
tech_width <- read_rds(file.path("data", "capability", "tech_width.rds"))
min_year <- min(tech_width$year); max_year <- max(tech_width$year)
tech_width <- tech_width |>
  complete(gvkey, year = min_year:max_year) |>
  mutate(across(-c(gvkey, year), ~ replace_na(., 0.1)))
rm(min_year, max_year)
```

#### Historical patent output {#sec-patent-hist}

To extend coverage backward, a historical patent file (1981–2012) drawn from the
NBER/Searle dynamic-assignee data supplements the WRDS counts.[^hist-data] The two
sources are reconciled into a single patent-count series, preferring the historical
count where both exist.

[^hist-data]: The two large input files (`dynass_nber_searle.dta` and
`patents_conveyance.dta`) are too large to redistribute here but are publicly
archived; the code reconstructs the firm-year patent counts from them.

```{r}
#| eval: false
patent_output_hist <-
  rio::import(file.path("data", "capability", "patents_conveyance.dta")) |>
  select(h_assignee_code, grant_date, patentid) |>
  na.omit() |>
  inner_join(
    rio::import(file.path("data", "capability", "dynass_nber_searle.dta")) |>
      select(gvkey1, h_assignee_code) |> na.omit(),
    by = join_by(h_assignee_code)) |>
  mutate(year = year(grant_date)) |>
  rename(gvkey = gvkey1) |>
  group_by(gvkey, year) |>
  summarise(patent_count = n(), .groups = "drop") |>
  complete(gvkey, year = range(year)) |>
  mutate(across(-c(gvkey, year), ~ replace_na(., 0)))

patent_output_hist |>
  write_rds(file.path("data", "capability", "patent_output_hist.rds"))
```

```{r}
patent_output_hist <-
  read_rds(file.path("data", "capability", "patent_output_hist.rds"))

patent_output_total <- patent_output |>
  full_join(patent_output_hist, by = join_by(gvkey, year)) |>
  mutate(pat_count = case_when(
    !is.na(patent_count) ~ patent_count,
    !is.na(pat_count)    ~ pat_count,
    TRUE                 ~ 0
  )) |>
  select(-patent_count) |>
  arrange(gvkey, year) |>
  filter(year >= 2000)
```

#### Building the input stocks {#sec-build-stocks}

The final construction step applies the Koyck transformation of @eq-koyck (with
$\lambda = 0.4$) to each investment flow, producing the patent, technology,
marketing, advertising, installed-base, and R&D stocks, and imputes the
cost-of-capital and labor-cost inputs from industry–year means where firm-level
values are missing. The cost of capital is proxied by interest expense over
current debt ($\text{xint}/\text{dlc}$) and labor cost by per-employee staff
expense ($\text{xlr}/\text{emp}$); both are filled hierarchically from four-digit
down to one-digit SIC industry averages so that no firm-year is dropped for a
single missing input.

```{r}
#| eval: false
lambda <- 0.4

df_cap <- capability |>
  mutate(sic1 = substr(sic, 1, 1),
         sic2 = substr(sic, 1, 2),
         sic3 = substr(sic, 1, 3)) |>
  filter(xad >= 0, xrd >= 0) |>
  left_join(tech_innv,           by = join_by(gvkey, year)) |>
  left_join(tech_width,          by = join_by(gvkey, year)) |>
  left_join(patent_output_total, by = join_by(gvkey, year)) |>
  mutate(
    pat_count     = if_else(is.na(pat_count), 0, pat_count),
    dlc           = if_else(dlc == 0, NA, dlc),
    costofcapital = xint / dlc,            # interest expense / current debt
    emp           = if_else(emp == 0, NA, emp),
    xlr           = if_else(xlr <= 0, NA, xlr),
    laborcost     = xlr / emp
  )

# Hierarchical industry-year imputation of cost of capital and labor cost.
impute_grp <- function(df, ...) {
  df |>
    group_by(...) |>
    mutate(
      costofcapital = ifelse(is.na(costofcapital),
                             mean(costofcapital, na.rm = TRUE), costofcapital),
      laborcost     = ifelse(is.na(laborcost),
                             mean(laborcost, na.rm = TRUE), laborcost)
    ) |>
    ungroup()
}
df_cap <- df_cap |>
  impute_grp(sic, year)  |> impute_grp(sic3, year) |>
  impute_grp(sic2, year) |> impute_grp(sic1, year)

# Koyck geometric-lag stocks (equation @eq-koyck) for each investment flow.
koyck_stock <- function(x, lambda) {
  map_dbl(seq_along(x), ~ {
    if (all(is.na(x[1:.x]))) NA_real_
    else sum(x[1:.x] * lambda ^ (.x - seq_along(x[1:.x])), na.rm = TRUE)
  })
}
df_cap <- df_cap |>
  group_by(gvkey) |>
  arrange(year, .by_group = TRUE) |>
  mutate(
    pat_stock     = koyck_stock(pat_count,      lambda),
    techbase_innv = koyck_stock(tech_inv_patent, lambda),
    techbase_width= koyck_stock(tech_width,     lambda),
    marstock      = koyck_stock(xsga,           lambda),
    adstock       = koyck_stock(xad,            lambda),
    installedbase = koyck_stock(sale,           lambda),
    rdstock       = koyck_stock(xrd,            lambda)
  ) |>
  ungroup()

df_cap |> write_rds(file.path("data", "capability", "df_cap.rds"))
```

### Estimating the Frontiers: @dutta1999 Replication {#sec-dutta-replication}

With the panel built, the three frontiers of @eq-sfa are estimated with the
`frontier` package. Each capability is estimated twice—once with the
innovativeness-adjusted technology measure and once with the width-adjusted
measure—so that the robustness of the recovered efficiencies to the
output-quality definition can be assessed. The marketing frontier is a production
frontier (maximize sales), the operations frontier is a cost frontier (minimize
cost of goods sold), and the recovered marketing efficiency is fed forward into
the R&D and operations frontiers, operationalizing the interdependence of
capabilities.

```{r}
#| message: false
#| warning: false
library(plm)
library(tidyverse)
df_cap <- read_rds(file.path("data", "capability", "df_cap.rds"))

df_cap_panel_dutta <- df_cap |>
  filter(year > 2010) |>
  pdata.frame(c("gvkey", "year"))
```

```{r}
#| message: false
#| warning: false
library(frontier)

# Marketing capability: production frontier for log sales.
cap_mar_innv <- frontier::sfa(
  log(sale) ~ log(adstock) + log(marstock) + log(techbase_innv) +
    log(rect) + log(installedbase) + sic1 + factor(year),
  ineffDecrease = TRUE,   # maximize: inefficiency lowers output
  timeEffect = TRUE,
  data = df_cap_panel_dutta
)
df_cap_panel_dutta$mar_eff_innv <-
  frontier::efficiencies(cap_mar_innv, asInData = TRUE)

cap_mar_width <- frontier::sfa(
  log(sale) ~ log(adstock) + log(marstock) + log(techbase_width) +
    log(rect) + log(installedbase) | sic1 + factor(year),
  ineffDecrease = TRUE, timeEffect = TRUE, data = df_cap_panel_dutta
)
df_cap_panel_dutta$mar_eff_width <-
  frontier::efficiencies(cap_mar_width, asInData = TRUE)

# R&D capability: marketing efficiency enters as an input (interdependence).
cap_rd_innv <- sfa(
  log(tech_inv_patent) ~ log(techbase_innv) + log(rdstock) +
    log(mar_eff_innv) + log(mar_eff_innv) * log(rdstock) |
    sic1 + factor(year),
  timeEffect = TRUE, ineffDecrease = TRUE, data = df_cap_panel_dutta
)
df_cap_panel_dutta$rd_eff_innv <-
  frontier::efficiencies(cap_rd_innv, asInData = TRUE)

cap_rd_width <- sfa(
  log(tech_width) ~ log(techbase_width) + log(rdstock) +
    log(mar_eff_width) + log(mar_eff_width) * log(rdstock) |
    sic1 + factor(year),
  timeEffect = TRUE, ineffDecrease = TRUE, data = df_cap_panel_dutta
)
df_cap_panel_dutta$rd_eff_width <-
  frontier::efficiencies(cap_rd_width, asInData = TRUE)

# Operations capability: cost frontier for log COGS (minimize).
cap_op_innv <- sfa(
  log(cogs) ~ log(invfg) + log(laborcost) + log(costofcapital) +
    log(techbase_innv) + log(mar_eff_innv) | sic1 + factor(year),
  timeEffect = TRUE, ineffDecrease = FALSE,  # minimize: inefficiency raises cost
  data = df_cap_panel_dutta
)
df_cap_panel_dutta$op_eff_innv <-
  frontier::efficiencies(cap_op_innv, asInData = TRUE)

cap_op_width <- sfa(
  log(cogs) ~ log(invfg) + log(laborcost) + log(costofcapital) +
    log(techbase_width) + log(mar_eff_width) | sic1 + factor(year),
  timeEffect = TRUE, ineffDecrease = FALSE, data = df_cap_panel_dutta
)
df_cap_panel_dutta$op_eff_width <-
  frontier::efficiencies(cap_op_width, asInData = TRUE)

df_cap_panel_dutta |>
  write_rds(file.path("data", "capability", "df_cap_panel_dutta.rds"))
```

The recovered efficiencies can then be correlated with each other and with sales to
check that they behave as capability measures should—positively associated with
performance and only moderately correlated with one another, since a firm strong in
marketing need not be strong in R&D.

```{r}
#| eval: false
df_cap_panel_dutta |> select(contains("eff")) |> na.omit() |> cor()
```

An alternative implementation uses the `sfa` package in place of `frontier`; the two
agree up to numerical tolerance.

```{r}
#| eval: false
library(sfa)
```

### Alternative Capability Specifications {#sec-capability-alt}

The Dutta frontier is one of several closely related specifications, and the panel
built above supports the others with minor changes to the input set. We give three
published variants, each differing in how output is measured and which stocks enter
the frontier; the substantive payoff is that the recovered capability efficiencies
are reassuringly correlated across specifications, lending the construct convergent
validity.

#### @saboo2017assessing {#sec-saboo}

This specification uses lagged sales as the installed-base proxy and patent counts
(rather than citation-weighted output) for the R&D frontier; the operations frontier
is a standard cost frontier in operating expense. It is computationally heavy
because of the lag structure.

```{r}
#| eval: false
library(frontier)
df_cap_panel_saboo <- df_cap |>
  group_by(gvkey) |> arrange(year) |>
  mutate(sale_t_1 = dplyr::lag(sale, n = 1)) |>
  ungroup() |>
  pdata.frame(c("gvkey", "year"))

cap_mar <- sfa(
  log(sale) ~ log(xsga) + log(rect) + log(sale_t_1) | sic1 + factor(year),
  ineffDecrease = TRUE, timeEffect = TRUE, data = df_cap_panel_saboo
)
df_cap_panel_saboo$mar_eff <- efficiencies(cap_mar, asInData = TRUE)

cap_rd <- sfa(
  log(pat_count) ~ log(rdstock) + log(pat_stock) | sic1 + factor(year),
  timeEffect = TRUE, ineffDecrease = TRUE, data = df_cap_panel_saboo
)
df_cap_panel_saboo$rd_eff <- efficiencies(cap_rd, asInData = TRUE)

cap_op <- sfa(
  log(xopr) ~ log(act) + log(ppegt) + log(emp) | sic1 + factor(year),
  timeEffect = TRUE, ineffDecrease = FALSE, data = df_cap_panel_saboo
)
df_cap_panel_saboo$op_eff <- efficiencies(cap_op, asInData = TRUE)
```

#### @elhelaly2023collaborating {#sec-elhelaly}

This variant (with Koyck weight 0.5) enriches each frontier with three-year
accumulated stocks. The marketing frontier adds both advertising and its stock,
both marketing expense and its stock, and both current and three-year receivables;
the R&D frontier uses three-year patent and R&D accumulations; and the operations
frontier is a cost frontier:

$$
\begin{aligned}
\log(\text{sales}) &= \log(\text{xad}) + \log(\text{adstock}) + \log(\text{xsga}) + \log(\text{marstock}) \\
&\quad + \log(\text{rect}) + \log(\text{rec}) + \log(\text{installedbase}),
\end{aligned}
$$ {#eq-elhelaly-mar}

where $\text{rec}$ is accounts receivable summed over the three years prior;

$$
\log(\text{patent}) = \log(\text{patstock}) + \log(\text{xrd}) + \log(\text{accumrd}),
$$ {#eq-elhelaly-rd}

where $\text{patstock}$ is the three-year patent total, $\text{xrd}$ is total R&D
expense, and $\text{accumrd}$ is the three-year R&D total; and

$$
\log(\text{cogs}) = \log(\text{output}) + \log(\text{laborcost}) + \log(\text{costofcapital}),
$$ {#eq-elhelaly-op}

where $\text{output}$ is the dollar value of output, $\text{laborcost}$ is
per-employee wages and benefits, and $\text{costofcapital}$ is the average
long-term interest rate.

```{r}
#| eval: false
df_cap_panel_elhelaly <- df_cap |>
  select(gvkey, year, contains("sic"), sale, xad, adstock, invfg, xsga,
         marstock, rect, installedbase, pat_count, pat_stock, xrd, cogs,
         laborcost, costofcapital) |>
  group_by(gvkey) |> arrange(gvkey, year) |>
  mutate(
    rec      = dplyr::lag(rect, 1) + dplyr::lag(rect, 2) + dplyr::lag(rect, 3),
    patstock = dplyr::lag(pat_count, 1) + dplyr::lag(pat_count, 2) + dplyr::lag(pat_count, 3),
    accumrd  = dplyr::lag(xrd, 1) + dplyr::lag(xrd, 2) + dplyr::lag(xrd, 3)
  ) |>
  pdata.frame(c("gvkey", "year"))
```

```{r}
#| warning: false
#| message: false
#| eval: false
library(frontier)

cap_mar <- frontier::sfa(
  log(sale) ~ log(xad) + log(adstock) + log(xsga) + log(marstock) +
    log(rect) + log(rec) + log(installedbase) | sic1 + factor(year),
  ineffDecrease = TRUE, timeEffect = TRUE, data = df_cap_panel_elhelaly
)
df_cap_panel_elhelaly$mar_eff <-
  frontier::efficiencies(cap_mar, asInData = TRUE)

cap_rd <- sfa(
  log(pat_count) ~ log(patstock) + log(xrd) + log(accumrd) | sic1 + factor(year),
  timeEffect = TRUE, ineffDecrease = TRUE, data = df_cap_panel_elhelaly
)
df_cap_panel_elhelaly$rd_eff <-
  frontier::efficiencies(cap_rd, asInData = TRUE)

cap_op <- sfa(
  log(cogs) ~ log(invfg) + log(laborcost) + log(costofcapital) | sic1 + factor(year),
  timeEffect = TRUE, ineffDecrease = FALSE, data = df_cap_panel_elhelaly
)
df_cap_panel_elhelaly$op_eff <-
  frontier::efficiencies(cap_op, asInData = TRUE)

df_cap_panel_elhelaly |>
  write_rds(file.path("data", "capability", "df_cap_panel_elhelaly.rds"))
```

```{r}
df_cap_panel_elhelaly <-
  read_rds(file.path("data", "capability", "df_cap_panel_elhelaly.rds"))
df_cap_panel_elhelaly |> select(contains("eff")) |> na.omit() |> cor()
```

#### @cao2023marketing {#sec-cao}

This specification uses a valuation outcome—Total Q in place of Tobin's Q—for the
marketing frontier and a lagged-R&D, lagged-patent-stock structure for the R&D
frontier:

$$
\log(\text{Total } Q_t) = \log(\text{xsga}_t) + \log(\text{xsga}_{t-1}) + \log(\text{xad}) + \log(\text{pat}),
$$ {#eq-cao-mar}

$$
\log(\text{pat}) = \log(\text{xrd}_t) + \log(\text{xrd}_{t-1}) + \log(\text{patstock}_{t-1}),
$$ {#eq-cao-rd}

where $\text{pat}$ is the number of patents.

```{r}
#| eval: false
df_cap_panel_cao <- df_cap |>
  left_join(totalq, by = join_by(gvkey, year)) |>
  select(gvkey, year, contains("sic"), q_tot, xsga, xad,
         pat_count, xrd, pat_stock) |>
  group_by(gvkey) |> arrange(gvkey, year) |>
  mutate(
    xsga_t_1      = dplyr::lag(xsga, n = 1),
    xrd_t_1       = dplyr::lag(xrd, n = 1),
    pat_stock_t_1 = dplyr::lag(pat_stock, n = 1)
  ) |>
  pdata.frame(c("gvkey", "year"))
```

```{r}
#| eval: false
library(frontier)

cap_mar <- frontier::sfa(
  log(q_tot) ~ log(xsga) + log(xsga_t_1) + log(xad) + log(pat_count) |
    sic1 + factor(year),
  ineffDecrease = TRUE, timeEffect = TRUE, data = df_cap_panel_cao
)
df_cap_panel_cao$mar_eff <-
  frontier::efficiencies(cap_mar, asInData = TRUE)

cap_rd <- sfa(
  log(pat_count) ~ log(xrd) + log(xrd_t_1) + log(pat_stock_t_1) |
    sic1 + factor(year),
  timeEffect = TRUE, ineffDecrease = TRUE, data = df_cap_panel_cao
)
df_cap_panel_cao$rd_eff <-
  frontier::efficiencies(cap_rd, asInData = TRUE)

df_cap_panel_cao |>
  write_rds(file.path("data", "capability", "df_cap_panel_cao.rds"))
```

```{r}
df_cap_panel_cao <-
  read_rds(file.path("data", "capability", "df_cap_panel_cao.rds"))
df_cap_panel_cao |> select(contains("eff")) |> na.omit() |> cor()
```

### The Broader Marketing-Capability Literature {#sec-marketing-cap-lit}

The frontier-estimation tradition above is one strand of a much larger literature
that decomposes marketing capability into functional sub-capabilities and links each
to performance. The foundational input weights used in the stocks above—0.5 for
marketing expenditure and 0.4 for R&D—come from @dutta2005, and the
brand/advertising-stock weighting from @peles1971 and @wang2017. Building on these
primitives, scholars have isolated specialized capabilities: marketing capability
broadly [@bahadir2008; @xiong2013; @wiles2012; @mishra2016; @dinner2018], marketing
*alliance* capability [@swaminathan2009a], digitized selling capability
[@johnson2005], big-data capability [@johnson2017], social-CRM capability
[@trainor2014], and customer-relationship capability [@wang2017].

The single most influential disaggregation is @morgan2009linking, who link three
core marketing capabilities to *profit growth* and uncover a tension invisible to
coarser analyses. Although profit growth is a primary driver of stock price, the
mechanism by which marketing capabilities feed it was poorly understood. Using a
cross-industry sample of 114 firms, @morgan2009linking decompose marketing
capability into **market sensing**, **brand management**, and **customer
relationship management** (CRM), and decompose profit growth into **revenue growth**
and **margin growth**. The capabilities exert both direct and synergistic effects on
the two growth components—but, critically, brand-management and CRM capabilities can
*counteract* each other across the components: a capability that lifts revenue growth
may simultaneously depress margin growth, and vice versa. A surface-level analysis
that examines only aggregate profit growth would miss these offsetting effects and
mis-state the true relationship between capabilities and performance. The
methodological moral reinforces the theme of this chapter: the level of aggregation
at which a metric is defined determines what relationships it can reveal.

A generic frontier implementation, parameterized by a `sub_market` grouping, ties
the strands together; the `sfaR` package offers a cross-sectional alternative when
the panel structure is unavailable.

```{r}
#| eval: false
library(frontier)
mar_cap <- sfa(
  log(sales) ~ sub_market +
    log(adstock) + log(marketingstock) + log(techbase) +
    log(receivable) + log(installedbase),
  data = capability,
  ineffDecrease = TRUE,  # production frontier
  timeEffect = TRUE      # time-varying error-components frontier
)
efficiencies(mar_cap)
```

```{r}
#| eval: false
library(sfaR)
mar_cap <- sfacross(
  log(sales) ~ sub_market +
    log(adstock) + log(marketingstock) + log(techbase) +
    log(receivable) + log(installedbase),
  data = capability
)
efficiencies(mar_cap)
```

#### Digital, Social-Media, and Organizational Capabilities {#sec-digital-cap}

The capability lens has been extended to the digital domain. Survey-based studies of
**digital marketing capability** [@wang2020; @homburg2022] and a synthesizing review
[@herhausen2020] map the organizational routines through which firms convert digital
assets into performance, while @nguyen2015a treats **social-media strategic
capability** as a distinct competence.

A complementary line generalizes from functional capabilities to the *embeddedness*
of capabilities in the organization. @grewal2007embeddedness argue that managers must
deploy scarce resources to build durable capabilities, and that neglecting the
underlying processes obscures how capabilities translate into competitive advantage.
Their central construct is **capability embeddedness**—the depth at which a
capability is ingrained in the organization, itself a consequence of managerial
resource-allocation decisions. Methodologically, they introduce a hierarchical
composed-error framework (a multilevel generalization of the stochastic frontier)
that applies to cross-sectional and panel data alike, and they show in a retailing
application that capability embeddedness directly improves performance even after
controlling for tangible and intangible resources. The framework's payoff is
diagnostic: recognizing whether the objectives of different capabilities are
*convergent* or *divergent* tells a manager whether deepening embeddedness will
amplify or undercut firm performance—an organizational echo of the revenue/margin
tension in @morgan2009linking.

## Management Metrics {#sec-management}

A final, smaller family of metrics characterizes the firm's strategy and leadership,
which moderate how marketing investments convert into performance. Two constructs
have proven especially measurable from archival data. **Founding strategy**, and
specifically the degree of *market differentiation* a firm adopts at founding, can be
recovered from the language of its early communications and product positioning;
@guzman2023measuring develop a scalable measure of this founding differentiation.
**CEO overconfidence**—a behavioral trait with consequences for innovation and
investment—is measured from executives' revealed reluctance to exercise in-the-money
stock options, following the option-based approach @galasso2011ceo apply to study
innovation. Both illustrate the chapter's recurring method: a psychological or
strategic construct, once thought to require surveys or inside access, recovered at
scale from the residue firms leave in archival and market data.

## Key Takeaways {#sec-metrics-takeaways}

A metric is a definition plus an identification argument. The financial measures of
@sec-metrics-finance are easy to compute and hard to interpret causally: ROI and
ROMI summarize an effect whose incremental numerator is itself the object of
inference, and EVA and MVA price value creation only as well as invested capital and
market expectations are measured. The accounting covariates are largely settled in
their construction—@tbl-compustat-core and @tbl-kahle catalogue the conventions—but
their role is to *neutralize* confounding, and they earn that role only when chosen
to match the marketing decision under study. The marketing constructs of
@sec-metrics-marketing exemplify the field's methodological frontier: sentiment,
purchase intention, trust, and reputation, once the province of surveys, are now
recovered from text and images by pre-trained models whose accuracy and biases must
themselves be validated. Capability, the subject of @sec-capability, is the hardest
case and the template for the rest: a latent efficiency, identified only under
explicit distributional and exogeneity assumptions, whose credibility rests entirely
on the care taken in building the input stocks and specifying the frontier. Across
all three families, the lesson is the same—report the metric, but report the
assumptions that make it mean what you claim.
