# Data {#sec-data}

Empirical marketing research is only as credible as the data behind it, and the
data a researcher can obtain increasingly determine which questions are even
askable. This chapter is a working guide to *acquiring* the raw material of
marketing science—firm financials, advertising and engagement metrics, household
expenditure, and the digital traces of consumer attention—and, just as
importantly, to understanding what each source actually measures. Acquisition is
never neutral: every dataset embeds a sampling frame, a measurement model, and a
set of selection mechanisms that propagate into every downstream estimate. A
panel that surveys 6,000 households quarterly, an index that rescales search
volume to its own maximum, and an API that silently throttles after 1,400 calls
each impose structure on what the analyst can learn.

We organize the material by source, but the through-line is measurement. For each
source we state, as precisely as the source allows, the *estimand*—the population
quantity of interest—and the *observable* the source actually delivers, because
the gap between the two is where most empirical mistakes live. Search interest is
not search volume; a relative index is not an absolute count; a name-based
demographic prediction is a posterior probability, not a fact. Treating these
distinctions casually is how plausible pipelines produce biased coefficients.

The chapter is also deliberately reproducible. Connection recipes, query
patterns, and rescaling algorithms appear as runnable code so that a reader can
move from "this dataset exists" to "I have it in a tidy frame" without reverse
engineering undocumented APIs. Where a third-party platform's own documentation
is the authoritative reference—rate limits, OAuth scopes, schema—we point to the
mechanism rather than reproduce ephemeral details. Throughout, the goal is the
one stated in the style guide: intuition first, then the formal object, then code
that runs.

@fig-data-landscape situates the sources covered here along the two dimensions
that matter most for research design: the unit of observation (firm, household,
or individual consumer) and whether the data are *behavioral* (revealed in
actions) or *survey-elicited* (reported by respondents).

```{mermaid}
%%| label: fig-data-landscape
%%| fig-cap: "The data sources in this chapter, organized by unit of observation and by whether they capture revealed behavior or elicited reports. Behavioral traces are abundant and timely but noisy proxies for latent constructs; survey instruments measure constructs directly but are costly, lagged, and subject to reporting bias."
flowchart TB
  subgraph FIRM["Firm-level"]
    WRDS["WRDS / Compustat<br/>financials, R&D, advertising"]
  end
  subgraph HH["Household-level"]
    CE["Consumer Expenditure Survey<br/>elicited expenditure (panel + diary)"]
  end
  subgraph IND["Individual / keyword-level"]
    GT["Google Trends<br/>relative search interest"]
    BD["Baidu Index<br/>search volume"]
    YT["YouTube<br/>views, comments, engagement"]
    NAME["Name-based inference<br/>gender / age / nationality"]
  end
  WRDS -->|"revealed (accounting)"| OUT["Research design:<br/>estimand vs. observable"]
  CE -->|"elicited (survey)"| OUT
  GT -->|"revealed (behavior)"| OUT
  BD -->|"revealed (behavior)"| OUT
  YT -->|"revealed (behavior)"| OUT
  NAME -->|"inferred (model)"| OUT
```

## Firm Financial Data: WRDS {#sec-data-wrds}

Most marketing-finance research (@sec-marketing-finance) and a large share of
strategy work rest on standardized firm financials. Wharton Research Data Services
(WRDS) is the dominant academic gateway to these: it federates Compustat
(accounting fundamentals), CRSP (security prices and returns), I/B/E/S (analyst
forecasts), and dozens of other libraries behind a single SQL interface. For the
marketing researcher the attraction is concrete. Advertising expense (`xad`) and
research-and-development expense (`xrd`) in Compustat's fundamentals-annual file
(`funda`) are the canonical proxies for a firm's marketing and innovation
intensity, and they join cleanly to market-based outcomes such as Tobin's $q$ or
abnormal returns used throughout @sec-marketing-finance.

The connection is a standard PostgreSQL session; WRDS exposes its warehouse over
the wire, so the same `DBI` idioms used for any relational database apply.
Credentials should never be hard-coded—read them from the environment so that the
source file is safe to share.

```{r wrds-connect, message=FALSE, eval=FALSE}
library(RPostgres)
library(tidyverse)

wrds <- dbConnect(
  Postgres(),
  host    = "wrds-pgdata.wharton.upenn.edu",
  port    = 9737,
  dbname  = "wrds",
  sslmode = "require",
  user    = Sys.getenv("wrds_user"),
  pass    = Sys.getenv("wrds_pass")
)
```

The warehouse is self-describing through the `information_schema`, which is the
disciplined way to discover what is available rather than guessing table names.
Three nested queries answer the three questions a new user has: which *libraries*
(schemas) exist, which *tables* live in a chosen library, and which *columns* a
chosen table holds.

```{r wrds-explore, eval=FALSE}
# Libraries (schemas) available to your subscription
dbGetQuery(wrds, "
  select distinct table_schema
  from information_schema.tables
  order by table_schema")

# Tables inside one library
dbGetQuery(wrds, "
  select distinct table_name
  from information_schema.columns
  where table_schema = 'comp_na_daily_all'
  order by table_name")

# Columns inside one table
dbGetQuery(wrds, "
  select column_name
  from information_schema.columns
  where table_schema = 'comp_na_daily_all'
    and table_name   = 'funda'
  order by column_name")
```

A representative extraction pulls advertising and R&D for a decade, then enriches
each firm-year with its industry classification by joining on the firm identifier
`gvkey`. Pushing the filters (`fyear`, non-null `xad`/`xrd`) into the `WHERE`
clause rather than filtering in R keeps the transfer small; this matters when
`funda` runs to millions of rows.

```{r wrds-extract, eval=FALSE}
fin <- dbGetQuery(wrds, "
  select gvkey, fyear, xad, xrd
  from comp_na_daily_all.funda
  where fyear between 2000 and 2010
    and xad is not null
    and xrd is not null") |>
  distinct()

industry <- dbGetQuery(wrds, "
  select gvkey, gind, gsubind, naics, sic
  from comp_na_daily_all.names")

panel <- fin |>
  left_join(industry, by = "gvkey")

head(panel)
```

Two cautions recur in practice and shape identification downstream. First,
`xad` is *reported* advertising; under accounting rules many firms expense
marketing without breaking it out, so missingness is non-random—it correlates
with firm size, industry, and disclosure regime. Treating `is null` as "no
advertising" rather than "not separately disclosed" silently selects the sample
on a variable correlated with the outcome, the textbook recipe for selection
bias. Second, `funda` ships in multiple *formats* (consolidated vs. parent-only,
standardized vs. restated) keyed by the `indfmt`, `datafmt`, `popsrc`, and
`consol` flags; the conventional screen for North American industrial firms is
`indfmt='INDL' and datafmt='STD' and popsrc='D' and consol='C'`, and omitting it
double-counts firm-years. Neither caution is exotic, but both are invisible until
an estimate looks wrong.

::: {.callout-note}
Beyond WRDS, the **Consumer Expenditure Survey** (below) and commercial scanner
panels (e.g., IRI, Nielsen) cover demand-side behavior that firm financials
cannot. The general principle—match the unit of observation to the estimand—is
developed in @sec-measurement-scales.
:::

## Local Document Stores: MongoDB {#sec-data-mongo}

Not all marketing data are relational. Social-media posts, scraped reviews, and
API responses arrive as nested, schema-flexible JSON, and a document database
such as MongoDB stores them without forcing a rectangular schema up front. When
ingesting at scale the first operational question is simply *where the data physically
live*, which the running server reports through an administrative command issued
in the `mongo` shell:

```
db.adminCommand("getCmdLineOpts")
```

The returned document includes the `storage.dbPath`, the on-disk location of the
data files—useful for capacity planning, backups, and confirming that a large
ingest is writing where you expect. The broader point is architectural: document
stores trade the relational guarantees of WRDS-style sources (a fixed schema,
referential integrity, declarative joins) for flexibility on write. That trade is
right for heterogeneous, evolving web data and wrong for the tidy panels demanded
by most econometric estimators, which is why a common pipeline *lands* raw
captures in MongoDB and *projects* a flattened, analysis-ready frame into R or a
relational store before modeling.

## Search Interest: Google Trends {#sec-data-gtrends}

Aggregate search behavior is among the most useful behavioral signals in
marketing because it is timely, granular, and a leading indicator of attention.
The motivating regularity is that public interest and its institutional
reflections move together but on different clocks: online search volume and news
reporting are strongly correlated, whereas academic publishing lags public
interest because of the delay inherent in peer review and publication
[@nghiem2016]. Read structurally, news media act as a conductor between the
research community and the public, and online search is the most immediate
barometer of the public's side of that exchange.

### What Google Trends Measures

Google Trends does **not** report search volume. It reports a *relative search
interest index*, and the distinction is the single most important thing to
understand about the source. For a keyword $k$, a geography $g$, and a chosen time
window, let $v_{k,t}$ be the underlying volume of queries for $k$ in period $t$
(itself a sampled, privacy-filtered quantity). Trends returns the rescaled series

$$
I_{k,t} \;=\; 100 \times \frac{v_{k,t}}{\displaystyle\max_{s \in W} v_{k,s}},
$$ {#eq-gtrends-index}

where $W$ is the requested window. Three consequences follow directly from
@eq-gtrends-index and trip up the unwary.

First, the index is **window-dependent**. The denominator is the maximum *within
the requested window*, so the same keyword queried over two different windows
generally returns different index values for the overlapping dates—the series is
not anchored to any fixed scale. Second, the index is **co-normalized across the
keywords in a single request**: when several keywords are submitted together they
share one denominator (the global maximum across all of them), which is exactly
what makes cross-keyword comparison valid *within* a request and invalid *across*
separate requests. Third, the data are **sampled**, so repeated identical queries
return slightly different indices; for stable estimates one averages several
draws.

The window also dictates the *granularity* Google will return, which is fixed by
the length of the requested span:

| Data granularity | Window length |
|------------------|------------------------------|
| Hourly           | Last 7 days |
| Daily            | Less than 9 months |
| Weekly           | 9 months to 5 years |
| Monthly          | Longer than 5 years |

: Native temporal granularity returned by Google Trends as a function of the requested window length. Longer windows are silently coarsened, which is why reconstructing a long daily series requires the stitching procedure in @sec-data-gtrends-daily. {#tbl-gtrends-granularity}

A first pass pulls weekly relative interest for two brands over a multi-year
window. The `gtrendsR` package wraps the public endpoint; `interest_over_time`
holds the index series.

```{r gtrends-basic, eval=FALSE}
library(gtrendsR)
library(tidyverse)

trends <- gtrends(
  keyword = c("7eleven", "3m"),
  geo     = "US",
  time    = "2010-01-01 2012-01-30",  # earliest supported start is 2004
  gprop   = "web"                       # web | news | images | froogle | youtube
)

time_trend <- trends$interest_over_time   # weekly index for this window
```

Because @eq-gtrends-index produces a relative, sampled series, plotting it raw
shows both signal and seasonal noise. A smoother separates the trend component
from within-year seasonality without committing to a parametric seasonal model.

```{r gtrends-plot, eval=FALSE}
library(ggplot2)

ggplot(time_trend, aes(x = date, y = hits, group = keyword, colour = keyword)) +
  geom_line(alpha = 0.4) +
  geom_smooth(span = 0.5, se = FALSE) +
  labs(x = "Time", y = "Relative interest (weekly)",
       title = "Google search interest", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

### Reconstructing a Long Daily Series {#sec-data-gtrends-daily}

The granularity rule in @tbl-gtrends-granularity creates a real problem:
researchers often need *daily* resolution over *multi-year* spans, but a
multi-year request returns only weekly (or monthly) data. The fix exploits the
co-normalization property of @eq-gtrends-index. Two reconstruction strategies
exist, and they differ in how they reconcile the high-frequency and
low-frequency series.

The **scaling (overlapping) method** is the more reliable of the two. It
queries short windows at daily resolution, queries the full span at lower
resolution to obtain period weights, and rescales the daily pieces so that each
month's daily indices are multiplied by that month's relative weight. Formally,
let $d_{k,t}$ be the daily index obtained from a short within-window query
covering month $m(t)$, and let $w_{k,m}$ be the month-$m$ weight derived from the
long-window query (the month's aggregate index divided by the maximum monthly
aggregate). The reconstructed daily series is the product

$$
\hat{v}_{k,t} \;=\; d_{k,t} \times w_{k,m(t)},
$$ {#eq-gtrends-scaling}

which restores a common scale across months while preserving within-month daily
shape. The **concatenation (normalization) method** instead stitches one-month
daily queries end to end and normalizes them against the weekly series; empirical
comparisons favor the scaling method, which is why @eq-gtrends-scaling is the
recommended estimator.[^gtrends-attrib] This reconstruction underlies published
applications that require daily search data over long horizons [@risteski2014],
and related approaches to extracting daily volume are documented elsewhere
[@liugoogle2019].

[^gtrends-attrib]: The overlapping-scaling procedure follows the approach
popularized by Alex Dyachenko; the concatenation variant mirrors the
`dailydata` routine in the `pytrends` library. Both predate—and are wrapped
by—the function below.

The function below implements @eq-gtrends-scaling end to end: it pulls a single
aggregated query to derive monthly multipliers `mult`, loops over months to pull
daily data, and joins the two to produce reconstructed daily estimates
`est_hits`. Note the rate-limit budget of roughly **1,400 requests per 4 hours**;
a long reconstruction must respect it or the loop will start returning empty
frames.

```{r gtrends-daily, eval=FALSE}
library(gtrendsR)
library(tidyverse)
library(lubridate)

get_daily_gtrend <- function(keyword = c("7eleven", "3M"),
                             geo  = "US",
                             from = "2013-01-01",
                             to   = "2013-02-15") {
  # Disallow a 'to' date in the current (incomplete) month
  if (ymd(to) >= floor_date(Sys.Date(), "month")) {
    to <- floor_date(ymd(to), "month") - days(1)
    if (to < from) stop("`to` date in the current month is not allowed")
  }

  # Long query -> monthly multipliers w_{k,m}
  aggregated <- gtrends(keyword = keyword, geo = geo, time = paste(from, to))
  if (is.null(aggregated$interest_over_time)) {
    message("No data in Google Trends for this query.")
    return(invisible(NULL))
  }

  mult_m <- aggregated$interest_over_time |>
    mutate(hits = as.integer(ifelse(hits == "<1", "0", hits))) |>
    group_by(month = floor_date(date, "month"), keyword) |>
    summarise(hits = sum(hits), .groups = "drop") |>
    mutate(ym = format(month, "%Y-%m"), mult = hits / max(hits)) |>
    select(month, ym, keyword, mult)

  # Per-month daily queries -> d_{k,t}
  windows <- tibble(
    s = seq(ymd(from), ymd(to), by = "month"),
    e = seq(ymd(from), ymd(to), by = "month") + months(1) - days(1)
  )

  raw_daily <- map2_dfr(windows$s, windows$e, function(s, e) {
    curr <- gtrends(keyword, geo = geo, time = paste(s, e))
    if (is.null(curr$interest_over_time)) return(tibble())
    curr$interest_over_time
  })

  trend_d <- raw_daily |>
    transmute(date, keyword,
              ym   = format(date, "%Y-%m"),
              hits = as.integer(ifelse(hits == "<1", "0", hits)))

  # Reconstruct: est_hits = d_{k,t} * w_{k,m(t)}   (eq-gtrends-scaling)
  trend_d |>
    left_join(mult_m, by = c("ym", "keyword")) |>
    mutate(est_hits = hits * mult, date = as.Date(date)) |>
    select(date, keyword, est_hits)
}

daily_trend <- get_daily_gtrend(
  keyword = c("7eleven", "3M"), geo = "US",
  from = "2013-01-01", to = "2013-02-01")
head(daily_trend)
```

### Absolute Volume

When the application genuinely requires *absolute* counts rather than the relative
index of @eq-gtrends-index—for example, to size a market rather than track its
trajectory—the relative-interest endpoint is the wrong source. Google publishes a
separate top-terms dataset through BigQuery that exposes absolute search figures,
and that warehouse, not the Trends index, is the appropriate input for level
estimates.

## Search Volume in China: Baidu Index {#sec-data-baidu}

Outside the markets Google serves, the analogous behavioral signal comes from the
dominant local search engine. In China that is Baidu, and the **Baidu Index**
plays the role Google Trends plays elsewhere. Because Baidu exposes no comparable
public API for bulk extraction, researchers have built crawlers to recover its
search-volume series; @liu2019 document a Java-based spider for exactly this
purpose. The same measurement caveats apply—an index is a rescaled, sampled
proxy for latent interest, not a count—so the estimand-versus-observable
discipline of @sec-data-gtrends carries over directly.

## Video Engagement: YouTube {#sec-data-youtube}

Video is now a primary advertising and engagement channel, and YouTube exposes
rich behavioral data—views, likes, comments, and watch metrics—at the video,
playlist, and channel level. Two access paths exist, and the choice between them
is a governance decision rather than a convenience one.

The first path is the **public Data API**, which serves *publicly visible*
statistics for any video or channel using a simple API key. The second is the
**Analytics and Reporting APIs**, which serve *owner-private* metrics (impressions,
audience retention, revenue) and therefore require OAuth authorization by the
channel or content owner that owns the requested data. @tbl-youtube-apis
contrasts the two; the operative rule is that anything a viewer cannot see
requires the owner's delegated consent.

| Dimension | Data API (key) | Analytics / Reporting (OAuth) |
|-----------|----------------|-------------------------------|
| Auth | API key | OAuth 2.0, owner-delegated |
| Scope | Public statistics | Owner-private metrics |
| Typical fields | views, likes, comments | impressions, retention, revenue |
| Who can call | anyone with a key | the channel/content owner |
| Use case | competitive/market scans | first-party performance analytics |

: Two paths to YouTube data. The boundary is consent: public statistics need only a key, whereas owner-private analytics require OAuth authorization by the data owner. {#tbl-youtube-apis}

### OAuth Path via `tuber`

The `tuber` package wraps the authenticated endpoints. Authorization is a
one-time OAuth handshake with credentials provisioned in the Google Cloud
console; thereafter the session token authorizes calls for video statistics,
details, captions, search, and comment threads.

```{r youtube-oauth, eval=FALSE}
library(tuber)

# Provision app_id / app_secret in the Google Cloud console, then:
yt_oauth(app_id = "YOUR_APP_ID", app_secret = "YOUR_APP_SECRET")

get_stats(video_id = "N708P-A45D0")          # public statistics
get_video_details(video_id = "N708P-A45D0")  # snippet metadata
get_captions(video_id = "yJXTXN4xrI8")       # caption track
yt_search(term = "test")                      # search the corpus
get_comment_threads(c(video_id = "N708P-A45D0"))  # top-level comments
get_all_comments(video_id = "a-UQz7fqR3w")    # comments + replies
```

A common research task is to assemble a panel of *every* video in a channel.
YouTube models a channel's uploads as a hidden playlist, so the recipe is: read
the channel's `relatedPlaylists$uploads` id, page through that playlist's items
to collect video ids, then map the statistics call over the ids and bind the
results into a frame.

```{r youtube-channel, eval=FALSE}
chan <- list_channel_resources(
  filter = c(channel_id = "UCT5Cx1l4IS3wHkJXNyuj4TA"),
  part   = "contentDetails")

uploads_id <- chan$items[[1]]$contentDetails$relatedPlaylists$uploads
vids       <- get_playlist_items(filter = c(playlist_id = uploads_id))
vid_ids    <- as.vector(vids$contentDetails.videoId)

stats_df <- do.call(rbind, lapply(vid_ids, function(id) data.frame(get_stats(id))))
head(stats_df)
```

### Key Path via the Data API

When only public statistics are needed, the raw REST endpoints are lighter than
an OAuth session. The pattern is a small set of helper functions that build the
request URL, parse the JSON, and return a tidy frame—one helper per resource
type (video, playlist, channel).

```{r youtube-api, eval=FALSE}
library(jsonlite)
library(dplyr)

API_key <- Sys.getenv("YOUTUBE_API_KEY")   # never hard-code keys

video_stats <- function(video_id, key) {
  url <- paste0("https://www.googleapis.com/youtube/v3/videos",
                "?part=snippet,statistics&id=", video_id, "&key=", key)
  res <- fromJSON(url)
  data.frame(
    name  = res$items$snippet$channelTitle,
    res$items$statistics,
    title = res$items$snippet$title,
    date  = res$items$snippet$publishedAt
  )
}

channel_stats <- function(channel_id, key) {
  url <- paste0("https://www.googleapis.com/youtube/v3/channels",
                "?part=snippet,contentDetails,statistics&id=", channel_id,
                "&key=", key)
  res <- fromJSON(url)
  data.frame(
    name = res$items$snippet$title,
    res$items$statistics,
    uploads = res$items$contentDetails$relatedPlaylists$uploads
  )
}
```

These frames feed directly into comparative visualizations—total channel views,
or the time path of comments and likes across competing channels—using the
`ggplot2` idioms already shown for Google Trends. Because the engagement counts
(`viewCount`, `commentCount`) arrive as character strings, they must be coerced
to numeric before plotting.

::: {.callout-warning}
A `requests`-and-`BeautifulSoup` scraper that pages through search-result HTML can
retrieve video ids without an API key, but scraping rendered pages is brittle,
violates many terms of service, and breaks whenever the front-end markup changes.
Prefer the official endpoints; treat HTML scraping as a last resort for
prototyping, never production.
:::

## Household Expenditure: The Consumer Expenditure Survey {#sec-data-ce}

When the research question concerns *what households buy*—budget shares,
category demand, the income elasticity of a product class—the canonical U.S.
source is the Consumer Expenditure (CE) Survey administered by the Bureau of
Labor Statistics. Its sampling design is worth understanding in detail, because
the survey's structure dictates which estimands it can support and at what
frequency.

The CE's unit of observation is the **consumer unit**, defined as a single person
or a group of persons who live together and share responsibility for most major
expenditures.

> A consumer unit is a single person, or a group of persons who live together and
> who share responsibilities for most major expenditures.

Within each unit, the **reference person** is the first individual named when the
respondent answers, "Who is responsible for owning or renting this home?"—a
convention that fixes the household head deterministically and makes
cross-survey linkage well defined.

The survey is not one instrument but two complementary ones, a design choice that
follows directly from a measurement trade-off between recall accuracy and
coverage. Large, infrequent purchases are remembered well but happen rarely;
small, frequent purchases are forgotten quickly but dominate the budget in count.
No single instrument measures both well, so the CE splits them:

- The **Interview Survey** captures big-ticket and recurring expenditures, plus
  global estimates for some categories. It is a rotating-panel survey: each
  consumer unit is interviewed once every three months for four quarters, with
  roughly 6,000 units in the sample each quarter. The panel structure supports
  within-unit change over the year, but the four-quarter cap means it cannot
  follow a household across years.
- The **Diary Survey** captures small-ticket, frequently purchased items that
  recall-based interviewing would miss. It is independent of the interview
  component and collects roughly 14,000 diaries from participating families each
  year.

Published tabulations come in several time formats: **annual** tables run
January–December (available since 1984) and, in a fiscal variant, July–June
(since 2013); **two-year** tables span January of year one through December of
year two (since 1986). Beyond the standard tables, BLS produces **experimental**
research tabulations—covering, for instance, income detail or generational groups
such as millennials and Gen X—available on request. The practical implication for
design is that the rotating four-quarter panel is the right frame for
within-year, within-household questions, while cross-year dynamics require
pooling repeated cross-sections rather than following the same units.

## Inferring Demographics from Names {#sec-data-names}

A recurring need in marketing research is to attach demographic attributes—gender,
age, nationality—to records that carry only a name, as when a researcher has a
roster of customers, reviewers, or social-media authors but no self-reported
demographics. A family of services performs exactly this inference: gender from
genderize.io, age from agify.io, and nationality from nationalize.io. Each
returns, for a queried name, a predicted attribute together with a confidence.

It is essential to read these outputs as what they are: **posterior probabilities
from a name-frequency classifier**, not ground-truth labels. The underlying logic
is a simple Bayesian classifier. Let $n$ be an observed name and $g$ a candidate
attribute value (say, gender). The services effectively report

$$
\Pr(g \mid n) \;=\; \frac{\Pr(n \mid g)\,\Pr(g)}{\displaystyle\sum_{g'} \Pr(n \mid g')\,\Pr(g')},
$$ {#eq-name-bayes}

estimated from large registries of names tagged with the attribute, where
$\Pr(n \mid g)$ is the frequency of name $n$ among individuals with value $g$ and
$\Pr(g)$ the base rate of $g$ in the reference population. Three properties of
@eq-name-bayes govern responsible use. The prediction inherits the **reference
population's composition**: a classifier trained predominantly on one region's
registries mispredicts systematically for names common elsewhere, so the base
rate $\Pr(g)$ is not universal. The reported confidence is the **posterior
itself**, so unisex or rare names yield diffuse, low-confidence posteriors that
should not be hardened into deterministic labels. And because the inferred
attribute is a *generated regressor*, using it as a covariate in a downstream
model imports classification error into that model—the resulting attenuation and
the need to propagate the first-stage uncertainty are treated in
@sec-measurement-scales. Used with those caveats, name-based inference is a cheap
way to recover coarse demographic structure; used naively, it manufactures
measurement error correlated with name origin.

## Pitfalls and Identification {#sec-data-pitfalls}

The sources in this chapter differ in unit, frequency, and provenance, but the
threats to valid inference share a small number of structural causes. @fig-data-pitfalls
maps each common pitfall to the source where it bites hardest and to its remedy.

```{mermaid}
%%| label: fig-data-pitfalls
%%| fig-cap: "Recurring threats to validity across the chapter's data sources and their remedies. Each threat is a gap between the estimand and the observable that the source actually delivers."
flowchart LR
  A["Non-random missingness<br/>(unreported xad in Compustat)"] --> A2["Model disclosure;<br/>do not treat NULL as zero"]
  B["Relative, window-dependent index<br/>(Google Trends, Baidu)"] --> B2["Rescale via overlapping<br/>method (eq-gtrends-scaling)"]
  C["Sampling noise in indices<br/>(Trends draws)"] --> C2["Average repeated queries"]
  D["Generated-regressor error<br/>(name-based demographics)"] --> D2["Propagate first-stage<br/>uncertainty"]
  E["Recall vs. coverage trade-off<br/>(CE survey)"] --> E2["Match instrument to estimand<br/>(interview vs. diary)"]
  F["Rate limits / TOS<br/>(APIs, scrapers)"] --> F2["Budget requests; prefer<br/>official endpoints"]
```

The unifying discipline is the one stated at the outset: name the estimand, name
the observable the source delivers, and treat the difference as a modeling problem
rather than a nuisance. Reported advertising is not advertising; a relative index
is not a volume; a name-based prediction is not a demographic. Each gap is
tractable once it is made explicit, and ruinous when it is not.

## Key Takeaways

- **Match the unit of observation to the estimand.** WRDS delivers firm-level
  financials, the CE Survey household expenditure, and search/engagement APIs
  individual behavioral traces; no source substitutes for another
  (@fig-data-landscape).
- **Reported is not measured.** Compustat's `xad` records *disclosed* advertising;
  missingness is non-random and treating `NULL` as zero selects the sample on a
  variable correlated with outcomes (@sec-data-wrds).
- **A relative index is not a volume.** Google Trends rescales to its window's
  maximum (@eq-gtrends-index); long daily series must be reconstructed by the
  overlapping-scaling method (@eq-gtrends-scaling), and absolute counts require a
  different source entirely.
- **Survey design encodes a recall–coverage trade-off.** The CE splits big-ticket
  recall (interview panel) from frequent-purchase coverage (diary), so the right
  instrument depends on the category (@sec-data-ce).
- **Name-based demographics are posteriors, not labels.** They follow a Bayesian
  name-frequency classifier (@eq-name-bayes) whose error is correlated with name
  origin and must be propagated downstream (@sec-data-names).
- **Respect the platform contract.** Owner-private metrics require OAuth, public
  ones a key (@tbl-youtube-apis); honor rate limits and prefer official endpoints
  over brittle scrapers.
