# Modeling in Marketing {#sec-modeling}

A *model* is a deliberately incomplete description of a marketing system, built to
answer a specific question. Models are not the world; they are arguments about the
world, made precise enough that their assumptions can be stated, their logic
checked, and their predictions confronted with data. Every quantitative chapter in
the methodology part of this book builds a model of some kind, and the chapters
differ less in subject matter than in the *kind* of question each model is licensed
to answer. This chapter is the map. It distinguishes the three modeling traditions
that organize quantitative marketing—**analytical** (game-theoretic), **empirical**
(econometric and structural), and **predictive** (statistical and machine
learning)—and gives the reader a principled way to decide which one a given problem
calls for.

The distinction matters because the three traditions answer different questions and
fail in different ways. An analytical model asks *what should happen* when rational
firms and consumers interact under stated rules; it trades realism for logical
transparency and yields comparative statics—signed predictions about how
equilibrium outcomes move with primitives. An empirical model asks *what is
happening, and why*; it recovers parameters of preferences, costs, or response
functions from data and, when those parameters are invariant to the policies under
study, supports counterfactual prediction. A predictive model asks *what will
happen next*; it learns a flexible mapping from features to outcomes and is judged
almost entirely by out-of-sample accuracy, with little claim on the mechanism that
generated the data. Confusing these aims is the most common and most expensive error
in applied marketing modeling: a model fit for prediction is routinely, and wrongly,
read as if it identified a causal effect, and a stylized analytical result is
routinely, and wrongly, taken as an empirical fact.

This chapter proceeds from the shared anatomy of a model to the three traditions in
turn, formalizing each with its canonical object, its notion of a "solution" or
"estimate," and the assumption whose failure is fatal. It closes with a decision
framework and a worked illustration that runs the *same* marketing question—how a
price change affects demand—through all three lenses, so the reader can see exactly
where they agree, where they diverge, and why. The goal is not to rank the
traditions but to make the reader fluent in choosing among them. Subsequent chapters
develop each in depth: analytical models in @sec-analytical-models and
@sec-industrial-organization, empirical and structural models in
@sec-empirical-models, @sec-structural-models, and @sec-marketing-mix-models, and
predictive methods in @sec-ai-ml.

## What a Model Is

Before partitioning models into traditions, it helps to fix the parts every model
shares. Strip away the subject matter and a quantitative marketing model is a tuple
of four objects: **primitives**, **structure**, **a solution or estimation concept**,
and **an empirical content**.

The *primitives* are the objects taken as given and not themselves explained:
consumer preferences, firm cost functions, the information each agent holds, the
timing of moves, the data-generating distribution. The *structure* is the set of
assumptions that link primitives to observable outcomes—a utility function, a demand
system, a conduct assumption about how firms compete, a likelihood. The *solution or
estimation concept* is the rule that turns structure into an answer: an equilibrium
notion in an analytical model, an estimator in an empirical one, a learning
algorithm and loss function in a predictive one. The *empirical content* is whatever
the model says about data we can observe—signs of comparative statics, point
estimates with standard errors, or a forecast with a quantified error.

> A model is a representation that is simpler than the system it represents, built so
> that reasoning about the model is a reliable guide to reasoning about the system,
> within a stated scope. Its value lies entirely in the gap between the assumptions
> we impose and the conclusions we can then defend.

The decisive property that separates the traditions is what each model holds fixed
when the environment changes. A **policy-invariant** parameter is one that does not
shift when the policy under study shifts—a consumer's taste for a product attribute,
say, which should not change merely because a firm reprices. A **policy-variant**
relationship is one that is itself a function of the policy environment—a reduced-form
regression of sales on price confounds the demand the analyst wants with the supply
behavior that set the price, so its coefficient changes if pricing conduct changes.
This distinction, central to the structural tradition, is the thread we follow
throughout: it explains why two models can fit the same data equally well yet give
opposite answers to a counterfactual question.

::: {.callout-note}
## Notation for this chapter

Following the book's master notation (@sec-data), scalars are italic ($p$, $x$),
vectors bold lowercase ($\mathbf{x}$), matrices bold uppercase ($\mathbf{X}$),
estimators carry hats ($\hat\beta$), and $\mathbb{E}[\cdot]$ denotes expectation.
A *parameter* $\theta$ is a feature of the data-generating process; an *estimate*
$\hat\theta$ is a function of a finite sample; a *prediction* $\hat y$ is a model's
output for a new input.
:::

## A Taxonomy of Marketing Models {#sec-modeling-taxonomy}

The three traditions can be placed on two axes that together explain most of their
differences in practice. The first axis is **the role of optimizing behavior**:
whether the model derives outcomes from agents solving stated optimization problems
(analytical, and the structural branch of empirical work) or treats outcomes as the
realization of a statistical process to be fit (reduced-form empirical and predictive
work). The second axis is **the target of inference**: explanation and counterfactual
prediction on one side, out-of-sample forecasting on the other. @fig-modeling-map
locates the traditions, and the table that follows gives the formal contrast.

```{mermaid}
%%| label: fig-modeling-map
%%| fig-cap: "Three modeling traditions in marketing, placed by the role of optimizing behavior and the primary target of inference. Structural empirical work shares the optimizing foundations of analytical models but, unlike them, takes parameters to data; reduced-form and predictive work relax optimization in exchange for flexibility."
flowchart TB
    Q["A marketing question"]
    Q --> A["Analytical / game-theoretic<br/>(what should happen?)"]
    Q --> E["Empirical / econometric<br/>(what is happening, and why?)"]
    Q --> P["Predictive / machine learning<br/>(what will happen next?)"]

    A --> A1["Primitives: payoffs, info, timing"]
    A --> A2["Solution: equilibrium"]
    A --> A3["Output: comparative statics"]

    E --> E1["Structural: utilities, costs (invariant)"]
    E --> E2["Reduced-form: response functions (variant)"]
    E --> E3["Output: estimates + counterfactuals"]

    P --> P1["Features and a loss function"]
    P --> P2["Solution: regularized fit"]
    P --> P3["Output: forecasts, scored by test error"]
```

| Dimension | Analytical | Empirical (structural) | Empirical (reduced-form) | Predictive (ML) |
|---|---|---|---|---|
| Core question | What *should* happen | Why does it happen | What *is* the effect | What happens next |
| Canonical object | Game $(N, S, u)$ | Likelihood from optimization | Conditional mean $\mathbb{E}[y\mid \mathbf{x}]$ | Mapping $\hat f:\mathbf{x}\mapsto \hat y$ |
| "Solution" concept | Equilibrium | (Quasi-)MLE / GMM | OLS / IV | Empirical-risk minimization |
| Data role | None (or calibration) | Disciplines parameters | Identifies an effect | Trains and tests $\hat f$ |
| Key virtue | Logical transparency | Counterfactual validity | Credible causal estimate | Predictive accuracy |
| Fatal assumption | Wrong game / behavior | Misspecified structure | Confounding / weak instrument | Distribution shift |
| Generalizes via | Theory | Invariant parameters | Design (exogenous variation) | i.i.d. test data |

: Formal contrast of the three modeling traditions. The empirical column is split because structural and reduced-form work, though both "empirical," differ exactly on policy invariance—the structural branch recovers invariant primitives and so supports counterfactuals, while the reduced-form branch estimates a possibly policy-variant response function and so supports causal claims only under the design that produced its variation. {#tbl-modeling-contrast}

The remaining sections take each tradition in turn. The reader should keep
@tbl-modeling-contrast in view: nearly every methodological decision in the rest of
the book is a choice among its columns.

## Analytical Models {#sec-modeling-analytical}

An analytical model derives marketing outcomes as the *equilibrium* of a game played
by rational agents. It contains no data. Its purpose is to establish, with the
certainty of a proof, that under stated assumptions a particular force operates in a
particular direction—that, say, increased competition lowers equilibrium price, or
that informative advertising can support a separating equilibrium in which only
high-quality firms advertise. The payoff is a *comparative static*: the sign, and
sometimes the magnitude, of the derivative of an equilibrium outcome with respect to
a primitive.

Formally, a game is a triple $(N, \{S_i\}_{i \in N}, \{u_i\}_{i \in N})$ where $N$ is
the set of players (firms, consumers), $S_i$ is player $i$'s strategy set (prices,
advertising levels, qualities), and $u_i: \prod_j S_j \to \mathbb{R}$ is $i$'s payoff
as a function of the chosen profile. A pure-strategy **Nash equilibrium**
[@nash1950] is a profile $\mathbf{s}^\* = (s_1^\*, \dots, s_n^\*)$ such that no player
can improve unilaterally,

$$
u_i(s_i^\*, \mathbf{s}_{-i}^\*) \;\ge\; u_i(s_i, \mathbf{s}_{-i}^\*)
\qquad \text{for all } s_i \in S_i,\ \text{all } i \in N,
$$ {#eq-nash}

where $\mathbf{s}_{-i}$ denotes the strategies of $i$'s rivals. The equilibrium is the
solution concept: it plays the role that an estimator plays in empirical work. To
*solve* the model is to characterize $\mathbf{s}^\*$ and then differentiate it with
respect to a primitive to obtain the comparative static of interest.

A concrete instance fixes the idea. Consider two firms choosing prices for
horizontally differentiated products located at the ends of a unit interval, with
consumers uniformly distributed and incurring a quadratic "transport" cost $t$ for
distance from their preferred variety—the Hotelling spatial-competition setup
[@hotelling1929; @dixit1980]. Each firm $i$ has constant marginal cost $c$ and earns
$\pi_i = (p_i - c)\, D_i(p_i, p_j)$, where the demand $D_i$ is the mass of consumers
who prefer firm $i$ at the posted prices. Solving the first-order conditions and
imposing symmetry yields the equilibrium price

$$
p^\* = c + t,
$$ {#eq-hotelling}

a closed-form result whose comparative statics are immediate and exact:
$\partial p^\* / \partial t > 0$ and $\partial p^\* / \partial c = 1$. Greater
differentiation (larger $t$) softens competition and raises equilibrium margins; the
result holds for *every* admissible parameter value, with no data and no estimation.
This is the characteristic strength of the analytical tradition—universally valid,
sign-definite conclusions—and its characteristic limitation: the conclusion is true
*of the model*, and whether the model resembles any real market is a separate
question the framework cannot answer. Marketing's analytical literature has used this
machinery to study product line design, channel structure, advertising, and pricing
[@moorthy1985; @moorthy1988; @moorthy1993; @narasimhan1988b; @villas-boas1999;
@iyer2005], and signaling games to rationalize how brands credibly convey unobservable
quality [@spence1973; @akerlof1970; @nelson1974].

The assumption whose failure is fatal to an analytical model is the specification of
the *game itself*: the players' rationality, the information they hold, the order of
moves, and the equilibrium concept invoked. If real firms do not best-respond, or
hold different beliefs, or the relevant equilibrium is not the one selected, the
comparative statics need not describe the world even approximately. Analytical models
therefore earn their keep as *theories*—sources of sharp, falsifiable hypotheses and
of the behavioral content that disciplines structural estimation—rather than as
descriptions to be read off as fact. @sec-analytical-models develops the solution
techniques and @sec-industrial-organization the equilibrium models of competition
that structural work later takes to data.

## Empirical Models {#sec-modeling-empirical}

An empirical model confronts structure with data to recover parameters. The
tradition splits along the policy-invariance line introduced above, and the split is
the single most important methodological distinction in the chapter.

### The Reduced-Form Branch

A **reduced-form** model specifies a response function relating an outcome to
marketing and control variables, and estimates it without deriving that function from
an explicit optimization problem. The workhorse is the regression of a sales or
choice outcome $y$ on marketing inputs $\mathbf{x}$ and controls $\mathbf{w}$,

$$
y_i = \mathbf{x}_i^\top \boldsymbol{\beta} + \mathbf{w}_i^\top \boldsymbol{\gamma}
       + \varepsilon_i,
$$ {#eq-25-reduced-form}

estimated by ordinary least squares (OLS) when the regressors are exogenous, or by
instrumental variables (IV) / two-stage least squares when they are not
[@greene2003econometric; @cameron2005microeconometrics]. The marketing-mix and
response-modeling literatures are largely reduced-form in this sense
[@Little_1970; @Little_1976; @lambin1970a; @clarke1976; @lilien1992marketing], and
@sec-marketing-mix-models treats them at length.

The defining hazard of @eq-25-reduced-form is **endogeneity**: a regressor correlated
with the error makes the OLS estimator inconsistent, $\mathbb{E}[\hat{\boldsymbol\beta}]
\neq \boldsymbol\beta$. Price is the canonical offender—firms set higher prices where
they observe (but the analyst does not) higher demand, so $\mathrm{Cov}(p_i,
\varepsilon_i) > 0$ and the naive price coefficient is biased toward zero, sometimes
to the point of the wrong sign. The remedies are an instrument $\mathbf{z}$ that
shifts $\mathbf{x}$ without entering the outcome directly, or a **control function**
that conditions on the endogenous portion of the regressor [@Petrin_2010], or a
research design that isolates exogenous variation. The marketing literature has
matured on exactly this point: modal recent work addresses both sample selection and
endogeneity rather than one in isolation [@frennea2018], and the rise of
quasi-experimental designs—exploiting borders, weather, regulatory changes, and
abrupt firm-policy shifts—gives reduced-form estimates a credible causal warrant
[@goldfarb2022]. A reduced-form estimate, properly identified, answers "what is the
effect of this intervention, here." It does *not*, in general, answer "what would
happen under a policy we have never observed," because the estimated $\boldsymbol\beta$
may itself be a function of the prevailing policy regime—it is policy-variant.

::: {.callout-warning}
## Reduced-form does not mean atheoretic

A reduced-form model can be every bit as causally credible as a structural one when
its identifying variation is genuinely exogenous; "reduced-form" is a statement about
where the estimating equation comes from, not about rigor. What reduced-form
estimates lack is *portability*: their coefficients are valid for counterfactuals
spanned by the variation that identified them, and silent about the rest.
:::

### The Structural Branch

A **structural** model imposes the optimization problem itself and estimates the
parameters of its primitives—utilities and costs—rather than a response function.
Following @erdem1996a, structural estimation "assumes and imposes a structure on the
consumer's maximization problem," and the recovered parameters are those of the
consumers' utility functions, which are taken to be **policy-invariant**. Because the
primitives do not move when the policy moves, the estimated model can be simulated
under counterfactual policies the data never contained—the property that justifies
the additional assumptions structural work requires [@reiss2011;
@chintagunta2006structural; @dube2002structural].

The canonical structural object in marketing is the **random-utility discrete-choice
model** [@mcfadden1986choice; @mcfadden2001economic]. Consumer $i$ obtains utility

$$
u_{ij} = \mathbf{x}_j^\top \boldsymbol{\beta}_i - \alpha_i p_j + \xi_j + \varepsilon_{ij}
$$ {#eq-25-rum}

from alternative $j$, where $\mathbf{x}_j$ are observed attributes, $p_j$ is price,
$\xi_j$ is a demand shock observed by firms and consumers but not the analyst, and
$\varepsilon_{ij}$ is idiosyncratic taste. Under a Type-I extreme-value assumption on
$\varepsilon_{ij}$ the choice probabilities take the multinomial-logit form, and with
random coefficients $(\boldsymbol\beta_i, \alpha_i)$ integrated over a population
distribution one obtains the mixed-logit / BLP demand system estimated by maximum
likelihood or GMM [@guadagni1983; @berry1994estimating; @berry1995; @rossi2014;
@dube2021random]. The structural-marketing program has extended @eq-25-rum to consumers
who *learn* about product quality through experience and forward-looking dynamics
[@erdem1996a; @erdem1998; @erdem2008impact; @chintagunta1998; @seetharaman1999;
@allenby1999dynamic; @gilbride2004; @bradlow2004learning; @sun2011learning], to firm
conduct and equilibrium pricing [@kadiyali2001structural; @misra2011structural], and
to dynamic competition [@gowrisankaran2012dynamics; @Borkovsky_2017]. The estimation
machinery—mixed logit, hierarchical Bayes, GMM—is the subject of
@sec-structural-models and @sec-choice-bayesian.

The crucial contrast between the two structural choice models in marketing is whether
the consumer is **myopic** or **forward-looking**. A myopic model has the consumer
choose to maximize *current-period* utility given current beliefs; a forward-looking
model has the consumer choose to maximize the expected present value of utility,
internalizing how today's choice affects tomorrow's information and options—relevant
precisely when learning about uncertain product characteristics drives the choice
process [@erdem1996a]. The two can rationalize the same observed choices yet imply
sharply different responses to, for example, a temporary promotion, which a
forward-looking consumer partly anticipates.

The assumption whose failure is fatal to a structural model is **specification**: if
the imposed utility, the error distribution, or the conduct assumption is wrong, the
estimated "primitives" absorb the misspecification and the counterfactuals built on
them are unreliable. Structural work thus buys counterfactual portability at the
price of strong, and contestable, functional-form and behavioral assumptions—the
exact trade reduced-form work declines to make. Neither dominates; the right choice
depends on the question, as @sec-modeling-decision argues.

## Predictive Models {#sec-modeling-predictive}

A predictive model learns a flexible mapping from features to an outcome and is
judged by how well it forecasts data it has not seen. It makes minimal commitments
about the mechanism that generated the data; in exchange it can fit relationships far
richer than the linear-in-parameters forms of @eq-25-reduced-form, and it scales to the
high-dimensional, unstructured data—text, images, clickstreams—that now pervade
marketing [@wedel2016; @athey2017digital]. The aim is not to estimate $\boldsymbol\beta$
but to construct an $\hat f$ that minimizes expected loss on future inputs.

Formally, given a loss $L(y, \hat y)$ and a joint distribution $\mathcal{P}$ over
$(\mathbf{x}, y)$, the target is the function minimizing **risk**, the expected loss,

$$
f^\* = \arg\min_{f \in \mathcal{F}} \; \mathbb{E}_{(\mathbf{x}, y)\sim\mathcal{P}}
        \big[\, L(y, f(\mathbf{x})) \,\big].
$$ {#eq-risk}

Because $\mathcal{P}$ is unknown, the model minimizes **empirical risk**—average loss
on a training sample—typically with a penalty that controls complexity,

$$
\hat f = \arg\min_{f \in \mathcal{F}} \;
        \frac{1}{n}\sum_{i=1}^{n} L\big(y_i, f(\mathbf{x}_i)\big)
        \;+\; \lambda\, \Omega(f),
$$ {#eq-erm}

where $\Omega(f)$ penalizes complexity (e.g., the $\ell_1$ norm of coefficients in
the lasso, tree depth in a gradient-boosted ensemble) and $\lambda \ge 0$ is a
regularization strength chosen by cross-validation. The whole apparatus exists to
manage the **bias–variance trade-off**: a model too simple to capture the signal is
biased; one flexible enough to memorize the training noise has high variance and
generalizes poorly. The discipline of the tradition is that performance is reported on
a held-out test set, never on the data used to fit—so the relevant guarantee is i.i.d.
generalization, not parameter recovery. Marketing applications span churn and
response scoring, recommendation, and the extraction of structured signal from text
and images for downstream marketing use [@netzer2008; @toubia2013; @ansari2018;
@tirunillai2014a; @buschken2016sentence; @hauser2006]; @sec-ai-ml develops the methods
and @sec-text-as-data their application to language.

The assumption whose failure is fatal to a predictive model is **distributional
stability**: @eq-risk presumes that future $(\mathbf{x}, y)$ are drawn from the same
$\mathcal{P}$ as the training data. When the deployment environment shifts—a new
competitor enters, a price moves outside its historical range, a recommendation
changes the very behavior it predicts—the i.i.d. assumption breaks and accuracy can
collapse without warning. This failure mode is the deep reason a predictive model,
however accurate, does not license a causal or counterfactual claim. A model that
predicts churn superbly from a feature that *includes* a retention discount cannot be
used to evaluate *changing* that discount: intervening on the feature changes the
distribution, and @eq-risk no longer applies. The point generalizes—predictive
accuracy and causal validity are distinct objectives, and optimizing one does not
deliver the other [@athey2017digital].

::: {.callout-note}
## Prediction policy versus parameter

A useful test of which tradition a problem belongs to: ask whether the decision
requires knowing the *effect of an action you will take*, or only an *accurate
forecast given actions already in train*. Choosing whom to target with a retention
offer, conditional on the offer policy, is a prediction problem. Choosing *whether to
change* the offer policy is a causal problem and needs an empirical or analytical
model. Many marketing tasks are prediction-policy problems—who, not whether—and for
those a predictive model is exactly right.
:::

## Choosing a Modeling Approach {#sec-modeling-decision}

The traditions are complements, not rivals, and mature marketing science routinely
chains them: theory generates a hypothesis (analytical), an estimate tests it and
quantifies a counterfactual (empirical), and a predictive model operationalizes the
resulting policy at scale. The choice for a given study turns on three questions,
asked in order.

First, **what must the answer support**? If the deliverable is a counterfactual—what
would profits be under a price the firm has never charged, a product it has never
launched, a market structure that does not yet exist—the model must recover
policy-invariant primitives, which points to structural estimation
(@sec-modeling-empirical) disciplined by analytical theory (@sec-modeling-analytical).
If the deliverable is the causal effect of an intervention *within* the observed
regime, a credibly identified reduced-form design suffices and is usually preferable
for its weaker assumptions [@goldfarb2022]. If the deliverable is an accurate forecast
or a ranking conditional on a fixed policy, a predictive model is the right tool
(@sec-modeling-predictive).

Second, **where does identifying variation come from**? A causal or structural answer
requires variation in the cause that is independent of the unobserved drivers of the
outcome—an instrument, a natural experiment, or a randomized trial. Absent such
variation, no amount of modeling sophistication recovers an effect; the honest move is
to retreat to prediction and say so. Predictive models, by contrast, need only that
the future resemble the past, a much weaker and more often satisfied condition.

Third, **what is the cost of a wrong assumption**? Structural models deliver the most
but assume the most, and a misspecified structure quietly corrupts every
counterfactual built on it. Reduced-form designs assume less and fail more visibly.
Predictive models assume least about mechanism but stake everything on distributional
stability. The right model is the one whose fatal assumption (the last column of
@tbl-modeling-contrast) is the most defensible in the application at hand—not the most
sophisticated available.

## Illustration: One Question, Three Lenses {#sec-modeling-illustration}

To make the contrast concrete, consider a single question—*how does a price change
affect demand?*—and run it through all three traditions on a common simulated market.
The exercise is deliberately small and fully reproducible; its point is conceptual,
not computational. We simulate a market in which firms set prices partly in response
to an unobserved demand shock, creating textbook price endogeneity, and then show how
each tradition treats it.

```{r modeling-dgp, message=FALSE, warning=FALSE}
set.seed(2025)
n <- 4000

# Unobserved (to the analyst) demand shifter xi; firms see it and price on it.
xi    <- rnorm(n, 0, 1)
cost  <- rnorm(n, 2, 0.4)              # exogenous cost shifter (a valid instrument)
# Pricing rule: price rises with both cost and the latent demand shock -> endogeneity
price <- 1.5 + 0.8 * cost + 0.6 * xi + rnorm(n, 0, 0.2)

# True structural demand: utility decreasing in price, increasing in xi.
alpha_true <- -1.2                     # the policy-invariant price coefficient
util  <- 3 + alpha_true * price + xi
prob  <- plogis(util)                  # logit choice probability
buy   <- rbinom(n, 1, prob)

dat <- data.frame(buy, price, cost, xi)
round(cor(dat[, c("price", "cost", "xi")]), 2)
```

The correlation matrix confirms the design: `price` is correlated with the
unobserved `xi`, so any model that regresses purchase on price *without* addressing
this correlation will recover a biased price effect. Now the three lenses.

**Predictive lens.** Fit a flexible classifier to predict `buy` from `price` and
`cost`, and evaluate it out of sample. The model is excellent at forecasting and says
nothing trustworthy about the *effect* of moving price.

```{r modeling-predictive, message=FALSE, warning=FALSE}
train <- 1:3000; test <- 3001:n
pred_fit <- glm(buy ~ poly(price, 2) + cost, data = dat[train, ],
                family = binomial)
phat <- predict(pred_fit, newdata = dat[test, ], type = "response")
# Out-of-sample area under the ROC curve (rank accuracy), computed from scratch.
pos  <- phat[dat$buy[test] == 1]; neg <- phat[dat$buy[test] == 0]
auc  <- mean(outer(pos, neg, ">")) + 0.5 * mean(outer(pos, neg, "=="))
cat("Out-of-sample AUC:", round(auc, 3), "\n")
```

**Reduced-form (naive) lens.** A logit of purchase on price recovers a coefficient
that is biased toward zero, because price proxies for the omitted positive demand
shock `xi`.

```{r modeling-reduced-naive, message=FALSE, warning=FALSE}
naive <- glm(buy ~ price, data = dat, family = binomial)
cat("Naive price coefficient:", round(coef(naive)["price"], 3),
    " (true:", alpha_true, ")\n")
```

**Structural / IV lens.** Exploit the exogenous cost shifter as an instrument via a
control function: regress the endogenous price on the instrument, take the residual,
and include it in the choice model to absorb the correlation with `xi`
[@Petrin_2010]. The price coefficient now recovers the policy-invariant primitive and
supports counterfactual repricing.

```{r modeling-structural-cf, message=FALSE, warning=FALSE}
first   <- lm(price ~ cost, data = dat)          # first stage on the instrument
dat$cfv <- resid(first)                          # control-function residual
struct  <- glm(buy ~ price + cfv, data = dat, family = binomial)
cat("Control-function price coefficient:",
    round(coef(struct)["price"], 3), " (true:", alpha_true, ")\n")
```

The three numbers tell the chapter's whole story. The predictive model attains high
out-of-sample accuracy yet its fitted price terms are not the demand slope and must
not be read as such. The naive reduced-form coefficient is attenuated by endogeneity,
a quantitative bias that no added flexibility would cure—only identifying variation
does. The control-function estimate, using the exogenous cost instrument, recovers the
true policy-invariant price coefficient and so, alone among the three, can answer the
counterfactual "what happens to demand if the firm reprices." Same data, same
question, three answers—because the three models target different things and rest on
different assumptions. Choosing among them is the skill this part of the book builds.

## Key Takeaways

- A model is an argument about a marketing system, valued for the gap between the
  assumptions it imposes and the conclusions it then licenses; the traditions differ
  in what they hold fixed when the environment changes (@sec-modeling-taxonomy).
- **Analytical** models derive equilibrium comparative statics from stated games
  [@nash1950]; they yield sign-definite, universally valid conclusions *about the
  model*, and are fatal to mis-specify the game (@sec-modeling-analytical).
- **Empirical** models split on **policy invariance**: structural estimation recovers
  invariant utility and cost primitives and so supports counterfactuals
  [@erdem1996a; @mcfadden1986choice], whereas reduced-form work estimates a possibly
  policy-variant response function and supports causal claims only under its
  identifying design [@goldfarb2022; @frennea2018] (@sec-modeling-empirical).
- **Predictive** models minimize out-of-sample risk over a flexible function class
  [@wedel2016; @athey2017digital]; they excel at forecasting under a fixed policy but
  collapse under distribution shift and never, by themselves, license a causal claim
  (@sec-modeling-predictive).
- The traditions are complements: choose by what the answer must support, where
  identifying variation comes from, and which fatal assumption is most defensible
  (@sec-modeling-decision); the same pricing question yields three different,
  internally valid answers depending on the lens (@sec-modeling-illustration).
