# Analytical Models {#sec-analytical-models}

Marketing science advances along two complementary tracks. **Analytical models**
reason about markets through pure mathematics: a researcher posits an environment
of consumers and firms, assumes that each agent optimizes, and derives the
behavior that must follow in equilibrium. **Empirical models** instead let data
speak, estimating relationships and testing the predictions that theory hands
them. This chapter is about the first track—how to build a stylized mathematical
world, solve it, and extract substantive claims about pricing, positioning,
advertising, channels, contracts, and the diffusion of new products—while keeping
one eye on the empirical counterparts that make the theory testable.

A *model*, in the working definition that organizes the field, is a representation
of the most important elements of a perceived real-world system. The art lies in
the word *important*: a model that retains everything is the territory, not the
map, and a model that retains too little predicts nothing. The discipline of
marketing modeling is therefore the discipline of principled omission. The reward
for doing it well is decision-relevant insight. @Little_1976 frames the goal as a
**decision calculus**—"a model-based set of procedures for processing data and
judgments to assist a manager in his decision making"—and argues that such a system
earns managerial adoption only when it is simple, robust, easy to control,
adaptive, as complete as possible, and easy to communicate with. Those six
criteria recur throughout this chapter as a standard against which competing
formulations are judged: a more elaborate model that a manager cannot interrogate
is often a worse model.

By the end of the chapter the reader should be able to classify a marketing
problem by the type of model it calls for, set up the corresponding game or
optimization, solve for equilibrium or the optimum, and state the assumptions whose
failure would overturn the result. We proceed from the philosophy and craft of
model building, through the workhorse spatial and oligopoly models, to the major
application domains—market response, channels, advertising, product quality under
asymmetric information, bargaining, search-theoretic pricing, promotions,
salesforce compensation, and diffusion—closing with the optimization models that
turn behavioral theory into resource-allocation decisions.

## A Taxonomy of Marketing Models

Marketing models improve decision-making in three broad families, and it is worth
fixing the distinctions before building anything, because the family determines
what counts as a valid test of the model.

**Econometric models** describe, predict, and simulate. A *descriptive* model
summarizes how a market behaves; a *predictive* model forecasts an outcome such as
sales; a *simulation* model traces how the system responds to a counterfactual
intervention. @leeflang2000 organizes the empirical tradition along exactly these
lines. Predictive models include sales models built on time-series data, trial-rate
models built on exponential growth, and the new-product growth model of
@Bass_1969. Descriptive models include purchase-incidence and purchase-timing
models built on the Poisson process, brand-choice models built on Markov or
learning processes, and pricing models for oligopolistic markets in the tradition
of @howard1968. **Normative** models, by contrast, prescribe: they maximize an
objective such as profit over price, advertising, and quality, the program studied
by @dorfman1976 and extended by @roberts1964 and @lambin1970a. The decision-calculus
program of @Little_1970 and the multinomial logit choice model brought to scanner
data by @guadagni1983 are landmarks in turning normative structure into operational
tools.

**Optimization models** take a market-response function—how outputs such as sales,
share, profit, and awareness respond to inputs such as price, advertising, and
selling effort—together with cost functions and constraints, and maximize an
objective. The candidate decisions that such models automate are familiar from
practice: promotion and pricing programs, media allocation, distribution, product
assortment, and direct-mail solicitation. **Quasi- and field-experimental
analyses** and **conjoint and choice experiments** complete the toolkit, supplying
the causal variation that observational data often lack.

Cutting across these families is the distinction between *decision-support* modeling
and *theoretical* modeling. The former describes how things work; the latter
prescribes how things *should* work. Theoretical, game-theoretic models in the
tradition of @moorthy1993 are best understood as **logical experimentation**: the
modeler specifies an environment through assumptions, deduces consequences, and
learns which assumptions drive which conclusions. @moorthy1993 separates two kinds
of assumption that play very different roles. *Mathematical* assumptions are made
for tractability and carry no empirical content; *substantive* assumptions encode
claims about the world and are the proper object of empirical testing. A classic
illustration is the design of a salesforce compensation package combining salary
and commission, which @moorthy1993 reads as a trade-off between insuring the
salesperson against income risk and motivating effort—a substantive tension we
return to in @sec-principal-agent. The internal and external validity of such a
model are questions about its boundary conditions, and Friedman's dictum looms
over the enterprise: theories are tested by their predictions, not by the realism
of their assumptions.

Because so much of analytical marketing is game theory, it helps to organize the
solution concepts by two attributes of the game: whether moves are *static*
(simultaneous) or *dynamic* (sequential), and whether information is *complete* or
*incomplete*. The four cells in @tbl-game-types name the equilibrium concept that
applies in each.

| Information | Static (simultaneous) | Dynamic (sequential) |
|---|---|---|
| **Complete** | Nash equilibrium | Subgame-perfect equilibrium |
| **Incomplete** | Bayesian Nash equilibrium (auctions) | Perfect Bayesian equilibrium (signaling) |

: Equilibrium concepts by game type and information structure {#tbl-game-types}

@moorthy1985 supplies the vocabulary that the rest of the chapter assumes.
*Rationality* is the maximization of subjective expected utility; *intelligence* is
the recognition that rival firms are themselves rational and will reason likewise.
The **rules of the game** comprise the number of firms, the feasible set of actions
for each, the utilities attached to every combination of moves, the sequence of
moves, and the information structure—who knows what, and when. *Incomplete*
information arises from unknown motivations, unknown abilities, or simply differing
knowledge of the world. A *pure strategy* is a complete plan of action; a *mixed
strategy* is a probability distribution over pure strategies; the *strategic-form*
representation lists each firm's strategy set together with its payoffs. An
**equilibrium** is a profile of strategies in which no firm would unilaterally
change its own—crucially, an equilibrium is a fixed point, not the outcome of a
dynamic adjustment process, a subtlety that separates equilibrium analysis from
the learning dynamics it is sometimes mistaken for.

These equilibrium ideas have a canonical set of applications that recur below:
oligopolistic competition in the quantity-setting tradition of Cournot and the
price-setting tradition of Bertrand; perfect competition as a limiting case;
spatial product competition following Hotelling; entry, with its first-mover
advantages and deterrence strategies; and distribution channels. Refinements of the
basic Nash concept—subgame perfectness, sequential rationality, and trembling-hand
perfectness—discipline which equilibria survive in dynamic and incomplete-information
settings, and they matter for substantive questions such as strategic entry
deterrence, implicit collusion through price-matching by leader firms, durable-goods
pricing by a monopolist, predatory and limit pricing, reputation and product
quality, and competitive bidding.

Competition itself does analytical work that bargaining cannot. @mcafee1996 shows
that competition operates under uncertainty and, in doing so, *reveals hidden
information*: in the independent-private-values benchmark, the selling price in a
standard auction equals the second-highest valuation, so the mechanism extracts
information that no single negotiation could. Sellers are generally better off
revealing what they know, because concealment invites cautious bidding driven by
the winner's curse. Competition is also computationally and strategically cheaper
than bargaining—it demands less commitment and less calculation—and it sharpens
effort incentives. These themes resurface when we treat auctions, bargaining, and
salesforce contests as alternative mechanisms for allocating scarce resources.

## The Craft of Building an Analytical Model

Before deriving any equilibrium it is worth making explicit how an analytical paper
is actually constructed, because the logic of discovery differs sharply from the
logic of presentation.[^craft] The sequence below distills the practice of
theoretical modeling in marketing.

[^craft]: The eight-step account follows lecture notes by Sajeesh Sajeesh and
reflects the conventions of the analytical-modeling seminar tradition rather than a
single published source.

The process begins with a *good idea*, drawn either from the literature or from
industry, and a hard-headed feasibility assessment: is the question interesting,
can it be told as a story, who is the target audience, and what is the opportunity
cost of pursuing it rather than something else? A counterintuitive piece of advice
follows: do not consult the literature too soon. Re-deriving a model that already
exists is not wasted effort—it trains the modeler's judgment, and the act of
building independently is what teaches which assumptions matter. The model itself
is then built from the simplest possible starting point—one period, two products, a
linear consumer utility function—with the formulation written out explicitly,
respecting the maxim that everything should be made as simple as possible, but no
simpler. Only after the simplest version is solved does one *generalize*, adding
complexity one assumption at a time, so that each new result can be attributed to a
specific relaxation. The literature search comes late and serves a sharpening
function: finding a related paper invites the productive question of why one did not
make the modeling choice the author made. The work concludes with a seminar, where
live objection stress-tests the assumptions, and finally with the written paper.

```{mermaid}
%%| label: fig-model-building
%%| fig-cap: "The analytical modeling workflow, from idea generation to written paper. Generalization and literature search are deliberately placed after a simplest-case solution is in hand."
flowchart TD
  A[Get a good idea<br/>literature or industry] --> B{Feasibility:<br/>interesting? story?<br/>audience? opp. cost?}
  B -->|yes| C[Resist the literature<br/>re-derive independently]
  C --> D[Build simplest model<br/>1 period, 2 products,<br/>linear utility]
  D --> E[Generalize:<br/>add complexity<br/>one assumption at a time]
  E --> F[Search the literature:<br/>why didn't I do<br/>what they did?]
  F --> G[Seminar:<br/>stress-test assumptions]
  G --> H[Write the paper]
```

## Spatial Competition: The Hotelling Model {#sec-26-hotelling}

Spatial models are the backbone of analytical marketing because they give *product
differentiation* a geometry. Consumers are arrayed along a line, each firm occupies
a location, and the distance between a consumer's ideal point and a firm's offering
becomes a disutility—literally a transportation cost, figuratively a measure of how
poorly the product matches the consumer's taste. The framework is flexible enough to
represent positioning, pricing, advertising, and entry within one consistent
geometry, which is why it reappears throughout the chapter.

@hotelling1929 launched this tradition with a study of *stability in competition*,
arguing—against Bertrand's critique of Cournot, which Edgeworth had elaborated—that
duopoly is inherently unstable, the instability traceable to Cournot's assumption of
absolutely identical products. Consider two sellers, A and B, on a street of length
$l$, located at distances $a$ and $b$ from the two ends, with a constant
transportation cost $c$ per unit distance. A consumer located between the firms is
*indifferent* between them when total cost—price plus travel—is equalized. Writing
$x$ and $y$ for the lengths of the market segments captured to the inside of A and
B respectively, the indifferent consumer satisfies

$$
p_1 + cx = p_2 + cy,
$$ {#eq-hotelling-indiff}

while the segments partition the street,

$$
a + x + y + b = l .
$$ {#eq-hotelling-length}

Solving @eq-hotelling-indiff and @eq-hotelling-length jointly gives the marginal
boundaries

$$
x = \tfrac{1}{2}\!\left(l - a - b + \frac{p_2 - p_1}{c}\right), \qquad
y = \tfrac{1}{2}\!\left(l - a - b + \frac{p_1 - p_2}{c}\right).
$$ {#eq-hotelling-xy}

Each firm's demand is the length of the street it captures, $q_1 = a + x$ and
$q_2 = b + y$, so with zero marginal cost the profit functions are

$$
\begin{aligned}
\pi_1 = p_1 q_1 &= \tfrac{1}{2}(l + a - b)\,p_1 - \frac{p_1^2}{2c} + \frac{p_1 p_2}{2c},\\
\pi_2 = p_2 q_2 &= \tfrac{1}{2}(l - a + b)\,p_2 - \frac{p_2^2}{2c} + \frac{p_1 p_2}{2c}.
\end{aligned}
$$ {#eq-hotelling-profit}

The price equilibrium follows from the first-order conditions
$\partial \pi_1/\partial p_1 = 0$ and $\partial \pi_2/\partial p_2 = 0$, namely

$$
\tfrac{1}{2}(l + a - b) - \frac{p_1}{c} + \frac{p_2}{2c} = 0, \qquad
\tfrac{1}{2}(l - a + b) - \frac{p_2}{c} + \frac{p_1}{2c} = 0,
$$ {#eq-hotelling-foc}

which solve to the equilibrium prices and quantities

$$
p_1 = c\!\left(l + \frac{a-b}{3}\right),\quad
p_2 = c\!\left(l - \frac{a-b}{3}\right),\qquad
q_1 = \tfrac{1}{2}\!\left(l + \frac{a-b}{3}\right),\quad
q_2 = \tfrac{1}{2}\!\left(l - \frac{a-b}{3}\right),
$$ {#eq-hotelling-eqm}

with the second-order conditions satisfied. The substantive payoff is a tension
between private and social incentives: in the choice of *locations*, central
planning ("socialism") outperforms decentralized competition ("capitalism"),
because firms have a private incentive to cluster that is socially wasteful.

@hotelling1929 conjectured a **Principle of Minimum Differentiation**—that firms
locate back-to-back at the market center. @daspremont1979 demonstrated that the
principle is *invalid* once the price subgame is treated rigorously, because the
linear-transport-cost demand is discontinuous: when prices diverge enough that one
firm could undercut and seize the entire market, the profit function jumps. Writing
the discontinuity explicitly,

$$
\pi_1(p_1, p_2) =
\begin{cases}
a p_1 + \tfrac{1}{2}(l-a-b)p_1 + \dfrac{p_1 p_2}{2c} - \dfrac{p_1^2}{2c} & \text{if } |p_1 - p_2| \le c(l-a-b),\\[2mm]
l\,p_1 & \text{if } p_1 < p_2 - c(l-a-b),\\[1mm]
0 & \text{if } p_1 > p_2 + c(l-a-b),
\end{cases}
$$ {#eq-dasp-pi1}

and symmetrically

$$
\pi_2(p_1, p_2) =
\begin{cases}
b p_2 + \tfrac{1}{2}(l-a-b)p_2 + \dfrac{p_1 p_2}{2c} - \dfrac{p_2^2}{2c} & \text{if } |p_1 - p_2| \le c(l-a-b),\\[2mm]
l\,p_2 & \text{if } p_2 < p_1 - c(l-a-b),\\[1mm]
0 & \text{if } p_2 > p_1 + c(l-a-b).
\end{cases}
$$ {#eq-dasp-pi2}

The discontinuity destroys existence of a pure-strategy price equilibrium when firms
are close together, so the minimum-differentiation result cannot hold; @daspremont1979
restore existence by replacing linear with *quadratic* transport costs, which yields
maximal differentiation instead. The methodological lesson is general: the shape of
the transport-cost function is a substantive assumption with first-order
consequences, not an innocuous mathematical convenience. @kim2006 later embeds a
preference for variety in a location model, foreshadowing the multi-purchase
extensions discussed next.

## Positioning Models {#sec-positioning}

Building on the Hotelling skeleton, a stream of work asks where firms locate when the
auxiliary assumptions are relaxed. @tabuchi1995 relaxes the assumption that consumers
are uniformly distributed. With consumers spread over $[0,1]$ according to a
cumulative distribution $F(x)$ with $F(1)=1$, two densities are compared: the
traditional uniform $f(x)=1$ and a triangular $f(x) = 2 - 2|2x - 1|$ that
concentrates consumers toward the center. Transportation cost is quadratic in
distance. The marginal consumer located between firms at $x_1 < x_2$ with prices
$p_1, p_2$ is

$$
\bar{x} = \frac{p_2 - p_1 + x_2^2 - x_1^2}{2(x_2 - x_1)},
$$ {#eq-tabuchi-marginal}

so that profits are $\Pi_1 = p_1 F(\bar{x})$ and $\Pi_2 = p_2[1 - F(\bar{x})]$ when
$x_1 < x_2$, with the roles reversed for $x_1 > x_2$ and a Bertrand subgame when
$x_1 = x_2$. @tabuchi1995 establishes that when firms choose locations
*simultaneously* and then prices *simultaneously*, and the density is log-concave, a
unique Nash price equilibrium exists. Under the uniform density firms locate as far
apart as possible—consistent with shopping centers sited away from city centers,
though at the cost of forcing consumers to buy products far from their ideal. Under
the triangular density no symmetric location equilibrium exists, but two asymmetric
equilibria do, with lower profits for both firms. When locations are instead chosen
*sequentially* and prices simultaneously, the first entrant locates at the market
center under both densities—a first-mover result that contrasts with the maximal-
differentiation outcome of simultaneous choice.

@sajeesh2010 turns to *variety-seeking*, modeling satiation as a relative reduction
in willingness to pay for the previously purchased brand—a form of negative state
dependence. The received wisdom held that variety-seeking consumers let firms charge
higher prices and earn higher profits; @sajeesh2010 shows the opposite, that average
prices and profits are *lower*, because firms cut second-period prices to deter
switching. The timing is a three-stage game: firms choose locations simultaneously
in period 0, prices simultaneously in period 1, and prices again simultaneously in
period 2.

Several papers vary the order and structure of competition. @moorthy1988 has two
identical firms choose product *quality* first and then price, a two-stage structure
we formalize under vertical differentiation in @sec-quality. @tyagi2000 extends
@hotelling1929, @tyagi1999, and @tabuchi1995 to firms that enter *sequentially* with
*different cost structures*, and—against the usual presumption that moving first is
advantageous—demonstrates a *second-mover advantage*. @kim2006 allows consumers to
make *multiple* purchases, so that some are loyal to a single brand while others
consume more than one product, enriching the demand side beyond unit demand.
@shreay2015 shows that quantity surcharges across different package sizes of the same
product—an apparent violation of quantity discounting—can be rationalized by
consumer preferences when the sizes are imperfect substitutes, i.e., differentiated
products.

## Oligopoly: Market Structure and Frameworks {#sec-oligopoly}

The spatial models above derive demand from geometry. The oligopoly tradition begins
instead from an aggregate demand function and asks how firms compete given it. Three
canonical structures organize the field: **Bertrand** competition, in which firms set
prices; **Cournot** competition, in which firms set quantities; and **Stackelberg**
competition, the leader–follower structure in which one firm commits before the
other. Because everything flows from the demand function, it matters where that
function comes from; @shubik1980 derives aggregate demand from primitives rather than
positing it, supplying the microfoundation that the reduced-form models take as
given. As noted above, @moorthy1988 studies how two firms compete on product quality
and price under both simultaneous and sequential timing, bridging the spatial and
oligopoly traditions.

### Cournot Competition: Simultaneous Quantity Setting {#sec-cournot}

Let two firms produce a homogeneous good with constant marginal costs $c_1, c_2$, so
total cost is $TC_i = c_i q_i$. Inverse demand is linear, $P(Q) = a - bQ$ with
$Q = q_1 + q_2$. Each firm chooses its quantity to maximize profit,

$$
\pi_1 = [a - b(q_1 + q_2)]q_1 - c_1 q_1, \qquad
\pi_2 = [a - b(q_1 + q_2)]q_2 - c_2 q_2 .
$$ {#eq-cournot-profit}

The first-order conditions define each firm's **reaction** (best-response) function,

$$
\frac{\partial \pi_1}{\partial q_1} = a - 2 b q_1 - b q_2 - c_1 = 0, \qquad
\frac{\partial \pi_2}{\partial q_2} = a - 2 b q_2 - b q_1 - c_2 = 0,
$$ {#eq-cournot-foc}

which rearrange to

$$
q_1 = \frac{a - c_1}{2b} - \frac{q_2}{2} \equiv R_1(q_2), \qquad
q_2 = \frac{a - c_2}{2b} - \frac{q_1}{2} \equiv R_2(q_1).
$$ {#eq-cournot-reaction}

The Cournot–Nash equilibrium is the intersection of the two reaction functions.
Substituting $R_2$ into $R_1$ gives
$q_1 = \frac{a-c_1}{2b} - \frac{a-c_2}{4b} + \frac{q_1}{4}$, and solving,

$$
q_1^* = \frac{a - 2c_1 + c_2}{3b}, \qquad q_2^* = \frac{a - 2c_2 + c_1}{3b},
$$ {#eq-cournot-eqm}

with total quantity $Q = \frac{2a - c_1 - c_2}{3b}$ and equilibrium price
$P = a - bQ = \frac{a + c_1 + c_2}{3}$. The structure is transparent: a firm with
lower marginal cost produces more, and the equilibrium price is a simple average of
the demand intercept and the two cost levels.

### Stackelberg Competition: Sequential Quantity Setting {#sec-stackelberg}

Now suppose Firm 1 (the *leader*) chooses quantity in stage 1, and Firm 2 (the
*follower*) responds in stage 2, with common marginal cost $c_1 = c_2 = c$. Solving
by backward induction, the follower's stage-2 best response is
$R_2(q_1) = \frac{a-c}{2b} - \frac{q_1}{2}$. The leader anticipates this and chooses
$q_1$ to maximize

$$
\pi_1 = \left[a - b\!\left(q_1 + \frac{a-c}{2b} - \frac{q_1}{2}\right)\right] q_1 - c q_1 ,
$$ {#eq-stackelberg-leader}

whose first-order condition $\frac{a+c}{2} - b q_1 - c = 0$ yields the
**Stackelberg equilibrium**

$$
q_1^* = \frac{a - c}{2b}, \qquad q_2^* = \frac{a - c}{4b}.
$$ {#eq-stackelberg-eqm}

Comparing with the symmetric Cournot outcome $q_1 = q_2 = \frac{a-c}{3b}$, the
leader produces *more* and the follower *less* than under simultaneous competition.
The leader's ability to commit first—to move down the follower's reaction
function—is worth real profit, which is the analytical content of "first-mover
advantage" in quantity competition. @tbl-cournot-stackelberg summarizes the
contrast.

| Quantity | Cournot (simultaneous) | Stackelberg (sequential) |
|---|---|---|
| Leader / Firm 1 | $\dfrac{a-c}{3b}$ | $\dfrac{a-c}{2b}$ |
| Follower / Firm 2 | $\dfrac{a-c}{3b}$ | $\dfrac{a-c}{4b}$ |
| Total $Q$ | $\dfrac{2(a-c)}{3b}$ | $\dfrac{3(a-c)}{4b}$ |

: Equilibrium quantities under Cournot and Stackelberg with common marginal cost $c$ {#tbl-cournot-stackelberg}

### Differentiation and the Value of Independence {#sec-differentiation-beta}

When products are differentiated rather than homogeneous, a firm's profit depends on
how substitutable its product is for its rival's. Let $\beta$ index substitutability
(higher $\beta$ meaning closer substitutes) and let $d$ index a second structural
parameter such as a demand or cost asymmetry, with $\bar d$ a threshold. For a
"weak" firm $W$, the comparative static

$$
\frac{d \pi_W^*}{d \beta} < 0 \quad \text{for all } d < \bar{d}
$$ {#eq-beta-weak}

shows that profit falls as products become closer substitutes, so the weak firm
prefers a *lower* $\beta$—it wants a more independent, more differentiated product.
For a "strong" firm $S$, the relationship is non-monotone in $d$:

$$
\frac{d \pi_S^*}{d \beta} < 0 \quad \text{over a range } d < \bar{d},
$$ {#eq-beta-strong}

so that the strong firm's profit rises as $\beta$ falls when $d$ is small, but rises
as $\beta$ increases when $d$ is large. The substantive reading is that strong firms
can benefit from *less* differentiation (closer substitutes) precisely when the
structural asymmetry $d$ is large, because their strength lets them win the more
intense competition that proximity creates—whereas weak firms always seek the shelter
of differentiation.

## Entry, Retail Structure, and Conspicuous Consumption {#sec-entry-structure}

The market-structure machinery extends naturally to questions of entry and the
vertical organization of retailing. @dixit1980 reconsiders the **Bain–Sylos
postulate**, under which an incumbent builds enough capacity to render entry
unprofitable. The key insight is one of *credibility*: investment in capacity deters
entry only if it is an irreversible commitment, because if the incumbent can costlessly
change capacity after entry, the threat is empty and the incumbent cannot in fact
deter entry. Commitment, not capacity per se, is what does the strategic work—a theme
shared with the Stackelberg leader's first-mover commitment.

@tyagi1999a decomposes the effect of an additional retailer on prices. The entry of
the $(n+1)$-st retailer sets two forces against each other: a **competition effect**
that lowers retail prices, and an **input-cost effect** on the wholesale price.
Manufacturers, gaining bargaining power as the number of retailers grows, may raise
the wholesale price; because each retailer's quantity choice depends on the wholesale
price, this can *raise* downstream prices, partially or wholly offsetting the
competition effect. More retailers therefore need not mean lower consumer prices.
@jerath2016 reaches a related conclusion through a different channel: when an organized
retailer enters a market, inefficient unorganized retailers exit, and the surviving
unorganized retailers *raise* their prices, leaving customers worse off.

Conspicuous consumption introduces consumption *externalities* into the Hotelling
framework. @amaldoss2005 models two opposing motives in the pricing of conspicuous
goods—a *desire for uniqueness* and a desire for *conformism*—and shows they push
prices in opposite directions: greater desire for uniqueness raises equilibrium
prices and profits, whereas greater desire for conformity lowers them. For a *snob*
who dislikes crowding, utility from the two firms is

$$
U_A = V - p_A - \theta t_s - \lambda_s(n_A), \qquad
U_B = V - p_B - (1-\theta) t_s - \lambda_s(n_B),
$$ {#eq-snob}

while for a *conformist* who values crowding, the externality enters with the
opposite sign,

$$
U_A = V - p_A - \theta t_s + \lambda_c(n_A), \qquad
U_B = V - p_B - (1-\theta) t_s + \lambda_c(n_B),
$$ {#eq-conformist}

where $\theta$ is the consumer's Hotelling position, $t_s$ the transport cost,
$\lambda_s$ and $\lambda_c$ the sensitivities to the externality, and $n_A, n_B$ the
numbers buying each brand. The novel solution concept is **rational-expectations
equilibrium**: the number of buyers a consumer *expects* must equal the number that
in fact materializes. Let $\theta_s$ mark the marginal snob and $\theta_c$ the
marginal conformist, with $\beta$ the fraction of snobs in the market. Setting
$U_A = U_B$ for each segment pins down $\theta_s$ and $\theta_c$, and the
expectations consistency conditions are

$$
\beta \theta_s + (1-\beta)\theta_c = n_A, \qquad
\beta(1-\theta_s) + (1-\beta)(1-\theta_c) = n_B,
$$ {#eq-ree}

where $\beta \theta_s$ is the mass of snobs buying from A, $(1-\beta)\theta_c$ the
mass of conformists buying from A, and the remaining terms the masses buying from B.
The equilibrium expectation is *realized*: what consumers anticipate is what occurs.
This expectation is not an empirical expectation in the statistical sense
$\mathbb{E}[\cdot]$—it carries no measurement error—but an internal-consistency
requirement of the model. The striking comparative static, confirmed both
analytically and in laboratory tests, is that a price *increase* can raise snobs'
desire to buy, because higher price enhances the exclusivity the snob values.

Limited editions and exclusivity round out the theme. @balachander2009 shows that
adding a limited-edition product has a positive *direct* effect on profits, through
the higher willingness to pay it commands, but a negative *strategic* effect, through
the intensified price competition it triggers between brands. Under vertical (quality)
differentiation, the high-quality brand gains from limited editions; under horizontal
(taste) differentiation, the negative strategic effect lowers equilibrium profits for
both brands, yet both still introduce limited editions because not doing so is
dominated—a prisoners'-dilemma structure. @sajeesh2020 distinguishes
*functionality-oriented* from *exclusivity-oriented* consumer segments and shows that
firms increase value enhancements when functionality-oriented consumers perceive
greater product differentiation, but decrease them when exclusivity-oriented consumers
do, because exclusivity is undercut by enhancements that everyone can enjoy.

## Market Response Models {#sec-market-response}

Optimization of marketing spend requires a function mapping inputs to outputs. The
**market-response model** is that function. Its inputs are marketing instruments—
selling effort, advertising spending, promotional spending—and its outputs are
sales, share, profit, and awareness. The modeling problem is to choose a functional
form flexible enough to capture the empirical regularities of how output responds to
input, yet parsimonious enough to estimate and to optimize.

```{mermaid}
%%| label: fig-response-io
%%| fig-cap: "The input–output logic of a market-response model. Marketing actions are the controllable inputs; observed market outcomes are the outputs; the response function is the mapping the modeler must specify and estimate."
flowchart LR
  subgraph Inputs[Marketing actions: inputs]
    A1[Selling effort]
    A2[Advertising spend]
    A3[Promotional spend]
  end
  R[(Response<br/>function)]
  subgraph Outputs[Observed market outcomes]
    O1[Sales]
    O2[Share]
    O3[Profit]
    O4[Awareness]
  end
  A1 --> R
  A2 --> R
  A3 --> R
  R --> O1
  R --> O2
  R --> O3
  R --> O4
```

A good dynamic response model captures several stylized facts. Sales response is
dynamic, with a *growth* rate and a *decay* rate that differ; the steady-state
response can be concave or S-shaped, with positive sales even at zero advertising;
competitive effects matter; advertising effectiveness itself shifts with media, copy,
and other factors; and sales continue to rise or fall even while advertising is held
constant. Saunders codified a finer list of phenomena any response form might need to
reproduce: zero output at zero input; a linear input–output relationship; decreasing
returns to scale; an output ceiling (saturation); increasing returns to scale; an
S-shaped pattern in which returns first increase then decrease; a threshold below
which input produces no output; and a supersaturation point beyond which additional
input *reduces* output.

### A Catalogue of Aggregate Response Forms {#sec-response-forms}

No single functional form reproduces every phenomenon, so the modeler chooses among a
catalogue, each member of which buys some flexibility at the cost of some. The
**linear** model $Y = a + bX$ is the simplest but admits only constant returns and so
cannot represent concave, convex, or S-shaped response. The **power-series /
polynomial** model $Y = a + bX + cX^2 + dX^3 + \cdots$ adds curvature but cannot
capture saturation or a threshold. The **fractional-root / power** model $Y = a + bX^c$
fixes the exponent $c$ in advance: $c = \tfrac12$ gives the *square-root* model,
$c = -1$ the *reciprocal* model, $c = 1$ the linear case, $c < 1$ decreasing returns,
$c > 1$ increasing returns; when $a = 0$, $c$ is the constant elasticity. The
**semilog** model $Y = a + b\ln X$ suits situations in which a constant *percentage*
increase in effort yields a constant *absolute* increase in sales. The **exponential**
model $Y = a e^{bX}$ (for $X > 0$) gives increasing, convex returns when $b > 0$ and
decreasing returns with saturation when $b < 0$. The **modified exponential**
$Y = a(1 - e^{-bX}) + c$ delivers decreasing returns and saturation, with upper bound
$a + c$ and lower bound $c$, and is the workhorse for selling-effort response. The
**logistic** $Y = \frac{a}{a + e^{-(b + cX)}} + d$ produces increasing-then-decreasing
returns—an S-shape—with saturation at $a + d$, and is preferred when both saturation
and an S-shape are required; the **Gompertz** model is a related asymmetric S-curve.

The **ADBUDG** model of @Little_1970,

$$
Y = b + (a - b)\frac{X^c}{d + X^c},
$$ {#eq-adbudg}

is the most versatile of the single-instrument forms: it is S-shaped when $c > 1$ and
concave with a saturation effect when $0 < c < 1$, with upper bound $a$ and lower
bound $b$. It passes through the origin or not as specified, handles concavity,
saturation, and S-shapes within one expression, and is standard for advertising and
selling-effort response. Multiple instruments are accommodated **additively**,
$Y = a f(X_1) + b g(X_2)$; **multiplicatively**, $Y = a X_1^b X_2^c$ with $b$ and $c$
as elasticities, or more generally $Y = a f(X_1)\cdot b g(X_2)$; or in a combined
**multiplicative-and-additive** form $Y = a f(X_1) + b g(X_2) + c f(X_1) g(X_2)$ that
allows interaction between instruments. Finally, the **dynamic** response model
$Y_t = a_0 + a_1 X_t + \lambda Y_{t-1}$ separates the *current* effect $a_1$ from the
*carry-over* effect $\lambda$, the lagged influence of past sales on present sales.

### Dynamic Effects and Adstock {#sec-adstock}

Marketing effects rarely land all at once. The **carry-over effect** captures how
current expenditure influences *future* sales—advertising *adstock* is precisely this
lagged effect of advertising on sales. Related dynamics include the *delayed-response*
effect (a lag between investment and impact), *customer-holdout* effects, *hysteresis*
(persistent effects that do not reverse when spend reverses), *new-trier* and
*wear-out* effects, and *stocking* effects whereby promotions shift purchase timing.
The simplest decay model writes adstock as a geometrically weighted sum of current and
lagged spend,

$$
A_t = T_t + \lambda T_{t-1}, \qquad t = 1, 2, \dots,
$$ {#eq-26-adstock}

where $A_t$ is adstock at time $t$, $T_t$ is advertising spend at $t$, and $\lambda$ is
the decay (lag-weight) parameter governing how slowly past advertising depreciates.

Response models can be classified along six axes that, together, determine which form
is appropriate: the number of marketing variables; whether competition is included;
the nature of the input–output relationship (linear versus S-shaped); whether the
setting is static or dynamic; whether response is individual or aggregate; and the
level of demand analyzed (sales versus market share).

### From Category Sales to Brand Share {#sec-attraction}

Brand sales factor into category demand and within-category share, $Y = M \times V$,
where $Y$ is brand sales, $V$ is product-class (category) sales, and $M$ is market
share. The dominant family for the share component is the **attraction** model, in
which a brand's share equals its *attractiveness* relative to the total
attractiveness in the category,

$$
M_i = \frac{A_i}{A_1 + \cdots + A_n},
$$ {#eq-attraction}

with $A_i$ the attractiveness of brand $i$. Attraction models guarantee shares that
are non-negative and sum to one—a logical-consistency property that ad hoc share
regressions lack.

At the individual level, the **multinomial logit** model gives the probability that
consumer $i$ chooses brand $l$ as

$$
P_{il} = \frac{e^{A_{il}}}{\sum_j e^{A_{ij}}},
$$ {#eq-26-mnl}

where $A_{ij} = \sum_k w_k b_{ijk}$ is the attractiveness of product $j$ to consumer
$i$, $b_{ijk}$ is consumer $i$'s evaluation of product $j$ on attribute $k$, the sum
running over the products $i$ considers, and $w_k$ is the importance weight on
attribute $k$. The logit is the individual-level microfoundation of the aggregate
attraction model and the bridge to the choice-modeling tradition of @guadagni1983; we
treat its estimation in the conjoint and choice-modeling discussion below.

### Worked Illustration: Estimating an ADBUDG Response Curve {#sec-adbudg-code}

To make the response-modeling apparatus concrete, the following code simulates an
advertising response that is genuinely S-shaped, then recovers the ADBUDG parameters
by nonlinear least squares and plots the fit. The example is seeded for
reproducibility.

```{r}
#| label: fig-adbudg-fit
#| fig-cap: "Simulated advertising response with an S-shaped ADBUDG curve recovered by nonlinear least squares. Points are noisy observations; the line is the fitted response function."
#| message: false
#| warning: false
set.seed(2026)

# True ADBUDG parameters: Y = b + (a - b) * X^c / (d + X^c)
a_true <- 100   # upper bound (saturation)
b_true <- 5     # lower bound (sales at zero advertising)
c_true <- 2.2   # S-shape (c > 1)
d_true <- 40

adbudg <- function(X, a, b, c, d) b + (a - b) * X^c / (d + X^c)

X <- seq(0, 10, length.out = 60)
Y_mean <- adbudg(X, a_true, b_true, c_true, d_true)
Y <- Y_mean + rnorm(length(X), sd = 4)

fit <- nls(
  Y ~ b + (a - b) * X^c / (d + X^c),
  start = list(a = 90, b = 10, c = 1.5, d = 30)
)

round(coef(fit), 2)

plot(X, Y, pch = 16, col = "grey50",
     xlab = "Advertising spend (X)", ylab = "Sales response (Y)")
lines(X, predict(fit, newdata = list(X = X)), lwd = 2)
```

## Conjoint and Augmented Conjoint Analysis {#sec-conjoint-models}

Conjoint analysis estimates how consumers trade off product attributes, and the
analytical-modeling literature has extended it from preference measurement toward
managerially decisive quantities such as reservation prices. The core of conjoint is
treated in @sec-conjoint; here the concern is the augmentations that connect
attribute part-worths to pricing and demand.

@jedidi2002 augments conjoint to estimate consumer **reservation prices**—the maximum
a consumer would pay—rather than mere preference rankings. Starting from a rating-based
utility specification

$$
\text{Utility (rating)} = \alpha + \sum_i \beta_i\, \text{Attribute}_i,
$$ {#eq-conjoint-utility}

with $\alpha$ a baseline constant and $\beta_i$ the part-worth of attribute $i$, the
method maps estimated part-worths into reservation prices for a product in a category.
The resulting estimates support decisions about product introduction, the switching
effect (consumers drawn from rival brands), the cannibalization effect (consumers
drawn from the firm's own existing products), and the market-expansion effect (new
demand created). @netzer2011 decomposes conjoint into a sequence of constant-sum
paired-comparison questions, which not only estimates attribute importances but also
yields *standard errors* for each importance weight—a statistical refinement that lets
the analyst distinguish attributes that genuinely matter from those whose apparent
importance is noise.

## Distribution Channels {#sec-channels}

A *channel* is a vertical sequence of firms—manufacturers upstream, retailers
downstream—each setting prices or quantities to maximize its own profit. The central
analytical phenomenon is **double marginalization**: when an upstream firm marks up
over cost and a downstream firm marks up again over the wholesale price, the final
price exceeds the integrated-monopoly price and channel profit falls. Whether to
integrate, and how to coordinate, are the questions the channel literature answers.

@mcguire1983 is the foundational treatment. Two manufacturers produce differentiated,
competing products and reach consumers through downstream retailers. Three vertical
structures are compared: a **decentralized** system with two manufacturers and two
independent retailers (four players); a **vertically integrated** system in which each
manufacturer owns its retailing (two players); and a **mixed** system with one
integrated manufacturer and one manufacturer–independent-retailer pair (three
players). Each retail outlet faces a downward-sloping demand $q_i = f_i(p_1, p_2)$;
under the decentralized system the Nash-equilibrium retail demands become functions of
the wholesale prices, $q_i^* = g_i(w_1, w_2)$. The decision variables are prices, not
quantities, and retail demand is assumed linear in prices.

```{mermaid}
%%| label: fig-channel-structures
%%| fig-cap: "Three vertical channel structures in @mcguire1983. Left: fully decentralized (four players). Center: mixed (three players). Right: fully integrated (two players). The optimal structure depends on the degree of retail-level substitutability."
flowchart TD
  subgraph Dec[Decentralized: 4 players]
    M1[Mfr 1] --> R1[Retailer 1]
    M2[Mfr 2] --> R2[Retailer 2]
  end
  subgraph Mix[Mixed: 3 players]
    M3[Mfr 1] --> R3[Retailer 1]
    M4[Mfr 2 integrated store]
  end
  subgraph Int[Integrated: 2 players]
    M5[Mfr 1 integrated store]
    M6[Mfr 2 integrated store]
  end
```

The demand system, after specifying linear retail demands, is

$$
q_1' = \mu S\!\left[1 - \frac{\beta}{1-\theta}\,p_1' + \frac{\beta\theta}{1-\theta}\,p_2'\right],
$$ {#eq-mcguire-q1}

$$
q_2' = (1-\mu) S\!\left[1 + \frac{\beta\theta}{1-\theta}\,p_1' - \frac{\beta}{1-\theta}\,p_2'\right],
$$ {#eq-mcguire-q2}

with $0 \le \mu, \theta \le 1$ and $\beta, S > 0$. Here $S$ is a scale factor equal to
industry demand $q' \equiv q_1' + q_2'$ at zero prices, $\mu$ governs the asymmetry in
demand between firms, and $\theta$ measures **substitutability**—the ratio of the
cross-price sensitivity to the own-price sensitivity. At $\theta = 0$ demands are
independent and each firm is a monopolist; as $\theta \to 1$ the products become
maximally substitutable. Feasibility requires that prices cover fixed manufacturing
and selling costs $m'$ and $s'$ and that demands stay non-negative,

$$
P = \{\, p_1', p_2' \mid p_i' - m' - s' \ge 0,\ i=1,2;\ (1-\theta) - \beta p_1' + \beta\theta p_2' \ge 0,\ (1-\theta) + \beta\theta p_1' - \beta p_2' \ge 0 \,\},
$$ {#eq-mcguire-feasible}

which forces $\beta \le \frac{1}{m' + s'}$; and for industry demand not to rise with
either price, $\frac{\theta}{1+\theta} \le \mu \le \frac{1}{1+\theta}$. After
rescaling, industry demand is $q = 2(1-\theta)(p_1 + p_2)$.

The central result connects channel design to substitutability. When each manufacturer
is a monopolist ($\theta = 0$), selling through a *private* (independent) channel is
twice as profitable as integrating. When demand is maximally driven by rival retailers'
actions ($\theta \to 1$), private dealers are *three times* as profitable. The
break-even degree of substitutability is $\theta \approx 0.708$. The conclusion is that
the optimal distribution system depends on the degree of substitutability at the retail
level: decentralization, by softening downstream price competition, can be a strategic
asset rather than merely a source of double marginalization.

A complementary literature studies channel *coordination* devices. @jeuland2008
catalogues four rationales for quantity discounts—cost-based economies of scale,
demand-based reasons (large buyers are more price-sensitive), strategic single-sourcing,
and channel coordination—and contributes the last, showing how a quantity-discount
schedule can align an independent retailer's incentives with the manufacturer's.
@moorthy1987 analyzes second-degree price discrimination as a channel and screening
device, while @geylani2007 and @jerath2010 extend the channel framework to the
strategic interaction between powerful retailers and manufacturers.

## Advertising Models {#sec-26-advertising}

Advertising enters consumer utility through three distinct mechanisms, and the choice
among them is the central modeling decision. **Informative** advertising increases
overall demand for the brand by informing more consumers of its existence;
**persuasive** advertising shifts demand toward the brand by raising perceived value or
loyalty; **comparative** advertising shifts demand *away* from a competitor. The
spatial Hotelling framework hosts all three cleanly. With $n$ consumers distributed
uniformly along the line—an apt assumption for a mature market with stable demand—base
utilities are

$$
U_A = V - p_A - t x, \qquad U_B = V - p_B - t(1-x),
$$ {#eq-ad-base}

where $t$ is the transport cost, reinterpretable here as a return-on-advertising
parameter.

**Persuasive** advertising can be modeled as scaling the value term, $U_A = A_A V - p_A
- tx$, or as raising the reservation price through a concave transformation, $U_A =
\sqrt{Ad_A}\,V - p_A - tx$, or as effectively pulling the firm's product closer to the
consumer's ideal, $U_A = V - p_A - \frac{tx}{\sqrt{Ad_A}}$. **Comparative** advertising
instead inflates the *rival's* effective distance,

$$
U_A = V - p_A - t\sqrt{Ad_B}\,x, \qquad U_B = V - p_B - t\sqrt{Ad_A}\,(1-x).
$$ {#eq-ad-comparative}

The marginal consumer solves $U_A = U_B$, giving

$$
x = \frac{1}{t\sqrt{Ad_A} + t\sqrt{Ad_B}}\left(-p_A + p_B + t\sqrt{Ad_A}\right),
$$ {#eq-ad-marginal}

so the (concave) profit functions, net of advertising cost, are

$$
\pi_A = p_A\, x\, n - \phi\, Ad_A, \qquad \pi_B = p_B (1-x)\, n - \phi\, Ad_B,
$$ {#eq-ad-profit}

with $\phi$ the per-unit cost of advertising (higher for television than for online,
say) and $t$ interpreted as the return on advertising. Solving the price subgame
conditional on advertising levels yields

$$
p_A = \tfrac{2}{3} t\sqrt{Ad_A} + \tfrac{1}{3} t\sqrt{Ad_B}, \qquad
p_B = \tfrac{1}{3} t\sqrt{Ad_A} + \tfrac{2}{3} t\sqrt{Ad_B}.
$$ {#eq-ad-prices}

Prices rise with advertising intensity—invest more, charge more—and each firm's price
rises in its *rival's* advertising as well. Optimizing advertising given these prices,
the symmetric competitive equilibrium is

$$
Ad_A = Ad_B = \frac{25\, t^2 n^2}{576\, \phi^2}, \qquad p_A = p_B = \frac{5\, t^2 n}{24\, \phi}.
$$ {#eq-ad-eqm}

As the cost of advertising $\phi$ rises, firms advertise less; as the return $t$ rises,
they benefit more. Notably, equilibrium prices are *higher* with advertising than
without it. Because colluding on prices is illegal and easily detected while colluding
on advertising is hard to observe, firms may tacitly coordinate on advertising—pulsing,
for instance.[^aluf] Setting a common collusive level $Ad_A = Ad_B = Ad$,

$$
Ad_A = Ad_B = \frac{t^2 n^2}{16\, \phi^2} > \frac{25\, t^2 n^2}{576\, \phi^2},
$$ {#eq-ad-collusive}

the collusive advertising level exceeds the competitive one, supporting higher prices
and leaving consumers worse off. A model combining comparative and persuasive
advertising writes the rival's relative advertising into the transport term,

$$
U_A = V - p_A - tx\,\frac{\sqrt{Ad_B}}{\sqrt{Ad_A}}, \qquad
U_B = V - p_B - t(1-x)\,\frac{\sqrt{Ad_A}}{\sqrt{Ad_B}} .
$$ {#eq-ad-combined}

**Informative** advertising, by contrast, works by increasing the number $n$ of
potential customers—the natural specification for a new product whose pool of aware
consumers is still growing. A subtle assumption throughout is that the advertising
decision precedes the pricing decision, which is reasonable because prices are easy to
adjust while advertising budgets are committed at the start of a period. Consumption
itself responds to availability: consumers consume more when stock is plentiful and
less when it is scarce [@ailawadi1998].

[^aluf]: The collusion-through-advertising mechanism connects to a broader literature
on pulsing and tacit coordination in advertising spend.

The empirical and theoretical literature refines these mechanisms. @villas-boas1993
shows that under monopoly, firms do better to **pulse**—alternating between a minimum
and an efficient level of advertising—because the advertising response function is
S-shaped; the analysis assumes Markov strategies, so that current actions depend only
on payoff-relevant state rather than on history, and establishes that if the loss from
letting the consideration level lapse exceeds the efficient advertising expenditure,
the unique Markov-perfect equilibrium has both firms always advertising.

@nelson1974 supplies the information-economics foundation that distinguishes advertising
by *what consumers can verify*. **Search qualities** are knowable before purchase;
**experience qualities** are not. Misleading advertising risks credibility and imposes
the cost of processing non-buying customers, and there is an inverse association between
quality produced and utility-adjusted price. Firms that wish to sell more advertise
more, and they advertise to their natural audience—those whose tastes a brand best
serves are the most likely to see its advertisement. Advertising for experience
qualities conveys information *indirectly* (chiefly by signaling that the seller finds
it worthwhile to advertise), whereas advertising for search qualities conveys it
*directly*. Goods are accordingly classified by whether their quality variation is
search- or experience-based, into experience durables, experience nondurables, and
search goods. Experience goods are advertised more, because advertising raises sales by
raising the seller's reputability; the marginal revenue of advertising is nonetheless
higher for search goods, which concentrate in newspapers and magazines while experience
goods migrate to other media. For experience goods, word of mouth is a better
information source than advertising, and purchase frequency moderates the contrast—low-
frequency purchases lean on word of mouth. When laws against deception are *moderately*
enforced, deceptive advertising flourishes: too little enforcement and consumers
distrust all claims, too much and deception is deterred, but a moderate regime lets
consumers believe and advertisers cheat, which is why experience goods—where
enforcement concentrates—carry more deceptive advertising.

Targeting sharpens these effects. @iyer2005 shows that firms advertising to their
*targeted* market—consumers with a strong preference for their product—rather than to
competitors' loyalists endogenously *increase* market differentiation and raise
equilibrium profits. Targeted advertising is more valuable than targeted pricing:
targeted advertising raises profits whether or not firms also price-discriminate,
whereas targeted pricing merely intensifies competition for comparison shoppers without
improving equilibrium profits. The mass of comparison shoppers is $s = 1 - 2h$, where
$h$ is the size of each firm's loyal segment and thus also indexes the differentiation
between firms; $A$ denotes the cost of advertising to the entire market and $r$ the
reservation price. @chen2009 distinguishes *combative* from *constructive* advertising
and integrates the informative, complementary, and persuasive mechanisms: informative
advertising raises awareness, lowers search costs, and increases differentiation;
complementary advertising raises utility by signaling social prestige; persuasive
advertising lowers price sensitivity, with combative advertising a special case.
Consumer response moderates the effect of combative advertising on price competition—it
generally *decreases* competition, but it *increases* competition when consumer
preferences are biased toward advertising firms and disfavored firms cannot advertise
and can only respond on price, because an advertising war then spills into a price war.

## Product Differentiation, Quality, and Asymmetric Information {#sec-quality}

Differentiation comes in two forms with very different welfare logic. **Horizontal**
differentiation means different consumers prefer different products—there is no
universal ranking—and is naturally represented by a quadratic taste loss,

$$
U = V - p - t(\theta - a)^2,
$$ {#eq-horizontal}

where $\theta$ is the consumer's ideal point and $a$ the product's location.
**Vertical** differentiation means one product is unambiguously *better*: all consumers
agree on the ranking and differ only in willingness to pay for quality. With qualities
$s_A, s_B$ and a taste parameter $\theta$,

$$
U_A = \theta s_A - p_A, \qquad U_B = \theta s_B - p_B,
$$ {#eq-vertical}

where $\theta$ is the consumer's position on the quality-valuation line and a consumer
buys only if utility is non-negative. Taking $B$ as the higher-quality product, the
consumer indifferent between the two is found from $\theta s_B - p_B = \theta s_A -
p_A$, giving the boundary

$$
\bar{\theta} = \frac{p_B - p_A}{s_B - s_A}.
$$ {#eq-vertical-boundary}

At equal prices every consumer prefers the higher quality, the *characteristics*
approach treating a product as the bundle of its characteristics.

@moorthy1988 closes the vertical-differentiation model with quality-dependent costs.
With profits $\pi_A = (p_A - c s_A^2)\,\text{(share}_A)$ and $\pi_B = (p_B - c s_B^2)\,
\text{(share}_B)$ and the lower boundary $\bar\theta_2 = p_A / s_A$ marking the consumer
indifferent between buying the low-quality product and not buying, the two firms choose
qualities first and prices second, and differentiate their qualities to relax price
competition. @wauthy1996 characterizes when the market is *covered*: with $\theta$
drawn over a support of width $b/a$, full coverage requires

$$
2 \le \frac{b}{a} \le \frac{2 s_2 + s_1}{s_2 - s_1}.
$$ {#eq-wauthy}

A modeling subtlety follows: with $\theta \in [0,1]$ one cannot simultaneously have full
market coverage, so the support is often re-specified, for instance

$$
\theta \in
\begin{cases}
[1,4] & \text{if } b/a = 4,\\
[1,2] & \text{if } b/a = 2,
\end{cases}
$$ {#eq-wauthy-support}

to obtain coverage while preserving the vertical structure.

### Asymmetric Information: Adverse Selection and Signaling {#sec-asymmetric}

Quality is often *unobservable* at the point of purchase, and the consequences are
profound. Two failures must be distinguished by their timing relative to the contract.
**Adverse selection** is a *pre*-contract problem: a relevant characteristic is
uncertain before agreement. **Moral hazard** is a *post*-contract problem: intentions
or actions are hidden after agreement. The adverse-selection problem can unravel a
market entirely.

A transparent setup adapts Akerlof's used-car market. Car quality is $\theta \in [0,1]$;
a seller owning a car of type $\theta$ values it at $\theta$, while a buyer's willingness
to pay is $\tfrac{3}{2}\theta$, so trade is mutually beneficial whenever it occurs. If
quality is *observable* to both sides, any price $p(\theta) \in [\theta, \tfrac{3}{2}
\theta]$ supports trade and both parties gain. If quality is *unobservable* but
symmetrically so—neither side knows $\theta$, which is uniform—then $\mathbb{E}[\theta] =
\tfrac12$, the seller's expected value is $\tfrac12$, the buyer's is $\tfrac{3}{2}\cdot
\tfrac12 = \tfrac34$, and trade occurs at any $p \in [\tfrac12, \tfrac34]$. But under
genuine *asymmetric* information—only the seller knows $\theta$—a seller offers a car
only if $p \ge \theta$, so the cars on offer have quality in $[0, p]$, and a buyer
reasoning correctly computes $\mathbb{E}[\theta \mid \theta \le p] = \tfrac12 p$, giving
buyer utility $\tfrac34 p < p$. The buyer's value falls short of the price for every
$p$, and *no trade occurs*. Asymmetric information, not uncertainty per se, is what
destroys the market.

@akerlof1970 is the founding analysis of this adverse-selection mechanism, framed as the
relationship between quality and uncertainty in the automobile market across a
$2\times 2$ of used-versus-new and good-versus-bad. Let $q$ be the probability of a good
car and $1-q$ the probability of a *lemon*. Sellers know the probability that a given car
is bad; buyers do not, and pay the same price for a lemon as for a good car. The result
contrasts with Gresham's law: bad money drives out good only because the exchange rate is
fixed and the two are distinguishable, whereas the buyer cannot tell a good car from a
lemon at all. Formally, demand depends on price and average quality, $Q^d = D(p, \mu)$;
supply depends on price, $S = S(p)$; and average quality is itself a function of price,
$\mu = \mu(p)$, with equilibrium $S(p) = D(p, \mu(p))$—and the punchline is that *at no
price will any trade happen*.

@akerlof1970 makes this concrete with two trader groups. The first has utility $U_1 = M +
\sum_{i=1}^n x_i$ and the second $U_2 = M + \sum_{i=1}^n \tfrac{3}{2} x_i$, where $M$ is
consumption of other goods, $x_i$ is the quality of car $i$, and $n$ the number of cars,
with incomes $Y_1, Y_2$. Group-1 demand is

$$
D_1 =
\begin{cases}
Y_1 / p & \text{if } \mu/p > 1,\\
0 & \text{if } \mu/p < 1,
\end{cases}
$$ {#eq-akerlof-d1}

and with quality uniform, the supply offered by group 1 is $S_1 = pN/2$ for $p \le 2$,
carrying average quality $\mu = p/2$. Group-2 demand is

$$
D_2 =
\begin{cases}
Y_2 / p & \text{if } \tfrac{3}{2}\mu > p,\\
0 & \text{if } \tfrac{3}{2}\mu < p,
\end{cases}
$$ {#eq-akerlof-d2}

with $S_2 = 0$. Total demand under asymmetric information is therefore

$$
D(p, \mu) =
\begin{cases}
(Y_1 + Y_2)/p & \text{if } p < \mu,\\
Y_2 / p & \text{if } \mu < p < \tfrac{3}{2}\mu,\\
0 & \text{if } p > \tfrac{3}{2}\mu,
\end{cases}
$$ {#eq-akerlof-total}

and since average quality at price $p$ is $\mu = p/2$, the middle and upper branches bind
and no trade occurs. Under *symmetric* information the market revives: with quality
uniform on $[0,2]$, supply is $S(p) = N$ for $p > 1$ and zero otherwise, demand is
$(Y_1+Y_2)/p$ for $p<1$, $Y_2/p$ for $1<p<\tfrac32$, and zero above, and equilibrium price
is

$$
p =
\begin{cases}
1 & \text{if } Y_2 < N,\\
Y_2 / N & \text{if } \tfrac{2}{3}Y_2 < N < Y_2,\\
\tfrac{3}{2} & \text{if } N < \tfrac{2}{3}Y_2,
\end{cases}
$$ {#eq-akerlof-sym}

so trade resumes once quality is commonly observed. The mechanism generalizes to
insurance for the elderly, the employment of minorities, the cost of dishonesty, and
credit markets in developing economies, and the institutional remedies—guarantees,
brand names, retail chains, and licensing—are precisely the devices that re-establish
trust where quality is uncertain.

@spence1973 builds the *signaling* response to adverse selection on Akerlof's foundation.
Two worker types produce one and two units respectively, with a fraction $\alpha$ of type
1 and $1-\alpha$ of type 2, so average productivity is $\mathbb{E}[P] = \alpha + 2(1-
\alpha) = 2 - \alpha$. Workers can signal type through education, whose cost is $E$ for
type 1 and $E/2$ for the more able type 2. A *separating* equilibrium requires the signal
to be incentive-compatible: type 1 must find it not worthwhile to acquire the education,
$2 - E < 1 \Rightarrow E > 1$, while type 2 must find it worthwhile, $2 - E/2 > 1
\Rightarrow E < 2$. Any cost $1 < E < 2$ therefore sustains a separating equilibrium in
which education credibly distinguishes the types even though it adds nothing to
productivity—the canonical demonstration that a costly, otherwise-useless signal can
convey information when its cost is differentially borne.

@moorthy1995 carries the signaling logic into marketing through **money-back guarantees**.
A guarantee signals quality, the transaction cost being whatever the buyer or seller pays
to redeem it; the guarantee refunds the purchase price without compensating the buyer's
own return expense. If signaling were costless, money-back guarantees and price would be
interchangeable signals. Because signaling is costly, the results sharpen: with
homogeneous buyers, a low-quality seller cannot profitably mimic a high-quality seller's
guarantee. With heterogeneous buyers, the picture is richer—when transaction costs are
very high the seller forgoes the guarantee or signals through price instead, and when they
are moderate there is a critical value of seller transaction cost below which the
high-quality seller's profit *increases* in transaction cost and above which it
*decreases*. A striking corollary is that uninformative advertising—"money-burning"
expenditure that does not shift demand directly—is *never* needed when a guarantee is
available. Moral hazard remains a caveat: consumers may exhaust the product's value within
the guarantee window before claiming a refund. @tbl-moorthy-quality lays out the seller's
cost structure, the engine of the separating equilibrium.

| | High-quality seller ($h$) | Low-quality seller ($l$) |
|---|---|---|
| Unit cost ($c_h > c_l$) | $c_h$ | $c_l$ |

: Cost asymmetry underlying the money-back-guarantee signaling model {#tbl-moorthy-quality}

## Bargaining {#sec-bargaining}

Bargaining is the allocation mechanism for the thinnest of markets: one buyer and one
seller. @tbl-allocation places it among the alternatives—many buyers and many sellers
constitute a traditional market, many buyers and one seller an auction, and one buyer and
one seller a bargaining problem. Before invoking game theory at all, a share may be fixed
by an arbitrator, by meeting in the middle, or by a forced-final take-it-or-leave-it move.
Negotiation is the art; bargaining is the science; game theory's contribution is to the
*rules* of the encounter, and the area remains fertile for research.

| Buyers and sellers | Allocation type |
|---|---|
| Many buyers and many sellers | Traditional market |
| Many buyers and one seller | Auction |
| One buyer and one seller | Bargaining |

: Allocation mechanisms by market thickness {#tbl-allocation}

### Non-cooperative Bargaining {#sec-noncoop-bargaining}

The non-cooperative approach specifies the rules explicitly and solves for the
subgame-perfect outcome. The simplest rule is a **take-it-or-leave-it** offer over a cake:
if the responder accepts, the parties trade; if she rejects, no one eats. Under perfect
information this has a trivial rollback equilibrium. More realistically, bargaining is
"take-it-or-counteroffer," and *time has value*: both parties prefer to trade sooner, so
delay—labor negotiations dragging into strikes and work stoppages—destroys surplus.

Consider **two-stage** bargaining over a cake. In the first period I offer you a share,
keeping proportion $p$; if you reject, you counteroffer, but a fraction $\delta$ of the
cake melts in the interim. Payoffs are $(1-p, p)$ if you accept immediately and
$\big((1-\delta)(1-p),\,(1-\delta)p\big)$ in the second period. Because period 2 is the
final period, it is itself a take-it-or-leave-it offer: you will offer me the smallest
slice I will accept, keeping essentially all of the remaining $1-\delta$. Rolling back, in
period 1 I must give you at least the surplus you would secure by waiting. Your second-period
surplus is $1-\delta$, so you accept a first-period offer only if your share $p \ge 1 -
\delta$; I therefore offer you exactly $1-\delta$ and we settle immediately. The comparative
static is sharp: the more *patient* you are—the slower the cake melts, the smaller is
$\delta$—the more you receive now. Whether the first or second mover is advantaged depends
on $\delta$: a fast-melting cake (high $\delta$) favors the first mover, a slow-melting one
(low $\delta$) the second. Either way, if both parties reason, agreement is reached in the
first period, and the general lesson is to strike a deal as early as possible.

Why, then, do real negotiations delay? Reputation-building and a lack of information.
Information asymmetry is the deeper answer—uncertainty about the likelihood of success, as
in civil litigation. Each side has a *reservation price* (in a lawsuit, its expectation of
winning) that is unknown to the other; one must probabilistically choose a best offer,
accepting that the very probability implies a chance that no bargain is struck. The rules of
the game uniquely determine the outcome; which rules favor whom depends on patience and
information; and persistent delay means *someone must be wrong*.

A worked screening example shows information at work. A company negotiates with a union. The
firm is "highly profitable" (annual profit \$200k) with probability $p$; a strike costs the
firm 10% of annual profits. If the union demands a \$150k wage it is surely accepted; if it
demands \$200k it is accepted only with probability $p$, for an expected wage of $200p$. At
$p = 0.9$ the best opening demand is \$200k, worth $0.9 \times 200 = 180$; at $p = 0.1$ a
\$200k demand is worth only $0.1 \times 200 = 20$, so the union does better to ask \$150k—it
is not worth the risk. Now suppose a first-period offer can be rejected at the cost of a
strike. Because a strike costs a high-profit firm *more* than a low-profit firm, the union
can use time to **screen**. If the union asks \$170k in period 1, a low-profit firm (worth
\$150k) rejects because it cannot afford to accept, while a high-profit firm weighs
rejection—at best the union strikes and then asks only \$140k, but the strike has meanwhile
cost the firm \$20k—and accepts. The result is a **separating equilibrium**: only high-profit
firms accept in the first period, and a rejection reveals a low-profit firm, against which the
union then asks \$140k. The union's apparent "giving in" after a rejection is not weakness
but screening—different firm types value the future differently, and time is the screening
device.

### Cooperative Bargaining and the Nash Solution {#sec-nash-bargaining}

The cooperative approach abstracts from the procedural rules and asks what division *should*
result, given axioms the solution must satisfy. Two people divide a sum: if they disagree each
gets nothing, and they cannot divide more than the whole. @nash1950 founds the axiomatic theory
for this bilateral-monopoly, nonzero-sum two-person game, in which no action by one party,
without the other's consent, can affect the other's gain. The assumptions are rational
individuals who maximize gain, full knowledge of tastes and preferences, transitive ordering
($A \succ C$ whenever $A \succ B$ and $B \succ C$), and continuity. Utility is ordinal-plus:
$u(A) > u(B)$ means $A$ is preferred, and the linearity property requires that for $0 \le p \le
1$, $u(pA + (1-p)B) = p\,u(A) + (1-p)\,u(B)$, extending to two persons as $p[A,B] + (1-p)[C,D] =
[pA + (1-p)C,\ pB + (1-p)D]$. An *anticipation* $pA + (1-p)B$ is a lottery giving event $A$ with
probability $p$.

Let $u_1, u_2$ be the players' utility functions and $c(S)$ the solution point selected from a
bargaining set $S$ that is compact, convex, and contains the disagreement point at the origin.
@nash1950 characterizes $c(S)$ by three axioms. *Efficiency*: if some $\beta \in S$ gives both
players strictly more than $\alpha \in S$, then $\alpha \neq c(S)$—players maximize utility.
*Independence of irrelevant alternatives*: if $S \subseteq T$ and $c(T) \in S$, then $c(T) =
c(S)$. *Symmetry*: if $S$ is symmetric about the line $u_1 = u_2$, then $c(S)$ lies on that line,
encoding equal bargaining power. These axioms uniquely select the division that maximizes the
product of the players' gains over their disagreement payoffs—the Nash bargaining solution.

The bargaining lens illuminates channel power. @iyer2003 shows that the presence of a powerful
retailer such as Walmart can benefit *all* channel members, not merely the retailer, by
reshaping the terms of trade. @desai2004 analyzes haggling: with two consumer segments—hagglers
and non-hagglers—a haggling policy can be more profitable than fixed pricing when the proportion
of non-hagglers is sufficiently high, because the firm extracts surplus from hagglers without
sacrificing the non-hagglers who pay the posted price.

## Pricing and Search Theory {#sec-search}

If all consumers were perfectly informed, the law of one price would hold and price dispersion
would be a puzzle. Search theory resolves the puzzle by recognizing that *information is costly*,
and that the resulting mix of informed and uninformed consumers can sustain dispersion as an
*equilibrium* phenomenon rather than a disequilibrium artifact.

@varian1980 builds an equilibrium theory of price dispersion on the foundation laid by Stigler
and developed by @salop1977 and @stiglitz1982. Dispersion comes in two flavors: *spatial*
dispersion, arising when uninformed and informed consumers coexist—though it need not persist,
since consumers learn from experience over time—and *temporal* dispersion, manifest as sales.
@varian1980 follows @stiglitz1982, who assumes informed consumers who buy at the lowest-price
store and uninformed consumers who choose a store at random, and @shilony1977, who studies
randomized pricing strategies. Let $I > 0$ be the number of informed consumers, $M > 0$ the
number of uninformed, and $n$ the number of stores, so $U = M/n$ uninformed consumers patronize
each store. Each store draws its price from a density $f(p)$: it *succeeds*—charging the lowest
of the $n$ prices—and then serves $I + U$ customers, or it *fails* and serves only $U$; stores
tying at the lowest price split the informed segment equally. With cost curve $c(q)$, the price
$p^* = c(I+U)/(I+U)$ is the average cost at the maximal customer count. @varian1980 establishes a
sequence of properties of the equilibrium mixed strategy: the density vanishes outside $[p^*, r]$
where $r$ is the reservation price (Prop. 1); no symmetric pure-price equilibrium exists (Prop.
2); the equilibrium has no point masses (Prop. 3); where $f(p) > 0$, the indifference condition

$$
\pi_s(p)\,(1 - F(p))^{n-1} + \pi_f(p)\,\big[1 - (1 - F(p))^{n-1}\big] = 0
$$ {#eq-varian-indiff}

holds, equating the expected profit from a successful and a failed price draw (Prop. 4); the
quantity $\pi_f(p)\,(\pi_f(p) - \pi_s(p))$ is strictly decreasing in $p$ (Prop. 5); the
distribution puts positive mass just above $p^*$ and below $r$ (Props. 6–7); and there are no gaps
in the support (Prop. 8). Whether a consumer chooses to *become* informed is itself endogenous,
turning on the "full price" of search—search costs plus any fixed cost of information.

@lazear1984 turns to retail pricing and clearance sales, asking how a good's characteristics, the
thinness of its market, and its time on the shelf shape its price path. In a single-period model,
let $V$ be the valuation of the lone potential buyer, with prior density $f(V)$ and distribution
$F(V)$; the firm chooses price $R$ to maximize $R[1 - F(R)]$, and with $V$ uniform on $[0,1]$ this
gives $R = \tfrac12$ and expected profit $\tfrac14$. The dynamics emerge in a two-period model. A
failed sale in period 1 reveals $V < R_1$, so by Bayes' rule the period-2 posterior is uniform on
$[0, R_1]$ with $F_2(V) = V/R_1$; the first-period price thus does double duty, affecting both
period-1 sales and period-2 information. The firm maximizes

$$
\max_{R_1, R_2}\ R_1[1 - F(R_1)] + R_2[1 - F_2(R_2)]\,F(R_1),
$$ {#eq-lazear-two}

solving the period-2 subproblem $\max_{R_2} R_2[1 - F_2(R_2)]$ with posterior

$$
F_2(R_2) =
\begin{cases}
F(R_2)/F(R_1) & \text{for } R_2 < R_1,\\
1 & \text{otherwise.}
\end{cases}
$$ {#eq-lazear-posterior}

The first-order conditions imply that the second-period price always lies below the first, and
expected profit exceeds the one-period benchmark because the option to mark down raises the
probability of an eventual sale. With a concrete uniform example, the two-period prices are $R_1 =
\tfrac23$ and $R_2 = \tfrac13$ (since $R_2 = R_1/2$ from the second-period first-order condition),
against the one-period $R_1 = \tfrac12$. The model abstracts from brand recognition and from
contagion or network effects.

Thin markets and consumer heterogeneity enrich the price path. Suppose $N$ customers inspect the
good, each a *shopper* (valuation $0$) with prior probability $P$ or a *buyer* (valuation $V$) with
probability $1-P$; a person learns which she is only upon inspecting. Three types coexist:
*customers*, all who inspect; *buyers*, whose value equals $V$; and *shoppers*, whose value is
zero. The firm maximizes

$$
\max_{R_1, R_2}\ R_1(1 - F(R_1))(1 - P^N) + R_2\big\{(1 - F_2(R_2))(1 - P^N)\big\}\big\{1 - [(1 - F(R_1))(1 - P^N)]\big\},
$$ {#eq-lazear-thin}

with period-2 density, by Bayes' rule,

$$
f_2(V) =
\begin{cases}
\dfrac{1}{R_1(1 - P^N) + P^N} & \text{for } V \le R_1,\\[3mm]
\dfrac{P^N}{R_1(1 - P^N) + P^N} & \text{for } V > R_1.
\end{cases}
$$ {#eq-lazear-thin-density}

As $P^N \to 1$—almost everyone is a shopper—there is little information to be gleaned, so the
two-period problem collapses to two independent one-period problems and prices are roughly constant.
When $P^N$ is small, prices start high and fall rapidly as time unsold accumulates. The composite
$P^N$ thus stands in for search cost and information. Observable price–quantity patterns follow from
three primitives: the number of customers $N$, the proportion of shoppers $P$, and the firm's prior
on $V$. Where prices fall quickly, the chance the good goes unsold is low; a high *initial* price
signals a low $P^N$—few shoppers—and so predicts a likely sale. Greater prior dispersion raises the
expected price for a given mean but, through longer shelf time, can lower expected revenue. The more
obsolescence-prone ("fashion") the good, the more anxious the seller and the steeper the markdowns;
"classic" goods carry higher initial prices that are less sensitive to inventory. A constant discount
rate is irrelevant to the pricing condition, unlike an increasing obsolescence rate, and for a
non-unique good the solution reduces to the one-period problem. Costs themselves fall through three
distinct channels worth keeping separate: *economies of scale*, $c(\text{units})$; *economies of
scope*, $c(\text{types of products})$, typically via knowledge transfer; and the *experience effect*,
$c(\text{cumulative output over time})$, a superset of scale economies.

The Internet sharpened the search question. Conventional wisdom held that lower search costs intensify
price competition, but the analytical literature complicates this. @lal1999 distinguishes *digital*
product attributes, communicable over the Internet, from *nondigital* attributes that are not, on a
demand side where consumers vary, and a supply side where firms operate both traditional and Internet
stores. Monopoly pricing can survive—prices stay high and search is discouraged—when a high proportion
of consumers are Internet users, nondigital attributes are not overwhelming, consumers favor familiar
brands, and shopping is destination-driven; stores serve to *acquire* customers while the Internet
*retains* loyal ones. @kuksov2004 shows the design margin matters: for products whose design cannot be
easily changed, lower search costs *raise* price competition, but for products whose design is flexible,
lower search costs induce firms to *increase* differentiation, which *softens* price competition, lowers
social welfare, and raises industry profits—so the welfare effect of cheaper search hinges on whether
firms can respond by redesigning. @salop1977 supplies the foundational model of costly search underlying
this stream.

## Pricing and Promotions {#sec-promotions}

Pricing promotions are among the most heavily studied phenomena in marketing, and the analytical lens
reframes "loyalty" in the language of information: an uninformed consumer behaves like a loyal one, and
an informed consumer like a non-loyal switcher. The strategic questions are whether to run everyday-low-
pricing or high–low pricing, and how to design short-term discounts—trade deals to the channel, and
consumer promotions such as shelf-price discounts (used by everyone) and cents-off coupons (used by
consumers whose time is relatively cheap).

The decomposition of what a price cut actually does is well established. @gupta1988 attributes the bulk of
the measured sales response to **brand switching** (about 84% of the effect), with **purchase
acceleration** (about 14%) and **quantity** increases (about 2%) accounting for the rest, and finds the
elasticity of short-term price changes to be an order of magnitude larger than other elasticities. Beyond
the measured effects, promotions can drive general trial, encourage consumers to hold inventory and thereby
raise consumption, lift sales of complementary products, and produce small store-switching effects—with an
*asymmetric* pattern in which stronger brands benefit more, except against store brands. Against these
benefits stand persistent costs: consumers come to *expect* future promotions, their *reference price*
ratchets down, their price sensitivity rises, and a *post-promotion dip* follows as stockpiled demand is
repaid.

Trade discounts—short-term discounts offered to the channel rather than the consumer—incentivize retailers
to push the product and capture the sales force's attention, but they may not be passed through to consumers,
they invite *forward buying* that disrupts production planning, they complicate demand forecasting, and they
breed an expectation of future discounts that becomes a cost of doing business. *Scanback* arrangements,
which tie the discount to verified retail sales, raise pass-through. Pass-through is higher when consumer
elasticity is higher, the promoting brand is stronger, the demand function has a favorable shape, and
promotions are infrequent. On shelf-price discounts, @raju1990 shows that a stronger brand can discount
*infrequently* precisely because weaker brands cannot predict when it will promote and so must promote more
often to defend their position.

Coupons illustrate the design subtleties. Little over 1% of coupons are redeemed annually, their
price-discrimination power has eroded with their ubiquity, and the sales lift required to make
free-standing-insert coupons profitable is often unattainable. Coupon *design*—expiration dates and
distribution method—therefore matters: stronger brands should set *shorter* expiration windows, because a
larger share of their loyal base will use the coupon regardless; in-store distribution dominates
through-package and targeted distribution on average. Package coupons embody an acquisition–retention
trade-off across three types: *peel-off* coupons are used by many but yield the lowest firm profit;
*in-pack* coupons attract fewer first-period buyers; and *on-pack* coupons, redeemed on a subsequent
purchase, are typically the best approach. The summary verdict is that trade and consumer promotions are
both necessary, that consumer promotion should lean toward package coupons rather than shelf-price discounts
or newspaper coupons, and that the strong interaction between advertising and promotion remains an open
research frontier.

It helps to organize discrimination by *degree*: first-degree discrimination prices on individual willingness
to pay, second-degree on quantity, third-degree on observable membership, and fourth-degree on cost-to-serve.

@narasimhan1988b formalizes promotions as **mixed-strategy** pricing driven by brand loyalty. Marketing tools
to promote a product—advertising, trade promotions, and consumer promotions—shade into pricing instruments
such as price deals, cents-off labels, coupons, and rebates, and brand loyalty explains the variation in
prices observed in competitive markets, as firms trade off attracting switchers against the profit forgone on
loyal customers; a deviation below the maximum price *is* a promotion. The model assumes firms with identical
products and (constant or declining) costs playing a non-cooperative game, a common reservation price, and
three consumer segments: loyals to firm 1 of size $\alpha_1$, loyals to firm 2 of size $\alpha_2$ (with
$0 < \alpha_2 < \alpha_1$, an asymmetric duopoly), and switchers of size $\beta = 1 - \alpha_1 - \alpha_2$, with
costless price changes and no intertemporal effects. Switchers may be modeled with an individual switching cost
$d \in (-b, a)$, so that a consumer buys brand 1 if $P_1 \le P_2 - d$ and brand 2 otherwise; with identical
switchers (common $d$); or with $d = 0$ (extreme price sensitivity). The first case admits a pure-strategy
equilibrium; the latter two admit only mixed strategies. For the $d = 0$ case, the profit function is

$$
\Pi_i(P_i, P_j) = \alpha_i P_i + \delta_{ij}\,\beta P_i, \qquad
\delta_{ij} =
\begin{cases}
1 & \text{if } P_i < P_j,\\
\tfrac12 & \text{if } P_i = P_j,\\
0 & \text{if } P_i > P_j,
\end{cases}
$$ {#eq-narasimhan-profit}

for $i = 1, 2$, $i \neq j$. There is **no pure-strategy Nash equilibrium**: each firm randomizes its price.
Writing the expected profit under randomization,

$$
\Pi_i(P_i) = \alpha_i P_i + \Pr(P_j > P_i)\,\beta P_i + \Pr(P_j = P_i)\,\tfrac{\beta}{2} P_i,
$$ {#eq-narasimhan-mixed}

each firm chooses a pricing distribution $F_i$ to solve

$$
\max_{F_i}\ \mathbb{E}(\Pi_i) = \int \Pi_i(P_i)\, dF_i(P_i)
$$ {#eq-narasimhan-program}

subject to a participation floor $\Pi_i \ge \alpha_i r$ (where $r$ is the reservation price), $\int dF_i(P_i) =
1$, and support $P_i \in S_i^*$. The equilibrium mixed strategy is the analytical signature of promotional
pricing: firms do not promote on a schedule but randomize, making their discount timing unpredictable to rivals.

Two extensions broaden the promotional toolkit. @balachander2010 shows that *bundle discounts* can dominate
price promotions in a competitive market, because the bundle raises loyalty and thereby softens the intensity of
promotional competition. @goi2011 studies *cross-market discounts*, in which a purchase in a source market earns
a discount redeemable in a target market, raising both prices and sales in the source market.

## Market Entry and New-Product Diffusion {#sec-diffusion}

How a new product spreads through a population is the province of **diffusion** models, which sit at the boundary
between descriptive and predictive modeling and feed directly into entry-timing decisions. The product life
cycle—the rise, peak, and decline of category sales—is the empirical regularity these models explain.

@Bass_1969 is the canonical formulation. Sales arise from two sources: *innovators*, who adopt for the product's
intrinsic benefits, and *imitators*, who adopt after interacting with prior adopters. Let $p$ be the coefficient
of innovation (the fraction of the untapped market that adopts spontaneously each period), $q$ the coefficient of
imitation (the fraction of interactions that lead to adoption), $M$ the market potential, $N(t)$ cumulative sales
through time $t$, and $M - N(t)$ the untapped market. Sales in any period are spontaneous adoptions plus
interaction-driven adoptions,

$$
S(t) = p\,[M - N(t)] + q\,\frac{N(t)}{M}\,[M - N(t)] = pM + (q - p)\,N(t) - \frac{q}{M}\,[N(t)]^2,
$$ {#eq-bass}

a relation from which $p$, $q$, and $M$ can be estimated from sales data. When $q > p$—imitation dominates
innovation—the model generates the bell-shaped life cycle with an interior sales peak; when $p$ dominates, sales
decline monotonically from launch.

```{r}
#| label: fig-bass-diffusion
#| fig-cap: "Bass diffusion: cumulative adoption (left axis logic) and per-period sales (the bell-shaped curve) for an imitation-dominated market ($q > p$). The interior peak is the signature of word-of-mouth-driven diffusion."
#| message: false
#| warning: false
set.seed(2026)
bass_sales <- function(t, p, q, M) {
  # closed-form Bass non-cumulative adoption f(t) * M
  num <- (p + q)^2 * exp(-(p + q) * t)
  den <- p * (1 + (q / p) * exp(-(p + q) * t))^2
  M * num / den
}
t  <- seq(0, 20, by = 0.1)
p  <- 0.03   # coefficient of innovation
q  <- 0.38   # coefficient of imitation (q > p => bell shape)
M  <- 1000   # market potential
S  <- bass_sales(t, p, q, M)

plot(t, S, type = "l", lwd = 2,
     xlab = "Time", ylab = "Per-period adoptions S(t)",
     main = "")
abline(v = t[which.max(S)], lty = 2, col = "grey50")
```

Pioneering is not the unalloyed advantage folklore suggests. @golder1993 overturns earlier estimates that relied on
limited databases such as PIMS and ASSESSOR [@urban1986], that excluded non-survivors, and that depended on
single-informant self-reports. With a dataset free of these biases, @golder1993 finds that about half of market
pioneers *fail* and that their mean market share is far lower than previously believed, while *early market
leaders*—who enter on average thirteen years after the pioneers—enjoy greater long-term success. The taxonomy is
precise: an *inventor* develops the patent or core technology of a new category; a *product pioneer* builds the first
working model; a *market pioneer* is first to sell; and a *product category* is a group of close substitutes.
@boulding2003 sharpens the timing story: first-to-market confers an initial profit *advantage* that lasts roughly
twelve to fourteen years before turning into a long-term *disadvantage*, with consumer learning, market position, and
patent protection moderating the effect.

@golder2004 studies the shape of the life cycle itself. Consumer durables typically grow about 45% per year over an
eight-year growth stage, then *slow down*—sales decline by about 15% and remain below the previous peak for some five
years—with the slowdown usually arriving when the product has penetrated 35–50% of the market. Larger initial sales
increases beget larger declines at slowdown; leisure-enhancing products grow faster but for shorter stages, while
time-saving products grow slower but longer; and a lower likelihood of slowdown correlates with steeper price
reductions, lower penetration, and higher economic growth. A hazard model predicts both takeoff and slowdown
reasonably well. @vandenbulte2007 refines the social structure of diffusion by splitting the market into
*influentials*, who are aware of new developments and affect imitators, and *imitators*, who model their behavior on
influentials. This two-segment structure is consistent with observed features such as the dip between the early and
later phases of the diffusion curve, and it matters for inference: erroneously fitting a mixed-influence model to a
process in which influentials act independently can generate the systematic parameter drift reported in earlier
research. The two-segment model outperforms the standard mixed-influence, Gamma/Shifted-Gompertz, and Weibull-Gamma
specifications and performs comparably to the Karmeshu–Goswami mixed-influence model.

## Principal–Agent Models and Salesforce Compensation {#sec-principal-agent}

Salesforce compensation is the marketing application of **principal–agent** theory. A firm (the principal) cannot
perfectly observe the effort of its salesperson (the agent); it must design a wage contract that induces effort while
respecting the agent's outside option. Two questions organize the field: how to ensure agents exert effort, and how to
design compensation so that they exert *high* effort. The answer depends on whether effort is observable and on how
costly monitoring is. The timing is canonical: the manager designs the contract, offers it, the worker accepts or
walks, the worker chooses effort, and finally the outcome is observed and the wage paid.

```{mermaid}
%%| label: fig-agency-timing
%%| fig-cap: "Timing of the principal–agent contracting game. The agent's effort choice is unobservable to the principal, which is the source of the moral-hazard problem the contract must solve."
flowchart LR
  S1[1. Manager designs<br/>contract] --> S2[2. Manager offers;<br/>worker accepts or rejects]
  S2 --> S3[3. Worker chooses<br/>effort level]
  S3 --> S4[4. Outcome observed;<br/>wage paid]
```

### Building Intuition: Three Scenarios {#sec-agency-scenarios}

The mechanics are clearest in a binary-effort example. The worker chooses effort $e \in \{0, 2\}$—high effort costs
2 units of disutility, shirking costs 0—against a reservation utility of 10 available elsewhere. The agent's utility is

$$
U =
\begin{cases}
w - e & \text{if he exerts effort } e,\\
10 & \text{if he takes the outside option,}
\end{cases}
$$ {#eq-agency-utility}

and revenue depends on effort, $R(e) = H$ if $e = 2$ and $R(e) = L$ if $e = 0$, with the contract specifying wages
$w^H$ and $w^L$ contingent on observed revenue.

**Scenario 1 — certainty and observable output.** With a deterministic link from effort to revenue, two constraints
bind. The *participation* constraint $w^H - 2 \ge 10$ keeps the worker in the market, and the *incentive-compatibility*
constraint $w^H - 2 \ge w^L - 0$ ensures effort is preferred to shirking. The solution is $w^H = 12$, $w^L = 10$: when
output reveals effort, the contract is simple.

**Scenario 2 — uncertainty.** Now effort shifts the *distribution* of revenue rather than determining it. High effort
yields $H$ with probability 0.8 and $L$ with 0.2; shirking yields $H$ with probability 0.4 and $L$ with 0.6. With the
worker risk-neutral, the participation and incentive-compatibility constraints become

$$
0.8\,w^H + 0.2\,w^L - 2 \ge 10, \qquad 0.8\,w^H + 0.2\,w^L - 2 \ge 0.4\,w^H + 0.6\,w^L,
$$ {#eq-agency-uncertain}

solving to $w^H = 13$, $w^L = 8$. The expected wage bill $13 \times 0.8 + 8 \times 0.2 = 12$ equals the certainty cost:
*under risk neutrality, uncertainty alone does not raise the wage bill*.

**Scenario 3 — asymmetric beliefs and risk aversion.** A risk-averse worker assigns more pessimistic odds to high
output than the risk-neutral manager—say the worker perceives $H$ with probability 0.7 under high effort where the
manager believes 0.8, a gap the manager understands. The participation constraint becomes $0.7\,w^H + 0.3\,w^L - 2 \ge
10$ and the incentive-compatibility constraint (taking $R(0)$ from Scenario 2) becomes $0.6\,w^H + 0.3\,w^L - 2 \ge
0.4\,w^H + 0.6\,w^L$, which reduce to

$$
0.7\,w^H + 0.3\,w^L = 12, \qquad 0.3\,w^H - 0.3\,w^L = 2,
$$ {#eq-agency-asym}

with solution $w^H = 14$, $w^L = 22/3$. The manager's expected wage bill is now $14 \times 0.8 + (22/3)\times 0.2 \approx
12.67$, *higher* than under risk neutrality: the worker's risk aversion forces the firm to pay a risk premium. The
qualitative map of optimal contracts appears in @tbl-compensation.

| Effort | Risk-neutral salesperson | Risk-averse salesperson |
|---|---|---|
| **Observable** | Any salary–commission mix; desired effort | All salary; desired effort |
| **Not observable** | All commission; desired effort | Specific salary–commission mix; salesperson shirks at the margin |

: Optimal compensation structure by observability of effort and salesperson risk attitude {#tbl-compensation}

### The Basu–Lal–Srinivasan–Staelin Framework {#sec-bls}

@grossman1986 is the landmark formal treatment of the principal–agent problem; @basu1985 brings it into marketing and
is the reference point for the salesforce-compensation literature. The relevant compensation plans are those that are
independent of performance (straight salary), partly dependent on output (salary plus commission), defined relative to
others (sales contests), or chosen by the salesperson from a menu. Within the first two categories the three canonical
schemes are straight salary, straight commission, and a combination. @tbl-comp-types summarizes when each is preferred.

| Compensation type | Best when | Limitation |
|---|---|---|
| Straight salary | Long-term objectives; performance hard to measure | Induces less effort |
| Straight commission | Performance easy to measure | Imposes high risk on the salesperson |
| Combination | Intermediate cases | Requires tuning the salary–commission split |

: Compensation schemes, their ideal conditions, and limitations {#tbl-comp-types}

The conceptual advance of @basu1985 is to replace the deterministic effort–sales relationship of earlier work with a
*stochastic* one. The firm is risk-neutral and maximizes expected profit; the salesperson is risk-averse, with
diminishing marginal utility of income ($U(s) \ge 0$, $U'(s) > 0$, $U''(s) < 0$) and an additively separable utility
$U(s) - V(t)$ in salary $s$ and effort $t$, where the disutility of effort is increasing and convex ($V(t) \ge 0$,
$V'(t) > 0$, $V''(t) > 0$). Production has constant marginal cost $c$, both parties know the utility and sales-response
functions, and dollar sales $x$ follow a Gamma (or Binomial) distribution with density $f(x \mid t)$. The firm solves

$$
\max_{s(x)}\ \int \big[(1-c)x - s(x)\big] f(x \mid t)\, dx
$$ {#eq-bls-firm}

subject to the salesperson's *participation* constraint, that expected utility weakly exceed the outside option $m$,

$$
\int U(s(x))\, f(x \mid t)\, dx - V(t) \ge m,
$$ {#eq-bls-participation}

and the salesperson's own optimization—the *incentive-compatibility* condition—that effort be chosen to maximize utility,

$$
\max_t\ \int U(s(x))\, f(x \mid t)\, dx - V(t).
$$ {#eq-bls-agent}

The optimal commission schedule is generally a *nonlinear* (curvilinear) function of sales.

A sequence of papers probes and simplifies this framework. @lal1986 shows that a *menu* of compensation plans, from which
the salesperson self-selects, is optimal precisely when @basu1985's homogeneity assumptions fail—when salespeople differ
in risk characteristics, reservation utility, or information about the environment. If the firm can otherwise distinguish
high from low performers (or build a selection mechanism), a single contract in the style of @basu1985 suffices; if it
cannot, the menu does the screening. @raju1996 compares a *quota-based* plan to the @basu1985 curvilinear plan and finds
the simpler quota plan close to optimal: it differs materially only in special cases—about 1% of simulated scenarios—where
two sources of loss appear, a *shape-induced* nonoptimality (the quota plan's total-compensation-versus-sales curve is not
the general curvilinear form) and a *heterogeneity-induced* nonoptimality (a common commission rate is applied across the
salesforce). Despite these, the quota plan loses little in optimality and is far more robust to a salesperson switching
territories or to territorial change in business conditions, because the firm can adjust the *quota* per territory rather
than re-engineering the entire commission structure as @basu1985 would require. Heterogeneity in this literature stems from
the salesperson side—effectiveness, risk level, disutility for effort, and alternative opportunities—and the territory
side—sales potential and volatility—and adjusting quotas accommodates both. The nonoptimality assessment follows the
risk-analytic apparatus of @basu1990.

@raju1996 grounds its sales technology in an explicit specification. Expected sales per period are linear in effort,

$$
\mathbb{E}[x_i \mid t_i] = h_i + k_i t_i, \qquad h_i > 0,\ k_i > 0,
$$ {#eq-raju-sales}

where $h_i$ is the base sales level and $k_i$ the effectiveness of effort, with $1/\sqrt{c}$ the coefficient of
variation (uncertainty) in sales, so that $c \to \infty$ is perfect certainty. The salesperson's overall utility takes the
constant-relative-risk-aversion-plus-power-disutility form

$$
U_i[s_i(x_i)] - V_i(t_i) = \frac{1}{\delta_i}\,[s_i(x_i)]^{\delta_i} - d_i\, t_i^{\gamma_i},
$$ {#eq-raju-utility}

with $0 < \delta_i < 1$ (larger $\delta_i$ meaning less risk aversion) and $\gamma_i > 1$ (larger $\gamma_i$ meaning a
steeper effort cost), and $V_i(t_i) = d_i t_i^{\gamma_i}$ the convex, increasing disutility of effort. The deep
moral-hazard problem—the firm cannot observe the salesperson's true effort—is what makes all of this necessary.
@gerstner1987 and @joseph1998 extend the compensation literature, and @simester2010 highlights a further tension between
motivating managerial effort and eliciting truthful information-sharing, a trade-off that complicates the design of
incentives when agents control information as well as effort.

## Branding, Bundling, and Mixed Strategies {#sec-branding-bundling}

Three shorter analytical threads round out the firm's strategic toolkit. On **branding**, @wernerfelt1988 studies
*umbrella branding*, in which a brand name spanning multiple products signals quality across them, and @chu1994 examines
how *retailer reputation* substitutes for manufacturer signals in conveying quality to consumers.

**Bundling** is the practice of selling two goods together at a single price, and its logic turns on the heterogeneity and
*negative correlation* of valuations. Consider a firm selling equipment and installation to two equally numerous customer
types whose valuations appear in @tbl-bundling. Pricing the components separately, the firm sets the equipment price to the
lower valuation (\$5,000) and the installation price likewise (\$2,000) to sell to both types, earning \$14,000 across the
four sales. *Bundling* equipment and installation into a single package lets the firm charge each type its total valuation
(\$10,000 from type 1, \$8,000 from type 2), and since both types value the bundle at no less than \$8,000, charging \$8,000
for the bundle earns \$16,000—strictly more. Bundling works here because valuations are *negatively correlated* across the
two goods: the type that values equipment more values installation less, so the bundle compresses the dispersion of total
willingness to pay and lets the firm extract more surplus.

| | Equipment | Installation | Bundle total |
|---|---|---|---|
| Customer type 1 | \$8,000 | \$2,000 | \$10,000 |
| Customer type 2 | \$5,000 | \$3,000 | \$8,000 |

: Valuations illustrating the bundling advantage under negatively correlated willingness to pay {#tbl-bundling}

Whether to bundle therefore depends on the mix of customer types and on the correlation of valuations, which is why bundling
is not universal. *Information products* are the limiting case: with marginal cost near zero, bundling is almost costless and
the surplus-extraction logic is unimpeded, so information products are almost always bundled.

**Mixed strategies** are the equilibrium resolution of games with no pure-strategy equilibrium, and the existence guarantee
is foundational: any game with finitely many players and finitely many strategies per player has at least one Nash
equilibrium, possibly in mixed strategies. The intuition is cleanest in a tennis serve-and-return game. Suppose the server
serves left a fraction $p$ of the time. The receiver's expected payoff is $p \cdot \tfrac34 + (1-p)\cdot \tfrac14$ from
defending left and $p \cdot \tfrac14 + (1-p)\cdot \tfrac34$ from defending right, so she defends left whenever

$$
p \cdot \tfrac34 + (1-p)\cdot \tfrac14 > p \cdot \tfrac14 + (1-p)\cdot \tfrac34,
$$ {#eq-mixed-receiver}

that is, whenever $p > \tfrac12$. Symmetrically, if the receiver goes left with probability $q$, the server serves left
whenever $\tfrac14 q + \tfrac34 (1-q) > \tfrac34 q + \tfrac14 (1-q)$, i.e., whenever $q < \tfrac12$. A **mixed-strategy
equilibrium** is a pair of mixed strategies that are mutual best responses—here, each player choosing a 50–50 mixture.
The deep principle is that a player randomizes so as to make the opponent *indifferent* among the opponent's options:
each player earns the same expected payoff from every pure strategy played with positive probability, and—an important and
often-surprising property—when a player's own payoff from a pure strategy rises or falls, that player's own equilibrium
*mixture does not change*, since the mixture is pinned down by the rival's indifference, not by one's own payoffs.

## Marketing Resource-Allocation Models {#sec-resource-allocation}

The response models of @sec-market-response become decision tools when embedded in an optimization that allocates a fixed
budget across submarkets. The treatment here follows @Mantrala_1992, comparing an *informed* allocator who optimizes
against naive fixed-proportion rules across response-function shapes and investor risk attitudes.

**Concave response.** Let submarket $i$ have a concave (modified-exponential) sales response

$$
s_i = k_i\,(1 - e^{-b_i x_i}),
$$ {#eq-mantrala-response}

where $s_i$ is current-period sales response in dollars per period, $x_i$ the resource allocated to submarket $i$, $b_i$ the
rate at which sales approach saturation, and $k_i$ the sales potential. A *fixed-proportion* allocator simply splits the
budget by weights, $\hat{x}_i = w_i R$ with $\sum_t w_t = 1$ and $0 < w_t < 1$, regardless of the response parameters. An
*informed* allocator instead maximizes contribution by marginal analysis,

$$
\max\ C = m \sum_{i=1}^2 k_i\,(1 - e^{-b_i x_i}) \quad \text{s.t.}\quad x_1 + x_2 \le R,\ x_i \ge 0,
$$ {#eq-mantrala-objective}

equalizing marginal returns across submarkets. With two submarkets the optimal allocations are

$$
x_1 = \frac{1}{b_1 + b_2}\left(b_2 R + \ln\!\frac{k_1 b_1}{k_2 b_2}\right), \qquad
x_2 = \frac{1}{b_1 + b_2}\left(b_2 R + \ln\!\frac{k_2 b_2}{k_1 b_1}\right),
$$ {#eq-mantrala-optimal}

so that the submarket with higher potential or steeper saturation rate receives proportionately more, and the optimal split
shifts with the total budget $R$ rather than staying fixed—the central contrast with the proportional rule.

**S-shaped response and stochastic response.** When the response function is S-shaped, the optimal-versus-proportional
comparison again favors the informed allocator, but the threshold region of the S-curve creates the possibility of corner
solutions—pouring the entire budget into one submarket—that the concave case never produces. When response is *stochastic*,
of quadratic form, the optimal allocation depends on the investor's risk attitude, diverging between a risk-averse and a
risk-neutral allocator in a way the deterministic models cannot capture. The general lesson of @Mantrala_1992 is that the
*shape* of the response function and the *risk attitude* of the decision-maker jointly determine how far optimal allocation
departs from naive proportional rules—and that the departure is often large enough to matter for profit.

```{r}
#| label: fig-allocation
#| fig-cap: "Optimal versus fixed-proportion budget allocation across two submarkets with concave (modified-exponential) response. The informed allocator equalizes marginal returns; the proportional rule splits the budget by fixed weights and leaves contribution on the table."
#| message: false
#| warning: false
set.seed(2026)
m  <- 1          # contribution margin
k1 <- 120; b1 <- 0.010   # submarket 1: high potential, slow saturation
k2 <- 80;  b2 <- 0.030   # submarket 2: lower potential, fast saturation
R  <- 200                # total budget

contribution <- function(x1) {
  x2 <- R - x1
  m * (k1 * (1 - exp(-b1 * x1)) + k2 * (1 - exp(-b2 * x2)))
}
x1_grid <- seq(0, R, by = 1)
C_grid  <- sapply(x1_grid, contribution)
x1_star <- x1_grid[which.max(C_grid)]

plot(x1_grid, C_grid, type = "l", lwd = 2,
     xlab = "Budget to submarket 1 (x1)", ylab = "Total contribution",
     main = "")
abline(v = x1_star, lty = 2, col = "grey50")          # optimal split
abline(v = R / 2,   lty = 3, col = "red")             # naive 50/50 split
legend("bottomright", lty = c(2, 3), col = c("grey50", "red"),
       legend = c("Optimal", "Proportional (50/50)"), bty = "n")
```

## Frontiers and Further Reading {#sec-frontiers}

Several active areas extend the models above and reward further study. **Technology markets** raise questions of
compatibility and standards, where network effects make a product's value depend on its installed base—@amaldoss2005,
@balachander2009, and @sajeesh2020 supply the conspicuous-consumption and exclusivity building blocks that this literature
adapts to network goods. **Meta-analyses of econometric marketing models** synthesize estimated elasticities across studies
to recover generalizable response parameters, complementing the single-market response forms of @sec-market-response.
**Dynamic advertising** models extend the adstock and pulsing logic of @sec-26-advertising and @sec-adstock into continuous-time
optimal-control formulations of spending over the product life cycle. **Marketing-mix optimization** scales the
resource-allocation logic of @sec-resource-allocation to many instruments and channels, incorporating carryover and shape
effects in media-mix models.[^mmm] **New-product diffusion** continues to refine the @Bass_1969 and @Van_den_Bulte_2001
tradition with richer social structure and competitive entry, and **two-sided platform** models bring matching and
cross-side network effects into the analytical fold, a setting in which the pricing, advertising, and channel logic of this
chapter all reappear under the additional constraint that the platform must serve two interdependent customer groups at once.

[^mmm]: Practical implementations of carryover and shape effects in media-mix modeling, including the
geometric-adstock and Hill-saturation transforms, are widely available in open-source R and Python toolkits.

### Key Takeaways {#sec-key-takeaways}

Three threads run through the chapter. First, the *assumptions are the model*: the shape of a transport-cost function
[@daspremont1979], the observability of effort [@basu1985], or the correlation of valuations across bundled goods each
overturns a headline result, so a model's contribution is inseparable from the assumption that drives it. Second,
*information structure determines outcomes*: asymmetric information can destroy a market [@akerlof1970], be repaired by costly
signals [@spence1973; @moorthy1995], sustain equilibrium price dispersion [@varian1980], and force firms into mixed-strategy
promotions [@narasimhan1988b]. Third, *commitment and timing create strategic value*: the Stackelberg leader, the capacity-
committing incumbent [@dixit1980], the early-mover in positioning [@tabuchi1995], and the patient bargainer all profit from
the structure of moves rather than from any advantage in fundamentals. These principles, more than any single closed-form
solution, are what the analytical-modeling tradition contributes to marketing science.
