# Empirical Models {#sec-empirical-models}

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

The preceding chapters develop constructs—brand equity, customer value, advertising
response—largely from the firm's strategic vantage. This chapter turns to the
empirical machinery that a working analyst deploys once transaction logs,
clickstreams, and customer journeys arrive as data. The unifying question is one of
*credit and value assignment*: given that a customer touched many channels, passed
through several funnel stages, and bought repeatedly over time, how much of the
realized outcome should be attributed to each touchpoint, each customer, and each
moment? Answering that question well is what separates a defensible marketing-mix
or budget decision from a vanity dashboard.

The models here share a common pedigree in applied probability and game theory, and
a common hazard: each imposes structure that, if violated, silently biases the
credit it assigns. We therefore lead each method with its intuition, state its formal
object, give the estimator and the assumptions that identify it, and—just as
importantly—name the conditions under which identification fails. The chapter
proceeds through five families of empirical models. **Attribution models**
(@sec-attribution) allocate conversion credit across marketing channels using
cooperative-game and Markov-chain machinery. **Sales-funnel models**
(@sec-sales-funnel) attribute credit instead to stages of a purchase process and
visualize where prospects leak. **Recency–frequency–monetary (RFM) scoring**
(@sec-rfm) compresses transaction histories into a behavioral summary. **Customer
segmentation** (@sec-segmentation) builds on RFM to partition customers into
managerially actionable groups and to track them over their lifecycle. Finally,
**basket and sequence analysis** (@sec-basket) and **geodemographic classification**
(@sec-geodemographic) model what customers buy together, in what order, and where
they live. Every method is implemented in reproducible, seeded R.

## Attribution Models {#sec-attribution}

When a customer converts after a sequence of exposures—a display ad, then a paid
search click, then an email, then a direct visit—the firm observes a single
outcome but many candidate causes. **Attribution** is the problem of distributing
the credit for that outcome across the touchpoints that preceded it. The stakes are
budgetary: channel spend is reallocated in proportion to attributed credit, so a
biased attribution rule mechanically misallocates money.

The naïve solutions are *heuristic* rules that assign all credit to a single
position in the path. **First-touch** attribution credits the channel that
initiated the journey; **last-touch** attribution credits the channel that
immediately preceded conversion; **linear-touch** attribution splits credit equally
across all touchpoints. These rules are transparent and require no estimation, but
each encodes an indefensible causal assumption—that position in the sequence is a
sufficient statistic for contribution. Last-touch attribution, the long-standing
industry default, systematically over-credits lower-funnel channels (branded search,
retargeting) that harvest demand other channels created. The two model-based
alternatives developed below—the Shapley value and the Markov removal effect—replace
positional heuristics with a principled notion of marginal contribution.
@fig-attribution-taxonomy organizes these two families, contrasting the positional
heuristics with the model-based rules that estimate marginal contribution.

```{mermaid}
%%| label: fig-attribution-taxonomy
%%| fig-cap: "Two families of attribution rule. Heuristic rules assign credit by position in the path; model-based rules assign credit by marginal contribution, estimated either from cooperative game theory (Shapley) or from a Markov model of the journey (removal effect)."
flowchart TD
    A[Conversion path:<br/>ordered touchpoints] --> B{Attribution rule}
    B --> C[Heuristic]
    B --> D[Model-based]
    C --> C1[First touch]
    C --> C2[Last touch]
    C --> C3[Linear touch]
    D --> D1[Shapley value<br/>cooperative game theory]
    D --> D2[Markov removal effect<br/>absorbing chain]
    D1 --> E[Channel credit allocation]
    D2 --> E
    C1 --> E
    C2 --> E
    C3 --> E
```

### The Shapley Value {#sec-shapley}

The Shapley value imports a solution concept from cooperative game theory into
attribution [@Zhao_2018]. Treat each marketing channel as a *player* and a
conversion as the *payoff* of a coalition of channels acting together. The question
"how much credit does channel $i$ deserve?" becomes "what is channel $i$'s average
marginal contribution to the payoff, taken over every order in which the channels
could have joined the coalition?"

Formally, let $N = \{1, \dots, n\}$ be the set of channels and let
$v: 2^{N} \to \mathbb{R}$ be a *characteristic function* assigning to each coalition
$S \subseteq N$ the conversions (or conversion value) realized when exactly the
channels in $S$ are present. The marginal contribution of channel $i$ to a coalition
$S$ that does not contain it is $v(S \cup \{i\}) - v(S)$. The **Shapley value**
averages this marginal contribution over all $n!$ orderings of the players:

$$
\phi_i(v) = \sum_{S \subseteq N \setminus \{i\}}
\frac{|S|!\,\bigl(n - |S| - 1\bigr)!}{n!}
\bigl[v(S \cup \{i\}) - v(S)\bigr].
$$ {#eq-shapley}

The combinatorial weight in @eq-shapley counts the permutations consistent with
channel $i$ joining immediately after coalition $S$ has formed. The Shapley value is
the *unique* allocation satisfying four axioms that any fair attribution scheme
should arguably obey: **efficiency** (the credits sum to the total payoff,
$\sum_i \phi_i = v(N)$), **symmetry** (two channels making identical marginal
contributions receive equal credit), the **null-player** property (a channel that
never changes any coalition's payoff receives zero), and **additivity** (credit is
linear in the payoff across independent games). This axiomatic uniqueness is what
distinguishes the Shapley value from heuristic rules: it is not one allocation among
many but the only one consistent with these fairness requirements.

Two channel-specific features of the marketing problem deserve emphasis. First, the
characteristic function encodes the *marginal contribution* of each channel.
Second, the averaging over orderings encodes the *sequence in which channels join
the coalition*—the empirical analogue of the customer journey. The ordered Shapley
extension of @Zhao_2018 preserves information about the order of touchpoints rather
than treating the path as an unordered set, which matters when the same channels in
different orders convert at different rates.

::: {.callout-note}
## When the Shapley value breaks down

Three failure modes recur in practice. **Computational burden**: @eq-shapley sums
over $2^{n}$ coalitions and implicitly over $n!$ orderings, so exact computation is
infeasible beyond a couple dozen channels and must be approximated by Monte Carlo
sampling of permutations. **Coalitional completeness**: efficiency forces all credit
to be divided among the *observed* channels; if a relevant channel is unmeasured
(offline word of mouth, a competitor's action), its contribution is silently
redistributed onto the measured channels, biasing every estimate. **No ground
truth**: because the conversion is observed only once per customer, there is no
held-out outcome against which to validate the allocation, so model comparison
across attribution schemes is inherently difficult and rests on face validity rather
than predictive accuracy.
:::

The `GameTheory` package provides the coalitional-game primitives—characteristic
functions and Shapley-value solvers—on which an attribution pipeline is built.

```{r}
library("GameTheory")
```

In a production setting the characteristic function $v(S)$ is estimated from the log
of observed paths: for each coalition $S$, one tallies conversions among customers
whose journeys used exactly (or at least) the channels in $S$. Sparse coalitions—
combinations of channels that few customers ever experience—are the practical limit
on reliability, since $v(S)$ for an unobserved $S$ must be imputed or smoothed.

### Markov-Chain Attribution {#sec-markov-attribution}

The Markov approach reframes the customer journey as a random walk over a graph
whose nodes are channels and whose edges carry transition probabilities. Conversion
and non-conversion are *absorbing* states; once a journey reaches either, it stops.
Channel importance is then measured by how much the conversion probability *falls*
when a channel is deleted from the graph—the **removal effect**. The appeal is that
this construction yields a single internally consistent model of the journey from
which credit, expected duration, and counterfactual scenarios all follow.

A Markov chain is specified by three objects: a **state space**, the set of all
states the process can occupy (here, the channels plus the absorbing conversion and
null states); a **transition operator**, the probabilities of moving from each state
to each other state; and a **current-state distribution**, the probability of
occupying each state at the start. The defining restriction is the **Markov
property**: the next state depends only on the current state, not on the full
history. The transition probabilities satisfy

$$
w_{ij} = P\bigl(X_t = s_j \mid X_{t-1} = s_i\bigr),
\qquad 0 \le w_{ij} \le 1,
\qquad \sum_{j=1}^{N} w_{ij} = 1 \;\; \forall i.
$$ {#eq-markov-transition}

The three conditions in @eq-markov-transition state, respectively, that $w_{ij}$ is
the probability of moving to state $s_j$ given current state $s_i$; that each such
probability is a genuine probability bounded in $[0,1]$; and that the rows of the
transition matrix sum to one, because from any state the process must move
*somewhere* (including possibly back to itself or into an absorbing state).

The credit a channel earns is its **removal effect**. Let $\pi$ denote the
probability that a random walk starting from the `(start)` node is eventually
absorbed in `(conversion)` under the full transition matrix. Deleting channel $i$
means rerouting every path that would have passed through $i$ directly to the
`(null)` absorbing state; let $\pi_{-i}$ be the conversion probability of the
mutilated chain. The removal effect of channel $i$ is the relative drop in
conversion probability,

$$
\text{RE}_i = \frac{\pi - \pi_{-i}}{\pi},
$$ {#eq-removal-effect}

and attributed conversions are obtained by normalizing the removal effects to sum to
one and multiplying by total conversions. The removal effect is therefore a
counterfactual: it answers "what fraction of conversions would we lose if this
channel disappeared?"

#### The Order of the Chain {#sec-markov-order}

The chain's **order** controls how much history conditions the next transition, and
the choice trades bias against variance.

| Order | Memory | Conditioning set for the next state |
|------:|:-------|:------------------------------------|
| 0 | none | unconditional probability of entering any state |
| 1 | current state only | the state you are in now (*memory-free*) |
| 2 | one prior state | where you came from and where you are |
| 3 | two prior states | the last two states and where you are |
| 4 | three prior states | the last three states and where you are |

: Markov-chain order and the history it conditions on. A first-order ("memory-free")
chain is the workhorse: the probability of the next state depends only on the
current state. {#tbl-markov-order}

A first-order chain is *memory-free*: the probability of reaching the next state
depends only on the state currently occupied, as @tbl-markov-order records. Higher
orders capture sequence dependence—the same channel may convert differently
depending on what preceded it—at the cost of a combinatorial explosion in the number
of estimated transition probabilities and corresponding data sparsity. As we
demonstrate below, in the *first*-order chain consecutive duplicate channels leave
the transition matrix and hence the attribution unchanged, which justifies the common
practice of collapsing duplicates; this invariance no longer holds at second order
and above.

The terminology recurs throughout: each node is a **transition state**, the
probability of moving between two channels is a **transition probability**, and the
counterfactual deletion of a node yields its removal effect.

#### Worked Example: Heuristics versus Markov {#sec-markov-example-1}

The first worked example contrasts the heuristic rules with a first-order Markov
model on a channel-attribution dataset, following the implementation popularized by
Analytics Vidhya. The raw data encode each customer's path as a sequence of channel
identifiers, where codes 1–19 are marketing channels, 20 marks device selection, 21
marks the final purchase, and 22 marks an undecided customer.

```{r}
#Load the libraries
library("ChannelAttribution")
library("ggplot2")
library("reshape")
library("dplyr")
library("plyr")
library("reshape2")
library("plotly")

#Read the data into R
channel = read.csv("images/Channel_attribution.csv", header = T) %>%
    select(-c(Output))
head(channel, n = 2)
```

The preprocessing step reconstructs a single delimited path string per customer and
marks conversion wherever the purchase code 21 appears, then truncates each path at
the point of purchase so that post-conversion states do not leak credit.

```{r}
for (row in 1:nrow(channel)){
    if (21 %in% channel[row,]){
        channel$convert = 1
    }
}

column = colnames(channel)
channel$path = do.call(paste, c(channel, sep = " > "))
head(channel$path)
```

```{r}
for(row in 1:nrow(channel)){
  channel$path[row] = strsplit(channel$path[row], " > 21")[[1]][1]
}
channel_fin = channel[,c(22,21)]
channel_fin = ddply(channel_fin,~path,summarise, conversion= sum(convert))
head(channel_fin)

Data = channel_fin
head(Data)
```

The heuristic models assign credit by position. First-touch credits the path's
initiator, last-touch credits the closer, and linear-touch divides credit equally—
the three positional baselines against which the Markov model is judged.

```{r}
H <- heuristic_models(Data, 'path', 'conversion', var_value='conversion')
H
```

The first-order Markov model estimates the transition matrix from the observed paths
and computes removal effects via @eq-removal-effect.

```{r}
M <- markov_model(Data, 'path', 'conversion', var_value='conversion', order = 1)
M
```

Merging the heuristic and Markov outputs on the channel name aligns the four
allocations for comparison.

```{r}
# Merges the two data frames on the "channel_name" column.
R <- merge(H, M, by='channel_name')

# Select only relevant columns
R1 <- R[, (colnames(R) %in% c('channel_name', 'first_touch_conversions', 'last_touch_conversions', 'linear_touch_conversions', 'total_conversion'))]

# Transforms the dataset into a data frame that ggplot2 can use to plot the outcomes
R1 <- melt(R1, id='channel_name')
```

```{r}
#| label: fig-markov-heuristic-conversions
#| fig-cap: "Total conversions attributed to each channel under first-touch, last-touch, linear-touch, and first-order Markov rules. Divergence between the Markov bar and the positional bars localizes channels that heuristics systematically mis-credit."
# Plot the total conversions
ggplot(R1, aes(channel_name, value, fill = variable)) +
  geom_bar(stat='identity', position='dodge') +
  ggtitle('TOTAL CONVERSIONS') +
  theme(axis.title.x = element_text(vjust = -2)) +
  theme(axis.title.y = element_text(vjust = +2)) +
  theme(title = element_text(size = 16)) +
  theme(plot.title=element_text(size = 20)) +
  ylab("")
```

Where the Markov bar in @fig-markov-heuristic-conversions stands well above a
channel's last-touch bar, that channel is an *assist* whose contribution last-touch
attribution would discard; where it stands below, the channel is a *closer* that
last-touch attribution over-credits.

#### Worked Example: Transition Matrix and Model Comparison {#sec-markov-example-2}

The second example, following Sergey Bryl', simulates a realistic journey log and
exposes the transition matrix directly, along with the `(start)`, `(conversion)`,
and `(null)` bookkeeping states that make the chain absorbing. The first block builds
a minimal three-channel example to illustrate the data structure the
`markov_model()` function expects—a path string, a conversion flag, and a
non-conversion flag—and to extract the raw transition matrix and the dummy absorbing
states needed to plot the journey graph.

```{r, eval=FALSE}
library(dplyr)
library(reshape2)
library(ggplot2)
library(ggthemes)
library(ggrepel)
library(RColorBrewer)
library(ChannelAttribution)
library(markovchain)

##### simple example #####
# creating a data sample
df1 <- data.frame(path = c('c1 > c2 > c3', 'c1', 'c2 > c3'), conv = c(1, 0, 0), conv_null = c(0, 1, 1))

# calculating the model
mod1 <- markov_model(df1,
                    var_path = 'path',
                    var_conv = 'conv',
                    var_null = 'conv_null',
                    out_more = TRUE)

# extracting the results of attribution
df_res1 <- mod1$result

# extracting a transition matrix
df_trans1 <- mod1$transition_matrix
df_trans1 <- dcast(df_trans1, channel_from ~ channel_to, value.var = 'transition_probability')

### plotting the Markov graph ###
df_trans <- mod1$transition_matrix

# adding dummies in order to plot the graph
df_dummy <- data.frame(channel_from = c('(start)', '(conversion)', '(null)'),
                       channel_to = c('(start)', '(conversion)', '(null)'),
                       transition_probability = c(0, 1, 1))
df_trans <- rbind(df_trans, df_dummy)

# ordering channels
df_trans$channel_from <- factor(df_trans$channel_from,levels = c('(start)','(conversion)', '(null)', 'c1', 'c2', 'c3'))
df_trans$channel_to <- factor(df_trans$channel_to,levels = c('(start)', '(conversion)', '(null)', 'c1', 'c2', 'c3'))
df_trans <- dcast(df_trans, channel_from ~ channel_to, value.var ='transition_probability')

# creating the markovchain object
trans_matrix <- matrix(data = as.matrix(df_trans[, -1]),nrow = nrow(df_trans[, -1]), ncol = ncol(df_trans[, -1]),dimnames = list(c(as.character(df_trans[,1])),c(colnames(df_trans[, -1]))))
trans_matrix[is.na(trans_matrix)] <- 0
```

The next block simulates 5,000 touchpoints across ten channels for up to 1,000
customers, aggregates them into per-customer paths, and fits the Markov model
alongside manually computed first- and last-touch baselines.

```{r}
# simulating the "real" data
set.seed(354)
df2 <- data.frame(client_id = sample(c(1:1000), 5000, replace = TRUE),
                  date = sample(c(1:32), 5000, replace = TRUE),
                  channel = sample(c(0:9), 5000, replace = TRUE,
                                   prob = c(0.1, 0.15, 0.05, 0.07, 0.11, 0.07, 0.13, 0.1, 0.06, 0.16)))
df2$date <- as.Date(df2$date, origin = "2015-01-01")
df2$channel <- paste0('channel_', df2$channel)

# aggregating channels to the paths for each customer
df2 <- df2 %>%
        arrange(client_id, date) %>%
        group_by(client_id) %>%
        summarise(path = paste(channel, collapse = ' > '),
                  # assume that all paths were finished with conversion
                  conv = 1,
                  conv_null = 0) %>%
        ungroup()

# calculating the models (Markov and heuristics)
mod2 <- markov_model(df2,
                     var_path = 'path',
                     var_conv = 'conv',
                     var_null = 'conv_null',
                     out_more = TRUE)

# heuristic_models() function doesn't work for me, therefore I used the manual calculations
# instead of:
#h_mod2 <- heuristic_models(df2, var_path = 'path', var_conv = 'conv')

df_hm <- df2 %>%
        mutate(channel_name_ft = sub('>.*', '', path),
               channel_name_ft = sub(' ', '', channel_name_ft),
               channel_name_lt = sub('.*>', '', path),
               channel_name_lt = sub(' ', '', channel_name_lt))
# first-touch conversions
df_ft <- df_hm %>%
        group_by(channel_name_ft) %>%
        summarise(first_touch_conversions = sum(conv)) %>%
        ungroup()
# last-touch conversions
df_lt <- df_hm %>%
        group_by(channel_name_lt) %>%
        summarise(last_touch_conversions = sum(conv)) %>%
        ungroup()

h_mod2 <- merge(df_ft, df_lt, by.x = 'channel_name_ft', by.y = 'channel_name_lt')

# merging all models
all_models <- merge(h_mod2, mod2$result, by.x = 'channel_name_ft', by.y = 'channel_name')
colnames(all_models)[c(1, 4)] <- c('channel_name', 'attrib_model_conversions')
```

The transition-matrix heatmap renders the estimated $w_{ij}$ as a colored grid,
making the dominant journey paths and the channels that most often precede
conversion visually legible.

```{r}
#| label: fig-transition-heatmap
#| fig-cap: "Estimated first-order transition matrix as a heatmap. Each cell is the probability of moving from the row channel to the column channel; bright cells trace the most-traveled edges of the journey graph."
library("RColorBrewer")
library("ggthemes")
library("ggrepel")
############## visualizations ##############
# transition matrix heatmap for "real" data
df_plot_trans <- mod2$transition_matrix

cols <- c("#e7f0fa", "#c9e2f6", "#95cbee", "#0099dc", "#4ab04a", "#ffd73e", "#eec73a",
          "#e29421", "#e29421", "#f05336", "#ce472e")
t <- max(df_plot_trans$transition_probability)

ggplot(df_plot_trans, aes(y = channel_from, x = channel_to, fill = transition_probability)) +
        theme_minimal() +
        geom_tile(colour = "white", width = .9, height = .9) +
        scale_fill_gradientn(colours = cols, limits = c(0, t),
                             breaks = seq(0, t, by = t/4),
                             labels = c("0", round(t/4*1, 2), round(t/4*2, 2), round(t/4*3, 2), round(t/4*4, 2)),
                             guide = guide_colourbar(ticks = T, nbin = 50, barheight = .5, label = T, barwidth = 10)) +
        geom_text(aes(label = round(transition_probability, 2)), fontface = "bold", size = 4) +
        theme(legend.position = 'bottom',
              legend.direction = "horizontal",
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              plot.title = element_text(size = 20, face = "bold", vjust = 2, color = 'black', lineheight = 0.8),
              axis.title.x = element_text(size = 24, face = "bold"),
              axis.title.y = element_text(size = 24, face = "bold"),
              axis.text.y = element_text(size = 8, face = "bold", color = 'black'),
              axis.text.x = element_text(size = 8, angle = 90, hjust = 0.5, vjust = 0.5, face = "plain")) +
        ggtitle("Transition matrix heatmap")

# models comparison
all_mod_plot <- reshape2::melt(all_models, id.vars = 'channel_name', variable.name = 'conv_type')
all_mod_plot$value <- round(all_mod_plot$value)
# slope chart
pal <- colorRampPalette(brewer.pal(10, "Set1"))
ggplot(all_mod_plot, aes(x = conv_type, y = value, group = channel_name)) +
        theme_solarized(base_size = 18, base_family = "", light = TRUE) +
        scale_color_manual(values = pal(10)) +
        scale_fill_manual(values = pal(10)) +
        geom_line(aes(color = channel_name), size = 2.5, alpha = 0.8) +
        geom_point(aes(color = channel_name), size = 5) +
        geom_label_repel(aes(label = paste0(channel_name, ': ', value), fill = factor(channel_name)),
                         alpha = 0.7,
                         fontface = 'bold', color = 'white', size = 5,
                         box.padding = unit(0.25, 'lines'), point.padding = unit(0.5, 'lines'),
                         max.iter = 100) +
        theme(legend.position = 'none',
              legend.title = element_text(size = 16, color = 'black'),
              legend.text = element_text(size = 16, vjust = 2, color = 'black'),
              plot.title = element_text(size = 20, face = "bold", vjust = 2, color = 'black', lineheight = 0.8),
              axis.title.x = element_text(size = 24, face = "bold"),
              axis.title.y = element_text(size = 16, face = "bold"),
              axis.text.x = element_text(size = 16, face = "bold", color = 'black'),
              axis.text.y = element_blank(),
              axis.ticks.x = element_blank(),
              axis.ticks.y = element_blank(),
              panel.border = element_blank(),
              panel.grid.major = element_line(colour = "grey", linetype = "dotted"),
              panel.grid.minor = element_blank(),
              strip.text = element_text(size = 16, hjust = 0.5, vjust = 0.5, face = "bold", color = 'black'),
              strip.background = element_rect(fill = "#f0b35f")) +
        labs(x = 'Model', y = 'Conversions') +
        ggtitle('Models comparison') +
        guides(colour = guide_legend(override.aes = list(size = 4)))
```

The slope chart above connects each channel's attributed
conversions across the four rules; channels whose lines cross steeply are precisely
those whose ranking depends on the attribution method, and therefore those for which
the choice of model has real budgetary consequences.

#### Practical Refinements to Markov Attribution {#sec-markov-refinements}

A production attribution pipeline must confront several complications that the
textbook chain ignores. The following refinements, again following Bryl', address
journey staging, missing data, single-channel paths, chain order, non-converting
paths, and journey duration. We begin by simulating a richer log of 80,000
touchpoints across eight channels for 20,000 customers, with conversions sprinkled
stochastically and one channel deliberately set to missing to exercise the
missing-data logic.

```{r}
library(tidyverse)
library(reshape2)
library(ggthemes)
library(ggrepel)
library(RColorBrewer)
library(ChannelAttribution)
library(visNetwork)
library(expm)
library(stringr)
library(purrr)
library(purrrlyr)

##### simulating the "real" data #####
set.seed(454)
df_raw <- data.frame(customer_id = paste0('id', sample(c(1:20000), replace = TRUE)), date = as.Date(rbeta(80000, 0.7, 10) * 100, origin = "2016-01-01"), channel = paste0('channel_', sample(c(0:7), 80000, replace = TRUE, prob = c(0.2, 0.12, 0.03, 0.07, 0.15, 0.25, 0.1, 0.08))) ) %>%
        group_by(customer_id) %>%
        mutate(conversion = sample(c(0, 1), n(), prob = c(0.975, 0.025), replace = TRUE)) %>%
        ungroup() %>%
        dmap_at(c(1, 3), as.character) %>%
        arrange(customer_id, date)

df_raw <- df_raw %>%
        mutate(channel = ifelse(channel == 'channel_2', NA, channel))
head(df_raw, n = 2)
```

**Journey staging.** A customer occupies a different position in the funnel after
each conversion: a first-time buyer's path looks nothing like a tenth-time buyer's,
who will not, for instance, begin at the landing page. The analyst therefore splits
each customer's history into separate *paths* indexed by the cumulative number of
prior conversions, and may model each stage separately. The serial path number is the
lagged cumulative sum of the conversion flag.

```{r}
##### splitting paths #####
df_paths <- df_raw %>%
        group_by(customer_id) %>%
        mutate(path_no = ifelse(is.na(lag(cumsum(conversion))), 0, lag(cumsum(conversion))) + 1) %>% # add the path's serial number by using the lagged cumulative sum of conversion binary marks
        ungroup()
head(df_paths)
```

```{r}
df_paths_1 <- df_paths %>%
        filter(path_no == 1) %>%
        select(-path_no)
```

**Missing and non-attributable channels.** Channels may be unrecorded, or the firm
may choose not to attribute certain touchpoints (a direct visit, say, that merely
reflects existing intent). Two remedies are available: drop the offending touchpoint,
or impute it with the previous channel in the path. A useful invariance justifies the
casual treatment of duplicates here: *in a first-order chain, consecutive duplicate
channels do not change the transition matrix*, so substituting a previous channel for
a missing one leaves first-order attribution unaffected. The code removes missing
values, then replaces a designated direct channel (`channel_6`) with the preceding
touchpoint except when it opens the path, in which case it is dropped.

```{r}
##### replace some channels #####
df_path_1_clean <- df_paths_1 %>%
        # removing NAs
        filter(!is.na(channel)) %>%

        # adding order of channels in the path
        group_by(customer_id) %>%
        mutate(ord = c(1:n()),
               is_non_direct = ifelse(channel == 'channel_6', 0, 1),
               is_non_direct_cum = cumsum(is_non_direct)) %>%

        # removing Direct (channel_6) when it is the first in the path
        filter(is_non_direct_cum != 0) %>%

        # replacing Direct (channel_6) with the previous touch point
        mutate(channel = ifelse(channel == 'channel_6', channel[which(channel != 'channel_6')][is_non_direct_cum], channel)) %>%

        ungroup() %>%
        select(-ord, -is_non_direct, -is_non_direct_cum)
```

**Single-channel versus multi-channel paths.** Because the removal effects do not
sum to one, the package reweights each channel's importance by the total. This
creates a subtle bias: for a path consisting of a single unique channel, that
channel's removal effect and importance are exactly one, but pooling such paths with
multi-channel paths and reweighting *dilutes* the single-channel credit, so channels
that frequently appear alone are underestimated. The logic of the fix is direct—when
a path contains only one channel we know with certainty which channel earned the
conversion, and that certainty should not be redistributed. The procedure is
therefore to (i) split the data into one-channel and multi-channel paths, (ii) count
conversions directly for one-channel paths while fitting the Markov model only on
multi-channel paths, and (iii) sum the two contributions per channel.

```{r}
##### one- and multi-channel paths #####
df_path_1_clean <- df_path_1_clean %>%
        group_by(customer_id) %>%
        mutate(uniq_channel_tag = ifelse(length(unique(channel)) == 1, TRUE, FALSE)) %>%
        ungroup()

df_path_1_clean_uniq <- df_path_1_clean %>%
        filter(uniq_channel_tag == TRUE) %>%
        select(-uniq_channel_tag)

df_path_1_clean_multi <- df_path_1_clean %>%
        filter(uniq_channel_tag == FALSE) %>%
        select(-uniq_channel_tag)

### experiment ###
# attribution model for all paths
df_all_paths <- df_path_1_clean %>%
        group_by(customer_id) %>%
        summarise(path = paste(channel, collapse = ' > '),
                  conversion = sum(conversion)) %>%
        ungroup() %>%
        filter(conversion == 1)

mod_attrib <- markov_model(df_all_paths,
                           var_path = 'path',
                           var_conv = 'conversion',
                           out_more = TRUE)
mod_attrib$removal_effects
mod_attrib$result
d_all <- data.frame(mod_attrib$result)

# attribution model for splitted multi and unique channel paths
df_multi_paths <- df_path_1_clean_multi %>%
        group_by(customer_id) %>%
        summarise(path = paste(channel, collapse = ' > '),
                  conversion = sum(conversion)) %>%
        ungroup() %>%
        filter(conversion == 1)

mod_attrib_alt <- markov_model(df_multi_paths,
                           var_path = 'path',
                           var_conv = 'conversion',
                           out_more = TRUE)
mod_attrib_alt$removal_effects
mod_attrib_alt$result

# adding unique paths
df_uniq_paths <- df_path_1_clean_uniq %>%
        filter(conversion == 1) %>%
        group_by(channel) %>%
        summarise(conversions = sum(conversion)) %>%
        ungroup()

d_multi <- data.frame(mod_attrib_alt$result)

d_split <- full_join(d_multi, df_uniq_paths, by = c('channel_name' = 'channel')) %>%
        mutate(result = total_conversions + conversions)

sum(d_all$total_conversions)
sum(d_split$result)
```

**Higher-order chains and duplicate channels.** The first-order invariance noted
above fails at second order: once the next transition conditions on two prior states,
a duplicated channel changes the conditioning history and hence the estimated
probabilities. To verify that the practical effect on first-order attribution is nil,
one can compute the transition matrix and removal effects "manually"—the package
collapses duplicates automatically, so a hand-rolled calculation is needed to compare.
The script below builds the transition matrix by tabulating adjacent channel pairs,
then computes attribution by iterating the mutilated transition matrix to its
stationary behavior via the matrix power `%^%`.

```{r, eval = FALSE}
##### Higher order of Markov chains and consequent duplicated channels in the path #####

# computing transition matrix - 'manual' way
df_multi_paths_m <- df_multi_paths %>%
        mutate(path = paste0('(start) > ', path, ' > (conversion)'))
m <- max(str_count(df_multi_paths_m$path, '>')) + 1 # maximum path length

df_multi_paths_cols <- reshape2::colsplit(string = df_multi_paths_m$path, pattern = ' > ', names = c(1:m))
colnames(df_multi_paths_cols) <- paste0('ord_', c(1:m))
df_multi_paths_cols[df_multi_paths_cols == ''] <- NA

df_res <- vector('list', ncol(df_multi_paths_cols) - 1)

for (i in c(1:(ncol(df_multi_paths_cols) - 1))) {

        df_cache <- df_multi_paths_cols %>%
                select(num_range("ord_", c(i, i+1))) %>%
                na.omit() %>%
                group_by_(.dot = c(paste0("ord_", c(i, i+1)))) %>%
                summarise(n = n()) %>%
                ungroup()

        colnames(df_cache)[c(1, 2)] <- c('channel_from', 'channel_to')
        df_res[[i]] <- df_cache
}

df_res <- do.call('rbind', df_res)

df_res_tot <- df_res %>%
        group_by(channel_from, channel_to) %>%
        summarise(n = sum(n)) %>%
        ungroup() %>%
        group_by(channel_from) %>%
        mutate(tot_n = sum(n),
               perc = n / tot_n) %>%
        ungroup()

df_dummy <- data.frame(channel_from = c('(start)', '(conversion)', '(null)'),
                       channel_to = c('(start)', '(conversion)', '(null)'),
                       n = c(0, 0, 0),
                       tot_n = c(0, 0, 0),
                       perc = c(0, 1, 1))

df_res_tot <- rbind(df_res_tot, df_dummy)

# comparing transition matrices
trans_matrix_prob_m <- dcast(df_res_tot, channel_from ~ channel_to, value.var = 'perc', fun.aggregate = sum)
trans_matrix_prob <- data.frame(mod_attrib_alt$transition_matrix)
trans_matrix_prob <- dcast(trans_matrix_prob, channel_from ~ channel_to, value.var = 'transition_probability')

# computing attribution - 'manual' way
channels_list <- df_path_1_clean_multi %>%
        filter(conversion == 1) %>%
        distinct(channel)
channels_list <- c(channels_list$channel)

df_res_ini <- df_res_tot %>% select(channel_from, channel_to)
df_attrib <- vector('list', length(channels_list))

for (i in c(1:length(channels_list))) {

        channel <- channels_list[i]

        df_res1 <- df_res %>%
                mutate(channel_from = ifelse(channel_from == channel, NA, channel_from),
                       channel_to = ifelse(channel_to == channel, '(null)', channel_to)) %>%
                na.omit()

        df_res_tot1 <- df_res1 %>%
                group_by(channel_from, channel_to) %>%
                summarise(n = sum(n)) %>%
                ungroup() %>%

                group_by(channel_from) %>%
                mutate(tot_n = sum(n),
                       perc = n / tot_n) %>%
                ungroup()

        df_res_tot1 <- rbind(df_res_tot1, df_dummy) # adding (start), (conversion) and (null) states

        df_res_tot1 <- left_join(df_res_ini, df_res_tot1, by = c('channel_from', 'channel_to'))
        df_res_tot1[is.na(df_res_tot1)] <- 0

        df_trans1 <- dcast(df_res_tot1, channel_from ~ channel_to, value.var = 'perc', fun.aggregate = sum)

        trans_matrix_1 <- df_trans1
        rownames(trans_matrix_1) <- trans_matrix_1$channel_from
        trans_matrix_1 <- as.matrix(trans_matrix_1[, -1])

        inist_n1 <- dcast(df_res_tot1, channel_from ~ channel_to, value.var = 'n', fun.aggregate = sum)
        rownames(inist_n1) <- inist_n1$channel_from
        inist_n1 <- as.matrix(inist_n1[, -1])
        inist_n1[is.na(inist_n1)] <- 0
        inist_n1 <- inist_n1['(start)', ]

        res_num1 <- inist_n1 %*% (trans_matrix_1 %^% 100000)

        df_cache <- data.frame(channel_name = channel,
                               conversions = as.numeric(res_num1[1, 1]))

        df_attrib[[i]] <- df_cache
}

df_attrib <- do.call('rbind', df_attrib)

# computing removal effect and results
tot_conv <- sum(df_multi_paths_m$conversion)

df_attrib <- df_attrib %>%
        mutate(tot_conversions = sum(df_multi_paths_m$conversion),
               impact = (tot_conversions - conversions) / tot_conversions,
               tot_impact = sum(impact),
               weighted_impact = impact / tot_impact,
               attrib_model_conversions = round(tot_conversions * weighted_impact)
        ) %>%
        select(channel_name, attrib_model_conversions)
```

Because the removal effects and attribution results coincide whether or not
duplicates are collapsed in the first-order chain, the package's default of skipping
duplicates is innocuous, and we adopt it in practice.

**Non-converting paths.** A complete model must include the paths that *failed* to
convert, since they are exactly what the `(null)` absorbing state represents. Adding a
null-conversion flag and refitting yields the *generic probabilistic model*, whose
transition matrix can be visualized as an interactive network graph in which green
nodes are the start and conversion states, the red node is the null state, and yellow
nodes are channels.

```{r}
##### Generic Probabilistic Model #####
df_all_paths_compl <- df_path_1_clean %>%
        group_by(customer_id) %>%
        summarise(path = paste(channel, collapse = ' > '),
                  conversion = sum(conversion)) %>%
        ungroup() %>%
        mutate(null_conversion = ifelse(conversion == 1, 0, 1))

mod_attrib_complete <- markov_model(
        df_all_paths_compl,
        var_path = 'path',
        var_conv = 'conversion',
        var_null = 'null_conversion',
        out_more = TRUE
)

trans_matrix_prob <- mod_attrib_complete$transition_matrix %>%
        dmap_at(c(1, 2), as.character)

##### viz #####
edges <-
        data.frame(
                from = trans_matrix_prob$channel_from,
                to = trans_matrix_prob$channel_to,
                label = round(trans_matrix_prob$transition_probability, 2),
                font.size = trans_matrix_prob$transition_probability * 100,
                width = trans_matrix_prob$transition_probability * 15,
                shadow = TRUE,
                arrows = "to",
                color = list(color = "#95cbee", highlight = "red")
        )

nodes <- data_frame(id = c( c(trans_matrix_prob$channel_from), c(trans_matrix_prob$channel_to) )) %>%
        distinct(id) %>%
        arrange(id) %>%
        mutate(
                label = id,
                color = ifelse(
                        label %in% c('(start)', '(conversion)'),
                        '#4ab04a',
                        ifelse(label == '(null)', '#ce472e', '#ffd73e')
                ),
                shadow = TRUE,
                shape = "box"
        )


# Interactive htmlwidget: HTML output only. In PDF/EPUB, webshot2 cannot rasterize
# a percentage-width widget, so render it only when the output format is HTML.
if (knitr::pandoc_to("html")) {
  visNetwork(nodes,
             edges,
             height = "2000px",
             width = "100%",
             main = "Generic Probabilistic model's Transition Matrix") %>%
          visIgraphLayout(randomSeed = 123) %>%
          visNodes(size = 5) %>%
          visOptions(highlightNearest = TRUE)
}
```

The transition matrix can also be raised to successive powers to *simulate forward*
the distribution of customers across states after a given number of steps—a
counterfactual that answers, for instance, where 1,000 customers seeded at one
channel will have dispersed after five steps versus at the stationary limit.

```{r eval = FALSE}
##### modeling states and conversions #####
# transition matrix preprocessing
trans_matrix_complete <- mod_attrib_complete$transition_matrix
trans_matrix_complete <- rbind(trans_matrix_complete, df_dummy %>%
                                       mutate(transition_probability = perc) %>%
                                       select(channel_from, channel_to, transition_probability))
trans_matrix_complete$channel_to <- factor(trans_matrix_complete$channel_to, levels = c(levels(trans_matrix_complete$channel_from)))
trans_matrix_complete <- dcast(trans_matrix_complete, channel_from ~ channel_to, value.var = 'transition_probability')
trans_matrix_complete[is.na(trans_matrix_complete)] <- 0
rownames(trans_matrix_complete) <- trans_matrix_complete$channel_from
trans_matrix_complete <- as.matrix(trans_matrix_complete[, -1])


# creating empty matrix for modeling
model_mtrx <- matrix(data = 0,
                     nrow = nrow(trans_matrix_complete), ncol = 1,
                     dimnames = list(c(rownames(trans_matrix_complete)), '(start)'))
# adding modeling number of visits
model_mtrx['channel_5', ] <- 1000

c(model_mtrx) %*% (trans_matrix_complete %^% 5) # after 5 steps
c(model_mtrx) %*% (trans_matrix_complete %^% 100000) # after 100000 steps
```

**Customer-journey duration.** The model is silent on *time*, but the firm needs a
rule for declaring a path dead. Computing the elapsed time from first touch to
conversion, and the gap since the most recent touch, yields empirical distributions
whose upper quantiles define a churn cutoff. The histograms and empirical cumulative
distribution functions below locate the 95th percentile of each duration.

```{r}
##### Customer journey duration #####
# computing time lapses from the first contact to conversion/last contact
df_multi_paths_tl <- df_path_1_clean_multi %>%
        group_by(customer_id) %>%
        summarise(path = paste(channel, collapse = ' > '),
                  first_touch_date = min(date),
                  last_touch_date = max(date),
                  tot_time_lapse = round(as.numeric(last_touch_date - first_touch_date)),
                  conversion = sum(conversion)) %>%
        ungroup()

# distribution plot
ggplot(df_multi_paths_tl %>% filter(conversion == 1), aes(x = tot_time_lapse)) +
        theme_minimal() +
        geom_histogram(fill = '#4e79a7', binwidth = 1)

# cumulative distribution plot
ggplot(df_multi_paths_tl %>% filter(conversion == 1), aes(x = tot_time_lapse)) +
        theme_minimal() +
        stat_ecdf(geom = 'step', color = '#4e79a7', size = 2, alpha = 0.7) +
        geom_hline(yintercept = 0.95, color = '#e15759', size = 1.5) +
        geom_vline(xintercept = 23, color = '#e15759', size = 1.5, linetype = 2)
```

```{r}
### for generic probabilistic model ###
df_multi_paths_tl_1 <- reshape2::melt(df_multi_paths_tl[c(1:50), ] %>% select(customer_id, first_touch_date, last_touch_date, conversion),
                    id.vars = c('customer_id', 'conversion'),
                    value.name = 'touch_date') %>%
        arrange(customer_id)
rep_date <- as.Date('2016-01-10', format = '%Y-%m-%d')

ggplot(df_multi_paths_tl_1, aes(x = as.factor(customer_id), y = touch_date, color = factor(conversion), group = customer_id)) +
        theme_minimal() +
        coord_flip() +
        geom_point(size = 2) +
        geom_line(size = 0.5, color = 'darkgrey') +
        geom_hline(yintercept = as.numeric(rep_date), color = '#e15759', size = 2) +
        geom_rect(xmin = -Inf, xmax = Inf, ymin = as.numeric(rep_date), ymax = Inf, alpha = 0.01, color = 'white', fill = 'white') +
        theme(legend.position = 'bottom',
              panel.border = element_blank(),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.ticks.x = element_blank(),
              axis.ticks.y = element_blank()) +
        guides(colour = guide_legend(override.aes = list(size = 5)))
```

```{r}
df_multi_paths_tl_2 <- df_path_1_clean_multi %>%
        group_by(customer_id) %>%
        mutate(prev_touch_date = lag(date)) %>%
        ungroup() %>%
        filter(conversion == 1) %>%
        mutate(prev_time_lapse = round(as.numeric(date - prev_touch_date)))

# distribution
ggplot(df_multi_paths_tl_2, aes(x = prev_time_lapse)) +
        theme_minimal() +
        geom_histogram(fill = '#4e79a7', binwidth = 1)

# cumulative distribution
ggplot(df_multi_paths_tl_2, aes(x = prev_time_lapse)) +
        theme_minimal() +
        stat_ecdf(geom = 'step', color = '#4e79a7', size = 2, alpha = 0.7) +
        geom_hline(yintercept = 0.95, color = '#e15759', size = 1.5) +
        geom_vline(xintercept = 12, color = '#e15759', size = 1.5, linetype = 2)
```

Reading the two cutoffs together yields an operational churn rule: a customer whose
*first* contact was more than 23 days ago and whose *most recent* contact was more
than 12 days ago is on a fruitless path and may be excluded or re-targeted. The
final block operationalizes exactly this flag.

```{r}
# extracting data for generic model
df_multi_paths_tl_3 <- df_path_1_clean_multi %>%
        group_by(customer_id) %>%
        mutate(prev_time_lapse = round(as.numeric(date - lag(date)))) %>%
        summarise(path = paste(channel, collapse = ' > '),
                  tot_time_lapse = round(as.numeric(max(date) - min(date))),
                  prev_touch_tl = prev_time_lapse[which(max(date) == date)],
                  conversion = sum(conversion)) %>%
        ungroup() %>%
        mutate(is_fruitless = ifelse(conversion == 0 & tot_time_lapse > 20 & prev_touch_tl > 10, TRUE, FALSE)) %>%
        filter(conversion == 1 | is_fruitless == TRUE)
```

**Channel value comparison.** Finally, supplying `markov_model()` with a `var_value`
argument attributes *gross margin* rather than raw conversions, so channels are
ranked by the revenue they help generate rather than by conversion counts alone—the
managerially relevant currency.

#### Worked Example: Value Attribution {#sec-markov-example-3}

The third example, following Bounteous, exercises the value-attribution path on the
package's bundled `PathData`. Here each path carries four quantities: the channel
sequence, the conversion count, the monetary value of each conversion, and the count
of exits (the null variable).

```{r}
# Load these libraries (every time you start RStudio)
library(ChannelAttribution)
library(reshape)
library(ggplot2)

# This loads the demo data. You can load your own data by importing a dataset or reading in a file
data(PathData)
```

The heuristic and Markov models are fit on both conversions and conversion value, so
that two parallel comparisons—credit and dollars—can be drawn.

```{r}
H <- heuristic_models(Data, 'path', 'total_conversions', var_value='total_conversion_value')
```

```{r}
M <- markov_model(Data, 'path', 'total_conversions', var_value='total_conversion_value', order = 1)
```

```{r}
# Merges the two data frames on the "channel_name" column.
R <- merge(H, M, by='channel_name')

# Selects only relevant columns
R1 <- R[, (colnames(R)%in%c('channel_name', 'first_touch_conversions', 'last_touch_conversions', 'linear_touch_conversions', 'total_conversion'))]

# Renames the columns
colnames(R1) <- c('channel_name', 'first_touch', 'last_touch', 'linear_touch', 'markov_model')

# Transforms the dataset into a data frame that ggplot2 can use to graph the outcomes
R1 <- melt(R1, id='channel_name')
```

```{r}
#| label: fig-value-conversions
#| fig-cap: "Total conversions attributed to each channel under the four rules on the bundled PathData."
ggplot(R1, aes(channel_name, value, fill = variable)) +
  geom_bar(stat='identity', position='dodge') +
  ggtitle('TOTAL CONVERSIONS') +
  theme(axis.title.x = element_text(vjust = -2)) +
  theme(axis.title.y = element_text(vjust = +2)) +
  theme(title = element_text(size = 16)) +
  theme(plot.title=element_text(size = 20)) +
  ylab("")
```

@fig-value-conversions reports how many conversions each method assigns to each
channel. The companion chart in @fig-value-monetary repeats the exercise on
*monetary value*, which can reorder the channels when high-converting channels close
low-value purchases and vice versa.

```{r}
#| label: fig-value-monetary
#| fig-cap: "Total conversion value attributed to each channel. Comparing against the conversion-count chart reveals channels that convert often but cheaply, or rarely but at high value."
R2 <- R[, (colnames(R)%in%c('channel_name', 'first_touch_value', 'last_touch_value', 'linear_touch_value', 'total_conversion_value'))]

colnames(R2) <- c('channel_name', 'first_touch', 'last_touch', 'linear_touch', 'markov_model')

R2 <- melt(R2, id='channel_name')

ggplot(R2, aes(channel_name, value, fill = variable)) +
  geom_bar(stat='identity', position='dodge') +
  ggtitle('TOTAL VALUE') +
  theme(axis.title.x = element_text(vjust = -2)) +
  theme(axis.title.y = element_text(vjust = +2)) +
  theme(title = element_text(size = 16)) +
  theme(plot.title=element_text(size = 20)) +
  ylab("")
```

The contrast between @fig-value-conversions and @fig-value-monetary is the practical
payoff of value attribution: a channel that converts often but cheaply and one that
converts rarely but at high value receive very different budget recommendations,
and only the monetary view distinguishes them.

## Sales-Funnel Models {#sec-sales-funnel}

Attribution credits *channels*; funnel analysis credits *stages*. The classical
**AIDA** funnel posits that a prospect moves through awareness, interest, desire, and
action in sequence,

$$
\text{Awareness} \to \text{Interest} \to \text{Desire} \to \text{Action},
$$ {#eq-aida}

and that the firm's task is to maximize the conversion probability at each transition
while minimizing leakage. Where Markov attribution treats channels as nodes, the
funnel model treats the @eq-aida stages as nodes and asks which stage a channel most
helps a customer clear.

```{mermaid}
%%| label: fig-funnel-stages
%%| fig-cap: "The operationalized AIDA funnel. Each stage is defined by an observable on-site event, and the step probability is the share of customers who advance from one stage to the next."
flowchart TD
    A["Site visit<br/>(necessary condition)"] --> B["Awareness<br/>two pages visited"]
    B --> C["Interest<br/>product page reviewed"]
    C --> D["Desire<br/>add to cart"]
    D --> E["Action<br/>purchase complete"]
```

### Funnel-Based Attribution {#sec-funnel-attribution}

The first funnel model, following Sergey Bryl, operationalizes the abstract AIDA
stages of @eq-aida as observable on-site events, as @fig-funnel-stages depicts: a
first site visit is the necessary entry condition; visiting two pages marks
*awareness*; reviewing a product page marks *interest*; adding to cart marks
*desire*; and completing the purchase marks *action*. The data are simulated as a
clickstream in which each later event is conditional on the previous one, so the
funnel structure is built into the generating process.

```{r}
library(tidyverse)
library(purrrlyr)
library(reshape2)

##### simulating the "real" data #####
set.seed(454)
df_raw <-
  data.frame(
    customer_id = paste0('id', sample(c(1:5000), replace = TRUE)),
    date = as.POSIXct(
      rbeta(10000, 0.7, 10) * 10000000,
      origin = '2017-01-01',
      tz = "UTC"
    ),
    channel = paste0('channel_', sample(
      c(0:7),
      10000,
      replace = TRUE,
      prob = c(0.2, 0.12, 0.03, 0.07, 0.15, 0.25, 0.1, 0.08)
    )),
    site_visit = 1
  ) %>%

  mutate(
    two_pages_visit = sample(c(0, 1), 10000, replace = TRUE, prob = c(0.8, 0.2)),
    product_page_visit = ifelse(
      two_pages_visit == 1,
      sample(
        c(0, 1),
        length(two_pages_visit[which(two_pages_visit == 1)]),
        replace = TRUE,
        prob = c(0.75, 0.25)
      ),
      0
    ),
    add_to_cart = ifelse(
      product_page_visit == 1,
      sample(
        c(0, 1),
        length(product_page_visit[which(product_page_visit == 1)]),
        replace = TRUE,
        prob = c(0.1, 0.9)
      ),
      0
    ),
    purchase = ifelse(add_to_cart == 1,
                      sample(
                        c(0, 1),
                        length(add_to_cart[which(add_to_cart == 1)]),
                        replace = TRUE,
                        prob = c(0.02, 0.98)
                      ),
                      0)
  ) %>%
  dmap_at(c('customer_id', 'channel'), as.character) %>%
  arrange(date) %>%
  mutate(session_id = row_number()) %>%
  arrange(customer_id, session_id)
df_raw <-
  reshape2::melt(
    df_raw,
    id.vars = c('customer_id', 'date', 'channel', 'session_id'),
    value.name = "trigger",
    variable.name = 'event'
  ) %>%
  filter(trigger == 1) %>%
  select(-trigger) %>%
  arrange(customer_id, date)
```

Because the analysis assumes every customer is a first-time buyer, only the *first*
occurrence of each event per customer is retained; any later repurchase event is
discarded.

```{r}
### removing not first events ###
df_customers <- df_raw %>%
  group_by(customer_id, event) %>%
  filter(date == min(date)) %>%
  ungroup()
```

The core quantities are the **step probabilities**. Let $c_k$ be the number of
customers reaching stage $k$. The cumulative funnel probability is
$c_k / c_{\text{site\_visit}}$, the step probability is $c_k / c_{k-1}$, and the
*importance* of clearing step $k$ is $1 - c_k / c_{k-1}$—the share of customers lost
at that transition, which is exactly the leakage the firm wants to plug. Normalizing
the importances to sum to one yields stage weights that play the role the removal
effect played for channels.

```{r}
### Sales Funnel probabilities ###
sf_probs <- df_customers %>%

    group_by(event) %>%
    summarise(customers_on_step = n()) %>%
    ungroup() %>%

    mutate(
        sf_probs = round(customers_on_step / customers_on_step[event == 'site_visit'], 3),
        sf_probs_step = round(customers_on_step / lag(customers_on_step), 3),
        sf_probs_step = ifelse(is.na(sf_probs_step) == TRUE, 1, sf_probs_step),
        sf_importance = 1 - sf_probs_step,
        sf_importance_weighted = sf_importance / sum(sf_importance)
    )
```

The journey visualization in @fig-funnel-journeys overlays every customer's path on the
funnel and annotates each stage with its cumulative conversion probability, so that the
width of the leakage at each step is immediately visible.

```{r}
#| label: fig-funnel-journeys
#| fig-cap: "All customer journeys overlaid on the funnel, colored by acquisition channel, with cumulative conversion probability annotated at each stage."
### Sales Funnel visualization ###
df_customers_plot <- df_customers %>%

    group_by(event) %>%
    arrange(channel) %>%
    mutate(pl = row_number()) %>%
    ungroup() %>%

    mutate(
        pl_new = case_when(
            event == 'two_pages_visit' ~ round((max(pl[event == 'site_visit']) - max(pl[event == 'two_pages_visit'])) / 2),
            event == 'product_page_visit' ~ round((max(pl[event == 'site_visit']) - max(pl[event == 'product_page_visit'])) / 2),
            event == 'add_to_cart' ~ round((max(pl[event == 'site_visit']) - max(pl[event == 'add_to_cart'])) / 2),
            event == 'purchase' ~ round((max(pl[event == 'site_visit']) - max(pl[event == 'purchase'])) / 2),
            TRUE ~ 0
        ),
        pl = pl + pl_new
    )

df_customers_plot$event <-
    factor(
        df_customers_plot$event,
        levels = c(
            'purchase',
            'add_to_cart',
            'product_page_visit',
            'two_pages_visit',
            'site_visit'
        )
    )

# color palette
cols <- c(
    '#4e79a7',
    '#f28e2b',
    '#e15759',
    '#76b7b2',
    '#59a14f',
    '#edc948',
    '#b07aa1',
    '#ff9da7',
    '#9c755f',
    '#bab0ac'
)

ggplot(df_customers_plot, aes(x = event, y = pl)) +
    theme_minimal() +
    scale_colour_manual(values = cols) +
    coord_flip() +
    geom_line(aes(group = customer_id, color = as.factor(channel)), size = 0.05) +
    geom_text(
        data = sf_probs,
        aes(
            x = event,
            y = 1,
            label = paste0(sf_probs * 100, '%')
        ),
        size = 4,
        fontface = 'bold'
    ) +
    guides(color = guide_legend(override.aes = list(size = 2))) +
    theme(
        legend.position = 'bottom',
        legend.direction = "horizontal",
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        plot.title = element_text(
            size = 20,
            face = "bold",
            vjust = 2,
            color = 'black',
            lineheight = 0.8
        ),
        axis.title.y = element_text(size = 16, face = "bold"),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_text(
            size = 8,
            angle = 90,
            hjust = 0.5,
            vjust = 0.5,
            face = "plain"
        )
    ) +
    ggtitle("Sales Funnel visualization - all customers journeys")
```

Channel attribution then follows by crediting each channel with the weighted
importance of the stages a converting customer cleared while exposed to it: customers
who never purchased are dropped, the stage weights are joined on, and the weighted
importances are summed by channel.

```{r}
### computing attribution ###
df_attrib <- df_customers %>%
    # removing customers without purchase
    group_by(customer_id) %>%
    filter(any(as.character(event) == 'purchase')) %>%
    ungroup() %>%

    # joining step's importances
    left_join(., sf_probs %>% select(event, sf_importance_weighted), by = 'event') %>%

    group_by(channel) %>%
    summarise(tot_attribution = sum(sf_importance_weighted)) %>%
    ungroup()
```

### Funnel Visualization {#sec-funnel-viz}

The second funnel example, also following Bryl, is a presentation device rather than
an estimator: it renders a funnel chart in which each stage's width is proportional
to the number of customers present, with conversion rates annotated between stages.
The data combine *content* nodes (pages mapped to AIDA stages) with *customer*
lifecycle nodes (new, engaged, loyal), and the plotting code constructs the
symmetric trapezoidal silhouette by adding white "dummy" segments that center each
bar. @fig-funnel-chart presents the result, with stage widths proportional to customer
counts and the step conversion rates annotated between stages.

```{r}
#| label: fig-funnel-chart
#| fig-cap: "Funnel chart with stage widths proportional to customer counts and step conversion rates annotated between stages, combining content stages with customer-lifecycle stages."
library(dplyr)
library(ggplot2)
library(reshape2)

# creating a data samples
# content
df.content <- data.frame(
    content = c(
        'main',
        'ad landing',
        'product 1',
        'product 2',
        'product 3',
        'product 4',
        'shopping cart',
        'thank you page'
    ),
    step = c(
        'awareness',
        'awareness',
        'interest',
        'interest',
        'interest',
        'interest',
        'desire',
        'action'
    ),
    number = c(150000, 80000,
               80000, 40000, 35000, 25000,
               130000,
               120000)
)
# customers
df.customers <- data.frame(
    content = c('new', 'engaged', 'loyal'),
    step = c('new', 'engaged', 'loyal'),
    number = c(25000, 40000, 55000)
)
# combining two data sets
df.all <- rbind(df.content, df.customers)

# calculating dummies, max and min values of X for plotting
df.all <- df.all %>%
    group_by(step) %>%
    mutate(totnum = sum(number)) %>%
    ungroup() %>%
    mutate(dum = (max(totnum) - totnum) / 2,
           maxx = totnum + dum,
           minx = dum)

# data frame for plotting funnel lines
df.lines <- df.all %>%
    distinct(step, maxx, minx)

# data frame with dummies
df.dum <- df.all %>%
    distinct(step, dum) %>%
    mutate(content = 'dummy',
           number = dum) %>%
    select(content, step, number)

# data frame with rates
conv <- df.all$totnum[df.all$step == 'action']

df.rates <- df.all %>%
    distinct(step, totnum) %>%
    mutate(
        prevnum = lag(totnum),
        rate = ifelse(
            step == 'new' | step == 'engaged' | step == 'loyal',
            round(totnum / conv, 3),
            round(totnum / prevnum, 3)
        )
    ) %>%
    select(step, rate)
df.rates <- na.omit(df.rates)

# creting final data frame
df.all <- df.all %>%
    select(content, step, number)

df.all <- rbind(df.all, df.dum)

# defining order of steps
df.all$step <-
    factor(
        df.all$step,
        levels = c(
            'loyal',
            'engaged',
            'new',
            'action',
            'desire',
            'interest',
            'awareness'
        )
    )
df.all <- df.all %>%
    arrange(desc(step))
list1 <- df.all %>% distinct(content) %>%
    filter(content != 'dummy')
df.all$content <-
    factor(df.all$content, levels = c(as.character(list1$content), 'dummy'))

# calculating position of labels
df.all <- df.all %>%
    arrange(step, desc(content)) %>%
    group_by(step) %>%
    mutate(pos = cumsum(number) - 0.5 * number) %>%
    ungroup()

# creating custom palette with 'white' color for dummies
cols <- c(
    "#fec44f",
    "#fc9272",
    "#a1d99b",
    "#fee0d2",
    "#2ca25f",
    "#8856a7",
    "#43a2ca",
    "#fdbb84",
    "#e34a33",
    "#a6bddb",
    "#dd1c77",
    "#ffffff"
)

# plotting chart
ggplot() +
    theme_minimal() +
    coord_flip() +
    scale_fill_manual(values = cols) +
    geom_bar(
        data = df.all,
        aes(x = step, y = number, fill = content),
        stat = "identity",
        width = 1
    ) +
    geom_text(
        data = df.all[df.all$content != 'dummy',],
        aes(
            x = step,
            y = pos,
            label = paste0(content, '-', number / 1000, 'K')
        ),
        size = 4,
        color = 'white',
        fontface = "bold"
    ) +
    geom_ribbon(data = df.lines,
                aes(
                    x = step,
                    ymax = max(maxx),
                    ymin = maxx,
                    group = 1
                ),
                fill = 'white') +
    geom_line(
        data = df.lines,
        aes(x = step, y = maxx, group = 1),
        color = 'darkred',
        size = 4
    ) +
    geom_ribbon(data = df.lines,
                aes(
                    x = step,
                    ymax = minx,
                    ymin = min(minx),
                    group = 1
                ),
                fill = 'white') +
    geom_line(
        data = df.lines,
        aes(x = step, y = minx, group = 1),
        color = 'darkred',
        size = 4
    ) +
    geom_text(
        data = df.rates,
        aes(
            x = step,
            y = (df.lines$minx[-1]),
            label = paste0(rate * 100, '%')
        ),
        hjust = 1.2,
        color = 'darkblue',
        fontface = "bold"
    ) +
    theme(
        legend.position = 'none',
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank()
    )
```

## Recency, Frequency, Monetary (RFM) {#sec-rfm}

RFM scoring is the oldest and most durable behavioral-segmentation device in direct
marketing. The premise is that three sufficient statistics of a customer's
transaction history predict future value: how *recently* they purchased, how
*frequently* they purchase, and how much *monetary* value they have generated.
Customers who bought recently, buy often, and spend heavily are the most responsive
to the next offer.

Formally, for customer $i$ observed up to an analysis date $t_0$, define recency
$R_i = t_0 - \max_j t_{ij}$ as the elapsed time since the most recent of their
transactions $\{t_{ij}\}$, frequency $F_i = n_i$ as the transaction count, and
monetary value $M_i = \sum_j m_{ij}$ as cumulative revenue. Each raw measure is
mapped to an ordinal score, conventionally a quintile $\{1,\dots,5\}$, by binning the
empirical distribution: recency is reverse-coded so that more recent buyers score
higher, while frequency and monetary value score higher with larger values. The
composite **RFM score** concatenates or sums the three sub-scores and is the basis for
the segmentation rules of @sec-segmentation.

The `rfm` package implements this pipeline from either customer-level summaries or
raw order logs. The customer-level table requires a customer identifier, a
transaction count, total revenue, and days since the last visit; the order-level
table requires an identifier, a transaction date, and an amount, from which
`rfm_table_order()` derives the summaries.

```{r}
library("rfm")
rfm_data_customer

rfm_data_orders # to generate data_orders, use rfm_table_order()
```

Supplying an explicit analysis date $t_0$ makes recency well defined; the
`rfm_table_customer()` call below scores the customer-level data and returns, for
each customer, the most-recent-visit date, recency in days, transaction count, total
amount, the three sub-scores, and the composite.

```{r}
analysis_date <- lubridate::as_date('2007-01-01')
rfm_result <-
    rfm_table_customer(
        rfm_data_customer,
        customer_id,
        number_of_orders,
        recency_days,
        revenue,
        analysis_date
    )
rfm_result
```

### Visualizing RFM {#sec-rfm-viz}

A battery of standard diagnostics interrogates the scored data. The heatmap in
@fig-rfm-heatmap displays average monetary value across the recency–frequency grid,
revealing whether high-value customers cluster in the recent-and-frequent corner as the
premise predicts.

```{r}
#| label: fig-rfm-heatmap
#| fig-cap: "Average monetary value across the recency-by-frequency grid. Concentration in the high-recency, high-frequency corner confirms the RFM premise that recent frequent buyers are the most valuable."
rfm_heatmap(rfm_result)
```

Three further marginal views complement the heatmap: @fig-rfm-bar shows the distribution
of customers across recency scores within frequency bins, @fig-rfm-hist gives the
marginal histograms of the three measures, and @fig-rfm-orders reports the distribution
of customers by order count.

```{r}
#| label: fig-rfm-bar
#| fig-cap: "Distribution of customers across recency scores within frequency bins."
rfm_bar_chart(rfm_result)
```

```{r}
#| label: fig-rfm-hist
#| fig-cap: "Marginal histograms of the recency, frequency, and monetary measures."
rfm_histograms(rfm_result)
```

```{r}
#| label: fig-rfm-orders
#| fig-cap: "Distribution of customers by number of orders."
rfm_order_dist(rfm_result)
```

The bivariate scatter plots in @fig-rfm-scatter, recency against monetary, frequency
against monetary, and recency against frequency, expose the pairwise dependence structure
that the composite score collapses.

```{r}
#| label: fig-rfm-scatter
#| fig-cap: "Bivariate scatter plots of the three RFM dimensions, exposing the pairwise dependence the composite score collapses."
# the rfm package's rfm_rm_plot/rfm_fm_plot/rfm_rf_plot helpers error on this version,
# so we draw the three pairwise scatters directly from the scored RFM table
library(ggplot2)
rfm_df <- rfm_result$rfm
p_rm <- ggplot(rfm_df, aes(amount, recency_days)) + geom_point(alpha = .4) +
    labs(x = "Monetary", y = "Recency (days)", title = "Recency vs Monetary") + theme_bw()
p_fm <- ggplot(rfm_df, aes(amount, transaction_count)) + geom_point(alpha = .4) +
    labs(x = "Monetary", y = "Frequency", title = "Frequency vs Monetary") + theme_bw()
p_rf <- ggplot(rfm_df, aes(transaction_count, recency_days)) + geom_point(alpha = .4) +
    labs(x = "Frequency", y = "Recency (days)", title = "Recency vs Frequency") + theme_bw()
gridExtra::grid.arrange(p_rm, p_fm, p_rf, ncol = 3)
```

### Adding Clumpiness: RFMC {#sec-rfmc}

The three RFM moments summarize *how much* and *how recently* a customer transacts
but are blind to *how the events are spaced in time*. Two customers with identical
frequency over a window can differ sharply in whether their purchases were evenly
spread or arrived in a tight burst. **Clumpiness** (the "C" in RFMC) measures this
spacing, defined by @Zhang_2015 as the degree of nonconformity to equal spacing of
events.

> Clumpiness is the serial dependence or non-constant propensity—specifically,
> temporary elevations of propensity, i.e., periods during which one event is more
> likely to occur than the average level [@Zhang_2013].

The construct migrated into marketing from two prior literatures. In finance,
clumpiness can flag high growth potential coupled with elevated risk, so it informs
firm-acquisition decisions; the phenomenon itself originates in the sports "hot hand"
effect, in which success appears to breed further success. A well-behaved clumpiness
measure satisfies three properties [@Zhang_2015]: it attains its **minimum** when
events are equally spaced and its **maximum** when they cluster tightly together;
it varies **continuously** in the event times; and it **converges** appropriately as
the number of events grows. Appending a clumpiness score to the RFM triple sharpens
churn prediction, because a customer whose previously bursty buying has gone quiet is
a stronger churn signal than recency alone conveys.

## Customer Segmentation {#sec-segmentation}

Segmentation partitions the customer base into groups that are internally homogeneous
and managerially distinct, so that each group can receive a tailored treatment. RFM
scores furnish a natural partition; lifecycle and cohort views add the time dimension
that a static score omits.

### RFM-Based Segments {#sec-rfm-segments}

The most direct segmentation reads named segments off the RFM score by specifying,
for each segment, the admissible ranges of the recency, frequency, and monetary
sub-scores. A "Premium" customer scores high on all three; an "At Risk" customer
combines low recency with high historical frequency and value; a "Lost" customer
scores low everywhere. The `rfm_segment()` function maps every customer to the first
segment whose score box contains them.

```{r, eval = FALSE}
segment_names <-
    c(
        "Premium",
        "Loyal Customers",
        "Potential Loyalist",
        "New Customers",
        "Promising",
        "Need Attention",
        "About To Churn",
        "At Risk",
        "High Value Churners/Resurrection",
        "Low Value Churners"
    )

recency_lower <- c(4, 2, 3, 4, 3, 2, 2, 1, 1, 1)
recency_upper <- c(5, 5, 5, 5, 4, 3, 3, 2, 1, 2)
frequency_lower <- c(4, 3, 1, 1, 1, 2, 1, 2, 4, 1)
frequency_upper <- c(5, 5, 3, 1, 1, 3, 2, 5, 5, 2)
monetary_lower <- c(4, 3, 1, 1, 1, 2, 1, 2, 4, 1)
monetary_upper <- c(5, 5, 3, 1, 1, 3, 2, 5, 5, 2)

rfm_segments <-
    rfm_segment(
        rfm_result,
        segment_names,
        recency_lower,
        recency_upper,
        frequency_lower,
        frequency_upper,
        monetary_lower,
        monetary_upper
    )

head(rfm_segments, n = 5)


rfm_segments %>%
    count(rfm_segments$segment) %>%
    arrange(desc(n)) %>%
    rename(Count = n)

# median recency
rfm_plot_median_recency(rfm_segments)

# median frequency
rfm_plot_median_frequency(rfm_segments)

# Median Monetary Value
rfm_plot_median_monetary(rfm_segments)
```

### LifeCycle Grids {#sec-lifecycle-grids}

A score-box segmentation is static. The **lifecycle grid**, following Sergey,
restores dynamics by cross-tabulating recency bands against frequency bands and
populating each cell with the customers it contains, so that movement through the
grid over time traces a customer's lifecycle. We simulate an order log with products,
clients, gender, and order dates, then collapse it to one row per client carrying
frequency and recency.

```{r}
# loading libraries
library(dplyr)
library(reshape2)
library(ggplot2)

# creating data sample
set.seed(10)
data <- data.frame(
    orderId = sample(c(1:1000), 5000, replace = TRUE),
    product = sample(
        c('NULL', 'a', 'b', 'c'),
        5000,
        replace = TRUE,
        prob = c(0.15, 0.65, 0.3, 0.15)
    )
)
order <- data.frame(orderId = c(1:1000),
                    clientId = sample(c(1:300), 1000, replace = TRUE))
gender <- data.frame(clientId = c(1:300),
                     gender = sample(
                         c('male', 'female'),
                         300,
                         replace = TRUE,
                         prob = c(0.40, 0.60)
                     ))
date <- data.frame(orderId = c(1:1000),
                   orderdate = sample((1:100), 1000, replace = TRUE))
orders <- merge(data, order, by = 'orderId')
orders <- merge(orders, gender, by = 'clientId')
orders <- merge(orders, date, by = 'orderId')
orders <- orders[orders$product != 'NULL',]
orders$orderdate <- as.Date(orders$orderdate, origin = "2012-01-01")
rm(data, date, order, gender)
```

```{r}
#| label: fig-lcg-explore
#| fig-cap: "Marginal distributions of purchase frequency and recency in the simulated order log."
# reporting date
today <- as.Date('2012-04-11', format = '%Y-%m-%d')

# processing data
orders <-
    dcast(
        orders,
        orderId + clientId + gender + orderdate ~ product,
        value.var = 'product',
        fun.aggregate = length
    )

orders <- orders %>%
    group_by(clientId) %>%
    mutate(frequency = n(),
           recency = as.numeric(today - orderdate)) %>%
    filter(orderdate == max(orderdate)) %>%
    filter(orderId == max(orderId)) %>%
    ungroup()

# exploratory analysis
ggplot(orders, aes(x = frequency)) +
    theme_bw() +
    scale_x_continuous(breaks = c(1:10)) +
    geom_bar(alpha = 0.6, width = 1) +
    ggtitle("Distribution by frequency")

ggplot(orders, aes(x = recency)) +
    theme_bw() +
    geom_bar(alpha = 0.6, width = 1) +
    ggtitle("Distribution by recency")
```

The two panels above plot the marginal distributions of purchase frequency and recency in
the simulated log. The continuous frequency and recency are then bucketed into ordered
bands, and the band factors are ordered so the grid reads from most-recent-and-frequent
to least, with a last-cart feature capturing which products the customer last bought.

```{r}
orders.segm <- orders %>%
    mutate(segm.freq = ifelse(between(frequency, 1, 1), '1',
                              ifelse(
                                  between(frequency, 2, 2), '2',
                                  ifelse(between(frequency, 3, 3), '3',
                                         ifelse(
                                             between(frequency, 4, 4), '4',
                                             ifelse(between(frequency, 5, 5), '5', '>5')
                                         ))
                              ))) %>%
    mutate(segm.rec = ifelse(
        between(recency, 0, 6),
        '0-6 days',
        ifelse(
            between(recency, 7, 13),
            '7-13 days',
            ifelse(
                between(recency, 14, 19),
                '14-19 days',
                ifelse(
                    between(recency, 20, 45),
                    '20-45 days',
                    ifelse(between(recency, 46, 80), '46-80 days', '>80 days')
                )
            )
        )
    )) %>%
    # creating last cart feature
    mutate(cart = paste(
        ifelse(a != 0, 'a', ''),
        ifelse(b != 0, 'b', ''),
        ifelse(c != 0, 'c', ''),
        sep = ''
    )) %>%
    arrange(clientId)

# defining order of boundaries
orders.segm$segm.freq <-
    factor(orders.segm$segm.freq, levels = c('>5', '5', '4', '3', '2', '1'))
orders.segm$segm.rec <-
    factor(
        orders.segm$segm.rec,
        levels = c(
            '>80 days',
            '46-80 days',
            '20-45 days',
            '14-19 days',
            '7-13 days',
            '0-6 days'
        )
    )
```

Faceting customer counts by the two band factors produces the grid itself, shown in
@fig-lcg-basic: each panel is one recency–frequency cell, and its height encodes how many
customers presently occupy it.

```{r}
#| label: fig-lcg-basic
#| fig-cap: "Lifecycle grid: customer counts faceted by recency band (columns) and frequency band (rows)."
lcg <- orders.segm %>%
    group_by(segm.rec, segm.freq) %>%
    summarise(quantity = n()) %>%
    mutate(client = 'client') %>%
    ungroup()
lcg.matrix <-
    dcast(lcg,
          segm.freq ~ segm.rec,
          value.var = 'quantity',
          fun.aggregate = sum)

ggplot(lcg, aes(x = client, y = quantity, fill = quantity)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    geom_bar(stat = 'identity', alpha = 0.6) +
    geom_text(aes(y = max(quantity) / 2, label = quantity), size = 4) +
    facet_grid(segm.freq ~ segm.rec) +
    ggtitle("LifeCycle Grids")
```

Coarsening the grid into recent/not-recent and frequent/infrequent quadrants, and
splitting by gender and last cart, turns the grid into a strategic map of where value
concentrates and which sub-populations occupy each region, as the enriched grids below display.

```{r}
#| label: fig-lcg-advanced
#| fig-cap: "Lifecycle grids enriched with quadrant coloring (recent/frequent), and split by gender and last-cart composition, turning the grid into a strategic map of the customer base."
lcg.adv <- lcg %>%
    mutate(
        rec.type = ifelse(
            segm.rec %in% c("> 80 days", "46 - 80 days", "20 - 45 days"),
            "not recent",
            "recent"
        ),
        freq.type = ifelse(segm.freq %in% c(" >
          5", "5", "4"), "frequent", "infrequent"),
        customer.type = interaction(rec.type, freq.type)
    )

ggplot(lcg.adv, aes(x = client, y = quantity, fill = customer.type)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    facet_grid(segm.freq ~ segm.rec) +
    geom_bar(stat = 'identity', alpha = 0.6) +
    geom_text(aes(y = max(quantity) / 2, label = quantity), size = 4) +
    ggtitle("LifeCycle Grids")

# with background
ggplot(lcg.adv, aes(x = client, y = quantity, fill = customer.type)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    geom_rect(
        aes(fill = customer.type),
        xmin = -Inf,
        xmax = Inf,
        ymin = -Inf,
        ymax = Inf,
        alpha = 0.1
    ) +
    facet_grid(segm.freq ~ segm.rec) +
    geom_bar(stat = 'identity', alpha = 0.7) +
    geom_text(aes(y = max(quantity) / 2, label = quantity), size = 4) +
    ggtitle("LifeCycle Grids")

lcg.sub <- orders.segm %>%
    group_by(gender, cart, segm.rec, segm.freq) %>%
    summarise(quantity = n()) %>%
    mutate(client = 'client') %>%
    ungroup()

ggplot(lcg.sub, aes(x = client, y = quantity, fill = gender)) +
    theme_bw() +
    scale_fill_brewer(palette = 'Set1') +
    theme(panel.grid = element_blank()) +
    geom_bar(stat = 'identity',
             position = 'fill' ,
             alpha = 0.6) +
    facet_grid(segm.freq ~ segm.rec) +
    ggtitle("LifeCycle Grids by gender (proportion)")

ggplot(lcg.sub, aes(x = gender, y = quantity, fill = cart)) +
    theme_bw() +
    scale_fill_brewer(palette = 'Set1') +
    theme(panel.grid = element_blank()) +
    geom_bar(stat = 'identity',
             position = 'fill' ,
             alpha = 0.6) +
    facet_grid(segm.freq ~ segm.rec) +
    ggtitle("LifeCycle Grids by gender and last cart (proportion)")
```

#### CLV and CAC on the Grid {#sec-clv-cac}

Customer counts alone do not reveal whether a segment is *profitable*. Overlaying
**customer lifetime value (CLV)**—the gross margin a customer has generated—and
**customer acquisition cost (CAC)**—what was spent to acquire them—on the grid shows
which cells return their acquisition investment. A cell where average CLV exceeds
average CAC is profitable; a cell where it does not is a candidate for reduced
acquisition spend. We attach simulated per-client CAC and product gross margins,
recompute the grid, and plot CLV against CAC both in aggregate and per client in
the grids below.

```{r}
#| label: fig-lcg-clv-cac
#| fig-cap: "Lifecycle grids overlaying customer lifetime value against acquisition cost, in total and per client. Cells where CLV exceeds CAC return their acquisition investment."
# loading libraries
library(dplyr)
library(reshape2)
library(ggplot2)

# creating data sample
set.seed(10)
data <- data.frame(
    orderId = sample(c(1:1000), 5000, replace = TRUE),
    product = sample(
        c('NULL', 'a', 'b', 'c'),
        5000,
        replace = TRUE,
        prob = c(0.15, 0.65, 0.3, 0.15)
    )
)
order <- data.frame(orderId = c(1:1000),
                    clientId = sample(c(1:300), 1000, replace = TRUE))
gender <- data.frame(clientId = c(1:300),
                     gender = sample(
                         c('male', 'female'),
                         300,
                         replace = TRUE,
                         prob = c(0.40, 0.60)
                     ))
date <- data.frame(orderId = c(1:1000),
                   orderdate = sample((1:100), 1000, replace = TRUE))
orders <- merge(data, order, by = 'orderId')
orders <- merge(orders, gender, by = 'clientId')
orders <- merge(orders, date, by = 'orderId')
orders <- orders[orders$product != 'NULL', ]
orders$orderdate <- as.Date(orders$orderdate, origin = "2012-01-01")

# creating data frames with CAC and Gross margin
cac <-
    data.frame(clientId = unique(orders$clientId),
               cac = sample(c(10:15), 288, replace = TRUE))
gr.margin <-
    data.frame(product = c('a', 'b', 'c'),
               grossmarg = c(1, 2, 3))

rm(data, date, order, gender)

# reporting date
today <- as.Date('2012-04-11', format = '%Y-%m-%d')

# calculating customer lifetime value
orders <- merge(orders, gr.margin, by = 'product')

clv <- orders %>%
    group_by(clientId) %>%
    summarise(clv = sum(grossmarg)) %>%
    ungroup()

# processing data
orders <-
    dcast(
        orders,
        orderId + clientId + gender + orderdate ~ product,
        value.var = 'product',
        fun.aggregate = length
    )

orders <- orders %>%
    group_by(clientId) %>%
    mutate(frequency = n(),
           recency = as.numeric(today - orderdate)) %>%
    filter(orderdate == max(orderdate)) %>%
    filter(orderId == max(orderId)) %>%
    ungroup()

orders.segm <- orders %>%
    mutate(segm.freq = ifelse(between(frequency, 1, 1), '1',
                              ifelse(
                                  between(frequency, 2, 2), '2',
                                  ifelse(between(frequency, 3, 3), '3',
                                         ifelse(
                                             between(frequency, 4, 4), '4',
                                             ifelse(between(frequency, 5, 5), '5', '>5')
                                         ))
                              ))) %>%
    mutate(segm.rec = ifelse(
        between(recency, 0, 6),
        '0-6 days',
        ifelse(
            between(recency, 7, 13),
            '7-13 days',
            ifelse(
                between(recency, 14, 19),
                '14-19 days',
                ifelse(
                    between(recency, 20, 45),
                    '20-45 days',
                    ifelse(between(recency, 46, 80), '46-80 days', '>80 days')
                )
            )
        )
    )) %>%
    # creating last cart feature
    mutate(cart = paste(
        ifelse(a != 0, 'a', ''),
        ifelse(b != 0, 'b', ''),
        ifelse(c != 0, 'c', ''),
        sep = ''
    )) %>%
    arrange(clientId)

# defining order of boundaries
orders.segm$segm.freq <-
    factor(orders.segm$segm.freq, levels = c('>5', '5', '4', '3', '2', '1'))
orders.segm$segm.rec <-
    factor(
        orders.segm$segm.rec,
        levels = c(
            '>80 days',
            '46-80 days',
            '20-45 days',
            '14-19 days',
            '7-13 days',
            '0-6 days'
        )
    )

orders.segm <- merge(orders.segm, cac, by = 'clientId')
orders.segm <- merge(orders.segm, clv, by = 'clientId')

lcg.clv <- orders.segm %>%
    group_by(segm.rec, segm.freq) %>%
    summarise(quantity = n(),
              # calculating cumulative CAC and CLV
              cac = sum(cac),
              clv = sum(clv)) %>%
    ungroup() %>%
    # calculating CAC and CLV per client
    mutate(cac1 = round(cac / quantity, 2),
           clv1 = round(clv / quantity, 2))

lcg.clv <-
    reshape2::melt(lcg.clv, id.vars = c('segm.rec', 'segm.freq', 'quantity'))

ggplot(lcg.clv[lcg.clv$variable %in% c('clv', 'cac'), ], aes(x = variable, y =
                                                                 value, fill = variable)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    geom_bar(stat = 'identity', alpha = 0.6, aes(width = quantity / max(quantity))) +
    geom_text(aes(y = value, label = value), size = 4) +
    facet_grid(segm.freq ~ segm.rec) +
    ggtitle("LifeCycle Grids - CLV vs CAC (total)")
ggplot(lcg.clv[lcg.clv$variable %in% c('clv1', 'cac1'), ], aes(x = variable, y =
                                                                   value, fill = variable)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    geom_bar(stat = 'identity', alpha = 0.6, aes(width = quantity / max(quantity))) +
    geom_text(aes(y = value, label = value), size = 4) +
    facet_grid(segm.freq ~ segm.rec) +
    ggtitle("LifeCycle Grids - CLV vs CAC (average)")
```

#### Cohort Analysis {#sec-cohort}

A **cohort** groups customers by a shared origin—most often their acquisition
month—so that groups acquired under different conditions can be compared on equal
footing as they age. Cohort analysis combines customers through common
characteristics into homogeneous groups and tracks each group's value trajectory over
its lifetime. The setup below simulates a larger order log and assembles, per client,
frequency, recency, the average gap between purchases, the first-purchase cohort
label, realized CLV to date, an acquisition campaign with its CAC, and a predicted
*potential* CLV.

```{r}
# loading libraries
library(dplyr)
library(reshape2)
library(ggplot2)
library(networkD3)

# helper: render a from/to/weight link table as an embeddable D3 Sankey diagram.
# networkD3 produces a self-contained htmlwidget; googleVis's gvisSankey only opens a
# browser window and cannot embed in the rendered book, so we replace it here.
make_sankey <- function(links, from = "from", to = "to", weight = "n", font = 11) {
  # htmlwidget: HTML output only. PDF/EPUB cannot rasterize the percentage-width
  # widget via webshot2, so skip rendering it for non-HTML formats.
  if (!knitr::pandoc_to("html")) return(invisible(NULL))
  nodes <- data.frame(name = unique(c(links[[from]], links[[to]])), stringsAsFactors = FALSE)
  links$.src <- match(links[[from]], nodes$name) - 1L
  links$.tgt <- match(links[[to]],   nodes$name) - 1L
  networkD3::sankeyNetwork(Links = links, Nodes = nodes, Source = ".src",
                           Target = ".tgt", Value = weight, NodeID = "name",
                           fontSize = font, nodeWidth = 20)
}

set.seed(10)
# creating orders data sample
data <- data.frame(
    orderId = sample(c(1:5000), 25000, replace = TRUE),
    product = sample(
        c('NULL', 'a', 'b', 'c'),
        25000,
        replace = TRUE,
        prob = c(0.15, 0.65, 0.3, 0.15)
    )
)
order <- data.frame(orderId = c(1:5000),
                    clientId = sample(c(1:1500), 5000, replace = TRUE))
date <- data.frame(orderId = c(1:5000),
                   orderdate = sample((1:500), 5000, replace = TRUE))
orders <- merge(data, order, by = 'orderId')
orders <- merge(orders, date, by = 'orderId')
orders <- orders[orders$product != 'NULL',]
orders$orderdate <- as.Date(orders$orderdate, origin = "2012-01-01")
rm(data, date, order)
# creating data frames with CAC, Gross margin, Campaigns and Potential CLV
gr.margin <-
    data.frame(product = c('a', 'b', 'c'),
               grossmarg = c(1, 2, 3))
campaign <- data.frame(clientId = c(1:1500),
                       campaign = paste('campaign', sample(c(1:7), 1500, replace = TRUE), sep =
                                            ' '))
cac <-
    data.frame(campaign = unique(campaign$campaign),
               cac = sample(c(10:15), 7, replace = TRUE))
campaign <- merge(campaign, cac, by = 'campaign')
potential <- data.frame(clientId = c(1:1500),
                        clv.p = sample(c(0:50), 1500, replace = TRUE))
rm(cac)

# reporting date
today <- as.Date('2013-05-16', format = '%Y-%m-%d')
```

The four auxiliary tables play distinct roles: *campaign* carries each customer's
acquisition campaign and its acquisition cost; *margin* gives the gross margin per
product; *potential* holds the predicted CLV per client; and *orders* records the
transactions. The next block computes the per-client features and cohort labels.

```{r}
# calculating CLV, frequency, recency, average time lapses between purchases and defining cohorts

orders <- merge(orders, gr.margin, by = 'product')

customers <- orders %>%
    # combining products and summarising gross margin
    group_by(orderId, clientId, orderdate) %>%
    summarise(grossmarg = sum(grossmarg)) %>%
    ungroup() %>%
    # calculating frequency, recency, average time lapses between purchases and defining cohorts
    group_by(clientId) %>%
    mutate(
        frequency = n(),
        recency = as.numeric(today - max(orderdate)),
        av.gap = round(as.numeric(max(orderdate) - min(orderdate)) / frequency, 0),
        cohort = format(min(orderdate), format = '%Y-%m')
    ) %>%
    ungroup() %>%
    # calculating CLV to date
    group_by(clientId, cohort, frequency, recency, av.gap) %>%
    summarise(clv = sum(grossmarg)) %>%
    arrange(clientId) %>%
    ungroup()
# calculating potential CLV and CAC
customers <- merge(customers, campaign, by = 'clientId')
customers <- merge(customers, potential, by = 'clientId')
# leading the potential value to more or less real value
customers$clv.p <-
    round(customers$clv.p / sqrt(customers$recency) * customers$frequency,
          2)

rm(potential, gr.margin, today)
# adding segments
customers <- customers %>%
    mutate(segm.freq = ifelse(between(frequency, 1, 1), '1',
                              ifelse(
                                  between(frequency, 2, 2), '2',
                                  ifelse(between(frequency, 3, 3), '3',
                                         ifelse(
                                             between(frequency, 4, 4), '4',
                                             ifelse(between(frequency, 5, 5), '5', '>5')
                                         ))
                              ))) %>%
    mutate(segm.rec = ifelse(
        between(recency, 0, 30),
        '0-30 days',
        ifelse(
            between(recency, 31, 60),
            '31-60 days',
            ifelse(
                between(recency, 61, 90),
                '61-90 days',
                ifelse(
                    between(recency, 91, 120),
                    '91-120 days',
                    ifelse(between(recency, 121, 180), '121-180 days', '>180 days')
                )
            )
        )
    ))

# defining order of boundaries
customers$segm.freq <-
    factor(customers$segm.freq, levels = c('>5', '5', '4', '3', '2', '1'))
customers$segm.rec <-
    factor(
        customers$segm.rec,
        levels = c(
            '>180 days',
            '121-180 days',
            '91-120 days',
            '61-90 days',
            '31-60 days',
            '0-30 days'
        )
    )
```

##### First-Purchase-Date Cohorts {#sec-fpd-cohorts}

Grouping by first-purchase month and laying the result on the lifecycle grid lets the
analyst compare cohorts on profitability (average CLV minus CAC), on combined realized
plus potential value, and on purchasing pace. The customer-flow analysis tracks how a
single cohort redistributes across recency–frequency states at successive reporting
dates and renders the migration as a Sankey diagram.

```{r, eval = FALSE}
lcg.coh <- customers %>%
    group_by(cohort, segm.rec, segm.freq) %>%
    # calculating cumulative values
    summarise(
        quantity = n(),
        cac = sum(cac),
        clv = sum(clv),
        clv.p = sum(clv.p),
        av.gap = sum(av.gap)
    ) %>%
    ungroup() %>%
    # calculating average values
    mutate(
        av.cac = round(cac / quantity, 2),
        av.clv = round(clv / quantity, 2),
        av.clv.p = round(clv.p / quantity, 2),
        av.clv.tot = av.clv + av.clv.p,
        av.gap = round(av.gap / quantity, 2),
        diff = av.clv - av.cac
    )

# 1. Structure of averages and comparison cohorts

ggplot(lcg.coh, aes(x = cohort, fill = cohort)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    geom_bar(aes(y = diff), stat = 'identity', alpha = 0.5) +
    geom_text(aes(y = diff, label = round(diff, 0)), size = 4) +
    facet_grid(segm.freq ~ segm.rec) +
    theme(axis.text.x = element_text(
        angle = 90,
        hjust = .5,
        vjust = .5,
        face = "plain"
    )) +
    ggtitle("Cohorts in LifeCycle Grids - difference between av.CLV to date and av.CAC")

ggplot(lcg.coh, aes(x = cohort, fill = cohort)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    geom_bar(aes(y = av.clv.tot), stat = 'identity', alpha = 0.2) +
    geom_text(aes(
        y = av.clv.tot + 10,
        label = round(av.clv.tot, 0),
        color = cohort
    ), size = 4) +
    geom_bar(aes(y = av.clv), stat = 'identity', alpha = 0.7) +
    geom_errorbar(aes(y = av.cac, ymax = av.cac, ymin = av.cac),
                  color = 'red',
                  size = 1.2) +
    geom_text(
        aes(y = av.cac, label = round(av.cac, 0)),
        size = 4,
        color = 'darkred',
        vjust = -.5
    ) +
    facet_grid(segm.freq ~ segm.rec) +
    theme(axis.text.x = element_text(
        angle = 90,
        hjust = .5,
        vjust = .5,
        face = "plain"
    )) +
    ggtitle("Cohorts in LifeCycle Grids - total av.CLV and av.CAC")

# 2. Analyzing customer flows
# customers flows analysis (FPD cohorts)

# defining cohort and reporting dates
coh <- '2012-09'
report.dates <- c('2012-10-01', '2013-01-01', '2013-04-01')
report.dates <- as.Date(report.dates, format = '%Y-%m-%d')

# defining segments for each cohort's customer for reporting dates
df.sankey <- data.frame()

for (i in 1:length(report.dates)) {
    orders.cache <- orders %>%
        filter(orderdate < report.dates[i])

    customers.cache <- orders.cache %>%
        select(-product,-grossmarg) %>%
        unique() %>%
        group_by(clientId) %>%
        mutate(
            frequency = n(),
            recency = as.numeric(report.dates[i] - max(orderdate)),
            cohort = format(min(orderdate), format = '%Y-%m')
        ) %>%
        ungroup() %>%
        select(clientId, frequency, recency, cohort) %>%
        unique() %>%
        filter(cohort == coh) %>%
        mutate(segm.freq = ifelse(
            between(frequency, 1, 1),
            '1 purch',
            ifelse(
                between(frequency, 2, 2),
                '2 purch',
                ifelse(
                    between(frequency, 3, 3),
                    '3 purch',
                    ifelse(
                        between(frequency, 4, 4),
                        '4 purch',
                        ifelse(between(frequency, 5, 5), '5 purch', '>5 purch')
                    )
                )
            )
        )) %>%
        mutate(segm.rec = ifelse(
            between(recency, 0, 30),
            '0-30 days',
            ifelse(
                between(recency, 31, 60),
                '31-60 days',
                ifelse(
                    between(recency, 61, 90),
                    '61-90 days',
                    ifelse(
                        between(recency, 91, 120),
                        '91-120 days',
                        ifelse(between(recency, 121, 180), '121-180 days', '>180 days')
                    )
                )
            )
        )) %>%
        mutate(
            cohort.segm = paste(cohort, segm.rec, segm.freq, sep = ' : '),
            report.date = report.dates[i]
        ) %>%
        select(clientId, cohort.segm, report.date)

    df.sankey <- rbind(df.sankey, customers.cache)
}

# processing data for Sankey diagram format
df.sankey <-
    dcast(df.sankey,
          clientId ~ report.date,
          value.var = 'cohort.segm',
          fun.aggregate = NULL)
write.csv(df.sankey, 'customers_path.csv', row.names = FALSE)
df.sankey <- df.sankey %>% select(-clientId)

df.sankey.plot <- data.frame()
for (i in 2:ncol(df.sankey)) {
    df.sankey.cache <- df.sankey %>%
        group_by(df.sankey[, i - 1], df.sankey[, i]) %>%
        summarise(n = n()) %>%
        ungroup()

    colnames(df.sankey.cache)[1:2] <- c('from', 'to')

    df.sankey.cache$from <-
        paste(df.sankey.cache$from, ' (', report.dates[i - 1], ')', sep = '')
    df.sankey.cache$to <-
        paste(df.sankey.cache$to, ' (', report.dates[i], ')', sep = '')

    df.sankey.plot <- rbind(df.sankey.plot, df.sankey.cache)
}

# plotting
make_sankey(df.sankey.plot)

# purchasing pace

ggplot(lcg.coh, aes(x = cohort, fill = cohort)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    geom_bar(aes(y = av.gap), stat = 'identity', alpha = 0.6) +
    geom_text(aes(y = av.gap, label = round(av.gap, 0)), size = 4) +
    facet_grid(segm.freq ~ segm.rec) +
    theme(axis.text.x = element_text(
        angle = 90,
        hjust = .5,
        vjust = .5,
        face = "plain"
    )) +
    ggtitle("Cohorts in LifeCycle Grids - average time lapses between purchases")
```

##### Campaign Cohorts {#sec-campaign-cohorts}

Reindexing the same machinery by *acquisition campaign* rather than first-purchase
month answers a different question: which campaigns recruited customers whose realized
value justifies their acquisition cost, and at what purchasing pace. The plots below
report, by campaign, the CLV-minus-CAC gap, total
realized-plus-potential value against CAC, and the average gap between purchases.

```{r}
#| label: fig-campaign-cohorts
#| fig-cap: "Campaign cohorts on the lifecycle grid: profitability gap, total value against acquisition cost, and purchasing pace by acquisition campaign."
# campaign cohorts
lcg.camp <- customers %>%
    group_by(campaign, segm.rec, segm.freq) %>%
    # calculating cumulative values
    summarise(
        quantity = n(),
        cac = sum(cac),
        clv = sum(clv),
        clv.p = sum(clv.p),
        av.gap = sum(av.gap)
    ) %>%
    ungroup() %>%
    # calculating average values
    mutate(
        av.cac = round(cac / quantity, 2),
        av.clv = round(clv / quantity, 2),
        av.clv.p = round(clv.p / quantity, 2),
        av.clv.tot = av.clv + av.clv.p,
        av.gap = round(av.gap / quantity, 2),
        diff = av.clv - av.cac
    )

ggplot(lcg.camp, aes(x = campaign, fill = campaign)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    geom_bar(aes(y = diff), stat = 'identity', alpha = 0.5) +
    geom_text(aes(y = diff, label = round(diff, 0)), size = 4) +
    facet_grid(segm.freq ~ segm.rec) +
    theme(axis.text.x = element_text(
        angle = 90,
        hjust = .5,
        vjust = .5,
        face = "plain"
    )) +
    ggtitle("Campaigns in LifeCycle Grids - difference between av.CLV to date and av.CAC")

ggplot(lcg.camp, aes(x = campaign, fill = campaign)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    geom_bar(aes(y = av.clv.tot), stat = 'identity', alpha = 0.2) +
    geom_text(aes(
        y = av.clv.tot + 10,
        label = round(av.clv.tot, 0),
        color = campaign
    ), size = 4) +
    geom_bar(aes(y = av.clv), stat = 'identity', alpha = 0.7) +
    geom_errorbar(aes(y = av.cac, ymax = av.cac, ymin = av.cac),
                  color = 'red',
                  size = 1.2) +
    geom_text(
        aes(y = av.cac, label = round(av.cac, 0)),
        size = 4,
        color = 'darkred',
        vjust = -.5
    ) +
    facet_grid(segm.freq ~ segm.rec) +
    theme(axis.text.x = element_text(
        angle = 90,
        hjust = .5,
        vjust = .5,
        face = "plain"
    )) +
    ggtitle("Campaigns in LifeCycle Grids - total av.CLV and av.CAC")

ggplot(lcg.camp, aes(x = campaign, fill = campaign)) +
    theme_bw() +
    theme(panel.grid = element_blank()) +
    geom_bar(aes(y = av.gap), stat = 'identity', alpha = 0.6) +
    geom_text(aes(y = av.gap, label = round(av.gap, 0)), size = 4) +
    facet_grid(segm.freq ~ segm.rec) +
    theme(axis.text.x = element_text(
        angle = 90,
        hjust = .5,
        vjust = .5,
        face = "plain"
    )) +
    ggtitle("Campaigns in LifeCycle Grids - average time lapses between purchases")
```

##### Retention Rate {#sec-retention-rate}

The **retention rate** is the share of a cohort still active after a given lifetime
interval. For cohort $c$ with $N_{c,0}$ members at acquisition and $N_{c,\tau}$ active
at lifetime $\tau$, the retention rate is $N_{c,\tau}/N_{c,0}$, and the
period-over-period rate is $N_{c,\tau}/N_{c,\tau-1}$. Plotting retention as a cycle
plot, with seasonality (a simulated Black Friday bump) baked into the data, separates
genuine cohort-quality differences from calendar effects, as the cycle plots below show.

```{r}
#| label: fig-retention-cycle
#| fig-cap: "Customer retention rate as a cycle plot and as bars of cohort size by year of lifetime, with simulated seasonality. Smoothed trends separate cohort-quality differences from calendar effects."
# loading libraries
library(dplyr)
library(reshape2)
library(ggplot2)
library(scales)
library(gridExtra)
# creating data sample
set.seed(10)
cohorts <-
    data.frame(
        cohort = paste('cohort', formatC(
            c(1:36),
            width = 2,
            format = 'd',
            flag = '0'
        ), sep = '_'),
        Y_00 = sample(c(1300:1500), 36, replace = TRUE),
        Y_01 = c(sample(c(800:1000), 36, replace = TRUE)),
        Y_02 = c(sample(c(600:800), 24, replace = TRUE), rep(NA, 12)),
        Y_03 = c(sample(c(400:500), 12, replace = TRUE), rep(NA, 24))
    )
# simulating seasonality (Black Friday)
cohorts[c(11, 23, 35), 2] <-
    as.integer(cohorts[c(11, 23, 35), 2] * 1.25)
cohorts[c(11, 23, 35), 3] <-
    as.integer(cohorts[c(11, 23, 35), 3] * 1.10)
cohorts[c(11, 23, 35), 4] <-
    as.integer(cohorts[c(11, 23, 35), 4] * 1.07)

# calculating retention rate and preparing data for plotting
df_plot <-
    reshape2::melt(
        cohorts,
        id.vars = 'cohort',
        value.name = 'number',
        variable.name = "year_of_LT"
    )

df_plot <- df_plot %>%
    group_by(cohort) %>%
    arrange(year_of_LT) %>%
    mutate(number_prev_year = lag(number),
           number_Y_00 = number[which(year_of_LT == 'Y_00')]) %>%
    ungroup() %>%
    mutate(
        ret_rate_prev_year = number / number_prev_year,
        ret_rate = number / number_Y_00,
        year_cohort = paste(year_of_LT, cohort, sep = '-')
    )

##### The first way for plotting cycle plot via scaling
# calculating the coefficient for scaling 2nd axis
k <-
    max(df_plot$number_prev_year[df_plot$year_of_LT == 'Y_01'] * 1.15) / min(df_plot$ret_rate[df_plot$year_of_LT == 'Y_01'])

# retention rate cycle plot
ggplot(
    na.omit(df_plot),
    aes(
        x = year_cohort,
        y = ret_rate,
        group = year_of_LT,
        color = year_of_LT
    )
) +
    theme_bw() +
    geom_point(size = 4) +
    geom_text(
        aes(label = percent(round(ret_rate, 2))),
        size = 4,
        hjust = 0.4,
        vjust = -0.6,
        fontface = "plain"
    ) +
    # smooth method can be changed (e.g. for "lm")
    geom_smooth(
        size = 2.5,
        method = 'loess',
        color = 'darkred',
        aes(fill = year_of_LT)
    ) +
    geom_bar(aes(y = number_prev_year / k, fill = year_of_LT),
             alpha = 0.2,
             stat = 'identity') +
    geom_bar(aes(y = number / k, fill = year_of_LT),
             alpha = 0.6,
             stat = 'identity') +
    geom_text(
        aes(y = 0, label = cohort),
        color = 'white',
        angle = 90,
        size = 4,
        hjust = -0.05,
        vjust = 0.4
    ) +
    geom_text(
        aes(y = number_prev_year / k, label = number_prev_year),
        angle = 90,
        size = 4,
        hjust = -0.1,
        vjust = 0.4
    ) +
    geom_text(
        aes(y = number / k, label = number),
        angle = 90,
        size = 4,
        hjust = -0.1,
        vjust = 0.4
    ) +
    theme(
        legend.position = 'none',
        plot.title = element_text(size = 20, face = "bold", vjust = 2),
        axis.title.x = element_text(size = 18, face = "bold"),
        axis.title.y = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    ) +
    labs(x = 'Year of Lifetime by Cohorts', y = 'Number of Customers / Retention Rate') +
    ggtitle("Customer Retention Rate - Cycle plot")

##### The second way for plotting cycle plot via multi-plotting
# plot #1 - Retention rate
p1 <-
    ggplot(
        na.omit(df_plot),
        aes(
            x = year_cohort,
            y = ret_rate,
            group = year_of_LT,
            color = year_of_LT
        )
    ) +
    theme_bw() +
    geom_point(size = 4) +
    geom_text(
        aes(label = percent(round(ret_rate, 2))),
        size = 4,
        hjust = 0.4,
        vjust = -0.6,
        fontface = "plain"
    ) +
    geom_smooth(
        size = 2.5,
        method = 'loess',
        color = 'darkred',
        aes(fill = year_of_LT)
    ) +
    theme(
        legend.position = 'none',
        plot.title = element_text(size = 20, face = "bold", vjust = 2),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 18, face = "bold"),
        axis.text = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    ) +
    labs(y = 'Retention Rate') +
    ggtitle("Customer Retention Rate - Cycle plot")

# plot #2 - number of customers
p2 <-
    ggplot(na.omit(df_plot),
           aes(x = year_cohort, group = year_of_LT, color = year_of_LT)) +
    theme_bw() +
    geom_bar(aes(y = number_prev_year, fill = year_of_LT),
             alpha = 0.2,
             stat = 'identity') +
    geom_bar(aes(y = number, fill = year_of_LT),
             alpha = 0.6,
             stat = 'identity') +
    geom_text(
        aes(y = number_prev_year, label = number_prev_year),
        angle = 90,
        size = 4,
        hjust = -0.1,
        vjust = 0.4
    ) +
    geom_text(
        aes(y = number, label = number),
        angle = 90,
        size = 4,
        hjust = -0.1,
        vjust = 0.4
    ) +
    geom_text(
        aes(y = 0, label = cohort),
        color = 'white',
        angle = 90,
        size = 4,
        hjust = -0.05,
        vjust = 0.4
    ) +
    theme(
        legend.position = 'none',
        plot.title = element_text(size = 20, face = "bold", vjust = 2),
        axis.title.x = element_text(size = 18, face = "bold"),
        axis.title.y = element_text(size = 18, face = "bold"),
        axis.text = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    ) +
    scale_y_continuous(limits = c(0, max(df_plot$number_Y_00 * 1.1))) +
    labs(x = 'Year of Lifetime by Cohorts', y = 'Number of Customers')

# multiplot
grid.arrange(p1, p2, ncol = 1)
# retention rate bubble chart
ggplot(na.omit(df_plot),
       aes(
           x = cohort,
           y = ret_rate,
           group = cohort,
           color = year_of_LT
       )) +
    theme_bw() +
    scale_size(range = c(15, 40)) +
    geom_line(size = 2, alpha = 0.3) +
    geom_point(aes(size = number_prev_year), alpha = 0.3) +
    geom_point(aes(size = number), alpha = 0.8) +
    geom_smooth(
        linetype = 2,
        size = 2,
        method = 'loess',
        aes(group = year_of_LT, fill = year_of_LT),
        alpha = 0.2
    ) +
    geom_text(
        aes(label = paste0(
            number, '/', number_prev_year, '\n', percent(round(ret_rate, 2))
        )),
        color = 'white',
        size = 3,
        hjust = 0.5,
        vjust = 0.5,
        fontface = "plain"
    ) +
    theme(
        legend.position = 'none',
        plot.title = element_text(size = 20, face = "bold", vjust = 2),
        axis.title.x = element_text(size = 18, face = "bold"),
        axis.title.y = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        axis.text.x = element_text(
            size = 10,
            angle = 90,
            hjust = .5,
            vjust = .5,
            face = "plain"
        ),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    ) +
    labs(x = 'Cohorts', y = 'Retention Rate by Year of Lifetime') +
    ggtitle("Customer Retention Rate - Bubble chart")

# retention rate falling drops chart
ggplot(df_plot,
       aes(
           x = cohort,
           y = ret_rate,
           group = cohort,
           color = year_of_LT
       )) +
    theme_bw() +
    scale_size(range = c(15, 40)) +
    scale_y_continuous(limits = c(0, 1)) +
    geom_line(size = 2, alpha = 0.3) +
    geom_point(aes(size = number), alpha = 0.8) +
    geom_text(
        aes(label = paste0(number, '\n', percent(round(
            ret_rate, 2
        )))),
        color = 'white',
        size = 3,
        hjust = 0.5,
        vjust = 0.5,
        fontface = "plain"
    ) +
    theme(
        legend.position = 'none',
        plot.title = element_text(size = 20, face = "bold", vjust = 2),
        axis.title.x = element_text(size = 18, face = "bold"),
        axis.title.y = element_text(size = 18, face = "bold"),
        axis.text = element_text(size = 16),
        axis.text.x = element_text(
            size = 10,
            angle = 90,
            hjust = .5,
            vjust = .5,
            face = "plain"
        ),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    ) +
    labs(x = 'Cohorts', y = 'Retention Rate by Year of Lifetime') +
    ggtitle("Customer Retention Rate - Falling Drops chart")
```

##### Retention Charts {#sec-retention-charts}

The canonical **retention triangle** arranges cohorts in rows and lifetime months in
columns, so that each diagonal corresponds to a calendar period and each row traces a
cohort's decay. The panels below assemble these views. Normalizing every row by
its first-month count converts raw counts to retention ratios; the dynamics,
second-month, and cycle-plot views then interrogate the triangle from complementary
angles.

```{r}
#| label: fig-retention-triangle
#| fig-cap: "Retention triangle analyses: ratio dynamics across lifetime months, a single-month cross-cohort comparison, and a cycle plot highlighting two reference cohorts."
# libraries
library(dplyr)
library(ggplot2)
library(reshape2)

cohort.clients <- data.frame(
    cohort = c(
        'Cohort01',
        'Cohort02',
        'Cohort03',
        'Cohort04',
        'Cohort05',
        'Cohort06',
        'Cohort07',
        'Cohort08',
        'Cohort09',
        'Cohort10',
        'Cohort11',
        'Cohort12'
    ),
    M01 = c(11000, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
    M02 = c(1900, 10000, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
    M03 = c(1400, 2000, 11500, 0, 0, 0, 0, 0, 0, 0, 0, 0),
    M04 = c(1100, 1300, 2400, 13200, 0, 0, 0, 0, 0, 0, 0, 0),
    M05 = c(1000, 1100, 1400, 2400, 11100, 0, 0, 0, 0, 0, 0, 0),
    M06 = c(900, 900, 1200, 1600, 1900, 10300, 0, 0, 0, 0, 0, 0),
    M07 = c(850, 900, 1100, 1300, 1300, 1900, 13000, 0, 0, 0, 0, 0),
    M08 = c(850, 850, 1000, 1200, 1100, 1300, 1900, 11500, 0, 0, 0, 0),
    M09 = c(800, 800, 950, 1100, 1100, 1250, 1000, 1200, 11000, 0, 0, 0),
    M10 = c(800, 780, 900, 1050, 1050, 1200, 900, 1200, 1900, 13200, 0, 0),
    M11 = c(750, 750, 900, 1000, 1000, 1180, 800, 1100, 1150, 2000, 11300, 0),
    M12 = c(740, 700, 870, 1000, 900, 1100, 700, 1050, 1025, 1300, 1800, 20000)
)

cohort.clients.r <- cohort.clients #create new data frame
totcols <-
    ncol(cohort.clients.r) #count number of columns in data set
for (i in 1:nrow(cohort.clients.r)) {
    #for loop for shifting each row
    df <- cohort.clients.r[i,] #select row from data frame
    df <- df[, !df[] == 0] #remove columns with zeros
    partcols <-
        ncol(df) #count number of columns in row (w/o zeros)
    #fill columns after values by zeros
    if (partcols < totcols)
        df[, c((partcols + 1):totcols)] <- 0
    cohort.clients.r[i,] <- df #replace initial row by new one
}
# Retention ratio = # clients in particular month / # clients in 1st month of life-time

#calculate retention (1)
x <- cohort.clients.r[, c(2:13)]
y <- cohort.clients.r[, 2]
reten.r <- apply(x, 2, function(x)
    x / y)
reten.r <- data.frame(cohort = (cohort.clients.r$cohort), reten.r)

#calculate retention (2)
c <- ncol(cohort.clients.r)
reten.r <- cohort.clients.r
for (i in 2:c) {
    reten.r[, (c + i - 1)] <- reten.r[, i] / reten.r[, 2]
}
reten.r <- reten.r[,-c(2:c)]
colnames(reten.r) <- colnames(cohort.clients.r)




#charts
reten.r <- reten.r[,-2] #remove M01 data because it is always 100%
#dynamics analysis chart
cohort.chart1 <- melt(reten.r, id.vars = 'cohort')
colnames(cohort.chart1) <- c('cohort', 'month', 'retention')
cohort.chart1 <- filter(cohort.chart1, retention != 0)
p <-
    ggplot(cohort.chart1,
           aes(
               x = month,
               y = retention,
               group = cohort,
               colour = cohort
           ))
p + geom_line(size = 2, alpha = 1 / 2) +
    geom_point(size = 3, alpha = 1) +
    geom_smooth(
        aes(group = 1),
        method = 'loess',
        size = 2,
        colour = 'red',
        se = FALSE
    ) +
    labs(title = "Cohorts Retention ratio dynamics")

#second month analysis chart
cohort.chart2 <-
    filter(cohort.chart1, month == 'M02') #choose any month instead of M02
p <-
    ggplot(cohort.chart2, aes(x = cohort, y = retention, colour = cohort))
p + geom_point(size = 3) +
    geom_line(aes(group = 1), size = 2, alpha = 1 / 2) +
    geom_smooth(
        aes(group = 1),
        size = 2,
        colour = 'red',
        method = 'lm',
        se = FALSE
    ) +
    labs(title = "Cohorts Retention ratio for 2nd month")
#cycle plot
cohort.chart3 <- cohort.chart1
cohort.chart3 <-
    mutate(cohort.chart3, month_cohort = paste(month, cohort))
p <-
    ggplot(cohort.chart3,
           aes(
               x = month_cohort,
               y = retention,
               group = month,
               colour = month
           ))
#choose any cohorts instead of Cohort07 and Cohort06
m1 <- filter(cohort.chart3, cohort == 'Cohort07')
m2 <- filter(cohort.chart3, cohort == 'Cohort06')
p + geom_point(size = 3) +
    geom_line(aes(group = month), size = 2, alpha = 1 / 2) +
    labs(title = "Cohorts Retention ratio cycle plot") +
    geom_line(
        data = m1,
        aes(group = 1),
        colour = 'blue',
        size = 2,
        alpha = 1 / 5
    ) +
    geom_line(
        data = m2,
        aes(group = 1),
        colour = 'blue',
        size = 2,
        alpha = 1 / 5
    ) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

#### Lifecycle-Phase Sequence Analysis {#sec-lifecycle-sequence}

The final lifecycle device borrows **sequence analysis** from the social sciences to
study the *order* in which customers pass through lifecycle states (new, engaged,
best, and their at-risk and former counterparts). The goal is to identify the path
patterns characteristic of each cohort and to surface the cohorts that acquired
customers on the trajectory the firm prefers, so offers can be targeted accordingly.
Each customer is assigned a lifecycle state at a sequence of reporting dates, the
states are encoded as a sequence object with `TraMineR::seqdef()`, and the most
frequent sequences are plotted by campaign and by first-purchase cohort.

```{r, eval=FALSE}
library(TraMineR)

min.date <- min(orders$orderdate)
max.date <- max(orders$orderdate)

l <-
    c(seq(0, as.numeric(max.date - min.date), 10), as.numeric(max.date - min.date))

df <- data.frame()
for (i in l) {
    cur.date <- min.date + i
    print(cur.date)

    orders.cache <- orders %>%
        filter(orderdate <= cur.date)

    customers.cache <- orders.cache %>%
        select(-product,-grossmarg) %>%
        unique() %>%
        group_by(clientId) %>%
        mutate(frequency = n(),
               recency = as.numeric(cur.date - max(orderdate))) %>%
        ungroup() %>%
        select(clientId, frequency, recency) %>%
        unique() %>%

        mutate(segm =
                   ifelse(
                       between(frequency, 1, 2) & between(recency, 0, 60),
                       'new customer',
                       ifelse(
                           between(frequency, 1, 2) &
                               between(recency, 61, 180),
                           'under risk new customer',
                           ifelse(
                               between(frequency, 1, 2) & recency > 180,
                               '1x buyer',

                               ifelse(
                                   between(frequency, 3, 4) &
                                       between(recency, 0, 60),
                                   'engaged customer',
                                   ifelse(
                                       between(frequency, 3, 4) &
                                           between(recency, 61, 180),
                                       'under risk engaged customer',
                                       ifelse(
                                           between(frequency, 3, 4) & recency > 180,
                                           'former engaged customer',

                                           ifelse(
                                               frequency > 4 & between(recency, 0, 60),
                                               'best customer',
                                               ifelse(
                                                   frequency > 4 &
                                                       between(recency, 61, 180),
                                                   'under risk best customer',
                                                   ifelse(frequency > 4 &
                                                              recency > 180, 'former best customer', NA)
                                               )
                                           )
                                       )
                                   )
                               )
                           )
                       )
                   )) %>%

        mutate(report.date = i) %>%
        select(clientId, segm, report.date)

    df <- rbind(df, customers.cache)
}

# converting data to the sequence format
df <-
    dcast(df,
          clientId ~ report.date,
          value.var = 'segm',
          fun.aggregate = NULL)
df.seq <- seqdef(df,
                 2:ncol(df),
                 left = 'DEL',
                 right = 'DEL',
                 xtstep = 10)

# creating df with first purch.date and campaign cohort features
feat <- df %>% select(clientId)
feat <- merge(feat, campaign[, 1:2], by = 'clientId')
feat <- merge(feat, customers[, 1:2], by = 'clientId')

par(mar = c(1, 1, 1, 1))

# plotting the 10 most frequent sequences based on campaign
seqfplot(df.seq, border = NA, group = feat$campaign)

# plotting the 10 most frequent sequences based on campaign
seqfplot(
    df.seq,
    border = NA,
    group = feat$campaign,
    cex.legend = 0.9
)

# plotting the 10 most frequent sequences based on first purch.date cohort
coh.list <- sort(unique(feat$cohort))
# defining cohorts for plotting
feat.coh.list <- feat[feat$cohort %in% coh.list[1:6] ,]
df.coh <- df %>% filter(clientId %in% c(feat.coh.list$clientId))
df.seq.coh <-
    seqdef(
        df.coh,
        2:ncol(df.coh),
        left = 'DEL',
        right = 'DEL',
        xtstep = 10
    )
seqfplot(
    df.seq.coh,
    border = NA,
    group = feat.coh.list$cohort,
    cex.legend = 0.9
)
```

## Basket and Sequence Analysis {#sec-basket}

The models so far summarize *whether* and *when* customers buy. Basket analysis turns
to *what* they buy together, and sequence analysis to *in what order* they buy across
visits. Both inform cross-sell, merchandising, and churn detection.

### Multi-Layer Pie Chart of Baskets {#sec-basket-pie}

The first device, following Sergey Bryl, visualizes the co-occurrence structure of
products within carts as a set of concentric rings—an innermost ring for
single-product carts and outer rings for each additional product—so that the most
common product combinations are read off as the widest arcs. We simulate orders of up
to five products and reduce each order to a unique set of products.

```{r}
# loading libraries
library(dplyr)
library(tidyverse)
library(reshape2)
library(plotrix)

# Simulate of orders
set.seed(15)
df <- data.frame(
    orderId = sample(c(1:1000), 5000, replace = TRUE),
    product = sample(
        c('NULL', 'a', 'b', 'c', 'd'),
        5000,
        replace = TRUE,
        prob = c(0.15, 0.65, 0.3, 0.15, 0.1)
    )
)
df <- df[df$product != 'NULL',]
head(df)
```

Each order is collapsed into a delimited *cart* string with a count of distinct
products, and carts are tallied, so the analysis operates on cart *types* rather than
individual orders.

```{r}
# processing initial data
# we need to be sure that product's names are unique
df$product <- paste0("#", df$product, "#")

prod.matrix <- df %>%
    # removing duplicated products from each order (exclude the effect of quantity)
    group_by(orderId, product) %>%
    arrange(product) %>%
    unique() %>%
    # combining products to cart and calculating number of products
    group_by(orderId) %>%
    summarise(cart = paste(product, collapse = ";"),
              prod.num = n()) %>%
    # calculating number of carts
    group_by(cart, prod.num) %>%
    summarise(num = n()) %>%
    ungroup()

head(prod.matrix)
```

Splitting single-product from multi-product carts sets up the layered geometry of
@fig-basket-pie: the inner ring is reserved for one-product carts and successive outer
rings encode the presence of each additional product.

```{r}
# calculating total number of orders/carts
tot <- sum(prod.matrix$num)

# spliting orders for sets with 1 product and more than 1 product
one.prod <- prod.matrix %>% filter(prod.num == 1)

sev.prod <- prod.matrix %>%
    filter(prod.num > 1) %>%
    arrange(desc(prod.num))
```

```{r}
# defining parameters for pie chart
iniR <- 0.2 # initial radius
cols <- c(
    "#ffffff",
    "#fec44f",
    "#fc9272",
    "#a1d99b",
    "#fee0d2",
    "#2ca25f",
    "#8856a7",
    "#43a2ca",
    "#fdbb84",
    "#e34a33",
    "#a6bddb",
    "#dd1c77",
    "#ffeda0",
    "#756bb1"
)
prod <- df %>%
    select(product) %>%
    arrange(product) %>%
    unique()
prod <- c('NO', c(prod$product))
colors <- as.list(setNames(cols[c(1:(length(prod)))], prod))
```

```{r}
#| label: fig-basket-pie
#| fig-cap: "Multi-layer pie chart of shopping carts. The inner ring shows single-product carts; each outer ring encodes the presence of an additional product, so the widest arcs are the most common combinations."
# 0 circle: blank
pie(
    1,
    radius = iniR,
    init.angle = 90,
    col = c('white'),
    border = NA,
    labels = ''
)

# drawing circles from last to 2nd
for (i in length(prod):2) {
    p <- grep(prod[i], sev.prod$cart)
    col <- rep('NO', times = nrow(sev.prod))
    col[p] <- prod[i]
    floating.pie(
        0,
        0,
        c(sev.prod$num, tot - sum(sev.prod$num)),
        radius = (1 + i) * iniR,
        startpos = pi / 2,
        col = as.character(colors [c(col, 'NO')]),
        border = "#44aaff"
    )
}

# 1 circle: orders with 1 product
floating.pie(
    0,
    0,
    c(tot - sum(one.prod$num), one.prod$num),
    radius = 2 * iniR,
    startpos = pi / 2,
    col = as.character(colors [c('NO', one.prod$cart)]),
    border = "#44aaff"
)

# legend
legend(
    1.5,
    2 * iniR,
    gsub("_", " ", names(colors)[-1]),
    col = as.character(colors [-1]),
    pch = 19,
    bty = 'n',
    ncol = 1
)
```

A companion statistics table records each cart type's count and share, ranking the
combinations by prevalence.

```{r}
# creating a table with the stats
stat.tab <- prod.matrix %>%
    select(-prod.num) %>%
    mutate(share = num / tot) %>%
    arrange(desc(num))

library(scales)
stat.tab$share <-
    percent(stat.tab$share) # converting values to percents
```

### Sankey Diagram of Order Sequences {#sec-basket-sankey}

A **Sankey diagram** visualizes flows: here, how customers move from one order's cart
composition to the next across successive purchases, with band width proportional to
the number of customers following each transition. Carts are simulated, concatenated
into per-order strings, pivoted so each customer's successive orders sit in adjacent
columns, and the transitions between consecutive orders are tallied as flow weights and
rendered in @fig-basket-sankey.

```{r}
#| label: fig-basket-sankey
#| fig-cap: "Sankey diagram of cart composition across successive orders. Band width is proportional to the number of customers making each cart-to-cart transition."
# loading libraries
library(networkD3)
library(dplyr)
library(reshape2)

# creating an example of orders
set.seed(15)
df <- data.frame(
    orderId = c(1:1000),
    clientId = sample(c(1:300), 1000, replace = TRUE),
    prod1 = sample(
        c('NULL', 'a'),
        1000,
        replace = TRUE,
        prob = c(0.15, 0.5)
    ),
    prod2 = sample(
        c('NULL', 'b'),
        1000,
        replace = TRUE,
        prob = c(0.15, 0.3)
    ),
    prod3 = sample(
        c('NULL', 'c'),
        1000,
        replace = TRUE,
        prob = c(0.15, 0.2)
    )
)

# combining products
df$cart <- paste(df$prod1, df$prod2, df$prod3, sep = ';')
df$cart <- gsub('NULL;|;NULL', '', df$cart)
df <- df[df$cart != 'NULL',]

df <- df %>%
    select(orderId, clientId, cart) %>%
    arrange(clientId, orderId, cart)

head(df)

orders <- df %>%
    group_by(clientId) %>%
    mutate(n.ord = paste('ord', c(1:n()), sep = '')) %>%
    ungroup()

head(orders)

orders <-
    dcast(orders,
          clientId ~ n.ord,
          value.var = 'cart',
          fun.aggregate = NULL)


# choose a number of carts/orders in the sequence we want to analyze
orders <- orders %>%
    select(ord1, ord2, ord3, ord4, ord5)

orders.plot <- data.frame()

for (i in 2:ncol(orders)) {
    ord.cache <- orders %>%
        group_by(orders[, i - 1], orders[, i]) %>%
        summarise(n = n()) %>%
        ungroup()

    colnames(ord.cache)[1:2] <- c('from', 'to')

    # adding tags to carts
    ord.cache$from <- paste(ord.cache$from, '(', i - 1, ')', sep = '')
    ord.cache$to <- paste(ord.cache$to, '(', i, ')', sep = '')

    orders.plot <- rbind(orders.plot, ord.cache)

}

make_sankey(orders.plot)
```

### In-Depth Sequence Analysis {#sec-basket-sequence}

Sequence analysis proper, again following Bryl, treats each customer's ordered series
of carts (interleaved with a synthetic *no-purchase* spell to mark dormancy) as a
categorical sequence amenable to the descriptive and information-theoretic tools of
`TraMineR`. The aim is to understand purchasing behavior and churn: whether a customer
has lapsed, and how long the gaps between purchases run. We simulate orders with
products, clients, gender, and dates, then build per-client cart sequences and append
a no-purchase spell whose length is keyed to the customer's own average inter-purchase
interval.

```{r}
library(dplyr)
library(TraMineR)
library(reshape2)
library(networkD3)

# creating an example of shopping carts
set.seed(10)
data <- data.frame(
    orderId = sample(c(1:1000), 5000, replace = TRUE),
    product = sample(
        c('NULL', 'a', 'b', 'c'), # assume we have only 3 products
        5000,
        replace = TRUE,
        prob = c(0.15, 0.65, 0.3, 0.15)
    )
)

# we also know customers' purchase order
order <- data.frame(orderId = c(1:1000),
                    clientId = sample(c(1:300), 1000, replace = TRUE))

# suppose we know customers' gender
sex <- data.frame(clientId = c(1:300),
                  sex = sample(
                      c('male', 'female'),
                      300,
                      replace = TRUE,
                      prob = c(0.40, 0.60)
                  ))


date <- data.frame(orderId = c(1:1000),
                   orderdate = sample((1:90), 1000, replace = TRUE))
orders <- merge(data, order, by = 'orderId')
orders <- merge(orders, sex, by = 'clientId')
orders <- merge(orders, date, by = 'orderId')
orders <- orders[orders$product != 'NULL',]
orders$orderdate <- as.Date(orders$orderdate, origin = "2012-01-01")
rm(data, date, order, sex)

head(orders)

# combining products to the cart (include cases where customers make 2 visits in a day)
df <- orders %>%
    arrange(product) %>%
    select(-orderId) %>%
    unique() %>%
    group_by(clientId, sex, orderdate) %>%
    summarise(cart = paste(product, collapse = ";")) %>%
    ungroup()

head(df)



max.date <- max(df$orderdate) + 1
ids <- unique(df$clientId)
df.new <- data.frame()

for (i in 1:length(ids)) {
    df.cache <- df %>%
        filter(clientId == ids[i])

    ifelse(nrow(df.cache) == 1,
           av.dur <- 30,
           av.dur <-
               round(((
                   max(df.cache$orderdate) - min(df.cache$orderdate)
               ) / (
                   nrow(df.cache) - 1
               )) * 1.5, 0))

    df.cache <-
        rbind(
            df.cache,
            data.frame(
                clientId = df.cache$clientId[nrow(df.cache)],
                sex = df.cache$sex[nrow(df.cache)],
                orderdate = max(df.cache$orderdate) + av.dur,
                cart = 'nopurch'
            )
        )
    ifelse(max(df.cache$orderdate) > max.date,
           df.cache$orderdate[which.max(df.cache$orderdate)] <- max.date,
           NA)

    df.cache$to <- c(df.cache$orderdate[2:nrow(df.cache)] - 1, max.date)

    # order# for Sankey diagram
    df.cache <- df.cache %>%
        mutate(ord = paste('ord', c(1:nrow(df.cache)), sep = ''))

    df.new <- rbind(df.new, df.cache)
}
# filtering dummies
df.new <- df.new %>%
    filter(cart != 'nopurch' | to != orderdate)
rm(orders, df, df.cache, i, ids, max.date, av.dur)
```

A Sankey diagram of the first few carts per customer gives a flow view of how cart
composition evolves, with no-purchase states propagated forward once a customer
lapses; @fig-sequence-sankey traces these flows across the first four orders.

```{r}
#| label: fig-sequence-sankey
#| fig-cap: "Sankey flow of cart composition across the first four orders per customer, with no-purchase states carried forward once a customer becomes dormant."
##### Sankey diagram #######

df.sankey <- df.new %>%
    select(clientId, cart, ord)

df.sankey <-
    dcast(df.sankey,
          clientId ~ ord,
          value.var = 'cart',
          fun.aggregate = NULL)

df.sankey[is.na(df.sankey)] <- 'unknown'

# chosing a length of sequence
df.sankey <- df.sankey %>%
    select(ord1, ord2, ord3, ord4)

# replacing NAs after 'nopurch' for 'nopurch'
df.sankey[df.sankey[, 2] == 'nopurch', 3] <- 'nopurch'
df.sankey[df.sankey[, 3] == 'nopurch', 4] <- 'nopurch'

df.sankey.plot <- data.frame()
for (i in 2:ncol(df.sankey)) {
    df.sankey.cache <- df.sankey %>%
        group_by(df.sankey[, i - 1], df.sankey[, i]) %>%
        summarise(n = n()) %>%
        ungroup()

    colnames(df.sankey.cache)[1:2] <- c('from', 'to')

    # adding tags to carts
    df.sankey.cache$from <-
        paste(df.sankey.cache$from, '(', i - 1, ')', sep = '')
    df.sankey.cache$to <- paste(df.sankey.cache$to, '(', i, ')', sep = '')

    df.sankey.plot <- rbind(df.sankey.plot, df.sankey.cache)
}

make_sankey(df.sankey.plot)

rm(df.sankey, df.sankey.cache, df.sankey.plot, i)
```

Recasting the spells into `TraMineR`'s state-sequence format unlocks a battery of
descriptive tools, collected in the panels below: state-distribution plots over
time (overall and by gender), frequency tables of the most common sequences, mean time
spent in each state, and a per-sequence **entropy** that quantifies how varied a
customer's state history is.
Entropy is the information-theoretic dispersion of the states a customer visits; a
low-entropy customer stays in one or two states, while a high-entropy customer churns
through many, and comparing its distribution by gender exposes behavioral
heterogeneity.

```{r}
#| label: fig-sequence-analysis
#| fig-cap: "State-sequence analyses: state distribution over time overall and by gender, most-frequent sequences, mean time per state, and the distribution of within-customer sequence entropy."
df.new <- df.new %>%
    # chosing a length of sequence
    filter(ord %in% c('ord1', 'ord2', 'ord3', 'ord4')) %>%
    select(-ord)

# converting dates to numbers
min.date <- as.Date(min(df.new$orderdate), format = "%Y-%m-%d")
df.new$orderdate <- as.numeric(df.new$orderdate - min.date + 1)
df.new$to <- as.numeric(df.new$to - min.date + 1)

df.form <-
    seqformat(
        as.data.frame(df.new),
        id = 'clientId',
        begin = 'orderdate',
        end = 'to',
        status = 'cart',
        from = 'SPELL',
        to = 'STS',
        process = FALSE
    )

df.seq <-
    seqdef(df.form,
           left = 'DEL',
           right = 'unknown',
           xtstep = 10) # xtstep - step between ticks (days)
summary(df.seq)

df.feat <- unique(df.new[, c('clientId', 'sex')])

# distribution analysis
seqdplot(df.seq, border = NA, withlegend = 'right')
seqdplot(df.seq, border = NA, group = df.feat$sex) # distribution based on gender


seqstatd(df.seq)

df.seq <- seqdef(df.form,
                 left = 'DEL',
                 right = 'DEL',
                 xtstep = 10)


# the 10 most frequent sequences
seqfplot(df.seq, border = NA, withlegend = 'right')
# the 10 most frequent sequences based on gender
seqfplot(df.seq, group = df.feat$sex, border = NA)

# returning the frequency stats
seqtab(df.seq) # frequency table
seqtab(df.seq[, 1:30]) # frequency table for 1st month

# mean time spent in each state
seqmtplot(df.seq, title = 'Mean time', withlegend = 'right')
seqmtplot(df.seq, group = df.feat$sex, title = 'Mean time')

statd <-
    seqistatd(df.seq) #function returns for each sequence the time spent in the different states
apply(statd, 2, mean) #We may be interested in the mean time spent in each state


# calculating entropy
df.ient <- seqient(df.seq)
hist(df.ient,
     col = 'cyan',
     main = NULL,
     xlab = 'Entropy') # plot an histogram of the within entropy of the sequences
# entrophy distribution based on gender
df.ent <- cbind(df.seq, df.ient)
boxplot(
    Entropy ~ df.feat$sex,
    data = df.ent,
    xlab = 'Gender',
    ylab = 'Sequences entropy',
    col = 'cyan'
)
```

## Geodemographic Classification {#sec-geodemographic}

Geodemographic classification segments customers by *where they live*, on the premise
that neighborhood composition predicts consumption. National statistical agencies
publish small-area classifications—in the United Kingdom, the Output Area
Classification (OAC) groups every census output area into named supergroups such as
"Blue Collar Communities" or "City Living"—that marketers join to behavioral data to
profile and target geographies. The example, following Nick Bearman, profiles
newspaper readership by OAC supergroup and then maps the classification onto Liverpool.

The first step expresses each supergroup's broadsheet readership as a percentage
deviation from the national average, the natural scale for spotting over- and
under-indexing audiences, plotted in @fig-geo-broadsheets.

```{r}
#| label: fig-geo-broadsheets
#| fig-cap: "Broadsheet readership by OAC supergroup, expressed as percentage deviation from the national average. Positive bars mark over-indexing audiences."
#Load libraries
library(scales)
library(ggplot2)


#set up data and data frame
oac_names <-
    c(
        "Blue Collar Communities",
        "City Living",
        "Countryside",
        "Prospering Suburbs",
        "Constrained by Circumstances",
        "Typical Traits",
        "Multicultural"
    )

broadsheets <- c(73.2, 144, 103.9, 109.1, 78.2, 97.1, 120.2)
oac_broadsheets <- data.frame(oac_names, broadsheets)
#convert the percentage values (e.g. 144%) to decimal increase or decrease (e.g. 0.44)
oac_broadsheets$broadsheets <-  broadsheets / 100 - 1

oac_broadsheets

#select the colours we are going to use
my_colour <-
    c("#33A1C9",
      "#FFEC8B",
      "#A2CD5A",
      "#CD7054",
      "#B7B7B7",
      "#9F79EE",
      "#FCC08F")

#plot the graph - this has several bits to it
#the first three lines setup the data and type of graph
ggplot(oac_broadsheets, aes(oac_names, broadsheets)) +
    geom_bar(stat = "identity",
             fill = my_colour,
             position = "identity") +
    theme(axis.text.x = element_text(
        angle = 90,
        hjust = 1,
        vjust = 1,
        size = 12
    )) +
    #this line add the lables to each bar
    geom_text(aes(
        label = paste(round(broadsheets * 100, digits = 0), "%"),
        vjust = ifelse(broadsheets >= 0,-0.5, 1.5)
    ), size = 3) +
    #these lines as the axis labels and these fonts
    theme(axis.title.x = element_text(size = 12)) +
    theme(axis.title.y = element_text(size = 12)) +
    scale_y_continuous("Difference from national average for broadsheet", labels = percent_format()) +
    scale_x_discrete("OAC SuperGroups")
```

Repeating the exercise for tabloid readership yields the complementary profile in
@fig-geo-tabloids, so the two newspaper types can be contrasted across the same
supergroups.

```{r}
#| label: fig-geo-tabloids
#| fig-cap: "Tabloid readership by OAC supergroup as deviation from the national average, complementing the broadsheet profile."
tabloids <- c(110.8, 82.2, 104.9, 94.5, 108.4, 96.4, 96.0)
oac_tabloids <- data.frame(oac_names, tabloids)

#convert the percentage values (e.g. 144%) to decimal increase or decrease (e.g. 0.44)
oac_tabloids$tabloids <-  tabloids / 100 - 1

oac_tabloids


# plot
ggplot(oac_tabloids, aes(oac_names, tabloids)) +
    geom_bar(stat = "identity",
             fill = my_colour,
             position = "identity") +
    theme(axis.text.x = element_text(
        angle = 90,
        hjust = 1,
        vjust = 1,
        size = 12
    )) +
    #this line add the lables to each bar
    geom_text(aes(
        label = paste(round(tabloids * 100, digits = 0), "%"),
        vjust = ifelse(tabloids >= 0, -0.5, 1.5)
    ), size = 3) +
    #these lines as the axis labels and these fonts
    theme(axis.title.x = element_text(size = 12)) +
    theme(axis.title.y = element_text(size = 12)) +
    scale_y_continuous("Difference from national average for tabloids", labels = percent_format()) +
    scale_x_discrete("OAC SuperGroups")
```

To see how the classification distributes across a real city, the analysis loads a
Liverpool output-area shapefile.

```{r geo-load-shapefile, message=FALSE, warning=FALSE}
# `sf` is the modern, maintained replacement for the retired maptools/sp stack
library(sf)
library(tidyverse)

# the Liverpool output-area shapefile is bundled in images/; unzip to a temp dir
oa_dir <- file.path(tempdir(), "liverpool_OA")
unzip("images/liverpool_OA.zip", exdir = oa_dir)
shp <- list.files(oa_dir, pattern = "[.]shp$", recursive = TRUE, full.names = TRUE)[1]

# read the geometry and set the British National Grid CRS (EPSG:27700)
liverpool <- st_read(shp, quiet = TRUE)
st_crs(liverpool) <- 27700
```

The OAC lookup—mapping each output area code to its supergroup, group, and subgroup—is
read in and joined onto the shapefile by area code.

```{r}
library(tidyverse)

#read in OAC by OA csv file
OAC <- rio::import("images/2011oacclustersandnamescsvv2.zip") %>%
    select(
        "Output Area Code",
        "Supergroup Name",
        "Supergroup Code",
        "Group Name",
        "Group Code",
        "Subgroup Name",
        "Subgroup Code"
    ) %>%
    rename("OA_CODE" = "Output Area Code")
```

```{r geo-join}
# join the OAC classification onto the output-area geometry by area code
liverpool <- liverpool %>%
    left_join(OAC, by = c("OA01CD" = "OA_CODE")) %>%
    drop_na(`Supergroup Code`)

# show the attribute head (geometry dropped for display)
liverpool %>%
    st_drop_geometry() %>%
    select(OA01CD, `Supergroup Name`, `Supergroup Code`) %>%
    head()
```

Coloring each output area by its supergroup produces the choropleth map in
@fig-geo-map, turning the abstract classification into a spatial portrait of the city's
neighborhoods.

```{r}
#| label: fig-geo-map
#| fig-cap: "Choropleth map of Liverpool with each output area colored by its OAC supergroup, rendering the geodemographic classification as a spatial portrait of the city."
#Define a set of colours, one for each of the OAC supergroups
my_colour <-
    c("#33A1C9",
      "#FFEC8B",
      "#A2CD5A",
      "#CD7054",
      "#B7B7B7",
      "#9F79EE",
      "#FCC08F")
# build the choropleth: one fill colour per OAC supergroup, ordered by code
liverpool %>%
    mutate(Supergroup = fct_reorder(`Supergroup Name`, `Supergroup Code`)) %>%
    ggplot() +
    geom_sf(aes(fill = Supergroup), color = NA) +
    scale_fill_manual(values = my_colour, name = "OAC supergroup") +
    labs(title = "OAC Supergroup Map of Liverpool") +
    theme_void() +
    theme(legend.position = "right")
```

## Key Takeaways {#sec-empirical-takeaways}

The empirical models in this chapter are all answers to credit-and-value assignment,
and they share a discipline worth restating. **Attribution** is best treated as a
problem of *marginal* contribution rather than position: the Shapley value
(@sec-shapley) supplies the axiomatically unique allocation but at exponential cost
and under a fragile completeness assumption, while the Markov removal effect
(@sec-markov-attribution) supplies a tractable counterfactual whose first-order
duplicate-invariance justifies the standard preprocessing. **Funnel models**
(@sec-sales-funnel) reallocate credit from channels to stages and localize leakage.
**RFM and its clumpiness extension** (@sec-rfm) compress histories into behavioral
sufficient statistics, and **segmentation** (@sec-segmentation) renders those
statistics actionable through lifecycle grids, cohort and retention analysis, and
sequence models that respect order. **Basket and sequence analysis**
(@sec-basket) and **geodemographic classification** (@sec-geodemographic) round out
the toolkit on the product and spatial dimensions.

The recurring caution is identification. Every model attributes credit only among the
factors it can see, so an unobserved channel, an unmodeled stage, or a missing
covariate does not vanish—its effect is silently redistributed onto the measured
quantities. Because conversions are observed once per customer, there is rarely a
held-out ground truth against which to validate an attribution rule, so face validity
and sensitivity to modeling choices must substitute for out-of-sample accuracy. The
prudent analyst reports results under more than one rule and treats large divergences
across rules as a signal that the budgetary stakes of the modeling assumption are
high.

## Full-semester seminar reading map {#sec-empirical-seminar-reading-map}

The empirical models above are the working parts of a much larger doctoral
curriculum. This section reframes the chapter as the methodological spine of a
~13–14 week PhD seminar in empirical/quantitative marketing, mapping each week to
its canonical readings, methods, and the live debate that organizes class
discussion. The reading map is deliberately structured the way top programs
converge on the material: the demand-estimation workhorse first, the twin threats
of heterogeneity and endogeneity next, and the dynamic and causal frontier last.
Readings carry a Crossref-style bibliographic line with a verbatim DOI link; works
already in this book's bibliography are cited by key, and works not (yet) in the
bibliography are named in full so the citation can be added later. Each reading is
tagged **[F]** foundational (stable canon, taught nearly everywhere) or **[R]**
frontier (last ~10–15 years, more program-specific).

**Semester arc.** The quantitative-models seminar moves from *what we measure* to
*how we identify it* to *how we model behavior over time*. The first third installs
the discrete-choice demand workhorse — the multinomial and nested logit on
individual scanner-panel data (Guadagni–Little), the aggregate random-coefficients
logit on market shares (Berry; BLP; Nevo) — because nearly every downstream question
(pricing, promotion, advertising, product line, welfare) is answered through a demand
system. Students learn to read a model as a set of identifying assumptions, not a
regression: what variation in the data pins down the price coefficient, and what
would bias it. The middle third confronts the two threats that define the field's
empirical standards: **unobserved heterogeneity** and **endogeneity**. Heterogeneity
is handled either by integrating it out (random coefficients, mixed logit) or by
estimating it hierarchically (hierarchical Bayes, the Allenby–Rossi program), and the
seminar treats the Bayesian and classical (GMM / simulated maximum likelihood) camps
as complementary toolkits rather than rival faiths. Endogeneity — prices and
marketing-mix variables set in response to demand shocks the econometrician does not
see — motivates instrumental variables, the BLP contraction-and-GMM machinery, and
control-function approaches, alongside a candid critique of weak instruments in
marketing (Rossi 2014). The final third turns dynamic and causal: **state dependence
vs. heterogeneity** (is brand loyalty real switching cost or just persistent taste?),
**forward-looking consumers and firms** (dynamic discrete choice à la Rust; learning
models à la Erdem–Keane; dynamic advertising), and the **field-experiment /
design-based** turn that reframed advertising-effect measurement (Lewis–Rao,
Blake–Nosko–Tadelis) and pushed the field toward heterogeneous-treatment-effect
estimation and ML-assisted targeting. The course closes on applied syntheses —
attribution and customer-lifetime-value modeling — where the semester's tools are
assembled into managerially decision-relevant systems. Throughout, weekly problem
sets are estimation-heavy: students code a logit and a random-coefficients logit from
scratch, run a Gibbs sampler, solve a small dynamic program, and replicate a
published result.

### Week 1 — Foundations: the demand-estimation paradigm & identification thinking {#sec-em-week01}

**Topic.** What "structural" means and why a demand system is the organizing object
of empirical marketing.

**Subtopics.** Demand systems as the organizing object; identification vs.
estimation; reading a paper for its identifying variation; the logit inversion that
turns a demand model into an estimating equation.

**Methods.** Utility-maximization micro-foundations; the share inversion; GMM vs.
likelihood preview; extracting identifying assumptions from a published paper.

**Key readings**

- Berry, S. T. (1994), "Estimating Discrete-Choice Models of Product Differentiation," *RAND Journal of Economics* 25(2), pp. 242–262, [doi:10.2307/2555829](https://doi.org/10.2307/2555829) — @berry1994discretechoice. The share-inversion idea that turns a demand model into a linear-in-mean-utility estimating equation; the conceptual hinge of the course. [F]
- Nevo, A. (2000), "A Practitioner's Guide to Estimation of Random-Coefficients Logit Models of Demand," *Journal of Economics & Management Strategy* 9(4), pp. 513–548, [doi:10.1111/j.1430-9134.2000.00513.x](https://doi.org/10.1111/j.1430-9134.2000.00513.x). The "practitioner's guide" everyone assigns to make BLP estimable; sets the computational template. [F]
- Rossi, P. E. & Allenby, G. M. (2003), "Bayesian Statistics and Marketing," *Marketing Science* 22(3), pp. 304–328, [doi:10.1287/mksc.22.3.304.17739](https://doi.org/10.1287/mksc.22.3.304.17739). Frames the Bayesian alternative early so students see the two estimation paradigms side by side. [F]

**Debate.** Structural ("estimate primitives, do counterfactuals") vs.
reduced-form / design-based ("estimate a credible causal effect") — introduced here,
revisited all semester.

### Week 2 — Discrete choice I: logit/probit on individual panel data {#sec-em-week02}

**Topic.** The multinomial logit on scanner-panel data and the econometrics of
individual choice.

**Subtopics.** MNL and the IIA property; nested logit; the scanner-panel data
environment; loyalty / last-purchase variables.

**Methods.** Coding an MNL likelihood; IIA tests; conditional / fixed-effects logit;
interpreting marginal effects and elasticities.

**Key readings**

- Guadagni, P. M. & Little, J. D. C. (1983), "A Logit Model of Brand Choice Calibrated on Scanner Data," *Marketing Science* 2(3), pp. 203–238, [doi:10.1287/mksc.2.3.203](https://doi.org/10.1287/mksc.2.3.203) — @guadagni1983logit. The paper that launched scanner-data choice modeling in marketing; the canonical MNL-with-loyalty specification. [F]
- Chamberlain, G. (1980), "Analysis of Covariance with Qualitative Data," *Review of Economic Studies* 47(1), pp. 225–238, [doi:10.2307/2297110](https://doi.org/10.2307/2297110). Conditional (fixed-effects) logit — how panel structure controls persistent unobserved heterogeneity; the econometric backbone. [F]

**Debate.** Is the Guadagni–Little loyalty variable capturing structural state
dependence or merely soaking up heterogeneity? (Foreshadows Week 9.)

### Week 3 — Unobserved heterogeneity I: mixed logit & random coefficients {#sec-em-week03}

**Topic.** Modeling taste heterogeneity so substitution patterns are not straitjacketed
by IIA.

**Subtopics.** The IIA problem and substitution patterns; random-coefficients logit;
simulated maximum likelihood; finite-mixture (latent-class) models.

**Methods.** Monte Carlo integration / SML; latent-class EM; comparing fixed vs.
random heterogeneity specifications.

**Key readings**

- Gönül, F. & Srinivasan, K. (1993), "Modeling Multiple Sources of Heterogeneity in Multinomial Logit Models: Methodological and Managerial Issues," *Marketing Science* 12(3), pp. 213–229, [doi:10.1287/mksc.12.3.213](https://doi.org/10.1287/mksc.12.3.213). Multiple sources of heterogeneity in MNL; the managerial-vs-methodological framing marketing students need. [F]
- Allenby, G. M. & Rossi, P. E. (1998), "Marketing Models of Consumer Heterogeneity," *Journal of Econometrics* 89(1–2), pp. 57–78, [doi:10.1016/s0304-4076(98)00055-4](https://doi.org/10.1016/s0304-4076(98)00055-4) — @allenby1998heterogeneity. The case for modeling continuous heterogeneity and its consequences for pricing/targeting. [F]

**Debate.** Parametric mixing distribution vs. nonparametric / finite-mixture; how
much heterogeneity is "real" vs. an artifact of misspecification.

### Week 4 — Hierarchical Bayes & MCMC estimation {#sec-em-week04}

**Topic.** Hierarchical Bayes as the marketing-native engine for individual-level
inference.

**Subtopics.** The hierarchical model; Gibbs sampling and data augmentation;
individual-level posterior estimates; shrinkage and borrowing strength.

**Methods.** Writing a Gibbs sampler for a hierarchical multinomial logit/probit;
convergence diagnostics; posterior predictive checks.

**Key readings**

- Rossi, P. E., McCulloch, R. E. & Allenby, G. M. (1996), "The Value of Purchase History Data in Target Marketing," *Marketing Science* 15(4), pp. 321–340, [doi:10.1287/mksc.15.4.321](https://doi.org/10.1287/mksc.15.4.321) — @rossi1996purchasehistory. The flagship demonstration that household-level Bayesian estimates have direct target-marketing value. [F]
- Rossi, P. E. & Allenby, G. M. (2003), "Bayesian Statistics and Marketing," *Marketing Science* 22(3), pp. 304–328, [doi:10.1287/mksc.22.3.304.17739](https://doi.org/10.1287/mksc.22.3.304.17739). The agenda-setting review of Bayesian methods for marketing decision problems. [F]

**Debate.** Bayes vs. classical for heterogeneity — individual-level inference and
decision-theoretic loss vs. frequentist guarantees.

### Week 5 — Aggregate demand: the BLP random-coefficients model {#sec-em-week05}

**Topic.** Estimating demand from market shares with endogenous prices — the BLP
machinery.

**Subtopics.** Demand from market shares; the contraction mapping; supply side and
markups; micro-moments / combining macro and micro data.

**Methods.** The BLP contraction; nested GMM; computing own/cross price elasticities
and implied markups.

**Key readings**

- Berry, S., Levinsohn, J. & Pakes, A. (1995), "Automobile Prices in Market Equilibrium," *Econometrica* 63(4), pp. 841–890, [doi:10.2307/2171802](https://doi.org/10.2307/2171802) — @berry1995. The foundational aggregate random-coefficients model with endogenous prices; the field's most-imitated method. [F]
- Nevo, A. (2001), "Measuring Market Power in the Ready-to-Eat Cereal Industry," *Econometrica* 69(2), pp. 307–342, [doi:10.1111/1468-0262.00194](https://doi.org/10.1111/1468-0262.00194). The canonical applied BLP used to teach implementation and the market-power counterfactual. [F]
- Nevo, A. (2000), "A Practitioner's Guide to Estimation of Random-Coefficients Logit Models of Demand," *Journal of Economics & Management Strategy* 9(4), pp. 513–548, [doi:10.1111/j.1430-9134.2000.00513.x](https://doi.org/10.1111/j.1430-9134.2000.00513.x). The estimation recipe students actually code against. [F]

**Debate.** Aggregate vs. individual data — what is gained/lost by estimating from
shares; numerical fragility and the "BLP is hard to get right" literature.

### Week 6 — Endogeneity I: instrumental variables & control functions {#sec-em-week06}

**Topic.** Price endogeneity and the two main fixes — IV/GMM and control functions.

**Subtopics.** Price endogeneity from unobserved product/demand shocks; cost-shifter
and Hausman/BLP instruments; the control-function alternative for nonlinear models.

**Methods.** 2SLS/GMM with logit; control-function (residual-inclusion) estimation;
instrument-relevance/validity diagnostics.

**Key readings**

- Villas-Boas, J. M. & Winer, R. S. (1999), "Endogeneity in Brand Choice Models," *Management Science* 45(10), pp. 1324–1338, [doi:10.1287/mnsc.45.10.1324](https://doi.org/10.1287/mnsc.45.10.1324) — @villas-boas1999. The paper that put price endogeneity on the marketing field's agenda for brand-choice models. [F]
- Petrin, A. & Train, K. (2010), "A Control Function Approach to Endogeneity in Consumer Choice Models," *Journal of Marketing Research* 47(1), pp. 3–13, [doi:10.1509/jmkr.47.1.3](https://doi.org/10.1509/jmkr.47.1.3) — @Petrin_2010. The control-function approach — the practical alternative to BLP-style IV inside choice models. [F]
- Chintagunta, P., Dubé, J.-P. & Goh, K. Y. (2005), "Beyond the Endogeneity Bias: The Effect of Unmeasured Brand Characteristics on Household-Level Brand Choice Models," *Management Science* 51(5), pp. 832–849, [doi:10.1287/mnsc.1040.0323](https://doi.org/10.1287/mnsc.1040.0323) — @chintagunta2005. Endogeneity from unobserved brand characteristics at the household level; bridges micro and aggregate treatments. [F]

**Debate.** IV vs. control function; what makes a credible marketing instrument.

### Week 7 — Endogeneity II: the IV critique & the credibility standard {#sec-em-week07}

**Topic.** When instruments fail, and the credibility standard the field now demands.

**Subtopics.** Weak/invalid instruments; the bias–variance trade-off of IV; when to
prefer direct measurement or design over instrumenting.

**Methods.** First-stage F-stats and weak-instrument diagnostics; overidentification
(J) tests; sensitivity analysis.

**Key readings**

- Rossi, P. E. (2014), "Even the Rich Can Make Themselves Poor: A Critical Examination of IV Methods in Marketing Applications," *Marketing Science* 33(5), pp. 655–672, [doi:10.1287/mksc.2014.0860](https://doi.org/10.1287/mksc.2014.0860) — @rossi2014. The influential critique arguing strong, valid instruments are rare in marketing. [R]
- Hansen, L. P. (1982), "Large Sample Properties of Generalized Method of Moments Estimators," *Econometrica* 50(4), pp. 1029–1054, [doi:10.2307/1912775](https://doi.org/10.2307/1912775). GMM foundations — the inferential machinery (and overidentification test) underneath IV and BLP. [F]

**Debate.** The "credibility revolution" reaches marketing — instrument-based vs.
experimental/design-based identification; how much to trust structural counterfactuals
built on weak IVs.

### Week 8 — Panel data, fixed effects & dynamic panels {#sec-em-week08}

**Topic.** Within vs. between variation and the econometrics of nonlinear panels.

**Subtopics.** Within vs. between variation; fixed vs. random effects; the
incidental-parameters problem; lagged-dependent-variable bias.

**Methods.** FE/RE estimation and Hausman testing; handling initial-conditions and
incidental-parameters problems; clustering/inference in panels.

**Key readings**

- Chamberlain, G. (1980), "Analysis of Covariance with Qualitative Data," *Review of Economic Studies* 47(1), pp. 225–238, [doi:10.2307/2297110](https://doi.org/10.2307/2297110). Conditional-likelihood estimation that differences out fixed effects in nonlinear panels. [F]
- Keane, M. P. (1997), "Modeling Heterogeneity and State Dependence in Consumer Choice Behavior," *Journal of Business & Economic Statistics* 15(3), pp. 310–327, [doi:10.1080/07350015.1997.10524709](https://doi.org/10.1080/07350015.1997.10524709) — @keane1997statedependence. The reference treatment of heterogeneity vs. state dependence in panel choice data; the methodological heart of the course's dynamics half. [F]

**Debate.** Fixed effects (robust, costly in degrees of freedom) vs.
random effects / hierarchical (efficient, requires distributional assumptions).

### Week 9 — State dependence vs. heterogeneity (consumer inertia) {#sec-em-week09}

**Topic.** The field's signature identification problem: real switching costs vs.
persistent taste.

**Subtopics.** Structural state dependence vs. persistent heterogeneity; loyalty as
switching cost; initial conditions; identification from price/promotion variation.

**Methods.** Specifying lagged-choice effects; controlling for unobserved
heterogeneity while estimating state dependence; exploiting moves/shocks for
identification.

**Key readings**

- Keane, M. P. (1997), "Modeling Heterogeneity and State Dependence in Consumer Choice Behavior," *Journal of Business & Economic Statistics* 15(3), pp. 310–327, [doi:10.1080/07350015.1997.10524709](https://doi.org/10.1080/07350015.1997.10524709) — @keane1997statedependence. The framing paper: disentangling true state dependence from spurious persistence. [F]
- Dubé, J.-P., Hitsch, G. J. & Rossi, P. E. (2010), "State Dependence and Alternative Explanations for Consumer Inertia," *RAND Journal of Economics* 41(3), pp. 417–445, [doi:10.1111/j.1756-2171.2010.00106.x](https://doi.org/10.1111/j.1756-2171.2010.00106.x). The modern reference distinguishing genuine switching costs from alternative explanations for inertia. [R]
- Bronnenberg, B. J., Dubé, J.-P. & Gentzkow, M. (2012), "The Evolution of Brand Preferences: Evidence from Consumer Migration," *American Economic Review* 102(6), pp. 2472–2508, [doi:10.1257/aer.102.6.2472](https://doi.org/10.1257/aer.102.6.2472). Brand capital and persistent preferences identified from consumer migration — a clean natural-experiment take on "why preferences persist." [R]

**Debate.** What variation separates habit/loyalty from taste, and why it matters for
pricing and promotion policy.

### Week 10 — Dynamic discrete choice: forward-looking consumers & firms {#sec-em-week10}

**Topic.** Estimating choice when agents are forward-looking.

**Subtopics.** The dynamic-programming choice framework; conditional value functions;
nested fixed-point estimation; stockpiling and learning.

**Methods.** Solving a discrete DP by value-function iteration; the nested
fixed-point algorithm; likelihood for forward-looking choices.

**Key readings**

- Rust, J. (1987), "Optimal Replacement of GMC Bus Engines: An Empirical Model of Harold Zurcher," *Econometrica* 55(5), pp. 999–1033, [doi:10.2307/1911259](https://doi.org/10.2307/1911259). The foundational dynamic-discrete-choice model; the estimation template (NFXP) for forward-looking agents. [F]
- Erdem, T. & Keane, M. P. (1996), "Decision-Making Under Uncertainty: Capturing Dynamic Brand Choice Processes in Turbulent Consumer Goods Markets," *Marketing Science* 15(1), pp. 1–20, [doi:10.1287/mksc.15.1.1](https://doi.org/10.1287/mksc.15.1.1) — @erdem1996learning. Brings dynamic decision-making under uncertainty (Bayesian learning) into marketing brand choice. [F]
- Erdem, T., Keane, M. P. & Sun, B. (2008), "A Dynamic Model of Brand Choice When Price and Advertising Signal Product Quality," *Marketing Science* 27(6), pp. 1111–1125, [doi:10.1287/mksc.1080.0362](https://doi.org/10.1287/mksc.1080.0362). Price and advertising as quality signals in a dynamic learning model — frontier integration of dynamics, information, and marketing mix. [R]

**Debate.** Myopic vs. forward-looking specifications; the heavy
computational/identification cost of full-solution dynamics vs. two-step (CCP)
estimators.

### Week 11 — Advertising & marketing-mix dynamics; aggregate response {#sec-em-week11}

**Topic.** Dynamic advertising and aggregate marketing-mix response.

**Subtopics.** Advertising carryover/goodwill stocks; dynamic firm advertising
policy; aggregate marketing-mix response; sponsored-search/digital dynamics.

**Methods.** Estimating Koyck/goodwill carryover; dynamic optimization of a marketing
instrument; aggregate response identification.

**Key readings**

- Dubé, J.-P., Hitsch, G. J. & Manchanda, P. (2005), "An Empirical Model of Advertising Dynamics," *Quantitative Marketing and Economics* 3(2), pp. 107–144, [doi:10.1007/s11129-005-0334-2](https://doi.org/10.1007/s11129-005-0334-2). Estimating advertising goodwill dynamics and rationalizing pulsing — the modern dynamic-advertising reference. [R]
- Yao, S. & Mela, C. F. (2011), "A Dynamic Model of Sponsored Search Advertising," *Marketing Science* 30(3), pp. 447–468, [doi:10.1287/mksc.1100.0626](https://doi.org/10.1287/mksc.1100.0626). A dynamic structural model of sponsored-search advertising; ports dynamics to digital marketplaces. [R]

**Debate.** Structural dynamic models of advertising vs. design-based field-experiment
measurement (sets up Week 12).

### Week 12 — Causal inference & field experiments in marketing {#sec-em-week12}

**Topic.** The experimental turn in advertising-effect measurement.

**Subtopics.** The experimental turn in advertising measurement; regression
discontinuity and natural experiments; selection and endogenous targeting in
observational ad data.

**Methods.** Designing/analyzing large field experiments; power analysis for ad
effects; RD estimation and validity checks.

**Key readings**

- Lewis, R. A. & Rao, J. M. (2015), "The Unfavorable Economics of Measuring the Returns to Advertising," *Quarterly Journal of Economics* 130(4), pp. 1941–1973, [doi:10.1093/qje/qjv023](https://doi.org/10.1093/qje/qjv023) — @Lewis_2015. Why observational and even experimental advertising-ROI estimates are so noisy — recalibrated the field's expectations. [R]
- Blake, T., Nosko, C. & Tadelis, S. (2015), "Consumer Heterogeneity and Paid Search Effectiveness: A Large-Scale Field Experiment," *Econometrica* 83(1), pp. 155–174, [doi:10.3982/ecta12423](https://doi.org/10.3982/ecta12423) — @blake2015. The eBay paid-search field experiment showing large observational estimates collapse under randomization. [R]
- Hartmann, W., Nair, H. S. & Narayanan, S. (2011), "Identifying Causal Marketing Mix Effects Using a Regression Discontinuity Design," *Marketing Science* 30(6), pp. 1079–1097, [doi:10.1287/mksc.1110.0670](https://doi.org/10.1287/mksc.1110.0670). Regression-discontinuity identification of marketing-mix effects — quasi-experimental design inside a marketing setting. [R]

**Debate.** Experiments vs. structural models — internal validity and causal
credibility vs. counterfactual richness and external extrapolation.

### Week 13 — Heterogeneous treatment effects, targeting & ML-meets-econometrics {#sec-em-week13}

**Topic.** Estimating who responds, and turning that into a targeting policy.

**Subtopics.** CATE estimation; causal forests / honest trees; ML for policy
targeting; robustness to real data challenges.

**Methods.** Estimating CATEs; constructing and evaluating targeting policies;
cross-fitting and honesty; out-of-sample policy value.

**Key readings**

- Wager, S. & Athey, S. (2018), "Estimation and Inference of Heterogeneous Treatment Effects Using Random Forests," *Journal of the American Statistical Association* 113(523), pp. 1228–1242, [doi:10.1080/01621459.2017.1319839](https://doi.org/10.1080/01621459.2017.1319839). Causal forests for heterogeneous-treatment-effect estimation with valid inference — the bridge from ML to causal targeting. [R]
- Simester, D., Timoshenko, A. & Zoumpoulis, S. I. (2020), "Targeting Prospective Customers: Robustness of Machine-Learning Methods to Typical Data Challenges," *Management Science* 66(6), pp. 2495–2522, [doi:10.1287/mnsc.2019.3308](https://doi.org/10.1287/mnsc.2019.3308). Field-experiment evaluation of ML targeting methods under typical marketing-data pathologies. [R]

**Debate.** Prediction vs. causal targeting; can off-the-shelf ML deliver valid
policies, and when does it break?

### Week 14 — Applied synthesis: attribution & customer-lifetime-value models {#sec-em-week14}

**Topic.** Assembling the semester's tools into CLV and attribution systems.

**Subtopics.** Buy-till-you-die / Pareto-NBD models; RFM and CLV; multichannel
attribution; carryover and spillover across touchpoints; social interactions as a
recurring identification challenge.

**Methods.** Fitting Pareto/NBD and BG/NBD; computing CLV; estimating attribution
with carryover/spillover; recognizing the reflection problem.

**Key readings**

- Schmittlein, D. C., Morrison, D. G. & Colombo, R. (1987), "Counting Your Customers: Who Are They and What Will They Do Next?," *Management Science* 33(1), pp. 1–24, [doi:10.1287/mnsc.33.1.1](https://doi.org/10.1287/mnsc.33.1.1). The Pareto/NBD model — foundation of probabilistic CLV. [F]
- Fader, P. S., Hardie, B. G. S. & Lee, K. L. (2005), "RFM and CLV: Using Iso-Value Curves for Customer Base Analysis," *Journal of Marketing Research* 42(4), pp. 415–430, [doi:10.1509/jmkr.2005.42.4.415](https://doi.org/10.1509/jmkr.2005.42.4.415). RFM-to-CLV via iso-value curves — the bridge from the academic model to managerial customer-base analysis. [F]
- Li, H. (A.) & Kannan, P. K. (2014), "Attributing Conversions in a Multichannel Online Marketing Environment: An Empirical Model and a Field Experiment," *Journal of Marketing Research* 51(1), pp. 40–56, [doi:10.1509/jmr.13.0050](https://doi.org/10.1509/jmr.13.0050) — @li2014. Multichannel attribution combining an empirical model with a field experiment — the capstone integrating choice, dynamics, and experiments. [R]
- Hartmann, W. R., Manchanda, P., Nair, H., Bothner, M., Dodds, P., Godes, D., Hosanagar, K. & Tucker, C. (2008), "Modeling Social Interactions: Identification, Empirical Methods and Policy Implications," *Marketing Letters* 19(3–4), pp. 287–304, [doi:10.1007/s11002-008-9048-z](https://doi.org/10.1007/s11002-008-9048-z). Identification of social interactions — the reflection problem students must recognize whenever "network" effects appear. [R]

**Debate.** Correlation vs. causation in attribution and "viral" effects; model-based
CLV vs. machine-learned churn/value scoring.

### Foundational vs. frontier at a glance {#sec-em-foundational-vs-frontier}

- **Demand/choice.** *Foundational:* logit/MNL (Guadagni–Little); the Berry inversion;
  BLP; Nevo's practitioner's guide. *Frontier:* combining micro and macro data;
  ML-based demand.
- **Heterogeneity.** *Foundational:* random coefficients; hierarchical Bayes
  (Allenby–Rossi). *Frontier:* nonparametric heterogeneity; deep/embedding models.
- **Endogeneity.** *Foundational:* IV/GMM (Hansen); control functions (Petrin–Train).
  *Frontier:* the IV critique (Rossi 2014); design-based identification.
- **Panel/dynamics.** *Foundational:* fixed-effects/conditional logit (Chamberlain);
  state dependence (Keane); dynamic discrete choice (Rust). *Frontier:* Bayesian
  learning (Erdem–Keane); CCP/two-step estimators.
- **Causal/experiments.** *Foundational:* newer to the canon. *Frontier:* field
  experiments (Lewis–Rao; Blake–Nosko–Tadelis); regression discontinuity;
  heterogeneous-treatment-effect / causal forests (Wager–Athey).
- **Applied synthesis.** *Foundational:* Pareto/NBD (Schmittlein–Morrison–Colombo);
  CLV (Fader–Hardie–Lee). *Frontier:* multichannel attribution (Li–Kannan); ML
  targeting (Simester et al.).

The first ~8 weeks are highly stable across programs and decades; weeks 12–13 are
where syllabi differ most (Stanford/Berkeley/MIT lean experiments and ML; Ohio
State/Wharton lean Bayes; Chicago/Yale lean structural IO). A 13-week version
typically merges Weeks 6–7 (endogeneity) or Weeks 11–12 (dynamics/experiments).

### How this reading map expands {#sec-em-reading-map-expands}

This reading map is a living scaffold meant to grow alongside the runnable code in
the rest of the chapter:

1. **Estimation walk-throughs as worked code.** Each foundational module should ship a
   reproducible estimator in the book's house style: a hand-coded MNL log-likelihood,
   a random-coefficients logit via simulated ML, a Gibbs sampler for a hierarchical
   probit, a small value-function iteration for a Rust-style dynamic program, and a
   BLP contraction. These are the chapter's highest-value, longest-lived assets.
2. **Identification call-outs.** For every model, add a boxed "what identifies the key
   parameter" note and a "what breaks it" note — the seminar's central skill, and what
   differentiates the chapter from a methods cookbook.
3. **The structural-vs-design debate as a through-line.** Rather than confining it to a
   single week, thread the tension (counterfactual richness vs. causal credibility)
   across demand, endogeneity, dynamics, and experiments, and close with a synthesis
   on when each approach is decision-relevant.
4. **Frontier refresh cadence.** Weeks 12–13 (experiments, HTE/ML, targeting) and the
   digital-dynamics material age fastest; budget a periodic pass to add current
   top-journal work on platform/algorithmic demand, privacy-constrained measurement,
   and LLM/embedding-based demand models.
5. **Cross-references.** Wire the map to sibling chapters — conjoint/preference
   measurement (choice foundations), advertising and marketing-mix modeling
   (Weeks 11–12), the CLV material (Week 14, and @sec-rfm here), structural models
   (Weeks 5 and 10), and the data/estimation craft chapters.
6. **Replication appendix.** Add a "replicate a classic" exercise (a teaching-scale BLP
   on public market-share data; a Pareto/NBD on a CDNOW-style dataset) so students
   leave having reproduced a published result.

As bibliography entries are added for the works named in full above (Nevo 2000/2001,
Rossi–Allenby 2003, Chamberlain 1980, Gönül–Srinivasan 1993, Hansen 1982,
Dubé–Hitsch–Rossi 2010, Bronnenberg–Dubé–Gentzkow 2012, Rust 1987,
Erdem–Keane–Sun 2008, Dubé–Hitsch–Manchanda 2005, Yao–Mela 2011,
Hartmann–Nair–Narayanan 2011, Wager–Athey 2018, Simester–Timoshenko–Zoumpoulis 2020,
Schmittlein–Morrison–Colombo 1987, Fader–Hardie–Lee 2005, and the Hartmann et al. 2008
social-interactions review), each prose citation can be upgraded to an `@`-key citation
without changing the surrounding text.
