# Structural Models {#sec-structural-models}

A *structural model* is an empirical model whose parameters are the primitives of an
economic theory: the preferences of consumers, the technologies and costs of firms,
the information agents hold, and the rules of the game they play. Rather than fit a
flexible function from covariates to an outcome, the analyst writes down how agents
*decide*—typically as the solution to an optimization problem under an equilibrium
concept—and then estimates the deep parameters that rationalize the observed data.
The payoff is that those parameters are, by construction, invariant to the policies
the analyst wants to study, so the model can answer questions about worlds that were
never observed: a price never charged, a product never launched, a compensation plan
never offered.

This is a different ambition from prediction or even from causal inference. A
reduced-form analysis asks *what is the effect of X on Y?* and answers it, ideally,
with a research design that isolates exogenous variation. A structural analysis asks
*how does the system work?*—and, having recovered the mechanism, can simulate the
effect of any intervention the mechanism admits, including ones for which no
experiment or natural experiment exists. The cost of that generality is steep:
structural models impose strong assumptions (functional forms, equilibrium, the
information environment), and when those assumptions are wrong the conclusions are
wrong in ways that are hard to detect. Much of the craft lies in choosing
assumptions that buy identification without buying bias, and in being honest about
which results lean on which assumptions.

This chapter develops the structural approach for a reader who already understands
regression and causal inference and wants to see what changes when economic theory
is moved from the discussion section into the likelihood. We proceed from intuition
to formalism: first the anatomy of a structural model and its relationship to
causal inference (@sec-sm-vs-causal); then a worked supply-and-demand example that
exposes the endogeneity problem at the heart of the field (@sec-sm-demand-supply);
then the two workhorse estimators—maximum likelihood and the generalized method of
moments—and the discrete-choice and dynamic models that dominate marketing
applications (@sec-sm-estimation through @sec-sm-dynamic); and finally the
identification threats that decide whether any of it can be believed
(@sec-sm-identification). Throughout, the running theme is the one stated above:
structural parameters are valuable precisely to the extent that they are policy
invariant, and they are policy invariant only to the extent that the model is
correct.

## What Makes a Model "Structural" {#sec-sm-vs-causal}

The defining feature of a structural model is that its parameters have an
interpretation *outside* the data-generating environment in which they were
estimated. @chintagunta2006structural, in a field review, characterize structural
modeling as the practice of using economic theory to discipline the specification of
an empirical model so that the estimated coefficients correspond to taste,
technology, or strategic primitives rather than to environment-specific
correlations. The distinction matters most when the goal is counterfactual: a
demand elasticity estimated as a structural taste parameter can be carried into a
pricing simulation, whereas a regression slope estimated in one price regime carries
no guarantee of stability when the regime changes.

It is useful to state the contrast formally. Suppose an outcome $y$ is generated by
agents who choose actions to maximize an objective indexed by a structural parameter
vector $\boldsymbol{\theta}$ (e.g., utility weights, marginal costs) given a state
$\mathbf{s}$ and a policy $\mathbf{p}$ (prices, assortment, incentives):
$$
y = m(\mathbf{s}, \mathbf{p}; \boldsymbol{\theta}) + \varepsilon ,
$$ {#eq-sm-structural}
where $m(\cdot)$ is the *decision rule*—the mapping from states and policy to
behavior implied by the agent's optimization and the market equilibrium. A
reduced-form analysis estimates a projection of $y$ on $(\mathbf{s},\mathbf{p})$
that is valid only on the support and policy regime observed. A structural analysis
estimates $\boldsymbol{\theta}$, then *re-solves* @eq-sm-structural under a new
policy $\mathbf{p}'$ to predict $y' = m(\mathbf{s}, \mathbf{p}';
\boldsymbol{\theta})$. The structural object is reusable; the reduced-form object,
strictly, is not.

> Causal inference estimates the effect of a treatment that occurred; structural
> modeling estimates the mechanism that would produce the effect of a treatment that
> did not. The first is a statement about a realized intervention, the second a
> statement about a class of interventions the model can simulate.

The two traditions are complements rather than rivals, and the modern literature
treats them as such. Credible causal estimates discipline structural models—a
quasi-experimental demand elasticity is a moment a structural demand system should
reproduce—and structural reasoning, in turn, tells the causal analyst *where* the
endogeneity is and *what* would constitute a valid instrument. @conlon2013demand
illustrate the round trip: they build a structural demand model to handle
endogenous product availability (stockouts that are correlated with demand shocks)
and then use the structural estimates to run counterfactuals that no reduced-form
design could deliver. The remainder of this section makes the three pillars of any
structural exercise—behavioral assumptions, equilibrium, and counterfactuals—precise
before we turn to estimation.

@fig-sm-pipeline summarizes the workflow that distinguishes the structural pipeline
from a reduced-form one.

```{mermaid}
%%| label: fig-sm-pipeline
%%| fig-cap: "The structural modeling pipeline. Economic theory specifies agents' objectives; optimization and an equilibrium concept yield a decision rule whose deep parameters are estimated by MLE or GMM; the estimated rule is then re-solved under new policies for counterfactual analysis. The dashed path is the reduced-form alternative, which stops at an environment-specific association."
flowchart TD
    A[Economic theory:<br/>preferences, costs,<br/>information, game] --> B[Agent optimization<br/>+ equilibrium concept]
    B --> C["Decision rule<br/>m(s, p; θ)"]
    C --> D[Estimation:<br/>MLE / GMM<br/>recover θ]
    D --> E[Re-solve under<br/>new policy p′]
    E --> F[Counterfactual /<br/>policy simulation]
    G[(Observed data)] -.->|reduced form| H[Environment-specific<br/>association]
    G --> D
```

### Behavioral and Equilibrium Assumptions

Structural models earn their counterfactual power by committing to assumptions of
four kinds, and the credibility of any application rests on how defensible those
commitments are.

The first is **functional form**: the shape of utility, demand, or cost. Is demand
linear in price, or constant-elasticity? Is utility linear in attributes, or does it
exhibit diminishing marginal utility? These choices are not innocuous, because the
counterfactual extrapolates along the assumed curvature.

The second is the **decision protocol**: the rule by which agents map information to
actions. The canonical assumption is utility maximization—each consumer chooses the
alternative delivering the highest utility—but the model must also say whether agents
are myopic or forward looking, risk neutral or risk averse, and whether they learn.

The third is the **information environment**. Do consumers know all product
attributes when choosing, or do they face uncertainty and update beliefs?
@erdem2008impact, for instance, build a model in which advertising influences price
sensitivity precisely *because* consumers are uncertain about quality in experience-
good markets and advertising conveys information; the substantive conclusion is
inseparable from the informational assumption.

The fourth is the **equilibrium concept**. In a market model the analyst must say how
agents' choices are mutually consistent: that the market clears, or that firms play a
Nash equilibrium in prices, or that a dynamic game is in Markov-perfect equilibrium.
@kadiyali2001structural review how equilibrium assumptions let the analyst infer
unobserved competitive conduct—whether firms behave more collusively or more
competitively than Bertrand—from the comovement of prices and quantities, an
inference impossible without the equilibrium restriction.

These assumptions are what make the model *tractable and estimable*, and they are
simultaneously the model's principal vulnerability: a misspecified information
environment or an incorrect conduct assumption biases the deep parameters and, with
them, every counterfactual. The discipline of the field is to state each assumption
explicitly, to test those that are testable, and to flag the counterfactuals that
hinge on the rest.

### Counterfactuals: the Reason to Pay the Price

The justification for the structural apparatus is the counterfactual. Once
$\boldsymbol{\theta}$ is in hand, the analyst re-solves the model under a policy that
was never implemented and reads off the predicted outcome—*what would sales have
been at a price we never charged?* or *what compensation plan maximizes profit?*
@gowrisankaran2020framework provide a framework for exactly this kind of policy
evaluation, and the counterfactual logic is what separates a structural paper from a
descriptive one. The validity of a counterfactual is conditional on two things: the
parameters are policy invariant (the Lucas-critique requirement), and the new policy
lies within the class the model can represent. A model estimated under linear
pricing cannot credibly simulate nonlinear tariffs; a static model cannot simulate a
policy whose whole point is to change dynamic incentives. Stating the admissible
counterfactual class is therefore part of specifying the model, not an afterthought.

## A First Structural Model: Demand and Supply {#sec-sm-demand-supply}

The simplest model that exhibits the structural way of thinking is linear
simultaneous supply and demand. It is worth working through in full because it
already contains the central econometric problem of the entire field—the endogeneity
of price—and the standard solution.

Let quantity demanded and supplied at price $p$ be
$$
Q^{d} = \alpha_0 - \alpha_1 p + u^{d}, \qquad
Q^{s} = \beta_0 + \beta_1 p + u^{s},
$$ {#eq-sm-demand-supply}
with $\alpha_1, \beta_1 > 0$ and structural shocks $u^{d}, u^{s}$ (a demand shifter
such as a taste shock; a supply shifter such as an input-cost shock). The
*structural* parameters are the slopes $\alpha_1$ (how strongly quantity responds to
price along the demand curve) and $\beta_1$ (the analogous supply response); these
are the policy-invariant objects a manager would want for pricing. Market clearing,
$Q^{d} = Q^{s}$, is the equilibrium condition, and it determines the observed price
$$
p^{\star} = \frac{\alpha_0 - \beta_0 + u^{d} - u^{s}}{\alpha_1 + \beta_1}.
$$ {#eq-sm-equilibrium}

@eq-sm-equilibrium is the crux. Because equilibrium price depends on the demand
shock $u^{d}$, price is **endogenous**: $\mathrm{Cov}(p^{\star}, u^{d}) \neq 0$. A
naive regression of observed quantity on observed price does not recover $\alpha_1$;
it recovers a hybrid of the demand and supply slopes—the textbook simultaneity bias.
Intuitively, the observed price–quantity pairs are intersections of shifting demand
and supply curves, so the scatter traces out neither curve. This single fact—that
prices are set by agents who observe demand conditions the econometrician does
not—motivates the instrumental-variable and control-function machinery that
pervades structural demand estimation.

Identification comes from **exclusion restrictions**: a variable that shifts supply
but not demand traces out the demand curve, and vice versa. A cost shifter $z$
entering supply but excluded from demand instruments for price in the demand
equation. @montgomery1999estimating exploit precisely this logic to estimate price
elasticities at the store and household level, recovering structural demand slopes
that a single-equation regression would contaminate. The simulation below makes the
bias concrete and shows that an instrument recovers the truth.

```{r sm-simultaneity, message=FALSE, warning=FALSE}
set.seed(29)
n  <- 5000
a0 <- 10; a1 <- 1.5      # demand intercept, slope (truth)
b0 <- 2;  b1 <- 1.0      # supply intercept, slope (truth)

z  <- runif(n, 0, 4)     # cost shifter: enters supply, excluded from demand
ud <- rnorm(n)           # demand shock
us <- rnorm(n)           # supply shock

# Equilibrium price and quantity (solve Qd = Qs with a cost shifter in supply)
# Qd = a0 - a1 p + ud ;  Qs = b0 + b1 p + g*z + us
g  <- 1.2
p  <- (a0 - b0 - g * z + ud - us) / (a1 + b1)
q  <- a0 - a1 * p + ud

# (1) Naive OLS of quantity on price: biased for the demand slope -a1
ols  <- coef(lm(q ~ p))["p"]

# (2) Two-stage least squares using the cost shifter z as an instrument
phat <- fitted(lm(p ~ z))
iv   <- coef(lm(q ~ phat))["phat"]

cat("True demand slope (-a1):      ", -a1, "\n")
cat("OLS estimate (biased):        ", round(ols, 3), "\n")
cat("IV / 2SLS estimate (consistent):", round(iv, 3), "\n")
```

The OLS slope is attenuated and unstable; the instrumented slope recovers $-\alpha_1$.
This linear example scales, conceptually, all the way to modern differentiated-
products demand systems, where the same endogeneity reappears (price correlates with
unobserved product quality) and is handled with the same idea—cost or competitor-
based instruments—inside a far richer choice model, to which we now turn.

## Estimation: Maximum Likelihood and GMM {#sec-sm-estimation}

Structural parameters are rarely obtained by ordinary least squares. The two
dominant estimators are maximum likelihood (ML) and the generalized method of
moments (GMM); the standard graduate references develop both in depth
[@greene2003econometric; @cameron2005microeconometrics; @hensher2005applied], and
the *Handbook of Econometrics* chapters on the econometric evaluation of policy
collect the identification theory [@heckman2007econometric1;
@heckman2007econometric; @abbring2007econometric3]. The choice between ML and GMM is
chiefly a choice about how much of the data-generating process one is willing to
specify.

**Maximum likelihood** specifies the full conditional distribution of the data. If
the model implies a density $f(y_i \mid \mathbf{x}_i; \boldsymbol{\theta})$, the
estimator maximizes the log-likelihood,
$$
\hat{\boldsymbol{\theta}}_{\mathrm{ML}}
  = \arg\max_{\boldsymbol{\theta}}
    \sum_{i=1}^{n} \log f(y_i \mid \mathbf{x}_i; \boldsymbol{\theta}),
$$ {#eq-sm-mle}
which is efficient when the distributional assumption is correct but inconsistent
when it is not. ML is the natural estimator for discrete-choice models, where the
choice probabilities *are* the likelihood contributions.

**The generalized method of moments** specifies only a set of population moment
conditions implied by the theory—typically that the structural error is orthogonal
to a vector of instruments $\mathbf{z}_i$,
$$
\mathbb{E}\!\left[\, \mathbf{z}_i \, g(y_i, \mathbf{x}_i; \boldsymbol{\theta}) \,\right] = \mathbf{0},
$$ {#eq-sm-gmm-moment}
where $g(\cdot)$ is the structural residual. GMM minimizes the empirical analog of
these moments in a quadratic form,
$$
\hat{\boldsymbol{\theta}}_{\mathrm{GMM}}
  = \arg\min_{\boldsymbol{\theta}}
    \bar{g}_n(\boldsymbol{\theta})^{\top}\, \mathbf{W}\, \bar{g}_n(\boldsymbol{\theta}),
\qquad
\bar{g}_n(\boldsymbol{\theta}) = \frac{1}{n}\sum_{i=1}^{n}
  \mathbf{z}_i\, g(y_i, \mathbf{x}_i; \boldsymbol{\theta}),
$$ {#eq-sm-gmm}
with weighting matrix $\mathbf{W}$; the efficient choice sets $\mathbf{W}$ to the
inverse of the moment covariance. GMM is the estimator of choice when the analyst
trusts a few orthogonality conditions (a cost instrument, an equilibrium first-order
condition) but does not wish to commit to a full likelihood—exactly the situation in
differentiated-products demand, where the famous contraction-mapping estimator
inverts market shares to demand shocks and stacks them against instruments in
@eq-sm-gmm. @tbl-sm-mle-gmm contrasts the two.

| Dimension | Maximum likelihood | Generalized method of moments |
|---|---|---|
| Specifies | Full conditional density of the data | A set of moment/orthogonality conditions |
| Efficiency | Efficient if the distribution is correct | Efficient given the chosen moments; generally less than full ML |
| Robustness | Inconsistent under distributional misspecification | Consistent under far weaker assumptions |
| Endogeneity | Handled via control functions or joint likelihood | Handled natively through instrument moments |
| Typical use | Discrete choice, dynamic discrete choice | Demand systems, Euler-equation and conduct models |
| Failure mode | Wrong likelihood $\Rightarrow$ biased everything | Weak or invalid instruments $\Rightarrow$ weak identification |

: Maximum likelihood versus GMM for structural estimation. {#tbl-sm-mle-gmm}

## Discrete Choice as the Workhorse {#sec-sm-discrete-choice}

A large share of structural marketing models are built on **random-utility discrete
choice**, because most consumer decisions—which brand, which channel, whether to buy
at all—are discrete, and because the framework yields choice probabilities that are
both behaviorally founded and computationally convenient. The reviews by
@dube2002structural and @heiss2002structural lay out the family and its estimation;
the model below is the canonical entry point.

Consumer $i$ choosing among alternatives $j = 1,\dots,J$ derives utility
$$
U_{ij} = V_{ij} + \varepsilon_{ij}
       = \mathbf{x}_{ij}^{\top}\boldsymbol{\beta} - \alpha p_{ij} + \xi_j + \varepsilon_{ij},
$$ {#eq-sm-rum}
where $\mathbf{x}_{ij}$ are observed attributes, $p_{ij}$ is price, $\xi_j$ is an
unobserved (to the analyst) product-quality term, and $\varepsilon_{ij}$ is an
idiosyncratic taste shock. The consumer picks $j$ if $U_{ij} \ge U_{ik}$ for all
$k$. When $\varepsilon_{ij}$ is i.i.d. type-I extreme value, the choice
probabilities take the closed-form **multinomial logit**,
$$
P_{ij} = \frac{\exp(V_{ij})}{\sum_{k=1}^{J}\exp(V_{ik})},
$$ {#eq-sm-logit}
which is the likelihood contribution maximized in @eq-sm-mle. Two structural
warnings travel with @eq-sm-logit. First, the logit imposes **independence of
irrelevant alternatives** (IIA): the relative odds of two alternatives are
unaffected by any third, an implausible substitution pattern that the **nested
logit** of @heiss2002structural relaxes by grouping close substitutes into nests.
Second, the quality term $\xi_j$ is correlated with price (better products are
priced higher), so $\alpha$ is **endogenous**—the same simultaneity as in
@sec-sm-demand-supply, now inside a choice model—and is corrected with cost or
competitor-price instruments.

A practical route to estimation, due to Berry, *inverts* the share equation. With an
outside option $j=0$ whose mean utility is normalized to zero, @eq-sm-logit implies
the linear relation
$$
\ln s_j - \ln s_0 = \mathbf{x}_{j}^{\top}\boldsymbol{\beta} - \alpha p_{j} + \xi_j,
$$ {#eq-sm-berry-inversion}
so the left-hand side—an observed function of market shares—equals mean utility, and
the demand parameters can be estimated by regressing it on attributes and price. The
example below builds one market of many products, computes the inverted shares of
@eq-sm-berry-inversion, and contrasts a naive regression (which ignores the
correlation between price and $\xi_j$) with a two-stage least-squares estimator that
instruments price with a cost shifter.

```{r sm-logit, message=FALSE, warning=FALSE}
set.seed(29)
J    <- 200; N <- 4e5               # products in one market; consumers sampled
x    <- rnorm(J)                    # observed product attribute
xi   <- rnorm(J)                    # unobserved quality -> source of endogeneity
cost <- runif(J, 0, 2)             # cost shifter -> instrument for price
beta_true <- 1.0; alpha_true <- 1.5

# Price rises in unobserved quality (endogeneity) and in the cost shifter
price <- 1.0 + 0.8 * xi + cost + rnorm(J, 0, 0.3)
delta <- x * beta_true - alpha_true * price + xi   # mean utility of products 1..J

# Logit shares with an outside good (utility 0); add sampling noise via multinomial
num    <- exp(delta)
s      <- num / (1 + sum(num)); s0 <- 1 / (1 + sum(num))
counts <- rmultinom(1, N, c(s0, s))[, 1]
share  <- counts[-1] / N; share0 <- counts[1] / N

y <- log(share) - log(share0)       # Berry inversion: y_j = delta_j

ols  <- coef(lm(y ~ x + price))                 # ignores xi: endogenous, biased
phat <- fitted(lm(price ~ x + cost))            # first stage: cost instruments price
iv   <- coef(lm(y ~ x + phat))                  # 2SLS price coefficient

cat("True price coef (-alpha):", -alpha_true, "\n")
cat("Naive logit price coef:  ", round(ols["price"], 3), "\n")
cat("IV logit price coef:     ", round(iv["phat"], 3), "\n")
```

The naive coefficient is badly attenuated toward zero—the unobserved quality $\xi_j$
raises both utility and price, masking the true price sensitivity—while the
instrumented estimate recovers $-\alpha$. This is the lesson that recurs across every
structural demand paper: the price coefficient is the parameter most contaminated by
the unobserved quality term, and a valid instrument is the remedy.

## Dynamics, Learning, and Forward-Looking Agents {#sec-sm-dynamic}

Many marketing decisions are intertemporal—when to buy a durable, whether a
salesperson exerts effort today for a bonus tomorrow, how a relationship evolves—and
a static choice model cannot represent them. **Dynamic structural models** treat the
agent as solving a sequential optimization, and they are the technically hardest but
most rewarding members of the family.

The canonical object is a forward-looking agent who, in state $s_t$, chooses action
$a_t$ to maximize the expected present value of a stream of payoffs,
$$
V(s_t) = \max_{a_t}\;
  \Big\{ u(s_t, a_t; \boldsymbol{\theta})
       + \delta\, \mathbb{E}\!\left[\, V(s_{t+1}) \mid s_t, a_t \,\right] \Big\},
$$ {#eq-sm-bellman}
the **Bellman equation**, where $\delta$ is the discount factor, $u(\cdot)$ the
per-period payoff, and the expectation runs over the law of motion of the state.
Estimation recovers the payoff primitives $\boldsymbol{\theta}$ from observed
state–action sequences, typically by nesting the solution of @eq-sm-bellman inside a
likelihood. The substantive range is wide. @allenby1999dynamic model purchase
*timing* in direct marketing as a dynamic decision, recovering when a household will
next buy. @gowrisankaran2012dynamics let consumer preferences evolve and apply the
machinery to durable-goods markets, where today's purchase forecloses tomorrow's.
@hitsch2006empirical study optimal product launch and exit under demand uncertainty,
treating the firm as a forward-looking decision maker learning about an uncertain
market. @netzer2008hidden model the dynamics of customer–firm relationships as a
**hidden Markov** process, in which a latent relationship state migrates over time
and governs observable purchase behavior—a structural account of relationship
"states" rather than a reduced-form churn score.

Forward-looking *firms* appear on the compensation side of the same coin.
@misra2011structural estimate a structural model of sales-force incentives in which
the salesperson chooses unobserved effort given the compensation plan, the firm
designs the plan anticipating that response, and the recovered primitives (the cost
of effort, the agent's risk attitude) support counterfactual plans the firm never
tried—work that @kim2022structural extend to a multitasking sales force allocating
effort across acquisition and retention. On the demand side over the customer
lifecycle, @gupta2006modeling synthesize structural and other models of customer
lifetime value, and @kamakura2005choice embed choice models in customer-relationship
strategy. @fig-sm-taxonomy organizes the marketing applications cited in this
chapter by the decision they model.

```{mermaid}
%%| label: fig-sm-taxonomy
%%| fig-cap: "A taxonomy of structural marketing applications by the decision being modeled. Static discrete-choice and demand models sit on the left; dynamic, forward-looking models on the right. Each leaf names a representative study from this chapter."
flowchart LR
    R[Structural marketing models] --> S[Static]
    R --> D[Dynamic / forward-looking]
    S --> S1[Demand & price elasticity<br/>montgomery1999estimating]
    S --> S2[Discrete choice<br/>dube2002structural · heiss2002structural]
    S --> S3[Competitive conduct<br/>kadiyali2001structural]
    D --> D1[Purchase timing & durables<br/>allenby1999dynamic · gowrisankaran2012dynamics]
    D --> D2[Customer relationships<br/>netzer2008hidden · gupta2006modeling]
    D --> D3[Salesforce incentives<br/>misra2011structural · kim2022structural]
    D --> D4[Product launch & exit<br/>hitsch2006empirical]
```

The dynamic apparatus also extends to settings where behavior and content
co-evolve. @goh2013social model the co-evolution of user behavior in social media,
and @rutz2011modeling use a structural model to disentangle the *indirect* effects
of paid search advertising—how a search click today shifts later behavior—an effect
invisible to a single-period attribution model. Price promotions, too, have dynamic
consequences: @elberg2019dynamic study how promotions move purchases across time
rather than merely expanding the category, a stockpiling response that only a
dynamic model represents correctly.

## Identification: What Can and Cannot Be Recovered {#sec-sm-identification}

A model is **identified** if distinct parameter values imply distinct distributions
of the observable data; without identification, no estimator—however clever—can
recover the truth, and a perfectly converged optimizer may report a number that the
data do not pin down. Because structural models lean on assumptions for
identification, naming the threats is as important as writing the likelihood.

The first and most pervasive threat is **endogeneity**, already met twice: price is
set by agents who observe demand shocks the analyst does not, so the price
coefficient is identified only through an exclusion restriction—a cost or competitor
instrument that shifts the choice set without entering utility directly. When no
valid instrument exists, the elasticity is not identified, full stop.

The second is **functional-form dependence**. In a static logit the substitution
pattern is dictated by the IIA assumption rather than estimated, so a counterfactual
that turns on cross-elasticities is identified by assumption, not by data; relaxing
to nested or random-coefficient logit buys realism at the cost of additional
parameters that themselves require variation to identify.

The third is the **discount factor** in dynamic models. The forward-looking payoff
in @eq-sm-bellman and the per-period payoff are notoriously hard to separate: a
patient agent with low effort costs can mimic an impatient agent with high effort
costs, so $\delta$ is typically *fixed* rather than estimated, with the
counterfactual's credibility contingent on that choice.

The fourth is the **equilibrium and conduct assumption**: inferring competitive
conduct from price–quantity comovement [@kadiyali2001structural] identifies conduct
only if the maintained equilibrium concept is correct, and a misspecified conduct
parameter contaminates the recovered costs.

The pattern across all four is the same and is worth internalizing as the chapter's
governing principle: structural identification trades *data variation* for
*assumptions*, and every counterfactual inherits the assumptions that identified the
parameters it uses. The honest structural paper, accordingly, reports which results
are robust to weakening which assumption—and the practitioner reading one should ask,
of every headline number, *what assumption is this resting on?*

## Getting Started

For readers building toward this literature, the graduate econometrics texts of
@greene2003econometric and @cameron2005microeconometrics supply the estimation
foundations; @hensher2005applied is the standard reference for the discrete-choice
models that anchor most marketing applications; @diamantopoulos2013quantitative
surveys quantitative modeling in marketing more broadly; and the *Handbook of
Econometrics* policy-evaluation chapters [@heckman2007econometric1;
@heckman2007econometric; @abbring2007econometric3] connect structural estimation to
the identification of treatment effects. A productive path is to reproduce a linear
demand-and-supply system (@sec-sm-demand-supply), then a logit with endogenous price
(@sec-sm-discrete-choice), before attempting a dynamic model—each step adds one
layer of structure, and the endogeneity intuition learned in the first carries
unchanged into the last.

## Key Takeaways

- A model is **structural** when its parameters are economic primitives—tastes,
  costs, information, conduct—and are therefore policy invariant; the payoff is
  counterfactual simulation of policies never observed (@eq-sm-structural).
- Structural modeling and causal inference are complements: causal estimates
  discipline structural models, and structural reasoning locates the endogeneity a
  causal design must address.
- **Endogeneity of price** is the central econometric problem, visible already in
  linear supply and demand (@eq-sm-equilibrium) and recurring inside discrete-choice
  demand; it is resolved only by a credible exclusion restriction.
- **MLE** (full likelihood, efficient if correct) and **GMM** (moment conditions,
  robust) are the two workhorse estimators; the choice is about how much of the
  data-generating process one is willing to specify (@tbl-sm-mle-gmm).
- **Dynamic** models recover forward-looking primitives via the Bellman equation
  (@eq-sm-bellman) but face hard identification of the discount factor.
- Every counterfactual inherits the assumptions that identified its parameters; the
  credible structural paper reports which conclusions survive weakening which
  assumption (@sec-sm-identification).
