# Measurement Scales {#sec-measurement-scales}

Most of the constructs marketing cares about—brand equity, satisfaction,
involvement, trust, perceived authenticity—cannot be read off a receipt. They are
*latent*: properties of a consumer's mind that leave only indirect traces in
answers to survey questions, in choices, and in behavior. A **measurement scale**
is the instrument that turns those traces into numbers we can analyze, and the
**measurement model** is the formal theory that says how the numbers relate to the
construct they are supposed to capture. Getting the measurement model wrong is not
a cosmetic error. It silently biases every downstream coefficient, invalidates the
reliability and validity tests a reviewer will demand, and—most consequentially—can
reverse a managerial recommendation, because the model dictates whether a manager
should strengthen a coherent latent disposition or intervene on specific,
possibly uncorrelated, drivers.

This chapter develops measurement as a modeling problem rather than a checklist.
We begin with the classical theory that underwrites almost all psychometrics—**true-score
theory**—and use it to define reliability precisely. We then draw the central
distinction of the modern literature, between **reflective** and **formative**
measurement, and make it formal with structural equations, including the
identification conditions that decide whether a model can be estimated at all. With
that machinery in place we walk through **scale development** as a disciplined
sequence, define the standard battery of **reliability and validity** statistics
and give their estimators and assumptions, and confront **common-method bias**—the
contamination that arises when the same respondent supplies both predictor and
outcome through the same instrument. Throughout, the goal is the one set in
@sec-branding when it asked whether brand equity is reflective or formative: to let
the reader decide a construct's measurement model on principled grounds and defend
that decision to a skeptical referee.

A note on scope. We treat *psychometric* measurement—scales built from multiple
survey items—because that is where the reflective/formative question is sharpest and
where the field's two leading handbooks of marketing scales live [@bearden2011;
@stewart1993]. The logic transfers directly to behavioral and text-derived measures,
which are developed in @sec-text-as-data.

## True-Score Theory and the Definition of Reliability

The foundation of classical test theory is disarmingly simple. An observed score
$X$ on an item is the sum of a **true score** $T$—the construct value we wish we
could see—and a **measurement error** $E$:

$$
X = T + E .
$$ {#eq-truescore}

The theory's content is in three assumptions about $E$: it has mean zero,
$\mathbb{E}[E] = 0$, so $T = \mathbb{E}[X]$ is *defined* as the expected observed
score; it is uncorrelated with the true score, $\mathrm{Cov}(T, E) = 0$; and errors
on distinct items are mutually uncorrelated. Under these assumptions the observed
variance decomposes additively,

$$
\sigma_X^2 = \sigma_T^2 + \sigma_E^2 ,
$$ {#eq-varsplit}

and **reliability** is *defined* as the share of observed variance that is true
variance,

$$
\rho_{XX'} \;=\; \frac{\sigma_T^2}{\sigma_X^2} \;=\; \frac{\sigma_T^2}{\sigma_T^2 + \sigma_E^2} \in [0,1].
$$ {#eq-30-reliability}

Reliability is therefore not "consistency" in a loose sense; it is a variance ratio,
equal to the squared correlation between observed and true scores and to the
correlation between two *parallel* measurements of the same construct (hence the
notation $\rho_{XX'}$). It says nothing about whether $T$ measures the *right*
construct—that is validity, and the two must never be conflated. A bathroom scale
that always reads three kilograms heavy is perfectly reliable and invalid.

Reliability matters beyond bookkeeping because unreliable measurement *attenuates*
relationships. If a predictor $X$ measures a true $T$ with reliability $\rho_{XX}$
and an outcome $Y$ measures $U$ with reliability $\rho_{YY}$, the observable
correlation is shrunk toward zero relative to the construct-level correlation:

$$
\mathrm{Corr}(X, Y) \;=\; \mathrm{Corr}(T, U)\,\sqrt{\rho_{XX}\,\rho_{YY}} .
$$ {#eq-30-attenuation}

@eq-30-attenuation is the *correction for attenuation*. It is why a true effect can be
buried by a noisy scale, and why latent-variable models—which estimate $\mathrm{Corr}(T,U)$
directly—are preferred to regressions on summed scores when reliability is modest.

> Multiple-item measurement exists almost entirely to beat down $\sigma_E^2$. By the
> Spearman–Brown logic, averaging $k$ interchangeable items whose errors are
> independent divides the error variance of the composite by $k$ while leaving the
> true variance intact, so a composite of many noisy items can be far more reliable
> than any single item.[^sb] This is the engine behind multi-item scales—and, as we
> will see, it is exactly the property a *formative* index does not have.

[^sb]: The Spearman–Brown prophecy formula gives the reliability of a test lengthened
by a factor $k$ as $\rho_k = k\rho_1 / [1 + (k-1)\rho_1]$, where $\rho_1$ is the
reliability of the original test. It assumes the added items are *parallel*—equal
true-score loadings and equal error variances—an assumption formative indicators
violate by construction. <!-- CITE-NEEDED: Spearman 1910 / Brown 1910 prophecy formula -->

## Reflective versus Formative Measurement

The single most consequential decision in scale construction is the direction of
causality between the latent construct and its indicators. The choice is not a
matter of taste; it is a substantive claim about how the world works, and it
determines which model is even identified [@coltman2008; @diamantopoulos2013quantitative].

### The Reflective Model

In a **reflective** (effect-indicator) model, the latent construct *causes* its
indicators: the items are manifestations, symptoms, or reflections of an underlying
common cause. Let $\eta$ be a scalar latent construct and $x_1,\dots,x_K$ its
observed indicators. The model is the common-factor model,

$$
x_k \;=\; \lambda_k\, \eta \;+\; \varepsilon_k, \qquad k = 1,\dots,K,
$$ {#eq-30-reflective}

with $\mathbb{E}[\varepsilon_k]=0$, $\mathrm{Cov}(\eta,\varepsilon_k)=0$, and
$\mathrm{Cov}(\varepsilon_j,\varepsilon_k)=0$ for $j\neq k$ (uncorrelated errors).
The arrows run *from* $\eta$ *to* each $x_k$. Three properties follow and are
diagnostic. First, because the indicators share a single common cause, they must be
**positively correlated**; the off-diagonal covariances are $\mathrm{Cov}(x_j,x_k)=\lambda_j\lambda_k\,\sigma_\eta^2$,
all of the same sign as the loadings. Second, the indicators are
**interchangeable**: dropping one changes the construct's domain coverage but not
its meaning, since each is a noisy replica of the same $\eta$. Third, internal
consistency (Cronbach's alpha, composite reliability) is the *appropriate* quality
criterion, precisely because @eq-30-reflective is the model under which @eq-30-reliability
was derived.

A reflective construct is the right model when the construct is a real, unitary
attribute that radiates into measurable symptoms—the way a fever radiates into a
high thermometer reading, a flushed face, and chills. Brand love
[@batra2012; @bagozzi2016], satisfaction, and attitude are conventionally modeled
this way: a more satisfied customer answers *every* satisfaction item more
favorably, so the items co-move.

### The Formative Model

In a **formative** (causal-indicator) model, the indicators *cause* the construct:
the latent variable is a composite *formed by* its indicators, which are defining
ingredients rather than reflections. The structural equation reverses the arrows,

$$
\eta \;=\; \sum_{k=1}^{K} \gamma_k\, x_k \;+\; \zeta,
$$ {#eq-30-formative}

where $\gamma_k$ are formative weights and $\zeta$ is a disturbance capturing the
part of $\eta$ not explained by the chosen indicators, with
$\mathrm{Cov}(x_k,\zeta)=0$. Everything that was diagnostic for the reflective model
is now reversed. The indicators need **not** be correlated—socioeconomic status
formed from income, education, and occupation does not require that high earners be
highly educated. The indicators are **not interchangeable**: removing one (say,
income) deletes part of the construct's *definition*, not merely part of its
sampling. And internal-consistency reliability is **inapplicable**: there is no
common cause to make the items co-vary, so a low alpha is no defect. Quality is
judged instead by indicator weights, their significance, and **multicollinearity**
among indicators (high collinearity makes the weights unstable and
uninterpretable), typically diagnosed with the variance inflation factor.

A construct should be formative when it is constituted by a combination of distinct
facets that need not move together. Classic examples are an index of *life stress*
(a job loss and a marriage are both stressors but are uncorrelated causes), a
*marketing capability* composite, or a *deprivation* index. @sec-branding raised the
live debate over whether **brand equity** and **brand authenticity** are reflective
or formative; the two leading authenticity scales take opposite stances—@morhart2015
reflective, @nunes2021 formative—and the disagreement is exactly a disagreement about
the direction of @eq-30-reflective versus @eq-30-formative.

### A Formal Contrast and the MIMIC Bridge

@fig-meas-models contrasts the two path structures, and the table that follows
states the operational consequences side by side.

```{mermaid}
%%| label: fig-meas-models
%%| fig-cap: "Reflective versus formative measurement. In the reflective model the latent construct η causes the indicators (arrows out, each indicator carries its own error ε). In the formative model the indicators cause the construct (arrows in, a single disturbance ζ attaches to the construct). A MIMIC model combines both: formative causes and reflective indicators of the same latent variable."
flowchart LR
  subgraph R["Reflective"]
    direction TB
    ETA1((η)) --> X1[x1]
    ETA1 --> X2[x2]
    ETA1 --> X3[x3]
    X1 --- E1((ε1))
    X2 --- E2((ε2))
    X3 --- E3((ε3))
  end
  subgraph F["Formative"]
    direction TB
    Z1[x1] --> ETA2((η))
    Z2[x2] --> ETA2
    Z3[x3] --> ETA2
    ZETA((ζ)) --- ETA2
  end
```

| Property | Reflective (@eq-30-reflective) | Formative (@eq-30-formative) |
|---|---|---|
| Causal direction | Construct → indicators | Indicators → construct |
| Indicator correlation | High, same sign (required) | Not required; any sign |
| Interchangeability | Indicators interchangeable | Indicators define the construct |
| Dropping an indicator | Narrows sampling of the domain | Changes the construct's meaning |
| Error term | One per indicator ($\varepsilon_k$) | One per construct ($\zeta$) |
| Reliability criterion | Internal consistency (alpha, CR, AVE) | Indicator weights; collinearity (VIF) |
| Identification | Standard (factor model) | Needs external structure (e.g., MIMIC) |

: Operational consequences of the reflective–formative distinction. {#tbl-refl-form}

The deepest difference in @tbl-refl-form is the last row. A pure formative model is
**not identified** in isolation: @eq-30-formative places no testable restrictions on the
indicator covariances, the scale and location of $\eta$ are arbitrary, and the
disturbance variance $\sigma_\zeta^2$ cannot be separated from the weights without
further information. Identification requires that the formative construct emit *at
least two reflective indicators* (or two paths to downstream constructs), which fixes
its metric and disturbance. The resulting hybrid is the **multiple-indicators,
multiple-causes (MIMIC)** model: formative causes $x_1,\dots,x_K$ feed a latent
$\eta$, which in turn emits reflective indicators $y_1,\dots,y_M$,

$$
\eta \;=\; \sum_{k=1}^{K}\gamma_k x_k + \zeta,
\qquad
y_m \;=\; \lambda_m \eta + \varepsilon_m, \quad m = 1,\dots,M .
$$ {#eq-mimic}

The MIMIC structure is the standard way marketing accommodates a genuinely formative
construct inside an estimable structural-equation model [@diamantopoulos2013quantitative].
It also clarifies a common error: estimating @eq-30-formative by ordinary least squares
treats $\eta$ as observed and conflates the formative weights with a regression of an
*index* on its parts, discarding the disturbance and the very latency that motivated
the construct. <!-- CITE-NEEDED: Jarvis, MacKenzie & Podsakoff 2003 on misspecification of formative vs reflective constructs -->
<!-- CITE-NEEDED: Bollen & Lennox 1991 conventional wisdom on measurement -->

### Why Misspecification Is Costly

Treating a formative construct as reflective (or the reverse) is not a venial sin.
Forcing reflective machinery onto formative indicators discards items that fail to
correlate—dropping low-alpha items *narrows the construct's content*, biasing it
toward whatever facet happens to be internally consistent. Conversely, modeling a
reflective construct formatively throws away the error structure that
@eq-30-attenuation depends on and inflates apparent validity. Simulation and analytic
work show the resulting structural-coefficient bias can be large and of either sign,
which is why the direction-of-causality decision belongs at the *start* of scale
development, justified on conceptual grounds, not discovered after the data are in.
<!-- CITE-NEEDED: MacKenzie, Podsakoff & Jarvis 2005 on measurement-model misspecification consequences -->

## Scale Development

Scale development is the disciplined translation of a construct definition into a
validated instrument. The field's canonical paradigm is a staged procedure that
moves from conceptual specification to a purified, validated multi-item scale, and
modern treatments differ mainly in how much weight they place on the formative
possibility [@bearden2011]. The stages below are reflective by default; we flag where
a formative construct diverges.

1. **Construct definition and domain specification.** State precisely what the
   construct is and is not, its conceptual boundaries, and—critically—the *direction
   of measurement* (@tbl-refl-form). This step decides everything downstream and
   cannot be repaired later by statistics. <!-- CITE-NEEDED: Churchill 1979 paradigm for developing marketing measures -->
2. **Item generation.** Draft a large pool of candidate items spanning the domain,
   from theory, qualitative interviews, and existing scales. For reflective
   constructs the pool should oversample each facet (items are replaceable); for
   formative constructs the pool must *exhaust* the facets (each is constitutive).
3. **Content validity / expert judging.** Have domain experts rate each item's
   relevance and representativeness, retaining items with high agreement. This is a
   qualitative check that no statistic substitutes for.
4. **Purification on a calibration sample.** Collect data and trim items. For
   reflective scales, drop items with low item–total correlations or weak loadings
   and inspect exploratory factor structure for dimensionality. For formative
   indices, *do not* trim on inter-item correlation; trim only on collinearity and
   weight significance.
5. **Confirmatory validation on a fresh sample.** Fit the hypothesized measurement
   model (confirmatory factor analysis for reflective constructs; a MIMIC model for
   formative) and assess fit, reliability, and the validity battery of
   @sec-reliability-validity below. Cross-validation on data not used for
   purification guards against capitalizing on chance.
6. **Norming and replication.** Establish means, variances, and stability across
   populations and over time, and demonstrate the nomological network—correlations
   with antecedents and consequences in the predicted pattern.

@fig-scale-dev renders the pipeline. The arrows back from validation to item
generation are essential: scale development is iterative, and a construct that fails
discriminant validity often needs its *definition* sharpened, not just its items
re-fitted.

```{mermaid}
%%| label: fig-scale-dev
%%| fig-cap: "The scale-development pipeline. Conceptual specification fixes the measurement direction; item generation and content judging are qualitative; purification and confirmatory validation are quantitative and run on separate samples. Validation failures feed back to redefinition, not merely re-estimation."
flowchart TB
  A[Define construct\n& measurement direction] --> B[Generate item pool]
  B --> C[Expert content judging]
  C --> D[Purify on\ncalibration sample]
  D --> E[Confirm on\nfresh sample]
  E --> F[Norm & replicate\nnomological network]
  E -. fails discriminant/convergent .-> A
  D -. weak items .-> B
```

## Reliability and Validity {#sec-reliability-validity}

A scale must be both **reliable** (low error variance) and **valid** (measures the
intended construct). We give the standard estimators, their assumptions, and what
breaks them.

### Internal-Consistency Reliability

**Cronbach's alpha** estimates reliability from a single administration of $K$
items as

$$
\alpha \;=\; \frac{K}{K-1}\left(1 - \frac{\sum_{k=1}^{K}\sigma_{x_k}^2}{\sigma_{X_{\text{tot}}}^2}\right),
$$ {#eq-30-alpha}

where $\sigma_{x_k}^2$ is item variance and $\sigma_{X_{\text{tot}}}^2$ the variance
of the summed score. Alpha is a *lower bound* on reliability and equals it only under
**tau-equivalence**—all items load equally on the common factor. When loadings differ
(the congeneric case), alpha understates reliability, and **composite reliability**
(CR), computed from the CFA loadings $\hat\lambda_k$ and error variances
$\hat\theta_k$,

$$
\mathrm{CR} \;=\; \frac{\left(\sum_{k}\hat\lambda_k\right)^2}{\left(\sum_{k}\hat\lambda_k\right)^2 + \sum_{k}\hat\theta_k},
$$ {#eq-cr}

is preferred because it does not assume equal loadings. Both assume a *reflective*
unidimensional model; **neither is meaningful for a formative index**, where the
items are not expected to covary. Conventional thresholds put acceptable internal
consistency at roughly $0.70$ and above, though the number is a heuristic, not a law.
<!-- CITE-NEEDED: Cronbach 1951 coefficient alpha -->

### Convergent and Discriminant Validity

**Construct validity** asks whether the scale measures its target construct.
@fornell1981a give the workhorse criteria for reflective constructs estimated by CFA.
**Convergent validity**—indicators of one construct converge—is evidenced when the
**average variance extracted** (AVE),

$$
\mathrm{AVE} \;=\; \frac{\sum_{k}\hat\lambda_k^2}{\sum_{k}\hat\lambda_k^2 + \sum_{k}\hat\theta_k},
$$ {#eq-30-ave}

exceeds $0.50$, i.e., the construct explains more than half the variance in its
indicators. **Discriminant validity**—distinct constructs are empirically
distinct—was traditionally assessed by the **Fornell–Larcker criterion**: each
construct's $\sqrt{\mathrm{AVE}}$ must exceed its correlation with every other
construct [@fornell1981a]. This criterion is intuitive but has poor statistical
power in Monte Carlo simulations: the cross-loadings criterion—checking that each
indicator loads more strongly on its own construct than on others—has near-zero
sensitivity in variance-based SEM [@henseler2015]. The current recommended criterion
for variance-based SEM is the **heterotrait–monotrait ratio of correlations (HTMT)**
and its improved successor **HTMT2** [@henseler2015; @roemer2021].

The HTMT is defined as the ratio of (a) the average of all cross-construct
inter-indicator correlations (heterotrait–heteromethod correlations) to (b) the
geometric mean of the average within-construct inter-indicator correlations
(monotrait–heteromethod correlations),

$$
\mathrm{HTMT} = \frac{\bar{r}_{\text{HT}}}{\sqrt{\bar{r}_{A} \cdot \bar{r}_{B}}},
$$ {#eq-htmt}

where $\bar{r}_{\text{HT}}$ is the mean of all correlations between indicators of
different constructs and $\bar{r}_A$, $\bar{r}_B$ are the mean within-construct
correlations for each construct. A value of HTMT below $0.85$ (or $0.90$ in
exploratory work) is evidence of discriminant validity; values approaching 1 indicate
that the two constructs are empirically indistinguishable. HTMT2 [@roemer2021]
replaces the arithmetic mean in the denominator with a geometric mean of the
individual within-construct correlations, producing a consistent estimator of the
inter-construct correlation under congeneric (heterogeneous loading) models; an
816,000-condition Monte Carlo study found HTMT2 less biased than HTMT when loadings
are heterogeneous. For covariance-based SEM (CFA), discriminant validity is better
assessed by a $\chi^2$ difference test or confidence-interval test comparing a
free-correlation model to one constraining the inter-construct correlation to 1.

A construct that fails discriminant validity against a neighbor is not separately
measurable from it, and the usual remedy is conceptual: merge the constructs or
sharpen their definitions. These criteria are reflective; for formative
constructs, validity is assessed through the nomological net and the significance of
formative weights instead.

### Criterion and Nomological Validity

**Criterion validity** is the scale's correlation with an external criterion it
should predict—*concurrent* when measured contemporaneously, *predictive* when the
criterion is future behavior. **Nomological validity**, the most demanding, asks
whether the construct relates to *other* constructs as theory dictates: a valid brand-equity
scale should predict price premium and choice share in the direction
@sec-branding specifies. Validity, unlike reliability, is never established by a
single coefficient; it is the accumulated weight of a construct behaving as its
theory says it should.

### Inter-Rater Reliability for Coded Data

When measurement comes from human coders rather than respondents—coding ad content,
classifying reviews, labeling images—reliability is *agreement* between raters
corrected for chance. **Cohen's kappa** [@Cohen_1960] for two raters is

$$
\kappa \;=\; \frac{p_o - p_e}{1 - p_e},
$$ {#eq-30-kappa}

where $p_o$ is observed agreement and $p_e$ is the agreement expected if raters
labeled independently at their marginal rates. $\kappa = 1$ is perfect agreement,
$\kappa = 0$ is chance-level; conventional benchmarks read $0.41$–$0.60$ as
moderate and above $0.60$ as substantial [@Landis_1977]. For continuous ratings, the
**intraclass correlation coefficient** (ICC) plays the analogous role, partitioning
variance into rater, target, and error components [@Shrout_1979]. The worked example
below estimates kappa for a two-coder labeling task.

```{r kappa-example}
#| echo: true
#| message: false
#| warning: false
set.seed(42)

# Two coders label 200 reviews as positive (1) or negative (0).
# Simulate a shared latent truth plus idiosyncratic coder noise.
n      <- 200
truth  <- rbinom(n, 1, 0.5)
flip_a <- rbinom(n, 1, 0.10)          # coder A errs 10% of the time
flip_b <- rbinom(n, 1, 0.12)          # coder B errs 12% of the time
coder_a <- ifelse(flip_a == 1, 1 - truth, truth)
coder_b <- ifelse(flip_b == 1, 1 - truth, truth)

tab <- table(coder_a, coder_b)
p_o <- sum(diag(tab)) / sum(tab)                       # observed agreement
p_e <- sum(rowSums(tab) * colSums(tab)) / sum(tab)^2   # chance agreement
kappa <- (p_o - p_e) / (1 - p_e)

cat("Observed agreement p_o:", round(p_o, 3), "\n")
cat("Chance agreement   p_e:", round(p_e, 3), "\n")
cat("Cohen's kappa        :", round(kappa, 3), "\n")
```

### A Worked Reflective Measurement Model

To make @eq-30-reflective through @eq-30-ave concrete, we simulate a single reflective
construct with four congeneric indicators, recover the loadings by factor analysis,
and compute composite reliability and AVE from the estimates.

```{r reflective-cfa}
#| echo: true
#| message: false
#| warning: false
set.seed(2024)

n       <- 500
lambda  <- c(0.85, 0.80, 0.75, 0.70)   # true (heterogeneous) loadings
theta   <- 1 - lambda^2                 # error variances (unit-variance items)
eta     <- rnorm(n)                     # latent construct, Var = 1

# Generate indicators under the reflective model x_k = lambda_k * eta + e_k
X <- sapply(seq_along(lambda), function(k) lambda[k] * eta + rnorm(n, sd = sqrt(theta[k])))
colnames(X) <- paste0("x", seq_along(lambda))

# Recover a one-factor solution; loadings are the std. correlations with the factor
fa <- factanal(X, factors = 1, scores = "none")
lhat <- as.numeric(fa$loadings[, 1])     # estimated loadings
ehat <- 1 - lhat^2                       # implied error variances

CR  <- sum(lhat)^2 / (sum(lhat)^2 + sum(ehat))
AVE <- sum(lhat^2) / (sum(lhat^2) + sum(ehat))

# Cronbach's alpha from the item covariance matrix
S       <- cov(X)
K       <- ncol(X)
alpha   <- (K / (K - 1)) * (1 - sum(diag(S)) / sum(S))

cat("Estimated loadings:", round(lhat, 3), "\n")
cat("Cronbach's alpha  :", round(alpha, 3), "\n")
cat("Composite reliab. :", round(CR, 3), "\n")
cat("AVE               :", round(AVE, 3), "(> 0.50 => convergent)\n")
```

Because the loadings are heterogeneous, alpha will sit *below* composite reliability
in this example—the tau-equivalence gap of @eq-30-alpha made visible. AVE above $0.50$
confirms convergent validity for the simulated construct.

## Common-Method Bias

A pervasive threat to survey-based measurement is **common-method variance** (CMV):
variance attributable to the *measurement method* rather than the constructs the
measures represent. When the same respondent rates both an independent and a
dependent variable, on the same scale, at the same sitting, shared method
factors—consistency motifs, social desirability, acquiescence, common scale anchors,
transient mood—inject a spurious component into every item. The bias, **common-method
bias** (CMB), is the distortion this induces in observed relationships, and it can
inflate *or* deflate correlations [@mackenzie1986].

### A Formal Model of the Bias

Augment the reflective model @eq-30-reflective with a method factor $M$ common to all
items measured by the shared method:

$$
x_k \;=\; \lambda_k\, \eta \;+\; \omega_k\, M \;+\; \varepsilon_k,
$$ {#eq-method-factor}

with $M \perp \eta$, $M \perp \varepsilon_k$, and method loadings $\omega_k$. Two
items measuring *different* substantive constructs $\eta$ and $\eta'$ now share the
method factor, so their observed covariance picks up a method term that has nothing
to do with their true relationship:

$$
\mathrm{Cov}(x_j, y_l) \;=\; \underbrace{\lambda_j \lambda_l\,\mathrm{Cov}(\eta,\eta')}_{\text{substantive}}
\;+\; \underbrace{\omega_j \omega_l\,\sigma_M^2}_{\text{method bias}} .
$$ {#eq-cmb-bias}

When predictor and outcome share the method, the second term is added to *every*
cross-construct covariance, biasing structural estimates—usually upward, because
method loadings of the same sign inflate correlations, though differential signs can
deflate. The contaminant is most dangerous in single-source, single-instrument
designs: exactly the cross-sectional self-report survey that dominates applied
marketing.

### Procedural and Statistical Remedies

Defenses fall into two families. **Procedural** remedies design the bias out before
data collection: obtain predictor and outcome measures from *different sources* (e.g.,
attitudes from the consumer, behavior from the firm's records) or at different times;
separate the measurement of constructs psychologically or temporally;
counterbalance item order; protect respondent anonymity to blunt social desirability;
and write items to reduce ambiguity and common scale anchoring. Separating the source
of the predictor from the source of the outcome severs the $M$ that links them in
@eq-cmb-bias and is the single most effective remedy. <!-- CITE-NEEDED: Podsakoff, MacKenzie, Lee & Podsakoff 2003 common method biases -->

**Statistical** remedies detect or partial out the method factor after the fact, and
none is a substitute for good design. Harman's single-factor test—loading all items
on one unrotated factor and checking that it explains less than half the
variance—is widely reported but weak, detecting only gross contamination. Stronger is
to estimate @eq-method-factor directly: include an unmeasured **latent method
factor** in the CFA, let all indicators load on it alongside their substantive
constructs, and compare structural estimates with and without it. A **marker-variable**
approach uses a theoretically unrelated marker construct as a proxy for $M$ and
partials its correlation out of the substantive correlations. The diagnostic below
implements the marker-variable logic on simulated data containing a known method
component.

```{r cmb-marker}
#| echo: true
#| message: false
#| warning: false
set.seed(7)

n   <- 800
eta1 <- rnorm(n); eta2 <- rnorm(n)            # two substantive constructs
M    <- rnorm(n)                               # shared method factor

# Substantive correlation between constructs is set to ZERO here:
# any observed X-Y correlation is pure method bias.
w <- 0.5                                        # common method loading
x <- 0.8 * eta1 + w * M + rnorm(n, sd = 0.3)    # predictor (one item)
y <- 0.8 * eta2 + w * M + rnorm(n, sd = 0.3)    # outcome   (one item)
marker <- 0.0 * eta1 + w * M + rnorm(n, sd = 0.3)  # theoretically unrelated marker

raw_xy <- cor(x, y)

# Marker-variable correction: subtract the method-induced correlation,
# proxied by the smallest observed correlation with the marker.
r_xm <- cor(x, marker); r_ym <- cor(y, marker)
r_m  <- min(r_xm, r_ym)                         # conservative marker correlation
adj_xy <- (raw_xy - r_m) / (1 - r_m)            # partial out shared method

cat("Raw X-Y correlation (contaminated):", round(raw_xy, 3), "\n")
cat("Marker correlation used           :", round(r_m, 3), "\n")
cat("Method-adjusted X-Y correlation   :", round(adj_xy, 3), "\n")
cat("True substantive correlation      : 0.000\n")
```

The raw correlation is spuriously positive—pure method bias, since the substantive
correlation was set to zero—and the marker adjustment pulls the estimate back toward
the truth. The lesson is the one @eq-cmb-bias formalizes: when a relationship is
estimated from a single self-report instrument, part of it may be the instrument
talking to itself, and the credible defense is procedural separation of sources, with
statistical partialling as a fallback.

## Key Takeaways

- Measurement is a *modeling* problem. Under true-score theory (@eq-truescore),
  reliability is a variance ratio (@eq-30-reliability), and unreliability *attenuates*
  every estimated relationship (@eq-30-attenuation).
- The reflective–formative distinction is a claim about the **direction of causality**
  between construct and indicators (@eq-30-reflective vs. @eq-30-formative) and dictates the
  entire validity battery; it must be decided conceptually, *before* data collection
  (@tbl-refl-form).
- A pure formative model is **not identified**; it needs external structure, most
  commonly the **MIMIC** specification (@eq-mimic), to fix its metric and disturbance.
- Internal-consistency reliability (alpha, CR, AVE) apply to **reflective**
  constructs only; formative indices are judged by weights and collinearity.
- The **Fornell–Larcker criterion** and cross-loadings assessment have poor
  sensitivity for detecting discriminant validity failures in variance-based SEM;
  **HTMT** and **HTMT2** are the current recommended criteria, with thresholds
  below 0.85 indicating acceptable discriminant validity [@henseler2015; @roemer2021].
- **Common-method bias** arises when one instrument supplies both predictor and
  outcome (@eq-cmb-bias); the strongest remedy is procedural—separate the sources—not
  a post-hoc statistical patch.

## Further Reading

The two standard compendia of validated marketing instruments are the
*Handbook of Marketing Scales* [@bearden2011] and the *Marketing Scales Handbook*
[@stewart1993]; both organize scales by construct with reliability and validity
evidence. For the reflective–formative debate, @coltman2008 and
@diamantopoulos2013quantitative give the theoretical and operational decision rules,
and @edwards2010 examines the construct-validity consequences of the choice.
@fornell1981a remains the reference for convergent and discriminant validity with
unobservable variables; @henseler2015 introduced HTMT as the preferred discriminant
validity criterion for variance-based SEM and @roemer2021 refined it to HTMT2.
@mackenzie1986 is the standard reference for method effects in marketing
measurement. The branding constructs whose measurement models are actively contested—brand
love [@batra2012; @bagozzi2016] and authenticity [@morhart2015; @nunes2021]—are
treated substantively in @sec-branding.
