# Surveys and Experiments {#sec-surveys}

Most of what marketing researchers want to know—what consumers intend to buy, how
much they value an attribute, whether a campaign caused a sale—cannot be read
directly off transactional data. It must be *elicited* or *manufactured*: elicited
through **surveys** that ask people to report latent states, or manufactured through
**experiments** that intervene on the world and observe the response. This chapter
treats the two as a matched pair. Both are instruments for recovering a quantity
that the analyst cannot observe, and both fail in characteristic, formalizable ways.
The survey's central problem is *error*—the gap between what respondents report and
the population truth—decomposed into how respondents are selected and how they
answer. The experiment's central problem is *validity*—whether the measured effect
is the causal effect, generalizes beyond the lab, and operationalizes the construct
the researcher intended.

The two instruments also interact, and not benignly. The very act of surveying a
consumer can change that consumer's later behavior, so a survey designed to *measure*
intentions can *create* the intentions it claims to observe. Recognizing such
feedback is the first step toward using either instrument credibly. By the end of the
chapter the reader should be able to state the total-survey-error decomposition,
build an unbiased estimator from a known sampling design, correct cross-respondent
incomparability with anchoring vignettes, design an experiment whose causal claim
survives scrutiny, calculate a defensible sample size, and apply the open-science
toolkit that guards against false discovery.

## Surveys as Measurement

A survey is a measurement device whose output is a set of self-reports. Two
properties make it powerful and dangerous in equal measure. It is *cheap*, so it
scales to populations and to questions—intentions, attitudes, satisfaction—that
leave no behavioral trace. And it is *reactive*: the instrument can perturb the
state it measures.

### Self-Generated Validity: When Measurement Changes Behavior

The canonical reactivity result concerns purchase intentions. Researchers routinely
ask "How likely are you to buy product $X$ in the next month?" and treat the answer
as a noisy reading of a pre-existing propensity. @chandon2005 show that the reading
is not passive. Merely asking the intentions question increases the *correspondence*
between stated intentions and subsequent purchase—a phenomenon they call
**self-generated validity**. The respondent, prompted to articulate an intention,
forms or crystallizes an attitude that did not fully exist before the question and
then acts on it. The intention–behavior link is therefore endogenous to the
measurement: it is, on average, **58% stronger among surveyed consumers than among
otherwise-comparable nonsurveyed consumers** [@chandon2005]. The implication is
sharp. An intentions survey does not merely forecast demand; it is a light-touch
*intervention* that nudges demand. A forecast calibrated on surveyed respondents will
overstate the predictability of an unsurveyed population.

> **Self-generated validity.** The increase in the predictive validity of a stated
> intention that is *caused by the act of stating it*. Asking the question
> reactively strengthens the very intention–behavior correlation the researcher
> intends to measure passively [@chandon2005].

### The Intention–Behavior Correlation

How tightly do intentions track behavior once reactivity is acknowledged? The
meta-analytic benchmark from the theory of reasoned action places the
intention–behavior correlation near $\rho \approx 0.53$ across domains
[@sheppard1988]. This is a substantial but far-from-deterministic association: stated
intentions explain on the order of $\rho^2 \approx 28\%$ of the variance in
subsequent behavior. The residual is the space in which forecasting models earn their
keep. @morwitz1992 show that **segmenting before forecasting** materially improves
predictive power: a model that uses historical purchase behavior together with stated
intention, and that allows the mapping from intention to purchase to differ across
consumer segments, forecasts sales better than a pooled model that imposes a single
intention–purchase slope. The intuition is that the intention–behavior gap is
heterogeneous—deliberate planners realize their stated intentions far more reliably
than impulsive or constrained consumers—so pooling averages over groups whose mapping
functions genuinely differ.

The practical sequence is summarized in @fig-intent-pipeline.

```{mermaid}
%%| label: fig-intent-pipeline
%%| fig-cap: "From elicited intentions to a sales forecast. The dashed feedback edge is self-generated validity: the act of measurement perturbs the behavior being forecast."
flowchart LR
    A["Elicit stated<br/>intention"] --> B["Segment respondents<br/>(planners vs. impulsive,<br/>constrained, etc.)"]
    B --> C["Estimate segment-specific<br/>intention to purchase mapping"]
    C --> D["Aggregate to<br/>sales forecast"]
    A -. "self-generated<br/>validity" .-> E["Actual purchase<br/>behavior"]
    D --> E
```

## The Total Survey Error Framework

A survey statistic can miss the truth for two broad reasons: the *people* who answer
are not the people the analyst wanted, and the *answers* they give are not the values
the analyst wanted. The **total survey error** (TSE) framework organizes every way a
survey can go wrong under these two headings and, orthogonally, separates systematic
error (**bias**) from sampling-driven dispersion (**variance**) [@Groves_2010].
Formally, for an estimator $\hat\theta$ of a population target $\theta$, the relevant
loss is mean squared error,
$$
\mathrm{MSE}(\hat\theta) = \underbrace{\bigl(\mathbb{E}[\hat\theta] - \theta\bigr)^2}_{\text{bias}^2}
                          + \underbrace{\operatorname{Var}(\hat\theta)}_{\text{variance}},
$$ {#eq-tse-mse}
and TSE decomposes both terms by *source*. The first decomposition is by stage of the
survey process:
$$
\text{Total Survey Error} = \underbrace{\text{Measurement Error}}_{\text{the answers}}
                          + \underbrace{\text{Representation Error}}_{\text{the people}}.
$$ {#eq-tse-decomp}

The **measurement** branch tracks a single response from the target *construct* the
analyst cares about, to the *measurement* the question actually taps, to the
*response* the person gives, to the *edited datum* stored after processing—accruing
validity error, response error, and processing error along the way. The
**representation** branch tracks the *units* from the target *population*, to the
*sampling frame* (the list from which units are drawn), to the *sample*, to the
*respondents* who actually answer—accruing coverage error, sampling error, and
nonresponse error. @fig-tse renders both branches.

```{mermaid}
%%| label: fig-tse
%%| fig-cap: "The total survey error framework: a measurement (left) and a representation (right) branch, each accumulating distinct error sources from construct/population down to the final respondent-level datum. Adapted in original form from the decomposition in @Groves_2010."
flowchart TB
    subgraph M["Measurement (the answers)"]
        C1["Construct"] -->|validity| C2["Measurement"]
        C2 -->|response error| C3["Response"]
        C3 -->|processing error| C4["Edited response"]
    end
    subgraph R["Representation (the people)"]
        P1["Target population"] -->|coverage error| P2["Sampling frame"]
        P2 -->|sampling error| P3["Sample"]
        P3 -->|nonresponse error| P4["Respondents"]
    end
    C4 --> S["Survey statistic"]
    P4 --> S
```

The framework is more than a taxonomy: it tells the analyst *where to spend the next
dollar*. A statistic can be unbiased yet imprecise (a clean design with a small
sample), or precise yet badly biased (a large but unrepresentative convenience
panel). Because the two failure modes call for opposite remedies—more units versus a
better frame—naming the dominant source is the prerequisite to fixing it.

### Three Eras of Survey Practice

The TSE trade-offs have not been static; the technology of contacting respondents has
reshaped which errors bind. @tbl-eras sketches the three eras now standard in survey
methodology. The trajectory is one of falling cost and falling response rates: each
era widened reach and cut expense while making the representation branch harder to
control, culminating in non-probability online panels that must lean on modeling
rather than design to justify inference.

| Era | Sampling | Interview mode | Data environment |
|-----|----------|----------------|------------------|
| First | Area probability | Face-to-face | Stand-alone |
| Second | Random-digit-dial probability | Telephone | Stand-alone |
| Third | Non-probability | Computer-administered | Linked |

: Three eras of survey research, after the design taxonomy popularized in survey methodology. {#tbl-eras}

## Probability and Non-Probability Sampling

The representation branch lives or dies on the **sampling design**. A sample is a
**probability sample** if every unit in the frame has a *known, non-zero* probability
of inclusion. This single property is what licenses design-based inference: because
the inclusion probabilities are known, the analyst can *reweight* the sample to
recover an unbiased estimate of the population total even when the design
oversamples some units.

### The Horvitz–Thompson Estimator

Let $\pi_i = \Pr(i \in s)$ be unit $i$'s inclusion probability and let $y_i$ be the
quantity of interest. The **Horvitz–Thompson** estimator of the population mean
weights each sampled observation by the inverse of its inclusion probability,
$$
\hat{\bar y}_{\text{HT}} = \frac{1}{N}\sum_{i \in s} \frac{y_i}{\pi_i},
$$ {#eq-ht}
where $N$ is the population size and $s$ the realized sample. Inverse-probability
weighting is the formal content of the slogan "weighting recovers
representativeness": a unit sampled at half the average rate stands in for twice as
many population members and so receives twice the weight. The estimator is
*design-unbiased*—$\mathbb{E}[\hat{\bar y}_{\text{HT}}] = \bar y$ under the
randomization that generated $s$—for *any* design with strictly positive $\pi_i$,
which is precisely why the positivity requirement is non-negotiable.[^positivity]

[^positivity]: If $\pi_i = 0$ for some unit, that unit can never enter the sample and
no weight can resurrect it; the corresponding stratum is structurally invisible and
no amount of reweighting recovers it. This is the formal statement of coverage error
in @eq-tse-decomp.

A short simulation makes the bias correction concrete. We construct a population in
which the outcome is correlated with the inclusion probability—the worst case for a
naive sample mean—and verify that the Horvitz–Thompson estimator recovers the truth
while the unweighted mean does not.

```{r ht-estimator, message=FALSE, warning=FALSE}
set.seed(31)

N <- 100000
# A covariate that drives BOTH the outcome and the sampling probability,
# so the naive sample mean is biased and weighting is required.
x      <- runif(N)
y      <- 2 + 3 * x + rnorm(N, sd = 0.5)        # population outcome
pop_mean <- mean(y)

# Unequal-probability design: high-x units are sampled more often.
pi_i   <- 0.02 + 0.08 * x                         # inclusion probability in (0.02, 0.10)
in_sample <- runif(N) < pi_i

y_s    <- y[in_sample]
pi_s   <- pi_i[in_sample]

naive_mean <- mean(y_s)                           # ignores the design -> biased
ht_mean    <- sum(y_s / pi_s) / N                 # Horvitz-Thompson -> unbiased

cat("Population mean:        ", round(pop_mean, 4), "\n")
cat("Naive sample mean:      ", round(naive_mean, 4), "  (biased upward)\n")
cat("Horvitz-Thompson mean:  ", round(ht_mean, 4),  "  (design-unbiased)\n")
```

### The Non-Response Problem

The third-era pivot to non-probability and online sampling breaks the clean logic
above, because the operative inclusion probability is no longer the *design*
probability the analyst set but the *response* probability the respondent chose. Let
$R_i \in \{0,1\}$ indicate that unit $i$ both is selected and chooses to respond, with
response propensity $\rho_i = \Pr(R_i = 1)$. The respondent-based estimator is
unbiased only if the analyst knows $\rho_i$, and in practice $\rho_i$ must be
*estimated*—typically by modeling response as a function of observed covariates. This
is where non-response error enters: if response depends on the outcome $y_i$ even
after conditioning on observed covariates (non-response that is *not* missing at
random), no reweighting on observables removes the bias, and the survey is identified
only under an untestable assumption about the missingness mechanism. The TSE framework
thus locates the modern survey's hardest problem precisely: representation error
migrates from a *design* quantity the analyst controls to a *behavioral* quantity the
analyst must model [@Groves_2010].

## Online Panels and Data Quality {#sec-online-panels}

The practical home of modern marketing surveys is the online panel. Understanding
platform quality is therefore a prerequisite, not an afterthought: the choice of
panel determines which errors from @eq-tse-decomp dominate and can render even
technically sound analyses uninterpretable.

### The Platform Hierarchy

A now-extensive empirical literature benchmarks the major platforms. The headline
finding is unambiguous: **Prolific and CloudResearch substantially outperform Amazon
Mechanical Turk (MTurk) on virtually every quality metric** [@douglas2023;
@peer2022]. In the largest direct comparison, @douglas2023 recruited $n = 2{,}729$
respondents across five platforms and embedded objective quality tests—video-recall
accuracy, attention checks, unique device fingerprints, completion time, and
meaningful open-ended responses. The results are striking. On a composite criterion
requiring all quality tests to pass, **Prolific achieved 67.9% high-quality
participants and CloudResearch 61.9%, versus only 26.4% for MTurk**. Video recall
accuracy told the same story: 83.5% on Prolific and 81.8% on CloudResearch, compared
to 52.2% on MTurk. @peer2022 corroborate this using a different design, finding that
CloudResearch's Approved participant pool produced implausible open-ended responses at
3.0%, versus 18.8% for a standard MTurk sample.

The source of MTurk's decline is largely documented by @kennedy2021, who analyzed 38
studies ($n = 24{,}930$ total, 2013–2018) and traced the quality crisis to a sharp
rise in fraudulent respondents—non-US workers using virtual private servers (VPS) or
VPN connections to circumvent US-only restrictions—from roughly 15% in 2015 to over
20% by late 2018, with individual studies hitting 26.9%. The quality damage is
concentrated: legitimate respondents fail quality checks at 2.8%, whereas VPS users
fail at 23.9% and verified non-US respondents at 32.0%. Critically, the traditional
MTurk quality filter—requiring a 95% HIT acceptance rate and 100+ prior HITs
completed—no longer provides meaningful protection, because fraudulent workers are
experienced enough to satisfy these criteria.

The practical implication for researchers is direct: **MTurk should not be the
default for behavioral and attitudinal studies** unless the researcher can verify
participant authenticity through additional screening. Prolific, which pays at or
above minimum wage and requires verified participant identities, or CloudResearch's
Approved pool, which screens out blocked participants, are the current recommended
alternatives. For studies requiring nationally representative samples, Lucid and
similar router-based panels offer quota matching but introduce their own quality
heterogeneity and require careful vendor-specific screening.

### Attention Checks and Quality Filters

Even on high-quality platforms, a fraction of respondents engages carelessly.
**Attention checks**—items designed to detect inattentive responding—are the most
common data-quality tool. Three types are in wide use:

1. **Instructed response items.** The question instructs the respondent to select a
   specific answer (e.g., "For quality control, please select 'Strongly agree'
   below"). Failure rates on legitimate platforms are typically low (2–10%) but
   higher on MTurk.
2. **Reading comprehension checks.** A paragraph of text is followed by questions
   whose correct answers appear verbatim in the passage. Inattentive respondents
   cannot answer correctly without reading.
3. **Screener traps / infrequency items.** Items with no plausible true-positive
   response (e.g., "I have visited the moon") flag extreme agreement as careless
   or dishonest.

A key design question is whether to warn respondents of attention checks in advance.
Forewarning increases *apparent* pass rates but may induce strategic compliance
rather than genuine engagement; researchers who forewarned found effects shrank
relative to un-forewarned conditions in some paradigms. Current practice favors
embedding checks unobtrusively without forewarning, placing them early and late in
the survey to capture fatigue, and using multiple check types to reduce the chance
that any single type creates a systematic artifact.

Two additional quality filters have become standard:

- **Completion time.** Respondents who complete a survey far below the median time
  (e.g., less than one-third of median completion) are unlikely to have read all
  items. A conservative cutoff is removing the fastest 5–10% of completions after
  verifying the distribution is not bimodal (which would indicate a speeder mode
  distinct from the main distribution).
- **Duplicate detection.** IP-address deduplication catches the same person
  completing multiple times; browser fingerprinting and geolocation checks catch
  coordinate fraud (many responses from the same physical location).

### Survey Satisficing and Careless Responding

Beneath outright fraud lies a subtler quality threat: **satisficing**—the tendency
to provide "good enough" rather than accurate answers [@krosnick1991]. Satisficing
arises when the cognitive cost of careful responding exceeds the respondent's
motivation. It manifests in several behavioral signatures:

- **Straight-lining.** Giving the same response to every item in a matrix question,
  ignoring item content. Detectable by computing the within-respondent standard
  deviation across a battery: a zero or near-zero value is diagnostic.
- **Non-differentiation.** Using only a narrow range of scale points (e.g.,
  responding 4 or 5 exclusively on a 1–7 scale) in a way that does not reflect
  genuine construct-level agreement.
- **Acquiescence bias.** The tendency to agree with statements regardless of their
  content—sometimes called *yea-saying*. Acquiescence is partly a trait (stable
  across items) and partly a function of item wording and survey length. It can be
  diagnosed by including matched pairs of items worded in opposite directions: if the
  respondent agrees with both the item and its negation, they are acquiescing.
- **Midpoint endorsement / don't-know overuse.** Disproportionate use of the scale
  midpoint or neutral category, often as a low-effort response rather than a genuine
  expression of ambivalence.

Recommended detection pipeline: flag respondents with (a) a within-row SD of zero
on any matrix battery longer than four items; (b) total survey completion below a
reasonable minimum; and (c) an implausible open-ended response (e.g., a single
character, gibberish, or a copied prompt). Each flag alone may be innocuous; two or
more together strongly suggest careless responding. The analyst should report the
number of flagged observations and sensitivity-check key results with and without
them.

```{r satisficing-flags, message=FALSE, warning=FALSE}
library(tidyverse)
set.seed(42)

n_resp <- 500
n_items <- 8

# Simulate a matrix battery with some careless respondents
# Careful respondents: random responses around their true position
# Straight-liners: same value repeated (within-row SD = 0)
careful  <- matrix(sample(1:7, (n_resp - 50) * n_items, replace = TRUE),
                   nrow = n_resp - 50)
careless <- matrix(rep(sample(1:7, 50, replace = TRUE), each = n_items),
                   nrow = 50, byrow = TRUE)
battery  <- rbind(careful, careless)
colnames(battery) <- paste0("q", seq_len(n_items))

df <- as_tibble(battery) %>%
  mutate(respondent = row_number(),
         row_sd     = apply(battery, 1, sd),
         straightline = row_sd == 0)

cat("Straight-line flags detected:", sum(df$straightline),
    "of", n_resp, "respondents\n")
cat("Expected (planted):", 50, "\n")
```

## Response Styles and Systematic Measurement Biases

Beyond careless responding, **response styles** are systematic tendencies that
distort scores even when respondents are fully engaged. They contribute to measurement
error (@eq-tse-decomp) because they introduce variance that is *not* attributable to
the construct of interest.

### Social Desirability

**Social desirability bias** is the tendency to give answers that present oneself
favorably to others, regardless of one's true attitudes or behaviors. It is
especially consequential for sensitive constructs—health behavior, financial
decisions, ethical attitudes, political views—where the socially approved answer
diverges from the true answer. Two components are distinguished in the psychometric
literature: *impression management* (conscious self-presentation) and *self-deception
enhancement* (a genuine but inflated self-image). The **Marlowe–Crowne Social
Desirability Scale** [@crowne1960] and its abbreviated versions are the most widely
used instruments for measuring and controlling social desirability as a covariate.

Design-based defenses are generally more effective than post-hoc statistical
partialling. The **bogus pipeline** technique convinced respondents that a
physiological device could detect dishonest answers, which substantially reduced
social desirability in early studies but is now rarely used due to ethical concerns
and the ease of disbelief. More practical alternatives include:

- **Anonymity assurances** communicated prominently before sensitive questions.
- **Third-person phrasing** ("Some people find that they... Do you?"), which
  reduces the normative force of the question.
- **Randomized response technique (RRT)**, where a random device (e.g., a coin flip
  unobserved by the researcher) determines whether the respondent answers the
  sensitive question or an innocuous one. The researcher can recover population
  prevalence estimates without knowing any individual's true answer.
- **Indirect questioning** through projective measures or implicit attitude tests,
  though the psychometric properties of many implicit measures remain debated.

### Acquiescence Bias

As noted above, acquiescence systematically inflates agreement with positively worded
items and deflates agreement with negatively worded items, creating a spurious
construct mean and attenuating structural relationships. The standard remedy is
**balanced item wording**: including an equal number of items worded in each
direction and reverse-scoring the negatively worded items before averaging. Balanced
scales do not eliminate acquiescence entirely—a persistent acquiescer will agree with
both poles—but they ensure that the scale *mean* is acquiescence-neutral. A clean
diagnostic is the **polychoric correlation** between matched positive–negative item
pairs: strongly negative correlations confirm that the items are tapping the same
underlying construct; near-zero correlations suggest the items are measuring
unrelated things or that acquiescence is negligible.

### Extreme Response Style

**Extreme response style (ERS)** is the tendency to endorse the scale endpoints (1
or 7 on a 7-point scale) regardless of item content. It is most pronounced in
cross-cultural research, where ERS varies substantially across countries and can
create spurious cross-national differences. Mitigation strategies include:

- Using **forced-choice** or **best–worst scaling** formats that eliminate the
  option to cluster at extreme points.
- **Response style corrections** that model ERS as a latent nuisance factor alongside
  the substantive construct, analogous to the common-method factor in @sec-measurement-scales.
- **Ranking formats** that force differentiation across items.

## Anchoring Vignettes

The measurement branch has a subtler pathology than noisy answers: respondents may
use the *same* response scale to mean *different* things. Ask consumers to rate their
satisfaction "on a 1–5 scale" and a lenient respondent's 4 may encode the same
underlying experience as a demanding respondent's 2. This **interpersonal
incomparability**—formally, **differential item functioning** (DIF)—means that
differences across respondents partly reflect differences in how they map a latent
state onto categories, not differences in the latent state itself.

**Anchoring vignettes** solve this by adding, alongside the self-assessment, a small
set of fixed hypothetical scenarios that every respondent rates on the same scale
[@king2004; @king2007; @hopkins2010]. The key move is that the analyst *authors* the
vignettes, so their true level is **fixed across respondents by construction**. Any
variation in how respondents rate the *same* vignette is therefore pure DIF—a direct
reading of each respondent's idiosyncratic scale use—which can then be netted out of
the self-assessment.

Two assumptions identify the correction [@king2004]:

- **Response consistency.** Each respondent applies the *same* mapping from latent
  state to response category when rating the vignettes as when rating themselves.
- **Vignette equivalence.** The latent level a given vignette describes is perceived
  identically by all respondents; only its *translation* into a category varies.

### Nonparametric Correction

The nonparametric estimator requires no distributional assumptions. For each
respondent, the analyst recodes the self-assessment $y_i$ relative to where it falls
among that respondent's *ordered* vignette ratings $z_{i1} \le z_{i2} \le \dots \le
z_{iJ}$. The rescaled outcome $C_i$ records the self-rating's rank position within
the respondent's own vignette thresholds—how good the respondent is relative to
anchors whose true level is known—so it is comparable across respondents by
construction. Order **inconsistencies**, where a respondent's vignette ratings are
non-monotone in the vignettes' known true order, are absorbed as **ties**.

### Parametric Correction

The parametric estimator embeds the same idea in a likelihood. Self-assessment and
vignettes are jointly modeled with an **ordered probit** in which the *cut points*
vary across respondents as functions of covariates, with a respondent **random
effect** capturing residual scale heterogeneity. Identification improves as the
number of vignettes $J$ grows, but there is a bias–variance trade-off: each
additional vignette is measured with error, so adding vignettes trades sharper
threshold identification against more measurement noise [@hopkins2010].

```{r anchors-demo, message=FALSE, warning=FALSE}
library(tidyverse)

# Three respondents with identical true ability differ only in rating strictness.
demo <- tribble(
  ~respondent, ~self, ~v1, ~v2, ~v3,
  "lenient",      5L,   3L,  4L,  5L,
  "moderate",     4L,   2L,  3L,  4L,
  "harsh",        3L,   1L,  2L,  3L
)

anchor_pos <- function(self, v) {
  v <- sort(v)
  paste0(sum(self > v), " above, ", sum(self == v), " tied")
}

demo %>%
  rowwise() %>%
  mutate(anchored = anchor_pos(self, c(v1, v2, v3))) %>%
  ungroup() %>%
  select(respondent, raw_self = self, v1, v2, v3, anchored)
```

The raw self-ratings (5, 4, 3) look different, but every respondent rates themselves
"2 above, 1 tied" relative to their own vignettes—anchoring reveals the apparent gap
was pure rating strictness.

## Experiments

Where a survey *observes* a latent state, an experiment *manufactures* variation in a
cause and reads off its effect. The experiment's payoff is causal identification: by
randomly assigning the treatment, the analyst makes the treatment statistically
independent of every confounder, observed or not, so the difference in mean outcomes
estimates the average treatment effect rather than a confounded association. The cost
is that the analyst must defend four distinct kinds of **validity**, manage the
heterogeneity that averages conceal, choose a design appropriate to the research
question, and respect the ethics of intervening on people.

### Four Validities

A credible experiment clears four hurdles, summarized in @tbl-validity. They are
ordered roughly from "is there an effect at all" to "does it mean what we claim" to
"does it travel."

| Validity | Question it answers | Characteristic threat |
|----------|--------------------|-----------------------|
| Statistical conclusion | Is the treatment–outcome covariation real, not noise? | Low power; fishing; violated test assumptions |
| Internal | Is the covariation *causal* within the study? | Confounding; attrition; failed randomization |
| Construct | Does the operationalization capture the intended concept? | Mono-operation bias; demand effects |
| External | Does the effect *generalize* beyond this sample and setting? | Non-representative subjects; artificial context |

: The four validities an experiment must defend, in roughly increasing order of generality. {#tbl-validity}

The four trade off against one another. Tight laboratory control buys internal
validity at the expense of external validity (realism), while a field experiment
buys realism at the expense of control. The same tension recurs in the practical
design levers—**cost, control, realism, and ethics**—that no single design
optimizes simultaneously.

### Lab, Field, and Survey Experiments

Three broad settings host experimental research in marketing, each occupying a
distinct position in the internal-vs.-external-validity trade-off.

**Laboratory experiments** (including online survey experiments) offer maximum
control. The researcher assigns stimuli, controls timing, eliminates distractors, and
can probe mechanism directly with manipulation checks and mediator measures. The
price is reduced realism: stimuli are often hypothetical, settings unfamiliar, and
participant pools (student samples, crowdsourcing panels) may not represent the
population of interest. Most published causal claims in marketing rest on lab or
online experiments, which makes their external validity an ongoing empirical question
rather than an assumption.

**Field experiments** embed the treatment in an actual market. A/B tests on email
campaigns, retail price promotions deployed to random zip codes, and randomized
product launches are field experiments. Internal validity is preserved by
randomization; external validity is substantially stronger because the setting, the
stakes, and the population are real. The drawbacks are cost, limited measurement
(typically only behavioral outcomes, not psychological processes), ethical exposure
(firm and consumer stakes are real), and the difficulty of isolating mechanisms. The
complementarity of lab and field is explicit: lab experiments establish that an
effect *can* exist and identify the mechanism; field experiments establish that it
*does* exist in the target environment [@gerber2012].

**Survey experiments** embed random assignment within an otherwise-standard survey
questionnaire, assigning respondents to see different question wordings, orderings,
or vignette descriptions. They are cheap and fast but inherit the survey's
weaknesses: the outcome is self-reported, respondents may recognize the experimental
structure (demand effects), and stakes are hypothetical. Factorial survey experiments
(*vignette studies*) present each respondent with a profile constructed by randomly
varying several attributes simultaneously, allowing estimation of all main effects
and interactions within a single study.

### Demand Effects and Deception

**Experimenter demand effects** arise when participants infer the researcher's
hypothesis and adjust their responses accordingly—either cooperating (out of
helpfulness) or doing the opposite (reactance). Demand effects constitute a construct
validity threat: the observed treatment effect partly reflects participants responding
to *perceived* social expectations rather than to the construct the treatment was
designed to manipulate.

Diagnosing demand effects requires measuring participants' awareness of the
hypothesis. The **Awareness Questionnaire** approach (often called the *funnel
debriefing*) asks participants, at the end of the study and in progressively more
explicit terms, (a) what they thought the study was about, (b) what they thought the
researcher's hypothesis was, and (c) whether that suspicion influenced their
responses. Effect-size estimates from participants who report hypothesis awareness can
then be compared to those who do not; a large discrepancy is evidence of demand.

Three design strategies reduce demand effects:

1. **Cover stories.** Present the study under a plausible but innocuous description
   that conceals the hypothesis. The cover story must be convincing enough not to
   trigger suspicion, and debriefing is required afterward if the cover constitutes
   deception.
2. **Between-subjects designs.** Participants who see only one level of the
   independent variable cannot infer what the manipulation is comparing, making
   hypothesis detection harder. This advantage comes at a power cost (see
   @sec-power).
3. **Behavioral outcomes.** Demand effects are primarily a self-report pathology:
   participants cannot as easily adjust their actual behavior (how many M&Ms they
   take from a bowl, how much they bid at auction) as they can their stated opinions.
   When a behavioral outcome is available, it is preferable.

**Deception** is the deliberate use of false information—a cover story, a
confederate, a fictitious scenario—to suppress demand or achieve experimental realism
that cannot otherwise be manufactured. Deception is ethically permissible under
standard IRB/ethics-review frameworks when (a) the research question cannot be
answered by non-deceptive means, (b) the deception causes no lasting harm, and (c)
participants are **debriefed** promptly and fully after data collection. The
debriefing must not merely reveal the deception passively: it should explain why the
deception was necessary, allow participants to withdraw their data with no penalty,
and verify that they understand and are not disturbed. Some ethics frameworks
(especially in Europe under GDPR-adjacent guidance) prohibit certain forms of
deception outright regardless of debriefing.

### Within-Person versus Between-Person Designs

The most fundamental structural decision in an experiment is whether each participant
sees *one* condition (**between-subjects** or **independent-groups** design) or
*multiple* conditions (**within-subjects** or **repeated-measures** design).

**Between-subjects** designs assign each participant to exactly one treatment cell.
They are clean, immune to carryover, and eliminate demand effects that arise from
participants contrasting conditions. Their weakness is statistical: to detect the
same effect size, they require substantially more participants, because each person's
response is used exactly once. Variance between people (individual differences) is
not isolated and becomes part of the error term.

**Within-subjects** designs have each participant experience multiple conditions (in
counterbalanced or randomized order). By measuring the *same person* under each
condition, the analyst can difference out individual-level constants, isolating the
within-person treatment effect from stable individual differences. The statistical
gain is large: within-person designs can detect the same effect with far fewer
participants because the person serves as their own control. The costs are:

- **Carryover effects.** Exposure to condition A may affect responses to condition B
  (e.g., fatigue, sensitization, contrast effects). Counterbalancing across
  participants controls for first-order carryover but not interactions.
- **Demand effects.** Participants who experience multiple conditions can deduce the
  comparison, increasing hypothesis awareness.
- **Practice and fatigue.** Performance may change systematically across trials
  regardless of the treatment.

The trade-off dictates: use within-subjects designs when the effect is expected to be
small (power is paramount) and carryover is unlikely; use between-subjects when
carryover or demand is a plausible confound.

**Experience sampling methods (ESM)**, also called ecological momentary assessment
(EMA), extend within-person measurement to the natural environment. Participants
complete brief surveys multiple times per day, often triggered by a smartphone
prompt, capturing states (mood, hunger, context) and behaviors as they occur. ESM is
ideal for constructs that are highly state-dependent or that would be biased by
retrospective recall (e.g., in-the-moment affect, impulse purchase triggers, media
consumption). The design produces a multilevel data structure—many observations
nested within each participant—requiring mixed-effects models that partition variance
into within-person and between-person components and are analyzed in @sec-modeling.

### Conjoint and Discrete-Choice Experiments {#sec-conjoint}

When the research question concerns how multiple product attributes jointly determine
consumer preference, the workhorse tool is the **conjoint experiment**. Respondents
are shown a set of product profiles described by combinations of attributes (brand,
price, warranty, green certification, etc.) and either rate each profile or choose
among them. Because the attribute combinations are researcher-designed and
orthogonally or near-orthogonally varied, the analyst can recover each attribute's
**part-worth utility**—its marginal contribution to overall preference—as if running
a separate experiment on each attribute while holding others constant.

Two formulations dominate:

1. **Rating-based conjoint (traditional full-profile conjoint).** Each profile is
   rated on a preference scale. Part-worths are recovered by regressing ratings on
   attribute dummies. Simple to implement but subject to rating-scale
   inconsistencies and demand effects if the attribute structure is transparent.

2. **Choice-based conjoint (CBC).** Respondents choose their preferred option from a
   set of competing profiles, including an "opt-out" option. Choice is more realistic
   than rating—it mirrors actual purchase decisions—and the opt-out preserves
   revealed-preference logic. Part-worths are estimated by conditional logit or
   mixed logit, which allows heterogeneous preferences across respondents. The market
   simulation step maps part-worths to market-share predictions under alternative
   product configurations, making CBC the standard tool for new-product development
   and pricing research.

A critical design decision in any conjoint study is **orthogonality**: the attribute
levels shown to any respondent should be uncorrelated across profiles, so that
part-worths are identified without confounding. Fractional factorial designs achieve
near-orthogonality with far fewer profiles than full-factorial enumeration would
require, trading off some high-order interaction estimation for feasibility.

**Best–worst scaling (BWS)**, or MaxDiff, is a simpler cousin: respondents choose
the *best* and *worst* item from a set, producing a pair of anchored choices that is
more discriminating than ratings and avoids the extreme response style biases of
Likert-type scales. BWS produces individual-level utility scores and is increasingly
preferred in applied research because it forces genuine differentiation.

### Heterogeneity and Mechanisms

A single number—the **average treatment effect** (ATE)—can be an artifact. If the
effect is positive for one segment and negative for another, the average may be near
zero, near either extreme, or anywhere between, depending only on segment mix.
Writing the conditional average treatment effect as $\tau(x) = \mathbb{E}[Y(1) -
Y(0)\mid X=x]$, the reported ATE is the mix-weighted average $\int \tau(x)\,dF(x)$,
which is silent about the sign and size of $\tau(x)$ for any particular $x$. The
same caution that motivated segmentation in forecasting (@sec-surveys) applies to
experiments: heterogeneous effects demand that the analyst report the distribution of
$\tau(x)$, not only its mean, and that any claimed **mechanism**—the causal pathway
through which the treatment acts—be argued rather than assumed.

Mediation analyses are the standard vehicle for mechanistic claims, but they carry
assumptions of their own. A significant indirect effect ($a \times b$ in the
Baron–Kenny path notation) identifies the mechanism only if (a) there is no
unmeasured common cause of the mediator and the outcome, and (b) the mediator is not
on a back-door path from treatment to outcome. These assumptions are untestable from
observational data on mediators and outcomes and motivate the use of **sequential
experimental designs** that manipulate both the treatment and the putative mediator
in separate studies to establish causal-chain evidence.

### Power Analysis {#sec-power}

Every experiment must answer the question: *how many participants are needed to detect
the effect of interest if it exists?* Underpowered studies waste resources and
expose participants to risk without a realistic chance of a conclusive result—a
direct violation of the Reduce principle from the 3Rs framework (see below). They
also produce inflated effect-size estimates when they do yield significant results
(the "winner's curse"), because only the largest noise-amplified estimates clear the
significance threshold.

**Conventional power analysis** requires three inputs and produces a sample size:

- $\delta$: the effect size, usually Cohen's $d$ for means or $f^2$ for regression,
- $\alpha$: the Type I error rate (usually 0.05), and
- $1-\beta$: the desired power (usually 0.80 or 0.90).

The mechanical calculation is available in R (`pwr` package) but is only as good as
its inputs. The critical debate is about $\delta$: what effect size should power
calculations assume?

The **smallest effect size of interest (SESOI)** approach [@lakens2018] replaces
"what effect size will I find?" with "what is the smallest effect that would be
theoretically or practically meaningful?" The SESOI is set on substantive grounds
before data collection—for example, a standardized mean difference smaller than $d =
0.2$ may be too small to justify the cost of a marketing intervention—and the study
is powered to detect *that* threshold. This framing avoids the common error of
powering studies on prior literature effect sizes that are themselves upward-biased
by publication selection.

**Simulation-based power analysis** is more flexible than analytic formulas when the
design is complex (multilevel, repeated-measures, multiple endpoints, or non-normal
outcomes). The simulation encodes the assumed data-generating process, applies the
planned analysis, and counts the fraction of simulated studies that return $p < \alpha$.

```{r power-simulation, message=FALSE, warning=FALSE}
set.seed(2024)

# Simulation-based power for a between-subjects t-test
# SESOI: d = 0.30 (small but practically meaningful)

sim_power <- function(n_per_group, d, alpha = 0.05, nsim = 2000) {
  sig <- replicate(nsim, {
    group0 <- rnorm(n_per_group, mean = 0,   sd = 1)
    group1 <- rnorm(n_per_group, mean = d,   sd = 1)
    t.test(group1, group0, var.equal = TRUE)$p.value < alpha
  })
  mean(sig)
}

# Sweep over sample sizes
ns    <- c(50, 100, 150, 200, 300)
pows  <- sapply(ns, sim_power, d = 0.30)

results <- data.frame(n_per_group = ns, power = round(pows, 3))
print(results)
cat("\nTarget: 80% power for d = 0.30 requires ~", ns[which(pows >= 0.80)[1]],
    "per group\n")
```

The simulation makes the researcher's assumptions explicit and auditable. Standard
practice is to pre-register the power analysis alongside the hypotheses, so the
sample-size decision is on record before any data are seen. Reporting power analysis
in published work has increased substantially in recent years—from roughly 9.5% of
empirical psychology articles in 2015–2016 to 30% by 2020–2021 across 24 APA-listed
journals [@adamkovicova2024].

### Sequential Testing {#sec-sequential}

A limitation of the conventional null-hypothesis testing framework is that the $\alpha$
level is calibrated for a *fixed* sample: collecting data until significance is
reached, then stopping, inflates the true Type I error rate far above $\alpha$.
**Sequential testing** addresses this by specifying in advance when interim analyses
will be conducted and how much $\alpha$ each interim analysis "spends."

Two approaches are in common use:

1. **Group sequential designs** with formal **$\alpha$-spending functions** (e.g.,
   O'Brien–Fleming, Pocock bounds). The analyst specifies the number of interim
   looks and a monotone spending function that allocates cumulative $\alpha$ across
   looks; the critical value at each look is adjusted upward so that the overall Type
   I error remains at $\alpha$. These designs are standard in clinical trials and
   increasingly used in A/B testing platforms.

2. **Sequential probability ratio tests (SPRT)** and **anytime-valid inference**.
   The SPRT computes a likelihood ratio at every new observation and crosses a
   pre-specified boundary when evidence is sufficiently strong for either hypothesis.
   Recent extensions provide **e-values** and **anytime-valid $p$-values** that
   remain valid under optional stopping without adjustment, enabling "keep collecting
   data until we decide" designs with formal error control [@ramdas2023].

For marketing A/B testing with continuous data collection, the practical alternative
is the **always-valid inference** framework, implemented in commercial platforms
(e.g., Optimizely's Stats Engine) and available in R. The analyst sets a minimum
detectable effect and a maximum sample size, and the platform signals when the
boundary is crossed in either direction or when the maximum is reached without
crossing.

### Ethics: The 3Rs

Experiments intervene on living subjects, and the governing ethical framework is the
**3Rs** of Russell and Burch (1959):[^threers] *Replace* an invasive procedure with a
less invasive (or non-animal, or natural) alternative; *Refine* procedures to reduce
harm and distress; and *Reduce* the number of participants to the minimum consistent
with a valid answer. The Reduce principle has a statistical edge: the minimum sample
is set by the power calculation, so an underpowered study is not only wasteful but
also *unethical*, exposing participants to risk without a credible chance of a
conclusive result. The 3Rs underlie modern institutional review board (IRB) and
research ethics committee requirements; any study involving human participants must
secure approval before data collection and adhere to the approved protocol.

[^threers]: Russell, W. M. S. and Burch, R. L. (1959), *The Principles of Humane
Experimental Technique*. The 3Rs framework predates and underlies most modern
research-ethics review.

## Open Science {#sec-open-science}

The reproducibility crisis of the 2010s—in which a substantial fraction of landmark
behavioral science findings failed to replicate in independent studies—forced a
reckoning with the research practices that inflated false-positive rates. The result
was an open-science infrastructure that is now the expected standard for credible
empirical work.

### Pre-Registration

**Pre-registration** time-stamps the hypotheses, design, and analysis plan *before*
data collection begins, in a public registry (AsPredicted, OSF Registries). By
separating the confirmatory analysis from the exploratory analysis, pre-registration
prevents the two most damaging forms of analytical flexibility:

- **Hypothesis-after-results (HARKing)**: presenting an exploratory finding as if it
  were a pre-specified hypothesis, inflating the apparent prior probability that any
  given test was of theoretical interest.
- **Outcome switching**: testing multiple outcomes and reporting only those that
  reached significance.

Pre-registration does not make studies immune to problems: it cannot ensure that
the theory is correct, the manipulation is valid, or the measure is reliable.
As @simmons2021 argue—with an analogy to random assignment—pre-registration is
"a game changer" precisely because it is narrow: like random assignment, which
controls confounds but cannot guarantee external validity, pre-registration controls
analytical flexibility but cannot substitute for good measurement and adequate power.
Non-pre-registered studies conducted transparently can still be credible; pre-registered
studies with flawed designs are not credible because of the time-stamp.

The pre-registration should specify at minimum: the primary hypothesis and its
directionality, the key measures and their operationalizations, the sample size and
how it was determined, the statistical test and any covariates included, and the
exclusion criteria applied before analysis. Deviations from the plan in the
published paper should be disclosed and justified.

### The $p$-Curve

The **$p$-curve** is the distribution of statistically significant $p$-values across
a set of studies testing the same effect [@simonsohn2014]. Under a true effect, that
distribution is *right-skewed*—many very small $p$-values, because a true effect
makes extreme test statistics likely. Under no effect but selective reporting
($p$-hacking), it is *flat or left-skewed*—a pile-up just below 0.05, because
researchers publish only the highest values that clear the threshold. Inspecting the
curve thus diagnoses whether a literature reflects real effects or selective analysis.

> A right-skewed $p$-curve is the signature of a true effect; a flat or left-skewed
> curve is the signature of $p$-hacking. The latter does not mean no individual
> study is valid—it means the *literature as collected* cannot be trusted at face
> value.

The $p$-curve also supports **effect-size estimation**: the underlying effect can be
estimated from the distribution of significant $p$-values alone, without access to
unpublished studies (the file drawer). The estimate is biased upward because it
conditions on significance, but the bias is lower than the naive published-studies
mean.

### Specification Curve Analysis

Pre-registration disciplines a single analysis. **Specification curve analysis (SCA)**
asks what happens across *all* defensible analyses [@simonsohn2020]. The researcher
enumerates every combination of analysis choices that could reasonably be defended—
different covariate sets, outcome transformations, sample exclusions, model
families—and estimates the effect for each specification. The output is a graphical
"specification curve" that plots every estimate in decreasing order, annotated with
the choices that produced it.

SCA has two inferential uses. Descriptively, it reveals whether the focal result is
robust—if the effect is consistently in the same direction and significant across the
full curve, robustness is strong—or fragile, concentrated in a narrow slice of
specifications. Inferentially, a permutation test under the null shuffles treatment
assignment, recomputes the curve, and assesses whether the observed median estimate
exceeds the permutation distribution; this tests whether the pattern of results is
consistent with a null effect across the universe of reasonable analyses.

SCA is implemented in R (the `specr` package) and Stata and is increasingly expected
for papers where specification choices are numerous or contentious. It is the
multivariate generalization of a sensitivity analysis: instead of varying one choice
at a time, it varies all choices simultaneously and characterizes the full
consequence space.

### Equivalence Testing

Standard null-hypothesis tests answer "is there *any* effect?" Equivalence testing
answers "is the effect smaller than a meaningful threshold?" The **two one-sided
tests (TOST)** procedure [@lakens2017] tests the null hypothesis that the effect is
*at least* as large as the SESOI in absolute value. Both one-sided tests must reject
at $\alpha$ for the analyst to conclude equivalence to zero within the equivalence
bounds. Formally, for effect size $\hat\delta$ and equivalence bounds $[-\Delta,
+\Delta]$:

$$
H_{0,1}: \delta \le -\Delta \qquad H_{0,2}: \delta \ge +\Delta.
$$

Rejecting both $H_{0,1}$ and $H_{0,2}$ establishes that the true effect falls within
$(-\Delta, +\Delta)$—practically equivalent to zero by the researcher's pre-specified
standard. TOST is the appropriate analysis when a researcher wants to claim that a
proposed replication failed or that an intervention has no practically meaningful
effect. It is *not* the same as "failing to reject the null": a non-significant
standard $t$-test is uninformative about whether the effect is small; TOST is
informative precisely because it has power against the specific alternative that the
effect exceeds $\Delta$.

```{r tost-example, message=FALSE, warning=FALSE}
# TOST equivalence test for a between-subjects comparison
# H0: effect >= delta (either side)
# Equivalence bound: d = 0.20 (effect below this is "practically zero")
set.seed(99)

n     <- 200
group0 <- rnorm(n, mean = 0.00, sd = 1)
group1 <- rnorm(n, mean = 0.05, sd = 1)   # true d ≈ 0.05, near zero

delta  <- 0.20   # equivalence bound in raw units (same SD = 1)

# Two one-sided t-tests
t_lower <- t.test(group1 - group0, mu = -delta, alternative = "greater")
t_upper <- t.test(group1 - group0, mu =  delta, alternative = "less")

cat("TOST lower bound p-value:", round(t_lower$p.value, 4), "\n")
cat("TOST upper bound p-value:", round(t_upper$p.value, 4), "\n")
cat("Equivalence established (both p < .05):",
    t_lower$p.value < 0.05 & t_upper$p.value < 0.05, "\n")
```

### Registered Reports

The most radical open-science reform is the **registered report** format, now offered
by over 300 journals. In a registered report, the introduction, literature review,
and methods are peer-reviewed *before* data collection; acceptance is provisional on
the approved methodology. After data collection, the results and discussion are added
and the paper is published regardless of whether the outcome is significant, null, or
mixed—provided the registered methods were followed.

Registered reports eliminate publication bias at the source by decoupling acceptance
from results. They also incentivize well-powered designs with valid measures, because
reviewers evaluate only the pre-data sections. Meta-analytic evidence comparing
registered versus non-registered studies in the same journals finds that registered
reports show substantially smaller effect sizes on average, consistent with the
hypothesis that the larger effects in the conventional literature are upward-biased
by selective publication. For marketing researchers, a growing number of journals in
the field (including the *Journal of Marketing Research* and *Journal of Consumer
Research*) now accept registered reports.

### Open Data, Materials, and Code

Pre-registration governs what was planned; open data, materials, and code govern what
was done. The **FAIR principles** (Findable, Accessible, Interoperable, Reusable)
provide an operational standard for data sharing. Platforms include:

- **Open Science Framework (OSF)**: the dominant repository for behavioral science,
  supporting pre-registration, data hosting, and materials sharing under versioned
  DOIs.
- **ResearchBox**: a packaging system that bundles data, analysis code, and
  materials in a single citable artifact, making exact replication straightforward.
- **AsPredicted**: focused specifically on pre-registration with a minimal interface.

Open materials mean that a reviewer or reader can audit whether the stimuli,
measures, and procedures match the paper's description. Open analysis code means that
every reported number traces to a reproducible computation. Together, these practices
address the "computational reproducibility" dimension of the crisis: the finding
that a substantial fraction of published statistical results cannot be reproduced from
the reported data even without any scientific dispute about the methods.

## Key Takeaways

- Surveys and experiments are both *instruments for recovering an unobserved
  quantity*; surveys fail through **error** (@eq-tse-decomp) and experiments through
  threats to **validity** (@tbl-validity).
- Measurement is **reactive**: asking about purchase intentions strengthens the
  intention–behavior link by roughly 58%, so intentions surveys are mild
  interventions, not passive forecasts [@chandon2005].
- Total survey error separates **measurement** error (the answers) from
  **representation** error (the people), and bias from variance; naming the dominant
  source dictates the fix [@Groves_2010].
- Probability designs license design-based inference through the
  **Horvitz–Thompson** estimator (@eq-ht); non-probability and non-response designs
  shift the burden from *design* to *modeling* the inclusion mechanism.
- **Online panels are not interchangeable**: Prolific and CloudResearch substantially
  outperform MTurk on composite quality metrics (67.9%/61.9% vs. 26.4% high-quality
  participants), and traditional MTurk filters—95% HIT acceptance, 100+ HITs—no
  longer recover historical quality [@douglas2023; @peer2022; @kennedy2021].
- **Satisficing** (straight-lining, acquiescence, speeding) degrades measurement
  quality even on high-quality platforms; a multi-indicator quality pipeline (SD
  flags, time filters, open-ended checks) should be applied before analysis.
- **Response styles**—social desirability, acquiescence bias, extreme response
  tendency—introduce systematic variance unrelated to the construct; balanced item
  wording and design-based remedies (anonymity, RRT, behavioral outcomes) are
  preferred over post-hoc statistical corrections.
- **Anchoring vignettes** purge interpersonal incomparability by netting out
  differential item functioning, identified by response consistency and vignette
  equivalence [@king2004].
- Causal claims rest on four validities; demand effects threaten construct validity
  and are mitigated by cover stories, between-subjects designs, and behavioral
  outcomes.
- **Conjoint experiments** (especially choice-based CBC) recover attribute
  part-worths under realistic competitive choice contexts; best–worst scaling avoids
  extreme response biases in preference measurement.
- **Within-subjects designs** dramatically increase power by removing individual
  differences from the error term; experience sampling extends within-person
  measurement to naturalistic settings via repeated smartphone surveys.
- Power analysis should target the **smallest effect size of interest (SESOI)**,
  not the expected effect; simulation-based power makes distributional assumptions
  explicit [@lakens2018].
- **Sequential testing** with $\alpha$-spending functions or anytime-valid
  inference allows data collection to stop early when evidence is sufficient,
  without inflating Type I error.
- **Pre-registration** eliminates HARKing and outcome switching; **specification
  curve analysis** characterizes robustness across all defensible specifications
  [@simonsohn2020]; **TOST** establishes practical equivalence to zero [@lakens2017];
  and **registered reports** eliminate publication bias at the acceptance stage.

## Further Reading

For total survey error the canonical reference is @Groves_2010; @krosnick1991 is the
foundational treatment of satisficing and cognitive aspects of survey response.
@peer2022 and @douglas2023 are the current benchmarks for online panel quality
comparisons; @kennedy2021 documents the MTurk fraud timeline. For experiment design,
@gerber2012 is the definitive treatment of field experiments; @shadish2002 covers the
four-validity framework in depth. The open-science toolkit is developed in
@simonsohn2014 ($p$-curve), @simonsohn2020 (specification curve), @lakens2017 and
@lakens2018 (equivalence testing and SESOI), and @simmons2021 (pre-registration as
game changer). Conjoint methods are treated comprehensively in @orme2020 and in the
academic review of @green1990. For within-person and ESM designs, @bolger2013
provides the multilevel framework. The measurement scales literature that underlies
survey item development is treated in @sec-measurement-scales.
