# Surveys and Experiments {#sec-surveys}

Most of what marketing researchers want to know—what consumers intend to buy, how
much they value an attribute, whether a campaign caused a sale—cannot be read
directly off transactional data. It must be *elicited* or *manufactured*: elicited
through **surveys** that ask people to report latent states, or manufactured through
**experiments** that intervene on the world and observe the response. This chapter
treats the two as a matched pair. Both are instruments for recovering a quantity
that the analyst cannot observe, and both fail in characteristic, formalizable ways.
The survey's central problem is *error*—the gap between what respondents report and
the population truth—decomposed into how respondents are selected and how they
answer. The experiment's central problem is *validity*—whether the measured effect
is the causal effect, generalizes beyond the lab, and operationalizes the construct
the researcher intended.

The two instruments also interact, and not benignly. The very act of surveying a
consumer can change that consumer's later behavior, so a survey designed to *measure*
intentions can *create* the intentions it claims to observe. Recognizing such
feedback is the first step toward using either instrument credibly. By the end of the
chapter the reader should be able to state the total-survey-error decomposition,
build an unbiased estimator from a known sampling design, correct cross-respondent
incomparability with anchoring vignettes, and lay out an experiment whose causal
claim survives scrutiny.

## Surveys as Measurement

A survey is a measurement device whose output is a set of self-reports. Two
properties make it powerful and dangerous in equal measure. It is *cheap*, so it
scales to populations and to questions—intentions, attitudes, satisfaction—that
leave no behavioral trace. And it is *reactive*: the instrument can perturb the
state it measures.

### Self-Generated Validity: When Measurement Changes Behavior

The canonical reactivity result concerns purchase intentions. Researchers routinely
ask "How likely are you to buy product $X$ in the next month?" and treat the answer
as a noisy reading of a pre-existing propensity. @chandon2005 show that the reading
is not passive. Merely asking the intentions question increases the *correspondence*
between stated intentions and subsequent purchase—a phenomenon they call
**self-generated validity**. The respondent, prompted to articulate an intention,
forms or crystallizes an attitude that did not fully exist before the question and
then acts on it. The intention–behavior link is therefore endogenous to the
measurement: it is, on average, **58% stronger among surveyed consumers than among
otherwise-comparable nonsurveyed consumers** [@chandon2005]. The implication is
sharp. An intentions survey does not merely forecast demand; it is a light-touch
*intervention* that nudges demand. A forecast calibrated on surveyed respondents will
overstate the predictability of an unsurveyed population.

> **Self-generated validity.** The increase in the predictive validity of a stated
> intention that is *caused by the act of stating it*. Asking the question
> reactively strengthens the very intention–behavior correlation the researcher
> intends to measure passively [@chandon2005].

### The Intention–Behavior Correlation

How tightly do intentions track behavior once reactivity is acknowledged? The
meta-analytic benchmark from the theory of reasoned action places the
intention–behavior correlation near $\rho \approx 0.53$ across domains
[@sheppard1988]. This is a substantial but far-from-deterministic association: stated
intentions explain on the order of $\rho^2 \approx 28\%$ of the variance in
subsequent behavior. The residual is the space in which forecasting models earn their
keep. @morwitz1992 show that **segmenting before forecasting** materially improves
predictive power: a model that uses historical purchase behavior together with stated
intention, and that allows the mapping from intention to purchase to differ across
consumer segments, forecasts sales better than a pooled model that imposes a single
intention–purchase slope. The intuition is that the intention–behavior gap is
heterogeneous—deliberate planners realize their stated intentions far more reliably
than impulsive or constrained consumers—so pooling averages over groups whose mapping
functions genuinely differ.

The practical sequence is summarized in @fig-intent-pipeline.

```{mermaid}
%%| label: fig-intent-pipeline
%%| fig-cap: "From elicited intentions to a sales forecast. The dashed feedback edge is self-generated validity: the act of measurement perturbs the behavior being forecast."
flowchart LR
    A["Elicit stated<br/>intention"] --> B["Segment respondents<br/>(planners vs. impulsive,<br/>constrained, etc.)"]
    B --> C["Estimate segment-specific<br/>intention to purchase mapping"]
    C --> D["Aggregate to<br/>sales forecast"]
    A -. "self-generated<br/>validity" .-> E["Actual purchase<br/>behavior"]
    D --> E
```

## The Total Survey Error Framework

A survey statistic can miss the truth for two broad reasons: the *people* who answer
are not the people the analyst wanted, and the *answers* they give are not the values
the analyst wanted. The **total survey error** (TSE) framework organizes every way a
survey can go wrong under these two headings and, orthogonally, separates systematic
error (**bias**) from sampling-driven dispersion (**variance**) [@Groves_2010].
Formally, for an estimator $\hat\theta$ of a population target $\theta$, the relevant
loss is mean squared error,
$$
\mathrm{MSE}(\hat\theta) = \underbrace{\bigl(\mathbb{E}[\hat\theta] - \theta\bigr)^2}_{\text{bias}^2}
                          + \underbrace{\operatorname{Var}(\hat\theta)}_{\text{variance}},
$$ {#eq-tse-mse}
and TSE decomposes both terms by *source*. The first decomposition is by stage of the
survey process:
$$
\text{Total Survey Error} = \underbrace{\text{Measurement Error}}_{\text{the answers}}
                          + \underbrace{\text{Representation Error}}_{\text{the people}}.
$$ {#eq-tse-decomp}

The **measurement** branch tracks a single response from the target *construct* the
analyst cares about, to the *measurement* the question actually taps, to the
*response* the person gives, to the *edited datum* stored after processing—accruing
validity error, response error, and processing error along the way. The
**representation** branch tracks the *units* from the target *population*, to the
*sampling frame* (the list from which units are drawn), to the *sample*, to the
*respondents* who actually answer—accruing coverage error, sampling error, and
nonresponse error. @fig-tse renders both branches.

```{mermaid}
%%| label: fig-tse
%%| fig-cap: "The total survey error framework: a measurement (left) and a representation (right) branch, each accumulating distinct error sources from construct/population down to the final respondent-level datum. Adapted in original form from the decomposition in @Groves_2010."
flowchart TB
    subgraph M["Measurement (the answers)"]
        C1["Construct"] -->|validity| C2["Measurement"]
        C2 -->|response error| C3["Response"]
        C3 -->|processing error| C4["Edited response"]
    end
    subgraph R["Representation (the people)"]
        P1["Target population"] -->|coverage error| P2["Sampling frame"]
        P2 -->|sampling error| P3["Sample"]
        P3 -->|nonresponse error| P4["Respondents"]
    end
    C4 --> S["Survey statistic"]
    P4 --> S
```

The framework is more than a taxonomy: it tells the analyst *where to spend the next
dollar*. A statistic can be unbiased yet imprecise (a clean design with a small
sample), or precise yet badly biased (a large but unrepresentative convenience
panel). Because the two failure modes call for opposite remedies—more units versus a
better frame—naming the dominant source is the prerequisite to fixing it.

### Three Eras of Survey Practice

The TSE trade-offs have not been static; the technology of contacting respondents has
reshaped which errors bind. @tbl-eras sketches the three eras now standard in survey
methodology. The trajectory is one of falling cost and falling response rates: each
era widened reach and cut expense while making the representation branch harder to
control, culminating in non-probability online panels that must lean on modeling
rather than design to justify inference.

| Era | Sampling | Interview mode | Data environment |
|-----|----------|----------------|------------------|
| First | Area probability | Face-to-face | Stand-alone |
| Second | Random-digit-dial probability | Telephone | Stand-alone |
| Third | Non-probability | Computer-administered | Linked |

: Three eras of survey research, after the design taxonomy popularized in survey methodology. {#tbl-eras}

## Probability and Non-Probability Sampling

The representation branch lives or dies on the **sampling design**. A sample is a
**probability sample** if every unit in the frame has a *known, non-zero* probability
of inclusion. This single property is what licenses design-based inference: because
the inclusion probabilities are known, the analyst can *reweight* the sample to
recover an unbiased estimate of the population total even when the design
oversamples some units.

### The Horvitz–Thompson Estimator

Let $\pi_i = \Pr(i \in s)$ be unit $i$'s inclusion probability and let $y_i$ be the
quantity of interest. The **Horvitz–Thompson** estimator of the population mean
weights each sampled observation by the inverse of its inclusion probability,
$$
\hat{\bar y}_{\text{HT}} = \frac{1}{N}\sum_{i \in s} \frac{y_i}{\pi_i},
$$ {#eq-ht}
where $N$ is the population size and $s$ the realized sample. Inverse-probability
weighting is the formal content of the slogan "weighting recovers
representativeness": a unit sampled at half the average rate stands in for twice as
many population members and so receives twice the weight. The estimator is
*design-unbiased*—$\mathbb{E}[\hat{\bar y}_{\text{HT}}] = \bar y$ under the
randomization that generated $s$—for *any* design with strictly positive $\pi_i$,
which is precisely why the positivity requirement is non-negotiable.[^positivity]

[^positivity]: If $\pi_i = 0$ for some unit, that unit can never enter the sample and
no weight can resurrect it; the corresponding stratum is structurally invisible and
no amount of reweighting recovers it. This is the formal statement of coverage error
in @eq-tse-decomp.

A short simulation makes the bias correction concrete. We construct a population in
which the outcome is correlated with the inclusion probability—the worst case for a
naive sample mean—and verify that the Horvitz–Thompson estimator recovers the truth
while the unweighted mean does not.

```{r ht-estimator, message=FALSE, warning=FALSE}
set.seed(31)

N <- 100000
# A covariate that drives BOTH the outcome and the sampling probability,
# so the naive sample mean is biased and weighting is required.
x      <- runif(N)
y      <- 2 + 3 * x + rnorm(N, sd = 0.5)        # population outcome
pop_mean <- mean(y)

# Unequal-probability design: high-x units are sampled more often.
pi_i   <- 0.02 + 0.08 * x                         # inclusion probability in (0.02, 0.10)
in_sample <- runif(N) < pi_i

y_s    <- y[in_sample]
pi_s   <- pi_i[in_sample]

naive_mean <- mean(y_s)                           # ignores the design -> biased
ht_mean    <- sum(y_s / pi_s) / N                 # Horvitz-Thompson -> unbiased

cat("Population mean:        ", round(pop_mean, 4), "\n")
cat("Naive sample mean:      ", round(naive_mean, 4), "  (biased upward)\n")
cat("Horvitz-Thompson mean:  ", round(ht_mean, 4),  "  (design-unbiased)\n")
```

### The Non-Response Problem

The third-era pivot to non-probability and online sampling breaks the clean logic
above, because the operative inclusion probability is no longer the *design*
probability the analyst set but the *response* probability the respondent chose. Let
$R_i \in \{0,1\}$ indicate that unit $i$ both is selected and chooses to respond, with
response propensity $\rho_i = \Pr(R_i = 1)$. The respondent-based estimator is
unbiased only if the analyst knows $\rho_i$, and in practice $\rho_i$ must be
*estimated*—typically by modeling response as a function of observed covariates. This
is where non-response error enters: if response depends on the outcome $y_i$ even
after conditioning on observed covariates (non-response that is *not* missing at
random), no reweighting on observables removes the bias, and the survey is identified
only under an untestable assumption about the missingness mechanism. The TSE framework
thus locates the modern survey's hardest problem precisely: representation error
migrates from a *design* quantity the analyst controls to a *behavioral* quantity the
analyst must model [@Groves_2010].

## Anchoring Vignettes

The measurement branch has a subtler pathology than noisy answers: respondents may
use the *same* response scale to mean *different* things. Ask consumers to rate their
satisfaction "on a 1–5 scale" and a lenient respondent's 4 may encode the same
underlying experience as a demanding respondent's 2. This **interpersonal
incomparability**—formally, **differential item functioning** (DIF)—means that
differences across respondents partly reflect differences in how they map a latent
state onto categories, not differences in the latent state itself. Agreement on the
exact meaning of an abstract scale point is, in practice, nearly impossible to
secure, and the problem is acute for constructs that can only be conveyed by example.

**Anchoring vignettes** solve this by adding, alongside the self-assessment, a small
set of fixed hypothetical scenarios that every respondent rates on the same scale
[@king2004; @king2007; @hopkins2010]. The key move is that the analyst *authors* the
vignettes, so their true level is **fixed across respondents by construction**. Any
variation in how respondents rate the *same* vignette is therefore pure DIF—a direct
reading of each respondent's idiosyncratic scale use—which can then be netted out of
the self-assessment. The logic is to *measure as usual, then subtract the
incomparable portion*, using each respondent's own vignette ratings to recalibrate
that same respondent's self-rating.

Two assumptions identify the correction [@king2004]:

- **Response consistency.** Each respondent applies the *same* mapping from latent
  state to response category when rating the vignettes as when rating themselves.
  Without this, the vignette ratings carry no information about the respondent's
  self-rating scale.
- **Vignette equivalence.** The latent level a given vignette describes is perceived
  identically by all respondents; only its *translation* into a category varies. This
  is what makes cross-respondent differences in vignette ratings attributable to DIF
  rather than to genuine disagreement about the scenario.

### Nonparametric Correction

The nonparametric estimator requires no distributional assumptions. For each
respondent, the analyst recodes the self-assessment $y_i$ relative to where it falls
among that respondent's *ordered* vignette ratings $z_{i1} \le z_{i2} \le \dots \le
z_{iJ}$. The rescaled outcome $C_i$ records the self-rating's rank position within the
respondent's own vignette thresholds—how good the respondent is relative to anchors
whose true level is known—so it is comparable across respondents by construction.
Order **inconsistencies**, where a respondent's vignette ratings are non-monotone in
the vignettes' known true order, are absorbed as **ties** (interval-valued
categories). The recoded $C_i$ is then modeled, for example with an ordered probit,
purged of the respondent-specific scale shifts.

### Parametric Correction

The parametric estimator embeds the same idea in a likelihood. Self-assessment and
vignettes are jointly modeled with an **ordered probit** in which the *cut points*
(category thresholds) are allowed to vary across respondents as functions of
covariates, with a respondent **random effect** capturing residual scale
heterogeneity. Identification improves as the number of vignettes $J$ grows—more
anchors pin down each respondent's thresholds more tightly—but there is a bias–variance
trade-off: each additional vignette is itself measured with error, so adding vignettes
trades sharper threshold identification against more measurement noise [@hopkins2010].
The `anchors` package that implemented these estimators was archived from CRAN, but the
nonparametric recoding is simple to reproduce directly. The illustration below takes three
respondents with *identical* true ability who differ only in how harshly they rate, and
shows that anchoring collapses their incomparable raw self-ratings to the same position.

```{r anchors-demo, message=FALSE, warning=FALSE}
library(tidyverse)

# Each respondent rates three vignettes of known increasing quality (v1 < v2 < v3) plus
# themselves, on a 1-5 scale. They share the same true level but differ in strictness.
demo <- tribble(
  ~respondent, ~self, ~v1, ~v2, ~v3,
  "lenient",      5L,   3L,  4L,  5L,
  "moderate",     4L,   2L,  3L,  4L,
  "harsh",        3L,   1L,  2L,  3L
)

# Nonparametric anchoring ("C"): locate each self-rating relative to the respondent's
# OWN vignette ratings (ties -> interval categories); the personal scale shift cancels.
anchor_pos <- function(self, v) {
  v <- sort(v)
  paste0(sum(self > v), " above, ", sum(self == v), " tied")
}

demo %>%
  rowwise() %>%
  mutate(anchored = anchor_pos(self, c(v1, v2, v3))) %>%
  ungroup() %>%
  select(respondent, raw_self = self, v1, v2, v3, anchored)
```

The raw self-ratings (5, 4, 3) look different, but every respondent rates themselves
"2 above, 1 tied" relative to their own vignettes—anchoring reveals the apparent gap was
pure rating strictness, exactly what the parametric ordered-probit version corrects for.

## Experiments

Where a survey *observes* a latent state, an experiment *manufactures* variation in a
cause and reads off its effect. The experiment's payoff is causal identification: by
randomly assigning the treatment, the analyst makes the treatment statistically
independent of every confounder, observed or not, so the difference in mean outcomes
estimates the average treatment effect rather than a confounded association. The cost
is that the analyst must defend four distinct kinds of **validity**, manage the
heterogeneity that averages conceal, and respect the ethics of intervening on people.

### Four Validities

A credible experiment clears four hurdles, summarized in @tbl-validity. They are
ordered roughly from "is there an effect at all" to "does it mean what we claim" to
"does it travel."

| Validity | Question it answers | Characteristic threat |
|----------|--------------------|-----------------------|
| Statistical conclusion | Is the treatment–outcome covariation real, not noise? | Low power; fishing; violated test assumptions |
| Internal | Is the covariation *causal* within the study? | Confounding; attrition; failed randomization |
| Construct | Does the operationalization capture the intended concept? | Mono-operation bias; demand effects |
| External | Does the effect *generalize* beyond this sample and setting? | Non-representative subjects; artificial context |

: The four validities an experiment must defend, in roughly increasing order of generality. {#tbl-validity}

The four trade off against one another. Tight laboratory control buys internal
validity at the expense of external validity (realism), while a field experiment buys
realism at the expense of control. The same tension recurs in the practical design
levers—**cost, control, realism, and ethics**—that no single design optimizes
simultaneously.

### Heterogeneity and Mechanisms

A single number—the **average treatment effect** (ATE)—can be an artifact. If the
effect is positive for one segment and negative for another, the average may be near
zero, near either extreme, or anywhere between, depending only on segment mix. Writing
the conditional average treatment effect as $\tau(x) = \mathbb{E}[Y(1) - Y(0)\mid X=x]$,
the reported ATE is the mix-weighted average $\int \tau(x)\,dF(x)$, which is silent
about the sign and size of $\tau(x)$ for any particular $x$. The same caution that
motivated segmentation in forecasting (@sec-surveys) applies to experiments:
heterogeneous effects demand that the analyst report the distribution of $\tau(x)$,
not only its mean, and that any claimed **mechanism**—the causal pathway through which
the treatment acts—be argued rather than assumed.

### Ethics: The 3Rs

Experiments intervene on living subjects, and the governing ethical framework is the
**3Rs** of Russell and Burch (1959):[^threers] *Replace* an invasive procedure with a
less invasive (or non-animal, or natural) alternative; *Refine* procedures to reduce
harm and distress; and *Reduce* the number of participants to the minimum consistent
with a valid answer. The Reduce principle has a statistical edge: the minimum sample
is set by the power calculation, so an underpowered study is not only wasteful but
also *unethical*, exposing participants to risk without a credible chance of a
conclusive result.

[^threers]: Russell, W. M. S. and Burch, R. L. (1959), *The Principles of Humane
Experimental Technique*. The 3Rs framework predates and underlies most modern
research-ethics review.

### Pre-Registration and the Replication Crisis

The credibility of an experimental finding depends not only on its design but on
whether the analysis was specified *before* the data were seen. **Pre-registration**
time-stamps the hypotheses, design, and analysis plan in advance, so a reviewer can
distinguish a confirmatory test from a hypothesis read off the data after the
fact—the practice that inflates false-positive rates and fuels the replication crisis
in the behavioral sciences. Public registries such as AsPredicted provide the
time-stamp; open-data and open-materials platforms such as the Open Science Framework
and ResearchBox let others re-run the analysis.

A complementary forensic tool is the **$p$-curve**: the distribution of statistically
significant $p$-values across a body of studies. Under a true effect, that
distribution is *right-skewed* (many very small $p$-values); under no effect but
selective reporting ($p$-hacking), it is *left-skewed* (a pile-up just below 0.05).
Inspecting the curve thus diagnoses whether a literature reflects real effects or
selective analysis.[^pcurve] A final design caution: testing for **interaction**
effects requires controlling for the *interactions* of the covariates, not merely
their main effects—controlling only for main effects leaves the interaction estimate
confounded.

[^pcurve]: The $p$-curve was developed and popularized through the Data Colada
research collective; the underlying logic is that selective reporting leaves a
detectable signature in the shape of the significant-$p$ distribution.

## Key Takeaways

- Surveys and experiments are both *instruments for recovering an unobserved
  quantity*; surveys fail through **error** (@eq-tse-decomp) and experiments through
  threats to **validity** (@tbl-validity).
- Measurement is **reactive**: asking about purchase intentions strengthens the
  intention–behavior link by roughly 58%, so intentions surveys are mild
  interventions, not passive forecasts [@chandon2005].
- Total survey error separates **measurement** error (the answers) from
  **representation** error (the people), and bias from variance; naming the dominant
  source dictates the fix [@Groves_2010].
- Probability designs license design-based inference through the
  **Horvitz–Thompson** estimator (@eq-ht); non-probability and non-response designs
  shift the burden from *design* to *modeling* the inclusion mechanism.
- **Anchoring vignettes** purge interpersonal incomparability by netting out
  differential item functioning, identified by response consistency and vignette
  equivalence [@king2004].
- Causal claims from experiments rest on four validities, demand attention to
  effect **heterogeneity** beyond the average, and gain credibility from
  **pre-registration** and open science.
