# Preference Measurement {#sec-preference-measurement}

*Preference measurement* is the problem of recovering, from observable choices or
judgments, the latent valuation a consumer places on a product and on each of the
attributes that compose it. It is the empirical bridge between what people *do*—rate,
rank, click, buy—and what firms need to *know*: how much a feature is worth, what a
segment will pay, which configuration to launch, and at what price. Where
@sec-branding asks what a brand is worth, this chapter asks the more granular
question of what each *part* of an offering is worth, and supplies the estimators
that turn survey or transaction data into those numbers.

The construct that anchors the chapter is the **reservation price** (also called
willingness to pay, WTP): the maximum amount a consumer would pay for a product
before preferring to forgo it. Reservation price is not a primitive that can simply
be asked for—people are reluctant to reveal price sensitivity, and direct questions
bias the answer downward—so it must be *inferred* from a model of utility. The
dominant inferential machinery is **conjoint analysis**, a family of designed
experiments and discrete-choice models that decompose overall preference into
attribute-level *partworths* and, with a price attribute included, express those
partworths in dollars.

The chapter proceeds from intuition to formalism to estimation. We first define
reservation price and derive it from an additive utility model, making explicit the
assumptions that license the derivation. We then develop conjoint analysis as the
general framework: its experimental-design logic, its measurement variants
(full-profile, choice-based, adaptive, hybrid, self-explicated, MaxDiff), the
random-utility model that underlies the modern choice-based form, and the estimators
(OLS, multinomial and mixed logit, hierarchical Bayes) that recover heterogeneous
partworths. Throughout we flag what breaks identification—confounded designs,
scale–heterogeneity confounds, hypothetical bias—because a partworth that is not
identified is a number without a referent. Worked, reproducible R code accompanies
each major method.

## Reservation Price and the Additive Utility Model

The commercial questions branded around pricing—segmentation, positioning,
product-line design, the go/no-go decision on a feature—reduce to a single
quantity per consumer: how many dollars of value the consumer attaches to a given
configuration. Two routes lead to that quantity. The **direct** route asks the
consumer outright, but self-reported reservation prices are systematically biased
downward: respondents understate WTP because they do not wish to appear price
insensitive, and because a hypothetical question carries no budget consequence.[^direct]
The **indirect** route, which this chapter develops, recovers reservation price as a
*derived* quantity from a fitted utility model, most commonly the coefficients of a
conjoint study.

[^direct]: Direct elicitation is not worthless—open-ended WTP, the Becker–DeGroot–
Marschak mechanism, and contingent valuation all have their uses—but for
multi-attribute products the indirect route dominates because it yields attribute-level
dollar values, not just a single product-level number, and it embeds the valuation in
a choice context that disciplines hypothetical bias.

The indirect route rests on one structural assumption, which should be stated before
any result: **additive separability** of utility across attributes. Consumption
utility is taken to equal the sum of the utilities contributed by the product's
attributes, with no interaction among them. Let a product be described by a price $P$
and by $N$ non-price attributes, each taking one of several discrete levels. Index the
attributes by $k = 1,\dots,N$ and let $A_k$ denote the level of attribute $k$. The
additive model writes a consumer's utility (or rating) as

$$
U = \beta_0 + \beta_p P + \sum_{k=1}^{N} \beta_k A_k + \epsilon,
$$ {#eq-additive-utility}

where the $\beta$'s are *utility weights* (partworths in conjoint language),
$\beta_0$ absorbs consumer-specific shifts in the rating scale, $\beta_p$ is the
marginal utility of price (expected to be negative), and $\epsilon$ is an idiosyncratic
error. Equation @eq-additive-utility is the linear, additive special case of the
random-utility model formalized in @sec-rum; the categorical attributes $A_k$ are in
practice expanded into level dummies, so each attribute contributes one coefficient per
non-baseline level.

The dollar value of one *util* follows immediately. Because price enters linearly,
the marginal rate of substitution between money and utility is the negative reciprocal
of the price coefficient. If $\Delta p$ is the price difference between two levels of
the price attribute, the dollar value of a unit of utility is

$$
\text{\$ per util} = -\frac{\Delta p}{\beta_p},
$$ {#eq-dollar-per-util}

and the utility a consumer derives from attribute $k$, expressed per unit of that
attribute, is $(\beta_k / \Delta A_k)\,A_k$, where $\Delta A_k$ is the spacing between
adjacent levels of attribute $k$. The **reservation price** for a configuration is then
the total utility of its non-price attributes converted to dollars:

$$
r(P) = -\frac{\Delta p}{\beta_p}\sum_{k=1}^{N}\frac{\beta_k}{\Delta A_k}\,A_k.
$$ {#eq-reservation-price}

The interpretation is exactly the intuition with which we began. A consumer's
reservation price for a product is the sum of the dollar values she places on each of
its attributes—each attribute's partworth, scaled by the dollar value of a util in
@eq-dollar-per-util. Two assumptions make @eq-reservation-price legitimate, and both
fail in identifiable circumstances. First, **additivity**: if attributes interact
(a fast processor is worth more in a laptop with a good screen), the cross-terms
omitted from @eq-additive-utility are loaded onto the main effects and the partworths
are biased. Second, **linearity of the price disutility**: if $\beta_p$ is not
constant across the price range—if marginal price sensitivity rises near a budget
ceiling—then $\Delta p / \beta_p$ is not a single number and @eq-dollar-per-util is
only a local approximation. We return to both threats once the design machinery is in
place.

::: {.callout-note}
## Sign and scale conventions
We write $\beta_p < 0$, so the dollar-per-util conversion in @eq-dollar-per-util carries
an explicit negative sign that makes $r(P)$ positive for value-adding attributes. In a
ratings model the scale of $U$ is set by the response scale and $\beta_0$ absorbs
per-respondent scale heterogeneity; in a choice model (@sec-rum) the scale is fixed by
normalizing the error variance, which has consequences for cross-respondent comparison
that we take up under the scale–heterogeneity confound.
:::

## Conjoint Analysis {#sec-conjoint}

Conjoint analysis—originally *conjoint measurement*, a term inherited from mathematical
psychology and psychometrics—is the workhorse for estimating the partworths in
@eq-additive-utility. Its premise is that consumers evaluate products holistically, as
bundles, and that the analyst can recover the weight on each attribute by observing
evaluations of systematically varied bundles. The name encodes the logic: the joint
("con-joint") effect of attributes considered together is decomposed into separable
contributions.

> Conjoint analysis is a decompositional method that estimates the structure of a
> consumer's preferences given her overall evaluations of a set of alternatives that
> are prespecified in terms of levels of different attributes.

The method's enduring appeal is twofold. It improves the analyst's ability to assess
customers' true wants—above all their price sensitivity, which direct questioning
distorts—and it dissolves the "everything is important" problem that plagues
self-reported importance ratings, forcing trade-offs that reveal which attributes
genuinely move choice. @netzer2008 survey the field and organize it around three
managerial tasks that conjoint serves—product development, pricing and segmentation,
and positioning—together with parallel uses for consumers (recommendation agents),
policymakers, and academic researchers; the review traces the field's movement from
designs that maximize statistical efficiency toward designs and adaptive elicitation
that also respect managerial and respondent constraints.

A foundational distinction motivates why preference, not perception, is the target.
*Perception*—beliefs about what products exist and what they are like—is largely
**homogeneous** across consumers: people broadly agree that one car is faster than
another. *Preference*—which product a consumer actually wants—is **heterogeneous**:
people who agree on the facts still choose differently [@netzer2008]. Conjoint is a tool
for the second object, and its modern forms are built to estimate the *distribution* of
partworths across consumers, not merely their average. The original combinatorial
treatment of @greenwind1975 already contained the core recipe: use an orthogonal array
to construct a balanced subset of attribute combinations so that each attribute's
contribution is estimated independently of the others, then collect rankings or ratings
over that reduced set. Their catalog of applications—new-product development, package
design, pricing and brand alternatives, service design, and industrial (organizational)
buying—remains an accurate map of where conjoint is used today.

### Designing the Attribute Space

The quality of a conjoint study is fixed before any data are collected, in the choice of
attributes and levels. Four design principles recur, and each protects a specific
inferential property rather than being mere good practice. Attributes must be
**actionable**—a partworth on a feature the firm cannot build is uninterpretable as a
decision input. Their number must be **bounded**, because respondent cognitive load
grows with the dimensionality of each profile and noisy responses inflate $\epsilon$ in
@eq-additive-utility. Levels must span a **feasible and relevant range**: every level
should be one the firm could actually offer and one the consumer would plausibly
encounter, since extrapolating partworths outside the studied range is unsupported.
Finally, the **number of levels** per attribute is itself a design variable with a known
artifact—the *number-of-levels effect*, whereby an attribute is estimated to be more
important simply because it is split into more levels, independent of its true
importance.[^nlevels]

[^nlevels]: The number-of-levels effect is partly a halo of the level count and partly an
ordering artifact; it means importance comparisons across attributes with different
numbers of levels are not clean. The defensive design is to equalize level counts where
possible, or to interpret importances within, not across, attributes.

Under additivity and independence, the number of profiles the design must support equals
the number of estimated dummy variables. When the analyst suspects an **interaction**—a
violation of the additive separability that @eq-additive-utility assumes—two remedies are
available, and they trade respondent effort against bias. One can fold the interacting
pair into a single higher-level "super-attribute" whose levels enumerate the relevant
combinations, absorbing the interaction into main effects; or one can enlarge the design
to include enough profiles to estimate the interaction term explicitly. The first is
cheaper but coarsens the attribute space; the second is faithful but costly in profiles.

The mechanism that makes a reduced design estimable is the **fractional factorial**: a
carefully chosen subset of the full factorial that preserves orthogonality among main
effects so that partworths remain independently identified. A full factorial over $N$
attributes with $L$ levels each contains $L^N$ profiles—infeasible for any realistic $N$—
so the analyst selects a fraction in which the columns of the design matrix remain
uncorrelated. @fig-conjoint-pipeline summarizes the workflow from attribute definition
through to managerial action.

```{mermaid}
%%| label: fig-conjoint-pipeline
%%| fig-cap: "The conjoint analysis pipeline, from attribute specification to managerial decision. The estimation step varies by conjoint type (see @tbl-conjoint-types)."
flowchart TD
    A[Define attributes & levels<br/>actionable, feasible, bounded] --> B[Construct design<br/>fractional factorial / orthogonal array]
    B --> C[Choose elicitation<br/>rate · rank · choose · MaxDiff]
    C --> D[Collect responses]
    D --> E[Estimate partworths<br/>OLS · MNL · mixed logit · HB]
    E --> F[Derive quantities<br/>WTP · importance · share-of-preference]
    F --> G[Managerial action<br/>design · price · segment · position]
```

Conjoint is rarely run in isolation. It pairs naturally with **factor analysis** for
reducing a large attribute set to underlying dimensions, with **perceptual mapping**
(multidimensional scaling) for positioning the estimated preferences in a perceptual
space, and with **cluster analysis** for grouping consumers by their partworth vectors
into actionable segments—the empirical expression of preference heterogeneity.

### The Random-Utility Foundation {#sec-rum}

The ratings model in @eq-additive-utility treats $U$ as an observed (or directly
reported) quantity. Modern conjoint, especially its choice-based form, instead treats
$U$ as **latent** and observes only the consumer's *choice* among alternatives. This is
the random-utility model (RUM), the foundation McFadden built for discrete choice
[@mcfadden1986choice; @mcfadden2001economic]. Consumer $i$ derives utility from
alternative $j$ in choice set $\mathcal{S}_t$,

$$
U_{ijt} = \mathbf{x}_{jt}^{\top}\boldsymbol{\beta}_i + \varepsilon_{ijt},
$$ {#eq-32-rum}

where $\mathbf{x}_{jt}$ collects the attribute levels of alternative $j$ (including
price), $\boldsymbol{\beta}_i$ is consumer $i$'s partworth vector, and $\varepsilon_{ijt}$
is an unobserved component. The consumer chooses the alternative of greatest utility, so
the analyst observes $y_{it} = \arg\max_{j \in \mathcal{S}_t} U_{ijt}$. The systematic
part $\mathbf{x}_{jt}^{\top}\boldsymbol{\beta}_i$ is exactly the additive partworth sum of
@eq-additive-utility; what RUM adds is an explicit stochastic structure on $\varepsilon$
that delivers a likelihood for *choices*.

If the $\varepsilon_{ijt}$ are independent and identically distributed type-I extreme
value, the choice probabilities take the **multinomial logit** (MNL) form,

$$
\Pr(y_{it} = j) = \frac{\exp(\mathbf{x}_{jt}^{\top}\boldsymbol{\beta}_i)}
{\sum_{l \in \mathcal{S}_t}\exp(\mathbf{x}_{lt}^{\top}\boldsymbol{\beta}_i)}.
$$ {#eq-32-mnl}

The aggregate logit of @eq-32-mnl (with $\boldsymbol{\beta}_i = \boldsymbol{\beta}$ common
to all consumers) is the model behind the @guadagni1983 scanner-data application that
launched choice modeling in marketing, and it is the workhorse for demand estimation more
broadly [@berry1995]. Two identification facts must be stated. First, **only utility
differences are identified**: adding a constant to every alternative's utility leaves
choices unchanged, so one alternative's intercept is normalized to zero. Second, the
**scale is not separately identified** from the partworths: the error variance is
normalized (to $\pi^2/6$ in the logit), which means estimated $\boldsymbol{\beta}$ are
confounded with scale. This scale normalization is the root of the scale–heterogeneity
confound discussed below. MNL also carries the **independence of irrelevant alternatives**
(IIA) property—the odds of choosing $j$ over $l$ are independent of other alternatives—
which is implausible when alternatives are close substitutes and which the richer models
below relax.

### Heterogeneity: Mixed Logit and Hierarchical Bayes

Because preference is heterogeneous, a single $\boldsymbol{\beta}$ misrepresents the
market and, worse, can produce biased aggregate elasticities. The **mixed (random-
coefficients) logit** lets partworths vary across consumers, $\boldsymbol{\beta}_i \sim
g(\boldsymbol{\beta}\mid\boldsymbol{\theta})$, and integrates them out of the likelihood:

$$
\Pr(y_{it} = j) = \int \frac{\exp(\mathbf{x}_{jt}^{\top}\boldsymbol{\beta})}
{\sum_{l \in \mathcal{S}_t}\exp(\mathbf{x}_{lt}^{\top}\boldsymbol{\beta})}\,
g(\boldsymbol{\beta}\mid\boldsymbol{\theta})\,d\boldsymbol{\beta}.
$$ {#eq-mixed-logit}

Mixed logit relaxes IIA and accommodates flexible substitution, at the cost of a
likelihood with no closed form that must be simulated. In marketing the dominant route to
@eq-mixed-logit is **hierarchical Bayes** (HB): a Gaussian (or richer) population prior
$\boldsymbol{\beta}_i \sim \mathcal{N}(\bar{\boldsymbol{\beta}}, \boldsymbol{\Sigma})$ is
placed over individual partworths, and Markov chain Monte Carlo recovers a full posterior
for *each respondent's* partworth vector by borrowing strength across the panel
[@rossi2014; @allenby1999dynamic]. HB is what makes individual-level WTP estimable from a
handful of choices per respondent—too few to fit a separate model per person—and it is the
engine behind commercial choice-based conjoint. The same Bayesian decomposition recovers
brand value from choice data by splitting utility into attribute-driven and name-driven
components [@kamakura1993; @jedidi2003].

::: {.callout-warning}
## The scale–heterogeneity confound
Because the logit fixes the error scale, what looks like heterogeneity in partworths can
instead be heterogeneity in *scale*—how deterministically a respondent chooses. Two
respondents with identical relative preferences but different choice consistency will
appear to have proportionally different $\boldsymbol{\beta}_i$. WTP, a ratio of two
coefficients ($-\beta_k/\beta_p$), partially cancels scale, which is one reason WTP is
often more stable across respondents than raw partworths—but only if scale is common to
the numerator and denominator. Mis-attributing scale to preference is a standard threat to
identification in choice-based conjoint.
:::

### Variants of Conjoint {#sec-conjoint-variants}

Conjoint is a family, not a single method. The variants differ in how preferences are
*elicited* and therefore in what they assume about the respondent and what they can
estimate. @tbl-conjoint-types contrasts them.

| Variant | Elicitation task | Estimator | When it fits |
|---|---|---|---|
| **Full-profile** | Rate or rank complete profiles | OLS / regression on partworths | Few attributes; ratings interpretable |
| **Choice-based (CBC)** | Choose one profile from a set | MNL / mixed logit / HB | Realistic; recovers WTP and shares |
| **Adaptive (ACA)** | Tailored questions by prior answers | Self-explicated + tradeoff, often HB | Many attributes; shorten survey |
| **Hybrid** | Self-explicated stage + full-profile/choice | Two-stage / HB | Many attributes with efficiency |
| **Self-explicated** | Rate level desirability and attribute importance | Weighted compositional sum | Many attributes; minimal load |
| **MaxDiff** | Pick most- and least-preferred items | Best–worst MNL | Importance ranking of many items |

: Conjoint variants by elicitation task and estimator. Ranking and rating full-profile
conjoint yield substantively equivalent partworths [@kalish1991]. {#tbl-conjoint-types}

**Full-profile** conjoint presents complete product descriptions for rating or ranking and
estimates the partworths of @eq-additive-utility by regression. A classic result is that
ranking and rating data, though they look different, recover the same preference structure
and need not be treated as fundamentally distinct response modes [@kalish1991].

**Choice-based conjoint** (CBC), also called discrete-choice conjoint, replaces the rating
with a *choice* among profiles, often including a no-choice option. It is the most
behaviorally realistic variant—people choose in markets, they do not rate—and it is
estimated with the RUM machinery of @sec-rum, yielding WTP and market-share simulations
directly. CBC is now the default in commercial practice.

**Adaptive conjoint** (ACA) varies the choice sets dynamically, conditioning each question
on the respondent's earlier answers to concentrate information where the respondent's
preferences are most uncertain, thereby shortening the survey for high-dimensional
attribute spaces. The polyhedral methods of @toubia2013 formalize adaptive question
selection as iteratively shrinking the set of partworth vectors consistent with a
respondent's answers, and incentive-aligned and incentive-compatible variants address the
hypothetical-bias problem at the elicitation stage. **Hybrid** conjoint combines a
self-explicated stage with full-profile or choice tasks, and **self-explicated** conjoint
dispenses with profiles entirely, asking respondents to rate each level's desirability and
each attribute's importance and composing partworths from the product—cheap and scalable
but vulnerable to the "everything is important" inflation that holistic methods avoid.
**MaxDiff** (best–worst scaling) asks respondents to identify the most- and least-preferred
items in a set, a task that produces sharper discrimination among many items than rating
scales and is estimated as a best–worst extension of MNL.

A persistent practical problem is **missing data** in partial profiles—respondents do not
see every attribute or level. @bradlow2004learning develop a learning-based model that
imputes the missing levels by treating the partial profile as informative about the
respondent's inferences, rather than discarding incomplete responses; modern machine-
learning approaches extend this to learn preferences from sparse, adaptively collected data
[@ding2015learning]. The broader frontier integrates conjoint with text and image data and
with structural demand models so that designed-experiment partworths and revealed-preference
market data can be reconciled [@netzer2008hidden; @rao2017; @wedel2016].

## Worked Examples

The examples below use the `conjoint` and `AlgDesign` packages, which implement the
orthogonal-array design and OLS-based partworth estimation of full-profile conjoint. They
are deliberately small and reproducible; production CBC studies would substitute the HB
estimator of @sec-rum.

### Generating an Efficient Design

The first task is to reduce a full factorial to an estimable fractional design whose
columns remain (near-)orthogonal. The example below specifies a four-attribute product and
extracts an orthogonal fraction, then verifies that the selected profiles are mutually
near-uncorrelated—the property that keeps partworths independently identified.

```{r conjoint-design}
#| message: false
#| warning: false
library(conjoint)

set.seed(123)

# A four-attribute product; full factorial = 2 x 2 x 2 x 2 = 16 profiles
experiment <- expand.grid(
  Weight   = c("light", "heavy"),
  Price    = c("low", "high"),
  Warranty = c("2yr", "3yr"),
  Battery  = c("10h", "20h")
)

# Reduce to an orthogonal fraction
design  <- caFactorialDesign(data = experiment, type = "orthogonal")
encoded <- caEncodedDesign(design)

# Orthogonality check: off-diagonal correlations should be near zero
round(cor(encoded), 2)
```

### Estimating Partworths and Importances

With a design and a respondent's preferences in hand, OLS recovers the partworths. The
example uses the `tea` data shipped with the `conjoint` package: preferences of
respondents over tea profiles described by price, variety, kind, and aroma. We estimate one
respondent's partworths, then the aggregate importances—the share of total preference range
each attribute commands.

```{r conjoint-estimate}
#| message: false
#| warning: false
library(conjoint)
data(tea)

# Single-respondent partworths via OLS on the encoded design
caModel(y = tprefm[1, ], x = tprof)

# Aggregate attribute importances across all respondents
caImportance(y = tpref, x = tprof)
```

The importances sum to 100% and quantify how much of the preference variation each
attribute explains. They are the empirical answer to "what matters?"—and, unlike
self-reported importance ratings, they are forced to respect trade-offs, which is precisely
why conjoint dissolves the "everything is important" pathology.

### From Partworths to Willingness to Pay

The managerial payoff is dollar-valued WTP, obtained by dividing each attribute's partworth
by the (negative) price partworth, as in @eq-dollar-per-util. The synthetic example below
fits an aggregate model with a numeric price attribute so that the price coefficient is a
single slope, then converts non-price partworths to dollars.

```{r conjoint-wtp}
#| message: false
#| warning: false
set.seed(2024)

# Synthetic full-profile ratings for a 3-attribute product with numeric price
profiles <- expand.grid(
  price    = c(10, 20, 30),          # dollars
  battery  = c(0, 1),                # 0 = 10h, 1 = 20h
  warranty = c(0, 1)                 # 0 = 2yr, 1 = 3yr
)

# True partworths (utils): price hurts, battery and warranty help
beta_true <- c(intercept = 5, price = -0.15, battery = 1.2, warranty = 0.6)
profiles$rating <- with(profiles,
  beta_true["intercept"] + beta_true["price"] * price +
    beta_true["battery"] * battery + beta_true["warranty"] * warranty +
    rnorm(nrow(profiles), sd = 0.1)
)

fit <- lm(rating ~ price + battery + warranty, data = profiles)
b   <- coef(fit)

# Dollar value of one util = -1 / price coefficient  (Delta p = 1 here)
dollars_per_util <- -1 / b["price"]

wtp <- c(
  battery_20h_vs_10h = b["battery"]  * dollars_per_util,
  warranty_3yr_vs_2yr = b["warranty"] * dollars_per_util
)
round(wtp, 2)
```

The recovered WTP—the dollar premium a consumer would pay to upgrade battery life or extend
the warranty—is the reservation-price increment of @eq-reservation-price, attribute by
attribute. Summing these increments across a configuration's upgraded attributes yields the
configuration's total reservation price relative to the baseline product, which is the input
that pricing and product-line decisions actually consume.

## Pitfalls and Identification

Three threats recur, each tracing to an assumption made explicit above.

**Hypothetical bias.** Stated-preference partworths can diverge from revealed behavior
because survey choices carry no consequence. Incentive-aligned designs—where the respondent
faces a real chance of receiving the chosen option—and the reconciliation of conjoint
partworths against market data are the standard defenses [@toubia2017; @ding2015learning].
WTP, being a ratio, is somewhat insulated, but level effects on the absolute scale are not.

**Confounded designs.** If the fractional factorial is poorly chosen, attribute columns are
correlated and their partworths are not separately identified—the design, not the data,
determines the estimates. The orthogonality check in the worked example is not decorative;
it is the diagnostic that the partworths in @eq-additive-utility are recoverable at all.

**Heterogeneity and scale.** Treating a heterogeneous market with a single aggregate logit
biases elasticities through composition; the mixed-logit and HB estimators of @sec-rum exist
to fix this. But heterogeneity in choice *consistency* (scale) masquerades as heterogeneity
in *preference*, so individual-level partworths must be interpreted with the scale–
heterogeneity confound in mind. Segmenting on partworth vectors via cluster analysis is sound
only after this confound is addressed.

A final, structural caveat returns to additivity. Equation @eq-additive-utility and the
reservation-price formula @eq-reservation-price assume attributes contribute independently.
Where they do not—where a feature's value depends on the presence of another—the analyst must
either model the interaction explicitly (more profiles) or absorb it into a higher-level
attribute, as discussed in @sec-conjoint. A partworth estimated under a falsely additive model
is biased toward the configurations over-represented in the design, and the WTP derived from it
inherits that bias.

## Key Takeaways

- **Reservation price is derived, not asked.** Direct WTP questions bias downward; the
  defensible route recovers WTP from a fitted utility model as the partworth-to-price ratio in
  @eq-dollar-per-util.
- **Additive separability is the load-bearing assumption.** Equations @eq-additive-utility and
  @eq-reservation-price hold only when attributes do not interact and price disutility is
  locally linear; both fail in identifiable ways.
- **Choice-based conjoint with random utility is the modern default.** The MNL/mixed-logit/HB
  stack of @sec-rum estimates heterogeneous, individual-level partworths from realistic choice
  tasks [@mcfadden1986choice; @rossi2014].
- **Design determines identification.** Orthogonal fractional factorials keep partworths
  separately estimable; a confounded design produces numbers without referents.
- **Mind the confounds.** Hypothetical bias, the number-of-levels effect, and the
  scale–heterogeneity confound each threaten the interpretation of partworths and the WTP
  derived from them.
