# Qualitative Research {#sec-qualitative-research}

Qualitative research generates evidence that is not born as numbers: interview
transcripts, open-ended survey responses, field notes, social-media posts,
product reviews, and the imagery and video that increasingly dominate consumer
expression. To make such evidence cumulative and defensible, researchers
*code* it—assigning categorical or ordinal labels to passages, respondents, or
images according to a coding scheme. The credibility of every downstream claim
then hinges on a single question: would a second, independent analyst, applying
the same scheme, have produced the same labels? If coding is idiosyncratic to
the coder, the "findings" are an artifact of who did the reading, not a property
of the data. **Inter-rater reliability** (IRR), also called inter-coder
agreement, is the family of statistics that answers this question, and it is the
methodological hinge on which qualitative rigor turns.

This concern is not parochial to ethnography. The same machinery governs the
quality of any human-labeled dataset: relevance judgments in information
retrieval, sentiment labels that train a classifier, diagnostic ratings in
medicine, and—of growing importance in marketing—the gold-standard annotations
used to validate text- and image-mining pipelines (@sec-measurement-scales).
Whenever a learned model is benchmarked against "ground truth," that ground
truth is itself a set of human codes whose reliability must be quantified before
the benchmark means anything. Reliability is necessary but not sufficient for
validity: two coders can agree perfectly on a scheme that measures the wrong
construct. The agreement coefficients in this chapter certify *consistency*; the
*meaning* of the codes is established by the construct-validity arguments of
@sec-measurement-scales.

Before any of that, though, a boundary has to be drawn, because "qualitative
research" in marketing names two enterprises that do not share a standard of proof.
This chapter treats the first at length and the second in outline, and the reader
should know which is which.

## Two Qualitative Traditions {#sec-qual-two-traditions}

The machinery developed in the rest of this chapter belongs to what is best called
**qualitative content analysis**: text or images are treated as data, a coding scheme
imposes categories on them, and the coding is defensible to the degree that
independent analysts converge. Reliability is the governing criterion, and the
coefficients below are how it is certified. This is the tradition that supplies
gold-standard labels for the text- and image-mining pipelines of
@sec-text-as-data and @sec-image-processing, and it is the one most doctoral students
in quantitative marketing will actually use.

The second tradition is **interpretive**, and it does not accept that criterion.
Consumer culture theory (CCT) — the label @arnould2005culturetheory gave a body of
work that had accumulated for two decades without a name — studies consumption as
culturally constituted meaning rather than as behavior to be classified. Its object
is the marketplace as a site of identity projects, marketplace cultures, and
ideology, and its evidence is the depth of a reading rather than the reproducibility
of a label. On this view a second coder producing the same categories would
demonstrate only that the scheme was coarse enough to be mechanical; the analytic
value lies precisely in the interpretation a naive coder could not have produced.
@thompson1997interprethermeneut sets out the hermeneutic framework most often
invoked: meaning is built by moving iteratively between the part and the whole of a
consumption narrative, and between the informant's account and its cultural context,
with the researcher's own frame acknowledged rather than eliminated.

The traditions are not, however, symmetric in their looseness, and it is a mistake to
read the interpretive turn as licence. @spiggle1994interpretqualitati specifies the
operations that interpretive analysis actually performs — categorization,
abstraction, comparison, dimensionalization, integration, iteration, refutation —
and holds the analyst to demonstrating them, which is a demanding standard even
though it is not a reliability coefficient. @askegaard2011epistemolculture presses
further, arguing that CCT's habit of invoking "context" has become an epistemological
placeholder that needs the same scrutiny the tradition applies to everything else.

What the interpretive tradition buys, when it works, is access to phenomena a coding
scheme cannot represent because the categories do not exist in advance.
@price2000olderdispositi trace how older consumers dispose of special possessions and
find the transfer of meaning across generations governing decisions that no
transactional model anticipates. @giesler2012doppelganbrand follows Botox over a
decade and shows a brand image being contested and reconstructed by activists,
consumers, and the firm together — a market-creation process, not a positioning
choice. Findings of this shape are the tradition's justification.

For most readers the practical question is which standard applies to their own work,
and the answer follows from the claim being made. A claim that *these texts fall into
these categories at this rate* is a reliability claim and must clear the bar the rest
of this chapter sets. A claim that *this is what the practice means to the people
performing it* is an interpretive claim, and reporting a kappa for it is a category
error that borrows the appearance of rigor from a standard the work is not being held
to. Mixed designs are common and legitimate — an interpretive study that codes a
subset for descriptive reporting should meet both standards, each where it applies.

The chapter develops the IRR coefficients as a single coherent system rather
than a catalog. We begin with the deceptively simple idea of percent agreement
and show precisely why it fails. Correcting that failure—subtracting the
agreement expected by chance—yields the chance-corrected family that includes
Cohen's and Fleiss's kappa, Krippendorff's $\alpha$, and Kendall's $W$. We then
turn to the intraclass correlation coefficient (ICC) for continuous ratings,
which embeds reliability inside a variance-components model and exposes the
design assumptions the kappa family leaves implicit. Throughout, intuition
leads, the estimator and its assumptions follow in full, and every coefficient
is computed on reproducible data.

```{r}
#| label: setup-qual
#| message: false
#| warning: false
library(irr)        # agree, kappa2, kappam.fleiss, kappam.light, kripp.alpha, icc
library(DescTools)  # KendallW, ICC, CohenKappa with CIs
```

## The Chance-Correction Problem

Every chance-corrected agreement coefficient in this chapter is a special case
of one idea. Let $p_o$ be the proportion of items on which raters *observed*
agreement and let $p_e$ be the proportion of agreement *expected* if raters
labeled items independently at random. The corrected coefficient rescales
observed agreement against the headroom that chance leaves:

$$
S = \frac{p_o - p_e}{1 - p_e}
   = 1 - \frac{1 - p_o}{1 - p_e}.
$$ {#eq-chance-correction}

The numerator $p_o - p_e$ is agreement beyond chance; the denominator $1 - p_e$
is the maximum agreement beyond chance attainable. Thus $S = 1$ for perfect
agreement, $S = 0$ when observed agreement equals chance, and $S < 0$ when raters
agree *less* than chance—a diagnostic signal of systematic disagreement, not
mere noise. The coefficients differ only in how they define $p_o$ and, above all,
$p_e$: how chance is modeled, how many raters and categories are allowed, whether
categories are nominal or ordinal, and how partial credit is assigned to near
misses. Holding @eq-chance-correction fixed and varying its two inputs organizes
the entire family, summarized in @tbl-irr-map.

```{r}
#| label: tbl-irr-map
#| echo: false
irr_map <- data.frame(
  Coefficient = c("Percent agreement", "Cohen's κ", "Weighted κ",
                  "Fleiss's κ", "Krippendorff's α", "Kendall's W",
                  "ICC"),
  Raters = c("2+", "2", "2", "3+", "2+", "3+", "2+"),
  Scale = c("Any", "Nominal", "Ordinal", "Nominal/Ordinal",
            "Nominal–Ratio", "Ordinal (ranks)", "Interval/Continuous"),
  `Chance corrected` = c("No", "Yes", "Yes", "Yes", "Yes", "Yes (ties)", "Yes"),
  `Handles missing` = c("No", "No", "No", "Limited", "Yes", "No", "Limited"),
  check.names = FALSE
)
knitr::kable(
  irr_map,
  caption = "A map of inter-rater reliability coefficients. All chance-corrected
  measures instantiate Equation @eq-chance-correction; they differ in the rating
  scale they assume, the number of raters they admit, and how they model chance
  agreement and missing data."
)
```

### Why Raw Agreement Misleads

The naive measure of reliability is **percent agreement**: the share of items two
raters label identically,
$$
p_o = \frac{\text{number of agreements}}{\text{number of items}} \times 100.
$$
Its appeal is transparency; its flaw is fatal. Percent agreement credits the
agreement two raters would reach even if both labeled items by coin flip, and
that baseline grows with category imbalance. Consider a rare event coded present
in 5% of items. Two raters who blindly code everything "absent" agree 90% of the
time and never once detect the event—yet percent agreement applauds them. The
statistic conflates skill with the base rate, so it is uninterpretable across
studies with different category distributions. Chance correction
(@eq-chance-correction) exists precisely to strip out this base-rate-driven
floor.

```{r}
#| label: percent-agreement
data("diagnoses", package = "irr")  # 30 patients, 6 raters, psychiatric dx
head(diagnoses, 5)
agree(diagnoses[, 1:2])             # raw percent agreement, raters 1 and 2
```

The `diagnoses` data record six psychiatrists independently assigning each of 30
patients to one of five diagnostic categories—a canonical nominal-coding problem
isomorphic to coding open-ended responses into themes. Raw agreement looks
reassuring, but we cannot yet say whether it reflects genuine reliability or the
prevalence of common diagnoses.

## Cohen's Kappa: Two Raters, Nominal Categories

@Cohen_1960 supplied the first widely adopted correction for two raters and
nominal categories. Cohen's $\kappa$ takes $p_o$ as the observed proportion of
exact agreement and models chance under the assumption that each rater applies
their *own* marginal distribution of category usage independently of the other:

$$
\kappa = \frac{p_o - p_e}{1 - p_e},
\qquad
p_e = \sum_{k=1}^{K} \hat{p}_{1k}\,\hat{p}_{2k},
$$ {#eq-cohen-kappa}

where $\hat{p}_{1k}$ and $\hat{p}_{2k}$ are the marginal proportions with which
rater 1 and rater 2 use category $k$, and $K$ is the number of categories. The
product $\hat{p}_{1k}\,\hat{p}_{2k}$ is the chance probability that both land on
category $k$ when acting independently; summing over $k$ gives total expected
agreement. Substituting into @eq-chance-correction yields the coefficient. Cohen's
$\kappa$ presumes a *fixed* pair of raters (the marginals are specific to those
two coders) and *exhaustive, mutually exclusive* categories, and it treats every
disagreement as equally severe—appropriate when categories are nominal but
wasteful when they are ordinal.

To read a $\kappa$ value, the field relies on the verbal benchmarks of
@Landis_1977, reproduced in @tbl-landis. These thresholds are conventions, not
laws: they ignore $K$ (a given $\kappa$ is harder to achieve with few categories)
and the cost of disagreement, so they should anchor judgment rather than replace
it.

```{r}
#| label: tbl-landis
#| echo: false
landis <- data.frame(
  `Kappa range` = c("< 0.00", "0.00 – 0.20", "0.21 – 0.40",
                    "0.41 – 0.60", "0.61 – 0.80", "0.81 – 1.00"),
  Interpretation = c("Poor (worse than chance)", "Slight", "Fair", "Moderate",
                     "Substantial", "Almost perfect"),
  check.names = FALSE
)
knitr::kable(
  landis,
  caption = "Benchmarks for chance-corrected agreement following @Landis_1977.
  The bands are interpretive conventions, not significance thresholds, and do not
  adjust for the number of categories."
)
```

### Weighting Ordinal Disagreements

When categories are ordered—Likert intensities, severity grades, sentiment from
*very negative* to *very positive*—treating a one-step miss as harshly as a
five-step miss discards information. **Weighted kappa** generalizes
@eq-cohen-kappa by attaching a disagreement weight $w_{jk}$ to each cell of the
$K \times K$ rating table, with $w_{kk} = 0$ on the diagonal and weights
increasing in the distance $|j-k|$ off it. Both observed and expected
disagreement are then weighted:

$$
\kappa_w = 1 - \frac{\sum_{j,k} w_{jk}\, p_{o,jk}}{\sum_{j,k} w_{jk}\, p_{e,jk}},
$$ {#eq-weighted-kappa}

where $p_{o,jk}$ is the observed and $p_{e,jk} = \hat{p}_{1j}\hat{p}_{2k}$ the
chance proportion of the $(j,k)$ cell. Two weighting schemes dominate. *Linear*
weights, $w_{jk} = |j-k|$, penalize disagreement in proportion to the number of
steps; *quadratic* weights, $w_{jk} = (j-k)^2$, penalize large discrepancies far
more steeply. The choice is substantive: quadratic weights are appropriate when a
two-category error is much worse than two one-category errors, and—usefully—the
quadratically weighted $\kappa$ coincides with a two-way intraclass correlation
under mild conditions, linking the kappa and ICC families.[^kappa-icc]

[^kappa-icc]: The numerical equivalence of quadratically weighted $\kappa$ and
the two-way ICC holds when the marginal rating distributions are equal across
raters; it fails as the marginals diverge, which is one reason ICC is preferred
when raters use the scale with systematically different generosity (a *prevalence*
or *bias* problem the unweighted $\kappa$ also suffers, the so-called kappa
paradox).

```{r}
#| label: cohen-kappa
dx2 <- diagnoses[, c("rater1", "rater2")]
kappa2(dx2, weight = "unweighted")  # nominal: all disagreements equal
kappa2(dx2, weight = "equal")       # linear weights for ordinal scales
kappa2(dx2, weight = "squared")     # quadratic weights penalize large gaps
```

The accompanying $z$-test reports whether $\kappa$ exceeds zero; a $p$-value
below 0.05 indicates raters agree more than chance. That is a low bar—rejecting
$\kappa = 0$ is not evidence of *good* reliability—so the point estimate read
against @tbl-landis, with a confidence interval, remains the substantive summary.
`DescTools::CohenKappa()` returns that interval directly.

```{r}
#| label: cohen-kappa-ci
DescTools::CohenKappa(table(dx2$rater1, dx2$rater2), conf.level = 0.95)
```

## Fleiss's Kappa: Many Raters

Cohen's $\kappa$ is structurally limited to two raters. Coding teams routinely
use three or more, and—as in the `diagnoses` study—the particular raters
assigned to an item may vary. @eq-cohen-kappa cannot accommodate this because its
$p_e$ is built from two specific raters' marginals. **Fleiss's kappa** generalizes
to any number of raters by redefining agreement *per item* rather than per rater
pair, and by assuming the raters for each item are a random draw from a larger
pool, so a single category distribution describes chance for all items.

For $n$ items, $m$ raters per item, and $K$ categories, let $n_{ik}$ be the number
of raters who assign item $i$ to category $k$. The per-item observed agreement is
the proportion of agreeing rater pairs,
$$
P_i = \frac{1}{m(m-1)} \left( \sum_{k=1}^{K} n_{ik}^2 - m \right),
$$
and $\bar{P} = \tfrac{1}{n}\sum_i P_i$ is mean observed agreement. Expected
agreement uses the pooled category proportions
$\bar{p}_k = \tfrac{1}{nm}\sum_i n_{ik}$, giving
$$
\bar{P}_e = \sum_{k=1}^{K} \bar{p}_k^2,
\qquad
\kappa_{\text{Fleiss}} = \frac{\bar{P} - \bar{P}_e}{1 - \bar{P}_e}.
$$ {#eq-fleiss-kappa}

This is again @eq-chance-correction, now with multi-rater definitions of observed
and expected agreement. The key identifying assumption—that raters are
exchangeable draws from a common pool—is what permits a single $\bar{p}_k$ to
stand in for every item's chance baseline; it is appropriate when coders are
interchangeable but not when specific, non-substitutable experts are deliberately
paired.

```{r}
#| label: fleiss-kappa
kappam.fleiss(diagnoses)  # all six raters; no fixed-pairing assumption
```

A related multi-rater measure, **Light's kappa**, instead averages Cohen's
$\kappa$ over every pair of raters [@Cohen_1960]. It retains Cohen's pairwise,
fixed-rater chance model while summarizing across the team, and is preferable when
raters are *not* exchangeable—each pair's marginals are honored before averaging.

```{r}
#| label: lights-kappa
kappam.light(diagnoses)  # mean of pairwise Cohen's kappa
```

## Krippendorff's Alpha: A General Reliability Coefficient

The kappa family carries assumptions that real coding violates: it presumes
complete data (every rater codes every item), a fixed measurement level, and—for
Fleiss—exchangeable raters. **Krippendorff's $\alpha$** dispenses with these. It
is the most general agreement coefficient: it accommodates any number of raters,
missing values, unequal numbers of raters per item, and any measurement
level—nominal, ordinal, interval, or ratio—through a single pluggable difference
function. Rather than correcting observed *agreement*, $\alpha$ is built on
observed and expected *disagreement*:

$$
\alpha = 1 - \frac{D_o}{D_e},
$$ {#eq-krippendorff}

which is @eq-chance-correction rearranged: writing $S = 1 - (1-p_o)/(1-p_e)$ and
identifying $1 - p_o$ with observed disagreement $D_o$ and $1 - p_e$ with expected
disagreement $D_e$ makes the equivalence explicit. The observed disagreement
$D_o$ aggregates, over all pairs of ratings of the same item, a metric
$\delta^2(c,c')$ measuring how far apart categories $c$ and $c'$ are; the expected
disagreement $D_e$ aggregates the same metric over all pairs drawn from the data
*irrespective* of item, i.e., under independence. The genius of the construction
is that the measurement level enters only through $\delta^2$: it is an indicator
$\mathbb{1}[c \neq c']$ for nominal data, a squared rank distance for ordinal
data, and $(c-c')^2$ for interval data. One estimator therefore spans all scales,
and because $D_o$ and $D_e$ are computed over *coincidences* (pairs of available
ratings) rather than complete records, missing data are handled natively.

For interpretation, @Shelley_1984 recommend treating $\alpha \geq 0.80$ as
acceptable reliability and $\alpha \geq 0.667$ as the lowest threshold at which
tentative conclusions may be drawn—a deliberately conservative bar reflecting the
coefficient's use as a gatekeeper for publishable coding.

> Reliability data must be reproducible: a coefficient certifies that the data
> would arise again under independent coding by other competent coders applying
> the same instructions. By this standard, variables with $\alpha < 0.667$ should
> not be analyzed, and only $\alpha \geq 0.80$ licenses confident substantive
> claims [@Shelley_1984].

The following example—four observers rating twelve units, with missing entries
coded `NA`—computes $\alpha$ under each measurement level so the role of
$\delta^2$ is visible: the same coincidences yield different coefficients purely
because the distance between categories is defined differently.

```{r}
#| label: krippendorff
ratings_mat <- matrix(
  c(1, 1, NA, 1, 2, 2, 3, 2, 3, 3, 3, 3,
    3, 3, 3, 3, 2, 2, 2, 2, 1, 2, 3, 4,
    4, 4, 4, 4, 1, 1, 2, 1, 2, 2, 2, 2,
    NA, 5, 5, 5, NA, NA, 1, 1, NA, NA, 3, NA),
  nrow = 4, byrow = TRUE      # 4 observers (rows) x 12 units (columns)
)
kripp.alpha(ratings_mat, method = "nominal")   # delta = 1[c != c']
kripp.alpha(ratings_mat, method = "ordinal")   # rank-distance metric
kripp.alpha(ratings_mat, method = "interval")  # squared difference
kripp.alpha(ratings_mat, method = "ratio")     # ratio-scale metric
```

Moving from `nominal` to `interval` typically *raises* $\alpha$ when most
disagreements are between adjacent categories, because the metric awards partial
credit for near misses—exactly the ordinal logic of weighted kappa, generalized.
Reporting the measurement level alongside the coefficient is therefore essential:
an unqualified "$\alpha = 0.74$" is incomplete.

## Kendall's W: Agreement Among Rankings

Some qualitative tasks ask raters not to *categorize* items but to *rank* them:
ordering brand concepts by appeal, prioritizing themes by salience, or sorting
stimuli. Agreement is then concordance among orderings, and the natural
coefficient is **Kendall's coefficient of concordance** $W$. For $m$ raters each
ranking $n$ objects, let $R_i$ be the sum of ranks object $i$ receives. If raters
agree, some objects accumulate consistently high rank sums and others
consistently low, so the $R_i$ are *dispersed*; if raters are random, every object
draws a middling sum and the $R_i$ cluster. $W$ normalizes the realized dispersion
by its maximum:

$$
W = \frac{12 \sum_{i=1}^{n} \left(R_i - \bar{R}\right)^2}{m^2\,(n^3 - n)},
\qquad
\bar{R} = \frac{m(n+1)}{2},
$$ {#eq-kendall-w}

where the denominator $m^2(n^3-n)/12$ is the variance of the rank sums under
perfect agreement, so $W \in [0,1]$ with $1$ denoting unanimous ordering and $0$
denoting no association.[^kendall-corr] Under the null of no concordance,
$m(n-1)W$ is approximately $\chi^2_{n-1}$, giving a significance test; the
correction for tied ranks reduces the denominator accordingly.

[^kendall-corr]: $W$ relates linearly to the mean Spearman rank correlation
$\bar{r}_s$ across all rater pairs via $W = [(m-1)\bar{r}_s + 1]/m$. Because $W$
is bounded in $[0,1]$ it cannot represent *systematic disagreement* (negative
association), which $\bar{r}_s$ can—a limitation to keep in mind when raters may be
genuinely opposed rather than merely noisy.

```{r}
#| label: kendall-w
rankings <- cbind(
  rater1 = c(1, 6, 3, 2, 5, 4),
  rater2 = c(1, 5, 6, 2, 4, 3),
  rater3 = c(2, 3, 6, 5, 4, 1)
)
DescTools::KendallW(rankings, test = TRUE)
```

## Intraclass Correlation: Reliability as Variance Decomposition

When ratings are *continuous*—perceived quality on a 0–100 scale, sentiment
scores, expert intensity judgments—agreement becomes a question about variance.
The **intraclass correlation coefficient** (ICC) of @Shrout_1979 asks: of all the
variation in the ratings, how much is true between-subject signal versus rater
noise? Intuitively, a measurement is reliable when subjects differ from one
another far more than raters differ on the same subject. The ICC formalizes this
as the share of total variance attributable to subjects.

The ICC is defined inside a variance-components model. Let $y_{ij}$ be the rating
of subject $i$ by rater $j$. The two-way model decomposes it as
$$
y_{ij} = \mu + \alpha_i + \beta_j + \varepsilon_{ij},
\qquad
\alpha_i \sim \mathcal{N}(0, \sigma_\alpha^2),\;
\beta_j \sim \mathcal{N}(0, \sigma_\beta^2),\;
\varepsilon_{ij} \sim \mathcal{N}(0, \sigma_\varepsilon^2),
$$ {#eq-icc-model}

where $\mu$ is the grand mean, $\alpha_i$ the subject effect (the signal we want
to measure reliably), $\beta_j$ the rater effect (systematic leniency or severity
of rater $j$), and $\varepsilon_{ij}$ residual error. The variance components are
estimated from the rating table's mean squares by analysis of variance. The ICC
is then a ratio of variance components, but *which* ratio depends on three design
decisions that must be stated before estimation, because they change the
estimand—not merely its estimate:

1. **Model.** A *one-way* model treats only subjects as random and folds rater
   and error into a single residual; it applies when each subject is rated by a
   *different*, randomly chosen set of raters. A *two-way* model adds a rater
   effect $\beta_j$ and applies when the *same* raters score *every* subject.
2. **Definition: agreement vs. consistency.** *Absolute agreement* counts rater
   bias $\sigma_\beta^2$ as error—two raters who are perfectly correlated but
   offset by ten points do not *agree*. *Consistency* excludes $\sigma_\beta^2$,
   asking only whether raters rank-order subjects alike. Use agreement when the
   rating's absolute level is interpreted; consistency when only relative
   standing matters.
3. **Unit: single vs. average.** Report reliability of a *single* rater's score
   when one coder will rate future items, or of the *average* of $k$ raters when
   the mean of the panel is the operative measurement. Averaging suppresses error,
   so the average-measure ICC is mechanically higher (the Spearman–Brown
   relationship).

The two-way, absolute-agreement, single-measure ICC makes these choices concrete:
$$
\text{ICC}(\text{agreement}, 1) =
\frac{\sigma_\alpha^2}
     {\sigma_\alpha^2 + \sigma_\beta^2 + \sigma_\varepsilon^2}.
$$ {#eq-icc-agreement}
The consistency variant simply drops $\sigma_\beta^2$ from the denominator,
mechanically raising the coefficient; the gap between the two quantifies how much
rater bias inflates apparent agreement. @fig-icc-tree maps the decisions to the
estimand.

```{mermaid}
%%| label: fig-icc-tree
%%| fig-cap: "Choosing an ICC. The three design decisions of @Shrout_1979—model, definition, and unit—jointly determine which variance ratio is estimated. Each choice changes the estimand, so it must be justified by the rating design, not chosen to maximize the coefficient."
flowchart TD
  A["Continuous ratings"] --> B{"Same raters score<br/>every subject?"}
  B -->|"No: raters vary<br/>by subject"| C["One-way model<br/>ICC(1,&middot;)"]
  B -->|"Yes: fixed panel"| D{"Is the absolute<br/>level interpreted?"}
  D -->|"Yes"| E["Two-way<br/>Absolute agreement"]
  D -->|"No: only ranking<br/>matters"| F["Two-way<br/>Consistency"]
  C --> G{"Score future items<br/>with one rater<br/>or a panel mean?"}
  E --> G
  F --> G
  G -->|"One rater"| H["Single-measure ICC"]
  G -->|"Panel mean of k"| I["Average-measure ICC<br/>(Spearman&ndash;Brown higher)"]
```

The `anxiety` data—three raters scoring the same twenty subjects—illustrate the
two-way, absolute-agreement, single-measure case.

```{r}
#| label: icc-anxiety
#| message: false
data("anxiety", package = "irr")
icc(anxiety,
    model = "twoway",      # same raters score all subjects
    type  = "agreement",   # absolute level matters; counts rater bias as error
    unit  = "single")      # reliability of one rater's score
```

`DescTools::ICC()` reports all six ICC forms of @Shrout_1979 at once, which is the
disciplined way to expose how sensitive the conclusion is to the design
assumptions: a reliability that looks "substantial" as a consistency,
average-measure coefficient can collapse as a single-rater, absolute-agreement
coefficient.

```{r}
#| label: icc-desctools
panel <- cbind(
  rater1 = c(9, 6, 8, 7, 10, 6),
  rater2 = c(2, 1, 4, 1,  5, 2),
  rater3 = c(5, 3, 6, 2,  6, 4),
  rater4 = c(8, 2, 8, 6,  9, 7)
)
DescTools::ICC(panel)
```

Here the raters track one another's *ordering* of subjects well but sit at very
different absolute levels (rater 1 is systematically generous, rater 2 severe).
Consistency ICCs are accordingly high while agreement ICCs are low—a textbook case
of why @eq-icc-agreement, not the consistency form, is the honest choice when the
score's level will be interpreted.

## Choosing and Reporting a Coefficient

The coefficients are not interchangeable, and selecting one post hoc to maximize
the reported number is a form of specification search. The choice is dictated by
the data's structure—measurement level, number of raters, presence of missing
values, and whether the rating's absolute level carries meaning—as mapped in
@tbl-irr-map and @fig-icc-tree. Krippendorff's $\alpha$ is the safest default for
categorical coding because it degrades gracefully to the others' use cases while
tolerating missing data and any measurement level; the kappa coefficients remain
standard and expected in many literatures; the ICC is mandatory for continuous
ratings.

Three reporting disciplines separate credible IRR from box-checking. First,
report the *coefficient with a confidence interval*, not a bare point estimate: a
$\kappa$ of 0.70 from 20 items is far weaker evidence than the same value from
500. Second, *state every assumption that defines the estimand*—the measurement
level and difference metric for $\alpha$, the model/definition/unit triple for the
ICC—because, as the examples show, the same data yield materially different
coefficients under different assumptions. Third, remember that reliability bounds
but does not establish validity: a perfectly reliable code can still measure the
wrong construct, and the validity argument belongs to @sec-measurement-scales.
High agreement earns the right to interpret the codes; it does not, by itself,
make the interpretation correct.

## Key Takeaways

- Inter-rater reliability is the methodological hinge of qualitative coding and of
  every human-labeled "ground truth" used to validate text- and image-mining
  models (@sec-measurement-scales).
- Raw percent agreement is uninterpretable because it credits agreement expected
  from category base rates; all defensible coefficients chance-correct via
  @eq-chance-correction.
- The kappa family (@eq-cohen-kappa, @eq-weighted-kappa, @eq-fleiss-kappa) handles
  nominal and ordinal categories; weighting awards partial credit for near misses
  on ordered scales [@Cohen_1960; @Landis_1977].
- Krippendorff's $\alpha$ (@eq-krippendorff) is the general coefficient—any number
  of raters, missing data, any measurement level—and its thresholds gatekeep
  publishable coding [@Shelley_1984]; Kendall's $W$ (@eq-kendall-w) handles
  rankings.
- For continuous ratings, the ICC (@eq-icc-model, @eq-icc-agreement) embeds
  reliability in a variance-components model whose model/definition/unit choices
  change the estimand and must be justified by the design, not the result
  [@Shrout_1979].
- Report coefficients with confidence intervals and explicit assumptions, and
  never mistake reliability for validity.
