# Images as Data {#sec-image-processing}

Marketing has always been a visual discipline. A logo, a package, an
advertisement, an influencer's selfie, a product photograph on a retail site—each
is a deliberately constructed image meant to move attention, belief, and demand.
Until recently those images were data only in a metaphorical sense: researchers
*looked* at them, coded them by hand, and ran small experiments on a handful of
manipulated stimuli. What has changed is that an image can now be turned into
*numbers*—high-dimensional, machine-readable features—at the scale of millions of
photographs, and those numbers can be entered into the same demand systems, choice
models, and regressions that the rest of this book develops. This chapter is about
how to do that responsibly: how to extract brand, aesthetic, and content features
from images, what the deep-learning machinery underneath actually computes, and
how to deploy the resulting features in advertising and social-media research
without fooling oneself.

The intellectual move mirrors the one made for text (@sec-text-as-data). There,
unstructured language becomes a document–term matrix or a sequence of embeddings;
here, an unstructured raster of pixels becomes a feature vector. In both cases the
representation is *lossy* and *learned*, and in both cases the central empirical
danger is the same: the features are generated by a model whose errors are
correlated with the very outcomes the researcher wants to explain, so naive
regression confounds measurement with effect. The payoff for getting it right is
large. Image features let the analyst measure constructs that were previously
locked inside qualitative judgment—a brand's visual identity, an ad's aesthetic
appeal, the "warmth" of a product photo—and relate them to clicks, engagement,
sales, and firm value.

The chapter proceeds from pixels to constructs to applications. It first fixes
what an image *is* as a mathematical object and what we mean by an image feature.
It then develops the workhorse of modern computer vision—the convolutional neural
network—giving its estimator, its loss, and the assumptions under which a feature
extracted from it is meaningful. With that in hand it turns to the three families
of marketing-relevant features (brand, aesthetic, content) and to the
econometrics of using generated image features as regressors. It closes with
applications to advertising and social media, and with the identification
pitfalls that separate a credible image-as-data study from a decorative one.

## What an Image Is

Formally, a digital image is a function sampled on a grid. A color image of height
$H$ and width $W$ is a third-order tensor

$$
\mathbf{X} \in [0,1]^{H \times W \times C}, \qquad C = 3,
$$ {#eq-image-tensor}

where the three channels $C$ hold red, green, and blue intensities and each entry
$\mathbf{X}_{ijc}$ is a normalized pixel value. A modest $224 \times 224$ RGB
image—the canonical input size for many vision models—already lives in a space of
dimension $224 \times 224 \times 3 \approx 1.5 \times 10^{5}$. This is the
**curse of dimensionality** in its rawest form: the number of pixels vastly
exceeds the number of labeled examples in any marketing dataset, and pixels are
individually almost meaningless. The pixel at position $(112, 60)$ carries no
stable interpretation; what matters is the *arrangement* of pixels into edges,
textures, objects, and scenes.

Two properties of images dictate everything about how they are modeled. First,
**locality**: meaningful structure (an edge, a corner) is built from nearby
pixels, so a useful representation should aggregate local neighborhoods before
global ones. Second, **translation structure**: a logo is the same logo whether it
sits in the top-left or the center of the frame, so a useful representation should
respond similarly to a pattern regardless of where it appears. A model that
ignored these properties—say, a fully connected network treating each pixel as an
unrelated input—would have to relearn "what an edge looks like" separately at every
location and would need astronomically more data to do so. The architectures that
dominate computer vision are precisely those that build locality and translation
structure in by construction.

::: {.callout-note}
An **image feature** is any function $f: [0,1]^{H \times W \times C} \to
\mathbb{R}^{d}$ that maps a raw image to a lower-dimensional vector intended to
capture a construct of interest. Features range from hand-engineered and
interpretable (mean saturation, number of detected faces, fraction of the frame
occupied by a brand logo) to learned and opaque (the 2{,}048-dimensional
penultimate-layer activations of a deep network). The art of images-as-data is
choosing features whose dimensions a marketing theory can speak about.
:::

## Classical Features: Color, Composition, and Hand-Engineering

Before deep learning, vision in marketing relied on **hand-engineered features**:
quantities a researcher computes with an explicit formula and can defend to a
referee line by line. They remain valuable precisely because they are transparent,
and several map directly onto long-standing aesthetic theory.

Color is the most tractable. RGB is poor for human-meaningful description because
its axes (red, green, blue intensity) do not correspond to how people talk about
color, so analysts transform to the **HSV** space (hue, saturation, value), in
which *hue* is the dominant wavelength, *saturation* the colorfulness, and *value*
the brightness. From an image one can compute the mean and dispersion of each
channel, the share of warm versus cool hues, and the **colorfulness** index, and
relate these to response. This connects to a classical account of aesthetic
preference: @Berlyne_1960 argued that hedonic value is an inverted-U function of
*arousal potential*—stimuli that are too simple bore and too complex overwhelm,
with moderate complexity, novelty, and contrast preferred. Color statistics,
visual entropy, and edge density are all operationalizations of arousal potential,
and the inverted-U is a recurring empirical shape in this literature.

Composition features quantify *where* content sits and *how much* of it there is.
**Visual complexity**—the amount and variety of detail in an image—has a long
pedigree in advertising research, where @pieters2010 distinguish *feature
complexity* (irregular, dense visual elements) from *design complexity* (the
elaborateness of the deliberate arrangement) and show the two have opposite
effects on attention and attitude: feature complexity hurts brand attention while
design complexity helps it. Earlier, @pieters2007 established that the
**eye-trackable** structure of an ad—how gaze is distributed across the brand,
the pictorial, and the text elements—predicts memory. Visual complexity is
commonly proxied by file size after compression, by edge density from a Sobel or
Canny filter, or by the entropy of the color histogram. The point of cataloguing
these is not nostalgia: hand-engineered features are still the right tool when the
construct is simple, the sample is small, or interpretability is paramount, and
they make excellent *controls* alongside learned features.

A compact illustration computes interpretable color and complexity features for a
synthetic image and shows how they vary with content.

```{r image-handcrafted-features, message=FALSE, warning=FALSE}
set.seed(34)

# Build three synthetic 64x64 RGB images with known properties:
# (a) a calm, low-saturation gray scene; (b) a vivid warm scene;
# (c) a high-complexity noisy scene. Each is an array [H, W, C] in [0,1].
make_image <- function(base, noise_sd) {
  arr <- array(base, dim = c(64, 64, 3))
  arr <- arr + array(rnorm(64 * 64 * 3, 0, noise_sd), dim = dim(arr))
  pmin(pmax(arr, 0), 1)
}
img_calm  <- make_image(c(0.55, 0.55, 0.55), 0.02)  # near-gray, low noise
img_warm  <- make_image(c(0.85, 0.35, 0.15), 0.04)  # warm (red/orange)
img_busy  <- make_image(c(0.50, 0.50, 0.50), 0.25)  # high-variance "busy"

# RGB -> HSV saturation and a colorfulness proxy (Hasler-Susstrunk style).
sat_value <- function(im) {
  mx <- pmax(im[, , 1], im[, , 2], im[, , 3])
  mn <- pmin(im[, , 1], im[, , 2], im[, , 3])
  mean(ifelse(mx > 0, (mx - mn) / mx, 0))            # mean saturation
}
colorfulness <- function(im) {
  rg <- im[, , 1] - im[, , 2]
  yb <- 0.5 * (im[, , 1] + im[, , 2]) - im[, , 3]
  sqrt(sd(rg)^2 + sd(yb)^2) + 0.3 * sqrt(mean(rg)^2 + mean(yb)^2)
}
# Edge density as a translation-invariant complexity proxy: mean absolute
# horizontal+vertical gradient of the luminance channel.
edge_density <- function(im) {
  lum <- 0.299 * im[, , 1] + 0.587 * im[, , 2] + 0.114 * im[, , 3]
  gx <- abs(lum[-1, ] - lum[-nrow(lum), ])
  gy <- abs(lum[, -1] - lum[, -ncol(lum)])
  mean(gx) + mean(gy)
}

features <- data.frame(
  image        = c("calm", "warm", "busy"),
  saturation   = c(sat_value(img_calm),  sat_value(img_warm),  sat_value(img_busy)),
  colorfulness = c(colorfulness(img_calm), colorfulness(img_warm), colorfulness(img_busy)),
  edge_density = c(edge_density(img_calm), edge_density(img_busy), edge_density(img_busy))
)
features[, -1] <- round(features[, -1], 3)
knitr::kable(features, caption = "Interpretable color and complexity features for three synthetic images.")
```

The warm image scores high on saturation and colorfulness; the busy image scores
high on edge density—the digital analogue of @Berlyne_1960's arousal potential.
These features are cheap, reproducible, and interpretable, but they are blind to
*meaning*: they cannot tell a logo from a face. For meaning, the field turns to
learned representations.

## Deep Representations: The Convolutional Neural Network

The central tool of modern computer vision is the **convolutional neural network**
(CNN). Its design is a direct response to the two properties of images named
above. Rather than connect every pixel to every hidden unit, a CNN slides small
learnable filters across the image, so each unit sees only a local neighborhood
(locality) and the *same* filter is applied at every position (translation
structure, and an enormous reduction in parameters).

### The Convolution Operation

A **convolutional layer** applies a bank of small filters (kernels) to its input.
Let the input be a feature map $\mathbf{Z} \in \mathbb{R}^{H \times W \times
C_{\text{in}}}$ and let a single filter be $\mathbf{K} \in \mathbb{R}^{k \times k
\times C_{\text{in}}}$ with bias $b$. The output at spatial location $(i,j)$ is

$$
(\mathbf{Z} * \mathbf{K})_{ij}
= \sigma\!\left(
b + \sum_{u=1}^{k}\sum_{v=1}^{k}\sum_{c=1}^{C_{\text{in}}}
\mathbf{K}_{uvc}\,\mathbf{Z}_{\,i+u,\,j+v,\,c}
\right),
$$ {#eq-convolution}

where $\sigma(\cdot)$ is a nonlinearity, almost always the **rectified linear
unit** $\sigma(z) = \max(0, z)$. A layer holds $C_{\text{out}}$ such filters,
producing an output tensor with $C_{\text{out}}$ channels. Three properties make
@eq-convolution the right primitive. **Parameter sharing**: one filter, with
$k^2 C_{\text{in}} + 1$ parameters, is reused at every location, so a layer learns
"detect this pattern anywhere" rather than memorizing locations. **Local
connectivity**: each output depends only on a $k \times k$ window, encoding
locality. **Translation equivariance**: shifting the input shifts the output
identically, so a logo detector fires wherever the logo appears.

Convolutional layers are interleaved with **pooling** layers that downsample—most
commonly *max pooling*, which reports the maximum activation in each small
window—building tolerance to small shifts and shrinking the spatial resolution
while the channel dimension grows. Stacking these operations yields a
representational hierarchy that has been verified empirically: early layers
respond to oriented edges and color blobs, middle layers to textures and motifs,
and late layers to object parts and whole objects [@zeiler2014visualizing]. @fig-cnn-pipeline
sketches the pipeline from raw pixels to a task head.

```{mermaid}
%%| label: fig-cnn-pipeline
%%| fig-cap: "A convolutional network as a feature extractor with a task-specific head. Early layers encode generic low-level structure; deep layers encode semantic content. Marketing applications usually freeze the backbone and read off the penultimate-layer embedding."
flowchart LR
  A["Raw image<br/>H × W × 3"] --> B["Conv + ReLU<br/>(edges, color)"]
  B --> C["Pool<br/>(downsample)"]
  C --> D["Conv + ReLU<br/>(textures, motifs)"]
  D --> E["Conv + ReLU<br/>(object parts)"]
  E --> F["Embedding z ∈ ℝ^d<br/>(penultimate layer)"]
  F --> G["Task head"]
  G --> H1["Classification<br/>(brand present?)"]
  G --> H2["Detection<br/>(where is the logo?)"]
  G --> H3["Regression<br/>(aesthetic score)"]
```

### The Estimator and Its Loss

A CNN with parameters $\boldsymbol{\theta}$ (all filter weights and biases) defines
a map $g_{\boldsymbol{\theta}}: \mathbf{X} \mapsto \hat{\mathbf{y}}$. For a
classification task with $L$ labels (e.g., "contains a car," "contains a dog"), the
final layer produces a probability vector via the **softmax**,
$\hat{p}_\ell = \exp(s_\ell)/\sum_{m}\exp(s_m)$, where $s_\ell$ are the network's
output scores (logits). Given labeled training data $\{(\mathbf{X}_n,
\mathbf{y}_n)\}_{n=1}^{N}$, the parameters minimize the **cross-entropy** loss with
weight penalty,

$$
\hat{\boldsymbol{\theta}}
= \arg\min_{\boldsymbol{\theta}}\;
-\frac{1}{N}\sum_{n=1}^{N}\sum_{\ell=1}^{L}
y_{n\ell}\,\log \hat{p}_{n\ell}(\boldsymbol{\theta})
\;+\; \lambda \lVert \boldsymbol{\theta} \rVert_2^2 ,
$$ {#eq-crossentropy}

solved by **stochastic gradient descent**: gradients are computed on
mini-batches by backpropagation and the parameters stepped against them. The
penalty $\lambda \lVert \boldsymbol{\theta}\rVert_2^2$ (weight decay) is one of
several regularizers—dropout, data augmentation, and early stopping are
others—that the over-parameterized regime makes essential. This is the same
empirical-risk-minimization template developed for prediction in @sec-ai-ml; what
is special here is only the architecture of $g_{\boldsymbol{\theta}}$.

::: {.callout-warning}
A CNN trained by @eq-crossentropy minimizes *predictive* loss, not the recovery of
any causal or structural quantity. Its outputs are calibrated to the training
distribution and labels, nothing more. Treating a predicted label or an embedding
as if it were a ground-truth measurement—rather than an estimate with
distribution-dependent error—is the original sin of images-as-data, and
@sec-image-econometrics shows what it costs.
:::

### Transfer Learning: Why Marketing Rarely Trains From Scratch

No marketing dataset is large enough to estimate the tens of millions of
parameters in a modern CNN from scratch. The field instead relies on **transfer
learning**: take a backbone network pre-trained on a massive general-purpose
corpus (canonically ImageNet, with roughly a million labeled images across a
thousand object categories), discard its task head, and reuse its learned
representation. The canonical backbones trace the field's progress—the deep CNN that
launched it [@krizhevsky2017imagenet], the deeper inception architecture
[@szegedy2015googlenet], and the residual networks [@he2016resnet] that remain a default
image encoder—and any of them can be downloaded pre-trained and reused. Two modes are common. In **feature extraction**, the backbone is
*frozen* and the penultimate-layer activations $\mathbf{z} =
h_{\boldsymbol{\theta}}(\mathbf{X}) \in \mathbb{R}^{d}$—the **embedding**—are
treated as a fixed, off-the-shelf feature vector fed to a simple downstream model
(logistic regression, gradient boosting) trained on the marketing labels. In
**fine-tuning**, the backbone weights are *unfrozen* and updated, usually with a
small learning rate, so the representation adapts to the target domain.

The justification is the hierarchy of @fig-cnn-pipeline: early- and middle-layer
features (edges, textures, parts) are nearly universal across natural images, so
they transfer; only the late, task-specific layers must be relearned. The
practical rule is to *fine-tune more layers the larger and more domain-specific
the target data, and freeze more the smaller and more generic*. The same logic
underlies the brand-image work of @liu2020, who train a multi-label
convolutional network on consumer-created images to recover perceptual brand
attributes (see @sec-image-brand). For most marketing studies—where the labeled
sample numbers in the thousands, not millions—frozen-backbone feature extraction
is both the safest and the most reproducible choice.

```{r transfer-embedding, message=FALSE, warning=FALSE}
set.seed(34)

# Simulate the *output* of a frozen CNN backbone: each image is represented by a
# d-dimensional embedding z. In practice z = h_theta(image) from a pretrained
# network; here we generate embeddings whose geometry encodes two latent
# "visual styles" so a downstream classifier can separate them.
d <- 16; n_per <- 150
style_A_mean <- rnorm(d, 0, 1)            # e.g., "minimalist product shot"
style_B_mean <- style_A_mean + rnorm(d, 0, 1.2)  # e.g., "lifestyle scene"

Z_A <- matrix(rnorm(n_per * d), n_per, d) + matrix(style_A_mean, n_per, d, byrow = TRUE)
Z_B <- matrix(rnorm(n_per * d), n_per, d) + matrix(style_B_mean, n_per, d, byrow = TRUE)
Z   <- rbind(Z_A, Z_B)
y   <- factor(rep(c("minimalist", "lifestyle"), each = n_per))

# Downstream task: a simple logistic head on the frozen embedding (no CNN
# retraining). This is "feature extraction" transfer learning in miniature.
df  <- data.frame(y = y, Z)
idx <- sample(nrow(df), 0.7 * nrow(df))
fit <- glm(y ~ ., data = df[idx, ], family = binomial)
pred <- ifelse(predict(fit, df[-idx, ], type = "response") > 0.5,
               "minimalist", "lifestyle")
acc <- mean(pred == df[-idx, "y"])
cat("Held-out accuracy of logistic head on frozen embedding:",
    round(acc, 3), "\n")
```

The example deliberately separates the expensive, transferable part (the
embedding, treated as given) from the cheap, task-specific part (a logistic head
the analyst actually estimates)—the division of labor that makes images-as-data
feasible in marketing.

## Three Families of Marketing Image Features

Marketing-relevant image features fall into three families—**brand**,
**aesthetic**, and **content**—distinguished by the construct they measure and the
machinery they require. @tbl-feature-families summarizes them; the subsections that
follow develop each in turn.

```{r}
#| label: tbl-feature-families
#| tbl-cap: "Three families of marketing image features, the construct each targets, and how it is measured."
#| message: false
#| warning: false
#| echo: false
ff <- data.frame(
  Family    = c("Brand", "Aesthetic", "Content"),
  Construct = c("Presence/prominence of brand marks; visual brand identity",
                "Beauty, professionalism, arousal potential, style",
                "Objects, scenes, faces, emotions, activities depicted"),
  Method    = c("Logo detection; learned brand-attribute classifiers",
                "Aesthetic regressors; hand-engineered color/complexity",
                "Object detection; scene & face recognition"),
  Output    = c("Logo box, area share, brand-attribute scores",
                "Continuous aesthetic / arousal score",
                "Object labels & counts, scene class, facial affect")
)
knitr::kable(ff)
```

### Brand Features {#sec-image-brand}

Brand features answer two questions: *is the brand here, and how prominently?* The
first is **logo detection**—an object-detection task that returns bounding boxes
and confidence scores for known marks—and it underwrites a quantity of direct
managerial interest: the **share of visual voice**, the fraction of image area (or
of impressions) a brand's marks occupy across a corpus of user- or
sponsor-generated images. This matters because brand exposure on social media is
increasingly *incidental*, embedded in content the firm does not control, so
counting hashtags understates true exposure while logo detection captures it. The
notion of **prominence**—the conspicuousness of a brand's mark—carries
status-signaling consequences developed in @sec-branding, and image data let it be
measured at scale rather than coded by hand. The *design* of the mark itself is
consequential: @luffarelli2019logo show that descriptive logos (those that visually
signal what the brand does) raise brand equity, and @mahmood2019logo find that complex
visual logo cues shape equity-crowdfunding outcomes---logo features that computer vision
can now extract at scale. How consumers depict brands in their own photos is itself a
measurable typology: @hartmann2021brandselfies use computer vision to distinguish
"brand selfies" (the consumer in frame with the product) from "packshots" and show the
two drive engagement differently.

The second, deeper question is whether an image conveys a brand's *perceptual
identity*. @liu2020 answer it directly: their BrandImageNet trains a multi-label
CNN to detect abstract perceptual attributes (e.g., "glamorous," "rugged,"
"fun") in ordinary consumer-created images, and the machine-derived portrait of a
brand's image tracks survey-based perceptions in near real time. Complementary approaches
elicit brand perception by having consumers select images rather than rate scales
[@dzyabura2021visual] and mine brand attributes from the social-network structure around a
brand [@culotta2016]. This is a
conceptual advance over logo counting—it measures what a brand *means* visually,
not merely whether it is shown—and it links image data to the brand-meaning
construct of @batra2019, who locate visual cues (advertising, packaging) among the
primary sources from which brand associations are built. Crucially, such attribute
classifiers are learned, so their outputs are estimates with brand- and
context-dependent error; this is exactly the generated-regressor problem
formalized below.

### Aesthetic Features

Aesthetic features score an image on beauty, professionalism, or arousal potential.
Two routes exist. The **hand-engineered** route assembles the color and complexity
statistics of the previous section into an interpretable index grounded in the
@Berlyne_1960 inverted-U. The **learned** route trains a CNN to regress human
aesthetic ratings onto images, yielding a single continuous "aesthetic score" per
image. Each route has its place: learned scores predict human judgment far better
but are opaque and may encode spurious correlates of the training raters'
tastes; hand-engineered indices are weaker predictors but transparent and
manipulable, which matters when the goal is *advice* ("raise saturation") rather
than mere ranking.

Aesthetics is not decoration. In advertising, the visual *pleasure* of an ad is a
proximate driver of attention and attitude, mediated by the complexity mechanism of
@pieters2010 and by the gaze patterns of @pieters2007, and the rhetorical *figuration*
of an image—visual metaphor, pun, and other tropes—shapes persuasion in ways parallel to
verbal rhetoric [@mcquarrie1999visualrhetoric]. Specific perceptual choices carry meaning:
color hue and saturation move brand personality and downstream response [@labrecque2012red;
@labrecque2013palette], and aesthetic styling trades off against perceived functionality
rather than improving evaluation monotonically [@hagtvedt2014overstyling]. In social
commerce, image quality is a documented driver of demand: professionally composed product
and listing photos lift engagement and conversion, and @zhang2022goodimage show, using
interpretable deep-learning image features on Airbnb listings, *which* properties of a
photo (composition, brightness, depth) actually raise demand. The
aesthetic score is therefore a feature with a theory behind it, not a black box to
be regressed on outcomes for its own sake.

### Content Features

Content features identify *what is depicted*. **Object detection** returns the
set, count, and location of recognizable objects (a person, a beverage, a beach);
**scene recognition** classifies the overall setting (kitchen, gym, nightclub);
**face analysis** locates faces and, more controversially, infers attributes such
as apparent emotion, age, or gaze direction. Content features connect images to
substantive marketing theory: the presence of *people* versus *products* in a post,
the warmth of depicted facial expressions, and the activities shown all map onto
constructs—self-presentation, social proof, lifestyle congruence—that the
social-media and influencer literatures (@sec-influencer-marketing) treat as
drivers of engagement.

::: {.callout-warning}
Face- and demographic-inference features are the most fraught in this chapter.
Apparent-emotion and inferred-demographic classifiers are trained on labeled data
that encode the annotators' cultural assumptions; their errors are systematically
correlated with skin tone, age, and gender, raising both validity and fairness
concerns. The privacy implications of inferring protected attributes from images
are governed by the regimes in @sec-privacy. Researchers should treat such
features as low-confidence proxies, audit error rates by subgroup, and prefer
coarse, well-validated labels (face present / absent) over fine-grained inferences
(specific emotion, exact age) wherever the theory permits.
:::

## The Econometrics of Generated Image Features {#sec-image-econometrics}

The defining methodological problem of images-as-data is that the features are
**generated regressors**: they are not observed but *predicted* by a first-stage
model (the CNN), and that prediction carries error. Suppose the structural object
of interest is

$$
y_i = \alpha + \beta\, f_i^{\ast} + \mathbf{w}_i^{\top}\boldsymbol{\gamma}
+ \varepsilon_i,
$$ {#eq-structural-image}

where $y_i$ is an outcome (clicks, likes, sales), $f_i^{\ast}$ is the *true* image
feature (true aesthetic appeal, true logo prominence), and $\mathbf{w}_i$ are
controls. The analyst does not observe $f_i^{\ast}$; the CNN supplies an estimate
$\hat f_i = f_i^{\ast} + u_i$. Regressing $y_i$ on $\hat f_i$ instead of
$f_i^{\ast}$ raises three distinct hazards.

**Attenuation from classical measurement error.** If $u_i$ is mean-zero noise
independent of $f_i^{\ast}$ and $\varepsilon_i$, the ordinary-least-squares
estimand is biased toward zero by the reliability ratio,

$$
\operatorname{plim}\hat\beta_{\text{OLS}}
= \beta \cdot \frac{\operatorname{Var}(f^{\ast})}
{\operatorname{Var}(f^{\ast}) + \operatorname{Var}(u)} \;<\; \beta .
$$ {#eq-34-attenuation}

The noisier the classifier (the smaller its reliability), the more the true effect
is understated. A nonsignificant coefficient on an image feature can therefore mean
"no effect" *or* "a noisy detector," and the two are not distinguishable without an
estimate of $\operatorname{Var}(u)$.

**Non-classical, correlated error.** Worse, CNN error is rarely classical. A logo
detector may systematically miss small or occluded logos; an aesthetic regressor
may rate professionally lit photos higher *and* such photos may be posted by
firms that also buy promotion. When $\operatorname{Cov}(u_i, \varepsilon_i) \neq
0$—because the same unobserved factor (a firm's production budget) drives both the
classifier's error and the outcome—$\hat\beta$ is biased in an *unknown* direction
and no reliability correction recovers $\beta$. This is the image analogue of the
endogeneity problems treated throughout @sec-causal-inference, and it is the
reason image features must be validated against, and where possible instrumented or
held fixed within an experiment.

**Forbidden-regression and look-elsewhere risks.** Because embeddings are
high-dimensional, an analyst can search across hundreds of learned dimensions for
one that correlates with the outcome and report it as a "discovered visual driver."
Without pre-registration or a held-out test, this is overfitting dressed as
inference. The discipline that text-as-data imposes—split the sample, fix the
representation before looking at outcomes, report out-of-sample fit—applies with
full force here.

Three remedies recur. First, **validate the feature** against human labels on a
held-out subsample and report the classifier's accuracy or correlation with
ground truth, so the reader can gauge $\operatorname{Var}(u)$. Second,
**experimentally manipulate** the feature rather than only observe it: the cleanest
images-as-data designs randomize which image a user sees (an A/B test on ad
creative) so that $f_i$ is assigned, not estimated from confounded content, which
severs $\operatorname{Cov}(u_i, \varepsilon_i)$ by design. Third, **correct or
bound** the bias—correcting attenuation with an estimated reliability ratio when
error is plausibly classical, and otherwise reporting the OLS estimate as a
conservative (attenuated) bound on a positive effect. The simulation below makes
the attenuation in @eq-34-attenuation concrete and shows that a reliability
correction recovers the true slope when error is classical.

```{r generated-regressor-attenuation, message=FALSE, warning=FALSE}
set.seed(34)
n <- 4000
beta_true <- 1.5

# True (latent) image feature f* and outcome y generated from it.
f_star <- rnorm(n, 0, 1)
y <- 2 + beta_true * f_star + rnorm(n, 0, 1)

# A noisy CNN measures f* with classical error: reliability rho = Var(f*)/Var(fhat).
sigma_u <- 0.9
f_hat <- f_star + rnorm(n, 0, sigma_u)
rho   <- var(f_star) / var(f_hat)                # estimable from a validation set

naive  <- coef(lm(y ~ f_hat))["f_hat"]           # attenuated OLS
corrected <- naive / rho                         # reliability-ratio correction

cat("True slope beta          :", beta_true, "\n")
cat("Reliability ratio (rho)  :", round(rho, 3), "\n")
cat("Naive OLS on f_hat       :", round(naive, 3),
    " (attenuated toward 0)\n")
cat("Reliability-corrected    :", round(corrected, 3), "\n")
```

The naive slope is pulled toward zero by exactly the reliability ratio; dividing by
$\rho$—which a validation set identifies—restores the truth. The correction works
*only* under classical error; when the detector's mistakes correlate with the
outcome, no such fix exists and design-based identification is the only honest
route.

## Applications in Advertising and Social Media

### Advertising Creative

The richest application is the measurement and optimization of advertising
**creative**. The visual content of an ad—its complexity, color, brand prominence,
and pacing—drives whether viewers attend, remember, and respond, and image (and
video-frame) features make these properties measurable at the scale of entire ad
libraries. The construct lineage is clear: visual complexity and gaze
(@pieters2007; @pieters2010), the arousal-potential account of aesthetic
preference (@Berlyne_1960), and the visual sources of brand meaning (@batra2019)
all become *features* once a CNN is in the loop. Video advertising extends this
from stills to sequences. The dynamics of attention within a video ad—when a brand
appears, how scenes are cut, what holds the viewer—predict skipping and recall;
@teixeira2010 show that inducing *joy* and *surprise* sustains attention to
online video ads, and @teixeira2014 that emotional and attentional dynamics jointly
govern whether viewers watch through or zap. Per-frame image features are precisely
how those dynamics are operationalized at scale, a theme picked up in the
video-marketing literature (@rajaram2020video; @leyva2021video) and developed
further in @sec-advertising. The methodological caveat from @sec-image-econometrics
binds hardest here: observational creative comparisons confound the image with the
campaign that bought it, so the credible designs hold the audience fixed and
randomize the creative.

### Social Media and User-Generated Images

On social platforms, the image *is* the post. Engagement, reach, and
brand-relevant outcomes depend on visual content the firm often does not produce,
which makes image features the only way to measure exposure and appeal across
user-generated corpora. Three measurement problems recur. **Incidental brand
exposure** is captured by logo detection and the share-of-visual-voice it yields,
revealing brand presence that text-only listening (@sec-online-environments)
misses entirely. **Image-driven engagement** is studied by relating content and
aesthetic features—people versus products, warmth, professionalism, complexity—to
likes, shares, and comments; @Li_2019 show empirically which image-content features
raise social-media engagement, and @hartmann2021power and @hartmann2023more
develop image-mining pipelines for brand-relevant social-media content
and relate visual elements to engagement. @majumdar2026image push this strand from
description to *prescription*: using deep learning to separate a post's **primary**
features (its subject) from the **secondary** ones a designer can add at an extraction
cost (background detail, visual complexity), they estimate a nonlinear, platform-specific
map from features to engagement and embed it in an optimization framework that
allocates design effort across features and platforms—beating rules of thumb that
budget by a platform's audience size alone or ignore the controllable secondary
features entirely. **Influencer aesthetics** connect a
creator's visual style to follower growth and sponsorship value
(@sec-influencer-marketing), with content features quantifying the self-presentation
strategies that drive parasocial response. A recurring, sobering finding across
this work is that *what is depicted* (content) and *how it looks* (aesthetics) carry
largely separable effects, so a study that conflates them—loading both onto a single
"image quality" score—misattributes one to the other.

::: {.callout-note}
The text-and-image fusion frontier. Marketing stimuli are rarely image-only: an
ad, a post, or a listing pairs a picture with words. **Multimodal** models learn a
joint embedding in which images and the text that co-occurs with them are mapped to
the same space, so a post can be represented by a single vector capturing both. This
lets the analyst ask whether the image and caption are *congruent*—and incongruence
is itself a measurable, theory-relevant feature—and connects images-as-data to the
text methods of @sec-text-as-data and the broader machine-learning toolkit of
@sec-ai-ml. @witte2026language show that transformer-based vision-language models can
classify marketing images directly from natural-language prompts, often without
task-specific training, and the same model class can now *generate* marketing visuals,
raising the question of whether it reaches human-level creative quality
[@hartmann2025generative]. The commercial stakes are clearest where the listing *is*
the market: @sikdar2026artwork show that for novice artwork on Etsy, the photographs
and the written description carry price-relevant signal beyond the structured
attributes a seller fills into a form, which is what makes fusion worth the trouble
rather than merely fashionable (@sec-mmf-artpricing). The fusion of image with text and
other modalities is the subject of @sec-multimodal-fusion.
:::

## Pitfalls and Identification

Five recurring failures separate credible image-as-data work from decorative work.

The first is **treating predictions as measurements**, ignoring the generated-
regressor problem of @sec-image-econometrics. Always report classifier validation
and, where error may be non-classical, prefer experimental assignment of the image.

The second is **dataset shift**. A backbone pre-trained on ImageNet encodes the
visual statistics of web photos circa its training era; applied to a domain it never
saw—X-ray packaging, niche product categories, a new platform's aesthetic—its
features may be uninformative or biased. Validate on in-domain data, and fine-tune
when the target distribution diverges.

The third is **spurious cues and shortcut learning**. CNNs latch onto whatever
predicts the label in training, including artifacts—a watermark, a background, a
camera type—that correlate with the outcome by accident. A classifier that appears
to "detect luxury" may have learned to detect studio lighting. Probe what the model
actually responds to before interpreting a feature substantively.

The fourth is **conflating content and aesthetics**, collapsing *what is shown* and
*how it looks* into one score when the social-media evidence says their effects are
separable. Keep the two families of features distinct and let the data assign their
effects.

The fifth is **ethics and privacy**, especially for facial and demographic
inference. Subgroup-correlated error and the inference of protected attributes raise
fairness and legal exposure (@sec-privacy); coarse, validated labels are preferable
to fine-grained inferences, and audits by subgroup are not optional.

Underlying all five is a single discipline borrowed from the rest of this book:
fix the representation before looking at outcomes, validate the features against
ground truth, prefer designs that randomize the image over those that merely
observe it, and report the bias that remains. An image feature is a *measurement*,
and like every measurement in @sec-measurement-scales it must be shown to be
reliable and valid before its coefficient can be believed. Images are one branch of
the broader unstructured-data program [@balducci2018unstructured]; for a survey
focused on the visual channel and its methods, see @dzyabura2021imageanalytics.

::: {.callout-tip}
## Replication resources: image analytics

The hand-engineered color/complexity features and the frozen-embedding transfer-learning
demonstration in this chapter run on open R/Python tooling (`magick`/`imager` in R; Pillow,
OpenCV, and pretrained `torchvision`/`timm` backbones in Python). The canonical backbones
ship reference implementations—ResNet at `github.com/KaimingHe/deep-residual-networks`
[@he2016resnet]; ImageNet-pretrained AlexNet/Inception weights [@krizhevsky2017imagenet;
@szegedy2015googlenet] are bundled in every deep-learning framework—and SIFT
[@lowe2004sift] is in OpenCV. The empirical marketing studies cited here
(e.g., @liu2020, @hartmann2021brandselfies, @zhang2022goodimage) generally rely on
proprietary image corpora; confirm any code/data release on the article page rather than
assuming one.
:::

## Key Takeaways

- An image is a high-dimensional tensor (@eq-image-tensor); useful image analysis
  replaces raw pixels with **learned or hand-engineered features** whose dimensions
  a marketing theory can interpret.
- The **convolutional network** (@eq-convolution) builds locality and translation
  structure in by construction; in marketing it is almost always used via
  **transfer learning**, with a frozen pre-trained backbone supplying an embedding
  to a small downstream model.
- Marketing image features divide into **brand** (logo and perceptual identity, as
  in @liu2020), **aesthetic** (beauty and arousal potential, grounded in
  @Berlyne_1960), and **content** (objects, scenes, faces) families.
- Image features are **generated regressors**: classical error attenuates effects
  by the reliability ratio (@eq-34-attenuation), and non-classical, outcome-correlated
  error biases them unpredictably. Validate, and prefer designs that randomize the
  image.
- The strongest applications—advertising creative and social-media engagement—pair
  image features with **experimental assignment** of the visual stimulus, severing
  the confound between what an image looks like and the campaign that produced it.
