# Industrial Organization {#sec-industrial-organization}

Industrial organization (IO) studies how firms behave in markets, how that behavior
maps into prices, product variety, and welfare, and how researchers can recover the
deep parameters—preferences, costs, conduct—that generate observed market outcomes.
For marketing, IO is the structural backbone: it supplies the demand systems that
price a product line, the equilibrium concepts that turn one firm's pricing problem
into an industry forecast, and the identification arguments that let an analyst run
*counterfactuals*—what would happen to prices, shares, and surplus if two firms
merged, if a new product entered, or if a regulator capped a fee. Where reduced-form
marketing analytics describe correlations, structural IO builds an economic model
whose estimated primitives remain invariant under the policy change being studied,
so that the model can be re-solved under the new regime.

This chapter develops the toolkit in the order a modeler actually uses it. It opens
with the methodological frame: the core questions of IO, the modeling tools
(demand, constrained profit maximization, game theory, comparative statics), and the
discipline of turning a question into a solvable model. It then builds **consumer
demand** from two directions. The *classical* approach treats a representative
consumer choosing continuous quantities, yielding the elegant Slutsky/Roy machinery,
the CES and quasi-linear special cases, diversion ratios, and the homothetic-surplus
family that makes multiproduct pricing tractable. The *discrete-choice* approach
treats heterogeneous consumers each picking one option from a menu, which is how
modern empirical IO and quantitative marketing estimate demand for differentiated
products. We give each model its formal definition, its estimator, its identifying
assumptions, and the failures that break identification—price endogeneity,
zero-valued shares, selection on unobservables—and we close with the frontier
applications in marketing: dynamic pricing of new products, behavioral price
discrimination in channels, email contact policy, and multi-sided platforms. Code is
runnable and seeded throughout. The companion structural-econometrics machinery
appears in @sec-marketing-finance and in the demand-estimation discussions elsewhere
in the book; here the emphasis is on the economic models themselves.

## The IO Research Program

Three questions organize the field. First, **why do firms behave as we observe**—why
this price, this product line, this bundle, this contract? Second, **what are the
effects** of that behavior on consumers, rivals, and employees, and against which
welfare standard should those effects be judged? Third, **how can we design better
strategy** for firms and better policy for regulators? The first question is
positive, the second normative, and the third prescriptive; a complete IO study
usually touches all three, because a counterfactual that improves firm profit and one
that improves consumer surplus are different optimizations over the same estimated
model.

Three methodological approaches answer these questions and are complementary rather
than competing. The **theoretical** approach builds a model and derives predictions;
the **empirical** approach estimates the model from data, tests its predictions, and
runs counterfactual and welfare analysis; the **experimental and behavioral**
approach manipulates conditions to isolate causal mechanisms and to discipline the
behavioral assumptions the other two impose. The structural empirical tradition that
dominates this chapter fuses the first two: it writes down a theoretical model of
demand and conduct, then estimates its primitives so the model can be re-solved under
hypothetical policies.

### Building an IO Model

A study begins with an *idea*—a question drawn from the literature in IO and adjacent
fields, or from the world (a news item, a consulting engagement, an unexplained
pattern in data). The idea is then expressed through a small set of **modeling
tools** that recur across nearly every IO paper, summarized in @fig-io-workflow.

```{mermaid}
%%| label: fig-io-workflow
%%| fig-cap: "The structural IO modeling workflow, from question to welfare statement. Each stage constrains the next: the demand system feeds the firm's first-order conditions, which define equilibrium, whose comparative statics answer the counterfactual."
flowchart TD
    A[Idea / Question<br/>literature or world] --> B[Model consumer demand]
    B --> C[Profit maximization<br/>subject to constraints]
    C --> D[Game-theoretic equilibrium<br/>among firms]
    D --> E[Comparative statics<br/>& counterfactuals]
    E --> F[Welfare & policy<br/>evaluation]
    F -.->|do results make<br/>intuitive sense?| B
```

The first tool is a **model of consumer demand** for the products or services in
question. The modeler decides whether consumers choose among single or multiple
products, single or multiple units, and whether products are differentiated—and if
so, whether they are substitutes or complements, and whether differentiation is
*vertical* (everyone agrees more is better, as with quality) or *horizontal*
(consumers disagree about the ideal variety, as with location or flavor). Demand
*complementarities* deserve special attention because they reshape strategy: when
products are complements, raising one price depresses demand for both, so a
monopolist who internalizes the cross-effect charges *lower* prices than two
independent firms would, and the complementarity can itself become a barrier to
entry. Network effects, multi-sided markets, and ecosystems are demand
complementarities operating across consumers or across platform sides rather than
within a single basket.

The second tool is **profit maximization subject to constraints**. The modeler
specifies choice variables and an objective, then layers on the constraints that make
the problem economically interesting: technology constraints from the cost side;
consumer *participation* constraints (the consumer must prefer buying to not buying),
which rationalize bundling and tied selling and let contracts act as entry barriers;
*arbitrage* constraints that limit price discrimination across consumers who can
resell; and *incentive* constraints in principal–agent settings, which generate
second-degree price discrimination (self-selection across a menu) and franchise
contracts. The third tool is **game theory**, treated next. The fourth is
**comparative static analysis**: with exogenous parameters and policy variables on one
side and endogenous equilibrium variables on the other, the modeler asks how the
optimum or equilibrium shifts as a parameter moves—the formal content of every
"what if" the study can answer.

### Game-Theoretic Models

Strategic interaction among firms is modeled non-cooperatively or cooperatively.
**Non-cooperative** models come in two forms. *Normal-form* (simultaneous-move) games
specify a set of players, a strategy set for each, and a payoff function for each, and
solve for Nash equilibrium in whatever strategic variable is relevant—price
(Bertrand), quantity (Cournot), capacity, advertising, or location. *Extensive-form*
(sequential-move) games capture first-mover advantages, entry deterrence, and
collusion sustained through repeated interaction, where the threat of future
punishment supports prices above the static equilibrium. **Cooperative** models, used
heavily in channel and platform settings, center on **Nash bargaining**: two parties
split the gains from agreement according to their bargaining power, relative to a
*threat point* (the disagreement payoff). The **Nash-in-Nash** refinement embeds
bilateral Nash bargains inside a larger non-cooperative game—each upstream–downstream
pair bargains taking all other negotiated prices as given—and has become the workhorse
for modeling negotiated wholesale prices in vertical markets.

> A profile of strategies is a **Nash equilibrium** if no player can raise its own
> payoff by unilaterally deviating, holding all rivals' strategies fixed. In a pricing
> game this means each firm's price is a best response to its rivals' prices, so the
> equilibrium price vector solves all firms' first-order conditions simultaneously.

The output of this machinery is a set of results, after which—per the discipline of
the field—the modeler pauses to ask whether the results make intuitive sense before
trusting them. A written IO study then motivates the question (with examples,
statistics, prior work, and policy stakes), states the approach and headline
findings, constructs the model, derives the equilibrium and its comparative statics,
and closes with limitations and extensions.

## Consumer Demand {#sec-io-demand}

The central modeling primitive is a **demand system**: a description of how the
quantities consumers buy respond to firms' offers—prices, qualities, versions,
advertising, warranties. Three questions drive the demand modeler. How should demand
be formally specified given the product and the market? What properties must a demand
*system* (more than one product) satisfy to be internally consistent? And how can the
system be estimated so that own-price elasticities, cross-price elasticities, and
diversion ratios can be computed and fed into a pricing or merger analysis?

Demand is fundamentally a **matching process** between a population of consumers—each
with observable and unobservable characteristics—and a set of products with their own
observable and unobservable characteristics. The microeconomics and IO literatures
offer three routes through this matching problem: *classical consumer theory* with
continuous choices; *discrete-choice models* with heterogeneous tastes, which
accommodate horizontal differentiation, vertical differentiation, or both; and *hybrid
discrete–continuous* models in which the consumer first chooses which product and then
how much. The next two parts develop the classical and discrete-choice routes in
turn.

## Classical Consumer Demand {#sec-io-classical}

A representative consumer with utility $U(q_0, q_1, \dots, q_n)$ and income $I$, facing
prices $\mathbf{p} = (p_1, \dots, p_n)$, chooses continuous quantities
$\mathbf{q} = (q_1, \dots, q_n)$ to solve

$$
\max_{q_0, \dots, q_n} \; U(q_0, q_1, \dots, q_n)
\quad \text{s.t.} \quad q_0 + \sum_{i=1}^n p_i q_i \le I,
$$ {#eq-um-classical}

where $q_0$ is a numeraire good with normalized price $p_0 = 1$. The solution is the
**Walrasian** (or Marshallian) demand system $q_i = D_i(\mathbf{p}, I)$, and
substituting it back into $U$ gives the **indirect utility** $V(\mathbf{p}, I)$.

This system is not an arbitrary set of functions; it inherits a rigid structure from
optimization, and that structure is what makes the classical approach powerful. Two
identities encode it. **Roy's identity** recovers demand from the indirect utility,

$$
q_i(\mathbf{p}, I) = -\,\frac{\partial V / \partial p_i}{\partial V / \partial I},
$$ {#eq-roy}

so the entire demand system is implicit in a single scalar function $V$. The **Slutsky
equation** decomposes the response of demand to a price change into a substitution
effect (the compensated, or Hicksian, derivative $\partial h_i / \partial p_j$) and an
income effect,

$$
\frac{\partial h_i(\mathbf{p}, u)}{\partial p_j}
= \frac{\partial q_i(\mathbf{p}, I)}{\partial p_j}
+ \frac{\partial q_i(\mathbf{p}, I)}{\partial I}\, q_j(\mathbf{p}, I).
$$ {#eq-slutsky}

The matrix of compensated derivatives—the **Slutsky substitution matrix**—is symmetric
and negative semi-definite. Symmetry is the testable fingerprint of utility
maximization: absent income effects, the cross-price response of good $i$ to $p_j$
equals that of good $j$ to $p_i$. Negative semi-definiteness encodes the law of
demand, that compensated own-price effects are non-positive.

### CES Sub-Utility and the Taste for Variety

A leading special case separates the numeraire from a bundle of differentiated goods,
$U = U(q_0, Q)$, with the differentiated goods aggregated by a **constant elasticity
of substitution** (CES) sub-utility. Taking $U = q_0 Q$ with

$$
Q = \Big( \sum_{i=1}^n q_i^{\rho} \Big)^{1/\rho}, \qquad \rho \le 1,
$$ {#eq-ces-Q}

delivers the CES demand system. The parameter $\rho$ controls substitutability: the
elasticity of substitution is $1/(1-\rho)$, so $\rho \in (0,1)$ makes the goods
imperfect substitutes (elasticity above one), $\rho \to 1$ makes them perfect
substitutes, and $\rho \to 0$ approaches Cobb–Douglas. The CES form is the standard
tool for studying product *variety*, because the consumer values having many distinct
varieties even at equal total quantity—the "love of variety" that justifies entry of
differentiated competitors.

To derive demand, write the consumer problem over $n+1$ goods with the multiplicative
specification

$$
U = q_0 \Big[ \sum_{i=1}^n q_i^{\rho} \Big]^{1/\rho}, \qquad \rho \in (0,1),
$$

subject to the budget $P_0 q_0 + \sum_{i=1}^n P_i q_i = M$. The Lagrangian is

$$
\mathcal{L} = q_0 \Big[ \sum_{i=1}^n q_i^{\rho} \Big]^{1/\rho}
+ \lambda \Big( M - P_0 q_0 - \sum_{i=1}^n P_i q_i \Big).
$$

The first-order condition for the numeraire,
$\partial \mathcal{L}/\partial q_0 = \big[\sum_i q_i^{\rho}\big]^{1/\rho} - P_0\lambda = 0$,
identifies the marginal utility of income with the aggregate $Q$ scaled by price,
$\big[\sum_i q_i^{\rho}\big]^{1/\rho} = P_0 \lambda$. The condition for a
differentiated good,

$$
\frac{\partial \mathcal{L}}{\partial q_i}
= q_0 \Big[ \sum_{i=1}^n q_i^{\rho} \Big]^{1/\rho - 1} q_i^{\rho - 1} - P_i \lambda = 0,
$$

gives $q_0 \big[\sum_i q_i^{\rho}\big]^{1/\rho - 1} q_i^{\rho-1} = P_i \lambda$.
Dividing the two conditions eliminates $\lambda$ and yields the *direct* CES demand;
solving instead for $P_i$ yields the *inverse* demand. The economic reading is the one
above: with $\rho \in (0,1)$ the varieties are imperfect substitutes.

### Quasi-Linear Utility and the Absence of Income Effects

A second special class drops income effects entirely. With **quasi-linear** utility,

$$
U = q_0 + u(q_1, \dots, q_n),
$$ {#eq-quasilinear}

substituting the budget constraint turns the consumer problem into

$$
\max_{q_1, \dots, q_n} \; u(q_1, \dots, q_n) - \sum_{i=1}^n p_i q_i + I,
$$

with first-order conditions $\partial u / \partial q_i = p_i$. Because $I$ enters
additively, the demands for the $n$ differentiated goods are **independent of income**
(provided income is large enough that $q_0 > 0$). All income is absorbed by the
numeraire, and the maximized utility is $V = v(p_1, \dots, p_n) + I$, where
$v(\cdot)$ is the consumer-surplus function. This additive separability is what makes
quasi-linear utility the default for welfare analysis in IO: changes in consumer
surplus are read directly off $v$, untangled from income.

The structure delivers three results used repeatedly below. First, demand is the
negative gradient of surplus, $-\partial v / \partial p_i = q_i(\mathbf{p})$, a special
case of Roy's identity (@eq-roy). Second, the **Jacobian** of the demand system,
$J = \big( \partial q_i(\mathbf{p}) / \partial p_j \big)_{n \times n}$, is symmetric and
negative semi-definite—the Slutsky matrix (@eq-slutsky) stripped of income terms.
Third, the construction can be **reverse-engineered**: if an observed demand system
has a symmetric, negative-definite Jacobian, it is rationalizable by a representative
consumer maximizing a concave quasi-linear utility, recovered by substituting the
inverse demand into the first-order conditions and integrating the resulting system of
partial differential equations $\partial u / \partial q_i = p_i(q)$. Recovering $u$ in
turn enables exact welfare statements.

#### Substitutes, Complements, and Diversion {#sec-io-diversion}

With $n = 2$, the sign of the cross-effect classifies the goods. They are
**demand-substitutes** if good $j$'s demand rises with $p_i$ and **demand-complements**
if it falls. Applying Cramer's rule to the first-order conditions of quasi-linear
maximization,

$$
|J|\, \frac{\partial q_1}{\partial p_2}
= |J|\, \frac{\partial q_2}{\partial p_1}
= -\,u_{12} = -\,u_{21},
$$ {#eq-cramer-cross}

where $|J| > 0$ by negative definiteness (strict concavity of $u$). The products are
therefore demand-substitutes when $u_{12} = u_{21} < 0$ and demand-complements when
$u_{12} = u_{21} > 0$: the *sign of the utility cross-partial* governs the substitution
relationship.

The **diversion ratio** quantifies substitution and is the single most important
demand statistic in merger analysis. For imperfect substitutes, it measures how much
of the demand lost from raising $p_1$ is recaptured by product 2,

$$
d_{12} = \frac{\partial q_2 / \partial p_1}{-\,\partial q_1 / \partial p_1} > 0.
$$ {#eq-diversion}

Intuitively, if a consumer prefers coffee brand A but buys brand B when A's price
rises (or A is out of stock), sales "divert" from A to B; $d_{12}$ is the fraction of
A's lost sales that lands on B. Negative definiteness of the substitution matrix
imposes a discipline on the pair of diversion ratios,

$$
\frac{\partial q_1}{\partial p_1}\frac{\partial q_2}{\partial p_2}
- \frac{\partial q_2}{\partial p_1}\frac{\partial q_1}{\partial p_2} > 0
\quad \Longleftrightarrow \quad d_{12}\, d_{21} < 1,
$$

so two goods cannot each divert most of their lost sales to the other. The empirically
estimable version, for products $A$ and $B$ with quantities and prices
$(Q_A, P_A, Q_B, P_B)$, is

$$
D_{AB} = \frac{\Delta Q_B / \Delta P_A}{\Delta Q_A / \Delta P_A},
$$ {#eq-diversion-emp}

the change in $B$'s quantity per unit change in $A$'s price, normalized by $A$'s own
quantity response. The following seeded example simulates two products and estimates
the diversion ratio from first differences.

```{r io-diversion-sim, message=FALSE, warning=FALSE}
set.seed(123)
n <- 100
price_A <- runif(n, 5, 10)
price_B <- runif(n, 5, 10)
quantity_A <- 100 - 5 * price_A + rnorm(n)
quantity_B <-  80 - 4 * price_B + rnorm(n)

# Diversion ratio from first differences (eq. of D_AB)
delta_QA <- diff(quantity_A)
delta_PA <- diff(price_A)
delta_QB <- diff(quantity_B)

D_AB <- mean(delta_QB / delta_PA) / mean(delta_QA / delta_PA)
D_AB
```

```{r io-diversion-plot, message=FALSE, warning=FALSE}
#| fig-cap: "Change in quantity of each product against the change in product A's price. Product A (blue) responds steeply to its own price; product B (red) is comparatively unresponsive, so little demand diverts from A to B."
library(ggplot2)
plot_df <- data.frame(delta_PA, delta_QA, delta_QB)
ggplot(plot_df, aes(x = delta_PA)) +
  geom_point(aes(y = delta_QA), color = "blue", alpha = 0.5) +
  geom_point(aes(y = delta_QB), color = "red",  alpha = 0.5) +
  geom_smooth(aes(y = delta_QA), method = "lm", color = "blue", se = FALSE) +
  geom_smooth(aes(y = delta_QB), method = "lm", color = "red",  se = FALSE) +
  labs(x = "Change in price of A", y = "Change in quantity",
       title = "Change in quantity vs. change in price of A",
       subtitle = "Blue: product A   Red: product B") +
  theme_minimal()
```

### The Linear Demand System from Quadratic Utility

The most-used closed-form demand system comes from a quasi-linear **quadratic**
utility. Over two differentiated goods,

$$
U(q_0, q_1, q_2) = q_0 + \alpha_1 q_1 + \alpha_2 q_2
- \tfrac{1}{2}\big[ \beta_1 q_1^2 + 2\gamma\, q_1 q_2 + \beta_2 q_2^2 \big],
$$ {#eq-quad-utility}

with $\alpha_i > 0$, $\beta_i > 0$, and $\beta_1 \beta_2 - \gamma^2 > 0$ (concavity).
The first-order conditions are the **inverse** (linear) demands,

$$
p_1 = \alpha_1 - \beta_1 q_1 - \gamma q_2, \qquad
p_2 = \alpha_2 - \beta_2 q_2 - \gamma q_1,
$$ {#eq-linear-inverse}

and inverting this linear system gives the direct demands with constant diversion
ratios. The cross-partial of utility is exactly $\partial^2 U / \partial q_1 \partial q_2
= \gamma$, so $\gamma$ *is* the substitution parameter: under the sign convention of
@eq-cramer-cross, a positive cross-partial signals complements and a negative one
signals substitutes. Two parameterization styles appear in practice. Treating $\gamma$
as a free parameter captures the relationship directly and flexibly. Tying the
curvature to substitution by setting $\beta_1 = \beta_2 = 1 - \gamma$ is more
parsimonious but restrictive: it forces own-curvature to move with the cross-effect,
which can be arbitrary when theory offers no guidance.

#### The Shubik–Levitan Normalization

A subtle defect of the raw quadratic form is that adding products mechanically changes
the scale of demand. @shubik1980 propose a normalization that separates the *number*
of products from their *substitutability*. With

$$
u(q_1, \dots, q_n) = \alpha \sum_{i=1}^n q_i
- \frac{n}{2(1+\gamma)}\Big[ \sum_{i=1}^n q_i^2
+ \frac{\gamma}{n}\Big( \sum_{i=1}^n q_i \Big)^2 \Big],
$$ {#eq-shubik-utility}

where $\alpha > 0$ and $\gamma \in [0, \infty)$ indexes substitutability, the demand
system becomes

$$
q_i = \frac{1}{n}\Big[ \alpha - p_i(1+\gamma) + \frac{\gamma}{n} \sum_{j=1}^n p_j \Big].
$$ {#eq-shubik-demand}

This form has two properties that make it the standard for oligopoly-variety models.
The aggregate demand $Q = \sum_i q_i = \alpha - \tfrac{1}{n}\sum_j p_j$ does not depend
on the degree of substitution $\gamma$, and under symmetric prices $p_i = p$ the
aggregate $Q = \alpha - p$ does not depend on the number of products $n$. Holding the
"market" fixed while varying competition and variety is exactly what one wants when
asking how entry or differentiation reshapes equilibrium.

The normalization supports two instructive applications. In the first, a downstream
**buyer** with revenue $u(q_1, q_2)$ faces a *market-share restriction* imposed by the
seller of product 1: $q_1/(q_1 + q_2) \ge s$ or $q_1 = 0$, where $s \in [0,1]$ is the
required share and $s = 1$ is an exclusive contract. The buyer solves
$\max_{q_1, q_2} u(q_1, q_2) - p_1 q_1 - p_2 q_2$ subject to that constraint, and the
linear-quadratic utility yields a constrained demand $q(p, s)$ whose comparative
statics in $p$ and $s$ reveal how share-based contracts distort purchasing. In the
second, a demand system satisfies the **strong-complements** property if the ratio
$q_i^*(p)/q_1^*(p)$ is independent of $p_1$ for $i = 2, \dots, n$. This is equivalent
to the statement that, for any constant marginal costs $c_i > 0$, the monopolist's
problem $\max_p \sum_i (p_i - c_i) q_i^*(p)$ sets $p_i^* = c_i$ for $i = 2, \dots, n$:
when goods are strong complements, the multiproduct monopolist prices the
complementary goods at marginal cost and earns its margin entirely on the base good.
Writing $v(p) = u(q^*(p)) - \sum_i p_i q_i^*(p)$ for indirect utility, the maximum
profit conditional on delivering the consumer at least $u$,
$\pi(u) = \max_p \{ \sum_i (p_i - c_i) q_i^*(p) : v(p) \ge u \}$, is decreasing and
concave in the promised surplus.

### Equilibrium Pricing and Cost Pass-Through {#sec-io-bertrand}

The demand system is only half the model; the other half is firm conduct. Suppose two
single-product firms with constant marginal costs $c_1, c_2$ set prices
simultaneously. Firm 1 maximizes $\pi_1(p_1, p_2) = (p_1 - c_1) q_1(p_1, p_2)$, with
first-order condition

$$
q_1 + (p_1 - c_1)\frac{\partial q_1}{\partial p_1} = 0,
$$ {#eq-bertrand-foc}

and symmetrically for firm 2. The pair of conditions defines the **Bertrand–Nash**
equilibrium prices $(p_1^*, p_2^*)$. A key comparative static is **cost pass-through**
(CPT)—how an equilibrium price responds to a marginal-cost shock—with own and cross
rates

$$
CPT_{11} = \frac{\partial p_1^*}{\partial c_1}, \quad
CPT_{22} = \frac{\partial p_2^*}{\partial c_2}, \quad
CPT_{12} = \frac{\partial p_1^*}{\partial c_2}, \quad
CPT_{21} = \frac{\partial p_2^*}{\partial c_1}.
$$ {#eq-cpt}

Pass-through is the quantity a regulator or analyst needs to predict how a tax,
tariff, or input-cost shock reaches consumers, and it depends on the *curvature* of
demand, not merely its slope.

When the two firms **merge or collude**, the combined entity becomes a multiproduct
monopolist maximizing joint profit
$\pi(p_1, p_2) = (p_1 - c_1) q_1 + (p_2 - c_2) q_2$, with first-order conditions

$$
\begin{aligned}
\frac{\partial \pi}{\partial p_1}
&= q_1 + (p_1 - c_1)\frac{\partial q_1}{\partial p_1}
+ (p_2 - c_2)\frac{\partial q_2}{\partial p_1} = 0,\\[4pt]
\frac{\partial \pi}{\partial p_2}
&= q_2 + (p_2 - c_2)\frac{\partial q_2}{\partial p_2}
+ (p_1 - c_1)\frac{\partial q_1}{\partial p_2} = 0.
\end{aligned}
$$ {#eq-monopoly-foc}

The monopolist's conditions differ from the competitive ones (@eq-bertrand-foc) by the
extra cross-effect term $(p_j - c_j)\,\partial q_j / \partial p_i$: the merged firm
**internalizes** the demand it steals from itself. For substitutes this raises both
prices above the Bertrand level—the classic unilateral-effects concern in merger
review—while pass-through rates $CPT^M$ generally differ from the competitive case
because the monopolist accounts for cross-effects when passing through cost shocks.
@tbl-bertrand-monopoly contrasts the two regimes.

| Feature | Bertrand competition | Multiproduct monopoly / collusion |
|---|---|---|
| First-order condition | own margin × own slope only | own margin + cross-margin × cross-slope |
| Internalizes cross-effects | no | yes |
| Equilibrium prices (substitutes) | lower | higher |
| Cost pass-through | own-cost driven | reflects internalized cross-effects |
| Treatment of rival product | competitor | own product line |

: Bertrand competition versus joint pricing under the same linear demand system. The single structural difference—internalizing cross-demand—drives the price and pass-through divergence. {#tbl-bertrand-monopoly}

The consumer side of this model is closed by **consumer surplus**. Integrating each
inverse demand from @eq-linear-inverse below the price gives
$CS_i = \int_0^{q_i} p_i(q_i', q_{-i})\,dq_i' - p_i q_i$, and total surplus is
$CS = CS_1 + CS_2$; this is the welfare object compared across the competitive and
monopoly regimes.

### Properties of Multiproduct Demand and the Symmetry of Cross-Effects

For a single product, demand $q = D(p)$ is downward-sloping, $q'(p) < 0$, with price
elasticity $\epsilon(p) = -q'(p)\,p/q$. With multiple products the new content is the
*cross*-price structure. A linear system

$$
\begin{aligned}
q_1 &= \alpha_1 - \beta_{11} p_1 + \gamma_{12} p_2 + \cdots,\\
q_2 &= \alpha_2 - \beta_{22} p_2 + \gamma_{21} p_1 + \cdots,
\end{aligned}
$$

has own-price effects $\partial q_i / \partial p_i < 0$ and cross-price effects whose
sign is positive for substitutes and negative for complements. A foundational result
is that the cross-effects are **symmetric**,
$\partial q_1 / \partial p_2 = \partial q_2 / \partial p_1$, whenever consumers have
quasi-linear utility (no income effect). The derivation makes the source of symmetry
explicit. From $\max_{q_1, q_2} u(q_1, q_2) - p_1 q_1 - p_2 q_2$ the first-order
conditions $u_1(q_1^*, q_2^*) = p_1$ and $u_2(q_1^*, q_2^*) = p_2$ hold; differentiating
the system with respect to $p_1$,

$$
\begin{aligned}
u_{11}\frac{\partial q_1^*}{\partial p_1} + u_{12}\frac{\partial q_2^*}{\partial p_1} &= 1,\\
u_{21}\frac{\partial q_1^*}{\partial p_1} + u_{22}\frac{\partial q_2^*}{\partial p_1} &= 0,
\end{aligned}
$$

and solving yields

$$
\frac{\partial q_2^*}{\partial p_1} = -\,\frac{u_{21}}{u_{11}u_{22} - u_{12}u_{21}},
\qquad
\frac{\partial q_1^*}{\partial p_2} = -\,\frac{u_{12}}{u_{11}u_{22} - u_{12}u_{21}}.
$$

Because $u_{12} = u_{21}$ (Young's theorem on the symmetry of cross-partials), the two
cross-effects are equal. This is the Slutsky symmetry of @eq-slutsky with the income
terms switched off.

When income effects are present, symmetry of the *uncompensated* cross-effects no
longer follows automatically. Slutsky symmetry holds for the *compensated* derivatives,

$$
\frac{\partial h_i}{\partial p_j}
= \frac{\partial q_i}{\partial p_j} + \frac{\partial q_i}{\partial I} q_j
= \frac{\partial h_j}{\partial p_i}
= \frac{\partial q_j}{\partial p_i} + \frac{\partial q_j}{\partial I} q_i,
$$

so the observed (uncompensated) cross-effects are equal only if the income terms
balance, i.e. if $\frac{\partial q_i}{\partial I}/q_i = \frac{\partial q_j}{\partial I}/q_j$.
The empirical reading is useful: if the data show *equal* uncompensated cross-price
effects, that is consistent evidence for either no income effect, or income effects
that scale the two goods' consumption proportionally—an additional rupee of income
expands consumption of both goods in the same relative measure. Two normalizations make
the cross-versus-own comparison concrete: assuming
$-\partial q_i / \partial p_i > \partial q_j / \partial p_i$ keeps the diversion ratio
below one, and assuming $-\partial q_i / \partial p_i > \partial q_i / \partial p_j$
keeps the *indirect* diversion ratio below one. Own- and cross-price elasticities are
defined analogously by scaling each derivative by $p/q$.

The classical, continuous approach has real strengths: the tight algebraic structure
of the demand system delivers clean pricing strategies and exact welfare statements.
But it pays little attention to product *characteristics*—quality, location, timing,
availability, the consumer's information about a product's existence and quality—and it
does not formally model the *heterogeneity* of consumer preferences. Discrete-choice
models, developed in @sec-io-discrete, are designed precisely to address these gaps.

### Tractable Multiproduct Demand: Homothetic Surplus {#sec-io-armstrong}

A frontier question is which multiproduct demand systems are simultaneously rich
enough to be realistic and tractable enough to solve. @armstrong2018multiproduct
answer it by characterizing the demand systems whose *consumer surplus* is homothetic,
a property that collapses a high-dimensional pricing problem into a one-dimensional
one. Begin with the representative consumer's quasi-linear problem: utility $u(q)$ with
$u(0) = 0$, increasing and strictly concave, choosing $q$ to solve $\max_q u(q) - p
\cdot q$. The inverse demand is the gradient $p(q) = \nabla u(q)$, total revenue is
$R(q) = q \cdot \nabla u(q)$, and consumer surplus is

$$
s(q) = u(q) - q \cdot \nabla u(q).
$$ {#eq-armstrong-surplus}

> **Proposition** [@armstrong2018multiproduct]. Consumer surplus $s(q)$ is homothetic
> in $q$ if and only if utility takes the form
> $$ u(q) = h(q) + g\big(f(q)\big), $$
> where $h$ and $f$ are homogeneous of degree one and $g$ is concave with $g(0) = 0$.

This is a far wider class than the systems whose surplus is homothetic in *prices*
(the case $h = 0$), and it exposes exactly **three degrees of freedom**—the functions
$f(q)$, $h(q)$, and $g(\cdot)$—that the modeler can choose to match data while
preserving tractability. The "if" direction is instructive. Inverse demand is
$p(q) = \nabla h(q) + g'(f(q))\,\nabla f(q)$, and since $q \cdot \nabla h(q) \equiv
h(q)$ for a degree-one homogeneous function, revenue is $R(q) = h(q) + g'(f(q)) f(q)$,
so surplus reduces to $s(q) = g(f(q)) - g'(f(q)) f(q)$—a monotone function of the
single composite $f(q)$.

The proof technique is a change to "polar coordinates" that pays off conceptually.
Write $q = f(q) \cdot \tfrac{q}{f(q)}$, where $q/f(q)$ is homogeneous of degree zero and
depends only on the *ray* from the origin (the mix of products), while $f(q)$—the
**composite quantity**—measures how far along that ray the bundle lies. The consumer's
problem then *separates* into two stages. Surplus
$u(q) - p\cdot q = g(f(q)) - f(q)\frac{p\cdot q - h(q)}{f(q)}$ is maximized, for any
fixed $f(q)$, by minimizing $\frac{p\cdot q - h(q)}{f(q)}$; define the **composite
price**

$$
\phi(p) \equiv \min_{q \ge 0} \frac{p \cdot q - h(q)}{f(q)},
$$ {#eq-armstrong-phi}

which is increasing and concave in $p$. By the envelope theorem the optimal *relative
quantities* are $q^r(p) \equiv \nabla \phi(p)$. Given those, the consumer chooses the
composite quantity $Q$ to maximize the concave objective $g(Q) - Q\,\phi(p)$, whose
solution $\hat{Q}(\phi(p))$ is the demand for the composite. The full demand function
factors as $q(p) = \hat{Q}(\phi(p)) \times q^r(p)$: the consumer picks a *mix* from
relative prices and a *scale* from the composite price, and all prices with the same
$\phi(p)$ induce the same composite quantity through the inverse relation
$\phi(p) = g'(\hat{Q})$.

Three examples show the family's reach. The **linear** system $u(q) = aq -
\tfrac{1}{2} q^\top B q$ with $a > 0$ and $B$ positive definite gives $p(q) = a - Bq$
and decomposes as $h(q) = aq$, $f(q) = \sqrt{q^\top B q}$, $g(f) = -\tfrac{1}{2}f^2$,
and $s(q) = \tfrac{1}{2} f(q)^2$. The **logit** system
$q_i(p) = \frac{e^{(a_i - p_i)/\mu}}{k + \sum_j e^{(a_j - p_j)/\mu}}$ corresponds to
surplus $v(p) = \mu \log\!\big(1 + \tfrac{1}{k}\sum_j e^{(a_j - p_j)/\mu}\big)$, with
the decomposition $f(q) = \sum_j q_j$ (total quantity),
$g(f) = -\mu\big(f\log f + (1-f)\log(1-f)\big)$, and $s(q) = -\mu\log(1 - f(q))$. The
**complementary-products** example captures hardware-plus-services: if $q_1$ is the
number of access units (hardware) and each requires services $y = q_2/q_1$, gross
utility is $u(q_1, q_2) = q_1 \hat{u}(q_2/q_1) + g(q_1)$, an instance with $f(q) = q_1$,
so surplus depends only on the number of access units. The ratio $q_2/q_1$ is
independent of $p_1$—precisely the strong-complements property of @sec-io-classical—
with the implication that the monopolist prices services to extract the base-good
margin.

## Vertical Differentiation: Quality {#sec-io-vertical}

Differentiation is **vertical** when augmenting a product characteristic benefits *all*
consumers: everyone prefers higher quality, faster delivery, longer battery life. The
canonical model gives a consumer of type $\theta$ utility

$$
u =
\begin{cases}
\theta s - p & \text{if she buys quality } s \text{ at price } p,\\
0 & \text{if she does not buy,}
\end{cases}
$$ {#eq-vertical-utility}

where $s > 0$ is quality (common knowledge), utility is separable in quality and
price, and $\theta > 0$ is a taste parameter—the consumer's marginal willingness to pay
for quality—distributed across the population with density $f(\theta)$ and cdf
$F(\theta)$ on $[\underline{\theta}, \bar{\theta}]$. Firms know the distribution but not
any individual's $\theta$. The defining feature is unanimity: every consumer prefers
higher $s$, in contrast to horizontal differentiation where consumers disagree about
the ideal variety.

**One quality.** When a single quality $s$ is offered at price $p$, a type $\theta$
buys iff $\theta s - p \ge 0$, i.e. $\theta \ge p/s$. With $N$ identical consumers,
aggregate demand is

$$
D(p) = N\big[1 - F(p/s)\big],
$$ {#eq-vertical-demand1}

which is downward-sloping. Under the uniform $F$, $D(p) = N(1 - p/s)$ for $p \le s$ and
zero otherwise. Unlike classical demand, quality $s$ enters *explicitly*, so the model
can speak to quality choice, not just price.

**Two qualities.** Offering a product line $s_1 < s_2$ (with $p_1 < p_2$) lets the firm
screen consumers and extract more surplus—the logic of second-degree price
discrimination. A type $\theta$ buys $s_1$ iff $\theta s_1 - p_1 \ge 0$ and
$\theta s_1 - p_1 \ge \theta s_2 - p_2$; she buys $s_2$ iff $\theta s_2 - p_2 \ge 0$ and
$\theta s_2 - p_2 \ge \theta s_1 - p_1$. Three cutoffs—$p_1/s_1$, $p_2/s_2$, and the
indifference type $\theta^* \equiv (p_2 - p_1)/(s_2 - s_1)$—partition consumers across
not buying, buying low, and buying high. The configuration depends on which good offers
more *quality per dollar*. If the high good dominates on quality-per-dollar
($p_2/s_2 \le p_1/s_1$), the low good is driven out and only $s_2$ sells; otherwise both
sell, with the low good serving an interior band of types. Collecting both cases,

$$
D_1(p_1, p_2) =
\begin{cases}
0 & \text{if } \tfrac{p_1}{p_2} \ge \tfrac{s_1}{s_2},\\[4pt]
N\big[F(\tfrac{p_2 - p_1}{s_2 - s_1}) - F(\tfrac{p_1}{s_1})\big] & \text{if } \tfrac{p_1}{p_2} < \tfrac{s_1}{s_2},
\end{cases}
$$ {#eq-vertical-D1}

$$
D_2(p_1, p_2) =
\begin{cases}
N\big[1 - F(\tfrac{p_2}{s_2})\big] & \text{if } \tfrac{p_1}{p_2} \ge \tfrac{s_1}{s_2},\\[4pt]
N\big[1 - F(\tfrac{p_2 - p_1}{s_2 - s_1})\big] & \text{if } \tfrac{p_1}{p_2} < \tfrac{s_1}{s_2}.
\end{cases}
$$ {#eq-vertical-D2}

Under the uniform special case $F(\theta) = \theta$ on $[0,1]$, the interior cross-price
derivatives are $\partial D_1 / \partial p_2 = \partial D_2 / \partial p_1 = N/(s_2 -
s_1) > 0$: the two qualities are **substitutes**, and the strength of substitution
scales inversely with the quality gap $s_2 - s_1$. Widening the quality ladder softens
competition between the firm's own tiers—the structural reason a multiproduct firm
spaces its versions apart.

The taste parameter $\theta$ admits a measurable interpretation: it can stand for
willingness to pay for quality and is often linked to **income**, the most observable
axis of consumer heterogeneity. If consumers differ in income and utility is
$U = u(I - p) + s$, a first-order Taylor expansion of $u$ around $I$ recovers the
quasi-linear form $\theta s - p$ with $\theta$ tied to the marginal utility of income,
making the abstract taste parameter empirically grounded.

## Discrete-Choice Demand {#sec-io-discrete}

Discrete-choice models replace the continuous-quantity consumer with one who selects a
single alternative from a menu, which is how modern empirical IO and quantitative
marketing estimate demand for differentiated products. The payoff is twofold: a small
number of *preference* parameters generates the full matrix of own- and cross-price
elasticities, and demand can be predicted for *new* or hypothetical products by
projecting them into characteristic space. @fig-discrete-lineage traces the lineage of
models this section develops, from the workhorse logit through its corrections for
heterogeneity and endogeneity to the frontier estimators for zero shares and
multi-sided markets.

```{mermaid}
%%| label: fig-discrete-lineage
%%| fig-cap: "Lineage of discrete-choice demand models. Each arrow relaxes a limitation of the parent: random coefficients break the IIA restriction, BLP adds price endogeneity, and the frontier estimators handle zero shares and platform externalities."
flowchart TD
    A[Random utility<br/>McFadden] --> B[Multinomial logit MNL<br/>closed form, but IIA]
    B --> C[Nested logit / GEV<br/>relax IIA across groups]
    B --> D[Multinomial probit<br/>flexible correlation]
    B --> E[Random-coefficients logit<br/>heterogeneous tastes]
    E --> F[BLP<br/>+ price endogeneity, share inversion]
    F --> G[Zero-share estimator<br/>pairwise differencing]
    F --> H[Multi-sided markets<br/>+ cross-side externalities]
```

### Random Utility and the Multinomial Logit {#sec-io-mnl}

@mcfadden1986choice frames market research through **economic choice theory**: a
consumer is an "optimizing black box" mapping inputs (product attributes, socioeconomic
features, market information, history, constraints) to outputs (purchase decisions,
consumption levels). The behavioral premise is that individuals maximize a preference
order influenced by perceptions, attitudes, and past decisions, with commodities
represented as bundles of attributes. The intellectual debt is explicit: Thurstone's
1927 *random utility* idea—choice of the momentarily most beneficial alternative—was
revived by Luce in 1959 and married, by the mid-1960s, to a *multidimensional
attribute* representation of goods, producing the multinomial logit. McFadden's program
is to model the cognitive process inside the black box using experimental data on
consumer cognition, because historical field data may not reveal how consumers would
respond to genuinely new product designs. The constructs that populate the box—product
perceptions, general attitudes, preferences, decision protocols, and behavioral
intentions—are measured by psychometric instruments (multidimensional scaling for
perceptions, factor analysis for tastes, verbal protocols for decision-making, conjoint
analysis for preferences), and the central challenge is to turn those psychometric
indicators into market forecasts that can be validated. Laboratory elicitation is
attractive for its precise control, ability to study not-yet-marketed innovations, and
low cost, but only field validation is fully convincing, and the literature warns that
incentive-poor experiments may mispredict real behavior.

The **multinomial logit** (MNL) is the workhorse random-utility model. With a choice
set $C = \{1, \dots, M\}$ and scale values (strict utilities) $V_i$, the probability of
choosing $i$ is

$$
P_C(i) = \frac{\exp(V_i)}{\sum_{j \in C} \exp(V_j)}.
$$ {#eq-35-mnl}

The scale values are functions of alternatives' attributes, can interact with
individual characteristics and choice-set features, and usually take an additively
separable linear form $V_i = \mathbf{X}_i \boldsymbol{\beta}$. The closed form follows
from assuming the unobserved utility component is i.i.d. type-I extreme value (double
exponential). The MNL inherits Luce's **independence of irrelevant alternatives**
(IIA) axiom,

$$
\frac{P_A(i)}{P_A(j)} = \frac{P_C(i)}{P_C(j)}
\qquad \forall\, i, j \in A \subseteq C,
$$ {#eq-35-iia}

which states that the relative odds of two alternatives do not depend on what else is
available. IIA is a double-edged property. It buys real advantages: inference about
many alternatives from paired comparisons, easy demand forecasting for new
alternatives, a clean link between choice probabilities and ranking data (ideal for
conjoint), and the closed form itself. But it is often unrealistic—it forces a uniform
substitution response to any one alternative's attribute change, at odds with the
clustered similarity patterns of real markets (the "red bus / blue bus" problem), and
this defect, too, traces to the extreme-value error assumption.

The literature addresses IIA in three ways: test for it empirically; modify the MNL so
scale values depend on choice-set features; or switch models. The alternatives include
**multinomial probit** (MNP), which permits arbitrary correlation patterns through
normally distributed utilities at the cost of high-dimensional integration; **nested
MNL**, the marketing standard, which groups similar alternatives so IIA holds only
*within* nests; **generalized extreme value**; and elimination-by-aspects variants
(HEBA, elimination by strategy). Consumer heterogeneity is handled by random-
coefficients MNL (the marketing standard for breaking IIA), fixed-coefficient
specifications, or hierarchical Bayesian models. Estimation is normally by maximum
likelihood, though richer models demand more computation: hierarchical models arrange
alternatives in a preference tree and choose by eliminating clusters, while MNP and
random-coefficients MNP rely on **Monte Carlo simulation** for both estimation and
forecasting. A low-dimensional **factor-analytic** structure on the covariance matrix
keeps MNP tractable, and the *hybrid* model of Westin and Talvitie makes choice
probabilities logit with scale values linear in random taste parameters. The
**simulated moments** estimator of @mcfadden1986method makes high-dimensional MNP
practical: it is consistent and asymptotically normal while requiring only one Monte
Carlo draw per observation, and it supplies computationally feasible instruments and a
covariance estimator for hypothesis testing. Market demand is then the sum of
individual choice probabilities, approximated either by **microsimulation** (Monte
Carlo sampling over the population, the more flexible route, which reflects population
heterogeneity when the model carries fixed-effect parameters) or by segmenting the
population and summing representative probabilities. Forecasting additionally requires
external projections of prices, attributes, demographics, and population size.

A distinctive strand of @mcfadden1986choice incorporates **psychometric data** directly
into the choice model. Perceptions and attitudes—scaled metrically—shape the utility
function and the product-attribute scales, formalized in a latent-variable system that
links measured causes and indicators to a utility-maximizing choice protocol.
Psychometricians favor *fixed-effect* descriptions of individual heterogeneity;
marketing favors *random-effect* models for their parsimony, forecast accuracy, and
flexibility. A practical sequential estimator first regresses the psychometric
indicators to recover the latent variables, then estimates the choice model on the
fitted latents; a more efficient estimator integrates over the latent-variable
distribution by maximum likelihood, with the simulated-moments estimator as a lighter
alternative. **Conjoint analysis** is the experimental engine for this program: it
gathers laboratory choice data in simulated markets plus self-explicated attribute
importances, and—because everything is exogenous by design—it can be used much like
real market data for forecasting, with *choice-based* conjoint the gold standard for
mimicking reality. Its validity hinges on the stability of preferences across task
framings and incentives, the structure of the preference map, and the cognitive
congruence of experimental and market tasks. Because the MNL's IIA structure is a
maintained assumption when conjoint data are summarized this way, it must be tested:
auxiliary regressions detect omitted variables and interactions, and the tests of
McFadden–Tye–Train and Hausman–McFadden gauge IIA, with power depending on the
variables already in the base model and the capacity to discriminate among nesting
structures.

### Differentiated-Products Demand and Price Endogeneity {#sec-io-berry}

The defining obstacle in estimating differentiated-products demand is **price
endogeneity** arising from *unobserved* product characteristics. @berry1994estimating
gives the structural interpretation: a demand error represents unobserved demand
influences—style, design, perceived quality, durability—that the econometrician cannot
measure but that firms observe and price against. Because high-$\xi$ products carry
both higher demand and higher prices, the correlation between price and the demand
error biases naive estimates and can even produce upward-sloping fitted demand curves,
exactly the anomaly Trajtenberg documents in CT scanners and Berry, Levinsohn, and
Pakes confirm in automobiles. For *homogeneous* goods this is a standard endogeneity
problem fixable with instrumental variables on a constant-elasticity system
$\ln(q_j) = \alpha_j + \sum_k \eta_{jk}\ln(p_k) + \epsilon_j$, where $\eta_{jk}$ is the
elasticity of good $j$ to price $k$. But in discrete-choice models price and $\xi$ enter
demand *non-linearly*, so linear IV cannot be applied directly.

There is also a dimensionality problem. A system of $N$ goods has $N^2$ elasticities,
infeasible to estimate freely. Imposing *a priori* restrictions (setting some
cross-elasticities to zero or to common values) is arbitrary and weakly grounded in
theory. The discrete-choice solution is to project demand onto a low-dimensional
**characteristic space**: consumer utility depends on product characteristics and a few
random taste parameters, market demand aggregates individual choices, and all
elasticities are derived from the handful of utility parameters—which also lets the
model predict demand for new and dissimilar products and move freely between aggregate
demand and the underlying utility. The cost is the discrete-choice straitjacket: it
rules out buying multiple items, struggles with dynamics, and imposes parametric
structure. This program descends from McFadden (1974) and Bresnahan's (1987) vertical
model of autos, both of which model utility over characteristics, allow an *outside
good*, and use price-setting firms' first-order conditions—while, like most of the
empirical literature, treating characteristics as exogenous even as prices are
determined within the model.

@berry1994estimating resolves the non-linear IV problem with the **mean-utility
inversion**. The data are $R$ independent markets; market $r$ has $N_r$ single-product
firms; product $j$ in market $r$ has observed characteristics $z_{jr} \in
\mathbb{R}^{K_z}$ (affecting demand $x_j$ and marginal cost $w_j$) and unobserved
characteristics $(\xi_j, \omega_j)$, where $\xi_j$ shifts demand and $\omega_j$ shifts
cost. The unobservables are assumed mean-independent of observed characteristics and
independent across markets, so $(\mathbf{z}, \boldsymbol{\xi}, \boldsymbol{\omega})$ are
exogenous to pricing. Consumer $i$'s utility for product $j$ is $U(x_j, \xi_j, p_j,
\nu_i, \theta_d)$, with $\nu_i$ a consumer-specific term. The leading **random-
coefficients** specification is

$$
u_{ij} = x_j \tilde{\beta} - \alpha p_j + \xi_j + \epsilon_{ij},
$$ {#eq-35-blp-utility}

where the taste for characteristic $k$ is $\tilde{\beta}_k = \beta_k + \sigma_k
\zeta_{ik}$ with $\beta_k$ the mean taste and $\zeta_{ik}$ i.i.d. standard normal across
consumers and characteristics. Substituting,

$$
u_{ij} = x_j \beta + \xi_j - \alpha p_j + \nu_{ij},
\qquad
\nu_{ij} = \sum_k x_{jk}\,\sigma_k\,\zeta_{ik} + \epsilon_{ij},
$$ {#eq-blp-decomp}

where $\nu_{ij}$ is a mean-zero, heteroskedastic error capturing the random tastes.
Collecting the terms common to all consumers defines the **mean utility**

$$
\delta_j \equiv x_j \beta - \alpha p_j + \xi_j.
$$ {#eq-blp-mean-utility}

The estimator's insight is that observed market shares can be *inverted* to recover the
$\delta_j$, which are linear in the demand error $\xi_j$; instrumental variables can
then be applied to that linear equation. We make the inversion explicit in
@sec-io-zeroshares, where the zero-share problem is also confronted. The companion
papers extend the framework to equilibrium: @berry1995automobile estimates automobile
prices in market equilibrium with supply-side first-order conditions, and
@berry2004differentiated combine micro (individual) and macro (aggregate) data to
sharpen identification of the new-car demand system.

### Zero-Valued Market Shares {#sec-io-zeroshares}

The BLP machinery assumes every product has a strictly positive share, because the
share inversion takes logs. Real retail data violate this constantly:
@dube2021random report that in scanner data, products with literally zero sales are
ubiquitous—30.7% of store-day-product observations in their ice-cream application
record no sale—and these zeros persist even in large consumer populations, so they are
not sampling noise. The standard fixes are *ad hoc* and biased: imputing a tiny share
(say $s_j = 10^{-12}$) makes results sensitive to the arbitrary small number, dropping
zero-share observations discards information and, worse, induces **selection bias**
when zeros are systematic, and common data sources (IRI, Nielsen) silently exclude
zero-sales products. @dube2021random build an estimator that treats the zeros as
informative through a regime-switching selection model and a pairwise-difference GMM.

The demand side is a **random-coefficients logit**. With a consideration set $J = \{1,
\dots, J+1\}$ where $J+1$ is the outside option ($p_{J+1} = 0$), household $h$ with
budget $y$ maximizes $U(\mathbf{q}, y; \theta^h) = \sum_{j=1}^{J+1} (\chi_j^h q_j)
\exp(\alpha^h q_0)$ over discrete consumption, where perceived quality is $\chi_j^h =
\exp(x_j'\beta^h + \xi_j + \epsilon_j^h)$, $x_j$ are exogenous attributes, $\xi_j$ is an
unobserved vertical characteristic, $\beta^h$ are taste parameters, and $\alpha^h$ is
the marginal value of the numeraire. The choice-specific indirect utilities are

$$
u_j^h = \alpha^h(y - p_j) + x_j'\beta^h + \xi_j + \epsilon_j^h,
\qquad
u_{J+1}^h = \alpha^h y + \epsilon_{J+1}^h,
$$ {#eq-dube-utility}

with $\epsilon_j^h$ i.i.d. extreme value and preferences $\theta^h = (\alpha^h,
\beta^h)' \sim F_\theta(\cdot; \Theta)$. The predicted share integrates the logit
probability over the taste distribution,

$$
\pi_j(\mathbf{p}, \mathbf{x}, \xi; \Theta)
= \int \frac{\exp(\alpha^h p_j + x_j'\beta^h + \xi_j)}
{1 + \sum_{k=1}^J \exp(\alpha^h p_k + x_k'\beta^h + \xi_k)}\,
dF_\theta(\theta^h; \Theta),
$$ {#eq-dube-share}

and is always strictly positive, which is exactly why the model alone cannot rationalize
an observed zero.

The fix is a **product-selection** stage. For each product an indicator $d_j = \mathbf{1}
\big(\phi(p_j, x_j, \tilde{\mathbf{w}}_j) + \eta_j > 0\big)$ governs whether $j$ enters
the consumer's choice set, where $\tilde{\mathbf{w}}_j$ are selection-shifting factors
(stock-outs, feature ads, in-store displays) that move a product's *consideration* but
not its consumption value [@mehta2003price], and $\eta_j$ is a continuous selection
shock. Observed shares then follow a **regime switch**,

$$
\pi_j(\mathbf{p}, \mathbf{x}, \xi; \Theta) =
\begin{cases}
\displaystyle \int \frac{\exp(\alpha^h p_j + x_j'\beta^h + \xi_j)}
{1 + \sum_{k \in C} \exp(\alpha^h p_k + x_k'\beta^h + \xi_k)}\, dF_\theta & d_j = 1,\\[10pt]
0 & d_j = 0,
\end{cases}
$$ {#eq-dube-regime}

with $C$ the realized choice set. The econometric danger is **selection on
unobservables**: demand and selection shocks can be correlated, $\mathrm{cov}(\xi_j,
\eta_j) \ne 0$. The motivating example is a locally sourced premium item (high $\xi_j$)
that retailers give scarce shelf space and stock thinly (high non-selection), so quality
and consideration move together; if shelf-space and inventory data are unavailable, the
correlation is unavoidable.

Identification rests on **BLP's moment condition** $E(\xi_j \mid \mathbf{z}_j) = 0$,
where $\mathbf{z}_j$ stacks attributes $x_j$ and excluded price shifters
$\tilde{\mathbf{r}}_j$. @berry1994estimating proves the share map $s_j = \pi_j(\mathbf{p},
\mathbf{x}, \xi; \Theta)$ is invertible in the demand shocks, $\xi(\Theta) = \pi^{-1}
(\mathbf{x}, \mathbf{s}; \Theta)$, and @berry1995 propose the GMM estimator built on the
empirical analogue of $E(\xi_j(\Theta) \mid \mathbf{z}_j) = 0$. With any zero share the
inversion fails. Dropping zeros invokes the weaker condition $E[\xi_j \mid \mathbf{z}_j,
d_j = 1] = 0$, valid only under selection-at-random ($\mathrm{cov}(\xi_j, \eta_j) = 0$);
under selection on unobservables it fails and GMM is inconsistent. Reduced-form prices
are modeled additively, $p_j = \Pi(\mathbf{r}_j; g) + \upsilon_j$ with supply shock
$\upsilon_j$, which is reasonable under cost-plus grocery pricing and supports a
**control-function** correction for endogenous prices.

The solution is **pairwise differencing**. Assume the demand shock decomposes through
the non-selection propensity $\mu_j$ as $\xi_j = \iota(\mu_j) + \zeta_j$ with $E[\zeta_j
\mid \mu_j, \mathbf{z}] = 0$ for a smooth increasing $\iota(\cdot)$. Then for a pair of
products with equal propensities the selection terms cancel, yielding the moment

$$
E[\xi_i - \xi_j \mid \mathbf{z}_i, \mathbf{z}_j, \mu_i = \mu_j, d_i = d_j = 1] = 0.
$$ {#eq-dube-moment}

In the homogeneous-logit case the inverted demand is $\ln(s_j/s_0) = \alpha p_j + x_j'
\beta + \iota(\mu_j) + \zeta_j$, and differencing two products with $\mu_i = \mu_j$
eliminates the unknown selection function,

$$
\ln(s_i/s_0) - \ln(s_j/s_0) = \alpha(p_i - p_j) + (x_i - x_j)'\beta + (\zeta_i - \zeta_j),
$$ {#eq-dube-diff}

so the conditional moment $E[\zeta_i - \zeta_j \mid \cdots] = E[\xi_i - \xi_j \mid \cdots]
= 0$ identifies $(\alpha, \beta)$ without ever logging a zero share. The price is that the
utility *intercept* is differenced away and cannot be identified.

Estimation is a three-step procedure. **Stage 1** estimates the pricing equation $p_j =
\Pi(\mathbf{r}_j; g) + \upsilon_j$ (known functional form) with a $\sqrt{n}$-consistent
$\hat{g}$, forms residuals $\hat{\upsilon}_j = p_j - \Pi(\mathbf{r}_j; \hat{g})$, and
builds the control vector $\mathbf{w}_j' = (\hat{\upsilon}_j, \tilde{\mathbf{w}}_j')$;
under the distributional exclusion restriction, $p_j$ and $x_j$ are excluded from
$\mathbf{w}_j$. **Stage 2** estimates the selection propensity nonparametrically by
Nadaraya–Watson kernel regression,

$$
\hat{\mu}(\mathbf{w}_j) = 1 - \frac{\sum_k d_k\, H\!\big(\tfrac{\mathbf{w}_k - \mathbf{w}_j}{b_n}\big)}
{\sum_k H\!\big(\tfrac{\mathbf{w}_k - \mathbf{w}_j}{b_n}\big)},
$$ {#eq-dube-nw}

with kernel $H$ and bandwidth $b_n$. **Stage 3** is the **pairwise-difference weighted
GMM** (PDWGMM) estimator. Writing $\Delta$ for the differencing operator over product
pairs, $g_{i,j}(\Theta) = \Delta \mathbf{z}_{i,j}\,\Delta \xi_{i,j}(\Theta)$, kernel
weights $\hat{w}_{i,j} = h_{n_d}^{-1}\kappa\big((\hat{\mu}_i - \hat{\mu}_j)/h_{n_d}\big)$
that upweight pairs with similar propensities, and a GMM weight matrix $\Phi$, the
estimator solves

$$
\Theta^{\text{PDWGMM}} = \arg\min_{(\Theta, \xi)}
\frac{2}{n(n-1)} \sum_{i=1}^{n-1} \sum_{j=i+1}^{n}
\hat{w}_{i,j}\, \mathbf{g}_{i,j}(\Theta)'\, \Phi\, \mathbf{g}_{i,j}(\Theta)
\quad \text{s.t.}\quad \mathbf{s} = \pi(\mathbf{p}, \mathbf{x}, \xi; \Theta),
$$ {#eq-dube-pdwgmm}

with the share constraint imposed only where shares are positive.

Monte Carlo evidence sharpens the stakes. Under **selection on unobserved demand**
(scenario a), PDWGMM recovers the mean parameters accurately while the drop-zero
estimator overstates the price coefficient by roughly 150%. When zeros are **missing at
random** (scenario b) both estimators are accurate, and PDWGMM loses no precision. With
**strictly positive shares** (scenario c) PDWGMM needs no first two stages and matches
the drop-zero estimates. In the ice-cream application—nonpremium brands (Breyers,
Dean's, Edy's, store brand), the 48-oz size that is 98.5% of sales, two stores,
Thursdays, January 2013–June 2014, 14,907 product-store observations—price endogeneity
is instrumented by unit cost and its square, and consideration is instrumented by UPC-
and brand-level feature advertising. Across six estimators (random-coefficients and
homogeneous logit, each with PDWGMM, drop-zero, and where applicable Heckman and impute
corrections), the imputed-share correction is unstable, PDWGMM's mean price coefficient
is significantly *lower* (demand more elastic) than drop-zero's, the hypothesis of equal
mean elasticities is rejected at 5%, and the Heckman two-step resembles drop-zero. The
methodological contribution is to accommodate zero shares within aggregate
discrete-choice demand—extensible to CES demand—rather than papering over them.

### Multi-Sided Markets {#sec-io-platforms}

Platforms—marketplaces, payment networks, operating systems—create value through
*cross-side network externalities*: a buyer's utility rises with the number of sellers,
and vice versa. @tan2021effects extend the discrete-choice framework to such settings by
letting a consumer's utility incorporate an externality term $\phi_i(\mathbf{x}^k)$ that
depends on participation on the platform's other side. This converts the demand system
into a fixed point—each side's demand depends on the other side's realized
participation—so entry and competition analysis must account for feedback that is absent
in single-sided markets. Adding a competitor or a new platform shifts not only the
direct substitution among options but also the externality each side confers on the
other, which is why competition and entry in multi-sided markets can have welfare
effects that single-sided intuition gets backwards.

## Vertical Relations and Negotiated Wholesale Prices {#sec-io-bargaining}

Every supply-side model so far has taken each firm's marginal cost $c_j$ as an
exogenous primitive to be recovered from a pricing first-order condition
(@eq-bertrand-foc). In a distribution channel that premise is wrong in a way that
matters. A grocery retailer's marginal cost of a national brand is a **wholesale price
negotiated bilaterally with the manufacturer**, and it moves when anything that
changes either side's outside option changes—an entrant, a patent expiry, a new
private label on the retailer's own shelf. Treating $w_j$ as a technological constant
throws away exactly the variation a channel study is about.

The textbook alternative is not much better. The classical channel model makes the
manufacturer a Stackelberg leader who posts $w_j$ and lets the retailer respond, which
produces double marginalization and assigns *all* channel power to the upstream firm
by assumption. Real negotiations are neither take-it-or-leave-it nor costlessly
efficient, and the position of a given pair on that spectrum is precisely what an
empirical channel study wants to measure. The **Nash-in-Nash** framework introduced in
passing above is the workhorse that makes channel power an estimated parameter rather
than a modelling assumption. @horn1988bilateral supplies the original bilateral-
monopoly formulation, @collardwexler2019nash supplies the non-cooperative
microfoundation, @draganska2010larger brings it into marketing to measure who captures
the channel margin in German coffee, and @ellickson2026private uses it to answer a
question the reduced-form literature could not: *how much* of a retailer's gain from
launching a private label comes from the private label's own margin, and how much from
the improved terms it wins on the national brands it still carries.

### Setup and Notation {#sec-io-nn-setup}

A single retailer sells $J$ differentiated products, one from each upstream supplier
$j = 1, \dots, J$, to a unit mass of consumers. Demand $q_j(\mathbf{p})$ comes from
the discrete-choice system estimated in @sec-io-berry: with mean utilities
$\delta_j$ (@eq-blp-mean-utility) and price coefficient $\alpha$, logit shares are
$q_j(\mathbf{p}) = \exp(\delta_j - \alpha p_j) / [1 + \sum_k \exp(\delta_k - \alpha p_k)]$.
Supplier $j$ produces at constant marginal cost $c_j$ and sells to the retailer at the
**linear wholesale price** $w_j$; the retailer incurs an additional handling cost
$c^R_j$ and sets the retail price $p_j$. Profits are

$$
\Pi^R(\mathbf{p}; \mathbf{w}) = \sum_{j} (p_j - w_j - c^R_j)\, q_j(\mathbf{p}),
\qquad
\Pi^S_j(\mathbf{p}; \mathbf{w}) = (w_j - c_j)\, q_j(\mathbf{p}).
$$ {#eq-nn-profits}

The retailer is a multiproduct monopolist over its own shelf, so its pricing
first-order condition is @eq-monopoly-foc applied to all $J$ products with $w_j + c^R_j$
in place of $c_j$. In matrix form, with $\Omega_{jk} \equiv -\partial q_k / \partial p_j$,
the retail equilibrium given $\mathbf{w}$ solves

$$
\mathbf{p}^*(\mathbf{w}) = \mathbf{w} + \mathbf{c}^R + \Omega^{-1}\mathbf{q}.
$$ {#eq-nn-retail-foc}

Two derived objects carry all the economics. The **gains from trade** on agreement $j$
are what each side would lose if that one negotiation broke down,

$$
G^R_j(\mathbf{w}) = \Pi^R\!\left(\mathbf{p}^*(\mathbf{w}); \mathbf{w}\right) - \Pi^R_{-j}(\mathbf{w}_{-j}),
\qquad
G^S_j(\mathbf{w}) = \Pi^S_j\!\left(\mathbf{p}^*(\mathbf{w}); \mathbf{w}\right) - \underline{\pi}_j,
$$ {#eq-nn-gft}

where $\Pi^R_{-j}$ is the retailer's profit when product $j$ is off the shelf and all
remaining retail prices are **re-optimised**, and $\underline{\pi}_j$ is the supplier's
outside option (normalised to zero when the retailer is its only route to market).
The second object is the **effective quantity response to a wholesale price**,

$$
\Gamma_j \equiv \frac{d\, q_j}{d\, w_j}
= \sum_{k} \frac{\partial q_j}{\partial p_k}\,\frac{\partial p^*_k(\mathbf{w})}{\partial w_j} \;<\; 0,
$$ {#eq-nn-gamma}

which is *not* the own-price slope $\partial q_j / \partial p_j$: it is that slope
composed with the retailer's pass-through matrix from @eq-cpt. Confusing the two is the
most common arithmetic error in applied bargaining work.

### Assumptions {#sec-io-nn-assumptions}

**Assumption 1 (Two-stage timing).** Wholesale prices are agreed in stage one; the
retailer then sets $\mathbf{p}$ in stage two taking $\mathbf{w}$ as given, so realised
prices satisfy @eq-nn-retail-foc. *Substantively:* contracts are signed before the
pricing period and are not renegotiated within it. Violated by trade deals struck
mid-period in response to a competitor's promotion.

**Assumption 2 (Nash-in-Nash / passive beliefs).** Each pair $(R, j)$ chooses $w_j$ to
maximise the generalised Nash product of the two sides' gains from trade, taking all
*other* agreements $\mathbf{w}_{-j}$ as given and unchanged:

$$
w_j^* \;=\; \arg\max_{w_j}\;
\left[G^R_j(\mathbf{w})\right]^{\lambda_j}
\left[G^S_j(\mathbf{w})\right]^{1-\lambda_j},
\qquad \lambda_j \in (0,1).
$$ {#eq-nn-objective}

*Substantively:* a buyer who is renegotiating with one supplier does not infer from
that conversation anything about what its rivals are being offered. @collardwexler2019nash
show this is the limit of a non-cooperative alternating-offers game with simultaneous,
non-observable bilateral negotiations. Violated when a supplier can credibly threaten
to punish the retailer in *other* categories, or when most-favoured-nation clauses link
the agreements.

**Assumption 3 (Disagreement removes only the focal product).** If $(R,j)$ fails to
agree, product $j$ alone leaves the shelf; every other agreement stands and the
retailer re-optimises $\mathbf{p}_{-j}$. *Substantively:* the threat is delisting one
SKU or brand, not the end of the trading relationship. Violated by all-or-nothing
category-captain arrangements and by settings where a breakdown triggers renegotiation
of the surviving contracts.

**Assumption 4 (Linear wholesale prices).** The transfer is $w_j q_j$ with no fixed
fee, slotting allowance, or quantity discount. *Substantively:* the per-unit invoice
price is the whole contract. This is the assumption most likely to fail in grocery, and
@sec-io-nn-failure shows what it costs.

### The Estimating Equation and Its Derivation {#sec-io-nn-derivation}

Take logs of @eq-nn-objective and differentiate in $w_j$:

$$
\lambda_j\, \frac{\partial G^R_j / \partial w_j}{G^R_j}
\;+\;
(1-\lambda_j)\, \frac{\partial G^S_j / \partial w_j}{G^S_j}
\;=\; 0 .
$$ {#eq-nn-foc}

The retailer's derivative collapses by an **envelope argument**. Total profit at the
optimum is $\Pi^R(\mathbf{p}^*(\mathbf{w}); \mathbf{w})$, so

$$
\frac{d}{d w_j}\Pi^R\!\left(\mathbf{p}^*(\mathbf{w}); \mathbf{w}\right)
= \underbrace{\sum_k \frac{\partial \Pi^R}{\partial p_k}\frac{\partial p^*_k}{\partial w_j}}_{=\,0 \text{ by @eq-nn-retail-foc}}
\;+\; \frac{\partial \Pi^R}{\partial w_j}
\;=\; -\,q_j .
$$ {#eq-nn-envelope}

Each term in the first sum vanishes because $\mathbf{p}^*$ solves the retailer's own
first-order conditions; the only surviving effect of a wholesale-price increase is the
mechanical one, an extra dollar paid on each of the $q_j$ units already sold. Because
the disagreement profit $\Pi^R_{-j}$ does not depend on $w_j$, @eq-nn-gft gives
$\partial G^R_j / \partial w_j = -q_j$ exactly.

The supplier's derivative does *not* collapse, because the supplier does not choose
$\mathbf{p}$: writing $\mu_j \equiv w_j - c_j$ for the wholesale margin,

$$
\frac{\partial G^S_j}{\partial w_j} = q_j + \mu_j \Gamma_j .
$$ {#eq-nn-supplier-deriv}

Substituting @eq-nn-envelope and @eq-nn-supplier-deriv into @eq-nn-foc, using
$G^S_j = \mu_j q_j$, and solving for the margin yields the **bargaining-augmented
markup equation**

$$
\mu_j \;=\; w_j - c_j \;=\;
\frac{q_j}{\;-\Gamma_j \;+\; \dfrac{\lambda_j}{1-\lambda_j}\cdot\dfrac{q_j^{\,2}}{G^R_j}\;} .
$$ {#eq-nn-markup}

This single equation is the supply-side analogue of @eq-bertrand-foc for a negotiated
input, and it is what makes the framework empirically useful: given demand estimates
(which deliver $q_j$ and $\Gamma_j$) and an observed wholesale price, it maps a value
of $\lambda_j$ into a marginal cost $c_j$, or a known $c_j$ into a $\lambda_j$.

Two limits confirm that @eq-nn-markup nests what the book already teaches. As
$\lambda_j \to 0$ the second term in the denominator vanishes and
$\mu_j = q_j / (-\Gamma_j)$, which is exactly the first-order condition of a
manufacturer that posts $w_j$ unilaterally and anticipates the retailer's response—the
Stackelberg channel model, with its double marginalization. As $\lambda_j \to 1$ the
second term diverges and $\mu_j \to 0$, so $w_j \to c_j$: the retailer buys at marginal
cost and the channel replicates the vertically integrated outcome. The bargaining
weight is the estimable parameter that interpolates between the two textbook cases.

The comparative static that drives @ellickson2026private falls straight out of
@eq-nn-markup. Anything that raises the retailer's disagreement profit $\Pi^R_{-j}$—a
private label that keeps earning when a national brand is delisted—*lowers* $G^R_j$,
which *raises* $q_j^2/G^R_j$, which *lowers* $\mu_j$. A better outside option buys
better terms on the products you still carry. This is the "indirect" channel the paper
isolates, and it is invisible to any model that treats $w_j$ as a cost parameter.

### Intuition {#sec-io-nn-intuition}

Strip the algebra away and the bargain is a fight over a fixed pie whose *size* the
per-unit price itself distorts. Raising $w_j$ transfers surplus to the supplier
one-for-one on the units sold (the $q_j$ term), but it also raises the retail price and
shrinks the units (the $\mu_j \Gamma_j$ term), so the supplier's own gain is
concave in $w_j$. The retailer's loss, by contrast, is *linear* in $w_j$ at the margin,
because the retailer re-optimises prices and the envelope theorem kills the second-order
effect. Bargaining weight decides where on that trade-off the pair lands. What makes
the outcome estimable is that the two sides' gains from trade are *observable functions
of the demand system*: $G^R_j$ is computable by simply re-solving the retail pricing
problem with product $j$ deleted, which is why credible identification of $\lambda_j$
demands variation that shifts outside options rather than variation that shifts costs.

### Simulation {#sec-io-nn-simulation}

The simulation below builds a channel with one retailer and two national-brand
suppliers, solves the Nash-in-Nash fixed point exactly, and asks three questions: does
$\lambda$ really interpolate between the two textbook channel models; what does an
analyst lose by imposing the posted-price model the book teaches when the truth is a
negotiation; and how does a private label change the terms on the national brands.

```{r io-nn-setup, message=FALSE, warning=FALSE}
set.seed(35201)
alpha <- 1.5   # price coefficient; unit mass of consumers, so profits are per-consumer

shares <- function(p, delta) { v <- exp(delta - alpha * p); v / (1 + sum(v)) }

# Retail stage: multiproduct-monopoly FOC, p = w + cR + Omega^{-1} q   (eq-nn-retail-foc)
retail_prices <- function(w, delta, cR, tol = 1e-12, maxit = 1000) {
  p <- w + cR + 1 / alpha
  for (i in seq_len(maxit)) {
    q  <- shares(p, delta)
    Om <- -alpha * outer(q, q); diag(Om) <- alpha * q * (1 - q)
    p_new <- w + cR + solve(Om, q)
    if (max(abs(p_new - p)) < tol) { p <- p_new; break }
    p <- 0.5 * p + 0.5 * p_new          # damped iteration
  }
  p
}
retail_profit <- function(w, delta, cR) {
  p <- retail_prices(w, delta, cR); q <- shares(p, delta)
  sum((p - w - cR) * q)
}
# Disagreement (Assumption 3): product j leaves, survivors are re-priced.
# reoptimise = FALSE is the shortcut examined in the failure analysis below.
retail_profit_drop <- function(j, w, delta, cR, reoptimise = TRUE) {
  keep <- setdiff(seq_along(w), j)
  if (reoptimise) return(retail_profit(w[keep], delta[keep], cR[keep]))
  p <- retail_prices(w, delta, cR)
  sum((p[keep] - w[keep] - cR[keep]) * shares(p[keep], delta[keep]))
}
# Gamma_j = dq_j/dw_j through retail pass-through   (eq-nn-gamma)
Gamma <- function(j, w, delta, cR, h = 1e-5) {
  wp <- w; wp[j] <- w[j] + h; wl <- w; wl[j] <- w[j] - h
  (shares(retail_prices(wp, delta, cR), delta)[j] -
   shares(retail_prices(wl, delta, cR), delta)[j]) / (2 * h)
}
# Log-Nash-product FOC   (eq-nn-foc)
nash_foc <- function(j, w, delta, cR, cS, lam, reoptimise = TRUE) {
  q  <- shares(retail_prices(w, delta, cR), delta)
  gR <- retail_profit(w, delta, cR) - retail_profit_drop(j, w, delta, cR, reoptimise)
  mu <- w[j] - cS[j]; gS <- mu * q[j]
  if (gR <= 1e-10 || gS <= 1e-10) return(NA_real_)
  lam[j] * (-q[j]) / gR + (1 - lam[j]) * (q[j] + mu * Gamma(j, w, delta, cR)) / gS
}
# Nash-in-Nash equilibrium: solve each bilateral FOC taking w_{-j} as given (Assumption 2)
nash_in_nash <- function(delta, cR, cS, lam, reoptimise = TRUE,
                         fixed = NULL, w_fixed = NULL, tol = 1e-10, maxit = 400) {
  w <- cS + 0.5
  if (!is.null(fixed)) w[fixed] <- w_fixed
  free <- setdiff(seq_along(w), fixed)
  for (it in seq_len(maxit)) {
    w_old <- w
    for (j in free) {
      f <- function(x) { ww <- w; ww[j] <- x; nash_foc(j, ww, delta, cR, cS, lam, reoptimise) }
      lo <- cS[j] + 1e-4; hi <- cS[j] + 8
      flo <- f(lo); fhi <- f(hi)
      while ((is.na(fhi) || flo * fhi > 0) && hi > lo + 1e-3) {
        hi <- lo + 0.6 * (hi - lo); fhi <- f(hi)
      }
      if (is.na(flo) || is.na(fhi) || flo * fhi > 0) next
      w[j] <- uniroot(f, c(lo, hi), tol = 1e-12)$root
    }
    if (max(abs(w - w_old)) < tol) break
  }
  w
}
cons_surplus <- function(p, delta) log(1 + sum(exp(delta - alpha * p))) / alpha

delta <- c(4.0, 3.6)      # mean utilities of the two national brands
cS    <- c(1.00, 0.90)    # supplier marginal costs
cR    <- c(0.20, 0.20)    # retailer handling costs
```

The envelope result @eq-nn-envelope is not an approximation, and it is cheap to verify
numerically before relying on it.

```{r io-nn-envelope, message=FALSE, warning=FALSE}
w5 <- nash_in_nash(delta, cR, cS, rep(0.5, 2)); h <- 1e-5
wp <- w5; wp[1] <- w5[1] + h
wl <- w5; wl[1] <- w5[1] - h
c(numeric = (retail_profit(wp, delta, cR) - retail_profit(wl, delta, cR)) / (2 * h),
  minus_q = -shares(retail_prices(w5, delta, cR), delta)[1])
```

**Does $\lambda$ interpolate between the textbook models?** @tbl-nn-lambda traces the
equilibrium across bargaining weights, with the vertically integrated benchmark in the
last row.

```{r io-nn-lambda, message=FALSE, warning=FALSE}
#| label: tbl-nn-lambda
#| tbl-cap: "Nash-in-Nash equilibrium as the retailer's bargaining weight rises. The lambda = 0.01 row reproduces the manufacturer-Stackelberg channel; the final row is the vertically integrated benchmark. Channel profit is the sum of retailer and supplier profit, and the gap between it and the integrated benchmark is the double-marginalization loss."
tab_lambda <- do.call(rbind, lapply(c(0.01, 0.25, 0.50, 0.75, 0.99), function(L) {
  w <- nash_in_nash(delta, cR, cS, rep(L, 2))
  p <- retail_prices(w, delta, cR); q <- shares(p, delta)
  data.frame(lambda = L, w_NB1 = w[1], p_NB1 = p[1],
             pi_R = sum((p - w - cR) * q), pi_S = sum((w - cS) * q),
             channel = sum((p - cS - cR) * q), CS = cons_surplus(p, delta))
}))
p_vi <- retail_prices(cS, delta, cR); q_vi <- shares(p_vi, delta)
tab_lambda <- rbind(tab_lambda,
  data.frame(lambda = NA, w_NB1 = cS[1], p_NB1 = p_vi[1],
             pi_R = sum((p_vi - cS - cR) * q_vi), pi_S = 0,
             channel = sum((p_vi - cS - cR) * q_vi), CS = cons_surplus(p_vi, delta)))
knitr::kable(round(tab_lambda, 4), align = "r")
```

The two limits derived above appear exactly. At $\lambda = 0.01$ the wholesale price is
2.010 and channel profit 0.855; at $\lambda = 0.99$ the wholesale price is 1.009 against
a marginal cost of 1.00 and channel profit is 0.949, indistinguishable from the
integrated benchmark. Roughly 9.9% of attainable channel profit is destroyed by conduct
alone at the manufacturer-power extreme, and consumer surplus is 41% lower there
(0.348 versus 0.590) even though costs, preferences, and the number of products are
identical throughout. Channel power is not a distributional detail; it is an efficiency
parameter.

**What does imposing the posted-price model cost?** This is the comparison that
matters for practice, because the posted-price model is what @eq-bertrand-foc delivers
if an analyst treats $w_j$ as an observed cost and never models the negotiation.
Inverting @eq-nn-markup at an *assumed* $\lambda_a$ recovers a supplier marginal cost;
@tbl-nn-misspec sets $\lambda_a = 0$ and varies the truth.

```{r io-nn-misspec, message=FALSE, warning=FALSE}
#| label: tbl-nn-misspec
#| tbl-cap: "Recovering supplier marginal cost for national brand 1 under an assumed bargaining weight of zero (the posted-price channel model) when the truth is a negotiation. dp_true and dp_assumed are the retail price change predicted by the true and the misspecified model for a 25% supplier cost shock."
recover_c <- function(j, w, delta, cR, lam_a) {           # invert eq-nn-markup for c_j
  q  <- shares(retail_prices(w, delta, cR), delta)[j]
  G  <- Gamma(j, w, delta, cR)
  gR <- retail_profit(w, delta, cR) - retail_profit_drop(j, w, delta, cR)
  w[j] - q / (-G + (lam_a / (1 - lam_a)) * q^2 / gR)
}
tab_mis <- do.call(rbind, lapply(c(0.10, 0.30, 0.50, 0.70, 0.90), function(Lt) {
  w  <- nash_in_nash(delta, cR, cS, rep(Lt, 2))
  cS_hat <- sapply(1:2, function(j) recover_c(j, w, delta, cR, 1e-6))   # assume lambda = 0
  shock  <- 1.25
  p_true <- retail_prices(nash_in_nash(delta, cR, cS * shock, rep(Lt, 2)), delta, cR)[1]
  w_base <- nash_in_nash(delta, cR, cS_hat, rep(1e-6, 2))
  p_mis  <- retail_prices(nash_in_nash(delta, cR, cS_hat * shock, rep(1e-6, 2)), delta, cR)[1]
  data.frame(lambda_true = Lt, c_true = cS[1], c_hat = cS_hat[1],
             bias_pct = 100 * (cS_hat[1] - cS[1]) / cS[1],
             dp_true = p_true - retail_prices(w, delta, cR)[1],
             dp_assumed = p_mis - retail_prices(w_base, delta, cR)[1])
}))
knitr::kable(round(tab_mis, 4), align = "r")
c(RMSE_assumed_lambda0 = sqrt(mean((tab_mis$c_hat - tab_mis$c_true)^2)))
```

The damage is severe and systematic. At a true $\lambda = 0.5$ the recovered supplier
marginal cost is 0.353 against a truth of 1.00, a 64.7% understatement; the RMSE across
the grid is 0.733 on a cost of 1.00, against exactly zero when the correct $\lambda$ is
imposed. The mechanism is transparent from @eq-nn-markup: a negotiated wholesale price
is *lower* than a posted one, and a model that insists the supplier was a monopolist
must explain the low price by attributing a low cost to the supplier. At
$\lambda_{\text{true}} = 0.9$ the recovered cost is **negative** ($-0.137$), which is a
useful diagnostic in real work—negative implied costs in a channel study are usually
evidence of imposed conduct that is too manufacturer-favourable, not of subsidised
production.

The counterfactual columns show why this is not merely a labelling problem. For a 25%
supplier cost shock at $\lambda_{\text{true}} = 0.5$, the true model predicts national
brand 1's retail price rises by 0.124; the misspecified model predicts 0.037, understating
pass-through by a factor of 3.4. At $\lambda_{\text{true}} = 0.9$ the misspecified model
predicts a price *decrease* of 0.013—the sign is wrong—because a 25% increase applied to
a negative recovered cost moves it further below zero. Conduct misspecification does not
wash out of a counterfactual; it propagates through it.

**What does a private label do?** @ellickson2026private decompose the retailer's gain
from a private label into a direct margin effect and an indirect bargaining effect. The
decomposition is reproducible here by solving the three-product channel twice: once with
the national-brand wholesale prices frozen at their pre-entry values (direct effect only)
and once with everything renegotiated (total effect).

```{r io-nn-privatelabel, message=FALSE, warning=FALSE}
#| label: tbl-nn-privatelabel
#| tbl-cap: "Private-label entry with national-brand bargaining weight fixed at 0.5. dw_NB1 is the change in national brand 1's negotiated wholesale price relative to the no-private-label equilibrium. The indirect gain is the part of the retailer's profit increase that comes from renegotiated national-brand terms rather than from the private label's own margin."
lamNB <- 0.5
w0    <- nash_in_nash(delta, cR, cS, rep(lamNB, 2))
p0    <- retail_prices(w0, delta, cR)
base  <- sum((p0 - w0 - cR) * shares(p0, delta))

delta3 <- c(delta, 2.8); cS3 <- c(cS, 0.70); cR3 <- c(cR, 0.20)
tab_pl <- do.call(rbind, lapply(c(0.30, 0.50, 0.70, 0.90), function(Lpl) {
  lam3 <- c(lamNB, lamNB, Lpl)
  w_d  <- nash_in_nash(delta3, cR3, cS3, lam3, fixed = 1:2, w_fixed = w0)  # NB terms frozen
  p_d  <- retail_prices(w_d, delta3, cR3)
  pi_d <- sum((p_d - w_d - cR3) * shares(p_d, delta3))
  w_f  <- nash_in_nash(delta3, cR3, cS3, lam3)                            # all renegotiated
  p_f  <- retail_prices(w_f, delta3, cR3)
  pi_f <- sum((p_f - w_f - cR3) * shares(p_f, delta3))
  data.frame(lambda_PL = Lpl, w_PL = w_f[3], w_NB1 = w_f[1], dw_NB1 = w_f[1] - w0[1],
             gain_total = pi_f - base, gain_direct = pi_d - base,
             gain_indirect = pi_f - pi_d,
             indirect_pct = 100 * (pi_f - pi_d) / (pi_f - base))
}))
knitr::kable(round(tab_pl, 4), align = "r")
```

Both of @ellickson2026private's headline patterns reproduce. First, entry improves the
terms on the products the retailer still carries: national brand 1's wholesale price
falls by 0.041 at $\lambda_{PL} = 0.5$, purely because the retailer now loses less when
it threatens to delist. Second, the retailer's total gain rises monotonically in its
bargaining position against the *private-label supplier*—from 0.089 at
$\lambda_{PL} = 0.30$ to 0.158 at $\lambda_{PL} = 0.90$, a 78% difference driven entirely
by supply-side arrangements that a demand-side analysis cannot see. The indirect channel
is a minority of the gain but not a rounding error: 14.3% of the retailer's profit
increase at $\lambda_{PL} = 0.5$, or 0.0154 on a base of 0.688. Its *share* falls as
$\lambda_{PL}$ rises even though its level rises, because a cheaper private label grows
the direct margin faster than it grows the threat value.

### When the Assumptions Fail {#sec-io-nn-failure}

Two of the four assumptions are worth stress-testing, and they fail very differently.

```{r io-nn-failure, message=FALSE, warning=FALSE}
#| label: tbl-nn-failure
#| tbl-cap: "Two assumption failures. Panel A: the retailer's disagreement payoff computed without re-pricing the surviving products (Assumption 3 shortcut). Panel B: a share phi of the supplier's rent is paid as a fixed fee rather than a per-unit margin, violating Assumption 4; the linear-wholesale-price estimator is applied anyway. True lambda is 0.5 in Panel B."
recover_lambda <- function(j, w, delta, cR, cS, reoptimise = TRUE) {
  q  <- shares(retail_prices(w, delta, cR), delta)[j]
  G  <- Gamma(j, w, delta, cR); mu <- w[j] - cS[j]
  gR <- retail_profit(w, delta, cR) - retail_profit_drop(j, w, delta, cR, reoptimise)
  A  <- -q / gR; B <- (q + mu * G) / (mu * q)
  B / (B - A)
}
panelA <- do.call(rbind, lapply(c(0.10, 0.30, 0.50, 0.70, 0.90), function(Lt) {
  w <- nash_in_nash(delta, cR, cS, rep(Lt, 2))
  data.frame(panel = "A: no re-pricing", setting = Lt, lambda_true = Lt,
             lambda_hat = recover_lambda(1, w, delta, cR, cS, reoptimise = FALSE))
}))
mu_star <- nash_in_nash(delta, cR, cS, rep(0.5, 2)) - cS
panelB <- do.call(rbind, lapply(c(0, 0.25, 0.50, 0.75, 0.95), function(phi) {
  w_obs <- cS + (1 - phi) * mu_star     # share phi of the rent moved into a fixed fee
  data.frame(panel = "B: two-part tariff", setting = phi, lambda_true = 0.5,
             lambda_hat = recover_lambda(1, w_obs, delta, cR, cS))
}))
out <- rbind(panelA, panelB)
out$bias <- out$lambda_hat - out$lambda_true
out[, 2:5] <- round(out[, 2:5], 4)
knitr::kable(out, align = "r", row.names = FALSE)
c(RMSE_panelA = sqrt(mean((panelA$lambda_hat - panelA$lambda_true)^2)),
  RMSE_panelB = sqrt(mean((panelB$lambda_hat - panelB$lambda_true)^2)))
```

**Assumption 3 bends.** Computing $\Pi^R_{-j}$ without re-pricing the survivors—a
shortcut that saves one equilibrium solve per negotiation and is common in applied
code—biases $\hat\lambda$ upward by 0.004 to 0.016, an RMSE of 0.0117 on a parameter
bounded in $[0,1]$. The direction is predictable: skipping the re-optimisation
understates the disagreement profit, overstates the retailer's gains from trade, and so
attributes more of the observed wholesale discount to retailer power. The magnitude is
small enough that the shortcut is defensible in a first pass and indefensible in a
published estimate.

**Assumption 4 breaks.** Moving the supplier's rent from a per-unit margin into a fixed
fee leaves the *total* transfer and the split of surplus untouched, but drives the
observed per-unit price toward marginal cost. The linear-wholesale-price estimator reads
that as retailer power and nothing else: at $\phi = 0.95$ it returns
$\hat\lambda = 0.973$ against a truth of 0.5, a bias of 0.473 on a $[0,1]$ parameter, and
in the limit $\phi \to 1$ it returns $\hat\lambda \to 1$ for *every* true split. This is
not a precision problem that more data fixes—it is a failure of identification, and it is
why credible applications either observe contract terms directly or argue explicitly that
linear pricing describes the institution.

**What the simulation demonstrates:** the bargaining weight is the parameter that decides
whether a channel behaves like a Stackelberg manufacturer or a vertically integrated
firm; imposing the former when the truth lies between them understates supplier marginal
costs by up to 114%—far enough to drive the recovered cost below zero—and gets the sign
of a cost-shock counterfactual wrong; a private
label raises retailer profit both directly and by cheapening the national brands the
retailer still stocks; and the whole apparatus rests on linear pricing, which is the one
assumption whose failure destroys identification rather than merely degrading precision.

### Replication and Practical Guidance {#sec-io-nn-practice}

**Replication.** @ellickson2026private's estimates cannot be reproduced here. The paper
combines retailer scanner data with proprietary information on the *identity* of
private-label manufacturers—the feature that lets it measure disagreement payoffs
directly—and neither input is public. The simulation above reproduces the paper's
qualitative decomposition and comparative statics on synthetic data; the magnitudes are
not comparable to its published tables and should not be read as such.
@draganska2010larger use commercially available German coffee scanner data and report
bargaining weights that vary substantially across manufacturer–retailer pairs, which is
the closest public-data analogue.

**Estimation in practice.** The pipeline is nested three deep: a retail-equilibrium
solve (@eq-nn-retail-foc) inside a bargaining fixed point (@eq-nn-foc) inside the GMM
objective that also estimates demand. Cost grows with the number of *negotiations*
rather than products, because every $G^R_j$ requires an additional retail solve with
product $j$ deleted—so a category with 40 SKUs from 12 suppliers needs 13 equilibrium
solves per bargaining iteration, not 41. Estimate demand first and hold it fixed;
there is no R package for the supply side, so the bargaining solver is written by hand
(the ~60 lines above are the whole method), while demand estimation goes through
`PyBLP` in Python or a hand-rolled GMM routine in R.

**Identification.** $\lambda_j$ and $c_j$ are not separately identified from a single
cross-section of wholesale prices: @eq-nn-markup is one equation in two unknowns per
product. Applications break the tie with variation that shifts *disagreement payoffs*
while holding costs fixed—@ellickson2026private uses a patent expiry that coincided with
private-label entry, @grennan2013price uses staggered adoption of a competing medical
device, @crawford2012welfare uses channel carriage disputes. Cost-shifting instruments,
the workhorse of @sec-io-berry, are the wrong instrument here: they identify demand, not
conduct.

**Pitfalls.**

1. **Using $\partial q_j/\partial p_j$ where @eq-nn-gamma calls for $\Gamma_j$.** The
   correct object composes the demand slope with retail pass-through and is
   substantially smaller in magnitude; substituting the raw slope inflates the implied
   supplier markup.
2. **Fixed fees and slotting allowances.** See @sec-io-nn-failure. If the institution
   uses two-part tariffs, the per-unit price is close to marginal cost by construction
   and carries almost no information about the split.
3. **Treating a private label as just another product.** Its effect on the shelf runs
   through the *threat point*, not only through substitution. A demand-side model that
   adds a product and holds $\mathbf{w}$ fixed recovers 86% of the retailer's gain in
   @tbl-nn-privatelabel and misses the rest entirely.
4. **Reading $\lambda_j$ as a preference or a personality.** It is a reduced-form summary
   of an unmodelled negotiation protocol. @collardwexler2019nash show it maps to the
   probability of making the offer in an alternating-offers game, which means it can
   shift with the institutional environment and should not be held fixed across a
   counterfactual that changes that environment.

## Structural Applications in Marketing {#sec-io-applications}

The discrete-choice and dynamic-pricing apparatus underwrites a body of empirical
marketing work that estimates structural primitives and runs managerial
counterfactuals. Four applications illustrate the range.

**Dynamic pricing of new products.** @cosguner2022dynamic confront a limitation of the
Generalized Bass Model (GBM): although the Bass diffusion model is the standard new-
product sales forecaster, the GBM extension responds only to *percentage* changes in
marketing variables, not their levels, so it cannot prescribe an optimal *launch
price*. The authors propose utility-based generalizations—the Bass–Gumbel and
Bass–Logit diffusion models (BGDM, BLDM)—that ground diffusion in discrete choice and
therefore prescribe both the launch price and the post-launch price path. Across four
product categories both models outperform the GBM in predicting new-product sales, the
paper supplies methods to estimate them when no prior sales data exist, contrasts the
BLDM's optimal pricing policy with the GBM's across parameter ranges, and offers a
computationally light routine—extensible to competitive, multiproduct settings—for
managers to set dynamic pricing and advertising.

**Behavioral price discrimination in channels.** @cosguner2017behavioral study
behavioral price discrimination (BPD) when consumers face switching costs, so price
elasticity varies *over time* and not only across customers as in prior work. Estimating
a dynamic pricing model on cola-category data, they find that retailers profit more by
*delegating* coupon customization and data analytics to manufacturers than by managing
coupons themselves, that retailers can profit substantially by selling their customer
databases to manufacturers, and that—despite gaining data access—manufacturers earn
*lower* profits, evidence that customer information confers channel power on the
retailer. Strikingly, conditioning BPD on only the *most recent* purchase is a simple,
cheap policy that materially shifts profitability, so customer information drives channel
dynamics even when used coarsely.

**Email contact policy.** @zhang2017dynamically address an understudied but profitable
channel by modeling customer email-open behavior jointly with purchases in a unified
hidden-Markov and copula framework, using data from a U.S. home-improvement retailer.
Two findings matter for practice: active email readers are not necessarily active
buyers (and vice versa), and email *volume* affects short- and long-run profitability
*non-linearly*. Their decision-support system locates an interior optimum—around seven
emails—where deviating is costly: sending four rather than seven loses about 32% of
per-customer lifetime profit, and sending ten loses about 16%. Over- and under-
communicating both destroy value, so frequency must be tuned, not maximized.

**Online word of mouth and sales.** @chintagunta2010effects estimate the effect of
nationwide online user reviews on *local* box-office performance at the
designated-market-area (DMA) level, confronting three identification threats: spatial
aggregation across heterogeneous markets, serial correlation from staged film rollout,
and other unobserved serial correlation. Using daily ticket sales for 148 of 874 films
over 16 months with Yahoo! Movies ratings, and instrumenting ratings with the staged
release across markets, they find that review **valence** (sentiment), not volume,
drives local sales—reversing the volume-dominates conclusion of studies that use
*aggregated national* data, which their disaggregated design shows to be an artifact of
aggregation. The rollout-based instrument for review endogeneity generalizes to any
industry with sequential product releases.

These applications share a structural logic that distinguishes IO-style marketing from
reduced-form analysis: each estimates preference and cost primitives that are invariant
to the policy being studied, then re-solves the model under the counterfactual—a new
price path, a different coupon regime, an altered email cadence, a changed information
environment—to compute the managerially relevant outcome.

## Key Takeaways

- Industrial organization supplies the structural primitives—demand systems,
  equilibrium concepts, and identification arguments—that let an analyst run policy
  counterfactuals, in contrast to reduced-form analytics that describe correlations.
- The classical, continuous-quantity approach yields the Slutsky/Roy machinery, the CES
  and quasi-linear special cases, and clean diversion ratios (@eq-diversion); its
  symmetry of cross-effects (@eq-slutsky) is the empirical fingerprint of utility
  maximization, and the homothetic-surplus family of @armstrong2018multiproduct makes
  multiproduct pricing tractable by reducing it to a composite-price problem.
- Vertical differentiation makes quality explicit in demand (@eq-vertical-demand1) and
  shows why a multiproduct firm spaces its quality tiers apart to soften internal
  competition.
- Discrete-choice models recover the full elasticity matrix from a few preference
  parameters and predict demand for new products, but the multinomial logit's IIA
  restriction (@eq-35-iia) forces unrealistic substitution, motivating random coefficients,
  nested logit, and probit.
- Price endogeneity from unobserved characteristics is the central estimation obstacle;
  @berry1994estimating solves it by inverting market shares to recover mean utilities
  (@eq-blp-mean-utility) that are linear in the demand error and amenable to
  instrumental variables.
- Zero-valued market shares break the share inversion and, when driven by selection on
  unobservables, bias the drop-zero estimator severely; the pairwise-difference GMM of
  @dube2021random (@eq-dube-pdwgmm) restores consistency by differencing out the
  selection function.
- Structural marketing applications—dynamic new-product pricing, behavioral price
  discrimination in channels, email contact policy, and online word of mouth—estimate
  policy-invariant primitives and re-solve the model under counterfactuals to answer
  managerial questions.

## Further Reading

The structural-econometrics toolkit underlying this chapter—GMM, simulation estimation,
and the marketing–finance event-study machinery—is developed in @sec-marketing-finance.
The classical demand foundations connect to the welfare and pricing material throughout
the pricing chapters; the discrete-choice estimators here are the demand-side complement
to the supply-side conduct models. For the homothetic-surplus characterization and its
pricing implications see @armstrong2018multiproduct; for the canonical share-inversion
estimator and its equilibrium extensions, @berry1994estimating, @berry1995,
@berry1995automobile, and @berry2004differentiated; for the choice-theoretic foundations
of market research, @mcfadden1986choice and @mcfadden1986method; and for the frontier
treatment of zero shares and platform competition, @dube2021random and @tan2021effects.
The structural marketing applications—@cosguner2022dynamic, @cosguner2017behavioral,
@zhang2017dynamically, and @chintagunta2010effects—are entry points to the empirical
literature that puts these models to work.

## Full-semester seminar reading map {#sec-io-seminar-reading-map}

The discrete-choice and structural material developed above is, in most doctoral
programs, only the first movement of a full-semester sequence in structural
econometrics / empirical industrial organization (EIO). This section appends a
13–14 week seminar reading map that situates the chapter's BLP-style demand model
inside the broader EIO pipeline—demand, then supply and conduct, then
counterfactuals, taught first for static markets and then again for dynamic ones.
Every reading carries a Crossref-verified version-of-record DOI (copied verbatim from
the source syllabus, verified 2026-06-21); where a work has no entry in this book's
bibliography it is named in full rather than cited by key. Foundational canon is marked
**[F]** and frontier / actively-contested methods **[R]**.

### The semester arc {#sec-io-semester-arc}

The structural-IO seminar is organized as a single methodological pipeline taught
twice—once for *static* markets and once for *dynamic* ones. The spine is the
demand–supply equilibrium model of a differentiated-products oligopoly: students first
learn to estimate **demand** for differentiated products (logit, nested logit, and the
random-coefficients/BLP workhorse), then to layer a **supply side** (Bertrand–Nash
pricing, markups, conduct) on top, and finally to use the estimated structural
parameters to run **counterfactuals** (mergers, new goods, price regulation, taxes).
This static arc—roughly the first half of the term—is where the field's identification
arguments, instruments, and estimation algorithms (GMM, the BLP nested fixed point, and
MPEC) are taught most carefully, because every later model reuses them.

The second half generalizes the same logic to settings where agents are
**forward-looking**. Consumers become dynamic when goods are durable or storable; firms
become dynamic when they invest, enter, or exit. The methodological payload here is
**dynamic discrete choice**—Rust's nested fixed point and the conditional-choice-
probability (CCP) shortcut of Hotz–Miller—extended to dynamic games via two-step
estimators. The seminar then surveys the major application areas that have organized the
field's recent agenda: entry and market structure, auctions, two-sided/platform markets,
and consumer search and consideration. The term typically closes on the active frontier
where machine learning meets structural estimation (heterogeneous treatment effects,
text-as-data, high-dimensional demand) and on a recurring debate that runs through the
whole course: structural vs. reduced-form / "credible" design-based inference.
Pedagogically, most programs intertwine three threads every week: the economic *model*
and its equilibrium concept; the *econometrics* of what is identified, from what
variation, under what exclusion restrictions; and *computation*—students are expected to
code estimators (increasingly in Python with PyBLP, or in Julia/MATLAB) and replicate a
canonical paper. The marketing version of the seminar leans harder on individual-level
scanner/CRM data, consumer heterogeneity, micro-moments, and managerial counterfactuals
than the economics version, which leans toward antitrust and policy.

### Week 1 — Foundations: the structural approach and the modeling cycle {#sec-io-week01}

**Topic:** What the structural approach buys and what it costs, and the demand–supply–
counterfactual loop that organizes the rest of the term.

**Subtopics:** What "structural" buys you (welfare, counterfactuals, extrapolation) vs.
its costs (assumptions, computation); the demand–supply–counterfactual loop; the
structural vs. reduced-form debate.

**Methods:** Reading a structural paper; mapping primitives → estimating equations →
counterfactual.

**Key readings:**

- **[F]** @berry1994discretechoice — Berry, S. (1994), "Estimating Discrete-Choice Models
  of Product Differentiation," *RAND Journal of Economics* 25(2): 242–262.
  [doi:10.2307/2555829](https://doi.org/10.2307/2555829). The mean-utility inversion that
  makes aggregate differentiated-products demand estimable—the conceptual on-ramp to the
  whole field.
- **[F]** Nevo, A. (2000), "A Practitioner's Guide to Estimation of Random-Coefficients
  Logit Models of Demand," *Journal of Economics & Management Strategy* 9(4): 513–548.
  [doi:10.1162/105864000567954](https://doi.org/10.1162/105864000567954). The teaching
  reference that turns BLP into runnable code; a near-universal Week 1–3 reading.

**Debate:** When are structural assumptions worth their cost vs. a clean natural
experiment?

### Week 2 — Discrete-choice demand I: logit and nested logit {#sec-io-week02}

**Topic:** Random-utility demand and the substitution patterns that logit and nested
logit can and cannot represent.

**Subtopics:** Random utility; IIA and its failures; the logit "red bus/blue bus"
problem; nested logit and the substitution-pattern fix; aggregate vs. individual data.

**Methods:** Deriving choice probabilities; computing own- and cross-price elasticities;
handling the "outside good."

**Key readings:**

- **[F]** @berry1994estimating — Berry, S. (1994), *RAND J. Econ.* 25(2): 242–262.
  [doi:10.2307/2555829](https://doi.org/10.2307/2555829). (Re-used) the mean-utility
  inversion specializes cleanly to logit and nested logit.

**Debate:** Is IIA an acceptable approximation, or does it doom logit for policy work?

### Week 3 — Discrete-choice demand II: random coefficients / BLP {#sec-io-week03}

**Topic:** The random-coefficients demand-and-supply workhorse and the endogeneity
problem it solves.

**Subtopics:** Unobserved product characteristics (ξ) and price endogeneity; random
coefficients for flexible substitution; the contraction mapping; BLP instruments.

**Methods:** GMM with the BLP nested fixed point; constructing demand-side instruments;
micro-moments.

**Key readings:**

- **[F]** @berry1995automobile — Berry, S., Levinsohn, J., & Pakes, A. (1995),
  "Automobile Prices in Market Equilibrium," *Econometrica* 63(4): 841–890.
  [doi:10.2307/2171802](https://doi.org/10.2307/2171802). The foundational
  random-coefficients demand-and-supply paper; the workhorse of the field.
- **[F]** Nevo, A. (2001), "Measuring Market Power in the Ready-to-Eat Cereal Industry,"
  *Econometrica* 69(2): 307–342.
  [doi:10.1111/1468-0262.00194](https://doi.org/10.1111/1468-0262.00194). The canonical
  applied BLP study; introduces brand-specific intercepts and panel structure to control
  for ξ.
- **[R]** @berry2004differentiated — Berry, S., Levinsohn, J., & Pakes, A. (2004),
  "Differentiated Products Demand Systems from a Combination of Micro and Macro Data: The
  New Car Market," *Journal of Political Economy* 112(1): 68–105.
  [doi:10.1086/379939](https://doi.org/10.1086/379939). Shows how consumer-level
  micro-moments sharpen identification of heterogeneity; the template for modern marketing
  applications.

**Debate:** What variation actually identifies the random coefficients, and how fragile
is it to weak instruments?

### Week 4 — Identification & instruments in demand estimation {#sec-io-week04}

**Topic:** What differentiated-products demand identifies nonparametrically, and which
instruments deliver it.

**Subtopics:** Price endogeneity; cost-shifters, Hausman, and BLP-type instruments;
nonparametric identification of demand; instrument relevance and validity.

**Methods:** Exclusion restrictions; testing instrument strength; the "connected
substitutes" condition.

**Key readings:**

- **[R]** Berry, S. & Haile, P. (2014), "Identification in Differentiated Products Markets
  Using Market Level Data," *Econometrica* 82(5): 1749–1797.
  [doi:10.3982/ECTA9027](https://doi.org/10.3982/ECTA9027). Establishes nonparametric
  identification of differentiated-products demand and supply; reframes BLP's parametric
  choices as identification, not just estimation.

**Debate:** Are the standard instruments (rival characteristics, cost shifters) credible,
and what do they identify nonparametrically?

### Week 5 — Estimation & computation: GMM, NFP, and MPEC {#sec-io-week05}

**Topic:** How the BLP estimator is actually computed, and how much numerical choices move
published results.

**Subtopics:** GMM objective and weighting; the nested fixed-point (NFP) algorithm; MPEC
(constrained optimization) as an alternative; numerical pitfalls (tight inner loops,
local minima); modern software.

**Methods:** Coding the contraction; MPEC vs. NFP; optimal instruments; supply-side
moments; PyBLP.

**Key readings:**

- **[R]** Dubé, J.-P., Fox, J. T., & Su, C.-L. (2012), "Improving the Numerical
  Performance of Static and Dynamic Aggregate Discrete Choice Random Coefficients Demand
  Estimation," *Econometrica* 80(5): 2231–2267.
  [doi:10.3982/ecta8585](https://doi.org/10.3982/ecta8585). Reformulates BLP estimation as
  MPEC, showing the NFP inner-loop tolerance can bias estimates; reshaped how the field
  computes BLP.
- **[R]** @conlon2020 — Conlon, C. & Gortmaker, J. (2020), "Best Practices for
  Differentiated Products Demand Estimation with PyBLP," *RAND Journal of Economics* 51(4):
  1108–1161. [doi:10.1111/1756-2171.12352](https://doi.org/10.1111/1756-2171.12352). The
  current computational standard and the software most students now use; consolidates two
  decades of practical lessons.

**Debate:** How much do numerical choices (tolerances, optimizers, starting values) drive
published BLP results?

### Week 6 — Supply, conduct, and markups {#sec-io-week06}

**Topic:** Adding the supply side—recovering marginal costs and markups, and testing how
firms compete.

**Subtopics:** Bertrand–Nash pricing with differentiated products; recovering marginal
costs from the first-order conditions; testing conduct (Cournot vs. Bertrand vs.
collusion); identification of the conduct parameter.

**Methods:** Inverting FOCs for markups/marginal costs; ownership matrices; conduct
tests.

**Key readings:**

- **[F]** Bresnahan, T. F. (1982), "The Oligopoly Solution Concept Is Identified,"
  *Economics Letters* 10(1–2): 87–92.
  [doi:10.1016/0165-1765(82)90121-5](https://doi.org/10.1016/0165-1765(82)90121-5). The
  original result that demand rotations identify the conduct parameter; foundation of the
  "new empirical IO."
- **[F]** Nevo, A. (1998), "Identification of the Oligopoly Solution Concept in a
  Differentiated-Products Industry," *Economics Letters* 59(3): 391–395.
  [doi:10.1016/s0165-1765(98)00061-5](https://doi.org/10.1016/s0165-1765(98)00061-5).
  Extends conduct identification to differentiated products; the bridge from Bresnahan to
  BLP-style supply.
- **[F]** Nevo, A. (2001), *Econometrica* 69(2): 307–342.
  [doi:10.1111/1468-0262.00194](https://doi.org/10.1111/1468-0262.00194). (Re-used)
  recovers cereal markups and decomposes them into cost, markup, and portfolio effects.

**Debate:** Can conduct be credibly tested, or must it be assumed? (the "testing vs.
calibrating conduct" problem)

### Week 7 — Counterfactuals & merger / policy simulation {#sec-io-week07}

**Topic:** Using estimated demand and supply to recompute equilibria under mergers, new
goods, and policy changes.

**Subtopics:** Recomputing equilibrium prices under new ownership (mergers), new goods,
taxes/regulation; welfare and consumer surplus from logit/BLP; passthrough; the role of
supply-side assumptions in counterfactuals.

**Methods:** Solving for post-merger equilibrium; computing compensating variation;
sensitivity of counterfactuals to conduct.

**Key readings:**

- **[F]** Nevo, A. (2000), *J. Econ. & Mgmt. Strategy* 9(4): 513–548.
  [doi:10.1162/105864000567954](https://doi.org/10.1162/105864000567954). (Re-used) lays
  out the merger-simulation mechanics that follow directly from estimated demand and FOCs.
- **[F]** @petrin2002minivan — Petrin, A. (2002), "Quantifying the Benefits of New
  Products: The Case of the Minivan," *Journal of Political Economy* 110(4): 705–729.
  [doi:10.1086/340779](https://doi.org/10.1086/340779). The canonical "value of a new good"
  counterfactual; shows how micro-data discipline the welfare numbers.

**Debate:** How much do counterfactuals depend on un-tested supply assumptions and
out-of-sample extrapolation?

### Week 8 — Dynamic demand: durable and storable goods {#sec-io-week08}

**Topic:** Forward-looking consumers and the bias from ignoring durability or
stockpiling.

**Subtopics:** Forward-looking consumers; the durable-goods problem (Coase) and
intertemporal price discrimination; stockpiling and storable goods; dynamic demand
estimation.

**Methods:** Solving consumer dynamic programs; integrating dynamics into BLP-style
demand; managing the state space.

**Key readings:**

- **[R]** Gowrisankaran, G. & Rysman, M. (2012), "Dynamics of Consumer Demand for New
  Durable Goods," *Journal of Political Economy* 120(6): 1173–1219.
  [doi:10.1086/669540](https://doi.org/10.1086/669540). The workhorse framework for
  estimating dynamic demand for durables (digital camcorders); falling prices and
  forward-looking buyers.
- **[R]** Hendel, I. & Nevo, A. (2006), "Measuring the Implications of Sales and Consumer
  Inventory Behavior," *Econometrica* 74(6): 1637–1673.
  [doi:10.1111/j.1468-0262.2006.00721.x](https://doi.org/10.1111/j.1468-0262.2006.00721.x).
  Shows how ignoring consumer stockpiling biases demand elasticities; the canonical
  storable-goods paper.
- **[R]** Nair, H. (2007), "Intertemporal Price Discrimination with Forward-Looking
  Consumers: Application to the US Market for Console Video-Games," *Quantitative Marketing
  and Economics* 5(3): 239–292.
  [doi:10.1007/s11129-007-9026-4](https://doi.org/10.1007/s11129-007-9026-4). The
  marketing-canon dynamic-demand application; links forward-looking demand to optimal
  dynamic pricing.

**Debate:** How to handle expectations about future prices/products without overfitting
the dynamics?

### Week 9 — Dynamic discrete choice: Rust and CCP {#sec-io-week09}

**Topic:** Single-agent dynamic programming and the two-step shortcut that makes it
estimable.

**Subtopics:** Single-agent dynamic programming; the nested fixed-point (Rust) estimator;
conditional-choice-probability (Hotz–Miller) inversion; finite dependence; the curse of
dimensionality.

**Methods:** Value-function iteration; CCP estimation; simulation of dynamic programs.

**Key readings:**

- **[F]** Rust, J. (1987), "Optimal Replacement of GMC Bus Engines: An Empirical Model of
  Harold Zurcher," *Econometrica* 55(5): 999–1033.
  [doi:10.2307/1911259](https://doi.org/10.2307/1911259). The foundational single-agent
  dynamic discrete-choice model and nested fixed-point estimator.
- **[F]** Hotz, V. J. & Miller, R. A. (1993), "Conditional Choice Probabilities and the
  Estimation of Dynamic Models," *Review of Economic Studies* 60(3): 497–529.
  [doi:10.2307/2298122](https://doi.org/10.2307/2298122). The CCP inversion that avoids the
  full dynamic-programming solution; the basis for all modern two-step dynamic estimators.
- **[R]** Aguirregabiria, V. & Mira, P. (2010), "Dynamic Discrete Choice Structural
  Models: A Survey," *Journal of Econometrics* 156(1): 38–67.
  [doi:10.1016/j.jeconom.2009.09.007](https://doi.org/10.1016/j.jeconom.2009.09.007). The
  standard survey tying NFP, CCP, and two-step methods together; the module's roadmap.

**Debate:** NFP (full solution) vs. CCP (two-step)—efficiency vs. tractability and
robustness.

### Week 10 — Dynamic games: entry, exit, and investment {#sec-io-week10}

**Topic:** Extending dynamic discrete choice to oligopoly games with strategic
interaction.

**Subtopics:** Markov-perfect equilibrium; the Ericson–Pakes framework; two-step
estimators for dynamic games; multiplicity of equilibria.

**Methods:** Estimating policy functions; forward simulation; handling equilibrium
multiplicity.

**Key readings:**

- **[R]** Bajari, P., Benkard, C. L., & Levin, J. (2007), "Estimating Dynamic Models of
  Imperfect Competition," *Econometrica* 75(5): 1331–1370.
  [doi:10.1111/j.1468-0262.2007.00796.x](https://doi.org/10.1111/j.1468-0262.2007.00796.x).
  The general two-step (BBL) estimator for dynamic oligopoly games; a foundational frontier
  method.
- **[R]** Pakes, A., Ostrovsky, M., & Berry, S. (2007), "Simple Estimators for the
  Parameters of Discrete Dynamic Games (with Entry/Exit Examples)," *RAND Journal of
  Economics* 38(2): 373–399.
  [doi:10.1111/j.1756-2171.2007.tb00073.x](https://doi.org/10.1111/j.1756-2171.2007.tb00073.x).
  A tractable moment-based estimator for entry/exit games; widely taught alongside BBL.
- **[R]** Aguirregabiria, V. & Mira, P. (2007), "Sequential Estimation of Dynamic Discrete
  Games," *Econometrica* 75(1): 1–53.
  [doi:10.1111/j.1468-0262.2007.00731.x](https://doi.org/10.1111/j.1468-0262.2007.00731.x).
  The nested pseudo-likelihood (NPL) estimator for dynamic games; the third pillar of the
  estimation toolkit.

**Debate:** How to estimate games with multiple equilibria, and how credible are the
Markov/MPE assumptions?

### Week 11 — Static entry and market structure {#sec-io-week11}

**Topic:** What entry patterns alone reveal about competition when prices and quantities
are unobserved.

**Subtopics:** Entry as a revealed-preference inequality; ordered/sequential entry; what
entry thresholds reveal about competition and fixed costs; market structure without
price/quantity data.

**Methods:** Estimating entry models from count/threshold data; inequality-based
inference; identifying fixed costs.

**Key readings:**

- **[F]** Bresnahan, T. F. & Reiss, P. C. (1991), "Entry and Competition in Concentrated
  Markets," *Journal of Political Economy* 99(5): 977–1009.
  [doi:10.1086/261786](https://doi.org/10.1086/261786). The foundational entry-threshold
  approach: how the number of firms a market supports reveals competitive conduct.
- **[F]** Bresnahan, T. F. & Reiss, P. C. (1990), "Entry in Monopoly Markets," *Review of
  Economic Studies* 57(4): 531–553.
  [doi:10.2307/2298085](https://doi.org/10.2307/2298085). The companion paper formalizing
  entry decisions in small isolated markets; the empirical setup students replicate.

**Debate:** What can entry patterns identify about competition when prices and quantities
are unobserved?

### Week 12 — Auctions {#sec-io-week12}

**Topic:** Structural inversion from observed bids to the underlying distribution of
bidder valuations.

**Subtopics:** First-price sealed-bid auctions and the equilibrium bid function;
structural inversion from bids to valuations; nonparametric identification; private vs.
common values; risk aversion.

**Methods:** Inverting the bid function; kernel estimation of bid densities; testing for
common values.

**Key readings:**

- **[F]** Guerre, E., Perrigne, I., & Vuong, Q. (2000), "Optimal Nonparametric Estimation
  of First-Price Auctions," *Econometrica* 68(3): 525–574.
  [doi:10.1111/1468-0262.00123](https://doi.org/10.1111/1468-0262.00123). The GPV two-step
  nonparametric estimator that recovers the value distribution from observed bids; the
  canonical structural-auctions paper.
- **[R]** Campo, S., Guerre, E., Perrigne, I., & Vuong, Q. (2011), "Semiparametric
  Estimation of First-Price Auctions with Risk-Averse Bidders," *Review of Economic
  Studies* 78(1): 112–147.
  [doi:10.1093/restud/rdq001](https://doi.org/10.1093/restud/rdq001). Extends GPV to
  risk-averse bidders and the identification of preferences; the frontier reading for the
  module.

**Debate:** Private vs. common values and the limits of identification from bid data
alone.

### Week 13 — Two-sided markets and platforms {#sec-io-week13}

**Topic:** Structural estimation of indirect network effects and platform pricing across
two customer groups.

**Subtopics:** Network effects across two customer groups; platform pricing and the
"see-saw"; vertical integration and exclusivity; structural estimation of indirect
network effects.

**Methods:** Estimating cross-side externalities; modeling adoption on both sides;
platform counterfactuals.

**Key readings:**

- **[F]** Rysman, M. (2004), "Competition Between Networks: A Study of the Market for
  Yellow Pages," *Review of Economic Studies* 71(2): 483–512.
  [doi:10.1111/0034-6527.00512](https://doi.org/10.1111/0034-6527.00512). The foundational
  structural estimation of a two-sided market with indirect network effects.
- **[R]** Lee, R. S. (2013), "Vertical Integration and Exclusivity in Platform and
  Two-Sided Markets," *American Economic Review* 103(7): 2960–3000.
  [doi:10.1257/aer.103.7.2960](https://doi.org/10.1257/aer.103.7.2960). A structural model
  of platform competition (video-game consoles) with exclusive content; the modern
  platform-IO template.

**Debate:** Identifying network effects vs. unobserved quality, and the welfare effects of
platform exclusivity.

### Week 14 — Consumer search & consideration; ML + structural; synthesis {#sec-io-week14}

**Topic:** Limited demand from search and inattention, and where machine learning
complements or substitutes for economic structure.

**Subtopics:** Sequential vs. simultaneous search; consideration sets vs. full-information
demand; search/switching costs; machine learning for high-dimensional demand,
heterogeneity, and unstructured data; the structural-vs-design-based debate revisited.

**Methods:** Estimating search models; consideration-set inference; integrating ML
predictors and heterogeneity into structural objectives.

**Key readings:**

- **[R]** De los Santos, B., Hortaçsu, A., & Wildenbeest, M. R. (2012), "Testing Models of
  Consumer Search Using Data on Web Browsing and Purchasing Behavior," *American Economic
  Review* 102(6): 2955–2980.
  [doi:10.1257/aer.102.6.2955](https://doi.org/10.1257/aer.102.6.2955). Uses clickstream
  data to discriminate between sequential and fixed-sample search; the canonical empirical
  search paper.
- **[R]** Honka, E. (2014), "Quantifying Search and Switching Costs in the US Auto
  Insurance Industry," *RAND Journal of Economics* 45(4): 847–884.
  [doi:10.1111/1756-2171.12073](https://doi.org/10.1111/1756-2171.12073). A marketing-canon
  structural search-and-consideration model separating search costs from switching costs.
- **[R]** Gentzkow, M., Kelly, B., & Taddy, M. (2019), "Text as Data," *Journal of Economic
  Literature* 57(3): 535–574.
  [doi:10.1257/jel.20181020](https://doi.org/10.1257/jel.20181020). The reference survey for
  bringing unstructured text into structural/empirical work; anchors the ML-meets-structural
  discussion.
- **[R]** Wager, S. & Athey, S. (2018), "Estimation and Inference of Heterogeneous
  Treatment Effects Using Random Forests," *Journal of the American Statistical Association*
  113(523): 1228–1242.
  [doi:10.1080/01621459.2017.1319839](https://doi.org/10.1080/01621459.2017.1319839).
  Causal/generalized forests for flexible heterogeneity; the entry point for ML-based
  heterogeneity in demand and policy.

**Debate:** Search vs. preferences vs. inattention as competing explanations for "limited"
demand; and where ML complements vs. substitutes for structure.

### Foundational vs. frontier at a glance {#sec-io-foundational-vs-frontier}

**Foundational [F] — the canon every student must master.** Berry 1994 (the share
inversion); BLP 1995 (random coefficients); Nevo 2000/2001 (applied BLP, markups, merger
simulation); Bresnahan 1982 and Nevo 1998 (conduct identification); Rust 1987 (single-
agent DDC) and Hotz–Miller 1993 (CCP); Bresnahan–Reiss 1990/1991 (entry);
Guerre–Perrigne–Vuong 2000 (auctions); Rysman 2004 (two-sided markets); with Nevo's 2000
practitioner's guide and Petrin 2002 (counterfactuals / new goods) as the teaching and
applied backbone.

**Frontier [R] — methods still moving and actively contested.** Berry–Haile 2014
(nonparametric identification, which reframes what BLP identifies); Dubé–Fox–Su 2012
(MPEC) and Conlon–Gortmaker 2020 (PyBLP best practices), as computation keeps evolving;
Bajari–Benkard–Levin 2007, Pakes–Ostrovsky–Berry 2007, and Aguirregabiria–Mira 2007/2010
(dynamic games and two-step estimation); Gowrisankaran–Rysman 2012, Hendel–Nevo 2006, and
Nair 2007 (dynamic demand); Lee 2013 (platforms, bundling, vertical integration);
De los Santos–Hortaçsu–Wildenbeest 2012 and Honka 2014 (search/consideration); and
Gentzkow–Kelly–Taddy 2019 together with Wager–Athey 2018 (text-as-data and generalized
random forests) at the ML + structural frontier.

### How this reading map expands {#sec-io-reading-map-expansion}

The arc above is meant to deepen within modules rather than sprawl into disconnected
topics; the demand → supply/conduct → counterfactual spine, taught static-then-dynamic,
stays fixed. The most productive directions to extend it are:

1. **Demand frontier.** Add micro-BLP with consumer-level data and the newer PyBLP
   micro-moment workflow, plus nonparametric and "demand for characteristics" approaches
   building on Berry–Haile.
2. **Conduct-testing renaissance.** The post-2020 literature on testing firm conduct with
   instruments deserves its own subsection (paired with Bresnahan 1982 as the historical
   anchor) once those version-of-record DOIs are confirmed.
3. **Dynamic games at scale.** Add moment-inequality and machine-learning approaches to
   large state spaces, and reinforcement-learning solution methods for dynamic oligopoly.
4. **Platforms / digital markets.** Expand Week 13 into two weeks as the platform-IO and
   digital-advertising auction literatures mature (sponsored search, ad auctions,
   recommendation systems).
5. **Search & inattention.** Track the growing marketing literature on consideration sets,
   rational inattention, and search with rich clickstream/CRM data.
6. **ML + structural.** The fastest-moving module: double machine learning, causal forests,
   deep learning for demand, and LLM/text-as-data for product attributes and reviews, with
   the methodological debate about combining flexible prediction with economic structure.
7. **Marketing-specific counterfactuals.** Deepen managerial counterfactuals—dynamic
   pricing, advertising and targeting, product-line and assortment, and CRM/
   personalization—each tied back to the estimated structural model.

Every new reading added here should carry a Crossref-verified version-of-record DOI before
inclusion, and should attach to an existing module rather than open a new front.
