# Industrial Organization {#sec-industrial-organization}

Industrial organization (IO) studies how firms behave in markets, how that behavior
maps into prices, product variety, and welfare, and how researchers can recover the
deep parameters—preferences, costs, conduct—that generate observed market outcomes.
For marketing, IO is the structural backbone: it supplies the demand systems that
price a product line, the equilibrium concepts that turn one firm's pricing problem
into an industry forecast, and the identification arguments that let an analyst run
*counterfactuals*—what would happen to prices, shares, and surplus if two firms
merged, if a new product entered, or if a regulator capped a fee. Where reduced-form
marketing analytics describe correlations, structural IO builds an economic model
whose estimated primitives remain invariant under the policy change being studied,
so that the model can be re-solved under the new regime.

This chapter develops the toolkit in the order a modeler actually uses it. It opens
with the methodological frame: the core questions of IO, the modeling tools
(demand, constrained profit maximization, game theory, comparative statics), and the
discipline of turning a question into a solvable model. It then builds **consumer
demand** from two directions. The *classical* approach treats a representative
consumer choosing continuous quantities, yielding the elegant Slutsky/Roy machinery,
the CES and quasi-linear special cases, diversion ratios, and the homothetic-surplus
family that makes multiproduct pricing tractable. The *discrete-choice* approach
treats heterogeneous consumers each picking one option from a menu, which is how
modern empirical IO and quantitative marketing estimate demand for differentiated
products. We give each model its formal definition, its estimator, its identifying
assumptions, and the failures that break identification—price endogeneity,
zero-valued shares, selection on unobservables—and we close with the frontier
applications in marketing: dynamic pricing of new products, behavioral price
discrimination in channels, email contact policy, and multi-sided platforms. Code is
runnable and seeded throughout. The companion structural-econometrics machinery
appears in @sec-marketing-finance and in the demand-estimation discussions elsewhere
in the book; here the emphasis is on the economic models themselves.

## The IO Research Program

Three questions organize the field. First, **why do firms behave as we observe**—why
this price, this product line, this bundle, this contract? Second, **what are the
effects** of that behavior on consumers, rivals, and employees, and against which
welfare standard should those effects be judged? Third, **how can we design better
strategy** for firms and better policy for regulators? The first question is
positive, the second normative, and the third prescriptive; a complete IO study
usually touches all three, because a counterfactual that improves firm profit and one
that improves consumer surplus are different optimizations over the same estimated
model.

Three methodological approaches answer these questions and are complementary rather
than competing. The **theoretical** approach builds a model and derives predictions;
the **empirical** approach estimates the model from data, tests its predictions, and
runs counterfactual and welfare analysis; the **experimental and behavioral**
approach manipulates conditions to isolate causal mechanisms and to discipline the
behavioral assumptions the other two impose. The structural empirical tradition that
dominates this chapter fuses the first two: it writes down a theoretical model of
demand and conduct, then estimates its primitives so the model can be re-solved under
hypothetical policies.

### Building an IO Model

A study begins with an *idea*—a question drawn from the literature in IO and adjacent
fields, or from the world (a news item, a consulting engagement, an unexplained
pattern in data). The idea is then expressed through a small set of **modeling
tools** that recur across nearly every IO paper, summarized in @fig-io-workflow.

```{mermaid}
%%| label: fig-io-workflow
%%| fig-cap: "The structural IO modeling workflow, from question to welfare statement. Each stage constrains the next: the demand system feeds the firm's first-order conditions, which define equilibrium, whose comparative statics answer the counterfactual."
flowchart TD
    A[Idea / Question<br/>literature or world] --> B[Model consumer demand]
    B --> C[Profit maximization<br/>subject to constraints]
    C --> D[Game-theoretic equilibrium<br/>among firms]
    D --> E[Comparative statics<br/>& counterfactuals]
    E --> F[Welfare & policy<br/>evaluation]
    F -.->|do results make<br/>intuitive sense?| B
```

The first tool is a **model of consumer demand** for the products or services in
question. The modeler decides whether consumers choose among single or multiple
products, single or multiple units, and whether products are differentiated—and if
so, whether they are substitutes or complements, and whether differentiation is
*vertical* (everyone agrees more is better, as with quality) or *horizontal*
(consumers disagree about the ideal variety, as with location or flavor). Demand
*complementarities* deserve special attention because they reshape strategy: when
products are complements, raising one price depresses demand for both, so a
monopolist who internalizes the cross-effect charges *lower* prices than two
independent firms would, and the complementarity can itself become a barrier to
entry. Network effects, multi-sided markets, and ecosystems are demand
complementarities operating across consumers or across platform sides rather than
within a single basket.

The second tool is **profit maximization subject to constraints**. The modeler
specifies choice variables and an objective, then layers on the constraints that make
the problem economically interesting: technology constraints from the cost side;
consumer *participation* constraints (the consumer must prefer buying to not buying),
which rationalize bundling and tied selling and let contracts act as entry barriers;
*arbitrage* constraints that limit price discrimination across consumers who can
resell; and *incentive* constraints in principal–agent settings, which generate
second-degree price discrimination (self-selection across a menu) and franchise
contracts. The third tool is **game theory**, treated next. The fourth is
**comparative static analysis**: with exogenous parameters and policy variables on one
side and endogenous equilibrium variables on the other, the modeler asks how the
optimum or equilibrium shifts as a parameter moves—the formal content of every
"what if" the study can answer.

### Game-Theoretic Models

Strategic interaction among firms is modeled non-cooperatively or cooperatively.
**Non-cooperative** models come in two forms. *Normal-form* (simultaneous-move) games
specify a set of players, a strategy set for each, and a payoff function for each, and
solve for Nash equilibrium in whatever strategic variable is relevant—price
(Bertrand), quantity (Cournot), capacity, advertising, or location. *Extensive-form*
(sequential-move) games capture first-mover advantages, entry deterrence, and
collusion sustained through repeated interaction, where the threat of future
punishment supports prices above the static equilibrium. **Cooperative** models, used
heavily in channel and platform settings, center on **Nash bargaining**: two parties
split the gains from agreement according to their bargaining power, relative to a
*threat point* (the disagreement payoff). The **Nash-in-Nash** refinement embeds
bilateral Nash bargains inside a larger non-cooperative game—each upstream–downstream
pair bargains taking all other negotiated prices as given—and has become the workhorse
for modeling negotiated wholesale prices in vertical markets.

> A profile of strategies is a **Nash equilibrium** if no player can raise its own
> payoff by unilaterally deviating, holding all rivals' strategies fixed. In a pricing
> game this means each firm's price is a best response to its rivals' prices, so the
> equilibrium price vector solves all firms' first-order conditions simultaneously.

The output of this machinery is a set of results, after which—per the discipline of
the field—the modeler pauses to ask whether the results make intuitive sense before
trusting them. A written IO study then motivates the question (with examples,
statistics, prior work, and policy stakes), states the approach and headline
findings, constructs the model, derives the equilibrium and its comparative statics,
and closes with limitations and extensions.

## Consumer Demand {#sec-io-demand}

The central modeling primitive is a **demand system**: a description of how the
quantities consumers buy respond to firms' offers—prices, qualities, versions,
advertising, warranties. Three questions drive the demand modeler. How should demand
be formally specified given the product and the market? What properties must a demand
*system* (more than one product) satisfy to be internally consistent? And how can the
system be estimated so that own-price elasticities, cross-price elasticities, and
diversion ratios can be computed and fed into a pricing or merger analysis?

Demand is fundamentally a **matching process** between a population of consumers—each
with observable and unobservable characteristics—and a set of products with their own
observable and unobservable characteristics. The microeconomics and IO literatures
offer three routes through this matching problem: *classical consumer theory* with
continuous choices; *discrete-choice models* with heterogeneous tastes, which
accommodate horizontal differentiation, vertical differentiation, or both; and *hybrid
discrete–continuous* models in which the consumer first chooses which product and then
how much. The next two parts develop the classical and discrete-choice routes in
turn.

## Classical Consumer Demand {#sec-io-classical}

A representative consumer with utility $U(q_0, q_1, \dots, q_n)$ and income $I$, facing
prices $\mathbf{p} = (p_1, \dots, p_n)$, chooses continuous quantities
$\mathbf{q} = (q_1, \dots, q_n)$ to solve

$$
\max_{q_0, \dots, q_n} \; U(q_0, q_1, \dots, q_n)
\quad \text{s.t.} \quad q_0 + \sum_{i=1}^n p_i q_i \le I,
$$ {#eq-um-classical}

where $q_0$ is a numeraire good with normalized price $p_0 = 1$. The solution is the
**Walrasian** (or Marshallian) demand system $q_i = D_i(\mathbf{p}, I)$, and
substituting it back into $U$ gives the **indirect utility** $V(\mathbf{p}, I)$.

This system is not an arbitrary set of functions; it inherits a rigid structure from
optimization, and that structure is what makes the classical approach powerful. Two
identities encode it. **Roy's identity** recovers demand from the indirect utility,

$$
q_i(\mathbf{p}, I) = -\,\frac{\partial V / \partial p_i}{\partial V / \partial I},
$$ {#eq-roy}

so the entire demand system is implicit in a single scalar function $V$. The **Slutsky
equation** decomposes the response of demand to a price change into a substitution
effect (the compensated, or Hicksian, derivative $\partial h_i / \partial p_j$) and an
income effect,

$$
\frac{\partial h_i(\mathbf{p}, u)}{\partial p_j}
= \frac{\partial q_i(\mathbf{p}, I)}{\partial p_j}
+ \frac{\partial q_i(\mathbf{p}, I)}{\partial I}\, q_j(\mathbf{p}, I).
$$ {#eq-slutsky}

The matrix of compensated derivatives—the **Slutsky substitution matrix**—is symmetric
and negative semi-definite. Symmetry is the testable fingerprint of utility
maximization: absent income effects, the cross-price response of good $i$ to $p_j$
equals that of good $j$ to $p_i$. Negative semi-definiteness encodes the law of
demand, that compensated own-price effects are non-positive.

### CES Sub-Utility and the Taste for Variety

A leading special case separates the numeraire from a bundle of differentiated goods,
$U = U(q_0, Q)$, with the differentiated goods aggregated by a **constant elasticity
of substitution** (CES) sub-utility. Taking $U = q_0 Q$ with

$$
Q = \Big( \sum_{i=1}^n q_i^{\rho} \Big)^{1/\rho}, \qquad \rho \le 1,
$$ {#eq-ces-Q}

delivers the CES demand system. The parameter $\rho$ controls substitutability: the
elasticity of substitution is $1/(1-\rho)$, so $\rho \in (0,1)$ makes the goods
imperfect substitutes (elasticity above one), $\rho \to 1$ makes them perfect
substitutes, and $\rho \to 0$ approaches Cobb–Douglas. The CES form is the standard
tool for studying product *variety*, because the consumer values having many distinct
varieties even at equal total quantity—the "love of variety" that justifies entry of
differentiated competitors.

To derive demand, write the consumer problem over $n+1$ goods with the multiplicative
specification

$$
U = q_0 \Big[ \sum_{i=1}^n q_i^{\rho} \Big]^{1/\rho}, \qquad \rho \in (0,1),
$$

subject to the budget $P_0 q_0 + \sum_{i=1}^n P_i q_i = M$. The Lagrangian is

$$
\mathcal{L} = q_0 \Big[ \sum_{i=1}^n q_i^{\rho} \Big]^{1/\rho}
+ \lambda \Big( M - P_0 q_0 - \sum_{i=1}^n P_i q_i \Big).
$$

The first-order condition for the numeraire,
$\partial \mathcal{L}/\partial q_0 = \big[\sum_i q_i^{\rho}\big]^{1/\rho} - P_0\lambda = 0$,
identifies the marginal utility of income with the aggregate $Q$ scaled by price,
$\big[\sum_i q_i^{\rho}\big]^{1/\rho} = P_0 \lambda$. The condition for a
differentiated good,

$$
\frac{\partial \mathcal{L}}{\partial q_i}
= q_0 \Big[ \sum_{i=1}^n q_i^{\rho} \Big]^{1/\rho - 1} q_i^{\rho - 1} - P_i \lambda = 0,
$$

gives $q_0 \big[\sum_i q_i^{\rho}\big]^{1/\rho - 1} q_i^{\rho-1} = P_i \lambda$.
Dividing the two conditions eliminates $\lambda$ and yields the *direct* CES demand;
solving instead for $P_i$ yields the *inverse* demand. The economic reading is the one
above: with $\rho \in (0,1)$ the varieties are imperfect substitutes.

### Quasi-Linear Utility and the Absence of Income Effects

A second special class drops income effects entirely. With **quasi-linear** utility,

$$
U = q_0 + u(q_1, \dots, q_n),
$$ {#eq-quasilinear}

substituting the budget constraint turns the consumer problem into

$$
\max_{q_1, \dots, q_n} \; u(q_1, \dots, q_n) - \sum_{i=1}^n p_i q_i + I,
$$

with first-order conditions $\partial u / \partial q_i = p_i$. Because $I$ enters
additively, the demands for the $n$ differentiated goods are **independent of income**
(provided income is large enough that $q_0 > 0$). All income is absorbed by the
numeraire, and the maximized utility is $V = v(p_1, \dots, p_n) + I$, where
$v(\cdot)$ is the consumer-surplus function. This additive separability is what makes
quasi-linear utility the default for welfare analysis in IO: changes in consumer
surplus are read directly off $v$, untangled from income.

The structure delivers three results used repeatedly below. First, demand is the
negative gradient of surplus, $-\partial v / \partial p_i = q_i(\mathbf{p})$, a special
case of Roy's identity (@eq-roy). Second, the **Jacobian** of the demand system,
$J = \big( \partial q_i(\mathbf{p}) / \partial p_j \big)_{n \times n}$, is symmetric and
negative semi-definite—the Slutsky matrix (@eq-slutsky) stripped of income terms.
Third, the construction can be **reverse-engineered**: if an observed demand system
has a symmetric, negative-definite Jacobian, it is rationalizable by a representative
consumer maximizing a concave quasi-linear utility, recovered by substituting the
inverse demand into the first-order conditions and integrating the resulting system of
partial differential equations $\partial u / \partial q_i = p_i(q)$. Recovering $u$ in
turn enables exact welfare statements.

#### Substitutes, Complements, and Diversion {#sec-io-diversion}

With $n = 2$, the sign of the cross-effect classifies the goods. They are
**demand-substitutes** if good $j$'s demand rises with $p_i$ and **demand-complements**
if it falls. Applying Cramer's rule to the first-order conditions of quasi-linear
maximization,

$$
|J|\, \frac{\partial q_1}{\partial p_2}
= |J|\, \frac{\partial q_2}{\partial p_1}
= -\,u_{12} = -\,u_{21},
$$ {#eq-cramer-cross}

where $|J| > 0$ by negative definiteness (strict concavity of $u$). The products are
therefore demand-substitutes when $u_{12} = u_{21} < 0$ and demand-complements when
$u_{12} = u_{21} > 0$: the *sign of the utility cross-partial* governs the substitution
relationship.

The **diversion ratio** quantifies substitution and is the single most important
demand statistic in merger analysis. For imperfect substitutes, it measures how much
of the demand lost from raising $p_1$ is recaptured by product 2,

$$
d_{12} = \frac{\partial q_2 / \partial p_1}{-\,\partial q_1 / \partial p_1} > 0.
$$ {#eq-diversion}

Intuitively, if a consumer prefers coffee brand A but buys brand B when A's price
rises (or A is out of stock), sales "divert" from A to B; $d_{12}$ is the fraction of
A's lost sales that lands on B. Negative definiteness of the substitution matrix
imposes a discipline on the pair of diversion ratios,

$$
\frac{\partial q_1}{\partial p_1}\frac{\partial q_2}{\partial p_2}
- \frac{\partial q_2}{\partial p_1}\frac{\partial q_1}{\partial p_2} > 0
\quad \Longleftrightarrow \quad d_{12}\, d_{21} < 1,
$$

so two goods cannot each divert most of their lost sales to the other. The empirically
estimable version, for products $A$ and $B$ with quantities and prices
$(Q_A, P_A, Q_B, P_B)$, is

$$
D_{AB} = \frac{\Delta Q_B / \Delta P_A}{\Delta Q_A / \Delta P_A},
$$ {#eq-diversion-emp}

the change in $B$'s quantity per unit change in $A$'s price, normalized by $A$'s own
quantity response. The following seeded example simulates two products and estimates
the diversion ratio from first differences.

```{r io-diversion-sim, message=FALSE, warning=FALSE}
set.seed(123)
n <- 100
price_A <- runif(n, 5, 10)
price_B <- runif(n, 5, 10)
quantity_A <- 100 - 5 * price_A + rnorm(n)
quantity_B <-  80 - 4 * price_B + rnorm(n)

# Diversion ratio from first differences (eq. of D_AB)
delta_QA <- diff(quantity_A)
delta_PA <- diff(price_A)
delta_QB <- diff(quantity_B)

D_AB <- mean(delta_QB / delta_PA) / mean(delta_QA / delta_PA)
D_AB
```

```{r io-diversion-plot, message=FALSE, warning=FALSE}
#| fig-cap: "Change in quantity of each product against the change in product A's price. Product A (blue) responds steeply to its own price; product B (red) is comparatively unresponsive, so little demand diverts from A to B."
library(ggplot2)
plot_df <- data.frame(delta_PA, delta_QA, delta_QB)
ggplot(plot_df, aes(x = delta_PA)) +
  geom_point(aes(y = delta_QA), color = "blue", alpha = 0.5) +
  geom_point(aes(y = delta_QB), color = "red",  alpha = 0.5) +
  geom_smooth(aes(y = delta_QA), method = "lm", color = "blue", se = FALSE) +
  geom_smooth(aes(y = delta_QB), method = "lm", color = "red",  se = FALSE) +
  labs(x = "Change in price of A", y = "Change in quantity",
       title = "Change in quantity vs. change in price of A",
       subtitle = "Blue: product A   Red: product B") +
  causalverse::ama_theme()
```

### The Linear Demand System from Quadratic Utility

The most-used closed-form demand system comes from a quasi-linear **quadratic**
utility. Over two differentiated goods,

$$
U(q_0, q_1, q_2) = q_0 + \alpha_1 q_1 + \alpha_2 q_2
- \tfrac{1}{2}\big[ \beta_1 q_1^2 + 2\gamma\, q_1 q_2 + \beta_2 q_2^2 \big],
$$ {#eq-quad-utility}

with $\alpha_i > 0$, $\beta_i > 0$, and $\beta_1 \beta_2 - \gamma^2 > 0$ (concavity).
The first-order conditions are the **inverse** (linear) demands,

$$
p_1 = \alpha_1 - \beta_1 q_1 - \gamma q_2, \qquad
p_2 = \alpha_2 - \beta_2 q_2 - \gamma q_1,
$$ {#eq-linear-inverse}

and inverting this linear system gives the direct demands with constant diversion
ratios. The cross-partial of utility is exactly $\partial^2 U / \partial q_1 \partial q_2
= \gamma$, so $\gamma$ *is* the substitution parameter: under the sign convention of
@eq-cramer-cross, a positive cross-partial signals complements and a negative one
signals substitutes. Two parameterization styles appear in practice. Treating $\gamma$
as a free parameter captures the relationship directly and flexibly. Tying the
curvature to substitution by setting $\beta_1 = \beta_2 = 1 - \gamma$ is more
parsimonious but restrictive: it forces own-curvature to move with the cross-effect,
which can be arbitrary when theory offers no guidance.

#### The Shubik–Levitan Normalization

A subtle defect of the raw quadratic form is that adding products mechanically changes
the scale of demand. @shubik1980 propose a normalization that separates the *number*
of products from their *substitutability*. With

$$
u(q_1, \dots, q_n) = \alpha \sum_{i=1}^n q_i
- \frac{n}{2(1+\gamma)}\Big[ \sum_{i=1}^n q_i^2
+ \frac{\gamma}{n}\Big( \sum_{i=1}^n q_i \Big)^2 \Big],
$$ {#eq-shubik-utility}

where $\alpha > 0$ and $\gamma \in [0, \infty)$ indexes substitutability, the demand
system becomes

$$
q_i = \frac{1}{n}\Big[ \alpha - p_i(1+\gamma) + \frac{\gamma}{n} \sum_{j=1}^n p_j \Big].
$$ {#eq-shubik-demand}

This form has two properties that make it the standard for oligopoly-variety models.
The aggregate demand $Q = \sum_i q_i = \alpha - \tfrac{1}{n}\sum_j p_j$ does not depend
on the degree of substitution $\gamma$, and under symmetric prices $p_i = p$ the
aggregate $Q = \alpha - p$ does not depend on the number of products $n$. Holding the
"market" fixed while varying competition and variety is exactly what one wants when
asking how entry or differentiation reshapes equilibrium.

The normalization supports two instructive applications. In the first, a downstream
**buyer** with revenue $u(q_1, q_2)$ faces a *market-share restriction* imposed by the
seller of product 1: $q_1/(q_1 + q_2) \ge s$ or $q_1 = 0$, where $s \in [0,1]$ is the
required share and $s = 1$ is an exclusive contract. The buyer solves
$\max_{q_1, q_2} u(q_1, q_2) - p_1 q_1 - p_2 q_2$ subject to that constraint, and the
linear-quadratic utility yields a constrained demand $q(p, s)$ whose comparative
statics in $p$ and $s$ reveal how share-based contracts distort purchasing. In the
second, a demand system satisfies the **strong-complements** property if the ratio
$q_i^*(p)/q_1^*(p)$ is independent of $p_1$ for $i = 2, \dots, n$. This is equivalent
to the statement that, for any constant marginal costs $c_i > 0$, the monopolist's
problem $\max_p \sum_i (p_i - c_i) q_i^*(p)$ sets $p_i^* = c_i$ for $i = 2, \dots, n$:
when goods are strong complements, the multiproduct monopolist prices the
complementary goods at marginal cost and earns its margin entirely on the base good.
Writing $v(p) = u(q^*(p)) - \sum_i p_i q_i^*(p)$ for indirect utility, the maximum
profit conditional on delivering the consumer at least $u$,
$\pi(u) = \max_p \{ \sum_i (p_i - c_i) q_i^*(p) : v(p) \ge u \}$, is decreasing and
concave in the promised surplus.

### Equilibrium Pricing and Cost Pass-Through {#sec-io-bertrand}

The demand system is only half the model; the other half is firm conduct. Suppose two
single-product firms with constant marginal costs $c_1, c_2$ set prices
simultaneously. Firm 1 maximizes $\pi_1(p_1, p_2) = (p_1 - c_1) q_1(p_1, p_2)$, with
first-order condition

$$
q_1 + (p_1 - c_1)\frac{\partial q_1}{\partial p_1} = 0,
$$ {#eq-bertrand-foc}

and symmetrically for firm 2. The pair of conditions defines the **Bertrand–Nash**
equilibrium prices $(p_1^*, p_2^*)$. A key comparative static is **cost pass-through**
(CPT)—how an equilibrium price responds to a marginal-cost shock—with own and cross
rates

$$
CPT_{11} = \frac{\partial p_1^*}{\partial c_1}, \quad
CPT_{22} = \frac{\partial p_2^*}{\partial c_2}, \quad
CPT_{12} = \frac{\partial p_1^*}{\partial c_2}, \quad
CPT_{21} = \frac{\partial p_2^*}{\partial c_1}.
$$ {#eq-cpt}

Pass-through is the quantity a regulator or analyst needs to predict how a tax,
tariff, or input-cost shock reaches consumers, and it depends on the *curvature* of
demand, not merely its slope.

When the two firms **merge or collude**, the combined entity becomes a multiproduct
monopolist maximizing joint profit
$\pi(p_1, p_2) = (p_1 - c_1) q_1 + (p_2 - c_2) q_2$, with first-order conditions

$$
\begin{aligned}
\frac{\partial \pi}{\partial p_1}
&= q_1 + (p_1 - c_1)\frac{\partial q_1}{\partial p_1}
+ (p_2 - c_2)\frac{\partial q_2}{\partial p_1} = 0,\\[4pt]
\frac{\partial \pi}{\partial p_2}
&= q_2 + (p_2 - c_2)\frac{\partial q_2}{\partial p_2}
+ (p_1 - c_1)\frac{\partial q_1}{\partial p_2} = 0.
\end{aligned}
$$ {#eq-monopoly-foc}

The monopolist's conditions differ from the competitive ones (@eq-bertrand-foc) by the
extra cross-effect term $(p_j - c_j)\,\partial q_j / \partial p_i$: the merged firm
**internalizes** the demand it steals from itself. For substitutes this raises both
prices above the Bertrand level—the classic unilateral-effects concern in merger
review—while pass-through rates $CPT^M$ generally differ from the competitive case
because the monopolist accounts for cross-effects when passing through cost shocks.
@tbl-bertrand-monopoly contrasts the two regimes.

| Feature | Bertrand competition | Multiproduct monopoly / collusion |
|---|---|---|
| First-order condition | own margin × own slope only | own margin + cross-margin × cross-slope |
| Internalizes cross-effects | no | yes |
| Equilibrium prices (substitutes) | lower | higher |
| Cost pass-through | own-cost driven | reflects internalized cross-effects |
| Treatment of rival product | competitor | own product line |

: Bertrand competition versus joint pricing under the same linear demand system. The single structural difference—internalizing cross-demand—drives the price and pass-through divergence. {#tbl-bertrand-monopoly}

The consumer side of this model is closed by **consumer surplus**. Integrating each
inverse demand from @eq-linear-inverse below the price gives
$CS_i = \int_0^{q_i} p_i(q_i', q_{-i})\,dq_i' - p_i q_i$, and total surplus is
$CS = CS_1 + CS_2$; this is the welfare object compared across the competitive and
monopoly regimes.

### Properties of Multiproduct Demand and the Symmetry of Cross-Effects

For a single product, demand $q = D(p)$ is downward-sloping, $q'(p) < 0$, with price
elasticity $\epsilon(p) = -q'(p)\,p/q$. With multiple products the new content is the
*cross*-price structure. A linear system

$$
\begin{aligned}
q_1 &= \alpha_1 - \beta_{11} p_1 + \gamma_{12} p_2 + \cdots,\\
q_2 &= \alpha_2 - \beta_{22} p_2 + \gamma_{21} p_1 + \cdots,
\end{aligned}
$$

has own-price effects $\partial q_i / \partial p_i < 0$ and cross-price effects whose
sign is positive for substitutes and negative for complements. A foundational result
is that the cross-effects are **symmetric**,
$\partial q_1 / \partial p_2 = \partial q_2 / \partial p_1$, whenever consumers have
quasi-linear utility (no income effect). The derivation makes the source of symmetry
explicit. From $\max_{q_1, q_2} u(q_1, q_2) - p_1 q_1 - p_2 q_2$ the first-order
conditions $u_1(q_1^*, q_2^*) = p_1$ and $u_2(q_1^*, q_2^*) = p_2$ hold; differentiating
the system with respect to $p_1$,

$$
\begin{aligned}
u_{11}\frac{\partial q_1^*}{\partial p_1} + u_{12}\frac{\partial q_2^*}{\partial p_1} &= 1,\\
u_{21}\frac{\partial q_1^*}{\partial p_1} + u_{22}\frac{\partial q_2^*}{\partial p_1} &= 0,
\end{aligned}
$$

and solving yields

$$
\frac{\partial q_2^*}{\partial p_1} = -\,\frac{u_{21}}{u_{11}u_{22} - u_{12}u_{21}},
\qquad
\frac{\partial q_1^*}{\partial p_2} = -\,\frac{u_{12}}{u_{11}u_{22} - u_{12}u_{21}}.
$$

Because $u_{12} = u_{21}$ (Young's theorem on the symmetry of cross-partials), the two
cross-effects are equal. This is the Slutsky symmetry of @eq-slutsky with the income
terms switched off.

When income effects are present, symmetry of the *uncompensated* cross-effects no
longer follows automatically. Slutsky symmetry holds for the *compensated* derivatives,

$$
\frac{\partial h_i}{\partial p_j}
= \frac{\partial q_i}{\partial p_j} + \frac{\partial q_i}{\partial I} q_j
= \frac{\partial h_j}{\partial p_i}
= \frac{\partial q_j}{\partial p_i} + \frac{\partial q_j}{\partial I} q_i,
$$

so the observed (uncompensated) cross-effects are equal only if the income terms
balance, i.e. if $\frac{\partial q_i}{\partial I}/q_i = \frac{\partial q_j}{\partial I}/q_j$.
The empirical reading is useful: if the data show *equal* uncompensated cross-price
effects, that is consistent evidence for either no income effect, or income effects
that scale the two goods' consumption proportionally—an additional rupee of income
expands consumption of both goods in the same relative measure. Two normalizations make
the cross-versus-own comparison concrete: assuming
$-\partial q_i / \partial p_i > \partial q_j / \partial p_i$ keeps the diversion ratio
below one, and assuming $-\partial q_i / \partial p_i > \partial q_i / \partial p_j$
keeps the *indirect* diversion ratio below one. Own- and cross-price elasticities are
defined analogously by scaling each derivative by $p/q$.

The classical, continuous approach has real strengths: the tight algebraic structure
of the demand system delivers clean pricing strategies and exact welfare statements.
But it pays little attention to product *characteristics*—quality, location, timing,
availability, the consumer's information about a product's existence and quality—and it
does not formally model the *heterogeneity* of consumer preferences. Discrete-choice
models, developed in @sec-io-discrete, are designed precisely to address these gaps.

### Tractable Multiproduct Demand: Homothetic Surplus {#sec-io-armstrong}

A frontier question is which multiproduct demand systems are simultaneously rich
enough to be realistic and tractable enough to solve. @armstrong2018multiproduct
answer it by characterizing the demand systems whose *consumer surplus* is homothetic,
a property that collapses a high-dimensional pricing problem into a one-dimensional
one. Begin with the representative consumer's quasi-linear problem: utility $u(q)$ with
$u(0) = 0$, increasing and strictly concave, choosing $q$ to solve $\max_q u(q) - p
\cdot q$. The inverse demand is the gradient $p(q) = \nabla u(q)$, total revenue is
$R(q) = q \cdot \nabla u(q)$, and consumer surplus is

$$
s(q) = u(q) - q \cdot \nabla u(q).
$$ {#eq-armstrong-surplus}

> **Proposition** [@armstrong2018multiproduct]. Consumer surplus $s(q)$ is homothetic
> in $q$ if and only if utility takes the form
> $$ u(q) = h(q) + g\big(f(q)\big), $$
> where $h$ and $f$ are homogeneous of degree one and $g$ is concave with $g(0) = 0$.

This is a far wider class than the systems whose surplus is homothetic in *prices*
(the case $h = 0$), and it exposes exactly **three degrees of freedom**—the functions
$f(q)$, $h(q)$, and $g(\cdot)$—that the modeler can choose to match data while
preserving tractability. The "if" direction is instructive. Inverse demand is
$p(q) = \nabla h(q) + g'(f(q))\,\nabla f(q)$, and since $q \cdot \nabla h(q) \equiv
h(q)$ for a degree-one homogeneous function, revenue is $R(q) = h(q) + g'(f(q)) f(q)$,
so surplus reduces to $s(q) = g(f(q)) - g'(f(q)) f(q)$—a monotone function of the
single composite $f(q)$.

The proof technique is a change to "polar coordinates" that pays off conceptually.
Write $q = f(q) \cdot \tfrac{q}{f(q)}$, where $q/f(q)$ is homogeneous of degree zero and
depends only on the *ray* from the origin (the mix of products), while $f(q)$—the
**composite quantity**—measures how far along that ray the bundle lies. The consumer's
problem then *separates* into two stages. Surplus
$u(q) - p\cdot q = g(f(q)) - f(q)\frac{p\cdot q - h(q)}{f(q)}$ is maximized, for any
fixed $f(q)$, by minimizing $\frac{p\cdot q - h(q)}{f(q)}$; define the **composite
price**

$$
\phi(p) \equiv \min_{q \ge 0} \frac{p \cdot q - h(q)}{f(q)},
$$ {#eq-armstrong-phi}

which is increasing and concave in $p$. By the envelope theorem the optimal *relative
quantities* are $q^r(p) \equiv \nabla \phi(p)$. Given those, the consumer chooses the
composite quantity $Q$ to maximize the concave objective $g(Q) - Q\,\phi(p)$, whose
solution $\hat{Q}(\phi(p))$ is the demand for the composite. The full demand function
factors as $q(p) = \hat{Q}(\phi(p)) \times q^r(p)$: the consumer picks a *mix* from
relative prices and a *scale* from the composite price, and all prices with the same
$\phi(p)$ induce the same composite quantity through the inverse relation
$\phi(p) = g'(\hat{Q})$.

Three examples show the family's reach. The **linear** system $u(q) = aq -
\tfrac{1}{2} q^\top B q$ with $a > 0$ and $B$ positive definite gives $p(q) = a - Bq$
and decomposes as $h(q) = aq$, $f(q) = \sqrt{q^\top B q}$, $g(f) = -\tfrac{1}{2}f^2$,
and $s(q) = \tfrac{1}{2} f(q)^2$. The **logit** system
$q_i(p) = \frac{e^{(a_i - p_i)/\mu}}{k + \sum_j e^{(a_j - p_j)/\mu}}$ corresponds to
surplus $v(p) = \mu \log\!\big(1 + \tfrac{1}{k}\sum_j e^{(a_j - p_j)/\mu}\big)$, with
the decomposition $f(q) = \sum_j q_j$ (total quantity),
$g(f) = -\mu\big(f\log f + (1-f)\log(1-f)\big)$, and $s(q) = -\mu\log(1 - f(q))$. The
**complementary-products** example captures hardware-plus-services: if $q_1$ is the
number of access units (hardware) and each requires services $y = q_2/q_1$, gross
utility is $u(q_1, q_2) = q_1 \hat{u}(q_2/q_1) + g(q_1)$, an instance with $f(q) = q_1$,
so surplus depends only on the number of access units. The ratio $q_2/q_1$ is
independent of $p_1$—precisely the strong-complements property of @sec-io-classical—
with the implication that the monopolist prices services to extract the base-good
margin.

## Vertical Differentiation: Quality {#sec-io-vertical}

Differentiation is **vertical** when augmenting a product characteristic benefits *all*
consumers: everyone prefers higher quality, faster delivery, longer battery life. The
canonical model gives a consumer of type $\theta$ utility

$$
u =
\begin{cases}
\theta s - p & \text{if she buys quality } s \text{ at price } p,\\
0 & \text{if she does not buy,}
\end{cases}
$$ {#eq-vertical-utility}

where $s > 0$ is quality (common knowledge), utility is separable in quality and
price, and $\theta > 0$ is a taste parameter—the consumer's marginal willingness to pay
for quality—distributed across the population with density $f(\theta)$ and cdf
$F(\theta)$ on $[\underline{\theta}, \bar{\theta}]$. Firms know the distribution but not
any individual's $\theta$. The defining feature is unanimity: every consumer prefers
higher $s$, in contrast to horizontal differentiation where consumers disagree about
the ideal variety.

**One quality.** When a single quality $s$ is offered at price $p$, a type $\theta$
buys iff $\theta s - p \ge 0$, i.e. $\theta \ge p/s$. With $N$ identical consumers,
aggregate demand is

$$
D(p) = N\big[1 - F(p/s)\big],
$$ {#eq-vertical-demand1}

which is downward-sloping. Under the uniform $F$, $D(p) = N(1 - p/s)$ for $p \le s$ and
zero otherwise. Unlike classical demand, quality $s$ enters *explicitly*, so the model
can speak to quality choice, not just price.

**Two qualities.** Offering a product line $s_1 < s_2$ (with $p_1 < p_2$) lets the firm
screen consumers and extract more surplus—the logic of second-degree price
discrimination. A type $\theta$ buys $s_1$ iff $\theta s_1 - p_1 \ge 0$ and
$\theta s_1 - p_1 \ge \theta s_2 - p_2$; she buys $s_2$ iff $\theta s_2 - p_2 \ge 0$ and
$\theta s_2 - p_2 \ge \theta s_1 - p_1$. Three cutoffs—$p_1/s_1$, $p_2/s_2$, and the
indifference type $\theta^* \equiv (p_2 - p_1)/(s_2 - s_1)$—partition consumers across
not buying, buying low, and buying high. The configuration depends on which good offers
more *quality per dollar*. If the high good dominates on quality-per-dollar
($p_2/s_2 \le p_1/s_1$), the low good is driven out and only $s_2$ sells; otherwise both
sell, with the low good serving an interior band of types. Collecting both cases,

$$
D_1(p_1, p_2) =
\begin{cases}
0 & \text{if } \tfrac{p_1}{p_2} \ge \tfrac{s_1}{s_2},\\[4pt]
N\big[F(\tfrac{p_2 - p_1}{s_2 - s_1}) - F(\tfrac{p_1}{s_1})\big] & \text{if } \tfrac{p_1}{p_2} < \tfrac{s_1}{s_2},
\end{cases}
$$ {#eq-vertical-D1}

$$
D_2(p_1, p_2) =
\begin{cases}
N\big[1 - F(\tfrac{p_2}{s_2})\big] & \text{if } \tfrac{p_1}{p_2} \ge \tfrac{s_1}{s_2},\\[4pt]
N\big[1 - F(\tfrac{p_2 - p_1}{s_2 - s_1})\big] & \text{if } \tfrac{p_1}{p_2} < \tfrac{s_1}{s_2}.
\end{cases}
$$ {#eq-vertical-D2}

Under the uniform special case $F(\theta) = \theta$ on $[0,1]$, the interior cross-price
derivatives are $\partial D_1 / \partial p_2 = \partial D_2 / \partial p_1 = N/(s_2 -
s_1) > 0$: the two qualities are **substitutes**, and the strength of substitution
scales inversely with the quality gap $s_2 - s_1$. Widening the quality ladder softens
competition between the firm's own tiers—the structural reason a multiproduct firm
spaces its versions apart.

The taste parameter $\theta$ admits a measurable interpretation: it can stand for
willingness to pay for quality and is often linked to **income**, the most observable
axis of consumer heterogeneity. If consumers differ in income and utility is
$U = u(I - p) + s$, a first-order Taylor expansion of $u$ around $I$ recovers the
quasi-linear form $\theta s - p$ with $\theta$ tied to the marginal utility of income,
making the abstract taste parameter empirically grounded.

## Discrete-Choice Demand {#sec-io-discrete}

Discrete-choice models replace the continuous-quantity consumer with one who selects a
single alternative from a menu, which is how modern empirical IO and quantitative
marketing estimate demand for differentiated products. The payoff is twofold: a small
number of *preference* parameters generates the full matrix of own- and cross-price
elasticities, and demand can be predicted for *new* or hypothetical products by
projecting them into characteristic space. @fig-discrete-lineage traces the lineage of
models this section develops, from the workhorse logit through its corrections for
heterogeneity and endogeneity to the frontier estimators for zero shares and
multi-sided markets.

```{mermaid}
%%| label: fig-discrete-lineage
%%| fig-cap: "Lineage of discrete-choice demand models. Each arrow relaxes a limitation of the parent: random coefficients break the IIA restriction, BLP adds price endogeneity, and the frontier estimators handle zero shares and platform externalities."
flowchart TD
    A[Random utility<br/>McFadden] --> B[Multinomial logit MNL<br/>closed form, but IIA]
    B --> C[Nested logit / GEV<br/>relax IIA across groups]
    B --> D[Multinomial probit<br/>flexible correlation]
    B --> E[Random-coefficients logit<br/>heterogeneous tastes]
    E --> F[BLP<br/>+ price endogeneity, share inversion]
    F --> G[Zero-share estimator<br/>pairwise differencing]
    F --> H[Multi-sided markets<br/>+ cross-side externalities]
```

### Random Utility and the Multinomial Logit {#sec-io-mnl}

@mcfadden1986choice frames market research through **economic choice theory**: a
consumer is an "optimizing black box" mapping inputs (product attributes, socioeconomic
features, market information, history, constraints) to outputs (purchase decisions,
consumption levels). The behavioral premise is that individuals maximize a preference
order influenced by perceptions, attitudes, and past decisions, with commodities
represented as bundles of attributes. The intellectual debt is explicit: Thurstone's
1927 *random utility* idea—choice of the momentarily most beneficial alternative—was
revived by Luce in 1959 and married, by the mid-1960s, to a *multidimensional
attribute* representation of goods, producing the multinomial logit. McFadden's program
is to model the cognitive process inside the black box using experimental data on
consumer cognition, because historical field data may not reveal how consumers would
respond to genuinely new product designs. The constructs that populate the box—product
perceptions, general attitudes, preferences, decision protocols, and behavioral
intentions—are measured by psychometric instruments (multidimensional scaling for
perceptions, factor analysis for tastes, verbal protocols for decision-making, conjoint
analysis for preferences), and the central challenge is to turn those psychometric
indicators into market forecasts that can be validated. Laboratory elicitation is
attractive for its precise control, ability to study not-yet-marketed innovations, and
low cost, but only field validation is fully convincing, and the literature warns that
incentive-poor experiments may mispredict real behavior.

The **multinomial logit** (MNL) is the workhorse random-utility model. With a choice
set $C = \{1, \dots, M\}$ and scale values (strict utilities) $V_i$, the probability of
choosing $i$ is

$$
P_C(i) = \frac{\exp(V_i)}{\sum_{j \in C} \exp(V_j)}.
$$ {#eq-mnl}

The scale values are functions of alternatives' attributes, can interact with
individual characteristics and choice-set features, and usually take an additively
separable linear form $V_i = \mathbf{X}_i \boldsymbol{\beta}$. The closed form follows
from assuming the unobserved utility component is i.i.d. type-I extreme value (double
exponential). The MNL inherits Luce's **independence of irrelevant alternatives**
(IIA) axiom,

$$
\frac{P_A(i)}{P_A(j)} = \frac{P_C(i)}{P_C(j)}
\qquad \forall\, i, j \in A \subseteq C,
$$ {#eq-iia}

which states that the relative odds of two alternatives do not depend on what else is
available. IIA is a double-edged property. It buys real advantages: inference about
many alternatives from paired comparisons, easy demand forecasting for new
alternatives, a clean link between choice probabilities and ranking data (ideal for
conjoint), and the closed form itself. But it is often unrealistic—it forces a uniform
substitution response to any one alternative's attribute change, at odds with the
clustered similarity patterns of real markets (the "red bus / blue bus" problem), and
this defect, too, traces to the extreme-value error assumption.

The literature addresses IIA in three ways: test for it empirically; modify the MNL so
scale values depend on choice-set features; or switch models. The alternatives include
**multinomial probit** (MNP), which permits arbitrary correlation patterns through
normally distributed utilities at the cost of high-dimensional integration; **nested
MNL**, the marketing standard, which groups similar alternatives so IIA holds only
*within* nests; **generalized extreme value**; and elimination-by-aspects variants
(HEBA, elimination by strategy). Consumer heterogeneity is handled by random-
coefficients MNL (the marketing standard for breaking IIA), fixed-coefficient
specifications, or hierarchical Bayesian models. Estimation is normally by maximum
likelihood, though richer models demand more computation: hierarchical models arrange
alternatives in a preference tree and choose by eliminating clusters, while MNP and
random-coefficients MNP rely on **Monte Carlo simulation** for both estimation and
forecasting. A low-dimensional **factor-analytic** structure on the covariance matrix
keeps MNP tractable, and the *hybrid* model of Westin and Talvitie makes choice
probabilities logit with scale values linear in random taste parameters. The
**simulated moments** estimator of @mcfadden1986method makes high-dimensional MNP
practical: it is consistent and asymptotically normal while requiring only one Monte
Carlo draw per observation, and it supplies computationally feasible instruments and a
covariance estimator for hypothesis testing. Market demand is then the sum of
individual choice probabilities, approximated either by **microsimulation** (Monte
Carlo sampling over the population, the more flexible route, which reflects population
heterogeneity when the model carries fixed-effect parameters) or by segmenting the
population and summing representative probabilities. Forecasting additionally requires
external projections of prices, attributes, demographics, and population size.

A distinctive strand of @mcfadden1986choice incorporates **psychometric data** directly
into the choice model. Perceptions and attitudes—scaled metrically—shape the utility
function and the product-attribute scales, formalized in a latent-variable system that
links measured causes and indicators to a utility-maximizing choice protocol.
Psychometricians favor *fixed-effect* descriptions of individual heterogeneity;
marketing favors *random-effect* models for their parsimony, forecast accuracy, and
flexibility. A practical sequential estimator first regresses the psychometric
indicators to recover the latent variables, then estimates the choice model on the
fitted latents; a more efficient estimator integrates over the latent-variable
distribution by maximum likelihood, with the simulated-moments estimator as a lighter
alternative. **Conjoint analysis** is the experimental engine for this program: it
gathers laboratory choice data in simulated markets plus self-explicated attribute
importances, and—because everything is exogenous by design—it can be used much like
real market data for forecasting, with *choice-based* conjoint the gold standard for
mimicking reality. Its validity hinges on the stability of preferences across task
framings and incentives, the structure of the preference map, and the cognitive
congruence of experimental and market tasks. Because the MNL's IIA structure is a
maintained assumption when conjoint data are summarized this way, it must be tested:
auxiliary regressions detect omitted variables and interactions, and the tests of
McFadden–Tye–Train and Hausman–McFadden gauge IIA, with power depending on the
variables already in the base model and the capacity to discriminate among nesting
structures.

### Differentiated-Products Demand and Price Endogeneity {#sec-io-berry}

The defining obstacle in estimating differentiated-products demand is **price
endogeneity** arising from *unobserved* product characteristics. @berry1994estimating
gives the structural interpretation: a demand error represents unobserved demand
influences—style, design, perceived quality, durability—that the econometrician cannot
measure but that firms observe and price against. Because high-$\xi$ products carry
both higher demand and higher prices, the correlation between price and the demand
error biases naive estimates and can even produce upward-sloping fitted demand curves,
exactly the anomaly Trajtenberg documents in CT scanners and Berry, Levinsohn, and
Pakes confirm in automobiles. For *homogeneous* goods this is a standard endogeneity
problem fixable with instrumental variables on a constant-elasticity system
$\ln(q_j) = \alpha_j + \sum_k \eta_{jk}\ln(p_k) + \epsilon_j$, where $\eta_{jk}$ is the
elasticity of good $j$ to price $k$. But in discrete-choice models price and $\xi$ enter
demand *non-linearly*, so linear IV cannot be applied directly.

There is also a dimensionality problem. A system of $N$ goods has $N^2$ elasticities,
infeasible to estimate freely. Imposing *a priori* restrictions (setting some
cross-elasticities to zero or to common values) is arbitrary and weakly grounded in
theory. The discrete-choice solution is to project demand onto a low-dimensional
**characteristic space**: consumer utility depends on product characteristics and a few
random taste parameters, market demand aggregates individual choices, and all
elasticities are derived from the handful of utility parameters—which also lets the
model predict demand for new and dissimilar products and move freely between aggregate
demand and the underlying utility. The cost is the discrete-choice straitjacket: it
rules out buying multiple items, struggles with dynamics, and imposes parametric
structure. This program descends from McFadden (1974) and Bresnahan's (1987) vertical
model of autos, both of which model utility over characteristics, allow an *outside
good*, and use price-setting firms' first-order conditions—while, like most of the
empirical literature, treating characteristics as exogenous even as prices are
determined within the model.

@berry1994estimating resolves the non-linear IV problem with the **mean-utility
inversion**. The data are $R$ independent markets; market $r$ has $N_r$ single-product
firms; product $j$ in market $r$ has observed characteristics $z_{jr} \in
\mathbb{R}^{K_z}$ (affecting demand $x_j$ and marginal cost $w_j$) and unobserved
characteristics $(\xi_j, \omega_j)$, where $\xi_j$ shifts demand and $\omega_j$ shifts
cost. The unobservables are assumed mean-independent of observed characteristics and
independent across markets, so $(\mathbf{z}, \boldsymbol{\xi}, \boldsymbol{\omega})$ are
exogenous to pricing. Consumer $i$'s utility for product $j$ is $U(x_j, \xi_j, p_j,
\nu_i, \theta_d)$, with $\nu_i$ a consumer-specific term. The leading **random-
coefficients** specification is

$$
u_{ij} = x_j \tilde{\beta} - \alpha p_j + \xi_j + \epsilon_{ij},
$$ {#eq-blp-utility}

where the taste for characteristic $k$ is $\tilde{\beta}_k = \beta_k + \sigma_k
\zeta_{ik}$ with $\beta_k$ the mean taste and $\zeta_{ik}$ i.i.d. standard normal across
consumers and characteristics. Substituting,

$$
u_{ij} = x_j \beta + \xi_j - \alpha p_j + \nu_{ij},
\qquad
\nu_{ij} = \sum_k x_{jk}\,\sigma_k\,\zeta_{ik} + \epsilon_{ij},
$$ {#eq-blp-decomp}

where $\nu_{ij}$ is a mean-zero, heteroskedastic error capturing the random tastes.
Collecting the terms common to all consumers defines the **mean utility**

$$
\delta_j \equiv x_j \beta - \alpha p_j + \xi_j.
$$ {#eq-blp-mean-utility}

The estimator's insight is that observed market shares can be *inverted* to recover the
$\delta_j$, which are linear in the demand error $\xi_j$; instrumental variables can
then be applied to that linear equation. We make the inversion explicit in
@sec-io-zeroshares, where the zero-share problem is also confronted. The companion
papers extend the framework to equilibrium: @berry1995automobile estimates automobile
prices in market equilibrium with supply-side first-order conditions, and
@berry2004differentiated combine micro (individual) and macro (aggregate) data to
sharpen identification of the new-car demand system.

### Zero-Valued Market Shares {#sec-io-zeroshares}

The BLP machinery assumes every product has a strictly positive share, because the
share inversion takes logs. Real retail data violate this constantly:
@dube2021random report that in scanner data, products with literally zero sales are
ubiquitous—30.7% of store-day-product observations in their ice-cream application
record no sale—and these zeros persist even in large consumer populations, so they are
not sampling noise. The standard fixes are *ad hoc* and biased: imputing a tiny share
(say $s_j = 10^{-12}$) makes results sensitive to the arbitrary small number, dropping
zero-share observations discards information and, worse, induces **selection bias**
when zeros are systematic, and common data sources (IRI, Nielsen) silently exclude
zero-sales products. @dube2021random build an estimator that treats the zeros as
informative through a regime-switching selection model and a pairwise-difference GMM.

The demand side is a **random-coefficients logit**. With a consideration set $J = \{1,
\dots, J+1\}$ where $J+1$ is the outside option ($p_{J+1} = 0$), household $h$ with
budget $y$ maximizes $U(\mathbf{q}, y; \theta^h) = \sum_{j=1}^{J+1} (\chi_j^h q_j)
\exp(\alpha^h q_0)$ over discrete consumption, where perceived quality is $\chi_j^h =
\exp(x_j'\beta^h + \xi_j + \epsilon_j^h)$, $x_j$ are exogenous attributes, $\xi_j$ is an
unobserved vertical characteristic, $\beta^h$ are taste parameters, and $\alpha^h$ is
the marginal value of the numeraire. The choice-specific indirect utilities are

$$
u_j^h = \alpha^h(y - p_j) + x_j'\beta^h + \xi_j + \epsilon_j^h,
\qquad
u_{J+1}^h = \alpha^h y + \epsilon_{J+1}^h,
$$ {#eq-dube-utility}

with $\epsilon_j^h$ i.i.d. extreme value and preferences $\theta^h = (\alpha^h,
\beta^h)' \sim F_\theta(\cdot; \Theta)$. The predicted share integrates the logit
probability over the taste distribution,

$$
\pi_j(\mathbf{p}, \mathbf{x}, \xi; \Theta)
= \int \frac{\exp(\alpha^h p_j + x_j'\beta^h + \xi_j)}
{1 + \sum_{k=1}^J \exp(\alpha^h p_k + x_k'\beta^h + \xi_k)}\,
dF_\theta(\theta^h; \Theta),
$$ {#eq-dube-share}

and is always strictly positive, which is exactly why the model alone cannot rationalize
an observed zero.

The fix is a **product-selection** stage. For each product an indicator $d_j = \mathbf{1}
\big(\phi(p_j, x_j, \tilde{\mathbf{w}}_j) + \eta_j > 0\big)$ governs whether $j$ enters
the consumer's choice set, where $\tilde{\mathbf{w}}_j$ are selection-shifting factors
(stock-outs, feature ads, in-store displays) that move a product's *consideration* but
not its consumption value [@mehta2003price], and $\eta_j$ is a continuous selection
shock. Observed shares then follow a **regime switch**,

$$
\pi_j(\mathbf{p}, \mathbf{x}, \xi; \Theta) =
\begin{cases}
\displaystyle \int \frac{\exp(\alpha^h p_j + x_j'\beta^h + \xi_j)}
{1 + \sum_{k \in C} \exp(\alpha^h p_k + x_k'\beta^h + \xi_k)}\, dF_\theta & d_j = 1,\\[10pt]
0 & d_j = 0,
\end{cases}
$$ {#eq-dube-regime}

with $C$ the realized choice set. The econometric danger is **selection on
unobservables**: demand and selection shocks can be correlated, $\mathrm{cov}(\xi_j,
\eta_j) \ne 0$. The motivating example is a locally sourced premium item (high $\xi_j$)
that retailers give scarce shelf space and stock thinly (high non-selection), so quality
and consideration move together; if shelf-space and inventory data are unavailable, the
correlation is unavoidable.

Identification rests on **BLP's moment condition** $E(\xi_j \mid \mathbf{z}_j) = 0$,
where $\mathbf{z}_j$ stacks attributes $x_j$ and excluded price shifters
$\tilde{\mathbf{r}}_j$. @berry1994estimating proves the share map $s_j = \pi_j(\mathbf{p},
\mathbf{x}, \xi; \Theta)$ is invertible in the demand shocks, $\xi(\Theta) = \pi^{-1}
(\mathbf{x}, \mathbf{s}; \Theta)$, and @berry1995 propose the GMM estimator built on the
empirical analogue of $E(\xi_j(\Theta) \mid \mathbf{z}_j) = 0$. With any zero share the
inversion fails. Dropping zeros invokes the weaker condition $E[\xi_j \mid \mathbf{z}_j,
d_j = 1] = 0$, valid only under selection-at-random ($\mathrm{cov}(\xi_j, \eta_j) = 0$);
under selection on unobservables it fails and GMM is inconsistent. Reduced-form prices
are modeled additively, $p_j = \Pi(\mathbf{r}_j; g) + \upsilon_j$ with supply shock
$\upsilon_j$, which is reasonable under cost-plus grocery pricing and supports a
**control-function** correction for endogenous prices.

The solution is **pairwise differencing**. Assume the demand shock decomposes through
the non-selection propensity $\mu_j$ as $\xi_j = \iota(\mu_j) + \zeta_j$ with $E[\zeta_j
\mid \mu_j, \mathbf{z}] = 0$ for a smooth increasing $\iota(\cdot)$. Then for a pair of
products with equal propensities the selection terms cancel, yielding the moment

$$
E[\xi_i - \xi_j \mid \mathbf{z}_i, \mathbf{z}_j, \mu_i = \mu_j, d_i = d_j = 1] = 0.
$$ {#eq-dube-moment}

In the homogeneous-logit case the inverted demand is $\ln(s_j/s_0) = \alpha p_j + x_j'
\beta + \iota(\mu_j) + \zeta_j$, and differencing two products with $\mu_i = \mu_j$
eliminates the unknown selection function,

$$
\ln(s_i/s_0) - \ln(s_j/s_0) = \alpha(p_i - p_j) + (x_i - x_j)'\beta + (\zeta_i - \zeta_j),
$$ {#eq-dube-diff}

so the conditional moment $E[\zeta_i - \zeta_j \mid \cdots] = E[\xi_i - \xi_j \mid \cdots]
= 0$ identifies $(\alpha, \beta)$ without ever logging a zero share. The price is that the
utility *intercept* is differenced away and cannot be identified.

Estimation is a three-step procedure. **Stage 1** estimates the pricing equation $p_j =
\Pi(\mathbf{r}_j; g) + \upsilon_j$ (known functional form) with a $\sqrt{n}$-consistent
$\hat{g}$, forms residuals $\hat{\upsilon}_j = p_j - \Pi(\mathbf{r}_j; \hat{g})$, and
builds the control vector $\mathbf{w}_j' = (\hat{\upsilon}_j, \tilde{\mathbf{w}}_j')$;
under the distributional exclusion restriction, $p_j$ and $x_j$ are excluded from
$\mathbf{w}_j$. **Stage 2** estimates the selection propensity nonparametrically by
Nadaraya–Watson kernel regression,

$$
\hat{\mu}(\mathbf{w}_j) = 1 - \frac{\sum_k d_k\, H\!\big(\tfrac{\mathbf{w}_k - \mathbf{w}_j}{b_n}\big)}
{\sum_k H\!\big(\tfrac{\mathbf{w}_k - \mathbf{w}_j}{b_n}\big)},
$$ {#eq-dube-nw}

with kernel $H$ and bandwidth $b_n$. **Stage 3** is the **pairwise-difference weighted
GMM** (PDWGMM) estimator. Writing $\Delta$ for the differencing operator over product
pairs, $g_{i,j}(\Theta) = \Delta \mathbf{z}_{i,j}\,\Delta \xi_{i,j}(\Theta)$, kernel
weights $\hat{w}_{i,j} = h_{n_d}^{-1}\kappa\big((\hat{\mu}_i - \hat{\mu}_j)/h_{n_d}\big)$
that upweight pairs with similar propensities, and a GMM weight matrix $\Phi$, the
estimator solves

$$
\Theta^{\text{PDWGMM}} = \arg\min_{(\Theta, \xi)}
\frac{2}{n(n-1)} \sum_{i=1}^{n-1} \sum_{j=i+1}^{n}
\hat{w}_{i,j}\, \mathbf{g}_{i,j}(\Theta)'\, \Phi\, \mathbf{g}_{i,j}(\Theta)
\quad \text{s.t.}\quad \mathbf{s} = \pi(\mathbf{p}, \mathbf{x}, \xi; \Theta),
$$ {#eq-dube-pdwgmm}

with the share constraint imposed only where shares are positive.

Monte Carlo evidence sharpens the stakes. Under **selection on unobserved demand**
(scenario a), PDWGMM recovers the mean parameters accurately while the drop-zero
estimator overstates the price coefficient by roughly 150%. When zeros are **missing at
random** (scenario b) both estimators are accurate, and PDWGMM loses no precision. With
**strictly positive shares** (scenario c) PDWGMM needs no first two stages and matches
the drop-zero estimates. In the ice-cream application—nonpremium brands (Breyers,
Dean's, Edy's, store brand), the 48-oz size that is 98.5% of sales, two stores,
Thursdays, January 2013–June 2014, 14,907 product-store observations—price endogeneity
is instrumented by unit cost and its square, and consideration is instrumented by UPC-
and brand-level feature advertising. Across six estimators (random-coefficients and
homogeneous logit, each with PDWGMM, drop-zero, and where applicable Heckman and impute
corrections), the imputed-share correction is unstable, PDWGMM's mean price coefficient
is significantly *lower* (demand more elastic) than drop-zero's, the hypothesis of equal
mean elasticities is rejected at 5%, and the Heckman two-step resembles drop-zero. The
methodological contribution is to accommodate zero shares within aggregate
discrete-choice demand—extensible to CES demand—rather than papering over them.

### Multi-Sided Markets {#sec-io-platforms}

Platforms—marketplaces, payment networks, operating systems—create value through
*cross-side network externalities*: a buyer's utility rises with the number of sellers,
and vice versa. @tan2021effects extend the discrete-choice framework to such settings by
letting a consumer's utility incorporate an externality term $\phi_i(\mathbf{x}^k)$ that
depends on participation on the platform's other side. This converts the demand system
into a fixed point—each side's demand depends on the other side's realized
participation—so entry and competition analysis must account for feedback that is absent
in single-sided markets. Adding a competitor or a new platform shifts not only the
direct substitution among options but also the externality each side confers on the
other, which is why competition and entry in multi-sided markets can have welfare
effects that single-sided intuition gets backwards.

## Structural Applications in Marketing {#sec-io-applications}

The discrete-choice and dynamic-pricing apparatus underwrites a body of empirical
marketing work that estimates structural primitives and runs managerial
counterfactuals. Four applications illustrate the range.

**Dynamic pricing of new products.** @cosguner2022dynamic confront a limitation of the
Generalized Bass Model (GBM): although the Bass diffusion model is the standard new-
product sales forecaster, the GBM extension responds only to *percentage* changes in
marketing variables, not their levels, so it cannot prescribe an optimal *launch
price*. The authors propose utility-based generalizations—the Bass–Gumbel and
Bass–Logit diffusion models (BGDM, BLDM)—that ground diffusion in discrete choice and
therefore prescribe both the launch price and the post-launch price path. Across four
product categories both models outperform the GBM in predicting new-product sales, the
paper supplies methods to estimate them when no prior sales data exist, contrasts the
BLDM's optimal pricing policy with the GBM's across parameter ranges, and offers a
computationally light routine—extensible to competitive, multiproduct settings—for
managers to set dynamic pricing and advertising.

**Behavioral price discrimination in channels.** @cosguner2017behavioral study
behavioral price discrimination (BPD) when consumers face switching costs, so price
elasticity varies *over time* and not only across customers as in prior work. Estimating
a dynamic pricing model on cola-category data, they find that retailers profit more by
*delegating* coupon customization and data analytics to manufacturers than by managing
coupons themselves, that retailers can profit substantially by selling their customer
databases to manufacturers, and that—despite gaining data access—manufacturers earn
*lower* profits, evidence that customer information confers channel power on the
retailer. Strikingly, conditioning BPD on only the *most recent* purchase is a simple,
cheap policy that materially shifts profitability, so customer information drives channel
dynamics even when used coarsely.

**Email contact policy.** @zhang2017dynamically address an understudied but profitable
channel by modeling customer email-open behavior jointly with purchases in a unified
hidden-Markov and copula framework, using data from a U.S. home-improvement retailer.
Two findings matter for practice: active email readers are not necessarily active
buyers (and vice versa), and email *volume* affects short- and long-run profitability
*non-linearly*. Their decision-support system locates an interior optimum—around seven
emails—where deviating is costly: sending four rather than seven loses about 32% of
per-customer lifetime profit, and sending ten loses about 16%. Over- and under-
communicating both destroy value, so frequency must be tuned, not maximized.

**Online word of mouth and sales.** @chintagunta2010effects estimate the effect of
nationwide online user reviews on *local* box-office performance at the
designated-market-area (DMA) level, confronting three identification threats: spatial
aggregation across heterogeneous markets, serial correlation from staged film rollout,
and other unobserved serial correlation. Using daily ticket sales for 148 of 874 films
over 16 months with Yahoo! Movies ratings, and instrumenting ratings with the staged
release across markets, they find that review **valence** (sentiment), not volume,
drives local sales—reversing the volume-dominates conclusion of studies that use
*aggregated national* data, which their disaggregated design shows to be an artifact of
aggregation. The rollout-based instrument for review endogeneity generalizes to any
industry with sequential product releases.

These applications share a structural logic that distinguishes IO-style marketing from
reduced-form analysis: each estimates preference and cost primitives that are invariant
to the policy being studied, then re-solves the model under the counterfactual—a new
price path, a different coupon regime, an altered email cadence, a changed information
environment—to compute the managerially relevant outcome.

## Key Takeaways

- Industrial organization supplies the structural primitives—demand systems,
  equilibrium concepts, and identification arguments—that let an analyst run policy
  counterfactuals, in contrast to reduced-form analytics that describe correlations.
- The classical, continuous-quantity approach yields the Slutsky/Roy machinery, the CES
  and quasi-linear special cases, and clean diversion ratios (@eq-diversion); its
  symmetry of cross-effects (@eq-slutsky) is the empirical fingerprint of utility
  maximization, and the homothetic-surplus family of @armstrong2018multiproduct makes
  multiproduct pricing tractable by reducing it to a composite-price problem.
- Vertical differentiation makes quality explicit in demand (@eq-vertical-demand1) and
  shows why a multiproduct firm spaces its quality tiers apart to soften internal
  competition.
- Discrete-choice models recover the full elasticity matrix from a few preference
  parameters and predict demand for new products, but the multinomial logit's IIA
  restriction (@eq-iia) forces unrealistic substitution, motivating random coefficients,
  nested logit, and probit.
- Price endogeneity from unobserved characteristics is the central estimation obstacle;
  @berry1994estimating solves it by inverting market shares to recover mean utilities
  (@eq-blp-mean-utility) that are linear in the demand error and amenable to
  instrumental variables.
- Zero-valued market shares break the share inversion and, when driven by selection on
  unobservables, bias the drop-zero estimator severely; the pairwise-difference GMM of
  @dube2021random (@eq-dube-pdwgmm) restores consistency by differencing out the
  selection function.
- Structural marketing applications—dynamic new-product pricing, behavioral price
  discrimination in channels, email contact policy, and online word of mouth—estimate
  policy-invariant primitives and re-solve the model under counterfactuals to answer
  managerial questions.

## Further Reading

The structural-econometrics toolkit underlying this chapter—GMM, simulation estimation,
and the marketing–finance event-study machinery—is developed in @sec-marketing-finance.
The classical demand foundations connect to the welfare and pricing material throughout
the pricing chapters; the discrete-choice estimators here are the demand-side complement
to the supply-side conduct models. For the homothetic-surplus characterization and its
pricing implications see @armstrong2018multiproduct; for the canonical share-inversion
estimator and its equilibrium extensions, @berry1994estimating, @berry1995,
@berry1995automobile, and @berry2004differentiated; for the choice-theoretic foundations
of market research, @mcfadden1986choice and @mcfadden1986method; and for the frontier
treatment of zero shares and platform competition, @dube2021random and @tan2021effects.
The structural marketing applications—@cosguner2022dynamic, @cosguner2017behavioral,
@zhang2017dynamically, and @chintagunta2010effects—are entry points to the empirical
literature that puts these models to work.
