# Qualitative Research

## Inter-rate reliability methods

Calculation to judge the degree of agreement between the choices made by two or more independent judges

Other packages are

-   `vcd` for visualization

-   `DescTools`

### Percent Agreement

$$
\frac{\text{number of agreement}}{\text{number of total}} \times 100
$$

```{r}
library("irr")
data("diagnoses", package = "irr")
data("anxiety", package = "irr")
head(diagnoses,10)
agree(diagnoses)

# library(vcd)
# # Create the plot
# p <- agreementplot(anxiety)
```

### Cohen's Kappa

[@Cohen_1960]

$$
k = \frac{p_o - p_e}{1 - p_e} = 1 - \frac{1 - p_o}{1- p_e}
$$

where

-   $p_o$ = relative observed agreement among raters

-   $p_e$ = hypothetical probability of chance agreement

strict agreements between raters

appropriate in case of 2 ordinal or nominal varibles

Based on [@Landis_1977]'s guide, we have

| Degree       | Decision                            |
|--------------|-------------------------------------|
| 0.01 -- 0.20 | slight agreement                    |
| 0.21 -- 0.40 | fair agreement                      |
| 0.41 -- 0.60 | moderate agreement                  |
| 0.61 -- 0.80 | substantial agreement               |
| 0.81 -- 1.00 | almost perfect or perfect agreement |

```{r}
# Unweighted kappa for 2 nominal or 2 ordinal categorical
kappa2(diagnoses[, c("rater1", "rater2")], weight = "unweighted") # two ordinal variables only, allows partial agreement


# Weighted kappa ordinal scales
kappa2(diagnoses[, c("rater1", "rater2")], weight = "equal") # linear weightes of the differences

kappa2(diagnoses[, c("rater1", "rater2")], weight = "squared") # squared weightes of the differences



```

p-value less than 0.05, mean that raters agree more than what you would expect by chance.

### Fleiss'kappa

for two or more categorical variables (nominal or ordinal)

for three or more raters

-   0 = no agreement

-   1 = perfect agreement

```{r}
# no assumption of same raters for all subjects
kappam.fleiss(diagnoses[, 1:3])
```

## Krippendorff's Alpha

calculates disagreement among raters.

$$
\frac{\text{observed disagreement}}{\text{expected disagreement}}
$$

Rules of thumb by [@Shelley_1984]

-   $\alpha \ge 0.08$ = good

-   $\alpha \ge 0.667$ = acceptable lower limit

```{r}
nmm <-
    matrix(
        c(
            1,
            1,
            NA,
            1,
            2,
            2,
            3,
            2,
            3,
            3,
            3,
            3,
            3,
            3,
            3,
            3,
            2,
            2,
            2,
            2,
            1,
            2,
            3,
            4,
            4,
            4,
            4,
            4,
            1,
            1,
            2,
            1,
            2,
            2,
            2,
            2,
            NA,
            5,
            5,
            5,
            NA,
            NA,
            1,
            1,
            NA,
            NA,
            3,
            NA
        ),
nrow = 4
)
 # first assume the default nominal classification
 kripp.alpha(nmm)
 # now use the same data with the other three methods
 kripp.alpha(nmm, "ordinal")
 kripp.alpha(nmm, "interval")
 kripp.alpha(nmm, "ratio") 
```

### Kendall's W

Continuous ordinal ratings

```{r}
rtr1 <- c(1, 6, 3, 2, 5, 4)
rtr2 <- c(1, 5, 6, 2, 4, 3)
rtr3 <- c(2, 3, 6, 5, 4, 1)
ratings <- cbind(rtr1, rtr2, rtr3)
library(DescTools)
KendallW(ratings, test=TRUE)
```

### Intraclass correlation coefficients

[@Shrout_1979]

#### Continuous scales

Decision:

-   model:

    -   `"oneway"` when subjects are random effects

    -   `"twoway"` when subjects and raters are random effects

-   type:

    -   `"agreement"` differences in mean ratings among raters are of interest

    -   `"consistency"` is default

-   unit

    -   `"single"` single value

    -   `"average"` mean of several ratings

```{r, message=FALSE}
icc(anxiety,
    model = "twoway", # can be "oneway"
    type = "agreement", # can be "consistency"
    unit = "single" # can be "average"
    )
```

`DescTools` package

```{r}
rtr1 <- c(9, 6, 8, 7, 10, 6)
rtr2 <- c(2, 1, 4, 1,  5, 2)
rtr3 <- c(5, 3, 6, 2,  6, 4)
rtr4 <- c(8, 2, 8, 6,  9, 7)
ratings <- cbind(rtr1, rtr2, rtr3, rtr4)
DescTools::ICC(ratings)
```

### Light's kappa

-   average of Cohen's Kappa if using more than two categorical variables

```{r}
# Light’s kappa: multiple raters
kappam.light(diagnoses[, 1:3])
```
