# Analytical Modeling Seminar {#sec-analytical-modeling-seminar}

*Analytical modeling* in marketing is the practice of building stylized
game-theoretic models—small enough to solve in closed form, rich enough to
isolate one strategic force—and reading their equilibria for managerial and
policy implications. Where the empirical chapters of this book estimate
parameters from data, this chapter studies the logic that generates the data:
why competing firms randomize discounts, when a manufacturer wants its retailers
to compete, why personalization can hurt the firm doing it. The unifying method
is the same throughout. State the players, their information, and the timing;
solve backward for subgame-perfect (or perfect Bayesian) equilibrium; and
characterize how the equilibrium moves with the primitives.

This chapter is built as a **full-semester PhD seminar**—the analytical /
game-theoretic (economic-theory) track taught in the quantitative-modeling
sequence of top doctoral programs. It opens with *tools* before *topics*: a
demanding refresher in non-cooperative game theory and the solution concepts the
literature actually uses (Nash, subgame perfection, Perfect Bayesian
Equilibrium, and the mechanism-design / information-design vocabulary). It then
walks the canonical application areas in the order theory developed them—price
competition, promotions, channels, advertising, product line and signaling,
information and disclosure, salesforce contracting—before turning to the dynamic
and behavioral frontier. An analytical model is *not* a forecast; it is a
controlled thought experiment whose value is the comparative static it isolates.
The seminar habit worth internalizing is to ask, of any result, *which single
assumption is doing the work*—because that is the assumption a referee will
attack and the one a practitioner must check against their own market.

::: {.callout-note}
## Presenting analytical work
Seminar craft and modeling craft are different skills. Three conventions reduce
the friction of presenting theory to a skeptical audience: lead with the
research question and the headline result, not the findings table, so the room
is oriented before it can attack; never deny a known limitation—state it and
give the reason the model still isolates the force of interest; and hold
extensions in reserve, surfacing them only if the audience does not raise them
first. These are rhetorical, not scientific, but they materially change how a
correct result is received.
:::

## Semester arc {#sec-analytical-semester-arc}

The analytical-modeling seminar trains students to *build and solve equilibrium
models of marketing phenomena*—to take a managerial puzzle, strip it to a
tractable game, solve for equilibrium, and extract a testable, often
counterintuitive comparative-static result. The semester therefore opens with
**tools** before **topics**: a fast, demanding refresher in non-cooperative game
theory and the solution concepts the literature uses. Students arriving from a
first-year microeconomic-theory sequence find weeks 1–2 a consolidation; those
from an empirical background find them the steepest climb of the course.

The middle of the semester walks the **canonical application areas in the order
theory developed them**: price competition and promotions; vertical channels and
distribution; advertising as persuasion and as information; product line,
quality, and signaling; information, disclosure, and reviews; and salesforce /
principal–agent contracting. The ordering is pedagogically deliberate—each block
reuses and sharpens the tools introduced earlier (promotions motivate mixed
strategies; channels motivate double-marginalization and contracting machinery;
signaling motivates PBE; salesforce motivates moral hazard and the revelation
principle). Throughout, the seminar insists on the distinction between *modeling
for insight* (the marketing-science tradition: simple models whose comparative
statics overturn a managerial intuition) and *modeling for measurement* (the
structural / empirical-IO tradition of the adjacent seminar). The recurring
refrain is Moorthy's and Wernerfelt's question: *what is the simplest model that
produces this effect, and is the effect robust to the obvious extensions?*

The final third turns to **dynamics and frontier structure**: durable-goods and
forward-looking-consumer pricing; behavior-based price discrimination and
customer recognition; network effects, platforms, and two-sided markets; and
behavioral industrial organization. The course closes with a synthesis week on
research craft—how to find a modeling question, how to know when a model is
"done," how analytical and empirical work speak to each other, and how to
referee theory. Assessment is uniformly a **referee report plus an original
model proposal**, because the terminal skill is not reading theory but producing
it.

## Week 1 — Game-Theory Tools I: Static Games, Nash, and the Modeling Stance {#sec-analytical-week01}

**Topic.** Normal-form games and the "modeling for insight" stance that defines
analytical marketing. *Foundational.*

**Subtopics.** Normal-form games and Nash equilibrium; pure vs. mixed
strategies; best-response and reaction functions; Bertrand vs. Cournot as the
two workhorses of marketing competition; what makes a good marketing-theory
question.

**Methods.** Writing down a game (players, actions, payoffs, timing,
information); solving for pure- and mixed-strategy Nash; comparative statics.

**Key readings.**

- Moorthy, K. S. (1985), "Using Game Theory to Model Competition," *Journal of Marketing Research* 22(3), pp. 262–282. *(DOI to verify)* — the field's manifesto for analytical marketing modeling. [F]
- Tirole, J. (1988), *The Theory of Industrial Organization*, chs. 0–5, MIT Press. (book) — standard IO background on competition and the solution concepts the literature reuses. [F]
- Fudenberg, D. and Tirole, J. (1991), *Game Theory*, selected chapters, MIT Press. (book) — the reference grammar for everything that follows. [F]

**Debate.** Insight vs. realism; the role of functional-form assumptions; when a
"result" is an artifact of the demand specification rather than a feature of the
strategic environment.

## Week 2 — Game-Theory Tools II: Dynamic Games, Information, and Refinement {#sec-analytical-week02}

**Topic.** Extensive-form games, incomplete information, and the equilibrium
refinements modern marketing theory relies on. *Foundational.*

**Subtopics.** Extensive form and backward induction; subgame perfection; games
of incomplete information and Bayesian Nash equilibrium; Perfect Bayesian and
sequential equilibrium; signaling and belief refinements (intuitive criterion);
a first look at mechanism/information design as the modern toolkit.

**Methods.** Solving signaling games; constructing and refining off-path
beliefs; reading an information-design argument.

**Key readings.**

- Fudenberg, D. and Tirole, J. (1991), *Game Theory* (dynamic games, PBE, refinements), MIT Press. (book) — the canonical treatment of refinement. [F]
- Iyer, G. and Singh, S. (2022), "Persuasion Contest: Disclosing Own and Rival Information," *Marketing Science* 41(4), pp. 682–709, [doi:10.1287/mksc.2021.1333](https://doi.org/10.1287/mksc.2021.1333) — a frontier illustration of how Bayesian-persuasion / information-design ideas enter competitive marketing; the "where the tools are going" capstone. [R]

**Debate.** Multiplicity of equilibria and the legitimacy of refinements;
whether information design is a genuinely new toolkit or a repackaging of
signaling.

## Week 3 — Price Competition and Differentiation {#sec-analytical-week03}

**Topic.** How firms soften price competition through differentiation.
*Foundational.*

**Subtopics.** The Bertrand paradox and its resolutions; horizontal
differentiation (Hotelling / spatial competition) and the principle of
(minimum/maximum) differentiation; vertical differentiation and quality; the
strategic value of *softening* price competition.

**Methods.** Solving location-then-price two-stage games; deriving
differentiation as an equilibrium device.

**Key readings.**

- Hotelling, H. (1929), "Stability in Competition," *Economic Journal* 39(153), pp. 41–57. *(DOI to verify)* — the origin of spatial differentiation. [F]
- Iyer, G. and Kuksov, D. (2012), "Competition in Consumer Shopping Experience," *Marketing Science* 31(6), pp. 913–933, [doi:10.1287/mksc.1120.0734](https://doi.org/10.1287/mksc.1120.0734) — shows when a non-price instrument (experience) differentiates vs. triggers escalation. [R]
- Pazgal, A., Soberman, D., and Thomadsen, R. (2016), "Profit-Increasing Asymmetric Entry," *International Journal of Research in Marketing* 33(1), pp. 107–122, [doi:10.1016/j.ijresmar.2015.08.002](https://doi.org/10.1016/j.ijresmar.2015.08.002) — counterintuitive: entry can raise incumbent profit by relaxing price competition. [R]

**Debate.** Do firms differentiate to avoid competition or to match
heterogeneous tastes? Is "minimum differentiation" robust to the pricing stage?

## Week 4 — Promotions, Price Discrimination, and Mixed-Strategy Pricing {#sec-analytical-week04}

**Topic.** Why competing firms randomize discounts, and when price
discrimination is constrained. *Foundational (Narasimhan) → frontier
(price-matching, fairness).*

**Subtopics.** Why competing firms randomize discounts (loyal vs. switcher
segments); temporal price discrimination and Hi-Lo vs. EDLP; price-matching
guarantees as discrimination vs. collusion; behavioral limits on price
discrimination.

**Methods.** Constructing mixed-strategy equilibria; segment-based
price-discrimination models.

**Key readings.**

- Narasimhan, C. (1988), "Competitive Promotional Strategies," *Journal of Business* 61(4), pp. 427–449, [doi:10.1086/296442](https://doi.org/10.1086/296442) — the mixed-strategy rationale for randomized promotions [@narasimhan1988b]. [F]
- Chen, Y., Narasimhan, C., and Zhang, Z. J. (2001), "Consumer Heterogeneity and Competitive Price-Matching Guarantees," *Marketing Science* 20(3), pp. 300–314, [doi:10.1287/mksc.20.3.300.9766](https://doi.org/10.1287/mksc.20.3.300.9766) — reframes a common practice as price discrimination, not collusion. [R]
- Feinberg, F. M., Krishna, A., and Zhang, Z. J. (2002), "Do We Care What Others Get? A Behaviorist Approach to Targeted Promotions," *Journal of Marketing Research* 39(3), pp. 277–291, [doi:10.1509/jmkr.39.3.277.19108](https://doi.org/10.1509/jmkr.39.3.277.19108) — social comparison / fairness constrains targeted discounting. [R]
- Varian, H. R. (1980), "A Model of Sales," *American Economic Review* 70(4), pp. 651–659. *(DOI to verify)* — the canonical equilibrium-dispersion model of promotions. [F]

**Debate.** Are promotions discrimination, inventory/competition artifacts, or
behavioral cues? Do price-match guarantees raise or lower prices?

## Week 5 — Channels and Distribution I: Coordination and Double Marginalization {#sec-analytical-week05}

**Topic.** Why decentralized channels underperform and which contracts fix it.
*Foundational.*

**Subtopics.** Double marginalization and channel coordination; two-part
tariffs, quantity discounts, and resale price maintenance as coordinating
contracts; coordinating channels under price *and* non-price competition;
manufacturer control of retail competition.

**Methods.** Vertical-contracting models; designing wholesale contracts to align
incentives; manufacturer-Stackelberg leadership.

**Key readings.**

- Jeuland, A. P. and Shugan, S. M. (1983), "Managing Channel Profits," *Marketing Science* 2(3), pp. 239–272. *(DOI to verify)* — the foundational channel-coordination paper. [F]
- Iyer, G. (1998), "Coordinating Channels Under Price and Nonprice Competition," *Marketing Science* 17(4), pp. 338–355, [doi:10.1287/mksc.17.4.338](https://doi.org/10.1287/mksc.17.4.338) — the multi-instrument result: the optimal degree of retail competition differs across price and service. [F]
- Moorthy, K. S. (1987), "Managing Channel Profits: Comment," *Marketing Science* 6(4), pp. 375–379. *(DOI to verify)* — the canonical refinement of the coordination argument. [F]

**Debate.** Which contractual instruments achieve coordination, and why is
coordination so often *not* observed in practice?

## Week 6 — Channels and Distribution II: Power, Conflict, and the Retail-Media Frontier {#sec-analytical-week06}

**Topic.** Channel power, bargaining, and the modern retail-media turn.
*Foundational → frontier.*

**Subtopics.** Channel power and bargaining; vertical restraints and
exclusivity; private labels and the manufacturer–retailer game; the modern
"retail media" / in-store-media turn.

**Methods.** Bargaining solutions in vertical settings; dynamic channel pricing.

**Key readings.**

- Dukes, A. and Liu, Y. (2010), "In-Store Media and Distribution Channel Coordination," *Marketing Science* 29(1), pp. 94–107, [doi:10.1287/mksc.1080.0483](https://doi.org/10.1287/mksc.1080.0483) — in-store-media revenue alters retailer incentives and can mitigate double marginalization (an early theory of retail media networks). [R]
- Cosguner, K., Chan, T. Y., and Seetharaman, P. B. (2018), "Dynamic Pricing in a Distribution Channel in the Presence of Switching Costs," *Management Science* 64(3), pp. 1212–1229, [doi:10.1287/mnsc.2016.2649](https://doi.org/10.1287/mnsc.2016.2649) — bridges channels and dynamics: inertial demand lets the retailer raise margins. [R]
- McGuire, T. W. and Staelin, R. (1983), "An Industry Equilibrium Analysis of Downstream Vertical Integration," *Marketing Science* 2(2), pp. 161–191. *(DOI to verify)* — the classic case for decentralization as a strategic commitment. [F]

**Debate.** Who holds channel power and how is it modeled (Nash bargaining vs.
Stackelberg)? Is retail media a coordination device or a rent-extraction tool?

## Week 7 — Advertising: Persuasion, Information, and Targeting {#sec-analytical-week07}

**Topic.** Advertising as persuasion vs. information, and the strategic logic of
targeting. *Foundational (Nelson) → frontier (targeting, cheap talk).*

**Subtopics.** Advertising as persuasion vs. information vs. complement;
informative advertising and price competition; the *targeting* of advertising
and its effect on differentiation; advertising as (cheap) talk.

**Methods.** Equilibrium advertising-intensity models; cheap-talk and
verifiable-disclosure games.

**Key readings.**

- Iyer, G., Soberman, D., and Villas-Boas, J. M. (2005), "The Targeting of Advertising," *Marketing Science* 24(3), pp. 461–476, [doi:10.1287/mksc.1050.0117](https://doi.org/10.1287/mksc.1050.0117) — the modern result that advertising less to switchers raises differentiation and softens price competition [@iyer2005]. [F]
- Gardete, P. M. (2013), "Cheap-Talk Advertising and Misrepresentation in Vertically Differentiated Markets," *Marketing Science* 32(4), pp. 609–621, [doi:10.1287/mksc.2013.0772](https://doi.org/10.1287/mksc.2013.0772) — when unverifiable quality claims can still credibly transmit information. [R]
- Grossman, G. M. and Shapiro, C. (1984), "Informative Advertising with Differentiated Products," *Review of Economic Studies* 51(1), pp. 63–81. *(DOI to verify)* — the workhorse model of informative advertising and price competition. [F]

**Debate.** Is advertising informative, persuasive, or a complementary good?
When does targeting intensify vs. soften competition?

## Week 8 — Product Line, Quality, and Signaling {#sec-analytical-week08}

**Topic.** Designing a product line and signaling unobservable quality.
*Foundational.*

**Subtopics.** Product-line design and self-selection (second-degree price
discrimination / damaged goods); quality choice; price and advertising as
*signals* of quality; product variety with consumer evaluation costs.

**Methods.** Mechanism-design / self-selection (incentive-compatibility,
single-crossing); separating vs. pooling signaling equilibria.

**Key readings.**

- Villas-Boas, J. M. (2009), "Product Variety and Endogenous Pricing with Evaluation Costs," *Management Science* 55(8), pp. 1338–1346, [doi:10.1287/mnsc.1090.1024](https://doi.org/10.1287/mnsc.1090.1024) — fewer products can commit the firm not to extract all surplus when evaluation is costly. [F]
- Moorthy, K. S. (1984), "Market Segmentation, Self-Selection, and Product Line Design," *Marketing Science* 3(4), pp. 288–307. *(DOI to verify)* — the foundational self-selection model of the product line. [F]
- Milgrom, P. and Roberts, J. (1986), "Price and Advertising Signals of Product Quality," *Journal of Political Economy* 94(4), pp. 796–821. *(DOI to verify)* — the canonical model of dissipative signaling of quality. [F]

**Debate.** When is dissipative signaling (e.g., uninformative ad spend)
credible? How fine should a product line be, and is cannibalization a bug or a
feature?

## Week 9 — Information, Disclosure, and Online Reviews {#sec-analytical-week09}

**Topic.** Strategic disclosure, manipulated word of mouth, and information
design in markets. *Frontier (with foundational disclosure roots).*

**Subtopics.** Verifiable disclosure and unraveling; strategic / selective
disclosure by firms and media; manipulation of word of mouth and reviews;
information design / Bayesian persuasion in markets.

**Methods.** Disclosure/unraveling arguments; persuasion as choice of a signal
structure.

**Key readings.**

- Mayzlin, D. (2006), "Promotional Chat on the Internet," *Marketing Science* 25(2), pp. 155–163, [doi:10.1287/mksc.1050.0137](https://doi.org/10.1287/mksc.1050.0137) — firms post disguised promotional chat; rational consumers partially undo it. [F]
- Zhu, Y. and Dukes, A. (2015), "Selective Reporting of Factual Content by Commercial Media," *Journal of Marketing Research* 52(1), pp. 56–76, [doi:10.1509/jmr.12.0379](https://doi.org/10.1509/jmr.12.0379) — endogenous selective (not false) reporting by ad-funded media. [R]
- Iyer, G. and Singh, S. (2022), "Persuasion Contest: Disclosing Own and Rival Information," *Marketing Science* 41(4), pp. 682–709, [doi:10.1287/mksc.2021.1333](https://doi.org/10.1287/mksc.2021.1333) — competitive information design. [R]
- Kamenica, E. and Gentzkow, M. (2011), "Bayesian Persuasion," *American Economic Review* 101(6), pp. 2590–2615. *(DOI to verify)* — the foundational statement of persuasion as the choice of a signal structure. [F]

**Debate.** Does information unravel in equilibrium? When can a sender benefit
from committing to a noisy signal?

## Week 10 — Salesforce, Incentives, and Principal–Agent Theory {#sec-analytical-week10}

**Topic.** Optimal sales compensation under moral hazard, and the structural
turn. *Foundational (Basu–Lal–Srinivasan–Staelin) → frontier (structural
compensation).*

**Subtopics.** Moral hazard and optimal sales-compensation contracts; quotas,
commissions, and bonuses; multitasking and contract uniformity; the
"estimate-then-optimize-then-implement" structural turn.

**Methods.** Principal–agent contracting (IC and IR constraints); multitasking;
bridging theory to estimable dynamic models.

**Key readings.**

- Misra, S. and Nair, H. S. (2011), "A Structural Model of Sales-Force Compensation Dynamics: Estimation and Field Implementation," *Quantitative Marketing and Economics* 9(3), pp. 211–257, [doi:10.1007/s11129-011-9096-1](https://doi.org/10.1007/s11129-011-9096-1) — recovers agents' dynamic response to nonlinear incentives and redesigns the plan in the field [@misra2011structural]. [R]
- Daljord, Ø., Misra, S., and Nair, H. S. (2016), "Homogeneous Contracts for Heterogeneous Agents: Aligning Sales Force Composition and Compensation," *Journal of Marketing Research* 53(2), pp. 161–182, [doi:10.1509/jmr.14.0018](https://doi.org/10.1509/jmr.14.0018) — why uniform contracts can be an equilibrium given selection. [R]
- Basu, A. K., Lal, R., Srinivasan, V., and Staelin, R. (1985), "Salesforce Compensation Plans: An Agency Theoretic Perspective," *Marketing Science* 4(4), pp. 267–291. *(DOI to verify)* — the foundational agency model of sales compensation. [F]

**Debate.** How rich should real contracts be relative to theory's predictions?
Why are observed plans simpler than the optimum?

## Week 11 — Dynamic Models I: Durable Goods, Forward-Looking Consumers, and Learning {#sec-analytical-week11}

**Topic.** Pricing against a forward-looking buyer and the seller's own future.
*Foundational (Coase, Bass) → frontier (consumer learning, dynamic pricing).*

**Subtopics.** The Coase conjecture and durable-goods pricing; intertemporal
price discrimination with forward-looking consumers; new-product diffusion and
dynamic pricing; consumer learning and endogenous loyalty.

**Methods.** Dynamic optimization / Markov-perfect equilibrium; the difference
between open-loop and closed-loop strategies.

**Key readings.**

- Villas-Boas, J. M. (2004), "Consumer Learning, Brand Loyalty, and Competition," *Marketing Science* 23(1), pp. 134–145, [doi:10.1287/mksc.1030.0044](https://doi.org/10.1287/mksc.1030.0044) — forward-looking marginal consumers can be *less* price-sensitive than myopic ones. [F]
- Nair, H. (2007), "Intertemporal Price Discrimination with Forward-Looking Consumers: Application to the US Market for Console Video-Games," *Quantitative Marketing and Economics* 5(3), pp. 239–292, [doi:10.1007/s11129-007-9026-4](https://doi.org/10.1007/s11129-007-9026-4) — the structural counterpart; quantifies declining-price paths for durables. [R]
- Coase, R. H. (1972), "Durability and Monopoly," *Journal of Law and Economics* 15(1), pp. 143–149. *(DOI to verify)* — the conjecture that a durable-goods monopolist competes with its own future. [F]

**Debate.** How binding is the Coase conjecture with commitment devices? Myopic
vs. rational-expectations consumers in pricing models.

## Week 12 — Dynamic Models II: Behavior-Based Price Discrimination and Customer Recognition {#sec-analytical-week12}

**Topic.** Recognizing past customers and the strategic perils of
personalization. *Frontier.*

**Subtopics.** Customer recognition and price cycles; behavior-based price
discrimination (BBPD); poaching vs. retention; behavior-based *advertising*; the
strategic perils of personalization.

**Methods.** Two-period (and infinite-horizon) recognition games; mapping
privacy/data regimes to equilibrium prices.

**Key readings.**

- Villas-Boas, J. M. (2004), "Price Cycles in Markets with Customer Recognition," *RAND Journal of Economics* 35(3), pp. 486–501, [doi:10.2307/1593704](https://doi.org/10.2307/1593704) — endogenous price cycles over the customer relationship. [F]
- Zhang, J. (2011), "The Perils of Behavior-Based Personalization," *Marketing Science* 30(1), pp. 170–186, [doi:10.1287/mksc.1100.0607](https://doi.org/10.1287/mksc.1100.0607) — personalization can intensify competition and erode profits. [R]
- Chen, Y., Narasimhan, C., and Zhang, Z. J. (2001), "Individual Marketing with Imperfect Targetability," *Marketing Science* 20(1), pp. 23–41, [doi:10.1287/mksc.20.1.23.10201](https://doi.org/10.1287/mksc.20.1.23.10201) — when better targeting is "win-win" vs. cut-throat. [R]
- Shen, Q. and Villas-Boas, J. M. (2018), "Behavior-Based Advertising," *Management Science* 64(5), pp. 2047–2064, [doi:10.1287/mnsc.2016.2719](https://doi.org/10.1287/mnsc.2016.2719) — BBPD's advertising analogue. [R]

**Debate.** Does data-driven targeting help or hurt the firms that use it? What
is the welfare and privacy verdict on BBPD?

## Week 13 — Network Effects, Platforms, and Two-Sided Markets {#sec-analytical-week13}

**Topic.** Network effects, tipping, and the platform-design problem. *Frontier
(with foundational network-effects roots).*

**Subtopics.** Direct and indirect network effects; tipping and winner-take-all
dynamics; two-sided pricing and the platform-design problem; commercial-network
formation.

**Methods.** Two-sided pricing models; coordination games and tipping;
network-formation equilibria.

**Key readings.**

- Dubé, J.-P., Hitsch, G. J., and Chintagunta, P. K. (2010), "Tipping and Concentration in Markets with Indirect Network Effects," *Marketing Science* 29(2), pp. 216–249, [doi:10.1287/mksc.1090.0541](https://doi.org/10.1287/mksc.1090.0541) — measures tipping from indirect network effects (video-game consoles). [R]
- Katona, Z. and Sarvary, M. (2008), "Network Formation and the Structure of the Commercial World Wide Web," *Marketing Science* 27(5), pp. 764–778, [doi:10.1287/mksc.1070.0349](https://doi.org/10.1287/mksc.1070.0349) — equilibrium link-buying explains web specialization [@katona2008]. [R]
- Wilbur, K. C. (2008), "A Two-Sided, Empirical Model of Television Advertising and Viewing Markets," *Marketing Science* 27(3), pp. 356–378, [doi:10.1287/mksc.1070.0303](https://doi.org/10.1287/mksc.1070.0303) — the two-sided media-market modeling benchmark. [R]
- Katz, M. L. and Shapiro, C. (1985), "Network Externalities, Competition, and Compatibility," *American Economic Review* 75(3), pp. 424–440. *(DOI to verify)* — the foundational statement of network externalities [@katz1985network]. [F]

**Debate.** When do markets tip? Who should subsidize which side, and does
platform competition discipline pricing?

## Week 14 — Behavioral Industrial Organization and Research Craft (Synthesis) {#sec-analytical-week14}

**Topic.** Non-standard consumers in equilibrium, and the craft of producing and
refereeing theory. *Frontier.*

**Subtopics.** Models with non-standard consumers—reference dependence, fairness,
limited attention, naïveté/present bias; how behavioral assumptions change
equilibrium predictions; **research craft**: finding a question, knowing when a
model is "done," theory ↔ empirics, and refereeing.

**Methods.** Embedding behavioral primitives in equilibrium models; reading and
writing a referee report; constructing an original modeling proposal.

**Key readings.**

- Zhu, Y. and Dukes, A. (2017), "Prominent Attributes Under Limited Attention," *Marketing Science* 36(5), pp. 683–698, [doi:10.1287/mksc.2017.1037](https://doi.org/10.1287/mksc.2017.1037) — firms make attributes prominent to exploit limited attention, reshaping price competition. [R]
- Misra, K., Schwartz, E. M., and Abernethy, J. (2019), "Dynamic Online Pricing with Incomplete Information Using Multiarmed Bandit Experiments," *Marketing Science* 38(2), pp. 226–252, [doi:10.1287/mksc.2018.1129](https://doi.org/10.1287/mksc.2018.1129) — learning-and-earning pricing where the firm, not just the consumer, is the learner. [R]
- Dubé, J.-P. and Misra, S. (2023), "Personalized Pricing and Consumer Welfare," *Journal of Political Economy* 131(1), pp. 131–189, [doi:10.1086/720793](https://doi.org/10.1086/720793) — a capstone tying theory, ML, and welfare. [R]
- DellaVigna, S. (2009), "Psychology and Economics: Evidence from the Field," *Journal of Economic Literature* 47(2), pp. 315–372. *(DOI to verify)* — the survey that frames behavioral primitives for equilibrium modeling. [F]

**Debate.** Do behavioral models *predict* better or merely *rationalize*? How
should analytical and empirical marketing inform each other?

## Foundational vs. frontier at a glance {#sec-analytical-foundational-frontier}

| Block | Foundational core (the "must-know") | Frontier extensions (active research) |
|---|---|---|
| Tools (Wk 1–2) | Nash, SPE, PBE, signaling refinements | Bayesian persuasion / information design |
| Pricing & promotions (Wk 3–4) | Hotelling differentiation; Narasimhan mixed-strategy promotions | fairness / social-comparison limits; price-matching as discrimination |
| Channels (Wk 5–6) | Jeuland–Shugan / Iyer coordination; double marginalization | retail media; dynamic channel pricing with switching costs |
| Advertising (Wk 7) | Iyer–Soberman–Villas-Boas targeting; informative-advertising models | cheap-talk / disclosure advertising |
| Product line & signaling (Wk 8) | Moorthy self-selection; Milgrom–Roberts signaling | endogenous variety with evaluation costs |
| Information & disclosure (Wk 9) | unraveling; Mayzlin promotional chat | competitive information design |
| Salesforce / P-A (Wk 10) | Basu–Lal–Srinivasan–Staelin; Holmström | structural compensation, field implementation |
| Dynamics (Wk 11–12) | Coase, Bass; Villas-Boas learning & price cycles | forward-looking-consumer structural models; BBPD & personalization perils |
| Platforms (Wk 13) | Katz–Shapiro / Rochet–Tirole network effects | empirical tipping; two-sided media models |
| Behavioral IO (Wk 14) | reference dependence, limited attention | prominence/attention competition; bandit pricing; personalized-pricing welfare |

**Reading the map.** A student must be fluent in the left column to pass
qualifying exams; the right column is where dissertations are written. The
seminar's value-add over a pure economics IO course is the marketing-specific
frontier—channels, salesforce, targeting, reviews/WOM, and the
personalization/privacy interface—where marketing scholars, not economists, set
the agenda.

## How this chapter expands {#sec-analytical-expands}

The week-by-week map above is the seminar's backbone; the worked models below are
its frontier face. The remaining sections develop a connected set of
**platform-design** models in full detail—players, information, equilibrium
concept, and the single assumption that breaks each result—so that students see
how the Week 5–6 channel machinery, the Week 13 two-sided-market logic, and the
Week 14 personalization frontier combine in current research. They are
deliberately deeper than a syllabus entry: each is a template for the kind of
original model a seminar referee report and proposal must produce.

Three priorities should guide future revisions. First, **verify and promote the
named-but-uncited canon**: works flagged *(DOI to verify)* above—Moorthy (1984,
1985, 1987), Jeuland–Shugan (1983), McGuire–Staelin (1983),
Basu–Lal–Srinivasan–Staelin (1985), Milgrom–Roberts (1986), Coase (1972),
Hotelling (1929), Varian (1980), Grossman–Shapiro (1984), Kamenica–Gentzkow
(2011), Katz–Shapiro (1985), and DellaVigna (2009)—are the textbook spine of the
seminar and should be confirmed against the Crossref version of record and added
to the bibliography. Second, **track the fast-moving frontier modules**:
information design, behavioral IO, platforms/two-sided markets, and the
personalization–privacy interface each turn over 2–3 version-of-record papers per
revision cycle. Third, **consider promoting two modules to their own chapters**:
platforms / two-sided markets and the analytics–theory bridge (bandit pricing,
ML-driven personalized pricing) are both large enough to stand alone.

## Worked Frontier: Platform Competition and Design {#sec-ams-platforms}

A *platform* is a firm that intermediates transactions between two or more
groups whose participation decisions are interdependent—the canonical
**two-sided market** [@katz1985network]. Modern marketplaces complicate the
classical picture in a way that organizes most of the recent theory: the
platform is not a neutral intermediary but a strategic actor that can host
third-party sellers, compete against them as a first-party retailer, *and*
control the information consumers receive through search and recommendation. The
platform's design levers—fees, entry, ranking, bundling, personalization—are the
object of study.

@fig-ams-platform-roles fixes the three roles a platform plays in the models
that follow. The same firm can occupy more than one node simultaneously, and the
strategic interest of the literature lies precisely in the conflicts of interest
that arise when it does.

```{mermaid}
%%| label: fig-ams-platform-roles
%%| fig-cap: "The three roles a platform can occupy simultaneously. Conflicts of interest arise on every edge: the platform-as-retailer competes with the sellers it hosts; the platform-as-intermediary controls the information consumers use to choose among those sellers."
flowchart TD
    P["Platform"]
    P -->|"hosts, charges fee f or commission φ"| M["Marketplace<br/>(third-party sellers)"]
    P -->|"may enter and sell directly"| R["Retailer<br/>(first-party product)"]
    P -->|"ranks, filters, recommends"| I["Information intermediary<br/>(search & display)"]
    M --> C["Consumers"]
    R --> C
    I --> C
    C -->|"reviews, clicks, purchases"| P
```

### The Threat of Platform Entry: Amazon's Mid-Tail {#sec-ams-midtail}

The defining tension of a hybrid platform is that hosting a successful seller
generates a per-unit fee today but reveals a profitable market the platform may
enter and capture tomorrow. @jiang2011 formalize this as a signaling problem
under the heading of the *mid-tail*: products too popular to be left to
independent sellers (the platform will take them) yet not so commoditized that
the platform sells them itself. The counterintuitive equilibrium behavior is
that a high-demand seller may *suppress* its own sales to avoid revealing demand
and triggering entry.

Consider one seller facing a platform with fixed entry cost $F > 0$. In each of
two periods the seller sets a price $p$ and a service level $e$, and demand is
$$
q(p, e) = \gamma + e - b\,p,
$$ {#eq-ams-demand}
where $\gamma$ is the demand intercept, $b > 0$ the price sensitivity, and
service is costly at the convex rate $s(e) = k e^{2}$. The seller's *type* is its
demand intercept: high, $\gamma = \gamma_H$, with prior probability $\theta$, or
low, $\gamma = \gamma_L$, with probability $1 - \theta$. The seller knows its
type; the platform does not. The platform charges a per-unit fee $f$, constant
across the two periods, and after observing first-period outcomes decides whether
to pay $F$ and enter.

The timing is a two-stage signaling game. In stage one, nature draws the type,
the platform commits to $f$, and the seller chooses whether to sell and, if so,
$(e^{(1)}, p^{(1)})$. In stage two, the platform updates its belief about the
type from observed first-period sales and decides whether to enter; the seller,
if still active, chooses $(e^{(2)}, p^{(2)})$. The solution concept is **perfect
Bayesian equilibrium** (PBE): strategies are sequentially rational given beliefs,
and beliefs are derived from strategies by Bayes' rule wherever possible.

The economics turn on two thresholds. If $F$ is large, entry is unprofitable
regardless of demand and the platform commits to *not* entering; absent the
threat, both seller types provide the same (high) service in both periods,
because there is nothing to hide. If $F$ is small enough that the platform would
enter upon learning $\gamma = \gamma_H$, the high type faces a dynamic
disincentive: high first-period sales betray high demand and invite entry, which
expropriates its second-period profit. Three regimes result.

| Regime | Platform information | Equilibrium behavior |
|---|---|---|
| Full information | Platform knows the type | Platform enters iff demand is high and $F$ is low; first-best service in both periods |
| Commitment not to enter | Type irrelevant to entry | Both types provide high service in both periods; no signaling distortion |
| Incomplete information, entry possible | Platform infers type from sales | Separating: only the high type stays, earns in period 1, zero in period 2. Pooling: the high type mimics the low type's restrained sales to *mask* its demand |

: Three information regimes in the mid-tail entry game of @jiang2011. {#tbl-ams-midtail}

The managerial reading inverts naive intuition. A high-demand mid-tail seller can
be better off *under* the threat of entry than under a platform that commits
never to enter, because the threat lets the platform justify a low per-unit fee
$f$ in expectation—the platform forgoes fee revenue to keep the option of
entering alive—and the low fee benefits both types ex ante. Consumer surplus
follows a predictable arc: entry, when it occurs, reduces surplus early (the
platform exploits its informational advantage) and raises it later. Finally,
**consumer reviews** that proxy for service quality sharpen the platform's
inference and let it raise the optimal fee when the prior probability of a high
type is low, and lower it when that probability is high. The result that breaks
the clean separating equilibrium is the cost of mimicry: pooling survives only
when the high type's loss from restraining first-period sales is smaller than its
expected loss from triggering entry.

### Search Neutrality and Personalized Ranking {#sec-ams-search-neutrality}

When the platform also ranks the sellers it hosts, its ranking rule is a
competitive instrument. *Search neutrality*—a policy requiring the platform to
rank by relevance alone, without favoring its own or any seller's
listing—sounds pro-consumer by analogy with net neutrality. @zou2021 show the
analogy fails: with **personalized** search, neutrality can *soften* price
competition and reduce consumer surplus, even as it improves the relevance of
what consumers see.

The mechanism is that personalized ranking, unconstrained, lets the platform
steer each consumer toward the listing that best fits them, which intensifies
price competition because sellers know a low price can win prominence. Neutrality
removes that lever: rankings become relevance-only and price loses its power to
buy position, so sellers compete less aggressively and prices rise. Crucially the
effect is specific to *online personalized* search; a neutrality rule applied to
an offline store with a uniform display has no analogous price effect, because
there is no personalization to suppress.

Formally, two consumer segments differ in their taste for first- versus
third-party products and search sequentially. Search has a pre-search phase, in
which consumers costlessly learn expected valuations, and a costly inspection
phase, in which they learn exact match values. A consumer's action set at any
point is to buy the prominent product, to stop and not purchase, or to pay the
search cost to inspect a non-prominent product. The game has three stages:
sellers set prices, the platform sets personalized rankings, and consumers
purchase. The platform trades commission earned on a relevance-maximizing ranking
against the sales it can divert by promoting its own first-party product, under
the maintained assumption that first-party sales are more profitable.

The policy lesson is cautionary on two fronts. Search neutrality is hard to
implement for the same reason fair-lending rules are hard to implement without
reference to protected attributes: the prohibited behavior (biased ranking) is
defined by an intent that is not directly observable in the output. And because
neutrality can raise prices while raising relevance, welfare analysis cannot
stop at the relevance margin—the price-competition channel can dominate.

### Strategic Listing with Sponsored and Organic Slots {#sec-ams-strategic-listing}

@long2022 enrich the ranking problem by letting the platform learn product
quality from the *advertising auction* and use it to design the organic results.
Two sellers compete for a query that returns one sponsored and one organic slot.
Consumer preferences combine a **vertical** component (product quality $q_i \sim
U(0,1)$, known only to seller $i$) and a **horizontal** component (personal fit
$\lambda_{ki} \sim U(0,1)$, known only to the platform). The match probability of
product $i$ for consumer $k$ is
$$
m_{ki} = \theta_k\, q_i + (1 - \theta_k)\,\lambda_{ki},
$$ {#eq-ams-match}
where $\theta_k \sim U(0,1)$ is the consumer's private weight on quality versus
fit. The seller knows $q_i$, the platform knows $\lambda_{ki}$, and the consumer
knows $\theta_k$—a three-sided information structure.

The platform's design choice is an **information-sharing weight** $T \in [0,1]$.
It places in the organic slot the seller with the higher
$T\,\tilde q_i + (1 - T)\,\lambda_{ki}$, where $\tilde q_i$ is the quality the
platform *infers* from the seller's bid. The key analytical step is that, in the
seller's optimal bidding equilibrium, the bid $b(q_i)$ is strictly increasing in
$q_i$; therefore the bid is a credible signal and the platform can recover $q_i$
exactly. Consumers, anticipating this, read the sponsored slot as the
higher-quality seller and the organic slot as the seller with the higher
weighted index. The timeline runs: the platform commits to $(T, \phi)$ for
information weight and commission rate $\phi$; sellers join; private types
realize; sellers bid; the platform infers $\tilde q_i$ and ranks; consumers
search and buy.

The central comparative static, illustrated in @fig-ams-listing-revenue,
decomposes the platform's revenue into two channels with opposite shapes in $T$.
Commission revenue, $\phi\,D(T)$, is **concave** in $T$ through an *information*
effect: too small a $T$ overvalues fit and wastes the quality information in
bids, while too large a $T$ overvalues quality and degrades the match. Advertising
revenue, $2\int_0^1\!\int_0^{q_i} b(q_i)\,\mathrm{d}q_i\,\mathrm{d}q_j$, is
**increasing** in $T$ through a *competition* effect: a higher $T$ raises each
seller's probability of winning both slots, intensifying the bidding.

```{r ams-listing-revenue}
#| label: fig-ams-listing-revenue
#| fig-cap: "Stylized revenue channels from the strategic-listing model of Long et al. (2022). Commission revenue is concave in the information weight T (an information effect), while advertising revenue rises in T (a competition effect). The total is maximized at an interior weight."
#| message: false
#| warning: false
T <- seq(0, 1, length.out = 201)
# Stylized functional forms matching the qualitative results: concave
# commission revenue, increasing convex-ish advertising revenue.
commission <- 0.9 - 1.6 * (T - 0.55)^2     # concave, interior peak
advertising <- 0.15 + 0.6 * T^1.3          # increasing in T
total <- commission + advertising

plot(T, total, type = "l", lwd = 2, ylim = range(c(commission, advertising, total)),
     xlab = "Information weight T", ylab = "Platform revenue (stylized)",
     main = "Strategic listing: commission vs. advertising revenue")
lines(T, commission, col = "steelblue", lwd = 2, lty = 2)
lines(T, advertising, col = "firebrick", lwd = 2, lty = 3)
abline(v = T[which.max(total)], col = "grey60", lty = 4)
legend("topleft", bty = "n",
       legend = c("Total", "Commission (concave)", "Advertising (increasing)"),
       col = c("black", "steelblue", "firebrick"), lwd = 2, lty = c(1, 2, 3))
```

Two further results sharpen the design problem. Fixing $T$, total revenue rises
with the commission rate $\phi$ because the commission applies to sales from
*both* slots and so dominates the advertising loss confined to the sponsored
slot. Allowing both levers to vary, and giving sellers an outside option $u_0$,
the optimal commission is $\phi^\star = 1 - u_0 / U(T, 0)$, which *falls* as $T$
rises: a larger $T$ raises bids and advertising fees, so the platform extracts
less through commission. When sellers' outside options are intermediate, the
platform sets $\phi = 0$ and competes purely on the information weight; when they
are large, sellers do not join. Comparison with a benchmark that bans strategic
listing (organic results ignore auction information) shows the platform benefits
from strategic listing overall—lower commission revenue but higher advertising
revenue—provided the commission rate is not too high.

### Personalized Pricing under Network Effects {#sec-ams-personalized-pricing}

Personalization on the *price* side interacts with **network effects** in a way
that reverses the textbook welfare intuition. Absent network effects,
price discrimination lowers the price offered to the low-valuation segment, as
standard. @hajihashemi2021 show that with network effects the sign flips: under a
**fulfilled-expectations equilibrium**—consumers' purchase decisions depend on
the network size they rationally anticipate, and that anticipation must be
correct in equilibrium [@katz1985network]—personalized pricing can *raise* the
equilibrium price for the low-valuation consumer. The intuition is that the
network externality makes each segment's demand depend on the other's
participation, so the firm internalizes a cross-segment complementarity that
standard discrimination ignores; when the network effect is strong, the firm
prices the low segment up to seed the network. The effect vanishes as the network
effect goes to zero, because then the marginal effect of a price cut on the
anticipated reservation value is negligible. A welfare corollary follows:
personalization can *decrease* the unconditional expected surplus of the
low-valuation segment, precisely the group price discrimination is usually
presumed to help.

### Subscription Platforms and Channel Bargaining {#sec-ams-subscription}

A different platform problem arises when the platform bundles *physical services*
from independent vendors into a subscription—ClassPass for fitness, or the
hospitality passes that bundle hotels and resorts. @gal-or2022 ask when such a
platform can profitably enter, and the answer is a channel-coordination result:
entry succeeds only when the platform has a contractual instrument that *softens*
the price competition its presence would otherwise create.

The setup places two vendors on a Hotelling line of unit length with
transportation cost $t$ per unit distance, facing variety-seeking consumers. A
consumer derives gross utility $v$ from one service and $(1+\alpha)v$ from the
variety of both, where $\alpha \in (0,1]$ indexes the preference for variety.
Vendor $i$ charges $p_i$ directly; the platform charges a subscription fee $s$
and may negotiate a quality level $k_i \in [0,1]$ that vendor $i$ delivers to
platform subscribers. A consumer who buys only from vendor $i$ obtains
$$
U(x_i, \text{only } i) = \max\!\big\{\, v - t x_i - p_i,\ \ k_i v - t x_i - s \,\big\},
$$ {#eq-ams-util-single}
choosing between the direct channel and the platform, and a consumer who wants
both obtains
$$
U(x_i, \text{both}) = \max\!\Big\{\, \tfrac{k_1 + k_2}{2}(1+\alpha) v - t - s,\ \
(1+\alpha) v - t - (p_1 + p_2) \,\Big\}.
$$ {#eq-ams-util-both}
The model assumes $(1+\alpha)v - t > 0$ for tractability, so the variety-seeking
consumer always derives positive gross utility net of travel. Vendor $i$'s
market share, derived from the indifferent consumer, is
$$
\mathrm{MS}_i = \frac{t + s - p_i - v\big[\tfrac{k_1 + k_2}{2}(1+\alpha) - 1\big]}{t},
$$ {#eq-ams-ms-vendor}
and the platform's share is the residual demand from variety-seekers,
$$
\mathrm{MS}_p = \frac{v\big[(k_1 + k_2)(1+\alpha) - 2\big] - t - 2s + p_1 + p_2}{t}.
$$ {#eq-ams-ms-platform}

The game has three stages. The platform decides whether to negotiate; it then
bargains *separately* and simultaneously with each vendor over the contract; and
finally all three parties set prices noncooperatively. Each bilateral
negotiation is resolved by the **generalized Nash bargaining solution**, which
requires the outside options. The platform's outside option is zero—it cannot
form without an agreement—while each vendor's outside option $w_i$ is its profit
in the no-platform duopoly, which depends on $(\alpha, v, t)$ through the regime
of competition.

::: {.callout-note}
## Why outside options must be solved first
In bilateral bargaining the surplus to be split is *total surplus minus the sum
of disagreement payoffs*. The vendors' no-platform duopoly profits $w_i$ are
therefore not a side detail but the denominator of the whole analysis: they
determine whether a platform can ever offer both vendors more than they would get
alone. @gal-or2022 select, among multiple no-platform equilibria, the one that
maximizes aggregate vendor profit, because side payments make the vendors care
about the joint pie.
:::

The no-platform duopoly profit $w_i$ takes four forms depending on the
transportation cost relative to valuation, summarized in @tbl-ams-outside.

| Parameter region | Outside option $w_i$ | Competitive reading |
|---|---|---|
| $\tfrac{1}{1+\alpha} < \tfrac{v}{t} < 1$ (market not covered) | $\dfrac{v^2}{4t}$ | Each vendor is a local monopoly; high travel cost insulates them |
| Covered market, single-vendor purchase | $\tfrac{1}{2}\big(v - \tfrac{t}{2}\big)$ | Multiple equilibria with $p_1 + p_2 = 2v - t$ |
| Low travel cost, low variety taste | $\dfrac{t}{2}$ | Intense price competition |
| Most-competitive region | $\dfrac{(\alpha v)^2}{4t}$ | Differentiation collapses to the variety margin |

: No-platform vendor outside options $w_i$ across competitive regimes [@gal-or2022]. {#tbl-ams-outside}

With outside options in hand, each bilateral negotiation maximizes the Nash
product of the parties' net gains. Under a **variable transfer fee** regime, in
which vendor $i$ pays the platform a per-customer fee $m_i$ plus a fixed transfer
$F_i$,
$$
\max_{k_i,\, m_i,\, F_i}\ NB_{ip} =
\big[\mathrm{MS}_i p_i + m_i \mathrm{MS}_p + F_i - w_i\big]^{\beta}
\big[\mathrm{MS}_p(s - m_i - m_j) - F_i - F_j\big]^{1-\beta},
$$ {#eq-ams-nash-transfer}
where $\beta$ is the vendor's bargaining power and $1 - \beta$ the platform's.
Under a **quantity restriction** regime, in which vendor $i$ caps the number of
platform customers it serves at $y_i$,
$$
\max_{k_i,\, y_i,\, F_i}\ NB_{ip} =
\big[\mathrm{MS}_i p_i + F_i - w_i\big]^{\beta}
\big[\mathrm{MS}_p\, s - F_i - F_j\big]^{1-\beta}.
$$ {#eq-ams-nash-quantity}

The headline impossibility result is sharp. With only the two instruments
$(k_i, F_i)$—service quality and a fixed transfer—no agreement is reachable,
because adding the platform's bundle intensifies price competition between the
vendors and destroys the surplus the platform would need to capture
(Proposition 1). The platform can enter only if it adds *one* of two
competition-softening instruments: a per-customer transfer fee $m_i$, or a
quantity cap $y_i$. Both raise the equilibrium prices that would otherwise be
competed down; the per-customer fee does so by raising the vendor's effective
marginal cost of a platform sale, and the quantity cap by directly limiting
platform volume. In equilibrium the negotiated quality is
$k_1 = k_2 = 1$—neither side wants to degrade quality, because doing so only
cannibalizes the channel—and the market is fully covered. Entry requires a
sufficiently high variety parameter $\alpha$ *and* weak vendor competition; these
are not the same condition, because two strongly differentiated vendors can be
local monopolies with no incentive to join unless the platform opens an
otherwise-unserved variety-seeking segment. A final comparison (Proposition 4)
ranks the two regimes: the variable-transfer-fee regime yields higher industry
profit when $\tfrac{9}{4 + 6\alpha} < \tfrac{v}{t}$—low travel cost and high
variety taste—and the quantity restriction dominates otherwise. The general
managerial lesson generalizes beyond subscriptions: a platform that bundles
substitutes can coordinate the channel only when it holds an instrument that
prices the competitive externality its bundle creates.

## Worked Frontier: Intertemporal Pricing and Bundling {#sec-ams-intertemporal}

The second family of models studies a seller's pricing *across time*. The
strategic tension is not between competing firms but between the seller and its
own future: a seller who will cut prices tomorrow teaches forward-looking
consumers to wait today. Three forces recur—the heterogeneity of consumer
willingness to pay, the split between **myopic** consumers (who optimize each
period independently) and **strategic** consumers (who optimize across periods),
and the seller's ability or inability to commit to future prices.

### Bundling across Time {#sec-ams-bundling}

A monopolist sells two products and faces a static choice of bundling strategy:
**pure components** (PC, sell separately), **pure bundling** (PB, sell only the
bundle), or **mixed bundling** (MB, offer both)—the last a form of second-degree
price discrimination. Bundling homogenizes heterogeneous willingness to pay (the
classic Stigler logic) while separate sales price-discriminate; the optimal mix
depends on marginal cost and on the share of strategic consumers. @prasad2017
embed this choice in a two-period problem and characterize when each
*sequence* of strategies is optimal.

Let $\alpha$ be the fraction of myopic consumers and $1 - \alpha$ the strategic
fraction, with the bundle's marginal cost equal to the sum of component costs
$c$. Under a pure-components-then-pure-components (PC–PC) policy, first-period
myopic demand for each product is $(1 - P)\alpha$; second-period myopic demand
at the discounted price $P - \delta$ is $\alpha\delta$; and strategic demand,
which waits for the discount, is $(1 - \alpha)(1 - P + \delta)$. The seller's
first-period profit is $2(P - c)(1 - P)\alpha$ and second-period profit is
$2(P - \delta - c)\big[\alpha\delta + (1 - \alpha)(1 - P + \delta)\big]$.
Optimizing the two prices yields
$$
P_1 = \frac{1 + c}{2} + \frac{(2 - \alpha)(1 - c)}{2(4 - \alpha)},
\qquad
P_2 = \frac{1 + c}{2} - \frac{\alpha(1 - c)}{2(4 - \alpha)},
$$ {#eq-ams-bundle-prices}
with total profit
$$
\Pi = \frac{2(1 - c)^2}{4 - \alpha}.
$$ {#eq-ams-bundle-profit}

The comparative statics deliver the policy map: pure bundling in both periods is
optimal when marginal costs are low; pure components in both periods when costs
are high and consumers are mostly myopic; pure bundling then pure components when
costs are moderate and consumers are mostly strategic; and mixed bundling in both
periods when the market is dominated by strategic consumers. A notable
implication is that the optimal *bundle* price can exceed the sum of optimal
individual prices—bundling is not a discount device but a discrimination device.

A related but distinct problem is the **event bundle**: two events with different
popularity sold over two periods. Events differ from durable products in being
perishable and temporally ordered, and consumers may be uncertain about their
valuation of a future event and about the seller's commitment to its future
price. @kolay2022 let consumer valuations of the less-popular event be
$v_L \sim U(0,1)$ and of the more-popular event $v_H \sim U(0, 1+x)$, where $x$
indexes the popularity gap. The seller chooses to sell the events independently,
as a pure bundle, or under full or partial mixed bundling. The ordering result is
memorable: when the seller controls the sequence, popular events should be saved
for *later*—"save the best for last"—mirroring the result that lower-quality
products should be introduced before higher-quality ones to manage intertemporal
self-cannibalization [@moorthy1992].

### Fairness Concerns and Intertemporal Prices {#sec-ams-fairness-pricing}

A behavioral wrinkle on intertemporal pricing is the consumer's sense of price
*fairness*: a price increase across periods can be perceived as unfair, and that
perception enters the demand function. @diao2019 show that consumer fairness
concerns lower retail prices in *both* periods, while the wholesale price may
*rise* in the first period. The often-cited intuition that a retailer raises the
first-period price and cuts it later holds only under a centralized channel
(manufacturer selling directly to consumers), not under a decentralized one. The
model represents heterogeneity with a fraction $\alpha$ of fairness-concerned
consumers whose disutility from an unfair increase enters the second-period
demand function. A counterintuitive welfare consequence is that second-period
unit sales are *higher* with fairness concerns than without, because the fairness
constraint disciplines the channel and thereby alleviates the
double-marginalization that otherwise depresses output.

### Inventory Controls and Price Rigidity {#sec-ams-inventory}

Why are observed prices often flat over time even when demand shifts? @dana2022
provide an oligopoly answer: with fixed capacity chosen before prices, firms set
prices that are flat or uniform across periods in equilibrium, and **inventory
controls**—sales limits that ration availability—are what make genuine
intertemporal price discrimination possible. Firms first choose capacity, then
compete in prices over two periods. The mechanism is that inventory controls
mitigate price competition by committing each firm to withhold supply, which is
profitable when demand becomes more inelastic over time. The model rationalizes a
pervasive empirical regularity—price rigidity under demand variation—as an
equilibrium outcome rather than a friction.

## Worked Frontier: Algorithmic Fairness as an Economic Constraint {#sec-ams-fairness}

When a firm uses a machine-learning algorithm to screen consumers—approving loans,
say—a *fairness constraint* on the algorithm is a normative input with economic
consequences. @fu2021 compare two constraints widely discussed in the fairness
literature: **equal treatment** (ET), which forbids using group membership in the
decision rule, and **equal opportunity** (EO), which equalizes the true-positive
rate across groups. The firm's strategic choice is how much costly *learning
effort* to invest in a better classifier.

The setup has a risk-neutral firm that accepts or rejects applicants drawn from a
**regular** and a **protected** group, each of unit measure, with a fraction $d$
of "bad" (defaulting) consumers. A good consumer yields $\alpha$, a bad consumer
costs $\beta$, so the group-level expected gains are
$$
\alpha_g = \alpha(1 - d_g), \quad \beta_g = \beta\, d_g, \qquad g \in \{r, p\},
$$ {#eq-ams-fairness-gains}
with $r$ the regular and $p$ the protected group. The firm pays a learning cost
to raise the *separability* of good from bad applicants—indexed by efficiencies
$\gamma_r, \gamma_p$—and sets a group-specific approval threshold subject to
either the ET or the EO constraint.

The results are uniformly unfavorable to the more demanding constraint. The firm
is *less* motivated to invest in learning under EO than under ET, because EO
caps the value of better discrimination in the harder-to-separate protected
group; profit is lower under EO for the same reason; and EO can make *everyone*
worse off—the regular group faces a higher threshold under EO than under ET, and
the firm invests less in the algorithm overall, degrading accuracy for both
groups. The optimal threshold and profit hinge on the protected-group bad-rate
term $\beta_p$. The chapter's general point is methodological: a fairness
criterion is not a free constraint laid over a fixed algorithm but an argument in
the firm's optimization, and equilibrium adjustment of *learning effort* can
reverse the criterion's intended distributional effect.

## Worked Frontier: Decentralized Channels {#sec-ams-channels}

Most platform and pricing models above sit atop a **distribution channel**: an
upstream supplier or manufacturer and a downstream retailer, whose separately
optimized markups create the **double-marginalization** problem. Two contract
forms organize the modern literature—the **wholesale** contract, in which the
supplier sets a wholesale price and the retailer marks it up, and the **agency**
contract, in which the retailer takes a revenue share and the supplier sets the
final price.

### Reactive Capacity and Pull Contracts {#sec-ams-capacity}

@jiang2022effects study a **pull** contract, in which a brand-owning retailer
designs product quality and sets wholesale and retail prices, but the *supplier*
bears the risk of excess production. The question is how an improvement in the
supplier's *reactive capacity*—its ability to produce during the selling
season—is divided between supplier and retailer.

Before the season the retailer designs quality $q$ and offers a wholesale price
$w$, outsourcing production at the supplier's convex cost $k q^2$. At the start of
the season the retailer orders $Q$ units and sets retail price $p$; a consumer of
willingness-to-pay $\theta$ obtains utility $u = q\theta - p$, and a fraction
$\alpha$ of consumers are high-type ($\theta = \theta_H$). Market uncertainty is
about the *composition* of demand—the fraction $\alpha$—not its size. The
supplier's lead time is
$$
T = T_S + \frac{Q}{c},
$$ {#eq-ams-leadtime}
a quantity-independent component $T_S$ plus a quantity-dependent component, and
production must satisfy $T \le A$ for an acceptable delivery time $A$. Reactive
capacity is therefore $\bar Q = (A - T_S)\,c$: it rises with raw capacity $c$ and
with flexibility (lower $T_S$).

The equilibrium partitions the parameter space by the level of reactive capacity.
With **high** reactive capacity the supplier can fulfill in-season orders, so the
retailer anticipates flexible supply and targets only high-valuation consumers at
a high price. With **medium** reactive capacity two opportunistic-targeting
regimes appear: under *stock-out* the retailer sets the wholesale price at the
supplier's marginal cost and the supplier produces a risk-free pre-season
inventory; under *overstock* the retailer raises the wholesale price to induce a
larger pre-season inventory. The extreme case $\alpha_b \ge \theta_L / \theta_H$
(even the bad market has enough high-valuation consumers) makes high-price
targeting optimal regardless of the state. The model's value is in locating
exactly where the inventory risk sits in each regime and how reactive capacity
shifts surplus between the channel partners.

### Wholesale versus Agency {#sec-ams-agency}

@hu2022agency compare the channel efficiency of the two contract forms directly.
With $n \ge 2$ suppliers selling substitutable products through a dominant online
retailer, demand for product $i$ is $q_i(p)$, downward-sloping and symmetric, with
substitutability $\gamma > 0$. Define the demand-sensitivity measure
$$
\lambda(p) = -\frac{q(p)}{q'(p)},
$$ {#eq-ams-lambda}
and assume profit functions are quasi-concave and the elasticity of demand
increases with price. The two contracts differ in who moves and who sets the
final price, summarized in @tbl-ams-contracts.

| | Wholesale contract | Agency contract |
|---|---|---|
| First mover | Supplier announces wholesale price $w_i$ | Retailer sets revenue-share rate $1 - \alpha$ |
| Price setter | Retailer sets retail price $p_i$ | Supplier sets retail price $p_i$ directly |
| Margin structure | Double marginalization (two markups) | Single markup; retailer takes a share |
| Pass-through | Through the retailer's markup | Through the supplier's price |

: Wholesale versus agency contracts in a decentralized channel [@hu2022agency]. {#tbl-ams-contracts}

The analytical payoff is that the agency contract eliminates one of the two
markups—the supplier sets the consumer price directly and the retailer's revenue
share is a proportional, not additive, wedge—so its efficiency relative to the
wholesale contract turns on the curvature of demand through $\lambda(p)$ and on
the intensity of inter-supplier competition $\gamma$. The comparison is the
analytical counterpart to the empirical pass-through literature and explains why
dominant online retailers have migrated many categories from wholesale to agency
terms.

## Worked Frontier: Consumer Search {#sec-ams-search}

The final family of models is the engine beneath most of the platform results:
**sequential consumer search** in the tradition of @wolinsky1986. A consumer who
cannot observe match values without effort searches sellers one at a time, paying
a cost $\tau$ per inspection, and stops by an optimal reservation rule. The
platform's pricing of search—through a referral fee, a filtering tool, or a
blind-buy option—is what connects search theory to platform design.

### Search, Filtering, and the Referral Fee {#sec-ams-filtering}

@jiang2020 study a platform that charges sellers a percentage referral fee and
offers consumers a *filtering* tool. Products are horizontally differentiated; a
consumer learns price and match value only after paying the search cost. The
match value decomposes into a **filterable** and an **unfilterable** component,
$$
M_{ij} = \mu_{ij} + m_{ij}, \qquad \mu_{ij} = \delta\,\hat\mu_{ij},
$$ {#eq-ams-matchvalue}
where $\mu_{ij}$ is the filterable component ($\delta > 0$ a scale factor,
$\hat\mu$ a mean-zero discrete shock with $K$ realizations) and $m_{ij}$ the
unfilterable component; consumer utility is $u_{ij} = M_{ij} - p_i$. The consumer
knows her outside option and the filterable value $\mu_{ij}$ before searching but
learns $m_{ij}$ and the price only by paying $\tau$. The game runs: the platform
sets the referral fee $r \in (0,1)$; sellers set prices to maximize
$(1 - r)p_i - c$; consumers search and buy.

The optimal stopping rule is the standard reservation rule [@wolinsky1986]: the
consumer continues until the expected gain from one more search falls below
$\tau$. The acceptance threshold $\bar m(\tau)$ for the unfilterable component
solves
$$
\int_{\bar m(\tau)}^{m_{\max}} \big(m - \bar m(\tau)\big)\,\mathrm{d}F(m) = \tau,
$$ {#eq-ams-reservation}
and is decreasing in $\tau$: cheaper search makes consumers pickier. The code
below solves @eq-ams-reservation for a uniform match distribution and confirms the
monotonicity.

```{r ams-reservation}
#| label: fig-ams-reservation
#| fig-cap: "The sequential-search reservation threshold m-bar as a function of the search cost tau, for a uniform match-value distribution on [0,1]. Lower search cost raises the acceptance threshold (pickier consumers), reproducing Result 2 of Jiang et al. (2020)."
#| message: false
#| warning: false
set.seed(39)
# Reservation rule for U(0,1) match: integral_{m̄}^{1} (m - m̄) dF = tau
# For F uniform on [0,1]: (1 - m̄)^2 / 2 = tau  =>  m̄ = 1 - sqrt(2 tau).
reservation_threshold <- function(tau) pmax(0, 1 - sqrt(2 * tau))

tau_grid <- seq(0.001, 0.5, length.out = 200)
m_bar <- reservation_threshold(tau_grid)

plot(tau_grid, m_bar, type = "l", lwd = 2, col = "steelblue",
     xlab = expression(paste("Search cost ", tau)),
     ylab = expression(paste("Acceptance threshold ", bar(m), "(", tau, ")")),
     main = "Sequential-search reservation threshold")
abline(h = 0, col = "grey70", lty = 3)
```

The platform-design results then follow. When the referral fee is exogenous,
lower search cost intensifies seller competition and lowers prices, but expands
the market; the platform's profit can move either way (Lemma 2). When the fee is
chosen optimally, a decrease in search cost raises the platform's profit
(Proposition 1), and the platform should *raise* its referral fee if search does
not increase demand elasticity and *lower* it if search increases elasticity
(Proposition 2). Filtering is the more robustly beneficial tool: it raises the
consumer's acceptance threshold for aggregate match value by less than $\mu_K$
(Lemma 3), increases the probability of buying after searching, and reduces the
expected number of searches (Proposition 3). When filtering reveals only a small
share of match-value variation and the fee is exogenous, it increases platform
demand, seller profit, platform profit, and consumer surplus simultaneously
(Proposition 5)—a rare Pareto improvement. The contrast with a search-cost
reduction is the chapter's sharpest dichotomy: lower search cost makes consumers
search *more* and intensifies competition, whereas filtering makes consumers
search *fewer, better-targeted* products and softens competition, raising prices.

### Blind Buying {#sec-ams-blind}

@chen2021 add a realistic online wrinkle: the consumer can *blind-buy*—purchase
without inspecting the match value and without paying the search cost. Two firms
sell horizontally differentiated products in price competition; the random
utility of consumer $l$ from firm $i$ is
$$
\tilde u_{il} = \tilde\eta_{il} + \tilde\epsilon_{il} - p_i,
$$ {#eq-ams-blind-util}
where $\tilde\eta_{il}$ is the observable match component and
$\tilde\epsilon_{il}$ the unobservable one. The blind-buy option gives the
consumer, at any stage, the expected utility
$E(\tilde u_{il}) = \eta_{il} + E(\tilde\epsilon_{il}) - p_i$ without incurring
$\tau$. The game runs: firms set prices; consumers observe prices and prior
values and may blind-buy immediately (ending the game); otherwise they inspect
and choose among buying the inspected product, blind-buying the other, or
inspecting further. The novel comparative static is that an increase in the
*first-sample* search cost—which has no effect on prices in standard
price-directed search—now *lowers* equilibrium prices, because the blind-buy
outside option disciplines the firms. With a symmetric match distribution,
consumers and firms are indifferent to search order conditional on searching.

### Superior Knowledge and Customer Inspection {#sec-ams-inspection}

@li2022 close the loop by giving the *firm* superior knowledge of the consumer's
preferences while the consumer must inspect to learn her own. A monopolist serves
a unit continuum of customers with heterogeneous match values, high with prior
probability $\alpha$. Customers know only their expected match value ex ante; the
firm knows both its product characteristics and each customer's preferences. The
timing is: each customer's match is realized; the firm offers a *personalized*
price drawn from a chosen distribution; observing the price, the customer decides
whether to pay an inspection cost to learn her true preference; and, if she
inspects, she decides whether to buy. The model's interest is the interaction
between the firm's price *signal* and the consumer's inspection *decision*: the
personalized price both extracts surplus and conveys information, so the firm
randomizes its price to manage what the customer infers and whether she finds it
worthwhile to inspect. The result connects price discrimination to information
design—the firm's pricing policy is simultaneously a screening device and a
signal.

## Reading and Refereeing Analytical Work {#sec-ams-refereeing}

A seminar on analytical modeling is also training in evaluation, and the
evaluative standard for theory differs from that for empirical work. To *reject*
a theory paper, a referee must find a flaw that is *major*—an assumption that
drives the headline result and is indefensible, an equilibrium that does not
exist or is not the one claimed, or a comparative static that does not survive a
natural perturbation. Proposing an alternative specification or a robustness
check obliges the referee to give the rationale for why it is *necessary*, not
merely possible; an analytical model is deliberately stylized, and "the model
omits $X$" is not a defect unless $X$ overturns the mechanism.

The same standard guides authors handling editorial outcomes. A useful taxonomy
runs from *reject* (do not resubmit) through *reject-and-resubmit* (treated as a
new paper if specified conditions are met), *major revision* (substantive
problems with an uncertain outcome—roughly half are rejected in the second
round), *minor revision* (easily solvable problems with low residual
uncertainty), *conditional acceptance* (wording and small fixes, essentially no
rejection risk), and *accept as is*. The practical discipline—let a decision sit
before responding, separate the substantive concerns from the affective sting,
and address the load-bearing objections first—is craft rather than science, but
it is the craft that turns a correct model into a published one.

## Key Takeaways {#sec-ams-takeaways}

- The seminar is built **tools-first**: weeks 1–2 establish Nash, subgame
  perfection, PBE, and signaling refinements before any application, because every
  later block reuses them (@sec-analytical-semester-arc).
- A **hybrid platform** is simultaneously marketplace, retailer, and information
  intermediary (@fig-ams-platform-roles); its conflicts of interest—entry against
  its own sellers, ranking that favors first-party listings—are the engine of the
  modern analytical literature (@sec-ams-platforms).
- The **threat of platform entry** is a signaling problem: a high-demand seller
  may suppress sales to mask its type, and sellers can prefer the threat of entry
  to a no-entry commitment because it justifies a lower per-unit fee
  (@sec-ams-midtail).
- **Search neutrality** can *raise* prices and lower consumer surplus by removing
  the price-for-prominence lever that personalized ranking creates—policy intuition
  from net neutrality does not transfer (@sec-ams-search-neutrality).
- A **subscription platform** that bundles substitutes can coordinate the channel
  only if it holds a competition-softening instrument—a per-customer fee or a
  quantity cap; quality and a fixed fee alone yield no agreement
  (@sec-ams-subscription).
- **Intertemporal pricing** pits the seller against forward-looking strategic
  consumers; the optimal bundling sequence depends on marginal cost and the
  strategic share, and "save the best for last" governs event ordering
  (@sec-ams-intertemporal).
- A **fairness constraint** on an algorithm is an argument in the firm's
  optimization, not a free overlay: equilibrium adjustment of learning effort can
  reverse the constraint's intended distributional effect (@sec-ams-fairness).
- **Agency** contracts remove one of the two channel markups that the
  **wholesale** contract creates; the efficiency comparison turns on demand
  curvature and competitive intensity (@sec-ams-agency).
- **Sequential search** with an optimal reservation rule underlies the platform
  models; *lowering search cost* intensifies competition while *filtering* softens
  it, and a *blind-buy* option disciplines prices through an outside option
  (@sec-ams-search).
