# Scientific Writing {#sec-scientific-writing}

A research paper is the instrument through which a discovery becomes knowledge.
A result that is not written down does not exist scientifically; a result written
badly exists only for the handful of readers willing to excavate it. This chapter
treats scientific writing not as cosmetic polish applied after the "real" work but
as an integral part of the research itself—a design problem with its own
constraints, its own failure modes, and its own measurable quality. The audience
is the marketing scholar who has a finding and now faces the harder task of
convincing a skeptical editor, two or three adversarial reviewers, and an
indifferent field that the finding is correct, novel, and important.

The economics of the situation are unforgiving. The top journals in marketing—the
*Journal of Marketing* (JM), the *Journal of Marketing Research* (JMR),
*Marketing Science*, and the *Journal of Consumer Research* (JCR), collectively
the "top-4"—desk-reject a large share of submissions and ultimately accept only a
small fraction of what is sent to them. <!-- CITE-NEEDED: top-4 marketing journal acceptance/desk-reject rates -->
A reviewer reads dozens of papers a year as unpaid labor and forms a judgment
within the first few pages. The paper that survives is not merely the one with the
best result; it is the one whose argument the reviewer can reconstruct without
effort and whose contribution the reviewer can state in a sentence. Writing is
therefore a competitive technology, and this chapter develops it as one.

The chapter proceeds from architecture to sentence. We begin with the **structure**
of a paper—the canonical section sequence and what each section is for—and formalize
the underlying logical scaffold. We then treat the hardest and most consequential
task, **framing the contribution**, and then the two genres the hypo-deductive
template serves badly—the **descriptive paper**, which contributes a pattern, and
the **abductive paper**, which reasons from a pattern to the best available
explanation. Next comes the part of the paper where framing
lives or dies: the **front end** (title, abstract, introduction). We turn next to
the specific demands of writing for the **top-4**, where the contribution bar is
highest. We close with **clarity and revision**: the sentence-level craft that
makes an argument legible, the measurement of readability, and the disciplined
process of responding to reviewers. Throughout, the stance is the house stance—
intuition first, then the formal structure that makes the intuition operational.

## The Structure of a Paper {#sec-sw-structure}

### The canonical sections and their tenses

Empirical marketing papers share a near-universal skeleton, sometimes abbreviated
**IMRaD** (Introduction, Methods, Results, and Discussion) with literature review
and conclusions bracketing the core. Each section answers one question and only
one, and—this is a point novices routinely miss—each section has a *default verb
tense* that signals the epistemic status of its claims. Completed actions take the
**simple past** ("we estimated", "respondents rated"); established knowledge and
the paper's own ongoing argument take the **present** ("the effect is", "Table 2
reports"); prior literature as a stream that continues into the present takes the
**present perfect** ("researchers have shown"). @tbl-sw-sections collects the
sections, their purposes, approximate lengths, and tense conventions.

```{r}
#| label: tbl-sw-sections
#| tbl-cap: "Canonical sections of an empirical marketing paper, with purpose, indicative length, and default verb tense. Lengths are rules of thumb for a full-length top-4 submission, not hard limits."
#| echo: false
#| message: false
#| warning: false
library(knitr)
sw_sections <- data.frame(
  Section = c("Title", "Abstract", "Introduction", "Literature / Theory",
              "Methods", "Results", "Discussion", "Conclusion"),
  `Words` = c("8–15", "200–250", "800–1200", "1500–3000",
              "1000–2500", "1000–2500", "1500–2500", "400–800"),
  Purpose = c("Name the contribution",
              "Self-contained summary: objective, method, finding, implication",
              "Why the paper exists; the gap and the contribution",
              "Position vs. prior work; derive hypotheses or model",
              "What was done, reproducibly",
              "What was found, with the data doing the talking",
              "What it means, limits, boundary conditions",
              "Restate contribution; implications and future work"),
  `Default tense` = c("—", "Past + present", "Present + present perfect",
                      "Present + present perfect", "Past", "Past + present",
                      "Present + past", "Present"),
  check.names = FALSE
)
kable(sw_sections)
```

The discipline of one-question-per-section is what gives a paper its navigability.
A reader who wants to know *what you did* turns to Methods and finds only that; a
reader who wants to know *why it matters* turns to the Discussion and is not
detained by procedure. Violations—results smuggled into the methods, mechanism
speculation buried in the results—are the most common structural defect flagged in
review, because they force the reader to hold the paper's logic in their head
rather than read it off the page.

> A scientific paper is not a chronological record of what the researcher did. It
> is a reconstructed argument, organized so that the reader can verify the
> conclusion with minimum effort. The order of discovery and the order of
> exposition are different things, and conflating them is the surest route to an
> unreadable manuscript.

### The argument as a logical object

Underneath the section skeleton sits a logical scaffold that the prose merely
dresses. Make it explicit. A confirmatory empirical paper advances a chain

$$
\underbrace{P}_{\text{problem}} \;\Rightarrow\;
\underbrace{G}_{\text{gap}} \;\Rightarrow\;
\underbrace{H}_{\text{hypotheses}} \;\Rightarrow\;
\underbrace{D}_{\text{design}} \;\Rightarrow\;
\underbrace{R}_{\text{results}} \;\Rightarrow\;
\underbrace{C}_{\text{contribution}}
$$ {#eq-sw-chain}

in which every arrow is a claim the reader can refuse to grant. The problem $P$
must be real (someone cares about the answer); the gap $G$ must be genuine (the
answer is not already known); the hypotheses $H$ must follow from a theory and be
falsifiable; the design $D$ must be capable of discriminating $H$ from its
negation; the results $R$ must actually bear on $H$; and the contribution $C$ must
be what $R$ licenses, not more. A paper fails at whichever arrow is weakest, and
the writing task is to make each arrow as short and as forced as possible. Many
rejections that are phrased as "not enough contribution" are in fact broken
arrows: a gap that turns out to be already filled ($G$), or a design that cannot
separate the focal hypothesis from a confound ($D \not\Rightarrow R$).

The conceptual or theory paper substitutes a different chain—a construct is
defined, related to existing constructs, and shown to do explanatory work—but the
demand for unbroken arrows is identical. @fig-sw-paper-flow renders the canonical
flow and shows where each link in @eq-sw-chain is forged.

```{mermaid}
%%| label: fig-sw-paper-flow
%%| fig-cap: "The logical flow of an empirical paper. The front end (title, abstract, introduction) carries the problem, gap, and promised contribution; the body discharges the promise. A reviewer reads the front end to decide whether the rest is worth reading."
flowchart TD
    A["Title<br/><i>names the contribution</i>"] --> B["Abstract<br/><i>P → R → C in 250 words</i>"]
    B --> C["Introduction<br/><i>problem P, gap G,<br/>promised contribution C</i>"]
    C --> D["Theory / Literature<br/><i>position; derive H</i>"]
    D --> E["Methods<br/><i>design D</i>"]
    E --> F["Results<br/><i>evidence R</i>"]
    F --> G["Discussion<br/><i>does R license C?</i>"]
    G --> H["Conclusion<br/><i>restate C; boundaries</i>"]
    C -.->|"promise"| G
    H -.->|"payoff"| C
    style C fill:#e8f0fe,stroke:#1a73e8
    style G fill:#e8f0fe,stroke:#1a73e8
```

The dashed promise–payoff loop in @fig-sw-paper-flow is the structural heart of a
paper. The introduction makes a promise—"we will show that $X$ causes $Y$ through
mechanism $M$"—and the discussion must redeem exactly that promise, no more and no
less. A paper that promises a causal claim and delivers a correlation, or that
delivers a richer finding than it promised and never updates the framing, reads as
incoherent even when every individual sentence is sound.

### Paper types

The skeleton flexes with the paper's type, and naming the type early orients both
writer and reviewer. Four types recur. A **research paper** tests hypotheses and
reports findings; its contribution lives in $H \Rightarrow R$. A **methods paper**
proposes a new estimator, design, or measurement instrument; its contribution is
that the method recovers something prior methods could not, so the burden is a
demonstration—often on synthetic data with known ground truth—that it works and a
comparison against incumbents. A **review paper** consolidates a domain, and its
contribution is organization: a taxonomy, an integrative framework, or an agenda
that practitioners and scholars did not previously have. A **conceptual or
discussion paper** advances or critiques theory, trading empirical evidence for
argumentative rigor. The reviewer evaluates each type against a different
standard, and a paper that does not signal its type invites evaluation against the
wrong one.

## Framing the Contribution {#sec-sw-contribution}

### What a contribution is

The single most important sentence in a paper is the one that states its
contribution, and the most common reason good work is rejected is that this
sentence is missing, vague, or overclaimed. Define the contribution formally.
Let $\mathcal{K}$ denote the field's existing knowledge—the set of claims a
competent reader already accepts before reading the paper. A paper's contribution
is the increment

$$
\Delta\mathcal{K} \;=\; \mathcal{K}_{\text{after}} \setminus \mathcal{K}_{\text{before}},
$$ {#eq-sw-contribution}

the set of claims that a reader is licensed to accept *after* reading the paper and
was not licensed to accept before. Three properties make $\Delta\mathcal{K}$
publishable, and reviewers probe each one explicitly:

- **Novelty** ($\Delta\mathcal{K} \neq \varnothing$): the increment is non-empty.
  If every claim in the paper already belonged to $\mathcal{K}_{\text{before}}$,
  there is no contribution, however well executed the study.
- **Validity** ($\Delta\mathcal{K}$ is warranted): the increment is actually
  licensed by the evidence and argument. Overclaiming—asserting elements of
  $\Delta\mathcal{K}$ that the design $D$ cannot support—is the fastest route to a
  reviewer's "the conclusions outrun the data".
- **Importance** ($|\Delta\mathcal{K}|$ matters): the increment is one the field
  cares about. A true, novel, but trivial claim clears the first two bars and fails
  the third, and "importance" is precisely the bar that rises from a field journal
  to the top-4.

Two broad species of contribution recur in marketing, and a paper should know which
it is making. A **substantive** contribution changes what we believe about a
marketing phenomenon—that brand prominence signals status through quiet versus loud
cues, say, or that responding to online reviews trades complaint quantity for depth
(see @sec-branding). A **methodological** contribution changes how we can learn
about phenomena—a new identification strategy, a scalable text measure, a more
efficient estimator. The strongest papers often pair a methodological advance with
the substantive finding it unlocks, but a paper that is fuzzy about which kind of
increment it offers will be read as offering neither.

### The contribution statement and the framing trap

A contribution should be expressible as a single declarative sentence of the form
"We show that [claim], which prior work did not establish because [reason], using
[design]." The clause "which prior work did not establish" is the gap $G$ of
@eq-sw-chain made explicit, and it must be *true*: nothing destroys credibility
faster than a literature-savvy reviewer who knows the claimed gap was filled years
ago. Conversely, a contribution can be undersold. Authors who have lived inside a
problem for years often state their finding in the narrow terms in which they
discovered it, missing the more general claim the evidence actually supports. The
discipline is to state the *most general* claim that $R$ licenses and not one inch
more.

Framing is the act of choosing which $\mathcal{K}_{\text{before}}$ to write against—
which conversation the paper joins—and it is a genuine strategic choice, not a
neutral description. The same result can be framed as a contribution to several
literatures, and the framing determines the reference set, the reviewers, and the
perceived importance. A study of how customers respond to a salesperson's language,
for instance, could be framed against the persuasion literature, the
frontline-service literature, or the computational-text-analysis literature, and
the right choice depends on where the increment $\Delta\mathcal{K}$ is largest and
the audience most receptive. The **framing trap** is to choose the literature in
which the result is *most surprising* rather than the one in which it is *most
defensible*; surprise attracts an editor's eye but invites reviewers who hold the
strongest priors against the claim.

## The Descriptive Paper {#sec-sw-descriptive}

The template assumed so far—theory, hypotheses, identification, test—is not the
only publishable shape, and treating it as such quietly rules out a class of
papers the field needs. A **descriptive paper** establishes what a phenomenon
*looks like* without claiming to have identified a causal effect within it. Its
increment $\Delta\mathcal{K}$ consists of pattern claims: this outcome
co-occurs with that antecedent, this population sorts into these categories,
this quantity has moved this way over this period.

The genre has explicit methodological warrant. @helfat2007stylized argues that
stylized facts are a legitimate and necessary input to theory development,
because a theory can only be built against a phenomenon whose shape is known.
@hambrick2007devotion goes further, arguing that the field's insistence on a
theoretical contribution in every paper suppresses precisely the descriptive
work theory needs to feed on. Neither position licenses atheoretical
data-dumping: @suttonstaw1995 is equally clear that data, variables, and
diagrams are not themselves theory. The reconciliation is that a descriptive
paper contributes a *pattern*, states it as a pattern, and lets the theoretical
payoff be the discipline the pattern imposes on subsequent explanation.

@choudhury2026abduction supply the positive account. Abduction—the mode of
inference that formulates candidate explanations for an unexplained
phenomenon—is a distinct research design rather than a deduction that failed to
find support. Their treatment distinguishes the *abductive trigger* (the
anomalous observation that starts the inquiry) from *creative* abduction
(generating candidate explanations) and *selective* abduction (winnowing them),
and is explicit that abduction admits no fixed template. That absence of a
template is exactly why descriptive papers need to make their design visible on
the page: a reader cannot infer it from convention.

### What makes a descriptive contribution defensible

Five design choices separate a descriptive paper that publishes from one that
reads as a set of cross-tabulations. @tbl-sw-descriptive states them alongside a
worked example: @moeen2026valuecapture, which maps founder knowledge origin onto
mode of economic value capture in U.S. medical device start-ups and is developed
substantively in @sec-entrepreneurship.

| Design choice | Why it matters | In the worked example |
|:--|:--|:--|
| The outcome is a **category**, not a magnitude | A mode question ("which market does the venture monetize in?") has no natural causal counterpart, so description is the right design rather than a consolation prize | Value captured in the market for technology, the market for product, or not at all |
| The antecedent variation is **institutionally sharp** | A boundary drawn by an institution rather than by the researcher forecloses the "you defined the groups to get the result" objection | Founders who were academic scientists vs. former employees of within-industry firms |
| The **typology is declared before the data** | A pattern claim can only be wrong—and therefore only be informative—if the correspondence it predicts was specified in advance | Scientific vs. commercial knowledge dimensions, crossed with founder origin |
| The outcome categories are **externally classified** | Descriptive work dies when the dependent variable rests on the author's judgment calls | IP transactions and FDA clearance leave a regulatory paper trail |
| The **off-diagonal case is reported** | The cell that violates the clean mapping is what distinguishes a mechanism from a stereotype; in a design with no treatment, it does the work an interaction term would do elsewhere | Founders bridging both dimensions shift to pioneering products rather than following their institution |

: Five design choices that make a descriptive contribution defensible, with a worked example. {#tbl-sw-descriptive}

The fifth row deserves emphasis because it is the one most often omitted. A
descriptive paper that reports only the clean correspondence has documented a
stereotype: academics do this, employees do that. The reader has no way to tell
whether the operative variable is the credential or the knowledge. Reporting the
cases that break the correspondence, and showing that they break it *in the
direction the proposed mechanism predicts*, converts the pattern into an
explanation candidate. This is selective abduction performed in public.

### How the contribution sentence changes

The contribution sentence of @sec-sw-contribution still applies, with the causal
verb removed and the boundary stated. Rather than "we show that $X$ causes $Y$,"
a descriptive paper claims "we establish that $X$ and $Y$ correspond in
population $P$, which prior work had not documented because it lacked
[data/classification/setting], and we show the correspondence tracks $M$ rather
than $X$ itself." The final clause is what makes the paper theory-relevant
without overclaiming, and it should be paired with an explicit statement of what
the design cannot support. Reviewers punish descriptive papers that reach; they
reward the ones that name their own ceiling. Stating "we do not claim founder
origin is exogenous" in the paper is strictly better than having a reviewer
state it for you.

### Norms differ by field, and the difference is not subtle

This genre travels much better in *Organization Science*, *Strategic Management
Journal*, and *Strategy Science* than in *JM*, *JMR*, or *Marketing Science*,
where the descriptive lane is narrower and more codified. In marketing it runs
mainly through three channels: empirical generalizations, in which a regularity
is established across many datasets and contexts; new-data-source papers, in
which the contribution is a measurement capability and the first look it
affords; and stylized-facts sections embedded inside otherwise structural or
causal papers, where description motivates the model rather than standing alone.
@bronnenberg2009 is the canonical marketing-side exemplar of the standalone
form—a persistent geographic pattern in brand shares, a scaffold for
interpreting it, and no causal overclaim.

The practical implication is a targeting decision made *before* writing. A
descriptive paper aimed at a marketing journal usually needs one of the three
channels above as its declared home; the same paper aimed at a management
journal can stand on the stylized-facts warrant directly. Authors who write the
management version and submit it to the marketing journal receive a predictable
review: interesting, but what is the identification?

A descriptive paper that stops at the pattern is one genre; a paper that goes on
to argue for a particular explanation of that pattern is another, and it needs a
reporting structure of its own. That is the subject of @sec-sw-peebi.

## The Abductive Paper and the PEEBI Structure {#sec-sw-peebi}

A descriptive paper stops at the pattern. A large class of empirical papers does
not: it commits to an *explanation* of the pattern that was constructed after the
pattern was seen, and that was never tested deductively because it did not exist
before the data did. This is abduction, or **inference to the best explanation**
(IBE), and it produces a great deal of what marketing knows. The difficulty is
not the inference—it is the *testimony*. @pillai2026peebi observe that although
scholars employ abduction routinely, there is no agreed-upon structure for
reporting it, and the hypo-deductive reporting template does not align with the
epistemology of abduction. They propose one, **PEEBI**, a five-section structure
in which the authors take prior knowledge and theories, establish the context and
observations worthy of interest, identify and evaluate candidate explanations for
the observed patterns, determine the best explanation and their reasoning for
accepting it, and abstract that explanation to a more generalizable theoretical
contribution. The structure "advances knowledge in a modest, stepwise fashion"
and is designed to foreground the author's judgment rather than hide it.

### Why the deductive template misreports abductive work

The template of @sec-sw-structure asks the reader to grant a chain that begins
with a hypothesis and ends with a test. Abductive work runs the other way: the
observation comes first, the candidate explanations are generated by the author
rather than handed down by prior theory, and the evidence that discriminates
among them is assembled afterward. @tbl-sw-abduction contrasts what each mode
actually asks of a reader.

| | Hypo-deductive paper | Abductive paper |
|:--|:--|:--|
| Where the claim comes from | Derived from theory *before* the data | Constructed to explain an observation *after* it |
| What the evidence does | Discriminates $H$ from $\neg H$ | Discriminates $E_k$ from the other candidates in a set the author assembled |
| What the reader grants | That the design could have falsified $H$ | That the candidate set is adequate and was evaluated evenhandedly |
| What failure looks like | The test does not reject | A rival explanation survives that the author did not consider |
| What must be disclosed | Design, measures, robustness | All of that, **plus** the order of discovery and the rejected candidates |
| Who does the judging | Nominally the data | Explicitly the author—and, if the paper is written well, the reader |

: What the two reporting modes ask of a reader. The last two rows are where the deductive template fails abductive work. {#tbl-sw-abduction}

Forcing abductive work into the deductive template requires a small fiction: the
explanation is presented as though it preceded the data. The field already has a
name for the vicious version of that fiction—HARKing—and the reason abduction is
*not* HARKing is disclosure: the order of discovery is stated, the rival
explanations are named, and the residual uncertainty is quantified or at least
bounded. The deductive template deletes precisely those three things. It does not
merely misfit the work; it removes the information that licenses the conclusion.

### The five acts

@tbl-sw-peebi states the five sections, what each one owes the reader, and where
it goes in a manuscript aimed at a marketing journal, since few marketing
editors will accept section headings named after an epistemology.

| Act | What the section does | What the reader is handed | Marketing manuscript home |
|:--|:--|:--|:--|
| **P**riors | States the prior knowledge and theories the inquiry begins from | The baseline against which the observation is surprising | Introduction and a short theory-background section |
| **E**stablish | Establishes the context and the observations worthy of scholarly interest | The phenomenon itself: measurement, setting, and why the pattern is not an artifact | "Empirical setting" plus a stylized-facts or model-free-evidence section |
| **E**valuate | Identifies *and* evaluates candidate explanations for the observed pattern | The full candidate set and the evidence bearing on each | "Alternative explanations", promoted from appendix to spine |
| **B**est | Determines the best explanation and the reasoning for accepting it | The author's judgment, stated as judgment, with its grounds | "Discriminating evidence" or the analysis section that adjudicates |
| **I**mplication | Abstracts the best explanation into a more generalizable theoretical contribution | The claim that travels beyond this setting, and its boundary | General discussion and theoretical implications |

: The five PEEBI acts of @pillai2026peebi, with the marketing-manuscript section that hosts each. The letters index the five sections in the order the published account states them; the short act names here are our labels for the functions that account describes. {#tbl-sw-peebi}

The move that does the work is the third act. In a conventional marketing paper,
alternative explanations are a defensive appendix written to preempt reviewer
objections; in an abductive paper they are the argument. A reader who is shown
four candidates, the evidence for and against each, and the author's reasoning
for preferring one has been given the material to disagree—which is the point.
@pillai2026peebi frame this as elevating the reader's role: the paper's job is to
supply the information a reader needs to reach a judgment, not to foreclose it.

### The abductive chain

The counterpart to @eq-sw-chain runs

$$
\underbrace{O}_{\text{observation}} \;\Rightarrow\;
\underbrace{\mathcal{E} = \{E_1, \dots, E_K\}}_{\text{candidate set}} \;\Rightarrow\;
\underbrace{V}_{\text{evaluation}} \;\Rightarrow\;
\underbrace{E^{*}}_{\text{best explanation}} \;\Rightarrow\;
\underbrace{T}_{\text{abstraction}}
$$ {#eq-sw-abduction}

and, as before, every arrow is a claim the reader may refuse to grant. The
observation $O$ must be real and genuinely surprising against the priors—a
pattern that everyone expected is not an abductive trigger. The candidate set
$\mathcal{E}$ must be *adequate*: it must contain the explanation a hostile
expert would propose, which is the arrow that breaks most often. The evaluation
$V$ must apply the same standard to every candidate; searching hard for evidence
that supports $E_1$ and casually for evidence against it is the abductive
analogue of a rigged specification search. The selection $E^{*}$ must dominate
the survivors on stated grounds rather than on the author's enthusiasm. And the
abstraction $T$ must not outrun the case: one setting supports a mechanism claim,
not a law.

Failure at each arrow has a diagnostic signature in reviews. A weak $O$ draws
"this is well known." A thin $\mathcal{E}$ draws "the authors have not ruled
out..." An asymmetric $V$ draws "the evidence for the preferred story is treated
differently from the evidence against it." An overreaching $T$ draws "the
contribution is not supported by a single context."

### What makes an explanation "best"

The word *best* carries the whole burden, and @pillai2024lovely unpack it along
two dimensions that do not coincide. An explanation is **lovely** when it is
useful, general, and provides meaning; it is **likely** when it is close to the
truth. Loveliness is what makes a mechanism worth having; likeliness is what
makes it true. A managerially elegant story about why a promotion worked can be
extremely lovely and quite unlikely, and a precisely estimated coefficient can be
likely and entirely unlovely. Their argument is that interpreting observational
results requires an understanding of context that statistical analysis alone
cannot supply, and that historical methods—hermeneutics, contextualization, and
source criticism—help scholars both *generate* new candidate explanations and
*judge* and balance the virtues that make one explanation better than another.

In marketing the contextual sources are close at hand and systematically
underused: category and brand histories, trade press, retailer and manufacturer
archives, advertising archives, regulatory filings, and interviews with the
people who made the decisions in the data. A researcher who knows that a category
was restructured by a distribution change in 1998 has a candidate explanation
that no amount of panel-data variance decomposition will hand them.

The practical artifact this yields is a **candidate-explanation table**, written
early and revised as evidence accumulates. @tbl-sw-candidates sketches its form
for a stylized marketing pattern—a brand's share is persistently higher in
markets it entered first—and it is the single most useful document a student can
produce before writing an abductive paper, because it makes the completeness of
$\mathcal{E}$ visible to the author.

| Candidate explanation | Lovely? (useful, general, meaningful) | Likely? (evidence for) | Evidence against | Discriminating test available? |
|:--|:--|:--|:--|:--|
| Consumer-side brand capital formed early and persisting | High—travels to any repeat-purchase category | Share gap survives entry of national rivals | Cannot separate from supply-side with market-level data alone | Yes: observe consumers who move between markets |
| Supply-side cost or distribution advantage in early markets | Moderate—context-specific | Early entrant has denser distribution | Gap persists after distribution converges | Yes: control distribution and re-measure |
| Retailer assortment and shelf-space lock-in | Moderate | Assortment correlates with share | Pattern holds in categories with weak retailer power | Partly: cross-retailer variation |
| Advertising stock accumulated over decades | High | Cumulative spend correlates with share | Gap holds where historical spend was symmetric | Yes: historical ad-spend series |

: A candidate-explanation table for a stylized persistence pattern. The final column is the one that turns an essay into a paper. {#tbl-sw-candidates}

### A worked example, reconstructed

Marketing has abductive exemplars; what it lacks is the habit of labeling them.
Read as a two-paper arc, @bronnenberg2009 and @bronnenberg2012migration
instantiate the chain in @eq-sw-abduction almost exactly—though neither paper
presents itself in these terms, and the reconstruction is ours.

The first paper *establishes*: brand shares in consumer packaged goods differ
persistently across U.S. geographic markets, and the differences track the
brand's early entry more than a century back. Against the priors of a field that
models share as a function of current marketing mix, that is a genuine surprise,
and the paper states it as a pattern rather than as an effect. The candidate set
is the obvious one—something on the demand side (consumers in these markets
simply prefer this brand, for reasons formed long ago) versus something on the
supply side (the early entrant enjoys durable cost, distribution, or retail
advantages there).

Market-level data cannot separate those candidates, and this is where an
abductive paper either finds a discriminating observation or admits it cannot.
The second paper finds one: consumers who *move* between markets carry their
preferences with them, converging only slowly toward their new market's shares.
Movers separate the two candidates cleanly, because a consumer-side stock travels
with the consumer while a supply-side advantage does not. That is act four
performed in public—the best explanation determined, with the reasoning for
accepting it visible—and the abstraction in act five is the one that made the work
matter beyond CPG: brand preference behaves as an accumulated consumer-side
capital stock with a very long half-life.

Two features of the arc are worth copying. The discriminating evidence was
*sought after* the candidates were named, not before—the honest order, and the
one PEEBI asks you to report. And the abstraction stops where the evidence
stops: a claim about how preferences persist and transfer, not a claim that
early entry causes share everywhere.

### Getting an abductive paper through a marketing review

The genre travels less easily in *JM*, *JMR*, and *Marketing Science* than in
*SMJ* or *Organization Science*, for the reason given in @sec-sw-descriptive: the
default reviewer question in marketing is "what is the identification?" Three
adjustments carry most of the load.

Declare the design in the introduction. A sentence—"we do not test a
pre-specified hypothesis; we document a pattern, enumerate the explanations
consistent with it, and present evidence that discriminates among them"—costs
nothing and prevents the paper from being evaluated against a template it is not
using. Reviewers punish papers that appear to have failed at deduction far more
than papers that never attempted it.

Keep the conventional headings and put PEEBI underneath them, as in the last
column of @tbl-sw-peebi. The structure is a reporting discipline, not a demand
for new section names, and a manuscript that renames its sections after an
epistemology spends its first reviewer paragraph defending a formatting choice.

State the ceiling before a reviewer does. The abductive paper's ceiling is that
$\mathcal{E}$ may be incomplete, and the correct response is to say so and to
name what evidence would change the ranking. Papers that specify the observation
that would overturn their preferred explanation read as confident; papers that
imply the candidate set is exhaustive read as naive. This is the same principle
that governs the R&R response in @sec-review-process—make the reader's job of
disagreeing with you *easy*, and they will usually decline the invitation.

@fig-sw-peebi summarizes the flow, including the loop that the deductive template
has no way to represent: evaluation frequently sends the author back for more
candidates or more context, and the paper reports the end state of that loop
rather than pretending it did not run.

```{mermaid}
%%| label: fig-sw-peebi
%%| fig-cap: "The PEEBI reporting structure for abductive work, after Pillai et al. (2026). The dashed arrows are the iteration the deductive template cannot represent: evaluation sends the author back for further candidates or further context, and the paper testifies to where that loop ended."
flowchart TD
    P["<b>P</b>riors<br/><i>prior knowledge and theories</i>"] --> E1["<b>E</b>stablish<br/><i>context and observations<br/>worthy of interest</i>"]
    E1 --> E2["<b>E</b>valuate<br/><i>identify and evaluate<br/>candidate explanations</i>"]
    E2 --> B["<b>B</b>est explanation<br/><i>and the reasoning<br/>for accepting it</i>"]
    B --> I["<b>I</b>mplication<br/><i>abstract to a generalizable<br/>theoretical contribution</i>"]
    E2 -.->|"candidate set inadequate"| E1
    B -.->|"no candidate dominates"| E2
    I -.->|"new priors for the next study"| P
    style E2 fill:#e8f0fe,stroke:#1a73e8
    style B fill:#e8f0fe,stroke:#1a73e8
```

### A checklist for the abductive manuscript

A reader finishing each act should be able to answer one question. If they
cannot, the act is not written yet.

- **After P**: what did we believe before, and why is the observation surprising
  against that belief?
- **After E (establish)**: is the pattern real—measured how, in what population,
  and robust to what obvious artifacts?
- **After E (evaluate)**: what else could produce this pattern, and what does
  the evidence say about each of those?
- **After B**: why this explanation rather than the runner-up, and what would
  change my mind?
- **After I**: what claim travels beyond this setting, and where does it stop?

The parallel exercise for a course is worth assigning directly: take a published
hypo-deductive paper in the student's area, reconstruct the abductive history
that almost certainly produced it, and rewrite the introduction in PEEBI form.
Students discover quickly that the reconstructed version is both more honest and
easier to read, which is the argument of @pillai2026peebi in miniature. The
seminar treatment of abduction as a mode of inference, alongside empirical
generalizations and stylized facts, is in @sec-theo-week05.

## The Decomposition Paper {#sec-sw-decomposition}

A third genre sits beside description and abduction, and it is the one most likely to
rescue a paper whose headline effect is small, imprecise, or already known. Its claim is
not "$X$ affects $Y$" but "$X$ affects $Y$ through these specific channels, which push in
opposite directions, in these proportions, under these conditions." The increment
$\Delta\mathcal{K}$ is the *anatomy* of an effect rather than its sign.

The genre is worth naming because the alternative is so common and so weak. A great many
papers—in marketing, economics, and management alike—end at a signed average: the
promotion raised sales, the technology cut employment, the entry helped the incumbent.
Signed averages age badly. They do not transfer to settings where the mixture of channels
differs, they cannot explain why two credible studies disagree, and they give a manager or
regulator nothing to act on beyond "more" or "less."

**The exemplar.** @clemens2026lowskill study a question with a long history of
irreconcilable estimates: does employing low-skill immigrants displace native workers? The
design is a natural firm-level randomization—the US H-2B visa quota is allocated in part
by federal lottery—paired with a pre-analysis plan and a purpose-built firm survey of 2021
lottery winners and losers. That alone would produce a signed average. What makes the paper
a template is that the signed average is *derived* rather than reported.

The authors write down a monopolistically competitive firm producing with immigrant labor
$I$, native low-skill labor $N$, capital $K$, and high-skill labor $H$ in a nested CES
technology [@ottaviano2012rethinking], facing demand elasticity $\eta$, with
immigrant–native substitution elasticity $\sigma$. Two mechanisms then run in opposite
directions. **Substitution**: immigrants and natives compete for the same tasks, so cheap
access to one displaces the other, and the force is governed by $\sigma$. **Scale**:
relaxing the hiring constraint lets the firm expand output, raising demand for every
input including native labor, and the force is governed by $\eta$. Stripping the model to
labor alone makes the arithmetic transparent: with $s_I$ and $s_N$ the immigrant and
native revenue shares, the effects of winning the lottery on revenue and on native
employment are
$$
\ln\frac{R_w}{R_\ell} \approx s_I \cdot
\frac{\eta - 1}{(\eta-1)(1-s_N) + (\sigma-1)s_N}\cdot \ln\frac{I_w}{I_\ell},
\qquad
\ln\frac{N_w}{N_\ell} \approx s_I \cdot
\frac{\eta - \sigma}{(\eta-1)(1-s_N) + (\sigma-1)s_N}\cdot \ln\frac{I_w}{I_\ell}.
$$ {#eq-sw-decomp}

The revenue effect is unambiguously non-negative. The employment effect has *indeterminate
sign*, positive when $\eta > \sigma$ and negative otherwise: whether immigrant hiring
crowds natives in or out is not a fact about immigrants but a race between two elasticities.
(With capital and high-skill labor restored, the crossing condition generalizes but keeps
the same structure.)

```{r sw-decomposition}
#| message: false
#| warning: false
# Effects of an exogenous increase in immigrant employment, from eq-sw-decomp.
s_I <- 0.20; s_N <- 0.65          # revenue shares of immigrant and native labor
grid <- expand.grid(eta = c(4, 8, 16), sigma = c(0.5, 1.5, 3, 6))

denom <- with(grid, (eta - 1) * (1 - s_N) + (sigma - 1) * s_N)
grid$revenue_effect  <- with(grid, s_I * (eta - 1)     / denom)
grid$native_effect   <- with(grid, s_I * (eta - sigma) / denom)
grid$verdict <- ifelse(grid$native_effect > 0, "crowds in", "crowds out")

round_df <- within(grid, {
  revenue_effect <- round(revenue_effect, 3)
  native_effect  <- round(native_effect, 3)
})
round_df[order(round_df$eta, round_df$sigma), ]
```

Every row has a positive revenue effect; the native-employment column flips sign inside the
same table. That single display carries more transferable knowledge than any average
treatment effect estimated from it, because a reader in a different industry can locate
their own $(\eta, \sigma)$ and read off the prediction. The paper's empirical estimates
then close the loop: losing the lottery cuts immigrant employment by 56% and contracts the
firm—revenue elasticity about $+0.16$, investment about $+1.03$ in the working-paper
version, with the published estimates in the range $0.20$--$0.22$ for production and
$1.5$--$2.1$ for investment—while the effect on US employment is zero or positive. Read
through @eq-sw-decomp, a null employment effect is not a non-result: it *identifies* the
substitution elasticity, which the authors back out at $0.8$--$2.2$, low relative to the
demand elasticity, which is exactly why scale wins.

**The four moves worth stealing.** The genre has a recognizable structure, and each move
is portable to a marketing paper.

1. **Write the model that makes the sign ambiguous.** The theory's job here is not to
   generate a directional hypothesis but to prove that direction *cannot* be signed a
   priori. That reframes a null or mixed result from a failure into an estimate of the
   parameter that decides the race.
2. **Turn the mechanism into a prespecified heterogeneity test.** The same model predicts
   *where* each channel should dominate—larger revenue effects for small firms facing more
   elastic demand, larger native-employment effects in rural areas where complementary
   workers have worse outside options—and both predictions were registered in advance and
   confirmed. Heterogeneity chosen after the fact is fishing; heterogeneity implied by the
   decomposition is a test.
3. **Recover the structural parameter from the reduced form.** The reduced-form
   coefficients are used to invert for $\sigma$, which is the quantity other studies argue
   about and the quantity that transfers.
4. **Say which quantity you estimated.** The authors distinguish the LATE recovered from
   the usual shift-share instruments—variation in immigrant *supply*, irrespective of
   current demand—from the policy-relevant treatment effect of a *policy-induced*
   restriction, and argue their lottery identifies the latter. Naming the estimand is
   itself a contribution when the literature has been silently averaging over several.

**Companion designs.** The decomposition move is not foreign to marketing; it is simply
underused as an explicit frame.

- Promotion analytics has run on decomposition for decades: a sales spike splits into
  brand switching, category expansion, purchase acceleration, and stockpiling, and the
  split—not the spike—determines incremental profit [@vanheerde2007; @guyt2014]
  (@sec-pricing).
- @biswas2026channel decompose the value of channel adoption by the *pathway* that
  produced it, isolating forward buying from habit persistence, and show the two mechanisms
  imply opposite corrections to incremental CLV (@sec-clv-pathways).
- @liu2026platform decompose the environmental effect of platform entry by market
  structure, finding that the entire average effect comes from dual-platform cities and
  none from single-platform ones (@sec-platform-entry-externality).
- @ellickson2023dml decompose a campaign's effect into the effects of its *components*
  with double machine learning, turning "the email worked" into which element of the email
  worked.
- In economics, the same instinct underwrites the task-based decomposition of automation
  into displacement and reinstatement effects, and the decomposition of trade shocks into
  substitution and scale channels; the shared feature is that the aggregate is written as
  a sum of forces with known signs before it is estimated.

**Where it fails.** Two failure modes recur. The first is decomposition by regression
mediation with no design behind it: adding a mediator to a regression and reporting an
"indirect effect" decomposes nothing if the mediator is endogenous, and this remains the
most common way marketing papers claim mechanism without earning it. The second is
decomposing into channels the data cannot separate; if two mechanisms imply identical
predictions in every observable dimension, naming them both is rhetoric, not evidence. The
test to apply before writing is simple: for each channel in the decomposition, state the
observation that would change its estimated share. If no such observation exists, the
channel is not identified and the decomposition should be presented as an interpretation
rather than a result.

## The Front End {#sec-sw-front-end}

The front end—title, abstract, and introduction—is the part of the paper most
people read and the only part most people read. It carries the entire promise–
payoff loop of @fig-sw-paper-flow in compressed form, and a reviewer's
accept-lean or reject-lean is often set before the methods section begins. The
front end deserves a disproportionate share of writing effort precisely because it
does a disproportionate share of the persuasive work.

### The title

A title's job is to name the contribution in eight to fifteen words such that the
right reader stops scrolling. It should contain the paper's key constructs (the
terms a searcher would type) and, where possible, signal the relationship the paper
establishes rather than merely the topic it addresses. "Brand Prominence and
Status Signaling" names a topic; "Signaling Status with Luxury Goods: The Role of
Brand Prominence" names a relationship and a mechanism. Question titles and
two-part titles (a hook before the colon, the substance after) are common in JCR
and JM and rarer in *Marketing Science*, reflecting the venues' different house
voices—a difference worth matching to the target journal.

### The abstract

The abstract is the most-read and least-revised 250 words in science, and the
imbalance is a mistake. It must be **self-contained**: a reader who sees only the
abstract should be able to state the paper's objective, method, principal finding,
and implication. The canonical structure mirrors @eq-sw-chain in miniature—one or
two sentences each for the problem, the approach, the result (with direction and,
where it fits, magnitude), and the "so what". Two failure modes dominate. The
**descriptive abstract** announces what the paper does ("we examine the effect of
X on Y") without ever stating what it *found*, leaving the reader no wiser; the
**overloaded abstract** crams in every robustness check and moderator, drowning the
headline. The fix for both is the same: lead with the finding.

### The introduction as a funnel

The introduction is where most rejections are decided, and it has a near-mechanical
structure: a **funnel** from the broad problem to the specific contribution. A
robust template runs problem $\rightarrow$ tension or puzzle $\rightarrow$ gap
$\rightarrow$ "in this paper we" $\rightarrow$ approach $\rightarrow$ findings
$\rightarrow$ contributions $\rightarrow$ roadmap. @fig-sw-funnel renders it.

```{mermaid}
%%| label: fig-sw-funnel
%%| fig-cap: "The introduction funnel. Each band narrows from the world's problem to this paper's specific increment to knowledge. The 'in this paper, we...' sentence is the pivot from motivation to contribution and should appear by the end of the second page."
flowchart TD
    P["The broad problem<br/><i>why anyone should care</i>"] --> T["The tension / puzzle<br/><i>what doesn't fit</i>"]
    T --> G["The gap<br/><i>what prior work left open</i>"]
    G --> Q["<b>In this paper, we…</b><br/><i>the pivot to contribution</i>"]
    Q --> A2["Approach<br/><i>data and method, one paragraph</i>"]
    A2 --> F["Findings<br/><i>headline results, with direction</i>"]
    F --> C2["Contributions<br/><i>ΔK, stated as claims</i>"]
    C2 --> R2["Roadmap<br/><i>optional, brief</i>"]
    style Q fill:#fce8e6,stroke:#d93025
    style C2 fill:#e6f4ea,stroke:#188038
```

Two discipline points govern the funnel. First, the pivot sentence—"In this paper,
we..."—should arrive early, by the end of the second page; an introduction that is
still motivating the problem on page four has lost the reviewer. Second, the
contributions should be stated as *claims the reader will be able to accept*, not
as activities the authors performed. "We contribute by examining the role of
arousal" is an activity; "We show that empathetic responses raise gratitude even
before a service failure is resolved, which the prior literature's
resolution-centric models cannot accommodate" is a claim and an explicit
$\Delta\mathcal{K}$.

### Positioning against the literature

The literature review is not a homage to everyone who has touched the topic; it is
the construction of $\mathcal{K}_{\text{before}}$, the precise baseline against
which the contribution is measured. Its house failure mode is the **annotated
bibliography**—a serial list of "Author (year) found X; Author (year) found Y"—which
demonstrates reading but not synthesis and leaves the reviewer to infer the gap.
The remedy is to organize the review around *tensions and open questions* rather
than around papers, so that the gap emerges as the natural next move in a
conversation the reader can now follow. The house style guide's instruction to
**weave, not list** citations is exactly this principle applied at the sentence
level: each citation should earn its place by advancing the argument, and the
strongest evidence per point should crowd out the duplicative.

## Writing for the Top-4 {#sec-sw-top4}

### What rises to a top journal

The top-4 differ from solid field journals less in the correctness of the work they
publish than in the size and generality of the increment $\Delta\mathcal{K}$. The
bar that rises is **importance**: a top-4 paper must change how a non-trivial
slice of the field thinks, not merely add a brick to a narrow wall. Editors and
reviewers operationalize this in recurring questions—Would I assign this in a
doctoral seminar? Does it change what I would tell a manager? Does it open new
research, or close a question others were pursuing? A paper can be flawless and
still be rejected for answering a question too few people were asking.

The venues also have distinguishable identities, and matching the paper to the
venue is part of the craft. *Marketing Science* and JMR lean toward
methodological rigor and formal models—a structural estimator or an analytical
game-theoretic result is at home there. JCR prizes psychological process and
theory, typically established across multiple experiments that triangulate a
mechanism. JM spans substantive and managerial questions with an explicit demand
for managerial relevance. The same finding, dressed for the wrong venue, draws
reviewers who want what the paper was never built to deliver.

### Genres of contribution: many roads to the top-4

The importance bar is not cleared in a single way. Marketing is methodologically
plural, and a top journal publishes several distinct *genres* of research, each with
its own logic of contribution and its own standard of proof. Naming the genre—to
oneself first, then to the reader—matters as much as naming the paper type earlier in
this chapter, because a reviewer recruited for one genre will judge by its rules. A
working map of the traditions a marketing scholar will meet:

- **Experimental and consumer-behavior** work identifies a psychological *mechanism*
  by controlled manipulation, usually triangulated across several studies that close
  off rival accounts one at a time. The contribution is a process: not merely that $X$
  moves $Y$ but *why*.
- **Analytical / game-theoretic** work builds a deliberately stylized model and
  extracts a non-obvious, general implication from its equilibrium—why a firm might
  rationally do the counterintuitive thing. The currency is generality, not data.
- **Empirical / structural** work estimates the primitives of behavior—preferences,
  costs—under an explicit identification argument so it can run *counterfactuals* the
  raw data never ran (@sec-sm-discrete-choice). The contribution is policy-relevant
  quantities and the "what if" they license.
- **Descriptive** work documents robust, previously unseen *patterns* in novel—often
  large-scale—data, and makes *no* causal claim. The contribution is the fact itself.
- **Prescriptive / optimization** work formulates a decision problem and solves it with
  a guarantee or a demonstrably better method (@sec-sm-socpd). The contribution is a
  solvable formulation and its performance.
- **Strategy and marketing-management** work links a marketing action to firm outcomes
  and firm value with an explicit demand for managerial relevance (the JM register
  above).

These are axes, not bins: a single paper is often empirical *and* descriptive, or
analytical *and* prescriptive, and the strongest work borrows across them. But the
reader must be told which claim bears the weight, because the standard of proof differs
by genre—an identification argument is decisive for a structural paper and beside the
point for an analytical one.

The genre students most often underrate is the **descriptive**, and it repays a closer
look because its revival is one of the field's live developments. For two decades the
credibility revolution (@sec-credibility) so prized causal identification that pure
description read as second class—a table on the way to a regression. That has reversed,
for a concrete reason: platform, transaction, and sensor data now expose patterns no
prior dataset could show, and a pattern no one had seen is a genuine increment to
$\mathcal{K}$ even with no causal arrow attached. The descriptive paper still clears the
top-4 bar, but on *different* terms. It does not owe internal validity for an effect it
never claims; it owes (i) genuinely novel data, (ii) patterns robust to how the data are
sliced, (iii) a *surprise*—a fact that overturns or sharpens a belief the field held—and
(iv) scrupulous candor about what is described versus explained. Overclaiming causation
is the descriptive paper's characteristic failure; under-selling importance ("we merely
describe") is its mirror-image one.

@chen2026counterfeits is a clean specimen. From millions of cross-border orders the
authors document a *U-shaped* relationship between income and counterfeit
consumption—both low- and high-income consumers buy more fakes than the middle, with the
tails differing in composition—advancing no instrument and estimating no treatment
effect. Its title announces "A Descriptive Analysis," and its contribution is precisely
the documented gradient, which complicates the tidy poseur account of counterfeit demand
(@sec-branding). That the top journals now run short-format *Frontiers* articles for
exactly this kind of high-value description—@kaiser2026chatgpt's map of a nascent LLM
sales channel is a companion case—is institutional evidence that the genre has a settled
place. The craft lesson mirrors the framing discipline of @sec-sw-contribution: know your
genre, promise the kind of contribution it actually makes, and defend it on its own terms
rather than borrowing the rhetoric of a genre you did not practice.

### Robustness as rhetoric

At the top-4, a single clean result rarely suffices; the paper must anticipate and
preempt the reviewer's alternative explanations. This is why top-4 empirical papers
carry batteries of robustness checks, alternative specifications, and—in
experimental work—multiple studies whose designs rule out competing accounts one
by one. The logic is the broken-arrow logic of @eq-sw-chain: every alternative
explanation is a way the arrow $D \Rightarrow R$ might fail, and each robustness
check is a patch that keeps the arrow intact. The writing task is to present this
defensive machinery without burying the headline—typically by stating the main
result cleanly, then organizing the defenses as a navigable sequence of named
threats and the evidence that disarms each.

> The reviewer is not your enemy; the reviewer is the proxy for every skeptical
> reader your paper will ever have. Every alternative explanation a reviewer
> raises is one a future reader would have raised silently and then disbelieved
> you. Writing the rebuttal into the paper *before* submission is simply doing the
> reviewer's job for them, and it is the single highest-return revision activity.

### The review process and the response

Top-4 papers are essentially never accepted on first submission; the modal good
outcome is a *revise and resubmit* (R&R), and the revision is where many papers are
won or lost. The response to reviewers is itself a document with its own craft. Its
governing principle is that the author's job is to make the editor's
accept-decision *easy to defend*: every reviewer comment receives a numbered,
verbatim restatement followed by a specific response and a pointer to the exact
change in the manuscript. Disagreement is permitted but must be argued with
evidence, courteously, and sparingly—an author who fights every point signals
defensiveness, while an author who concedes everything signals that the original
work was unconsidered. The art is to distinguish comments that improve the paper
(adopt them) from comments that reflect a misunderstanding (clarify the text so the
next reader does not misunderstand either) from the rare comment that is simply
wrong (rebut it with evidence and grace).

## Clarity and Revision {#sec-sw-clarity}

### Clarity as a property of the reader, not the writer

Clarity is not an ornament; it is the probability that a reader recovers the
intended meaning on the first pass. Writing that the author finds clear because the
author already knows the answer is a category error—the test of clarity is whether
a reader *who does not yet know the conclusion* can follow the argument. A few
sentence-level disciplines do most of the work, and they are mechanical enough to
apply as a checklist. Prefer the active voice and a concrete agent ("we estimate",
"the model predicts") over agentless passives that hide who did what. Put the
subject and verb early and close together so the reader is not made to hold an open
clause across half a line. Use one term for one concept—do not alternate "effect",
"impact", and "influence" for the same quantity, because the reader will hunt for a
distinction that is not there. Expand every acronym on first use and define every
construct before deploying it, a discipline the house style enforces throughout.

### Measuring readability

Readability can be measured, and while no formula substitutes for judgment, the
indices make a useful diagnostic. The **Gunning fog index** estimates the years of
formal education a reader needs to understand a passage on first reading
[@gunning1952technique]. Let a passage contain $W$ words, $S$ sentences, and $C$
"complex" words (three or more syllables, excluding common suffix inflections).
The index is

$$
\text{Fog} \;=\; 0.4\left(\frac{W}{S} \;+\; 100\,\frac{C}{W}\right),
$$ {#eq-sw-fog}

a weighted sum of average sentence length and the percentage of complex words.
The two terms encode the two main sources of difficulty—long sentences and
unfamiliar words—and the constant scales the result to a U.S. grade level, so a fog
score of 12 corresponds roughly to the reading level of a high-school senior.
Academic prose runs higher, but a methods section with a fog score in the high
teens is usually not "rigorous"; it is merely overlong in the sentence and
overloaded in the noun phrase. Textual-complexity measures of exactly this family
are now standard tools in the marketing and finance literatures for analyzing
corporate disclosures and other documents at scale [@loughran2020; @loughran2020a],
which is a useful reminder that the readability of *your* prose is itself a
measurable, and therefore improvable, quantity.

The estimator in @eq-sw-fog is a heuristic, not a model, and its assumptions are
worth stating because they tell you when it misleads. It assumes that syllable
count proxies word difficulty (false for short technical jargon like "prior" or
"yield", which are hard despite being monosyllabic) and that sentence length
proxies syntactic load (false for a long but perfectly parallel list). It is best
read as a *smoke detector*: a high score reliably indicates a problem somewhere,
but a low score does not certify clarity. The following chunk implements it and a
companion length diagnostic so the measure is reproducible rather than mystical;
@fig-sw-fog shows the score falling as the same sentence is progressively tightened.

```{r}
#| label: fig-sw-fog
#| fig-cap: "Gunning fog scores for three versions of the same sentence—a bloated original, a moderate edit, and a lean rewrite—showing how shortening sentences and replacing complex words lowers the estimated reading grade. The dashed line marks a fog score of 12 (high-school senior)."
#| echo: true
#| message: false
#| warning: false
set.seed(41)

# A deliberately rough syllable counter: count vowel groups per word.
# Heuristic, not linguistic ground truth -- adequate for a relative diagnostic.
count_syllables <- function(word) {
  word <- tolower(gsub("[^a-z]", "", word))
  if (nchar(word) == 0) return(0L)
  groups <- gregexpr("[aeiouy]+", word)[[1]]
  n <- if (groups[1] == -1L) 1L else length(groups)
  if (grepl("e$", word) && n > 1L) n <- n - 1L   # silent terminal 'e'
  max(n, 1L)
}

gunning_fog <- function(text) {
  sentences <- unlist(strsplit(text, "(?<=[.!?])\\s+", perl = TRUE))
  sentences <- sentences[nzchar(trimws(sentences))]
  words <- unlist(strsplit(text, "\\s+"))
  words <- words[nzchar(words)]
  W <- length(words); S <- max(length(sentences), 1L)
  syl <- vapply(words, count_syllables, integer(1))
  C <- sum(syl >= 3L)                              # "complex" words
  0.4 * (W / S + 100 * C / W)
}

versions <- c(
  bloated  = paste("The utilization of an excessively elaborate and",
                   "circumlocutory prose style, characterized by the",
                   "accumulation of subordinate clauses, demonstrably",
                   "attenuates comprehensibility for the readership."),
  moderate = paste("Using an elaborate, indirect prose style with many",
                   "subordinate clauses reduces how well readers understand",
                   "the argument."),
  lean     = "Dense prose is hard to read."
)

fog <- vapply(versions, gunning_fog, numeric(1))
print(round(fog, 1))

barplot(fog, col = "grey80", border = NA, ylab = "Gunning fog (grade level)",
        main = "Readability across three drafts of one claim",
        names.arg = names(versions))
abline(h = 12, lty = 2)
```

### Revision as a process

First drafts are for the writer; revision is for the reader. The empirical
regularity behind all writing advice is that meaning is discovered, not transcribed:
the act of writing reveals which arrows in @eq-sw-chain are actually weak, and the
first draft's value is largely diagnostic. Productive revision proceeds top-down—
structure before paragraph, paragraph before sentence, sentence before word—because
fixing a comma in a paragraph that will be deleted is wasted effort. A useful
sequence is to first verify that the contribution sentence is present and true,
then that each section answers its one question, then that each paragraph has a
single point announced in its first sentence, and only then to polish prose. The
"reverse outline"—reading the draft and writing in the margin the single claim each
paragraph makes—exposes structural defects (two paragraphs making the same point, a
claim with no supporting paragraph, a paragraph with no claim) that are invisible at
the sentence level.

A final discipline is distance. Prose that seems clear the night it is written is
frequently opaque a week later, when the author has forgotten the unstated
assumptions that made it cohere—and a week-later author is a closer proxy for the
real reader than the night-of author ever is. Where time does not permit distance,
a co-author or colleague reading cold is the next best instrument, and the most
valuable feedback is not "I disagree" but "I got lost here", because confusion
localizes a broken arrow that the author can no longer see.

## Key Takeaways

- A paper is a **reconstructed argument**, not a lab diary: the order of exposition
  is engineered for the reader, and each section answers exactly one question with
  its own default verb tense (@tbl-sw-sections).
- The argument is a chain of claims the reader may refuse to grant
  (@eq-sw-chain); a paper fails at its **weakest arrow**, and most "low
  contribution" rejections are in fact broken arrows—a filled gap or a design that
  cannot license the conclusion.
- A contribution is the warranted increment to knowledge $\Delta\mathcal{K}$
  (@eq-sw-contribution); it must be **novel, valid, and important**, and the
  top-4 bar that rises is importance. State the most general claim the evidence
  licenses, and not one inch more.
- The **front end** carries the promise; the discussion must redeem exactly that
  promise. Make the "in this paper, we..." pivot early and state contributions as
  claims, not activities (@fig-sw-funnel).
- Not every paper is hypo-deductive. A **descriptive** paper contributes a
  pattern and names its ceiling; an **abductive** paper reasons from the pattern
  to the best explanation and owes the reader the whole candidate set
  (@eq-sw-abduction). The **PEEBI** structure—priors, establish, evaluate, best
  explanation, implication (@tbl-sw-peebi)—reports that reasoning honestly where
  the deductive template would force a fiction about the order of discovery.
- **Robustness is rhetoric**: writing the reviewer's objections into the paper
  before submission is the highest-return revision activity, and the R&R response is
  a craft document whose job is to make the editor's decision easy to defend.
- **Clarity is a property of the reader**: measure it (@eq-sw-fog), revise
  top-down from structure to sentence, and use distance or a cold reader to localize
  the arrows you can no longer see are broken.

## Further Reading

For the structure of scientific argument and the discipline of revision, the
classic writing-craft literature is the natural starting point; for the specific
norms of the marketing top-4, the editorials and "from the editor" notes of JM,
JMR, *Marketing Science*, and JCR are the authoritative—and frequently updated—
statements of what each venue rewards. <!-- CITE-NEEDED: editorial/from-the-editor pieces on contribution standards in JM, JMR, Marketing Science, JCR --> The
On the reporting of non-deductive work, @pillai2026peebi is the direct statement
of the PEEBI structure and @pillai2024lovely the companion treatment of what
makes one explanation better than another; @choudhury2026abduction supplies the
design-side account of abduction that the two SMJ papers presuppose. Read
together they are the closest thing the field has to a manual for writing a paper
whose conclusion was discovered rather than predicted.

The
textual-analysis methods that let us measure the readability of our own and others'
prose are developed in the disclosure literature [@loughran2020; @loughran2020a]
and connect this chapter to the broader treatment of text as data elsewhere in the
book.
