# Scientific Writing {#sec-scientific-writing}

A research paper is the instrument through which a discovery becomes knowledge.
A result that is not written down does not exist scientifically; a result written
badly exists only for the handful of readers willing to excavate it. This chapter
treats scientific writing not as cosmetic polish applied after the "real" work but
as an integral part of the research itself—a design problem with its own
constraints, its own failure modes, and its own measurable quality. The audience
is the marketing scholar who has a finding and now faces the harder task of
convincing a skeptical editor, two or three adversarial reviewers, and an
indifferent field that the finding is correct, novel, and important.

The economics of the situation are unforgiving. The top journals in marketing—the
*Journal of Marketing* (JM), the *Journal of Marketing Research* (JMR),
*Marketing Science*, and the *Journal of Consumer Research* (JCR), collectively
the "top-4"—desk-reject a large share of submissions and ultimately accept only a
small fraction of what is sent to them. <!-- CITE-NEEDED: top-4 marketing journal acceptance/desk-reject rates -->
A reviewer reads dozens of papers a year as unpaid labor and forms a judgment
within the first few pages. The paper that survives is not merely the one with the
best result; it is the one whose argument the reviewer can reconstruct without
effort and whose contribution the reviewer can state in a sentence. Writing is
therefore a competitive technology, and this chapter develops it as one.

The chapter proceeds from architecture to sentence. We begin with the **structure**
of a paper—the canonical section sequence and what each section is for—and formalize
the underlying logical scaffold. We then treat the hardest and most consequential
task, **framing the contribution**, and the part of the paper where framing
lives or dies: the **front end** (title, abstract, introduction). We turn next to
the specific demands of writing for the **top-4**, where the contribution bar is
highest. We close with **clarity and revision**: the sentence-level craft that
makes an argument legible, the measurement of readability, and the disciplined
process of responding to reviewers. Throughout, the stance is the house stance—
intuition first, then the formal structure that makes the intuition operational.

## The Structure of a Paper {#sec-sw-structure}

### The canonical sections and their tenses

Empirical marketing papers share a near-universal skeleton, sometimes abbreviated
**IMRaD** (Introduction, Methods, Results, and Discussion) with literature review
and conclusions bracketing the core. Each section answers one question and only
one, and—this is a point novices routinely miss—each section has a *default verb
tense* that signals the epistemic status of its claims. Completed actions take the
**simple past** ("we estimated", "respondents rated"); established knowledge and
the paper's own ongoing argument take the **present** ("the effect is", "Table 2
reports"); prior literature as a stream that continues into the present takes the
**present perfect** ("researchers have shown"). @tbl-sw-sections collects the
sections, their purposes, approximate lengths, and tense conventions.

```{r}
#| label: tbl-sw-sections
#| tbl-cap: "Canonical sections of an empirical marketing paper, with purpose, indicative length, and default verb tense. Lengths are rules of thumb for a full-length top-4 submission, not hard limits."
#| echo: false
#| message: false
#| warning: false
library(knitr)
sw_sections <- data.frame(
  Section = c("Title", "Abstract", "Introduction", "Literature / Theory",
              "Methods", "Results", "Discussion", "Conclusion"),
  `Words` = c("8–15", "200–250", "800–1200", "1500–3000",
              "1000–2500", "1000–2500", "1500–2500", "400–800"),
  Purpose = c("Name the contribution",
              "Self-contained summary: objective, method, finding, implication",
              "Why the paper exists; the gap and the contribution",
              "Position vs. prior work; derive hypotheses or model",
              "What was done, reproducibly",
              "What was found, with the data doing the talking",
              "What it means, limits, boundary conditions",
              "Restate contribution; implications and future work"),
  `Default tense` = c("—", "Past + present", "Present + present perfect",
                      "Present + present perfect", "Past", "Past + present",
                      "Present + past", "Present"),
  check.names = FALSE
)
kable(sw_sections)
```

The discipline of one-question-per-section is what gives a paper its navigability.
A reader who wants to know *what you did* turns to Methods and finds only that; a
reader who wants to know *why it matters* turns to the Discussion and is not
detained by procedure. Violations—results smuggled into the methods, mechanism
speculation buried in the results—are the most common structural defect flagged in
review, because they force the reader to hold the paper's logic in their head
rather than read it off the page.

> A scientific paper is not a chronological record of what the researcher did. It
> is a reconstructed argument, organized so that the reader can verify the
> conclusion with minimum effort. The order of discovery and the order of
> exposition are different things, and conflating them is the surest route to an
> unreadable manuscript.

### The argument as a logical object

Underneath the section skeleton sits a logical scaffold that the prose merely
dresses. Make it explicit. A confirmatory empirical paper advances a chain

$$
\underbrace{P}_{\text{problem}} \;\Rightarrow\;
\underbrace{G}_{\text{gap}} \;\Rightarrow\;
\underbrace{H}_{\text{hypotheses}} \;\Rightarrow\;
\underbrace{D}_{\text{design}} \;\Rightarrow\;
\underbrace{R}_{\text{results}} \;\Rightarrow\;
\underbrace{C}_{\text{contribution}}
$$ {#eq-sw-chain}

in which every arrow is a claim the reader can refuse to grant. The problem $P$
must be real (someone cares about the answer); the gap $G$ must be genuine (the
answer is not already known); the hypotheses $H$ must follow from a theory and be
falsifiable; the design $D$ must be capable of discriminating $H$ from its
negation; the results $R$ must actually bear on $H$; and the contribution $C$ must
be what $R$ licenses, not more. A paper fails at whichever arrow is weakest, and
the writing task is to make each arrow as short and as forced as possible. Many
rejections that are phrased as "not enough contribution" are in fact broken
arrows: a gap that turns out to be already filled ($G$), or a design that cannot
separate the focal hypothesis from a confound ($D \not\Rightarrow R$).

The conceptual or theory paper substitutes a different chain—a construct is
defined, related to existing constructs, and shown to do explanatory work—but the
demand for unbroken arrows is identical. @fig-sw-paper-flow renders the canonical
flow and shows where each link in @eq-sw-chain is forged.

```{mermaid}
%%| label: fig-sw-paper-flow
%%| fig-cap: "The logical flow of an empirical paper. The front end (title, abstract, introduction) carries the problem, gap, and promised contribution; the body discharges the promise. A reviewer reads the front end to decide whether the rest is worth reading."
flowchart TD
    A["Title<br/><i>names the contribution</i>"] --> B["Abstract<br/><i>P → R → C in 250 words</i>"]
    B --> C["Introduction<br/><i>problem P, gap G,<br/>promised contribution C</i>"]
    C --> D["Theory / Literature<br/><i>position; derive H</i>"]
    D --> E["Methods<br/><i>design D</i>"]
    E --> F["Results<br/><i>evidence R</i>"]
    F --> G["Discussion<br/><i>does R license C?</i>"]
    G --> H["Conclusion<br/><i>restate C; boundaries</i>"]
    C -.->|"promise"| G
    H -.->|"payoff"| C
    style C fill:#e8f0fe,stroke:#1a73e8
    style G fill:#e8f0fe,stroke:#1a73e8
```

The dashed promise–payoff loop in @fig-sw-paper-flow is the structural heart of a
paper. The introduction makes a promise—"we will show that $X$ causes $Y$ through
mechanism $M$"—and the discussion must redeem exactly that promise, no more and no
less. A paper that promises a causal claim and delivers a correlation, or that
delivers a richer finding than it promised and never updates the framing, reads as
incoherent even when every individual sentence is sound.

### Paper types

The skeleton flexes with the paper's type, and naming the type early orients both
writer and reviewer. Four types recur. A **research paper** tests hypotheses and
reports findings; its contribution lives in $H \Rightarrow R$. A **methods paper**
proposes a new estimator, design, or measurement instrument; its contribution is
that the method recovers something prior methods could not, so the burden is a
demonstration—often on synthetic data with known ground truth—that it works and a
comparison against incumbents. A **review paper** consolidates a domain, and its
contribution is organization: a taxonomy, an integrative framework, or an agenda
that practitioners and scholars did not previously have. A **conceptual or
discussion paper** advances or critiques theory, trading empirical evidence for
argumentative rigor. The reviewer evaluates each type against a different
standard, and a paper that does not signal its type invites evaluation against the
wrong one.

## Framing the Contribution {#sec-sw-contribution}

### What a contribution is

The single most important sentence in a paper is the one that states its
contribution, and the most common reason good work is rejected is that this
sentence is missing, vague, or overclaimed. Define the contribution formally.
Let $\mathcal{K}$ denote the field's existing knowledge—the set of claims a
competent reader already accepts before reading the paper. A paper's contribution
is the increment

$$
\Delta\mathcal{K} \;=\; \mathcal{K}_{\text{after}} \setminus \mathcal{K}_{\text{before}},
$$ {#eq-sw-contribution}

the set of claims that a reader is licensed to accept *after* reading the paper and
was not licensed to accept before. Three properties make $\Delta\mathcal{K}$
publishable, and reviewers probe each one explicitly:

- **Novelty** ($\Delta\mathcal{K} \neq \varnothing$): the increment is non-empty.
  If every claim in the paper already belonged to $\mathcal{K}_{\text{before}}$,
  there is no contribution, however well executed the study.
- **Validity** ($\Delta\mathcal{K}$ is warranted): the increment is actually
  licensed by the evidence and argument. Overclaiming—asserting elements of
  $\Delta\mathcal{K}$ that the design $D$ cannot support—is the fastest route to a
  reviewer's "the conclusions outrun the data".
- **Importance** ($|\Delta\mathcal{K}|$ matters): the increment is one the field
  cares about. A true, novel, but trivial claim clears the first two bars and fails
  the third, and "importance" is precisely the bar that rises from a field journal
  to the top-4.

Two broad species of contribution recur in marketing, and a paper should know which
it is making. A **substantive** contribution changes what we believe about a
marketing phenomenon—that brand prominence signals status through quiet versus loud
cues, say, or that responding to online reviews trades complaint quantity for depth
(see @sec-branding). A **methodological** contribution changes how we can learn
about phenomena—a new identification strategy, a scalable text measure, a more
efficient estimator. The strongest papers often pair a methodological advance with
the substantive finding it unlocks, but a paper that is fuzzy about which kind of
increment it offers will be read as offering neither.

### The contribution statement and the framing trap

A contribution should be expressible as a single declarative sentence of the form
"We show that [claim], which prior work did not establish because [reason], using
[design]." The clause "which prior work did not establish" is the gap $G$ of
@eq-sw-chain made explicit, and it must be *true*: nothing destroys credibility
faster than a literature-savvy reviewer who knows the claimed gap was filled years
ago. Conversely, a contribution can be undersold. Authors who have lived inside a
problem for years often state their finding in the narrow terms in which they
discovered it, missing the more general claim the evidence actually supports. The
discipline is to state the *most general* claim that $R$ licenses and not one inch
more.

Framing is the act of choosing which $\mathcal{K}_{\text{before}}$ to write against—
which conversation the paper joins—and it is a genuine strategic choice, not a
neutral description. The same result can be framed as a contribution to several
literatures, and the framing determines the reference set, the reviewers, and the
perceived importance. A study of how customers respond to a salesperson's language,
for instance, could be framed against the persuasion literature, the
frontline-service literature, or the computational-text-analysis literature, and
the right choice depends on where the increment $\Delta\mathcal{K}$ is largest and
the audience most receptive. The **framing trap** is to choose the literature in
which the result is *most surprising* rather than the one in which it is *most
defensible*; surprise attracts an editor's eye but invites reviewers who hold the
strongest priors against the claim.

## The Front End {#sec-sw-front-end}

The front end—title, abstract, and introduction—is the part of the paper most
people read and the only part most people read. It carries the entire promise–
payoff loop of @fig-sw-paper-flow in compressed form, and a reviewer's
accept-lean or reject-lean is often set before the methods section begins. The
front end deserves a disproportionate share of writing effort precisely because it
does a disproportionate share of the persuasive work.

### The title

A title's job is to name the contribution in eight to fifteen words such that the
right reader stops scrolling. It should contain the paper's key constructs (the
terms a searcher would type) and, where possible, signal the relationship the paper
establishes rather than merely the topic it addresses. "Brand Prominence and
Status Signaling" names a topic; "Signaling Status with Luxury Goods: The Role of
Brand Prominence" names a relationship and a mechanism. Question titles and
two-part titles (a hook before the colon, the substance after) are common in JCR
and JM and rarer in *Marketing Science*, reflecting the venues' different house
voices—a difference worth matching to the target journal.

### The abstract

The abstract is the most-read and least-revised 250 words in science, and the
imbalance is a mistake. It must be **self-contained**: a reader who sees only the
abstract should be able to state the paper's objective, method, principal finding,
and implication. The canonical structure mirrors @eq-sw-chain in miniature—one or
two sentences each for the problem, the approach, the result (with direction and,
where it fits, magnitude), and the "so what". Two failure modes dominate. The
**descriptive abstract** announces what the paper does ("we examine the effect of
X on Y") without ever stating what it *found*, leaving the reader no wiser; the
**overloaded abstract** crams in every robustness check and moderator, drowning the
headline. The fix for both is the same: lead with the finding.

### The introduction as a funnel

The introduction is where most rejections are decided, and it has a near-mechanical
structure: a **funnel** from the broad problem to the specific contribution. A
robust template runs problem $\rightarrow$ tension or puzzle $\rightarrow$ gap
$\rightarrow$ "in this paper we" $\rightarrow$ approach $\rightarrow$ findings
$\rightarrow$ contributions $\rightarrow$ roadmap. @fig-sw-funnel renders it.

```{mermaid}
%%| label: fig-sw-funnel
%%| fig-cap: "The introduction funnel. Each band narrows from the world's problem to this paper's specific increment to knowledge. The 'in this paper, we...' sentence is the pivot from motivation to contribution and should appear by the end of the second page."
flowchart TD
    P["The broad problem<br/><i>why anyone should care</i>"] --> T["The tension / puzzle<br/><i>what doesn't fit</i>"]
    T --> G["The gap<br/><i>what prior work left open</i>"]
    G --> Q["<b>In this paper, we…</b><br/><i>the pivot to contribution</i>"]
    Q --> A2["Approach<br/><i>data and method, one paragraph</i>"]
    A2 --> F["Findings<br/><i>headline results, with direction</i>"]
    F --> C2["Contributions<br/><i>ΔK, stated as claims</i>"]
    C2 --> R2["Roadmap<br/><i>optional, brief</i>"]
    style Q fill:#fce8e6,stroke:#d93025
    style C2 fill:#e6f4ea,stroke:#188038
```

Two discipline points govern the funnel. First, the pivot sentence—"In this paper,
we..."—should arrive early, by the end of the second page; an introduction that is
still motivating the problem on page four has lost the reviewer. Second, the
contributions should be stated as *claims the reader will be able to accept*, not
as activities the authors performed. "We contribute by examining the role of
arousal" is an activity; "We show that empathetic responses raise gratitude even
before a service failure is resolved, which the prior literature's
resolution-centric models cannot accommodate" is a claim and an explicit
$\Delta\mathcal{K}$.

### Positioning against the literature

The literature review is not a homage to everyone who has touched the topic; it is
the construction of $\mathcal{K}_{\text{before}}$, the precise baseline against
which the contribution is measured. Its house failure mode is the **annotated
bibliography**—a serial list of "Author (year) found X; Author (year) found Y"—which
demonstrates reading but not synthesis and leaves the reviewer to infer the gap.
The remedy is to organize the review around *tensions and open questions* rather
than around papers, so that the gap emerges as the natural next move in a
conversation the reader can now follow. The house style guide's instruction to
**weave, not list** citations is exactly this principle applied at the sentence
level: each citation should earn its place by advancing the argument, and the
strongest evidence per point should crowd out the duplicative.

## Writing for the Top-4 {#sec-sw-top4}

### What rises to a top journal

The top-4 differ from solid field journals less in the correctness of the work they
publish than in the size and generality of the increment $\Delta\mathcal{K}$. The
bar that rises is **importance**: a top-4 paper must change how a non-trivial
slice of the field thinks, not merely add a brick to a narrow wall. Editors and
reviewers operationalize this in recurring questions—Would I assign this in a
doctoral seminar? Does it change what I would tell a manager? Does it open new
research, or close a question others were pursuing? A paper can be flawless and
still be rejected for answering a question too few people were asking.

The venues also have distinguishable identities, and matching the paper to the
venue is part of the craft. *Marketing Science* and JMR lean toward
methodological rigor and formal models—a structural estimator or an analytical
game-theoretic result is at home there. JCR prizes psychological process and
theory, typically established across multiple experiments that triangulate a
mechanism. JM spans substantive and managerial questions with an explicit demand
for managerial relevance. The same finding, dressed for the wrong venue, draws
reviewers who want what the paper was never built to deliver.

### Genres of contribution: many roads to the top-4

The importance bar is not cleared in a single way. Marketing is methodologically
plural, and a top journal publishes several distinct *genres* of research, each with
its own logic of contribution and its own standard of proof. Naming the genre—to
oneself first, then to the reader—matters as much as naming the paper type earlier in
this chapter, because a reviewer recruited for one genre will judge by its rules. A
working map of the traditions a marketing scholar will meet:

- **Experimental and consumer-behavior** work identifies a psychological *mechanism*
  by controlled manipulation, usually triangulated across several studies that close
  off rival accounts one at a time. The contribution is a process: not merely that $X$
  moves $Y$ but *why*.
- **Analytical / game-theoretic** work builds a deliberately stylized model and
  extracts a non-obvious, general implication from its equilibrium—why a firm might
  rationally do the counterintuitive thing. The currency is generality, not data.
- **Empirical / structural** work estimates the primitives of behavior—preferences,
  costs—under an explicit identification argument so it can run *counterfactuals* the
  raw data never ran (@sec-sm-discrete-choice). The contribution is policy-relevant
  quantities and the "what if" they license.
- **Descriptive** work documents robust, previously unseen *patterns* in novel—often
  large-scale—data, and makes *no* causal claim. The contribution is the fact itself.
- **Prescriptive / optimization** work formulates a decision problem and solves it with
  a guarantee or a demonstrably better method (@sec-sm-socpd). The contribution is a
  solvable formulation and its performance.
- **Strategy and marketing-management** work links a marketing action to firm outcomes
  and firm value with an explicit demand for managerial relevance (the JM register
  above).

These are axes, not bins: a single paper is often empirical *and* descriptive, or
analytical *and* prescriptive, and the strongest work borrows across them. But the
reader must be told which claim bears the weight, because the standard of proof differs
by genre—an identification argument is decisive for a structural paper and beside the
point for an analytical one.

The genre students most often underrate is the **descriptive**, and it repays a closer
look because its revival is one of the field's live developments. For two decades the
credibility revolution (@sec-credibility) so prized causal identification that pure
description read as second class—a table on the way to a regression. That has reversed,
for a concrete reason: platform, transaction, and sensor data now expose patterns no
prior dataset could show, and a pattern no one had seen is a genuine increment to
$\mathcal{K}$ even with no causal arrow attached. The descriptive paper still clears the
top-4 bar, but on *different* terms. It does not owe internal validity for an effect it
never claims; it owes (i) genuinely novel data, (ii) patterns robust to how the data are
sliced, (iii) a *surprise*—a fact that overturns or sharpens a belief the field held—and
(iv) scrupulous candor about what is described versus explained. Overclaiming causation
is the descriptive paper's characteristic failure; under-selling importance ("we merely
describe") is its mirror-image one.

@chen2026counterfeits is a clean specimen. From millions of cross-border orders the
authors document a *U-shaped* relationship between income and counterfeit
consumption—both low- and high-income consumers buy more fakes than the middle, with the
tails differing in composition—advancing no instrument and estimating no treatment
effect. Its title announces "A Descriptive Analysis," and its contribution is precisely
the documented gradient, which complicates the tidy poseur account of counterfeit demand
(@sec-branding). That the top journals now run short-format *Frontiers* articles for
exactly this kind of high-value description—@kaiser2026chatgpt's map of a nascent LLM
sales channel is a companion case—is institutional evidence that the genre has a settled
place. The craft lesson mirrors the framing discipline of @sec-sw-contribution: know your
genre, promise the kind of contribution it actually makes, and defend it on its own terms
rather than borrowing the rhetoric of a genre you did not practice.

### Robustness as rhetoric

At the top-4, a single clean result rarely suffices; the paper must anticipate and
preempt the reviewer's alternative explanations. This is why top-4 empirical papers
carry batteries of robustness checks, alternative specifications, and—in
experimental work—multiple studies whose designs rule out competing accounts one
by one. The logic is the broken-arrow logic of @eq-sw-chain: every alternative
explanation is a way the arrow $D \Rightarrow R$ might fail, and each robustness
check is a patch that keeps the arrow intact. The writing task is to present this
defensive machinery without burying the headline—typically by stating the main
result cleanly, then organizing the defenses as a navigable sequence of named
threats and the evidence that disarms each.

> The reviewer is not your enemy; the reviewer is the proxy for every skeptical
> reader your paper will ever have. Every alternative explanation a reviewer
> raises is one a future reader would have raised silently and then disbelieved
> you. Writing the rebuttal into the paper *before* submission is simply doing the
> reviewer's job for them, and it is the single highest-return revision activity.

### The review process and the response

Top-4 papers are essentially never accepted on first submission; the modal good
outcome is a *revise and resubmit* (R&R), and the revision is where many papers are
won or lost. The response to reviewers is itself a document with its own craft. Its
governing principle is that the author's job is to make the editor's
accept-decision *easy to defend*: every reviewer comment receives a numbered,
verbatim restatement followed by a specific response and a pointer to the exact
change in the manuscript. Disagreement is permitted but must be argued with
evidence, courteously, and sparingly—an author who fights every point signals
defensiveness, while an author who concedes everything signals that the original
work was unconsidered. The art is to distinguish comments that improve the paper
(adopt them) from comments that reflect a misunderstanding (clarify the text so the
next reader does not misunderstand either) from the rare comment that is simply
wrong (rebut it with evidence and grace).

## Clarity and Revision {#sec-sw-clarity}

### Clarity as a property of the reader, not the writer

Clarity is not an ornament; it is the probability that a reader recovers the
intended meaning on the first pass. Writing that the author finds clear because the
author already knows the answer is a category error—the test of clarity is whether
a reader *who does not yet know the conclusion* can follow the argument. A few
sentence-level disciplines do most of the work, and they are mechanical enough to
apply as a checklist. Prefer the active voice and a concrete agent ("we estimate",
"the model predicts") over agentless passives that hide who did what. Put the
subject and verb early and close together so the reader is not made to hold an open
clause across half a line. Use one term for one concept—do not alternate "effect",
"impact", and "influence" for the same quantity, because the reader will hunt for a
distinction that is not there. Expand every acronym on first use and define every
construct before deploying it, a discipline the house style enforces throughout.

### Measuring readability

Readability can be measured, and while no formula substitutes for judgment, the
indices make a useful diagnostic. The **Gunning fog index** estimates the years of
formal education a reader needs to understand a passage on first reading
[@gunning1952technique]. Let a passage contain $W$ words, $S$ sentences, and $C$
"complex" words (three or more syllables, excluding common suffix inflections).
The index is

$$
\text{Fog} \;=\; 0.4\left(\frac{W}{S} \;+\; 100\,\frac{C}{W}\right),
$$ {#eq-sw-fog}

a weighted sum of average sentence length and the percentage of complex words.
The two terms encode the two main sources of difficulty—long sentences and
unfamiliar words—and the constant scales the result to a U.S. grade level, so a fog
score of 12 corresponds roughly to the reading level of a high-school senior.
Academic prose runs higher, but a methods section with a fog score in the high
teens is usually not "rigorous"; it is merely overlong in the sentence and
overloaded in the noun phrase. Textual-complexity measures of exactly this family
are now standard tools in the marketing and finance literatures for analyzing
corporate disclosures and other documents at scale [@loughran2020; @loughran2020a],
which is a useful reminder that the readability of *your* prose is itself a
measurable, and therefore improvable, quantity.

The estimator in @eq-sw-fog is a heuristic, not a model, and its assumptions are
worth stating because they tell you when it misleads. It assumes that syllable
count proxies word difficulty (false for short technical jargon like "prior" or
"yield", which are hard despite being monosyllabic) and that sentence length
proxies syntactic load (false for a long but perfectly parallel list). It is best
read as a *smoke detector*: a high score reliably indicates a problem somewhere,
but a low score does not certify clarity. The following chunk implements it and a
companion length diagnostic so the measure is reproducible rather than mystical;
@fig-sw-fog shows the score falling as the same sentence is progressively tightened.

```{r}
#| label: fig-sw-fog
#| fig-cap: "Gunning fog scores for three versions of the same sentence—a bloated original, a moderate edit, and a lean rewrite—showing how shortening sentences and replacing complex words lowers the estimated reading grade. The dashed line marks a fog score of 12 (high-school senior)."
#| echo: true
#| message: false
#| warning: false
set.seed(41)

# A deliberately rough syllable counter: count vowel groups per word.
# Heuristic, not linguistic ground truth -- adequate for a relative diagnostic.
count_syllables <- function(word) {
  word <- tolower(gsub("[^a-z]", "", word))
  if (nchar(word) == 0) return(0L)
  groups <- gregexpr("[aeiouy]+", word)[[1]]
  n <- if (groups[1] == -1L) 1L else length(groups)
  if (grepl("e$", word) && n > 1L) n <- n - 1L   # silent terminal 'e'
  max(n, 1L)
}

gunning_fog <- function(text) {
  sentences <- unlist(strsplit(text, "(?<=[.!?])\\s+", perl = TRUE))
  sentences <- sentences[nzchar(trimws(sentences))]
  words <- unlist(strsplit(text, "\\s+"))
  words <- words[nzchar(words)]
  W <- length(words); S <- max(length(sentences), 1L)
  syl <- vapply(words, count_syllables, integer(1))
  C <- sum(syl >= 3L)                              # "complex" words
  0.4 * (W / S + 100 * C / W)
}

versions <- c(
  bloated  = paste("The utilization of an excessively elaborate and",
                   "circumlocutory prose style, characterized by the",
                   "accumulation of subordinate clauses, demonstrably",
                   "attenuates comprehensibility for the readership."),
  moderate = paste("Using an elaborate, indirect prose style with many",
                   "subordinate clauses reduces how well readers understand",
                   "the argument."),
  lean     = "Dense prose is hard to read."
)

fog <- vapply(versions, gunning_fog, numeric(1))
print(round(fog, 1))

barplot(fog, col = "grey80", border = NA, ylab = "Gunning fog (grade level)",
        main = "Readability across three drafts of one claim",
        names.arg = names(versions))
abline(h = 12, lty = 2)
```

### Revision as a process

First drafts are for the writer; revision is for the reader. The empirical
regularity behind all writing advice is that meaning is discovered, not transcribed:
the act of writing reveals which arrows in @eq-sw-chain are actually weak, and the
first draft's value is largely diagnostic. Productive revision proceeds top-down—
structure before paragraph, paragraph before sentence, sentence before word—because
fixing a comma in a paragraph that will be deleted is wasted effort. A useful
sequence is to first verify that the contribution sentence is present and true,
then that each section answers its one question, then that each paragraph has a
single point announced in its first sentence, and only then to polish prose. The
"reverse outline"—reading the draft and writing in the margin the single claim each
paragraph makes—exposes structural defects (two paragraphs making the same point, a
claim with no supporting paragraph, a paragraph with no claim) that are invisible at
the sentence level.

A final discipline is distance. Prose that seems clear the night it is written is
frequently opaque a week later, when the author has forgotten the unstated
assumptions that made it cohere—and a week-later author is a closer proxy for the
real reader than the night-of author ever is. Where time does not permit distance,
a co-author or colleague reading cold is the next best instrument, and the most
valuable feedback is not "I disagree" but "I got lost here", because confusion
localizes a broken arrow that the author can no longer see.

## Key Takeaways

- A paper is a **reconstructed argument**, not a lab diary: the order of exposition
  is engineered for the reader, and each section answers exactly one question with
  its own default verb tense (@tbl-sw-sections).
- The argument is a chain of claims the reader may refuse to grant
  (@eq-sw-chain); a paper fails at its **weakest arrow**, and most "low
  contribution" rejections are in fact broken arrows—a filled gap or a design that
  cannot license the conclusion.
- A contribution is the warranted increment to knowledge $\Delta\mathcal{K}$
  (@eq-sw-contribution); it must be **novel, valid, and important**, and the
  top-4 bar that rises is importance. State the most general claim the evidence
  licenses, and not one inch more.
- The **front end** carries the promise; the discussion must redeem exactly that
  promise. Make the "in this paper, we..." pivot early and state contributions as
  claims, not activities (@fig-sw-funnel).
- **Robustness is rhetoric**: writing the reviewer's objections into the paper
  before submission is the highest-return revision activity, and the R&R response is
  a craft document whose job is to make the editor's decision easy to defend.
- **Clarity is a property of the reader**: measure it (@eq-sw-fog), revise
  top-down from structure to sentence, and use distance or a cold reader to localize
  the arrows you can no longer see are broken.

## Further Reading

For the structure of scientific argument and the discipline of revision, the
classic writing-craft literature is the natural starting point; for the specific
norms of the marketing top-4, the editorials and "from the editor" notes of JM,
JMR, *Marketing Science*, and JCR are the authoritative—and frequently updated—
statements of what each venue rewards. <!-- CITE-NEEDED: editorial/from-the-editor pieces on contribution standards in JM, JMR, Marketing Science, JCR --> The
textual-analysis methods that let us measure the readability of our own and others'
prose are developed in the disclosure literature [@loughran2020; @loughran2020a]
and connect this chapter to the broader treatment of text as data elsewhere in the
book.
