# Reporting Research Results {#sec-reporting}

A research result that no one can read, reuse, or act on is not yet a result; it
is private knowledge. *Reporting* is the discipline that converts an estimate,
a test, or a model into an object that a reader can evaluate and a decision-maker
can use. This chapter treats reporting as a first-class methodological problem
rather than a clerical afterthought. The choices made here—which numbers reach a
table, how an effect is drawn, what is held back, how the whole pipeline can be
re-run—determine whether a finding survives scrutiny and whether it changes
behavior. Two audiences read marketing research and they read it differently: a
**scholarly** audience that wants identification, uncertainty, and the means to
replicate, and a **managerial** audience that wants a decision, its expected
payoff, and the risk of being wrong. A good report serves both without lying to
either.

The chapter proceeds from substance to surface. It first fixes the object of
communication—an estimate together with its uncertainty—and the principle that
governs all honest reporting: show the reader enough to second-guess you. It then
treats the two workhorses of static communication, **tables** and **figures**,
formally enough to say when each is the right instrument and what makes one
misleading. From there it builds the case for **reproducible reporting**, where
the document and the computation are one artifact, and gives runnable machinery
for it. It contrasts the **academic** and **managerial** registers, which differ
not in honesty but in what they foreground. It closes with **dashboards**—live,
interactive reporting—and the metric-governance problems they create, because a
number shown continuously to a manager is a number that will be optimized,
sometimes against the firm's interest.

## What Reporting Communicates

The atomic unit of a quantitative report is not a number but an **estimate with
its uncertainty**. Let $\theta$ be a population quantity of interest—an
elasticity, a treatment effect, a market share, a lift. An estimator
$\hat\theta$ computed from data is a random variable; a single report of
$\hat\theta$ alone is therefore incomplete, because it suppresses the sampling
distribution that tells the reader how seriously to take it. The minimal honest
report is the triple

$$
\bigl(\hat\theta,\ \widehat{\operatorname{se}}(\hat\theta),\ \text{the model and data that produced them}\bigr),
$$ {#eq-report-triple}

from which a reader can reconstruct confidence intervals, test statistics, and—
crucially—decide whether the precision is adequate for *their* decision, which is
rarely the author's decision. Reporting only $\hat\theta$ invites the reader to
treat noise as signal; reporting only a $p$-value invites them to treat
"significant" as "large," a confusion the field has spent two decades trying to
unlearn.[^pvalue] The governing principle is **sufficiency for second-guessing**:
a report should contain enough for a competent skeptic to redo the inference, vary
the assumptions, and reach their own verdict.

[^pvalue]: A $p$-value answers "how surprising is this data under the null?"—not
"how big is the effect?" and not "how probable is the hypothesis?" An effect can
be precisely estimated and economically trivial, or imprecisely estimated and
economically enormous. Reporting effect sizes with intervals, rather than stars,
keeps the magnitude in view. The replication and credibility debates in marketing
and the adjacent social sciences turn substantially on having confused the two.

This reframing has an immediate consequence for what belongs in a report.
Three quantities travel together and should be reported together: a **point
estimate** (the best single guess), a **measure of dispersion** (standard error,
posterior standard deviation, or interval), and a **specification** (the
estimator, the controls, the sample). A fourth—an **economic translation** of the
estimate into units the audience cares about (dollars, share points, conversions)
—is what separates a report that informs from one that merely records. The
managerial value of marketing research rests on making that translation
explicit and defensible [@hanssens2016].

> A statistical analysis, properly conducted, is a delicate dissection of
> uncertainties, a surgery of suppositions. The motto of the statistician should
> be *de omnibus dubitandum*—everything is to be doubted. Reporting is where that
> doubt is made visible to the reader rather than hidden from them.
>
> --- after the spirit of the exploratory-data-analysis tradition  [@tukey1977eda]

### The Two-Audience Problem, Formalized

Reporting is a communication channel, and like any channel it has a sender, a
message, and a receiver who decodes under noise (@sec-marketing-finance develops
the marketing–finance translation; the information-theoretic framing recurs in
the diffusion and word-of-mouth literatures). The sender (the researcher) holds a
posterior over $\theta$; the receiver (reader) holds a prior and a **loss
function** that prices the consequences of acting on a wrong belief. The two
canonical receivers price errors differently:

- The **scholarly receiver** is loss-averse over *false claims*. The cost of
  asserting an effect that is not there (a Type I error that enters the
  literature) is high and durable, because others will build on it. This receiver
  wants conservative inference, full disclosure of researcher degrees of freedom,
  and replicability—precisely so the channel does not transmit noise as if it were
  signal.
- The **managerial receiver** is loss-averse over *bad decisions*. The cost is the
  expected regret of choosing the wrong action under the estimate, integrated over
  the estimate's uncertainty. This receiver may rationally act on a noisy estimate
  if the downside of inaction is worse, and wants the estimate translated into the
  decision's own currency.

These are different loss functions, not different standards of honesty. The same
$(\hat\theta, \widehat{\operatorname{se}})$ feeds both; what changes is what is
**foregrounded** and what is **relegated to an appendix**. A report that confuses
the two—burying the decision under robustness tables for a manager, or
suppressing identification concerns to impress a manager—fails the audience it
was written for. The sophistication of marketing research has risen precisely as
the field learned to hold rigor and relevance simultaneously rather than trading
one for the other [@lehmann2011].

## Tables and Figures

Tables and figures are the two static encodings of quantitative results, and they
are not interchangeable. The operative distinction is between **lookup** and
**comparison**. A table is a lookup structure: it preserves exact values and lets
a reader retrieve a specific number ("what was the coefficient on price?"). A
figure is a comparison structure: it maps numbers to position, length, or color so
the reader's visual system can perceive *relations*—trends, gaps, interactions—
that a table forces them to compute by hand. The design rule follows directly:
**use a table when the reader needs values; use a figure when the reader needs to
see a pattern.**

### When a Table

A table is the right instrument when (i) exact magnitudes matter and will be
quoted or recomputed, (ii) the entries are heterogeneous in kind (estimates,
standard errors, fit statistics, sample sizes) and cannot share a single visual
scale, or (iii) the number of items is small enough that the eye can scan them.
Regression results are the paradigm case: a reader needs the coefficient, its
standard error, and the model's fit, to four meaningful digits, in a form they can
transcribe. The canonical estimation table reports, for each specification, the
coefficient vector $\hat{\boldsymbol\beta}$, dispersion beneath each estimate, and
goodness-of-fit at the foot, with specifications across columns so the reader can
trace how an estimate moves as controls are added—the single most informative
robustness display, because a coefficient that swings when a plausible control
enters is a coefficient that is not identified.

A table also imposes obligations. **Precision should be meaningful, not maximal**:
reporting eight digits of a quantity estimated to two is a form of false
confidence, and rounding to the precision the data support is itself an honesty
requirement. Units, sample size, and the estimation window belong in the caption
or notes, not in the reader's memory. And tables in a published artifact must be
generated by code, never typed, because a hand-transcribed table is a table with
an unverifiable error rate.

The example below fits two nested specifications and renders a publication-style
comparison table directly from the model objects, so the printed numbers are, by
construction, the numbers the model produced.

```{r table-setup, message=FALSE, warning=FALSE}
library(dplyr)
library(broom)
library(knitr)
library(kableExtra)
set.seed(43)

# Simulated brand-choice-style data: price, a loyalty cue, and a promotion flag.
n <- 600
promo   <- rbinom(n, 1, 0.4)
loyalty <- rnorm(n)
price   <- 3 + 0.5 * promo + rnorm(n, sd = 0.4)        # promos lower observed price below
# True data-generating process for log-sales:
log_sales <- 6 - 1.2 * price + 0.8 * loyalty + 0.5 * promo + rnorm(n, sd = 0.5)
dat <- data.frame(log_sales, price, loyalty, promo = factor(promo))

m1 <- lm(log_sales ~ price, data = dat)
m2 <- lm(log_sales ~ price + loyalty + promo, data = dat)
```

```{r table-render, message=FALSE, warning=FALSE}
fmt <- function(model, label) {
  # Keep cells plain text: the book renders to HTML, PDF, and EPUB from one
  # execution, so HTML-only markup (e.g., <br>) would leak into LaTeX.
  # "(Intercept)" is renamed because booktabs' \midrule treats a row that
  # begins with a literal "(" as a malformed trim spec, breaking the PDF.
  tidy(model) |>
    transmute(term = sub("^\\(Intercept\\)$", "Intercept", term),
              !!label := sprintf("%.3f (%.3f)", estimate, std.error))
}
tab <- full_join(fmt(m1, "Sparse"), fmt(m2, "Full"), by = "term")
gof <- tibble(term = "Adj. R-squared",
              Sparse = sprintf("%.3f", summary(m1)$adj.r.squared),
              Full   = sprintf("%.3f", summary(m2)$adj.r.squared))
bind_rows(tab, gof) |>
  kable(escape = FALSE, align = "lrr",
        caption = "Log-sales regressions. Standard errors in parentheses. Adding the loyalty cue and promotion control shifts the price coefficient, the first thing a careful reader checks.") |>
  kable_styling(full_width = FALSE)
```

Notice what the two-column layout buys: the price coefficient is visible
*before and after* controls enter, and its stability (or movement) is the
report's most honest robustness signal—far more informative than any single
starred number.

### When a Figure

A figure is the right instrument when the message is a *relationship*: a trend over
time, a dose–response curve, an interaction, a distribution, an uncertainty band.
The reason is perceptual. Position along a common scale is the visual encoding
humans judge most accurately; length and angle are next; area, color saturation,
and hue are progressively less accurate.[^cleveland] A figure that encodes its key
comparison as position on a shared axis lets the reader perceive the result;
one that encodes it as color or area forces them to estimate, badly. The design
corollary is that **the variable carrying the result should be encoded by
position**, and decoration that competes for that channel should be removed.

[^cleveland]: This ordering of *elementary perceptual tasks* by accuracy is the
empirical foundation of statistical graphics: encode the most important
comparison in the most accurately perceived channel.
@cleveland1984graphicalperceptio established the ranking experimentally: position along a common scale is judged most accurately, followed by position on identical non-aligned scales, then length, angle, area, and finally color saturation and volume. A
practical consequence is the suspicion of pie charts and 3-D bars: they encode
quantity as angle, area, or foreshortened length—exactly the channels the eye
reads worst.

Three failure modes recur and each has a name and a fix. A **truncated axis**—
starting a bar chart's value axis above zero—exaggerates differences by mapping a
small ratio onto a large length; bars must baseline at zero because their meaning
is length. **Overplotting**—thousands of points collapsing into an ink blob—hides
the distribution it purports to show; the fix is transparency, binning, or a
density encoding. **Uncertainty erasure**—drawing a point estimate as a clean line
with no band—is the graphical analogue of reporting $\hat\theta$ without
$\widehat{\operatorname{se}}(\hat\theta)$, and is arguably the most common
dishonesty in applied figures, because a clean line *feels* certain. The honest
default is to draw the interval, not just the estimate.

The figure below makes the last point operational. It plots the model-implied
relationship between price and log-sales together with a 95% confidence band, so
the reader sees both the estimated slope and how well-pinned it is.

```{r fig-uncertainty, message=FALSE, warning=FALSE}
#| fig-cap: "Model-implied price response with a 95% confidence band. The band, not the line, is the honest object: it shows the reader how precisely the slope is identified."
#| fig-width: 6
#| fig-height: 4
library(ggplot2)
grid <- data.frame(price = seq(min(dat$price), max(dat$price), length.out = 100),
                   loyalty = 0, promo = factor(0, levels = c(0, 1)))
pred <- predict(m2, newdata = grid, interval = "confidence")
grid <- cbind(grid, pred)

ggplot(grid, aes(price, fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
  geom_line(linewidth = 1) +
  geom_point(data = dat, aes(price, log_sales), alpha = 0.15, inherit.aes = FALSE) +
  labs(x = "Price", y = "Log-sales (predicted)") +
  theme_minimal(base_size = 12)
```

### Choosing Between Them

The choice is not stylistic but functional, and it can be stated as a short
decision rule. The table below contrasts the two instruments on the dimensions
that actually decide the call.

```{r tbl-table-vs-figure, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr); library(kableExtra)
data.frame(
  Dimension = c("Reader's task", "Encodes", "Best when", "Exact values",
                "Patterns/interactions", "Failure mode"),
  Table = c("Look up a value", "Symbols (digits)", "Few heterogeneous numbers",
            "Preserved", "Reader must compute", "False precision; hand-typed errors"),
  Figure = c("See a relationship", "Position, length, color",
             "Many values or a trend", "Approximate", "Perceived directly",
             "Truncated axis; hidden uncertainty"),
  check.names = FALSE
) |>
  kable(caption = "Tables versus figures: the choice follows from what the reader must do with the result.") |>
  kable_styling(full_width = FALSE)
```

A useful heuristic: if you find yourself asking the reader to *compare across rows
or columns* of a table, the comparison wants to be a figure; if you find yourself
*annotating a figure with exact numbers* the reader will quote, those numbers want
to be a table. Many reports need both—a figure for the pattern, a table in the
appendix for the values—and that redundancy is a feature, because it serves the
two audiences at once.

## Reproducible Reporting

The deepest reporting failure is not a bad table or a misleading axis; it is a
result no one can regenerate, including its author six months later. **Reproducible
reporting** is the practice of binding the *narrative* (prose, tables, figures) to
the *computation* (data, code, environment) in a single executable artifact, so
that the reported numbers are produced by running the document rather than pasted
into it. Formally, let a report be a function

$$
R = f(D, C, E),
$$ {#eq-repro}

where $D$ is the data, $C$ the analysis code, $E$ the computational environment
(language and package versions, seeds), and $f$ the build process. A report is
**reproducible** when re-evaluating $f$ on the same $(D, C, E)$ returns a
byte-identical $R$, and **robust** when small, defensible perturbations to $C$ or
$E$ leave the substantive conclusions intact. The two are distinct: reproducibility
is about the pipeline, robustness about the science. A result can be perfectly
reproducible and entirely fragile.

This framing dissolves a category of error that copy-and-paste workflows make
nearly inevitable: the **transcription gap**, where the number in the manuscript
no longer matches the number the code produces because the data updated, the
model changed, or a digit was mistyped. In a literate-programming workflow—the
model this book is written in, where prose and `{r}`/`{python}` chunks live in one
`.qmd` file and the build executes the code to insert results—the transcription
gap cannot open, because there is no transcription. Marketing's move into
data-rich, code-intensive analysis makes this non-negotiable: when the analysis is
a pipeline of joins, models, and simulations, the only credible report is one the
pipeline emits [@wedel2016].

@fig-repro-pipeline lays out the dependency structure such a pipeline must make
explicit and re-runnable.

```{mermaid}
%%| label: fig-repro-pipeline
%%| fig-cap: "A reproducible reporting pipeline. Each artifact depends only on declared inputs; re-running from raw data regenerates every downstream table and figure, closing the transcription gap."
flowchart LR
  A["Raw data<br/>(immutable)"] --> B["Cleaning / joins<br/>(scripted)"]
  B --> C["Analysis dataset<br/>(versioned)"]
  C --> D["Models & estimates<br/>(seeded)"]
  D --> E["Tables & figures<br/>(code-generated)"]
  E --> F["Rendered report<br/>(prose + results)"]
  G["Environment<br/>(versions, seeds)"] -.-> B
  G -.-> D
  style A fill:#e8e8e8,stroke:#333
  style F fill:#d8efd8,stroke:#333
```

### The Pillars of Reproducibility

Four practices make @eq-repro hold in practice, and each closes a specific leak.

**Determinism.** Every stochastic step is seeded (`set.seed()` in R, an explicit
RNG seed elsewhere), so that simulations, bootstraps, cross-validation splits, and
posterior draws return the same values on re-run. Without seeding, $f$ is not a
function and reproducibility is impossible by definition. Every stochastic example
in this book is seeded for exactly this reason.

**Immutable raw data and scripted transformation.** Raw data is never edited in
place; all cleaning is code that reads raw inputs and writes derived outputs, so
the path from source to analysis dataset is auditable and re-runnable. The
analysis dataset becomes a *computed artifact*, not a hand-curated one.

**Environment capture.** The language and package versions are recorded (a
lockfile, a container, or at minimum a `sessionInfo()` printed in the appendix),
because the same code can give different numbers under different versions. The
environment $E$ in @eq-repro is part of the report, not background.

**Single-source build.** The document and the computation are one file; rendering
*is* re-running. This is the practice that eliminates the transcription gap
outright rather than merely policing it.

The minimal, self-contained reproducibility footer below records seed and
environment so a reader can reconstruct $E$. It is deliberately boring; that is the
point.

```{r repro-footer}
set.seed(43)            # determinism: re-runs reproduce every stochastic result
cat("R version:", R.version.string, "\n")
cat("Key packages:\n")
for (p in c("dplyr", "ggplot2", "broom")) {
  if (requireNamespace(p, quietly = TRUE))
    cat(sprintf("  %-8s %s\n", p, as.character(packageVersion(p))))
}
```

A subtle but consequential point: reproducibility is a precondition for, not a
guarantee of, credibility. A reproducible analysis can still be **overfit to
researcher degrees of freedom**—the many defensible choices of sample, controls,
and functional form that, searched over, manufacture significance. The honest
defense is to report the **specification curve** or **multiverse**: the
distribution of the estimate across the reasonable analytic choices, so the reader
sees whether the result is a robust feature of the data or an artifact of one path
through it. Reproducibility makes the multiverse cheap to compute; reporting it is
what turns reproducibility into credibility.

## Academic versus Managerial Reporting

The same study, honestly reported, looks different to the two audiences because
they price errors differently (the formalization in @sec-reporting's two-audience
discussion above). The differences are systematic enough to tabulate, and the
tabulation is itself a reporting decision: it tells an author what to foreground.

```{r tbl-aud, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr); library(kableExtra)
data.frame(
  Element = c("Primary question", "What leads", "Uncertainty", "Identification",
              "Effect units", "Robustness", "Length", "Success criterion"),
  Academic = c("Is the effect real and why?", "Theory and contribution",
               "Intervals, tests, full disclosure", "Foregrounded and defended",
               "Standardized / elasticities", "Extensive, in-text",
               "Long; appendices expected", "Survives peer review & replication"),
  Managerial = c("What should we do, and what is the payoff?",
                 "The decision and its $ impact", "Risk of the recommended action",
                 "Stated as a caveat, not derived", "Dollars, share points, conversions",
                 "Summarized; details on request", "Short; one-page summary first",
                 "Changes a decision; ROI realized"),
  check.names = FALSE
) |>
  kable(caption = "Academic and managerial registers differ in foregrounding, not in honesty. The same estimate and its uncertainty feed both reports.") |>
  kable_styling(full_width = FALSE) |>
  column_spec(1, bold = TRUE)
```

The managerial register inverts the academic document's structure. An academic
paper builds to its contribution: motivation, theory, data, identification,
results, robustness, then implications. A managerial report **leads with the
answer**—the recommendation and its expected return—and relegates method to an
appendix, following the inverted-pyramid logic that the busiest reader should
extract the decision from the first paragraph. This is not dumbing down; it is
re-sequencing for a different loss function. The discipline of translating an
estimate into a decision with a quantified payoff is exactly what gives marketing
research standing in the firm, and its absence is why analytically sound work so
often fails to move resources [@hanssens2016]. The presence of a senior marketing
voice in the firm measurably improves performance, which raises the stakes of
reporting in a register that voice can use [@germann2015].

Three translation errors recur when crossing the academic–managerial boundary, and
each has a discipline-specific antidote.

First, **statistical significance is reported as managerial importance.** A
precisely estimated but tiny elasticity is "significant" and irrelevant; the
managerial report must state the effect in decision units and let the manager judge
materiality. Second, **a metric is reported without its construct validity.** Many
managerial dashboards rest on accounting approximations that do not measure what
they claim—Tobin's $q$ proxies and similar shortcuts can mislead precisely because
they are reported as if they were the construct of interest, when they are noisy,
biased stand-ins for it [@bendle2018; @bendle2018misuse]. Reporting the proxy
without flagging the gap transmits a measurement error as a managerial fact. Third,
**correlation is reported as a lever.** A managerial recommendation implicitly
claims that acting on a variable will move the outcome—a causal claim—so a report
that recommends action on the basis of a correlational estimate must say so plainly
and price the risk that the relationship is not causal. The honest managerial
report does not hide the identification problem; it states it as the risk attached
to the recommendation.

## Dashboards

A **dashboard** is reporting made continuous and interactive: a live interface that
renders current values of a fixed set of metrics, refreshed as data arrives, and
typically lets the user filter or drill down. Where a static report answers a
question once, a dashboard answers a standing question repeatedly, which changes
both its value and its hazards. Its value is **monitoring**—detecting when a metric
leaves its expected range fast enough to act. Its hazard is **optimization
pressure**: a number shown continuously to people whose performance is judged by it
will be moved, and not always by improving the underlying reality.

The design problem is to choose a small set of metrics that are *decision-relevant*,
*timely*, and *jointly hard to game*, and to present each with enough context that a
value is interpretable on sight. A bare number ("conversion: 3.2%") is nearly
useless; the same number against a target, a trend, and a confidence band is a
decision input. The reporting principle from @eq-report-triple survives the move to
real time: a live point estimate without a sense of its variability invites the user
to chase noise, mistaking the ordinary jitter of a ratio for a signal worth acting
on.

@fig-dashboard-anatomy sketches the layering that distinguishes a dashboard from a
wall of numbers.

```{mermaid}
%%| label: fig-dashboard-anatomy
%%| fig-cap: "Anatomy of a decision-oriented dashboard. Each tier narrows attention: from raw streams, to a few governed metrics, to the small set that triggers action when out of range."
flowchart TD
  S1["Data streams<br/>(sales, web, CRM, spend)"] --> M["Metric layer<br/>(defined, versioned KPIs)"]
  M --> C["Context layer<br/>(target, trend, uncertainty band)"]
  C --> A["Alert layer<br/>(out-of-range triggers)"]
  A --> D["Decision<br/>(act / investigate / ignore)"]
  M -.governance.-> G["Metric definitions<br/>owned & documented"]
  style D fill:#d8efd8,stroke:#333
  style G fill:#fff3cd,stroke:#333
```

### The Metric-Governance Problem

The defining risk of dashboards is captured by a regularity old enough to be
folklore and sharp enough to be a design constraint:

> When a measure becomes a target, it ceases to be a good measure.
>
> --- Goodhart's law, as commonly stated  [@strathern1997improving]

The mechanism is mundane. A metric is a *proxy* for an unobserved goal; agents
optimize the proxy; the gap between proxy and goal widens exactly where the
incentive bites. A dashboard that surfaces "leads generated" will produce more
leads, of falling quality; one that surfaces "average review rating" will produce
review-solicitation tactics that move the rating without moving satisfaction. The
defense is not to hide metrics but to **govern** them: pair every gameable metric
with a **guardrail metric** that moves in the opposite direction under gaming (leads
*and* lead-to-sale conversion; rating *and* response rate), so that gaming is
visible in the pairing. Reporting metrics in tension, not in isolation, is the
dashboard analogue of reporting an estimate with its uncertainty—both refuse to let
a single number stand unchallenged.

Three further governance practices keep a dashboard honest. **Definitions are owned
and versioned**: every metric has a single documented definition and a person
accountable for it, because the most common dashboard failure is two teams reading
the same label as two different quantities. **Latency is disclosed**: a metric that
is two days stale must say so, or it will be acted on as if current. And **the
metric set is small**: a dashboard that shows everything directs attention to
nothing, defeating the monitoring purpose that justified it. The discipline is the
same one that governs a good table—report what supports a decision and suppress what
merely fills space.

The simulation below makes the guardrail logic concrete. A team is incentivized on a
headline metric; over time they game it, lifting the headline while the guardrail
quietly deteriorates. A dashboard that showed only the headline would report
improvement; one that reports the pair reveals the truth.

```{r fig-goodhart, message=FALSE, warning=FALSE}
#| fig-cap: "Goodhart's law on a dashboard. Under incentive pressure the headline metric (leads) rises while the guardrail (lead-to-sale conversion) falls. Reporting the pair, not the headline alone, exposes the gaming."
#| fig-width: 6.2
#| fig-height: 4
set.seed(43)
weeks <- 1:26
gaming <- pmin((weeks - 1) / 25, 1)            # incentive pressure ramps in
leads  <- 100 + 60 * gaming + rnorm(26, 0, 4)  # headline climbs as it is gamed
conv   <- 0.22 - 0.12 * gaming + rnorm(26, 0, 0.01)  # quality erodes in step

dash <- data.frame(
  week  = rep(weeks, 2),
  value = c(leads / max(leads), conv / max(conv)),  # scaled to a common axis
  metric = rep(c("Leads (headline)", "Conversion (guardrail)"), each = 26)
)
library(ggplot2)
ggplot(dash, aes(week, value, linetype = metric)) +
  geom_line(linewidth = 1) +
  labs(x = "Week", y = "Scaled value", linetype = NULL,
       title = "What the headline hides") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### Static Report or Dashboard?

The choice between a one-time report and a standing dashboard is a choice about the
*question's tense*. A question asked once—"did this campaign lift sales?"—wants a
report: it permits depth, identification, and a considered narrative, and it is read
once and archived. A question asked continuously—"is the funnel healthy right now?"
—wants a dashboard: it trades depth for timeliness because its value is in early
detection. Confusing the two is a common and expensive error. A causal question
forced onto a dashboard becomes a wall of correlations that invites spurious action;
a monitoring question forced into a quarterly report arrives too late to matter. The
deeper analyses that justify *changing* what a dashboard monitors—why a metric moved,
whether an intervention worked—remain the province of the static, reproducible report
that this chapter began with, because they require the identification and uncertainty
machinery a glance-able interface cannot carry.

## Key Takeaways

- The unit of an honest report is an estimate *with its uncertainty and its
  specification* (@eq-report-triple), not a point estimate or a $p$-value alone; the
  governing principle is sufficiency for second-guessing.
- **Tables** serve lookup (exact, heterogeneous values); **figures** serve
  comparison (patterns via position). Encode the result in the most accurately
  perceived channel, baseline bars at zero, and never erase uncertainty.
- **Reproducible reporting** binds narrative to computation, $R = f(D, C, E)$
  (@eq-repro), eliminating the transcription gap; determinism, immutable data,
  environment capture, and single-source builds are its pillars. Reproducibility
  enables but does not guarantee credibility—report the multiverse.
- **Academic and managerial** reports differ in foregrounding, not honesty: the same
  uncertainty feeds both, but the managerial report leads with the decision and its
  payoff and translates effects into decision units [@hanssens2016].
- **Dashboards** make reporting continuous; their governing hazard is Goodhart's law.
  Pair gameable metrics with guardrails, govern definitions, disclose latency, and
  keep the set small. Use a dashboard for standing questions, a reproducible report
  for causal ones.

## Further Reading

The reproducibility and literate-programming practices in this chapter sit within
the broader move of marketing into data-rich, code-intensive analysis
[@wedel2016] and the long argument over how to demonstrate marketing's value to the
firm in a register management can act on [@hanssens2016; @germann2015]. The
measurement-validity cautions that dashboards make urgent connect to the
construct-and-scale material in @sec-measurement-scales and to documented misuses of
accounting-based proxies [@bendle2018; @bendle2018misuse]. The graphical-perception
and exploratory-data-analysis traditions that underwrite the figures section, and
the Goodhart/metric-governance literature behind the dashboard section, are flagged
above as citation gaps to be filled before publication.
