# Case Studies in Advertising {#sec-case-advertising}

The constructs developed in @sec-advertising—the persuasion chain, the two routes to
attitude change, wear-in and wear-out, and above all the *incremental effect above a
baseline*—are theoretical objects. An advertising case study is where those objects
meet a real budget, a real schedule, and a real measurement problem. This chapter does
not treat cases as decorative anecdotes about clever creative. It treats them as
**evidence**, and its central discipline is the question every empiricist must ask of a
campaign: *did the advertising cause the outcome, or did the firm advertise because it
already expected the outcome?* That question is not pedantic. It is the difference
between a celebrated campaign that paid and a celebrated campaign that merely coincided
with sales the firm would have made anyway.

Three cases anchor the chapter, chosen because each has been studied with unusual rigor
and each isolates a distinct lesson from @sec-advertising. The first is the long
program of **split-cable field experiments** run on consumer packaged goods, which
established—decades before the digital-experiments literature—both that advertising can
move sales and that its average effect is small, fragile, and easy to overstate
[@lodish1995]. The second is the **eBay paid-search experiment**, in which a firm that
believed search advertising was highly profitable discovered, under randomization, that
most of its measured return was selection rather than causation [@blake2015]. The third
is the **public-health advertising** case, in which the dependent variable is widened
from own-brand sales to societal behavior and a credible geographic baseline recovers a
large effect that own-sales elasticities would have missed [@ghosh2023societal]. Read
together, the three cases trace the same arc the parent chapter draws: advertising
works, most measured returns are illusory, and what separates the two is the quality of
the counterfactual.

By the end of this chapter the reader should be able to map any campaign onto the
formal apparatus of @sec-advertising; state the baseline a credible evaluation requires
and name the bias that contaminates the estimate when that baseline is missing; and
distinguish a campaign that *identified* an advertising effect from one that merely
*correlated* spending with sales. As with the branding cases of @sec-case-branding, the
organizing unit is the construct, not the firm: each case is read through one piece of
theory so that the theory is exercised and the case is disciplined.

## The Inferential Problem, Restated for Campaigns

@sec-advertising establishes that the naïve sales-on-advertising regression

$$
y_{it} = \beta\, a_{it} + \mathbf{x}_{it}^{\top}\boldsymbol{\gamma} + \varepsilon_{it}
$$ {#eq-case-naive}

recovers the causal effect $\beta$ only when $\mathbb{E}[\varepsilon_{it}\,a_{it}]=0$,
and that this condition fails whenever a firm sets advertising $a_{it}$ in anticipation
of a demand shock that enters $\varepsilon_{it}$. Because firms target favorable
conditions—their best markets, their peak seasons, their most promising launches—the
omitted shock loads positively onto $\hat\beta$, so observational case readings
*overstate* effectiveness [@Gordon_2017; @Lewis_2015]. A campaign that looks brilliant
in a before–after sales chart may simply have been scheduled into a quarter the firm
already knew would be strong.

The corrective concept, following @Gordon_2019, is the **incremental effect**: the lift
in behavior above the baseline the consumer would have exhibited absent the exposure.
Formally, with potential outcomes $Y_i(1)$ and $Y_i(0)$ under exposure and no exposure,
the target is the average treatment effect

$$
\tau = \mathbb{E}\!\left[\,Y_i(1) - Y_i(0)\,\right],
$$ {#eq-case-ate}

and the entire difficulty is that $Y_i(0)$—what would have happened without the ad—is
never observed for an exposed unit. A case study earns evidentiary value precisely to
the extent that it constructs a credible estimate of $Y_i(0)$: a randomized control
arm, a holdout of untreated markets, or a sharp discontinuity across an advertising
boundary. The three traps below recur across all three cases and are worth naming once.

| Trap | What it is | Antidote in a case |
|------|-----------|--------------------|
| Targeting endogeneity | Firm advertises where/when it expects to sell | Randomize exposure, or hold out comparable untreated units |
| Selection in attribution | Credited touchpoints are chosen by the consumer, not assigned | Compare to who would have converted anyway |
| Survivorship | Only the campaigns that "worked" become famous | Ask what comparable campaigns that failed did |

The arrangement of the chapter follows @fig-case-adv-framework: each case supplies
relatively clean variation in one identification strategy from @sec-advertising and
thereby teaches one construct with minimal confounding.

```{mermaid}
%%| label: fig-case-adv-framework
%%| fig-cap: "Mapping each advertising case (left) to the identification strategy it exemplifies (center) and the construct from the Advertising chapter it teaches most cleanly (right)."
flowchart LR
  C1[Split-cable CPG<br/>experiments] --> S1[Randomized<br/>field experiment]
  C2[eBay paid<br/>search] --> S2[Randomized<br/>geo experiment]
  C3[Public-health<br/>advertising] --> S3[Border /<br/>discontinuity design]
  S1 --> K1[Small, fragile<br/>elasticities & wear-out]
  S2 --> K2[Selection vs.<br/>causation in attribution]
  S3 --> K3[Incremental effect on a<br/>widened dependent variable]
```

## Case 1 — Split-Cable Experiments: How Big Is the Effect, Really?

The longest-running attempt to measure advertising's effect causally predates the
digital era. Through the 1980s and early 1990s, the IRI **BehaviorScan** system ran
**split-cable** field experiments: in test markets wired so that matched panels of
households could be served *different* television advertising while their purchases
were tracked through scanner panels, the advertiser's weight could be raised in a
randomly assigned treatment group and held fixed in a control group. This is, in the
language of @sec-advertising, a genuine randomized field experiment—the benchmark row
of @tbl-adv-identification—because exposure is assigned rather than chosen, so the
control panel furnishes a direct estimate of the unobserved $Y_i(0)$ in
@eq-case-ate. @lodish1995 analyze roughly 389 such tests and report the empirical
generalizations that have disciplined the field ever since.

> Increasing the *weight* of advertising alone, with no change in copy, media plan,
> or target, has no reliable effect on sales for established products; about half of
> the heavy-up tests show no significant sales effect at all. When advertising does
> work, it tends to be because *something other than weight*—new copy, a new media
> strategy, a new product, or a category-expansion message—has changed.

Two lessons from @sec-advertising fall directly out of this corpus. The first is that
**short-run advertising elasticities are small**, consistent with the meta-analytic
benchmark that the short-term advertising elasticity for established brands is on the
order of $0.1$ [@tellis1987], and with the broader finding that elasticities decline
over the product life cycle [@vakratsas1999]. An established brand's consumers already
know it; additional gross rating points clear no new stage of the persuasion chain of
@eq-mcguire-chain, because the presentation and attention stages were already cleared
in prior years. The second lesson is **wear-out**: where weight increases did move
sales, the effect was concentrated early and decayed with repetition, the field
analogue of the laboratory wear-out @batra1986 document when counterargument
production rises with repeated exposure. The managerial implication the @lodish1995
authors draw is therefore not "spend more" but "spend *differently*"—change the
message rather than the weight—precisely because weight operates on stages of the
chain that are already saturated for a mature brand.

The split-cable program also illustrates why famous before–after charts mislead. A
brand that ran a heavy fourth-quarter campaign and saw sales rise has not identified
$\tau$; it has measured $\mathbb{E}[Y_i(1)\mid \text{advertised in Q4}]$ against its
own pre-period, which mixes the ad effect with the seasonal shock the firm anticipated
when it chose Q4. The split-cable design defeats this by holding the *same* weeks fixed
across treatment and control panels in the *same* market, so the seasonal shock differences
out. The following simulation reproduces the core finding: a small true weight effect,
recoverable by the experiment but inflated by the naïve before–after read that targeting
contaminates.

```{r split-cable-sim, message=FALSE, warning=FALSE}
set.seed(46)

n_house   <- 4000                 # panel households
true_lift <- 0.03                 # true incremental effect of the heavy-up: +3%

# A seasonal demand shock the brand SEES and schedules its heavy-up into.
season <- rnorm(n_house, 0, 1)

# Naive "field" data: the brand heavies-up where the season looks strong, so
# exposure is correlated with the shock (targeting endogeneity).
heavy_obs <- as.integer(season + rnorm(n_house, 0, 0.5) > 0)
base_sales <- 10 + 2.0 * season
sales_obs  <- base_sales * (1 + true_lift * heavy_obs) + rnorm(n_house, 0, 1)

naive_lift <- (mean(sales_obs[heavy_obs == 1]) -
               mean(sales_obs[heavy_obs == 0])) / mean(base_sales)

# Split-cable: RANDOM assignment within the same market and weeks, so exposure is
# orthogonal to the season.
treat_rand <- rbinom(n_house, 1, 0.5)
sales_exp  <- base_sales * (1 + true_lift * treat_rand) + rnorm(n_house, 0, 1)
exp_lift   <- (mean(sales_exp[treat_rand == 1]) -
               mean(sales_exp[treat_rand == 0])) / mean(base_sales)

cat("True incremental lift      :", paste0(round(100 * true_lift, 1), "%"), "\n")
cat("Naive before/after read    :", paste0(round(100 * naive_lift, 1), "%"),
    " (inflated by targeting)\n")
cat("Split-cable experiment     :", paste0(round(100 * exp_lift, 1), "%"),
    " (recovers the truth)\n")
```

The naïve contrast overstates the effect because exposure correlates with the season
the brand targeted; randomization within the market severs that correlation and returns
the small true lift. This is the same mechanism the parent chapter simulates for the
geo-baseline case, here grounded in a real measurement system. The split-cable program's
enduring contribution is that it made the smallness of the average effect *credible*: it
came not from a skeptical regression but from hundreds of randomized tests run by
advertisers who expected to find large effects.

## Case 2 — eBay Paid Search: When the Return Is Mostly Selection

If split-cable taught that *weight* effects are small, the eBay paid-search experiment
taught the deeper lesson that an entire measured *return* can be an artifact of
**attribution selection**. @blake2015 study eBay's search-engine marketing across two
designs. For **branded keywords**—queries containing the word "eBay"—they exploit a
natural experiment created when eBay stopped bidding on its own brand terms on one
engine. For **non-branded keywords**, they run a large randomized geo experiment,
switching paid search off in a random subset of designated market areas and comparing
sales to the markets where it stayed on. Both designs construct $Y_i(0)$ honestly: the
turned-off markets and the post-shutdown period estimate what sales would have been
without the ads.

The result overturned the firm's prior. The branded-keyword shutdown produced **no
measurable drop in sales**: consumers who searched for "eBay" and would otherwise have
clicked the paid link simply clicked the adjacent organic link to the same site, so the
firm had been *paying for traffic it would have received for free*. The mechanism is
pure attribution selection—the consumer, not the advertiser, had already selected eBay
by typing its name, so the paid ad received conversion credit it did not cause. For
non-branded keywords the picture was more nuanced: returns were positive for new and
infrequent users but near zero or negative for the frequent users who dominate spend,
so the *average* return on the search budget was far below the firm's internal estimate
and, on the relevant margin, negative.

The case is a clinic in why @eq-case-naive is biased even when the regression is
sophisticated. The standard industry attribution model credits a sale to the last (or
some weighted set of) clicks that preceded it. But the clicks a consumer makes are
*selected by the consumer*, exactly the selection-in-attribution trap, so the
correlation between paid clicks and conversions confounds the consumers the ads
*caused* to convert with the far larger pool who would have converted regardless. The
randomized holdout is what separates the two, and it revealed that the second pool
dominated. The accounting is identical to the organic-versus-paid decomposition the
parent chapter formalizes for the fake-news case (@eq-fakenews): total demand is the
sum of organic and ad-referred components, and a credible evaluation must net out the
organic share that the advertising merely *intercepted* rather than created.

A small decision-theoretic calculation makes the managerial stakes concrete. Suppose a
fraction $\theta$ of paid-search conversions would have occurred organically anyway.
Then the true incremental cost per acquired customer is the reported cost-per-click
figure divided by $(1-\theta)$ of conversions that are genuinely incremental, and the
naïve return-on-ad-spend (ROAS) overstates the true ROAS by a factor of $1/(1-\theta)$.

```{r ebay-roas, message=FALSE, warning=FALSE}
set.seed(460)

clicks            <- 100000          # paid clicks purchased
conv_rate         <- 0.05            # conversions per click
margin_per_conv   <- 12              # gross margin per conversion ($)
cost_per_click    <- 0.40            # paid CPC ($)

conversions       <- clicks * conv_rate
ad_spend          <- clicks * cost_per_click
naive_roas        <- (conversions * margin_per_conv) / ad_spend

# theta: share of paid conversions that would have happened organically (the eBay
# branded-keyword lesson: for brand terms theta is close to 1).
theta_grid <- c(0.0, 0.3, 0.6, 0.9)
incremental_roas <- (conversions * (1 - theta_grid) * margin_per_conv) / ad_spend

roas_table <- data.frame(
  theta_organic    = theta_grid,
  naive_ROAS       = round(naive_roas, 2),
  incremental_ROAS = round(incremental_roas, 2),
  profitable       = ifelse(incremental_roas > 1, "yes", "no")
)
roas_table
```

When $\theta$ is large—as it is for branded keywords, where nearly every conversion
would have arrived organically—the incremental ROAS collapses below one and the spend
destroys value, even though the naïve ROAS looks handsomely positive. The eBay case is
the empirical demonstration that this is not a hypothetical: a sophisticated firm spent
for years on the high-$\theta$ margin because its attribution model could not see
$\theta$. The lesson generalizes far beyond search. Any channel evaluated by crediting
observed conversions, rather than by comparison to a randomized holdout, is exposed to
the same upward bias, and the bias is largest exactly where the channel reaches
consumers who were already going to buy.[^holdout]

[^holdout]: The methodological remedy is the *always-on holdout*: a small, randomly
withheld audience that never receives the channel, against which the treated audience
is continuously benchmarked. This operationalizes @eq-case-ate at the level of a live
marketing program and is the production-grade descendant of the split-cable design of
@lodish1995. The cost is a deliberately forgone slice of reach; the return is an
unbiased estimate of $\tau$ rather than a self-confirming attribution report.

## Case 3 — Advertising as Public Policy: Widening the Dependent Variable

The first two cases shrink the estimated effect of advertising by measuring it
honestly. The third shows the complementary move @sec-advertising recommends when
own-brand sales elasticities are small and noisy: **widen the dependent variable** and
find a setting where a credible baseline can be constructed for the broader outcome.
@ghosh2023societal study television advertising that carried COVID-19 narratives during
the early pandemic and ask whether brand-sponsored advertising changed *societal*
behavior—specifically, social distancing—rather than sales. The outcome is mobility
measured from mobile-device data, and the identification strategy is a **border design**
from @tbl-adv-identification: television-market (DMA) boundaries cut across county and
even state lines, so two adjacent counties can sit in different ad markets and receive
different advertising while sharing local conditions. Comparing mobility across that
discontinuity estimates the advertising effect while differencing out the local shocks
that confound a naïve comparison.

The estimated effect is large where one would least expect advertising to matter: TV
ads with COVID narratives raised social-distancing behavior, and the effect was about
**eleven times larger in counties without government stay-at-home mandates** than in
counties with them. The interaction is the substantively important result. Where public
policy already compelled the behavior, the marginal persuasive content of an ad added
little; where policy was absent, brand-sponsored messaging *substituted* for the missing
mandate. In the language of @sec-advertising, the advertising cleared the yielding and
behavior stages of the persuasion chain (@eq-mcguire-chain) for a population whose
behavior was not already determined by law, and added nothing for a population whose
behavior the law had already fixed.

The border design earns its credibility the way the split-cable and geo experiments
earn theirs: by constructing $Y_i(0)$. The untreated side of the DMA boundary estimates
what the treated side's mobility would have been absent the COVID advertising, and the
boundary's arbitrariness with respect to local conditions is the identifying assumption.
What would break it is the sorting flagged in @tbl-adv-identification—if counties on the
treated side differed systematically in some mobility-relevant way that also tracks the
DMA boundary—which is why the design restricts attention to narrow bands around the
border where such sorting is least plausible. The following simulation reproduces the
estimator and its central interaction.

```{r border-design-sim, message=FALSE, warning=FALSE}
set.seed(4600)

n_pairs   <- 500                  # matched cross-border county pairs
true_eff_nomandate <- 0.11        # ad effect on distancing where no mandate (+11pp)
true_eff_mandate   <- 0.01        # ad effect where a mandate already binds (+1pp)

# Each pair shares a local baseline; one side is in the "treated" DMA (ran COVID ads).
local_base <- rnorm(n_pairs, 0.50, 0.08)        # baseline distancing rate
mandate    <- rbinom(n_pairs, 1, 0.5)           # does a local mandate bind?

treated_side <- local_base +
  ifelse(mandate == 1, true_eff_mandate, true_eff_nomandate) +
  rnorm(n_pairs, 0, 0.03)
control_side <- local_base + rnorm(n_pairs, 0, 0.03)

# Border estimator: within-pair difference (treated DMA minus control DMA),
# split by whether a mandate binds.
within_diff <- treated_side - control_side
est_nomandate <- mean(within_diff[mandate == 0])
est_mandate   <- mean(within_diff[mandate == 1])

cat("Effect where NO mandate (truth 0.11):", round(est_nomandate, 3), "\n")
cat("Effect where mandate binds (truth 0.01):", round(est_mandate, 3), "\n")
cat("Ratio (no-mandate / mandate):",
    round(est_nomandate / est_mandate, 1), "x\n")
```

The within-pair difference recovers both effects and their roughly order-of-magnitude
ratio, illustrating how a discontinuity design isolates the advertising effect and the
condition that moderates it. The case completes the chapter's arc. Advertising's
own-sales effect is small and easily overstated (Case 1); much of its measured digital
return is selection that a holdout dissolves (Case 2); yet when the dependent variable
is widened to a behavior the firm genuinely moves and a sharp baseline is available, a
real and policy-relevant effect emerges (Case 3). In every instance the verdict turns
on the same thing: the credibility of $Y_i(0)$, not the fame of the campaign.

## Cross-Case Synthesis

Reading the three cases together produces a comparison that is more instructive than
any one of them. @tbl-case-adv-summary records, for each, the construct it teaches, the
strategy by which it builds a baseline, and the bias that would have contaminated the
estimate had that baseline been missing.

| Case | Construct taught | Baseline strategy | Bias absent the baseline |
|------|------------------|-------------------|--------------------------|
| Split-cable CPG [@lodish1995] | Small, decaying weight elasticities; wear-out [@tellis1987; @batra1986] | Randomized panels in-market | Targeting endogeneity inflates the effect |
| eBay paid search [@blake2015] | Selection vs. causation in attribution | Randomized geo / shutdown holdout | Organic traffic credited to paid ads |
| Public-health TV [@ghosh2023societal] | Incremental effect on a widened outcome | Cross-DMA border discontinuity | Local shocks confounded with exposure |

The single thread is @Gordon_2019's: an advertising effect is a credibly constructed
*increment above a baseline*, and the three cases differ only in how they build that
baseline—by randomizing within a market, by withholding a random set of markets, or by
exploiting an arbitrary advertising boundary. None relies on the sophistication of a
regression; all rely on the design of a counterfactual. The cases also rehearse the
**survivorship** caution of @sec-case-branding from the advertising side: campaigns that
*appear* to have worked are over-represented in trade lore precisely because apparent
success is what gets a campaign remembered, so a portfolio of celebrated campaigns is a
sample drawn on the outcome. The corrective is the same one the split-cable program
embodied at scale—measure against a control that was assembled *before* the result was
known.

A final note connects these cases back to the behavioral theory of @sec-advertising. The
smallness of weight effects for mature brands (Case 1) is exactly what the persuasion
chain of @eq-mcguire-chain predicts once a brand's consumers have already cleared its
early stages; the dominance of selection in branded search (Case 2) is what one expects
when the consumer has *already* yielded and is merely executing a known choice; and the
mandate interaction in the public-health case (Case 3) is the behavior stage made
visible—advertising moves behavior only where behavior is not already determined by
another force. The econometrics and the psychology are not rival accounts. The
identification design tells the analyst *how large* the effect is, and the behavioral
model tells her *why it is that large, and where*.

## Key Takeaways

- An advertising case is **evidence about an effect**, not proof that a campaign paid;
  its value is entirely in the credibility of the counterfactual $Y_i(0)$ it constructs
  (@eq-case-ate) [@Gordon_2019].
- Decades of randomized **split-cable** tests established that pure weight increases for
  established brands usually do not move sales, and that elasticities are small and
  decay with repetition—wear-out made visible at scale [@lodish1995; @tellis1987;
  @batra1986].
- Much of the measured return to **digital advertising is selection**: crediting
  observed conversions confounds customers the ad caused with customers who would have
  bought anyway, a bias that an experimental holdout dissolves [@blake2015].
- **Widening the dependent variable**—from own-brand sales to societal behavior—and
  exploiting a sharp geographic baseline can recover large, policy-relevant effects that
  own-sales elasticities miss [@ghosh2023societal].
- Across all three, the verdict turns on the **baseline, not the regression**; naïve
  before–after and last-click reads overstate effectiveness because firms and consumers
  both *select* exposure [@Gordon_2017; @Lewis_2015].
```