# Causal Inference and Field Experiments {#sec-causal-inference}

Marketing decisions are causal claims in disguise. "Spend another dollar on
search advertising and we will earn three" is a statement about a counterfactual
world that did not happen—the sales the firm *would* have realized had it not
spent the dollar. A correlation between advertising and sales, however precisely
estimated, does not license that statement, because the firm chose how much to
advertise, and it chose to advertise more precisely where and when it expected to
sell more. The gap between *what we measure* (an association in observational
data) and *what we want to know* (the effect of an intervention we control) is the
central problem of this chapter, and of empirical marketing more broadly.

This chapter develops the modern toolkit for closing that gap. It begins with the
**potential-outcomes framework**, the language in which a causal effect is defined
before any estimator is chosen, and shows why the naive comparison of treated and
untreated units is biased by **selection**. It then treats randomized experiments
and online **A/B tests** as the benchmark that solves selection by design, and the
four leading **quasi-experimental** designs—**difference-in-differences (DiD)**,
**instrumental variables (IV)**, **regression discontinuity (RD)**, and the
**synthetic control method (SCM)**—as strategies for recovering causal effects
when randomization is infeasible. Each method is presented the same way: the
intuition first, then the estimator, then the identifying assumptions, and—most
important—*what breaks identification* and how a reader can tell. Throughout, the
running applications are the two domains where causal inference has reshaped
marketing practice: **advertising measurement** and **pricing**.

The stakes are not academic. A decade of large-scale field experiments has shown
that observational and even quasi-experimental estimates of advertising's return
can be off not by a few percent but by an order of magnitude, and sometimes carry
the wrong sign [@Lewis_2015; @gordon2019]. A marketing scientist who cannot
distinguish a credible causal estimate from an incredible one is a liability to
the firm. By the end of this chapter the reader will be able to state the
estimand, choose a design, defend its assumptions, implement the estimator in R,
and diagnose the most common failures.

## Semester arc {#sec-causal-semester-arc}

This chapter is organized as a fourteen-week doctoral seminar in causal inference
and field experiments, tuned to quantitative-marketing applications—advertising
measurement, pricing, promotions, and digital/platform experimentation. The arc
matches how the topic is taught in the empirical track of top programs (MIT,
Chicago Booth, Stanford GSB, Berkeley, Wharton, Columbia, Kellogg, NYU Stern, and
the economics and statistics departments marketing students cross-register into).
The single organizing question is *what makes the treatment as-good-as-randomly
assigned, conditional on what we observe and how we designed the study?*

The **first third** establishes the framework (potential outcomes and the
assignment mechanism), the gold standard (randomized experiments and large-scale
A/B testing), and the realistic case where randomization is absent (selection on
observables, and why regression is not magic). The ordering is deliberate:
students internalize the experimental ideal before judging how far an
observational design falls short of it. The **middle third** is a tour of
quasi-experimental designs, each motivated by a different source of
as-good-as-random variation: instrumental variables, difference-in-differences,
regression discontinuity, synthetic control, panel/event-study methods, and
matching/weighting. A recurring theme is that the past decade *re-litigated the
workhorses*—most dramatically the two-way fixed-effects DiD estimator, long the
default, shown to be badly biased under staggered adoption and heterogeneous
effects. The seminar treats this episode as the field's central methodological
cautionary tale: a method can be standard, published thousands of times, and still
wrong. The **final third** turns to the machine-learning and marketing-platform
frontier: ML for heterogeneous treatment effects; advertising measurement when
effects are tiny relative to sales noise; the experimental designs platforms
actually use (ghost ads, PSA holdouts, geo experiments); incrementality, pricing,
and targeting-policy evaluation; and the interference/spillover problems that break
SUTVA in marketplaces and social networks. The course closes on the question every
marketing dissertation must answer: *given that the effect is real but small and
the platform is non-stationary, what is the credible research design, and can it
scale?*

The standing textbooks for the whole course are Imbens and Rubin's *Causal
Inference for Statistics, Social, and Biomedical Sciences* (Cambridge, 2015, the
potential-outcomes spine), Angrist and Pischke's *Mostly Harmless Econometrics*
(Princeton, 2009, the applied quasi-experimental workhorse), Cunningham's *Causal
Inference: The Mixtape* (Yale, 2021, a modern code-forward synthesis), and
Wooldridge's *Econometric Analysis of Cross Section and Panel Data*, 2nd ed. (MIT,
2010, the panel/fixed-effects reference). Each weekly module below names its
**topic**, **subtopics**, **methods**, **key readings** (with DOI links and a
one-line rationale; foundational works are marked **[F]**, frontier works **[R]**),
and **central debate**. The detailed derivations, R simulations, and worked
marketing examples that follow the weekly map—the potential-outcomes algebra, the
A/B-power simulation, the DiD/IV/RD/SCM code, and the design-comparison table—are
the seminar's lab sessions; they are kept intact and cross-referenced from the
weeks they support.

### Week 1 — Potential Outcomes and the Assignment Mechanism {#sec-causal-week01}

**Topic.** [F] The Neyman–Rubin potential-outcomes model as the language in which a
causal effect is defined before any estimator is chosen.

**Subtopics.** The fundamental problem of causal inference; SUTVA; the estimands
(ATE, ATT, CATE); the assignment mechanism as the central object; "no causation
without manipulation"; causal versus predictive questions.

**Methods.** Writing an estimand precisely; distinguishing
estimand / identification / estimation / inference; recognizing when a question is
causal versus descriptive. Worked algebra in
@sec-causal-potential-outcomes below.

**Key readings.**

- Rubin, D. B. (1974). "Estimating Causal Effects of Treatments in Randomized and
  Nonrandomized Studies." *Journal of Educational Psychology* 66(5), 688–701.
  [doi:10.1037/h0037350](https://doi.org/10.1037/h0037350) — **[F]** the founding
  statement of the potential-outcomes framework.
- Holland, P. W. (1986). "Statistics and Causal Inference." *Journal of the
  American Statistical Association* 81(396), 945–960.
  [doi:10.1080/01621459.1986.10478354](https://doi.org/10.1080/01621459.1986.10478354)
  — **[F]** formalizes the framework and the "no causation without manipulation"
  doctrine; the field's conceptual touchstone.
- Imbens, G. W. & Rubin, D. B. (2015). *Causal Inference for Statistics, Social,
  and Biomedical Sciences*, Part I. Cambridge University Press *(book)* — **[F]**
  potential outcomes and the assignment mechanism, in book length.

**Debate.** Is "manipulation" required for causality (what of immutable
characteristics)? ATE versus ATT versus policy-relevant estimands; design-based
versus model-based inference.

### Week 2 — Randomized Experiments and A/B Testing at Scale {#sec-causal-week02}

**Topic.** [F]→[R] Randomization as the assignment mechanism that identifies the
ATE, scaled up to industrial web experimentation.

**Subtopics.** Fisher's sharp null and randomization inference versus Neyman's
repeated-sampling inference; covariate balance, stratification, and blocking;
CUPED-style variance reduction; the engineering of web experimentation,
sample-ratio mismatch, peeking/optional stopping, and the experimentation platform
as an organizational artifact.

**Methods.** Running Fisherian permutation tests; computing Neyman variance;
designing stratified/clustered assignment; variance reduction; diagnosing
sample-ratio mismatch. Lab: the A/B-power simulation in
@sec-causal-rct below (@fig-ab-power).

**Key readings.**

- Kohavi, R., Longbotham, R., Sommerfield, D. & Henne, R. M. (2009). "Controlled
  Experiments on the Web: Survey and Practical Guide." *Data Mining and Knowledge
  Discovery* 18(1), 140–181.
  [doi:10.1007/s10618-008-0114-1](https://doi.org/10.1007/s10618-008-0114-1) —
  **[R]** the practitioner canon for industrial-scale A/B testing; the pitfalls
  students will actually hit.
- Bojinov, I. & Shephard, N. (2019). "Time Series Experiments and Causal
  Estimands: Exact Randomization Tests and Trading." *Journal of the American
  Statistical Association* 114(528), 1665–1682.
  [doi:10.1080/01621459.2018.1527225](https://doi.org/10.1080/01621459.2018.1527225)
  — **[R]** extends design-based randomization inference to single-unit,
  switchback-style time-series experiments (directly relevant to one-market
  pricing/ad tests).
- Imbens, G. W. & Rubin, D. B. (2015). *Causal Inference...*, Part II. Cambridge
  University Press *(book)* — **[F]** classical randomized experiments; Fisher and
  Neyman inference.

**Debate.** Fisher versus Neyman inference; fixed-horizon versus
sequential/always-valid testing; how much can web A/B testing actually learn given
non-stationarity and short windows?

### Week 3 — Regression and Selection on Observables {#sec-causal-week03}

**Topic.** [F] The conditional-independence (unconfoundedness) assumption and what
regression actually estimates under it.

**Subtopics.** Regression as a weighting/comparison estimator and what it estimates
under heterogeneity; "bad controls" and collider bias; the overlap/common-support
condition; omitted-variable bias and sensitivity.

**Methods.** Specifying a credible conditioning set; drawing causal DAGs to spot
bad controls and colliders; checking overlap; OVB sensitivity bounds. The
selection-bias decomposition is derived in
@sec-causal-naive-comparison below.

**Key readings.**

- Athey, S. & Imbens, G. W. (2017). "The State of Applied Econometrics: Causality
  and Policy Evaluation." *Journal of Economic Perspectives* 31(2), 3–32.
  [doi:10.1257/jep.31.2.3](https://doi.org/10.1257/jep.31.2.3) — **[R]** the modern
  map of identification strategies and where selection-on-observables sits among
  them; assigned as the course's strategic overview.
- Angrist, J. D. & Pischke, J.-S. (2009). *Mostly Harmless Econometrics*, ch. 3.
  Princeton University Press *(book)* — **[F]** regression and the
  conditional-independence assumption; the "bad control" problem.
- Imbens, G. W. & Rubin, D. B. (2015). *Causal Inference...*, Part III. Cambridge
  University Press *(book)* — **[F]** unconfoundedness and regression.

**Debate.** When is "control for everything" self-defeating? How much faith can
selection-on-observables ever earn without a design?

### Week 4 — Instrumental Variables {#sec-causal-week04}

**Topic.** [F] Recovering causal effects from an exogenous shifter of an endogenous
treatment.

**Subtopics.** The IV/2SLS estimator; the LATE theorem and the
monotonicity/exclusion/relevance assumptions; compliers and what LATE is the
effect *for*; weak instruments; judge/examiner and shift-share designs; IV in
marketing (price endogeneity, demand estimation).

**Methods.** 2SLS and its LATE interpretation; first-stage *F* / weak-IV
diagnostics; reasoning about exclusion and monotonicity in a real application. Lab:
the demand-estimation IV simulation in @sec-causal-iv below.

**Key readings.**

- Imbens, G. W. & Angrist, J. D. (1994). "Identification and Estimation of Local
  Average Treatment Effects." *Econometrica* 62(2), 467–475.
  [doi:10.2307/2951620](https://doi.org/10.2307/2951620) — **[F]** the LATE
  theorem; redefines what IV identifies under heterogeneity.
- Angrist, J. D., Imbens, G. W. & Rubin, D. B. (1996). "Identification of Causal
  Effects Using Instrumental Variables." *Journal of the American Statistical
  Association* 91(434), 444–455.
  [doi:10.1080/01621459.1996.10476902](https://doi.org/10.1080/01621459.1996.10476902)
  — **[F]** embeds IV in potential outcomes; the canonical compliers/defiers
  exposition.
- Angrist, J. D. & Pischke, J.-S. (2009). *Mostly Harmless Econometrics*, ch. 4.
  Princeton University Press *(book)* — **[F]** applied IV and weak-instrument
  diagnostics.

**Debate.** Is LATE a "policy-relevant" estimand or an artifact of the instrument?
Are most instruments' exclusion restrictions defensible?

### Week 5 — Difference-in-Differences I: Foundations and the 2×2 {#sec-causal-week05}

**Topic.** [F] The canonical 2×2 DiD and the parallel-trends assumption.

**Subtopics.** Event-study/pre-trend plots; clustered and serially correlated
standard errors; placebo and pre-trend tests; DiD for marketing launches, entry,
and policy/regulatory shocks.

**Methods.** Building event-study plots; testing parallel pre-trends; clustering
and wild-cluster bootstrap; placebo tests. Lab: the DiD panel simulation in
@sec-causal-did below.

**Key readings.**

- Bertrand, M., Duflo, E. & Mullainathan, S. (2004). "How Much Should We Trust
  Differences-in-Differences Estimates?" *The Quarterly Journal of Economics*
  119(1), 249–275.
  [doi:10.1162/003355304772839588](https://doi.org/10.1162/003355304772839588) —
  **[F]** exposes serial-correlation underestimation of DiD standard errors; the
  inference cautionary tale.
- Card, D. & Krueger, A. B. (2000). "Minimum Wages and Employment: A Case Study of
  the Fast-Food Industry in New Jersey and Pennsylvania: Reply." *American Economic
  Review* 90(5), 1397–1420.
  [doi:10.1257/aer.90.5.1397](https://doi.org/10.1257/aer.90.5.1397) — **[F]** the
  verified record for the canonical minimum-wage DiD program; the design that made
  DiD a default tool. (The original Card & Krueger 1994 *AER* article is *(DOI to
  verify)*.)
- Angrist, J. D. & Pischke, J.-S. (2009). *Mostly Harmless Econometrics*, ch. 5.
  Princeton University Press *(book)* — **[F]** DiD and fixed effects, the
  classical treatment.

**Debate.** Is parallel trends testable, or only falsifiable on pre-periods? How
much should inference (not point estimates) drive DiD credibility?

### Week 6 — Difference-in-Differences II: Staggered Adoption and the TWFE Critique {#sec-causal-week06}

**Topic.** [R] The field's central recent methodological episode: why two-way
fixed-effects DiD is biased under staggered timing and heterogeneous effects, and
the family of robust estimators that replaced it.

**Subtopics.** "Forbidden comparisons" and negative weights; decomposition
diagnostics; the new robust estimators (group-time ATTs, interaction-weighted event
studies, imputation); choosing among them; honest pre-trends.

**Methods.** Running Goodman-Bacon decompositions; estimating the
Callaway–Sant'Anna, Sun–Abraham, and Borusyak–Jaravel–Spiess estimators and
reconciling them; honest pre-trend (HonestDiD-style) sensitivity. The TWFE failure
is introduced in the DiD discussion at @sec-causal-did.

**Key readings.**

- Goodman-Bacon, A. (2021). "Difference-in-Differences with Variation in Treatment
  Timing." *Journal of Econometrics* 225(2), 254–277.
  [doi:10.1016/j.jeconom.2021.03.014](https://doi.org/10.1016/j.jeconom.2021.03.014)
  — **[R]** the decomposition showing TWFE is a weighted average of all 2×2s,
  including "forbidden" already-treated comparisons.
- de Chaisemartin, C. & d'Haultfœuille, X. (2020). "Two-Way Fixed Effects
  Estimators with Heterogeneous Treatment Effects." *American Economic Review*
  110(9), 2964–2996.
  [doi:10.1257/aer.20181169](https://doi.org/10.1257/aer.20181169) — **[R]** proves
  TWFE can put negative weights on treatment effects; offers a robust alternative.
- Callaway, B. & Sant'Anna, P. H. C. (2021). "Difference-in-Differences with
  Multiple Time Periods." *Journal of Econometrics* 225(2), 200–230.
  [doi:10.1016/j.jeconom.2020.12.001](https://doi.org/10.1016/j.jeconom.2020.12.001)
  — **[R]** the group-time ATT estimator that is now a default robust DiD.
- Sun, L. & Abraham, S. (2021). "Estimating Dynamic Treatment Effects in Event
  Studies with Heterogeneous Treatment Effects." *Journal of Econometrics* 225(2),
  175–199.
  [doi:10.1016/j.jeconom.2020.09.006](https://doi.org/10.1016/j.jeconom.2020.09.006)
  — **[R]** fixes contamination of event-study coefficients; the interaction-weighted
  estimator.
- Borusyak, K., Jaravel, X. & Spiess, J. (2024). "Revisiting Event-Study Designs:
  Robust and Efficient Estimation." *Review of Economic Studies* 92(5), 3253–3285.
  [doi:10.1093/restud/rdae007](https://doi.org/10.1093/restud/rdae007) — **[R]** the
  imputation (efficient) estimator and a unifying view of event-study designs.
- Roth, J., Sant'Anna, P. H. C., Bilinski, A. & Poe, J. (2023). "What's Trending in
  Difference-in-Differences? A Synthesis of the Recent Econometrics Literature."
  *Journal of Econometrics* 235(2), 2218–2244.
  [doi:10.1016/j.jeconom.2023.03.008](https://doi.org/10.1016/j.jeconom.2023.03.008)
  — **[R]** the synthesis/decision guide assigned as the week's anchor.

**Debate.** Which robust estimator, when? Is the efficient imputation estimator's
stronger modeling worth it? How much published DiD work is invalidated?

### Week 7 — Regression Discontinuity Designs {#sec-causal-week07}

**Topic.** [F]→[R] Identifying the local effect at a known assignment threshold,
with modern robust inference.

**Subtopics.** Sharp and fuzzy RDD; continuity-based versus local-randomization
views; bandwidth/kernel choice and the bias–variance tradeoff; manipulation/sorting
tests (density discontinuity); RDD in marketing (loyalty-tier thresholds, ad-auction
reserve prices, eligibility cutoffs).

**Methods.** Local-polynomial estimation; MSE-optimal and CER-optimal bandwidths;
density (McCrary-style) manipulation tests; placebo cutoffs. Lab: the loyalty-tier
RD simulation in @sec-causal-rd below (@fig-rd).

**Key readings.**

- Imbens, G. W. & Lemieux, T. (2008). "Regression Discontinuity Designs: A Guide to
  Practice." *Journal of Econometrics* 142(2), 615–635.
  [doi:10.1016/j.jeconom.2007.05.001](https://doi.org/10.1016/j.jeconom.2007.05.001)
  — **[F]** the practitioner's guide that standardized RDD practice.
- Calonico, S., Cattaneo, M. D. & Titiunik, R. (2014). "Robust Nonparametric
  Confidence Intervals for Regression-Discontinuity Designs." *Econometrica* 82(6),
  2295–2326. [doi:10.3982/ecta11757](https://doi.org/10.3982/ecta11757) — **[R]**
  robust bias-corrected inference and data-driven bandwidths; the modern default.

**Debate.** Continuity versus local-randomization framing; how local is "local,"
and external validity at the cutoff; robustness to bandwidth.

### Week 8 — Synthetic Control {#sec-causal-week08}

**Topic.** [F]→[R] Constructing a single treated unit's counterfactual from a
weighted average of donors.

**Subtopics.** The synthetic-control estimator as a data-driven weighted comparison;
placebo/permutation inference; pre-treatment fit and donor-pool selection;
extensions—generalized synthetic control (interactive fixed effects), synthetic
DiD, and Bayesian structural-time-series "CausalImpact" (geo/ad applications).

**Methods.** Constructing donor pools and weights; permutation/placebo inference;
choosing among SC / synthetic DiD / gsynth / CausalImpact for a marketing case.
Lab: the single-market SCM simulation in @sec-causal-scm below (@fig-scm).

**Key readings.**

- Abadie, A., Diamond, A. & Hainmueller, J. (2010). "Synthetic Control Methods for
  Comparative Case Studies: Estimating the Effect of California's Tobacco Control
  Program." *Journal of the American Statistical Association* 105(490), 493–505.
  [doi:10.1198/jasa.2009.ap08746](https://doi.org/10.1198/jasa.2009.ap08746) —
  **[F]** the canonical synthetic-control method and placebo inference.
- Abadie, A. (2021). "Using Synthetic Controls: Feasibility, Data Requirements, and
  Methodological Aspects." *Journal of Economic Literature* 59(2), 391–425.
  [doi:10.1257/jel.20191450](https://doi.org/10.1257/jel.20191450) — **[F]** the
  authoritative how-to/when-not-to review.
- Arkhangelsky, D., Athey, S., Hirshberg, D. A., Imbens, G. W. & Wager, S. (2021).
  "Synthetic Difference-in-Differences." *American Economic Review* 111(12),
  4088–4118. [doi:10.1257/aer.20190159](https://doi.org/10.1257/aer.20190159) —
  **[R]** unifies SC and DiD; the frontier estimator for panel comparisons.
- Xu, Y. (2017). "Generalized Synthetic Control Method: Causal Inference with
  Interactive Fixed Effects Models." *Political Analysis* 25(1), 57–76.
  [doi:10.1017/pan.2016.2](https://doi.org/10.1017/pan.2016.2) — **[R]** the
  SC ⇄ factor-model bridge for multiple treated units.
- Brodersen, K. H., Gallusser, F., Koehler, J., Remy, N. & Scott, S. L. (2015).
  "Inferring Causal Impact Using Bayesian Structural Time-Series Models." *The
  Annals of Applied Statistics* 9(1), 247–274.
  [doi:10.1214/14-aoas788](https://doi.org/10.1214/14-aoas788) — **[R]** the
  "CausalImpact" approach used widely for ad/geo impact at platforms.

**Debate.** Is placebo inference valid with few units? Convex-weight restriction
versus regression-based panel methods; overfitting pre-treatment fit.

### Week 9 — Panel and Event-Study Methods {#sec-causal-week09}

**Topic.** [F]→[R] The causal content of fixed-effects and event-study panel
estimators, and the design-based view of panel experiments.

**Subtopics.** Fixed-effects and first-difference estimators and their causal
content; event-study specification and dynamic treatment effects; the
finite-population, design-based view of panel/switchback experiments;
matrix-completion / factor approaches; serial-correlation-robust inference.

**Methods.** Specifying dynamic event studies correctly (cf. Week 6); design-based
versus model-based panel inference; matrix-completion estimation. Connects to the
TWFE specification at @sec-causal-did.

**Key readings.**

- Bojinov, I., Rambachan, A. & Shephard, N. (2021). "Panel Experiments and Dynamic
  Causal Effects: A Finite Population Perspective." *Quantitative Economics* 12(4),
  1171–1196. [doi:10.3982/qe1744](https://doi.org/10.3982/qe1744) — **[R]**
  design-based estimands for panel/switchback experiments; ties experiments to
  event studies (directly relevant to platform rollouts).
- Athey, S. & Imbens, G. W. (2017). "The State of Applied Econometrics: Causality
  and Policy Evaluation." *Journal of Economic Perspectives* 31(2), 3–32.
  [doi:10.1257/jep.31.2.3](https://doi.org/10.1257/jep.31.2.3) — **[R]** situates
  panel/matrix-completion approaches within the broader toolkit (read here for the
  panel sections; reprised from Week 3).
- Wooldridge, J. M. (2010). *Econometric Analysis of Cross Section and Panel Data*,
  2nd ed., panel chapters. MIT Press *(book)* — **[F]** the FE/FD reference
  treatment.

**Debate.** When is FE identifying a causal effect versus an artifact? Design-based
versus model-based panel inference; overlap with the staggered-DiD literature.

### Week 10 — Matching and Weighting {#sec-causal-week10}

**Topic.** [F]→[R] Balancing observables via the propensity score, up to
high-dimensional balancing.

**Subtopics.** The propensity score and its balancing property; matching,
subclassification, and inverse-propensity weighting; doubly-robust / AIPW
estimation; covariate-balancing and entropy/approximate-residual balancing in high
dimensions; observational ad-effect estimation and its failure modes.

**Methods.** Estimating and diagnosing propensity scores; assessing overlap and
balance; AIPW/doubly-robust estimation; balancing weights. Motivated by the
observational-versus-experimental ad-measurement evidence at
@sec-causal-ad-measurement.

**Key readings.**

- Rosenbaum, P. R. & Rubin, D. B. (1983). "The Central Role of the Propensity Score
  in Observational Studies for Causal Effects." *Biometrika* 70(1), 41–55.
  [doi:10.1093/biomet/70.1.41](https://doi.org/10.1093/biomet/70.1.41) — **[F]**
  defines the propensity score and the balancing theorem.
- Hirano, K., Imbens, G. W. & Ridder, G. (2003). "Efficient Estimation of Average
  Treatment Effects Using the Estimated Propensity Score." *Econometrica* 71(4),
  1161–1189.
  [doi:10.1111/1468-0262.00442](https://doi.org/10.1111/1468-0262.00442) — **[F]**
  efficient IPW; why the *estimated* score helps.
- Abadie, A. & Imbens, G. W. (2016). "Matching on the Estimated Propensity Score."
  *Econometrica* 84(2), 781–807.
  [doi:10.3982/ecta11293](https://doi.org/10.3982/ecta11293) — **[R]** correct
  inference for propensity-score matching (a long-standing error).
- Athey, S., Imbens, G. W. & Wager, S. (2018). "Approximate Residual Balancing:
  Debiased Inference of Average Treatment Effects in High Dimensions." *Journal of
  the Royal Statistical Society Series B* 80(4), 597–623.
  [doi:10.1111/rssb.12268](https://doi.org/10.1111/rssb.12268) — **[R]** frontier
  balancing for many covariates; bridges to ML (Week 11).

**Debate.** Matching versus weighting versus doubly-robust; can balancing ever
substitute for a design? The "propensity-score paradox" (over-trimming).

### Week 11 — Machine Learning for Heterogeneous Treatment Effects {#sec-causal-week11}

**Topic.** [R] Estimating the CATE and learning targeting policies with modern ML.

**Subtopics.** The CATE and policy learning; causal trees/forests and honest
splitting; double/debiased ML (Neyman-orthogonal moments, cross-fitting);
generalized random forests; targeting and best-linear-projection summaries of
heterogeneity.

**Methods.** Fitting causal forests / DML with cross-fitting; estimating CATEs and
the best-linear-projection of the CATE; constructing and validating targeting
policies. The CATE estimand is defined at @sec-causal-potential-outcomes.

**Key readings.**

- Athey, S. & Imbens, G. (2016). "Recursive Partitioning for Heterogeneous Causal
  Effects." *Proceedings of the National Academy of Sciences* 113(27), 7353–7360.
  [doi:10.1073/pnas.1510489113](https://doi.org/10.1073/pnas.1510489113) — **[F]**
  the "causal tree" with honest estimation; foundation of the HTE-by-ML program.
- Wager, S. & Athey, S. (2018). "Estimation and Inference of Heterogeneous
  Treatment Effects Using Random Forests." *Journal of the American Statistical
  Association* 113(523), 1228–1242.
  [doi:10.1080/01621459.2017.1319839](https://doi.org/10.1080/01621459.2017.1319839)
  — **[R]** causal forests with valid pointwise confidence intervals.
- Athey, S., Tibshirani, J. & Wager, S. (2019). "Generalized Random Forests." *The
  Annals of Statistics* 47(2), 1148–1178.
  [doi:10.1214/18-aos1709](https://doi.org/10.1214/18-aos1709) — **[R]** the general
  forest-as-local-estimator framework behind `grf`.
- Chernozhukov, V., Chetverikov, D., Demirer, M., Duflo, E., Hansen, C., Newey, W. &
  Robins, J. (2018). "Double/Debiased Machine Learning for Treatment and Structural
  Parameters." *The Econometrics Journal* 21(1), C1–C68.
  [doi:10.1111/ectj.12097](https://doi.org/10.1111/ectj.12097) — **[R]**
  orthogonalization plus cross-fitting; the dominant ML-for-causal-inference recipe.
- Athey, S. & Imbens, G. W. (2019). "Machine Learning Methods That Economists Should
  Know About." *Annual Review of Economics* 11, 685–725.
  [doi:10.1146/annurev-economics-080217-053433](https://doi.org/10.1146/annurev-economics-080217-053433)
  — **[R]** the survey assigned as the week's bridge from prediction to causal ML.

**Debate.** Inference after model selection; do flexible HTE estimates replicate?
Heterogeneity discovery versus confirmatory testing; orthogonality versus plug-in
bias.

### Week 12 — Platforms and Marketing I: Advertising Measurement {#sec-causal-week12}

**Topic.** [R] The marketing-specific core: measuring advertising effects when they
are tiny relative to sales variance.

**Subtopics.** The "unfavorable economics" of ad measurement; intent-to-treat
designs, PSA/holdout controls, and ghost ads; selection/endogeneity in ad delivery;
experiments versus observational methods for ad ROI; advertising-as-signal measured
experimentally.

**Methods.** Powering ad experiments and MDE calculations; ITT versus exposure
effects and the ghost-ad/PSA logic; benchmarking observational estimators against
RCTs. Worked motivation in @sec-causal-ad-measurement below.

**Key readings.**

- Lewis, R. A. & Rao, J. M. (2015). "The Unfavorable Economics of Measuring the
  Returns to Advertising." *The Quarterly Journal of Economics* 130(4), 1941–1973.
  [doi:10.1093/qje/qjv023](https://doi.org/10.1093/qje/qjv023) — **[F]** shows even
  huge experiments struggle to detect ad effects; sets the statistical-power agenda
  for the field [@Lewis_2015].
- Johnson, G. A., Lewis, R. A. & Nubbemeyer, E. I. (2017). "Ghost Ads: Improving the
  Economics of Measuring Online Ad Effectiveness." *Journal of Marketing Research*
  54(6), 867–884.
  [doi:10.1509/jmr.15.0297](https://doi.org/10.1509/jmr.15.0297) — **[R]** the
  ghost-ads design: a clean, low-cost counterfactual for would-be-exposed users
  [@johnson2017].
- Gordon, B. R., Zettelmeyer, F., Bhargava, N. & Chapsky, D. (2019). "A Comparison
  of Approaches to Advertising Measurement: Evidence from Big Field Experiments at
  Facebook." *Marketing Science* 38(2), 193–225.
  [doi:10.1287/mksc.2018.1135](https://doi.org/10.1287/mksc.2018.1135) — **[R]**
  observational methods badly mismeasure ad lift versus RCT benchmarks [@gordon2019].
- Gordon, B. R., Moakler, R. & Zettelmeyer, F. (2023). "Close Enough? A Large-Scale
  Exploration of Non-Experimental Approaches to Advertising Measurement." *Marketing
  Science* 42(4), 768–793.
  [doi:10.1287/mksc.2022.1413](https://doi.org/10.1287/mksc.2022.1413) — **[R]**
  when (rarely) observational ad measurement gets "close enough."
- Sahni, N. S., Zou, D. & Chintagunta, P. K. (2017). "Do Targeted Discount Offers
  Serve as Advertising? Evidence from 70 Field Experiments." *Management Science*
  63(8), 2688–2705.
  [doi:10.1287/mnsc.2016.2450](https://doi.org/10.1287/mnsc.2016.2450) — **[R]**
  promotions carry advertising-like (information) effects beyond redemption; bridges
  to Week 13.

**Debate.** Can advertising ROI ever be measured precisely enough to manage? Are
platform-run experiments trustworthy and replicable? Lift versus attribution.

### Week 13 — Platforms and Marketing II: Incrementality, Geo, and Pricing Experiments {#sec-causal-week13}

**Topic.** [R] Measuring incremental lift and evaluating targeting and pricing
policies from experiments.

**Subtopics.** Incrementality versus last-touch attribution; geo experiments and
matched-market/synthetic-control designs; pricing field experiments and price-ending
effects; coupon/promotion experiments; bandit "learning-and-earning" pricing;
targeting-policy *evaluation* from experiments.

**Methods.** Designing geo/matched-market experiments and CausalImpact-style
readouts (cf. Week 8); incrementality test design; off-policy/targeting-policy
evaluation; power for pricing/promo tests.

**Key readings.**

- Blake, T., Nosko, C. & Tadelis, S. (2015). "Consumer Heterogeneity and Paid Search
  Effectiveness: A Large-Scale Field Experiment." *Econometrica* 83(1), 155–174.
  [doi:10.3982/ecta12423](https://doi.org/10.3982/ecta12423) — **[R]** eBay's
  brand-keyword experiment: paid search was largely non-incremental for known
  customers; the incrementality cautionary tale [@blake2015].
- Anderson, E. T. & Simester, D. I. (2003). "Effects of \$9 Price Endings on Retail
  Sales: Evidence from Field Experiments." *Quantitative Marketing and Economics*
  1(1), 93–110.
  [doi:10.1023/a:1023581927405](https://doi.org/10.1023/a:1023581927405) — **[F]**
  a clean pricing field experiment isolating a behavioral price-ending effect.
- Simester, D., Timoshenko, A. & Zoumpoulis, S. I. (2020). "Efficiently Evaluating
  Targeting Policies: Improving on Champion vs. Challenger Experiments." *Management
  Science* 66(8), 3412–3424.
  [doi:10.1287/mnsc.2019.3379](https://doi.org/10.1287/mnsc.2019.3379) — **[R]** how
  to evaluate many targeting policies from a single experiment.
- Simester, D., Timoshenko, A. & Zoumpoulis, S. I. (2020). "Targeting Prospective
  Customers: Robustness of Machine-Learning Methods to Typical Data Challenges."
  *Management Science* 66(6), 2495–2522.
  [doi:10.1287/mnsc.2019.3308](https://doi.org/10.1287/mnsc.2019.3308) — **[R]** how
  ML targeting from experimental data holds up under real data pathologies.

**Debate.** Geo experiments (few large units, interference) versus user-level
randomization; attribution versus incrementality in industry practice; can targeting
policies be both learned and validated on the same data?

### Week 14 — Interference, Spillovers, and SUTVA Violations {#sec-causal-week14}

**Topic.** [R] Synthesis and research craft: causal inference when one unit's
treatment affects another.

**Subtopics.** SUTVA violations in marketplaces and social networks;
partial-interference and exposure-mapping estimands; cluster/graph-cluster
randomization to bound bias; the reflection problem and identifying peer/social
effects; marketplace equilibrium effects (treatment "cannibalizes" control);
research craft—feasible designs under interference, pre-analysis plans, refereeing.

**Methods.** Specifying exposure mappings; cluster/graph-cluster randomization
design; partial-interference estimands; writing a referee report and a pre-analysis
plan. SUTVA is introduced at @sec-causal-potential-outcomes and revisited in
@sec-causal-threats.

**Key readings.**

- Aronow, P. M. & Samii, C. (2017). "Estimating Average Causal Effects Under General
  Interference, with Application to a Social Network Experiment." *The Annals of
  Applied Statistics* 11(4), 1912–1947.
  [doi:10.1214/16-aoas1005](https://doi.org/10.1214/16-aoas1005) — **[R]**
  design-based estimation under interference via exposure mappings.
- Sävje, F., Aronow, P. M. & Hudgens, M. G. (2021). "Average Treatment Effects in
  the Presence of Unknown Interference." *The Annals of Statistics* 49(2), 673–701.
  [doi:10.1214/20-aos1973](https://doi.org/10.1214/20-aos1973) — **[R]** what
  experiments still estimate when interference is present but unmodeled.
- Eckles, D., Karrer, B. & Ugander, J. (2017). "Design and Analysis of Experiments
  in Networks: Reducing Bias from Interference." *Journal of Causal Inference* 5(1),
  20150021.
  [doi:10.1515/jci-2015-0021](https://doi.org/10.1515/jci-2015-0021) — **[R]**
  graph-cluster randomization to cut interference bias in social/ad experiments.
- Blake, T. & Coey, D. (2014). "Why Marketplace Experimentation Is Harder Than It
  Seems: The Role of Test-Control Interference." *Proceedings of the Fifteenth ACM
  Conference on Economics and Computation (EC '14)*, 567–582.
  [doi:10.1145/2600057.2602837](https://doi.org/10.1145/2600057.2602837) — **[R]**
  marketplace interference biases naive A/B tests (eBay); the applied motivation.
- Manski, C. F. (1993). "Identification of Endogenous Social Effects: The Reflection
  Problem." *The Review of Economic Studies* 60(3), 531–542.
  [doi:10.2307/2298123](https://doi.org/10.2307/2298123) — **[F]** the foundational
  impossibility result for separating endogenous from contextual peer effects.

**Debate.** When does interference bias dominate, and is cluster randomization worth
the power loss? Can peer effects ever be point-identified? How should marketing
referee non-replicable platform experiments?

## The Potential-Outcomes Framework {#sec-causal-potential-outcomes}

The foundational move of modern causal inference is to define the effect of a
treatment *for each unit* before worrying about how to estimate it. Following the
Neyman–Rubin potential-outcomes model, let $D_i \in \{0,1\}$ denote whether unit
$i$ (a consumer, a store, a geographic market, a designated market area) receives
a treatment—an ad campaign, a price cut, a new feature. Associate with each unit
two **potential outcomes**: $Y_i(1)$, the outcome that *would* occur under
treatment, and $Y_i(0)$, the outcome that *would* occur under control. The
**unit-level treatment effect** is their difference,

$$
\tau_i = Y_i(1) - Y_i(0).
$$ {#eq-unit-effect}

The defining difficulty—the *fundamental problem of causal inference*—is that for
any unit we observe only one potential outcome. The realized outcome is

$$
Y_i = D_i\,Y_i(1) + (1 - D_i)\,Y_i(0),
$$ {#eq-switching}

so the other potential outcome is a **counterfactual** that is missing by
construction. We never see $\tau_i$ for any single unit. Causal inference is
therefore, at root, a **missing-data problem**, and progress comes from estimating
*averages* of $\tau_i$ over populations rather than the unit-level effects
themselves.

> A causal effect is a comparison of two states of the world that differ only in
> the treatment: the outcome we observe under the treatment a unit received, and
> the outcome that same unit would have realized under the treatment it did not
> receive. Because only one of these is ever observed, every causal estimate rests
> on an assumption that makes the missing outcome recoverable from the data.

The chief estimand is the **average treatment effect (ATE)**,
$\tau_{\text{ATE}} = \mathbb{E}[Y_i(1) - Y_i(0)]$, the expected effect of moving a
randomly chosen unit from control to treatment. Two relatives matter in practice.
The **average treatment effect on the treated (ATT)**,
$\tau_{\text{ATT}} = \mathbb{E}[Y_i(1) - Y_i(0) \mid D_i = 1]$, is the effect for
the units that actually received treatment—the relevant quantity when asking
whether a campaign *that ran* paid off. The **conditional average treatment effect
(CATE)**, $\tau(\mathbf{x}) = \mathbb{E}[Y_i(1) - Y_i(0) \mid \mathbf{X}_i =
\mathbf{x}]$, is the effect for a subpopulation with covariates $\mathbf{x}$, the
target of modern heterogeneous-treatment-effect and targeting
methods.[^estimand-first]

[^estimand-first]: A discipline this chapter enforces: name the estimand *before*
choosing a method. Much confusion in applied marketing arises from estimating one
quantity (say, an ATT among self-selected adopters) and interpreting it as
another (the ATE the firm would obtain by treating everyone). The two coincide
only under constant effects.

### Why the Naive Comparison Fails {#sec-causal-naive-comparison}

The intuitive estimator of a treatment effect is the difference in mean outcomes
between treated and untreated units, $\hat\Delta = \bar Y_{D=1} - \bar Y_{D=0}$.
Decomposing its probability limit exposes exactly why it is biased. Adding and
subtracting $\mathbb{E}[Y_i(0) \mid D_i = 1]$ gives

$$
\underbrace{\mathbb{E}[Y_i \mid D_i = 1] - \mathbb{E}[Y_i \mid D_i = 0]}_{\text{naive difference}}
= \underbrace{\tau_{\text{ATT}}}_{\text{causal effect}}
+ \underbrace{\mathbb{E}[Y_i(0) \mid D_i = 1] - \mathbb{E}[Y_i(0) \mid D_i = 0]}_{\text{selection bias}}.
$$ {#eq-selection-bias}

The second term is **selection bias**: the difference in the *baseline* (untreated)
outcome between the groups that did and did not get treated. It is zero only if
treated and untreated units would have looked the same absent treatment. In
marketing they almost never do. A firm targets ads at consumers already inclined
to buy; it cuts prices on products whose demand is already softening; loyal
customers self-select into a rewards program. In each case
$\mathbb{E}[Y_i(0) \mid D_i = 1] \neq \mathbb{E}[Y_i(0) \mid D_i = 0]$, and the
naive difference confounds the effect of the treatment with the reasons the unit
was treated. The remainder of this chapter is a catalogue of designs that drive
the selection-bias term to zero—by randomization, or by an assumption that makes
the untreated a valid stand-in for the treated counterfactual.

Two assumptions are maintained throughout unless stated otherwise. **SUTVA** (the
stable-unit-treatment-value assumption) requires that one unit's potential outcomes
not depend on other units' treatment assignments (no interference) and that there
be a single version of the treatment. SUTVA is violated by spillovers—a price
promotion at one store cannibalizing a neighbor, an ad shown to one household
influencing another through word of mouth—which is why experimental units in
marketing are often *geographies* rather than individuals. **Overlap** (positivity)
requires $0 < \Pr(D_i = 1 \mid \mathbf{X}_i) < 1$ for all covariate values, so that
comparable treated and control units exist at every $\mathbf{x}$.

@fig-ladder situates the designs of this chapter on a spectrum from the strongest
identification (randomization, where assumptions hold by construction) to designs
that buy identification with progressively stronger, untestable assumptions.

```{mermaid}
%%| label: fig-ladder
%%| fig-cap: "A hierarchy of causal designs. Moving down the ladder relaxes the requirement of randomized assignment but substitutes increasingly strong, largely untestable identifying assumptions. The estimand each design targets is named in brackets."
flowchart TD
    A["<b>Randomized experiment / A-B test</b><br/>assignment independent of potential outcomes<br/>[ATE, by design]"]
    B["<b>Regression discontinuity</b><br/>treatment jumps at a known cutoff<br/>[local ATE at the cutoff]"]
    C["<b>Instrumental variables</b><br/>an exogenous shifter of treatment<br/>[LATE: effect on compliers]"]
    D["<b>Difference-in-differences</b><br/>parallel trends absent treatment<br/>[ATT]"]
    E["<b>Synthetic control</b><br/>weighted donors reproduce the counterfactual<br/>[ATT for one treated unit]"]
    A --> B --> C --> D --> E
    style A fill:#1b7837,color:#fff
    style B fill:#5aae61,color:#fff
    style C fill:#a6dba0,color:#000
    style D fill:#d9f0d3,color:#000
    style E fill:#f7f7f7,color:#000
```

## Randomized Experiments and A/B Tests {#sec-causal-rct}

Randomization is the gold standard because it eliminates selection bias *by
construction* rather than by assumption. If treatment $D_i$ is assigned
independently of the potential outcomes—$\{Y_i(0), Y_i(1)\} \perp D_i$—then the
untreated group is, in expectation, a valid counterfactual for the treated group:
$\mathbb{E}[Y_i(0) \mid D_i = 1] = \mathbb{E}[Y_i(0) \mid D_i = 0]$. The
selection-bias term in @eq-selection-bias vanishes, and the difference in means is
unbiased for the ATE,

$$
\hat\tau_{\text{ATE}} = \bar Y_{D=1} - \bar Y_{D=0}.
$$ {#eq-rct-estimator}

The identifying assumption—independence of assignment and potential
outcomes—holds because the experimenter, not the units, controls $D_i$. This is
the single most important idea in the chapter: randomization does not make treated
and control units *identical*, it makes them *exchangeable in expectation*, which
is all that @eq-rct-estimator requires.

In the digital economy this logic is operationalized at enormous scale as the
**A/B test** (online controlled experiment): users arriving at a site are
randomly bucketed into a control arm seeing the incumbent experience and one or
more treatment arms seeing a variant, and a metric—conversion, revenue per
session, retention—is compared across arms. The mechanics are simple; the
discipline is in the design. Three issues dominate practice.

**Power and the minimum detectable effect.** An experiment that cannot detect the
effect sizes that matter is worse than no experiment, because it invites the
fallacy of accepting the null. For a two-arm test of a mean with equal allocation,
the sample size per arm needed to detect a true difference $\delta$ at
significance level $\alpha$ and power $1-\beta$ is approximately

$$
n \;\approx\; \frac{2\,(z_{1-\alpha/2} + z_{1-\beta})^2\,\sigma^2}{\delta^2},
$$ {#eq-power}

where $\sigma^2$ is the outcome variance and $z_p$ the standard-normal quantile.
The quadratic dependence on $\delta$ is the crux of advertising measurement: sales
are so volatile relative to advertising's effect that the required sample is often
larger than even large firms can muster, a point developed below.

**The randomization unit and interference.** When treating a user can affect other
users—social spillovers, marketplace competition for the same inventory, supply
constraints—unit-level randomization violates SUTVA and biases the estimate.
Cluster-randomizing at the level of the interference (markets, regions, social
clusters) restores validity at the cost of effective sample size.

**Peeking and the validity of the stopping rule.** Repeatedly testing a running
experiment and stopping the first time $p < 0.05$ inflates the false-positive rate
far above the nominal level. Either fix the sample size in advance, or adopt a
sequential procedure (always-valid $p$-values, group-sequential boundaries) whose
error guarantee survives continuous monitoring.

The following simulation makes the bias-then-fix logic concrete: it first shows
that a confounded observational comparison recovers the wrong effect, then that
randomization recovers the truth.

```{r ab-test-sim}
#| echo: true
#| message: false
#| warning: false
#| label: fig-ab-power
#| fig-cap: "Sampling distribution of the difference-in-means estimator across 2,000 randomized experiments, each with 1,000 users per arm and a true lift of 0.20. The estimator is centered on the truth (dashed line); the naive observational estimator from the same data-generating process is biased upward."
set.seed(47)

# Data-generating process: a latent 'propensity' drives BOTH self-selection
# into treatment (observationally) AND the baseline outcome -> confounding.
make_population <- function(n, true_tau = 0.20) {
  propensity <- rnorm(n)                       # latent buying inclination
  Y0 <- 0.50 + 0.40 * propensity + rnorm(n)    # untreated potential outcome
  Y1 <- Y0 + true_tau                          # constant treatment effect
  data.frame(propensity, Y0, Y1)
}

# (1) OBSERVATIONAL: high-propensity users self-select into treatment.
n <- 4000
pop <- make_population(n)
D_obs <- as.integer(pop$propensity + rnorm(n) > 0)   # selection on propensity
Y_obs <- ifelse(D_obs == 1, pop$Y1, pop$Y0)
naive <- mean(Y_obs[D_obs == 1]) - mean(Y_obs[D_obs == 0])

# (2) RANDOMIZED: assignment independent of potential outcomes.
rct_estimate <- function(n_per_arm = 1000, true_tau = 0.20) {
  p <- make_population(2 * n_per_arm, true_tau)
  D <- rep(0:1, each = n_per_arm)
  D <- sample(D)                               # random assignment
  Y <- ifelse(D == 1, p$Y1, p$Y0)
  mean(Y[D == 1]) - mean(Y[D == 0])
}
rct_draws <- replicate(2000, rct_estimate())

cat(sprintf("True effect (ATE):              %.3f\n", 0.20))
cat(sprintf("Naive observational estimate:   %.3f  (biased)\n", naive))
cat(sprintf("Mean of randomized estimates:   %.3f  (unbiased)\n", mean(rct_draws)))

hist(rct_draws, breaks = 40, col = "grey80", border = "white",
     main = "", xlab = "Estimated treatment effect")
abline(v = 0.20, lwd = 2, lty = 2)
abline(v = naive, lwd = 2, col = "firebrick")
legend("topright", bty = "n",
       legend = c("true effect (0.20)", "naive observational"),
       lwd = 2, lty = c(2, 1), col = c("black", "firebrick"))
```

### Advertising Measurement: The Hard Case for Experiments {#sec-causal-ad-measurement}

Advertising is the canonical setting where experiments are both essential and
difficult. The difficulty is statistical, and @eq-power explains it. The effect of
a typical online display campaign on purchase probability is small in absolute
terms, while individual purchasing is enormously variable; substituting realistic
magnitudes into @eq-power implies that detecting advertising's true return on
investment with confidence can require *millions* of randomized users—more than
many campaigns ever reach [@Lewis_2015]. Lewis and Rao show that even well-powered
experiments deliver confidence intervals on ROI so wide that firms cannot
distinguish a profitable campaign from a money-losing one, and that the
observational regressions firms typically run fare far worse.

How much worse is documented by large-scale comparisons. Across digital
advertising experiments at scale, observational methods—matching, regression
adjustment, propensity scores—routinely yield estimates that diverge sharply from
the experimental benchmark, frequently overstating effects by large multiples and
occasionally reversing their sign, because no set of observed covariates fully
captures why a platform showed a given user a given ad [@gordon2019]. The lesson
is not that observational methods are useless but that, in advertising, the
selection-bias term in @eq-selection-bias is large relative to the effect itself,
so even small failures of the conditional-independence assumption swamp the signal.
A recurrent culprit is **activity bias**: users who are online and active are both
more likely to be served (and to "see") an ad and more likely to convert for
reasons unrelated to the ad, manufacturing a spurious correlation that
intention-to-treat experimental designs avoid [@lewis2011].

These findings reframe the practitioner's task. The question is rarely "what is
the point estimate of ROI?" but "is my experiment powered to answer the decision I
face?"—and often the honest answer is to redesign the test (larger, geo-based, or
focused on an intermediate outcome) rather than to trust an underpowered or
observational number.

## Difference-in-Differences {#sec-causal-did}

When randomization is impossible—a competitor enters a market, a regulator changes
a privacy rule, a firm rolls a price change out to some regions before others—the
analyst exploits *timing*. **Difference-in-differences** compares the change in
outcomes for a treated group, before versus after the intervention, against the
contemporaneous change for an untreated control group. Differencing over time
removes any fixed differences between the groups; differencing across groups
removes any common time trend. What remains, under one key assumption, is the
treatment effect.

Formally, with two periods (pre, post) and two groups (treated $T$, control $C$),
the DiD estimator is the difference of differences in group means,

$$
\hat\tau_{\text{DiD}}
= \big(\bar Y_{T,\text{post}} - \bar Y_{T,\text{pre}}\big)
- \big(\bar Y_{C,\text{post}} - \bar Y_{C,\text{pre}}\big).
$$ {#eq-47-did}

In panel-regression form, with unit fixed effects $\alpha_i$, time fixed effects
$\gamma_t$, and a treatment indicator $D_{it}$ equal to one for treated units in
post-treatment periods,

$$
Y_{it} = \alpha_i + \gamma_t + \tau\,D_{it} + \varepsilon_{it},
$$ {#eq-twfe}

and $\hat\tau$ is the DiD (two-way fixed-effects) estimator of the ATT.

**Identifying assumption: parallel trends.** DiD is unbiased for the ATT if, *absent
treatment*, the treated and control groups would have followed the same trajectory:

$$
\mathbb{E}[Y_{i,\text{post}}(0) - Y_{i,\text{pre}}(0) \mid T]
= \mathbb{E}[Y_{i,\text{post}}(0) - Y_{i,\text{pre}}(0) \mid C].
$$ {#eq-parallel-trends}

The groups may differ in *levels*—that is differenced away—but their untreated
*trends* must coincide. This assumption is fundamentally untestable, because the
treated group's untreated post-period trend is the missing counterfactual. What
*is* testable is its observable implication: if trends are parallel after
treatment, they were typically parallel before. A **pre-trends** (event-study)
check estimates leads and lags of the treatment and asks whether the pre-treatment
leads are jointly zero.

**What breaks identification.** Parallel trends fails whenever something other than
the treatment moves the groups differently over the window—a localized demand
shock, mean reversion when treatment is assigned to units with unusually low recent
outcomes (Ashenfelter's dip), or anticipation effects. A subtler hazard has emerged
in **staggered adoption**, where units are treated at different times: the standard
two-way fixed-effects estimator in @eq-twfe implicitly uses already-treated units
as controls for later-treated ones, and when effects vary over time this
"forbidden comparison" can produce a weighted average with *negative weights* and a
biased—even wrong-signed—estimate. Modern heterogeneity-robust estimators address
this, and a careful analyst reports both the two-way fixed-effects number and a
robust alternative.

```{r did-sim}
#| echo: true
#| message: false
#| warning: false
set.seed(47)

# Panel: 60 markets x 12 months. Treatment (a price cut) hits half the markets
# in month 7. The control group is constructed to satisfy parallel trends.
n_markets <- 60; n_months <- 12; treat_month <- 7; true_att <- 1.5
markets <- 1:n_markets
treated <- markets %in% sample(markets, n_markets / 2)

panel <- expand.grid(market = markets, month = 1:n_months)
panel$treated <- treated[panel$market]
market_fe <- rnorm(n_markets, 10, 2)[panel$market]   # level differences (ok)
time_fe   <- 0.20 * panel$month                       # common trend (ok)
post      <- panel$month >= treat_month
panel$D   <- as.integer(panel$treated & post)
panel$sales <- market_fe + time_fe + true_att * panel$D +
               rnorm(nrow(panel), 0, 0.8)

# DiD via two-way fixed effects (unit + time dummies).
fit <- lm(sales ~ factor(market) + factor(month) + D, data = panel)
att_hat <- coef(fit)["D"]
ci <- confint(fit)["D", ]
cat(sprintf("True ATT:        %.3f\n", true_att))
cat(sprintf("DiD estimate:    %.3f  [%.3f, %.3f]\n", att_hat, ci[1], ci[2]))
```

DiD is the workhorse of *advertising-spillover* and *pricing* studies in marketing,
precisely because firms so often roll changes out geographically and at staggered
times—exactly the structure DiD exploits, and exactly the structure that demands
the heterogeneity-robust care just described.

## Instrumental Variables {#sec-causal-iv}

Sometimes treatment is endogenous and no untreated group plausibly satisfies
parallel trends, yet there exists a third variable that shifts the treatment for
reasons unrelated to the outcome. **Instrumental variables** exploit that variable
to isolate the slice of treatment variation that is "as good as randomly assigned."
The motivating problem in marketing is the price endogeneity that defeats naive
demand estimation: observed prices and quantities are jointly determined by demand
and supply, so a regression of quantity on price recovers neither the demand curve
nor a causal price elasticity. An instrument that shifts *supply* (and hence price)
without shifting demand traces out the demand curve.

Let the structural ("second-stage") equation of interest be

$$
Y_i = \beta_0 + \beta_1 D_i + \mathbf{x}_i^{\top}\boldsymbol{\gamma} + u_i,
$$ {#eq-iv-structural}

where $D_i$ (e.g., price) is correlated with the error $u_i$
($\mathbb{E}[D_i u_i] \neq 0$), so OLS is biased. An instrument $Z_i$ identifies
$\beta_1$ if it satisfies three conditions:

1. **Relevance:** $Z_i$ shifts the treatment, $\operatorname{Cov}(Z_i, D_i) \neq 0$.
   This is testable—it is the first-stage regression.
2. **Exclusion:** $Z_i$ affects $Y_i$ *only* through $D_i$, not directly,
   $\operatorname{Cov}(Z_i, u_i) = 0$. This is untestable and must be argued from
   institutional knowledge.
3. **Exogeneity (independence):** $Z_i$ is as good as randomly assigned with respect
   to the unobservables driving $Y_i$.

The estimator is **two-stage least squares (2SLS)**: regress $D_i$ on $Z_i$ and
covariates to obtain fitted $\hat D_i$ (the exogenous part of treatment variation),
then regress $Y_i$ on $\hat D_i$. With a single instrument and no covariates this
reduces to the Wald ratio,

$$
\hat\beta_1^{\text{IV}}
= \frac{\operatorname{Cov}(Y_i, Z_i)}{\operatorname{Cov}(D_i, Z_i)}
= \frac{\widehat{\text{reduced form}}}{\widehat{\text{first stage}}}.
$$ {#eq-wald}

**What IV identifies.** With heterogeneous effects and a binary instrument, 2SLS
does not estimate the ATE but the **local average treatment effect (LATE)**: the
average effect among **compliers**—units whose treatment status responds to the
instrument—under a monotonicity condition (the instrument moves no unit in the
opposite direction). The effect on always-takers and never-takers is not
identified. This is a feature to interpret, not a bug to hide: the LATE answers
"what is the effect for the units a policy operating through this instrument would
actually move?"

**What breaks identification.** Two failures dominate. A **weak instrument** (low
first-stage correlation) inflates variance and, worse, biases 2SLS toward the OLS
estimate it was meant to fix; the conventional diagnostic is a first-stage
$F$-statistic well above the rule-of-thumb of 10, though that threshold is now
known to be optimistic. An **exclusion-restriction violation**—the instrument
affecting the outcome through any channel other than the treatment—biases the
estimate by an amount that *cannot* be detected from the data and must be defended
by argument. In demand estimation the classic instruments are cost shifters and,
following the differentiated-products tradition, functions of rival products'
characteristics that shift a product's markup without entering its demand
directly [@berry1995]. The credibility of any IV study lives or dies on the
narrative for the exclusion restriction.

```{r iv-sim}
#| echo: true
#| message: false
#| warning: false
set.seed(47)

# Demand estimation with price endogeneity. A demand shock 'xi' raises BOTH
# quantity and (through the firm's pricing) price -> OLS elasticity is biased
# toward zero. A cost shifter 'Z' instruments for price.
n <- 5000
xi   <- rnorm(n)                              # unobserved demand shock
Z    <- rnorm(n)                              # cost shifter (instrument)
price <- 2.0 - 0.8 * Z + 1.0 * xi + rnorm(n) # price rises with demand shock
true_elasticity <- -1.2
logq <- 5 + true_elasticity * price + 1.5 * xi + rnorm(n)  # xi enters demand

ols <- coef(lm(logq ~ price))["price"]

# 2SLS by hand: first stage, then second stage on fitted price.
first  <- lm(price ~ Z)
phat   <- fitted(first)
iv     <- coef(lm(logq ~ phat))["phat"]
F_stat <- summary(first)$fstatistic["value"]

cat(sprintf("True price elasticity:   %.3f\n", true_elasticity))
cat(sprintf("OLS (biased):            %.3f\n", ols))
cat(sprintf("2SLS (IV):               %.3f\n", iv))
cat(sprintf("First-stage F:           %.1f\n", F_stat))
```

## Regression Discontinuity {#sec-causal-rd}

A **regression discontinuity** design exploits rules that assign treatment based on
whether a continuous **running variable** crosses a known **cutoff**. A customer
becomes a "gold-tier" member at 1,000 loyalty points; a buyer earns free shipping
at a \$50 basket; an ad campaign targets ZIP codes above a median-income
threshold. Units just below and just above the cutoff are, in every respect other
than treatment, nearly identical—their assignment is determined by a sliver of the
running variable that is plausibly as good as random. Comparing outcomes just above
versus just below the threshold therefore identifies the treatment effect *at the
cutoff*.

Let $X_i$ be the running variable, $c$ the cutoff, and treatment $D_i = \mathbf{1}\{X_i \geq c\}$
in the **sharp** design. The estimand is the jump in the conditional expectation of
the outcome at the cutoff,

$$
\tau_{\text{RD}}
= \lim_{x \downarrow c} \mathbb{E}[Y_i \mid X_i = x]
- \lim_{x \uparrow c} \mathbb{E}[Y_i \mid X_i = x].
$$ {#eq-47-rd}

In practice $\tau_{\text{RD}}$ is estimated by fitting flexible regressions—usually
**local linear** regressions within a bandwidth $h$ around $c$—separately on each
side and reading off the gap at $c$. In the **fuzzy** design, crossing the cutoff
changes the *probability* of treatment rather than treatment itself; the jump in
outcomes is then scaled by the jump in treatment probability, making fuzzy RD a
local IV with cutoff-crossing as the instrument.

**Identifying assumption: continuity.** RD identifies the effect at the cutoff if
the potential outcomes $\mathbb{E}[Y_i(0) \mid X_i = x]$ and
$\mathbb{E}[Y_i(1) \mid X_i = x]$ are **continuous** in $x$ at $c$—that is, nothing
*other* than treatment jumps at the threshold. Then any discontinuity in the
observed outcome must be caused by the treatment.

**What breaks identification.** The fatal threat is **manipulation** of the running
variable: if units can precisely control which side of the cutoff they land on—a
customer nudging a basket to exactly \$50, a salesperson booking revenue to clear a
quota—then those who sort just above the cutoff differ systematically from those
just below, and continuity fails. The standard diagnostic is the **McCrary density
test**: a discontinuity in the *density* of $X$ at $c$ signals sorting. Two further
practices guard validity: confirm that pre-determined covariates do *not* jump at
the cutoff (a placebo RD on baseline characteristics), and show the estimate is
robust to bandwidth choice, since RD trades bias (wide bandwidth, more
extrapolation) against variance (narrow bandwidth, fewer observations). The estimand
is also intrinsically **local**: RD recovers the effect only at $c$ and need not
generalize to units far from the threshold.

```{r rd-sim}
#| echo: true
#| message: false
#| warning: false
#| label: fig-rd
#| fig-cap: "A sharp regression-discontinuity design. Treatment (a loyalty-tier upgrade) switches on at the cutoff of 1,000 points. Local linear fits on each side identify the jump in spending at the threshold; points are a 10% sample for legibility."
set.seed(47)

# Running variable: loyalty points. Treatment at >= 1000. Smooth baseline
# relationship between points and spending PLUS a true jump of 8 at the cutoff.
n <- 4000; cutoff <- 1000; true_jump <- 8
points  <- runif(n, 600, 1400)
D       <- as.integer(points >= cutoff)
spend   <- 20 + 0.03 * (points - cutoff) + true_jump * D + rnorm(n, 0, 4)

# Local linear RD within a bandwidth, separate slopes each side.
h <- 150
inb <- abs(points - cutoff) <= h
rd_fit <- lm(spend ~ D + I(points - cutoff) + D:I(points - cutoff),
             data = data.frame(spend, D, points), subset = inb)
jump_hat <- coef(rd_fit)["D"]
cat(sprintf("True jump at cutoff: %.3f\n", true_jump))
cat(sprintf("RD estimate:         %.3f\n", jump_hat))

s <- sample(n, 400)
plot(points[s], spend[s], pch = 16, col = "grey70", cex = 0.6,
     xlab = "Loyalty points (running variable)", ylab = "Spending")
abline(v = cutoff, lty = 3)
gl <- seq(cutoff - h, cutoff, length.out = 50)
gr <- seq(cutoff, cutoff + h, length.out = 50)
predl <- predict(rd_fit, data.frame(D = 0, points = gl))
predr <- predict(rd_fit, data.frame(D = 1, points = gr))
lines(gl, predl, lwd = 3); lines(gr, predr, lwd = 3)
```

## Synthetic Control {#sec-causal-scm}

Difference-in-differences needs a control group whose untreated trend matches the
treated group's. When the treated unit is a *single* large entity—one country where
a brand launched, one state that changed a privacy law, one market where a flagship
store opened—no single control unit is a credible match. The **synthetic control
method** constructs the counterfactual as a *weighted average* of untreated "donor"
units, choosing the weights so that the synthetic unit reproduces the treated
unit's pre-treatment outcome path (and predictors). The treatment effect is then the
gap between the treated unit and its synthetic twin after the intervention.

Let unit $1$ be treated at time $T_0$ and units $2,\dots,J+1$ be the donor pool.
Choose nonnegative weights $\mathbf{w} = (w_2,\dots,w_{J+1})$, with
$\sum_j w_j = 1$, to minimize pre-treatment discrepancy,

$$
\mathbf{w}^{\star} = \arg\min_{\mathbf{w}}
\sum_{t < T_0}\Big(Y_{1t} - \sum_{j=2}^{J+1} w_j\,Y_{jt}\Big)^2,
\quad w_j \geq 0,\ \textstyle\sum_j w_j = 1.
$$ {#eq-scm-weights}

The estimated effect at each post-period is the gap between the treated unit and
the weighted donors,

$$
\hat\tau_{1t} = Y_{1t} - \sum_{j=2}^{J+1} w_j^{\star}\,Y_{jt},
\qquad t \geq T_0.
$$ {#eq-scm-effect}

**Identifying assumptions.** SCM identifies the treated unit's ATT if (i) the
weighted donors reproduce the treated unit's *pre-treatment* outcomes closely
(good pre-fit), (ii) the donors are themselves untreated and free of spillovers
from the treatment, and (iii) the relationship that made the synthetic control
track the treated unit before treatment *continues to hold* absent treatment—the
synthetic-control analogue of parallel trends, but enforced on the full
pre-period path rather than a single trend. The convexity constraints
($w_j \geq 0$, summing to one) prevent extrapolation outside the donor support and
make the weights interpretable.

**What breaks identification.** Poor pre-treatment fit is disqualifying: if the
synthetic unit cannot track the treated unit *before* the intervention, the
post-period gap is not credibly causal. **Interpolation bias** arises when donors
are too dissimilar from the treated unit; **overfitting** to noise inflates
apparent effects when the pre-period is short. Because there is only one treated
unit, inference rests not on large-sample theory but on **placebo (permutation)
tests**: reassign the treatment to each donor in turn, compute the placebo gap, and
ask whether the true treated unit's post/pre gap ratio is extreme relative to the
placebo distribution.

```{r scm-sim}
#| echo: true
#| message: false
#| warning: false
#| label: fig-scm
#| fig-cap: "Synthetic control for a single treated market. Donor weights are chosen to match the treated market's pre-intervention sales (vertical line at the launch period). The post-period gap between the treated market and its synthetic twin estimates the treatment effect."
set.seed(47)

# One treated market + 20 donors over 30 periods; a brand campaign starts at t=21
# and lifts the treated market's sales by a growing amount.
Tn <- 30; T0 <- 20; J <- 20; true_post_effect <- 6
common <- cumsum(rnorm(Tn, 0.2, 0.4))                 # shared latent trend
donors <- sapply(1:J, function(j)
  3 + runif(1, 0.5, 1.5) * common + rnorm(Tn, 0, 0.6))
# Treated market is a convex blend of three donors, pre-treatment.
truew <- numeric(J); truew[c(2, 5, 11)] <- c(0.5, 0.3, 0.2)
treated <- as.numeric(donors %*% truew) + rnorm(Tn, 0, 0.4)
treated[(T0 + 1):Tn] <- treated[(T0 + 1):Tn] +
  true_post_effect * seq_len(Tn - T0) / (Tn - T0)

# Solve for nonnegative weights summing to 1 on the PRE period via a simple
# projected optimization (quadratic loss, simplex constraint).
pre <- 1:T0
loss <- function(w) sum((treated[pre] - donors[pre, ] %*% w)^2)
ui <- rbind(diag(J), rep(-1, J)); ci <- c(rep(0, J), -1)  # w>=0, sum(w)<=1
w0 <- rep(0.9 / J, J)   # interior start: sum < 1, so constrOptim's strict-feasibility check passes
opt <- constrOptim(w0, loss, grad = NULL, ui = ui, ci = ci)
w <- opt$par / sum(opt$par)

synth <- as.numeric(donors %*% w)
gap_post <- mean(treated[(T0 + 1):Tn] - synth[(T0 + 1):Tn])
cat(sprintf("Mean post-period effect (truth ~%.2f): %.3f\n",
            true_post_effect / 2, gap_post))

plot(1:Tn, treated, type = "l", lwd = 2, xlab = "Period", ylab = "Sales",
     ylim = range(c(treated, synth)))
lines(1:Tn, synth, lwd = 2, lty = 2, col = "firebrick")
abline(v = T0 + 0.5, lty = 3)
legend("topleft", bty = "n", lwd = 2, lty = c(1, 2),
       col = c("black", "firebrick"),
       legend = c("treated market", "synthetic control"))
```

## Choosing a Design

No design dominates; each buys identification with a different assumption, and the
right choice depends on what variation the world (or the firm) supplies. The
comparison below summarizes the trade-offs.

```{r design-table}
#| echo: false
#| message: false
#| warning: false
#| label: tbl-designs
#| tbl-cap: "Causal designs compared: the estimand each targets, the assumption that does the identifying work, the diagnostic that probes it, and the failure that breaks it."
designs <- data.frame(
  Design = c("Randomized / A-B test", "Difference-in-differences",
             "Instrumental variables", "Regression discontinuity",
             "Synthetic control"),
  Estimand = c("ATE", "ATT", "LATE (compliers)",
               "Local effect at cutoff", "ATT (one unit)"),
  `Key assumption` = c("Random assignment",
                       "Parallel trends",
                       "Exclusion + relevance",
                       "Continuity at cutoff",
                       "Donors match pre-trend"),
  Diagnostic = c("Balance / power checks",
                 "Pre-trend (event study)",
                 "First-stage F; over-ID",
                 "McCrary density; covariate jumps",
                 "Pre-fit; placebo permutation"),
  `What breaks it` = c("Interference; peeking",
                       "Differential shocks; staggering",
                       "Weak / invalid instrument",
                       "Manipulation of running var.",
                       "Poor pre-fit; spillovers"),
  check.names = FALSE
)
knitr::kable(designs, align = "lllll")
```

Three principles guide the choice in marketing practice. First, **prefer
experimentation where the cost of running it is bearable**; in advertising and
pricing, firms increasingly hold out geographies or randomize at the user level
precisely because the observational alternatives are so unreliable [@gordon2019].
Second, when an experiment is infeasible, **match the design to the source of
exogenous variation**: a sharp eligibility rule invites RD, a staggered rollout
invites DiD, a single large treated entity invites SCM, and a credible supply-side
shifter invites IV. Third, **triangulate**: where two designs with different
assumptions point to the same answer, the causal claim is far more credible than
any single estimate, because the assumptions fail for different reasons.

## Threats Common to All Designs {#sec-causal-threats}

Beyond design-specific failures, several threats recur and deserve a standing
checklist. **Spillovers and general equilibrium**: a treatment that changes one
unit's behavior may change untreated units' behavior (SUTVA failure), and a price
experiment that would be profitable at small scale may not survive a firm-wide
rollout once competitors and supply respond. **Attrition and survivorship**:
differential dropout between arms reintroduces selection even into a randomized
design. **Outcome windows and dynamics**: advertising and brand effects accumulate
and decay, so an effect measured over too short a window understates the true
return; carryover (the Koyck/adstock structure of @sec-advertising) must be modeled,
not assumed away. **Multiple comparisons**: testing many metrics, segments, or
variants inflates false positives, demanding pre-registration of primary outcomes or
explicit correction. **External validity**: every estimand here is local to a
population, a period, and a level of treatment intensity; a LATE for compliers, an
RD effect at a cutoff, or an SCM effect for one market need not transfer to the
firm's broader decision. Naming the estimand—the discipline this chapter
opened with—is the antidote, because it forces the analyst to state precisely whose
effect, under what intervention, has been estimated.

## Foundational vs. frontier at a glance {#sec-causal-foundational-frontier}

The seminar marks every method as foundational (qualifying-exam material the
left column) or frontier (where dissertations are written, the right column). The
marketing-specific value-add over a generic econometrics course is concentrated in
Weeks 12–14—advertising measurement, incrementality/geo/pricing experiments, and
marketplace interference—and in framing every method around small effects,
non-stationary platforms, and targeting decisions.

| Block | Foundational core [F] | Frontier extensions [R] |
|---|---|---|
| Framework (Wk 1) | Rubin 1974; Holland 1986; Imbens–Rubin | design- vs. model-based inference |
| Experiments / A/B (Wk 2) | Fisher/Neyman inference; Kohavi | time-series/switchback; always-valid inference |
| Selection on observables (Wk 3) | CIA, bad controls, overlap (Angrist–Pischke) | sensitivity analysis; Athey–Imbens overview |
| IV (Wk 4) | Imbens–Angrist LATE; Angrist–Imbens–Rubin | weak-IV; judge/shift-share designs |
| DiD I (Wk 5) | Card–Krueger; Bertrand–Duflo–Mullainathan | wild-cluster inference |
| DiD II — staggered (Wk 6) | the *critique itself* is now foundational | Goodman-Bacon; CS; dCDH; Sun–Abraham; BJS; Roth et al. |
| RDD (Wk 7) | Imbens–Lemieux guide | Calonico–Cattaneo–Titiunik robust inference |
| Synthetic control (Wk 8) | Abadie–Diamond–Hainmueller; Abadie JEL | synthetic DiD; gsynth; CausalImpact |
| Panel/event-study (Wk 9) | FE/FD (Wooldridge) | design-based panel experiments; matrix completion |
| Matching/weighting (Wk 10) | Rosenbaum–Rubin; Hirano–Imbens–Ridder | approximate residual balancing; PS-matching inference |
| ML for HTE (Wk 11) | Athey–Imbens causal trees | causal forests; GRF; double/debiased ML |
| Ad measurement (Wk 12) | Lewis–Rao unfavorable economics | ghost ads; experiments-vs-observational (Gordon et al.) |
| Incrementality/geo/pricing (Wk 13) | Blake–Nosko–Tadelis; Anderson–Simester | targeting-policy evaluation; bandit pricing |
| Interference (Wk 14) | Manski reflection problem | Aronow–Samii; Sävje et al.; graph-cluster designs |

Week 6 is the rare case where a *critique* has itself become foundational: no
student can use DiD today without knowing the staggered-adoption failure of the
two-way fixed-effects estimator. A paper is promoted from frontier to foundational
only once it is standard qualifying-exam material—the staggered-DiD critique is the
current example mid-transition.

## How this chapter expands {#sec-causal-expansion}

This chapter is a living syllabus. Future revisions should (1) reconcile this
canonical ordering against actual week-by-week reading lists pulled from live PhD
syllabi (MIT applied econometrics, Chicago Booth's causal-inference and
quantitative-marketing sequences, Stanford GSB/Econ causal-ML courses, Berkeley
digital-experiments, and the Wharton/Columbia/Kellogg marketing-methods seminars),
citing program and URL; (2) track Week 6, which moves fastest, as the robust-DiD
literature consolidates (continuous treatment, non-absorbing treatment,
triple-difference, design-based DiD); (3) deepen the marketing-platform core of
Weeks 12–14 with incrementality and geo-experiment methodology as version-of-record
DOIs appear (much of it currently lives in working papers and platform tech
reports), plus the growing experimentation-under-interference literature for
marketplaces and recommender systems; (4) add a power/design appendix on MDE/power
for small-lift ad experiments, variance reduction (CUPED), cluster-randomization
power loss, and sequential testing; (5) add a software/replication appendix mapping
methods to canonical packages (`grf`, `DoubleML`/`econml`, `did`, `fixest`/`sunab`,
`rdrobust`, `Synth`/`synthdid`/`gsynth`, `CausalImpact`); and (6) maintain the
foundational/frontier marking as the load-bearing pedagogical device.

## Key Takeaways

- A causal effect is a comparison of potential outcomes (@eq-unit-effect); because
  only one is observed per unit (@eq-switching), every estimate rests on an
  assumption that recovers the missing counterfactual.
- The naive treated-minus-untreated difference is biased by selection
  (@eq-selection-bias); randomization removes the bias by construction, which is
  why field experiments and A/B tests are the benchmark.
- In advertising, the effect is small relative to sales volatility, so experiments
  must be very large to be conclusive and observational substitutes can err by
  multiples or in sign [@Lewis_2015; @gordon2019].
- Each quasi-experimental design targets a *specific* estimand under a *specific*
  assumption—DiD/ATT under parallel trends, IV/LATE under exclusion, RD under
  continuity, SCM/ATT under donor pre-fit—and each has a named diagnostic and a
  named failure mode (@tbl-designs).
- Name the estimand before choosing a method, probe the identifying assumption with
  its diagnostic, and triangulate across designs whose assumptions fail for
  different reasons.

## Further Reading

The potential-outcomes framework and the estimators in this chapter are developed
at length in the standard graduate econometrics references
[@cameron2005microeconometrics; @greene2003econometric] and in the econometric
treatments of program evaluation [@heckman2007econometric; @abbring2007econometric3].
For demand estimation with endogenous prices and instruments built from rival
product characteristics, the differentiated-products literature is the entry point
[@berry1995]. The advertising-measurement evidence that motivates experimentation
over observational inference is most directly accessible through the large-scale
experimental comparisons cited above [@lewis2011; @Lewis_2015; @gordon2019], and the
dynamics of advertising response that complicate any single-window estimate are
treated in @sec-advertising.

<!-- CITE-NEEDED: Neyman-Rubin / Rubin causal model foundational potential-outcomes references -->
<!-- CITE-NEEDED: Angrist-Imbens LATE / compliers identification -->
<!-- CITE-NEEDED: Card-Krueger or Bertrand-Duflo-Mullainathan difference-in-differences inference -->
<!-- CITE-NEEDED: Goodman-Bacon / Callaway-Sant'Anna staggered DiD with heterogeneous effects -->
<!-- CITE-NEEDED: staiger-stock / stock-yogo weak instruments F-statistic -->
<!-- CITE-NEEDED: Imbens-Lemieux / Lee-Lemieux regression discontinuity practical guide -->
<!-- CITE-NEEDED: McCrary density manipulation test for RD -->
<!-- CITE-NEEDED: Abadie-Diamond-Hainmueller synthetic control method -->
<!-- CITE-NEEDED: marketing pricing field experiment exemplar (e.g., online price experimentation) -->
<!-- CITE-NEEDED: sequential testing / always-valid p-values for online experiments (peeking) -->
