# Artificial Intelligence and Machine Learning in Marketing {#sec-ai-ml}

*Machine learning* is the study of algorithms that improve their performance on a
task as they are exposed to more data, rather than by being explicitly programmed
with the rules of that task. *Artificial intelligence* (AI) is the broader project
of building systems that perform tasks we associate with human cognition;
contemporary AI in marketing is, almost entirely, applied machine learning at
scale. The distinction matters less than what the two share: a willingness to let
flexible, high-capacity function approximators discover structure in data that a
human analyst could neither specify in advance nor write down as equations. That
willingness is also the source of every pitfall in this chapter.

Marketing is an unusually hospitable host for these methods, for three reasons.
First, the field is awash in the kind of data—clickstreams, transactions, images,
reviews, search queries—on which modern learning algorithms thrive
[@wedel2016; @martin2017data]. Second, many marketing tasks are natively
*predictive*: which customer will churn, which creative will earn the click, which
product to surface next. Prediction is exactly what supervised learning does well.
Third, the economic stakes of small improvements are large, because marketing
decisions are made billions of times a day across recommendation, bidding, and
targeting systems, so a one-percent lift compounds into real money
[@varian2016; @gao2024big].

This chapter has two jobs. The first is to give a working, formal command of the
methods—supervised and unsupervised learning, recommender systems, deep learning,
and large language models—at a level that lets the reader state each method's
estimator, its assumptions, and the conditions under which it fails. The second is
to install a discipline that the hype around AI actively erodes: the distinction
between **prediction** and **inference**, and with it a sober catalogue of the ways
machine learning quietly breaks in deployment—leakage, distribution drift, and
unfairness. A model that predicts well in a notebook and harms the business in
production is the modal failure, not the exception, and most of those failures are
conceptual rather than computational. By the end, the reader should be able to map
a marketing problem onto the right learning paradigm, build and validate a model
that does not lie to them, and recognize when a predictive tool is being asked,
illegitimately, to answer a causal question.

We assume familiarity with the regression and choice-modeling machinery developed
earlier in the book (@sec-measurement-scales for measurement; the causal-inference
and marketing-mix chapters for identification), and we connect to them rather than
re-deriving them.

## Prediction Versus Inference: The Organizing Distinction

The single most consequential idea in this chapter is also the one most often
skipped. Consider a generic supervised relationship between an outcome $Y$ and
features $\mathbf{X}$,
$$
Y = f(\mathbf{X}) + \varepsilon, \qquad \mathbb{E}[\varepsilon \mid \mathbf{X}] = 0,
$$ {#eq-aiml-dgp}
with $\hat f$ an estimate learned from data. There are two fundamentally different
things one can want from $\hat f$, and conflating them is the root of most misuse
of machine learning in marketing.

> In a **prediction** problem the object of interest is $\hat f(\mathbf{x})$ itself:
> we want accurate values of $Y$ at new inputs and treat $\hat f$ as a black box. In
> an **inference** problem the object of interest is some functional of $f$—a
> coefficient, an elasticity, a treatment effect—and $\hat f$ is a means to learn
> *how* $\mathbf{X}$ relates to $Y$, not merely to forecast $Y$.

The two goals reward different choices. Prediction tolerates—indeed often
prefers—biased, uninterpretable, highly flexible estimators, because the only
scorecard is out-of-sample loss, $\mathbb{E}[L(Y, \hat f(\mathbf{X}))]$ on data the
model has never seen. The bias–variance trade-off is the governing law: expected
squared prediction error decomposes as
$$
\mathbb{E}\!\left[(Y - \hat f(\mathbf{x}_0))^2\right]
= \underbrace{\big(\mathbb{E}[\hat f(\mathbf{x}_0)] - f(\mathbf{x}_0)\big)^2}_{\text{bias}^2}
+ \underbrace{\operatorname{Var}\!\big(\hat f(\mathbf{x}_0)\big)}_{\text{variance}}
+ \underbrace{\sigma^2_\varepsilon}_{\text{irreducible}},
$$ {#eq-aiml-bv}
and a method that accepts some bias to cut variance can win. Inference, by contrast,
demands an estimator whose sampling distribution we understand—unbiasedness or a
known bias, valid standard errors, an identification argument linking the estimand
to features of the data-generating process. A random forest can have lower test
error than a linear regression while being *useless* for the question "what is the
effect of a \$1 price cut," because flexible learners trade interpretable,
consistent parameters for predictive accuracy [@varian2016].

The practical hazard is that a predictive model's coefficients—or its
feature-importance scores—look like answers to inference questions and are routinely
read as such. They are not. A churn model may load heavily on "number of support
tickets," but acting on that association by *suppressing* support tickets would be
disastrous; the feature predicts churn because both are caused by underlying
dissatisfaction. Predictive importance is not causal importance, and no amount of
test-set accuracy converts one into the other. When the marketing question is "what
will happen," supervised learning is the right tool; when it is "what should we do,"
the model must be embedded in a design that identifies a causal effect—a randomized
experiment, an instrument, or one of the causal-machine-learning estimators we reach
at the end of the chapter. @fig-aiml-pred-inf fixes the fork in the road.

```{mermaid}
%%| label: fig-aiml-pred-inf
%%| fig-cap: "The prediction–inference fork. The same data and even the same algorithm serve different goals, validated by different criteria. Reading a predictive model's internals as causal estimates is the central error this chapter warns against."
flowchart TD
  Q["Marketing question"] --> D{"What is the<br/>object of interest?"}
  D -->|"Accurate Y at new inputs"| P["PREDICTION<br/>(black-box f-hat)"]
  D -->|"A functional of f:<br/>effect, elasticity"| I["INFERENCE<br/>(structure of f)"]
  P --> PV["Validate by:<br/>out-of-sample loss,<br/>cross-validation"]
  I --> IV["Validate by:<br/>identification,<br/>standard errors, design"]
  PV --> PU["Use: scoring, ranking,<br/>forecasting, matching"]
  IV --> IU["Use: pricing, budget<br/>allocation, policy"]
  PU -.->|"Do NOT read feature<br/>importance as causal"| IU
```

## Supervised Learning

In **supervised learning** the training data are labeled pairs
$\{(\mathbf{x}_i, y_i)\}_{i=1}^n$, and the goal is to learn a mapping
$\hat f : \mathcal{X} \to \mathcal{Y}$ that generalizes to unlabeled inputs. When
$Y$ is continuous the task is *regression*; when $Y$ is categorical it is
*classification*. Almost every workhorse marketing model—propensity to buy, churn,
response, lifetime-value, lead scoring, ad click-through—is a supervised classifier
or regressor.

### The learning problem and regularization

Formally, learning chooses $\hat f$ to minimize *regularized empirical risk*,
$$
\hat f = \arg\min_{f \in \mathcal{F}} \; \frac{1}{n}\sum_{i=1}^{n} L\big(y_i, f(\mathbf{x}_i)\big) + \lambda\, \Omega(f),
$$ {#eq-aiml-erm}
where $L$ is a loss function (squared error for regression, log-loss/cross-entropy
for classification), $\mathcal{F}$ is the hypothesis class (linear functions, trees,
neural networks), $\Omega(f)$ is a complexity penalty, and $\lambda \ge 0$ tunes the
trade-off. Minimizing training loss alone ($\lambda = 0$, $\mathcal{F}$ rich) yields
*overfitting*: $\hat f$ memorizes noise and generalizes poorly, the high-variance
failure in @eq-aiml-bv. The penalty $\Omega$ buys generalization by shrinking the
effective complexity of $\hat f$. For linear models the two canonical choices are
the $\ell_2$ (ridge) penalty $\Omega(\boldsymbol\beta)=\|\boldsymbol\beta\|_2^2$,
which shrinks coefficients smoothly, and the $\ell_1$ (lasso) penalty
$\Omega(\boldsymbol\beta)=\|\boldsymbol\beta\|_1$, which sets some coefficients
exactly to zero and thereby performs variable selection—valuable when the feature
space is wide, as it almost always is with behavioral data [@varian2016].

The assumptions behind @eq-aiml-erm are easy to state and easy to violate. Empirical
risk minimization is consistent for the risk-minimizing $f$ only if the training
data are drawn from the *same distribution* as the deployment data (the
identically-distributed assumption) and if observations are *exchangeable* in the
way the validation scheme assumes (commonly independence). Both assumptions fail
routinely in marketing: deployment data drift away from training data over time
(@sec-aiml-drift), and observations are correlated within customers, sessions, and
time, which—if ignored—makes naïve cross-validation report accuracy the model will
never achieve in production.

### Tree ensembles: the marketing workhorse

For tabular marketing data—mixed numeric and categorical features, nonlinearities,
interactions, missingness—ensembles of decision trees are, empirically, the default
high-performer. A single regression tree partitions $\mathcal{X}$ into rectangular
regions $R_1,\dots,R_M$ and predicts the within-region mean, $\hat
f(\mathbf{x})=\sum_m c_m \mathbf{1}\{\mathbf{x}\in R_m\}$; it is interpretable but
high-variance. Two ensemble strategies tame the variance. **Random forests** average
many trees grown on bootstrap samples with randomly restricted split candidates,
reducing variance through decorrelation. **Gradient-boosted trees** instead fit
trees *sequentially*, each new tree $h_t$ targeting the gradient of the loss left by
the running ensemble,
$$
\hat f_t(\mathbf{x}) = \hat f_{t-1}(\mathbf{x}) + \nu\, h_t(\mathbf{x}), \qquad
h_t \approx -\,\frac{\partial L}{\partial \hat f_{t-1}},
$$ {#eq-aiml-boost}
with learning rate $\nu \in (0,1]$. Boosting reduces bias and variance jointly and
typically tops leaderboards on tabular data, at the cost of more careful tuning to
avoid overfitting (the number of trees becomes a regularization parameter set by
validation). The worked example below builds a churn classifier and—critically—shows
how to validate it honestly. <!-- CITE-NEEDED: gradient boosting / random forests canonical reference (Breiman; Friedman) not in allowlist -->

```{r aiml-churn, message=FALSE, warning=FALSE}
set.seed(48)

# --- Simulate a customer-churn dataset with a known structure ----------------
n <- 4000
tenure        <- rpois(n, lambda = 18)                       # months as customer
monthly_spend <- round(rgamma(n, shape = 2, scale = 25), 2)  # $ per month
support_calls <- rpois(n, lambda = 1 + 0.05 * (40 - pmin(tenure, 40)))
discount_user <- rbinom(n, 1, 0.35)

# True churn propensity: short tenure, low spend, many support calls raise risk.
lin <- -1.0 - 0.06 * tenure - 0.015 * monthly_spend +
        0.45 * support_calls + 0.30 * discount_user
prob_churn <- plogis(lin)
churn <- rbinom(n, 1, prob_churn)

dat <- data.frame(churn = factor(churn, labels = c("stay", "leave")),
                  tenure, monthly_spend, support_calls,
                  discount_user = factor(discount_user))

# --- Honest train/test split -------------------------------------------------
idx   <- sample(seq_len(n), size = floor(0.7 * n))
train <- dat[idx, ]
test  <- dat[-idx, ]

# --- Gradient-boosted trees (gbm); fall back to logistic if gbm absent -------
has_gbm <- requireNamespace("gbm", quietly = TRUE)
if (has_gbm) {
  fit <- gbm::gbm(I(as.integer(churn) - 1) ~ tenure + monthly_spend +
                    support_calls + discount_user,
                  data = train, distribution = "bernoulli",
                  n.trees = 600, interaction.depth = 3,
                  shrinkage = 0.03, bag.fraction = 0.7, verbose = FALSE)
  best <- gbm::gbm.perf(fit, plot.it = FALSE, method = "OOB")
  p_hat <- gbm::predict.gbm(fit, test, n.trees = best, type = "response")
} else {
  fit   <- glm(churn ~ tenure + monthly_spend + support_calls + discount_user,
               data = train, family = binomial())
  p_hat <- predict(fit, test, type = "response")
}

# --- Out-of-sample evaluation: AUC and a calibration check -------------------
auc <- function(score, label) {              # rank-based AUC, no extra packages
  pos <- score[label == "leave"]; neg <- score[label == "stay"]
  mean(outer(pos, neg, ">") + 0.5 * outer(pos, neg, "=="))
}
cat("Test AUC:", round(auc(p_hat, test$churn), 3), "\n")

# Calibration: do predicted probabilities match realized churn rates?
bins <- cut(p_hat, breaks = quantile(p_hat, 0:5/5), include.lowest = TRUE)
calib <- aggregate(as.integer(test$churn) - 1 ~ bins, FUN = mean)
calib$predicted <- tapply(p_hat, bins, mean)
names(calib) <- c("bin", "observed_churn", "mean_predicted")
calib
```

The example reports two diagnostics, not one. *Discrimination* (AUC) measures
whether the model ranks churners above non-churners; *calibration* measures whether
a predicted 30% churn probability corresponds to a 30% realized rate. A model can
discriminate well yet be badly calibrated, and marketing decisions that multiply
predicted probabilities by margins—expected-value targeting—need calibration, not
just ranking [@neumann2019frontiers]. Reporting only AUC is a common and costly
omission.

### Classification thresholds and the cost of errors

A classifier outputs a score $\hat p(\mathbf{x}) = \widehat{\Pr}(Y=1\mid\mathbf{x})$;
turning it into an action requires a threshold $\tau$ such that we treat customers
with $\hat p > \tau$. The optimal threshold is *not* $0.5$—it depends on the
asymmetric costs of false positives and false negatives. If contacting a non-churner
costs $c_{\text{FP}}$ and failing to retain a churner costs $c_{\text{FN}}$, the
expected-cost-minimizing rule acts when
$\hat p / (1-\hat p) > c_{\text{FP}} / c_{\text{FN}}$. This is the point at which the
predictive model meets the decision problem, and it is where the marketing economics
re-enter: the ROC and precision–recall curves exist precisely because the right
operating point is a business choice, not a statistical default.

## Unsupervised Learning

In **unsupervised learning** the data are unlabeled, $\{\mathbf{x}_i\}_{i=1}^n$, and
the goal is to discover latent structure—groups, dimensions, topics—without a target
variable to supervise the search. The two dominant marketing uses are
*segmentation* (clustering customers or products) and *dimension reduction*
(compressing high-dimensional behavior into interpretable factors). Because there is
no label, there is no test-set accuracy to adjudicate "correctness"; validation is
intrinsically harder and more judgmental, which is both the method's flexibility and
its danger.

### Clustering and the segmentation problem

The canonical objective is *$k$-means*, which partitions observations into $K$
clusters to minimize within-cluster squared distance,
$$
\min_{\{S_k\}}\; \sum_{k=1}^{K}\sum_{\mathbf{x}_i \in S_k} \big\|\mathbf{x}_i - \boldsymbol\mu_k\big\|_2^2,
\qquad \boldsymbol\mu_k = \frac{1}{|S_k|}\sum_{\mathbf{x}_i \in S_k}\mathbf{x}_i.
$$ {#eq-aiml-kmeans}
This connects directly to the *a priori* versus *post hoc* segmentation distinction
of @sec-segmentation: post hoc segmentation is precisely the application of
a clustering algorithm to behavioral or attitudinal data to *discover* segments
rather than impose them. The estimator (Lloyd's algorithm) alternates assigning
points to the nearest centroid and recomputing centroids; it converges to a *local*
optimum, so results depend on initialization and on the (analyst-chosen) number of
clusters $K$. Three assumptions break identification of a "true" segmentation, and
all three are routinely violated in practice: $k$-means presumes roughly spherical,
equal-variance clusters (Euclidean distance encodes this); it is not scale-invariant,
so features must be standardized or the largest-variance feature dominates; and $K$
is not learned but assumed. Model-based clustering via finite mixtures replaces the
hard geometry with a probabilistic generative model and lets information criteria
choose $K$, at the cost of distributional assumptions. The deeper caution is that a
clustering algorithm *always* returns clusters, whether or not the population is
actually clustered; the burden is on the analyst to show the segments are stable,
managerially distinguishable, and reproducible out of sample, not merely that the
algorithm ran.

```{r aiml-segmentation, message=FALSE, warning=FALSE}
set.seed(48)

# Three latent customer segments differing in recency, frequency, monetary value
make_seg <- function(n, r, f, m) data.frame(
  recency   = pmax(1, round(rnorm(n, r, 8))),
  frequency = pmax(1, round(rnorm(n, f, 3))),
  monetary  = pmax(5, round(rnorm(n, m, 40)))
)
rfm <- rbind(make_seg(300, 10, 14,  220),   # champions
             make_seg(300, 45,  4,   60),   # at-risk
             make_seg(300, 25,  8,  130))   # mainstream

# Standardize before clustering: k-means is NOT scale-invariant
rfm_z <- scale(rfm)
km <- kmeans(rfm_z, centers = 3, nstart = 25)

# Profile the recovered segments on the original (interpretable) scale
prof <- aggregate(rfm, by = list(segment = km$cluster), FUN = function(x) round(mean(x), 1))
prof$size <- as.integer(table(km$cluster))
prof
```

### Dimension reduction

When behavior is high-dimensional—thousands of SKUs, pages, or features—*dimension
reduction* finds a low-dimensional representation that preserves the information that
matters. Principal component analysis (PCA) projects $\mathbf{X}$ onto the orthogonal
directions of maximal variance, the leading eigenvectors of the covariance matrix;
the first few components often capture interpretable axes of behavior (e.g., overall
intensity, then category mix). Non-negative matrix factorization and, for
text, *topic models* such as latent Dirichlet allocation generalize the idea to
parts-based and probabilistic decompositions [@tirunillai2014a; @buschken2016sentence].
The methods double as a feature-engineering step for supervised models and as a
listening tool: factorizing the term–document matrix of online reviews recovers the
latent dimensions of quality consumers actually discuss, a structure managers cannot
specify in advance [@tirunillai2014a; @netzer2008].

## Recommender Systems

Recommender systems are the most economically consequential deployment of machine
learning in marketing: they choose which of millions of items to surface to each
user, and on platforms from retail to streaming they drive a large share of demand.
Formally, a recommender estimates a *utility* or *preference* score
$\hat r_{ui}$ for each user $u$ and item $i$, then ranks items by that score. The
data are a sparse user–item matrix $\mathbf{R}$, mostly missing, with observed
entries being ratings, clicks, or purchases.

Two paradigms, with a hybrid, organize the field. **Content-based** filtering
recommends items similar to those a user has liked, using item features; it handles
new items but cannot discover tastes outside a user's history. **Collaborative
filtering** ignores item content and exploits the wisdom of the crowd—users who
agreed in the past will agree in the future—and is the more powerful approach when
interaction data are dense. The dominant collaborative formulation is *matrix
factorization*, which embeds users and items in a shared latent space of dimension
$K$ and models preference as an inner product,
$$
\hat r_{ui} = \mathbf{p}_u^{\top}\mathbf{q}_i + b_u + b_i + \mu,
\qquad
\min_{\mathbf{P},\mathbf{Q},\mathbf{b}} \sum_{(u,i)\in\mathcal{K}}
\big(r_{ui} - \hat r_{ui}\big)^2 + \lambda\big(\|\mathbf{p}_u\|^2 + \|\mathbf{q}_i\|^2 + b_u^2 + b_i^2\big),
$$ {#eq-aiml-mf}
where $\mathbf{p}_u, \mathbf{q}_i \in \mathbb{R}^K$ are the learned user and item
factors, $b_u, b_i, \mu$ are bias terms, and the sum runs only over *observed*
entries $\mathcal{K}$. The $K$ latent dimensions are discovered, not specified, and
often correspond to interpretable axes of taste. The regularizer is essential
because $\mathbf{R}$ is extremely sparse. <!-- CITE-NEEDED: matrix factorization for recommender systems (Koren, Bell, Volinsky) not in allowlist -->

Three structural problems define the research frontier and the deployment risk.
The **cold-start** problem—no data for new users or items—forces a fallback to
content features or popularity until interaction data accrue. **Feedback loops** are
subtler and more dangerous: a recommender trained on logged interactions learns from
data its own past recommendations generated, so popular items get recommended,
become more popular, and crowd out the long tail, narrowing exposure in ways that can
entrench rather than reveal preferences [@zheng2023recommending]. And recommendations
that maximize predicted clicks need not maximize *incremental* value: an item the
user would have bought anyway earns the recommender credit it did not create, a
confound between prediction and causal lift that only experimentation resolves. The
example below factorizes a small implicit-feedback matrix.

```{r aiml-recsys, message=FALSE, warning=FALSE}
set.seed(48)

# --- Simulate implicit feedback from latent tastes (rank-2 truth) ------------
n_users <- 200; n_items <- 60; K_true <- 2
P_true <- matrix(rnorm(n_users * K_true), n_users)
Q_true <- matrix(rnorm(n_items * K_true), n_items)
logits <- P_true %*% t(Q_true)
R <- matrix(rbinom(n_users * n_items, 1, plogis(logits)), n_users)  # 1 = engaged

# Hide 15% of entries as a test set (missing-at-random for illustration)
mask <- matrix(runif(length(R)) > 0.15, n_users)   # TRUE = observed in training
Rtr  <- R; Rtr[!mask] <- NA

# --- Matrix factorization by regularized alternating least squares -----------
K <- 2; lambda <- 0.1; iters <- 30
P <- matrix(rnorm(n_users * K, sd = 0.1), n_users)
Q <- matrix(rnorm(n_items * K, sd = 0.1), n_items)
solve_factor <- function(fixed, target_row, obs) {   # ridge solve per row
  F <- fixed[obs, , drop = FALSE]; y <- target_row[obs]
  solve(t(F) %*% F + lambda * diag(ncol(fixed)), t(F) %*% y)
}
for (it in seq_len(iters)) {
  for (u in seq_len(n_users)) { o <- which(!is.na(Rtr[u, ])); if (length(o)) P[u, ] <- solve_factor(Q, Rtr[u, ], o) }
  for (i in seq_len(n_items)) { o <- which(!is.na(Rtr[, i])); if (length(o)) Q[i, ] <- solve_factor(P, Rtr[, i], o) }
}
R_hat <- P %*% t(Q)

# Evaluate ranking quality on held-out entries via AUC
test_idx <- which(!mask)
auc_rank <- {
  s <- R_hat[test_idx]; y <- R[test_idx]
  pos <- s[y == 1]; neg <- s[y == 0]
  mean(outer(pos, neg, ">") + 0.5 * outer(pos, neg, "=="))
}
cat("Held-out ranking AUC:", round(auc_rank, 3), "\n")
```

## Deep Learning

*Deep learning* refers to neural networks with many layers of learned, nonlinear
transformations. A feedforward network composes affine maps and nonlinearities,
$$
\hat f(\mathbf{x}) = \sigma_L\!\big(\mathbf{W}_L\,\sigma_{L-1}(\cdots \sigma_1(\mathbf{W}_1\mathbf{x}+\mathbf{b}_1)\cdots)+\mathbf{b}_L\big),
$$ {#eq-aiml-dnn}
where each $\mathbf{W}_\ell$ is a learned weight matrix, $\mathbf{b}_\ell$ a bias,
and $\sigma_\ell$ an elementwise nonlinearity (commonly the rectified linear unit,
$\sigma(z)=\max(0,z)$). The parameters are fit by gradient descent on the loss in
@eq-aiml-erm, with gradients computed by *backpropagation*—the chain rule applied
layer by layer—and the data scanned in mini-batches (stochastic gradient descent).
Depth matters because composition lets the network build features from features:
early layers learn simple patterns, later layers compose them into abstractions, so
the network learns its own representation rather than relying on hand-engineered
features. <!-- CITE-NEEDED: foundational deep learning reference (LeCun, Bengio, Hinton) not in allowlist -->

The marketing payoff is largest where the data are *unstructured*—precisely the
domains where hand-engineering features is hopeless. Convolutional networks read
images, enabling brand-perception measurement directly from consumer-generated
photos at a scale and speed no survey could match: @liu2020 train a multi-label
convolutional network to detect perceptual brand attributes in user images,
recovering survey-consistent perceptions in near real time, and image content
systematically shapes engagement [@Li_2019]. Recurrent and, later, attention-based
architectures read sequences—text, clickstreams, purchase histories—turning the
unstructured trace of customer behavior into predictive features
[@martin2017data]. <!-- CITE-NEEDED: deep sequence/representation models for clickstream and text in marketing -->. The price of this expressive power is steep:
deep models are data-hungry, computationally expensive, prone to overfitting without
heavy regularization (dropout, early stopping, weight decay), and—most relevant to
this chapter—they are *opaque*, which makes them excellent predictors and poor
instruments of inference. The temptation to read a neural network's learned
representations as explanations is the deep-learning incarnation of the
prediction–inference confusion.

For tabular marketing data of modest size, it bears emphasizing that deep learning
usually does *not* beat gradient-boosted trees; the deep-learning advantage is
specific to large, unstructured, high-signal data. Choosing a neural network for a
50-feature churn table is a common and avoidable error.

## Large Language Models

*Large language models* (LLMs) are deep neural networks—almost always
*transformers*, built on the self-attention mechanism—trained on internet-scale text
to predict the next token, and then adapted to follow instructions. Their relevance
to marketing is twofold. As *measurement instruments*, they convert the field's vast
unstructured text—reviews, social posts, support transcripts, open-ended
survey responses—into structured variables: sentiment, topics, stance, entities,
and the latent dimensions of consumer voice that earlier text methods recovered more
laboriously [@netzer2008hidden; @buschken2016sentence]. As
*generators*, they produce marketing content—copy, product descriptions,
personalized email, synthetic chat—at near-zero marginal cost, reshaping the economics
of content production across the funnel [@appel2020future].

The transformer's core operation is *attention*, which lets each token's
representation be a weighted average of the others, with weights computed from
learned query, key, and value projections,
$$
\operatorname{Attention}(\mathbf{Q},\mathbf{K},\mathbf{V}) =
\operatorname{softmax}\!\left(\frac{\mathbf{Q}\mathbf{K}^{\top}}{\sqrt{d_k}}\right)\mathbf{V},
$$ {#eq-aiml-attn}
where $\mathbf{Q},\mathbf{K},\mathbf{V}$ are linear projections of the input sequence
and $d_k$ is the key dimension. This mechanism, stacked and scaled, is what lets the
model condition each word on the entire context. <!-- CITE-NEEDED: transformer architecture (Vaswani et al., "Attention Is All You Need") not in allowlist -->

Three properties demand discipline when LLMs are used as research instruments.
First, **hallucination**: an LLM optimizes for plausible continuations, not truth,
and will fabricate confidently; outputs used as data must be validated against ground
truth, ideally with a human-labeled audit sample and an inter-rater reliability
check (@sec-measurement-scales). Second, **non-determinism and version drift**: the
same prompt can yield different outputs, and the underlying model changes under the
analyst's feet, so a measurement pipeline built on a hosted LLM is not automatically
reproducible—prompts, model versions, and decoding parameters must be logged like any
other instrument. Third, **contamination and circularity**: an LLM trained on the
open web may have seen the very reviews or constructs under study, so "predicting"
them is not out-of-sample, and using an LLM both to generate and to evaluate content
risks a closed loop that measures the model's preferences rather than consumers'.
Used as instruments, LLMs are powerful but require the same validity scaffolding as
any measure; used as generators, they require the brand-safety, factuality, and
fairness controls developed in the rest of this chapter.

## Machine Learning Across the Marketing Funnel

The methods above are not siloed by funnel stage; the same supervised, unsupervised,
and generative tools recur, retargeted at different objectives. Organizing them by
the customer journey clarifies *what* is being predicted at each step and *where* the
prediction–inference distinction bites. @tbl-aiml-funnel maps the terrain, and the
recurring lesson is that most funnel applications are predictive scoring problems,
while the decisions they feed—how much to spend, what to charge, whom to
target—are causal questions that prediction alone cannot answer.

| Funnel stage | Representative task | Learning paradigm | Pred. vs. inf. | Anchor |
|---|---|---|---|---|
| Awareness | Audience look-alike modeling; media-mix forecasting | Supervised classification; sequence models | Prediction (lift is causal) | @wedel2016 |
| Consideration | Search/social listening; brand-perception mining | Unsupervised; LLM/text | Prediction (measurement) | @netzer2008; @liu2020 |
| Conversion | Propensity-to-buy; dynamic creative; recommendation | Supervised; recommender | Prediction; *targeting is causal* | @neumann2019frontiers |
| Retention | Churn prediction; next-best-action | Supervised classification | Prediction; *action is causal* | @neumann2019frontiers |
| Advocacy | Review/UGC analysis; influencer identification | Unsupervised; LLM/text | Prediction (measurement) | @tirunillai2014a |

: Machine learning across the marketing funnel. Most applications are predictive scoring tasks; the decisions they inform are causal. {#tbl-aiml-funnel}

The "prediction (lift is causal)" and "action is causal" entries flag the same trap
in different clothes. A look-alike model predicts who *resembles* a converter, but
whether advertising to them *causes* incremental conversion is a question only a
holdout experiment answers. A churn model predicts who *will* leave, but the right
retention action depends on who will leave *because of, or despite,* the
intervention—an uplift, not a prediction, problem. <!-- CITE-NEEDED: Ascarza (2018, JMR) on retention-targeting / uplift; churn-prediction-vs-uplift distinction --> The funnel view
is useful precisely because it keeps surfacing the boundary the chapter is built
around.

## Pitfalls: How Machine Learning Quietly Breaks

A model that scores well offline and fails in production is the rule, not the
exception, and the failures cluster into three families: *leakage*, which inflates
offline performance with information that will not exist at decision time; *drift*,
which erodes performance as the world moves away from the training distribution; and
*unfairness*, which encodes and amplifies inequities the data inherited. None of the
three is a coding bug; all three are violations of the assumptions behind
@eq-aiml-erm. @fig-aiml-lifecycle situates them in the model lifecycle.

```{mermaid}
%%| label: fig-aiml-lifecycle
%%| fig-cap: "Where the three pitfalls strike in the machine-learning lifecycle. Leakage corrupts training and validation before deployment; drift degrades a correct model after it; fairness harms accrue at the decision boundary and feed back into future data."
flowchart LR
  A["Data collection"] --> B["Feature engineering<br/>& training"]
  B --> C["Offline validation"]
  C --> D["Deployment &<br/>decisions"]
  D --> E["Monitoring"]
  E -.->|"new data"| A
  B -. "LEAKAGE<br/>(future/target info<br/>leaks into features)" .-> C
  D -. "DRIFT<br/>(world moves away<br/>from training dist.)" .-> E
  D -. "FAIRNESS<br/>(disparate harm at<br/>the decision boundary)" .-> A
```

### Data leakage

**Leakage** occurs when information unavailable at prediction time contaminates the
training data, so the model "cheats" offline and collapses in production. The signature
is an offline metric that is too good to be true. Leakage takes several forms, each
common in marketing pipelines. *Target leakage* includes a feature that is a proxy
for, or a consequence of, the outcome—predicting purchase using "added-to-cart in the
same session," which is nearly the outcome itself. *Temporal leakage* uses
information from after the prediction timestamp—computing a customer's average order
value over a window that includes the very order being predicted. *Preprocessing
leakage* is the most insidious because it survives a careful feature audit: when
scaling, imputation, feature selection, or target encoding is fit on the *full*
dataset before the train/test split, statistics from the test set bleed into
training, and cross-validation reports an accuracy the deployed model cannot reach.

The discipline that prevents leakage is to fit *every* data-dependent transformation
inside the cross-validation fold, on training data only, and to respect time: when
data are temporally ordered—as marketing data almost always are—validation must use a
forward-chaining split that trains on the past and tests on the future, never a random
split that lets the model peek ahead. The example contrasts the two and quantifies the
illusion.

```{r aiml-leakage, message=FALSE, warning=FALSE}
set.seed(48)

# Time-ordered data where the signal is non-stationary (drifts over time)
N <- 1500
time  <- seq_len(N)
beta_t <- 1.5 - 1.0 * (time / N)          # the effect of x decays over time
x <- rnorm(N)
y <- as.integer(plogis(-0.2 + beta_t * x + rnorm(N, 0, 0.5)) > 0.5)
df <- data.frame(time, x, y)

auc <- function(score, label) {
  pos <- score[label == 1]; neg <- score[label == 0]
  if (!length(pos) || !length(neg)) return(NA_real_)
  mean(outer(pos, neg, ">") + 0.5 * outer(pos, neg, "=="))
}

# (a) RANDOM split: leaks future into training on non-stationary data
ridx  <- sample(N, 0.7 * N)
m_rand <- glm(y ~ x, df[ridx, ], family = binomial())
auc_rand <- auc(predict(m_rand, df[-ridx, ], type = "response"), df[-ridx, ]$y)

# (b) FORWARD-CHAINING split: train on past, test on future (honest)
cut    <- floor(0.7 * N)
m_fwd  <- glm(y ~ x, df[df$time <= cut, ], family = binomial())
auc_fwd <- auc(predict(m_fwd, df[df$time > cut, ], type = "response"),
               df[df$time > cut, ]$y)

cat("AUC, random split (optimistic):  ", round(auc_rand, 3), "\n")
cat("AUC, forward-chaining (honest):  ", round(auc_fwd, 3), "\n")
```

The random-split AUC overstates the performance the model will actually deliver on
future data, because random splitting on a non-stationary process lets the model
borrow the future to predict the past. The gap between the two numbers is a direct
measure of self-deception, and it is invisible to anyone who validates with a random
split—the default in most tutorials.

### Distribution drift {#sec-aiml-drift}

Even a leakage-free model decays, because the assumption that deployment data share
the training distribution expires. **Drift** comes in two flavors that demand
different responses. *Covariate shift* changes the input distribution while the
relationship is stable, $p_{\text{test}}(\mathbf{x}) \neq p_{\text{train}}(\mathbf{x})$
but $p(y\mid\mathbf{x})$ unchanged—a new acquisition channel brings customers unlike
the training population. *Concept drift* changes the relationship itself,
$p_{\text{test}}(y\mid\mathbf{x}) \neq p_{\text{train}}(y\mid\mathbf{x})$—a recession,
a competitor's launch, or a pandemic rewrites how features map to behavior, exactly
the non-stationarity simulated above. Marketing is a near-worst case for drift
because the environment is adversarial and reflexive: competitors react, consumer
tastes move, and—uniquely—the model's *own actions* change the data it next sees, so a
targeting model trained on one policy's data is evaluated under another. The defenses
are monitoring (track input distributions and live performance against a holdout, and
alarm on divergence), scheduled or triggered retraining, and—where decisions feed back
into data—maintaining randomized holdouts so the model never fully determines its own
training distribution.

### Fairness and algorithmic bias

Machine-learning models inherit and can amplify the biases latent in their training
data, with legal and ethical force when marketing decisions touch credit, housing,
employment, insurance, or protected groups. The mechanism is not malice but
statistics: if historical data reflect discrimination, a model that faithfully
predicts the historical target reproduces the discrimination, and *omitting* a
protected attribute does not fix it, because correlated proxies (postal code, device,
browsing history) reconstruct the attribute—a phenomenon known as redundant encoding.

Fairness must therefore be defined, measured, and traded off explicitly, and a basic
impossibility result disciplines expectations: several intuitive fairness
criteria—equal false-positive and false-negative rates across groups (equalized odds)
versus equal calibration across groups versus statistical parity in selection
rates—*cannot in general all hold simultaneously* unless base rates are equal or the
classifier is perfect. There is no single "fair" model; there is a choice among
incompatible fairness criteria, and that choice is normative, not technical. The
worked example audits a scorer for *disparate impact*—whether selection rates differ
across groups—and shows that proxy features carry bias even when the protected
attribute is excluded. <!-- CITE-NEEDED: algorithmic fairness foundations (Barocas, Hardt, Narayanan; Kleinberg et al. impossibility) not in allowlist -->

```{r aiml-fairness, message=FALSE, warning=FALSE}
set.seed(48)

n <- 6000
group <- rbinom(n, 1, 0.5)                       # protected attribute A in {0,1}
# A proxy correlated with the protected group (e.g., neighborhood), NOT the label cause
proxy <- rnorm(n, mean = 0.8 * group)
quality <- rnorm(n)                              # the legitimately relevant signal
# True outcome depends ONLY on quality (group is irrelevant to merit)...
y <- rbinom(n, 1, plogis(1.2 * quality))
df <- data.frame(y, quality, proxy, group)

# Model that EXCLUDES the protected attribute but KEEPS the correlated proxy
fit <- glm(y ~ quality + proxy, df, family = binomial())
df$score <- predict(fit, type = "response")
tau <- quantile(df$score, 0.7)                   # select top 30%
df$selected <- as.integer(df$score > tau)

# Disparate impact: ratio of selection rates across groups (4/5ths rule -> >= 0.8)
sel_rate <- tapply(df$selected, df$group, mean)
di_ratio <- min(sel_rate) / max(sel_rate)
cat("Selection rate by group:", round(sel_rate, 3), "\n")
cat("Disparate-impact ratio:  ", round(di_ratio, 3),
    if (di_ratio < 0.8) " (FAILS 4/5ths rule)" else " (passes)", "\n")
```

The disparate-impact ratio falls below the conventional four-fifths threshold even
though the protected attribute never entered the model and plays no role in the true
outcome—the proxy alone manufactures the gap. The lesson generalizes: fairness is a
property of a model-in-context that must be measured on outcomes, not assumed from the
feature list, and remediation (reweighting, constrained optimization, post-hoc
threshold adjustment by group) is an explicit, auditable design choice with its own
trade-offs against accuracy and against other fairness criteria.

## From Prediction to Decision: Causal Machine Learning

The chapter's organizing distinction has a constructive resolution. The right way to
use machine learning's flexibility for *inference* is not to read coefficients off a
predictive model but to embed flexible learners inside an estimator whose
identification comes from design. *Double/debiased machine learning* uses
nonparametric learners to flexibly absorb high-dimensional confounders while
preserving valid inference on a low-dimensional causal parameter, via
orthogonalization and cross-fitting that immunize the target estimate against the
nuisance learner's regularization bias. *Causal forests* and related heterogeneous-
treatment-effect estimators adapt tree ensembles to estimate how a treatment effect
$\tau(\mathbf{x}) = \mathbb{E}[Y(1) - Y(0)\mid \mathbf{X}=\mathbf{x}]$ varies across
customers—the *uplift* that retention and targeting decisions actually require, as
distinct from the *level* a churn model predicts. These methods, developed in the
causal-inference chapters, are the principled bridge from machine learning's
predictive power to the causal questions marketing decisions pose, and they are where
the field's frontier is moving [@varian2016]. <!-- CITE-NEEDED: double/debiased ML (Chernozhukov et al.) and causal forests (Wager & Athey; Athey & Imbens) not in allowlist -->

## Key Takeaways

- The **prediction–inference distinction** (@eq-aiml-dgp) is the organizing idea:
  predictive accuracy and causal interpretation are different goals validated by
  different criteria, and a model's internals are not causal estimates no matter how
  well it predicts.
- **Supervised learning** minimizes regularized empirical risk (@eq-aiml-erm); for
  tabular marketing data, gradient-boosted trees are the empirical default, and honest
  evaluation requires *both* discrimination and calibration, on a split that respects
  time.
- **Unsupervised learning** discovers structure without labels; clustering always
  returns clusters, so the burden of proof is on the analyst to show segments are
  stable and managerially meaningful, not merely that the algorithm converged.
- **Recommender systems** turn sparse user–item data into rankings via matrix
  factorization (@eq-aiml-mf), but cold-start, feedback loops, and the
  prediction-versus-incremental-lift gap make naïve click-maximization a trap.
- **Deep learning and LLMs** excel on unstructured data (images, text, sequences) and
  are powerful measurement and generation instruments, but their opacity makes them
  poor instruments of inference and demands explicit validity, reproducibility, and
  factuality controls.
- The three deployment pitfalls—**leakage**, **drift**, and **unfairness**—are
  assumption violations, not bugs: validate inside the fold and forward in time,
  monitor and retrain against a moving world, and measure fairness on outcomes because
  proxies encode protected attributes even when those attributes are excluded.
- The constructive resolution is **causal machine learning**, which embeds flexible
  learners inside design-based estimators to recover the *uplift* that marketing
  decisions require.
