# Artificial Intelligence and Machine Learning in Marketing {#sec-ai-ml}

*Machine learning* is the study of algorithms that improve their performance on a
task as they are exposed to more data, rather than by being explicitly programmed
with the rules of that task. *Artificial intelligence* (AI) is the broader project
of building systems that perform tasks we associate with human cognition;
contemporary AI in marketing is, almost entirely, applied machine learning at
scale. The distinction matters less than what the two share: a willingness to let
flexible, high-capacity function approximators discover structure in data that a
human analyst could neither specify in advance nor write down as equations. That
willingness is also the source of every pitfall in this chapter.

Marketing is an unusually hospitable host for these methods, for three reasons.
First, the field is awash in the kind of data—clickstreams, transactions, images,
reviews, search queries—on which modern learning algorithms thrive
[@wedel2016; @martin2017data]. Second, many marketing tasks are natively
*predictive*: which customer will churn, which creative will earn the click, which
product to surface next. Prediction is exactly what supervised learning does well.
Third, the economic stakes of small improvements are large, because marketing
decisions are made billions of times a day across recommendation, bidding, and
targeting systems, so a one-percent lift compounds into real money
[@varian2016; @gao2024big].

This chapter has two jobs. The first is to give a working, formal command of the
methods—supervised and unsupervised learning, recommender systems, deep learning,
and large language models—at a level that lets the reader state each method's
estimator, its assumptions, and the conditions under which it fails. The second is
to install a discipline that the hype around AI actively erodes: the distinction
between **prediction** and **inference**, and with it a sober catalogue of the ways
machine learning quietly breaks in deployment—leakage, distribution drift, and
unfairness. A model that predicts well in a notebook and harms the business in
production is the modal failure, not the exception, and most of those failures are
conceptual rather than computational. By the end, the reader should be able to map
a marketing problem onto the right learning paradigm, build and validate a model
that does not lie to them, and recognize when a predictive tool is being asked,
illegitimately, to answer a causal question.

We assume familiarity with the regression and choice-modeling machinery developed
earlier in the book (@sec-measurement-scales for measurement; the causal-inference
and marketing-mix chapters for identification), and we connect to them rather than
re-deriving them.

## Prediction Versus Inference: The Organizing Distinction

The single most consequential idea in this chapter is also the one most often
skipped. Consider a generic supervised relationship between an outcome $Y$ and
features $\mathbf{X}$,
$$
Y = f(\mathbf{X}) + \varepsilon, \qquad \mathbb{E}[\varepsilon \mid \mathbf{X}] = 0,
$$ {#eq-aiml-dgp}
with $\hat f$ an estimate learned from data. There are two fundamentally different
things one can want from $\hat f$, and conflating them is the root of most misuse
of machine learning in marketing.

> In a **prediction** problem the object of interest is $\hat f(\mathbf{x})$ itself:
> we want accurate values of $Y$ at new inputs and treat $\hat f$ as a black box. In
> an **inference** problem the object of interest is some functional of $f$—a
> coefficient, an elasticity, a treatment effect—and $\hat f$ is a means to learn
> *how* $\mathbf{X}$ relates to $Y$, not merely to forecast $Y$.

The two goals reward different choices. Prediction tolerates—indeed often
prefers—biased, uninterpretable, highly flexible estimators, because the only
scorecard is out-of-sample loss, $\mathbb{E}[L(Y, \hat f(\mathbf{X}))]$ on data the
model has never seen. The bias–variance trade-off is the governing law: expected
squared prediction error decomposes as
$$
\mathbb{E}\!\left[(Y - \hat f(\mathbf{x}_0))^2\right]
= \underbrace{\big(\mathbb{E}[\hat f(\mathbf{x}_0)] - f(\mathbf{x}_0)\big)^2}_{\text{bias}^2}
+ \underbrace{\operatorname{Var}\!\big(\hat f(\mathbf{x}_0)\big)}_{\text{variance}}
+ \underbrace{\sigma^2_\varepsilon}_{\text{irreducible}},
$$ {#eq-aiml-bv}
and a method that accepts some bias to cut variance can win. Inference, by contrast,
demands an estimator whose sampling distribution we understand—unbiasedness or a
known bias, valid standard errors, an identification argument linking the estimand
to features of the data-generating process. A random forest can have lower test
error than a linear regression while being *useless* for the question "what is the
effect of a \$1 price cut," because flexible learners trade interpretable,
consistent parameters for predictive accuracy [@varian2016].

The practical hazard is that a predictive model's coefficients—or its
feature-importance scores—look like answers to inference questions and are routinely
read as such. They are not. A churn model may load heavily on "number of support
tickets," but acting on that association by *suppressing* support tickets would be
disastrous; the feature predicts churn because both are caused by underlying
dissatisfaction. Predictive importance is not causal importance, and no amount of
test-set accuracy converts one into the other. When the marketing question is "what
will happen," supervised learning is the right tool; when it is "what should we do,"
the model must be embedded in a design that identifies a causal effect—a randomized
experiment, an instrument, or one of the causal-machine-learning estimators we reach
at the end of the chapter. @fig-aiml-pred-inf fixes the fork in the road.

```{mermaid}
%%| label: fig-aiml-pred-inf
%%| fig-cap: "The prediction–inference fork. The same data and even the same algorithm serve different goals, validated by different criteria. Reading a predictive model's internals as causal estimates is the central error this chapter warns against."
flowchart TD
  Q["Marketing question"] --> D{"What is the<br/>object of interest?"}
  D -->|"Accurate Y at new inputs"| P["PREDICTION<br/>(black-box f-hat)"]
  D -->|"A functional of f:<br/>effect, elasticity"| I["INFERENCE<br/>(structure of f)"]
  P --> PV["Validate by:<br/>out-of-sample loss,<br/>cross-validation"]
  I --> IV["Validate by:<br/>identification,<br/>standard errors, design"]
  PV --> PU["Use: scoring, ranking,<br/>forecasting, matching"]
  IV --> IU["Use: pricing, budget<br/>allocation, policy"]
  PU -.->|"Do NOT read feature<br/>importance as causal"| IU
```

## Supervised Learning

In **supervised learning** the training data are labeled pairs
$\{(\mathbf{x}_i, y_i)\}_{i=1}^n$, and the goal is to learn a mapping
$\hat f : \mathcal{X} \to \mathcal{Y}$ that generalizes to unlabeled inputs. When
$Y$ is continuous the task is *regression*; when $Y$ is categorical it is
*classification*. Almost every workhorse marketing model—propensity to buy, churn,
response, lifetime-value, lead scoring, ad click-through—is a supervised classifier
or regressor.

### The learning problem and regularization

Formally, learning chooses $\hat f$ to minimize *regularized empirical risk*,
$$
\hat f = \arg\min_{f \in \mathcal{F}} \; \frac{1}{n}\sum_{i=1}^{n} L\big(y_i, f(\mathbf{x}_i)\big) + \lambda\, \Omega(f),
$$ {#eq-aiml-erm}
where $L$ is a loss function (squared error for regression, log-loss/cross-entropy
for classification), $\mathcal{F}$ is the hypothesis class (linear functions, trees,
neural networks), $\Omega(f)$ is a complexity penalty, and $\lambda \ge 0$ tunes the
trade-off. Minimizing training loss alone ($\lambda = 0$, $\mathcal{F}$ rich) yields
*overfitting*: $\hat f$ memorizes noise and generalizes poorly, the high-variance
failure in @eq-aiml-bv. The penalty $\Omega$ buys generalization by shrinking the
effective complexity of $\hat f$. For linear models the two canonical choices are
the $\ell_2$ (ridge) penalty $\Omega(\boldsymbol\beta)=\|\boldsymbol\beta\|_2^2$,
which shrinks coefficients smoothly, and the $\ell_1$ (lasso) penalty
$\Omega(\boldsymbol\beta)=\|\boldsymbol\beta\|_1$, which sets some coefficients
exactly to zero and thereby performs variable selection—valuable when the feature
space is wide, as it almost always is with behavioral data [@varian2016].

The assumptions behind @eq-aiml-erm are easy to state and easy to violate. Empirical
risk minimization is consistent for the risk-minimizing $f$ only if the training
data are drawn from the *same distribution* as the deployment data (the
identically-distributed assumption) and if observations are *exchangeable* in the
way the validation scheme assumes (commonly independence). Both assumptions fail
routinely in marketing: deployment data drift away from training data over time
(@sec-aiml-drift), and observations are correlated within customers, sessions, and
time, which—if ignored—makes naïve cross-validation report accuracy the model will
never achieve in production.

### Tree ensembles: the marketing workhorse

For tabular marketing data—mixed numeric and categorical features, nonlinearities,
interactions, missingness—ensembles of decision trees are, empirically, the default
high-performer. A single regression tree partitions $\mathcal{X}$ into rectangular
regions $R_1,\dots,R_M$ and predicts the within-region mean, $\hat
f(\mathbf{x})=\sum_m c_m \mathbf{1}\{\mathbf{x}\in R_m\}$; it is interpretable but
high-variance. Two ensemble strategies tame the variance. **Random forests** average
many trees grown on bootstrap samples with randomly restricted split candidates,
reducing variance through decorrelation. **Gradient-boosted trees** instead fit
trees *sequentially*, each new tree $h_t$ targeting the gradient of the loss left by
the running ensemble,
$$
\hat f_t(\mathbf{x}) = \hat f_{t-1}(\mathbf{x}) + \nu\, h_t(\mathbf{x}), \qquad
h_t \approx -\,\frac{\partial L}{\partial \hat f_{t-1}},
$$ {#eq-aiml-boost}
with learning rate $\nu \in (0,1]$. Boosting reduces bias and variance jointly and
typically tops leaderboards on tabular data, at the cost of more careful tuning to
avoid overfitting (the number of trees becomes a regularization parameter set by
validation). The worked example below builds a churn classifier and—critically—shows
how to validate it honestly. The two workhorses of that toolkit are @breiman2001randomforests, which averages decorrelated trees to cut variance, and @friedman2001greedyfunction, which fits them sequentially to the residual and so cuts bias instead.

```{r aiml-churn, message=FALSE, warning=FALSE}
set.seed(48)

# --- Simulate a customer-churn dataset with a known structure ----------------
n <- 4000
tenure        <- rpois(n, lambda = 18)                       # months as customer
monthly_spend <- round(rgamma(n, shape = 2, scale = 25), 2)  # $ per month
support_calls <- rpois(n, lambda = 1 + 0.05 * (40 - pmin(tenure, 40)))
discount_user <- rbinom(n, 1, 0.35)

# True churn propensity: short tenure, low spend, many support calls raise risk.
lin <- -1.0 - 0.06 * tenure - 0.015 * monthly_spend +
        0.45 * support_calls + 0.30 * discount_user
prob_churn <- plogis(lin)
churn <- rbinom(n, 1, prob_churn)

dat <- data.frame(churn = factor(churn, labels = c("stay", "leave")),
                  tenure, monthly_spend, support_calls,
                  discount_user = factor(discount_user))

# --- Honest train/test split -------------------------------------------------
idx   <- sample(seq_len(n), size = floor(0.7 * n))
train <- dat[idx, ]
test  <- dat[-idx, ]

# --- Gradient-boosted trees (gbm); fall back to logistic if gbm absent -------
has_gbm <- requireNamespace("gbm", quietly = TRUE)
if (has_gbm) {
  fit <- gbm::gbm(I(as.integer(churn) - 1) ~ tenure + monthly_spend +
                    support_calls + discount_user,
                  data = train, distribution = "bernoulli",
                  n.trees = 600, interaction.depth = 3,
                  shrinkage = 0.03, bag.fraction = 0.7, verbose = FALSE)
  best <- gbm::gbm.perf(fit, plot.it = FALSE, method = "OOB")
  p_hat <- gbm::predict.gbm(fit, test, n.trees = best, type = "response")
} else {
  fit   <- glm(churn ~ tenure + monthly_spend + support_calls + discount_user,
               data = train, family = binomial())
  p_hat <- predict(fit, test, type = "response")
}

# --- Out-of-sample evaluation: AUC and a calibration check -------------------
auc <- function(score, label) {              # rank-based AUC, no extra packages
  pos <- score[label == "leave"]; neg <- score[label == "stay"]
  mean(outer(pos, neg, ">") + 0.5 * outer(pos, neg, "=="))
}
cat("Test AUC:", round(auc(p_hat, test$churn), 3), "\n")

# Calibration: do predicted probabilities match realized churn rates?
bins <- cut(p_hat, breaks = quantile(p_hat, 0:5/5), include.lowest = TRUE)
calib <- aggregate(as.integer(test$churn) - 1 ~ bins, FUN = mean)
calib$predicted <- tapply(p_hat, bins, mean)
names(calib) <- c("bin", "observed_churn", "mean_predicted")
calib
```

The example reports two diagnostics, not one. *Discrimination* (AUC) measures
whether the model ranks churners above non-churners; *calibration* measures whether
a predicted 30% churn probability corresponds to a 30% realized rate. A model can
discriminate well yet be badly calibrated, and marketing decisions that multiply
predicted probabilities by margins—expected-value targeting—need calibration, not
just ranking [@neumann2019frontiers]. Reporting only AUC is a common and costly
omission.

### Classification thresholds and the cost of errors

A classifier outputs a score $\hat p(\mathbf{x}) = \widehat{\Pr}(Y=1\mid\mathbf{x})$;
turning it into an action requires a threshold $\tau$ such that we treat customers
with $\hat p > \tau$. The optimal threshold is *not* $0.5$—it depends on the
asymmetric costs of false positives and false negatives. If contacting a non-churner
costs $c_{\text{FP}}$ and failing to retain a churner costs $c_{\text{FN}}$, the
expected-cost-minimizing rule acts when
$\hat p / (1-\hat p) > c_{\text{FP}} / c_{\text{FN}}$. This is the point at which the
predictive model meets the decision problem, and it is where the marketing economics
re-enter: the ROC and precision–recall curves exist precisely because the right
operating point is a business choice, not a statistical default.

## Unsupervised Learning

In **unsupervised learning** the data are unlabeled, $\{\mathbf{x}_i\}_{i=1}^n$, and
the goal is to discover latent structure—groups, dimensions, topics—without a target
variable to supervise the search. The two dominant marketing uses are
*segmentation* (clustering customers or products) and *dimension reduction*
(compressing high-dimensional behavior into interpretable factors). Because there is
no label, there is no test-set accuracy to adjudicate "correctness"; validation is
intrinsically harder and more judgmental, which is both the method's flexibility and
its danger.

### Clustering and the segmentation problem

The canonical objective is *$k$-means*, which partitions observations into $K$
clusters to minimize within-cluster squared distance,
$$
\min_{\{S_k\}}\; \sum_{k=1}^{K}\sum_{\mathbf{x}_i \in S_k} \big\|\mathbf{x}_i - \boldsymbol\mu_k\big\|_2^2,
\qquad \boldsymbol\mu_k = \frac{1}{|S_k|}\sum_{\mathbf{x}_i \in S_k}\mathbf{x}_i.
$$ {#eq-aiml-kmeans}
This connects directly to the *a priori* versus *post hoc* segmentation distinction
of @sec-segmentation: post hoc segmentation is precisely the application of
a clustering algorithm to behavioral or attitudinal data to *discover* segments
rather than impose them. The estimator (Lloyd's algorithm) alternates assigning
points to the nearest centroid and recomputing centroids; it converges to a *local*
optimum, so results depend on initialization and on the (analyst-chosen) number of
clusters $K$. Three assumptions break identification of a "true" segmentation, and
all three are routinely violated in practice: $k$-means presumes roughly spherical,
equal-variance clusters (Euclidean distance encodes this); it is not scale-invariant,
so features must be standardized or the largest-variance feature dominates; and $K$
is not learned but assumed. Model-based clustering via finite mixtures replaces the
hard geometry with a probabilistic generative model and lets information criteria
choose $K$, at the cost of distributional assumptions. The deeper caution is that a
clustering algorithm *always* returns clusters, whether or not the population is
actually clustered; the burden is on the analyst to show the segments are stable,
managerially distinguishable, and reproducible out of sample, not merely that the
algorithm ran.

```{r aiml-segmentation, message=FALSE, warning=FALSE}
set.seed(48)

# Three latent customer segments differing in recency, frequency, monetary value
make_seg <- function(n, r, f, m) data.frame(
  recency   = pmax(1, round(rnorm(n, r, 8))),
  frequency = pmax(1, round(rnorm(n, f, 3))),
  monetary  = pmax(5, round(rnorm(n, m, 40)))
)
rfm <- rbind(make_seg(300, 10, 14,  220),   # champions
             make_seg(300, 45,  4,   60),   # at-risk
             make_seg(300, 25,  8,  130))   # mainstream

# Standardize before clustering: k-means is NOT scale-invariant
rfm_z <- scale(rfm)
km <- kmeans(rfm_z, centers = 3, nstart = 25)

# Profile the recovered segments on the original (interpretable) scale
prof <- aggregate(rfm, by = list(segment = km$cluster), FUN = function(x) round(mean(x), 1))
prof$size <- as.integer(table(km$cluster))
prof
```

### Dimension reduction

When behavior is high-dimensional—thousands of SKUs, pages, or features—*dimension
reduction* finds a low-dimensional representation that preserves the information that
matters. Principal component analysis (PCA) projects $\mathbf{X}$ onto the orthogonal
directions of maximal variance, the leading eigenvectors of the covariance matrix;
the first few components often capture interpretable axes of behavior (e.g., overall
intensity, then category mix). Non-negative matrix factorization and, for
text, *topic models* such as latent Dirichlet allocation generalize the idea to
parts-based and probabilistic decompositions [@tirunillai2014a; @buschken2016sentence].
The methods double as a feature-engineering step for supervised models and as a
listening tool: factorizing the term–document matrix of online reviews recovers the
latent dimensions of quality consumers actually discuss, a structure managers cannot
specify in advance [@tirunillai2014a; @netzer2008].

## Recommender Systems

Recommender systems are the most economically consequential deployment of machine
learning in marketing: they choose which of millions of items to surface to each
user, and on platforms from retail to streaming they drive a large share of demand.
Formally, a recommender estimates a *utility* or *preference* score
$\hat r_{ui}$ for each user $u$ and item $i$, then ranks items by that score. The
data are a sparse user–item matrix $\mathbf{R}$, mostly missing, with observed
entries being ratings, clicks, or purchases.

Two paradigms, with a hybrid, organize the field. **Content-based** filtering
recommends items similar to those a user has liked, using item features; it handles
new items but cannot discover tastes outside a user's history. **Collaborative
filtering** ignores item content and exploits the wisdom of the crowd—users who
agreed in the past will agree in the future—and is the more powerful approach when
interaction data are dense. The dominant collaborative formulation is *matrix
factorization*, which embeds users and items in a shared latent space of dimension
$K$ and models preference as an inner product,
$$
\hat r_{ui} = \mathbf{p}_u^{\top}\mathbf{q}_i + b_u + b_i + \mu,
\qquad
\min_{\mathbf{P},\mathbf{Q},\mathbf{b}} \sum_{(u,i)\in\mathcal{K}}
\big(r_{ui} - \hat r_{ui}\big)^2 + \lambda\big(\|\mathbf{p}_u\|^2 + \|\mathbf{q}_i\|^2 + b_u^2 + b_i^2\big),
$$ {#eq-aiml-mf}
where $\mathbf{p}_u, \mathbf{q}_i \in \mathbb{R}^K$ are the learned user and item
factors, $b_u, b_i, \mu$ are bias terms, and the sum runs only over *observed*
entries $\mathcal{K}$. The $K$ latent dimensions are discovered, not specified, and
often correspond to interpretable axes of taste. The regularizer is essential
because $\mathbf{R}$ is extremely sparse. The standard remedy is to factor the sparse matrix into low-rank user and item embeddings, the approach that won the Netflix Prize and remains the baseline against which newer recommenders are measured [@koren2009matrixfactoriza].

Three structural problems define the research frontier and the deployment risk.
The **cold-start** problem—no data for new users or items—forces a fallback to
content features or popularity until interaction data accrue. **Feedback loops** are
subtler and more dangerous: a recommender trained on logged interactions learns from
data its own past recommendations generated, so popular items get recommended,
become more popular, and crowd out the long tail, narrowing exposure in ways that can
entrench rather than reveal preferences [@zheng2023recommending]. And recommendations
that maximize predicted clicks need not maximize *incremental* value: an item the
user would have bought anyway earns the recommender credit it did not create, a
confound between prediction and causal lift that only experimentation resolves. The
example below factorizes a small implicit-feedback matrix.

```{r aiml-recsys, message=FALSE, warning=FALSE}
set.seed(48)

# --- Simulate implicit feedback from latent tastes (rank-2 truth) ------------
n_users <- 200; n_items <- 60; K_true <- 2
P_true <- matrix(rnorm(n_users * K_true), n_users)
Q_true <- matrix(rnorm(n_items * K_true), n_items)
logits <- P_true %*% t(Q_true)
R <- matrix(rbinom(n_users * n_items, 1, plogis(logits)), n_users)  # 1 = engaged

# Hide 15% of entries as a test set (missing-at-random for illustration)
mask <- matrix(runif(length(R)) > 0.15, n_users)   # TRUE = observed in training
Rtr  <- R; Rtr[!mask] <- NA

# --- Matrix factorization by regularized alternating least squares -----------
K <- 2; lambda <- 0.1; iters <- 30
P <- matrix(rnorm(n_users * K, sd = 0.1), n_users)
Q <- matrix(rnorm(n_items * K, sd = 0.1), n_items)
solve_factor <- function(fixed, target_row, obs) {   # ridge solve per row
  F <- fixed[obs, , drop = FALSE]; y <- target_row[obs]
  solve(t(F) %*% F + lambda * diag(ncol(fixed)), t(F) %*% y)
}
for (it in seq_len(iters)) {
  for (u in seq_len(n_users)) { o <- which(!is.na(Rtr[u, ])); if (length(o)) P[u, ] <- solve_factor(Q, Rtr[u, ], o) }
  for (i in seq_len(n_items)) { o <- which(!is.na(Rtr[, i])); if (length(o)) Q[i, ] <- solve_factor(P, Rtr[, i], o) }
}
R_hat <- P %*% t(Q)

# Evaluate ranking quality on held-out entries via AUC
test_idx <- which(!mask)
auc_rank <- {
  s <- R_hat[test_idx]; y <- R[test_idx]
  pos <- s[y == 1]; neg <- s[y == 0]
  mean(outer(pos, neg, ">") + 0.5 * outer(pos, neg, "=="))
}
cat("Held-out ranking AUC:", round(auc_rank, 3), "\n")
```

### Bundle Recommendation at Scale {#sec-bundle-rec}

Matrix factorization scores single items. A distinct and commercially large problem is
recommending **bundles**—sets of products to buy together—which turns on a relationship
factorization ignores: whether two products are **complements** (a phone and a case,
bought together) or **substitutes** (two phones, bought instead of each other). Raw
co-occurrence cannot tell them apart, because substitutes co-occur too—shoppers view and
compare them—so a naive "frequently seen together" recommender fills bundles with
substitutes, which no one buys as a pair.

@kumar2026bundle solve this at scale with **dense product embeddings**. Products are
embedded from behavioral co-occurrence—purchases and clickstream consideration sets—so
that products used in similar ways land near one another, exactly as words do in a
language model. The geometry then supplies the two signals a bundle needs:
*substitutability* is proximity in the consideration-set space (interchangeable products
are considered together), while *complementarity* is co-purchase association net of
substitutability (products bought together that are not merely alternatives). Bundling a
focal product with its complements while screening out its substitutes, and then using a
field experiment to **learn the recommendation policy offline**—mapping bundle features
to purchase likelihood and optimizing over them (@sec-field-experiments-seminar)—their
policy generalizes across the retailer's assortment and raises bundle purchases by an
estimated 35% over the incumbent.

The replication below runs the whole pipeline in miniature. It simulates purchase
baskets (complements) and consideration sets (substitutes), learns product embeddings by
the standard PPMI-plus-SVD route—the closed-form cousin of word2vec—separates
complementarity from substitutability, and shows a bundle policy built on that separation
beating a naive co-occurrence recommender; it then tunes the substitute penalty by
offline policy evaluation, the paper's optimization step in miniature.

```{r aiml-bundle, message=FALSE, warning=FALSE}
set.seed(48)

# --- Catalog with latent complement/substitute structure ---------------------
n_prod <- 120; n_cat <- 12
cat_id <- rep(seq_len(n_cat), length.out = n_prod)      # each product's category
role   <- matrix(rnorm(n_prod * 3), n_prod, 3)          # latent "goes-with" role

# --- Baskets: complements co-occur (cross-category), with some substitute
#     contamination (people sometimes buy similar items too) -------------------
complement_of <- function(i) {
  cross <- which(cat_id != cat_id[i])
  aff   <- as.numeric(role[cross, ] %*% role[i, ])
  cross[sample.int(length(cross), 1, prob = {p <- exp(2 * (aff - max(aff))); p / sum(p)})]
}
substitute_of <- function(i) {
  same <- setdiff(which(cat_id == cat_id[i]), i)
  if (!length(same)) i else same[sample.int(length(same), 1)]
}
make_basket <- function() {
  items <- i <- sample.int(n_prod, 1)
  for (r in seq_len(rbinom(1, 3, 0.6))) items <- c(items, complement_of(i))
  if (runif(1) < 0.25) items <- c(items, substitute_of(i))    # contamination
  unique(items)
}
make_view <- function() {                                # consideration set: substitutes
  pool <- which(cat_id == sample.int(n_cat, 1))
  if (length(pool) < 2) pool else sample(pool, min(length(pool), 2 + rbinom(1, 3, 0.5)))
}
baskets <- replicate(6000, make_basket(), simplify = FALSE)
views   <- replicate(6000, make_view(),   simplify = FALSE)

# --- Co-occurrence -> PPMI -> dense embeddings (SVD, the word2vec cousin) -----
co_matrix <- function(sets, n) {
  M <- matrix(0, n, n)
  for (s in sets) if (length(s) > 1) M[s, s] <- M[s, s] + 1
  diag(M) <- 0; M
}
ppmi <- function(M) {
  tot <- sum(M); if (tot == 0) return(M)
  P <- log((M / tot + 1e-12) / (outer(rowSums(M), colSums(M)) / tot^2 + 1e-12))
  P[P < 0 | !is.finite(P)] <- 0; P
}
embed  <- function(M, k = 16) { sv <- svd(ppmi(M)); sv$u[, 1:k] %*% diag(sqrt(sv$d[1:k])) }
cosine <- function(E) { En <- E / sqrt(rowSums(E^2) + 1e-12); En %*% t(En) }

Cpur  <- co_matrix(baskets, n_prod)      # co-purchase (complementarity signal)
Cview <- co_matrix(views,   n_prod)      # co-view     (substitutability signal)
Craw  <- Cpur + Cview                    # what a naive co-occurrence recommender sees
Comp  <- ppmi(Cpur)                      # complementarity = co-purchase association
Sub   <- cosine(embed(Cview))            # substitutability = co-view embedding similarity

# --- Policies: naive co-occurrence vs. complement-minus-substitute -----------
naive    <- function(i, k = 3) { s <- Craw[i, ]; s[i] <- -Inf; order(s, decreasing = TRUE)[1:k] }
policy_w <- function(w) function(i, k = 3) {
  s <- Comp[i, ] * (1 - w * pmax(Sub[i, ], 0)); s[i] <- -Inf
  order(s, decreasing = TRUE)[1:k]
}

# --- Ground-truth attach & offline policy evaluation -------------------------
true_attach <- function(i, j) plogis(1.5 * sum(role[i, ] * role[j, ]) -
                                       1.2 * (cat_id[i] == cat_id[j]))
eval_policy <- function(policy, k = 3)
  mean(vapply(seq_len(n_prod),
              function(i) mean(vapply(policy(i, k), function(j) true_attach(i, j), 0)), 0))

# --- Offline policy learning: tune the substitute penalty w ------------------
ws     <- seq(0, 1.5, by = 0.25)
vals   <- vapply(ws, function(w) eval_policy(policy_w(w)), 0)
w_star <- ws[which.max(vals)]
base   <- eval_policy(naive)

cat(sprintf("Naive co-occurrence attach         : %.3f\n", base))
cat(sprintf("Embedding policy, no penalty (w=0)  : %.3f\n", vals[1]))
cat(sprintf("Optimized policy  w* = %.2f         : %.3f\n", w_star, max(vals)))
cat(sprintf("Lift of optimized policy over naive : %+.0f%%\n", 100 * (max(vals) - base) / base))
```

The naive recommender, ranking by raw co-occurrence, fills bundles with the very
substitutes that pollute that signal; switching to a purchase-based complementarity score
and then penalizing embedded substitutes lifts expected attach sharply. The magnitude in
this stylized market runs well above the field experiment's 35%—the simulation makes the
naive baseline deliberately poor—but the direction and, more to the point, the *reason*
are the paper's: the value is in separating complements from substitutes, which the
embedding geometry does almost for free, and in learning the policy from an experiment
rather than assuming it.

## Deep Learning

*Deep learning* refers to neural networks with many layers of learned, nonlinear
transformations. A feedforward network composes affine maps and nonlinearities,
$$
\hat f(\mathbf{x}) = \sigma_L\!\big(\mathbf{W}_L\,\sigma_{L-1}(\cdots \sigma_1(\mathbf{W}_1\mathbf{x}+\mathbf{b}_1)\cdots)+\mathbf{b}_L\big),
$$ {#eq-aiml-dnn}
where each $\mathbf{W}_\ell$ is a learned weight matrix, $\mathbf{b}_\ell$ a bias,
and $\sigma_\ell$ an elementwise nonlinearity (commonly the rectified linear unit,
$\sigma(z)=\max(0,z)$). The parameters are fit by gradient descent on the loss in
@eq-aiml-erm, with gradients computed by *backpropagation*—the chain rule applied
layer by layer—and the data scanned in mini-batches (stochastic gradient descent).
Depth matters because composition lets the network build features from features:
early layers learn simple patterns, later layers compose them into abstractions, so
the network learns its own representation rather than relying on hand-engineered
features. The general case for learned rather than hand-engineered representations is set out in @lecun2015deeplearning.

The marketing payoff is largest where the data are *unstructured*—precisely the
domains where hand-engineering features is hopeless. Convolutional networks read
images, enabling brand-perception measurement directly from consumer-generated
photos at a scale and speed no survey could match: @liu2020 train a multi-label
convolutional network to detect perceptual brand attributes in user images,
recovering survey-consistent perceptions in near real time, and image content
systematically shapes engagement [@Li_2019]. Recurrent and, later, attention-based
architectures read sequences—text, clickstreams, purchase histories—turning the
unstructured trace of customer behavior into predictive features
[@martin2017data]. The price of this expressive power is steep:
deep models are data-hungry, computationally expensive, prone to overfitting without
heavy regularization (dropout, early stopping, weight decay), and—most relevant to
this chapter—they are *opaque*, which makes them excellent predictors and poor
instruments of inference. The temptation to read a neural network's learned
representations as explanations is the deep-learning incarnation of the
prediction–inference confusion.

For tabular marketing data of modest size, it bears emphasizing that deep learning
usually does *not* beat gradient-boosted trees; the deep-learning advantage is
specific to large, unstructured, high-signal data. Choosing a neural network for a
50-feature churn table is a common and avoidable error.

### Representations as Measurement Instruments {#sec-aiml-representation-measurement}

The warning just issued—that learned representations make poor instruments of
inference—needs a companion, because a growing and distinctly different line of work
uses the representation as the *measurement itself*, and does so legitimately. The
distinction is worth stating precisely. Reading a network's internal activations as an
*explanation* of why it predicted what it did is the error. Constructing an embedding
space whose **geometry operationalizes a construct the researcher has defined in
advance**, then validating that geometry against outcomes, is measurement, and it is
subject to the ordinary rules of construct validity (@sec-measurement-scales) rather
than to the interpretability critique.

Two recent studies show the pattern and its payoff. @sozuer2026recipe embed recipe
ingredients from co-occurrence across more than 57,000 recipes and read two
prespecified quantities out of the resulting geometry—the internal **coherence** among
an idea's components and the **uniqueness** of those components for the category—then
show that coherence raises popularity, trial, and post-trial evaluation, while unique
ingredients *lower* trial but *raise* ratings among those who try. The construct is
creativity; the embedding is the instrument; and the validation is that the measure
predicts outcomes it was not fit to. @sikdar2026artwork fuse structured listing
attributes, images, and text to price novice artwork on Etsy, and pair the price model
with a time-to-sale hazard model so that the output is a decision rather than a
coefficient (@sec-mmf-artpricing).

Three features recur and are worth naming as a template. The construct is **defined
before the model is fit**, so the embedding is asked a specific question rather than
mined for whatever it will yield. The measure is **validated against downstream
outcomes** that were not part of its construction. And both studies end not in a table
of coefficients but in a **tool**—a generative recipe assistant, a price
recommender—that hands a decision-maker a concrete action. That last move is what most
distinguishes this line from conventional predictive modeling, and it is where the
prediction–inference distinction becomes genuinely productive rather than merely
cautionary: the goal was never inference, so the opacity objection loses its force,
while the construct-validity obligation remains fully in place.

The same pattern carries a standing risk. Because the representation is estimated, any
regression that uses it as a covariate faces the **generated-regressor** problem
(@sec-image-econometrics), and because the artifacts being embedded are *chosen* by
firms and creators, the features are correlated with unobserved sophistication. Neither
study can escape this, and both are appropriately associational in their language.

## Large Language Models

*Large language models* (LLMs) are deep neural networks—almost always
*transformers*, built on the self-attention mechanism—trained on internet-scale text
to predict the next token, and then adapted to follow instructions. Their relevance
to marketing is twofold. As *measurement instruments*, they convert the field's vast
unstructured text—reviews, social posts, support transcripts, open-ended
survey responses—into structured variables: sentiment, topics, stance, entities,
and the latent dimensions of consumer voice that earlier text methods recovered more
laboriously [@netzer2008hidden; @buschken2016sentence]. As
*generators*, they produce marketing content—copy, product descriptions,
personalized email, synthetic chat—at near-zero marginal cost, reshaping the economics
of content production across the funnel [@appel2020future].

The transformer's core operation is *attention*, which lets each token's
representation be a weighted average of the others, with weights computed from
learned query, key, and value projections,
$$
\operatorname{Attention}(\mathbf{Q},\mathbf{K},\mathbf{V}) =
\operatorname{softmax}\!\left(\frac{\mathbf{Q}\mathbf{K}^{\top}}{\sqrt{d_k}}\right)\mathbf{V},
$$ {#eq-aiml-attn}
where $\mathbf{Q},\mathbf{K},\mathbf{V}$ are linear projections of the input sequence
and $d_k$ is the key dimension. This mechanism, stacked and scaled, is what lets the
model condition each word on the entire context. The architecture that made this practical at scale is the transformer [@vaswani2017attention], which replaces recurrence with attention and so lets every position attend to every other in one step.

Three properties demand discipline when LLMs are used as research instruments.
First, **hallucination**: an LLM optimizes for plausible continuations, not truth,
and will fabricate confidently; outputs used as data must be validated against ground
truth, ideally with a human-labeled audit sample and an inter-rater reliability
check (@sec-measurement-scales). Second, **non-determinism and version drift**: the
same prompt can yield different outputs, and the underlying model changes under the
analyst's feet, so a measurement pipeline built on a hosted LLM is not automatically
reproducible—prompts, model versions, and decoding parameters must be logged like any
other instrument. Third, **contamination and circularity**: an LLM trained on the
open web may have seen the very reviews or constructs under study, so "predicting"
them is not out-of-sample, and using an LLM both to generate and to evaluate content
risks a closed loop that measures the model's preferences rather than consumers'.
Used as instruments, LLMs are powerful but require the same validity scaffolding as
any measure; used as generators, they require the brand-safety, factuality, and
fairness controls developed in the rest of this chapter.

One instrument use has grown prominent enough to name on its own: the LLM as a
*synthetic respondent*. Prompted to adopt a demographic or attitudinal persona, an LLM
will answer survey and choice questions as that person plausibly might—**silicon
sampling** [@argyle2023]—which dangles the prospect of cheap, instant, fatigue-free
subjects. The evidence is mixed by design and by domain: LLMs recover some
individual-level preferences and reproduce aggregate patterns [@goli2024frontiers], yet
their synthetic distributions are too narrow, shift with the wording of a prompt, and
drift over time, so they are unreliable for the variance-dependent inference that survey
data exist to support [@bisbee2024synthetic]. The productive posture is therefore
*augmentation* rather than replacement—spending a little human data to debias a lot of
LLM data [@wang2026llm]—which @sec-online-panels develops for survey respondents and
@sec-conjoint-variants for conjoint.

Both roles cast the LLM as a tool the firm wields. A third relevance is now emerging
that the firm does not control at all: the LLM as a *distribution channel*. When a
chat assistant answers a shopping question and links out to a retailer, it refers
traffic the way a search engine or an affiliate does—a channel that did not exist
before ChatGPT began emitting organic outbound links in August 2024.
@kaiser2026chatgpt measure it across 973 e-commerce sites—more than 50,000
LLM-referred transactions against 164 million from traditional channels—and find that
organic LLM traffic still converts below every traditional channel except paid social
(about 13% under organic search, 86% under affiliate) and earns less per session,
though the gap narrows in complex product categories and is closing over time.

The closing-over-time claim is where the paper's method rewards a look, because "how
fast is a brand-new channel maturing?" is easy to answer badly. Rather than
extrapolate a straight line, the authors fit a months-since-launch trend under
competing functional forms—linear against Gompertz and logistic S-curves that
*saturate*—so a nascent channel's trajectory cannot be projected off to implausible
values. Its data sparsity is handled not with anything exotic but with overdispersed
*quasibinomial* models for rate outcomes, linear models weighted by each cell's
session count for dollar outcomes, and weekly aggregation to blunt zero inflation. The
tools are standard—the same diffusion-curve logic marketing has long used for
new-product growth (@sec-growth-curves), here turned on a distribution channel; the
novelty is the channel, not the estimator. The two estimation ideas the paper leans
on—overdispersed rate regression and saturating trend curves—are general enough to
deserve their own treatment. The first belongs here, with inference; the second sits
with diffusion in @sec-growth-curves.

## Modeling Rates with Overdispersion: The Quasibinomial GLM {#sec-quasibinomial}

The channel comparison above is an *inference* question in the exact sense of
@sec-modeling-illustration's fork: not "predict whether this session converts" but
"does channel $A$ convert at a different rate than channel $B$, and how sure are we?"
The object of interest is a coefficient and its standard error, so the estimator must
have a sampling distribution we trust. Aggregate conversion data look deceptively easy
to model this way—$y_i$ conversions out of $n_i$ sessions is the textbook binomial—but
the textbook variance is almost always wrong for marketing rates, and taking it at
face value manufactures precision that is not there.

Write the workhorse **logistic GLM** for a proportion. With $y_i$ conversions in
$n_i$ sessions and covariates $\mathbf{x}_i$ (channel, website, device, month),
$$
y_i \sim \text{Binomial}(n_i, \mu_i), \qquad
\operatorname{logit}(\mu_i) = \mathbf{x}_i^{\top}\boldsymbol\beta, \qquad
\operatorname{Var}(y_i) = n_i\,\mu_i(1-\mu_i).
$$ {#eq-aiml-binom}
The last equality is not an assumption one gets to make lightly: it says the *only*
source of variation in the count is the coin-flip randomness of $n_i$ independent
sessions sharing a single rate $\mu_i$. Real channel-week-website cells violate it
routinely. Sessions within a cell are correlated (a viral post, a promotion, a bot
wave); the rate $\mu_i$ itself wanders with unmodeled heterogeneity across sites and
weeks; a sparse new channel like oLLM piles many near-empty cells beside a few large
ones. Each mechanism inflates the variance of $y_i$ beyond the binomial floor—it is
**overdispersed**—and the inflation is exactly what a naive logistic regression
assumes away.

The **quasibinomial** model repairs the inference without pretending to a full
likelihood. Following the quasi-likelihood idea [@wedderburn1974quasi;
@mccullagh1989glm], one specifies only the mean and a variance *proportional* to the
binomial one,
$$
\operatorname{Var}(y_i) = \phi\, n_i\,\mu_i(1-\mu_i), \qquad \phi \ge 1,
$$ {#eq-aiml-overdisp}
where $\phi$ is a **dispersion parameter** to be estimated rather than fixed at $1$.
This is precisely the variance @kaiser2026chatgpt write as
$\operatorname{Var}(y^{\text{txn}}_{it}) = \theta\, n^{\text{sess}}_{it}\,
p_{it}(1-p_{it})$: their $\theta$ is this $\phi$, and their weekly aggregation and
session weighting are complementary devices aimed at the same variance inflation.

Two facts make the quasibinomial attractive. First, because the estimating equations
for $\boldsymbol\beta$,
$$
\sum_i \frac{y_i - n_i\mu_i}{\phi\, n_i \mu_i(1-\mu_i)}\,
\frac{\partial \mu_i}{\partial \boldsymbol\beta} = \mathbf{0},
$$ {#eq-aiml-qscore}
carry $\phi$ as a constant multiplier, it cancels: the **point estimates
$\hat{\boldsymbol\beta}$ are identical to the ordinary binomial fit**. Overdispersion
does not bias the coefficients; it corrupts their *standard errors*. Second, $\phi$ is
estimated cheaply from the Pearson residuals,
$$
\hat\phi = \frac{1}{N-p}\sum_i
\frac{(y_i - n_i\hat\mu_i)^2}{n_i\hat\mu_i(1-\hat\mu_i)},
$$ {#eq-aiml-phihat}
and the corrected covariance is just the binomial one scaled,
$\widehat{\operatorname{Var}}_{\text{quasi}}(\hat{\boldsymbol\beta}) =
\hat\phi\,\widehat{\operatorname{Var}}_{\text{binom}}(\hat{\boldsymbol\beta})$, so every
standard error grows by $\sqrt{\hat\phi}$ and every naive $t$-statistic shrinks by the
same factor. A cell that reports $\hat\phi = 9$ was quoting standard errors three times
too small. The simulation below plants overdispersion deliberately and recovers it.

```{r qbinom-demo, message=FALSE, warning=FALSE}
set.seed(2024)

# Weekly conversions for two channels across many websites. Overdispersion is built
# in: each website carries its OWN latent conversion propensity (a Beta draw), so the
# aggregate counts vary far more than a single common Binomial(n, p) would permit.
sim_channel <- function(mean_p, rho, n_site = 500, mean_sess = 300) {
  kappa  <- (1 - rho) / rho                          # Beta concentration; rho = dispersion
  p_site <- rbeta(n_site, mean_p * kappa, (1 - mean_p) * kappa)
  n_sess <- rpois(n_site, mean_sess) + 1L
  data.frame(conv = rbinom(n_site, n_sess, p_site), sess = n_sess)
}

d <- rbind(
  transform(sim_channel(0.020, rho = 0.10), channel = "oLLM"),
  transform(sim_channel(0.023, rho = 0.10), channel = "organic_search")
)
d$channel <- relevel(factor(d$channel), ref = "oLLM")

m_bin  <- glm(cbind(conv, sess - conv) ~ channel, family = binomial,      data = d)
m_qbin <- glm(cbind(conv, sess - conv) ~ channel, family = quasibinomial, data = d)
phi    <- summary(m_qbin)$dispersion               # estimated phi-hat

compare <- data.frame(
  term     = names(coef(m_bin)),
  estimate = round(coef(m_bin), 4),                # identical across the two fits
  se_binom = round(sqrt(diag(vcov(m_bin))),  4),
  se_quasi = round(sqrt(diag(vcov(m_qbin))), 4)
)
compare$se_ratio <- round(compare$se_quasi / compare$se_binom, 2)

cat(sprintf("Estimated dispersion  phi-hat = %.1f   (sqrt = %.2f)\n", phi, sqrt(phi)))
print(compare, row.names = FALSE)
```

The coefficients are the same to the last digit; the quasibinomial standard errors are
larger by very close to $\sqrt{\hat\phi}$, exactly as @eq-aiml-phihat predicts. The
practical consequence is a matter of what survives: the organic-search-versus-oLLM
contrast that clears significance under the binomial's optimistic errors may not clear
it once the dispersion is admitted, which is why @kaiser2026chatgpt report their
channel gaps against dispersion-corrected errors rather than the raw binomial ones.

::: {.callout-note}
## Quasibinomial and its neighbors

The quasibinomial is the lightest of several overdispersion fixes, and the right
choice depends on the question. A **beta-binomial** model puts an explicit
distribution on the per-cell rate and yields a genuine likelihood (hence AIC/BIC and
likelihood-ratio tests), at the cost of a distributional assumption the quasibinomial
avoids. A **generalized linear mixed model** with a website random effect models the
heterogeneity structurally and returns shrinkage estimates per site, but it is heavier
and can be finicky with thousands of sparse cells. **Cluster-robust (sandwich)**
standard errors correct inference without touching the variance function and are the
natural move when the correlation is within known clusters. All four buy the same
thing—honest uncertainty for a rate—so the quasibinomial's appeal is that it delivers
it with one extra estimated scalar and no new distributional commitment. Reach past it
only when you need the likelihood machinery or the per-site effects it does not
provide.
:::

Note the tidy division of labor with the prediction–inference fork of
@sec-modeling-illustration. A pure *predictor* of "will this session convert" is
indifferent to $\phi$: overdispersion leaves the fitted probabilities $\hat\mu_i$
untouched, so out-of-sample loss is unchanged. It is only when the coefficient itself
is the deliverable—a channel comparison, an elasticity, a lift—that the dispersion
parameter becomes load-bearing. Overdispersion is an inference bug, not a prediction
bug, which is why it is invisible to a leaderboard and fatal to a claim.

## Disclosure and the Producer Side of Generative AI {#sec-aiml-disclosure}

Return to the LLM's *generator* role, and to the policy instrument that has grown up
around it. Regulators increasingly require that generative-AI involvement in creative
work be **disclosed**—a label attached to the artifact, or a content credential
travelling with it—on the theory that audiences deserve to know what they are looking
at. The research question this invites is an evaluation question: does a label change
how an image, an ad, or a piece of copy is judged? Nearly all of the evidence answers
that question, and it holds the artifact fixed while randomizing the label.

@jussupow2026disclosure show that this design misses most of the effect, because the
label reaches the *producer* first. In two nested mixed-methods experiments in which
participants collaborate with a text-to-image tool under different disclosure regimes,
creators who merely *anticipate* that their AI use will be disclosed change how they
work: the majority **withdraw from the creative process**, handing generation over to
the model rather than iterating with it. The mechanism the authors theorize, from
Goffman's account of impression management, is not embarrassment about using AI but a
threat to *validation as a creative self*—creators expect that the label will keep the
audience from recognizing their own creative agency, so the agency stops being worth
investing. The artifacts that result predominantly reflect computational rather than
human creativity, and audiences perceive and downgrade them accordingly, label or no
label. The authors name this the **indirect disclosure effect**, and its policy sting
is precise: a transparency rule intended to protect human creative agency can erode it
*before any audience ever sees the disclosure*.

Two lessons generalize past the creative setting. The first is a **research-design**
lesson with the structure of a general-equilibrium critique: when an intervention is
announced in advance, the artifact under evaluation is itself an outcome of the
intervention, so an experiment that randomizes labels across fixed stimuli estimates a
partial effect and can get the sign of the policy wrong. Anything that alters what
producers anticipate—AI labels, provenance metadata, review-verification badges, algorithmic
transparency requirements (@sec-privacy)—requires a design that lets production
respond. The second is a **measurement** lesson: the operative variable is a
*second-order belief*, what the creator expects the audience to infer, which is
neither the creator's own attitude toward AI nor the audience's actual reaction, and
which must be measured as its own construct (@sec-measurement-scales) rather than
assumed from either endpoint. Firms deploying generative tools internally face the same
mechanism in miniature: an attribution policy that marks AI-assisted work will shift
not only how that work is received but how much of themselves employees put into it.

The same producer-side logic scales up from the individual creator to the platform that
hosts them. @huang2026stances study two opposite-signed policy shocks on Chinese
visual-arts platforms—one platform launching its own AI image generator, another
prohibiting AI-generated artwork—and find creator activity falling in the first case and
rising in the second. As with disclosure, the operative variable is an expectation rather
than a feature: creators read the policy as a statement about AI's role, about the
platform's commitment to human work, and about the competition their work will face, with
the largest pullback among popular, multi-homing, and AI-averse creators. Put beside
@jussupow2026disclosure, the pair delimits the producer-side response to generative AI at
two levels—what a creator does *within* a task when a label is coming, and whether the
creator shows up at all when a platform declares a stance. Neither effect is visible to a
design that holds the artifact, or the creator population, fixed; the platform-governance
reading of the same evidence is developed in @sec-platform-governance.

### Replication: what a label-only design cannot see {#sec-aiml-disclosure-replication}

The claim that the standard design is biased is worth making precise rather than
asserting, because the bias has a definite structure and a measurable size. Write the
regime as $D \in \{0,1\}$—whether the creator knows in advance that the artifact will
carry a disclosure label. The regime reaches the audience's rating $Y$ through two
routes. The **direct** route is the label itself, seen at evaluation time. The
**indirect** route runs through production: $D$ raises the creator's second-order
belief $B$ that the audience will not recognize their agency, $B$ triggers withdrawal
from the collaboration, withdrawal lowers the human creative content $H$ of the
artifact, and $H$ is what the audience actually rewards.

```{mermaid}
%%| label: fig-indirect-disclosure
%%| fig-cap: "The indirect disclosure effect. A label-only experiment randomizes the dashed arrow while holding the artifact fixed, so it estimates the direct path alone. The producer-side chain is realized before the audience ever sees the label, and it is the larger of the two."
flowchart LR
    D["Anticipated disclosure\n(regime announced to creator)"]
    B["Second-order belief:\n'they will not see my agency'"]
    W["Withdrawal from\nthe collaboration"]
    H["Human creative content\nof the artifact"]
    Y["Audience evaluation"]
    L["Disclosure label\nshown at evaluation"]
    D --> B --> W --> H --> Y
    D --> L
    L -. "direct path (all a label-only design sees)" .-> Y
```

The simulation below instantiates that chain with a modest direct label penalty and a
withdrawal response calibrated to the paper's qualitative finding—a minority of
creators hand generation over to the model when nothing will be disclosed, a majority
when disclosure is anticipated. It then runs the two designs side by side on the same
population: the conventional experiment, which draws artifacts from the no-disclosure
regime and randomizes only the label; and the regime comparison, which lets production
respond. A third contrast strips the label from the audience entirely, to check whether
the artifact is already worse on its own.

```{r aiml-indirect-disclosure, message=FALSE, warning=FALSE}
set.seed(951)
n <- 3000

# --- Producer stage: creators randomized to ANTICIPATED disclosure ------------
disc <- rbinom(n, 1, 0.5)                       # will my AI use be labeled?

# Second-order belief: "the audience will not recognize my creative agency."
belief <- 1.0 * disc + rnorm(n, 0, 1)

# Withdrawal: accept the model's output instead of iterating with it.
withdraw <- rbinom(n, 1, plogis(-0.7 + 1.2 * belief))

# Creative involvement, and the human creativity it puts into the artifact.
iters <- pmax(0, rnorm(n, 6 - 3 * withdraw - 0.8 * belief, 1.5))
human <- 0.35 * iters + rnorm(n, 0, 1)          # human-creativity content

# --- Audience stage: rating responds to the artifact AND to the label --------
b_human <- 0.55     # audiences reward human creative content
b_label <- -0.25    # direct penalty from seeing the label
rate <- function(h, lab) 3.2 + b_human * h + b_label * lab + rnorm(length(h), 0, 1)

# (a) The standard label-only design: hold artifacts fixed at what creators make
#     when no disclosure is anticipated, then randomize the label.
pool  <- human[disc == 0]
lab_r <- rbinom(length(pool), 1, 0.5)
d_lab <- data.frame(y = rate(pool, lab_r), lab = lab_r)
est_labelonly <- coef(lm(y ~ lab, data = d_lab))["lab"]

# (b) The policy-relevant contrast: artifacts are produced under the regime and
#     then carry its label.
y_regime  <- rate(human, disc)
est_total <- coef(lm(y_regime ~ disc))["disc"]

# (c) Blind the audience to the label: is the artifact itself already worse?
y_blind   <- rate(human, rep(0, n))
est_blind <- coef(lm(y_blind ~ disc))["disc"]

# --- Decomposition: how much of the total runs through production? -----------
decomp <- function(idx) {
  h <- human[idx]; d <- disc[idx]
  fit <- lm(rate(h, d) ~ h + d)
  dh  <- mean(h[d == 1]) - mean(h[d == 0])            # regime -> artifact
  c(indirect = unname(dh * coef(fit)["h"]),           # artifact -> rating
    direct   = unname(coef(fit)["d"]))
}
boot <- replicate(400, decomp(sample.int(n, n, replace = TRUE)))
ci   <- apply(boot, 1, quantile, c(0.025, 0.975))

cat(sprintf("Withdrawal rate, no disclosure anticipated  : %.0f%%\n",
            100 * mean(withdraw[disc == 0])))
cat(sprintf("Withdrawal rate, disclosure anticipated     : %.0f%%\n",
            100 * mean(withdraw[disc == 1])))
cat(sprintf("Creative iterations, control vs. disclosure : %.2f vs %.2f\n\n",
            mean(iters[disc == 0]), mean(iters[disc == 1])))
cat(sprintf("Label-only design (artifacts held fixed)    : %+.3f\n", est_labelonly))
cat(sprintf("Regime effect on rating (label + response)  : %+.3f\n", est_total))
cat(sprintf("Regime effect with the label hidden         : %+.3f\n\n", est_blind))
cat(sprintf("  producer-side (indirect) path : %+.3f  [%+.3f, %+.3f]\n",
            mean(boot["indirect", ]), ci[1, "indirect"], ci[2, "indirect"]))
cat(sprintf("  audience-side (direct) path   : %+.3f  [%+.3f, %+.3f]\n",
            mean(boot["direct", ]), ci[1, "direct"], ci[2, "direct"]))
cat(sprintf("  share of the total a label-only design cannot see : %.0f%%\n",
            100 * mean(boot["indirect", ]) / est_total))
```

The label-only experiment is not *wrong*; it is answering a different question,
correctly. It recovers the direct penalty—what a label does to the perception of a
fixed artifact—and misses roughly half of the effect of the policy, because the other
half was realized upstream when creators, anticipating the label, stopped putting
themselves into the work. The blinded contrast is the diagnostic that makes this
visible: artifacts produced under an anticipated-disclosure regime rate lower *even
when the audience never sees a label*, which cannot happen if the label operates only
at the point of evaluation.

The structure generalizes to any intervention whose existence is known to the party
generating the outcome, which is most of marketing regulation and a great deal of
platform policy. Verified-review badges, provenance metadata, algorithmic-transparency
notices, and internal attribution rules all share the shape: announce the rule and the
stimulus itself becomes endogenous. The design implication is that the randomization
must be assigned at the level of the *regime*, before production, not at the level of
the artifact after it—and that the second-order belief, what the producer expects the
audience to infer, has to be measured as its own construct
(@sec-measurement-scales) rather than backed out from either side's behavior.

The producer-side response is not always a cost. Where an anticipated label makes
creators withdraw, a well-designed first interaction makes users lean in:
@jiang2026fulfillment show on a generative-AI co-creation platform that revealing
*part* of the co-created output before the registration wall raises both the user's
sense that their input mattered and their curiosity about what remains, and so
raises registration and return visits (@sec-platform-fulfillment). The same
mechanism—what the human expects the interaction to say about their own
contribution—runs in both cases; only the sign of the design changes.

## Machine Learning Across the Marketing Funnel

The methods above are not siloed by funnel stage; the same supervised, unsupervised,
and generative tools recur, retargeted at different objectives. Organizing them by
the customer journey clarifies *what* is being predicted at each step and *where* the
prediction–inference distinction bites. @tbl-aiml-funnel maps the terrain, and the
recurring lesson is that most funnel applications are predictive scoring problems,
while the decisions they feed—how much to spend, what to charge, whom to
target—are causal questions that prediction alone cannot answer.

| Funnel stage | Representative task | Learning paradigm | Pred. vs. inf. | Anchor |
|---|---|---|---|---|
| Awareness | Audience look-alike modeling; media-mix forecasting | Supervised classification; sequence models | Prediction (lift is causal) | @wedel2016 |
| Consideration | Search/social listening; brand-perception mining | Unsupervised; LLM/text | Prediction (measurement) | @netzer2008; @liu2020 |
| Conversion | Propensity-to-buy; dynamic creative; recommendation | Supervised; recommender | Prediction; *targeting is causal* | @neumann2019frontiers |
| Retention | Churn prediction; next-best-action | Supervised classification | Prediction; *action is causal* | @neumann2019frontiers |
| Advocacy | Review/UGC analysis; influencer identification | Unsupervised; LLM/text | Prediction (measurement) | @tirunillai2014a |

: Machine learning across the marketing funnel. Most applications are predictive scoring tasks; the decisions they inform are causal. {#tbl-aiml-funnel}

The "prediction (lift is causal)" and "action is causal" entries flag the same trap
in different clothes. A look-alike model predicts who *resembles* a converter, but
whether advertising to them *causes* incremental conversion is a question only a
holdout experiment answers. A churn model predicts who *will* leave, but the right
retention action depends on who will leave *because of, or despite,* the
intervention—an uplift, not a prediction, problem. @ascarza2018futility makes the point with unusual force: in two field experiments, targeting the customers a churn model ranked as highest-risk was no better than random, because risk of leaving and responsiveness to an intervention are different quantities and the industry standard optimizes the wrong one. The funnel view
is useful precisely because it keeps surfacing the boundary the chapter is built
around.

## Pitfalls: How Machine Learning Quietly Breaks

A model that scores well offline and fails in production is the rule, not the
exception, and the failures cluster into three families: *leakage*, which inflates
offline performance with information that will not exist at decision time; *drift*,
which erodes performance as the world moves away from the training distribution; and
*unfairness*, which encodes and amplifies inequities the data inherited. None of the
three is a coding bug; all three are violations of the assumptions behind
@eq-aiml-erm. @fig-aiml-lifecycle situates them in the model lifecycle.

```{mermaid}
%%| label: fig-aiml-lifecycle
%%| fig-cap: "Where the three pitfalls strike in the machine-learning lifecycle. Leakage corrupts training and validation before deployment; drift degrades a correct model after it; fairness harms accrue at the decision boundary and feed back into future data."
flowchart LR
  A["Data collection"] --> B["Feature engineering<br/>& training"]
  B --> C["Offline validation"]
  C --> D["Deployment &<br/>decisions"]
  D --> E["Monitoring"]
  E -.->|"new data"| A
  B -. "LEAKAGE<br/>(future/target info<br/>leaks into features)" .-> C
  D -. "DRIFT<br/>(world moves away<br/>from training dist.)" .-> E
  D -. "FAIRNESS<br/>(disparate harm at<br/>the decision boundary)" .-> A
```

### Data leakage

**Leakage** occurs when information unavailable at prediction time contaminates the
training data, so the model "cheats" offline and collapses in production. The signature
is an offline metric that is too good to be true. Leakage takes several forms, each
common in marketing pipelines. *Target leakage* includes a feature that is a proxy
for, or a consequence of, the outcome—predicting purchase using "added-to-cart in the
same session," which is nearly the outcome itself. *Temporal leakage* uses
information from after the prediction timestamp—computing a customer's average order
value over a window that includes the very order being predicted. *Preprocessing
leakage* is the most insidious because it survives a careful feature audit: when
scaling, imputation, feature selection, or target encoding is fit on the *full*
dataset before the train/test split, statistics from the test set bleed into
training, and cross-validation reports an accuracy the deployed model cannot reach.

The discipline that prevents leakage is to fit *every* data-dependent transformation
inside the cross-validation fold, on training data only, and to respect time: when
data are temporally ordered—as marketing data almost always are—validation must use a
forward-chaining split that trains on the past and tests on the future, never a random
split that lets the model peek ahead. The example contrasts the two and quantifies the
illusion.

```{r aiml-leakage, message=FALSE, warning=FALSE}
set.seed(48)

# Time-ordered data where the signal is non-stationary (drifts over time)
N <- 1500
time  <- seq_len(N)
beta_t <- 1.5 - 1.0 * (time / N)          # the effect of x decays over time
x <- rnorm(N)
y <- as.integer(plogis(-0.2 + beta_t * x + rnorm(N, 0, 0.5)) > 0.5)
df <- data.frame(time, x, y)

auc <- function(score, label) {
  pos <- score[label == 1]; neg <- score[label == 0]
  if (!length(pos) || !length(neg)) return(NA_real_)
  mean(outer(pos, neg, ">") + 0.5 * outer(pos, neg, "=="))
}

# (a) RANDOM split: leaks future into training on non-stationary data
ridx  <- sample(N, 0.7 * N)
m_rand <- glm(y ~ x, df[ridx, ], family = binomial())
auc_rand <- auc(predict(m_rand, df[-ridx, ], type = "response"), df[-ridx, ]$y)

# (b) FORWARD-CHAINING split: train on past, test on future (honest)
cut    <- floor(0.7 * N)
m_fwd  <- glm(y ~ x, df[df$time <= cut, ], family = binomial())
auc_fwd <- auc(predict(m_fwd, df[df$time > cut, ], type = "response"),
               df[df$time > cut, ]$y)

cat("AUC, random split (optimistic):  ", round(auc_rand, 3), "\n")
cat("AUC, forward-chaining (honest):  ", round(auc_fwd, 3), "\n")
```

The random-split AUC overstates the performance the model will actually deliver on
future data, because random splitting on a non-stationary process lets the model
borrow the future to predict the past. The gap between the two numbers is a direct
measure of self-deception, and it is invisible to anyone who validates with a random
split—the default in most tutorials.

### Distribution drift {#sec-aiml-drift}

Even a leakage-free model decays, because the assumption that deployment data share
the training distribution expires. **Drift** comes in two flavors that demand
different responses. *Covariate shift* changes the input distribution while the
relationship is stable, $p_{\text{test}}(\mathbf{x}) \neq p_{\text{train}}(\mathbf{x})$
but $p(y\mid\mathbf{x})$ unchanged—a new acquisition channel brings customers unlike
the training population. *Concept drift* changes the relationship itself,
$p_{\text{test}}(y\mid\mathbf{x}) \neq p_{\text{train}}(y\mid\mathbf{x})$—a recession,
a competitor's launch, or a pandemic rewrites how features map to behavior, exactly
the non-stationarity simulated above. Marketing is a near-worst case for drift
because the environment is adversarial and reflexive: competitors react, consumer
tastes move, and—uniquely—the model's *own actions* change the data it next sees, so a
targeting model trained on one policy's data is evaluated under another. The defenses
are monitoring (track input distributions and live performance against a holdout, and
alarm on divergence), scheduled or triggered retraining, and—where decisions feed back
into data—maintaining randomized holdouts so the model never fully determines its own
training distribution.

### Fairness and algorithmic bias

Machine-learning models inherit and can amplify the biases latent in their training
data, with legal and ethical force when marketing decisions touch credit, housing,
employment, insurance, or protected groups. The mechanism is not malice but
statistics: if historical data reflect discrimination, a model that faithfully
predicts the historical target reproduces the discrimination, and *omitting* a
protected attribute does not fix it, because correlated proxies (postal code, device,
browsing history) reconstruct the attribute—a phenomenon known as redundant encoding.

Fairness must therefore be defined, measured, and traded off explicitly, and a basic
impossibility result disciplines expectations: several intuitive fairness
criteria—equal false-positive and false-negative rates across groups (equalized odds)
versus equal calibration across groups versus statistical parity in selection
rates—*cannot in general all hold simultaneously* unless base rates are equal or the
classifier is perfect. There is no single "fair" model; there is a choice among
incompatible fairness criteria, and that choice is normative, not technical. The
worked example audits a scorer for *disparate impact*—whether selection rates differ
across groups—and shows that proxy features carry bias even when the protected
attribute is excluded. The impossibility results are worth confronting directly: except in degenerate cases, calibration within groups and equal false-positive and false-negative rates across groups cannot all hold at once, so a fairness criterion is a choice among incompatible goods rather than a box to tick [@kleinberg2017humandecisions].

```{r aiml-fairness, message=FALSE, warning=FALSE}
set.seed(48)

n <- 6000
group <- rbinom(n, 1, 0.5)                       # protected attribute A in {0,1}
# A proxy correlated with the protected group (e.g., neighborhood), NOT the label cause
proxy <- rnorm(n, mean = 0.8 * group)
quality <- rnorm(n)                              # the legitimately relevant signal
# True outcome depends ONLY on quality (group is irrelevant to merit)...
y <- rbinom(n, 1, plogis(1.2 * quality))
df <- data.frame(y, quality, proxy, group)

# Model that EXCLUDES the protected attribute but KEEPS the correlated proxy
fit <- glm(y ~ quality + proxy, df, family = binomial())
df$score <- predict(fit, type = "response")
tau <- quantile(df$score, 0.7)                   # select top 30%
df$selected <- as.integer(df$score > tau)

# Disparate impact: ratio of selection rates across groups (4/5ths rule -> >= 0.8)
sel_rate <- tapply(df$selected, df$group, mean)
di_ratio <- min(sel_rate) / max(sel_rate)
cat("Selection rate by group:", round(sel_rate, 3), "\n")
cat("Disparate-impact ratio:  ", round(di_ratio, 3),
    if (di_ratio < 0.8) " (FAILS 4/5ths rule)" else " (passes)", "\n")
```

The disparate-impact ratio falls below the conventional four-fifths threshold even
though the protected attribute never entered the model and plays no role in the true
outcome—the proxy alone manufactures the gap. The lesson generalizes: fairness is a
property of a model-in-context that must be measured on outcomes, not assumed from the
feature list, and remediation (reweighting, constrained optimization, post-hoc
threshold adjustment by group) is an explicit, auditable design choice with its own
trade-offs against accuracy and against other fairness criteria.


#### When the fairness policy needs the attribute it is not allowed to demand {#sec-aiml-parity-disclosure}

The audit above assumes the protected attribute is *known* to the analyst even when
it is excluded from the model. In recommendation settings it frequently is not. A
firm that wants to commit to **accuracy parity**—consistent recommendation quality
across demographic groups—generally has to rely on consumers disclosing their
identity voluntarily, and that turns the fairness policy into a game rather than a
constraint.

@smith2026fairai model it as one. Digital platforms recommending video, music, and
news deliver systematically worse recommendations to some consumer groups, and a
parity commitment is the natural remedy; but consumers choose whether to disclose,
the firm chooses what to recommend, and the firm also chooses how much to invest in
the AI's learning. Their game-theoretic model puts all three in one system, and it is
worth being clear about why they cannot be separated. Disclosure is what makes parity
enforceable, and it is also what determines the training data the firm's learning
investment operates on, so a consumer's disclosure decision is partly a decision
about how good the firm's model will become for people like them—while the firm's
parity commitment changes the payoff to disclosing in the first place. The
equilibrium characterization of that system is the paper's contribution and is left
to it here.

The methodological point for this chapter is that the impossibility results above
are stated for a *given* dataset with group membership observed. Once group
membership is elicited rather than observed, the feasible set is endogenous to the
policy, and a parity rule that looks costless in an audit can be undermined or
amplified by who chooses to be counted. Any marketing team designing an
accuracy-parity commitment should treat the disclosure margin as a first-order
design variable rather than a data-collection detail.

## Deploying Marketing AI: The Operations Interface {#sec-aiml-operations}

Read the three pitfalls again and a pattern emerges: none of them is fixed by a better
estimator. Leakage is a failure of pipeline discipline, drift is a failure of
monitoring and refresh cadence, and unfairness is a failure of specification and
audit. They are *operations* problems attached to a statistical object, and the
literature that studies them systematically is operations management rather than
machine learning. @dai2026aiops organize that literature into a triadic framework
whose three pillars translate cleanly onto marketing.

The first pillar, **AI for operations**, asks where AI improves an operational
process—in their vocabulary the design–procure–produce–deliver chain, in marketing the
campaign chain of targeting, creative generation, bidding, and service delivery. This
is the pillar the chapter has occupied so far, and its discipline is the one already
insisted upon: performance claims are empirical and context-dependent, so AI may
improve a task, leave it unchanged, or introduce new frictions, and the burden is on
measurement rather than assumption. The second pillar, **operations for AI**, inverts
the lens: the AI system is itself something produced and operated, so the tools built
for factories—workflow design, capacity management, statistical process control,
lot-sizing, continuous improvement—apply to the model lifecycle. The third pillar,
**human–AI interaction**, holds that the value realized is mediated by whether the
humans in the loop trust, override, or rubber-stamp the system, which makes interface
design, incentives, and supervisory capacity part of the estimator's performance.
@fig-aiml-pillars situates the three and their dependencies.

```{mermaid}
%%| label: fig-aiml-pillars
%%| fig-cap: "The three pillars of the AI-operations interface, translated to marketing. Deployed applications (left) create lifecycle demands that operations discipline must absorb (right); both pass through the human interface (bottom) that determines whether model output becomes a decision. Adapted from the framework in Dai and Swaminathan (2026)."
flowchart TB
  A["AI FOR MARKETING<br/>Targeting, creative generation,<br/>bidding, recommendation,<br/>service delivery"]
  B["MARKETING OPS FOR AI<br/>Data pipelines, retraining cadence,<br/>drift monitoring, audit,<br/>governance"]
  C["HUMAN-AI INTERACTION<br/>Trust calibration, override policy,<br/>supervisory capacity,<br/>accountability"]
  A -->|"generates lifecycle<br/>demands"| B
  B -->|"sustains deployable<br/>models"| A
  A -->|"output must be<br/>acted on"| C
  C -->|"adoption, overrides,<br/>and labeled feedback"| B
  B -->|"explanations, alerts,<br/>and thresholds"| C
```

### The model pipeline as a production line {#sec-aiml-mlops}

The productive analogy is exact enough to borrow results, not merely vocabulary. Data
are inventory, subject to quality inspection at receipt; training is production, with
each pass a lot whose size trades off setup against holding cost; deployment is
service delivery under a latency service-level agreement; and monitoring with
retraining is continuous improvement. The claim carrying the most weight for a
marketing analytics team is the one about *cadence*, because it is where practice most
often defaults to habit ("we retrain quarterly") rather than to a decision.

Formalize the lot-sizing analogy that @dai2026aiops draw. Let $K$ denote the fixed
cost of one refresh cycle—compute, validation, review, and the risk of a bad
deployment—and let staleness damage accrue at rate $c_d$ per period, so that a model
$a$ periods old costs $c_d a$ in foregone margin relative to a freshly fitted one.
Over a refresh cycle of length $T$, average cost per period is
$$
C(T) \;=\; \frac{K}{T} \;+\; \frac{c_d\,(T-1)}{2},
$$ {#eq-aiml-cadence}
whose minimizer is the economic-order-quantity square root,
$$
T^\* \;=\; \sqrt{\frac{2K}{c_d}}.
$$ {#eq-aiml-cadence-opt}

Two consequences follow that are not obvious from the machine-learning side. First,
optimal cadence is *flat* near the optimum and scales with the square root of the cost
ratio: halving the cost of a retraining cycle shortens the optimal interval by only
about 30%, so the case for automating the pipeline rests on reliability and on the
option to respond to shocks, not on the arithmetic of cadence alone. Second, the whole
calculation presumes a *constant* drift rate, which marketing violates—drift arrives
in regime breaks (a competitor's launch, a platform policy change, a seasonal
discontinuity) rather than as a steady trickle. Under lumpy drift a fixed schedule is
dominated by a *triggered* policy that retrains on a monitoring signal, which is the
operational content of the monitoring advice in @sec-aiml-drift.

```{r aiml-retrain-cadence, message=FALSE, warning=FALSE}
set.seed(4826)

weeks <- 520
K     <- 40      # fixed cost of one refresh cycle (compute + validation + review)

# Weekly margin lost to staleness. Most weeks the world moves slowly; ~4% of weeks
# bring a regime break (competitor launch, platform policy change, seasonal shift)
# that pushes the deployed model away from the world far faster.
regime_break <- rbinom(weeks, 1, 0.04)
drift <- ifelse(regime_break == 1, rgamma(weeks, 2, 0.4), rgamma(weeks, 2, 4))

# A model of age a costs the accumulated drift since its last refresh.
cost_scheduled <- function(T) {
  age <- (seq_len(weeks) - 1) %% T                 # weeks since last refresh
  dmg <- ave(drift, cumsum(age == 0), FUN = cumsum)
  (sum(dmg) + K * ceiling(weeks / T)) / weeks
}

grid   <- 2:52
c_sched <- vapply(grid, cost_scheduled, numeric(1))
T_best  <- grid[which.min(c_sched)]
T_eoq   <- sqrt(2 * K / mean(drift))               # Eq. for T* at the average drift rate

# Triggered policy: refresh when a noisily monitored staleness measure crosses h.
cost_triggered <- function(h) {
  acc <- 0; dmg <- 0; n <- 0
  for (t in seq_len(weeks)) {
    acc <- acc + drift[t]
    dmg <- dmg + acc
    if (acc + rnorm(1, 0, 0.3) > h) { acc <- 0; n <- n + 1 }   # monitor is imperfect
  }
  (dmg + K * n) / weeks
}
h_grid  <- seq(1, 40, by = 1)
c_trig  <- vapply(h_grid, cost_triggered, numeric(1))

cat("EOQ cadence at average drift:  ", round(T_eoq, 1), "weeks\n")
cat("Best fixed schedule:            every", T_best, "weeks,",
    "cost", round(min(c_sched), 2), "\n")
cat("Best triggered policy:          threshold", h_grid[which.min(c_trig)], ",",
    "cost", round(min(c_trig), 2), "\n")
cat("Cost reduction from monitoring:",
    paste0(round(100 * (min(c_sched) - min(c_trig)) / min(c_sched), 1), "%\n"))
```

The square-root formula lands close to the best fixed schedule, which is the point of
carrying it: it gives an order-of-magnitude answer from two quantities a team can
estimate—what a refresh costs and how fast holdout performance decays—without a
simulation. The triggered policy then buys a further reduction precisely because drift
is lumpy: it spends refreshes when the world moves and withholds them when it does
not. The managerial reading is that monitoring infrastructure is not overhead attached
to the model but a substitute for retraining capacity, and the exchange rate between
them is estimable.

The same lens rehabilitates a practice the machine-learning literature treats as
hygiene. Statistical process control asks whether a monitored series is *in control*,
and drift detection is that question asked of a model: an alarm is an out-of-control
signal, and the specification problem is choosing which characteristics to chart.
Accuracy alone is inadequate, because a marketing model can hold its aggregate hit
rate while its false-positive rate on a segment, its latency at the tail, or its
calibration in a new channel deteriorates. The discipline @dai2026aiops describe as a
"Six Sigma for AI"—define the failure modes, measure baseline rates, analyze root
causes, improve, and control through ongoing monitoring—is the manufacturing version
of the validation habits urged in @sec-aiml-drift, with the addition that the metric
set is chosen *before* deployment and charted continuously rather than reconstructed
after a complaint.

### Human supervision is a capacity constraint {#sec-aiml-fanout}

The pillar that most often goes unmodeled is the human one, and it has a capacity that
can be written down. Define the **neglect time** $T_N$ as how long an automated agent
runs unattended before its performance degrades, and the **interaction time** $T_I$ as
how long a human needs to review and resolve one exception. The **fan-out**
$$
F \;=\; \frac{T_N}{T_I}
$$ {#eq-aiml-fanout}
is the number of agents one operator can effectively supervise [@dai2026aiops]. Treat
the reviewer as a server and exceptions as arrivals and the constraint becomes a
queueing one: with arrival rate $\lambda$, service rate $\mu = 1/T_I$, and $k$
reviewers, stability requires $\rho = \lambda/(k\mu) < 1$, and because waiting times
explode as $\rho \to 1$, practice targets $\rho \lesssim 0.7$ to leave headroom for
alert storms.

```{r aiml-fanout, message=FALSE, warning=FALSE}
# A marketing-operations review desk supervising autonomous agents (campaign
# generation, bid adjustment, creative approval, escalation triage).
T_N <- 6.0      # hours an agent runs unattended before quality degrades
T_I <- 0.4      # hours to review and resolve one exception

fan_out <- T_N / T_I                     # agents one reviewer can hold (Eq. for F)
lambda  <- 1 / T_N                       # exceptions per agent-hour
mu      <- 1 / T_I                       # reviews per reviewer-hour
k       <- 2                             # reviewers on shift

agents <- c(10, 15, 20, 25, 30)
rho    <- agents * lambda / (k * mu)

cat("Fan-out per reviewer:", fan_out, "agents\n")
data.frame(
  agents,
  rho      = round(rho, 2),
  stable   = ifelse(rho < 1, "yes", "NO"),
  headroom = ifelse(rho <= 0.7, "buffered", "at risk")
)
```

The table makes the design levers concrete. A desk of two reviewers is stable up to
thirty agents only in the knife-edge sense; it is *robust* to roughly twenty, and the
way to supervise more is not to hire linearly but to raise $T_N$ (better agents,
narrower autonomy, batched or escalated alerts so that only $m$ concurrent anomalies
interrupt) or to cut $T_I$ (pre-filled context, one-click rollback, interfaces that do
not impose extraneous cognitive load).

Getting this wrong produces the two failure modes that bracket the human–AI literature.
Overload the desk and review degrades to rubber-stamping—**automation bias**, the
tendency to accept algorithmic recommendations without the scrutiny that justified
deploying a human reviewer at all. Under-deliver on reliability and the opposite
appears: **algorithm aversion**, the documented tendency to abandon a statistically
superior algorithm after seeing it err once, even when the human alternative errs more
often [@dietvorst2015]. Both are miscalibrations of trust rather than preferences
about automation, which is why the remedies are operational: calibrated exposure
through parallel human-and-model pilots, explanations that expose the drivers of a
recommendation rather than post hoc rationalizations, and interpretable models where
the stakes make opacity untenable. The design objective is that reliance should track
the model's actual conditional accuracy—high where it is reliable, low where it is
not—and neither uniform deference nor uniform suspicion achieves that.

### Supervision assumes an independent supervisor {#sec-aiml-anchoring}

@eq-aiml-fanout counts reviewers. It says nothing about whether a reviewer who *is*
looking is still exercising independent judgment, and that assumption turns out to be
the fragile one.

@lakhiwal2026transparency study algorithmically evaluated asynchronous video
interviews—a hiring process in which an AI scores recorded candidate responses—and
manipulate how *transparent* the system is about what the algorithm does. The two
sides of the interaction move in opposite directions. For the people being evaluated,
transparency helped: it reduced stress and reduced impression management, and
interview performance improved. For the human evaluators, transparency did something
else. When AI-generated scores were visible, evaluators **anchored on them and
discounted their own judgment**—even though the scores could not, in fact,
discriminate honest from deceptive candidates.

That pairing is the finding worth carrying. Transparency is normally argued for as a
remedy for automation's opacity, and on the subject side it behaves like one. On the
supervisor side, showing the score is not neutral disclosure: it is an anchor, and it
substitutes for the judgment the supervisor was retained to supply. A review desk
staffed to a comfortable $\rho$ can therefore still be a rubber stamp, and the audit
that would detect it has to compare reviewer decisions against ground truth rather
than against agreement rates with the model—agreement is exactly what anchoring
produces. Where the human is the control on the model, the interface should surface
the *evidence* rather than the *verdict*, and the model's output should arrive after
the human's, not before it.

### The flywheel makes today's policy tomorrow's training data {#sec-aiml-flywheel}

One feedback loop deserves separate statement because it changes the objective
function. Adoption of a model generates data, the data improve the model, and the
improvement drives further adoption—the **AI flywheel** whose contracting and pricing
implications @gurkan2022flywheel formalize. For marketing the implication is sharper
than for most settings, because the firm's action *is* the sampling design: a
targeting policy determines who is exposed and therefore whose responses are observed,
and a pricing policy determines which price points are ever sampled. A myopically
revenue-optimal policy is therefore generically suboptimal as a data-collection
policy, and the gap is the exploration term that bandit and adaptive-experimentation
methods (@sec-pricing) exist to price. This is the same feedback that
@sec-aiml-drift flags as a source of drift, seen from the design side rather than the
diagnostic side, and it supplies the affirmative case for the randomized holdout: the
holdout is not a tax on revenue but the mechanism that keeps the training distribution
from being fully determined by the model's own past decisions.

The synthesis @dai2026aiops offer is worth stating plainly, because it cuts against
the way AI initiatives are usually justified internally. Stalled pilots and degrading
models are typically diagnosed as model-quality problems and answered with better
models; the operations reading is that they are usually symptoms of brittle pipelines,
unclear ownership, weak monitoring, and unsupervisable autonomy. The complementarity
runs both ways—AI expands the set of feasible marketing decisions, and operations
discipline determines which of those decisions survive contact with deployment.

## From Prediction to Decision: Causal Machine Learning

The chapter's organizing distinction has a constructive resolution. The right way to
use machine learning's flexibility for *inference* is not to read coefficients off a
predictive model but to embed flexible learners inside an estimator whose
identification comes from design. *Double/debiased machine learning* uses
nonparametric learners to flexibly absorb high-dimensional confounders while
preserving valid inference on a low-dimensional causal parameter, via
orthogonalization and cross-fitting that immunize the target estimate against the
nuisance learner's regularization bias. *Causal forests* and related heterogeneous-
treatment-effect estimators adapt tree ensembles to estimate how a treatment effect
$\tau(\mathbf{x}) = \mathbb{E}[Y(1) - Y(0)\mid \mathbf{X}=\mathbf{x}]$ varies across
customers—the *uplift* that retention and targeting decisions actually require, as
distinct from the *level* a churn model predicts. These methods, developed in the
causal-inference chapters, are the principled bridge from machine learning's
predictive power to the causal questions marketing decisions pose, and they are where
the field's frontier is moving [@varian2016]. Two developments define that frontier. Double/debiased machine learning uses flexible first-stage predictions with cross-fitting and an orthogonalized moment so that regularization bias in the nuisance functions does not contaminate the treatment-effect estimate [@chernozhukov2018doubledebiased]. Causal forests adapt the same trees to estimate heterogeneous effects with valid confidence intervals rather than point predictions [@wager2018estimatioinference; @athey2016recursivepartition].

The generative turn adds a version of this problem that the causal-ML toolkit was not
built for: the treatment itself is now cheap to manufacture. When a model can write two
thousand candidate emails, the constraint moves from creation to *selection*, and
selection requires predicting the effect of copy that has never been deployed.
@ellickson2026unstructured build the response surface for exactly this—embedding past
content with a pretrained LLM, estimating how its features causally relate to outcomes in
the spirit of the component-effect estimates of @ellickson2023dml, and then screening
AI-generated candidates by rejection sampling so that the model refuses to score content
lying outside the support of what the firm has actually run. In a 3.3-million-observation
email application spanning 34 campaigns, the framework improved both held-out prediction
and deployed performance. The design principle generalizes past text: any
decision system that scores novel machine-generated artifacts needs an explicit
in-support / out-of-support boundary, because the alternative is a confident
extrapolation with no error bar attached to its confidence. @sec-tad-causal-prediction
develops the framework with a runnable illustration of the screen.

## What Firm-Level Adoption Data Say About AI and Jobs {#sec-aiml-adoption-jobs}

The most-asked question about this technology—does adopting it cost jobs?—has been hard
to answer because adoption itself was unmeasured. The dominant empirical strategy assigns
firms or workers an *exposure* score from occupational task content
[@brynjolfsson2025genai], which varies across jobs but cannot distinguish a firm that
adopted AI from an identical firm that did not.

@kharazian2026ai attack the measurement problem directly by linking observed corporate
card and bill-pay records identifying payments to AI vendors with monthly workforce
records for 21,559 US firms. Adoption is dated by first observed AI payment, and intensity
is measured by AI spend per employee per month; the preferred design compares adopters
against firms in the same intensity tercile that have *not yet* adopted, using the
Callaway–Sant'Anna doubly robust estimator with sector fixed effects
[@callaway2021did; @santanna2020drdid]. Over the 24 months after adoption, low-intensity
adopters show no detectable headcount change ($-0.006$ log points), while high-intensity
adopters grow total headcount by 0.097 log points (about 10.2%) and entry-level headcount
by 0.113 (about 12.0%), with gains spread across sales, administration, engineering, and
customer service and concentrated, so far, in the Information sector (about 13.4%). The
event-study path is a learning curve rather than a jump: essentially nothing in the
adoption month, 0.071 by month 6, 0.188 by month 12, 0.277 by month 18.

Three things make this worth reading closely, and one makes it worth reading carefully.
It replaces a proxy with an observed behavior, which is the highest-leverage move
available in most measurement-constrained literatures. It puts the identification burden
where it belongs, on *when* firms adopt rather than on whether they could have, and it
reports pre-period diagnostic flags alongside every estimate rather than only for the
headline. And it declines the tempting comparison: adopters versus never-adopters yields
larger effects (12.6% for high intensity) but fails 8 of 11 pre-period checks, and the
authors report it as a completeness exercise rather than as the result. The caution is
that this is industry research, not peer-reviewed work, on a sample of firms that use one
payments platform and are visible to one workforce-data vendor—selected toward younger,
technology-intensive companies—and that the design identifies the effect of adoption
*timing* among eventual adopters, not the effect of AI on the labor market as a whole.
Read against @sec-clv-pathways, the two studies make the same methodological point from
opposite substantive corners: who took an action, and when, is not a nuisance to be
differenced away but the variable that determines what the action was worth.

## Key Takeaways

- The **prediction–inference distinction** (@eq-aiml-dgp) is the organizing idea:
  predictive accuracy and causal interpretation are different goals validated by
  different criteria, and a model's internals are not causal estimates no matter how
  well it predicts.
- **Supervised learning** minimizes regularized empirical risk (@eq-aiml-erm); for
  tabular marketing data, gradient-boosted trees are the empirical default, and honest
  evaluation requires *both* discrimination and calibration, on a split that respects
  time.
- **Unsupervised learning** discovers structure without labels; clustering always
  returns clusters, so the burden of proof is on the analyst to show segments are
  stable and managerially meaningful, not merely that the algorithm converged.
- **Recommender systems** turn sparse user–item data into rankings via matrix
  factorization (@eq-aiml-mf), but cold-start, feedback loops, and the
  prediction-versus-incremental-lift gap make naïve click-maximization a trap.
- **Deep learning and LLMs** excel on unstructured data (images, text, sequences) and
  are powerful measurement and generation instruments, but their opacity makes them
  poor instruments of inference and demands explicit validity, reproducibility, and
  factuality controls.
- The three deployment pitfalls—**leakage**, **drift**, and **unfairness**—are
  assumption violations, not bugs: validate inside the fold and forward in time,
  monitor and retrain against a moving world, and measure fairness on outcomes because
  proxies encode protected attributes even when those attributes are excluded.
- Deployment is an **operations** problem, not a modeling one: the model pipeline is a
  production line whose refresh cadence obeys a lot-sizing trade-off
  (@eq-aiml-cadence-opt), whose monitoring is statistical process control, and whose
  human reviewers have a finite fan-out (@eq-aiml-fanout) that bounds how much
  autonomy can safely be deployed [@dai2026aiops].
- The constructive resolution is **causal machine learning**, which embeds flexible
  learners inside design-based estimators to recover the *uplift* that marketing
  decisions require.
