# Text as Data {#sec-text-as-data}

Most of what consumers tell firms is now written down. Product reviews, social-media
posts, search queries, customer-service chats, call-center transcripts, open-ended
survey responses, and the firm's own filings constitute a stream of unstructured
text that dwarfs the structured panels marketing science was built on. *Text as data*
is the program of turning that stream into quantities a model can ingest: counts,
vectors, topic loadings, sentiment scores, extracted entities. The premise is that
language carries measurable information about constructs marketers care about—product
quality, brand sentiment, unmet needs, persuasion, emotion—and that this information
can be recovered at scale, cheaply, and in near real time [@netzer2011; @hartmann2023more].

The promise is genuine but the discipline is unforgiving. Text is high-dimensional,
sparse, and context-dependent; the same word means different things in different
sentences, and the same meaning is expressed in countless surface forms. Every step
of a text pipeline—how a document is *represented*, which *model* maps that
representation to a target, how the output is *validated*—embeds assumptions that
can quietly determine the answer. A sentiment lexicon that scores tokens at face
value will misread sarcasm (@sec-sarcasm); a topic model with the wrong number of
topics will split or merge constructs; a classifier trained on one platform will
degrade on another. Treating text as data therefore means treating measurement
seriously: defining the construct first, choosing a representation whose assumptions
are defensible for that construct, and validating against ground truth before any
downstream regression is run.

This chapter develops the pipeline in the order a working researcher confronts it.
It begins with **representation**—how to turn a document into a vector, from the
bag-of-words and its term-weighting refinements through dense neural embeddings. It
then covers **unsupervised discovery** via topic models, the marketing workhorse
being latent Dirichlet allocation and its supervised and structural descendants.
From there it turns to **measurement of attitudes**—sentiment, emotion, and the
harder problem of *stance*—and to **supervised text classification**, where a labeled
sample trains a map from text to category. It closes with **large-language-model
(LLM) extraction**, the current frontier, in which a general-purpose model is
prompted to read documents and emit structured fields, and with the validation and
identification problems that recur across all of these methods. Throughout, the
marketing applications are concrete: mining reviews for quality and demand signals,
listening to social media for brand health, and extracting structure from sales and
service calls.

## The Text-as-Data Pipeline {#sec-tad-pipeline}

Before any method, it helps to fix the shape of the problem. A *corpus* is a
collection of $N$ documents $\{d_1,\dots,d_N\}$; a *document* is a sequence of tokens
drawn from a *vocabulary* $V$ of size $|V|$. The analyst's job is to map each
document to a representation $\mathbf{x}_i \in \mathbb{R}^p$ and then to a target—a
topic distribution, a sentiment score, a category label, or an extracted field. The
generic pipeline runs in four stages (@fig-tad-pipeline), and most of the
consequential modeling choices are made in the first two, not the last.

```{mermaid}
%%| label: fig-tad-pipeline
%%| fig-cap: "The text-as-data pipeline. Representation and modeling choices made early (left) constrain everything downstream; validation against human labels closes the loop."
flowchart LR
  A[Raw corpus<br/>reviews, posts, calls] --> B[Preprocess<br/>tokenize, normalize,<br/>filter]
  B --> C[Represent<br/>BoW / TF-IDF /<br/>embeddings]
  C --> D{Modeling goal}
  D -->|discover| E[Topic models<br/>LDA, STM]
  D -->|measure| F[Sentiment /<br/>stance]
  D -->|classify| G[Supervised<br/>classifier]
  D -->|extract| H[LLM<br/>extraction]
  E --> V[Validate vs.<br/>human labels]
  F --> V
  G --> V
  H --> V
  V -->|downstream| I[Regression /<br/>causal model]
```

*Preprocessing* is the unglamorous step that fixes the unit of analysis: lowercasing,
tokenization, removal of *stop words* (high-frequency function words such as *the*,
*of*, *and* that carry little topical content), *stemming* or *lemmatization*
(collapsing inflected forms to a root), and pruning rare tokens. Each choice trades
information against noise and is not innocuous—stemming *running*, *runs*, and *ran*
to *run* helps a topic model but destroys tense information a stance model might need.
The guiding principle is that preprocessing should be chosen to preserve the signal
for the *specific construct* being measured, not applied as a reflex.

## Representing Text {#sec-representation}

A model cannot consume words; it consumes vectors. How a document becomes a vector
is the first and most consequential modeling decision, because the representation
fixes what the model can possibly learn. Three families dominate, in increasing order
of expressiveness and opacity: the bag-of-words, term-weighted variants such as
TF-IDF, and dense embeddings.

### The Bag-of-Words {#sec-bow}

The *bag-of-words* (BoW) representation discards word order and records only which
terms appear and how often. Document $d_i$ becomes a vector of term counts over the
vocabulary,
$$
\mathbf{x}_i = (c_{i1}, c_{i2}, \dots, c_{i|V|}), \qquad
c_{ij} = \#\{\text{occurrences of term } j \text{ in } d_i\},
$$ {#eq-bow}
and the corpus becomes a sparse $N \times |V|$ *document–term matrix* (DTM). The
modeling assumption is *exchangeability*: a document is treated as an unordered
multiset of words, so "service slow but food great" and "food slow but service great"
map to the *same* vector even though they say opposite things. This is plainly false
about language, yet it is a workable approximation for tasks that depend on *topical
composition* rather than syntax, and it underpins the entire family of count-based
topic models in @sec-topic-models.

The BoW's virtues are transparency and tractability: every dimension corresponds to a
human-readable term, and the resulting matrix, though wide, is sparse and amenable to
linear methods. Its costs are equally clear. Word order is gone, so negation
("not good") and figurative inversion (@sec-sarcasm) are invisible. Dimensionality
equals vocabulary size, which grows with corpus size, producing the *curse of
dimensionality*: in $\mathbb{R}^{|V|}$ almost every pair of documents is nearly
orthogonal, and naive distances become uninformative. And the representation is
*context-free*—*bank* (river) and *bank* (finance) collapse to one dimension. Partial
remedies include *$n$-grams* (contiguous token sequences of length $n$, so *bigrams*
recover local phrases like "not good") and collocation detection, but these inflate
dimensionality further without solving the underlying context problem.

### Term Weighting and TF-IDF {#sec-tfidf}

Raw counts overweight ubiquitous words. A term appearing in every document
discriminates between none of them, however frequent. *Term frequency–inverse
document frequency* (TF-IDF) corrects this by reweighting each count by how
*distinctive* the term is across the corpus. Let $\mathrm{tf}_{ij}$ be the frequency
of term $j$ in document $i$ and let $\mathrm{df}_j = \#\{i : c_{ij} > 0\}$ be the
number of documents containing term $j$. The TF-IDF weight is
$$
w_{ij} \;=\; \mathrm{tf}_{ij} \times \log\!\frac{N}{\mathrm{df}_j},
$$ {#eq-tfidf}
where the second factor, the *inverse document frequency*, is large for rare terms
and approaches zero for terms appearing in every document. The construction has an
information-theoretic reading: $\log(N/\mathrm{df}_j)$ is, up to a constant, the
self-information of observing term $j$ under a uniform document-occurrence model, so
TF-IDF upweights surprising terms [@Shannon_1948; @weaver1963mathematical]. In
practice term frequency is often dampened (e.g., $1 + \log \mathrm{tf}_{ij}$) to keep
a term that appears fifty times from counting fifty times as much as one appearing
once, and the resulting document vectors are length-normalized so that comparisons do
not merely reflect document length.

The natural similarity measure on TF-IDF vectors is the *cosine similarity*,
$$
\mathrm{sim}(d_i, d_k) = \frac{\mathbf{w}_i^{\top} \mathbf{w}_k}
{\lVert \mathbf{w}_i \rVert \, \lVert \mathbf{w}_k \rVert},
$$ {#eq-cosine}
which measures the angle between document vectors and is invariant to length. Cosine
similarity over TF-IDF is the backbone of information retrieval and of marketing
applications that need a *map* of how products, brands, or texts relate—@netzer2011
build exactly such a similarity structure from online forum text to recover a
market's competitive map, inferring which brands consumers mention together and how
the perceptual structure of a category is organized. The following code builds a DTM,
applies TF-IDF weighting, and computes the pairwise cosine similarity of a small set
of reviews.

```{r tfidf-cosine, message=FALSE, warning=FALSE}
set.seed(49)
reviews <- c(
  "battery life is short the phone dies fast and charging is slow",
  "great camera the photos are sharp and the screen is bright",
  "the battery drains quickly and charging takes forever",
  "excellent screen and the camera takes beautiful sharp photos"
)

# 1. Tokenize and build a document-term matrix (bag-of-words counts)
stop_words <- c("is", "the", "and", "are", "a", "of")
tokens <- lapply(strsplit(tolower(reviews), "\\s+"),
                 function(w) w[!w %in% stop_words])
vocab  <- sort(unique(unlist(tokens)))
dtm <- t(sapply(tokens, function(w)
  as.integer(table(factor(w, levels = vocab)))))
colnames(dtm) <- vocab

# 2. TF-IDF weighting  (eq-tfidf): tf * log(N / df)
N  <- nrow(dtm)
df <- colSums(dtm > 0)
idf <- log(N / df)
tfidf <- sweep(dtm, 2, idf, `*`)

# 3. Cosine similarity (eq-cosine) between documents
norms <- sqrt(rowSums(tfidf^2))
cosine <- (tfidf %*% t(tfidf)) / outer(norms, norms)
round(cosine, 2)
```

The similarity matrix recovers the latent structure a human would: reviews 1 and 3
(both about battery and charging) are close, as are reviews 2 and 4 (both about camera
and screen), while across-theme pairs are nearly orthogonal—exactly the perceptual map
logic at scale.

> TF-IDF is not a model of language; it is a model of *distinctiveness*. It will
> rank a term as informative whenever it is rare in the corpus, regardless of whether
> rarity tracks meaning. In a corpus of camera reviews, "camera" carries near-zero
> IDF and disappears—useful for retrieval, fatal if "camera" is the construct of
> interest. The corpus defines the contrast; choose it deliberately.

### Dimensionality Reduction {#sec-lsa}

Even weighted, the DTM is wide and collinear: synonyms ("photo", "picture", "image")
load on separate dimensions that mean the same thing. *Latent semantic analysis* (LSA)
compresses the DTM by a truncated singular value decomposition,
$\mathbf{X} \approx \mathbf{U}_k \boldsymbol{\Sigma}_k \mathbf{V}_k^{\top}$, projecting
documents into a $k$-dimensional *semantic space* in which synonymous terms collapse
onto shared directions. LSA is the linear-algebraic ancestor of the topic models in
@sec-topic-models and of the embeddings in @sec-embeddings: all three answer the same
complaint—that the BoW's dimensions are too many, too sparse, and too literal—but
they differ in whether the latent dimensions are interpretable (topics) or merely
predictive (embeddings).

### Word and Document Embeddings {#sec-embeddings}

The deepest limitation of count representations is that they are *one-hot* at the term
level: every word is a distinct dimension, equidistant from every other, so the model
has no prior that *excellent* is closer to *great* than to *battery*. *Embeddings*
replace this with *distributed* representations in which each term is a dense vector
$\mathbf{v}_w \in \mathbb{R}^d$ ($d$ typically 50–1000) learned so that geometric
proximity encodes semantic similarity. The learning principle is the *distributional
hypothesis*—words appearing in similar contexts have similar meanings—operationalized
by training a model to predict a word from its neighbors (or vice versa) across a
large corpus. Static word embeddings such as those of the word2vec and GloVe families
famously place synonyms near one another and encode analogies as vector offsets, and
they enter marketing as a way to measure semantic constructs: @hartmann2021power use
the geometry of language to study how powerful versus powerless brand communication
is perceived, and embedding-based features improve the prediction of consumer response
over BoW baselines [@hartmann2023more].

Two limitations of static embeddings motivate the modern default. First, they assign
one vector per word *type*, so the two senses of *bank* still collapse. Second, a
document is more than its words. *Contextual* embeddings from transformer models (the
BERT and GPT families) solve the first by producing token vectors that depend on the
*surrounding sentence*, so *bank* by a river and *bank* with interest receive
different representations. *Sentence and document embeddings* solve the second by
pooling token representations into a single vector $\mathbf{e}_i$ for the whole
document, which can then be fed to any downstream model exactly where a TF-IDF vector
would go—but carrying word order, negation, and context that the BoW threw away. This
is the representation underneath most current sentiment, classification, and retrieval
systems, and it is what makes LLM-based extraction (@sec-llm-extraction) possible.

@tbl-representations summarizes the trade-offs. The progression is monotone in
expressiveness and in opacity: each step captures more of language and explains less
of itself.

```{r}
#| label: tbl-representations
#| tbl-cap: "Text representations from sparse and transparent to dense and opaque. Expressiveness and opacity rise together."
#| message: false
#| warning: false
library(knitr)
repr <- data.frame(
  Representation = c("Bag-of-words", "TF-IDF", "Static embedding",
                     "Contextual embedding"),
  `Captures order` = c("No", "No", "No", "Yes"),
  `Captures context` = c("No", "No", "Partly", "Yes"),
  Interpretable = c("High", "High", "Low", "Low"),
  `Dimensionality` = c("|V| (sparse)", "|V| (sparse)", "d ~ 300", "d ~ 768+"),
  check.names = FALSE
)
kable(repr)
```

## Topic Models {#sec-topic-models}

Often the goal is not to predict a known label but to *discover* the latent themes a
corpus is about—the dimensions of quality consumers discuss, the issues a brand's
mentions cluster into, the topics a thousand call transcripts span. *Topic models* are
unsupervised generative models that posit a small set of latent *topics*, each a
probability distribution over the vocabulary, and explain each document as a mixture
of those topics. They turn a wide, sparse DTM into a narrow, interpretable
document–topic matrix, and they have become a marketing workhorse for reviews, social
media, and open-ended text [@tirunillai2014a; @buschken2016sentence].

### Latent Dirichlet Allocation {#sec-lda}

*Latent Dirichlet allocation* (LDA) is the canonical topic model. Its generative story
is the source of both its power and its assumptions. Fix $K$ topics. Each topic $k$ is
a distribution $\boldsymbol{\beta}_k$ over the $|V|$ vocabulary terms, drawn from a
Dirichlet prior. Each document $i$ has its own distribution $\boldsymbol{\theta}_i$
over the $K$ topics, also Dirichlet. Then every *word* in the document is generated by
first drawing a topic and then drawing a term from that topic's vocabulary
distribution:

$$
\begin{aligned}
\boldsymbol{\theta}_i &\sim \mathrm{Dirichlet}(\alpha), &
\boldsymbol{\beta}_k &\sim \mathrm{Dirichlet}(\eta), \\
z_{in} \mid \boldsymbol{\theta}_i &\sim \mathrm{Categorical}(\boldsymbol{\theta}_i), &
w_{in} \mid z_{in}, \boldsymbol{\beta} &\sim
\mathrm{Categorical}(\boldsymbol{\beta}_{z_{in}}),
\end{aligned}
$$ {#eq-lda}

for word positions $n = 1,\dots,N_i$, where $z_{in}$ is the (latent) topic assignment
of word $n$ and $w_{in}$ the observed term. Three assumptions are doing the work, and
each is a place identification can break. (i) *Bag-of-words*: words are exchangeable
within a document, so LDA inherits the BoW's blindness to order and negation. (ii) The
*Dirichlet prior* on $\boldsymbol{\theta}_i$ controls how concentrated documents are
on few topics; its hyperparameter $\alpha$ is a researcher choice that shapes the
solution. (iii) The *number of topics* $K$ is fixed in advance and is not identified
by the model—too few topics merge distinct themes, too many shatter one theme across
several, and there is no purely statistical oracle for the right $K$.

Estimation targets the posterior over the latent quantities
$p(\boldsymbol{\theta}, \boldsymbol{\beta}, \mathbf{z} \mid \mathbf{w})$, which is
intractable in closed form. Two estimators dominate. *Collapsed Gibbs sampling*
integrates out $\boldsymbol{\theta}$ and $\boldsymbol{\beta}$ and samples each word's
topic assignment $z_{in}$ conditional on all others, with the update probability
proportional to how often topic $k$ is used in document $i$ times how strongly topic
$k$ favors term $w_{in}$. *Variational inference* instead replaces the posterior with
a tractable factorized family and optimizes it to be as close as possible, trading
exactness for speed at scale. Both recover, for each document, an estimated topic
mixture $\hat{\boldsymbol{\theta}}_i$—a $K$-vector that *is* the document's reduced
representation—and, for each topic, its top-weighted terms, which the analyst reads to
*name* the topic.

That last step is where rigor is won or lost. **Topics are not labels; they are
distributions an analyst must interpret and validate.** A topic is identified only up
to the researcher's willingness to call its high-probability terms a coherent theme,
and two failure modes recur: a *junk* topic dominated by corpus-specific boilerplate,
and a *blended* topic mixing two themes that better $K$ or better priors would
separate. The honest practice is to choose $K$ by a combination of held-out
*perplexity* (predictive fit on unseen documents), *topic coherence* (whether a
topic's top terms co-occur in the corpus more than chance), and human inspection—and
to report sensitivity to that choice rather than presenting a single $K$ as given.

```{r lda-example, message=FALSE, warning=FALSE}
set.seed(49)
# Small synthetic corpus with two latent themes: "battery/charging" and "camera/screen"
docs <- c(
  "battery charging power drain battery slow charging",
  "camera photo screen bright sharp camera photo",
  "battery power slow charging drain battery",
  "screen camera sharp photo bright screen camera",
  "battery charging drain power slow battery charging",
  "photo camera screen sharp bright camera photo"
)
corp  <- strsplit(docs, "\\s+")
vocab <- sort(unique(unlist(corp)))
dtm <- t(sapply(corp, function(w)
  as.integer(table(factor(w, levels = vocab)))))
colnames(dtm) <- vocab

if (requireNamespace("topicmodels", quietly = TRUE)) {
  library(topicmodels)
  lda <- LDA(dtm, k = 2, control = list(seed = 49))
  # Top terms per topic (beta): what each topic is "about"
  terms_by_topic <- terms(lda, 4)
  print(terms_by_topic)
  # Document-topic mixtures (theta): the reduced representation
  print(round(posterior(lda)$topics, 2))
} else {
  message("Install 'topicmodels' to run this example.")
}
```

The estimated topics separate the battery/charging theme from the camera/screen theme,
and each document's $\hat{\boldsymbol{\theta}}_i$ places it on the corresponding
mixture—an interpretable, low-dimensional summary that a downstream regression can use.

### Extensions: Supervised and Structural Topic Models {#sec-stm}

Plain LDA is *unsupervised* and *unconditioned*: it ignores any document metadata and
any outcome. Two extensions matter for marketing. *Supervised* topic models attach a
response variable—a star rating, a sales figure—to each document and estimate topics
that are *predictive* of that response, so the discovered themes are the ones that move
the outcome rather than merely the ones that are frequent. *Structural topic models*
(STM) let topic *prevalence* and topic *content* depend on observed covariates—brand,
date, reviewer type—so the model can estimate, for example, how the share of
discussion devoted to "battery" shifts across product generations or differs between
verified and unverified reviewers. The marketing payoff is direct: @buschken2016sentence
build a sentence-level topic model for product reviews that respects the fact that a
single review discusses several attributes with different valence, recovering a more
faithful attribute-level structure than a document-level bag-of-words allows, and
@tirunillai2014a use latent-topic structure on user-generated content to extract the
dimensions of brand perception consumers actually talk about. The general lesson is
that conditioning the topic model on the structure of the marketing problem—attributes,
metadata, outcomes—buys both interpretability and downstream validity.

## Sentiment, Emotion, and Stance {#sec-49-sentiment}

A large share of text analytics in marketing reduces to one question: *how does the
writer feel?* The answer is layered. *Sentiment* (or *valence*) is the positive–
negative polarity of a text. *Emotion* is the finer affective state—joy, anger, fear,
sadness—often modeled as discrete categories or as continuous arousal and dominance
dimensions [@hartmann2021power]. *Stance* is the writer's *position toward a target*
(for or against a brand, a policy, a claim), which is distinct from sentiment: "I'm
furious that they discontinued it" is negative in valence but pro-brand in stance.
Conflating these three is a common and consequential error.

### Lexicon Methods {#sec-lexicon}

The simplest sentiment estimator is a *lexicon* (or dictionary): a list of words with
pre-assigned valence scores. The document score is an aggregation—typically the sum or
mean of its tokens' scores,
$$
s_i = \frac{1}{N_i} \sum_{n=1}^{N_i} \mathrm{val}(w_{in}),
$$ {#eq-lexicon-score}
where $\mathrm{val}(\cdot)$ looks each token up in the lexicon and unscored tokens
contribute zero. Lexicon scoring is transparent, fast, reproducible, and requires no
training data, which is why it remains the default for large-scale "social listening"
and for studies that need an auditable measure [@schweidel2014listening]. Its
assumptions are also its weaknesses, and they are severe. Scoring is *compositional in
the wrong way*: it sums token valences and so cannot represent *negation* ("not good"
scores positive), *intensification* ("extremely good" scores like "good"), or *
figurative inversion*—sarcasm and irony flip the intended meaning while leaving the
tokens positive, biasing the estimate in a *signed*, construct-correlated way rather
than merely adding noise (@sec-sarcasm). The lexicon is also *domain-blind*: "unpredictable"
is praise for a thriller and damnation for a car. The defensible use of a lexicon is as
a transparent, validated baseline whose biases are *bounded*, not as a black box whose
output is taken at face value.

### Supervised and Model-Based Sentiment {#sec-supervised-sentiment}

When labeled data exist—star ratings as a proxy, or human-coded sentiment—a *supervised*
model learns the mapping from representation to valence directly, absorbing negation,
intensification, and domain-specific usage from the data rather than from a fixed list.
This is simply the supervised-classification problem of @sec-classification with a
sentiment target, and contextual embeddings (@sec-embeddings) are what let modern
sentiment models read "not good" correctly. The marketing literature has repeatedly
shown that the *dimensions* of consumer affect—not just polarity—predict outcomes:
@tirunillai2012 link the valence and volume of user-generated content to abnormal stock
returns and trading volume, finding that *negative* word of mouth is the part that
moves firm value, and review-level valence correlates with sales [@chevalier2003measuring].
Word-of-mouth research further distinguishes *what* is said from *how* it is said:
emotional and self-referential language travels differently from neutral description
[@packard2016], and the linguistic features of a message—not only its sentiment—shape
its persuasive and diagnostic value [@packard2013; @melumad2020].

> **Sentiment is a construct, not a number.** Before scoring a corpus, the analyst
> must decide whether the target is valence, emotion, or stance; whether it is measured
> at the document, sentence, or aspect level; and against what human ground truth the
> measure will be validated. A pipeline that skips these decisions produces a column of
> numbers with no defensible interpretation.

### Aspect-Based Sentiment {#sec-aspect-sentiment}

A single review rarely has one sentiment. "The camera is superb but the battery is
hopeless" is positive about one attribute and negative about another, and a
document-level score averages the signal away. *Aspect-based sentiment analysis* (ABSA)
decomposes a document into (aspect, sentiment) pairs, asking not "is this review
positive?" but "what does it say, and how does it feel, about *each* attribute?" This
is the natural marriage of the topic models of @sec-topic-models with the sentiment
models of this section, and it is where the marketing value concentrates: aspect-level
structure tells a firm *which* attribute to fix, maps to the product's feature
hierarchy, and supports the perceptual maps of @sec-tfidf. Sentence-level and
aspect-level models are precisely the response to document-level averaging
[@buschken2016sentence].

## Supervised Text Classification {#sec-classification}

When the categories are known in advance—spam vs. ham, complaint vs. compliment, on-
vs. off-topic, fake vs. genuine review—the problem is *supervised classification*: learn
a function $f: \mathbf{x}_i \mapsto y_i$ from a *labeled* training sample
$\{(\mathbf{x}_i, y_i)\}_{i=1}^{n}$ and apply it to unlabeled documents. The
representation $\mathbf{x}_i$ is any of those in @sec-representation; the estimator can
be as simple as a linear model or as expressive as a fine-tuned transformer. What makes
text classification its own subject is the combination of high dimensionality, sparse
labels, and the need to validate against human judgment.

### From Naive Bayes to Regularized Logistic Regression {#sec-nb-logit}

The classical baseline is *multinomial naive Bayes*, which models the probability of a
class given a document by assuming words are conditionally independent given the class:
$$
p(y = c \mid d_i) \;\propto\; p(c) \prod_{j=1}^{|V|} p(\text{term}_j \mid c)^{\,c_{ij}}.
$$ {#eq-naive-bayes}
The independence assumption is false—words are correlated—but naive Bayes is fast,
needs little data, and is a stubbornly strong baseline. Its successor in most pipelines
is *$\ell_1$- or $\ell_2$-regularized logistic regression* on TF-IDF features, which
relaxes independence, handles correlated terms, and—through the penalty—survives the
high-dimensional regime where $|V|$ exceeds $n$. The regularizer is not optional
hygiene here; without it a model with more features than documents will fit the
training noise perfectly and generalize not at all. At the frontier, fine-tuned
contextual embeddings replace the linear model when enough labeled data exist and word
order matters, but the regularized linear baseline remains the right first model: it is
transparent, calibratable, and hard to beat on topical classification.

### Training, Evaluation, and the Threats to Validity {#sec-classification-validity}

The discipline of supervised text classification is mostly the discipline of honest
evaluation. Three threats recur. First, *leakage*: any preprocessing fit on the full
corpus—vocabulary selection, TF-IDF IDF weights, embeddings—must be learned on the
training fold only, or the held-out estimate of accuracy is optimistic. Second, *class
imbalance*: when one class is rare (fake reviews, churners, fraud), raw accuracy is
useless—a model that always predicts the majority class scores well and detects nothing.
The relevant metrics are *precision* (of the items flagged, how many are truly
positive), *recall* (of the truly positive items, how many are flagged), and their
harmonic mean the *F1* score, reported per class. Third, *distribution shift*: a
classifier trained on one platform, product category, or time period degrades on
another, because the joint distribution of language and label is not stable. The only
defense is out-of-distribution validation and periodic re-labeling—an estimate of
accuracy on data drawn like the deployment data, not like the training data.

```{r classification-eval, message=FALSE, warning=FALSE}
# Confusion matrix and the metrics that matter under class imbalance.
# 1000 reviews, 5% genuinely fake; a model flags some as fake.
set.seed(49)
truth <- factor(c(rep("fake", 50), rep("genuine", 950)))
# Simulated classifier: catches 35 of 50 fakes, falsely flags 40 genuine
pred <- truth
pred[sample(which(truth == "fake"), 15)]    <- "genuine"  # missed fakes
pred[sample(which(truth == "genuine"), 40)] <- "fake"     # false alarms

cm <- table(Predicted = pred, Actual = truth)
tp <- cm["fake", "fake"]; fp <- cm["fake", "genuine"]
fn <- cm["genuine", "fake"]
precision <- tp / (tp + fp)
recall    <- tp / (tp + fn)
f1 <- 2 * precision * recall / (precision + recall)
accuracy  <- sum(diag(cm)) / sum(cm)

cat("Accuracy :", round(accuracy, 3),
    " (a 'predict genuine always' model scores", round(950/1000, 3), ")\n")
cat("Precision:", round(precision, 3),
    " Recall:", round(recall, 3),
    " F1:", round(f1, 3), "\n")
```

Accuracy of 0.945 looks excellent yet conceals that the model misses a third of the
fakes—exactly the diagnosis precision, recall, and F1 are designed to surface and that
raw accuracy hides. A worked application of human-coded annotation as the ground truth
for such a classifier is developed in @sec-annotation.

## LLM-Based Extraction {#sec-llm-extraction}

The newest tool collapses much of the pipeline. A *large language model* (LLM) is a
transformer trained on web-scale text to predict the next token; the surprising
consequence is that, suitably *prompted*, it performs many text tasks—classification,
sentiment, summarization, entity and attribute extraction—with little or no
task-specific training data [@hartmann2023more]. The marketing use that matters most is
*structured extraction*: prompting the model to read an unstructured document (a review,
a call transcript, a complaint) and emit a structured record—attributes mentioned, their
sentiment, the customer's stated intent, whether a competitor was named—directly usable
in a downstream model. Where supervised classification needs a labeled training set per
task, an LLM can often produce a usable first pass *zero-shot* (from instructions alone)
or *few-shot* (from a handful of in-context examples), which is transforming the cost
structure of coding large corpora. @fig-llm-extraction sketches this schema-bearing
prompt-to-record workflow.

```{mermaid}
%%| label: fig-llm-extraction
%%| fig-cap: "LLM-based structured extraction: an unstructured document plus a schema-bearing prompt yields a structured record, which still requires validation against human labels before use."
flowchart LR
  A[Unstructured doc<br/>review / call / post] --> C[LLM]
  B[Prompt:<br/>schema + instructions<br/>+ few-shot examples] --> C
  C --> D[Structured record<br/>aspect, sentiment,<br/>intent, entities]
  D --> E{Validate vs.<br/>human labels}
  E -->|adequate| F[Downstream<br/>analysis]
  E -->|inadequate| B
```

The capability is real, but treating LLM output as data demands more discipline, not
less, because the failure modes are different and less visible than a misfit
regression. Four cautions are first-order. (i) *Hallucination*: an LLM can emit
confident, well-formed fields that are not supported by the document; extracted values
must be checked against the source, not trusted because they parse. (ii)
*Non-determinism and prompt sensitivity*: the same document under a slightly different
prompt, or the same prompt on a different day, can yield different output, so the prompt
and model version are part of the measurement instrument and must be fixed and reported.
(iii) *Train–test contamination and circularity*: if the construct being measured is
itself derived from the kind of text the model was trained on, the "measurement" may be
recovering the model's priors rather than the document's content. (iv) *Validation
remains mandatory*: an LLM extractor is an unvalidated classifier until its output has
been compared, on a held-out human-labeled sample, against the ground truth—the same
precision/recall/F1 discipline of @sec-classification-validity applies unchanged.
Marketing research using LLMs and machine learning on consumer text is accelerating
[@hartmann2023more; @ananthakrishnan2025book; @gao2024big], and the methodological
center of gravity is precisely this: the model is a powerful, cheap annotator whose
output must still earn its place in a regression by being validated like any other
measure. (For the LLM/API specifics—model selection, prompting, and structured-output
mechanics—the appropriate provider documentation should be consulted; this chapter
treats LLMs as measurement instruments, not as an engineering topic.)
<!-- CITE-NEEDED: foundational LLM / transformer references (e.g., attention, BERT, GPT) are CS-venue papers not in the marketing allowlist -->

## Marketing Applications {#sec-tad-applications}

The methods above are general; their value is in what they let a marketer measure.
Three settings dominate.

**Online reviews** are the most-mined consumer text, because they pair language with a
star rating and often with sales. Review *volume* and *valence* track product quality
and demand and move firm value, with negative content the most diagnostic
[@tirunillai2012; @chevalier2003measuring; @godes2004]. Aspect-based sentiment turns the
review corpus into an attribute-level scorecard that tells the firm which feature to fix
[@buschken2016sentence], and topic models recover the dimensions of brand perception
consumers actually discuss [@tirunillai2014a]. Reviews also raise their own measurement
hazards: figurative language biases naive sentiment (@sec-sarcasm), and the threat of
*fake* reviews makes the classification machinery of @sec-classification a quality-
control necessity rather than an academic exercise.

**Social media** supplies a continuous, unsolicited signal of brand health that firms
mine for "social listening" [@schweidel2014listening]. The text is short, noisy, and
laden with slang, sarcasm, and emoji, which strains lexicon methods and rewards
context-aware representations. The constructs of interest go beyond valence to *emotion*
and *stance*—how aroused, how powerful, how aligned the writer is toward the brand
[@hartmann2021power]—and to network-mediated diffusion, where *what* and *how* something
is said shapes whether it spreads (@sec-virality; @packard2016). Social text also feeds
back into measurement of constructs treated elsewhere: blog and post text predict and
explain marketing outcomes [@gopinath2013blogs; @gopinath2014investigating], and
user-generated images extend text measurement into the visual channel [@liu2020].

**Calls and conversations**—sales calls, service interactions, support chats—are the
frontier, because they are long, dyadic, and rich in the dynamics of persuasion and
emotion that static reviews lack. Transcribed and analyzed at scale, they let a firm
measure how a frontline agent's language drives outcomes, where in a conversation
sentiment turns, and which conversational moves resolve a complaint—linking directly to
the frontline-service and salesperson-value constructs of @sec-service and @sec-sales.
This is the natural home of LLM extraction (@sec-llm-extraction): a transcript is exactly
the kind of long, unstructured document from which a prompted model can pull a structured
record—intent, objections raised, resolution, sentiment trajectory—that no fixed lexicon
or bag-of-words could recover.

## Pitfalls and Identification {#sec-tad-pitfalls}

Text-as-data inherits every threat to validity that afflicts measurement, plus several
of its own. Naming them is the precondition for credible inference.

The deepest is that **the representation is an identifying assumption**. A bag-of-words
*cannot* represent negation or sarcasm, so any inference from it about valence is
conditional on those phenomena being rare or randomly distributed—an assumption that is
often false and rarely tested (@sec-sarcasm). The choice of representation is therefore
not a tuning detail but a substantive claim about what the text means.

A second hazard is **selection in who writes**. The corpus is generated by people who
chose to review, post, or call, and that selection is correlated with the outcome:
reviewers are disproportionately the delighted and the furious, producing the well-known
J-shaped ratings distribution, and social-media posters are not a random sample of
customers. Text measured on a selected sample estimates the sentiment of *writers*, not
of *customers*, and the gap is a bias, not noise. Methods that debias the poster
population are the appropriate response, not a larger but equally selected corpus.

A third is **endogeneity in downstream regressions**. When a text-derived measure—topic
share, sentiment, an LLM-extracted field—is used as a regressor explaining sales or firm
value, it is a *generated regressor* measured with error and often correlated with
omitted drivers. Plugging $\hat{s}_i$ into a regression as if it were observed
understates standard errors and can bias coefficients; the measurement model and the
outcome model should be treated as one system, and the error in the text measure
propagated. The marketing-finance and metrics chapters (@sec-marketing-finance,
@sec-metrics) develop the downstream inference that text measures feed.

Finally, **validation is not optional and is not a formality**. Every text measure—
lexicon score, topic label, classifier output, LLM extraction—is a claim about an
unobserved construct and must be validated against human-coded ground truth on a sample,
with an explicit reliability statistic, before it enters an analysis (@sec-annotation).
A pipeline that reports its accuracy on the data it was tuned on, or that never compares
its output to human judgment at all, has produced numbers, not measurements. The
discipline that distinguishes text-as-data from text mining is precisely this insistence
that a column of model output earn the status of data.

## Key Takeaways {#sec-tad-takeaways}

- **Representation is the first and most consequential choice.** Bag-of-words and TF-IDF
  are transparent but order- and context-blind (@eq-bow, @eq-tfidf); embeddings capture
  context at the cost of interpretability (@sec-embeddings). Match the representation to
  the construct, not to fashion.
- **Topic models discover structure but do not name it.** LDA's outputs are
  distributions an analyst must interpret and validate; $K$ and the priors are
  researcher choices, not data, and the solution is sensitive to them (@sec-lda).
- **Sentiment, emotion, and stance are distinct constructs.** Lexicons are auditable but
  cannot represent negation or sarcasm, biasing estimates in a *signed* way
  (@sec-lexicon, @sec-sarcasm); decide which construct, at which level, validated how,
  before scoring.
- **Supervised classification lives or dies by honest evaluation.** Under class
  imbalance, report precision, recall, and F1, guard against leakage, and validate
  out-of-distribution (@sec-classification-validity).
- **LLMs are powerful, cheap annotators—and unvalidated classifiers until proven
  otherwise.** Hallucination, prompt sensitivity, and contamination make validation
  *more* necessary, not less (@sec-llm-extraction).
- **Selection, generated-regressor error, and validation against ground truth** are the
  identification frontier; text measured on who chose to write estimates writers, not
  customers (@sec-tad-pitfalls).
