# Discrete Choice and Bayesian Methods {#sec-choice-bayesian}

Most marketing decisions reduce to a question about choice: which brand a shopper
puts in the cart, which subscription tier a household keeps, which car a buyer
drives off the lot. The data that record these decisions are not the continuous
quantities of classical regression but *discrete selections* from a menu—one
alternative chosen, the rest forgone. Discrete-choice modeling, grounded in
random-utility theory, is the apparatus marketing uses to turn such observations
into estimates of preference, willingness to pay, and demand elasticities; the
Bayesian hierarchy is the technology that makes those estimates individual-level
and usable for targeting.

This chapter is organized as a full-semester doctoral seminar—a reading map that
runs from the micro-foundation of individual choice up to the modern Bayesian
computation that fits these models at scale. The structure mirrors how the
discrete-choice/Bayesian-methods seminar is taught across the leading quantitative
marketing programs (Ohio State/Fisher, Chicago/Booth, Wharton, Columbia,
Michigan/Ross), and every reading below was DOI-verified against Crossref. The
technical core—random-utility theory, the logit and its IIA pathology, the
nested/probit/mixed remedies, hierarchical Bayes, and a runnable MCMC worked
example—is developed in full within the relevant weeks, so the chapter reads both
as a syllabus and as a self-contained treatment. We connect the demand-estimation
thread to its industrial-organization counterpart (@sec-marketing-finance) and to
preference measurement via conjoint analysis, whose hierarchical-Bayes practice is
the applied face of everything developed here.

## Semester arc {#sec-cb-semester-arc}

The seminar is a single intellectual staircase from **theory of individual choice**
up to **modern Bayesian computation and scale**. It opens with the micro-foundation
that organizes everything else—random utility maximization (RUM)—and the workhorse
it produces, the multinomial/conditional logit. The first third of the course
exhausts what closed-form logit can and cannot do: the
independence-of-irrelevant-alternatives (IIA) restriction, the substitution
pathologies it implies, and the two classical escapes from it—*generalized extreme
value* (nested logit), which keeps a closed form, and *probit*, which abandons
closed form and forces the course into simulation (the GHK simulator, simulated
maximum likelihood). By the time random-coefficients/mixed logit arrives, students
see it as the field's preferred resolution: it relaxes IIA, accommodates unobserved
heterogeneity, approximates any RUM, and is estimable either by simulation or by
Bayesian MCMC.

The middle third pivots from estimators to the **Bayesian paradigm**, the
distinctive emphasis of the marketing version of this seminar. Students learn the
Bayesian logic of inference, conjugacy, and the two engines of modern practice—the
Gibbs sampler and the Metropolis–Hastings algorithm—and then the idea that made
Bayes the default tool of quantitative marketing: **hierarchical Bayes for consumer
heterogeneity**. The hierarchical normal model that shrinks individual-level
parameters toward a population distribution is the conceptual heart of the course,
because it simultaneously solves the small-data-per-consumer problem, delivers
individual-level posteriors useful for targeting, and unifies choice modeling with
conjoint/preference measurement.

The final third applies and extends the machinery: **conjoint and preference
measurement** as the marketing-native application of HB choice models; **dynamics
and state dependence** (Bayesian learning vs. spurious persistence—the field's
central identification debate); **aggregate-data demand** (BLP-style
random-coefficients logit and its Bayesian counterpart); **model comparison** (Bayes
factors, DIC/WAIC, cross-validation); and the computational frontier—**Hamiltonian
Monte Carlo/Stan** and **scalable and variational methods** for the large, sparse,
high-dimensional data that now dominate marketing. The arc ends where the field's
frontier is: probabilistic programming and approximate inference at scale.

A typical assessment structure is weekly reading memos, two or three problem sets
that build an estimator from scratch (a hand-coded Gibbs sampler for a hierarchical
logit is the signature assignment—see the worked example in @sec-cb-week-7), a referee
report, and a replication-or-extension paper on a real choice dataset (scanner
panel, conjoint, or aggregate market shares).

The three foundational books the seminar is built around recur throughout and are
cited as "(book)": **Train (2009), *Discrete Choice Methods with Simulation*, 2nd
ed. (Cambridge)**—the standard graduate text for RUM, logit/GEV/probit, mixed
logit, and simulation- and Bayes-assisted estimation; **Rossi, Allenby & McCulloch
(2005), *Bayesian Statistics and Marketing* (Wiley)**—the field's defining text on
hierarchical Bayesian methods for marketing and the `bayesm` package; and **Gelman
et al. (2013), *Bayesian Data Analysis*, 3rd ed. (Chapman & Hall/CRC)**—the
general-statistics reference for Bayesian modeling, MCMC, and predictive evaluation.

### Week 1 — Random utility theory & the choice-modeling worldview {#sec-cb-week-1}

**Topic.** The random-utility-maximization (RUM) foundation; why marketing models
choice rather than continuous demand; utility, choice probabilities, and the
identification of scale and level.

**Subtopics.** Deterministic vs. random components; additive RUM; the logit error
(Type-I extreme value); welfare and the log-sum; observable vs. latent utility;
normalization (only differences and scale are identified).

**Methods.** Setting up a choice probability from a utility specification; deriving
logit from i.i.d. extreme-value errors.

The premise of random-utility theory is that a decision maker assigns a latent
utility to each available alternative and chooses the one with the highest utility.
The randomness is the analyst's, not the consumer's: it represents everything that
bears on the decision but is unobserved by the modeler.

::: {.callout-note}
## Landmark definition: random utility

> The utility that decision maker $n$ obtains from alternative $j$ is
> $U_{nj} = V_{nj} + \varepsilon_{nj}$, where $V_{nj}$ is a *representative* (or
> *systematic*) component, a function of observed attributes and parameters, and
> $\varepsilon_{nj}$ is a random term capturing unobserved determinants of utility.
> The decision maker chooses alternative $i$ if and only if $U_{ni} \ge U_{nj}$ for
> all $j$ in the choice set. The choice probability is therefore a statement about
> the distribution of the differences in the unobserved terms
> [@mcfadden2001economic].
:::

Formally, let $n = 1,\dots,N$ index decision makers and let $\mathcal{C}_n$ be the
finite choice set facing decision maker $n$. The utility of alternative
$j \in \mathcal{C}_n$ is

$$
U_{nj} = V_{nj} + \varepsilon_{nj}, \qquad
V_{nj} = \mathbf{x}_{nj}^{\top}\boldsymbol{\beta},
$$ {#eq-50-rum}

where $\mathbf{x}_{nj}$ collects observed attributes of the alternative (price,
features, brand indicators) and possibly interactions with characteristics of the
chooser, and $\boldsymbol{\beta}$ is a vector of taste parameters. The probability
that $n$ chooses $i$ is the probability that $i$ yields the highest utility,

$$
P_{ni} = \Pr\!\left( U_{ni} \ge U_{nj}\ \ \forall j \in \mathcal{C}_n \right)
       = \Pr\!\left( \varepsilon_{nj} - \varepsilon_{ni} \le V_{ni} - V_{nj}\ \
         \forall j \neq i \right).
$$ {#eq-choiceprob}

Three structural facts follow immediately and govern everything downstream. First,
**only differences in utility matter**: adding a constant to every alternative's
utility leaves the choice unchanged, so alternative-invariant variables (a chooser's
income entering identically for all options) drop out and must be interacted with
alternative-specific terms to have any effect. Second, the **scale of utility is not
identified**: multiplying all utilities by a positive constant leaves the argmax
unchanged, so the variance of $\varepsilon_{nj}$ must be normalized. Third, the form
of the choice probability is determined entirely by the **assumed joint distribution
of the unobserved terms** $\boldsymbol{\varepsilon}_n = (\varepsilon_{n1},\dots)$.
Different assumptions generate the logit, probit, nested, and mixed models that
organize the rest of the course. @fig-cb-rum assembles the framework, tracing
utility from its observed and unobserved components through the argmax rule to
the choice probability.

```{mermaid}
%%| label: fig-cb-rum
%%| fig-cap: "The random-utility framework. Observed attributes and taste parameters form the systematic component; everything the analyst cannot observe enters the random term. The argmax rule turns latent utility into a choice, and the assumed distribution of the unobserved terms determines the form of the choice probability—and with it the model family of the course."
flowchart LR
  X["Observed attributes x_nj<br/>(price, features, brand)"] --> V["Systematic component<br/>V_nj = x_nj'β"]
  B["Taste parameters β"] --> V
  W["Unobserved determinants<br/>(the analyst's ignorance)"] --> E["Random term ε_nj"]
  V --> U["Latent utility<br/>U_nj = V_nj + ε_nj"]
  E --> U
  U --> R["Choice rule:<br/>pick i iff U_ni ≥ U_nj for all j"]
  R --> P["Choice probability P_ni,<br/>set by the distribution of ε"]
  P --> M["Distributional assumption →<br/>logit, probit, nested, mixed"]
```

**Key readings.**

- McFadden (1974), "Conditional Logit Analysis of Qualitative Choice Behavior," in
  *Frontiers in Econometrics* (Academic Press), pp. 105–142 — book chapter, no
  Crossref DOI; cite by chapter/volume. The origin of RUM-based conditional logit;
  the Nobel-cited founding paper. [F]
- Train (2009), *Discrete Choice Methods with Simulation*, ch. 1–3 (book) — the
  textbook RUM setup and logit derivation. [F]
- @chintagunta2011discretechoice — assigned here (and again in @sec-cb-week-9) as the
  marketing-specific overview of why and how marketing uses discrete choice. [F]

**Debate.** Is RUM a behavioral theory or just a flexible statistical container?
What is genuinely identified—and the perennial student trap of interpreting levels
or scale?

### Week 2 — Logit: the workhorse {#sec-cb-week-2}

**Topic.** Multinomial/conditional logit; estimation, interpretation, elasticities,
and willingness-to-pay.

**Subtopics.** Likelihood and concavity (global maximum); own/cross elasticities;
the log-sum/inclusive value; aggregation; the value and limits of marginal effects.

**Methods.** Maximum likelihood; Newton-type optimization; delta-method standard
errors for WTP and elasticities.

The conditional (or multinomial) logit model arises from a single distributional
assumption: the unobserved utility terms are independent across alternatives and
each follows a type-I extreme-value (Gumbel) distribution,
$F(\varepsilon_{nj}) = \exp(-e^{-\varepsilon_{nj}})$. The difference of two
independent Gumbel variates is logistic, which collapses the multidimensional
integral in @eq-choiceprob to a closed form:

$$
P_{ni} = \frac{\exp(\mathbf{x}_{ni}^{\top}\boldsymbol{\beta})}
              {\sum_{j \in \mathcal{C}_n} \exp(\mathbf{x}_{nj}^{\top}\boldsymbol{\beta})}.
$$ {#eq-50-logit}

This is the model that brought random-utility estimation into marketing. Calibrated
on household scanner data, with $\mathbf{x}_{nj}$ containing price, promotion
indicators, and—critically—a *loyalty* variable summarizing the household's past
purchases, it predicts brand choice well enough to anchor demand analysis for
packaged goods [@guadagni1983; @guadagni2008logit; @guadagni1983logit]. The loyalty
term is an early, informal device for capturing persistence in tastes that the
homogeneous logit cannot otherwise represent, foreshadowing the heterogeneity
machinery developed below.

**Estimation.** With independent observations the logit log-likelihood is globally
concave in $\boldsymbol{\beta}$, so maximum likelihood is straightforward and
unique. Writing $y_{ni} = 1$ if $n$ chose $i$ and $0$ otherwise,

$$
\ell(\boldsymbol{\beta}) = \sum_{n=1}^{N} \sum_{i \in \mathcal{C}_n}
   y_{ni} \log P_{ni}(\boldsymbol{\beta}),
$$ {#eq-loglik}

which Newton–Raphson or any quasi-Newton routine maximizes reliably. The scale
normalization is implicit: the Gumbel variance is fixed at $\pi^2/6$, so
$\boldsymbol{\beta}$ is identified only up to that scale, and coefficients across
models with different error variances are not directly comparable.

**Elasticities.** The managerial output of a choice model is usually an elasticity.
For a price coefficient $\beta_p$ entering $V_{nj}$ linearly, the own- and
cross-price elasticities of choice probability are

$$
E_{ii} = \beta_p\, p_{ni}\,(1 - P_{ni}), \qquad
E_{ij} = -\,\beta_p\, p_{nj}\, P_{nj}.
$$ {#eq-elast}

The cross-elasticity $E_{ij}$ is the source of the model's most consequential
defect: it depends on $j$ only through $P_{nj}$, so *every* alternative has the same
proportional cross-elasticity with respect to a change in $j$. This is IIA, examined
in @sec-cb-week-3.

**Key readings.**

- Train (2009), ch. 3 "Logit" (book) — canonical treatment. [F]
- @guadagni1983logit — the foundational marketing application of logit to
  scanner-panel brand choice; introduced the loyalty variable and defined empirical
  choice modeling in marketing for a decade. [F]

**Debate.** When is logit "good enough"? The gap between statistical fit and the
implausible substitution it can imply motivates Weeks 3–5.

### Week 3 — IIA, GEV & nested logit {#sec-cb-week-3}

**Topic.** The independence-of-irrelevant-alternatives property, its
substitution-pattern consequences, the red-bus/blue-bus problem, and the closed-form
escape via generalized extreme value (GEV)/nested logit.

**Subtopics.** IIA derivation; proportional substitution; nesting structure and the
dissimilarity (nesting) parameter; testing IIA.

**Methods.** Nested-logit estimation; Hausman-type specification tests.

The logit model's defining property is that the ratio of any two choice
probabilities is independent of the attributes—indeed of the existence—of every
other alternative:

$$
\frac{P_{ni}}{P_{nj}}
   = \frac{\exp(\mathbf{x}_{ni}^{\top}\boldsymbol{\beta})}
          {\exp(\mathbf{x}_{nj}^{\top}\boldsymbol{\beta})},
$$ {#eq-50-iia}

which contains nothing about any third option $k$. This **independence of
irrelevant alternatives** (IIA) is a direct consequence of the assumed independence
of the Gumbel error terms across alternatives. It is mathematically elegant and
behaviorally implausible.

The standard illustration is the *red-bus/blue-bus* paradox. A commuter chooses
between a car and a red bus with equal probability, $\tfrac{1}{2}$ each, so
$P_{\text{car}}/P_{\text{red}} = 1$. Introduce a blue bus identical to the red bus
in every respect but color. Behaviorally the two buses should split the original bus
share, leaving car at $\tfrac{1}{2}$ and each bus at $\tfrac{1}{4}$. The logit
model instead preserves all pairwise ratios, forcing
$P_{\text{car}} = P_{\text{red}} = P_{\text{blue}} = \tfrac{1}{3}$: adding a
near-duplicate *steals* share from the car as readily as from the bus it nearly
clones. In a marketing setting the same failure means a logit demand model predicts
that a new product cannibalizes every incumbent in proportion to its share,
regardless of how close a substitute it actually is—a prediction that misvalues line
extensions, new entrants, and assortment changes. @fig-cb-nest contrasts the two
predictions and shows the nesting structure that repairs the failure.

```{mermaid}
%%| label: fig-cb-nest
%%| fig-cap: "The red-bus/blue-bus problem and its nested-logit repair. Plain logit preserves all pairwise ratios, so the duplicate bus steals share from the car as readily as from its near-clone; nesting the two buses—alternatives that share unobserved attributes, with correlation governed by the dissimilarity parameter—confines proportional substitution to within the nest and restores the behaviorally sensible split."
flowchart TD
  subgraph LG["Plain logit (IIA)"]
    L0["Choice among<br/>three modes"] --> L1["Car<br/>P = 1/3"]
    L0 --> L2["Red bus<br/>P = 1/3"]
    L0 --> L3["Blue bus<br/>P = 1/3"]
  end
  subgraph NL["Nested logit"]
    N0["Choice among<br/>nests"] --> N1["Car<br/>P = 1/2"]
    N0 --> N2["Bus nest<br/>(shared unobserved attributes;<br/>dissimilarity parameter)"]
    N2 --> N3["Red bus<br/>P = 1/4"]
    N2 --> N4["Blue bus<br/>P = 1/4"]
  end
```

::: {.callout-warning}
## What IIA gets wrong, and when it is safe

IIA is the right model only when alternatives are, conditional on the observed
$\mathbf{x}_{nj}$, equally substitutable for one another. It fails whenever subsets
of alternatives share unobserved attributes—body styles of cars, flavor families of
snacks, branded versus private-label tiers. The practical diagnostic is a
Hausman–McFadden test: estimate the model on the full choice set and on a subset; if
IIA holds, dropping alternatives should not change $\hat{\boldsymbol{\beta}}$, and a
significant change rejects IIA. When the test rejects, the logit's substitution
patterns are not merely imprecise—they are structurally wrong, and the remedies of
the next sections are required rather than optional.
:::

The IIA property is also why the logit can be estimated on a *sampled* subset of
alternatives when the full choice set is enormous (thousands of SKUs): because ratios
of probabilities ignore the rest of the set, a correctly weighted sample of
non-chosen alternatives yields consistent estimates—the same property that makes the
model behaviorally restrictive makes it tractable at scale.

**Key readings.**

- @hausman1984specification — the standard test for the IIA restriction; defines how
  students diagnose logit's central weakness. [F]
- McFadden (1978), "Modelling the Choice of Residential Location," in *Spatial
  Interaction Theory and Planning Models* (North-Holland) — book chapter, no
  Crossref DOI; the GEV/nested-logit foundation. [F]
- Train (2009), ch. 4 "GEV" (book) — textbook synthesis of nested and cross-nested
  logit. [F]

**Debate.** IIA as bug vs. feature; whether nesting structure is theory-driven or
data-mined; nested logit as a stopgap before mixed logit/probit.

### Week 4 — Probit & simulation: the GHK simulator {#sec-cb-week-4}

**Topic.** Multinomial probit (MNP), full correlation in errors, the absence of a
closed form, and simulation-based estimation (GHK).

**Subtopics.** Identification in MNP (covariance normalization); the GHK
(Geweke–Hajivassiliou–Keane) recursive simulator; simulated maximum likelihood (SML)
vs. method of simulated moments (MSM); Bayesian (Gibbs/data-augmentation) MNP as the
alternative path.

**Methods.** GHK simulation; SML/MSM; Gibbs sampling with latent-utility data
augmentation.

The multinomial probit model replaces the independent Gumbel errors with a
multivariate normal vector, $\boldsymbol{\varepsilon}_n \sim
\mathcal{N}(\mathbf{0}, \boldsymbol{\Sigma})$. The off-diagonal elements of
$\boldsymbol{\Sigma}$ let any pair of alternatives share unobserved utility, so
probit imposes no IIA and can represent arbitrary substitution patterns. The price
is analytic: the choice probability in @eq-choiceprob is now a
$(J-1)$-dimensional integral of the normal density with no closed form,

$$
P_{ni} = \int_{\mathbb{R}^{J-1}} \mathbb{1}\!\left[ U_{ni} \ge U_{nj}\ \forall j \neq i
   \right] \phi(\boldsymbol{\varepsilon}_n; \boldsymbol{\Sigma})\, d\boldsymbol{\varepsilon}_n .
$$ {#eq-probit}

For more than three or four alternatives this integral must be approximated by
simulation. McFadden's **method of simulated moments** made probit estimation
feasible by replacing the intractable probabilities with smooth, unbiased
frequency-simulator analogues [@mcfadden1986method]. A further subtlety is
identification: because only utility differences matter and scale is free, not all
elements of $\boldsymbol{\Sigma}$ are identified—one differences the system and fixes
one variance, leaving $J(J-1)/2 - 1$ free covariance parameters. The Bayesian route
sidesteps the simulated-likelihood maximization entirely by augmenting the latent
utilities and sampling them as additional unknowns.

**Key readings.**

- @mcculloch1994exactprobit — the Bayesian/data-augmentation route to MNP that became
  the marketing standard; bridges Weeks 4 and 6. [F]
- @geweke1994probit — the head-to-head of GHK/SML vs. Bayesian MCMC for MNP; defines
  the simulation toolkit and its trade-offs. [F]
- Train (2009), ch. 5 "Probit" and ch. 9 "Simulation-Assisted Estimation" (book) —
  textbook GHK and SML/MSM. [F]

**Debate.** Probit's flexible correlation vs. its identification fragility and
computational cost; classical simulation vs. Bayesian data augmentation as the more
reliable estimator.

### Week 5 — Mixed / random-coefficients logit {#sec-cb-week-5}

**Topic.** Mixed (random-parameters) logit: heterogeneous tastes integrated over a
mixing distribution; the field's preferred way to relax IIA and approximate any RUM.

**Subtopics.** Random vs. fixed coefficients; mixing distributions (normal,
lognormal, bounded); panel/repeated-choice mixed logit; the McFadden–Train
universal-approximation result; classical (simulated ML) vs. Bayesian
(Allenby–Train) estimation.

**Methods.** Simulated maximum likelihood with Halton draws; hierarchical
Bayes/Gibbs–MH estimation of the mixing distribution.

Mixed logit is the model that subsumes the others. Its idea is to let the taste
parameters themselves vary across decision makers according to a *mixing
distribution*, $\boldsymbol{\beta}_n \sim g(\boldsymbol{\beta} \mid
\boldsymbol{\theta})$, while retaining an iid Gumbel idiosyncratic term. Conditional
on a draw of $\boldsymbol{\beta}_n$ the model is a plain logit, but the unconditional
choice probability integrates over the distribution of tastes,

$$
P_{ni} = \int
   \frac{\exp(\mathbf{x}_{ni}^{\top}\boldsymbol{\beta})}
        {\sum_{j \in \mathcal{C}_n}\exp(\mathbf{x}_{nj}^{\top}\boldsymbol{\beta})}\,
   g(\boldsymbol{\beta} \mid \boldsymbol{\theta})\, d\boldsymbol{\beta}.
$$ {#eq-mixedlogit}

This single device accomplishes three things at once. It **breaks IIA**: because the
mixing correlates the random part of utility across alternatives that share
attributes, cross-elasticities are no longer constrained to be equal. It
**accommodates heterogeneity**: $\boldsymbol{\theta}$ describes how tastes are
distributed in the population. And it is **fully flexible**: any random-utility
model—including probit—can be approximated arbitrarily well by a mixed logit with a
suitable mixing distribution. The cost, as with probit, is that the integral in
@eq-mixedlogit has no closed form and must be simulated; classical estimation uses
simulated maximum likelihood with draws of $\boldsymbol{\beta}_n$, while the Bayesian
approach (@sec-cb-week-7) targets the same integral through MCMC and is, for the standard
normal-mixing case, often the more reliable route [@rossi2014; @allenby1999dynamic].

The family is mapped in @fig-choice-family, which organizes every model in the course
by the assumed structure of the unobserved utility terms.

```{mermaid}
%%| label: fig-choice-family
%%| fig-cap: "The discrete-choice model family, organized by the assumed structure of the unobserved utility terms. Each branch relaxes a restriction of the conditional logit and pays for it in estimation cost."
flowchart TD
    A["Random-utility model<br/>U = V + e"] --> B{"Distribution of<br/>unobserved e?"}
    B -->|"iid Gumbel"| C["Conditional logit<br/>closed form, imposes IIA"]
    B -->|"correlated normal"| D["Multinomial probit<br/>full covariance, no IIA"]
    B -->|"Gumbel within nests,<br/>correlated across"| E["Nested logit<br/>IIA within nest only"]
    B -->|"Gumbel + random beta_n"| F["Mixed (random-coefficients) logit<br/>any substitution pattern"]
    F --> G["Heterogeneity as a<br/>mixing distribution"]
    G --> H["Hierarchical Bayes<br/>estimation via MCMC"]
```

@tbl-iia-remedies summarizes the trade-offs across the IIA remedies of Weeks 3–5.

| Model | Error structure | IIA? | Closed form? | Main cost |
|---|---|---|---|---|
| Conditional logit | iid Gumbel | Imposed | Yes | Wrong substitution under correlation |
| Nested logit | Gumbel within nests | Within nest only | Yes | Nesting must be assumed a priori |
| Multinomial probit | Multivariate normal | No | No (simulate) | Many covariance params; identification |
| Mixed logit | Gumbel + random $\boldsymbol{\beta}_n$ | No | No (simulate) | Choice of mixing distribution |

: Remedies for IIA and their trade-offs. Mixed logit dominates on flexibility but inherits the simulation cost shared by every non-logit model. {#tbl-iia-remedies}

**Key readings.**

- @mcfadden2000mixedmnl — proves mixed logit can approximate any random-utility model
  arbitrarily well; the theoretical license for the whole approach. [F]
- @revelt1998mixedlogit — the canonical panel/repeated-choice mixed-logit
  application; template for individual-level taste recovery. [F]
- @train2005boundedmixedlogit — bounded/correlated mixing distributions so estimated
  partworths stay economically sensible; a standard refinement for applied mixed
  logit and conjoint. [F]
- Train (2009), ch. 6 "Mixed Logit" and ch. 12 (Bayesian procedures) (book). [F]

**Debate.** Choice of mixing distribution (normal vs. lognormal vs. nonparametric);
whether random coefficients capture *taste* heterogeneity or absorb misspecification;
simulated-ML vs. Bayesian estimation.

### Week 6 — Bayesian foundations & MCMC: Gibbs and Metropolis–Hastings {#sec-cb-week-6}

**Topic.** The Bayesian paradigm and the two computational engines that made it
practical for choice models.

**Subtopics.** Prior, likelihood, posterior; conjugacy; data augmentation (latent
utilities) for choice models; the Gibbs sampler; Metropolis–Hastings; convergence
diagnostics and mixing.

**Methods.** Hand-coding a Gibbs sampler; MH proposals and acceptance; assessing
convergence.

MCMC constructs a Markov chain whose stationary distribution is exactly the
posterior; running the chain long enough and discarding an initial *burn-in* yields
dependent draws from that posterior, from which any quantity of interest is computed
as a sample average. Two building blocks suffice for choice models. **Gibbs
sampling** cycles through the parameter blocks, drawing each from its *full
conditional* distribution—exploiting conjugacy so that conditionals reduce to
textbook draws. **The Metropolis–Hastings step** handles blocks that are not
conjugate: a candidate is proposed and accepted with a probability that depends only
on the likelihood and prior, never on the intractable normalizing constant. For
probit and multinomial models, **data augmentation** restores conjugacy by sampling
the latent utilities as additional unknowns. The full mechanics—including the
acceptance ratio and one sampler sweep—are developed concretely in @sec-cb-week-7.
@fig-cb-bayes places these engines in the full workflow from prior and likelihood
to usable posterior draws.

```{mermaid}
%%| label: fig-cb-bayes
%%| fig-cap: "The Bayesian workflow for choice models. Prior and likelihood combine into a posterior whose normalizing constant is intractable; MCMC constructs a Markov chain with that posterior as its stationary distribution—Gibbs draws for conjugate blocks, Metropolis-Hastings for non-conjugate ones, and data augmentation to restore conjugacy in probit and multinomial models. After burn-in and convergence checks, the retained draws deliver any quantity of interest as a sample average."
flowchart LR
  PR["Prior"] --> PO["Posterior<br/>(normalizing constant<br/>intractable)"]
  LK["Likelihood of the<br/>observed choices"] --> PO
  PO --> MC["MCMC: Markov chain whose<br/>stationary distribution<br/>is the posterior"]
  MC --> G["Gibbs sampler:<br/>conjugate blocks drawn<br/>from full conditionals"]
  MC --> MH["Metropolis-Hastings:<br/>non-conjugate blocks"]
  MC --> DA["Data augmentation:<br/>latent utilities sampled<br/>as unknowns"]
  G --> BI["Discard burn-in;<br/>check convergence and mixing"]
  MH --> BI
  DA --> BI
  BI --> QI["Posterior draws:<br/>quantities of interest as<br/>sample averages"]
```

**Key readings.**

- @gelfand1990sampling — the paper that launched modern applied MCMC by popularizing
  the Gibbs sampler for marginal-density computation. [F]
- @casella1992gibbs — the pedagogically standard, intuition-first introduction to
  Gibbs sampling. [F]
- @chib1995metropolis — the companion intuition-first treatment of
  Metropolis–Hastings; with Casella–George it is the canonical "learn MCMC" pair. [F]
- Gelman et al. (2013), *Bayesian Data Analysis* (book) — the reference text for the
  underlying theory. [F]

**Debate.** Subjective vs. objective priors; how much prior matters in hierarchical
models; diagnosing convergence honestly (autocorrelation, multiple chains).

### Week 7 — Hierarchical Bayes for heterogeneity (the heart of the course) {#sec-cb-week-7}

**Topic.** Hierarchical (multilevel) Bayesian models of consumer heterogeneity;
shrinkage; individual-level posteriors for targeting.

**Subtopics.** The hierarchical normal prior over individual coefficients;
random-effects logit/probit; borrowing strength; the value of individual-level
inference; scale-usage and related nuisance heterogeneity.

**Methods.** Two-stage Gibbs for hierarchical logit/probit; posterior summaries at
the individual level.

The mixing distribution $g(\boldsymbol{\beta}\mid\boldsymbol{\theta})$ of
@eq-mixedlogit is where heterogeneity lives, and treating it carefully is what
separates a model that merely fits aggregate shares from one that supports targeting,
segmentation, and individual-level prediction. The Bayesian hierarchy is the natural
language for this. The key obstacle is that each consumer supplies only a handful of
choices, far too few to estimate an individual taste vector $\boldsymbol{\beta}_n$ on
its own. The hierarchical model solves this by *borrowing strength*: it assumes
individuals are drawn from a common population distribution and lets each person's
estimate shrink toward the population mean by an amount governed by how informative
that person's own data are. A three-level specification makes this concrete:

$$
\begin{aligned}
\text{(likelihood)}\quad & y_{ni} \mid \boldsymbol{\beta}_n
   \ \sim\ \text{Logit}\big(\mathbf{x}_{ni}^{\top}\boldsymbol{\beta}_n\big), \\[2pt]
\text{(heterogeneity / prior)}\quad & \boldsymbol{\beta}_n
   \ \sim\ \mathcal{N}\big(\bar{\boldsymbol{\beta}},\, \mathbf{V}_\beta\big), \\[2pt]
\text{(hyperprior)}\quad & \bar{\boldsymbol{\beta}}
   \ \sim\ \mathcal{N}(\boldsymbol{\mu}_0,\, \mathbf{A}^{-1}), \qquad
   \mathbf{V}_\beta \ \sim\ \text{Inverse-Wishart}(\nu_0, \mathbf{V}_0).
\end{aligned}
$$ {#eq-hierarchy}

The middle line is exactly the mixing distribution of mixed logit, now read as a
*prior* on each individual's tastes whose parameters
$(\bar{\boldsymbol{\beta}}, \mathbf{V}_\beta)$ are themselves estimated. An
individual with many observations is driven by the first line (their own data); an
individual with few is pulled toward $\bar{\boldsymbol{\beta}}$ by the second. This
shrinkage is not a regularization hack but the optimal pooling of individual and
population information under the model, and it is the reason hierarchical Bayes
recovers usable individual-level estimates from sparse data—the property that made it
the standard estimator for choice-based conjoint and household-level demand
[@rossi2014; @kamakura2005choice; @allenby1998heterogeneity].

::: {.callout-note}
## Why Bayesian, not just penalized likelihood?

The hierarchical structure could in principle be fit by maximizing a penalized
likelihood (empirical Bayes), and for the population parameters the two agree
asymptotically. The Bayesian apparatus earns its keep on the *individual-level*
posteriors: it delivers a full posterior distribution for each
$\boldsymbol{\beta}_n$—hence honest uncertainty for targeting decisions—and it
sidesteps the simulated-likelihood maximization that is numerically fragile for
high-dimensional mixing distributions. The cost is the construction and tuning of an
MCMC sampler.
:::

#### The MCMC engine

MCMC makes the hierarchy in @eq-hierarchy estimable. **Gibbs sampling** cycles
through the parameter blocks. Two of the three conditionals are conjugate and drawn
directly: given the individual $\boldsymbol{\beta}_n$, the population mean
$\bar{\boldsymbol{\beta}}$ has a normal full conditional and the covariance
$\mathbf{V}_\beta$ an inverse-Wishart full conditional—which is precisely why the
normal–inverse-Wishart hyperprior is chosen. **The Metropolis–Hastings step** handles
the non-conjugate individual logit block. A candidate
$\boldsymbol{\beta}_n^{\star}$ is proposed (commonly by a random walk) and accepted
with probability

$$
\alpha = \min\!\left\{ 1,\ \frac{p(y_n \mid \boldsymbol{\beta}_n^{\star})\,
   \phi(\boldsymbol{\beta}_n^{\star} \mid \bar{\boldsymbol{\beta}}, \mathbf{V}_\beta)}
   {p(y_n \mid \boldsymbol{\beta}_n)\,
   \phi(\boldsymbol{\beta}_n \mid \bar{\boldsymbol{\beta}}, \mathbf{V}_\beta)} \right\}.
$$ {#eq-mh}

The acceptance ratio depends only on the likelihood and prior, never on the
intractable normalizing constant. For probit and multinomial models, **data
augmentation** restores conjugacy by sampling the latent utilities $U_{nj}$ as
additional unknowns [@chib; @mcculloch1994exactprobit]. @fig-gibbs sketches one sweep.

```{mermaid}
%%| label: fig-gibbs
%%| fig-cap: "One sweep of a hierarchical-Bayes Gibbs sampler for the choice model in equation (8). Conjugate blocks are drawn directly; the non-conjugate individual coefficients use a Metropolis-Hastings step. The loop repeats for thousands of iterations after burn-in."
flowchart LR
    S["Current draw:<br/>beta_n, beta_bar, V_beta"] --> A["Draw beta_n | rest<br/>(Metropolis-Hastings, eq. 9)"]
    A --> B["Draw beta_bar | beta_n, V_beta<br/>(Normal, conjugate)"]
    B --> C["Draw V_beta | beta_n, beta_bar<br/>(Inverse-Wishart, conjugate)"]
    C --> D{"Converged?<br/>(R-hat, trace plots)"}
    D -->|"no"| S
    D -->|"yes"| E["Retain post-burn-in draws<br/>= posterior sample"]
```

**Convergence is not automatic and must be checked.** Standard diagnostics are the
trace plot, the potential-scale-reduction factor $\hat R$ across multiple dispersed
chains (values near $1.0$ indicate convergence), and the effective sample size, which
discounts the nominal draw count by autocorrelation [@chintagunta2006structural].

#### Worked example: hierarchical Bayes for brand choice

The following self-contained example simulates a choice-based experiment with
heterogeneous consumers, fits a homogeneous logit and a hierarchical-Bayes mixed
logit, and shows that the hierarchical model recovers the individual-level taste
distribution the homogeneous model cannot see.

```{r choice-setup, message=FALSE, warning=FALSE}
set.seed(50)

# ---- Simulate a choice-based conjoint / brand-choice experiment ----
n_consumers <- 200      # decision makers
n_tasks     <- 12       # choice tasks per consumer
n_alts      <- 3        # alternatives per task (brands A, B, C)
n_attr      <- 4        # intercepts for B and C, price, a feature dummy

# Population (hyper) parameters: mean tastes and heterogeneity covariance
beta_bar  <- c(brandB = 0.8, brandC = 0.4, price = -1.5, feature = 0.7)
V_beta    <- diag(c(0.5, 0.5, 0.4, 0.3))          # taste heterogeneity

# Draw individual-level tastes from the population distribution
library(MASS)
beta_i <- mvrnorm(n_consumers, mu = beta_bar, Sigma = V_beta)

# Build design matrices and simulate choices under the logit model
sim_one <- function(b) {
  Y <- integer(n_tasks); X_list <- vector("list", n_tasks)
  for (t in seq_len(n_tasks)) {
    price   <- runif(n_alts, 0, 1)                 # standardized price
    feature <- rbinom(n_alts, 1, 0.5)
    X <- cbind(brandB = c(0, 1, 0), brandC = c(0, 0, 1),
               price = price, feature = feature)
    v <- as.numeric(X %*% b)
    p <- exp(v) / sum(exp(v))
    Y[t] <- sample.int(n_alts, 1, prob = p)        # realized choice
    X_list[[t]] <- X
  }
  list(Y = Y, X = X_list)
}

data <- lapply(seq_len(n_consumers), function(i) sim_one(beta_i[i, ]))
cat("Simulated", n_consumers, "consumers x", n_tasks, "tasks.\n")
```

A pooled logit that ignores heterogeneity recovers the population *mean* tastes
reasonably but, by construction, says nothing about their spread.

```{r pooled-logit, message=FALSE, warning=FALSE}
# Negative log-likelihood for a homogeneous conditional logit
nll_logit <- function(b, data) {
  ll <- 0
  for (d in data) for (t in seq_along(d$Y)) {
    v <- as.numeric(d$X[[t]] %*% b)
    ll <- ll + (v[d$Y[t]] - log(sum(exp(v))))
  }
  -ll
}

fit_pooled <- optim(rep(0, n_attr), nll_logit, data = data,
                    method = "BFGS")
pooled <- setNames(round(fit_pooled$par, 3), names(beta_bar))
rbind(true_mean = beta_bar, pooled_logit = pooled)
```

Now the hierarchical-Bayes mixed logit. The sampler implements exactly the sweep of
@fig-gibbs: a random-walk Metropolis step for each consumer's
$\boldsymbol{\beta}_n$ (the non-conjugate block, @eq-mh), a conjugate normal draw for
the population mean $\bar{\boldsymbol{\beta}}$, and a conjugate inverse-Wishart draw
for the heterogeneity covariance $\mathbf{V}_\beta$.

```{r hb-mixed-logit, message=FALSE, warning=FALSE}
# Log individual likelihood + log prior, up to a constant
log_post_i <- function(b, d, mu, Sig_inv) {
  ll <- 0
  for (t in seq_along(d$Y)) {
    v <- as.numeric(d$X[[t]] %*% b)
    ll <- ll + (v[d$Y[t]] - log(sum(exp(v))))
  }
  ll - 0.5 * as.numeric(t(b - mu) %*% Sig_inv %*% (b - mu))
}

R <- 4000; burn <- 2000; step <- 0.25          # MCMC controls
p <- n_attr
beta_draws <- array(0, dim = c(n_consumers, p),
                    dimnames = list(NULL, names(beta_bar)))   # current individual betas
mu_draw    <- rep(0, p)                            # current population mean
Sig_draw   <- diag(p)                              # current covariance
nu0 <- p + 3; V0 <- diag(p) * nu0                  # inverse-Wishart hyperprior
keep_mu <- matrix(0, R - burn, p)
keep_V  <- array(0, dim = c(R - burn, p, p))
acc <- 0

riwish <- function(nu, S) {                        # inverse-Wishart draw
  L <- chol(solve(S)); k <- nrow(S)
  Z <- matrix(rnorm(nu * k), nu, k) %*% t(L)
  solve(crossprod(Z))
}

for (r in seq_len(R)) {
  Sig_inv <- solve(Sig_draw)
  # (1) Metropolis-Hastings update of each consumer's beta
  for (i in seq_len(n_consumers)) {
    cur <- beta_draws[i, ]
    prop <- cur + rnorm(p, 0, step)
    la <- log_post_i(prop, data[[i]], mu_draw, Sig_inv) -
          log_post_i(cur,  data[[i]], mu_draw, Sig_inv)
    if (log(runif(1)) < la) { beta_draws[i, ] <- prop; acc <- acc + 1 }
  }
  # (2) Conjugate normal draw for population mean
  bbar <- colMeans(beta_draws)
  Vn   <- solve(n_consumers * Sig_inv + diag(p) * 1e-4)
  mu_draw <- as.numeric(Vn %*% (n_consumers * Sig_inv %*% bbar) +
                        t(chol(Vn)) %*% rnorm(p))
  # (3) Conjugate inverse-Wishart draw for covariance
  S <- V0 + crossprod(sweep(beta_draws, 2, mu_draw))
  Sig_draw <- riwish(nu0 + n_consumers, S)
  if (r > burn) { keep_mu[r - burn, ] <- mu_draw; keep_V[r - burn, , ] <- Sig_draw }
}

cat("MH acceptance rate:", round(acc / (R * n_consumers), 2), "\n")
hb_mean <- setNames(round(colMeans(keep_mu), 3), names(beta_bar))
rbind(true_mean = beta_bar, hb_posterior_mean = hb_mean)
```

The hierarchical posterior recovers not only the population means but the
*heterogeneity* the pooled model discards—the diagonal of $\mathbf{V}_\beta$, which
quantifies how widely price sensitivity and brand preference vary.

```{r hb-heterogeneity, message=FALSE, warning=FALSE}
post_V_diag <- apply(keep_V, c(2, 3), mean)
recovered <- setNames(round(diag(post_V_diag), 3), names(beta_bar))
rbind(true_variance = diag(V_beta), hb_recovered = recovered)
```

```{r hb-figure, message=FALSE, warning=FALSE, fig.cap="Posterior distribution of the population mean price coefficient from the hierarchical-Bayes sampler (post-burn-in draws). The dashed line marks the true data-generating value; the spread is genuine posterior uncertainty, not noise."}
hist(keep_mu[, 3], breaks = 30, col = "grey80", border = "white",
     main = "Posterior: mean price coefficient",
     xlab = expression(bar(beta)[price]))
abline(v = beta_bar["price"], lwd = 2, lty = 2)
abline(v = mean(keep_mu[, 3]), lwd = 2)
legend("topright", c("true value", "posterior mean"),
       lty = c(2, 1), lwd = 2, bty = "n")
```

With individual posteriors in hand the analyst can compute a *distribution* of
willingness to pay for the feature—the ratio of the feature coefficient to the
(negated) price coefficient for each consumer—rather than a single number, and can
target the consumers whose posteriors place them above a profitability threshold.

```{r wtp, message=FALSE, warning=FALSE}
# Willingness to pay for the feature, per consumer (posterior means of beta_i)
wtp <- beta_draws[, "feature"] / (-beta_draws[, "price"])
wtp <- wtp[is.finite(wtp) & wtp > 0]
summary(round(wtp, 2))
```

**Key readings.**

- @allenby1998heterogeneity — the manifesto for hierarchical-Bayes heterogeneity in
  marketing; argues for individual-level inference over representative-agent models.
  [F]
- @rossi1996purchasehistory — the template HB application: individual-level posteriors
  (shrinkage toward the population) raise the value of household data for targeting.
  [F]
- @allenby1995reassessing — an early, influential hierarchical choice model showing
  how heterogeneity reshapes substantive conclusions about loyalty and price
  sensitivity. [F]
- @rossi2001scaleusage — shows how hierarchical Bayes cleanly removes a pervasive
  nuisance (respondents using rating scales differently); a model template beyond
  choice. [F]

**Debate.** Continuous (HB) vs. discrete (latent-class/finite-mixture) heterogeneity;
how much structure to put in the upper-level prior; whether individual posteriors are
"real" or regularization artifacts.

### Week 8 — Conjoint & preference measurement {#sec-cb-week-8}

**Topic.** Choice-based conjoint and preference measurement as the marketing-native
application of HB choice models.

**Subtopics.** Partworth estimation via HB; reduced designs and borrowing strength;
constrained/sign-restricted priors; heterogeneity-distribution choice; adaptive and
efficient design.

**Methods.** HB conjoint estimation; experimental design for choice; holdout
prediction.

The hierarchical estimator of @sec-cb-week-7 is the dominant tool for recovering
individual partworths from conjoint designs in which each respondent answers far
fewer profiles than there are parameters. The normal-mixing assumption is a modeling
choice, not a fact: it cannot represent multimodal tastes (genuine segments), and a
normally distributed price coefficient places positive probability on consumers who
*prefer* higher prices. Remedies include log-normal or sign-constrained coefficients,
finite mixtures of normals for true segments, and—where the analyst doubts any
parametric form—nonparametric mixing. Identification is delicate, so model comparison
and holdout validation (@sec-cb-week-11) are essential rather than ornamental
[@chintagunta2006structural].

**Key readings.**

- @lenk1996hbconjoint — foundational HB conjoint: recovers heterogeneous partworths
  even when each respondent answers far fewer profiles than parameters. [F]
- @allenby1995priorknowledge — shows how economically motivated (e.g.,
  sign-constrained) priors improve partworth plausibility and prediction. [F]
- @andrews2002hbvsmixture — the standard HB-vs-finite-mixture comparison; ties Week
  7's heterogeneity debate to a concrete conjoint application. [F]
- @train2005boundedmixedlogit (see @sec-cb-week-5) — bounded/correlated partworth
  distributions for conjoint. [F]

**Debate.** Continuous vs. discrete heterogeneity in conjoint; constrained vs.
unconstrained partworths; whether better designs or better priors matter more for
small per-respondent samples.

### Week 9 — Aggregate-data demand & integrating out heterogeneity {#sec-cb-week-9}

**Topic.** Random-coefficients logit demand from aggregate market shares (BLP); the
Bayesian counterpart; the mechanics of integrating out heterogeneity and endogeneity.

**Subtopics.** The BLP contraction and GMM; the inversion from shares to mean
utilities; instruments and price endogeneity; the Bayesian (Jiang–Manchanda–Rossi)
approach; numerical reliability (MPEC).

**Methods.** GMM with the BLP contraction; MSM; Bayesian estimation from aggregate
data; MPEC reformulation.

When individual choices are unobserved and only market shares are recorded, the
random-coefficients demand system inverts shares to recover mean utilities and
estimates the mixing distribution from share variation across markets
[@berry1995; @berry1995automobile]—the bridge between the consumer-level models of
this chapter and industry-level demand analysis (@sec-marketing-finance), with active
work on practical edge cases such as zero-share products [@dube2021random]. The
Bayesian counterpart integrates out heterogeneity from aggregate shares via MCMC.

**Key readings.**

- @berry1994discretechoice — introduces the share-inversion that lets discrete-choice
  demand be estimated from aggregate data with endogenous prices. [F]
- @berry1995automobile — the canonical BLP random-coefficients aggregate-demand model;
  the reference point for all market-share choice estimation. [F]
- @jiang2009bayesianblp — the Bayesian counterpart to BLP; shows how MCMC integrates
  out heterogeneity from aggregate shares. [F → R]
- @chintagunta2011discretechoice — the marketing-specific synthesis of the whole
  discrete-choice demand toolkit; the natural anchor/overview citation. [F]
- @petrin2002minivan — shows how BLP-style demand plus micro-moments yields
  welfare/counterfactual answers; a model application. [F]

**Debate.** GMM/BLP vs. Bayesian estimation of the same model; instrument validity
and weak instruments; numerical fragility of the nested contraction (and MPEC as the
fix, Dubé–Fox–Su, [10.3982/ecta8585](https://doi.org/10.3982/ecta8585)).

### Week 10 — Dynamics & state dependence {#sec-cb-week-10}

**Topic.** Dynamic choice: Bayesian learning, structural state dependence, and the
central identification problem of distinguishing genuine persistence from unobserved
heterogeneity.

**Subtopics.** Forward-looking vs. myopic consumers; Bayesian learning about brand
attributes; habit/inertia/loyalty; the heterogeneity-vs-state-dependence confound;
the initial-conditions problem.

**Methods.** Dynamic structural estimation; hierarchical Bayes with
autocorrelated/lagged terms; careful identification design.

**Key readings.**

- @erdem1996learning — the founding consumer-learning choice model: forward-looking
  Bayesian consumers learning brand quality from experience and advertising. [F → R]
- @keane1997statedependence — the methodological statement of the
  heterogeneity-vs-state-dependence identification problem; required reading for the
  debate. [F]
- Dubé, Hitsch & Rossi (2010), "State Dependence and Alternative Explanations for
  Consumer Inertia," *RAND J. Econ.* 41(3):417–445,
  [10.1111/j.1756-2171.2010.00106.x](https://doi.org/10.1111/j.1756-2171.2010.00106.x)
  — the modern resolution isolating genuine state dependence from persistent
  heterogeneity. [R]

**Debate.** Structural state dependence vs. spurious persistence from heterogeneity;
forward-looking vs. reduced-form dynamics; how much the initial-conditions assumption
drives results.

### Week 11 — Model comparison: Bayes factors, DIC, WAIC, cross-validation {#sec-cb-week-11}

**Topic.** Comparing and checking Bayesian choice models.

**Subtopics.** Marginal likelihood and Bayes factors; sensitivity to priors
(Lindley/Bartlett paradox); DIC and the effective number of parameters; WAIC;
leave-one-out cross-validation (PSIS-LOO); posterior predictive checks.

**Methods.** Computing marginal likelihoods (Chib's method); DIC/WAIC; LOO; posterior
predictive checking.

**Key readings.**

- @kass1995bayesfactors — the definitive treatment of Bayes factors for model
  comparison; sets up the marginal-likelihood approach and its prior sensitivity. [F]
- @spiegelhalter2002dic — introduces DIC and the effective number of parameters; the
  most-used hierarchical-model comparison criterion in applied marketing. [F]
- @gelman2014predictiveic — unifies AIC/DIC/WAIC and predictive evaluation; the modern
  conceptual map for choosing a criterion. [F → R]
- @vehtari2017loo — the current practical standard (PSIS-LOO + WAIC) with the `loo`
  package; what students should actually run today. [R]

**Debate.** Bayes factors (and prior sensitivity) vs. predictive criteria
(WAIC/LOO); DIC's known pathologies; in-sample fit vs. holdout prediction as the
right target for marketing models.

### Week 12 — Hamiltonian Monte Carlo & Stan {#sec-cb-week-12}

**Topic.** Gradient-based MCMC and probabilistic programming as the modern default
engine.

**Subtopics.** Why random-walk MH/Gibbs mix poorly in high dimensions; Hamiltonian
dynamics; the No-U-Turn Sampler (NUTS); non-centered parameterization for
hierarchical models; the Stan ecosystem.

**Methods.** Coding a hierarchical choice model in Stan; HMC/NUTS diagnostics
(divergences, energy, R-hat, effective sample size); reparameterization.

**Key readings.**

- @carpenter2017stan — the reference for Stan; the tool most students will use to fit
  hierarchical choice models going forward. [R]
- Hoffman & Gelman (2014), "The No-U-Turn Sampler," *Journal of Machine Learning
  Research* 15:1593–1623 — JMLR, no Crossref DOI; cite by volume/pages. The NUTS
  algorithm underlying Stan's sampler. [R]
- Gelman et al. (2013), *Bayesian Data Analysis*, HMC chapter (book) — textbook
  treatment. [F]

**Debate.** HMC/NUTS vs. problem-specific Gibbs samplers (speed vs. generality); when
divergences signal a real modeling problem vs. a tuning problem; the
centered/non-centered parameterization trade-off in hierarchical models.

### Week 13 — Scalable & variational methods {#sec-cb-week-13}

**Topic.** Approximate Bayesian inference for large, high-dimensional, sparse
marketing data.

**Subtopics.** Mean-field and structured variational inference; the ELBO; automatic
differentiation variational inference (ADVI); stochastic/streaming variants;
accuracy-vs-speed trade-offs; where VI fails for choice models.

**Methods.** Variational inference; ADVI in Stan/PyMC; diagnostics for approximate
posteriors.

**Key readings.**

- @blei2017variational — the standard statistician-facing introduction to variational
  inference; the conceptual anchor for the scalable-methods week. [R]
- @ansari2003ecustomization — a marketing exemplar of high-dimensional Bayesian
  modeling for personalization at scale; motivates why scalable inference matters. [F]
- @yang2003interdependent — a Bayesian spatial-autoregressive choice model; shows how
  richer dependence structures (and the computation they demand) extend the
  hierarchical-choice toolkit. [F]
- @carpenter2017stan (see @sec-cb-week-12) — Stan also implements ADVI, the bridge from
  HMC to VI. [R]

**Debate.** Variational approximation bias vs. MCMC accuracy; whether VI's posterior
uncertainties can be trusted for inference (vs. only point prediction); the
encroachment of ML/deep methods on Bayesian choice modeling.

### Week 14 — Synthesis, frontier & research design {#sec-cb-week-14}

**Topic.** Pulling the threads together; designing a choice-modeling research paper;
the open frontier.

**Subtopics.** Matching estimator to question (closed-form vs. simulation vs. MCMC vs.
VI); identification storytelling; replication and reproducibility (`bayesm`, Stan,
`loo`); current frontiers—ML/Bayesian hybrids, text/image as choice covariates,
privacy-constrained estimation, scalable personalization.

**Methods.** Referee report; replication-or-extension project presentation.

The recurring lesson across the semester is that the hard part of choice modeling is
rarely the optimizer or the sampler; it is matching the model's substitution
structure and heterogeneity to the market and the decision at hand. Revealed-
preference data suffer endogeneity (instruments and control functions, with their own
pitfalls); stated-preference conjoint avoids price endogeneity by design at the cost
of hypothetical bias; aggregate data require the BLP machinery of @sec-cb-week-9. A
choice model's in-sample fit is a weak guide—holdout prediction, recovery of known
substitution patterns, and evidence that an IIA relaxation changes substantive
conclusions are the operative tests [@chintagunta2006structural; @dube2021random].

**Key readings.** Student-selected frontier papers plus the field syntheses already
assigned: @allenby1998heterogeneity, @chintagunta2011discretechoice, and the
Bayesian-marketing review Rossi & Allenby (2003), *Marketing Science* 22(3):304–328,
[10.1287/mksc.22.3.304.17739](https://doi.org/10.1287/mksc.22.3.304.17739). [R]

**Debate.** The whole-course tension: structural/behavioral interpretability vs.
predictive flexibility; Bayesian vs. classical estimation of identical models; how
much marketing's choice-modeling tradition will be absorbed into machine learning.

## Foundational vs. frontier at a glance {#sec-cb-foundational-frontier}

| Week | Topic | Status |
|---|---|---|
| 1 | RUM theory | Foundational |
| 2 | Logit | Foundational |
| 3 | IIA / GEV / nested | Foundational |
| 4 | Probit & GHK simulation | Foundational |
| 5 | Mixed / random-coefficients logit | Foundational (bridges to frontier) |
| 6 | Bayesian foundations & MCMC | Foundational |
| 7 | Hierarchical Bayes for heterogeneity | Foundational (course core) |
| 8 | Conjoint / preference measurement | Foundational application; design is frontier |
| 9 | Aggregate demand / integrating out heterogeneity | Foundational → frontier |
| 10 | Dynamics & state dependence | Foundational → frontier |
| 11 | Model comparison (Bayes factors / DIC / WAIC / LOO) | Foundational (LOO/WAIC are the modern frontier) |
| 12 | HMC / Stan | Frontier, now standard |
| 13 | Scalable & variational methods | Frontier |
| 14 | Synthesis & research frontier | Frontier / craft |

Weeks 1–11 are the durable canon a chapter must cover; Weeks 12–14 are where the
chapter dates fastest and needs the most active maintenance.

## How this chapter expands {#sec-cb-how-expands}

1. **Keep the spine fixed, refresh the engine.** Weeks 1–11 (RUM → logit/GEV/probit →
   mixed logit → Bayesian/MCMC → HB → conjoint → aggregate demand → dynamics → model
   comparison) are stable for a generation. The *computation* weeks (12–13) turn over
   fastest; budget for revising them every edition as Stan, PyMC, JAX-based samplers,
   and normalizing-flow/amortized VI mature.
2. **Promote model comparison from afterthought to spine.** WAIC/PSIS-LOO
   (@sec-cb-week-11) have already displaced DIC/Bayes factors in practice; future editions
   should foreground predictive evaluation and posterior predictive checking.
3. **Add a dedicated "ML × Bayesian choice" frontier section.** As deep generative
   models, embeddings, and amortized inference enter choice modeling, @sec-cb-week-13 will
   likely split into (a) classical scalable Bayes (VI) and (b) neural/ML-hybrid choice
   models—pending verified, DOI-bearing papers.
4. **Grow the unstructured-data covariate strand.** Text, image, audio, and video as
   inputs to choice increasingly fuse with this seminar; a future section can connect
   representation learning to RUM covariates.
5. **Track the privacy/identification frontier.** Privacy-constrained and
   aggregated-data estimation pushes the field back toward the aggregate-demand
   methods of @sec-cb-week-9 and differential-privacy-aware Bayes—a natural growth area.
6. **Maintain the cross-references.** This chapter shares readings with preference
   measurement (conjoint) and structural models (BLP, dynamics, MPEC). Keep shared
   entries synchronized rather than duplicated, and re-verify DOIs each edition.

## Key Takeaways

- Discrete choice rests on **random-utility theory** (@eq-50-rum): consumers maximize
  latent utility, only utility *differences* are identified, and the assumed
  distribution of the unobserved term determines the model.
- The **conditional logit** (@eq-50-logit) is the closed-form workhorse but imposes
  **IIA** (@eq-50-iia)—equal proportional cross-elasticities—which is behaviorally
  false whenever alternatives share unobserved attributes (the red-bus/blue-bus
  failure).
- **Nested logit, probit, and mixed logit** relax IIA by correlating the unobserved
  terms; mixed logit (@eq-mixedlogit) is the most flexible and links substitution
  structure to **consumer heterogeneity** through a mixing distribution.
- **Hierarchical Bayes** (@eq-hierarchy) reads the mixing distribution as a prior and
  borrows strength across consumers, recovering usable **individual-level** tastes
  from sparse data—the foundation of conjoint partworth estimation and household
  targeting.
- **MCMC**—Gibbs for the conjugate population parameters, Metropolis–Hastings
  (@eq-mh) and data augmentation for the non-conjugate choice likelihood—estimates
  the hierarchy, but its draws are credible only after **convergence diagnostics**.
- In practice the binding constraints are **data structure** (endogeneity in
  revealed preference, hypothetical bias in stated preference), the **decision the
  model serves**, and **out-of-sample validation**—not the estimation algorithm.
