# Discrete Choice and Bayesian Methods {#sec-choice-bayesian}

Most marketing decisions reduce to a question about choice: which brand a shopper
puts in the cart, which subscription tier a household keeps, which car a buyer
drives off the lot. The data that record these decisions are not the continuous
quantities of classical regression but *discrete selections* from a menu—one
alternative chosen, the rest forgone. A demand model that respects this structure
must explain why one option was picked over its rivals, predict how that
probability shifts when a price moves or a feature is added, and do so from data in
which each person reveals only the tip of their preference. Discrete-choice
modeling, grounded in random-utility theory, is the apparatus marketing uses to
turn such observations into estimates of preference, willingness to pay, and
demand elasticities.

This chapter builds that apparatus from first principles and then confronts the two
problems that dominate its practice. The first is *substitution structure*: the
workhorse logit model embeds an assumption—independence of irrelevant alternatives
(IIA)—that is empirically false in most markets and produces nonsensical demand
predictions when violated. The second is *heterogeneity*: consumers differ, and a
model that estimates a single "representative" taste vector both misstates
aggregate demand and is useless for targeting. The modern resolution of both
problems runs through the same machinery—mixing distributions over individual
tastes—and the dominant estimation technology for that machinery is Bayesian:
hierarchical models fit by Markov chain Monte Carlo (MCMC). By the end of the
chapter the reader can specify a random-utility model, recognize when IIA breaks
it, choose among nested, probit, and mixed-logit remedies, and estimate
individual-level preferences by hierarchical Bayes with runnable code.

The intellectual lineage is worth stating up front. The random-utility foundation
and the conditional logit estimator are due to McFadden, whose program connected
econometric choice models to consumer theory and earned the field its modeling
language [@mcfadden1986choice; @mcfadden2001economic]. The transplant of those
methods into marketing began with the logit model of brand choice calibrated on
supermarket scanner data [@guadagni1983; @guadagni2008logit], which remains the
template for how the discipline links observed purchases to a utility
specification. The Bayesian turn—treating heterogeneity as a distribution to be
inferred rather than a nuisance to be differenced away—reorganized the field around
hierarchical models and simulation-based estimation [@rossi2014]. We connect the
demand-estimation thread to its industrial-organization counterpart
(@sec-marketing-finance) and to preference measurement via conjoint analysis, whose
hierarchical-Bayes practice is the applied face of everything developed here.

## Random-Utility Theory

The premise of random-utility theory is that a decision maker assigns a latent
utility to each available alternative and chooses the one with the highest utility.
The analyst does not observe utility; the analyst observes only the choice and some
attributes of the alternatives and the chooser. The randomness in "random utility"
is the analyst's, not the consumer's: it represents everything that bears on the
decision but is unobserved by the modeler.

::: {.callout-note}
## Landmark definition: random utility

> The utility that decision maker $n$ obtains from alternative $j$ is
> $U_{nj} = V_{nj} + \varepsilon_{nj}$, where $V_{nj}$ is a *representative* (or
> *systematic*) component, a function of observed attributes and parameters, and
> $\varepsilon_{nj}$ is a random term capturing unobserved determinants of utility.
> The decision maker chooses alternative $i$ if and only if $U_{ni} \ge U_{nj}$ for
> all $j$ in the choice set. The choice probability is therefore a statement about
> the distribution of the differences in the unobserved terms
> [@mcfadden2001economic].
:::

Formally, let $n = 1,\dots,N$ index decision makers and let $\mathcal{C}_n$ be the
finite choice set facing decision maker $n$. The utility of alternative
$j \in \mathcal{C}_n$ is

$$
U_{nj} = V_{nj} + \varepsilon_{nj}, \qquad
V_{nj} = \mathbf{x}_{nj}^{\top}\boldsymbol{\beta},
$$ {#eq-rum}

where $\mathbf{x}_{nj}$ collects observed attributes of the alternative (price,
features, brand indicators) and possibly interactions with characteristics of the
chooser, and $\boldsymbol{\beta}$ is a vector of taste parameters. The probability
that $n$ chooses $i$ is the probability that $i$ yields the highest utility,

$$
P_{ni} = \Pr\!\left( U_{ni} \ge U_{nj}\ \ \forall j \in \mathcal{C}_n \right)
       = \Pr\!\left( \varepsilon_{nj} - \varepsilon_{ni} \le V_{ni} - V_{nj}\ \
         \forall j \neq i \right).
$$ {#eq-choiceprob}

Three structural facts follow immediately and govern everything downstream. First,
**only differences in utility matter**: adding a constant to every alternative's
utility leaves the choice unchanged, so alternative-invariant variables (a chooser's
income entering identically for all options) drop out and must be interacted with
alternative-specific terms to have any effect. Second, the **scale of utility is not
identified**: multiplying all utilities by a positive constant leaves the argmax
unchanged, so the variance of $\varepsilon_{nj}$ must be normalized. Third, the form
of the choice probability is determined entirely by the **assumed joint distribution
of the unobserved terms** $\boldsymbol{\varepsilon}_n = (\varepsilon_{n1},\dots)$.
Different assumptions about that distribution generate the logit, probit, nested,
and mixed models that organize the rest of this section.

## The Conditional Logit Model

The conditional (or multinomial) logit model arises from a single distributional
assumption: the unobserved utility terms are independent across alternatives and
each follows a type-I extreme-value (Gumbel) distribution,
$F(\varepsilon_{nj}) = \exp(-e^{-\varepsilon_{nj}})$. This assumption is convenient
because the difference of two independent Gumbel variates is logistic, which
collapses the multidimensional integral in @eq-choiceprob to a closed form. The
choice probability becomes

$$
P_{ni} = \frac{\exp(\mathbf{x}_{ni}^{\top}\boldsymbol{\beta})}
              {\sum_{j \in \mathcal{C}_n} \exp(\mathbf{x}_{nj}^{\top}\boldsymbol{\beta})}.
$$ {#eq-logit}

This is the model that brought random-utility estimation into marketing. Calibrated
on household scanner data, with $\mathbf{x}_{nj}$ containing price, promotion
indicators, and—critically—a *loyalty* variable summarizing the household's past
purchases, it predicts brand choice well enough to anchor demand analysis for
packaged goods [@guadagni1983; @guadagni2008logit]. The loyalty term is an early,
informal device for capturing persistence in tastes that the homogeneous logit
cannot otherwise represent, foreshadowing the heterogeneity machinery developed
below.

**Estimation.** With independent observations the logit log-likelihood is globally
concave in $\boldsymbol{\beta}$, so maximum likelihood is straightforward and
unique. Writing $y_{ni} = 1$ if $n$ chose $i$ and $0$ otherwise,

$$
\ell(\boldsymbol{\beta}) = \sum_{n=1}^{N} \sum_{i \in \mathcal{C}_n}
   y_{ni} \log P_{ni}(\boldsymbol{\beta}),
$$ {#eq-loglik}

which Newton–Raphson or any quasi-Newton routine maximizes reliably. The scale
normalization is implicit: the Gumbel variance is fixed at $\pi^2/6$, so
$\boldsymbol{\beta}$ is identified only up to that scale, and coefficients across
models with different error variances are not directly comparable—a subtlety that
resurfaces when comparing choice data sources.

**Elasticities.** The managerial output of a choice model is usually an elasticity.
For a price coefficient $\beta_p$ entering $V_{nj}$ linearly, the own- and
cross-price elasticities of choice probability are

$$
E_{ii} = \beta_p\, p_{ni}\,(1 - P_{ni}), \qquad
E_{ij} = -\,\beta_p\, p_{nj}\, P_{nj}.
$$ {#eq-elast}

The cross-elasticity $E_{ij}$ is the source of the model's most consequential
defect: it depends on $j$ only through $P_{nj}$, so *every* alternative has the same
proportional cross-elasticity with respect to a change in $j$. This is IIA, examined
next.

## The Independence of Irrelevant Alternatives

The logit model's defining property is that the ratio of any two choice
probabilities is independent of the attributes—indeed of the existence—of every
other alternative:

$$
\frac{P_{ni}}{P_{nj}}
   = \frac{\exp(\mathbf{x}_{ni}^{\top}\boldsymbol{\beta})}
          {\exp(\mathbf{x}_{nj}^{\top}\boldsymbol{\beta})},
$$ {#eq-iia}

which contains nothing about any third option $k$. This **independence of
irrelevant alternatives** (IIA) is a direct consequence of the assumed
independence of the Gumbel error terms across alternatives. It is mathematically
elegant and behaviorally implausible.

The standard illustration is the *red-bus/blue-bus* paradox. A commuter chooses
between a car and a red bus with equal probability, $\tfrac{1}{2}$ each, so
$P_{\text{car}}/P_{\text{red}} = 1$. Introduce a blue bus identical to the red bus
in every respect but color. Behaviorally the two buses should split the original bus
share, leaving car at $\tfrac{1}{2}$ and each bus at $\tfrac{1}{4}$. The logit
model instead preserves all pairwise ratios, forcing
$P_{\text{car}} = P_{\text{red}} = P_{\text{blue}} = \tfrac{1}{3}$: adding a near-duplicate
*steals* share from the car as readily as from the bus it nearly clones. In a
marketing setting the same failure means a logit demand model predicts that a new
product cannibalizes every incumbent in proportion to its share, regardless of how
close a substitute it actually is—a prediction that misvalues line extensions, new
entrants, and assortment changes.

::: {.callout-warning}
## What IIA gets wrong, and when it is safe

IIA is the right model only when alternatives are, conditional on the observed
$\mathbf{x}_{nj}$, equally substitutable for one another. It fails whenever subsets
of alternatives share unobserved attributes—body styles of cars, flavor families of
snacks, branded versus private-label tiers. The practical diagnostic is a
Hausman–McFadden test: estimate the model on the full choice set and on a subset; if
IIA holds, dropping alternatives should not change $\hat{\boldsymbol{\beta}}$, and a
significant change rejects IIA. When the test rejects, the logit's substitution
patterns are not merely imprecise—they are structurally wrong, and the remedies of
the next sections are required rather than optional.
:::

The IIA property is also the reason the logit can be estimated on a *sampled* subset
of alternatives when the full choice set is enormous (thousands of SKUs): because
ratios of probabilities ignore the rest of the set, a correctly weighted sample of
non-chosen alternatives yields consistent estimates. This is a genuine computational
benefit—the same property that makes the model behaviorally restrictive makes it
tractable at scale.

## Relaxing IIA: Probit, Nested, and Mixed Logit

Every remedy for IIA works by abandoning the assumption of independent, identically
distributed Gumbel errors. They differ in *how* they let the unobserved terms
correlate across alternatives, and that difference determines the substitution
patterns the model can represent and the cost of estimating it. @fig-choice-family
maps the family.

```{mermaid}
%%| label: fig-choice-family
%%| fig-cap: "The discrete-choice model family, organized by the assumed structure of the unobserved utility terms. Each branch relaxes a restriction of the conditional logit and pays for it in estimation cost."
flowchart TD
    A["Random-utility model<br/>U = V + e"] --> B{"Distribution of<br/>unobserved e?"}
    B -->|"iid Gumbel"| C["Conditional logit<br/>closed form, imposes IIA"]
    B -->|"correlated normal"| D["Multinomial probit<br/>full covariance, no IIA"]
    B -->|"Gumbel within nests,<br/>correlated across"| E["Nested logit<br/>IIA within nest only"]
    B -->|"Gumbel + random beta_n"| F["Mixed (random-coefficients) logit<br/>any substitution pattern"]
    F --> G["Heterogeneity as a<br/>mixing distribution"]
    G --> H["Hierarchical Bayes<br/>estimation via MCMC"]
```

### Multinomial Probit

The multinomial probit model replaces the independent Gumbel errors with a
multivariate normal vector, $\boldsymbol{\varepsilon}_n \sim
\mathcal{N}(\mathbf{0}, \boldsymbol{\Sigma})$. The off-diagonal elements of
$\boldsymbol{\Sigma}$ let any pair of alternatives share unobserved utility, so
probit imposes no IIA and can represent arbitrary substitution patterns. The price
is analytic: the choice probability in @eq-choiceprob is now a
$(J-1)$-dimensional integral of the normal density with no closed form,

$$
P_{ni} = \int_{\mathbb{R}^{J-1}} \mathbb{1}\!\left[ U_{ni} \ge U_{nj}\ \forall j \neq i
   \right] \phi(\boldsymbol{\varepsilon}_n; \boldsymbol{\Sigma})\, d\boldsymbol{\varepsilon}_n .
$$ {#eq-probit}

For more than three or four alternatives this integral must be approximated by
simulation. McFadden's **method of simulated moments** made probit estimation
feasible by replacing the intractable probabilities with smooth, unbiased
frequency-simulator analogues and matching simulated to observed choice frequencies,
preserving consistency as the number of draws grows [@mcfadden1986method]. A further
subtlety is identification: because only utility differences matter and scale is
free, not all elements of $\boldsymbol{\Sigma}$ are identified—one differences the
system and fixes one variance, leaving $J(J-1)/2 - 1$ free covariance parameters.
Overlooking this leaves the likelihood flat in unidentified directions. In marketing
practice probit's flexibility is most valuable when correlation structure is itself
of interest, but the mixed logit below typically achieves the same flexibility with
a more interpretable and more scalable parameterization, which is why probit is now
less common than it once was.

### Nested Logit

Nested logit is the parsimonious middle ground. It groups alternatives into *nests*
within which IIA still holds but across which it does not, capturing the leading case
of correlated unobservables—subsets of close substitutes—without a full covariance
matrix. The bus example resolves cleanly: place the red and blue buses in one nest
and the car in another, and a share gain by one bus is drawn disproportionately from
the other bus rather than uniformly from all alternatives.

The model follows from a generalized-extreme-value error structure. With nests
indexed by $k$ and the alternative $j$ belonging to nest $B_k$, the probability
factors into a within-nest conditional choice and a between-nest marginal choice,

$$
P_{nj} = \underbrace{\frac{\exp\!\big(V_{nj}/\lambda_k\big)}
                          {\sum_{m \in B_k}\exp\!\big(V_{nm}/\lambda_k\big)}}_{\text{within nest } k}
         \;\times\;
         \underbrace{\frac{\exp\!\big(\lambda_k\, I_{nk}\big)}
                          {\sum_{\ell}\exp\!\big(\lambda_\ell\, I_{n\ell}\big)}}_{\text{across nests}},
\qquad
I_{nk} = \log\!\sum_{m \in B_k}\exp\!\big(V_{nm}/\lambda_k\big),
$$ {#eq-nested}

where $I_{nk}$ is the *inclusive value* (the expected utility of nest $k$) and
$\lambda_k \in (0,1]$ is the **log-sum coefficient** governing within-nest
correlation. The boundary $\lambda_k = 1$ recovers the plain logit (no extra
within-nest correlation); $\lambda_k \to 0$ implies near-perfect substitution within
the nest. A value *outside* $(0,1]$ is inconsistent with random-utility
maximization, so an estimate of $\hat\lambda_k > 1$ is a warning that the nesting
structure is misspecified rather than a substantive finding. Nested logit's
limitations are that the analyst must impose the nesting structure a priori—the data
do not choose it—and that an alternative can sit in only one nest, so it cannot
represent cross-cutting similarities (a product that is both a "premium" item and a
"family-size" item).

### Mixed (Random-Coefficients) Logit

Mixed logit is the model that subsumes the others. Its idea is to let the taste
parameters themselves vary across decision makers according to a *mixing
distribution*, $\boldsymbol{\beta}_n \sim g(\boldsymbol{\beta} \mid \boldsymbol{\theta})$,
while retaining an iid Gumbel idiosyncratic term. Conditional on a draw of
$\boldsymbol{\beta}_n$ the model is a plain logit, but the unconditional choice
probability integrates over the distribution of tastes,

$$
P_{ni} = \int
   \frac{\exp(\mathbf{x}_{ni}^{\top}\boldsymbol{\beta})}
        {\sum_{j \in \mathcal{C}_n}\exp(\mathbf{x}_{nj}^{\top}\boldsymbol{\beta})}\,
   g(\boldsymbol{\beta} \mid \boldsymbol{\theta})\, d\boldsymbol{\beta}.
$$ {#eq-mixedlogit}

This single device accomplishes three things at once. It **breaks IIA**: because the
mixing correlates the random part of utility across alternatives that share
attributes, cross-elasticities are no longer constrained to be equal. It
**accommodates heterogeneity**: $\boldsymbol{\theta}$ describes how tastes are
distributed in the population. And it is **fully flexible**: any random-utility
model—including probit—can be approximated arbitrarily well by a mixed logit with a
suitable mixing distribution. The cost, as with probit, is that the integral in
@eq-mixedlogit has no closed form and must be simulated; classical estimation uses
simulated maximum likelihood with draws of $\boldsymbol{\beta}_n$, while the
Bayesian approach developed below targets the same integral through MCMC and is, for
the standard normal-mixing case, often the more reliable route
[@rossi2014; @allenby1999dynamic].

@tbl-iia-remedies summarizes the trade-offs.

| Model | Error structure | IIA? | Closed form? | Main cost |
|---|---|---|---|---|
| Conditional logit | iid Gumbel | Imposed | Yes | Wrong substitution under correlation |
| Nested logit | Gumbel within nests | Within nest only | Yes | Nesting must be assumed a priori |
| Multinomial probit | Multivariate normal | No | No (simulate) | Many covariance params; identification |
| Mixed logit | Gumbel + random $\boldsymbol{\beta}_n$ | No | No (simulate) | Choice of mixing distribution |

: Remedies for IIA and their trade-offs. Mixed logit dominates on flexibility but inherits the simulation cost shared by every non-logit model. {#tbl-iia-remedies}

## Heterogeneity and the Hierarchical Bayesian Approach

The mixing distribution $g(\boldsymbol{\beta}\mid\boldsymbol{\theta})$ in
@eq-mixedlogit is where heterogeneity lives, and treating it carefully is what
separates a model that merely fits aggregate shares from one that supports
targeting, segmentation, and individual-level prediction. Two facts make
heterogeneity unavoidable. First, *ignoring* it biases even aggregate quantities: a
single representative-agent logit systematically misstates demand elasticities
because the curvature of the share function differs between a homogeneous population
and a mixed one. Second, marketing's most valuable outputs—who to target, which
household values a feature, what each segment will pay—are statements about the
*distribution* of tastes, not its mean.

The Bayesian hierarchy is the natural language for this. The key obstacle is that
each consumer supplies only a handful of choices, far too few to estimate an
individual taste vector $\boldsymbol{\beta}_n$ on its own. The hierarchical model
solves this by *borrowing strength*: it assumes individuals are drawn from a common
population distribution and lets each person's estimate shrink toward the population
mean by an amount governed by how informative that person's own data are. A
three-level specification makes this concrete:

$$
\begin{aligned}
\text{(likelihood)}\quad & y_{ni} \mid \boldsymbol{\beta}_n
   \ \sim\ \text{Logit}\big(\mathbf{x}_{ni}^{\top}\boldsymbol{\beta}_n\big), \\[2pt]
\text{(heterogeneity / prior)}\quad & \boldsymbol{\beta}_n
   \ \sim\ \mathcal{N}\big(\bar{\boldsymbol{\beta}},\, \mathbf{V}_\beta\big), \\[2pt]
\text{(hyperprior)}\quad & \bar{\boldsymbol{\beta}}
   \ \sim\ \mathcal{N}(\boldsymbol{\mu}_0,\, \mathbf{A}^{-1}), \qquad
   \mathbf{V}_\beta \ \sim\ \text{Inverse-Wishart}(\nu_0, \mathbf{V}_0).
\end{aligned}
$$ {#eq-hierarchy}

The middle line is exactly the mixing distribution of mixed logit, now read as a
*prior* on each individual's tastes whose parameters
$(\bar{\boldsymbol{\beta}}, \mathbf{V}_\beta)$ are themselves estimated. An
individual with many observations is driven by the first line (their own data); an
individual with few is pulled toward $\bar{\boldsymbol{\beta}}$ by the second. This
shrinkage is not a regularization hack but the optimal pooling of individual and
population information under the model, and it is the reason hierarchical Bayes
recovers usable individual-level estimates from sparse data—the property that made
it the standard estimator for choice-based conjoint and household-level demand
[@rossi2014; @kamakura2005choice]. The same hierarchy underwrites dynamic and
multi-category extensions, where the within-person dependence across purchase
occasions further sharpens individual posteriors [@allenby1999dynamic; @chib].

::: {.callout-note}
## Why Bayesian, not just penalized likelihood?

The hierarchical structure could in principle be fit by maximizing a penalized
likelihood (empirical Bayes), and for the population parameters the two agree
asymptotically. The Bayesian apparatus earns its keep on the *individual-level*
posteriors: it delivers a full posterior distribution for each
$\boldsymbol{\beta}_n$—hence honest uncertainty for targeting decisions—and it
sidesteps the simulated-likelihood maximization that is numerically fragile for
high-dimensional mixing distributions. The cost is the construction and tuning of an
MCMC sampler, the subject of the next section.
:::

A caveat on the mixing distribution itself: the normal assumption in
@eq-hierarchy is a modeling choice, not a fact. It cannot represent multimodal
tastes (genuine segments), and a normally distributed price coefficient places
positive probability on consumers who *prefer* higher prices, which is economically
incoherent. Remedies include log-normal or sign-constrained coefficients, finite
mixtures of normals for true segments, and—where the analyst doubts any parametric
form—nonparametric mixing. Identification of the mixing distribution is delicate:
flexible mixtures can fit in-sample while being poorly pinned down, so model
comparison and holdout validation are essential rather than ornamental
[@chintagunta2006structural].

## Markov Chain Monte Carlo Estimation

MCMC is the engine that makes the hierarchy in @eq-hierarchy estimable. The idea is
to construct a Markov chain whose stationary distribution is exactly the posterior
$p(\{\boldsymbol{\beta}_n\}, \bar{\boldsymbol{\beta}}, \mathbf{V}_\beta \mid
\text{data})$; running the chain long enough and discarding an initial *burn-in*
yields dependent draws from that posterior, from which any quantity of interest—a
posterior mean, a credible interval, an elasticity—is computed as a sample average.
Two building blocks suffice for choice models.

**Gibbs sampling** cycles through the parameter blocks, drawing each from its *full
conditional* distribution—its distribution given the data and the current values of
all other parameters. The hierarchy is built so that two of its three conditionals
are conjugate and therefore drawn directly: given the individual $\boldsymbol{\beta}_n$,
the population mean $\bar{\boldsymbol{\beta}}$ has a normal full conditional and the
covariance $\mathbf{V}_\beta$ an inverse-Wishart full conditional. This conjugacy is
precisely why the normal–inverse-Wishart hyperprior is chosen: it turns two of the
three sampling steps into textbook draws.

**The Metropolis–Hastings step** handles the one block that is not conjugate. The
individual logit likelihood is not conjugate to the normal prior on
$\boldsymbol{\beta}_n$, so its full conditional cannot be sampled directly. Instead a
candidate $\boldsymbol{\beta}_n^{\star}$ is proposed (commonly by a random walk,
$\boldsymbol{\beta}_n^{\star} = \boldsymbol{\beta}_n + \boldsymbol{\eta}$ with
$\boldsymbol{\eta} \sim \mathcal{N}(\mathbf{0}, s\,\mathbf{V}_\beta)$) and accepted
with probability

$$
\alpha = \min\!\left\{ 1,\ \frac{p(y_n \mid \boldsymbol{\beta}_n^{\star})\,
   \phi(\boldsymbol{\beta}_n^{\star} \mid \bar{\boldsymbol{\beta}}, \mathbf{V}_\beta)}
   {p(y_n \mid \boldsymbol{\beta}_n)\,
   \phi(\boldsymbol{\beta}_n \mid \bar{\boldsymbol{\beta}}, \mathbf{V}_\beta)} \right\}.
$$ {#eq-mh}

The acceptance ratio depends only on the likelihood and prior, never on the
intractable normalizing constant, which is what makes the method work for models
with no closed-form posterior. For probit and multinomial models a complementary
trick, **data augmentation**, restores conjugacy by sampling the latent utilities
$U_{nj}$ as additional unknowns; conditional on the imputed latent utilities the
model becomes a linear-normal regression with standard conjugate draws, a device
central to the Bayesian treatment of multivariate and multi-category choice
[@chib]. @fig-gibbs sketches one sweep of the resulting sampler.

```{mermaid}
%%| label: fig-gibbs
%%| fig-cap: "One sweep of a hierarchical-Bayes Gibbs sampler for the choice model in equation (8). Conjugate blocks are drawn directly; the non-conjugate individual coefficients use a Metropolis-Hastings step. The loop repeats for thousands of iterations after burn-in."
flowchart LR
    S["Current draw:<br/>beta_n, beta_bar, V_beta"] --> A["Draw beta_n | rest<br/>(Metropolis-Hastings, eq. 9)"]
    A --> B["Draw beta_bar | beta_n, V_beta<br/>(Normal, conjugate)"]
    B --> C["Draw V_beta | beta_n, beta_bar<br/>(Inverse-Wishart, conjugate)"]
    C --> D{"Converged?<br/>(R-hat, trace plots)"}
    D -->|"no"| S
    D -->|"yes"| E["Retain post-burn-in draws<br/>= posterior sample"]
```

**Convergence is not automatic and must be checked.** Because the draws are
autocorrelated, an apparently long run can carry little information. Standard
diagnostics are the trace plot (the chain should mix rather than drift), the
potential-scale-reduction factor $\hat R$ comparing within- to between-chain
variance across multiple chains started at dispersed values (values near $1.0$
indicate convergence), and the effective sample size, which discounts the nominal
draw count by autocorrelation. Reporting a posterior without these diagnostics is
the Bayesian analogue of reporting a maximum-likelihood estimate without checking
that the optimizer converged. The structural-modeling literature in marketing treats
this diligence as a precondition for credible inference [@chintagunta2006structural].

## Worked Example: Hierarchical Bayes for Brand Choice

The following self-contained example simulates a choice-based experiment with
heterogeneous consumers, fits a homogeneous logit and a hierarchical-Bayes mixed
logit, and shows that the hierarchical model recovers the individual-level taste
distribution the homogeneous model cannot see. The data-generating process gives
each consumer a price sensitivity and brand preferences drawn from a population
distribution; the hierarchical estimator's job is to recover that distribution from
short individual choice sequences.

```{r choice-setup, message=FALSE, warning=FALSE}
set.seed(50)

# ---- Simulate a choice-based conjoint / brand-choice experiment ----
n_consumers <- 200      # decision makers
n_tasks     <- 12       # choice tasks per consumer
n_alts      <- 3        # alternatives per task (brands A, B, C)
n_attr      <- 4        # intercepts for B and C, price, a feature dummy

# Population (hyper) parameters: mean tastes and heterogeneity covariance
beta_bar  <- c(brandB = 0.8, brandC = 0.4, price = -1.5, feature = 0.7)
V_beta    <- diag(c(0.5, 0.5, 0.4, 0.3))          # taste heterogeneity

# Draw individual-level tastes from the population distribution
library(MASS)
beta_i <- mvrnorm(n_consumers, mu = beta_bar, Sigma = V_beta)

# Build design matrices and simulate choices under the logit model
sim_one <- function(b) {
  Y <- integer(n_tasks); X_list <- vector("list", n_tasks)
  for (t in seq_len(n_tasks)) {
    price   <- runif(n_alts, 0, 1)                 # standardized price
    feature <- rbinom(n_alts, 1, 0.5)
    X <- cbind(brandB = c(0, 1, 0), brandC = c(0, 0, 1),
               price = price, feature = feature)
    v <- as.numeric(X %*% b)
    p <- exp(v) / sum(exp(v))
    Y[t] <- sample.int(n_alts, 1, prob = p)        # realized choice
    X_list[[t]] <- X
  }
  list(Y = Y, X = X_list)
}

data <- lapply(seq_len(n_consumers), function(i) sim_one(beta_i[i, ]))
cat("Simulated", n_consumers, "consumers x", n_tasks, "tasks.\n")
```

A pooled logit that ignores heterogeneity recovers the population *mean* tastes
reasonably but, by construction, says nothing about their spread—and its implied
substitution patterns inherit IIA.

```{r pooled-logit, message=FALSE, warning=FALSE}
# Negative log-likelihood for a homogeneous conditional logit
nll_logit <- function(b, data) {
  ll <- 0
  for (d in data) for (t in seq_along(d$Y)) {
    v <- as.numeric(d$X[[t]] %*% b)
    ll <- ll + (v[d$Y[t]] - log(sum(exp(v))))
  }
  -ll
}

fit_pooled <- optim(rep(0, n_attr), nll_logit, data = data,
                    method = "BFGS")
pooled <- setNames(round(fit_pooled$par, 3), names(beta_bar))
rbind(true_mean = beta_bar, pooled_logit = pooled)
```

Now the hierarchical-Bayes mixed logit. The sampler implements exactly the sweep of
@fig-gibbs: a random-walk Metropolis step for each consumer's
$\boldsymbol{\beta}_n$ (the non-conjugate block, @eq-mh), a conjugate normal draw for
the population mean $\bar{\boldsymbol{\beta}}$, and a conjugate inverse-Wishart draw
for the heterogeneity covariance $\mathbf{V}_\beta$.

```{r hb-mixed-logit, message=FALSE, warning=FALSE}
# Log individual likelihood + log prior, up to a constant
log_post_i <- function(b, d, mu, Sig_inv) {
  ll <- 0
  for (t in seq_along(d$Y)) {
    v <- as.numeric(d$X[[t]] %*% b)
    ll <- ll + (v[d$Y[t]] - log(sum(exp(v))))
  }
  ll - 0.5 * as.numeric(t(b - mu) %*% Sig_inv %*% (b - mu))
}

R <- 4000; burn <- 2000; step <- 0.25          # MCMC controls
p <- n_attr
beta_draws <- array(0, dim = c(n_consumers, p))   # current individual betas
mu_draw    <- rep(0, p)                            # current population mean
Sig_draw   <- diag(p)                              # current covariance
nu0 <- p + 3; V0 <- diag(p) * nu0                  # inverse-Wishart hyperprior
keep_mu <- matrix(0, R - burn, p)
keep_V  <- array(0, dim = c(R - burn, p, p))
acc <- 0

riwish <- function(nu, S) {                        # inverse-Wishart draw
  L <- chol(solve(S)); k <- nrow(S)
  Z <- matrix(rnorm(nu * k), nu, k) %*% t(L)
  solve(crossprod(Z))
}

for (r in seq_len(R)) {
  Sig_inv <- solve(Sig_draw)
  # (1) Metropolis-Hastings update of each consumer's beta
  for (i in seq_len(n_consumers)) {
    cur <- beta_draws[i, ]
    prop <- cur + rnorm(p, 0, step)
    la <- log_post_i(prop, data[[i]], mu_draw, Sig_inv) -
          log_post_i(cur,  data[[i]], mu_draw, Sig_inv)
    if (log(runif(1)) < la) { beta_draws[i, ] <- prop; acc <- acc + 1 }
  }
  # (2) Conjugate normal draw for population mean
  bbar <- colMeans(beta_draws)
  Vn   <- solve(n_consumers * Sig_inv + diag(p) * 1e-4)
  mu_draw <- as.numeric(Vn %*% (n_consumers * Sig_inv %*% bbar) +
                        t(chol(Vn)) %*% rnorm(p))
  # (3) Conjugate inverse-Wishart draw for covariance
  S <- V0 + crossprod(sweep(beta_draws, 2, mu_draw))
  Sig_draw <- riwish(nu0 + n_consumers, S)
  if (r > burn) { keep_mu[r - burn, ] <- mu_draw; keep_V[r - burn, , ] <- Sig_draw }
}

cat("MH acceptance rate:", round(acc / (R * n_consumers), 2), "\n")
hb_mean <- setNames(round(colMeans(keep_mu), 3), names(beta_bar))
rbind(true_mean = beta_bar, hb_posterior_mean = hb_mean)
```

The hierarchical posterior recovers not only the population means but the
*heterogeneity* the pooled model discards—the diagonal of $\mathbf{V}_\beta$, which
quantifies how widely price sensitivity and brand preference vary across the
population.

```{r hb-heterogeneity, message=FALSE, warning=FALSE}
post_V_diag <- apply(keep_V, c(2, 3), mean)
recovered <- setNames(round(diag(post_V_diag), 3), names(beta_bar))
rbind(true_variance = diag(V_beta), hb_recovered = recovered)
```

```{r hb-figure, message=FALSE, warning=FALSE, fig.cap="Posterior distribution of the population mean price coefficient from the hierarchical-Bayes sampler (post-burn-in draws). The dashed line marks the true data-generating value; the spread is genuine posterior uncertainty, not noise."}
hist(keep_mu[, 3], breaks = 30, col = "grey80", border = "white",
     main = "Posterior: mean price coefficient",
     xlab = expression(bar(beta)[price]))
abline(v = beta_bar["price"], lwd = 2, lty = 2)
abline(v = mean(keep_mu[, 3]), lwd = 2)
legend("topright", c("true value", "posterior mean"),
       lty = c(2, 1), lwd = 2, bty = "n")
```

The economic payoff is that with individual posteriors in hand the analyst can
compute a *distribution* of willingness to pay for the feature—the ratio of the
feature coefficient to the (negated) price coefficient for each consumer—rather than
a single number, and can target the consumers whose posteriors place them above a
profitability threshold. This individual-level inference is what the homogeneous
logit cannot deliver and what makes the hierarchical Bayes machinery worth its
computational cost.

```{r wtp, message=FALSE, warning=FALSE}
# Willingness to pay for the feature, per consumer (posterior means of beta_i)
wtp <- beta_draws[, "feature"] / (-beta_draws[, "price"])
wtp <- wtp[is.finite(wtp) & wtp > 0]
summary(round(wtp, 2))
```

## Estimation in Marketing Practice

The methods above are not interchangeable, and choosing among them is itself a
modeling decision with commercial stakes. Three considerations recur.

The first is the **source and structure of the choice data**. Revealed-preference
data—scanner panels, transaction logs, clickstreams—record actual purchases under
real prices and are the gold standard for demand elasticities, but they suffer
endogeneity: prices and promotions are set by firms that observe demand shifters the
analyst does not, biasing price coefficients toward zero. The standard corrections
are instrumental variables and control-function approaches, though these carry their
own pitfalls in marketing applications, where weak or invalid instruments can do
more harm than the bias they purport to fix [@rossi2014]. Stated-preference data
from conjoint experiments avoid price endogeneity by *designing* the attribute
variation, at the cost of hypothetical bias; the hierarchical-Bayes estimator
developed here is the dominant tool for recovering individual partworths from such
designs. Aggregate market-share data, where individual choices are unobserved,
require the random-coefficients demand system that inverts shares to recover
mean utilities and estimates the mixing distribution from share variation across
markets [@berry1995; @berry1995automobile]—the bridge between the consumer-level
models of this chapter and industry-level demand analysis, with active work on the
practical edge cases such as zero-share products [@dube2021random].

The second is **what the model is for**. If the deliverable is an aggregate
elasticity for a pricing decision, a carefully estimated mixed logit or
random-coefficients demand model suffices and the individual posteriors are a
by-product. If the deliverable is a targeting policy—who receives the offer—the
individual-level posteriors are the product, and the full hierarchical Bayes
apparatus is warranted. The choice between continuous (normal-mixing) heterogeneity
and discrete latent-class segmentation likewise turns on use: continuous mixing
predicts better when tastes vary smoothly, while latent classes are more
interpretable and align with managerial segmentation when genuine clusters exist
[@kamakura2005choice].

The third is **disciplined validation**. A choice model's in-sample fit is a weak
guide; the operative tests are holdout choice prediction, the recovery of known
substitution patterns, and—for any model relaxing IIA—evidence that the relaxation
changes substantive conclusions and not merely the likelihood. The structural
tradition in marketing has codified these standards, insisting that identification
be argued from the data and the design rather than asserted, and that the economic
content of estimates be interrogated before they inform a decision
[@chintagunta2006structural; @dube2021random]. The recurring lesson is that the
hard part of choice modeling is rarely the optimizer or the sampler; it is matching
the model's substitution structure and heterogeneity to the market and the decision
at hand.

## Key Takeaways

- Discrete choice rests on **random-utility theory** (@eq-rum): consumers maximize
  latent utility, only utility *differences* are identified, and the assumed
  distribution of the unobserved term determines the model.
- The **conditional logit** (@eq-logit) is the closed-form workhorse but imposes
  **IIA** (@eq-iia)—equal proportional cross-elasticities—which is behaviorally
  false whenever alternatives share unobserved attributes (the red-bus/blue-bus
  failure).
- **Nested logit, probit, and mixed logit** relax IIA by correlating the unobserved
  terms; mixed logit (@eq-mixedlogit) is the most flexible and links substitution
  structure to **consumer heterogeneity** through a mixing distribution.
- **Hierarchical Bayes** (@eq-hierarchy) reads the mixing distribution as a prior and
  borrows strength across consumers, recovering usable **individual-level** tastes
  from sparse data—the foundation of conjoint partworth estimation and household
  targeting.
- **MCMC**—Gibbs for the conjugate population parameters, Metropolis–Hastings
  (@eq-mh) and data augmentation for the non-conjugate choice likelihood—estimates
  the hierarchy, but its draws are credible only after **convergence diagnostics**.
- In practice the binding constraints are **data structure** (endogeneity in
  revealed preference, hypothetical bias in stated preference), the **decision the
  model serves**, and **out-of-sample validation**—not the estimation algorithm.
