# Platforms and Two-Sided Markets {#sec-platforms}

A *platform* is an intermediary that creates value primarily by enabling direct
interactions between two or more distinct groups of users. The defining feature is
not that the firm sells a product but that it sells *access*: a credit-card network
sells merchants access to cardholders and cardholders access to merchants; a
ride-hailing app sells riders access to drivers and drivers access to riders; a
videogame console sells gamers access to titles and developers access to an installed
base. What makes platforms a distinct object of study—rather than ordinary resellers
who buy and resell—is that the value each side derives depends on *how many* and
*which* users join the other side. Demand is interdependent across groups, and that
interdependence rewrites the firm's pricing, product, and competitive problems.

This chapter treats the platform as two tightly coupled objects: an **economic**
structure (a market with cross-group externalities that the platform internalizes
through prices and governance) and a **strategic** asset (an installed base whose
growth is self-reinforcing and whose tipping dynamics can hand a market to a single
winner). A serious account must connect the two, because the central managerial
decisions—what to charge each side, whom to subsidize, how to solve the
chicken-and-egg launch problem, how to govern the interactions that occur on the
platform—follow directly from the externality structure. The economics is due in
large part to the two-sided-market theory of Rochet and Tirole, and to the
platform-pricing analyses of Parker–Van Alstyne and of Armstrong, who worked out the
structure logic;<!-- CITE-NEEDED: Rochet & Tirole two-sided markets (2003 JEEA; 2006 RAND); Parker & Van Alstyne (2005 Mgmt Sci); Armstrong (2006 RAND) --> the
strategic machinery traces to the network-effects literature begun by @katz1985network
and its companion analyses of standards and compatibility.<!-- CITE-NEEDED: Katz & Shapiro technology adoption / network externalities and compatibility -->

The chapter proceeds from the inside out. It begins with the raw force that
distinguishes platforms—**network effects** and their cross-side cousins—and defines
them formally. It then builds the **pricing structure** problem, derives the
canonical Rochet–Tirole condition, and confronts what the structure implies for which
side pays and which is subsidized. From there it treats the dynamic problems
managers actually face: the **chicken-and-egg** launch problem, **tipping** and
**multihoming**, **governance** of the interactions a platform hosts, and
**competition** among platforms. It closes with measurement—how an analyst estimates
network effects from data, what identification assumptions that requires, and what
breaks them—and supplies reproducible code.

## Network Effects and Cross-Side Externalities

The primitive is the **network effect** (or *network externality*): a good exhibits a
network effect when a user's utility from it increases in the number of other users.
The telephone is the textbook case—a phone is useless if no one else owns one, and
its value rises with every additional subscriber. Formally, let $u_i$ denote user
$i$'s utility and $n$ the number of adopters. A pure (within-side, or *same-side*)
network effect holds when $\partial u_i / \partial n > 0$. The effect is **direct**
when utility depends on the raw count of fellow users and **indirect** when it
operates through a complementary product whose supply rises with the user base (more
console owners induce more game titles, which in turn attract more console owners).

Platforms add a second, asymmetric channel. Let the two sides be $A$ and $B$ with
participation $n_A$ and $n_B$. A **cross-side network effect**, or **cross-group
externality**, holds when a user on side $A$ benefits from participation on side
$B$:

$$
\frac{\partial u_A}{\partial n_B} > 0,
$$ {#eq-cross-side}

and symmetrically for side $B$. Cross-side effects are the engine of the two-sided
market: merchants value more cardholders, cardholders value more accepting merchants,
and the platform sits between them internalizing both. Crucially, cross-side effects
need not be symmetric in sign or magnitude. In media markets the effect can be
*negative* in one direction—viewers dislike advertisers even as advertisers value
viewers—so a television network or a search engine maximizes by restraining the ad
side to protect the audience side.<!-- CITE-NEEDED: Rochet & Tirole, negative cross-side (advertiser-audience) externality in media platforms --> Within-side effects can likewise be
negative: more merchants on a marketplace intensifies competition among them, and
more drivers on a ride app thins each driver's earnings.

A landmark statement frames the construct:

> A market is two-sided if the platform can affect the volume of transactions by
> charging more to one side of the market and reducing the price paid by the other
> side by an equal amount; in other words, the price structure matters, and platforms
> must design it so as to bring both sides on board.
>
> --- after Rochet and Tirole's characterization of two-sided markets<!-- CITE-NEEDED: Rochet & Tirole (2006), "Two-sided markets: a progress report," RAND J. Econ. -->

The italicized claim—that the *structure* of prices, not merely their level, affects
volume—is what separates a two-sided market from an ordinary one and is the source of
nearly everything distinctive in platform strategy. We make it precise below.

It is worth distinguishing two phenomena that are often conflated. A **network
effect** is a property of demand: it concerns how users' valuations move with
participation. A **scale economy** is a property of cost: it concerns how average cost
moves with output. The two generate superficially similar "bigger is better"
dynamics, but they are not the same and have different policy and competitive
implications—a platform can enjoy strong demand-side network effects while running at
roughly constant marginal cost, and a manufacturer can enjoy steep scale economies
with no network effect at all. Conflating them leads analysts to attribute to network
effects market structure that is really driven by fixed costs, and vice versa.^[A
further subtlety: indirect network effects mediated by a competitively supplied
complement can be re-described as a pecuniary externality rather than a true
technological externality. For most marketing purposes the distinction is second
order, but it matters for welfare analysis, because pecuniary externalities are
transfers and do not by themselves justify intervention.]

## Pricing Structure

The defining managerial decision in a two-sided market is not the *level* of price
but its *structure*—how the total price is allocated across the two sides. Because a
user on side $A$ confers a benefit on side $B$ (and conversely), the platform can
profitably charge one side below its own marginal cost, even below zero, recouping the
subsidy from the side that values access more. This is why so many platforms give one
side away free: free consumer search funded by advertisers, free operating systems
funded by application developers, free ride-app signup funded by per-trip commissions.

### The Rochet–Tirole Condition

Consider a monopoly platform facing sides $A$ and $B$. Let $p_A$ and $p_B$ be the
per-interaction prices, and suppose the *volume* of interactions $V$ depends on the
participation each price induces on each side, $V = D_A(p_A)\,D_B(p_B)$ in a reduced
form, with constant marginal cost $c$ per interaction. The platform chooses $(p_A,
p_B)$ to maximize profit

$$
\pi = (p_A + p_B - c)\,V(p_A, p_B).
$$ {#eq-platform-profit}

The first-order conditions deliver a generalization of the Lerner rule in which the
markup charged to each side is governed by *that side's* demand elasticity. Writing
$\eta_A$ and $\eta_B$ for the (own-price) elasticities of participation, the optimal
structure satisfies

$$
\frac{p_A + p_B - c}{p_A} = \frac{1}{\eta_A}, \qquad
\frac{p_A + p_B - c}{p_B} = \frac{1}{\eta_B},
$$ {#eq-rochet-tirole}

so that the ratio of prices across sides is inversely related to the ratio of
elasticities.<!-- CITE-NEEDED: Rochet & Tirole (2003, 2006), the inverse-elasticity price-structure condition --> The intuition is sharp and is the central
takeaway of the entire pricing literature: **the side that is more elastic—more
reluctant to join, more price-sensitive, more easily lost to an outside option—pays
less, and the side that values the interaction more inelastically pays more.** The
platform allocates the burden to the side that will tolerate it, because every user
retained on the elastic side is worth more through the cross-side externality than the
revenue forgone from subsidizing them.

Two forces refine this. First, the *strength of the cross-side externality* each side
exerts pulls price the other way: a side that confers a large benefit on the other
side should be subsidized to bring it on board, independent of its own elasticity.<!-- CITE-NEEDED: Armstrong (2006), "Competition in two-sided markets," RAND J. Econ. --> A platform thus subsidizes the side that is either very
price-sensitive *or* very valuable to the opposite side—often the same side (consumers
are both fickle and the reason advertisers pay). Second, whether users **single-home**
(join one platform) or **multihome** (join several) reshapes the markup: when one side
multihomes and the other single-homes, the platform holds a *competitive bottleneck*
over access to the single-homing side and extracts rents from the multihoming side, a
result we return to under platform competition.<!-- CITE-NEEDED: Armstrong (2006), competitive bottleneck under single-homing/multihoming -->

### What the Structure Implies

Three implications of @eq-rochet-tirole organize how managers read platform pricing.

The first is **below-cost and negative pricing as an equilibrium, not a promotion.**
When one side is highly elastic and confers a strong externality, the profit-maximizing
$p_A$ can be negative—the platform pays users to join (sign-up bonuses, free hardware
sold below cost, cashback). This is not predatory pricing in the antitrust sense and
not a temporary loss leader; it is the structure the externalities call for, and it
persists in steady state. Misreading a subsidized side as evidence of below-cost
predation is a recurring error in both managerial and regulatory analysis of
platforms.

The second is that **the identity of the subsidized side is an empirical question, not
a convention.** It depends on relative elasticities and externality strengths, both of
which vary by market and over the life cycle. Newspapers historically subsidized
readers and charged advertisers; some digital outlets have inverted this with
reader paywalls as ad demand softened. The structure is not fixed by industry; it is
chosen, and re-chosen as conditions move.

The third concerns **price floors and the limits of the logic.** The subsidy to one
side is bounded by the threat that subsidized "users" are not genuine participants but
arbitrageurs who consume the subsidy without conferring the externality—fake accounts
harvesting sign-up bonuses, drivers who never complete trips. The platform's ability
to run an aggressive price structure is therefore inseparable from its ability to
*govern* who is on the platform, which we take up in @sec-platform-governance.

@tbl-pricing-structure summarizes how the structure resolves across canonical
platforms.

| Platform type | Subsidized side | Money side | Why (elasticity / externality) |
|---|---|---|---|
| Payment card network | Cardholders (rewards) | Merchants (fees) | Cardholders elastic and confer large externality on merchants |
| Search engine / ad-funded media | Users (free) | Advertisers (auctions) | Users elastic; advertisers value access inelastically |
| Console videogames | Gamers (hardware sold near cost) | Developers (royalties) | Installed base drives developer willingness to pay |
| Ride-hailing | Riders (low fares early) | Drivers, then riders (commission) | Both sides elastic at launch; structure shifts with maturity |
| B2B marketplace | Buyers (free to browse) | Sellers (listing/transaction fees) | Sellers value buyer access; buyers easily lost |

: How the price structure resolves across canonical two-sided markets. The
subsidized side is the one that is more price-sensitive or confers the larger
cross-side externality. {#tbl-pricing-structure}

```{mermaid}
%%| label: fig-two-sided
%%| fig-cap: "Anatomy of a two-sided platform. Solid arrows are cross-side network effects (each side's value rises with the other's participation); dashed arrows are the price structure the platform sets. The platform internalizes the externalities by choosing how to split the total price across sides."
flowchart LR
    A["Side A\n(e.g., riders, cardholders, users)"]
    P["PLATFORM\nsets price structure\n(p_A, p_B)"]
    B["Side B\n(e.g., drivers, merchants, advertisers)"]
    A -- "cross-side effect (+)" --> B
    B -- "cross-side effect (+)" --> A
    P -. "p_A (often subsidized)" .-> A
    P -. "p_B (money side)" .-> B
```

## The Chicken-and-Egg Problem

Cross-side externalities that drive a mature platform's profits are a curse at launch.
No user on side $A$ wants to join until side $B$ is present, and no user on side $B$
wants to join until side $A$ is present: each side's participation is a best response
to the other's, and the empty market—where neither side joins—is always an
equilibrium. This is the **chicken-and-egg problem** (also the **penguin problem**,
after penguins who will not be first into the water), and it is the central obstacle to
platform entry.

Formally, write expected utility for a representative side-$A$ user as $u_A(n_B^e)$,
increasing in the *expected* side-$B$ participation $n_B^e$, and symmetrically for side
$B$. An adoption equilibrium is a fixed point $(n_A^*, n_B^*)$ at which each side's
participation is consistent with the other's. Because the best-response curves both
slope upward, the system generically admits multiple equilibria, including the
no-adoption point $(0,0)$. Coordinating users away from the empty equilibrium toward a
high-participation one is the launch problem; expectations are self-fulfilling, so the
platform must *manufacture the belief* that the other side will show up.

The strategies platforms use to escape the bad equilibrium map cleanly onto this
structure:

- **Subsidize one side hard at launch**, even at a loss, to seed participation that
  makes joining a best response for the other side. This is the dynamic counterpart to
  the static pricing structure: early subsidies are an investment in expectations.
- **Solve one side's problem in a single-sided mode first**, then open the second
  side—the "come for the tool, stay for the network" pattern. A product that is
  standalone-useful to side $A$ bootstraps an installed base before any side-$B$
  participation is required, breaking the simultaneity.
- **Micromarket / zip-code launch**: ignite the network in one narrow geography or
  vertical where the critical mass needed for a viable interaction is small, then
  replicate. Density, not aggregate scale, is what makes a local marketplace tip, so a
  platform that achieves liquidity in one city before expanding faces a far smaller
  coordination problem than one launching everywhere thinly.
- **Seed the marquee side**: recruit a small number of high-value side-$B$
  participants (anchor merchants, marquee game titles, celebrity sellers) whose
  presence credibly signals that the other side will follow.

The unifying logic is that all four reduce the *expected* participation gap that any
joining user must bridge—either by directly buying participation, by removing the
cross-side dependence at launch, or by shrinking the critical mass needed for the
first viable interactions.

## Tipping, Critical Mass, and Multihoming

Network effects make platform markets prone to **tipping**: once one platform's
installed base pulls ahead, the cross-side externalities make it ever more attractive,
the lead widens, and the market collapses toward a single dominant platform or a
narrow oligopoly. The qualitative dynamic is a positive-feedback loop, and the
quantitative threshold above which it ignites is the platform's **critical mass**—the
participation level beyond which growth becomes self-sustaining without further
subsidy.

A minimal model makes the threshold visible. Suppose the *fraction* of a market that
adopts in the next period, $x_{t+1}$, responds to current adoption $x_t$ through an
S-shaped (logistic) best-response map driven by network value,

$$
x_{t+1} = \frac{1}{1 + e^{-\beta\,(x_t - \theta)}},
$$ {#eq-tipping}

where $\beta$ scales the strength of the network effect and $\theta$ is an adoption
cost or threshold parameter. The map has up to three fixed points: a low (often zero)
equilibrium, an unstable interior fixed point that is exactly the **critical mass**,
and a high equilibrium near full adoption. Starting below critical mass, the system
decays to the empty market; starting above it, the system tips to dominance. The
managerial content of the chicken-and-egg subsidy is precisely to push initial
adoption past the unstable interior point.

Whether a market actually tips, however, depends on three moderators that
practitioners frequently overlook:

1. **Multihoming.** If users cheaply join several platforms at once, no single
   platform monopolizes access to them, the winner-take-all force weakens, and the
   market sustains multiple platforms in equilibrium.<!-- CITE-NEEDED: Armstrong (2006), multihoming weakens tipping --> Tipping
   requires that at least one side predominantly **single-homes**. Drivers and riders
   who run several apps simultaneously are the reason ride-hailing has not tipped to a
   single platform in most cities, despite strong cross-side effects.

2. **Differentiation and heterogeneous needs.** When users have heterogeneous tastes
   and platforms differentiate, distinct platforms can serve distinct segments and
   coexist—niche professional networks alongside a general one—because the network
   benefit of the dominant platform does not dominate the fit benefit of the
   specialized one for every user.

3. **Capacity and congestion (negative same-side effects).** When more participation
   degrades the experience—congestion, thinner matches, intra-side competition—the
   positive feedback is damped and full tipping is resisted.

The practical upshot is that "network effects imply winner-take-all" is a half-truth.
Network effects create a *tendency* to tip; multihoming, differentiation, and
congestion determine whether the tendency is realized. An analyst who observes a
fragmented platform market should look first to these moderators rather than conclude
the network effects are weak.

## Platform Governance {#sec-platform-governance}

Because a platform's value is the interactions it hosts rather than a product it makes,
the platform's central operational task is **governance**: the rules, prices,
information, and enforcement that determine who may participate, what interactions are
permitted, and how disputes and quality are managed. Governance is to a platform what
manufacturing is to a product firm—the locus where value is actually produced or
destroyed—and it is the lever that makes an aggressive price structure (@sec-platforms)
sustainable by ensuring subsidized participants are genuine.

Three governance problems recur.

**Quality and adverse selection.** Open access invites low-quality participants whose
presence imposes a negative cross-side externality (counterfeit sellers, bad drivers,
spam advertisers). The platform's classic remedy is a **reputation system**—ratings
and reviews that aggregate private experience into a public signal, mitigating the
lemons problem that would otherwise unravel the market [@akerlof1970]. Reputation
systems are themselves a designed object: how reviews are solicited, displayed,
and responded to changes the information they convey. Hotels that begin responding to
reviews gain in average rating but receive fewer, longer negative reviews thereafter,
as dissatisfied users self-censor unjustified complaints under anticipated scrutiny
[@proserpio2017]—a reminder that governance interventions change participant behavior,
not merely measure it. <!-- CITE-NEEDED: platform reputation/feedback system design and its effect on transaction volume (e.g., online marketplace reputation) -->

**The platform's own role as competitor.** A platform that both hosts third-party
sellers and sells its own products faces a conflict: data on third-party demand can be
used to enter their niches, and ranking algorithms can be tilted toward house
products. This **self-preferencing** problem trades short-run platform profit against
long-run participation incentives, because sellers who fear expropriation invest less
or exit. <!-- CITE-NEEDED: platform self-preferencing / dual role of marketplace as host and competitor -->

**Openness versus control.** A platform chooses how *open* to be—how freely third
parties may build on it, transact, and access its users. More openness recruits more
complementors and accelerates indirect network effects; more control protects quality,
captures more value, and guards the core experience. The choice is not binary but a
governance gradient, and the optimal point shifts over the life cycle: openness to
bootstrap the network early, selective control to monetize and protect quality once the
network is established. <!-- CITE-NEEDED: platform openness vs. control trade-off; platform boundary/governance choices -->

## Platform Competition

When platforms compete, the network-effects logic interacts with the pricing-structure
logic to produce dynamics with no single-sided analogue. Three results organize the
field.

**Competitive bottlenecks.** When one side single-homes and the other multihomes, each
platform is a monopoly gatekeeper over its single-homing users, because reaching them
requires joining *that* platform. Competing platforms then compete fiercely for the
single-homing side (often subsidizing it heavily) and extract rents from the
multihoming side, which has no choice but to join all platforms to reach all of the
single-homing users.<!-- CITE-NEEDED: Armstrong (2006), competitive-bottleneck rent extraction from the multihoming side --> This explains why advertisers (who multihome
across media) pay, while audiences (who often single-home their attention) are
courted.

**Envelopment.** A platform in an adjacent market can attack an incumbent by
*bundling* an overlapping functionality into its own user base, leveraging shared users
to enter without solving the chicken-and-egg problem from scratch. Platform competition
is therefore frequently *cross-market* rather than within a narrowly defined product
market—a messaging app entering payments, a search engine entering maps—and an
incumbent's most dangerous rival is often a large platform from a neighboring market
rather than a direct entrant. <!-- CITE-NEEDED: platform envelopment / coring as a competitive strategy -->

**Compatibility and standards.** Competing platforms choose whether to be
**compatible**—to let their networks interconnect (as banks share an ATM network) or
remain proprietary. Compatibility converts a fragmented set of small networks into one
large network, eliminating the network-effect basis for competition and shifting
rivalry to price and features; incompatibility preserves the prize of tipping but
risks splitting the market. The strategic choice of compatibility is thus a choice
about *whether to compete on network size at all*, and the firm with the larger
installed base typically prefers incompatibility (to press its advantage) while the
smaller prefers compatibility (to neutralize it)—the classic asymmetry of standards
wars [@katz1985network].<!-- CITE-NEEDED: Katz & Shapiro (1986), standards/compatibility wars and the large-base-prefers-incompatibility asymmetry -->

## Measuring Network Effects

For both research and managerial decisions the quantity of interest is the *magnitude*
of the network effect: by how much does a user's adoption or value rise with the size
of the relevant network? The estimation problem is hard because the very feedback that
defines a network effect also confounds its measurement.

### The Reflection / Simultaneity Problem

Let adoption (or value) for a user $i$ on side $A$ depend on the participation of side
$B$,

$$
y_{iA} = \alpha + \gamma\, n_B + \mathbf{x}_{iA}'\boldsymbol{\beta} + \varepsilon_{iA},
$$ {#eq-network-reg}

where $\gamma$ is the cross-side network-effect coefficient we wish to recover. The
naive regression of $y_{iA}$ on $n_B$ is biased for three reasons, each fatal on its
own.

First, **simultaneity**: $n_B$ is itself determined by $n_A$ (which aggregates the
$y_{iA}$), so $n_B$ is correlated with $\varepsilon_{iA}$ through the very feedback
loop @eq-cross-side describes. This is the platform analogue of the **reflection
problem** in the study of social interactions—each side's behavior reflects the
other's, and the two cannot be disentangled without an exclusion restriction.

Second, **correlated unobservables**: a city that is attractive to riders (good
weather, dense nightlife) is also attractive to drivers, so $n_B$ and
$\varepsilon_{iA}$ share common demand shifters that masquerade as a network effect.

Third, **homophily / sorting**: users who join a popular platform may differ in
unobserved ways from those who do not, biasing the cross-sectional association.

### Identification Strategies

Credible estimates therefore lean on one of a few designs, each buying identification
with an explicit and falsifiable assumption.

*Instrumental variables.* Find a variable that shifts side-$B$ participation $n_B$ but
is excluded from side $A$'s utility @eq-network-reg given controls—a cost shock or
policy change affecting only side $B$. The estimator is two-stage least squares; the
identifying assumption is the exclusion restriction, which is *not* testable and must
be argued substantively. Weak instruments (a first stage that barely moves $n_B$)
deliver badly biased second-stage estimates, so the first-stage $F$-statistic must be
reported and large.

*Structural estimation of demand with network effects.* Specify utility with a network
term and estimate the system jointly, instrumenting for the endogenous installed base
within a discrete-choice demand model in the tradition of @berry1995. The payoff is a
fully specified model that supports counterfactuals (what if the platform changed its
price structure?); the cost is that identification now rests on the full set of
functional-form and distributional assumptions, and misspecification of the network
term contaminates every counterfactual. <!-- CITE-NEEDED: structural estimation of indirect network effects in a two-sided market (e.g., videogames / payment cards) -->

*Natural experiments and panel variation.* A discrete shock that adds or removes
participation on one side—an entry, an exit, a regulatory ban, a platform policy
change rolled out in some markets and not others—permits a difference-in-differences
estimate of the cross-side effect, with the parallel-trends assumption replacing the
exclusion restriction as the identifying premise.

### A Reproducible Illustration

The simulation below makes the simultaneity bias concrete. We generate a two-sided
market in which participation on each side genuinely responds to the other (a true
cross-side effect), plus a market-level demand shifter that raises participation on
*both* sides. We then show that the naive OLS regression of side-$A$ participation on
side-$B$ participation overstates the network effect, while an instrument that shifts
only side $B$ recovers the truth.

```{r platform-iv, message=FALSE, warning=FALSE}
set.seed(20260620)

n_markets <- 500
gamma_true <- 0.40   # true cross-side effect of n_B on n_A

# Market-level demand shifter raising BOTH sides (the confound)
demand_shock <- rnorm(n_markets)

# Side-B cost instrument: shifts n_B only (excluded from side-A utility)
z_B <- rnorm(n_markets)

# Side-B participation: driven by the instrument and the common demand shock
n_B <- 0.9 * z_B + 0.8 * demand_shock + rnorm(n_markets)

# Side-A participation: true response to n_B, plus the same demand shock
eps_A <- 0.8 * demand_shock + rnorm(n_markets)   # correlated unobservable
n_A   <- gamma_true * n_B + eps_A

dat <- data.frame(n_A, n_B, z_B)

# (1) Naive OLS: biased upward by the shared demand shock
ols <- lm(n_A ~ n_B, data = dat)

# (2) 2SLS by hand: instrument n_B with z_B
first  <- lm(n_B ~ z_B, data = dat)
n_B_hat <- fitted(first)
iv     <- lm(n_A ~ n_B_hat, data = dat)

cat("True cross-side effect (gamma):", gamma_true, "\n")
cat("Naive OLS estimate:           ", round(coef(ols)["n_B"], 3), "\n")
cat("2SLS (IV) estimate:           ", round(coef(iv)["n_B_hat"], 3), "\n")
cat("First-stage F-statistic:      ",
    round(summary(first)$fstatistic["value"], 1), "\n")
```

The OLS coefficient is inflated well above $\gamma = 0.40$ because the shared demand
shock loads onto both sides; the instrumented estimate recovers the true effect, and
the first-stage $F$ confirms the instrument is strong. The lesson generalizes: **any
estimate of a network effect that does not confront the simultaneity of the two sides
should be read as an upper bound, not an effect.**

The next chunk visualizes the tipping dynamics of @eq-tipping, showing the unstable
interior fixed point that is the platform's critical mass.

```{r platform-tipping, message=FALSE, warning=FALSE}
#| fig-cap: "Tipping dynamics. The S-shaped best-response map (solid) crosses the 45-degree line (dashed) at three fixed points; the middle crossing is the unstable critical mass. Adoption starting below it decays to the empty market; above it, the market tips to near-full adoption."
beta  <- 12   # strength of the network effect
theta <- 0.5  # adoption threshold
f <- function(x) 1 / (1 + exp(-beta * (x - theta)))

x <- seq(0, 1, length.out = 400)
plot(x, f(x), type = "l", lwd = 2,
     xlab = "Current adoption  x_t",
     ylab = "Next-period adoption  x_(t+1)",
     main = "Network tipping and critical mass")
abline(0, 1, lty = 2)              # 45-degree line: fixed points are crossings
# Locate the unstable interior fixed point (critical mass) near x = theta
g <- function(x) f(x) - x
crit <- uniroot(g, c(0.3, 0.7))$root
points(crit, crit, pch = 19)
text(crit, crit, "  critical mass", pos = 4)
```

## Marketplace and Digital-Platform Applications

The framework applies across a wide span of commercially important platforms, and the
same constructs—cross-side externalities, price structure, chicken-and-egg, tipping,
multihoming, governance—resolve differently as their parameters differ.

**Marketplaces** (horizontal commerce platforms matching buyers and sellers) live or
die on *liquidity*—the probability that a given listing finds a counterparty quickly.
Liquidity is a same-market density requirement, which is why marketplaces launch
narrow (one category, one city) and why the chicken-and-egg subsidy is usually aimed at
the supply side, whose presence is the binding constraint on the first viable
transactions. Governance—reputation, dispute resolution, fraud control—is the operating
core, because a single bad interaction imposes a negative cross-side externality on the
whole market.

**Ad-funded digital media** (search, social, video) are the canonical case of an
asymmetric, partly negative cross-side effect: users dislike ads even as advertisers
value users. The price structure gives the user side away free and monetizes the
advertiser side, typically through auctions that price advertiser access to attention,
and the central governance tension is how much advertising load to impose before the
negative externality erodes the audience the advertisers are paying for.

**Transaction platforms** with strong multihoming (ride-hailing, food delivery)
illustrate the limits of tipping: because both sides cheaply run multiple apps, these
markets sustain competition despite powerful cross-side effects, and the strategic
contest is over inducing single-homing (loyalty programs, exclusivity, subscription
tiers) rather than over a one-time tip to dominance.

**Hardware/software systems** (consoles, operating systems, smartphones) are the
classic *indirect* network-effect platform: the installed base of users drives
complementor (developer) entry, which drives more users. Here the price structure
famously subsidizes the hardware (sold near or below cost) and monetizes the software
side through royalties, and compatibility decisions—whether to support cross-platform
play, common standards, or proprietary lock-in—are the principal competitive lever.

Across all four, the discipline the framework imposes is the same: identify the sides,
sign and size the cross-side externalities, read the price structure off the
elasticities and externalities, ask whether the market will tip (and whether
multihoming will stop it), and recognize that governance is where the value actually
accrues.

## Key Takeaways

- A platform's distinguishing feature is **cross-side network effects**
  (@eq-cross-side): each side's value rises with the other's participation, so demand
  is interdependent and the firm sells *access*, not a product.
- The central decision is the **price structure**, not the price level. The
  Rochet–Tirole condition (@eq-rochet-tirole) says the more elastic side—and the side
  conferring the larger externality—is subsidized, often below cost or for free; this
  is an equilibrium, not a promotion.
- Cross-side effects create the **chicken-and-egg** launch problem: the empty market is
  always an equilibrium, and platforms escape it by subsidizing a side, going
  standalone-useful first, or igniting density in a narrow micromarket.
- Network effects create a *tendency* to **tip**, but **multihoming**,
  differentiation, and congestion determine whether it is realized—"network effects
  imply winner-take-all" is a half-truth.
- **Governance** (reputation systems, self-preferencing restraint, openness/control) is
  where platform value is produced and is what makes an aggressive price structure
  sustainable.
- **Measuring** a network effect requires confronting the simultaneity of the two
  sides; an estimate that ignores it is an upper bound, recoverable only with a valid
  side-specific instrument, a structural model, or a credible natural experiment.

## Further Reading

The two-sided market was formalized by Rochet and Tirole, with complementary pricing
treatments by Parker–Van Alstyne and by Armstrong;<!-- CITE-NEEDED: Rochet & Tirole (2003 JEEA, 2006 RAND); Parker & Van Alstyne (2005 Mgmt Sci); Armstrong (2006 RAND) --> the network-effects
foundations are due to @katz1985network and the Katz–Shapiro work on standards.<!-- CITE-NEEDED: Katz & Shapiro (1986, 1994) on compatibility and systems competition --> The
platform-competition and business-model literature is surveyed in
@casadesus2015competing, and the economics of digital platforms more broadly in
@goldfarb2022 and @tucker2019digital. Multiproduct and bundling considerations
relevant to envelopment appear in @armstrong2018multiproduct.
The branding, advertising, and marketing-finance machinery this chapter draws on is
developed in @sec-branding, @sec-advertising, and @sec-marketing-finance.
