# Platforms and Two-Sided Markets {#sec-platforms}

A *platform* is an intermediary that creates value primarily by enabling direct
interactions between two or more distinct groups of users. The defining feature is
not that the firm sells a product but that it sells *access*: a credit-card network
sells merchants access to cardholders and cardholders access to merchants; a
ride-hailing app sells riders access to drivers and drivers access to riders; a
videogame console sells gamers access to titles and developers access to an installed
base. What makes platforms a distinct object of study—rather than ordinary resellers
who buy and resell—is that the value each side derives depends on *how many* and
*which* users join the other side. Demand is interdependent across groups, and that
interdependence rewrites the firm's pricing, product, and competitive problems.

This chapter treats the platform as two tightly coupled objects: an **economic**
structure (a market with cross-group externalities that the platform internalizes
through prices and governance) and a **strategic** asset (an installed base whose
growth is self-reinforcing and whose tipping dynamics can hand a market to a single
winner). A serious account must connect the two, because the central managerial
decisions—what to charge each side, whom to subsidize, how to solve the
chicken-and-egg launch problem, how to govern the interactions that occur on the
platform—follow directly from the externality structure. The economics is due in
large part to the two-sided-market theory of Rochet and Tirole, and to the
platform-pricing analyses of Parker–Van Alstyne and of Armstrong, who worked out the
price-structure logic [@rochet2003platformcompetiti; @rochet2006sidedmarkets; @parker2005sidednetwork; @armstrong2006competitisided]; the
strategic machinery traces to the network-effects literature begun by @katz1985network
and its companion analyses of standards and compatibility [@katz1986technologadoption; @katz1994systemscompetiti].

The chapter proceeds from the inside out. It begins with the raw force that
distinguishes platforms—**network effects** and their cross-side cousins—and defines
them formally. It then builds the **pricing structure** problem, derives the
canonical Rochet–Tirole condition, and confronts what the structure implies for which
side pays and which is subsidized. From there it treats the dynamic problems
managers actually face: the **chicken-and-egg** launch problem, **tipping** and
**multihoming**, **governance** of the interactions a platform hosts, and
**competition** among platforms. It closes with measurement—how an analyst estimates
network effects from data, what identification assumptions that requires, and what
breaks them—and supplies reproducible code.

## Network Effects and Cross-Side Externalities

The primitive is the **network effect** (or *network externality*): a good exhibits a
network effect when a user's utility from it increases in the number of other users.
The telephone is the textbook case—a phone is useless if no one else owns one, and
its value rises with every additional subscriber. Formally, let $u_i$ denote user
$i$'s utility and $n$ the number of adopters. A pure (within-side, or *same-side*)
network effect holds when $\partial u_i / \partial n > 0$. The effect is **direct**
when utility depends on the raw count of fellow users and **indirect** when it
operates through a complementary product whose supply rises with the user base (more
console owners induce more game titles, which in turn attract more console owners).

Platforms add a second, asymmetric channel. Let the two sides be $A$ and $B$ with
participation $n_A$ and $n_B$. A **cross-side network effect**, or **cross-group
externality**, holds when a user on side $A$ benefits from participation on side
$B$:

$$
\frac{\partial u_A}{\partial n_B} > 0,
$$ {#eq-cross-side}

and symmetrically for side $B$. Cross-side effects are the engine of the two-sided
market: merchants value more cardholders, cardholders value more accepting merchants,
and the platform sits between them internalizing both. Crucially, cross-side effects
need not be symmetric in sign or magnitude. In media markets the effect can be
*negative* in one direction—viewers dislike advertisers even as advertisers value
viewers—so a television network or a search engine maximizes by restraining the ad
side to protect the audience side [@rochet2006sidedmarkets]. Television is the case
where this has been estimated rather than assumed: modeling viewing and advertising
demand jointly, @wilbur2008sidedempirical finds viewers are roughly twice as
ad-averse as advertisers are viewer-hungry, so the profit-maximizing network carries
materially less advertising than a one-sided reading of the ad market would imply.
Where exactly the boundary of a multi-sided platform lies is itself contested, and
the distinction between a platform and a reseller who simply buys and re-sells turns
on which side retains control rights rather than on the pattern of externalities
[@hagiu2015multisided]. Within-side effects can likewise be
negative: more merchants on a marketplace intensifies competition among them, and
more drivers on a ride app thins each driver's earnings.

A landmark statement frames the construct:

> A market is two-sided if the platform can affect the volume of transactions by
> charging more to one side of the market and reducing the price paid by the other
> side by an equal amount; in other words, the price structure matters, and platforms
> must design it so as to bring both sides on board.
>
> --- after Rochet and Tirole's characterization of two-sided markets [@rochet2006sidedmarkets]

The italicized claim—that the *structure* of prices, not merely their level, affects
volume—is what separates a two-sided market from an ordinary one and is the source of
nearly everything distinctive in platform strategy. We make it precise below.

It is worth distinguishing two phenomena that are often conflated. A **network
effect** is a property of demand: it concerns how users' valuations move with
participation. A **scale economy** is a property of cost: it concerns how average cost
moves with output. The two generate superficially similar "bigger is better"
dynamics, but they are not the same and have different policy and competitive
implications—a platform can enjoy strong demand-side network effects while running at
roughly constant marginal cost, and a manufacturer can enjoy steep scale economies
with no network effect at all. Conflating them leads analysts to attribute to network
effects market structure that is really driven by fixed costs, and vice versa.^[A
further subtlety: indirect network effects mediated by a competitively supplied
complement can be re-described as a pecuniary externality rather than a true
technological externality. For most marketing purposes the distinction is second
order, but it matters for welfare analysis, because pecuniary externalities are
transfers and do not by themselves justify intervention.]

## Pricing Structure

The defining managerial decision in a two-sided market is not the *level* of price
but its *structure*—how the total price is allocated across the two sides. Because a
user on side $A$ confers a benefit on side $B$ (and conversely), the platform can
profitably charge one side below its own marginal cost, even below zero, recouping the
subsidy from the side that values access more. This is why so many platforms give one
side away free: free consumer search funded by advertisers, free operating systems
funded by application developers, free ride-app signup funded by per-trip commissions.

### The Rochet–Tirole Condition

Consider a monopoly platform facing sides $A$ and $B$. Let $p_A$ and $p_B$ be the
per-interaction prices, and suppose the *volume* of interactions $V$ depends on the
participation each price induces on each side, $V = D_A(p_A)\,D_B(p_B)$ in a reduced
form, with constant marginal cost $c$ per interaction. The platform chooses $(p_A,
p_B)$ to maximize profit

$$
\pi = (p_A + p_B - c)\,V(p_A, p_B).
$$ {#eq-platform-profit}

The first-order conditions deliver a generalization of the Lerner rule in which the
markup charged to each side is governed by *that side's* demand elasticity. Writing
$\eta_A$ and $\eta_B$ for the (own-price) elasticities of participation, the optimal
structure satisfies

$$
\frac{p_A + p_B - c}{p_A} = \frac{1}{\eta_A}, \qquad
\frac{p_A + p_B - c}{p_B} = \frac{1}{\eta_B},
$$ {#eq-rochet-tirole}

so that the ratio of prices across sides is inversely related to the ratio of
elasticities [@rochet2003platformcompetiti; @rochet2006sidedmarkets]. The intuition is sharp and is the central
takeaway of the entire pricing literature: **the side that is more elastic—more
reluctant to join, more price-sensitive, more easily lost to an outside option—pays
less, and the side that values the interaction more inelastically pays more.** The
platform allocates the burden to the side that will tolerate it, because every user
retained on the elastic side is worth more through the cross-side externality than the
revenue forgone from subsidizing them.

Two forces refine this. First, the *strength of the cross-side externality* each side
exerts pulls price the other way: a side that confers a large benefit on the other
side should be subsidized to bring it on board, independent of its own elasticity [@armstrong2006competitisided]. A platform thus subsidizes the side that is either very
price-sensitive *or* very valuable to the opposite side—often the same side (consumers
are both fickle and the reason advertisers pay). Second, whether users **single-home**
(join one platform) or **multihome** (join several) reshapes the markup: when one side
multihomes and the other single-homes, the platform holds a *competitive bottleneck*
over access to the single-homing side and extracts rents from the multihoming side, a
result we return to under platform competition [@armstrong2006competitisided].

### What the Structure Implies

Three implications of @eq-rochet-tirole organize how managers read platform pricing.

The first is **below-cost and negative pricing as an equilibrium, not a promotion.**
When one side is highly elastic and confers a strong externality, the profit-maximizing
$p_A$ can be negative—the platform pays users to join (sign-up bonuses, free hardware
sold below cost, cashback). This is not predatory pricing in the antitrust sense and
not a temporary loss leader; it is the structure the externalities call for, and it
persists in steady state. Misreading a subsidized side as evidence of below-cost
predation is a recurring error in both managerial and regulatory analysis of
platforms.

The second is that **the identity of the subsidized side is an empirical question, not
a convention.** It depends on relative elasticities and externality strengths, both of
which vary by market and over the life cycle. Newspapers historically subsidized
readers and charged advertisers; some digital outlets have inverted this with
reader paywalls as ad demand softened. The structure is not fixed by industry; it is
chosen, and re-chosen as conditions move.

The third concerns **price floors and the limits of the logic.** The subsidy to one
side is bounded by the threat that subsidized "users" are not genuine participants but
arbitrageurs who consume the subsidy without conferring the externality—fake accounts
harvesting sign-up bonuses, drivers who never complete trips. The platform's ability
to run an aggressive price structure is therefore inseparable from its ability to
*govern* who is on the platform, which we take up in @sec-platform-governance.

@tbl-pricing-structure summarizes how the structure resolves across canonical
platforms.

| Platform type | Subsidized side | Money side | Why (elasticity / externality) |
|---|---|---|---|
| Payment card network | Cardholders (rewards) | Merchants (fees) | Cardholders elastic and confer large externality on merchants |
| Search engine / ad-funded media | Users (free) | Advertisers (auctions) | Users elastic; advertisers value access inelastically |
| Console videogames | Gamers (hardware sold near cost) | Developers (royalties) | Installed base drives developer willingness to pay |
| Ride-hailing | Riders (low fares early) | Drivers, then riders (commission) | Both sides elastic at launch; structure shifts with maturity |
| B2B marketplace | Buyers (free to browse) | Sellers (listing/transaction fees) | Sellers value buyer access; buyers easily lost |

: How the price structure resolves across canonical two-sided markets. The
subsidized side is the one that is more price-sensitive or confers the larger
cross-side externality. {#tbl-pricing-structure}

@fig-two-sided gives the general anatomy: two sides linked by cross-side network
effects, with the platform choosing how to split the total price across them.

```{mermaid}
%%| label: fig-two-sided
%%| fig-cap: "Anatomy of a two-sided platform. Solid arrows are cross-side network effects (each side's value rises with the other's participation); dashed arrows are the price structure the platform sets. The platform internalizes the externalities by choosing how to split the total price across sides."
flowchart LR
    A["Side A\n(e.g., riders, cardholders, users)"]
    P["PLATFORM\nsets price structure\n(p_A, p_B)"]
    B["Side B\n(e.g., drivers, merchants, advertisers)"]
    A -- "cross-side effect (+)" --> B
    B -- "cross-side effect (+)" --> A
    P -. "p_A (often subsidized)" .-> A
    P -. "p_B (money side)" .-> B
```

### Beyond Two Tiers: Ad Load and Fee as a Screening Menu {#sec-platform-gas}

@eq-rochet-tirole treats each side as facing a single price. Content platforms
increasingly do not: Netflix, Spotify, and YouTube fund the same catalogue through
some mix of subscription fees and advertising, and the mix is offered to viewers
as a *choice*. That turns the price-structure problem into a **mechanism design**
problem, and specifically into the screening problem of @sec-pricing—a menu whose
options must satisfy incentive compatibility because the platform cannot observe
which viewer is which.

The two-dimensional type is what makes the setting distinctive. A viewer is
characterized by how much the content is worth to them *and* by how much
advertising annoys them, and those two are not the same trait. The instrument has
two dimensions to match: an ad load and a fee. @gutt2026gas formalize the general
version—the **generalized ad-supported subscription** (GAS) mechanism, a menu
offering a continuum of (ad intensity, fee) combinations, which nests today's
ad-only, subscription-only, and two-tier "free-with-ads versus paid-ad-free"
models as special cases and is implementable as a slider or a short finite menu.

Their result is the one worth carrying into practice. GAS is revenue-optimal
within this broad class, but *the familiar two-tier menu is often nearly optimal*.
Calibrating to a video-on-demand setting with an incentive-compatible
willingness-to-pay survey, they find GAS raises revenue by 148% over ad-only, 20%
over subscription-only, and **1.8%** over the two-tier menu. The first two numbers
say the mix matters enormously; the third says the fine-grained menu mostly does
not. They also show that advertising-intensity caps of the kind the EU Digital
Services Act and Digital Fairness Act contemplate, and competitive pressure, both
shrink coverage of low-value viewers and push platforms toward less ad-intensive
offerings.

The simulation below builds the menu from scratch by incentive compatibility—
viewers self-select, the platform never observes types—and grows it greedily one
option at a time.

```{r platform-gas-menu}
set.seed(11)
N <- 4000
theta <- rlnorm(N, log(11), 0.55)   # $/month value of the content to a viewer
gamma <- rlnorm(N, log(1.1), 0.70)  # $ disutility per unit of ad load
A <- 8; adrev <- 1.25               # max ad load; platform ad revenue per unit

# Every (ad load, fee) pair the platform could put on a menu, and what one taker
# of that option is worth to the platform.
grid <- expand.grid(a = seq(0, A, by = 1), f = seq(0, 30, by = 1))
grid$yield <- grid$f + adrev * grid$a

# Consumers self-select: each takes the option maximizing theta - gamma*a - f,
# and abstains if the best option leaves negative surplus. This is the incentive-
# compatibility constraint doing its work -- the platform never sees (theta, gamma).
rev_of <- function(rows) {
  S <- matrix(theta, N, length(rows)) - outer(gamma, grid$a[rows]) -
       matrix(grid$f[rows], N, length(rows), byrow = TRUE)
  k <- max.col(cbind(0, S), ties.method = "first"); t <- k > 1
  c(rev = sum(grid$yield[rows][k[t] - 1]) / N, cover = mean(t))
}

# Greedily grow the menu one option at a time; the plateau is the whole point.
grow <- function(pool, steps = 4) {
  chosen <- integer(0); out <- NULL
  for (s in seq_len(steps)) {
    cand <- setdiff(pool, chosen)
    g <- vapply(cand, function(x) rev_of(c(chosen, x))["rev"], numeric(1))
    chosen <- c(chosen, cand[which.max(g)]); out <- rbind(out, rev_of(chosen))
  }
  list(path = out, menu = grid[chosen, c("a", "f")])
}

all_opts <- seq_len(nrow(grid))
ad_only  <- rev_of(which(grid$a == A & grid$f == 0))
sub_only <- {
  paid <- which(grid$a == 0)
  rev_of(paid[which.max(vapply(paid, function(g) rev_of(g)["rev"], numeric(1)))])
}
two_tier <- {
  free <- which(grid$a == A & grid$f == 0); paid <- which(grid$a == 0)
  b <- paid[which.max(vapply(paid, function(g) rev_of(c(free, g))["rev"], numeric(1)))]
  rev_of(c(free, b))
}
gas <- grow(all_opts)

tab <- rbind(`ad-only` = ad_only, `subscription-only` = sub_only, `two-tier` = two_tier,
             `GAS (2 options)` = gas$path[2, ], `GAS (3 options)` = gas$path[3, ],
             `GAS (4 options)` = gas$path[4, ])
tab <- cbind(tab, `gain over two-tier %` = 100 * (tab[, "rev"] / two_tier["rev"] - 1))
round(tab, 3)
```

The ordering reproduces theirs. Moving from either single-instrument mechanism to
a two-option menu is worth roughly twenty percent; moving from the two-option menu
to a richer one is worth about one percent, and the greedy path flattens
immediately—the third and fourth options add nothing, because two options already
separate the types that are separable. The reason the two-tier menu does so well
is that the incentive-compatibility constraints bind in one direction only: a
viewer who tolerates ads must not be tempted by the ad-free tier, and one cheap
option plus one clean option is nearly enough to arrange that.

Regulation acts on the menu by deleting its high-ad options, which is a different
intervention from capping the price.

```{r platform-gas-cap}
# A regulatory cap on advertising intensity (the EU Digital Services Act
# direction of travel) removes the high-ad options from the menu entirely.
capped <- grow(which(grid$a <= 3))
rbind(uncapped = gas$path[2, ], `ad load capped at 3` = capped$path[2, ])
cat("\nOptimal menu, uncapped (ad load, fee):\n");  print(gas$menu[1:2, ])
cat("\nOptimal menu, capped   (ad load, fee):\n");  print(capped$menu[1:2, ])
```

The cap costs revenue, as intended, but it also trims coverage: the option that
served the most ad-tolerant, lowest-willingness-to-pay viewers is the one removed,
and the replacement asks them for cash they were self-selecting away from. That is
the distributional footnote to any advertising cap, and it is invisible in an
analysis that looks only at the average ad load.
### Release Scheduling as a Complement to the Menu {#sec-platform-release}

The GAS menu governs the *terms* on which content is consumed. A content platform
also chooses the *schedule*, and that decision turns out to be a distinct lever on
the same catalogue. @zhao2026serialized study a digital platform for serialized
content—books released chapter by chapter—and ask whether the platform should drop
all chapters at once or meter them out.

The two obvious answers each capture half of the mechanism. Simultaneous release
enables **binge consumption**, which is what hooks a reader on a title.
Sequential release forces **return visits**, and a return visit is the occasion on
which a reader discovers some *other* title in the catalogue. Framed that way,
bingeing and exploration look like substitutes: every hour spent finishing one
book is an hour not spent browsing.

The paper's result is that a hybrid schedule makes them complements. Revenue is
maximized when roughly the first half of a book's chapters are released
simultaneously and the remainder sequentially. The simultaneous block does the
hooking; the sequential tail converts that attachment into repeated visits, and
the visits do the exploring. The managerial reading is that release policy is a
portfolio instrument rather than a title-level one—the schedule for any single
book is being chosen for what it does to consumption of the *rest* of the
catalogue, which is why a title-by-title optimization gets it wrong. It is the
same logic that makes the price structure of @eq-rochet-tirole a structure rather
than a set of independent prices.


## The Chicken-and-Egg Problem

Cross-side externalities that drive a mature platform's profits are a curse at launch.
No user on side $A$ wants to join until side $B$ is present, and no user on side $B$
wants to join until side $A$ is present: each side's participation is a best response
to the other's, and the empty market—where neither side joins—is always an
equilibrium. This is the **chicken-and-egg problem** (also the **penguin problem**,
after penguins who will not be first into the water), and it is the central obstacle to
platform entry.

Formally, write expected utility for a representative side-$A$ user as $u_A(n_B^e)$,
increasing in the *expected* side-$B$ participation $n_B^e$, and symmetrically for side
$B$. An adoption equilibrium is a fixed point $(n_A^*, n_B^*)$ at which each side's
participation is consistent with the other's. Because the best-response curves both
slope upward, the system generically admits multiple equilibria, including the
no-adoption point $(0,0)$. Coordinating users away from the empty equilibrium toward a
high-participation one is the launch problem; expectations are self-fulfilling, so the
platform must *manufacture the belief* that the other side will show up.

The strategies platforms use to escape the bad equilibrium map cleanly onto this
structure:

- **Subsidize one side hard at launch**, even at a loss, to seed participation that
  makes joining a best response for the other side. This is the dynamic counterpart to
  the static pricing structure: early subsidies are an investment in expectations.
- **Solve one side's problem in a single-sided mode first**, then open the second
  side—the "come for the tool, stay for the network" pattern. A product that is
  standalone-useful to side $A$ bootstraps an installed base before any side-$B$
  participation is required, breaking the simultaneity.
- **Micromarket / zip-code launch**: ignite the network in one narrow geography or
  vertical where the critical mass needed for a viable interaction is small, then
  replicate. Density, not aggregate scale, is what makes a local marketplace tip, so a
  platform that achieves liquidity in one city before expanding faces a far smaller
  coordination problem than one launching everywhere thinly.
- **Seed the marquee side**: recruit a small number of high-value side-$B$
  participants (anchor merchants, marquee game titles, celebrity sellers) whose
  presence credibly signals that the other side will follow.

The unifying logic is that all four reduce the *expected* participation gap that any
joining user must bridge—either by directly buying participation, by removing the
cross-side dependence at launch, or by shrinking the critical mass needed for the
first viable interactions.

## Designing the First Interaction: Fulfillment on a Co-Creation Platform {#sec-platform-fulfillment}

The launch tactics above buy participation with money, scope, or geography. A cheaper
lever sits inside the first session itself: *what the visitor is allowed to see before
being asked to commit*. This margin has become sharp on **generative-AI
content-generation platforms**, where a visitor types a prompt and the platform returns
a poster, a portrait, a song. The technology's headline virtue is immediacy—but
immediacy is also what the platform gives away for free, and a visitor who has already
received the thing has no reason to register for it.

@jiang2026fulfillment name the design variable **fulfillment**: how much of the
co-created output is revealed before the registration wall. Their theory, built on the
value-co-creation tradition [@vargo2004], is that fulfillment moves two motivational
states in *opposite* directions. **Value-in-use** is the visitor's recognition that
their own input shaped the output—it rises with fulfillment, because you cannot see
your fingerprints on something you have not seen. **Curiosity** is anticipatory
motivation, and by the information-gap account [@loewenstein1994curiosity] it exists
only while the experience remains perceptually open; a full reveal closes the gap and
extinguishes it. Since registration needs both, the optimum is interior: **partial
fulfillment**, which reveals enough to prove the visitor's contribution mattered while
withholding enough to leave something worth coming back for.

A randomized field experiment on a live platform, with a follow-up online study,
bears this out: partial fulfillment beats both full and no fulfillment on registration.
The design also crosses fulfillment with how the registration prompt is **framed**, and
finds a substitution: loss framing—emphasizing what the visitor forfeits by not
registering—lifts registration on average, but the lift attenuates under full
fulfillment, because a visitor holding the finished artifact has little left to lose.
Formal mediation shows why the interior optimum wins. Both partial and full fulfillment
raise value-in-use, so that pathway alone would prescribe showing everything; only
partial fulfillment sustains curiosity, and it is the joint activation of the two
that carries the effect. The advantage persists past the registration event into
subsequent engagement and return visits, and it appears only when the visitor
genuinely co-produced the output—fulfillment is not a generic teaser mechanic but a
lever on *co-creation*, which is why it grows with output quality rather than
substituting for it.

```{mermaid}
%%| label: fig-fulfillment-pathways
%%| fig-cap: "Why the optimum is interior. Value-in-use is increasing in how much of the co-created output is revealed; curiosity is single-peaked, because a full reveal closes the information gap. Registration needs both, so partial fulfillment dominates the corners."
flowchart LR
    N["No reveal\n(nothing shown)"]
    P["Partial reveal\n(some output shown)"]
    F["Full reveal\n(all output shown)"]
    V["Value-in-use:\n'my input shaped this'"]
    C["Curiosity:\nexperience stays open"]
    R["Registration,\nthen return visits"]
    N -- "low" --> V
    P -- "high" --> V
    F -- "highest" --> V
    N -- "moderate" --> C
    P -- "highest" --> C
    F -- "extinguished" --> C
    V --> R
    C --> R
```

The replication reconstructs the experiment: a three-by-two randomization of
fulfillment against message framing, two latent pathways with the curvature the theory
predicts, a binary registration outcome, and a post-registration return count. It
recovers the arm ranking, tests the framing substitution, and decomposes the
partial-versus-full gap into its two pathways with a bootstrap.

```{r platform-fulfillment, message=FALSE, warning=FALSE}
set.seed(1200)
n <- 9000

# --- 3 x 2 randomization: how much of the co-created output is revealed
#     before registration, crossed with how the registration prompt is framed.
fulf  <- factor(sample(c("none", "partial", "full"), n, TRUE),
                levels = c("none", "partial", "full"))
frame <- factor(sample(c("gain", "loss"), n, TRUE), levels = c("gain", "loss"))

# --- Two motivational pathways, moving in opposite directions ----------------
# Value-in-use rises with what you can see: proof your prompt shaped the output.
mu_viu <- c(none = 0.15, partial = 0.70, full = 0.85)[as.character(fulf)]
# Curiosity needs the experience to stay perceptually open: a full reveal closes it.
mu_cur <- c(none = 0.45, partial = 0.75, full = 0.20)[as.character(fulf)]
viu <- mu_viu + rnorm(n, 0, 0.35)
cur <- mu_cur + rnorm(n, 0, 0.35)

# --- Registration ------------------------------------------------------------
b_viu <- 1.30; b_cur <- 1.10
loss  <- as.integer(frame == "loss")
full  <- as.integer(fulf == "full")
eta   <- -1.20 + b_viu * viu + b_cur * cur +
          0.35 * loss - 0.30 * loss * full   # loss framing substitutes for a full reveal
reg   <- rbinom(n, 1, plogis(eta))

# Post-registration return behavior, driven by the same experiential state.
returns <- rpois(n, exp(-0.3 + 0.9 * viu + 0.6 * cur)) * reg

d <- data.frame(fulf, frame, viu, cur, reg, returns)

# --- (1) Registration by arm -------------------------------------------------
tab <- round(100 * tapply(d$reg, list(d$fulf, d$frame), mean), 1)

# --- (2) Main effect of fulfillment, and the framing interaction -------------
m_main <- lm(reg ~ fulf, data = d)
m_int  <- lm(reg ~ fulf * frame, data = d)

# --- (3) Parallel-mediator decomposition of partial vs. full -----------------
paths <- function(dd) {
  b  <- coef(lm(reg ~ viu + cur + fulf, data = dd))
  ap <- with(dd, c(mean(viu[fulf == "partial"]) - mean(viu[fulf == "full"]),
                   mean(cur[fulf == "partial"]) - mean(cur[fulf == "full"])))
  c(viu = unname(ap[1] * b["viu"]), curiosity = unname(ap[2] * b["cur"]))
}
bt <- replicate(400, paths(d[sample.int(n, n, TRUE), ]))
ci <- apply(bt, 1, quantile, c(0.025, 0.975))

cat("Registration rate (%) by fulfillment x framing\n"); print(tab); cat("\n")
cat(sprintf("Partial vs. none  : %+.1f pp\n", 100 * coef(m_main)["fulfpartial"]))
cat(sprintf("Full    vs. none  : %+.1f pp\n", 100 * coef(m_main)["fulffull"]))
cat(sprintf("Partial vs. full  : %+.1f pp\n\n",
            100 * (coef(m_main)["fulfpartial"] - coef(m_main)["fulffull"])))
cat(sprintf("Loss-framing lift -- no reveal      : %+.1f pp\n",
            100 * coef(m_int)["frameloss"]))
cat(sprintf("Loss-framing lift -- partial reveal : %+.1f pp\n",
            100 * (coef(m_int)["frameloss"] + coef(m_int)["fulfpartial:frameloss"])))
cat(sprintf("Loss-framing lift -- full reveal    : %+.1f pp  (substitution)\n",
            100 * (coef(m_int)["frameloss"] + coef(m_int)["fulffull:frameloss"])))
cat("\nWhy partial beats full -- decomposition of the partial-vs-full gap:\n")
cat(sprintf("  via value-in-use : %+.1f pp  [%+.1f, %+.1f]\n",
            100 * mean(bt["viu", ]), 100 * ci[1, "viu"], 100 * ci[2, "viu"]))
cat(sprintf("  via curiosity    : %+.1f pp  [%+.1f, %+.1f]\n",
            100 * mean(bt["curiosity", ]), 100 * ci[1, "curiosity"], 100 * ci[2, "curiosity"]))
cat(sprintf("\nReturn visits per 100 arrivals: none %.0f, partial %.0f, full %.0f\n",
            100 * mean(d$returns[d$fulf == "none"]),
            100 * mean(d$returns[d$fulf == "partial"]),
            100 * mean(d$returns[d$fulf == "full"])))
```

The decomposition is the part worth dwelling on. Judged on value-in-use alone, the
partial arm is *behind*—it shows less, so visitors are less certain their prompt
mattered—and a platform that optimized that single mediator would reveal everything.
The curiosity pathway more than repays the deficit, and it does so only in the interior
arm. This is the general hazard of single-mediator reasoning about experience design:
when two pathways respond to the same lever with opposite curvature, the sign of the
total effect is not recoverable from either one, and a mediation analysis that
instruments only the pathway the designer finds intuitive will prescribe a corner
solution.

Two boundaries deserve emphasis before the tactic is generalized. The effect is a
property of **co-creation**, not of withholding: it appears when the visitor's own
input shaped the output and strengthens with output quality, so a platform that
throttles a mediocre generator is not running this experiment—it is degrading the
product. And the framing interaction is a warning about stacking persuasion levers. A
loss-framed prompt and a full reveal are partial substitutes, so a team that A/B-tests
copy and reveal policy independently will double-count their gains and be surprised
when the combined launch underperforms the sum of its tests. The general point recurs
in @sec-platform-governance: platform design levers interact, and the interaction is
usually substitution rather than complementarity.

## Tipping, Critical Mass, and Multihoming

Network effects make platform markets prone to **tipping**: once one platform's
installed base pulls ahead, the cross-side externalities make it ever more attractive,
the lead widens, and the market collapses toward a single dominant platform or a
narrow oligopoly. The qualitative dynamic is a positive-feedback loop, and the
quantitative threshold above which it ignites is the platform's **critical mass**—the
participation level beyond which growth becomes self-sustaining without further
subsidy.

A minimal model makes the threshold visible. Suppose the *fraction* of a market that
adopts in the next period, $x_{t+1}$, responds to current adoption $x_t$ through an
S-shaped (logistic) best-response map driven by network value,

$$
x_{t+1} = \frac{1}{1 + e^{-\beta\,(x_t - \theta)}},
$$ {#eq-tipping}

where $\beta$ scales the strength of the network effect and $\theta$ is an adoption
cost or threshold parameter. The map has up to three fixed points: a low (often zero)
equilibrium, an unstable interior fixed point that is exactly the **critical mass**,
and a high equilibrium near full adoption. Starting below critical mass, the system
decays to the empty market; starting above it, the system tips to dominance. The
managerial content of the chicken-and-egg subsidy is precisely to push initial
adoption past the unstable interior point.

Whether a market actually tips, however, depends on three moderators that
practitioners frequently overlook:

1. **Multihoming.** If users cheaply join several platforms at once, no single
   platform monopolizes access to them, the winner-take-all force weakens, and the
   market sustains multiple platforms in equilibrium [@armstrong2006competitisided]. Tipping
   requires that at least one side predominantly **single-homes**. Drivers and riders
   who run several apps simultaneously are the reason ride-hailing has not tipped to a
   single platform in most cities, despite strong cross-side effects.

2. **Differentiation and heterogeneous needs.** When users have heterogeneous tastes
   and platforms differentiate, distinct platforms can serve distinct segments and
   coexist—niche professional networks alongside a general one—because the network
   benefit of the dominant platform does not dominate the fit benefit of the
   specialized one for every user.

3. **Capacity and congestion (negative same-side effects).** When more participation
   degrades the experience—congestion, thinner matches, intra-side competition—the
   positive feedback is damped and full tipping is resisted.

The practical upshot is that "network effects imply winner-take-all" is a half-truth.
Network effects create a *tendency* to tip; multihoming, differentiation, and
congestion determine whether the tendency is realized. An analyst who observes a
fragmented platform market should look first to these moderators rather than conclude
the network effects are weak.

## Platform Governance {#sec-platform-governance}

Because a platform's value is the interactions it hosts rather than a product it makes,
the platform's central operational task is **governance**: the rules, prices,
information, and enforcement that determine who may participate, what interactions are
permitted, and how disputes and quality are managed. Governance is to a platform what
manufacturing is to a product firm—the locus where value is actually produced or
destroyed—and it is the lever that makes an aggressive price structure (@sec-platforms)
sustainable by ensuring subsidized participants are genuine.

Six governance problems recur.

**Quality and adverse selection.** Open access invites low-quality participants whose
presence imposes a negative cross-side externality (counterfeit sellers, bad drivers,
spam advertisers). The platform's classic remedy is a **reputation system**—ratings
and reviews that aggregate private experience into a public signal, mitigating the
lemons problem that would otherwise unravel the market [@akerlof1970]. Reputation
systems are themselves a designed object: how reviews are solicited, displayed,
and responded to changes the information they convey. Hotels that begin responding to
reviews gain in average rating but receive fewer, longer negative reviews thereafter,
as dissatisfied users self-censor unjustified complaints under anticipated scrutiny
[@proserpio2017]—a reminder that governance interventions change participant behavior,
not merely measure it. The design question is old enough to have its own literature: the earliest systematic treatment of online feedback mechanisms already framed reputation as an engineered institution with identifiable failure modes rather than a naturally occurring signal [@dellarocas2003digitizatword].

**Matching technology and where its costs land.** The platform's search and
filtering machinery is a governance instrument, not a neutral convenience: it
decides which counterparties a participant ever sees, and therefore which
interactions are possible at all. Its usual justification is efficiency—cheaper
matching for the side doing the searching—and the usual evaluation stops at that
stage. @barach2026filtering ran the evaluation past it. In a randomized field
experiment on a large online labor market, employers given candidate filters
interviewed 3.4 percent fewer applicants, which is the intended saving. But
filters work by sorting on commonly preferred, easily observable attributes, so
every employer using them converges on the same shortlist. Using a machine-learning
proxy for quality unobservable to the employer, the interviewed pool thinned at
both ends—28.9 percent fewer low-quality applicants and 36.9 percent fewer
high-quality ones—and among employers who actually used the filters, offers were
8.9 percent more likely to be rejected and contracted wages 2.5 percent higher.
The screening cost the tool removed reappeared as bargaining power on the other
side of the market.

Two implications matter for platform design. First, a matching feature has a
cross-side externality of its own: concentrating demand on a visible subset of
supply raises that subset's outside options and depresses the discoverability of
everyone else, which is a governance choice about who gets liquidity, not a
retrieval optimization. Second, an efficiency metric measured at one stage of a
multi-stage transaction is not a welfare metric for the transaction. The same
caution applies to any ranking, recommendation, or filter that a marketplace ships
as a search improvement—the question is always whether the saved cost was
eliminated or relocated (@sec-policy-spillovers).

**Which objective the ranking serves.** @barach2026filtering show that a matching
tool can relocate cost rather than remove it. @greminger2026ranking asks the prior
question the platform has to answer before it ships any ranking at all: *what is
the ranking maximizing?* A marketplace can rank to maximize its own take
(commissions or markups), the number of transactions, or consumer welfare, and
these need not agree. The descriptive half of the paper shows that lower-priced
and higher-utility alternatives gain disproportionate demand from being ranked
higher—which implies that promoting them raises transactions and consumer surplus
while plausibly *lowering* commission revenue, since commissions scale with price.
That is the trade-off in its intuitive form, and it is the reason regulators treat
ranking as a competition-policy object.

The paper's contribution is to price the trade-off rather than assert it. It
estimates a structural demand model in which consumers search and discover
products—so a product's rank changes whether it is *seen*, not merely how
attractive it looks—and then constructs the optimal ranking under each objective
separately. The result is a negative one, and more useful for it: all three
objective-specific rankings improve on both a neutral benchmark and the platform's
status quo along *all three* metrics at once, and the differences among them are
small. The binding constraint is not that revenue and welfare pull apart; it is
that the status quo is leaving value on the table for everyone.

The simulation below reproduces the structure of that argument: position governs
discovery, three objectives imply three different orderings, and the three
orderings nonetheless land close together and well above the incumbent.

```{r platform-ranking-objectives}
set.seed(2026)
J <- 12; N <- 6000; commission <- 0.15
p <- round(runif(J, 40, 95))                   # prices: moderate dispersion
q <- rnorm(J, 0, 2.0)                          # match quality: wider dispersion
alpha <- rlnorm(N, log(0.035), 0.55)           # heterogeneous price sensitivity
V <- matrix(q, N, J, byrow = TRUE) - outer(alpha, p)

# Common random numbers: fix the taste shocks and the discovery draws once, so
# that comparing two rankings compares the rankings and not the noise.
gum <- function(n, m) -log(-log(matrix(runif(n * m), n, m)))
E <- gum(N, J); E0 <- 1.0 + gum(N, 1); U <- matrix(runif(N * J), N, J)

# Position governs DISCOVERY: rank r is examined with probability 0.85^(r-1).
# Rank changes what a consumer sees, not what it is worth once seen.
outcomes <- function(ord) {
  rk <- integer(J); rk[ord] <- seq_len(J)
  u <- V + E; u[U >= rep(0.85^(rk - 1), each = N)] <- -Inf
  k <- max.col(cbind(E0, u), ties.method = "first")   # column 1 = no purchase
  b <- k > 1; j <- k[b] - 1
  c(revenue = commission * sum(p[j]) / N,
    transactions = mean(b),
    welfare = sum(u[cbind(which(b), j)]) / N)
}

# Build each objective's ranking greedily, slot by slot: fill position 1 with
# whichever product maximizes the objective there, then position 2, and so on.
greedy <- function(obj) {
  rem <- seq_len(J); ord <- integer(0)
  for (s in seq_len(J)) {
    sc <- vapply(rem, function(x) outcomes(c(ord, x, rem[rem != x]))[obj], numeric(1))
    pick <- rem[which.max(sc)]
    ord <- c(ord, pick); rem <- rem[rem != pick]
  }
  ord
}

legacy <- q + rnorm(J, 0, 2.5)     # status quo: a stale relevance score
rankings <- list(
  `max revenue`      = greedy("revenue"),
  `max transactions` = greedy("transactions"),
  `max welfare`      = greedy("welfare"),
  `neutral (random)` = sample(J),
  `status quo`       = order(legacy, decreasing = TRUE))

round(t(sapply(rankings, outcomes)), 3)
```

Read the table by column. Each objective-specific ranking does best, or ties, on
its own column—but every one of them dominates both the neutral benchmark and the
status quo on *all three* columns at once, and the spread across the top three
rows is tiny next to the gap between them and the bottom two. A platform arguing
that it cannot promote cheaper or better-matched listings without sacrificing
revenue is, on this evidence, describing a constraint it has not tested.

Two caveats keep the exercise honest. The first is that the result is calibration-
dependent in an interpretable way: here match quality is more dispersed than price,
so the products that convert best are not systematically the cheapest, and the
objectives largely agree about which products belong at the top. Widen the price
distribution relative to the quality distribution and a genuine revenue-versus-
welfare tension reappears. The paper's contribution is that in its data the first
regime is the one that obtains—which is an empirical finding, not a theorem. The
second is the caution from @barach2026filtering: these are equilibrium-free
counterfactuals in which sellers do not reprice in response to the new ranking,
which is exactly the margin on which the relocated-cost story operated.


**Who actually sets the rules.** Everything above treats governance as something
the platform does *to* its participants. @huber2026counterorchestration document
the return channel. Tracing fifteen years of Apple's iOS ecosystem (2009–2024),
they show how large complementors—Spotify, Epic Games—forced changes to rules
including the commission structure and the anti-steering provisions, not through
market channels, which were closed to them, but through **nonmarket** ones:
public advocacy, litigation, and coalition-building.

They identify three mechanisms by which these "complementor giants" orchestrate
change. They create public occasions for other stakeholders to voice grievances;
they generate *confluence* across grievances that were separately too small to
matter; and they render the underlying tension salient enough that regulators,
media, users, other developers, and eventually the platform owner itself respond
in ways that legitimize the challenge. The pressure is cumulative rather than
episodic, which is why a fifteen-year window is the right unit of observation and
a single dispute is not.

Two things follow. For a complementor in a concentrated platform market,
bargaining power is built outside the market before it can be exercised inside
it—a strategic capability that looks nothing like the pricing and product levers
this chapter has otherwise been about. For a regulator, the finding points away
from prescribing platform-owner conduct directly and toward strengthening the
conditions under which complementors can advocate for themselves, which is a much
lighter-touch instrument than rule-writing and does not require the regulator to
know the efficient rule in advance.

**The platform's own role as competitor.** A platform that both hosts third-party
sellers and sells its own products faces a conflict: data on third-party demand can be
used to enter their niches, and ranking algorithms can be tilted toward house
products. This **self-preferencing** problem trades short-run platform profit against
long-run participation incentives, because sellers who fear expropriation invest less
or exit. The empirical record on Amazon shows the platform entering exactly the product spaces where third-party sellers had demonstrated demand, which is the sharpest available evidence that the conflict is real rather than hypothetical [@zhu2018competingcomplemen]; the theoretical counterpart shows when a search-neutrality rule raises or lowers welfare once the platform internalizes both roles [@zou2021].

**Openness versus control.** A platform chooses how *open* to be—how freely third
parties may build on it, transact, and access its users. More openness recruits more
complementors and accelerates indirect network effects; more control protects quality,
captures more value, and guards the core experience. The choice is not binary but a
governance gradient, and the optimal point shifts over the life cycle: openness to
bootstrap the network early, selective control to monetize and protect quality once the
network is established.  [@boudreau2010openplatform]

**Provenance of contributions.** Generative AI adds a governance problem the reputation
machinery was never designed for: a contribution can now be cheap, fluent, plausible, and
not the contributor's own knowledge, which corrodes the signal that a knowledge platform
exists to aggregate. Stack Overflow's December 2022 prohibition on ChatGPT-generated
questions and answers is the natural experiment, and @borwankar2026chatgpt read it with a
difference-in-differences design against the AskProgramming subreddit as a control.
The results split cleanly along the openness–control gradient. On the *quality* side, answers
after the restriction became longer, more linguistically complex, and more positive in
tone, and they drew more upvotes—the community judged them better. On the *volume* side,
both questions and answers posted fell, and first-time contributors declined; questions
themselves were unaffected, which localizes the effect to the answering side where the
policy bit. A ban on synthetic contributions, in other words, bought quality with
participation, exactly the trade the governance gradient predicts.

Two features make the evidence unusually credible for a platform-policy study. The first
is replication under an independent shock: Italy's temporary nationwide ChatGPT ban
removes the tool without any platform rule changing, and reproduces the pattern, which
rules out explanations resting on Stack Overflow's announcement rather than on AI
availability. The second is a mechanism test—a scenario-based experiment with 440
participants—yielding direct evidence of **compensatory knowledge signaling**: barred
from the tool, contributors invest in demonstrating human expertise. That mechanism is
what a purely aggregate DiD cannot deliver, and it matters managerially, because a rule
that works by changing what contributors want to signal will not survive a change in what
the audience finds impressive.

**Policy as stance signal.** The Stack Overflow case treats a rule as a constraint on
contributions, but a rule is also an *announcement*, and the announcement can move
participation on its own. @huang2026stances isolate that channel with two natural
experiments of opposite sign on leading Chinese visual-arts platforms. Lofter launched
its own AI image generator—a pro-AI stance—and creator activity **fell**. Graffiti
Kingdom prohibited AI-generated artwork—an anti-AI stance—and creator activity
**rose**. Neither movement is what a tool-features story predicts: adding a free
generator lowers the cost of producing on Lofter, and banning a technique removes an
option on Graffiti Kingdom. What the two experiments share is the stance the policy
communicated—about AI's role on the platform, about the platform's commitment to human
creators, and about the competitive environment those creators expect to face—and
multiple lines of evidence in the paper point to that signal rather than to the
mechanics of the tool or the enforcement.

The heterogeneity connects the result back to the price-structure and tipping logic of
this chapter. Reductions on Lofter were largest among higher-popularity,
multi-homing, and AI-averse creators: the creators with the most to lose from
synthetic substitutes, and—crucially—the ones already holding an outside option, since
multihoming (@sec-platforms) is what converts a discouraging signal into an exit rather
than a grumble. Analysis of creator posts names the concerns doing the work: replacement
risk, perceived low quality, and copyright infringement. For a platform whose subsidized
side *is* the human creator base, this makes a generative-AI product decision inseparable
from a governance decision. The feature is priced by the side that must interpret it, and
a launch communicated as augmenting creators is a different governance act from the same
launch communicated as substituting for them.

## Platform Competition

When platforms compete, the network-effects logic interacts with the pricing-structure
logic to produce dynamics with no single-sided analogue. Three results organize the
field.

**Competitive bottlenecks.** When one side single-homes and the other multihomes, each
platform is a monopoly gatekeeper over its single-homing users, because reaching them
requires joining *that* platform. Competing platforms then compete fiercely for the
single-homing side (often subsidizing it heavily) and extract rents from the
multihoming side, which has no choice but to join all platforms to reach all of the
single-homing users [@armstrong2006competitisided]. This explains why advertisers (who multihome
across media) pay, while audiences (who often single-home their attention) are
courted.

**Envelopment.** A platform in an adjacent market can attack an incumbent by
*bundling* an overlapping functionality into its own user base, leveraging shared users
to enter without solving the chicken-and-egg problem from scratch. Platform competition
is therefore frequently *cross-market* rather than within a narrowly defined product
market—a messaging app entering payments, a search engine entering maps—and an
incumbent's most dangerous rival is often a large platform from a neighboring market
rather than a direct entrant.  [@eisenmann2011platformenvelopme]

**Compatibility and standards.** Competing platforms choose whether to be
**compatible**—to let their networks interconnect (as banks share an ATM network) or
remain proprietary. Compatibility converts a fragmented set of small networks into one
large network, eliminating the network-effect basis for competition and shifting
rivalry to price and features; incompatibility preserves the prize of tipping but
risks splitting the market. The strategic choice of compatibility is thus a choice
about *whether to compete on network size at all*, and the firm with the larger
installed base typically prefers incompatibility (to press its advantage) while the
smaller prefers compatibility (to neutralize it)—the classic asymmetry of standards
wars [@katz1985network; @katz1986technologadoption].

### Market Structure and the Realized Externality {#sec-platform-entry-externality}

The three results above concern how competition splits rents. A fourth question, less
studied and often more consequential for policy, is how competition changes the *size of
the thing the platform does to the world outside it*. Platforms are routinely justified by
their externalities—ride-hailing reduces drunk driving, home-sharing spreads tourist
spending, bike sharing displaces short car trips—but those externalities are produced by
*usage*, not by the existence of an app. Between service availability and realized
externality sits adoption, and adoption depends on supply.

@liu2026platform put that chain to a test with the staggered rollout of the two largest
dockless bike-sharing platforms, Ofo and Mobike, across Chinese cities, using a stacked
difference-in-differences design with air quality as the outcome. The twin-platform
setting buys two analyses from one rollout. Treating the two firms as a single provider,
entry improves urban air quality by 3.82%, a 3.13-point reduction in the Air Quality
Index. Distinguishing market structure changes the picture: **dual-platform entry reduces
AQI by 6.47 points—more than twice the pooled estimate—while single-platform entry has no
statistically significant effect at all.** The heterogeneity analysis lines up with the
demand-side story: improvements are larger in cities with more domestic migrants, larger
well-educated populations, denser public transit, and better bike-lane infrastructure, and
smaller where the terrain is steep or rainfall frequent.

The mechanism is supply-side and cuts against the reflex that rivalry dissipates a
platform's value. Dockless fleets are capacity-constrained: the binding constraint on a
short trip is whether a usable bike is within a short walk of the origin. A second
platform does not split a fixed network the way a second social network splits an
audience; it adds bikes, densifies coverage, and lowers the expected walk, pushing
adoption past the level at which mode substitution actually happens. Multihoming is
nearly free (two apps, no switching cost, no content to migrate), so the two networks are
substitutes in the *user's* choice but complements in the *city's* realized service level.
The general condition is worth stating because it travels: when a platform's externality
depends on density of physical availability, and users multihome at near-zero cost,
competition can be complementary at the level of the outcome society cares about even
while it is rivalrous at the level of firm profit.

The design deserves as much attention as the finding. A pooled treatment indicator would
have reported a modest average effect and missed that the entire effect lives in
dual-entry cities. The chapter's earlier warning about network-effect estimation has a
counterpart here: **an average treatment effect estimated across heterogeneous market
structures is not the effect of the treatment in any market**. The template generalizes to
marketing settings with two rivals entering markets at different times—delivery apps,
streaming services, retail-media networks—wherever a researcher can code not only *when*
entry happened but *how many* platforms were present afterward.

Stacking is what makes the estimation honest under staggered timing. Rather than a single
two-way fixed-effects regression—which, with staggered adoption and heterogeneous effects,
uses already-treated units as controls and can return a weighted average with negative
weights—the analyst builds one clean sub-experiment per entry event, each with its own
treated cohort and not-yet-treated controls, then stacks them and estimates with
event-by-cohort fixed effects [@cengiz2019minimum; @callaway2021did]. The simulation below
reproduces both the estimation and the aggregation trap.

```{r platform-stacked-did}
#| message: false
#| warning: false
set.seed(51)

# --- Staggered entry across 180 cities, 24 periods ----------------------------
# A third of cities never get a platform; a third get one; a third get two.
n_city <- 180; n_per <- 24
structure_type <- rep(c("never", "single", "dual"), each = n_city / 3)
entry <- ifelse(structure_type == "never", Inf,
                sample(8:16, n_city, replace = TRUE))

# True effects: single entry does essentially nothing to the outcome;
# dual entry (denser fleet) moves it. AQI is coded so NEGATIVE = cleaner air.
tau <- c(never = 0, single = -0.4, dual = -6.5)

panel <- expand.grid(city = 1:n_city, period = 1:n_per)
panel$type   <- structure_type[panel$city]
panel$entry  <- entry[panel$city]
panel$post   <- as.integer(panel$period >= panel$entry)
panel$aqi    <- 100 + rnorm(n_city, 0, 8)[panel$city] +      # city level
                0.6 * panel$period +                          # common trend
                tau[panel$type] * panel$post +
                rnorm(nrow(panel), 0, 4)

# --- (1) Pooled TWFE: one indicator for "a platform is present" ----------------
pooled <- lm(aqi ~ post + factor(city) + factor(period), data = panel)

# --- (2) Stacked DiD: one clean sub-experiment per entry cohort ---------------
# For each entry period g, keep cohort-g cities plus not-yet-treated cities,
# in a window around g; stack with cohort-specific fixed effects.
cohorts <- sort(unique(entry[is.finite(entry)]))
stack <- do.call(rbind, lapply(cohorts, function(g) {
  keep <- subset(panel,
                 period >= g - 4 & period <= g + 4 &
                 (entry == g | entry > g + 4))
  keep$treat  <- as.integer(keep$entry == g)
  keep$after  <- as.integer(keep$period >= g)
  keep$cohort <- g
  keep
}))
stack$cell <- interaction(stack$cohort, stack$city, drop = TRUE)
stack$time <- interaction(stack$cohort, stack$period, drop = TRUE)

stacked_pooled <- lm(aqi ~ treat:after + cell + time, data = stack)

# --- (3) Stacked DiD with market structure ------------------------------------
stack$dual   <- as.integer(stack$type == "dual")
stacked_type <- lm(aqi ~ treat:after + treat:after:dual + cell + time, data = stack)

cat(sprintf("True effect      : single %.2f | dual %.2f\n", tau["single"], tau["dual"]))
cat(sprintf("Pooled TWFE      : %.2f (one 'platform present' indicator)\n",
            coef(pooled)["post"]))
cat(sprintf("Stacked, pooled  : %.2f\n", coef(stacked_pooled)["treat:after"]))
cat(sprintf("Stacked, single  : %.2f\n", coef(stacked_type)["treat:after"]))
cat(sprintf("Stacked, dual    : %.2f\n",
            coef(stacked_type)["treat:after"] + coef(stacked_type)["treat:after:dual"]))
```

The pooled specifications recover something like the average of a null effect and a large
one—a number that describes no city in the sample—while the structure-interacted stack
separates them. That is the estimation counterpart of the substantive claim: the
externality is not a property of the service being available, it is a property of the
market structure that determines how much of the service is actually used.

## Measuring Network Effects

For both research and managerial decisions the quantity of interest is the *magnitude*
of the network effect: by how much does a user's adoption or value rise with the size
of the relevant network? The estimation problem is hard because the very feedback that
defines a network effect also confounds its measurement.

### The Reflection / Simultaneity Problem

Let adoption (or value) for a user $i$ on side $A$ depend on the participation of side
$B$,

$$
y_{iA} = \alpha + \gamma\, n_B + \mathbf{x}_{iA}'\boldsymbol{\beta} + \varepsilon_{iA},
$$ {#eq-network-reg}

where $\gamma$ is the cross-side network-effect coefficient we wish to recover. The
naive regression of $y_{iA}$ on $n_B$ is biased for three reasons, each fatal on its
own.

First, **simultaneity**: $n_B$ is itself determined by $n_A$ (which aggregates the
$y_{iA}$), so $n_B$ is correlated with $\varepsilon_{iA}$ through the very feedback
loop @eq-cross-side describes. This is the platform analogue of the **reflection
problem** in the study of social interactions—each side's behavior reflects the
other's, and the two cannot be disentangled without an exclusion restriction.

Second, **correlated unobservables**: a city that is attractive to riders (good
weather, dense nightlife) is also attractive to drivers, so $n_B$ and
$\varepsilon_{iA}$ share common demand shifters that masquerade as a network effect.

Third, **homophily / sorting**: users who join a popular platform may differ in
unobserved ways from those who do not, biasing the cross-sectional association.

### Identification Strategies

Credible estimates therefore lean on one of a few designs, each buying identification
with an explicit and falsifiable assumption.

*Instrumental variables.* Find a variable that shifts side-$B$ participation $n_B$ but
is excluded from side $A$'s utility @eq-network-reg given controls—a cost shock or
policy change affecting only side $B$. The estimator is two-stage least squares; the
identifying assumption is the exclusion restriction, which is *not* testable and must
be argued substantively. Weak instruments (a first stage that barely moves $n_B$)
deliver badly biased second-stage estimates, so the first-stage $F$-statistic must be
reported and large.

*Structural estimation of demand with network effects.* Specify utility with a network
term and estimate the system jointly, instrumenting for the endogenous installed base
within a discrete-choice demand model in the tradition of @berry1995. The payoff is a
fully specified model that supports counterfactuals (what if the platform changed its
price structure?); the cost is that identification now rests on the full set of
functional-form and distributional assumptions, and misspecification of the network
term contaminates every counterfactual.  [@nair2004empiricalindirect; @rysman2004competitibetween]

*Natural experiments and panel variation.* A discrete shock that adds or removes
participation on one side—an entry, an exit, a regulatory ban, a platform policy
change rolled out in some markets and not others—permits a difference-in-differences
estimate of the cross-side effect, with the parallel-trends assumption replacing the
exclusion restriction as the identifying premise.

### A Reproducible Illustration

The simulation below makes the simultaneity bias concrete. We generate a two-sided
market in which participation on each side genuinely responds to the other (a true
cross-side effect), plus a market-level demand shifter that raises participation on
*both* sides. We then show that the naive OLS regression of side-$A$ participation on
side-$B$ participation overstates the network effect, while an instrument that shifts
only side $B$ recovers the truth.

```{r platform-iv, message=FALSE, warning=FALSE}
set.seed(20260620)

n_markets <- 500
gamma_true <- 0.40   # true cross-side effect of n_B on n_A

# Market-level demand shifter raising BOTH sides (the confound)
demand_shock <- rnorm(n_markets)

# Side-B cost instrument: shifts n_B only (excluded from side-A utility)
z_B <- rnorm(n_markets)

# Side-B participation: driven by the instrument and the common demand shock
n_B <- 0.9 * z_B + 0.8 * demand_shock + rnorm(n_markets)

# Side-A participation: true response to n_B, plus the same demand shock
eps_A <- 0.8 * demand_shock + rnorm(n_markets)   # correlated unobservable
n_A   <- gamma_true * n_B + eps_A

dat <- data.frame(n_A, n_B, z_B)

# (1) Naive OLS: biased upward by the shared demand shock
ols <- lm(n_A ~ n_B, data = dat)

# (2) 2SLS by hand: instrument n_B with z_B
first  <- lm(n_B ~ z_B, data = dat)
n_B_hat <- fitted(first)
iv     <- lm(n_A ~ n_B_hat, data = dat)

cat("True cross-side effect (gamma):", gamma_true, "\n")
cat("Naive OLS estimate:           ", round(coef(ols)["n_B"], 3), "\n")
cat("2SLS (IV) estimate:           ", round(coef(iv)["n_B_hat"], 3), "\n")
cat("First-stage F-statistic:      ",
    round(summary(first)$fstatistic["value"], 1), "\n")
```

The OLS coefficient is inflated well above $\gamma = 0.40$ because the shared demand
shock loads onto both sides; the instrumented estimate recovers the true effect, and
the first-stage $F$ confirms the instrument is strong. The lesson generalizes: **any
estimate of a network effect that does not confront the simultaneity of the two sides
should be read as an upper bound, not an effect.**

The next chunk visualizes the tipping dynamics of @eq-tipping, showing the unstable
interior fixed point that is the platform's critical mass.

```{r platform-tipping, message=FALSE, warning=FALSE}
#| fig-cap: "Tipping dynamics. The S-shaped best-response map (solid) crosses the 45-degree line (dashed) at three fixed points; the middle crossing is the unstable critical mass. Adoption starting below it decays to the empty market; above it, the market tips to near-full adoption."
beta  <- 12   # strength of the network effect
theta <- 0.5  # adoption threshold
f <- function(x) 1 / (1 + exp(-beta * (x - theta)))

x <- seq(0, 1, length.out = 400)
plot(x, f(x), type = "l", lwd = 2,
     xlab = "Current adoption  x_t",
     ylab = "Next-period adoption  x_(t+1)",
     main = "Network tipping and critical mass")
abline(0, 1, lty = 2)              # 45-degree line: fixed points are crossings
# Locate the unstable interior fixed point (critical mass) near x = theta
g <- function(x) f(x) - x
crit <- uniroot(g, c(0.3, 0.7))$root
points(crit, crit, pch = 19)
text(crit, crit, "  critical mass", pos = 4)
```

## Marketplace and Digital-Platform Applications

The framework applies across a wide span of commercially important platforms, and the
same constructs—cross-side externalities, price structure, chicken-and-egg, tipping,
multihoming, governance—resolve differently as their parameters differ.

**Marketplaces** (horizontal commerce platforms matching buyers and sellers) live or
die on *liquidity*—the probability that a given listing finds a counterparty quickly.
Liquidity is a same-market density requirement, which is why marketplaces launch
narrow (one category, one city) and why the chicken-and-egg subsidy is usually aimed at
the supply side, whose presence is the binding constraint on the first viable
transactions. Governance—reputation, dispute resolution, fraud control—is the operating
core, because a single bad interaction imposes a negative cross-side externality on the
whole market.

**Ad-funded digital media** (search, social, video) are the canonical case of an
asymmetric, partly negative cross-side effect: users dislike ads even as advertisers
value users. The price structure gives the user side away free and monetizes the
advertiser side, typically through auctions that price advertiser access to attention,
and the central governance tension is how much advertising load to impose before the
negative externality erodes the audience the advertisers are paying for.

**Transaction platforms** with strong multihoming (ride-hailing, food delivery)
illustrate the limits of tipping: because both sides cheaply run multiple apps, these
markets sustain competition despite powerful cross-side effects, and the strategic
contest is over inducing single-homing (loyalty programs, exclusivity, subscription
tiers) rather than over a one-time tip to dominance.

**Hardware/software systems** (consoles, operating systems, smartphones) are the
classic *indirect* network-effect platform: the installed base of users drives
complementor (developer) entry, which drives more users. Here the price structure
famously subsidizes the hardware (sold near or below cost) and monetizes the software
side through royalties, and compatibility decisions—whether to support cross-platform
play, common standards, or proprietary lock-in—are the principal competitive lever.

Across all four, the discipline the framework imposes is the same: identify the sides,
sign and size the cross-side externalities, read the price structure off the
elasticities and externalities, ask whether the market will tip (and whether
multihoming will stop it), and recognize that governance is where the value actually
accrues.

## Key Takeaways

- A platform's distinguishing feature is **cross-side network effects**
  (@eq-cross-side): each side's value rises with the other's participation, so demand
  is interdependent and the firm sells *access*, not a product.
- The central decision is the **price structure**, not the price level. The
  Rochet–Tirole condition (@eq-rochet-tirole) says the more elastic side—and the side
  conferring the larger externality—is subsidized, often below cost or for free; this
  is an equilibrium, not a promotion.
- Cross-side effects create the **chicken-and-egg** launch problem: the empty market is
  always an equilibrium, and platforms escape it by subsidizing a side, going
  standalone-useful first, or igniting density in a narrow micromarket.
- The **first session** is a design lever, not just a funnel. On generative-AI
  co-creation platforms, @jiang2026fulfillment show that *partial* fulfillment—
  revealing some but not all of the co-created output before the registration wall—
  beats both full and no reveal, because value-in-use rises with disclosure while
  curiosity is single-peaked (@sec-platform-fulfillment). Loss-framed prompts and a
  full reveal are substitutes, so the levers cannot be A/B-tested independently.
- Network effects create a *tendency* to **tip**, but **multihoming**,
  differentiation, and congestion determine whether it is realized—"network effects
  imply winner-take-all" is a half-truth.
- **Governance** (reputation systems, self-preferencing restraint, openness/control) is
  where platform value is produced and is what makes an aggressive price structure
  sustainable. Generative AI adds a provenance problem and a *signaling* one: an AI
  policy is read as a stance on human contributors, and the announcement moves
  participation on its own [@borwankar2026chatgpt; @huang2026stances].
- **Measuring** a network effect requires confronting the simultaneity of the two
  sides; an estimate that ignores it is an upper bound, recoverable only with a valid
  side-specific instrument, a structural model, or a credible natural experiment.

## Further Reading

The two-sided market was formalized by Rochet and Tirole, with complementary pricing
treatments by Parker–Van Alstyne and by Armstrong [@rochet2003platformcompetiti; @rochet2006sidedmarkets; @parker2005sidednetwork; @armstrong2006competitisided]; the network-effects
foundations are due to @katz1985network and the Katz–Shapiro work on standards [@katz1986technologadoption; @katz1994systemscompetiti]. The
platform-competition and business-model literature is surveyed in
@casadesus2015competing, and the economics of digital platforms more broadly in
@goldfarb2022 and @tucker2019digital. Multiproduct and bundling considerations
relevant to envelopment appear in @armstrong2018multiproduct.
The branding, advertising, and marketing-finance machinery this chapter draws on is
developed in @sec-branding, @sec-advertising, and @sec-marketing-finance.
