# Audio, Voice, and Speech {#sec-audio-voice}

Marketing has a soundtrack. A call-center recording, a podcast ad read, a
voice-assistant exchange, a thirty-second spot scored to a swelling string line,
the timbre of a spokesperson's voice: each is an acoustic artifact that carries
information about what a person said and, just as importantly, how they said it.
For most of the discipline's history that information was inaccessible at scale.
Audio could be played back and judged by a human coder, but it could not be turned
into the kind of numbers that enter a demand system, a choice model, or a
regression. What has changed is that a waveform can now be transcribed,
characterized, and embedded automatically across millions of recordings, so the
acoustic layer of marketing becomes data in the same operational sense that text
(@sec-text-as-data) and images (@sec-image-processing) already are.

The intellectual move is the one this part has made repeatedly. An unstructured
artifact, here a one-dimensional pressure signal sampled in time, is mapped into a
feature vector that a marketing theory can speak about. As with text and images,
the representation is lossy and learned, and the central econometric caution
carries over without modification: an acoustic feature is a generated regressor
whose measurement error can correlate with the outcome it is meant to explain, so
a naive regression confounds measurement with effect. Audio adds one distinctive
opportunity, which organizes much of this chapter. A spoken utterance has two
separable channels: the lexical channel (the words, recoverable by transcription
and then handed to the entire text-as-data pipeline) and the paralinguistic
channel (pitch, loudness, rhythm, voice quality, the acoustic carrier of emotion
and emphasis). The distinction between *what is said* and *how it is said* is the
payoff of treating audio as its own modality rather than as a noisy path to a
transcript.

The chapter proceeds from signal to construct to application. It first surveys
where audio enters marketing. It then fixes what a digital sound is and develops
the classic acoustic features that have powered speech and music analysis for
decades, with a runnable demonstration that synthesizes a voice-like waveform and
extracts a spectrogram, mel-frequency cepstral coefficients, a fundamental
frequency contour, and a spectral centroid from it. With those primitives in hand
it turns to automatic speech recognition and speech analytics, to vocal emotion
and prosody, and to music and sonic branding. It closes with the realities of
production practice and a short look at the frontier.

## Where Audio Enters Marketing {#sec-audio-applications}

Audio reaches marketing through several distinct streams, each with its own data
characteristics and its own measurement question.

The largest and least exploited stream is the *service and sales voice channel*.
Contact centers and outbound sales organizations record enormous volumes of calls,
historically for compliance and quality assurance, and these recordings hold both
a transcript and an acoustic signal. The transcript answers what was discussed:
the reason for the call, the objections raised, whether a resolution or a sale
occurred. The acoustic signal answers how the exchange unfolded: whether the
customer's voice tightened with frustration, whether the agent's pace and warmth
tracked or diverged from the customer's, where in the call the emotional tenor
turned. Vocal-tone analytics on this channel promises to predict churn,
satisfaction, conversion, and agent effectiveness from signals a transcript cannot
see. It is worth stating plainly, as the research underlying this chapter flags,
that a clean top-tier marketing study analyzing raw call-center acoustic features
at scale with a verifiable identification strategy is still scarce; the channel is
a first-class opportunity precisely because it is comparatively young.

A second stream is *voice-assistant and voice-commerce* interaction. Smart
speakers and phone assistants (Alexa, Siri, Google Assistant) turn shopping,
search, and brand contact into spoken dialogue, which raises questions both about
the consumer's relationship with the device and about the persuasive properties of
the assistant's synthesized voice. Schweitzer, Belk, Jordan, and Ortner (2019,
*Journal of Marketing Management*, doi:10.1080/0267257X.2019.1596970) frame the
consumer side: users build relationships with voice-controlled devices that range
from treating the device as a servant to treating it as a friend or even a master,
and that relational frame shapes adoption and reliance. On the persuasion side,
Flavián, Akdim, and Casaló (2023, *Psychology & Marketing*,
doi:10.1002/mar.21765) show experimentally that voice-based recommendations from a
virtual assistant can move consumer behavior more than text-based online reviews,
with credibility and usefulness mediating the effect. Voice commerce makes the
acoustic properties of a machine's voice a marketing-mix variable.

A third stream is *advertising and branded audio*: radio and television ad audio,
podcast host-read and programmatic insertions, and the music and sound design of
spots across media. Here the questions are classic advertising questions (recall,
attitude, persuasion) asked of an acoustic stimulus, and the literature on music
in advertising, reviewed below, is the oldest and most developed thread in the
modality.

A fourth stream is *sonic branding*: the deliberate design of a brand's auditory
identity, from the multi-second sound logo (the Intel chime, the Netflix
"ta-dum," a startup tone) to the chosen voice of a brand's spokes-character or
assistant. Sonic branding treats sound the way visual branding treats a logo or
color palette, as a recognizable, ownable, affect-laden asset.

Across all four streams the same architecture recurs and is worth adopting as a
default. Run automatic speech recognition to obtain a transcript and feed it to the
text pipeline; separately extract acoustic-prosodic features as their own signal;
and model the two channels jointly, never collapsing how-it-was-said into
what-was-said. The remainder of the chapter equips each half of that architecture.

## What a Sound Is, and Its Classic Features {#sec-audio-features}

Formally, a digital audio signal is a function of time sampled on a uniform grid.
A continuous sound pressure wave $x(t)$ is recorded as a sequence

$$
x[n] = x(n / f_s), \qquad n = 0, 1, \dots, N-1,
$$ {#eq-audio-sampling}

where $f_s$ is the sampling rate in samples per second (hertz) and the number of
samples $N = f_s \cdot T$ for a recording of duration $T$. The
Nyquist-Shannon sampling theorem requires $f_s$ to be at least twice the highest
frequency of interest, which is why telephone speech is adequately captured at 8
kHz (human speech intelligibility lives mostly below 4 kHz) while music is sampled
at 44.1 kHz to preserve content up to roughly 22 kHz. Each sample $x[n]$ is a
single amplitude value, the analogue of a pixel, and like a pixel it is almost
meaningless in isolation. What carries information is structure across many
samples: periodicity, spectral content, and how both evolve in time.

Because the informative structure is spectral and time-varying, the raw waveform
is almost never modeled directly. The foundational transform is the
short-time Fourier transform, which slides a short analysis window across the
signal and computes, in each window, the distribution of energy across frequency.
The squared magnitude of this transform, displayed as time on the horizontal axis,
frequency on the vertical, and energy as intensity, is the *spectrogram*, the
single most important visualization in audio analysis. A spectrogram of speech
reveals the horizontal bands of energy (formants) that distinguish vowels, the
broadband bursts of consonants, and the vertical striations of the glottal pulse
train whose spacing encodes pitch.

From the spectrogram, a small set of classic features has organized speech and
music analysis for four decades.

The *fundamental frequency* $f_0$ is the rate at which the vocal folds vibrate, the
acoustic correlate of perceived pitch. Its average level, its range, and the shape
of its contour over an utterance carry speaker identity, emphasis, question-versus
statement intonation, and a large share of vocal emotion.

*Mel-frequency cepstral coefficients* (MFCCs) are the workhorse representation of
the spectral envelope, that is, of the overall shape of the spectrum that
distinguishes one vowel or timbre from another independent of pitch. MFCCs were
introduced by Davis and Mermelstein (1980, *IEEE Transactions on Acoustics, Speech,
and Signal Processing*, doi:10.1109/TASSP.1980.1163420) and remain ubiquitous. They
are computed by warping the spectrum onto the perceptual *mel* scale (which spaces
frequencies the way the human ear resolves them, finely at low frequencies and
coarsely at high), taking logarithms of the energy in a bank of mel filters, and
applying a discrete cosine transform to decorrelate the result. The first dozen or
so coefficients compactly summarize timbre and are the standard input to speaker,
emotion, and (historically) speech recognition.

*Spectral shape* features summarize the distribution of energy across frequency in
a single number each. The most common is the *spectral centroid*, the
amplitude-weighted mean frequency, which corresponds to perceived brightness: a
bright, sharp voice or a cymbal has a high centroid, a dull or muffled sound a low
one. Related measures include spectral spread, rolloff, and flatness.

*Energy and rhythm* features (the short-time energy or loudness contour, the
speaking rate, the distribution of pause durations) capture intensity and timing,
which carry arousal, emphasis, and conversational dynamics.

*Voice-quality* features (jitter, the cycle-to-cycle variability in $f_0$; shimmer,
the variability in amplitude; and the harmonics-to-noise ratio) quantify the
roughness, breathiness, or strain of a voice, and are sensitive markers of stress
and emotion.

These features are interpretable, cheap, and defensible, the audio analogue of the
hand-engineered color and composition features of @sec-image-processing. They were
designed by acoustic phoneticians with explicit perceptual motivation, and they
remain the right tool when the construct is well understood, the sample is modest,
or interpretability is paramount.

### A Runnable Demonstration: From Waveform to Features {#sec-audio-demo}

The demonstration below synthesizes a short voice-like waveform with known
properties and then extracts the four classic features just described, using the
maintained R packages `tuneR` and `seewave`. The signal is deliberately simple and
fully simulated: a glottal source modeled as a fundamental plus harmonics, with a
slow vibrato so the pitch contour is non-trivial, shaped by an amplitude envelope
and contaminated with a little noise. Because the construction is known, the
extracted features can be checked against ground truth, which is the point of a
simulation. Everything here runs on synthetic data and uses only signal-processing
functions, not deep models.

```{r audio-setup, message=FALSE, warning=FALSE}
# Maintained CRAN packages for audio analysis in R.
library(tuneR)    # Wave objects, MFCCs (melfcc)
library(seewave)  # spectrogram, fundamental frequency, spectral measures

set.seed(52)
```

```{r audio-synthesize, message=FALSE, warning=FALSE}
# ---- Synthesize a voice-like waveform with KNOWN properties ----
fs   <- 16000                      # sampling rate (Hz); 16 kHz is standard for speech
dur  <- 1.2                        # seconds
t    <- seq(0, dur, by = 1 / fs)   # time grid
N    <- length(t)

f0_mean <- 140                     # mean fundamental frequency (Hz), a low-ish voice
vibrato <- 6 * sin(2 * pi * 4 * t) # +/- 6 Hz pitch wobble at 4 Hz (a gentle vibrato)
f0_t    <- f0_mean + vibrato       # instantaneous f0 over time

# Instantaneous phase = 2*pi * integral of f0(t); cumulative sum approximates it.
phase <- 2 * pi * cumsum(f0_t) / fs

# A glottal-like source: fundamental + decaying harmonics (gives a vowel-ish timbre).
harm_amp <- c(1.0, 0.6, 0.35, 0.2, 0.12)
sig <- rep(0, N)
for (h in seq_along(harm_amp)) {
  sig <- sig + harm_amp[h] * sin(h * phase)
}

# Amplitude envelope (fade in/out) and a little additive noise (breathiness).
env <- sin(pi * seq_len(N) / N)^0.5
sig <- env * sig + rnorm(N, 0, 0.03)

# Wrap as a normalized 16-bit mono Wave object (tuneR's core data structure).
wav <- Wave(left = sig, samp.rate = fs, bit = 16)
wav <- normalize(wav, unit = "16")
wav
```

The object `wav` is now an ordinary `Wave`, the same structure `tuneR` would
produce from reading a `.wav` file off disk, so every step that follows applies
identically to real recordings.

```{r audio-spectrogram, message=FALSE, warning=FALSE, fig.cap="Spectrogram of the synthesized voice-like signal. Horizontal bands are the harmonics of the fundamental; their gentle waviness is the simulated vibrato.", fig.width=6, fig.height=4}
# ---- Spectrogram: energy across time (x) and frequency (y) ----
# wl is the analysis-window length in samples; ovlp the percent overlap.
seewave::spectro(wav, f = fs, wl = 512, ovlp = 75,
                 collevels = seq(-60, 0, 5), flim = c(0, 2),
                 main = "Spectrogram (synthetic voice)")
```

The spectrogram shows a stack of horizontal bands, the fundamental near 140 Hz and
its harmonics above it, each rippling slightly with the 4 Hz vibrato that was built
in. This is exactly the harmonic structure a sustained vowel produces.

```{r audio-f0, message=FALSE, warning=FALSE, fig.cap="Estimated fundamental-frequency (pitch) contour. The estimate hovers near the 140 Hz that was synthesized and traces the vibrato.", fig.width=6, fig.height=3.5}
# ---- Fundamental frequency contour ----
# seewave::fund returns a matrix of (time, f0-in-kHz); we convert to Hz.
ff <- seewave::fund(wav, f = fs, wl = 512, ovlp = 75,
                    fmax = 600, plot = FALSE)
f0_hz <- ff[, 2] * 1000                     # kHz -> Hz
f0_hz <- f0_hz[is.finite(f0_hz) & f0_hz > 0]

plot(f0_hz, type = "l", xlab = "Frame", ylab = "f0 (Hz)",
     main = "Estimated fundamental frequency")
abline(h = f0_mean, col = "red", lty = 2)   # the true mean we synthesized

round(c(true_f0_mean = f0_mean,
        est_f0_median = median(f0_hz)), 1)
```

The estimated pitch track sits close to the dashed line at the true mean of 140 Hz,
recovering a property that was put into the signal by construction. In a real
recording this same contour is what carries question intonation, emphasis, and much
of the emotional signal developed below.

```{r audio-mfcc, message=FALSE, warning=FALSE}
# ---- Mel-frequency cepstral coefficients (timbre) ----
# tuneR::melfcc returns a matrix: one row per analysis frame, one column per coefficient.
mf <- melfcc(wav, sr = fs, numcep = 13,
             wintime = 0.025, hoptime = 0.010)  # 25 ms windows, 10 ms hop
cat("MFCC matrix dimensions (frames x coefficients):",
    paste(dim(mf), collapse = " x "), "\n")

# Per-utterance summary: the mean of each coefficient across frames.
# Aggregating frame-level features to a fixed-length vector is the standard way
# to turn a variable-length recording into one row of a design matrix.
mfcc_means <- round(colMeans(mf), 3)
names(mfcc_means) <- paste0("c", 0:12)
mfcc_means
```

The MFCC matrix has one row per short analysis frame and thirteen columns. A
recording of any length therefore yields a variable number of frames, and the
last lines show the standard device for handling that: summarize the frames (here
by their mean) into a single fixed-length vector that can become one row of a
regression or classifier design matrix. The first coefficient `c0` reflects
overall log-energy; the rest encode the spectral envelope, that is, the timbre.

```{r audio-centroid, message=FALSE, warning=FALSE}
# ---- Spectral centroid (perceived brightness) ----
# Compute the mean spectrum, then the amplitude-weighted mean frequency.
ms <- seewave::meanspec(wav, f = fs, wl = 512, plot = FALSE)
freq_khz <- ms[, 1]                   # frequency axis in kHz
amp      <- ms[, 2]                   # relative amplitude
centroid_hz <- sum(freq_khz * amp) / sum(amp) * 1000

# seewave::specprop packages the same and related descriptors; 'cent' is the
# centroid in Hz. We report both to show they agree.
sp <- seewave::specprop(ms, f = fs)
round(c(centroid_manual_hz = centroid_hz,
        centroid_specprop_hz = as.numeric(sp$cent)), 1)
```

The spectral centroid, computed both by hand as the amplitude-weighted mean
frequency and via `seewave::specprop`, lands in the low hundreds of hertz,
consistent with a low-pitched voice whose energy is concentrated in the
fundamental and its first few harmonics. A brighter sound (more high-frequency
energy) would push this number up. With four lines each, this demonstration has
turned a one-dimensional waveform into a pitch contour, a timbre vector, and a
brightness scalar, the same primitives that feed every application that follows.

## Automatic Speech Recognition and Speech Analytics {#sec-audio-asr}

The lexical channel of audio is unlocked by *automatic speech recognition* (ASR),
the task of transcribing speech to text. ASR is the bridge that lets the entire
text-as-data apparatus of @sec-text-as-data, that is, topic models, sentiment and
stance classifiers, embeddings, and large-language-model extraction, operate on
spoken marketing data. The history of ASR runs from the MFCC-plus-hidden-Markov
systems of the 1980s and 1990s, through hybrid deep-neural-network systems in the
2010s, to today's end-to-end sequence models trained on very large corpora. The
contemporary reference point is OpenAI's Whisper, an encoder-decoder transformer
trained on a large multilingual, multitask corpus that produces robust transcripts
across accents, noise, and domains and has become a common default for research
pipelines. Wav2vec 2.0 represents the parallel self-supervised line: it learns
speech representations from unlabeled audio that can then be fine-tuned for
recognition with relatively little labeled data. These deep-ASR systems are best
treated here as conceptual building blocks rather than something to run inline:
they require substantial compute and large pretrained weights, and unlike the
signal-processing demonstration above they are not lightweight enough to execute in
a book's render pipeline. The practical posture is to call them as a service or a
preprocessing step, then verify their output, because transcription error is a
measurement error that propagates into every downstream text feature.

Several capabilities sit alongside transcription and together constitute *speech
analytics*. *Speaker diarization* segments a recording into "who spoke when,"
essential for separating the agent from the customer in a service call or the host
from the guest in a podcast. *Speaker identification and verification* match a
voice to a known identity, underpinning voice biometrics and fraud detection.
*Keyword and intent spotting* flags compliance-relevant phrases, product mentions,
or competitor references without a full transcript. *Language and accent
identification* route and segment multilingual corpora. In a production speech-
analytics stack these run before or alongside ASR, and their errors, like ASR's,
are generated-feature errors that the downstream analysis must respect.

The strategic value of the lexical channel in marketing is that it converts the
spoken word, previously locked in un-searchable audio, into the same
representations that have made review text and social-media text so productive. A
corpus of sales-call transcripts can be mined for the objections that precede a
lost deal, the language that precedes a renewal, or the product features customers
actually ask about, exactly as a corpus of reviews is mined for product
attributes. But transcription alone discards the paralinguistic channel, and that
channel is where audio earns its place as a distinct modality.

## Vocal Emotion and Prosody: How It Is Said {#sec-audio-prosody}

The acoustic-prosodic channel carries information that the words do not. The same
sentence, "that's just great," can be sincere or sarcastic (@sec-sarcasm), and the
difference lives almost entirely in pitch, timing, and voice quality rather than in
the lexical content. *Prosody*, the melody and rhythm of speech, comprises the
$f_0$ contour, the loudness contour, speaking rate and pausing, and voice quality.
These are the features the demonstration above extracted, and they are the
substrate of vocal emotion.

The scientific anchor is Scherer (2003, *Speech Communication*,
doi:10.1016/S0167-6393(02)00084-5), whose review of vocal emotion communication
lays out both the production side (how emotions systematically modulate
acoustic parameters: anger raises and broadens pitch and energy, sadness lowers and
narrows them, and so on) and the recognition side (how listeners decode those
parameters). This is the theory that licenses treating acoustic features as
measures of an emotional construct rather than as arbitrary numbers. On the
engineering side, *speech emotion recognition* (SER) builds classifiers that map
acoustic features to emotion labels. El Ayadi, Kamel, and Karray (2011,
*Pattern Recognition*, doi:10.1016/j.patcog.2010.09.020) survey the classic
pipeline (hand-engineered prosodic, spectral, and voice-quality features feeding a
classifier) and the recurring difficulties: emotion is continuous and
context-dependent, labeled datasets are small and often acted rather than
spontaneous, and accuracy degrades sharply when a model trained on one corpus or
language meets another. Modern SER replaces or augments the hand-engineered
features with learned audio embeddings of the wav2vec family, but the validity
problems the survey names (label quality, acted-versus-spontaneous mismatch,
cross-corpus generalization) are exactly the construct-validity concerns
(@sec-construct-vs-variable) that this book insists on, and they do not disappear
when the feature extractor becomes a neural network.

Prosody also carries persuasion, not only emotion. Zoghaib (2019, *Recherche et
Applications en Marketing*, doi:10.1177/2051570719828687) manipulates a speaker's
voice along acoustic dimensions and finds that lower-pitched, smoother (less
rough), and duller (less bright) voices are the more persuasive, with effects that
interact with speaker gender. This is a direct demonstration that the very features
the demonstration computed, $f_0$ level and spectral centroid and voice-quality
roughness, are not bookkeeping quantities but marketing-relevant levers that shape
consumer response to the same words. The managerial reading is that a brand
choosing a spokesperson, a voice actor for an ad, or a timbre for a synthesized
assistant is choosing a persuasion parameter, and that choice can now be measured
acoustically rather than left to intuition.

The discipline's standing caution applies with force here. Acoustic-emotion
features are generated regressors, and the model that generates them was trained on
data whose emotional labels, speaker demographics, and recording conditions may
differ systematically from the marketing setting of interest. A vocal-frustration
score that is more accurate for some accents than others, or for studio audio than
for telephone audio, injects a bias that can correlate with the outcome. The
remedy is the one used throughout this part: validate the generated feature against
human judgment on a held-out sample from the target domain before trusting it in a
downstream model.

## Music and Sonic Branding {#sec-audio-music}

Music is the oldest and most developed thread in audio-as-marketing, and the
research is unusually clear that music in advertising is not decoration but a
processing variable. Hecker (1984, *Psychology & Marketing*,
doi:10.1002/mar.4220010303) gives the early statement of music's role in
advertising effect. The empirical core comes from a sequence of careful studies.
Milliman (1982, *Journal of Marketing*, doi:10.1177/002224298204600313) shows in a
field setting that the tempo of background music changes the pace at which
supermarket shoppers move and, with it, sales volume: slow music slows shoppers and
raises spending. North, Hargreaves, and McKendrick (1999, *Journal of Applied
Psychology*, doi:10.1037/0021-9010.84.2.271) demonstrate that the *style* of
in-store music biases product choice, with French music lifting French wine sales
and German music lifting German wine sales, an effect operating largely outside
shoppers' awareness. For advertising specifically, Kellaris and Cox (1989,
*Journal of Consumer Research*, doi:10.1086/209199) reassess earlier claims about
background music and persuasion, and Kellaris, Cox, and Cox (1993, *Journal of
Marketing*, doi:10.1177/002224299305700409) provide the contingency account that
organizes the field: music helps message reception when its attention-gaining
properties and its congruency with the message work together, and can hurt when
they conflict. The lesson is that music's effect is moderated by fit, not uniformly
positive, which is exactly why measuring musical content matters.

Measuring that content is the province of *music information retrieval* (MIR), which
extracts tempo and beat, key and mode (major versus minor, a strong correlate of
happy versus sad), timbre via the same MFCCs used for speech, and learned
embeddings of valence and arousal. These features let a researcher characterize the
music in thousands of ads or podcast segments and relate it to recall, attitude,
and sales at a scale the classic studies could not reach. The bridge from the
classic effects literature to MIR-based measurement at scale is one of the more
tractable open opportunities in the modality.

Sonic branding extends music's logic from the spot to the brand. Zoghaib,
Luffarelli, and Feiereisen (2023, *Psychology & Marketing*,
doi:10.1002/mar.21875) show that structural properties of a brand's music, an
irregular melodic contour and an unstable tonality, raise perceived brand
innovativeness and brand evaluations, with processing difficulty as the mechanism.
This connects sonic branding to the same aesthetic-complexity logic that governs
visual branding in @sec-image-processing: a moderate, productive difficulty in
processing a stimulus can enhance rather than diminish response. The sound logo, the
brand voice, and the scoring of a brand's content are, on this evidence, designable
assets whose acoustic properties carry measurable equity.

## Industry and Production Practice {#sec-audio-practice}

Bringing audio analysis into a working marketing organization confronts a set of
realities the academic framing can understate.

The first is *data access and consent*. Call recordings, voice-assistant logs, and
any audio containing identifiable voices are sensitive personal data. The human
voice is a biometric identifier, and many jurisdictions regulate it specifically;
two-party-consent recording laws, the need for a lawful basis to process voice
data, and rules on automated decision-making all constrain what can be collected
and modeled. The privacy and governance considerations of @sec-privacy are not an
afterthought for audio but a precondition for it, and they shape the dataset before
any feature is extracted.

The second is *audio quality and channel effects*. Telephone audio is band-limited
to roughly 300 to 3400 Hz and compressed by lossy codecs; far-field smart-speaker
audio carries room reverberation and background noise; podcast audio is studio-clean
but heavily processed. These channel differences alter the very features a model
relies on, so a pitch or emotion model trained on one channel can fail on another.
Robust practice fixes the sampling rate and channel, applies consistent
preprocessing (resampling, normalization, optional noise reduction), and validates
within the channel of deployment.

The third is the *pipeline architecture* that the chapter has advocated throughout,
now stated as engineering. A production system typically runs voice-activity
detection to find speech, diarization to separate speakers, ASR to transcribe, and
a parallel acoustic-feature extractor for prosody and emotion, then fuses the
lexical and paralinguistic streams for the downstream task. Each stage is a model
with its own error rate, and those errors compound, so monitoring and human spot-
checking at each stage are not optional.

The fourth is *build-versus-buy*. A mature ecosystem of commercial speech-to-text,
diarization, and emotion APIs (from the major cloud providers and specialized
vendors) makes it unnecessary to train ASR from scratch, while open-source models
(Whisper, wav2vec 2.0) and toolkits (`librosa` and `pyAudioAnalysis` in Python,
`openSMILE` for standardized acoustic feature sets, and `tuneR` and `seewave` in R
for the signal-processing primitives demonstrated above) make a transparent,
auditable in-house pipeline feasible. The choice turns on data sensitivity (a
regulated voice corpus may not leave the premises), the need for auditability in a
research setting, and cost at scale. For the construct-validity reasons this book
stresses, a research pipeline benefits from the transparency of open tooling even
when a commercial API would be more accurate out of the box.

## Frontier and Expansion {#sec-audio-frontier}

Audio is the youngest of the well-developed marketing data modalities, and several
frontiers are visible. The first is *self-supervised audio foundation models*. Just
as wav2vec 2.0 learns speech representations from unlabeled audio and Whisper
learns robust transcription from weak supervision at scale, general-purpose audio
embeddings increasingly provide a single representation that supports emotion,
speaker, and content tasks with light fine-tuning, lowering the barrier to
acoustic measurement in marketing much as pretrained image and text models did for
those modalities. The second is *multimodal fusion*, the natural endpoint of this
part: a podcast ad, a TikTok video (@sec-text-as-data and the image and video
threads), or a live-commerce stream is simultaneously audio, speech, on-screen
text, and image, and the acoustic features developed here are one stream to be
fused with the others rather than analyzed alone. The third is *generative and
synthetic voice*: text-to-speech has become good enough that brands synthesize
spokesperson voices and personalize audio at scale, which turns voice from a
measured variable into a designed one and raises fresh questions of authenticity,
disclosure, and consumer trust. The fourth is the still-open *call-analytics
opportunity* flagged at the start: the service and sales voice channel remains
comparatively under-studied in top-tier marketing research, and a credible
identification strategy applied to acoustic call features at scale is among the
clearer contributions the modality invites.

The synthesizing survey of machine learning and AI in marketing by Ma and Sun
(2020, *International Journal of Research in Marketing*,
doi:10.1016/j.ijresmar.2020.04.005) situates audio within the broader move to
connect computational representations of unstructured data to human marketing
insight, and it is the right place to read this chapter's modality back into the
whole. The consistent thread, from the MFCCs of 1980 to the foundation models of
today, is the one this part repeats for every modality: a sound becomes a learned,
lossy feature vector; that vector is a generated regressor; and the marketing
payoff comes from separating what is said from how it is said, then defending both
measurements before trusting either.
