# Video {#sec-video}

Video is the densest marketing artifact a consumer encounters and the hardest to
measure. A single fifteen-second TikTok ad is, simultaneously, a stream of images
(the visual frames), a stream of sound (speech, music, ambient audio), a stream of
language (the spoken script, the on-screen captions, the post caption and hashtags),
and a stream of *motion* (camera movement, cuts, the pace at which all of the above
changes). Every modality treated separately in the preceding chapters reappears here
at once, and a new dimension, time, organizes them. The marketing payoff of video is
precisely this richness: pacing, the moment a brand first appears, the emotional arc
from hook to call-to-action, and the synchrony of music with cuts are creative levers
that no still image or transcript can capture. The measurement burden is the same
richness seen from the other side: a thirty-frame-per-second clip is a high-dimensional
object whose informative structure lives in the *relationships* across frames and
across modalities, not in any frame alone.

This chapter treats video as an intrinsically multimodal, temporal data type and builds
the analyst's representation from the frame outward. It opens with the marketing
applications that make video a first-class modality (television and streaming ads,
short-form vertical video, livestream commerce, user-generated video). It then develops
the **frame-based representation**, the honest foundation on which all video analysis
rests, and demonstrates it with runnable code that samples a short synthetic clip,
extracts per-frame visual features, and computes temporal-difference signals that flag
shot changes. From there it layers the methods that make video more than a bag of
frames: scene and shot segmentation, action and motion recognition, and the temporal
architectures (3D convolutional networks, two-stream and SlowFast designs, video
transformers) that model dynamics directly. It then turns to **multimodal fusion**, the
joint modeling of frame, audio, and text streams that is the empirical heart of modern
video-as-data marketing research, and to **engagement and creative analytics**, where
extracted features predict watch time, completion, and conversion. It closes with the
industrial reality of how platforms and brands analyze video at scale, and with the
frontier of video-native foundation models. Throughout, the generated-regressor caution
that unifies @sec-text-as-data and @sec-image-processing applies with compounded force:
every video feature is a model output, and a video pipeline stacks several such models
in series, so the correlated-error problem accumulates across the stack.

## Why Video Is a Marketing Data Type {#video-applications}

Video earns a dedicated treatment because the marketing economy now runs on it, and
because the questions managers ask of video cannot be answered by analyzing its frames
or its transcript in isolation.

**Television and streaming advertising.** The oldest video-marketing question, what makes
a commercial work, has been transformed by the migration of the thirty-second spot from
broadcast to connected television and ad-supported streaming. The creative object is
unchanged in form but newly measurable at scale: streaming platforms log exposure,
completion, and downstream behavior at the household level, so the content of the ad
(its scenes, its pacing, when the brand appears, the emotional trajectory) can be linked
to response in a way that broadcast panels never permitted. The analyst's task is to turn
the creative into features and relate those features to lift.

**Short-form vertical video.** TikTok, Instagram Reels, and YouTube Shorts have made the
nine-to-sixty-second vertical clip the dominant unit of attention for large consumer
segments. Short-form video compresses the entire persuasive arc, hook, demonstration,
and call-to-action, into seconds, which puts a premium on pacing and on the first frames
(the "hook"). Influencer-produced short-form advertising is now a major channel, and the
flagship marketing study of it, @yang2024engagement on TikTok, shows that fusing the
video, audio, and text streams of influencer ads predicts product sales, and that
*engagement* signals mediate the path from creative to revenue. Short-form video is also
where algorithmic distribution is most aggressive: a clip's reach is determined by a
recommender that itself consumes the video's features, so creative analysis and
distribution analysis are entangled.

**Livestream and live commerce.** Live shopping, a host demonstrating and selling products
in real time to a chat-interacting audience, fuses video, continuous speech, and a
real-time text stream (the chat) with an immediate transactional outcome (purchases during
the stream). It is the most explicitly multimodal commercial video format, and it has begun
to attract rigorous measurement. @wen2026livestream analyze livestreaming e-commerce with
multimodal machine learning, decomposing a salesperson's on-screen effectiveness into
visual, vocal, and verbal components and using explainable-AI methods to attribute sales to
them; @xu2024multimodal and @xu2024memf build multimodal fusion frameworks that combine the
host's audio, visual, and textual signals (and, in the latter, multiple on-screen entities)
to predict product sales during a stream. Live commerce is distinctive because the outcome is contemporaneous with the content,
which sharpens both the opportunity (within-stream dynamics map directly onto within-stream
sales) and the identification challenge (the host adapts in real time to the chat).

**User-generated video: reviews, unboxings, hauls.** Consumers now narrate their product
experiences on camera. A video review carries information that a text review cannot: the
product shown in use, the reviewer's facial affect, vocal tone, the demonstration of fit or
scale. Unboxing and haul videos are a genre unto themselves with measurable engagement
consequences. For the firm, user-generated video is both a listening channel (what do
consumers actually do with the product, and how do they feel) and a demand signal (review
video volume and sentiment predict sales), extending the online word-of-mouth tradition of
review-to-sales research from text to the screen [@chevalier2006effect].

The common thread is that the marketing question is almost always about *content and its
dynamics*: which creative choices, appearing *when* in the clip, drive attention, engagement,
and conversion. Answering it requires turning the video into features that respect both its
multimodality and its temporal structure. The rest of the chapter builds that representation.

## The Frame-Based Representation {#video-frames}

A video is, at bottom, an ordered sequence of still images sampled at a frame rate, with
a synchronized audio track. The honest foundation of all video analysis is therefore the
*frame*: whatever sophistication follows (motion models, multimodal transformers) is built
on top of a decision about which frames to look at and what to extract from each. This
section develops that foundation and demonstrates it with code that genuinely runs.

### Frame sampling

Processing every frame of a clip is usually wasteful and sometimes infeasible: a one-minute
clip at thirty frames per second is eighteen hundred images, most of them near-duplicates of
their neighbors. The practical pipeline samples frames, and the sampling policy is a real
modeling choice with consequences:

- **Uniform sampling** takes one frame every $k$ frames (or every $t$ seconds). It is simple,
  reproducible, and unbiased with respect to content, but it can miss fast events that fall
  between samples and oversample static stretches.
- **Keyframe sampling** takes one representative frame per shot, detected by the shot-segmentation
  methods below. It aligns the sample with the video's own editorial structure and is efficient,
  but it presupposes a shot detector and discards within-shot motion.
- **Content-adaptive sampling** allocates more frames where the visual signal changes fastest,
  trading reproducibility for information density.

In tools, frame extraction is typically delegated to `ffmpeg` or `OpenCV`, which decode the
container and emit images at the chosen cadence. A marketing-specific recipe for the whole
frame-to-features workflow is given by @schwenzow2021videos, whose "understanding videos at
scale" pipeline shows how to sample frames, extract interpretable visual, audio, and textual
features, and aggregate them for business research—an accessible blueprint for the substrate this
section builds. For the conceptual and pedagogical purposes of
this chapter, we work with the maintained `magick` R package, which wraps ImageMagick and gives
us honest, runnable image manipulation. We will not pretend to run a deep video model locally;
instead we demonstrate the frame-level substrate, simulating a short clip, extracting interpretable
per-frame features, and computing the temporal signals that drive shot detection, all of which the
later sections build upon.

### Per-frame features

Once frames are in hand, each is an image and the entire apparatus of @sec-image-processing applies:
interpretable hand-engineered features (color, brightness, saturation, edge density / complexity)
when the construct is simple and interpretability matters, and learned CNN or vision-transformer
embeddings when meaning is required. The video-specific move is to treat each per-frame feature as a
*time series*: a clip becomes a matrix whose rows are frames and whose columns are features, and the
temporal structure of that matrix (how features rise, fall, and jump) carries the pacing and
editing information that distinguishes video from a photo album.

### A runnable frame-level demonstration

The code below is genuinely executed when the book is rendered. It uses `magick` to *synthesize* a
short sequence of frames, mimicking a clip with two scene changes (a calm opening, a vivid product
shot, a busy closing), then extracts per-frame brightness, colorfulness, and edge-density features,
and finally computes a frame-to-frame difference signal whose spikes localize the cuts. This is the
foundation, made concrete, on which the conceptual deep pipelines later in the chapter rest.
@fig-vid-frames diagrams that foundation, sampling frames, turning each into a feature vector, and
differencing consecutive frames to flag cuts, before the code below carries it out.

```{mermaid}
%%| label: fig-vid-frames
%%| fig-cap: "The frame-level foundation of video analysis. Frames are sampled from the clip, each becomes a vector of per-frame features, the features are read as a time series, and the frame-to-frame difference is thresholded to flag shot changes."
flowchart LR
  VID["Video clip"] --> SMP["Frame sampling<br/>(uniform, keyframe, adaptive)"]
  SMP --> FR["Sampled frames"]
  FR --> PF["Per-frame features<br/>(brightness, colorfulness, edge density)"]
  PF --> TSR["Feature time series<br/>(frames x features)"]
  TSR --> DF["Frame-to-frame<br/>difference"]
  DF --> SC["Threshold flags<br/>shot changes"]
```

```{r video-magick-frames, message=FALSE, warning=FALSE}
# Frame-level foundation with the maintained `magick` package.
# We simulate a short clip as a sequence of synthetic frames, extract
# per-frame visual features, and derive a temporal shot-change signal.
# This is a genuinely runnable demonstration of the substrate that
# real (deep) video pipelines sit on top of; it is NOT a deep model.

library(magick)
set.seed(53)

# A helper that draws one synthetic frame as a solid base color with
# additive noise (a stand-in for "scene content"). Returns a magick image.
make_frame <- function(r, g, b, noise = 0.04, size = 96) {
  base <- array(0, dim = c(size, size, 3))
  base[, , 1] <- r; base[, , 2] <- g; base[, , 3] <- b
  base <- base + array(rnorm(size * size * 3, 0, noise), dim = dim(base))
  base <- pmin(pmax(base, 0), 1)
  image_read(base)              # magick reads an [H, W, C] array in [0,1]
}

# Build a 9-frame "clip" with three scenes of three frames each:
#   frames 1-3  : calm gray opening (low saturation, low complexity)
#   frames 4-6  : vivid warm product shot (high saturation/colorfulness)
#   frames 7-9  : busy high-noise closing (high edge density)
scene_specs <- list(
  c(0.55, 0.55, 0.55, 0.02),   # calm
  c(0.55, 0.55, 0.55, 0.02),
  c(0.55, 0.55, 0.55, 0.02),
  c(0.85, 0.35, 0.15, 0.04),   # warm
  c(0.85, 0.35, 0.15, 0.04),
  c(0.85, 0.35, 0.15, 0.04),
  c(0.50, 0.50, 0.50, 0.22),   # busy
  c(0.50, 0.50, 0.50, 0.22),
  c(0.50, 0.50, 0.50, 0.22)
)
frames <- lapply(scene_specs, function(s) make_frame(s[1], s[2], s[3], s[4]))
clip   <- image_join(frames)   # an image sequence == our simulated video
length(clip)                   # number of frames in the clip
```

With the clip in hand, we extract interpretable per-frame features. We pull each frame back into a
numeric array via `magick`, then compute brightness, a colorfulness proxy, and an edge-density
complexity proxy, exactly the constructs introduced for still images, now indexed by frame.

```{r video-perframe-features, message=FALSE, warning=FALSE}
# Convert a single magick frame to an [H, W, 3] numeric array in [0,1].
frame_to_array <- function(im) {
  as.numeric(image_data(im, channels = "rgb")) -> v   # raw bytes 0-255
  arr <- array(v, dim = c(3, image_info(im)$width, image_info(im)$height))
  arr <- aperm(arr, c(3, 2, 1))                        # -> [H, W, C]
  arr / 255
}

brightness <- function(im) {
  a <- frame_to_array(im)
  mean(0.299 * a[, , 1] + 0.587 * a[, , 2] + 0.114 * a[, , 3])
}
colorfulness <- function(im) {
  a <- frame_to_array(im)
  rg <- a[, , 1] - a[, , 2]
  yb <- 0.5 * (a[, , 1] + a[, , 2]) - a[, , 3]
  sqrt(sd(rg)^2 + sd(yb)^2) + 0.3 * sqrt(mean(rg)^2 + mean(yb)^2)
}
edge_density <- function(im) {
  a   <- frame_to_array(im)
  lum <- 0.299 * a[, , 1] + 0.587 * a[, , 2] + 0.114 * a[, , 3]
  gx  <- abs(lum[-1, ] - lum[-nrow(lum), ])
  gy  <- abs(lum[, -1] - lum[, -ncol(lum)])
  mean(gx) + mean(gy)
}

feat <- data.frame(
  frame        = seq_len(length(clip)),
  brightness   = vapply(seq_len(length(clip)),
                        function(i) brightness(clip[i]), numeric(1)),
  colorfulness = vapply(seq_len(length(clip)),
                        function(i) colorfulness(clip[i]), numeric(1)),
  edge_density = vapply(seq_len(length(clip)),
                        function(i) edge_density(clip[i]), numeric(1))
)
feat[, -1] <- round(feat[, -1], 3)
knitr::kable(feat,
  caption = "Per-frame interpretable features for the simulated nine-frame clip.")
```

The feature matrix already exposes the editorial structure: colorfulness jumps when the warm product
scene begins (frame 4) and edge density jumps when the busy closing begins (frame 7). To localize
those transitions automatically, we compute a *temporal-difference* signal, the frame-to-frame change
in the feature vector, which is the elementary form of the shot-change detector developed in the next
section.

```{r video-shot-change, message=FALSE, warning=FALSE}
# Frame-to-frame difference in the standardized feature vector.
# A large difference between frame t and t-1 signals a shot change.
F  <- scale(as.matrix(feat[, c("brightness", "colorfulness", "edge_density")]))
d  <- c(NA, sqrt(rowSums((F[-1, ] - F[-nrow(F), ])^2)))   # Euclidean step
diff_signal <- data.frame(
  transition  = paste0(seq_len(length(clip) - 1), "->", 2:length(clip)),
  frame_delta = round(d[-1], 3)
)
# Flag transitions whose change exceeds a simple threshold (mean + 1 SD).
thr <- mean(d[-1]) + sd(d[-1])
diff_signal$shot_change <- diff_signal$frame_delta > thr
knitr::kable(diff_signal,
  caption = "Temporal-difference signal: large frame-to-frame changes flag the two scene cuts (3->4 and 6->7).")
```

The two flagged transitions, 3 to 4 and 6 to 7, are exactly the scene boundaries we built into the
simulated clip. This is the entire logic of classical shot detection in miniature: represent each
frame as a feature vector, difference consecutive frames, and threshold the difference. Everything
that follows, learned per-frame embeddings instead of hand-crafted features, learned temporal models
instead of a fixed threshold, is a sophistication of these three steps, not a departure from them. The
honesty of this demonstration matters: the deep pipelines below are the production reality, but they
are not runnable on a laptop without GPUs and large pretrained weights, so we present them as clearly
labeled conceptual code rather than as something this chapter pretends to execute.

## Temporal and Scene Methods {#video-temporal}

The frame-level demonstration treats the clip as a sequence of independent images differenced over
time. Real video understanding models the temporal dimension directly. This section develops the
methods in rough order of how much temporal structure they exploit.

### Shot and scene segmentation

A **shot** is an unbroken run of frames from a single camera take; a **scene** is a semantically
coherent group of shots. Segmenting a video into shots is the first structural operation in most
pipelines, because shots are the natural unit for keyframe sampling and for measuring *pacing*
(shots per minute, mean shot duration), a creative variable with direct marketing relevance: fast
cutting is a hallmark of high-energy short-form advertising. Classical shot-boundary detection is
exactly the thresholded frame-difference of the demonstration above, typically on color histograms
or edge maps, with refinements to distinguish *hard cuts* (a single large jump) from *gradual
transitions* (fades and dissolves, which spread the change over several frames and require detecting
a sustained ramp rather than a spike). Modern detectors learn the boundary from data, but the
construct is unchanged.

### Action and motion recognition

Beyond *where* the cuts are lies *what is happening* between them. Action recognition assigns labels
to motion (a person pouring, applying, unboxing, dancing), and it is the canonical task that forced
the field to model time rather than frames. The reference benchmark is **Kinetics**, introduced
alongside the inflated-3D-convolution (I3D) model by @carreira2017kinetics, which established that
large-scale video pretraining transfers to downstream action tasks much as ImageNet pretraining
transfers for images. For marketing, action recognition is what lets a pipeline detect *product use*
(the demonstration moment in a review or a livestream) as opposed to mere product *presence*, a
distinction that still-frame analysis cannot make.

### Temporal architectures

Three families of architecture model video dynamics, and they trade off temporal fidelity against
cost:

- **3D convolutional networks** extend the 2D convolution of an image CNN to a third, temporal axis,
  so a single learned filter spans several frames and captures short-range motion. I3D
  is the canonical instance, inflating a pretrained 2D image network into 3D [@carreira2017kinetics].
- **Two-stream and multi-rate designs** process appearance and motion on separate pathways. The
  **SlowFast** network of Feichtenhofer and colleagues runs a slow, high-resolution pathway to capture
  semantics (what is in the scene) alongside a fast, low-resolution pathway to capture rapid motion
  (how it moves), then fuses them, an architecture that maps cleanly onto the marketing intuition that
  *content* and *pacing* are distinct creative dimensions [@feichtenhofer2019slowfast].
- **Video transformers** apply self-attention across both space and time, treating a clip as a sequence
  of spatiotemporal patches. They are the video analogue of the vision transformer surveyed by
  @khan2022transformers, and they currently dominate large-scale video benchmarks at the cost of large
  data and compute requirements.

The following conceptual block sketches a temporal pipeline. It is deliberately written as *labeled
pseudocode in R syntax*, not as runnable analysis: a faithful version requires `ffmpeg`, a deep-learning
backend, and pretrained weights that this chapter does not assume are installed. The point is to show
the shape of the production pipeline honestly, not to hand-wave it.

```{r video-temporal-conceptual, eval=FALSE}
# CONCEPTUAL ONLY (not evaluated): a temporal video pipeline.
# Requires ffmpeg/OpenCV for decoding and a deep-learning backend
# (e.g. torch/keras) with pretrained video weights. Shown for shape.

# 1. Decode and sample frames at a fixed cadence (e.g. 8 fps).
frames <- decode_video("ad_clip.mp4", fps = 8)        # ~ tensor [T, H, W, 3]

# 2. Per-frame appearance embeddings from an image backbone (ResNet/ViT).
frame_emb <- image_backbone(frames)                   # [T, d_frame]

# 3. Optical flow between consecutive frames -> a motion stream.
flow      <- optical_flow(frames)                     # [T-1, H, W, 2]
motion_emb <- motion_backbone(flow)                   # [T-1, d_motion]

# 4. A temporal model over the per-frame embeddings:
#    a 3D-CNN / SlowFast / video-transformer that returns a clip embedding.
clip_emb  <- temporal_model(frame_emb, motion_emb)    # [d_clip]

# 5. Shot boundaries from a learned (or thresholded) frame-difference signal,
#    yielding pacing features (shots per minute, mean shot length).
shots     <- detect_shots(frame_emb)                  # list of [start, end]
pacing    <- shot_pacing_features(shots)              # interpretable controls
```

The deliberate split between the runnable `magick` foundation and this conceptual temporal block is the
methodological honesty the chapter insists on: the analyst should understand the frame-level signal
concretely and treat the deep temporal model as a powerful but heavyweight black box whose outputs are,
once again, generated features.

## Multimodal Video Models {#video-multimodal}

The defining empirical fact about marketing video is that meaning is distributed across modalities and
their *synchrony*. A product named in the script (text), shown on screen (frame), and underscored by a
music swell (audio) at the same instant is more persuasive than any one cue alone, and the persuasion
lives in the alignment. Single-modality analysis systematically misses this, which is why the leading
video-as-data marketing studies fuse modalities.

### The three streams

A marketing video decomposes into three analyzable streams, each handled by the apparatus of its own
chapter:

- **Visual (frames).** Sampled frames feed image models (@sec-image-processing) to yield per-frame and
  clip-level visual embeddings plus interpretable creative features (color, faces, brand/logo presence,
  shot pacing).
- **Audio.** The soundtrack splits, as in @sec-audio-voice, into *what is said* (automatic speech recognition
  producing a transcript) and *how it is said plus what is heard* (prosody, music tempo and valence,
  ambient sound), captured by acoustic embeddings.
- **Text.** Three textual sources arise: the ASR transcript of the speech, the on-screen text and
  captions (read by optical character recognition), and the platform metadata (post caption, hashtags,
  title), all fed to the text apparatus of @sec-text-as-data.

### Fusion strategies

Combining the streams is the modeling crux, and the choice of *when* to combine carries assumptions:

- **Late (decision-level) fusion** trains a separate model per modality and combines their predictions
  (averaging, stacking). It is robust and modular, tolerates missing modalities, and is the safe default,
  but it cannot represent cross-modal *interactions* (the music-swell-at-the-product-reveal effect).
- **Early (feature-level) fusion** concatenates per-modality embeddings into one vector before modeling.
  It can capture interactions but is sensitive to scaling and to one modality's dimensionality swamping
  another.
- **Hybrid and cross-attention fusion** lets modalities attend to one another, learning, for instance,
  which spoken words align with which on-screen moments. This is the architecture of modern multimodal
  transformers and the natural home for modeling *synchrony*, at a substantial data and compute cost.

@fig-vid-fusion sets out the three streams and the point at which each fusion strategy combines them.

```{mermaid}
%%| label: fig-vid-fusion
%%| fig-cap: "Multimodal video as three streams and a fusion choice. Visual, audio, and text streams each yield per-modality features, and late, early, or cross-attention fusion combines them before the outcome model of engagement or sales."
flowchart LR
  VDO["Marketing video"] --> VS["Visual stream<br/>(frames)"]
  VDO --> AS["Audio stream<br/>(speech, prosody, music)"]
  VDO --> TX["Text stream<br/>(transcript, captions, metadata)"]
  VS --> VE["Visual features"]
  AS --> AE["Acoustic features"]
  TX --> TE["Text features"]
  VE --> FUS["Fusion<br/>(late, early, or cross-attention)"]
  AE --> FUS
  TE --> FUS
  FUS --> OUT["Outcome model<br/>(engagement, sales)"]
```

The flagship demonstration in marketing is @yang2024engagement, who extract video, audio, and text
features from short-form influencer ads on TikTok and fuse them to predict product sales, finding that
the fused representation, mediated through engagement, carries genuine predictive content for revenue.
@wen2026livestream extend the logic to livestream commerce, decomposing a host's on-screen
effectiveness into visual, vocal, and verbal components with multimodal machine learning and attributing
sales to each via explainable-AI methods. Both studies exemplify the central design principle: model the
streams jointly, and let the data reveal which modality, and which cross-modal alignment, drives the
outcome.

The conceptual block below sketches a fusion pipeline, again as labeled pseudocode rather than runnable
code, to make the architecture explicit without pretending to execute three deep backbones locally.

```{r video-fusion-conceptual, eval=FALSE}
# CONCEPTUAL ONLY (not evaluated): multimodal fusion for a marketing video.

frames <- decode_video("ad.mp4", fps = 8)
audio  <- extract_audio("ad.mp4")

# --- Per-modality embeddings ---
visual_emb <- video_backbone(frames)                 # frames -> [d_v]
transcript <- asr(audio)                              # speech -> text
prosody    <- acoustic_features(audio)               # pitch/energy/tempo -> [d_a]
captions   <- ocr(frames)                             # on-screen text
meta_text  <- c(transcript, captions, "post caption + hashtags")
text_emb   <- text_encoder(meta_text)                # text -> [d_t]

# --- Fusion (choose one) ---
late  <- average(predict(m_v, visual_emb),
                 predict(m_a, prosody),
                 predict(m_t, text_emb))              # decision-level
early <- model(c(visual_emb, prosody, text_emb))     # feature-level
cross <- cross_attention_transformer(                # synchrony-aware
           visual_emb, prosody, text_emb)

# --- Outcome model ---
# engagement / sales ~ fused features (+ interpretable creative controls),
# with the generated-regressor caveat applied to EVERY feature.
```

The econometric warning that runs through the unstructured-data part reaches its peak here. A video
pipeline stacks a frame sampler, an image backbone, an ASR system, an OCR system, an acoustic model, a
text encoder, and a fusion network, each a fitted model with its own error. When the fused features enter
a downstream regression of sales or engagement on creative content, those features are *generated
regressors* and their errors may correlate with the outcome (a model that recognizes "product reveal"
better in successful ads will manufacture a spurious reveal-to-sales association). The remedies are the
familiar ones, validation against human-coded ground truth, holding out an independent sample for the
second stage, and, where possible, experimental variation in the creative, and they are *more* essential
for video because the error sources are more numerous.

## Engagement and Creative Analytics {#video-engagement}

The applied destination of video-as-data is usually a model relating extracted creative features to a
behavioral outcome: watch time, completion rate, like/share/comment engagement, click-through, or
conversion. This section organizes the work into the features that go in and the outcomes that come out.

### Creative-feature extraction

Marketing analysts extract a recurring vocabulary of creative features, some interpretable, some learned:

- **Pacing and structure.** Shots per minute, mean and variance of shot duration, time-to-first-cut, and
  the position of the "hook" (how quickly the opening establishes interest). These come directly from the
  shot-segmentation signal demonstrated above.
- **Visual content.** Color palette and brightness trajectory, presence and screen-time of faces, presence
  and timing of the brand or logo (brand *prominence* and time-to-first-brand-appearance), scene variety.
- **Audio content.** Music presence, tempo, and valence; speech rate and vocal energy; loudness dynamics
  and their synchrony with cuts.
- **Linguistic content.** Topics and sentiment of the transcript and captions, calls-to-action and their
  timing, and the alignment of spoken claims with on-screen demonstration.
- **Temporal arcs.** Rather than collapsing a feature to its clip average, the *trajectory* of arousal,
  emotion, or brand presence over the clip is itself a feature: a rising emotional arc, an early brand
  reveal, a late call-to-action.

@fig-vid-features collects this recurring creative vocabulary into the five families the analyst extracts.

```{mermaid}
%%| label: fig-vid-features
%%| fig-cap: "The recurring vocabulary of video creative features, grouped into five families that feed predictive models of engagement and conversion."
flowchart TB
  CF["Creative features"]
  CF --> PA["Pacing and structure"]
  CF --> VI["Visual content"]
  CF --> AU["Audio content"]
  CF --> LI["Linguistic content"]
  CF --> TA["Temporal arcs"]
  PA --> PA1["Shots per minute, shot duration,<br/>time-to-first-cut, hook position"]
  VI --> VI1["Color and brightness, faces,<br/>brand prominence, scene variety"]
  AU --> AU1["Music tempo and valence,<br/>speech rate, loudness, cut synchrony"]
  LI --> LI1["Transcript and caption topics,<br/>sentiment, calls-to-action"]
  TA --> TA1["Trajectories of arousal,<br/>emotion, brand presence"]
```

### Engagement prediction and its pitfalls

These features feed predictive models of engagement and conversion. The empirical lessons from the
marketing literature are consistent. The affective trajectory of a video ad drives engagement:
@teixeira2012emotionengagement use eye-tracking and automated facial-expression coding to show that
induced emotion (joy, surprise) raises attention and suppresses zapping, and @teixeira2014 establish,
via a web-based facial-tracking field study, *why, when, and how much* to entertain—the timing of
affect within the clip is itself a lever. @tellis2019 decompose what makes online video
content shared, isolating the roles of information, emotion, and brand prominence; brand prominence
in particular trades off against likeability, since too-early or too-heavy branding can depress
engagement even as it aids recall. For short-form specifically, @zhang2025audiovisual link the
audiovisual features of TikTok ads to engagement behaviors, and @leung2022 situate such creative
effects within the broader system of influencer-marketing effectiveness. The throughline is that
*fused* multimodal features outperform any single stream, as in @yang2024engagement. Three pitfalls
deserve emphasis:

1. **Engagement is not sales.** Platform engagement (views, likes) is an intermediate outcome that the
   distribution algorithm partly manufactures; @yang2024engagement treat engagement as a *mediator* on the
   path to sales rather than as the endpoint, which is the correct posture.
2. **Selection on the platform.** Observed videos are those a recommender chose to show, so the sample of
   "video and its engagement" is endogenously selected by an algorithm that consumed the very features
   under study, biasing naive content-to-engagement estimates.
3. **Generated-feature error.** As stressed above, every creative feature is a model output; validation and
   second-stage discipline are prerequisites, not niceties.

## Industry and Production Practice {#video-industry}

How video is analyzed in production differs from the research idealization in ways the analyst should
understand, because the data a firm can obtain is shaped by the platforms that host the video.

**Platform-side analysis at scale.** The large video platforms run video understanding as core
infrastructure, not as analysis after the fact. Uploaded video is automatically transcribed (ASR),
captioned, scanned for objects, scenes, and policy violations, and embedded into the same representation
space the recommender uses to match clips to viewers. The recommender is itself a video-multimodal model:
it consumes frame, audio, and text features along with behavioral signals to predict watch time and
engagement, which means the platform's *distribution* decision and the *creative features* of the video are
produced by one coupled system. For the outside analyst this is the central structural fact: observed
reach and engagement are jointly determined with content by a model the analyst cannot see.

**Brand- and agency-side analysis.** Advertisers and their agencies analyze video creative through three
broad routes.

::: {.callout-tip}
## Replication resources: video feature pipelines

The frame-level demonstration in this chapter runs on the maintained R package `magick`; a fuller
open R/Python workflow for marketing video combines `ffmpeg`/`OpenCV` (decoding), pretrained image
backbones such as ResNet (the official implementation accompanies @he2016resnet at
`github.com/KaimingHe/deep-residual-networks`), Whisper (ASR), and the action/temporal models of
@carreira2017kinetics and @feichtenhofer2019slowfast (whose authors release reference code). For a
marketing-tailored end-to-end recipe see @schwenzow2021videos. Code and data availability for the
empirical marketing studies cited here (e.g., @yang2024engagement, @wen2026livestream) vary—check the
journal's supplementary materials, since several rely on proprietary platform data that cannot be
redistributed.
:::

The first is *managed cloud services*, Google Cloud Video Intelligence, Amazon Rekognition
Video, Microsoft Azure Video Indexer, that return shot boundaries, labels, on-screen text, transcripts, and
face/celebrity detection as an API call, giving non-specialist teams a feature pipeline without training
models. The second is *creative analytics vendors* that specialize in linking extracted creative features to
ad performance across large libraries of past campaigns, effectively industrializing the
features-to-engagement regression of the previous section. The third is *in-house pipelines* at the largest
advertisers, built on the open-source stack (`ffmpeg`/`OpenCV` for decoding, pretrained image and video
backbones, ASR such as Whisper) for proprietary control and integration with first-party outcome data.

**Production constraints.** Three realities discipline the work. Video is *expensive* to process (decoding
and deep inference at frame rate over large libraries is compute-intensive), which is why frame *sampling*
and shot-based keyframing are not academic niceties but production necessities. Outcome data is *fragmented*
across walled platforms, so cross-platform creative-to-sales linkage is a data-integration problem before it
is a modeling problem. And the legal and ethical surface is large: scraping platform video runs into terms
of service, and the modality carries faces, voices, and other biometric-adjacent signals whose extraction
raises privacy obligations that the analyst must respect.

## Frontier and Expansion {#video-frontier}

Three developments are reshaping video-as-data marketing and define the near-term agenda.

**Video-native foundation models.** The trajectory that ran from per-frame CNNs to video transformers
[@khan2022transformers] is converging on large *multimodal* models that ingest video, audio, and text jointly
and emit natural-language descriptions, structured tags, or answers to questions about a clip. For the
analyst this collapses much of the bespoke pipeline into a prompt: rather than training a brand detector, one
can ask a video-language model whether and when a brand appears, recovering interpretable creative features
through instruction rather than supervised training, and extending to video the vision-language approach that
@witte2026language demonstrate for marketing images. Generative video models, in turn, make the creative
itself a model output: @hartmann2025generative ask whether generative AI can reach human-level visual
marketing content, a question that becomes sharper still in motion. The same caution applies in amplified form: a
foundation model's extracted feature is still a generated regressor, now produced by an opaque model whose
error structure is unknown and whose outputs require validation against human coding before they enter an
inference.

**Generative video and the synthetic-creative frontier.** Text-to-video generation makes the creative itself
a model output, enabling at-scale creative experimentation (generate many variants, measure response) but
also raising disclosure, authenticity, and brand-safety questions, and complicating the very notion of a
"natural" creative sample. The measurement opportunity is large: synthetic variation can supply the
experimental creative manipulation that observational video analysis lacks.

**Live, real-time, and causal video analytics.** Livestream commerce pushes analysis [@wen2026livestream]
toward *real-time* multimodal modeling, where features are extracted and related to contemporaneous purchases
as the stream unfolds, and where the host adapts to the audience, making the identification problem genuinely
dynamic. The broader frontier for the field, as for all of unstructured-data marketing surveyed by
@ma2020machine, is to move from *predicting* engagement from creative features to *causally* estimating which
creative choices drive outcomes, which will require pairing the rich feature pipelines of this chapter with
the experimental and quasi-experimental machinery of the methodology part, a transition the
unstructured-data program has called for since @balducci2018unstructured. Video is where the multimodal,
temporal, and causal challenges of marketing measurement meet most acutely, and where the payoff to getting
the measurement right is largest.
