# Networks and Graph Data {#sec-network-graph}

Most of the data structures the preceding chapters relied on are *rectangular*: a
customer is a row, an attribute is a column, and rows are treated as exchangeable draws
from a population. Network data violate that convenience at the root. The unit of
analysis is not the isolated individual but the *relation* between individuals, and the
informative quantity is the pattern of relations across the whole system. A customer
who buys, adopts, churns, or recommends does so embedded in a web of friends,
co-purchasers, followers, and referrers, and that embedding carries marketing signal
that no amount of demographic enrichment recovers. The object that encodes this
structure is a **graph**: a set of *nodes* (consumers, products, brands, sellers) and a
set of *edges* (friendships, co-purchases, referrals, co-mentions) connecting them.

This chapter treats the graph as a first-class marketing data type, parallel to text
(@sec-text-as-data) and images (@sec-image-processing) in the unstructured-and-
multimodal family. As with those modalities, the analytic arc is the same: an
unstructured artifact (here, a relational structure rather than a document or a photo)
is turned into machine-readable features, and those features are then used for
description, prediction, or causal claims. And as with those modalities, the central
econometric caution recurs in a network-specific and unusually severe form: a feature
extracted from the graph, such as a node's centrality or its neighbors' behavior, is a
*generated regressor* whose error is correlated with the outcome in ways that mimic the
very effect the analyst wants to measure. The identification of social influence is the
hardest version of this problem in all of empirical marketing, and it earns its own
section below.

The chapter proceeds from use to method to practice. It opens with the marketing
*applications* that motivate graph thinking. It then develops the formal
*representation* of a graph and the *centrality* measures that summarize a node's
position, with a runnable `igraph` demonstration on simulated data, and the
ego-network measures of *brokerage and structural holes*. It treats
*community detection*, the unsupervised discovery of cohesive subgroups. It then
confronts the *contagion-versus-homophily* identification problem head on, because no
honest account of network marketing can avoid it. It develops *diffusion* models and
simulates an independent-cascade process over a graph. It surveys the learned-
representation frontier, *graph embeddings* and *graph neural networks*, keeping the
deep-learning specifics conceptual but clearly labeled. It closes with *industry and
production* practice (recommendation, fraud, influencer selection) and a *frontier*
section on where the field is moving. @fig-net-workflow previews the shared analytic
arc these sections instantiate: relational data become a graph, the graph yields
features, and those features serve description, prediction, or causal claims.

```{mermaid}
%%| label: fig-net-workflow
%%| fig-cap: "The shared analytic arc of network data. Relational records become a graph of nodes and edges, the graph yields features such as centrality, community, and embeddings, and those features serve description, prediction, or causal claims."
flowchart LR
  RD["Relational data<br/>(friendships, co-purchases, referrals, co-mentions)"]
  RD --> GR["Graph<br/>(nodes and edges)"]
  GR --> FE["Network features<br/>(centrality, community, embeddings)"]
  FE --> US["Uses<br/>(description, prediction, causal claims)"]
```

## Why Networks Matter in Marketing {#sec-graph-applications}

The marketing relevance of graph structure is not a single idea but a family of them,
unified by the premise that a consumer's value, behavior, and reachability depend on
the consumers around them.

**Social networks, word of mouth, and contagion.** The oldest motivation is that
products spread through social ties. A consumer is more likely to adopt when adopters
sit nearby in the social graph, whether because of genuine influence, shared exposure,
or similarity. The word-of-mouth literature in marketing has long modeled this spread
formally; @goldenberg2001 used a complex-systems simulation of the underlying
person-to-person process to show how the structure of "weak" and "strong" ties governs
the macro-level diffusion curve, and a large empirical literature, surveyed below,
estimates contagion from observed adoption sequences over real ties. The contagion
premise is also what makes *misinformation* a network phenomenon: @vosoughi2018 showed
that false news diffuses faster, farther, and more broadly than true news through the
Twitter retweet graph, a finding that is fundamentally about diffusion structure rather
than content alone.

**Co-purchase and recommendation graphs.** A second, entirely different graph arises
without any social tie at all. Connect two products when they are frequently bought
together, or connect a customer to every product they purchased, and the resulting
co-purchase (or bipartite customer-product) graph is the substrate of modern
recommendation. "Customers who bought this also bought that" is a statement about the
neighborhood of a node in a co-purchase graph, and item-to-item collaborative
filtering, the algorithm that powered Amazon's recommendations, is literally a
computation over this graph (Linden, Smith and York 2003, *IEEE Internet Computing*
7(1), 76--80, doi:10.1109/MIC.2003.1167344). The graph here is a tool for *prediction
and retrieval*, and the identification anxieties of the social-contagion literature are
largely absent.

**Customer-referral networks and seeding.** Referral programs make the graph a managed
asset: a firm rewards existing customers for recruiting new ones, and the chain of
who-referred-whom is a directed network whose structure determines program value. The
adjacent question, *whom to seed* with free product or early access to maximize spread,
turns network position into a targeting variable. @hinz2011 compared seeding strategies
empirically and found that seeding well-connected "hubs" outperforms seeding random or
"fringe" customers, a result that operationalizes centrality as a marketing decision
variable. The structure of referral and affiliate networks on the commercial web has
itself been modeled as an outcome of strategic link formation by @katona2008.

**Influencer identification.** Closely related is the problem of finding the
individuals whose activity moves the system. The naive answer, "the most connected
person," is both intuitive and contested. @trusov2010 developed a method to identify
genuinely influential users in an internet social network by modeling whose activity
predicts others' subsequent activity, distinguishing influence from mere connectedness.
@watts2007influentials (Watts and Dodds 2007, *Journal of Consumer Research* 34(4),
441--458, doi:10.1086/518527) challenged the "influentials hypothesis" itself,
arguing through simulation that large cascades are often driven by a critical mass of
easily influenced ordinary people rather than by rare special individuals, a caution
that remains central to how marketers should read centrality.

**Customer value, retention, and CRM spillovers.** A consumer's economic value to a
firm is partly a network quantity. Social ties shape churn—a customer whose connections
defect is more likely to defect [@nitzan2011social]—and the value of participation in a
firm-sponsored community spills over into the spending of those who merely observe it,
the "social dollars" quantified by @manchanda2015social. CRM actions, too, propagate
along ties: @ascarza2017beyond show that a retention campaign aimed at target customers
also moves the behavior of their untreated social neighbors, so the campaign's true
return is a network effect that target-only accounting misses. The classic statement of
this logic is @hill2006networkbased, who show that consumers linked (by communication
ties) to existing customers are far likelier to adopt, making explicit consumer networks
a powerful targeting variable. Firm-generated content on social platforms engages users
through the same network substrate [@lee2018advertising].

**Referral economics and social-commerce value.** Referral programs turn the graph into
a managed asset whose returns can be measured: referred customers are more valuable and
loyal than others [@schmitt2011referral], and @vandenbulte2018customer decompose *why*,
separating the selection (homophily) and influence channels through which referral
programs convert social capital into economic capital. In social commerce, a seller's
position in the network generates value directly [@stephen2010deriving], and the motives
for contributing the content that builds these networks—intrinsic versus
status/image—are themselves measurable [@toubia2013]. Whether observed
clustering reflects genuine word of mouth or mere observational learning is itself an
identification question [@chen2011online], and the interplay of social influence with the
firm's own direct marketing unfolds dynamically over the adoption process
[@risselada2014dynamic]. The diffusion consequences of network position are not uniformly
positive: hubs play a distinctive role in adoption [@goldenberg2009role], yet network
externalities can also *chill* diffusion when consumers wait for others to commit first
[@goldenberg2010chilling].

**Brand co-mention and semantic networks.** Finally, graphs need not connect people or
products at all. Connect two brands when they are mentioned together in reviews, social
posts, or search sessions, and the resulting co-mention network is a map of *market
structure*: tightly connected brand clusters are competitive sets, and a brand's
position reveals its substitutes and complements. This is the network-analytic cousin
of the text-mining market-structure maps discussed in @sec-text-as-data, and it lets an
analyst recover perceptual structure from unstructured co-occurrence rather than from
surveys.

Across these applications the same primitive recurs: a position in a graph is a
feature, and the configuration of the graph is information. @fig-net-types sorts these
applications by the kind of graph each builds, since the nodes and edges differ sharply
across them.

```{mermaid}
%%| label: fig-net-types
%%| fig-cap: "A taxonomy of marketing networks by what their nodes and edges represent, with the marketing use each structure supports."
flowchart TB
  G["Marketing networks"]
  G --> SO["Social / word-of-mouth<br/>(people to people)"]
  G --> CP["Co-purchase / recommendation<br/>(product to product, customer to product)"]
  G --> RE["Referral / seeding<br/>(directed: who referred whom)"]
  G --> BC["Brand co-mention / semantic<br/>(brand to brand)"]
  SO --> SOu["Contagion, influence,<br/>churn spillover"]
  CP --> CPu["Recommendation, retrieval"]
  RE --> REu["Seeding, referral value"]
  BC --> BCu["Market-structure maps"]
```

The rest of the chapter develops the methods that turn that information into measures.

## Graph Representation and Centrality {#sec-graph-centrality}

A graph $G = (V, E)$ is a set of nodes $V$ and a set of edges $E \subseteq V \times V$.
Edges may be **undirected** (a Facebook friendship is mutual) or **directed** (a
Twitter follow is one-way; an A-referred-B link has an arrow), and they may be
**weighted** (the number of co-purchases, the strength of a tie). The workhorse
algebraic representation is the **adjacency matrix** $A$, an $n \times n$ matrix with
$A_{ij} = 1$ (or the edge weight) when an edge runs from $i$ to $j$ and $0$ otherwise.
Almost every structural quantity in this chapter is a function of $A$. For sparse
real-world graphs, where each node connects to a tiny fraction of all others, $A$ is
stored as an edge list rather than a dense matrix, but the mathematics is unchanged.

The most basic summary of a node is its **degree**, the number of edges incident to it
($d_i = \sum_j A_{ij}$). In a directed graph this splits into *in-degree* (how many
point to $i$; a proxy for received attention or popularity) and *out-degree* (how many
$i$ points to; a proxy for activity). Degree is the crudest notion of importance, and
its crudeness is exactly the point of the richer **centrality** measures, each of which
formalizes a different theory of what makes a node matter.

**Betweenness centrality** measures how often a node lies on shortest paths between
other pairs of nodes:
$$
C_B(i) = \sum_{s \ne i \ne t} \frac{\sigma_{st}(i)}{\sigma_{st}},
$$ {#eq-graph-betweenness}
where $\sigma_{st}$ is the number of shortest paths from $s$ to $t$ and
$\sigma_{st}(i)$ the number passing through $i$. A high-betweenness node is a *broker*
sitting on the bridges between otherwise separated regions; in marketing terms, a
broker controls the flow of information across communities and is a candidate "weak tie"
connector in the sense of the word-of-mouth literature.

Brokerage need not be defined on the graph of who knows whom. @mell2021brokerage
introduce **temporal brokerage**, a position in a team's *temporal* structure that
bridges subgroups sharing little or no overlap in working hours. It is not a formal
role and need not coincide with high betweenness in the communication graph, yet it
attracts the same coordination work, because the temporal broker is the only member
who can relay between subgroups that cannot speak to each other directly. The
consequences are two-sided, greater integrative complexity on one hand and a heavier
workload on the other, and they spill beyond the focal team into the broker's other
commitments. For measurement the lesson is that the adjacency matrix behind
@eq-graph-betweenness encodes one relation at a time, and the bridge that matters
substantively may be defined on a different one, such as overlap in working hours,
shared language, or shift schedule. Choosing the edge is choosing the theory.

**Eigenvector centrality** formalizes the recursive idea that a node is important if its
neighbors are important. It is the leading eigenvector of the adjacency matrix:
$$
A\,\mathbf{x} = \lambda_{\max}\,\mathbf{x},
$$ {#eq-graph-eigenvector}
so a node's score is proportional to the sum of its neighbors' scores. PageRank is a
directed, damped variant of the same idea. Eigenvector centrality captures *prestige*:
being connected to other well-connected nodes, not merely to many nodes. The three
measures can disagree sharply, and the disagreement is informative. A node can have
high degree but low betweenness (a popular member buried inside one tight cluster), or
modest degree but high betweenness (a lone bridge between two cliques). Choosing a
centrality measure is therefore choosing a theory of influence, not a neutral
preprocessing step.

The runnable demonstration below builds a simulated graph and computes all three. The
package is the maintained `igraph`, the standard for network analysis in R. The graph
is generated by *preferential attachment* (the Barabási-Albert model), which produces
the heavy-tailed degree distribution characteristic of real social and co-purchase
networks: a few hubs and a long tail of sparsely connected nodes.

```{r graph-centrality, message=FALSE, warning=FALSE}
#| fig-cap: "A simulated scale-free social/co-purchase graph of 80 nodes generated by preferential attachment. Node size is proportional to degree; color encodes betweenness (darker = higher). The hubs (large nodes) and the brokers (dark nodes) are not always the same vertices, which is the practical point of distinguishing centrality measures."
library(igraph)
set.seed(20260622)

# --- Build a simulated scale-free network (Barabasi-Albert) ---
# Interpretable as a social graph (who-follows-whom) or, with edges read as
# "frequently bought together," a product co-purchase graph.
g <- sample_pa(n = 80, power = 1.1, m = 2, directed = FALSE)
V(g)$name <- paste0("u", seq_len(vcount(g)))

# --- Centrality measures (all functions of the adjacency structure) ---
deg  <- degree(g)
btw  <- betweenness(g, normalized = TRUE)
eig  <- eigen_centrality(g)$vector

cent <- data.frame(
  node       = V(g)$name,
  degree     = deg,
  betweenness = round(btw, 3),
  eigenvector = round(eig, 3)
)

# Show the five highest-degree nodes and how they rank on the other measures
top_deg <- head(cent[order(-cent$degree), ], 5)
cat("Top-5 nodes by degree:\n")
print(top_deg, row.names = FALSE)

# The top node by EACH measure can differ:
cat("\nMost central node by measure:\n")
cat("  degree     :", cent$node[which.max(cent$degree)], "\n")
cat("  betweenness:", cent$node[which.max(cent$betweenness)], "\n")
cat("  eigenvector:", cent$node[which.max(cent$eigenvector)], "\n")

# Rank correlation between degree and betweenness: high but not 1,
# which is exactly why brokers and hubs are distinct managerial targets.
cat("\nSpearman corr(degree, betweenness):",
    round(cor(deg, btw, method = "spearman"), 3), "\n")

# --- Visualize ---
node_col <- colorRampPalette(c("#d9e3f0", "#1f3b73"))(100)[
  as.integer(cut(btw, breaks = 100, include.lowest = TRUE))]
plot(g,
     vertex.size   = 4 + 14 * (deg / max(deg)),
     vertex.color  = node_col,
     vertex.label  = NA,
     edge.color    = "grey75",
     layout        = layout_with_fr(g))
```

The output makes the conceptual point operational. The Spearman correlation between
degree and betweenness is high but well below one, and the single most central node
differs across measures. A seeding or influencer program that targets "the most
connected user" is implicitly committing to the degree theory of influence; if the real
mechanism is brokerage across communities, betweenness is the right target, and the two
will recommend different people.

## Brokerage and Structural Holes {#sec-graph-brokerage}

Betweenness (@eq-graph-betweenness) finds brokers by counting shortest paths across the
whole graph. A second tradition, older in sociology and now central in strategy and
organization research, asks a more local question: are a node's own contacts connected
to one another? Its starting point is @granovetter1973, who argued that the overlap of
two people's friendship networks rises with the strength of the tie between them, so
that weak ties are the ones that link otherwise separate groups and carry information
between them. @burt1992structural moved the argument from the tie to the position. A
**structural hole** is a disconnection between two of ego's nonredundant contacts, and
bridging such holes yields information benefits (access to nonredundant information) and
control benefits (the ability to deal with parties who cannot easily deal with each
other) [@burt1997contingent]. The control benefit is the *tertius gaudens* strategy of a
broker who plays contacts off against one another for his or her own benefit
[@obstfeld2005tertius]. The opposing view locates advantage in *closure*, a network in
which contacts are tied to one another: @coleman1988social describes how closure
facilitates two of the forms social capital takes, obligations and expectations on the
one hand and effective social norms on the other. Much of the subsequent literature is
an argument about when each view holds, and @burt2013advantage review it with network
advantage as the organizing lens, tracing advantage to information breadth, timing, and
arbitrage.

### Burt's measures

Burt's measures are computed on the ego network. Let $N(i)$ be ego $i$'s set of
contacts, $N_i = |N(i)|$, and let
$p_{ij} = (a_{ij} + a_{ji}) / \sum_{k}(a_{ik} + a_{ki})$ be the proportion of $i$'s
relational investment directed to $j$, which reduces to $p_{ij} = 1/N_i$ in a binary
network. **Constraint** measures how far ego's investment is concentrated, directly or
through mutual contacts, on contacts who are themselves connected:
$$
c_i = \sum_{j \in N(i)} \Big( p_{ij} + \sum_{q \ne i, j} p_{iq}\, p_{qj} \Big)^2 .
$$ {#eq-graph-constraint}
The first term inside the parentheses is ego's direct investment in $j$; the second is
the investment that reaches $j$ indirectly through a third contact $q$. Constraint is
therefore high when ego has few contacts, when tie strength is concentrated on a few of
them, and when those contacts are tied to one another, and lower constraint is read as
more structural holes [@borgatti2026beyond]. Burt's general formula also weights each
term by an exogenous concentration factor, which virtually all studies set to one
[@borgatti2026beyond]. In `igraph::constraint()` the proportions $p_{qj}$ are computed
over all of $q$'s ties (the "whole network" convention) rather than over $q$'s ties
inside ego's network; the two conventions coincide when a network contains only ego and
its contacts.

The second measure, *effective size*, is the number of contacts discounted for their
redundancy with one another:
$$
\mathrm{ES}_i = \sum_{j \in N(i)} \Big( 1 - \sum_{q \ne i, j} p_{iq}\, m_{jq} \Big)
\;=\; N_i - \frac{2\,t_i}{N_i} \quad \text{(binary data)},
$$ {#eq-graph-effsize}
where $m_{jq}$ is $j$'s tie to $q$ divided by $j$'s strongest tie to anyone and $t_i$
is the number of ties among ego's contacts. The binary simplification, due to Borgatti
(1997, *Connections* 20(1): 35--38), shows that effective size is
network size minus the average degree of the alters within the ego network
[@everett2020unpacking]. It equals $N_i$ when no contacts are tied and 1 when every
contact is tied to every other, and *efficiency* is effective size divided by $N_i$
[@burt1992structural].

A different formalization ties brokerage to group membership rather than to ego-network
density. @gould1989structures partition actors into mutually exclusive groups and
classify every open two-path $a \to b \to c$ (one in which $a$ has no direct tie to
$c$) by the group memberships of the three actors. The partition yields five distinct
roles for the broker $b$: a *coordinator* mediates between two members of its own
group; an *itinerant* broker mediates between two members of a single group to which it
does not belong; a *gatekeeper* passes a contact from an outsider into its own group; a
*representative* passes a contact from its own group to an outsider; and a *liaison*
connects two different groups, neither of them its own. Gould and Fernandez give
quantitative measures of each role and a test of whether occupancy of a role exceeds
what a random distribution of exchange relations would produce; the `sna` package's
`brokerage()` function implements the counts. The roles read naturally onto marketing
structures. A distributor that carries a manufacturer's offer into its own retail
network occupies the gatekeeper position, and a key-account manager who carries the
firm's offer to a customer's buying center acts as a representative.

### What the evidence shows

The canonical positive evidence is @burt2004goodideas. Among managers in a large
American electronics company, compensation, positive performance evaluations,
promotions, and good ideas were disproportionately held by people whose networks
spanned structural holes, and between-group brokers were more likely to express ideas,
less likely to have them dismissed, and more likely to have them judged valuable. The
mechanism is that opinion and behavior are more homogeneous within than between
groups, so a person connected across groups sees options that others do not. The value
of that position is contingent. @burt1997contingent shows that it is especially high for
managers with few peers doing the same work, and @burt2007secondhand finds that
brokerage benefits are concentrated in the immediate network, with secondhand brokerage
across indirect contacts often worth little. The content of ties matters too:
@podolny1997resources find that mobility is enhanced by a large, sparse network of
informal ties for acquiring information and resources, whereas well-defined performance
expectations are more likely to arise from a small, dense network.

The record is mixed once the outcome and the setting change. In the international
chemicals industry, @ahuja2000collaboration finds that direct and indirect collaborative
ties raise a firm's subsequent innovation output but that more structural holes in the
interfirm collaboration network lower it. Among Canadian mutual fund companies, by
contrast, innovative firms that also bridge structural holes earn a further performance
boost [@zaheer2005benefiting]. For investment banks advising on U.K. mergers and
acquisitions, structural holes increase status accumulation but dampen market
performance, because open networks supply information about new opportunities while
limiting information about partners' cooperativeness [@shipilov2008cake]. Cohesion and
brokerage are not simple opposites. @reagans2003network find that both social cohesion
around a tie and network range ease knowledge transfer, @fleming2007collaborative show
across 35,400 inventors that collaborative brokerage helps generate an idea but hampers
its diffusion and use by others, and @tortoriello2010activating find no advantage to
bridging per se among 276 R&D scientists and engineers, only to bridging ties that are
Simmelian, that is, embedded in a three-person clique. @aral2011diversity add a
diversity-bandwidth trade-off. A structurally diverse network raises novelty at the cost
of thinner information flow, so in e-mail and network data from an executive recruiting
firm bridging ties sometimes delivered less novel information, depending on how much
contacts' information overlaps, how deep their topical knowledge is, and how fast it
refreshes.

How brokers broker is a separate question. @obstfeld2005tertius contrasts the *tertius
gaudens* with the *tertius iungens* ("third who joins"), who introduces disconnected
people or coordinates connected ones, and finds that a tertius iungens orientation,
dense networks, and diverse social knowledge predict involvement in innovation in an
automotive engineering division. @quintane2016brokers reconcile the two strategies over
time with e-mail relational-event data. When brokering through short-term interactions
outside their long-term relationships, brokers are more likely than densely embedded
actors to intermediate the information flow themselves; when brokering through their
long-term ties, they are more likely to facilitate a direct exchange between the
brokered parties. Brokers also affect the people around them. In the French television
game show industry, hubs, the brokers who act as the main interface between their own
network community and others, generate positive externalities for neighbors in
creativity-focused roles and negative ones for neighbors in efficiency-focused roles
[@clement2018brokerage]. @stovel2012brokerage stress the same duality at the societal
level: brokerage eases interaction and economic activity, but it also breeds
exploitation, corruption, and the accumulation of power.

Finally, brokerage positions are contested and endogenous. In Italian television
production, the structural holes a team spans originate in the prior status and
centrality of the teams its members came from, and spanning holes is associated with
larger audiences [@zaheer2009origins]. In a game-theoretic model of network formation
in which every actor strives for structural holes, the predominant stable networks
distribute benefits evenly, so no one keeps a structural advantage in the long run
[@buskens2008dynamics]. @ryall2007brokers reach a parallel conclusion with a biform game:
a broker earns an advantage only if no substitutes exist for the connections it offers
or the value it creates, if it intermediates more than two parties, and if
interdependence does not lock it into a fixed pattern of exchange, and when actors form
relations strategically, profitable intermediation survives only under strict
assumptions of unilateral action. Substitutability is exactly what the next development
puts at the center of measurement.

### Opportunity versus exclusivity

Constraint has been the workhorse operationalization of structural holes, and
@johnson2026reassessing argue that it does not measure what it is usually taken to
measure.
Using formal, computational, and empirical analyses, they conclude that constraint
almost exclusively operationalizes dyadic constructs, namely available alternatives,
exchange-partner diversity, and concentration, rather than the broader network
structure that structural-hole theory invokes. As @borgatti2026beyond summarize the
evidence, in large unweighted simulated networks the variance of inverse degree alone
averages about 110 percent of the variance of constraint (possible because inverse
degree covaries negatively with the closure terms), and in replications of two earlier
studies, one of them the investment-bank study of @shipilov2008cake, constraint
correlates .94 and .99 with the Herfindahl index of ego's tie distribution.

@borgatti2026beyond accept that constraint has a measurement problem but locate it
elsewhere. Degree belongs in a brokerage measure by design, because contacts create the
pairs across which brokerage can occur: an ego with eight mutually disconnected contacts
has more brokerage occasions than one with three. The real failure is substitutability.
Their diagnostic holds degree fixed at eight. With no ties among the alters, ego's
constraint is .125. Tying one alter to every other alter creates a competing broker,
what @everett2020unpacking call a *shadow ego*, and constraint correctly rises to .455.
Adding a second, third, and fourth competitor, each of which weakens ego further,
lowers constraint to .452, .441, and .436, and even the fully closed ego network scores
.439, below the single-competitor case. This is the general pattern
@everett2020unpacking derived: with more than seven alters, maximum constraint occurs in
a sparse shadow-ego network rather than in a closed one, and constraint can exceed one
in small ego networks but cannot reach one in large ones.

The remedy is to take the brokered pair, not the individual alter, as the unit of
measurement. A *gap* is an unordered pair $\{j, k\}$ of ego's contacts that are not tied
to each other; let $\mathcal{G}_i$ be ego's set of gaps and $c_{jk}$ the number of
common neighbors of $j$ and $k$ in the whole network, which is at least one because ego
is among them. Then
$$
P_i = |\mathcal{G}_i|, \qquad
Q_i = \frac{1}{P_i} \sum_{\{j,k\} \in \mathcal{G}_i} \frac{1}{c_{jk}}, \qquad
P_i Q_i = \sum_{\{j,k\} \in \mathcal{G}_i} \frac{1}{c_{jk}} .
$$ {#eq-graph-pq}
$P$ is **opportunity volume**, the number of occasions on which ego can connect two
otherwise unconnected contacts. $Q$ is average *exclusivity*: $1/c_{jk}$ is ego's share
of the length-two brokerage between $j$ and $k$, equal to one when ego is the only
intermediary and to one half when a single competitor can make the same connection,
which can be read as ego's expected share of brokerage occasions when competing brokers
are equally capable. Their product $PQ$ is total exclusivity-weighted brokerage, and it
equals ego's betweenness restricted to geodesics of length two. An indicator version
scores a gap one only if $c_{jk} = 1$, so that $PQ_e$ counts the pairs ego connects
exclusively; it follows the value-capture logic that a broker with a substitute for a
given connection adds no value to it [@ryall2007brokers]. $P$ is closely tied to Burt's
measure: for fixed degree it is a monotone transformation of ego-network density, and
for binary data effective size equals $1 + 2P/N$ [@borgatti2026beyond].

Two refinements matter in practice. First, a gap with a single intermediary can still
be shallow if the two contacts are joined by a short path that avoids ego. Removing ego
and letting $d_{jk}$ be the length of the shortest remaining path between $j$ and $k$,
or letting $A_{-i}$ be the adjacency matrix with ego deleted, two distance-sensitive
pair scores are
$$
q^{\text{geo}}_{jk} = 1 - \rho^{\,d_{jk} - 1}, \qquad
q^{\text{walk}}_{jk} = \frac{1}{1 + r_{jk}}, \quad
r_{jk} = \Big[ A_{-i}^{2} \,(I - \beta A_{-i})^{-1} \Big]_{jk},
$$ {#eq-graph-hole-depth}
with $0 < \rho < 1$, $q^{\text{geo}}_{jk} = 1$ when no bypass exists, and
$0 < \beta < 1/\lambda_1$ for the largest eigenvalue $\lambda_1$ of $A_{-i}$. The
shortest-bypass score ignores how many bypasses there are, whereas the walk-discounted
score, which applies Katz-style walk counting to a single pair, penalizes both the
number and the length of ego-avoiding routes and approaches $1/c_{jk}$ as
$\beta \to 0$ [@borgatti2026beyond]. Second, the common-neighbor count must come from
the whole network. A rival need not be one of ego's contacts to weaken ego's position,
so with egocentric survey data $Q$ can only be computed locally and will overstate
exclusivity.

These measures share a monotonicity property that constraint lacks: a tie that does not
involve ego can close a gap, create a bypass, or add a competitor, but it can never
raise $P$ or $PQ$. $Q$ alone is not monotone, because closing a low-quality gap removes
it from the average, which is why it should be reported with $P$ rather than on its own.
The components also carry different strategic meaning. An ego spanning ten gaps, each
shared with four other brokers ($P = 10$, $Q = .20$), and an ego spanning two gaps
exclusively ($P = 2$, $Q = 1$) have the same $PQ = 2$; the first has more occasions for
transfer and recombination, the second more leverage over timing and bargaining. Hence
@borgatti2026beyond recommend specifying the structural-hole mechanism first.
Information access depends on $P$ (the diversity of what those contacts know has to be
measured separately, for instance from alter attributes), timing on exclusivity ($Q$ or
$PQ$), control on the absence of substitutes (the indicator or base $Q$, or $PQ$),
synthesis on $P$ and $Q$ jointly, and referral, which depends on contacts' ties beyond
ego's network, is captured by none of $P$, $Q$, or constraint. For empirical work they
suggest entering $P$ and $Q$ separately while controlling for a function of degree,
with $PQ$ acting as their interaction; in their reconstruction of a published industry
network the two components correlate only .201. All of these are measures of
opportunity, not of whether or how an actor actually brokers.

@fig-graph-brokerage contrasts three ego networks with the same three contacts. Closure
leaves no gaps, the open network gives ego three exclusive gaps, and in the contested
network a rival outside ego's network reaches the same contacts, which halves $Q$ and
$PQ$ while leaving constraint exactly where it was in the open case.

```{mermaid}
%%| label: fig-graph-brokerage
%%| fig-cap: "Three ego networks with the same three contacts. Under closure every pair of contacts is tied, so ego spans no gaps. In the open network ego spans all three gaps alone. In the contested network a rival broker outside ego's network (dashed ties) reaches the same contacts, so every gap is shared and exclusivity halves, yet constraint is identical to the open case because the rival is not one of ego's contacts."
flowchart TB
  subgraph CO["Contested: shared gaps"]
    direction TB
    E3["Ego<br/>P = 3, Q = 0.5, PQ = 1.5<br/>constraint = 0.333"]
    A3["A"]
    B3["B"]
    C3["C"]
    R3["Rival broker"]
    E3 --- A3
    E3 --- B3
    E3 --- C3
    R3 -.- A3
    R3 -.- B3
    R3 -.- C3
  end
  subgraph OP["Open: exclusive gaps"]
    direction TB
    E2["Ego<br/>P = 3, Q = 1, PQ = 3<br/>constraint = 0.333"]
    A2["A"]
    B2["B"]
    C2["C"]
    E2 --- A2
    E2 --- B2
    E2 --- C2
  end
  subgraph CL["Closure: no gaps"]
    direction TB
    E1["Ego<br/>P = 0, PQ = 0<br/>constraint = 0.926"]
    A1["A"]
    B1["B"]
    C1["C"]
    E1 --- A1
    E1 --- B1
    E1 --- C1
    A1 --- B1
    B1 --- C1
    A1 --- C1
  end
```

The first demonstration implements @eq-graph-effsize and @eq-graph-pq alongside
`igraph::constraint()` (@eq-graph-constraint) and reproduces the eight-alter shadow-ego
sequence of @borgatti2026beyond.

```{r}
#| label: fig-graph-brokerage-shadow
#| message: false
#| warning: false
#| fig-width: 8
#| fig-height: 3.8
#| fig-cap: "Burt's constraint and the pair-based measures along the eight-alter shadow-ego sequence, in which k of ego's alters become competing brokers tied to every other alter. Ego's position worsens with every added competitor. Constraint (left, plotted from k = 1 on a narrow vertical scale; it is 0.125 at k = 0) peaks at 0.455 with the first competitor and then falls, signalling an improvement that has not occurred; P and PQ (right) decline monotonically."
library(igraph)

# --- Pair-based brokerage measures (Borgatti and Everett 2026) ---
# For ego i, a brokered pair (a "gap") is an unordered pair {j, k} of i's
# alters that are not tied to each other. c_jk is the number of common
# neighbors of j and k in the WHOLE network; ego is always one, so c_jk >= 1.
#   P   = number of brokered pairs                  (opportunity volume)
#   Q   = mean of 1 / c_jk over brokered pairs      (base exclusivity)
#   PQ  = sum of 1 / c_jk                           (= two-step betweenness)
#   PQe = number of pairs with c_jk == 1            (indicator version)
brokerage_pq <- function(g, ego) {
  alters <- as.integer(neighbors(g, ego))
  none <- c(P = 0, Q = NA, PQ = 0, PQe = 0)
  if (length(alters) < 2) return(none)
  prs  <- combn(alters, 2)
  open <- !apply(prs, 2, function(p) are_adjacent(g, p[1], p[2]))
  if (!any(open)) return(none)
  prs  <- prs[, open, drop = FALSE]
  c_jk <- apply(prs, 2, function(p)
    length(intersect(neighbors(g, p[1]), neighbors(g, p[2]))))
  c(P = ncol(prs), Q = mean(1 / c_jk), PQ = sum(1 / c_jk),
    PQe = sum(c_jk == 1))
}

# Burt's effective size for binary data, N - 2t/N, where t is the number of
# ties among ego's N alters; efficiency is effective size divided by N.
effective_size <- function(g, ego) {
  alters <- as.integer(neighbors(g, ego))
  N <- length(alters)
  t <- ecount(induced_subgraph(g, alters))
  c(N = N, ES = N - 2 * t / N, eff = (N - 2 * t / N) / N)
}

# --- The eight-alter shadow-ego sequence ---
# Ego (vertex 1) has 8 alters (2..9), initially unconnected. Step k makes k
# alters competing brokers ("shadow egos") tied to every other alter. By
# structural-hole theory ego's position worsens monotonically in k.
shadow_net <- function(k, n_alters = 8) {
  g <- make_star(n_alters + 1, mode = "undirected", center = 1)
  alters <- 2:(n_alters + 1)
  for (s in alters[seq_len(k)]) {
    others <- setdiff(alters, s)
    new <- others[!sapply(others, function(o) are_adjacent(g, s, o))]
    if (length(new) > 0) g <- add_edges(g, as.vector(rbind(s, new)))
  }
  g
}

shadow <- do.call(rbind, lapply(0:7, function(k) {
  g <- shadow_net(k)
  m <- brokerage_pq(g, 1)
  data.frame(k          = k,
             constraint = round(constraint(g, nodes = 1), 3),
             eff_size   = effective_size(g, 1)[["ES"]],
             P          = m[["P"]],
             Q          = round(m[["Q"]], 3),
             PQ         = round(m[["PQ"]], 2),
             PQe        = m[["PQe"]],
             btw2       = round(betweenness(g, v = 1, cutoff = 2), 2))
}))
print(shadow, row.names = FALSE)

# --- Identities stated in the paper ---
cat("\nPQ equals two-step betweenness in every row:",
    isTRUE(all.equal(shadow$PQ, shadow$btw2)), "\n")
cat("Effective size equals 1 + 2P/N in every row:",
    isTRUE(all.equal(shadow$eff_size, 1 + 2 * shadow$P / 8)), "\n")

# --- Figure: constraint versus P and PQ along the sequence ---
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.4, 2.2, 1))
s1 <- shadow[shadow$k >= 1, ]
plot(s1$k, s1$constraint, type = "b", pch = 19, col = "#1f3b73",
     xlab = "Competing brokers (k)", ylab = "Burt's constraint",
     main = "Constraint (k = 0: 0.125)", cex.main = 0.95)
plot(shadow$k, shadow$P, type = "b", pch = 1, col = "grey40",
     ylim = c(0, 28), xlab = "Competing brokers (k)",
     ylab = "Count of gaps", main = "P and PQ", cex.main = 0.95)
lines(shadow$k, shadow$PQ, type = "b", pch = 19, col = "#b0413e")
legend("topright", c("P (gaps spanned)", "PQ (exclusivity-weighted)"),
       pch = c(1, 19), col = c("grey40", "#b0413e"), bty = "n", cex = 0.85)
par(op)
```

The printed table reproduces the published values: constraint runs .125, .455, .452,
.441, .436, .436, .438, .439, while $P$ falls from 28 to 0, $Q$ from 1 to 1/7, and
$PQ$ from 28 to 0. The indicator count $PQ_e$ drops from 28 to 0 at the very
first competitor, since from then on no remaining gap is ego's alone. The two printed
checks confirm the identities stated above, that $PQ$ is two-step betweenness
(`betweenness(g, cutoff = 2)` in `igraph`) and that effective size equals $1 + 2P/N$.

The second demonstration applies the same functions, plus the distance-sensitive scores
of @eq-graph-hole-depth, to the three toy networks of @fig-graph-brokerage and to a
simulated business-to-business referral network of 36 firms in three regional clusters,
with ties far more likely within clusters than across them.

```{r}
#| label: graph-brokerage-sim
#| message: false
#| warning: false
library(igraph)

# --- Three toy ego networks with the same three contacts A, B, C ---
# In the contested case the rival broker is tied to A, B, and C but is NOT
# one of ego's contacts.
toy <- list(
  closure   = graph_from_literal(Ego-A, Ego-B, Ego-C, A-B, B-C, A-C),
  open      = graph_from_literal(Ego-A, Ego-B, Ego-C),
  contested = graph_from_literal(Ego-A, Ego-B, Ego-C, Rival-A, Rival-B, Rival-C)
)
toy_tab <- do.call(rbind, lapply(names(toy), function(nm) {
  g <- toy[[nm]]; m <- brokerage_pq(g, "Ego")
  data.frame(case = nm,
             constraint = round(constraint(g, nodes = "Ego"), 3),
             P = m[["P"]], Q = round(m[["Q"]], 3), PQ = round(m[["PQ"]], 2))
}))
print(toy_tab, row.names = FALSE)

# --- Distance-sensitive quality scores (Borgatti and Everett 2026, App. B) ---
# Shortest bypass: 1 - rho^(d_jk - 1), with d_jk the j-k distance once ego is
# removed (no bypass: score 1). Walk-discounted: 1 / (1 + r_jk), where
# r = A^2 (I - beta A)^{-1} on the ego-deleted adjacency matrix A and
# beta < 1 / lambda_1 (here beta = 0.5 / lambda_1).
deep_quality <- function(g, ego, rho = 0.5, beta_frac = 0.5) {
  alters <- as.integer(neighbors(g, ego))
  if (length(alters) < 2) return(c(Qgeo = NA, Qwalk = NA))
  prs  <- combn(alters, 2)
  open <- !apply(prs, 2, function(p) are_adjacent(g, p[1], p[2]))
  if (!any(open)) return(c(Qgeo = NA, Qwalk = NA))
  prs  <- prs[, open, drop = FALSE]
  keep <- setdiff(seq_len(vcount(g)), ego)
  h    <- induced_subgraph(g, keep)
  idx  <- matrix(match(prs, keep), nrow = 2)      # pair ids inside h
  d    <- distances(h)[cbind(idx[1, ], idx[2, ])]
  geo  <- ifelse(is.finite(d), 1 - rho^(d - 1), 1)
  A    <- as_adjacency_matrix(h, sparse = FALSE)
  lam  <- max(abs(eigen(A, symmetric = TRUE, only.values = TRUE)$values))
  beta <- if (lam > 0) beta_frac / lam else 0
  R    <- A %*% A %*% solve(diag(nrow(A)) - beta * A)
  walk <- 1 / (1 + R[cbind(idx[1, ], idx[2, ])])
  c(Qgeo = mean(geo), Qwalk = mean(walk))
}

# --- A simulated B2B referral network: 36 firms in three regional clusters ---
set.seed(20261001)
pm  <- matrix(0.02, 3, 3); diag(pm) <- 0.30
g   <- sample_sbm(36, pref.matrix = pm, block.sizes = c(12, 12, 12))
blk <- rep(1:3, each = 12)                 # regional cluster of each firm
V(g)$name <- paste0("f", seq_len(vcount(g)))

meas <- do.call(rbind, lapply(seq_len(vcount(g)), function(i) {
  pq <- brokerage_pq(g, i); es <- effective_size(g, i); dq <- deep_quality(g, i)
  data.frame(firm = V(g)$name[i], degree = es[["N"]],
             clusters = length(unique(blk[as.integer(neighbors(g, i))])),
             eff_size   = round(es[["ES"]], 2),
             constraint = round(constraint(g, nodes = i), 3),
             P = pq[["P"]], Q = round(pq[["Q"]], 3), PQ = round(pq[["PQ"]], 2),
             Qgeo = round(dq[["Qgeo"]], 3), Qwalk = round(dq[["Qwalk"]], 3))
}))

cat("\nSix firms with the largest PQ:\n")
print(head(meas[order(-meas$PQ), ], 6), row.names = FALSE)
cat("\nLowest-constraint firm:", meas$firm[which.min(meas$constraint)],
    "| highest-PQ firm:", meas$firm[which.max(meas$PQ)], "\n")

cat("\nSpearman correlations across the 36 firms:\n")
print(round(cor(meas[, c("degree", "constraint", "P", "Q", "PQ")],
                method = "spearman", use = "pairwise.complete.obs"), 2))
```

The toy table confirms the figure: closure scores a constraint of .926 with no gaps,
and the open and contested networks share a constraint of .333 even though the rival
halves ego's exclusivity ($Q$ from 1 to .5, $PQ$ from 3 to 1.5). Constraint cannot see
the rival because the rival is not one of ego's contacts, whereas $Q$ counts common
neighbors over the whole network. The simulated network shows the same separation at
scale. Firms f33 and f21 have identical degree (5), effective size (4.60), number of
gaps ($P = 9$), and constraint (.262), yet f33, whose contacts lie in all three
clusters, spans every one of its gaps alone ($Q = 1$), while f21, whose contacts all
lie in its own cluster, shares some of its gaps with other firms in that cluster
($Q = .815$). Burt's measures cannot tell the two positions apart; $Q$, $PQ$, and both
distance-sensitive scores can. The correlation matrix makes the @johnson2026reassessing
point concrete, since constraint correlates $-.91$ with degree, and it shows why
separating $P$ from $Q$ is useful: $Q$ is nearly uncorrelated with constraint ($.14$),
so it carries information that constraint does not. The lowest-constraint firm (f27)
and the highest-$PQ$ firm (f13) also differ, so a program that targets "the firm with
the most structural holes" will choose different partners depending on the measure.
Note, too, that f13's contacts all sit in its own cluster. Structural measures count
gaps without knowing which group boundaries they cross, which is exactly the information
the Gould-Fernandez roles add.

### Implications for marketing

Seeding and opinion-leader programs have mostly targeted degree. @hinz2011 find that
seeding well-connected people is the most successful strategy because they are more
likely to participate in a campaign; they use their greater reach but are not more
influential on their peers than less-connected people. @goldenberg2009role show that
hubs tend to adopt early, with innovative hubs speeding adoption and follower hubs
enlarging the market, and @iyengar2011opinion find that sociometric and self-reported
opinion leadership are weakly correlated and capture different constructs. Structure
around the target matters as well: in an online social network, the density of
connections among a consumer's already-adopted contacts strongly raises the consumer's
adoption probability, and the average influence of adopters declines with their total
number of contacts [@katona2011network]. That clustering effect is a closure result, and
together with the finding that brokered ideas diffuse less [@fleming2007collaborative] it
cautions against equating "broker" with "best seed." The measurement lesson of this
section applies directly: if a campaign needs novel information to cross community
boundaries, the relevant quantity is opportunity volume ($P$, or betweenness); if the
firm needs a partner whose position cannot be bypassed, it is exclusivity ($Q$).

In business-to-business markets the evidence again favors balance over either pure
position. In an event study of 230 marketing-alliance announcements in the software
industry, the efficiency and density of a firm's alliance network had their strongest
positive effects on abnormal returns at moderate levels, while network centrality and
reputation had none [@swaminathan2009a]. What a firm gains from an alliance also
depends on the firm itself: in onshore drilling, slow firms gain the most speed from
partnering and keep the gain only with prior partnering experience [@li2026speed], so a
partner's network position and the focal firm's own capabilities jointly set the return
(@sec-ms-corpdev). In new product alliances, knowledge redundancy
diminishes information acquisition but enhances information utilization
[@rindfleisch2001b]. Inside the selling organization, relationship managers perform
best when wide-reaching, nonoverlapping ties in their informal network are combined
with a densely interconnected formal network, and the effects grow with the overlap
between the two networks [@gonzalez2014synergistic]. Salespeople's relational and
positional centrality in the intraorganizational network both contribute to objective
sales performance, but political skill predicts only the former
[@bolander2015socialnetworks].

For channel intermediaries, distributors, agencies, and platforms, @ryall2007brokers
supply the economic logic and $Q$ supplies the measure. An intermediary earns rents from
its position only while the parties it connects lack substitutes for the connection,
and the arrival of a rival intermediary, or of a direct route between buyer and seller,
erodes those rents without any change in the intermediary's own ties
[@borgatti2026beyond]. Because interorganizational networks such as alliances,
syndicates, and buyer-supplier relations are usually observed sociocentrically, the
common-neighbor counts that $Q$ requires can be computed on the full network rather than
approximated from ego-network surveys [@borgatti2026beyond]. The usual caution applies
with special force. Positions are not randomly assigned [@zaheer2009origins], so a
brokerage measure entered as a regressor is a generated feature subject to the
selection and identification problems of @sec-graph-identification, and the temporal
brokerage of @mell2021brokerage is a reminder that the relation on which a gap is
defined is itself a theoretical choice.

## Community Detection {#sec-graph-communities}

Real marketing graphs are not homogeneous. They decompose into **communities**:
subgroups of nodes more densely connected to one another than to the rest of the graph.
In a social network communities are friend groups, demographic clusters, or interest
tribes; in a co-purchase graph they are product categories or use-occasion bundles; in
a brand co-mention graph they are competitive sets. Detecting communities is an
unsupervised problem analogous to clustering or topic modeling, and it serves the same
ends: segmentation, market-structure mapping, and the construction of group-level
features.

The dominant family of methods maximizes **modularity**, a score that compares the
fraction of edges falling within proposed communities to the fraction expected if edges
were placed at random while preserving the degree sequence:
$$
Q = \frac{1}{2m} \sum_{i,j} \left( A_{ij} - \frac{d_i d_j}{2m} \right) \delta(c_i, c_j),
$$ {#eq-graph-modularity}
where $m$ is the number of edges, $d_i$ is node $i$'s degree, $c_i$ is its assigned
community, and $\delta$ is one when two nodes share a community and zero otherwise.
High modularity means communities capture far more internal edge density than chance.
The **Louvain** algorithm (Blondel, Guillaume, Lambiotte and Lefebvre 2008, *Journal of
Statistical Mechanics* P10008, doi:10.1088/1742-5468/2008/10/P10008) greedily and
hierarchically optimizes modularity and scales to graphs with millions of nodes, which
made it the field standard. Its successor, the **Leiden** algorithm (Traag, Waltman and
van Eck 2019, *Scientific Reports* 9, 5233, doi:10.1038/s41598-019-41695-z), repairs a
known defect in Louvain, that it can produce internally disconnected communities, and is
now the recommended default. Both are exposed in `igraph` through the `cluster_*`
family.

```{r graph-communities, message=FALSE, warning=FALSE}
#| fig-cap: "Community structure of a simulated graph with planted groups. Nodes are colored by detected community. The Louvain partition recovers the dense subgroups; modularity quantifies how much denser within-community ties are than chance."
library(igraph)
set.seed(20260622)

# --- Simulate a graph with PLANTED community structure ---
# Stochastic block model: 4 blocks (think 4 customer tribes / 4 product
# categories) with high within-block and low between-block edge probability.
block_sizes <- c(25, 25, 25, 25)
p_within  <- 0.18
p_between <- 0.01
pref <- matrix(p_between, nrow = 4, ncol = 4)
diag(pref) <- p_within
gsbm <- sample_sbm(n = sum(block_sizes),
                   pref.matrix = pref,
                   block.sizes = block_sizes)

# --- Louvain community detection (modularity maximization) ---
comm_louvain <- cluster_louvain(gsbm)

cat("Number of communities found:", length(comm_louvain), "\n")
cat("Community sizes:", paste(sizes(comm_louvain), collapse = ", "), "\n")
cat("Modularity Q:", round(modularity(comm_louvain), 3), "\n")

# Leiden is available in recent igraph; fall back to Louvain if not present,
# so the chapter renders across versions.
leiden_ok <- "cluster_leiden" %in% ls("package:igraph")
if (leiden_ok) {
  comm_leiden <- cluster_leiden(gsbm, objective_function = "modularity")
  cat("Leiden communities:", length(unique(membership(comm_leiden))), "\n")
}

# --- Visualize the Louvain partition ---
plot(comm_louvain, gsbm,
     vertex.size  = 6,
     vertex.label = NA,
     edge.color   = "grey80",
     layout       = layout_with_fr(gsbm))
```

A modularity score comfortably above zero (values around 0.3 to 0.7 are typical of
clear community structure) confirms that the detected partition captures genuine density
rather than noise. The marketing payoff is that the community label is now a feature:
it can segment a campaign, define a competitive set, or, as the next section warns,
serve as a (dangerous) proxy for the shared environment that confounds influence
estimation.

## Contagion, Homophily, and the Identification Challenge {#sec-graph-identification}

The most consequential marketing question one can ask of a network is causal: does a
consumer adopt *because* their neighbors adopted? If yes, the firm can engineer
cascades by seeding well-chosen nodes; if no, network targeting buys nothing beyond
ordinary lookalike targeting. This is also the question on which naive analysis fails
most spectacularly, and the failure has a precise structure.

Observed clustering of behavior in a network has at least three distinct generators that
are observationally similar in cross-sectional data. **Contagion** (or *induction*,
*influence*) is the causal mechanism of interest: $i$'s adoption changes $j$'s adoption
probability. **Homophily** is selection on similarity: $i$ and $j$ are connected
*because* they were already alike, so they adopt the same things for reasons that
predate and have nothing to do with the tie. **Common external shocks** (correlated
environment) are exposures shared by connected nodes, such as a local promotion, a
weather event, or a price change, that hit neighborhoods together. All three produce the
same surface pattern, that adoption is correlated across edges, yet only the first
justifies network-based seeding. Separating them is the central methodological problem
of the field: @aral2009distinguishing show, using dynamic matched sampling on a
large communication network, that homophily can account for more than half of what
naive models attribute to contagion, so that ignoring it overstates influence by a
factor of two or more.

The econometric core of the difficulty is the **reflection problem** named by Manski
(Manski 1993, *Review of Economic Studies* 60(3), 531--542, doi:10.2307/2298123). If
each person's behavior depends on the average behavior of their reference group, and the
reference group's behavior simultaneously depends on each person's, then the endogenous
social effect (true contagion), the exogenous or contextual effect (the group's
characteristics), and correlated effects (shared environment) cannot in general be
separated from a single cross-section: the regressors reflect each other. Manski's
result is a statement that the data, not merely the estimator, are uninformative about
the decomposition without further structure.

Three routes restore identification. The first is **experimental**: randomize the
treatment that is supposed to spread, breaking the link between a node's exposure and
its pre-existing similarity to its neighbors. The cleanest marketing demonstrations are
randomized field experiments. @aral2011creating randomized viral product-design features inside
a Facebook app and so measured peer influence free of homophily; in a companion study,
@aral2012identifying
identified *influential* and *susceptible* individuals by exploiting that
randomization, showing that influence and susceptibility are distinct traits and that
the most influential are not the most susceptible; and @aral2014tie further showed, in a
large networked experiment, how tie strength and structural embeddedness moderate how
far influence travels. Bapna and Umyarov (2015, *Management
Science* 61(8), 1902--1920, doi:10.1287/mnsc.2014.2059) ran a randomized field
experiment gifting premium subscriptions in a music social network and recovered a clean
causal peer-influence estimate, finding it substantial but far below naive
observational figures. Centola (2010, *Science* 329(5996), 1194--1197,
doi:10.1126/science.1185231) built artificial online networks with randomly assigned
topologies and showed that health behavior spread *farther and faster* in clustered
networks than in random ones, evidence for *complex contagion*, where adoption requires
reinforcement from multiple sources rather than a single contact. The complex-contagion
finding directly qualifies seeding advice: if reinforcement is needed, scattering seeds
to maximize reach can be worse than concentrating them to manufacture local density.

The second route is **observational with structure**. When experiments are infeasible,
identification can sometimes be recovered from the *architecture* of the network itself.
@bramoulle2009identification showed that when the network contains *intransitive
triads* (friends of friends who are not themselves friends), the characteristics of
those second-degree-only contacts serve as instruments that break the reflection
problem, partially identifying endogenous from contextual effects. Within marketing,
@iyengar2011opinion combined sociometric prescription data with self-reports to separate
opinion leadership from contagion in new-product diffusion, @iyengar2015 separated the
roles of opinion leadership and susceptibility in new-product
trial and repeat, and @iyengar2003 and the broader contagion literature deploy timing,
panel structure, and dynamic models to argue that influence, not merely similarity,
drives observed clustering. @peng2018 exploited *network overlap*, the extent to which
two users share common friends, to study content sharing, a structural feature that
itself shapes diffusion.

A third route matches on the *content* that travels rather than on the nodes. When the
object of transmission is a discrete, datable item—a paper, a design, a piece of
code—one can sometimes find near-duplicates of it and ask which version a given actor
picked up. @arts2026harnessing do this at the scale of the U.S. patent record,
identifying **paper twins**, pairs of scientific papers reporting nearly identical
findings published at nearly the same time by independent academic teams, and asking
which twin a corporate patent cites. Because the twins carry the same knowledge with
the same commercial value at the same moment, the quality of the idea cannot explain
the choice, and what remains is the citing inventor's social distance to each set of
authors—which turns out to matter only within two degrees of separation. The design is
the network analogue of a matched pair: rather than randomizing the tie or instrumenting
it, it holds the transmitted content fixed and lets the tie vary. Its scope condition is
that twins must exist and be findable, which is why it appears in science and patenting,
where the corpus is large enough that near-simultaneous duplicate discovery is common,
rather than in consumer settings where the adopted object is rarely duplicated.

@fig-net-identification shows why the naive reading fails and how the three routes
restore identification.

```{mermaid}
%%| label: fig-net-identification
%%| fig-cap: "The identification problem for social influence. Contagion, homophily, and common external shocks all produce the same observed clustering, the reflection problem blocks their separation from a single cross-section, and only randomization, architecture-based structure, or a content-matched comparison recovers identified contagion."
flowchart LR
  CO["Contagion<br/>(influence)"] --> OB["Observed clustering<br/>(adoption correlated across ties)"]
  HO["Homophily<br/>(selection on similarity)"] --> OB
  CS["Common external shocks<br/>(shared environment)"] --> OB
  OB --> RP["Reflection problem<br/>(not separable from a single cross-section)"]
  RP --> EX["Randomized experiment"]
  RP --> STR["Observational with structure<br/>(architecture-based instrument)"]
  RP --> CM["Content-matched comparison<br/>(near-duplicate transmitted objects)"]
  EX --> ID["Identified contagion<br/>(justifies seeding)"]
  STR --> ID
  CM --> ID
```

The discipline this section imposes is non-negotiable. A centrality score, a community
label, or a "fraction of neighbors who adopted" is a generated regressor saturated with
homophily and shared-environment variance. Reporting its coefficient as social influence
is the network analogue of the generated-regressor error that recurs throughout the
unstructured-data chapters [@balducci2018unstructured], and here it is at its most seductive because the story, "my
friends made me do it," is so plausible. The credible default is to treat any
observational network-effect estimate as an upper bound and to reach for randomization,
a structural model, or an architecture-based instrument before making a causal claim.

## Diffusion and the Independent-Cascade Process {#sec-graph-diffusion}

Whether or not contagion is cleanly identified, marketers need *models* of how something
spreads over a fixed graph, both to forecast a campaign's reach and to choose seeds.
Two classical families dominate. In the **independent cascade** model, each newly
activated node gets a single chance to activate each not-yet-active neighbor,
succeeding with an edge-specific probability $p_{ij}$; activation is a one-shot
push along each edge. In the **linear threshold** model, each node activates once the
weighted fraction of its already-active neighbors crosses a node-specific threshold,
which is the discrete analogue of the complex-contagion reinforcement Centola observed.
Both nest the aggregate-level Bass diffusion logic of @goldenberg2001 at the micro,
edge-by-edge level.

These micro-models are also the substrate of **influence maximization**, the
combinatorial problem of choosing a seed set of size $k$ that maximizes expected
eventual adoption. Kempe, Kleinberg and Tardos (2003, *Proc. 9th ACM SIGKDD*, 137--146,
doi:10.1145/956750.956769) proved that expected spread under both models is a
*submodular* (diminishing-returns) function of the seed set, so a simple greedy
algorithm, repeatedly adding the node with the largest marginal gain, achieves a
spread within a $(1 - 1/e) \approx 63\%$ factor of the optimum. This is the theoretical
backbone of principled seeding and the formal counterpart to the empirical seeding
comparisons in @hinz2011.

The simulation below runs an independent-cascade process over the scale-free graph from
the centrality section and compares three seeding rules, highest-degree, highest-
betweenness, and random, by Monte Carlo over many cascade realizations. It is a small,
self-contained, genuinely running implementation.

```{r graph-diffusion, message=FALSE, warning=FALSE}
#| fig-cap: "Independent-cascade diffusion over a simulated scale-free graph. Each bar is the mean final number of adopters (out of 80 nodes) across 200 cascade simulations, for three seeding rules with the same seed-set size. Degree- and betweenness-based seeding outperform random seeding, the simulated counterpart of the empirical 'seed the hubs' result."
library(igraph)
set.seed(20260622)

g <- sample_pa(n = 80, power = 1.1, m = 2, directed = FALSE)
A <- as_adjacency_matrix(g, sparse = FALSE)
n <- vcount(g)

# --- One independent-cascade realization from a given seed set ---
# Each active node gets ONE chance to activate each inactive neighbor,
# succeeding with probability p along that edge.
ic_spread <- function(adj, seeds, p = 0.12) {
  active   <- rep(FALSE, nrow(adj)); active[seeds] <- TRUE
  frontier <- seeds
  while (length(frontier) > 0) {
    new_active <- integer(0)
    for (u in frontier) {
      nbrs <- which(adj[u, ] == 1 & !active)
      if (length(nbrs) > 0) {
        success <- nbrs[runif(length(nbrs)) < p]
        new_active <- c(new_active, success)
      }
    }
    new_active <- unique(new_active)
    active[new_active] <- TRUE
    frontier <- new_active
  }
  sum(active)
}

# --- Monte Carlo: mean spread for each seeding rule ---
mc_spread <- function(seeds, reps = 200, p = 0.12) {
  mean(replicate(reps, ic_spread(A, seeds, p)))
}

k <- 3  # seed-set size
deg  <- degree(g)
btw  <- betweenness(g)

seeds_degree <- order(deg, decreasing = TRUE)[1:k]
seeds_betw   <- order(btw, decreasing = TRUE)[1:k]
set.seed(101)
seeds_random <- sample(seq_len(n), k)

results <- c(
  Degree      = mc_spread(seeds_degree),
  Betweenness = mc_spread(seeds_betw),
  Random      = mc_spread(seeds_random)
)

cat("Mean final adopters out of", n, "nodes (k =", k, "seeds, 200 sims):\n")
print(round(results, 1))

barplot(results,
        ylab = "Mean final adopters (of 80)",
        col  = c("#1f3b73", "#3f6fb0", "#cdd8e8"),
        main = "")
```

The simulation reproduces the qualitative seeding result in a controlled setting:
targeting high-centrality nodes meaningfully outperforms random seeding for the same
budget, because hubs reach more of the graph before the cascade dies. The size of the
advantage depends on the activation probability $p$ and the network topology, and the
reader is invited to vary both. The honest caveat is that this gain is a property of the
*assumed* contagion process; whether it materializes in a real campaign returns us to
the identification problem of the previous section, since a process that is mostly
homophily will not respond to seeding at all.

## Graph Embeddings and Graph Neural Networks {#sec-graph-embeddings}

Centrality and community labels are hand-designed features. The learned-representation
revolution that reshaped text and image analysis has a direct counterpart for graphs:
methods that *learn* a low-dimensional vector for each node such that nodes close in the
graph are close in the vector space, after which the vectors feed any ordinary
downstream model (a churn classifier, a recommender, a link predictor). The conceptual
move mirrors word embeddings exactly, with random walks on the graph playing the role
that sentences play in text.

**Random-walk embeddings.** The foundational methods generate many random walks from
each node, treat each walk as a "sentence" of node IDs, and run a word2vec-style
skip-gram model so that nodes appearing in the same walks receive similar vectors.
*DeepWalk* (Perozzi, Al-Rfou and Skiena 2014, *Proc. 20th ACM SIGKDD*, 701--710,
doi:10.1145/2623330.2623732) introduced this idea. *node2vec* (Grover and Leskovec 2016,
*Proc. 22nd ACM SIGKDD*, 855--864, doi:10.1145/2939672.2939754) generalized it with
biased random walks whose two parameters interpolate between breadth-first exploration,
which captures *structural-role* similarity such as "both are hubs," and depth-first
exploration, which captures *community* proximity. The practical value for marketing is
that a single 64- or 128-dimensional node vector compresses a consumer's entire
relational position into a feature usable in standard pipelines, sidestepping the need to
pre-commit to degree versus betweenness versus community as the relevant summary.

**Graph neural networks (conceptual).** The following is a *conceptual* description; the
heavy deep-learning machinery and its training are beyond this chapter's runnable scope
and are flagged as such. Graph neural networks (GNNs) generalize embeddings from a fixed
lookup table to a learned function of node and edge *attributes*. The core operation is
**neighborhood aggregation** (message passing): each node updates its representation by
combining its own features with a permutation-invariant aggregate (sum, mean, or
attention-weighted average) of its neighbors' representations, and stacking $L$ such
layers lets information flow from $L$ hops away. The graph convolutional network
formulates this as a spectral-motivated averaging over neighbors (Kipf and Welling,
"Semi-Supervised Classification with Graph Convolutional Networks," ICLR 2017, cited by
name; no canonical Crossref DOI). *GraphSAGE* (Hamilton, Ying and Leskovec,
"Inductive Representation Learning on Large Graphs," NeurIPS 2017, cited by name; no
canonical Crossref DOI) made the scheme *inductive*, learning an aggregation function
that generalizes to nodes unseen at training time, which is essential when the customer
graph grows daily. Graph attention networks weight neighbors by learned attention rather
than treating them uniformly. The marketing relevance is that GNNs jointly exploit graph
structure *and* node attributes (a customer's demographics, a product's category) in a
single supervised objective, which is why they now underpin large-scale industrial
recommendation and fraud systems.

The standard caution applies with full force to learned representations. A node
embedding is a generated feature, and its proximity structure encodes homophily and
shared environment along with whatever genuine influence exists. Using embeddings to
*predict* (who will churn, what to recommend) is on firm ground; using an embedding
coordinate as a regressor in a *causal* claim about social influence reintroduces every
hazard of @sec-graph-identification in a less transparent form.

## Industry and Production Practice {#sec-graph-production}

In production systems, graph methods are deployed overwhelmingly for *prediction and
ranking*, where the identification anxieties recede and the engineering challenges of
scale, latency, and freshness dominate.

**Recommendation.** The co-purchase and user-item interaction graph is the oldest and
largest commercial application. Item-to-item collaborative filtering computes, for each
item, the items most co-purchased or co-viewed with it, precisely the neighborhood
structure of the co-purchase graph, and serves them as "related items" (Linden, Smith
and York 2003, doi:10.1109/MIC.2003.1167344). Modern systems replace hand-built
similarity with learned node embeddings or GNNs over the bipartite user-item graph, and
graph-based candidate generation is now standard at web scale. Because the objective is
relevance rather than a causal effect, the right evaluation is held-out predictive
ranking (precision-at-k, recall, NDCG), not a structural parameter.

**Fraud and abuse detection.** Fraud is intrinsically relational: fraudulent accounts
collude, share devices and payment instruments, and form dense, anomalous subgraphs that
look nothing like organic user clusters. Community detection surfaces suspiciously
tight rings; centrality and *guilt-by-association* propagation spread suspicion from
known-bad nodes to their neighbors; and GNNs trained on labeled fraud generalize these
signals. The same machinery flags fake reviews, click farms, and bot-driven engagement,
which directly protects the integrity of the word-of-mouth and influencer signals that
the rest of marketing relies upon.

**Influencer selection.** Brands choosing influencers operationalize the centrality and
seeding ideas of this chapter, but production practice is more skeptical of raw reach
than the naive theory. Follower count is in-degree, the crudest centrality, and is
easily inflated by bots and cheaply purchased. Practitioners increasingly weight
*engagement* (whether the audience acts), *audience overlap* (whether two influencers
reach the same people, a network-overlap consideration in the spirit of @peng2018), and
*authenticity*, all of which are graph- and behavior-based rather than reach-based. The
@watts2007influentials caution, that ordinary susceptible audiences often matter more
than rare special individuals, and the @trusov2010 distinction between activity-driving
influence and mere connectedness, are the academic backbone of this more nuanced
practice. The video-advertising and platform chapters (@sec-platforms) develop the
influencer economy in its own right; here the point is narrowly that selection is a
network-position problem dressed in marketing clothes.

A cross-cutting production reality is that these graphs are enormous, dynamic, and often
privacy-sensitive. The relevant edges, who is friends with whom, who bought with whom,
may be governed by consent and platform policy, and the analyst's access to the true
graph is partial and changing. Methods that are inductive (generalizing to new nodes),
incremental (updating without full recomputation), and robust to missing edges are
favored not for elegance but for survival in deployment.

## The Frontier {#sec-graph-frontier}

Several directions are actively expanding what network data can do for marketing.

**Causal inference on networks** is the most consequential. The field is moving beyond
the simple contagion-versus-homophily dichotomy toward formal treatment of *interference*
and *spillovers*, the fact that in a connected population the stable-unit-treatment-value
assumption fails because one unit's treatment affects another's outcome. Designing
experiments that *measure* spillovers rather than assuming them away, through
cluster-randomized, graph-cluster, and ego-network designs, is an area where marketing,
platform experimentation, and econometrics are converging, and it is the natural
extension of the identification discipline in @sec-graph-identification.

**Dynamic and temporal graphs** treat edges as events with timestamps rather than a
static snapshot, letting models capture *when* a tie forms or fires, which is exactly the
information diffusion and influence estimation most need. Temporal GNNs and point-process
models over interaction streams are the methodological frontier here.

**Exponential random graph models (ERGMs)** model the *network itself* as the outcome,
specifying the probability of an observed graph as a function of local configurations
(edges, reciprocated ties, triangles, homophily terms), and so let an analyst test
*why* ties form, for example whether brand co-mentions cluster by category beyond
chance. The mature `ergm` software (Hunter, Handcock, Butts, Goodreau and Morris 2008,
*Journal of Statistical Software* 24(3), doi:10.18637/jss.v024.i03) makes these models
estimable, and they are the right tool when link *formation*, rather than node behavior,
is the question, complementing the strategic link-formation models of @katona2008.

**Multimodal graphs** fuse the relational structure of this chapter with the text,
image, and behavioral modalities of its sibling chapters: nodes carry text and image
attributes, edges carry interaction content, and the joint object is modeled with
attributed GNNs. This is where network data rejoins the broader unstructured-and-
multimodal program, and where the generated-feature caution must be enforced across all
modalities at once. The unifying thesis of the part holds here without exception: a
graph, like a document or a photograph, becomes a learned, lossy feature vector, and
that vector is a generated regressor whose error may correlate with the outcome the
analyst is trying to explain.

## Summary {#sec-graph-summary}

- Network data make the *relation* the unit of analysis. A node's position in a graph
  is a marketing feature, and the configuration of the graph is information that
  rectangular data cannot recover.
- The applications span social contagion and word of mouth, co-purchase and
  recommendation graphs, customer-referral and seeding networks, influencer
  identification, and brand co-mention market-structure maps.
- **Centrality** measures (degree, betweenness, eigenvector) each formalize a different
  theory of which node matters, and they can disagree; choosing one is choosing a theory
  of influence, as the runnable `igraph` demonstration shows.
- **Brokerage** is also measured locally. Burt's constraint and effective size summarize
  the structural holes in an ego network, the Gould-Fernandez roles tie brokerage to
  group boundaries, and the returns to brokerage depend on the outcome, the content of
  ties, and competition among brokers. Because constraint largely tracks inverse degree
  and can improve as rival brokers appear, the pair-based measures of @borgatti2026beyond
  separate opportunity volume ($P$) from exclusivity ($Q$); the runnable demonstration
  reproduces their shadow-ego result (@sec-graph-brokerage).
- **Community detection** via modularity maximization (Louvain, Leiden) recovers
  cohesive subgroups usable as segments, competitive sets, or group-level features.
- The **contagion-versus-homophily** identification problem, formalized by Manski's
  reflection problem, is the hardest causal problem in network marketing; credible
  answers require randomization, architecture-based instruments, or structural models,
  and observational network-effect estimates should be read as upper bounds.
- **Diffusion** models (independent cascade, linear threshold) and the submodularity of
  influence spread justify principled seeding; the simulation confirms that centrality-
  based seeding beats random seeding *under an assumed contagion process*.
- **Graph embeddings** (DeepWalk, node2vec) and **graph neural networks** learn node
  representations that power production recommendation and fraud systems, but a learned
  embedding is a generated feature carrying the same causal hazards in less transparent
  form.
- In production, graphs serve recommendation, fraud detection, and influencer selection,
  where prediction quality and scale dominate; the frontier is causal inference under
  interference, temporal graphs, ERGMs for link formation, and multimodal attributed
  graphs.

## Further Reading

The word-of-mouth and social-contagion foundations in marketing are developed in
@goldenberg2001, @godes2004, and @godes2009, with the seeding and influence-targeting
strand in @hinz2011, @trusov2009, @trusov2010, and @iyengar2015. The strategic formation
of commercial networks is treated in @katona2008, network-structure effects on idea
generation in @stephen2016, and network overlap and diffusion in @peng2018. The
misinformation-as-diffusion case is @vosoughi2018. For the identification machinery, the
canonical references cited by verified DOI in the text are Manski (1993) on the
reflection problem, Bramoullé, Djebbari and Fortin (2009) on identification through
network architecture, and the randomized-experiment line of Aral and Walker (2011,
2012), Bapna and Umyarov (2015), and Centola (2010); the homophily concept traces to
McPherson, Smith-Lovin and Cook (2001, *Annual Review of Sociology* 27, 415--444,
doi:10.1146/annurev.soc.27.1.415). The methods substrate, community detection (Blondel
et al. 2008; Traag et al. 2019), influence maximization (Kempe, Kleinberg and Tardos
2003), graph embeddings (Perozzi et al. 2014; Grover and Leskovec 2016), graph neural
networks (Kipf and Welling 2017; Hamilton, Ying and Leskovec 2017, both cited by name),
ERGMs (Hunter et al. 2008), and the Amazon item-to-item recommender (Linden, Smith and
York 2003), is referenced inline with verified DOIs where a canonical Crossref DOI
exists. This chapter sits alongside the text (@sec-text-as-data) and image
(@sec-image-processing) treatments in the unstructured-and-multimodal family, and its
platform and influencer-economy context is developed in @sec-platforms.

For brokerage and structural holes, the foundations are @granovetter1973,
@coleman1988social, @burt1992structural, and @gould1989structures, and the reviews by
@stovel2012brokerage and @burt2013advantage map the field. The measurement debate runs
from @everett2020unpacking through @johnson2026reassessing to @borgatti2026beyond, the
contingent returns to brokerage are documented in @burt2004goodideas,
@ahuja2000collaboration, @shipilov2008cake, and @aral2011diversity, and the competitive
logic of brokerage is formalized by @ryall2007brokers and @buskens2008dynamics.
Marketing applications of network position include @swaminathan2009a on alliance
networks and @gonzalez2014synergistic on relationship managers' networks.
