# Networks and Graph Data {#sec-network-graph}

Most of the data structures the preceding chapters relied on are *rectangular*: a
customer is a row, an attribute is a column, and rows are treated as exchangeable draws
from a population. Network data violate that convenience at the root. The unit of
analysis is not the isolated individual but the *relation* between individuals, and the
informative quantity is the pattern of relations across the whole system. A customer
who buys, adopts, churns, or recommends does so embedded in a web of friends,
co-purchasers, followers, and referrers, and that embedding carries marketing signal
that no amount of demographic enrichment recovers. The object that encodes this
structure is a **graph**: a set of *nodes* (consumers, products, brands, sellers) and a
set of *edges* (friendships, co-purchases, referrals, co-mentions) connecting them.

This chapter treats the graph as a first-class marketing data type, parallel to text
(@sec-text-as-data) and images (@sec-image-processing) in the unstructured-and-
multimodal family. As with those modalities, the analytic arc is the same: an
unstructured artifact (here, a relational structure rather than a document or a photo)
is turned into machine-readable features, and those features are then used for
description, prediction, or causal claims. And as with those modalities, the central
econometric caution recurs in a network-specific and unusually severe form: a feature
extracted from the graph, such as a node's centrality or its neighbors' behavior, is a
*generated regressor* whose error is correlated with the outcome in ways that mimic the
very effect the analyst wants to measure. The identification of social influence is the
hardest version of this problem in all of empirical marketing, and it earns its own
section below.

The chapter proceeds from use to method to practice. It opens with the marketing
*applications* that motivate graph thinking. It then develops the formal
*representation* of a graph and the *centrality* measures that summarize a node's
position, with a runnable `igraph` demonstration on simulated data. It treats
*community detection*, the unsupervised discovery of cohesive subgroups. It then
confronts the *contagion-versus-homophily* identification problem head on, because no
honest account of network marketing can avoid it. It develops *diffusion* models and
simulates an independent-cascade process over a graph. It surveys the learned-
representation frontier, *graph embeddings* and *graph neural networks*, keeping the
deep-learning specifics conceptual but clearly labeled. It closes with *industry and
production* practice (recommendation, fraud, influencer selection) and a *frontier*
section on where the field is moving.

## Why Networks Matter in Marketing {#sec-graph-applications}

The marketing relevance of graph structure is not a single idea but a family of them,
unified by the premise that a consumer's value, behavior, and reachability depend on
the consumers around them.

**Social networks, word of mouth, and contagion.** The oldest motivation is that
products spread through social ties. A consumer is more likely to adopt when adopters
sit nearby in the social graph, whether because of genuine influence, shared exposure,
or similarity. The word-of-mouth literature in marketing has long modeled this spread
formally; @goldenberg2001 used a complex-systems simulation of the underlying
person-to-person process to show how the structure of "weak" and "strong" ties governs
the macro-level diffusion curve, and a large empirical literature, surveyed below,
estimates contagion from observed adoption sequences over real ties. The contagion
premise is also what makes *misinformation* a network phenomenon: @vosoughi2018 showed
that false news diffuses faster, farther, and more broadly than true news through the
Twitter retweet graph, a finding that is fundamentally about diffusion structure rather
than content alone.

**Co-purchase and recommendation graphs.** A second, entirely different graph arises
without any social tie at all. Connect two products when they are frequently bought
together, or connect a customer to every product they purchased, and the resulting
co-purchase (or bipartite customer-product) graph is the substrate of modern
recommendation. "Customers who bought this also bought that" is a statement about the
neighborhood of a node in a co-purchase graph, and item-to-item collaborative
filtering, the algorithm that powered Amazon's recommendations, is literally a
computation over this graph (Linden, Smith and York 2003, *IEEE Internet Computing*
7(1), 76--80, doi:10.1109/MIC.2003.1167344). The graph here is a tool for *prediction
and retrieval*, and the identification anxieties of the social-contagion literature are
largely absent.

**Customer-referral networks and seeding.** Referral programs make the graph a managed
asset: a firm rewards existing customers for recruiting new ones, and the chain of
who-referred-whom is a directed network whose structure determines program value. The
adjacent question, *whom to seed* with free product or early access to maximize spread,
turns network position into a targeting variable. @hinz2011 compared seeding strategies
empirically and found that seeding well-connected "hubs" outperforms seeding random or
"fringe" customers, a result that operationalizes centrality as a marketing decision
variable. The structure of referral and affiliate networks on the commercial web has
itself been modeled as an outcome of strategic link formation by @katona2008.

**Influencer identification.** Closely related is the problem of finding the
individuals whose activity moves the system. The naive answer, "the most connected
person," is both intuitive and contested. @trusov2010 developed a method to identify
genuinely influential users in an internet social network by modeling whose activity
predicts others' subsequent activity, distinguishing influence from mere connectedness.
@watts2007influentials (Watts and Dodds 2007, *Journal of Consumer Research* 34(4),
441--458, doi:10.1086/518527) challenged the "influentials hypothesis" itself,
arguing through simulation that large cascades are often driven by a critical mass of
easily influenced ordinary people rather than by rare special individuals, a caution
that remains central to how marketers should read centrality.

**Brand co-mention and semantic networks.** Finally, graphs need not connect people or
products at all. Connect two brands when they are mentioned together in reviews, social
posts, or search sessions, and the resulting co-mention network is a map of *market
structure*: tightly connected brand clusters are competitive sets, and a brand's
position reveals its substitutes and complements. This is the network-analytic cousin
of the text-mining market-structure maps discussed in @sec-text-as-data, and it lets an
analyst recover perceptual structure from unstructured co-occurrence rather than from
surveys.

Across these applications the same primitive recurs: a position in a graph is a
feature, and the configuration of the graph is information. The rest of the chapter
develops the methods that turn that information into measures.

## Graph Representation and Centrality {#sec-graph-centrality}

A graph $G = (V, E)$ is a set of nodes $V$ and a set of edges $E \subseteq V \times V$.
Edges may be **undirected** (a Facebook friendship is mutual) or **directed** (a
Twitter follow is one-way; an A-referred-B link has an arrow), and they may be
**weighted** (the number of co-purchases, the strength of a tie). The workhorse
algebraic representation is the **adjacency matrix** $A$, an $n \times n$ matrix with
$A_{ij} = 1$ (or the edge weight) when an edge runs from $i$ to $j$ and $0$ otherwise.
Almost every structural quantity in this chapter is a function of $A$. For sparse
real-world graphs, where each node connects to a tiny fraction of all others, $A$ is
stored as an edge list rather than a dense matrix, but the mathematics is unchanged.

The most basic summary of a node is its **degree**, the number of edges incident to it
($d_i = \sum_j A_{ij}$). In a directed graph this splits into *in-degree* (how many
point to $i$; a proxy for received attention or popularity) and *out-degree* (how many
$i$ points to; a proxy for activity). Degree is the crudest notion of importance, and
its crudeness is exactly the point of the richer **centrality** measures, each of which
formalizes a different theory of what makes a node matter.

**Betweenness centrality** measures how often a node lies on shortest paths between
other pairs of nodes:
$$
C_B(i) = \sum_{s \ne i \ne t} \frac{\sigma_{st}(i)}{\sigma_{st}},
$$ {#eq-graph-betweenness}
where $\sigma_{st}$ is the number of shortest paths from $s$ to $t$ and
$\sigma_{st}(i)$ the number passing through $i$. A high-betweenness node is a *broker*
sitting on the bridges between otherwise separated regions; in marketing terms, a
broker controls the flow of information across communities and is a candidate "weak tie"
connector in the sense of the word-of-mouth literature.

**Eigenvector centrality** formalizes the recursive idea that a node is important if its
neighbors are important. It is the leading eigenvector of the adjacency matrix:
$$
A\,\mathbf{x} = \lambda_{\max}\,\mathbf{x},
$$ {#eq-graph-eigenvector}
so a node's score is proportional to the sum of its neighbors' scores. PageRank is a
directed, damped variant of the same idea. Eigenvector centrality captures *prestige*:
being connected to other well-connected nodes, not merely to many nodes. The three
measures can disagree sharply, and the disagreement is informative. A node can have
high degree but low betweenness (a popular member buried inside one tight cluster), or
modest degree but high betweenness (a lone bridge between two cliques). Choosing a
centrality measure is therefore choosing a theory of influence, not a neutral
preprocessing step.

The runnable demonstration below builds a simulated graph and computes all three. The
package is the maintained `igraph`, the standard for network analysis in R. The graph
is generated by *preferential attachment* (the Barabási-Albert model), which produces
the heavy-tailed degree distribution characteristic of real social and co-purchase
networks: a few hubs and a long tail of sparsely connected nodes.

```{r graph-centrality, message=FALSE, warning=FALSE}
#| fig-cap: "A simulated scale-free social/co-purchase graph of 80 nodes generated by preferential attachment. Node size is proportional to degree; color encodes betweenness (darker = higher). The hubs (large nodes) and the brokers (dark nodes) are not always the same vertices, which is the practical point of distinguishing centrality measures."
library(igraph)
set.seed(20260622)

# --- Build a simulated scale-free network (Barabasi-Albert) ---
# Interpretable as a social graph (who-follows-whom) or, with edges read as
# "frequently bought together," a product co-purchase graph.
g <- sample_pa(n = 80, power = 1.1, m = 2, directed = FALSE)
V(g)$name <- paste0("u", seq_len(vcount(g)))

# --- Centrality measures (all functions of the adjacency structure) ---
deg  <- degree(g)
btw  <- betweenness(g, normalized = TRUE)
eig  <- eigen_centrality(g)$vector

cent <- data.frame(
  node       = V(g)$name,
  degree     = deg,
  betweenness = round(btw, 3),
  eigenvector = round(eig, 3)
)

# Show the five highest-degree nodes and how they rank on the other measures
top_deg <- head(cent[order(-cent$degree), ], 5)
cat("Top-5 nodes by degree:\n")
print(top_deg, row.names = FALSE)

# The top node by EACH measure can differ:
cat("\nMost central node by measure:\n")
cat("  degree     :", cent$node[which.max(cent$degree)], "\n")
cat("  betweenness:", cent$node[which.max(cent$betweenness)], "\n")
cat("  eigenvector:", cent$node[which.max(cent$eigenvector)], "\n")

# Rank correlation between degree and betweenness: high but not 1,
# which is exactly why brokers and hubs are distinct managerial targets.
cat("\nSpearman corr(degree, betweenness):",
    round(cor(deg, btw, method = "spearman"), 3), "\n")

# --- Visualize ---
node_col <- colorRampPalette(c("#d9e3f0", "#1f3b73"))(100)[
  as.integer(cut(btw, breaks = 100, include.lowest = TRUE))]
plot(g,
     vertex.size   = 4 + 14 * (deg / max(deg)),
     vertex.color  = node_col,
     vertex.label  = NA,
     edge.color    = "grey75",
     layout        = layout_with_fr(g))
```

The output makes the conceptual point operational. The Spearman correlation between
degree and betweenness is high but well below one, and the single most central node
differs across measures. A seeding or influencer program that targets "the most
connected user" is implicitly committing to the degree theory of influence; if the real
mechanism is brokerage across communities, betweenness is the right target, and the two
will recommend different people.

## Community Detection {#sec-graph-communities}

Real marketing graphs are not homogeneous. They decompose into **communities**:
subgroups of nodes more densely connected to one another than to the rest of the graph.
In a social network communities are friend groups, demographic clusters, or interest
tribes; in a co-purchase graph they are product categories or use-occasion bundles; in
a brand co-mention graph they are competitive sets. Detecting communities is an
unsupervised problem analogous to clustering or topic modeling, and it serves the same
ends: segmentation, market-structure mapping, and the construction of group-level
features.

The dominant family of methods maximizes **modularity**, a score that compares the
fraction of edges falling within proposed communities to the fraction expected if edges
were placed at random while preserving the degree sequence:
$$
Q = \frac{1}{2m} \sum_{i,j} \left( A_{ij} - \frac{d_i d_j}{2m} \right) \delta(c_i, c_j),
$$ {#eq-graph-modularity}
where $m$ is the number of edges, $d_i$ is node $i$'s degree, $c_i$ is its assigned
community, and $\delta$ is one when two nodes share a community and zero otherwise.
High modularity means communities capture far more internal edge density than chance.
The **Louvain** algorithm (Blondel, Guillaume, Lambiotte and Lefebvre 2008, *Journal of
Statistical Mechanics* P10008, doi:10.1088/1742-5468/2008/10/P10008) greedily and
hierarchically optimizes modularity and scales to graphs with millions of nodes, which
made it the field standard. Its successor, the **Leiden** algorithm (Traag, Waltman and
van Eck 2019, *Scientific Reports* 9, 5233, doi:10.1038/s41598-019-41695-z), repairs a
known defect in Louvain, that it can produce internally disconnected communities, and is
now the recommended default. Both are exposed in `igraph` through the `cluster_*`
family.

```{r graph-communities, message=FALSE, warning=FALSE}
#| fig-cap: "Community structure of a simulated graph with planted groups. Nodes are colored by detected community. The Louvain partition recovers the dense subgroups; modularity quantifies how much denser within-community ties are than chance."
library(igraph)
set.seed(20260622)

# --- Simulate a graph with PLANTED community structure ---
# Stochastic block model: 4 blocks (think 4 customer tribes / 4 product
# categories) with high within-block and low between-block edge probability.
block_sizes <- c(25, 25, 25, 25)
p_within  <- 0.18
p_between <- 0.01
pref <- matrix(p_between, nrow = 4, ncol = 4)
diag(pref) <- p_within
gsbm <- sample_sbm(n = sum(block_sizes),
                   pref.matrix = pref,
                   block.sizes = block_sizes)

# --- Louvain community detection (modularity maximization) ---
comm_louvain <- cluster_louvain(gsbm)

cat("Number of communities found:", length(comm_louvain), "\n")
cat("Community sizes:", paste(sizes(comm_louvain), collapse = ", "), "\n")
cat("Modularity Q:", round(modularity(comm_louvain), 3), "\n")

# Leiden is available in recent igraph; fall back to Louvain if not present,
# so the chapter renders across versions.
leiden_ok <- "cluster_leiden" %in% ls("package:igraph")
if (leiden_ok) {
  comm_leiden <- cluster_leiden(gsbm, objective_function = "modularity")
  cat("Leiden communities:", length(unique(membership(comm_leiden))), "\n")
}

# --- Visualize the Louvain partition ---
plot(comm_louvain, gsbm,
     vertex.size  = 6,
     vertex.label = NA,
     edge.color   = "grey80",
     layout       = layout_with_fr(gsbm))
```

A modularity score comfortably above zero (values around 0.3 to 0.7 are typical of
clear community structure) confirms that the detected partition captures genuine density
rather than noise. The marketing payoff is that the community label is now a feature:
it can segment a campaign, define a competitive set, or, as the next section warns,
serve as a (dangerous) proxy for the shared environment that confounds influence
estimation.

## Contagion, Homophily, and the Identification Challenge {#sec-graph-identification}

The most consequential marketing question one can ask of a network is causal: does a
consumer adopt *because* their neighbors adopted? If yes, the firm can engineer
cascades by seeding well-chosen nodes; if no, network targeting buys nothing beyond
ordinary lookalike targeting. This is also the question on which naive analysis fails
most spectacularly, and the failure has a precise structure.

Observed clustering of behavior in a network has at least three distinct generators that
are observationally similar in cross-sectional data. **Contagion** (or *induction*,
*influence*) is the causal mechanism of interest: $i$'s adoption changes $j$'s adoption
probability. **Homophily** is selection on similarity: $i$ and $j$ are connected
*because* they were already alike, so they adopt the same things for reasons that
predate and have nothing to do with the tie. **Common external shocks** (correlated
environment) are exposures shared by connected nodes, such as a local promotion, a
weather event, or a price change, that hit neighborhoods together. All three produce the
same surface pattern, that adoption is correlated across edges, yet only the first
justifies network-based seeding.

The econometric core of the difficulty is the **reflection problem** named by Manski
(Manski 1993, *Review of Economic Studies* 60(3), 531--542, doi:10.2307/2298123). If
each person's behavior depends on the average behavior of their reference group, and the
reference group's behavior simultaneously depends on each person's, then the endogenous
social effect (true contagion), the exogenous or contextual effect (the group's
characteristics), and correlated effects (shared environment) cannot in general be
separated from a single cross-section: the regressors reflect each other. Manski's
result is a statement that the data, not merely the estimator, are uninformative about
the decomposition without further structure.

Two routes restore identification. The first is **experimental**: randomize the
treatment that is supposed to spread, breaking the link between a node's exposure and
its pre-existing similarity to its neighbors. The cleanest marketing demonstrations are
randomized field experiments. Aral and Walker (2011, *Management Science* 57(9),
1623--1639, doi:10.1287/mnsc.1110.1421) randomized viral product-design features inside
a Facebook app and so measured peer influence free of homophily; in a companion study,
Aral and Walker (2012, *Science* 337(6092), 337--341, doi:10.1126/science.1215842)
identified *influential* and *susceptible* individuals by exploiting that
randomization, showing that influence and susceptibility are distinct traits and that
the most influential are not the most susceptible. Bapna and Umyarov (2015, *Management
Science* 61(8), 1902--1920, doi:10.1287/mnsc.2014.2059) ran a randomized field
experiment gifting premium subscriptions in a music social network and recovered a clean
causal peer-influence estimate, finding it substantial but far below naive
observational figures. Centola (2010, *Science* 329(5996), 1194--1197,
doi:10.1126/science.1185231) built artificial online networks with randomly assigned
topologies and showed that health behavior spread *farther and faster* in clustered
networks than in random ones, evidence for *complex contagion*, where adoption requires
reinforcement from multiple sources rather than a single contact. The complex-contagion
finding directly qualifies seeding advice: if reinforcement is needed, scattering seeds
to maximize reach can be worse than concentrating them to manufacture local density.

The second route is **observational with structure**. When experiments are infeasible,
identification can sometimes be recovered from the *architecture* of the network itself.
Bramoullé, Djebbari and Fortin (2009, *Journal of Econometrics* 150(1), 41--55,
doi:10.1016/j.jeconom.2008.12.021) showed that when the network contains *intransitive
triads* (friends of friends who are not themselves friends), the characteristics of
those second-degree-only contacts serve as instruments that break the reflection
problem, partially identifying endogenous from contextual effects. Within marketing,
@iyengar2015 separated the roles of opinion leadership and susceptibility in new-product
trial and repeat, and @iyengar2003 and the broader contagion literature deploy timing,
panel structure, and dynamic models to argue that influence, not merely similarity,
drives observed clustering. @peng2018 exploited *network overlap*, the extent to which
two users share common friends, to study content sharing, a structural feature that
itself shapes diffusion.

The discipline this section imposes is non-negotiable. A centrality score, a community
label, or a "fraction of neighbors who adopted" is a generated regressor saturated with
homophily and shared-environment variance. Reporting its coefficient as social influence
is the network analogue of the generated-regressor error that recurs throughout the
unstructured-data chapters, and here it is at its most seductive because the story, "my
friends made me do it," is so plausible. The credible default is to treat any
observational network-effect estimate as an upper bound and to reach for randomization,
a structural model, or an architecture-based instrument before making a causal claim.

## Diffusion and the Independent-Cascade Process {#sec-graph-diffusion}

Whether or not contagion is cleanly identified, marketers need *models* of how something
spreads over a fixed graph, both to forecast a campaign's reach and to choose seeds.
Two classical families dominate. In the **independent cascade** model, each newly
activated node gets a single chance to activate each not-yet-active neighbor,
succeeding with an edge-specific probability $p_{ij}$; activation is a one-shot
push along each edge. In the **linear threshold** model, each node activates once the
weighted fraction of its already-active neighbors crosses a node-specific threshold,
which is the discrete analogue of the complex-contagion reinforcement Centola observed.
Both nest the aggregate-level Bass diffusion logic of @goldenberg2001 at the micro,
edge-by-edge level.

These micro-models are also the substrate of **influence maximization**, the
combinatorial problem of choosing a seed set of size $k$ that maximizes expected
eventual adoption. Kempe, Kleinberg and Tardos (2003, *Proc. 9th ACM SIGKDD*, 137--146,
doi:10.1145/956750.956769) proved that expected spread under both models is a
*submodular* (diminishing-returns) function of the seed set, so a simple greedy
algorithm, repeatedly adding the node with the largest marginal gain, achieves a
spread within a $(1 - 1/e) \approx 63\%$ factor of the optimum. This is the theoretical
backbone of principled seeding and the formal counterpart to the empirical seeding
comparisons in @hinz2011.

The simulation below runs an independent-cascade process over the scale-free graph from
the centrality section and compares three seeding rules, highest-degree, highest-
betweenness, and random, by Monte Carlo over many cascade realizations. It is a small,
self-contained, genuinely running implementation.

```{r graph-diffusion, message=FALSE, warning=FALSE}
#| fig-cap: "Independent-cascade diffusion over a simulated scale-free graph. Each bar is the mean final number of adopters (out of 80 nodes) across 200 cascade simulations, for three seeding rules with the same seed-set size. Degree- and betweenness-based seeding outperform random seeding, the simulated counterpart of the empirical 'seed the hubs' result."
library(igraph)
set.seed(20260622)

g <- sample_pa(n = 80, power = 1.1, m = 2, directed = FALSE)
A <- as_adjacency_matrix(g, sparse = FALSE)
n <- vcount(g)

# --- One independent-cascade realization from a given seed set ---
# Each active node gets ONE chance to activate each inactive neighbor,
# succeeding with probability p along that edge.
ic_spread <- function(adj, seeds, p = 0.12) {
  active   <- rep(FALSE, nrow(adj)); active[seeds] <- TRUE
  frontier <- seeds
  while (length(frontier) > 0) {
    new_active <- integer(0)
    for (u in frontier) {
      nbrs <- which(adj[u, ] == 1 & !active)
      if (length(nbrs) > 0) {
        success <- nbrs[runif(length(nbrs)) < p]
        new_active <- c(new_active, success)
      }
    }
    new_active <- unique(new_active)
    active[new_active] <- TRUE
    frontier <- new_active
  }
  sum(active)
}

# --- Monte Carlo: mean spread for each seeding rule ---
mc_spread <- function(seeds, reps = 200, p = 0.12) {
  mean(replicate(reps, ic_spread(A, seeds, p)))
}

k <- 3  # seed-set size
deg  <- degree(g)
btw  <- betweenness(g)

seeds_degree <- order(deg, decreasing = TRUE)[1:k]
seeds_betw   <- order(btw, decreasing = TRUE)[1:k]
set.seed(101)
seeds_random <- sample(seq_len(n), k)

results <- c(
  Degree      = mc_spread(seeds_degree),
  Betweenness = mc_spread(seeds_betw),
  Random      = mc_spread(seeds_random)
)

cat("Mean final adopters out of", n, "nodes (k =", k, "seeds, 200 sims):\n")
print(round(results, 1))

barplot(results,
        ylab = "Mean final adopters (of 80)",
        col  = c("#1f3b73", "#3f6fb0", "#cdd8e8"),
        main = "")
```

The simulation reproduces the qualitative seeding result in a controlled setting:
targeting high-centrality nodes meaningfully outperforms random seeding for the same
budget, because hubs reach more of the graph before the cascade dies. The size of the
advantage depends on the activation probability $p$ and the network topology, and the
reader is invited to vary both. The honest caveat is that this gain is a property of the
*assumed* contagion process; whether it materializes in a real campaign returns us to
the identification problem of the previous section, since a process that is mostly
homophily will not respond to seeding at all.

## Graph Embeddings and Graph Neural Networks {#sec-graph-embeddings}

Centrality and community labels are hand-designed features. The learned-representation
revolution that reshaped text and image analysis has a direct counterpart for graphs:
methods that *learn* a low-dimensional vector for each node such that nodes close in the
graph are close in the vector space, after which the vectors feed any ordinary
downstream model (a churn classifier, a recommender, a link predictor). The conceptual
move mirrors word embeddings exactly, with random walks on the graph playing the role
that sentences play in text.

**Random-walk embeddings.** The foundational methods generate many random walks from
each node, treat each walk as a "sentence" of node IDs, and run a word2vec-style
skip-gram model so that nodes appearing in the same walks receive similar vectors.
*DeepWalk* (Perozzi, Al-Rfou and Skiena 2014, *Proc. 20th ACM SIGKDD*, 701--710,
doi:10.1145/2623330.2623732) introduced this idea. *node2vec* (Grover and Leskovec 2016,
*Proc. 22nd ACM SIGKDD*, 855--864, doi:10.1145/2939672.2939754) generalized it with
biased random walks whose two parameters interpolate between breadth-first exploration,
which captures *structural-role* similarity such as "both are hubs," and depth-first
exploration, which captures *community* proximity. The practical value for marketing is
that a single 64- or 128-dimensional node vector compresses a consumer's entire
relational position into a feature usable in standard pipelines, sidestepping the need to
pre-commit to degree versus betweenness versus community as the relevant summary.

**Graph neural networks (conceptual).** The following is a *conceptual* description; the
heavy deep-learning machinery and its training are beyond this chapter's runnable scope
and are flagged as such. Graph neural networks (GNNs) generalize embeddings from a fixed
lookup table to a learned function of node and edge *attributes*. The core operation is
**neighborhood aggregation** (message passing): each node updates its representation by
combining its own features with a permutation-invariant aggregate (sum, mean, or
attention-weighted average) of its neighbors' representations, and stacking $L$ such
layers lets information flow from $L$ hops away. The graph convolutional network
formulates this as a spectral-motivated averaging over neighbors (Kipf and Welling,
"Semi-Supervised Classification with Graph Convolutional Networks," ICLR 2017, cited by
name; no canonical Crossref DOI). *GraphSAGE* (Hamilton, Ying and Leskovec,
"Inductive Representation Learning on Large Graphs," NeurIPS 2017, cited by name; no
canonical Crossref DOI) made the scheme *inductive*, learning an aggregation function
that generalizes to nodes unseen at training time, which is essential when the customer
graph grows daily. Graph attention networks weight neighbors by learned attention rather
than treating them uniformly. The marketing relevance is that GNNs jointly exploit graph
structure *and* node attributes (a customer's demographics, a product's category) in a
single supervised objective, which is why they now underpin large-scale industrial
recommendation and fraud systems.

The standard caution applies with full force to learned representations. A node
embedding is a generated feature, and its proximity structure encodes homophily and
shared environment along with whatever genuine influence exists. Using embeddings to
*predict* (who will churn, what to recommend) is on firm ground; using an embedding
coordinate as a regressor in a *causal* claim about social influence reintroduces every
hazard of @sec-graph-identification in a less transparent form.

## Industry and Production Practice {#sec-graph-production}

In production systems, graph methods are deployed overwhelmingly for *prediction and
ranking*, where the identification anxieties recede and the engineering challenges of
scale, latency, and freshness dominate.

**Recommendation.** The co-purchase and user-item interaction graph is the oldest and
largest commercial application. Item-to-item collaborative filtering computes, for each
item, the items most co-purchased or co-viewed with it, precisely the neighborhood
structure of the co-purchase graph, and serves them as "related items" (Linden, Smith
and York 2003, doi:10.1109/MIC.2003.1167344). Modern systems replace hand-built
similarity with learned node embeddings or GNNs over the bipartite user-item graph, and
graph-based candidate generation is now standard at web scale. Because the objective is
relevance rather than a causal effect, the right evaluation is held-out predictive
ranking (precision-at-k, recall, NDCG), not a structural parameter.

**Fraud and abuse detection.** Fraud is intrinsically relational: fraudulent accounts
collude, share devices and payment instruments, and form dense, anomalous subgraphs that
look nothing like organic user clusters. Community detection surfaces suspiciously
tight rings; centrality and *guilt-by-association* propagation spread suspicion from
known-bad nodes to their neighbors; and GNNs trained on labeled fraud generalize these
signals. The same machinery flags fake reviews, click farms, and bot-driven engagement,
which directly protects the integrity of the word-of-mouth and influencer signals that
the rest of marketing relies upon.

**Influencer selection.** Brands choosing influencers operationalize the centrality and
seeding ideas of this chapter, but production practice is more skeptical of raw reach
than the naive theory. Follower count is in-degree, the crudest centrality, and is
easily inflated by bots and cheaply purchased. Practitioners increasingly weight
*engagement* (whether the audience acts), *audience overlap* (whether two influencers
reach the same people, a network-overlap consideration in the spirit of @peng2018), and
*authenticity*, all of which are graph- and behavior-based rather than reach-based. The
@watts2007influentials caution, that ordinary susceptible audiences often matter more
than rare special individuals, and the @trusov2010 distinction between activity-driving
influence and mere connectedness, are the academic backbone of this more nuanced
practice. The video-advertising and platform chapters (@sec-platforms) develop the
influencer economy in its own right; here the point is narrowly that selection is a
network-position problem dressed in marketing clothes.

A cross-cutting production reality is that these graphs are enormous, dynamic, and often
privacy-sensitive. The relevant edges, who is friends with whom, who bought with whom,
may be governed by consent and platform policy, and the analyst's access to the true
graph is partial and changing. Methods that are inductive (generalizing to new nodes),
incremental (updating without full recomputation), and robust to missing edges are
favored not for elegance but for survival in deployment.

## The Frontier {#sec-graph-frontier}

Several directions are actively expanding what network data can do for marketing.

**Causal inference on networks** is the most consequential. The field is moving beyond
the simple contagion-versus-homophily dichotomy toward formal treatment of *interference*
and *spillovers*, the fact that in a connected population the stable-unit-treatment-value
assumption fails because one unit's treatment affects another's outcome. Designing
experiments that *measure* spillovers rather than assuming them away, through
cluster-randomized, graph-cluster, and ego-network designs, is an area where marketing,
platform experimentation, and econometrics are converging, and it is the natural
extension of the identification discipline in @sec-graph-identification.

**Dynamic and temporal graphs** treat edges as events with timestamps rather than a
static snapshot, letting models capture *when* a tie forms or fires, which is exactly the
information diffusion and influence estimation most need. Temporal GNNs and point-process
models over interaction streams are the methodological frontier here.

**Exponential random graph models (ERGMs)** model the *network itself* as the outcome,
specifying the probability of an observed graph as a function of local configurations
(edges, reciprocated ties, triangles, homophily terms), and so let an analyst test
*why* ties form, for example whether brand co-mentions cluster by category beyond
chance. The mature `ergm` software (Hunter, Handcock, Butts, Goodreau and Morris 2008,
*Journal of Statistical Software* 24(3), doi:10.18637/jss.v024.i03) makes these models
estimable, and they are the right tool when link *formation*, rather than node behavior,
is the question, complementing the strategic link-formation models of @katona2008.

**Multimodal graphs** fuse the relational structure of this chapter with the text,
image, and behavioral modalities of its sibling chapters: nodes carry text and image
attributes, edges carry interaction content, and the joint object is modeled with
attributed GNNs. This is where network data rejoins the broader unstructured-and-
multimodal program, and where the generated-feature caution must be enforced across all
modalities at once. The unifying thesis of the part holds here without exception: a
graph, like a document or a photograph, becomes a learned, lossy feature vector, and
that vector is a generated regressor whose error may correlate with the outcome the
analyst is trying to explain.

## Summary {#sec-graph-summary}

- Network data make the *relation* the unit of analysis. A node's position in a graph
  is a marketing feature, and the configuration of the graph is information that
  rectangular data cannot recover.
- The applications span social contagion and word of mouth, co-purchase and
  recommendation graphs, customer-referral and seeding networks, influencer
  identification, and brand co-mention market-structure maps.
- **Centrality** measures (degree, betweenness, eigenvector) each formalize a different
  theory of which node matters, and they can disagree; choosing one is choosing a theory
  of influence, as the runnable `igraph` demonstration shows.
- **Community detection** via modularity maximization (Louvain, Leiden) recovers
  cohesive subgroups usable as segments, competitive sets, or group-level features.
- The **contagion-versus-homophily** identification problem, formalized by Manski's
  reflection problem, is the hardest causal problem in network marketing; credible
  answers require randomization, architecture-based instruments, or structural models,
  and observational network-effect estimates should be read as upper bounds.
- **Diffusion** models (independent cascade, linear threshold) and the submodularity of
  influence spread justify principled seeding; the simulation confirms that centrality-
  based seeding beats random seeding *under an assumed contagion process*.
- **Graph embeddings** (DeepWalk, node2vec) and **graph neural networks** learn node
  representations that power production recommendation and fraud systems, but a learned
  embedding is a generated feature carrying the same causal hazards in less transparent
  form.
- In production, graphs serve recommendation, fraud detection, and influencer selection,
  where prediction quality and scale dominate; the frontier is causal inference under
  interference, temporal graphs, ERGMs for link formation, and multimodal attributed
  graphs.

## Further Reading

The word-of-mouth and social-contagion foundations in marketing are developed in
@goldenberg2001, @godes2004, and @godes2009, with the seeding and influence-targeting
strand in @hinz2011, @trusov2009, @trusov2010, and @iyengar2015. The strategic formation
of commercial networks is treated in @katona2008, network-structure effects on idea
generation in @stephen2016, and network overlap and diffusion in @peng2018. The
misinformation-as-diffusion case is @vosoughi2018. For the identification machinery, the
canonical references cited by verified DOI in the text are Manski (1993) on the
reflection problem, Bramoullé, Djebbari and Fortin (2009) on identification through
network architecture, and the randomized-experiment line of Aral and Walker (2011,
2012), Bapna and Umyarov (2015), and Centola (2010); the homophily concept traces to
McPherson, Smith-Lovin and Cook (2001, *Annual Review of Sociology* 27, 415--444,
doi:10.1146/annurev.soc.27.1.415). The methods substrate, community detection (Blondel
et al. 2008; Traag et al. 2019), influence maximization (Kempe, Kleinberg and Tardos
2003), graph embeddings (Perozzi et al. 2014; Grover and Leskovec 2016), graph neural
networks (Kipf and Welling 2017; Hamilton, Ying and Leskovec 2017, both cited by name),
ERGMs (Hunter et al. 2008), and the Amazon item-to-item recommender (Linden, Smith and
York 2003), is referenced inline with verified DOIs where a canonical Crossref DOI
exists. This chapter sits alongside the text (@sec-text-as-data) and image
(@sec-image-processing) treatments in the unstructured-and-multimodal family, and its
platform and influencer-economy context is developed in @sec-platforms.
