# Clickstream and Behavioral Data {#sec-clickstream}

Every action a user takes inside a digital property leaves a trace. A page is
requested, a button is tapped, a product is viewed, a search is run, an item is added
to a cart, a video is scrubbed, a form is abandoned. Logged with a timestamp and a
user or device identifier, these traces accumulate into a *clickstream*: a long,
ordered record of *who did what, where, and when*. Behavioral log data of this kind is
the connective tissue of digital marketing. It sits upstream of nearly every metric a
firm reports, every recommendation it serves, and every attribution claim it makes
about which channel "caused" a sale. Where survey data asks consumers what they
intend, and panel data records what they ultimately bought, clickstream data records
the *process in between*: the navigation, hesitation, comparison, and backtracking
through which an intention becomes (or fails to become) a purchase
[@montgomery2004modeling; @bucklin2003].

Clickstream data is distinctive among marketing data in three respects, and each shapes
the methods this chapter develops. First, it is *sequential*. The order of events
carries information that any aggregation into counts destroys: a session that goes
home page, search, product, cart, checkout is a different object from one that visits
the same five pages in a different order, even though both produce identical page-view
totals. Second, it is *sparse and heavy-tailed* at the user level. Most visitors view
a handful of pages and never return; a small minority generates the bulk of events and
revenue, so distributions of session length, dwell time, and visit frequency are
governed by their tails rather than their means [@johnson2004depth]. Third, it is
*observational and self-selected*. Users choose what to click, and that choice is
correlated with unobserved intent, so a naive regression of conversion on "pages
viewed" recovers a mixture of behavior and latent motivation rather than a causal
effect. The methods that follow are, in large part, disciplined responses to these
three features.

This chapter develops clickstream analysis in the order a practitioner confronts it.
It begins with **applications**, the marketing questions that behavioral logs are asked
to answer. It then fixes the **data structure**, defining events, sessions, and
sequences, and confronts the deceptively consequential problem of *sessionization*,
illustrated on a simulated event log we carry through the rest of the chapter. From
there it develops the two classical modeling traditions: **path and Markov models**,
which treat navigation as a stochastic walk over states and yield a principled notion
of channel attribution through the *removal effect*; and **sequence analysis**, which
treats each user's trajectory as a categorical sequence to be aligned, clustered, and
visualized. It then surveys **deep sequence models** (recurrent and transformer
architectures) that learn representations of behavior directly from raw event streams,
returns to **attribution** as a problem in its own right, and closes with **industry
and production practice**, the event pipelines and feature engineering that turn raw
logs into model inputs without leaking the future into the past. Throughout, the worked
code is genuinely runnable: we simulate an event log and analyze it with the
`markovchain` and `TraMineR` packages. @fig-clk-models maps the three modeling
traditions onto the shared object they consume.

```{mermaid}
%%| label: fig-clk-models
%%| fig-cap: "The chapter's three modeling traditions. Each consumes the same object, the ordered action sequence recovered by sessionization, and differs in what it extracts from it."
flowchart TB
  LOG["Sessionized log:<br/>ordered action sequences"]
  subgraph MK["Path and Markov models"]
    M1["Transition probabilities<br/>between states"] --> M2["Removal-effect<br/>attribution"]
  end
  subgraph SA["Sequence analysis"]
    S1["Optimal-matching<br/>distances"] --> S2["Clusters of<br/>behavioral types"]
  end
  subgraph DP["Deep sequence models"]
    D1["RNN, LSTM, GRU"]
    D2["Transformers,<br/>self-attention"]
    D3["Survival models,<br/>hazard of churn"]
  end
  LOG --> MK
  LOG --> SA
  LOG --> DP
```

## Applications {#sec-click-applications}

Behavioral log data underwrites a wide span of marketing decisions. It is worth fixing
the major application areas before the methods, because the choice of method is dictated
by the question.

*Web and app session analysis* is the foundational use. Aggregating events into
sessions yields the descriptive vocabulary of digital marketing: session counts, pages
per session, dwell time, bounce rate, entry and exit pages, and the conversion rate of
a session. These quantities are the raw material of dashboards and the denominators of
optimization, and getting them right depends entirely on defining a session correctly
(@sec-click-sessionization). Early marketing-science work established that browsing
itself is a modelable behavior, not merely a precursor to purchase: the depth and
dynamics of online search vary systematically across users and over the course of a
visit [@johnson2004depth], and a site visit can be represented as a probabilistic
walk whose transitions reveal goal-directed versus exploratory navigation
[@bucklin2003]. The conversion decision within and across visits is itself dynamic—the
probability of buying evolves with accumulated visits and within-session behavior
[@moe2004dynamic; @sismeiro2004modeling]—and browsing spans competing sites, so a
single-site log is a censored view of a cross-site process [@park2004modeling]. At the
fine grain of product search, clickstream reveals how consumers actually explore a
category before choosing [@bronnenberg2016zooming].

*Path-to-purchase analysis* asks how users move through the funnel: which sequences of
pages and channels precede conversion, where they stall, and how the path differs
between buyers and non-buyers. The path is rarely linear. Modern journeys loop across
sessions, devices, and channels, and a central empirical finding is that the apparent
"last step" before a purchase is a poor summary of what produced it
[@montgomery2004modeling].

*Attribution* is path analysis with a managerial edge: given a conversion preceded by
several marketing touchpoints (a paid-search click, a display impression, an email, an
organic visit), how much credit does each touchpoint deserve? Attribution is the budget
allocation problem of digital marketing, and it is the application where the choice
between naive heuristics and a model of the *path* matters most (@sec-click-attribution)
[@kannan2017].

*Recommendation* uses behavioral logs as the primary signal of preference.
Co-view and co-purchase patterns, session context, and the sequence of recent
interactions drive the "customers who viewed this also viewed" and "recommended for
you" systems that mediate a large share of digital demand. Sequential recommendation,
in particular, treats the next-item problem as a prediction over the user's event
sequence [@zheng2023recommending].

*Churn and retention prediction* reframes the log as a leading indicator of
disengagement. Declining session frequency, shortening visits, and shifts in the *mix*
of activities often precede outright lapse, so a model trained on behavioral features
can flag at-risk customers while intervention is still possible. Because lapse is a
timing phenomenon, the natural statistical frame is survival analysis
(@sec-click-deep), which models the *hazard* of churn as a function of behavior to date.

*Bot and fraud detection* turns the same sequential structure to a defensive purpose.
Automated traffic, click fraud, and scraping leave behavioral signatures that differ
from human browsing: implausibly regular inter-event timing, non-human navigation paths,
and event rates no person could sustain. Distinguishing genuine engagement from
synthetic traffic is a prerequisite for trustworthy metrics, since uncleaned bot traffic
inflates every downstream number and corrupts attribution and recommendation alike.

These applications share a substrate. Each begins from the same event log and differs
only in the target (a session metric, a path, a conversion, a next item, a lapse time,
a bot flag) and in the modeling machinery brought to bear. We therefore turn first to
the substrate itself.

## Data Structure and Sessionization {#sec-click-structure}

### Events, Sessions, and Sequences {#sec-click-events}

The atom of clickstream data is the *event*: a single recorded interaction. At minimum
an event carries three fields, an *actor* (a user or device identifier), an *action*
(what happened, often called the event type or page), and a *timestamp* (when, to the
second or millisecond). Real event records carry much more, a session or device
context, the referring channel, page or screen attributes, the product or item touched,
and device and geographic metadata, but the actor-action-time triple is the irreducible
core. Formally, the raw log is a set of tuples
$$
e = (u,\, a,\, t,\, \mathbf{x}),
$$ {#eq-click-event}
where $u$ indexes the actor, $a \in \mathcal{A}$ the action drawn from an event
vocabulary $\mathcal{A}$, $t$ the timestamp, and $\mathbf{x}$ any additional payload.

A *session* (or *visit*) groups a user's events into a single episode of activity. A
session is the unit at which most descriptive metrics are defined and the unit a path
model walks through. The grouping is not given by the data; it is *constructed* by a
sessionization rule, and that construction is the first place a clickstream analysis can
quietly go wrong (@sec-click-sessionization).

A *sequence* is the ordered list of actions within a session (or, sometimes, within a
user's entire history),
$$
s_u = (a_{u,1},\, a_{u,2},\, \dots,\, a_{u,L_u}),
$$ {#eq-click-sequence}
where the events are sorted by timestamp and $L_u$ is the session length. The sequence
is the object the methods of this chapter operate on. A Markov model treats it as a
realization of a stochastic walk over $\mathcal{A}$; sequence analysis treats it as a
categorical string to be aligned and clustered; a recurrent network treats it as an
ordered input to be encoded. The recurring theme is that *order is information*, and the
analytical leverage of clickstream data comes precisely from refusing to discard it.

### Simulating an Event Log {#sec-click-simulate}

To make the rest of the chapter concrete and reproducible, we simulate a clickstream
event log with realistic structure: many users, multiple sessions per user separated by
idle gaps, and within-session navigation governed by a plausible page-transition
process that culminates, for some sessions, in a purchase. We carry this single
simulated log through the sessionization, Markov, and sequence-analysis demonstrations
that follow. @fig-clk-walk sketches the page-transition process that generates each
session: a forward funnel from home toward checkout, with backtracking and a persistent
risk of exit at every page.

```{mermaid}
%%| label: fig-clk-walk
%%| fig-cap: "The page-transition process generating each simulated session. Edges show transition probabilities; self-loops and transitions below 0.10 are omitted for clarity. Every transient page risks exit, and checkout leads to exit with certainty, so exit is absorbing."
flowchart LR
  HM["home"] -->|"0.45"| SR["search"]
  HM -->|"0.25"| PD["product"]
  SR -->|"0.50"| PD
  PD -->|"0.30"| CT["cart"]
  PD -->|"0.20"| SR
  CT -->|"0.45"| CO["checkout"]
  CT -->|"0.20"| PD
  CO -->|"1.00"| EX["exit<br/>(absorbing)"]
  HM -->|"0.25"| EX
  SR -->|"0.25"| EX
  PD -->|"0.20"| EX
  CT -->|"0.20"| EX
```

```{r click-simulate}
#| message: false
#| warning: false
#| label: tbl-click-loghead
#| tbl-cap: "First rows of the simulated raw clickstream event log, sorted by user and time."
library(tidyverse)

set.seed(55)

# --- Event vocabulary and a plausible page-transition process -----------------
# States are pages a user can be on. "home", "search", "product", "cart",
# "checkout" form the conversion funnel; "exit" is an absorbing leave state.
pages <- c("home", "search", "product", "cart", "checkout", "exit")

# Row-stochastic transition matrix: P[from, to]. These probabilities encode a
# realistic funnel where most users browse and leave, a minority advance to cart,
# and fewer still check out. Each row sums to 1.
P <- matrix(0, nrow = length(pages), ncol = length(pages),
            dimnames = list(pages, pages))
P["home",     ] <- c(0.05, 0.45, 0.25, 0.00, 0.00, 0.25)
P["search",   ] <- c(0.05, 0.15, 0.50, 0.05, 0.00, 0.25)
P["product",  ] <- c(0.05, 0.20, 0.25, 0.30, 0.00, 0.20)
P["cart",     ] <- c(0.02, 0.03, 0.20, 0.10, 0.45, 0.20)
P["checkout", ] <- c(0.00, 0.00, 0.00, 0.00, 0.00, 1.00)  # checkout -> exit
P["exit",     ] <- c(0.00, 0.00, 0.00, 0.00, 0.00, 1.00)  # absorbing

stopifnot(all(abs(rowSums(P) - 1) < 1e-9))  # rows are proper distributions

# --- Walk a single session over the page-transition process ------------------
# A session starts on "home", steps through P until it hits "exit", and is
# capped at max_len steps to keep sessions finite.
simulate_session <- function(P, start = "home", max_len = 20) {
  states <- start
  cur <- start
  for (i in seq_len(max_len)) {
    nxt <- sample(colnames(P), size = 1, prob = P[cur, ])
    states <- c(states, nxt)
    cur <- nxt
    if (cur == "exit") break
  }
  states
}

# --- Generate users, sessions, and timestamped events ------------------------
n_users <- 400
out <- vector("list", 0)

for (u in seq_len(n_users)) {
  user_id <- sprintf("u%04d", u)
  n_sessions <- 1 + rpois(1, lambda = 1.2)          # 1+ sessions per user
  # First session begins at a random time within a 30-day window.
  clock <- as.POSIXct("2025-01-01 00:00:00", tz = "UTC") +
    runif(1, 0, 30 * 24 * 3600)
  for (sess in seq_len(n_sessions)) {
    visit <- simulate_session(P)
    # Inter-event gaps within a session: seconds to a few minutes.
    gaps <- c(0, round(rexp(length(visit) - 1, rate = 1 / 40)))
    ts <- clock + cumsum(gaps)
    out[[length(out) + 1]] <- tibble(
      user_id = user_id,
      page    = visit,
      ts      = ts
    )
    # Idle gap before the user's next session: hours to days.
    clock <- max(ts) + runif(1, 2 * 3600, 4 * 24 * 3600)
  }
}

events <- bind_rows(out) |>
  arrange(user_id, ts) |>
  mutate(event_id = row_number())

head(events, 10)
```

As @tbl-click-loghead shows, the log now has the actor-action-time structure of
@eq-click-event: a `user_id`, a `page` (the action), and a `ts` timestamp, with one row
per event. Its scale is modest
but its shape is realistic, with users contributing a variable number of sessions and
sessions of variable length governed by the funnel process.

```{r click-logsize}
#| label: click-logsize
# Basic shape of the simulated log.
events |>
  summarise(
    n_events = n(),
    n_users  = n_distinct(user_id),
    span_days = as.numeric(difftime(max(ts), min(ts), units = "days"))
  )
```

### Sessionization {#sec-click-sessionization}

The raw log records events but not sessions; the analyst must impose them. The
canonical rule is *inactivity-based*: a new session begins whenever the gap since a
user's previous event exceeds a timeout threshold, conventionally thirty minutes in web
analytics. The choice of threshold is consequential. Too short a timeout fragments a
single deliberation into multiple sessions and inflates session counts while deflating
pages-per-session; too long a timeout merges distinct visits and corrupts entry- and
exit-page statistics. There is no universally correct threshold, and a defensible
analysis either justifies the threshold from the empirical gap distribution or reports
sensitivity to it.

We sessionize the simulated log with a thirty-minute timeout. Because we simulated idle
gaps of hours to days *between* sessions and gaps of seconds to minutes *within* them,
the rule should recover the session structure we built in.

```{r click-sessionize}
#| label: click-sessionize
timeout_secs <- 30 * 60  # 30-minute inactivity timeout

sessions_df <- events |>
  arrange(user_id, ts) |>
  group_by(user_id) |>
  mutate(
    gap_secs   = as.numeric(difftime(ts, lag(ts), units = "secs")),
    new_session = is.na(gap_secs) | gap_secs > timeout_secs,
    session_seq = cumsum(new_session)
  ) |>
  ungroup() |>
  mutate(session_id = paste(user_id, session_seq, sep = "_"))

# Session-level summary: length, duration, and whether the funnel was completed.
session_summary <- sessions_df |>
  group_by(session_id, user_id) |>
  summarise(
    n_events   = n(),
    duration_s = as.numeric(difftime(max(ts), min(ts), units = "secs")),
    converted  = any(page == "checkout"),
    .groups = "drop"
  )

session_summary |>
  summarise(
    n_sessions       = n(),
    mean_events      = mean(n_events),
    median_duration  = median(duration_s),
    conversion_rate  = mean(converted)
  )
```

The summary reports the descriptive vocabulary of session analysis: the number of
sessions recovered, the mean events per session, the median session duration, and the
overall conversion rate. Each of these would shift if the timeout changed, which is why
sessionization is a modeling choice and not a clerical one.

### The Conversion Funnel {#sec-click-funnel}

A *funnel* is the canonical descriptive summary of path-to-purchase behavior: the share
of sessions that reach each successive stage of a defined sequence. Funnels make
drop-off visible, localizing where users are lost, and they are the first artifact a
practitioner builds before any model. We define a five-stage funnel over the page
vocabulary and compute, for each stage, the fraction of sessions that ever reached it
and the step-to-step retention.

```{r click-funnel}
#| label: tbl-click-funnel
#| tbl-cap: "Conversion funnel over the simulated sessions: reach and step-to-step retention by stage."
funnel_stages <- c("home", "search", "product", "cart", "checkout")

n_sessions <- n_distinct(sessions_df$session_id)

# For each funnel stage, count the sessions that ever reached it.
funnel_tbl <- tibble(stage = funnel_stages) |>
  mutate(
    sessions = map_int(stage, function(stg) {
      sessions_df |>
        filter(page == stg) |>
        distinct(session_id) |>
        nrow()
    }),
    reach          = sessions / n_sessions,
    step_retention = reach / lag(reach)
  )

funnel_tbl
```

In @tbl-click-funnel, the `reach` column is the share of sessions touching each stage;
`step_retention` is the conditional probability of advancing from the previous stage. The sharpest drop-offs
in `step_retention` are the points a conversion-rate optimization program would target.
Note that this funnel counts whether a stage was *ever* reached in a session, not
whether stages occurred in strict order; an order-sensitive funnel, which only credits a
stage if it followed the prior stage, is stricter and is the more honest summary when
backtracking is common.

## Path and Markov Models {#sec-click-markov}

### Navigation as a Stochastic Walk {#sec-click-markov-walk}

The defining structure of clickstream data is sequence, and the simplest faithful model
of a sequence is a *Markov chain*. A first-order Markov model treats the next page as
depending only on the current page, not on the full history before it,
$$
\Pr(a_{t+1} = j \mid a_t = i,\, a_{t-1},\, \dots) \;=\; \Pr(a_{t+1} = j \mid a_t = i)
\;=\; p_{ij},
$$ {#eq-click-markov}
where the *transition probabilities* $p_{ij}$ collect into a row-stochastic matrix
$\mathbf{P}$. The first-order (memoryless) assumption is strong, and a substantial
marketing-science literature relaxes it: higher-order chains condition on the last $k$
pages, hidden-Markov models posit latent states that govern observed navigation
[@netzer2008hidden], and dynamic models let transition probabilities evolve within a
visit [@montgomery2004modeling; @bucklin2003]. Yet the first-order chain is the right
starting point. It is interpretable, it is cheap to estimate from a transition-count
matrix, and, as we will see, it supports a clean and widely used notion of attribution.

### Estimating the Transition Matrix {#sec-click-markov-fit}

We estimate the first-order transition matrix from the sessionized log using the
`markovchain` package, whose tooling for fitting and manipulating discrete-time Markov
chains in R is documented by Spedicato [2017, *The R Journal* 9(2), DOI
`10.32614/rj-2017-036`]. The estimator is simply the matrix of empirical transition
frequencies, normalized to rows. We build, for each session, the ordered vector of
pages and pass the list of sequences to `markovchainFit`.

```{r click-markov-fit}
#| label: click-markov-fit
#| message: false
#| warning: false
library(markovchain)

# One ordered character vector of pages per session.
seq_list <- sessions_df |>
  arrange(user_id, ts) |>
  group_by(session_id) |>
  summarise(path = list(page), .groups = "drop") |>
  pull(path)

# Maximum-likelihood fit of a first-order chain from the list of sequences.
fit <- markovchainFit(data = seq_list, method = "mle")
mc   <- fit$estimate

# The estimated transition matrix (rounded for display).
round(mc@transitionMatrix, 3)
```

The fitted matrix recovers, up to sampling noise, the data-generating process of
@sec-click-simulate: high probability of leaving from `home`, a `search`-to-`product`
pathway, and the `cart`-to-`checkout` step that defines conversion. Because the
`markovchain` object knows that `exit` is absorbing, it exposes the chain's structure
directly.

```{r click-markov-props}
#| label: click-markov-props
# Which states are absorbing? (Self-probability 1.)
absorbingStates(mc)

# Steady-state behavior is degenerate here because every walk ends in "exit";
# the informative object is the matrix of transient transitions above, not a
# stationary distribution. We confirm the absorbing structure instead.
diag(mc@transitionMatrix)["exit"]
```

The estimated chain is a compact, interpretable model of navigation. It answers
questions a funnel cannot: not merely how many sessions reached `cart`, but the
probability that a user *in* the cart proceeds to checkout versus backtracks to keep
browsing. Those conditional probabilities are the substance of path analysis, and they
are exactly what attribution will exploit.

### Markov Attribution and the Removal Effect {#sec-click-removal}

Markov chains supply a principled answer to the attribution question through the
*removal effect*. The idea, developed for graph-based online attribution by Anderl,
Becker, von Wangenheim, and Schumann [2016, *International Journal of Research in
Marketing* 33(3), 457-474, DOI `10.1016/j.ijresmar.2016.03.001`], is to model journeys
as walks on a graph whose nodes are channels (or pages), with two special absorbing
nodes, *conversion* and *null* (no conversion). The value of a channel is then defined
by what happens when it is *removed* from the graph: the *removal effect* of channel
$c$ is the proportional drop in overall conversion probability when all paths through
$c$ are diverted to the null state,
$$
\text{RE}(c) \;=\; \frac{\pi_{\text{conv}} - \pi_{\text{conv}}^{(-c)}}{\pi_{\text{conv}}},
$$ {#eq-click-removal}
where $\pi_{\text{conv}}$ is the conversion probability of the full chain and
$\pi_{\text{conv}}^{(-c)}$ is the conversion probability after removing $c$. Channels
whose removal collapses conversion are credited heavily; channels whose removal barely
moves it are credited little. Normalizing removal effects across channels yields a
data-driven attribution that respects the *path* rather than mechanically crediting the
first or last touch (@sec-click-attribution).

We implement the removal effect on our page chain by treating `checkout` as conversion
and `exit`-without-checkout as the null state, then comparing the absorption probability
into conversion with and without each transient page. The absorption probability is
computed in closed form from the chain's transition matrix.

```{r click-removal}
#| label: tbl-click-removal
#| tbl-cap: "Markov removal effects for each transient page. Larger values indicate pages whose removal most reduces overall conversion probability."

tm <- mc@transitionMatrix

# Define the two absorbing outcomes. "checkout" is conversion. We add an explicit
# "null" outcome: from "exit", a session that never reached checkout is a non-
# conversion. In this chain "exit" already absorbs, so we treat reaching "exit"
# without having passed through "checkout" as the null outcome. For the removal-
# effect computation we work with the conversion-absorption probability from the
# entry state "home".

# Probability of being absorbed into "checkout" starting from "home", computed by
# solving the linear system for absorption probabilities of a Markov chain.
absorb_conv_prob <- function(tm, entry = "home", conv = "checkout",
                             absorbing = c("checkout", "exit")) {
  states <- rownames(tm)
  transient <- setdiff(states, absorbing)
  Q <- tm[transient, transient, drop = FALSE]   # transient-to-transient
  R <- tm[transient, absorbing, drop = FALSE]   # transient-to-absorbing
  N <- solve(diag(nrow(Q)) - Q)                 # fundamental matrix
  B <- N %*% R                                  # absorption probabilities
  B[entry, conv]
}

base_conv <- absorb_conv_prob(tm)

# Removal effect: zero out a transient page by diverting all its inflow to "exit"
# (the null outcome), then recompute conversion probability from "home".
transient_pages <- setdiff(rownames(tm), c("checkout", "exit", "home"))

removal <- map_dfr(transient_pages, function(pg) {
  tm2 <- tm
  # Send everything that would go TO page `pg` straight to "exit" instead,
  # then make `pg` itself a dead-end into "exit".
  tm2[, "exit"] <- tm2[, "exit"] + tm2[, pg]
  tm2[, pg] <- 0
  tm2[pg, ] <- 0
  tm2[pg, "exit"] <- 1
  conv_without <- absorb_conv_prob(tm2)
  tibble(page = pg,
         conv_without = conv_without,
         removal_effect = (base_conv - conv_without) / base_conv)
})

removal |>
  mutate(attribution = removal_effect / sum(removal_effect)) |>
  arrange(desc(removal_effect))
```

In @tbl-click-removal, the `removal_effect` column quantifies how much each page
contributes to conversion: a page with a large removal effect is one the conversion
paths depend on, so diverting its traffic to the null outcome sharply lowers the chance
of reaching checkout. The
`attribution` column normalizes these effects into shares that sum to one, the
quantities a budget-allocation exercise would use. The same machinery, applied to a
chain whose *states are marketing channels* rather than pages, is exactly the
graph-based attribution model of Anderl et al. (2016); the page-level version here is a
faithful, smaller illustration of the identical logic.

## Sequence Analysis {#sec-click-sequence}

Markov models compress each sequence into transition counts and, in doing so, discard
the identity of the individual trajectory. *Sequence analysis*, a tradition imported
into the social sciences from biology, takes the opposite stance: it treats each user's
ordered sequence of states as a holistic object, to be compared with other sequences,
aligned, clustered into behavioral types, and visualized as a population. The canonical
R toolkit is `TraMineR`, whose methods for representing, summarizing, and visualizing
categorical state sequences are documented by Gabadinho, Ritschard, Müller, and Studer
[2011, *Journal of Statistical Software* 40(4), DOI `10.18637/jss.v040.i04`].

### Representing Sequences {#sec-click-traminer-rep}

`TraMineR` operates on a *state-sequence object* in which each row is one unit's
trajectory over aligned positions. We build, for each session, the sequence of pages up
to a fixed number of positions, pad shorter sessions, and construct the sequence object.

```{r click-traminer-seq}
#| label: click-traminer-seq
#| message: false
#| warning: false
library(TraMineR)

max_pos <- 6  # align sessions to their first 6 events (the funnel depth)

seq_wide <- sessions_df |>
  arrange(user_id, ts) |>
  group_by(session_id) |>
  mutate(pos = row_number()) |>
  ungroup() |>
  filter(pos <= max_pos) |>
  mutate(pos = factor(pos, levels = 1:max_pos)) |>
  select(session_id, pos, page) |>
  pivot_wider(names_from = pos, values_from = page,
              names_prefix = "t", names_expand = TRUE) |>
  arrange(session_id)

# Columns t1..t6 hold the page at each position; NA past the session's end.
state_cols <- paste0("t", 1:max_pos)

# Define the alphabet and build the TraMineR sequence object.
click_seq <- seqdef(
  seq_wide,
  var      = state_cols,
  alphabet = pages,
  states   = pages,
  xtstep   = 1
)

# How many sequences, and a glance at the first few.
print(click_seq[1:5, ])
```

### State Distributions and Clustering {#sec-click-traminer-cluster}

A first, population-level view is the *state-distribution plot*: at each sequence
position, the share of sessions occupying each state. @fig-click-statedist shows how the
population flows through the funnel over event positions, with users draining into `exit`
as positions advance.

```{r click-traminer-state}
#| label: fig-click-statedist
#| fig-cap: "State-distribution plot of the simulated sessions. At each event position (t1..t6), the bar shows the share of sessions on each page; the population drains toward the absorbing exit state."
#| out-width: "85%"
seqdplot(click_seq, border = NA, with.legend = "right",
         main = NULL)
```

To move from description to typology, we cluster sequences by *dissimilarity*.
`TraMineR` computes pairwise distances between sequences, the classic choice being
*optimal matching* (OM), which defines the distance between two sequences as the minimum
cost of insertions, deletions, and substitutions needed to transform one into the other.
Clustering the resulting distance matrix groups sessions into behavioral archetypes,
short bouncers, deep browsers, converters, that a population of raw logs would otherwise
hide; @tbl-click-clusters reports the sizes of the four clusters so recovered.

```{r click-traminer-om}
#| label: tbl-click-clusters
#| tbl-cap: "Sizes of behavioral clusters recovered from optimal-matching distances between session sequences, via Ward hierarchical clustering into four groups."
#| message: false
#| warning: false
library(cluster)

# Substitution costs derived from observed transition rates; indel cost fixed.
sm <- seqsubm(click_seq, method = "TRATE")
om <- seqdist(click_seq, method = "OM", sm = sm, indel = 1)

# Ward clustering of the distance matrix into 4 behavioral types.
hc <- agnes(as.dist(om), method = "ward")
cl4 <- cutree(hc, k = 4)

table(cluster = cl4)
```

```{r click-traminer-clusterplot}
#| label: fig-click-clusterplot
#| fig-cap: "State-distribution plots by cluster. Each panel is a behavioral archetype recovered from the sequence distances: clusters differ in how quickly they exit and how far they advance into the funnel."
#| out-width: "95%"
seqdplot(click_seq, group = factor(cl4), border = NA,
         with.legend = "right")
```

The four panels of @fig-click-clusterplot are interpretable behavioral types: clusters
that exit almost immediately, clusters that linger in `search` and `product`, and clusters
that reach `cart` and `checkout`. This is the payoff of treating sequences holistically. Rather
than a single average transition matrix, sequence clustering surfaces a *heterogeneous
population* of navigation styles, which is the empirical reality of behavioral data and
the starting point for segmentation, targeting, and personalized intervention.

## Deep Sequence Models {#sec-click-deep}

The classical methods above impose structure: a Markov chain assumes limited memory, and
sequence analysis fixes a dissimilarity metric in advance. *Deep sequence models* invert
this, learning a representation of behavior directly from raw event streams with minimal
hand-specified structure. This section is conceptual; the models below are described
rather than fit, because training them at scale is beyond a runnable book example, but
the architectures are the production workhorses of large digital platforms and deserve
to be named precisely.

*Recurrent neural networks* (RNNs), and their gated variants the LSTM and GRU, process a
sequence one event at a time while maintaining a hidden state that summarizes the history
seen so far. Unlike a fixed-order Markov chain, an RNN can in principle carry information
across arbitrarily long gaps, learning, for example, that a search early in a session
conditions the value of a much later product view. RNNs were the first deep architecture
to deliver strong session-based recommendation, and sequential recommendation remains a
leading application of behavioral sequence modeling [@zheng2023recommending].

*Transformers* replace recurrence with *self-attention*, letting every event attend
directly to every other event in the sequence. Self-attention removes the sequential
bottleneck of RNNs, parallelizes training, and captures long-range dependencies more
effectively, which is why transformer architectures now dominate sequential
recommendation and behavioral modeling. The canonical architecture, Vaswani et al.,
"Attention Is All You Need" (NeurIPS 2017), has no canonical Crossref DOI and is cited by
name; its adaptation to user-behavior sequences underlies modern next-item prediction.

*Survival and hazard models* address a different target: not the next event but the
*timing* of an event, paradigmatically churn. A survival model treats the time until
lapse as the object of inference and models the *hazard*, the instantaneous risk of
lapsing at time $t$ given survival so far, as a function of behavioral covariates. This
is the natural frame for churn and retention precisely because lapse is censored: most
customers have not yet churned at the time of analysis, and a survival model uses that
censored information correctly where a binary classifier discards it. Behavioral
covariates, declining session frequency, shortening visits, shifting activity mix, enter
the hazard and yield a continuously updated risk score. The deep-learning extension
replaces the linear hazard with a neural network over the event sequence, fusing the
representational power of the architectures above with the censoring-aware machinery of
survival analysis.

A recurring caution governs all of these models, and it is the same caution that governs
text and image features elsewhere in this part: a learned representation of behavior is a
*generated regressor*. When the output of a sequence model (a churn score, a propensity,
an engagement embedding) is fed into a downstream regression or decision, its estimation
error can correlate with the outcome, and the standard errors of the downstream model
will understate uncertainty unless the generation step is accounted for. Behavioral
embeddings are powerful inputs, but they are model outputs first.

## Attribution {#sec-click-attribution}

Attribution deserves separate treatment because it is the application where method choice
most directly moves money, and where the gap between convenient heuristics and defensible
models is widest. The problem is to distribute credit for a conversion across the
touchpoints, channels, ad exposures, and site visits that preceded it.

The *heuristic* attribution rules require no model and are correspondingly crude.
*Last-touch* assigns all credit to the final touchpoint before conversion; *first-touch*
assigns it all to the first; *linear* splits credit equally; *time-decay* weights recent
touches more. Each encodes an untested assumption about how influence accrues along a
path, and last-touch in particular systematically overcredits the bottom-of-funnel
channels that users were always going to reach and undercredits the upper-funnel channels
that initiated the journey [@montgomery2004modeling].

The *data-driven* alternatives infer credit from the structure of observed paths. The
Markov removal-effect model of @sec-click-removal is the leading example: by defining a
channel's value as the conversion lost when its paths are diverted, it credits channels
by their causal role *within the observed path structure* rather than by their position
[@anderl2016mapping]. A complementary tradition
builds explicit measurement models of the multichannel journey: @li2014 model a
customer's consideration of channels, visits through them over time, and subsequent
purchase, and crucially validate the model against a field experiment, and Abhishek,
Fader, and Hosanagar ("The Long Road to Online Conversion," SSRN working paper 2012, DOI
`10.2139/ssrn.2158421`) model the path from media exposure to conversion with a
hidden-Markov structure over latent journey stages. Closely related work disentangles how
channels interact along the path: @dehaan2016effectiveness compare the conversion impact
of different online ad forms within a multichannel attribution framework, and
@kireyev2016display show that display advertising *dynamically* raises later search
activity, so crediting search alone double-undercounts display. Individual-level exposure
data let @zantedeschi2017 measure multichannel advertising response jointly across mail and
email. The same path logic extends to paid search itself, where keyword-level
clickstream supports both reduced-form [@ghose2009empirical] and dynamic structural
[@yao2011dynamic] models of sponsored-search performance. The broader digital-marketing
framework that situates attribution among the firm's measurement problems is reviewed by
@kannan2017.

Two cautions apply to every attribution model, heuristic or data-driven. First,
attribution is *correlational by default*: a model fit to observed paths recovers
associations between touchpoints and conversion, not the counterfactual conversion that
would have occurred absent a touchpoint, unless the design supports a causal reading. The
strongest attribution work therefore pairs the model with experimental variation, as Li
and Kannan (2014) do, rather than resting on observational paths alone. Second,
attribution is acutely sensitive to *what is logged*. Touchpoints that go unrecorded,
view-through impressions, cross-device journeys, offline exposures, are silently assigned
zero credit, biasing the attribution toward the well-instrumented channels. The
measurement boundary is the attribution boundary.

## Industry and Production Practice {#sec-click-production}

The methods above presuppose a clean event log. In practice that log is the *output* of a
substantial data-engineering pipeline, and the integrity of every downstream analysis
depends on how that pipeline is built. This section sketches the production reality that
academic treatments usually omit, because the failure modes it introduces are precisely
the ones that invalidate published-looking results. @fig-clk-pipeline maps the
pipeline end to end, from instrumentation to the point-in-time features a model
consumes.

```{mermaid}
%%| label: fig-clk-pipeline
%%| fig-cap: "The production event pipeline. Instrumented events flow through a collection layer into an append-only warehouse table; derived tables materialize sessionization and identity resolution; and the feature pipeline assembles training examples whose features use only information available at the prediction time, with labels observed after it."
flowchart LR
  IN["Instrumentation<br/>tracking plan on<br/>web and app clients"] --> CL["Collection layer<br/>streaming bus"]
  CL --> WH["Warehouse or lakehouse<br/>append-only event table"]
  WH --> DT["Derived tables<br/>sessionization and<br/>identity resolution"]
  DT --> FP["Feature pipeline<br/>features from history up<br/>to the prediction time"]
  FP --> TE["Training example<br/>features joined to a label<br/>observed after prediction time"]
  TE --> MD["Downstream models<br/>and metrics"]
```

The *event pipeline* originates in instrumentation: a *tracking plan* specifies which
events are emitted, with which properties, from web and app clients. Events flow through
a collection layer, often a streaming bus, into a *warehouse* or *lakehouse* where they
land as an append-only event table. Discipline at this stage, stable event names,
versioned schemas, consistent identifiers, is what makes the log analyzable at all; an
event whose semantics drift across an app release silently breaks every longitudinal
metric defined on it.

*Sessionization* (@sec-click-sessionization) is typically materialized in the pipeline
rather than recomputed per analysis, so the timeout rule becomes an organizational
commitment baked into derived tables. Identity resolution sits alongside it: stitching a
user's events across devices and login states determines whether a cross-device journey
appears as one user or several, with direct consequences for attribution and CLV.

The *warehouse-to-feature pipeline* turns the raw event table into the columns a model
consumes. For each user and a chosen *prediction time*, the pipeline aggregates the
history into features, counts, recencies, frequencies, sequence summaries, and embeddings,
and joins them to a label observed *after* the prediction time. The discipline that
governs this step is the avoidance of *look-ahead leakage*: a feature must be computable
using only information available at the prediction time, never information from the future
relative to it. Leakage is the dominant failure mode of behavioral modeling in practice
and is insidious because it inflates offline metrics, a churn model that accidentally
includes a post-churn event as a feature will look superb in backtests and fail in
production. The structural guard is a *point-in-time correct* feature store that, for any
training example, reconstructs feature values as they would have stood at that example's
prediction time, and never later.

Two further production realities shape the analysis. Bot and non-human traffic, discussed
in @sec-click-applications, must be filtered *before* metrics and models are computed,
since uncleaned synthetic traffic corrupts transition matrices, funnels, and attribution
alike. And privacy and consent regimes increasingly constrain what may be logged and
retained at all: cookie deprecation, consent-gated tracking, and on-device measurement
shrink the observable event log, which feeds directly back into the measurement-boundary
caution of @sec-click-attribution. The clickstream a model sees is not the behavior that
occurred; it is the behavior that was instrumented, permitted, and retained.

::: {.callout-tip}
## Replication resources: clickstream analysis

The sessionization, transition-matrix estimation, Markov removal-effect attribution, and
sequence-clustering demonstrations in this chapter run on open R packages—base R,
`data.table`/`dplyr`, the `markovchain`/`ChannelAttribution` family for path attribution,
and `TraMineR` for sequence analysis—which constitute a fully reproducible clickstream
toolchain. The empirical studies cited here (e.g., @montgomery2004modeling, @li2014,
@bronnenberg2016zooming) rest on proprietary firm clickstream that cannot be
redistributed, so public replication packages are rare; the methods, not the data, are
the transferable artifact, consistent with the unstructured-data program's emphasis on
validated, reproducible measurement [@balducci2018unstructured].
:::

## Frontier and Extensions {#sec-click-frontier}

Clickstream analysis is consolidating around a few active frontiers. The first is the
move from *hand-engineered behavioral features* to *learned behavioral embeddings*:
representing each user by a vector that a sequence model has learned from their raw event
history, then reusing that vector across churn, recommendation, and lifetime-value tasks.
This mirrors the trajectory of text and image data elsewhere in this part, and it inherits
the same generated-regressor caution (@sec-click-deep).

A second frontier is *causal* clickstream analysis. The observational, self-selected
nature of behavioral logs (@sec-click-applications) means that the most consequential
questions, does this navigation cause conversion, does this intervention reduce churn, are
causal questions that path models answer only under strong assumptions. Integrating
behavioral logs with experimental and quasi-experimental designs, so that the rich
sequence data supplies covariates and heterogeneity while the design supplies
identification, is where attribution and uplift modeling are heading.

A third frontier is *multimodal fusion*. The clickstream is one signal among several: a
session also generates text (search queries, reviews), images (products viewed), and,
increasingly, voice and video interactions. Joint models that fuse the behavioral sequence
with these other modalities, the subject of this part as a whole, promise representations
of the customer that no single log could yield. The unifying lesson is the one this part
returns to throughout: behavioral logs, like text and images, are unstructured traces that
become useful only once turned into learned, lossy features, and those features must be
defended as generated regressors before any claim is built on them.
