# Geospatial and Location Data {#sec-geospatial}

Marketing happens somewhere. A customer walks past a storefront, a delivery van
threads a route, a billboard faces one direction and not another, a smartphone
broadcasts its coordinates to an ad exchange, and a trade area swells or shrinks
as a competitor opens across the street. For most of the discipline's history this
*where* was treated as background: a control variable at best, a nuisance at worst.
What has changed is that location is now *data* in the same operational sense that
text (@sec-text-as-data) and images (@sec-image-processing) are data. A point on
the earth becomes a pair of coordinates, a customer becomes a point pattern, a
market becomes a polygon, and the spatial relations among them, who is near whom,
which areas border which, how demand decays with distance, become quantities a
model can ingest. This chapter is about turning the *where* into machine-readable
geometry and about the methods that respect what makes spatial data peculiar.

The peculiarity is worth stating at the outset, because it organizes everything
that follows. Spatial observations are not independent draws. Tobler's so-called
first law of geography puts it plainly: everything is related to everything else,
but near things are more related than distant things (Tobler 1970, *Economic
Geography* 46, 234-240, doi:10.2307/143141). Two stores a block apart cannibalize
each other; two census tracts that touch share commuters, climate, and culture; a
geo-targeted campaign in one zip code spills into the next. This **spatial
dependence** breaks the independence assumption underneath ordinary regression,
inflates apparent precision, and, if ignored, manufactures findings out of
geography itself. The same dependence is also the *signal* that trade-area models,
geodemographic segmentation, and geo-experiments are built to exploit. The
analyst's task is to represent geometry correctly, measure spatial relations
honestly, and then either model the dependence or design around it.

The chapter proceeds from applications to representation to method. It surveys the
marketing problems for which location is the central variable, fixes how spatial
data is stored and manipulated, and then develops the workhorse computations,
distance and trade-area analysis, spatial autocorrelation and spatial regression,
and geo-experiments, each with runnable `R`. It closes with the data sources and
privacy regime that govern location data in practice and with the frontier. The
runnable code uses the modern simple-features stack (`sf`), the current standard
for vector spatial data in `R` (Pebesma 2018, *The R Journal* 10(1), 439-446,
doi:10.32614/RJ-2018-009), and avoids the retired `sp`/`maptools` lineage entirely.

## Applications: Where Location Is the Variable {#sec-geo-applications}

Location data enters marketing wherever a decision turns on physical space. Eight
application families recur, and they motivate the methods in the rest of the
chapter.

**Trade-area analysis and site selection.** A store's *trade area* is the
geographic region from which it draws the bulk of its customers, and estimating it
is the oldest quantitative use of location in marketing. The foundational model is
Huff's gravity formulation, in which the probability that a consumer at location
$i$ patronizes store $j$ rises with the store's attractiveness (size, assortment)
and falls with the travel cost to reach it (Huff 1964, *Journal of Marketing*
28(3), 34-38, doi:10.2307/1249154; Huff 1963, *Land Economics* 39(1), 81-90,
doi:10.2307/3144521). Site selection inverts the problem: given candidate
locations, predict the trade area, the captured demand, and the cannibalization of
existing stores, then choose where to build.

**Geo-targeting and geo-experiments.** Digital advertising can be aimed at, and
withheld from, specific geographies, which makes geography both a *targeting*
instrument and an *experimental* one. This is now one of the best-developed empirical
literatures in marketing. Field experiments establish that *mobile targeting*—timing
and placing offers by a consumer's location—lifts redemption and sales
[@luo2014mobiletargeting], that effectiveness is *hyper-contextual*, rising with
situational factors such as physical crowdedness on a commute [@andrews2016crowdedness],
and that targeting on a consumer's *trajectory* (the sequence of places visited)
outperforms targeting on the current location alone [@ghose2019mobiletrajectory].
Promotions can be conditioned on proximity to a focal or a competitor's store, a tactic
known as geo-conquesting [@fong2015geoconquesting], and competitors can target each
other's customers with smartphone coupons, turning location into a strategic
price-discrimination instrument [@dube2017pricetargeting]. Which product categories
mobile advertising actually moves is itself an empirical question, the answer depending
on category and consumer context [@bart2014]. More broadly, location interacts with
targeting and obtrusiveness in display advertising [@goldfarb2011displaytargeting], with
physical colocation networks that predict ad response [@zubcsek2017colocation], with
interface design in location-based pull advertising [@molitor2020locationbased], and with
the cross-channel spillovers that link online and offline sales [@dinner2013crosschannel];
even broad targeted-discount programs function partly as advertising [@sahni2017targeteddiscounts].
Because regions can be randomized into treatment and control, geography also supports
clean field experiments at the market level, the basis of geo-lift designs
discussed in @sec-geo-experiments.

**Mobile location and footfall data.** Smartphones emit location through GPS,
cell-tower triangulation, Wi-Fi, and bluetooth beacons. Aggregated, these signals
measure *footfall*, the count of devices visiting a place over time, which serves
as a proxy for store visits, dwell time, and cross-shopping. Foot-traffic context
varies systematically across micro-locations and retail-center types (Philp,
Dolega, Singleton, and Green 2022, *Applied Spatial Analysis and Policy* 15(1),
161-187, doi:10.1007/s12061-021-09396-1). Mobile data also revealed that the
mobile internet is behaviorally distinct from the desktop internet: search costs
are higher, ranking effects stronger, and local activities more salient (Ghose,
Goldfarb, and Han 2013, *Information Systems Research* 24(3), 613-631,
doi:10.1287/isre.1120.0453).

**Store catchment and cannibalization.** Related to trade area but operationally
distinct, *catchment* assigns each customer or demand unit to a serving store,
usually the nearest or the most attractive, and then aggregates demand by store.
Catchment maps drive labor scheduling, inventory, and the cannibalization
accounting that any new-store decision requires.

**Geodemographics.** Geodemographic systems classify small areas (census tracts,
postal zones) into lifestyle segments by combining demographic, socioeconomic, and
behavioral data, on the premise that neighbors resemble one another. The approach
has a long applied history and a continuing research program spanning the United
States and United Kingdom (Singleton and Spielman 2014, *The Professional
Geographer* 66(4), 558-567, doi:10.1080/00330124.2013.848764). Geodemographic
codes feed direct mail, media planning, and site selection.

**Delivery and logistics.** Last-mile delivery, service-territory design, and
route optimization are spatial-marketing problems wherever delivery speed and cost
are part of the value proposition. The same distance and accessibility primitives
that define trade areas define delivery zones, surge regions, and the isochrones
(equal-travel-time contours) that quick-commerce firms advertise.

**Out-of-home advertising.** Billboards, transit panels, and place-based screens
are inherently geographic media: their value depends on the traffic that passes a
fixed location and the composition of that traffic. Programmatic out-of-home now
prices inventory using footfall and mobility data, turning a classically
untargeted medium into a location-data application.

**Local versus electronic competition.** Whether a consumer buys online or locally
depends on where the consumer lives relative to physical retail, an interaction
between geography and channel (Forman, Ghose, and Goldfarb 2009, *Management
Science* 55(1), 47-57, doi:10.1287/mnsc.1080.0932). Geography therefore shapes not
only *which* store but *which channel*, with direct consequences for pricing and
assortment.

::: {.callout-note}
The unifying abstraction is that a marketing object acquires a **geometry** (a
point, a line, or a polygon) and a **coordinate reference system** that fixes what
the coordinates mean. Once objects share a coordinate system, the spatial relations
among them, distance, containment, adjacency, overlap, become computable, and the
marketing question reduces to a query or a model over those relations.
:::

## Spatial Data and Its Representation {#sec-geo-representation}

Spatial structure is not a nuisance to be controlled away but a source of marketing
signal, and the agenda for *spatial models in marketing*—demand that is correlated
across nearby locations, spillovers between trade areas, and the joint modeling of
where consumers are and what they buy—was set out by @bradlow2005spatialmodels. Like
every modality in this part [@balducci2018unstructured], the work begins by turning the
raw artifact, here a coordinate, into a representation a model can speak about. Vector
spatial data has three ingredients: **geometry**, **attributes**, and a
**coordinate reference system** (CRS). The geometry is the shape, a point for a
store or a customer, a line for a road or a route, a polygon for a trade area or a
census tract. The attributes are the ordinary columns of a data frame: a store's
revenue, a tract's population. The CRS is the contract that says what the
coordinate numbers mean. Longitude-latitude pairs in degrees (the EPSG:4326
system, often called WGS84) live on a sphere, so Euclidean distance between them is
meaningless; a *projected* CRS maps the curved earth onto a plane in meters or feet
so that distances and areas can be computed with ordinary arithmetic. Mismatched or
missing CRS information is the single most common source of silent error in applied
spatial work, and the rule is simple: never compute a distance in degrees, and
never overlay two layers without confirming they share a CRS.

The `sf` package implements the Open Geospatial Consortium *simple features*
standard, storing a layer as a data frame with one list-column of geometries, so
that every `dplyr` and `ggplot2` idiom carries over (Pebesma 2018,
doi:10.32614/RJ-2018-009). The demonstration below simulates a small market: a set
of stores and a larger set of customers, each with coordinates, attached to a
projected CRS so that subsequent distances are in meters.

```{r geo-sf-setup, message=FALSE, warning=FALSE}
library(sf)
library(ggplot2)
library(dplyr)

set.seed(56)

# A synthetic metropolitan market on a projected grid measured in meters.
# We work directly in a planar CRS (a Universal Transverse Mercator zone)
# so that Euclidean distance in the coordinate units IS distance in meters.
utm <- 32616  # EPSG:32616, WGS84 / UTM zone 16N (planar, units = meters)

# Five stores placed across a roughly 30 km x 30 km area (coordinates in meters).
stores <- data.frame(
  store_id = paste0("S", 1:5),
  x = c( 5000, 22000, 12000, 27000,  8000),
  y = c( 8000,  6000, 20000, 24000, 26000),
  sqft = c(18000, 42000, 25000, 12000, 30000)  # store size (attractiveness)
)
stores_sf <- st_as_sf(stores, coords = c("x", "y"), crs = utm)

# 600 customers drawn from three spatial clusters (neighborhoods) plus noise,
# so the point pattern is realistically clumped rather than uniform.
n <- 600
centers <- data.frame(cx = c(8000, 24000, 15000),
                       cy = c(10000, 8000, 23000))
grp <- sample(1:3, n, replace = TRUE, prob = c(0.4, 0.35, 0.25))
cust <- data.frame(
  cust_id = 1:n,
  x = rnorm(n, centers$cx[grp], 3500),
  y = rnorm(n, centers$cy[grp], 3500),
  spend = round(rlnorm(n, log(60), 0.5), 2)  # monthly spend, skewed
)
cust_sf <- st_as_sf(cust, coords = c("x", "y"), crs = utm)

cat("Stores:", nrow(stores_sf), " Customers:", nrow(cust_sf), "\n")
cat("CRS units:", st_crs(utm)$units, "\n")
print(st_bbox(cust_sf))
```

The objects `stores_sf` and `cust_sf` are simple-features data frames: ordinary
columns plus a `geometry` column the package understands. Because the CRS is a
projected one in meters, every distance computed below is a real distance, not an
uninterpretable difference of degrees.

## Distance, Catchment, and Trade-Area Analysis {#sec-geo-distance}

With a shared CRS in place, the core spatial primitives are immediate. The
distance from each customer to each store is a single call; nearest-store
assignment is a row-wise minimum over that distance matrix; and aggregating spend
by serving store yields a catchment table. These three steps, distance,
assignment, aggregation, are the computational spine of trade-area analysis.

```{r geo-distance-assign, message=FALSE, warning=FALSE}
# Full customer-by-store distance matrix (meters). st_distance respects the CRS.
D <- st_distance(cust_sf, stores_sf)        # 600 x 5 matrix, units = m
D <- matrix(as.numeric(D), nrow = nrow(cust_sf))  # strip units for arithmetic

# Nearest-store assignment: the index of the minimum distance in each row.
nearest_idx <- apply(D, 1, which.min)
cust_sf$store_id  <- stores_sf$store_id[nearest_idx]
cust_sf$dist_m    <- D[cbind(seq_len(nrow(D)), nearest_idx)]

# Catchment table: customers, total spend, and median travel distance per store.
catchment <- cust_sf |>
  st_drop_geometry() |>
  group_by(store_id) |>
  summarise(customers   = n(),
            total_spend = sum(spend),
            median_dist_km = round(median(dist_m) / 1000, 2),
            .groups = "drop") |>
  arrange(desc(total_spend))

print(catchment)
```

Nearest-store assignment is the simplest catchment rule, and it implicitly draws a
*Voronoi tessellation*: the plane is partitioned into cells, one per store, each
cell being the set of locations closer to that store than to any other. Voronoi
catchments assume customers patronize the nearest store and nothing else, which is
a useful first approximation but ignores store attractiveness. The Huff gravity
model relaxes this by making patronage probabilistic and attractiveness-weighted.
For a customer at $i$ and store $j$ with attractiveness $A_j$ (here, square footage)
and distance $d_{ij}$,

$$
P_{ij} = \frac{A_j \, d_{ij}^{-\lambda}}{\sum_{k} A_k \, d_{ik}^{-\lambda}},
$$ {#eq-geo-huff}

where the distance-decay exponent $\lambda > 0$ governs how sharply patronage falls
with travel cost. The code below computes Huff probabilities and contrasts the
resulting *expected* demand with the hard nearest-store catchment, exposing how much
the attractiveness term redistributes demand toward larger stores.

```{r geo-huff, message=FALSE, warning=FALSE}
lambda <- 1.8                       # distance-decay exponent (calibrated elsewhere)
A <- stores_sf$sqft                 # attractiveness = store size

# Huff numerator A_j * d_ij^(-lambda), guarding against d = 0.
num <- sweep((pmax(D, 1))^(-lambda), 2, A, `*`)  # 600 x 5
P   <- num / rowSums(num)                          # patronage probabilities

# Expected spend each store captures = sum over customers of P_ij * spend_i.
huff_capture <- colSums(P * cust_sf$spend)
huff <- data.frame(store_id = stores_sf$store_id,
                   sqft = stores_sf$sqft,
                   huff_expected_spend = round(huff_capture, 0))

compare <- catchment |>
  select(store_id, nearest_spend = total_spend) |>
  left_join(huff, by = "store_id") |>
  arrange(desc(huff_expected_spend))
print(compare)
```

The two columns tell different stories. Nearest-store assignment is winner-take-all
within each Voronoi cell, so a small store can capture a large cell simply by being
isolated. The Huff model spreads each customer's demand across all stores in
proportion to attractiveness and inverse distance, so the largest stores claim a
disproportionate share even from customers who are nominally closer to a small
rival. Which model an analyst trusts is an empirical question settled by calibrating
$\lambda$ against observed patronage, but the contrast itself is the managerial
insight: a new large-format store does not merely capture its own cell, it siphons
probability mass from every store within its distance-decay reach, which is the
cannibalization that site selection must price.

::: {.callout-tip}
**Accessibility** generalizes distance. Straight-line (Euclidean) distance is a
crude proxy for travel cost when rivers, highways, and one-way streets intervene.
Production trade-area work replaces it with *network distance* or *travel time*
from a routing engine, and replaces circular buffers with *isochrones*, polygons of
equal travel time. The `sf` primitives are identical; only the distance metric
changes from a closed-form Euclidean formula to a routed query.
:::

## Aggregation to Areas and the Choropleth {#sec-geo-aggregation}

Marketing decisions are usually made over *areas*, sales territories, postal zones,
census tracts, not over raw points. Moving from points to areas requires a
**spatial join**: each point is matched to the polygon that contains it, and the
points are then aggregated within polygon. This points-in-polygon aggregation is the
operation that turns a scatter of customers into a thematic map. The example below
overlays a regular grid of analysis zones on the market, joins customers to zones,
and draws a **choropleth**, a map in which each area is shaded by an aggregated
value, here total customer spend.

```{r geo-choropleth, message=FALSE, warning=FALSE, fig.cap="Choropleth of total customer spend aggregated to a regular grid of analysis zones, with store locations overlaid. Darker cells capture more demand."}
# Build a 6 x 6 grid of square analysis zones covering the market bounding box.
grid <- st_make_grid(cust_sf, n = c(6, 6), what = "polygons")
grid <- st_sf(zone_id = seq_along(grid), geometry = grid)

# Spatial join: assign each customer to the grid cell that contains it,
# then aggregate spend and customer counts to the zone level.
joined <- st_join(cust_sf, grid, join = st_within)
zone_stats <- joined |>
  st_drop_geometry() |>
  group_by(zone_id) |>
  summarise(zone_spend = sum(spend), zone_n = n(), .groups = "drop")

grid <- grid |>
  left_join(zone_stats, by = "zone_id") |>
  mutate(zone_spend = ifelse(is.na(zone_spend), 0, zone_spend))

ggplot() +
  geom_sf(data = grid, aes(fill = zone_spend), color = "white", linewidth = 0.3) +
  geom_sf(data = stores_sf, shape = 24, fill = "red", color = "black", size = 3) +
  scale_fill_viridis_c(name = "Spend", option = "C") +
  labs(title = "Customer spend by analysis zone",
       subtitle = "Red triangles mark stores") +
  theme_minimal()
```

The map is more than decoration: it is the object that spatial-statistics methods
operate on. The grid's cells now carry an attribute (`zone_spend`) and an implicit
adjacency structure (which cells touch which), and the next two sections ask whether
the spatial pattern in that attribute is more clustered than chance, and how to
regress an outcome on covariates without letting the clustering corrupt the
inference.

## Spatial Autocorrelation and Spatial Regression {#sec-geo-autocorrelation}

The choropleth invites a question the eye cannot answer reliably: is the high-spend
region genuinely clustered, or could a pattern that *looks* clustered arise by
chance from independent draws? **Spatial autocorrelation** is the formal measure.
The most widely used global statistic is **Moran's $I$** (Moran 1950, *Biometrika*
37(1/2), 17-23, doi:10.2307/2332142). Given an attribute $x_a$ on areas
$a = 1,\dots,m$ and a spatial weights matrix $W$ with entries $w_{ab}$ encoding
which areas are neighbors,

$$
I = \frac{m}{\sum_{a}\sum_{b} w_{ab}}
\cdot
\frac{\sum_{a}\sum_{b} w_{ab}\,(x_a - \bar{x})(x_b - \bar{x})}
     {\sum_{a}(x_a - \bar{x})^2}.
$$ {#eq-geo-moran}

The statistic is a spatially weighted correlation of the attribute with itself
across neighboring areas. Values above the expectation $-1/(m-1)$ indicate positive
autocorrelation (neighbors resemble neighbors, the high-spend-near-high-spend
pattern Tobler's law predicts); values below it indicate a checkerboard. The weights
matrix $W$ is a modeling choice, not a given: contiguity (areas that share an edge),
$k$-nearest-neighbors, or distance-band schemes each encode a different theory of
what *neighbor* means, and the choice can move the conclusion. The code below builds
a contiguity weights matrix from the grid with base `sf` and computes Moran's $I$
together with a permutation reference distribution, so the result needs no spatial
econometrics package.

```{r geo-moran, message=FALSE, warning=FALSE}
# Contiguity weights: areas are neighbors if their polygons touch (share an edge
# or corner). st_intersects on polygons returns exactly this adjacency.
touch <- st_intersects(grid, grid)
m <- nrow(grid)
W <- matrix(0, m, m)
for (a in seq_len(m)) {
  nb <- setdiff(touch[[a]], a)      # neighbors of a, excluding itself
  if (length(nb) > 0) W[a, nb] <- 1
}
W <- W / pmax(rowSums(W), 1)        # row-standardize so weights per area sum to 1

x  <- grid$zone_spend
xc <- x - mean(x)
S0 <- sum(W)
moran_I <- (m / S0) * sum(W * outer(xc, xc)) / sum(xc^2)

# Permutation null: reshuffle attribute across areas, recompute I many times.
set.seed(56)
perm <- replicate(999, {
  xp <- sample(xc)
  (m / S0) * sum(W * outer(xp, xp)) / sum(xp^2)
})
p_perm <- (1 + sum(perm >= moran_I)) / (1 + length(perm))

cat("Moran's I          :", round(moran_I, 3), "\n")
cat("Expectation under H0:", round(-1 / (m - 1), 3), "\n")
cat("Permutation p-value :", round(p_perm, 3), "\n")
```

A Moran's $I$ comfortably above its null expectation, with a small permutation
$p$-value, says the spend surface is spatially clustered: knowing a zone's neighbors
tells you something about the zone. That diagnosis matters because it invalidates
ordinary regression. If the *residuals* of a model are spatially autocorrelated, the
ordinary-least-squares standard errors are wrong, typically too small, so $t$-ratios
are inflated and spurious significance follows. The remedy is **spatial regression**,
which writes the dependence into the model rather than leaving it in the error.

Two canonical specifications dominate (Anselin 1988, *Spatial Econometrics: Methods
and Models*, Springer, doi:10.1007/978-94-015-7799-1). The **spatial lag** model
adds a spatially weighted average of the outcome itself as a regressor,

$$
y = \rho W y + X\beta + \varepsilon,
$$ {#eq-geo-lag}

capturing genuine spatial *spillover* (an outcome in one area drives the outcome in
its neighbors, as when a store's promotion lifts traffic next door). The **spatial
error** model instead places the dependence in the disturbance,

$$
y = X\beta + u, \qquad u = \lambda W u + \varepsilon,
$$ {#eq-geo-error}

appropriate when the clustering reflects omitted, spatially smooth variables (local
income, weather, culture) rather than true interaction. Distinguishing the two is a
substantive question with a real managerial difference, spillover implies that
intervening in one area changes its neighbors, omitted-variable clustering does
not, and **local** indicators of spatial association (Anselin 1995, *Geographical
Analysis* 27(2), 93-115, doi:10.1111/j.1538-4632.1995.tb00338.x) decompose the
global Moran's $I$ into area-specific hot spots and cold spots that map where the
clustering lives. Estimation of @eq-geo-lag and @eq-geo-error uses maximum
likelihood or generalized method of moments and is the province of dedicated
spatial-econometrics packages; the modeling discipline, diagnose autocorrelation
first, then choose lag versus error on theory, is what the analyst owns.

::: {.callout-warning}
Spatial autocorrelation and the **modifiable areal unit problem** (MAUP) are
distinct hazards that compound. MAUP is the dependence of results on the arbitrary
boundaries and scale of the areal units: aggregate the same points to coarser zones,
or shift the grid, and correlations, regression coefficients, and even their signs
can change. The grid in @sec-geo-aggregation was a choice; a different grid would
give a different Moran's $I$. Robust spatial conclusions are checked across
sensible alternative aggregations, never reported from a single arbitrary
partition.
:::

## Geo-Experiments and Geo-Lift {#sec-geo-experiments}

The cleanest causal use of geography turns regions into experimental units. In a
**geo-experiment**, a set of geographic markets (designated market areas, metros,
or zip clusters) is randomized into treatment and control, the campaign runs only in
treated markets, and the incremental outcome, **geo-lift**, is the difference in
aggregated response between treated and control geographies, net of their baseline
trajectories. Geography is an attractive randomization unit precisely because of the
spillover this chapter has emphasized: an individual-level holdout leaks, since a
treated user talks to an untreated neighbor and a control user still sees the
billboard, whereas a market-level holdout contains most of the contamination inside
the treated region. The cost is statistical power, there are far fewer markets than
users, so geo-designs lean on pre-period data, matched market pairs, and
time-series counterfactuals to squeeze precision from a small number of large units.

The estimand is an incrementality, and the workhorse counterfactual is a synthetic
control or a Bayesian structural time-series model that predicts each treated
market's untreated trajectory from a weighted basket of control markets, attributing
the gap after launch to the campaign. The simulation below builds a minimal
geo-lift design, half the markets treated with a known multiplicative lift, and
recovers the effect with a difference-in-differences contrast on the pre/post market
panel, the simplest unbiased geo-lift estimator when assignment is random.

```{r geo-lift, message=FALSE, warning=FALSE}
set.seed(56)
G  <- 40                                   # number of geographic markets
Tn <- 24                                   # weeks; campaign launches at week 13
true_lift <- 0.08                          # +8% in treated markets, post-launch

market   <- rep(1:G, each = Tn)
week     <- rep(1:Tn, times = G)
treated  <- rep(rbinom(G, 1, 0.5), each = Tn)     # random assignment of markets
post     <- as.integer(week >= 13)
base     <- rep(rlnorm(G, log(1000), 0.4), each = Tn)  # market size heterogeneity

# Baseline sales: market level + common weekly trend + noise.
sales <- base * (1 + 0.01 * week) * exp(rnorm(G * Tn, 0, 0.05))
# Treatment effect: multiplicative lift applied only to treated markets post-launch.
sales <- sales * (1 + true_lift * treated * post)

geo <- data.frame(market, week, treated, post, sales)

# Difference-in-differences on log sales: the interaction is the geo-lift.
fit <- lm(log(sales) ~ treated + post + treated:post, data = geo)
est_lift <- exp(coef(fit)["treated:post"]) - 1

cat("True geo-lift      :", true_lift, "\n")
cat("Estimated geo-lift :", round(est_lift, 4), "\n")
cat("DiD interaction SE :",
    round(summary(fit)$coefficients["treated:post", "Std. Error"], 4), "\n")
```

The difference-in-differences interaction recovers the planted lift because random
assignment of markets makes treatment independent of the baseline trajectory, so the
common-trend assumption holds by construction. Real geo-experiments are harder: a
firm rarely controls enough comparable markets to randomize cleanly, market sizes
are wildly heterogeneous (the `base` term above), and contamination across adjacent
markets reintroduces the very spillover the design was meant to contain. These are
why production geo-lift leans on matched-market and synthetic-control machinery
rather than a bare difference of means, and why power analysis precedes any
geo-test. The conceptual payoff remains: geography converts an observational
targeting problem into a randomized one, complementing the user-level experimentation
of @sec-causal-inference at a spatial scale that respects spillover.

## Mobile-Location Data and Privacy {#sec-geo-privacy}

The richest contemporary location signal comes from mobile devices, and it is also
the most ethically and legally fraught. A smartphone reports position through GPS,
cell-tower triangulation, Wi-Fi and bluetooth scans, and the IP address of its
connection, at an accuracy ranging from a few meters outdoors to tens of meters
indoors. Software development kits embedded in apps stream these coordinates to data
brokers, who assemble device-level mobility traces, sequences of timestamped
locations, and resell them for footfall measurement, attribution, audience building,
and the hyper-contextual targeting that conditions an ad on where and when a person
is (Andrews, Luo, Fang, and Ghose 2016, *Marketing Science* 35(2), 218-233,
doi:10.1287/mksc.2015.0905; Luo, Andrews, Fang, and Phang 2014, *Management Science*
60(7), 1738-1756, doi:10.1287/mnsc.2013.1836).

The privacy hazard is acute and specific to location. Mobility traces are *uniquely
identifying* even when stripped of names: a handful of timestamped points, typically
home at night and work by day, single out an individual from millions, so
"anonymized" location data is anonymized in name only. Location also exposes
sensitive inference, visits to clinics, places of worship, protests, or a
competitor's headquarters, that consumers never consented to reveal. These
properties have made location a focus of regulation. The EU General Data Protection
Regulation treats precise location as personal data requiring a lawful basis and,
often, explicit consent; the California Consumer Privacy Act and its successors
grant rights to access and delete it; mobile platforms have tightened operating-system
permissions toward per-app, per-session, and coarse-location options; and
enforcement actions against brokers trading in sensitive-place visits are
accelerating. The methodological response is to design for privacy: aggregate to
areas large enough to prevent re-identification (the points-to-zones move of
@sec-geo-aggregation is itself a privacy control), add noise under formal
**differential privacy** guarantees as national statistical agencies now do for
census geographies, and prefer on-device or federated computation that never
centralizes raw traces. These tools connect directly to the broader treatment of
consumer privacy in @sec-privacy; here the point is that location data's analytical
power and its privacy risk are two faces of the same property, that *where* you are
is a near-complete description of *who* you are.

## Industry and Production Practice {#sec-geo-production}

Translating these methods into deployed systems means confronting the data-vendor
ecosystem and the engineering stack that supports spatial work at scale.

**Location-data vendors.** The supply chain has tiers. *Base-map and boundary*
providers supply the geometries (administrative boundaries, road networks, points of
interest) on which everything else is layered, ranging from open data such as
OpenStreetMap and national census bureaus to commercial map platforms.
*Geodemographic* vendors sell area-level lifestyle classifications and consumer
profiles keyed to postal geography. *Mobile-location and footfall* vendors sell
device-level traces and visit panels derived from app SDKs, used for foot-traffic
analytics and attribution. *Routing and accessibility* providers expose travel-time
and isochrone APIs that replace Euclidean distance with network reality. A recurring
production caution is **provenance and consent**: footfall panels vary enormously in
coverage, representativeness, and the legality of their consent chain, so a panel's
device sample must be weighted to the population it claims to represent and audited
for the lawful basis on which its locations were collected.

**The spatial stack.** Production geospatial analytics typically combines a
spatial database (PostgreSQL with the PostGIS extension, or the spatial functions
now native to cloud data warehouses) for storage and joins at scale, a tiling and
visualization layer for interactive maps, and an analytical layer in `R` (`sf`,
`terra`, `spdep`) or Python (`geopandas`, `shapely`, `pysal`) for modeling. The
`sf` package this chapter uses is the analytical front end of that stack and
interoperates with the same simple-features standard the database speaks, so a
trade-area model prototyped in `R` ports to a warehouse query with the geometry
semantics preserved (Bivand, Pebesma, and Gómez-Rubio 2013, *Applied Spatial Data
Analysis with R*, Springer, doi:10.1007/978-1-4614-7618-4).

**Engineering pitfalls.** Three errors dominate production incidents. The first is
**CRS mismatch**, overlaying or measuring across layers in different coordinate
systems, which the discipline of @sec-geo-representation prevents. The second is
**geocoding error**: converting addresses to coordinates is imperfect, and a
mis-geocoded customer silently corrupts every downstream catchment and choropleth,
so geocode quality must be scored and low-confidence matches flagged. The third is
**scale and aggregation sensitivity**, the modifiable areal unit problem of
@sec-geo-autocorrelation surfacing as a dashboard whose conclusions flip when the
zoom level or zone definition changes. Each is mundane and each routinely sinks an
otherwise sound spatial analysis.

::: {.callout-tip}
## Replication resources: spatial analysis

Every spatial computation in this chapter—distance matrices, catchment assignment, the
Huff model, choropleth aggregation, Moran's $I$ with a permutation null, and the
difference-in-differences geo-lift—runs on the open R packages `sf` (simple features),
`spdep` (spatial weights and autocorrelation), and `ggplot2`, the de facto standard
spatial stack and the reproducible core of the chapter. The empirical mobile-targeting
studies cited above (e.g., @ghose2019mobiletrajectory, @luo2014mobiletargeting,
@andrews2016crowdedness, @dube2017pricetargeting) rely on proprietary carrier or
platform location data that cannot be redistributed, so they rarely ship public
replication packages; their value here is as design templates, and any data/code link
should be confirmed on the article page rather than assumed.
:::

## Frontier and Expansion {#sec-geo-frontier}

Several directions are actively reshaping geospatial marketing. **Trajectory and
mobility modeling** moves beyond static visit counts to the *sequence* of places a
device traverses, enabling path-to-purchase analysis in physical space that mirrors
the clickstream models of online behavior, and inviting the sequence methods of the
text and behavioral-data literatures to operate over geographies rather than tokens.
**Spatial machine learning** brings graph neural networks and geographically
weighted models to demand prediction, treating areas as nodes in a spatial graph so
that the autocorrelation of @sec-geo-autocorrelation becomes an architectural
feature rather than a nuisance to be corrected. **Satellite and aerial imagery**
turns the image-as-data methods of @sec-image-processing onto the earth itself,
estimating economic activity, parking-lot fullness, and construction from pixels,
a remote-sensing channel that needs no consenting device and so sidesteps part of
the privacy problem while raising new ones. **Privacy-preserving geospatial
analytics**, differentially private aggregates, on-device computation, and synthetic
mobility data, is becoming a precondition rather than an enhancement as regulation
tightens. And the **fusion** of location with the other modalities of this part,
geotagged images, geolocated text, footfall paired with transaction logs, points
toward the multimodal models in which *where* is one channel among several, each a
generated feature carrying its own measurement error. The through-line is the one
this chapter opened with: location is data, near things are more related than
distant ones, and the methods that turn that fact into insight are the same ones
that, used carelessly, turn geography into spurious findings or consumers into
surveilled subjects. Handling both responsibly is what spatial marketing science now
requires.
