# Sensor, Biometric, and Neurophysiological Data {#sec-biometric}

Marketing has always wanted to measure what consumers will not, or cannot, tell it.
A questionnaire asks a respondent to introspect, to translate a fleeting reaction into
a number on a scale, and to do so honestly. Every step of that translation leaks
information: attention wanders before it is reported, arousal subsides before it is
named, social desirability rewrites the answer, and much of what actually drives
choice never reaches conscious access at all. Sensor, biometric, and
neurophysiological measurement is the attempt to bypass the self-report bottleneck by
recording the body and the brain directly while a consumer looks at an advertisement,
walks a store aisle, tastes a product, or decides whether to buy. The promise is a
window onto attention, emotion, and valuation that is continuous, time-resolved, and
less filtered by deliberation than anything a survey can deliver. The hazard, equally
real, is that the window is narrow, expensive, easily over-interpreted, and ringed
with ethical questions that a click-through rate never raised.

This chapter treats the family of physiological signals as one more modality of
unstructured marketing data, a sibling to text (@sec-text-as-data) and images
(@sec-image-processing). As with those modalities, the intellectual move is to turn
something that is not natively a number, a gaze trajectory, a trickle of sweat, a
voltage at the scalp, into a feature vector that can enter the demand systems, choice
models, and regressions the rest of this book develops. And as with those modalities,
the central discipline is measurement: defining the construct first, choosing a signal
and a set of features whose link to that construct is defensible, and validating
against behavior before any causal claim is made. The difference is that here the
sensor sits on the consumer's body, which raises the stakes for cost, consent, and
credibility far above what scraping reviews or product photos ever does.
@fig-bio-chain lays out that common measurement chain, the same steps every signal in
this chapter passes through.

```{mermaid}
%%| label: fig-bio-chain
%%| fig-cap: "The measurement chain common to every physiological signal. A raw recording is preprocessed, reduced to features, mapped to a psychological construct, and validated against behavior before any causal claim."
flowchart LR
  RS["Raw signal<br/>(gaze, skin conductance, scalp voltage)"]
  RS --> PP["Preprocessing"]
  PP --> FE["Feature extraction"]
  FE --> CI["Psychological construct<br/>(attention, emotion, valuation)"]
  CI --> VB["Validation<br/>against behavior"]
```

The umbrella term in industry is **neuromarketing**, though the toolkit is broader
than the brain. It spans the eyes (eye-tracking), the autonomic nervous system
(electrodermal activity, heart rate, facial muscle activity), and the central nervous
system (electroencephalography and functional magnetic resonance imaging), increasingly
supplemented by consumer wearables that stream heart rate and motion outside the lab.
The academic literature that anchors this chapter ranges from the foundational survey
of eye-tracking in marketing [Wedel and Pieters 2008, *Foundations and Trends in
Marketing*, doi:10.1561/1700000011] to the integrative demonstration that
neurophysiological measures can predict aggregate advertising response beyond
traditional metrics [Venkatraman et al. 2015, *Journal of Marketing Research*,
doi:10.1509/jmr.13.0593], with a sober counterweight in the review that named the
field's "hope and hype" problem [Ariely and Berns 2010, *Nature Reviews
Neuroscience*, doi:10.1038/nrn2795]. The maturation of consumer neuroscience as a
field is charted by a sequence of agenda-setting reviews—@yoon2012decisionneuro on the
bridge from decision neuroscience to consumer choice, @smidts2014advancing on the
field's research priorities, @plassmann2015consumerneuro on its applications and
methodological pitfalls, and @karmarkar2019pastpresentfuture on its past, present, and
future—which together situate these signals within the broader unstructured-data program
[@balducci2018unstructured] that organizes this part.

The chapter proceeds from the periphery inward and from the concrete to the
conceptual. It first lays out the applications that motivate the methods. It then takes
each signal in turn, eye-tracking, scroll and keystroke traces from ordinary devices,
autonomic measures, and the electroencephalogram, developing the features that matter and giving a small,
fully runnable simulation that builds the relevant quantities from scratch in base R
and ggplot2. It treats functional neuroimaging and modern neural decoding
conceptually, because those require hardware and data this book cannot simulate
honestly, and it labels that boundary clearly. It closes on the three questions that
decide whether a physiological study is worth running at all: is it valid, can it be
afforded, and is it ethical, before surveying what neuromarketing vendors actually
sell and where the credible line falls.

## What the Signals Measure and Why Marketing Wants Them {#sec-bio-applications}

The physiological toolkit maps onto a small set of marketing constructs that
self-report measures poorly. It is worth fixing that mapping before descending into any
one signal, because the recurring error in applied neuromarketing is to record a signal
and then reach for whichever construct tells the most flattering story.

**Visual attention** is the province of eye-tracking. Where the eyes land, in what
order, and for how long is the closest observable proxy for what a consumer is
processing at a given instant, on the working assumption, the eye-mind hypothesis,
that the point of gaze and the locus of attention largely coincide during active
viewing. Marketing cares because attention is the scarce resource that all advertising
competes for: an ad that is never fixated cannot persuade, a package that loses the
shelf-scan race is never considered, and a web layout that buries the call to action
forfeits conversions. The canonical applications are advertising layout, where
eye-tracking quantifies how brand, pictorial, and text elements capture and transfer
attention [Pieters and Wedel 2004, *Journal of Marketing*,
doi:10.1509/jmkg.68.2.36.27794]; packaging and shelf design, where gaze data reveal
which products win the first fixation and the considered set; and digital interfaces,
where it diagnoses whether users even see the elements designers intended.

**Emotional arousal**, the intensity dimension of affect, is read from the autonomic
nervous system. Electrodermal activity (EDA), historically the galvanic skin response
(GSR), tracks sweat-gland activity driven by sympathetic arousal and rises within a
couple of seconds of an emotionally engaging or surprising stimulus. It is cheap,
robust, and a clean index of *how much* a consumer is aroused, but it is blind to
*valence*: a delighted viewer and a disgusted one can show the same skin-conductance
spike. Heart rate and heart-rate variability add a slower autonomic channel. Arousal
matters to marketing because emotional intensity predicts memory encoding, ad
likeability, and the sharing of content; the EDA trace over a thirty-second spot shows
exactly which beats land and which fall flat [Poels and Dewitte 2006, *Journal of
Advertising Research*, doi:10.2501/s0021849906060041].

**Emotional valence**, the pleasant-unpleasant sign of affect, is where autonomic
measures need help. Facial electromyography (fEMG) supplies it by recording the
electrical activity of specific muscles: the *zygomaticus major* that pulls the mouth
into a smile, indexing positive affect, and the *corrugator supercilii* that knits the
brow into a frown, indexing negative affect. Automated facial-expression coding from
video pursues the same target without electrodes, trading sensitivity for scale and
unobtrusiveness. Pairing EDA (arousal) with fEMG or facial coding (valence) recovers
the two-dimensional affect space marketers actually want.

**Attention, preference, and valuation in the brain** are the domain of EEG and fMRI.
Electroencephalography records voltage fluctuations at the scalp with millisecond
resolution, ideal for timing-sensitive questions: did the brain register this brand at
200 milliseconds, did this ad frame elicit the attention-and-surprise signature, is
left-frontal activity, an approach-motivation marker, stronger for this design.
Functional MRI trades EEG's temporal precision for spatial precision, localizing
activity to deep structures EEG cannot see. Two of those structures carry outsized
marketing weight: the **ventral striatum** (especially the nucleus accumbens), whose
activity tracks anticipated reward and has repeatedly forecast purchasing and
population-level success, and the **medial prefrontal / orbitofrontal cortex**, which
encodes subjective value. The landmark demonstrations are that nucleus accumbens and
medial-prefrontal activity predict whether individuals will buy a product
[Knutson et al. 2007, *Neuron*, doi:10.1016/j.neuron.2006.11.010] and that marketing
actions such as a stated price literally modulate the neural representation of
experienced pleasantness [Plassmann et al. 2008, *Proceedings of the National Academy
of Sciences*, doi:10.1073/pnas.0706929105].

**Behavior in the wild** is the frontier opened by consumer **wearables**. Smartwatches
and fitness bands stream heart rate, motion, and increasingly EDA continuously, outside
any lab, at population scale. They blur the line between a controlled physiological
study and observational sensor data, and they import the same generated-feature and
privacy cautions that govern the rest of this part, now attached to the consumer's
pulse.

@fig-bio-modalities organizes the toolkit by construct, grouping the instruments under
what each one measures.

```{mermaid}
%%| label: fig-bio-modalities
%%| fig-cap: "The physiological toolkit organized by construct. Each marketing construct is measured by one or more instruments, and credible designs pair an arousal channel with a valence channel."
flowchart TB
  P["Physiological signals"]
  P --> AT["Visual attention"]
  P --> AR["Emotional arousal"]
  P --> VA["Emotional valence"]
  P --> BN["Brain: attention,<br/>value, timing"]
  AT --> ET["Eye-tracking<br/>(fixations, dwell, scanpath)"]
  AR --> EDA["Electrodermal activity"]
  AR --> HR["Heart rate / variability"]
  VA --> EMG["Facial EMG"]
  VA --> FC["Facial-expression coding"]
  BN --> EEG["EEG<br/>(ERP, band power, asymmetry)"]
  BN --> FMR["fMRI<br/>(ventral striatum, mPFC/OFC)"]
```

::: {.callout-note}
## The two-dimensional logic of affect
A useful organizing frame is the **circumplex** of affect: emotion as a point in a
plane of *arousal* (calm to excited) and *valence* (unpleasant to pleasant). Each
peripheral signal measures roughly one axis. EDA and heart rate read arousal; facial
EMG and facial-expression coding read valence. Neither alone locates an emotion;
together they do. Most credible biometric ad-testing designs therefore pair an
arousal channel with a valence channel rather than betting a conclusion on one signal.
:::

@tbl-bio-signals fixes the signal-to-construct mapping that organizes the rest of the
chapter, pairing each instrument with the marketing construct it measures, its
characteristic features, and a representative study from the verified literature.

| Signal | Construct | Characteristic features | Representative study |
|---|---|---|---|
| Eye-tracking | Visual attention | Fixations, dwell, time-to-first-fixation, scanpath | @pieters2004attentioncapture |
| Keystroke dynamics | Familiarity, cognitive load | Hesitation, key and field transitions | @kim2026areyouyou |
| Electrodermal activity | Arousal (intensity) | Tonic level, phasic SCRs | @teixeira2012emotionengagement |
| Facial EMG / coding | Valence (sign) | Zygomaticus / corrugator activity | @teixeira2014 |
| EEG | Attention, timing, approach | ERP (N100/P300), band power, frontal asymmetry | @venkatraman2015predicting |
| fMRI | Reward, subjective value | BOLD in nucleus accumbens, mPFC/OFC | @knutson2007neuralpurchases |

: The physiological toolkit: each signal, the construct it indexes, its characteristic features, and a representative marketing study. Credible designs combine an arousal channel with a valence channel. {#tbl-bio-signals}

## Eye-Tracking: Gaze, Fixations, and Areas of Interest {#sec-bio-eyetracking}

Eye-tracking is the most mature and most defensible of the physiological methods in
marketing, with the longest record of converging validity against downstream behavior
[Wedel and Pieters 2008, *Foundations and Trends in Marketing*,
doi:10.1561/1700000011]. The eye does not glide smoothly over a scene. It moves in a
sequence of **fixations**, brief pauses of roughly 200 to 300 milliseconds during which
gaze is held nearly still and visual information is actually taken in, separated by
**saccades**, ballistic jumps of 30 to 80 milliseconds during which vision is largely
suppressed. Modern remote eye-trackers sample point of gaze at 60 to 1000 Hz; an
event-detection algorithm then parses the raw gaze stream into fixations and saccades,
typically by thresholding velocity or dispersion. Everything downstream is built on
that parse.

The marketing-relevant features fall into two groups. The first describes individual
fixations and the saccades between them: **fixation count** (how many times a region
was looked at), **fixation duration** (how long each look lasted, an index of
processing depth), **saccade amplitude and direction**, and the ordered **scanpath**
itself. The second aggregates fixations over semantically meaningful regions of the
stimulus, the **areas of interest** (AOIs). An AOI is a rectangle (or polygon) drawn
around the brand logo, the product, the price, the headline. For each AOI the analyst
computes **dwell time** (total fixation duration inside it), **fixation count**, and
**time to first fixation** (TTFF), how long before the region first drew the eye, a
measure of bottom-up salience. These per-AOI metrics are the currency of applied
eye-tracking: they turn a gaze movie into a small table that says where attention went,
in what order, and for how long.

The runnable demonstration below builds this pipeline end to end. It simulates a
sequence of fixations over a unit-square stimulus whose looks cluster around four
design elements, defines AOI rectangles, classifies each fixation into an AOI, and
computes dwell time, fixation count, and time to first fixation per AOI. No
eye-tracking hardware or proprietary software is involved; the point is to make the
arithmetic of AOI analysis fully transparent.

```{r bio-eyetracking-sim, message=FALSE, warning=FALSE}
library(ggplot2)
set.seed(57)

## --- 1. Simulate a scanpath: fixations clustered around design elements ---
n_fix   <- 45
centers <- list(logo = c(0.20, 0.80), product = c(0.55, 0.45),
                price = c(0.80, 0.20), copy = c(0.35, 0.25))
probs   <- c(logo = 0.18, product = 0.42, price = 0.22, copy = 0.18)

draw_aoi  <- sample(names(probs), n_fix, replace = TRUE, prob = probs)
clamp01   <- function(z) pmin(pmax(z, 0), 1)
fix_x     <- clamp01(vapply(draw_aoi, function(a) rnorm(1, centers[[a]][1], 0.06), numeric(1)))
fix_y     <- clamp01(vapply(draw_aoi, function(a) rnorm(1, centers[[a]][2], 0.06), numeric(1)))
dur_ms    <- round(rgamma(n_fix, shape = 2, scale = 110))   # fixation durations (ms)

fix <- data.frame(order = seq_len(n_fix), x = fix_x, y = fix_y, dur_ms = dur_ms)

## --- 2. Define areas of interest (AOIs) as rectangles ---
aoi <- data.frame(
  aoi  = c("logo", "product", "price", "copy"),
  xmin = c(0.05, 0.40, 0.68, 0.18), xmax = c(0.35, 0.70, 0.92, 0.52),
  ymin = c(0.65, 0.30, 0.08, 0.10), ymax = c(0.95, 0.60, 0.32, 0.40)
)

## --- 3. Classify each fixation into the AOI that contains it ---
classify <- function(x, y) {
  hit <- which(x >= aoi$xmin & x <= aoi$xmax & y >= aoi$ymin & y <= aoi$ymax)
  if (length(hit)) aoi$aoi[hit[1]] else NA_character_
}
fix$aoi <- mapply(classify, fix$x, fix$y)

## --- 4. Per-AOI metrics: dwell time, fixation count, time to first fixation ---
metrics <- do.call(rbind, lapply(aoi$aoi, function(a) {
  idx <- which(fix$aoi == a)
  ttff <- if (length(idx)) sum(fix$dur_ms[seq_len(min(idx) - 1)]) else NA_real_
  data.frame(aoi = a, fix_count = length(idx),
             dwell_ms = sum(fix$dur_ms[idx]), ttff_ms = ttff)
}))
metrics
```

The table is the deliverable an eye-tracking study hands a brand manager: the product
region wins the most fixations and dwell time, the price is reached quickly because it
sits at high contrast in the lower right, and the body copy, fixated late and briefly,
is at risk of being skipped. Plotting the scanpath over the AOIs makes the same point
visually and is how such results are usually communicated.

```{r bio-eyetracking-plot, message=FALSE, warning=FALSE, fig.width=6, fig.height=5}
ggplot() +
  geom_rect(data = aoi,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = aoi),
            alpha = 0.20, colour = "grey40") +
  geom_path(data = fix, aes(x, y), colour = "grey50", linewidth = 0.4) +
  geom_point(data = fix, aes(x, y, size = dur_ms), alpha = 0.7) +
  geom_text(data = aoi, aes(x = (xmin + xmax)/2, y = ymax + 0.02, label = aoi),
            size = 3, fontface = "bold") +
  scale_size_continuous(name = "Fixation (ms)", range = c(1.5, 7)) +
  coord_fixed(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(title = "Simulated scanpath over four areas of interest",
       x = NULL, y = NULL) +
  theme_minimal() + theme(legend.position = "right")
```

Two cautions travel with every AOI analysis. First, AOI metrics are only as meaningful
as the AOIs are well drawn; gerrymandering a rectangle to capture or exclude fixations
is the eye-tracking equivalent of p-hacking, and AOIs should be fixed before the data
are seen. Second, gaze is necessary but not sufficient for persuasion. A consumer can
fixate a disclaimer without reading it and read a headline without believing it.
Eye-tracking measures the *opportunity* to process, which is why the strongest designs
pair gaze with a downstream measure, recall, choice, or sales, rather than treating
dwell time as the outcome of interest in itself.

## Scroll Tracking: Attention in the Wild {#sec-bio-scroll}

Eye-tracking is the gold standard for visual attention, but it needs a camera, a
calibration, and usually a lab—so it cannot follow the millions of choices people make
by thumb on a phone. **Scroll tracking** is the naturalistic proxy that can. On a
vertical mobile or web interface, the option in the viewport *is* the option being
attended, so the scroll trajectory—which alternative is centered at each moment, how long
it dwells there, how fast the user scrolls, how often they reverse direction—is a covert,
hardware-free recording of the same attention dynamics an eye-tracker would capture,
gathered in the wild during a real choice.

@leppanen2026attention develops scroll tracking as a process-tracing method and validates
it against the leading models of how attention shapes choice. Across binary choices
(gambles and consumer goods) presented on a scrollable interface, the scroll-derived
metrics—dwell asymmetry, direction reversals, scroll speed—predict not just the choice
but the decision maker's subjective valuations, latent individual differences, and risk
attitudes, and they improve every evidence-accumulation model tested.

The model they improve is the **attentional drift-diffusion model** (aDDM) of
@krajbich2010visual, the dominant account of how gaze and choice intertwine and the
engine behind the eye-tracking-in-marketing lineage of @shi2013information. In the aDDM a
relative decision value $\text{RDV}$ accumulates until it hits a boundary, but the drift
at each instant is governed by *what is currently attended*: the attended option's value
drives accumulation at full weight while the unattended option's is discounted by
$\theta\in(0,1)$,
$$
\Delta\text{RDV}_t = d\big(v_{a(t)} - \theta\, v_{-a(t)}\big)\,\Delta t + \varepsilon_t,
\qquad \varepsilon_t \sim \mathcal{N}(0, \sigma^2\Delta t),
$$ {#eq-bio-addm}
with a choice made when $\text{RDV}$ reaches $+B$ or $-B$, $a(t)$ the option attended at
time $t$, and $d$ a drift scale. The consequence is a **gaze cascade**: because looking
at an option boosts its accumulation, the option one dwells on is the option one tends to
choose, over and above its value. Scroll tracking supplies exactly the $a(t)$ sequence
@eq-bio-addm needs—which option is on screen—without an eye-tracker.

The demonstration below simulates the aDDM with attention alternating as a shopper scrolls
between two options, then puts the paper's two questions to the synthetic data: does the
scroll-attention asymmetry predict choice, and do scroll metrics improve a choice model
built on value alone?

```{r bio-scroll-addm, message=FALSE, warning=FALSE}
set.seed(57)

# Attention (which option is on screen, i.e., where the shopper has scrolled) modulates
# the drift of accumulated evidence. Parameters are illustrative, in the aDDM's range.
theta <- 0.35     # attentional discount: unattended option's value is down-weighted
d     <- 0.0006   # drift scaling (per ms);  sdw = accumulation noise;  B = boundary
sdw   <- 0.02; B <- 1; dt <- 10; Tmax <- 1000

sim_trial <- function(vL, vR) {
  loc <- character(0); cur <- sample(c("L", "R"), 1)      # alternating "scroll" dwells
  while (length(loc) < Tmax) {
    fd  <- max(1L, round(rgamma(1, shape = 2, scale = 25)))
    loc <- c(loc, rep(cur, fd)); cur <- if (cur == "L") "R" else "L"
  }
  loc <- loc[seq_len(Tmax)]
  mu  <- ifelse(loc == "L", d * dt * (vL - theta * vR), d * dt * (theta * vL - vR))
  rdv <- cumsum(mu + rnorm(Tmax, 0, sdw * sqrt(dt)))
  hit <- which(abs(rdv) >= B)[1]; if (is.na(hit)) hit <- Tmax
  c(choiceL = as.integer(rdv[hit] > 0),
    dwellL  = sum(loc[seq_len(hit)] == "L") * dt,
    dwellR  = sum(loc[seq_len(hit)] == "R") * dt)
}

n  <- 1500
vL <- runif(n, 0, 4); vR <- runif(n, 0, 4)
sim <- as.data.frame(t(vapply(seq_len(n), function(k) sim_trial(vL[k], vR[k]), numeric(3))))
sim$val_diff   <- vL - vR
sim$dwell_diff <- (sim$dwellL - sim$dwellR) / (sim$dwellL + sim$dwellR)  # scroll asymmetry

# 1. The scroll (gaze) cascade: the option dwelt on is chosen more, beyond its value.
cat(sprintf("P(choose left | dwelt more on LEFT)  = %.2f\n", mean(sim$choiceL[sim$dwell_diff > 0])))
cat(sprintf("P(choose left | dwelt more on RIGHT) = %.2f\n", mean(sim$choiceL[sim$dwell_diff < 0])))

# 2. Do scroll metrics improve a choice model over value alone?
m0 <- glm(choiceL ~ val_diff,              family = binomial, data = sim)  # value only
m1 <- glm(choiceL ~ val_diff + dwell_diff, family = binomial, data = sim)  # + scroll attention
mcf <- function(m) 1 - m$deviance / m$null.deviance
cat(sprintf("\nValue-only model    : pseudo-R2 %.3f\n", mcf(m0)))
cat(sprintf("+ scroll attention  : pseudo-R2 %.3f   (LR chi-sq %.0f, df 1)\n",
            mcf(m1), m0$deviance - m1$deviance))
```

The dwelt-on option is chosen well above chance regardless of value—the gaze cascade, now
read from scroll rather than gaze—and adding one scroll-attention feature to a value-only
choice model lifts its fit sharply, the miniature of the paper's finding that scroll
metrics improve every drift-diffusion model. Two cautions match the eye-tracking ones.
Attention and preference are *mutually* causal—we dwell on what we come to prefer as much
as we prefer what we dwell on—so a scroll metric predicts choice without cleanly
explaining it; and screen position is a coarser attention proxy than foveation, since a
user can leave an option on screen while thinking of something else. Because scroll
tracking is also *covert*, it inherits in sharper form the consent and manipulation
concerns of @sec-bio-validity-ethics: a method that reads deliberation from an interface
the firm itself designs is as usable for exploitation as for understanding.

## Keystroke Dynamics: Who Is Typing the Form? {#sec-bio-keystroke}

Scroll tracking reads attention from an interface the consumer already uses. Keystroke
dynamics does the same for cognition, and its most valuable marketing use is defensive:
detecting identity fraud at the moment a new customer fills in a sign-up form. The
problem is a marketing problem before it is a security one. Every bank card, buy-now-pay-later
line, sign-up bonus, referral reward, and first-order discount is an acquisition offer,
and fraudsters harvest those offers with stolen or synthetic identities. Every check
that stops them also costs genuine applicants time, and some of those applicants abandon
the funnel. The firm is choosing a point on a frontier between fraud loss and conversion.

@kim2026areyouyou attack that frontier with a design-science study of keystroke
dynamics during account setup. Their starting point is that knowledge-based
authentication, the "which of these streets have you lived on" quiz, no longer works:
after breaches such as Equifax's exposure of 147 million people's records in 2017, the
authors report that fraudsters pass these screens 92% of the time against 46% for
genuine customers. Their system instead scores how a first-time user types and moves
through the platform's existing onboarding form, so every applicant is screened and no
one is asked to do anything extra. They evaluated and refined the design across four
experiments. The university's news release on the work reports four studies with more
than 1,000 participants in total, and a system 95.5% effective at telling people entering
their own information from people entering someone else's. It also describes the intended use as flagging roughly the
most suspicious tenth of applications for extra verification
[@byu2023keystrokes]. (Only the published practice abstract, the open conference version
of the study, and that news release were accessible when this section was written; the
paper's own tables are not reported here, and its data and code are not public.)

### Why Typing Gives the Impostor Away

The mechanism is motor memory. A person has typed their own name, birthday, and ZIP
code hundreds of times, so entering them runs on the fast, automatic processing that
dual-process accounts call Type 1. An impostor has to read, recall, or work out each
answer, which is slow, deliberate Type 2 processing, and the extra load shows up as
longer pauses before a field and longer gaps between keys. The conference version of the
study makes the test concrete [@kim2023typing]. Its 202 participants, almost all aged
18 to 24, filled in the same short form twice, once with their own details and once with
a fictional person's, in random order. Welch t-tests found slower key transitions for
the fictional identity on the first name ($p = 0.033$), date of birth, and ZIP code
($p < 0.001$), though not on the last name ($p = 0.20$), and both key transitions and
field transitions were slower at the form level ($p < 10^{-14}$). The study belongs to a
line of information-systems work by overlapping authors that reads cognitive and
emotional states from mouse-cursor movements: negative emotion [@hibbeln2017userfeeling],
concealed information [@jenkins2019sleight], and fraudulent insurance claims
[@weinmann2022righteous]. Keystrokes extend the idea to the device every form requires.

Three quantities carry the signal. For the keydown times $d_1 < \dots < d_n$ of the
content keys typed into field $f$, with $\text{focus}_f$ and $\text{blur}_f$ the times
the cursor entered and left it,
$$
h_f = d_1 - \text{focus}_f, \qquad
\kappa_f = \operatorname{median}_i\,(d_{i+1} - d_i), \qquad
\tau_{f \to g} = d^{(g)}_1 - \text{blur}_f ,
$$ {#eq-bio-keystroke}
the hesitation before typing, the median key transition within the field, and the
field transition to the next field $g$. None of them requires knowing *which* keys were
pressed, which matters for privacy.

This differs from classic keystroke authentication in one decisive way. Authentication
asks whether a typist matches an enrolled template, a password typed fifty times at
sign-up, and the benchmark comparison of detectors for that task assumes the template
exists [@killourhy2009comparing]. A first-time applicant has no template. Kim et al.'s
system must decide whether the typing looks *familiar with the content*, judged
against how genuine people in general type their own details. That template-free design
is what lets it screen the very first interaction, and it is the design the replication
below puts to the test.

### Where It Applies in Marketing

The same screen fits wherever a consumer types claimed facts about themselves.

- **Account opening and credit.** Card, loan, and buy-now-pay-later applications are the
  setting Kim et al. target: synthetic identities that pass document checks but are
  typed like strangers' details.
- **Promotion and referral abuse.** New-customer discounts and referral bonuses invite
  one person to open many accounts. Each account's details are, by construction, not the
  typist's own.
- **Survey and panel fraud.** Respondents who fake a qualifying age, ZIP code, or
  occupation to get into a paid study are doing what the impostors in these experiments
  do, which makes timing a covert complement to the screening tools of
  @sec-online-panels.
- **Bots and scripted sign-ups.** Scripted form fills leave timing that is too regular,
  or no key events at all, the form-level version of the bot signatures in
  @sec-clickstream.
- **Lead-generation quality.** Forms submitted with someone else's contact details
  inflate cost-per-lead metrics. A familiarity score lets a buyer price leads by
  quality.

The commercial path is already travelled. NeuroID, founded in 2015 at the University of
Arizona by Valacich and Jenkins (two of Kim et al.'s authors) to sell behavioral
analytics for fraud screening, was acquired by the credit bureau Experian in August
2024 [@tla2024neuroid]. That makes this a rare case of marketing-relevant academic
measurement turned into a product at scale. Readers should also weigh the paper's claims
the way @sec-bio-industry recommends for any vendor-linked evidence: by asking what was
validated, against what ground truth, and out of sample.

### Replication on Open Data

Kim et al.'s data are not public, but one closely related study's are. @monaro2018covert
had 40 lab participants answer 18 identity questions by typing into a box, half
truthfully and half using a fake identity they had memorized from a mock ID card. The
questions came in three kinds: *control* questions everyone answered truthfully (hair
colour), *expected* questions printed on the card (name, birth year, city), and
*unexpected* questions that follow from the card but were never rehearsed (zodiac sign,
age spelled out, region of birth). They trained four WEKA classifiers on five
per-person features (the share of answers with conceptual errors, readability-adjusted
response latency, time from first key to Enter, typing time per character, and the pause
before Enter), then tested them on 20 new lab participants and 151 recruited online.
The authors posted the data, the feature files, and the exact classifier settings
[@monaro2018covertcode]. Our script re-runs their analysis from those files.

@tbl-bio-keystroke-ttests reproduces their univariate tests exactly, and the conceptual
error counts by question type also match to the unit: liars erred on 81 of 120
unexpected questions against 3 of 120 for truth-tellers, and on 3 of 160 expected
questions against none.

```{r}
#| label: tbl-bio-keystroke-ttests
#| tbl-cap: "Monaro et al.'s (2018) Table 2 re-run from the posted data (40 lab participants, Welch t-tests, truth-tellers minus liars; Cohen's d with pooled SD). Every value matches the published table."
ks_ttests <- data.frame(
  Feature = c("Conceptual errors", "Response latency", "Response latency, readability-adjusted",
              "Prompt to Enter"),
  Published = c("t(21) = -10.57, d = 3.34", "t(31) = -6.34, d = 2.00",
                "t(30) = -6.48, d = 2.05", "t(26) = -5.46, d = 1.73"),
  Reproduced = c("t(21) = -10.57, d = 3.34", "t(31) = -6.34, d = 2.00",
                 "t(30) = -6.48, d = 2.05", "t(26) = -5.46, d = 1.73"))
knitr::kable(ks_ttests)
```

@tbl-bio-keystroke-classifiers compares the classifiers. Logistic regression, SMO and
LMT reproduce exactly on both held-out samples, including the 86% to 91% accuracy on
the online sample that is the study's most demanding test. The results that depend on
random draws differ by a few points: the cross-validated accuracy of Logistic regression
and RandomForest, and RandomForest's online ROC area. That is expected, because the WEKA build bundled with R (3.9.3) draws
different random streams from the 3.8/3.9 build the authors used.

```{r}
#| label: tbl-bio-keystroke-classifiers
#| tbl-cap: "Classifier accuracy, published / reproduced. Trained on 40 lab participants with WEKA defaults; tested by 10-fold cross-validation, on 20 new lab participants, and on 151 participants recruited online (86 liars). ROC area is WEKA's class-weighted figure on the online sample."
ks_clf <- data.frame(
  Classifier = c("Logistic", "SMO (linear SVM)", "LMT", "RandomForest"),
  `10-fold CV, %`  = c("90 / 92.5", "95 / 95", "97.5 / 97.5", "92.5 / 90"),
  `20 lab, %`      = c("100 / 100", "90 / 90", "90 / 90", "95 / 95"),
  `151 online, %`  = c("86.1 / 86.1", "88.7 / 88.7", "90.1 / 90.1", "90.7 / 90.7"),
  `ROC, online`    = c("0.930 / 0.930", "0.897 / 0.897", "0.959 / 0.959", "0.980 / 0.973"),
  check.names = FALSE)
knitr::kable(ks_clf, align = "lcccc")
```

### What Carries the Signal

The replication licenses two questions the published tables do not answer, and both
matter for anyone deploying Kim et al.'s idea. First, how much of the accuracy comes
from keystroke *timing*, and how much from the *content* check, the error count that
requires knowing the right answer? Second, does the timing signal survive when no
fraudster is ever labelled, as in a template-free screen?

```{r}
#| label: tbl-bio-keystroke-signal
#| tbl-cap: "What carries the signal on the 151 online participants. Supervised rows: WEKA Logistic trained on the 40 lab participants. One-class rows: z-scores of log timings against the 20 lab truth-tellers only, so no liar is ever seen in training. AUC treats ties as one half."
ks_signal <- data.frame(
  Model = c("All five features (the paper's model)", "Conceptual errors only",
            "Timing only (four features, no error count)", "Response latency only",
            "One-class: response latency", "One-class: all four timing features"),
  Training = c(rep("Supervised, 40 lab", 4), rep("Truth-tellers only", 2)),
  Accuracy = c(0.861, 0.901, 0.828, 0.815, NA, NA),
  AUC      = c(0.925, 0.940, 0.877, 0.874, 0.874, 0.674))
knitr::kable(ks_signal, digits = 3)
```

The content check does most of the work. The error count alone (AUC 0.940) beats the
full five-feature model (0.925), and the full model's predicted probabilities collapse
to nine distinct values: anyone who made an error is scored near certainty. That leaves
timing as a ranking tool for everyone who made no error, and it is a useful one. Four
timing features alone reach an AUC of 0.877. Response latency carries nearly all of
that (0.874), and it transfers without any fraud labels: a one-class screen calibrated
only on lab truth-tellers scores the online sample with the same AUC of 0.874. Pooling
all four timing features in that screen *hurts* (0.674), because typing time and the
pause before Enter barely separate liars online (single-feature AUCs of 0.63 and 0.48)
and dilute the latency signal. Template-free screening works here, but only for a
feature that has been checked one at a time.

@tbl-bio-keystroke-qtype splits the lab data by question type, which is where the
mechanism becomes visible.

```{r}
#| label: tbl-bio-keystroke-qtype
#| tbl-cap: "Where the signal lives: subject-level AUC for separating liars from truth-tellers, by question type (60 lab participants; each person's mean over questions of that type). Control questions were answered truthfully by everyone and act as a placebo."
ks_qtype <- data.frame(
  `Question type` = c("Control (placebo)", "Expected (on the card)", "Unexpected (must be derived)"),
  Errors = c(0.500, 0.567, 0.995), `Response latency` = c(0.449, 0.810, 0.959),
  `Typing time per character` = c(0.504, 0.752, 0.794), `Pause before Enter` = c(0.520, 0.630, 0.701),
  Deletions = c(0.587, 0.621, 0.632), check.names = FALSE)
knitr::kable(ks_qtype, digits = 3)
```

The placebo row passes: on questions everyone answered truthfully, no feature separates
the groups. On expected questions, the closest analogue to Kim et al.'s own-versus-other
identity fields, latency (0.81) and typing speed (0.75) already separate liars who had
*rehearsed* the fake identity, which is the motor-memory effect in pure form. On
unexpected questions the gap widens sharply. Median adjusted latency is 2.1 seconds for
truth-tellers and 4.5 seconds for liars, against 1.8 and 2.1 seconds on expected
questions. The practical lesson for form design is that one cheap follow-up question an
owner knows and an impostor must compute (an age, given the birth date just typed; a
state, given the ZIP code) does more than any amount of passive timing on copied fields.
Kim et al.'s conference study also added follow-up questions after the form was
submitted [@kim2023typing].

### The Step-Up Budget

A score is used by sending the riskiest share of applicants to a slower check, such as a
document scan, a selfie match, or a call-back, while everyone else onboards without
friction. Two numbers describe the result: the share of fraud caught, and the share of
step-ups that are actually fraud. Because the ROC curve does not depend on prevalence,
the replication's online scores give both for any fraud rate $\pi$: flagging a share
$k$ of all applicants means choosing the threshold where
$\pi\,\text{TPR} + (1-\pi)\,\text{FPR} = k$. @fig-bio-keystroke-budget traces this
frontier at a 2% fraud rate.

```{r}
#| label: fig-bio-keystroke-budget
#| fig-cap: "Fraud caught against the share of all applicants sent to step-up verification, at a 2% fraud rate. Scores from the replication's Logistic models (trained on 40 lab participants, applied to 151 online). Ties in the score are broken at random, which is why the five-feature curve rises linearly until its tie block is exhausted."
#| fig-width: 7
#| fig-height: 4.5
ks_budget <- data.frame(
  flag   = seq(0, 0.30, by = 0.02),
  all5   = c(0, 0.210, 0.420, 0.630, 0.826, 0.849, 0.849, 0.872, 0.895, 0.919, 0.942,
             0.942, 0.965, 0.965, 0.965, 0.965),
  timing = c(0, 0.246, 0.477, 0.477, 0.535, 0.628, 0.721, 0.744, 0.744, 0.744, 0.837,
             0.837, 0.860, 0.860, 0.872, 0.884))
plot(100 * ks_budget$flag, 100 * ks_budget$all5, type = "l", lwd = 2.5, ylim = c(0, 100),
     xlab = "Applicants sent to step-up (%)", ylab = "Fraud caught (%)",
     main = "The step-up budget at a 2% fraud rate")
lines(100 * ks_budget$flag, 100 * ks_budget$timing, lwd = 2.5, lty = 2, col = "grey45")
abline(v = 10, col = "grey80", lty = 3)
legend("bottomright", bty = "n", lwd = 2.5, lty = c(1, 2), col = c("black", "grey45"),
       legend = c("Timing + error check (paper's five features)", "Timing only"))
```

At a 2% fraud rate, stepping up 10% of applicants catches 85% of fraud with the
five-feature model and 63% with timing alone. Each still pays for that in friction:
only 17% (or 13%) of the stepped-up applicants are fraudsters, so five to seven genuine
customers face the extra check for every impostor caught. Whether that is worth it is a
customer-lifetime-value calculation (@sec-clv). With fraud loss $L$ per undetected
impostor, and an expected cost $c$ of stepping up a genuine customer (abandonment
probability times the customer's lifetime value, plus handling cost), stepping an
applicant up pays when
$$
\Pr(\text{fraud} \mid s)\, L \;>\; \bigl(1 - \Pr(\text{fraud} \mid s)\bigr)\, c ,
$$ {#eq-bio-stepup}
so the budget $k$ should be set from the ratio $c/L$ and the calibrated score, not fixed
by habit. These figures come from Monaro et al.'s small samples, not from Kim et al.'s
data, and should be read as an illustration of the trade-off rather than as the
system's operating characteristics.

### Try It Yourself {#sec-bio-keystroke-demo}

::: {.content-visible when-format="html"}
The demo below runs entirely in your browser. Type your own details in the first form
and the assigned identity in the second, then compare the two rhythms. It measures key
timings only, never which keys you press, and it sends nothing anywhere. Its last panel
lets you vary the fraud rate and the step-up budget along the curves in
@fig-bio-keystroke-budget. The demo can also be
[opened in its own tab](replication/keystroke-identity-fraud/demo.html).

```{=html}
<iframe id="keystroke-demo" src="replication/keystroke-identity-fraud/demo.html"
        title="Keystroke identity demo" loading="lazy"
        style="width:100%; height:1900px; border:1px solid #dee2e6; border-radius:6px;"></iframe>
<script>
(function () {
  var f = document.getElementById("keystroke-demo");
  f.addEventListener("load", function () {
    try {
      var d = f.contentWindow.document;
      d.documentElement.setAttribute("data-theme", "light");   // match the book's light theme
      // body height is the content's own height; documentElement.scrollHeight would include
      // the iframe's viewport and grow on every resize
      var fit = function () { f.style.height = Math.ceil(d.body.getBoundingClientRect().height + 2) + "px"; };
      fit();
      // observe from the iframe's own window: an observer created here does not track
      // layout changes inside the frame
      var RO = f.contentWindow.ResizeObserver;
      if (RO) new RO(fit).observe(d.body);
      window.addEventListener("resize", fit);
    } catch (e) { /* cross-origin or file:// preview: keep the fixed height */ }
  });
})();
</script>
```
:::

::: {.content-hidden when-format="html"}
The online edition of this book embeds an interactive demo here: readers type their own
details and an assigned identity, compare the two keystroke rhythms, and vary the fraud
rate and step-up budget along the curves in @fig-bio-keystroke-budget. The demo's source
is `replication/keystroke-identity-fraud/demo.html` in the book's repository.
:::

### From Paper to Production

@fig-bio-keystroke-pipeline lays out the production design that follows from the
evidence above, and the book's repository implements it in
`replication/keystroke-identity-fraud/`. A dependency-free browser script records only
timings and key *classes* (letter, digit, deletion, shortcut), never key identities, so
the payload cannot reveal what was typed. The server turns each session into the
quantities of @eq-bio-keystroke and scores them against a reference built from
applicants later confirmed genuine, separately by device class. Scores are one-sided
robust z-scores, so only slower-than-reference typing raises risk, and each score comes
with the features that drove it. A threshold set at a budget quantile routes sessions to
*allow*, *step-up*, *no signal* (autofill, password managers, and paste are legitimate
and carry no timing evidence), or *automation review* (synthetic key events). Confirmed
outcomes feed back to refit the feature weights. Monitors track score drift and compare
step-up rates across segments, and both should be checked before any threshold goes
live. The folder also holds the demo in @sec-bio-keystroke-demo.

```{mermaid}
%%| label: fig-bio-keystroke-pipeline
%%| fig-cap: "A production keystroke-screening pipeline for first-time applicants. Capture is timing-only; scoring is server-side and template-free; the decision is a budgeted step-up rather than a denial; outcomes, drift, and segment parity feed back into calibration."
flowchart LR
  CAP["Browser capture<br/>(timings + key class only)"] --> FEAT["Session features<br/>(hesitation, key and<br/>field transitions)"]
  REF["Reference from<br/>known-genuine applicants<br/>(per device class)"] --> SC["One-sided robust score<br/>+ reason codes"]
  FEAT --> SC
  SC --> POL{"Budget<br/>threshold"}
  POL -->|"most sessions"| AL["Allow"]
  POL -->|"riskiest k%"| SU["Step-up check"]
  POL -->|"no keystrokes"| NS["No signal:<br/>standard path"]
  POL -->|"synthetic events"| AU["Automation review"]
  SU --> OUT["Outcomes<br/>(KYC, chargebacks)"]
  AL --> OUT
  OUT --> REF
  OUT --> MON["Monitors:<br/>drift, segment parity"]
```

```{r}
#| label: bio-keystroke-scoring
#| eval: false
# Scoring new applicants with the repository's keystroke_score.R (base R + jsonlite).
source("replication/keystroke-identity-fraud/keystroke_score.R")

good <- do.call(rbind, lapply(genuine_payloads, ks_features))   # sessions later confirmed genuine
ref  <- ks_calibrate(good, by_device = TRUE)                    # robust reference per device class

recent <- ks_score(do.call(rbind, lapply(last_week_payloads, ks_features)), ref)
thr    <- ks_threshold(recent$score, budget = 0.10)             # step up roughly 10%
ks_parity(ks_decide(recent, thr), recent$device)                # check before going live

ks_score_json(incoming_json, ref, thr)                          # JSON in, JSON out, per request
# returns score, decision (allow / step_up / no_signal / step_up_automation), the features
# that drove the score, device class, keystroke count, and the reference and threshold used
```

Four cautions follow. First, the evidence is small and mostly young: Kim et al.'s
conference study used students, and Monaro et al.'s classifiers were trained on 40
people and excluded online participants who used phones or tablets. Device, age,
language, and motor disability all slow typing for innocent reasons, so per-segment
calibration and the parity check are requirements, not refinements, and a step-up must
stay a verification path rather than an automatic denial. Second, fraud rings adapt:
pasting stolen details produces no keystrokes, which is why "no signal" must route to the
standard path while the *share* of no-signal sessions is itself monitored. Third, the
capture is covert by design. It inherits every consent obligation of
@sec-bio-validity-ethics and @sec-privacy, and keystroke timing used to verify identity
may be regulated as biometric data in some jurisdictions. Fourth, the headline accuracy
figures for Kim et al.'s system come from a press release, and two of the authors
co-founded the company that commercialized it. Out-of-sample validation by
independent researchers is the missing piece.

The full replication script below downloads Monaro et al.'s posted data and prints every
table and the figure data in this section. The book does not run it at render time.

```{r}
#| label: bio-keystroke-replication
#| eval: false
#| file: replication/keystroke-identity-fraud/replicate_monaro2018.R
```

## Autonomic Measures: Electrodermal Activity and Facial EMG {#sec-bio-autonomic}

Below the eyes lies the autonomic nervous system, whose sympathetic branch responds to
emotionally engaging stimuli within seconds and outside voluntary control. The two
workhorse channels are electrodermal activity and facial electromyography, the first
indexing arousal and the second valence.

### Electrodermal activity and phasic peak detection

EDA is recorded as skin conductance, in microsiemens, from electrodes on the fingers or
palm. The signal decomposes into two components. The slow-moving **tonic** level (the
skin conductance level, SCL) drifts over tens of seconds and reflects general arousal
and individual baseline. Riding on top are **phasic** bursts, the skin conductance
responses (SCRs), each a rapid rise within 1 to 3 seconds of an arousing event followed
by a slower exponential recovery. The marketing payoff is in the phasic component:
counting SCRs and locating them in time tells the analyst *which moments* of an
advertisement or experience triggered sympathetic arousal. The standard pipeline
therefore separates phasic from tonic and then detects peaks in the phasic signal.

The simulation below generates a sixty-second EDA trace at 20 Hz: a slow sinusoidal
tonic drift, five phasic SCRs at known event times each modeled as the difference of
two exponentials (a fast rise and a slow recovery), and a little measurement noise. It
then recovers the phasic component by subtracting a smoothed (moving-average) estimate
of the tonic baseline and applies a simple threshold-and-local-maximum peak detector.
Recovering the events that were planted is the validation that the detector works.

```{r bio-eda-sim, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
set.seed(57)

## --- 1. Simulate an EDA / GSR time series ---
fs <- 20; T_sec <- 60
t  <- seq(0, T_sec, by = 1/fs)

tonic   <- 3 + 0.4 * sin(2 * pi * t / 55)            # slow drift (microsiemens)
scr_at  <- c(8, 17, 26, 38, 49)                      # true SCR onset times (s)
scr_amp <- c(0.9, 0.6, 1.3, 0.5, 1.0)                # true SCR amplitudes
phasic  <- numeric(length(t))
for (i in seq_along(scr_at)) {
  d <- t - scr_at[i]; d[d < 0] <- NA                 # response only after onset
  bump <- scr_amp[i] * (exp(-d / 4) - exp(-d / 0.7)) # rise (0.7s) + recovery (4s)
  bump[is.na(bump)] <- 0
  phasic <- phasic + bump
}
eda <- tonic + phasic + rnorm(length(t), 0, 0.03)

## --- 2. Recover phasic component: subtract a smoothed tonic estimate ---
win        <- 41                                     # ~2 s moving-average window
tonic_hat  <- stats::filter(eda, rep(1/win, win), sides = 2)
tonic_hat  <- as.numeric(tonic_hat)
phasic_hat <- eda - tonic_hat
phasic_hat[is.na(phasic_hat)] <- 0

## --- 3. Simple phasic peak (SCR) detection: local maxima above a threshold ---
thr  <- 0.05
peak <- rep(FALSE, length(phasic_hat))
for (i in 2:(length(phasic_hat) - 1)) {
  if (phasic_hat[i] > phasic_hat[i - 1] &&
      phasic_hat[i] >= phasic_hat[i + 1] &&
      phasic_hat[i] > thr) peak[i] <- TRUE
}
detected <- data.frame(t = t[peak], amp = phasic_hat[peak])
detected
```

```{r bio-eda-plot, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
eda_df <- data.frame(t = t, eda = eda, phasic_hat = phasic_hat)
ggplot(eda_df, aes(t)) +
  geom_line(aes(y = eda), colour = "grey30") +
  geom_line(aes(y = phasic_hat + 2.5), colour = "steelblue") +
  geom_point(data = detected, aes(x = t, y = amp + 2.5),
             colour = "red", size = 2.4) +
  geom_vline(xintercept = scr_at, linetype = "dotted", colour = "grey60") +
  annotate("text", x = 1, y = 2.55, label = "recovered phasic",
           hjust = 0, size = 3, colour = "steelblue") +
  annotate("text", x = 1, y = max(eda), label = "raw EDA",
           hjust = 0, size = 3, colour = "grey30") +
  labs(title = "Simulated EDA with recovered phasic peaks (SCRs)",
       subtitle = "Dotted lines mark the true SCR onsets; red points are detected peaks",
       x = "Time (s)", y = "Skin conductance (microsiemens)") +
  theme_minimal()
```

The detector recovers the five planted responses near their onsets. The exercise also
exposes the judgment calls that make EDA analysis non-trivial in practice: the tonic
estimator's window length, the detection threshold, and a minimum inter-peak interval
all trade sensitivity against false alarms, and motion artifacts in real recordings
produce spurious spikes that no naive detector distinguishes from genuine SCRs. The
central interpretive limit remains valence-blindness. An SCR says the consumer was
aroused at second 26; it cannot say whether the arousal was delight or alarm. That is
the gap facial measures fill.

### Facial electromyography and expression coding

Facial EMG places electrodes over specific muscles and records their electrical
activity, typically the *zygomaticus major* (smiling, positive affect) and the
*corrugator supercilii* (frowning, negative affect). Because these muscles respond even
to expressions too faint to see, fEMG is sensitive to low-intensity and even covert
affective reactions, which is its advantage over visible-expression coding. Its cost is
intrusiveness: electrodes on the face are incompatible with naturalistic viewing and
unusable at scale.

Automated **facial-expression coding** from ordinary video pursues the same valence
target without electrodes. A computer-vision model (the image-as-data machinery of
@sec-image-processing) locates the face, tracks landmarks, and classifies action units
or basic emotions frame by frame, yielding a valence time series alongside the EDA
arousal trace. The trade is sensitivity and rigor for unobtrusiveness and scale: webcam
emotion coding can run in a panelist's living room across thousands of viewers, but it
misses the subthreshold activations fEMG catches and inherits every bias of the
training data behind the classifier. The signal-processing logic of feature extraction
and thresholding is identical to the EDA demonstration above; only the sensor and the
construct change, from sympathetic arousal to facial valence.

## Electroencephalography and the Event-Related Potential {#sec-bio-eeg}

The electroencephalogram records voltage fluctuations at the scalp produced by the
synchronized activity of large populations of cortical neurons, sampled at hundreds of
samples per second across a montage of electrodes. Its defining strength is temporal
resolution: EEG sees the brain's response unfold millisecond by millisecond, which
makes it the instrument of choice for questions about the *timing* and *sequence* of
processing that fMRI, with its sluggish hemodynamic signal, cannot answer. Its weakness
is spatial resolution; scalp voltage is a blurred mixture of many sources, so EEG
localizes poorly.

Two feature families dominate marketing EEG. The first is **spectral power** in
frequency bands, the rhythmic content of the ongoing signal. Frontal **alpha**
asymmetry (relative left-versus-right frontal power in the 8 to 13 Hz band) is read as
an index of approach versus avoidance motivation, and is used to gauge whether a
stimulus is appetitive. **Theta** and **beta** power are linked to engagement and
attention; @eijlers2020neuralarousal show that an EEG-derived index of neural arousal
relates to advertising success, a band-power analogue of the EDA arousal trace. The
second family is the **event-related potential** (ERP), the focus of the
runnable demonstration below.

An ERP is the brain's averaged electrical response time-locked to a repeated event,
the onset of a brand logo, a product image, a price. Any single trial is buried in
noise: ongoing alpha rhythm, muscle activity, and measurement noise dwarf the few
microvolts of stimulus-evoked signal. The foundational trick is **signal averaging**.
If the evoked response is identical across trials while the noise is independent and
zero-mean, then averaging $n$ epochs (short windows time-locked to the event) preserves
the response and shrinks the noise standard deviation by a factor of $\sqrt{n}$. The
averaged waveform reveals a stereotyped sequence of peaks, components named by polarity
and latency: an **N100** (a negative deflection near 100 ms, early sensory processing),
a **P200**, and a **P300** (a positive deflection near 300 ms, linked to attention,
surprise, and stimulus evaluation, and of obvious marketing interest as a candidate
signature of "this caught me").

The simulation builds an ERP from scratch. It defines a ground-truth evoked waveform (a
negative bump near 100 ms and a larger positive bump near 300 ms), generates sixty
noisy single-trial epochs by adding that waveform to independent noise and an
ongoing-alpha oscillation of random phase, and then averages the epochs to recover the
ERP. The contrast between a single messy trial and the clean average is the entire
logic of evoked-response analysis made visible.

```{r bio-eeg-sim, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
set.seed(57)

## --- 1. Define the ground-truth evoked response (an N100 and a P300) ---
fs2     <- 250                                   # samples per second
ep_t    <- seq(-0.1, 0.6, by = 1/fs2)            # epoch window: -100 to 600 ms
erp_true <- function(tt)                          # microvolts
  -2.5 * exp(-((tt - 0.10)^2) / (2 * 0.015^2)) +  # N100
   4.0 * exp(-((tt - 0.30)^2) / (2 * 0.050^2))    # P300
mu <- erp_true(ep_t)

## --- 2. Simulate noisy single-trial epochs: signal + alpha + white noise ---
n_tr   <- 60
epochs <- matrix(rnorm(n_tr * length(ep_t), 0, 6), nrow = n_tr)   # white noise
epochs <- sweep(epochs, 2, mu, "+")                                # add evoked response
for (i in seq_len(n_tr))                                           # add 10 Hz alpha
  epochs[i, ] <- epochs[i, ] + 3 * sin(2 * pi * 10 * ep_t + runif(1, 0, 2 * pi))

## --- 3. Average across trials to recover the ERP ---
erp_hat <- colMeans(epochs)

c(P300_peak_uV = round(max(erp_hat), 2),
  P300_lat_s   = round(ep_t[which.max(erp_hat)], 3),
  N100_peak_uV = round(min(erp_hat), 2),
  N100_lat_s   = round(ep_t[which.min(erp_hat)], 3))
```

```{r bio-eeg-plot, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}
erp_df <- data.frame(t_ms = ep_t * 1000,
                     single = epochs[1, ],
                     average = erp_hat,
                     truth = mu)
ggplot(erp_df, aes(t_ms)) +
  geom_line(aes(y = single), colour = "grey75") +
  geom_line(aes(y = average), colour = "steelblue", linewidth = 1) +
  geom_line(aes(y = truth), colour = "black", linetype = "dashed") +
  geom_vline(xintercept = 0, colour = "grey50") +
  annotate("text", x = 100, y = -3.2, label = "N100", size = 3) +
  annotate("text", x = 300, y =  4.5, label = "P300", size = 3) +
  labs(title = "From one noisy trial to an averaged ERP",
       subtitle = "Grey: a single epoch. Blue: average of 60 epochs. Dashed: ground truth.",
       x = "Time relative to stimulus (ms)", y = "Amplitude (microvolts)") +
  theme_minimal()
```

A single epoch (grey) is dominated by alpha and noise; the evoked components are
invisible. The sixty-trial average (blue) recovers the N100 and P300 and tracks the
ground-truth waveform (dashed) closely. This is why ERP designs require many repeated
presentations and tight time-locking, and why a marketing ERP claim built on a handful
of trials should be distrusted. The applied payoff is that ERP components index
processes self-report cannot time: the P300's amplitude to a brand or price reflects how
much attention and evaluative resource the stimulus commanded, and frontal asymmetry
over the same epochs indexes approach motivation. Both have been used to compare
creative executions and to forecast choice, as the next section's prediction results
show.

## Functional Neuroimaging and Neural Prediction {#sec-bio-fmri}

The methods to this point can be simulated honestly in a few lines of R because their
signals are one-dimensional time series with well-understood structure. Functional
magnetic resonance imaging and modern neural decoding cannot, and this section treats
them conceptually and clearly labels them as such. fMRI does not measure neural firing
directly; it measures the **blood-oxygen-level-dependent (BOLD)** signal, a slow
hemodynamic proxy for local neural activity, across roughly a hundred thousand volumetric
pixels (voxels) of the brain, every couple of seconds. The analytic workflow,
preprocessing, fitting a general linear model that regresses each voxel's time series
on the experimental design convolved with a hemodynamic response function, and
correcting for the massive multiple-comparisons problem of testing every voxel, is a
field of its own and is not reproduced here.

::: {.callout-important}
## Conceptual section: not simulated
The fMRI and deep neural-decoding material below is described, not demonstrated.
Unlike the eye-tracking, EDA, and ERP sections, there is no honest way to simulate a
BOLD volume or a trained decoder in base R without manufacturing a result that would
mislead. These methods require specialized hardware, large datasets, and dedicated
pipelines. Treat the claims here as pointers to a literature, each anchored to a
verified source, rather than as reproducible code.
:::

What fMRI buys marketing is access to deep valuation circuitry that scalp EEG and
peripheral sensors cannot see. Two structures recur. The **ventral striatum / nucleus
accumbens** tracks anticipated reward and gain; the **medial prefrontal and
orbitofrontal cortex** encode integrated subjective value. The foundational consumer
result is that activity in these regions, measured while people viewed products and
prices, predicted their subsequent purchase decisions trial by trial
[Knutson et al. 2007, *Neuron*, doi:10.1016/j.neuron.2006.11.010]. A second landmark
showed the causal porousness of "experienced value" itself: telling subjects a wine was
expensive raised both their reported pleasantness and the orbitofrontal activity
encoding it, so a marketing action reshaped the neural signal of experience, not merely
its report [Plassmann et al. 2008, *PNAS*, doi:10.1073/pnas.0706929105]. The critical
review that maps and tempers this literature is the "branding the brain" survey
[Plassmann, Ramsoy, and Milosavljevic 2012, *Journal of Consumer Psychology*,
doi:10.1016/j.jcps.2011.11.010], with the broader field overview in Hubert and Kenning
[2008, *Journal of Consumer Behaviour*, doi:10.1002/cb.251].

The most consequential idea in this section is **neuroforecasting**: using neural
activity measured in a small sample to predict *aggregate, population-level* market
outcomes, often outperforming the same small sample's self-reports. The logic is that
brain signals from a handful of people capture a shared valuation process that
generalizes to the market, even when those same people's stated preferences do not. The
evidence is striking and converging. Ventral-striatal activity in a small group of
adolescents predicted the *national* sales of songs years later, while the group's own
likeability ratings did not [Berns and Moore 2012, *Journal of Consumer Psychology*,
doi:10.1016/j.jcps.2011.05.001]. Brain responses to movie trailers predicted both
individual preferences and *population-wide* box-office success [Boksem and Smidts 2015,
*Journal of Marketing Research*, doi:10.1509/jmr.13.0572]. Nucleus-accumbens activity
forecast the *market-level* funding success of online loan and crowdfunding requests
beyond what the requests' features or subjects' choices predicted [Genevsky, Yoon, and
Knutson 2017, *Journal of Neuroscience*, doi:10.1523/jneurosci.1633-16.2017]. And in the
study that ties the whole toolkit to marketing practice, a model combining
neurophysiological measures (including fMRI ventral-striatum activity) predicted the
real-world, market-level elasticity of advertising response beyond what traditional
self-report and behavioral measures achieved [Venkatraman et al. 2015, *Journal of
Marketing Research*, doi:10.1509/jmr.13.0593; @venkatraman2015predicting]. The pattern
generalizes across domains: affective neural responses forecast the market-level success
of online microlending appeals [@genevsky2015microlending], and a small "neural focus
group" predicted population-level responses to a public-health media campaign that the
participants' own self-reports missed [@falk2012neuraltopopulation]. @knutson2018neuroforecasting
distill the common logic—neural signals from a handful of people recover a shared
valuation process that aggregates to the market—into a general account of *neuroforecasting
aggregate choice*, which is the single most defensible claim the modality makes because it
rests on out-of-sample prediction rather than reverse inference.

EEG-based prediction sits alongside the fMRI results and, unlike them, *can* be related
to the runnable ERP simulation above. EEG features have been shown to predict
individuals' subsequent product choices out of sample [Telpaz, Webb, and Levy 2015,
*Journal of Marketing Research*, doi:10.1509/jmr.13.0564], and EEG has been used to
study how social context shapes neural responses to branded luxury goods
[Pozharliev et al. 2015, *Journal of Marketing Research*, doi:10.1509/jmr.13.0560].

The frontier beyond mass-univariate fMRI is **neural decoding**: multivariate and
deep-learning models trained to read a distributed pattern of brain activity and output
the stimulus or the choice that produced it, reversing the usual forward direction of
analysis. In principle a decoder could read preference or attention from a brain-state
pattern; in practice marketing decoding is constrained by tiny sample sizes, low
signal-to-noise, and the ever-present risk of overfitting a flexible model to a handful
of subjects. These methods are powerful and genuinely advancing, and they are also the
easiest place in the entire toolkit to fool oneself, which is the cue to turn to
validity and ethics.

## Validity, Cost, and Ethics {#sec-bio-validity-ethics}

No part of marketing measurement is more prone to over-claiming than this one, and the
discipline that keeps it honest rests on three questions. @fig-bio-design arranges those
questions as sequential gates a study must clear before it is worth running.

```{mermaid}
%%| label: fig-bio-design
%%| fig-cap: "Three gates for a physiological study. A proposal must clear validity, then cost and external validity, then ethics, before it is run and triangulated with behavior and field data."
flowchart LR
  SP["Physiological study<br/>proposal"]
  SP --> VD["Validity<br/>(validated construct-signal mapping,<br/>predictive over reverse inference)"]
  VD --> CO["Cost and external validity<br/>(expense, sample size, realism)"]
  CO --> ETH["Ethics<br/>(consent, autonomy,<br/>inference beyond consent)"]
  ETH --> RN["Run and triangulate<br/>with behavior and field data"]
```

**Validity.** The first temptation is **reverse inference**: observing activity in a
brain region and inferring the mental state usually associated with it. Because the
nucleus accumbens activates in reward anticipation, it is tempting to read accumbens
activity as "the consumer wants this." But brain regions are many-to-one: the same
region participates in many processes, so its activation does not uniquely imply any one
of them. Reverse inference is a probabilistic argument that is only as strong as the
selectivity of the region for the inferred process, and it is routinely abused in
popular neuromarketing. The corresponding peripheral error is reading valence into a
valence-blind arousal signal, calling an EDA spike "excitement" when it might be
discomfort. The defenses are the ordinary ones of measurement: pre-register the
construct-to-signal mapping, validate physiological features against behavioral ground
truth, and prefer **predictive** validity (does the signal forecast held-out choice or
sales) over storytelling. The neuroforecasting results earn their credibility precisely
because they predict out-of-sample market outcomes rather than narrating a scan. The
field's own cautionary literature is essential reading here [Ariely and Berns 2010,
*Nature Reviews Neuroscience*, doi:10.1038/nrn2795; Plassmann, Ramsoy, and Milosavljevic
2012, *Journal of Consumer Psychology*, doi:10.1016/j.jcps.2011.11.010].

**Cost and external validity.** The methods are ordered roughly by expense and
inversely by ecological realism. Eye-tracking and EDA are relatively cheap, portable,
and tolerant of naturalistic settings; EEG is moderate in cost but demanding in setup;
fMRI is extremely expensive (scanner time, physicists, small samples) and maximally
unnatural, requiring a motionless supine subject in a noisy magnet. The expense forces
small samples, which limits power and inflates the danger of false positives, and the
artificiality threatens external validity: a brain response to an ad viewed
horizontally in a scanner may not transfer to the same ad glimpsed on a phone. A
credible program treats physiological measures as one converging line of evidence
triangulated with behavior and field data, not as an oracle that licenses skipping the
market test.

**Ethics.** Physiological and neural data are intimate in a way click logs are not, and
three concerns deserve explicit weight.

::: {.callout-warning}
## The ethics of measuring the body and brain
- **Informed consent and bodily privacy.** Biometric and neural signals are sensitive
  personal data. Subjects must understand what is recorded, how it is stored, and how it
  will be used, and consent must be genuine rather than buried in a panel agreement.
  Wearable streams collected in the field make consent harder, not easier, because the
  consumer rarely registers that a heart-rate trace is being repurposed for marketing.
- **The autonomy and manipulation worry.** The stated aspiration of some neuromarketing,
  to find a "buy button" and bypass deliberation, is ethically loaded. Measurement that
  is benign for diagnosing an ad's clarity becomes troubling when its purpose is to
  exploit non-conscious vulnerabilities, and most troubling for vulnerable populations
  such as children. The capability is routinely overstated by vendors, but the
  *intent* still warrants scrutiny.
- **Inference beyond consent and regulation.** Physiological data can leak information
  the subject never meant to share, health conditions, neurological markers, affective
  states, raising re-identification and discrimination risks. Such data fall under
  strict regimes (special-category data under GDPR, biometric-privacy statutes), and the
  governance, retention, and consent obligations of @sec-privacy apply with full force.
:::

The honest summary is that the ethical floor for physiological marketing research is
higher than for behavioral data, the validity bar is harder to clear, and the cost is
steep, all of which is why the credible academic literature is comparatively small and
the commercial literature comparatively loud.

## Industry Practice: What Vendors Sell and What Is Credible {#sec-bio-industry}

Neuromarketing is a commercial industry as much as an academic field, and the two
should not be confused. A tier of vendors sells ad and shopper testing built on the
methods of this chapter: webcam-based eye-tracking and facial-expression coding at panel
scale, EDA and EEG bundled into "neuro" ad-testing products, and shelf and
packaging studies in physical or virtual stores. The credible core of this offering is
real and useful. Eye-tracking for attention diagnosis, EDA and facial coding for
moment-by-moment emotional response, and EEG engagement metrics for creative comparison
are legitimate tools when used to inform design, and they earn their keep by catching
problems, an unseen call to action, a flat emotional arc, a confusing layout, that
self-report panels miss.

The line between credible and incredible practice can be drawn sharply.

**What is credible.** Aggregate, relative, predictive use. Comparing two ad executions
on attention and arousal, validating a physiological index against actual recall or
sales, using a large panel and reporting effect sizes, and treating the measure as one
input among several. The peer-reviewed neuroforecasting literature is the proof of
concept that, done carefully, neural and physiological signals can predict market
outcomes [Venkatraman et al. 2015, doi:10.1509/jmr.13.0593; Boksem and Smidts 2015,
doi:10.1509/jmr.13.0572; Genevsky, Yoon, and Knutson 2017,
doi:10.1523/jneurosci.1633-16.2017].

**What is not.** Single-subject "brain reading," the claim of a literal buy button,
reverse-inference storytelling ("this lit up the pleasure center, so consumers love
it"), tiny unreplicated samples sold as definitive, and proprietary black-box indices
with no published validation. A useful diagnostic when evaluating a vendor is to ask the
measurement questions from this book: what is the construct, what is the signal, how was
the mapping validated, against what ground truth, at what sample size, and does the
index predict anything out of sample. Vendors with credible answers exist; vendors who
deflect to the mystique of neuroscience do not.

The practical posture for a marketing scientist is therefore neither credulity nor
dismissal. The signals are real, the academic results are genuine, and the commercial
hype is also real. The job is to use the methods where their validity is established,
to triangulate them with behavior and field experiments, and to discount any claim that
the brain or the body has revealed a truth no market test could confirm.

::: {.callout-tip}
## Replication resources: physiological signals

The eye-tracking, EDA, and ERP demonstrations in this chapter are fully reproducible in
base R and `ggplot2`—they are the honest, runnable artifacts of the modality. For real
data, the open ecosystem is mature: EDA decomposition (`Ledalab`, `neurokit2` in Python),
eye-movement event detection and AOI analysis (`PyGaze`, `gazepoint`/`eyetrackingR`),
and EEG/ERP processing (`EEGLAB`, `MNE-Python`) are standard. The keystroke section is the
exception among the empirical studies: Monaro et al.'s posted data reproduce their published
tests and held-out accuracies exactly, and the repository's
`replication/keystroke-identity-fraud/` folder adds a capture script, a tested scoring
module, and an interactive demo. By contrast, the other empirical
neuro studies cited here—being fMRI/EEG work with small samples of identifiable human
subjects, much of it predating data-sharing mandates—rarely ship public replication
packages; some authors share stimuli or aggregate data on request or via OpenNeuro, but a
URL should be confirmed on the article page rather than assumed.
:::

## Frontier and Expansion {#sec-bio-frontier}

Several developments are pushing this modality from a niche lab method toward a scalable
data stream, and each imports the cautions developed above into a new setting.

**Consumer wearables and ambulatory sensing** are the largest shift. Smartwatches and
bands already stream heart rate, motion, and increasingly EDA continuously, from
millions of consumers, outside any laboratory. This collapses the distinction between a
controlled physiological study and observational sensor data, offering arousal and
activity signals at population scale while raising acute consent and privacy questions,
because a consumer who bought a fitness tracker did not obviously agree to have their
pulse mined for ad response.

**Webcam and remote sensing** remove the hardware barrier for the cheaper signals.
Browser-based eye-tracking and facial-expression coding run on ordinary laptop cameras,
trading precision for reach and enabling physiological-style measurement on large online
panels. The same image-as-data pipeline of @sec-image-processing supplies the
expression classifier, with the same dependence on training data and the same risk of
biased inference.

**Virtual and augmented reality** offer a route around the external-validity problem.
Immersive headsets can place a shopper in a photorealistic virtual store while
recording gaze (eye-tracking is now built into many headsets), and increasingly heart
rate and EDA, in a setting far more naturalistic than a scanner yet fully instrumented
and controllable. VR shelf tests are a plausible near-term bridge between the realism of
the field and the measurement density of the lab.

**Multimodal fusion** is the unifying frontier and the link back to the rest of this
part. The richest physiological designs already combine channels, gaze for attention,
EDA for arousal, facial coding for valence, EEG for timing, and the natural next
step is to fuse them with the text, image, audio, and video features of the surrounding
chapters and the foundation-model methods that learn joint representations across
modalities. A consumer's reaction to a video advertisement is at once a gaze pattern, an
arousal trace, a facial-valence series, and a neural response, and modeling them
jointly, against the ad's own multimodal content, is where this chapter meets the
multimodal-fusion capstone of the part.

The throughline, finally, is the one that governs every modality in this part. A gaze
movie, a skin-conductance trace, a scalp voltage, and a BOLD volume are all turned into
feature vectors by lossy, assumption-laden procedures, and those vectors are generated
regressors whose errors can correlate with the outcomes they are used to explain. The
body and the brain offer marketing a signal of unusual intimacy and unusual promise.
They reward the researcher who defines the construct first, validates the measure
against behavior, respects the consumer whose physiology is being recorded, and
distrusts any story the data tell too easily.
