# Multimodal Fusion and Foundation Models {#sec-multimodal-fusion}

The preceding chapters of this part took the unstructured modalities one at a time:
text, images, audio, video, and the family of behavioral, geospatial, network, and
sensor signals. Each chapter ended at the same place. A raw artifact, whatever its
modality, was passed through a learned encoder and emerged as a vector of features
that the book's downstream methods then consumed as if they were ordinary covariates.
That recurring move is not incidental. It is the organizing fact of the entire part,
and this capstone chapter makes it explicit, generalizes it, and confronts what
happens when the artifacts that marketing actually cares about arrive in several
modalities at once.

Real marketing objects are rarely unimodal. A TikTok advertisement is moving image,
synchronized audio, on-screen text, a written caption, and a stream of viewer
comments. A product listing is a set of photographs, a title, structured attributes,
bullet-point copy, and a corpus of reviews. A livestream-commerce session is video,
the host's speech, a scrolling chat, and a clickstream of purchases. To model these
objects faithfully, an analyst cannot treat each modality as a separate study and hope
the pieces reconcile. The modalities must be brought into a single representation that
a predictor, a demand model, or a causal estimator can use. That bringing-together is
**multimodal fusion**, and the engines that increasingly perform it are **foundation
models**: large, pretrained, general-purpose networks that map text, images, and
their combinations into shared representation spaces.

This chapter proceeds from the concrete to the conceptual. It begins with the case for
fusing modalities at all. It then lays out the three canonical fusion strategies and
demonstrates two of them in genuinely runnable R, comparing early fusion against late
fusion on simulated multimodal customer data. From there it treats the representational
backbone of modern fusion: contrastive image-text models in the style of CLIP, and
the multimodal large language models that have absorbed them. It examines the use of
those models as measurement instruments, including the contested practice of treating a
language model as a synthetic survey respondent. It turns to demand and response
estimation when the right-hand side contains text and image features, and to the
production machinery (feature stores, embedding pipelines, foundation-model APIs) that
makes any of this operational at scale. It closes by stating the throughline that has
run silently beneath every chapter of the part, and by surveying the frontier.

## The Case for Fusion {#sec-mmf-case}

Why fuse at all? The first answer is **complementarity**. Distinct modalities carry
non-redundant information about the same underlying object, and a model with access to
all of them can resolve ambiguities that no single modality can. A product photograph
reveals form, color, and finish that the title never states; the title encodes brand,
model, and category that pixels render only implicitly; the reviews supply experiential
attributes (durability, fit, smell) that neither image nor title contains. Each
modality is a different, lossy projection of a richer latent object, and fusion is the
attempt to invert several projections jointly rather than one at a time.

The second answer is **disambiguation and grounding**. Modalities discipline one
another. A caption that reads "absolutely sick" is praise or complaint depending on the
image it sits beside; a smiling face in a frame reads as warmth or as sarcasm depending
on the words spoken over it. Audio prosody separates a sincere "great service" from a
withering one (the *how it is said* versus *what is said* distinction introduced in the
audio chapter). When modalities are modeled jointly, each constrains the interpretation
of the others, and the fused representation is less brittle than any unimodal one.

The third answer is **robustness**. Modalities fail independently. A listing may have
no review text, a video may have no speech, an image may be missing or corrupted.
A model that has learned to draw on whatever modalities are present degrades gracefully
when one drops out, whereas a unimodal pipeline simply goes blind. Missingness is the
norm in marketing data, not the exception, and a fusion architecture that tolerates it
is worth more in production than a marginally more accurate one that does not.

The fourth answer is **measurement reach**. Many marketing constructs are inherently
multimodal and cannot be measured from one channel. Ad creative "quality," brand
"warmth," influencer "authenticity," and listing "appeal" are perceived by consumers
through the simultaneous arrival of sight, sound, and language. A measure built on text
alone, or images alone, captures a shadow of the construct. Fusion is what lets the
analyst measure the construct as the consumer experiences it.

Against these benefits stands a real cost, and it is the same caution the entire part
has been building toward. Every modality enters the model as a *generated feature*: the
output of an upstream encoder, estimated with error, and that error can correlate with
the outcome. Fusing modalities multiplies the channels through which generated-regressor
problems enter. The case for fusion is strong, but it raises the stakes on validation
and identification rather than lowering them, a point @sec-mmf-demand develops in full.

## Fusion Strategies {#sec-mmf-strategies}

Given several per-modality representations, where in the modeling pipeline should they
be combined? Three strategies span the design space, distinguished by *when* fusion
occurs relative to the modality-specific processing.

**Early fusion** (feature-level fusion) concatenates the per-modality feature vectors
into one long vector and fits a single model on the combined input. If text yields a
$d_T$-dimensional embedding, image a $d_I$-dimensional embedding, and behavior a
$d_B$-dimensional vector, early fusion forms the $(d_T + d_I + d_B)$-dimensional stack
and learns a single predictor over it. Its virtue is that the model can learn arbitrary
cross-modal interactions directly: the effect of an image feature is free to depend on
a text feature. Its costs are dimensionality (the concatenated vector can be very wide,
straining the sample size), sensitivity to differences in scale and noise across
modalities, and intolerance of missingness, since a single absent block leaves a hole
in every observation's input.

**Late fusion** (decision-level fusion) fits a separate model per modality and combines
their *outputs*, typically by averaging the predicted scores or by training a small
**meta-learner** (a stacking model) on the per-modality predictions. Its virtues mirror
early fusion's vices: each modality model can be tuned to its own structure, modalities
of wildly different dimension and scale never have to share a feature space, and a
missing modality simply drops one input to the combiner rather than corrupting a shared
vector. Its limitation is that cross-modal interactions are captured only insofar as the
combiner can recover them from the per-modality scores; rich interactions that require
seeing raw features from two modalities together are unavailable, because each base model
has already collapsed its modality to a scalar before fusion.

**Joint fusion** (intermediate or embedding-level fusion) sits between the two. Each
modality is passed through its own learnable encoder, the resulting intermediate
representations are combined (by concatenation, by summation, or by cross-attention
between modalities), and the entire stack, encoders and combiner together, is trained
end to end against the final objective. Joint fusion is the architecture of essentially
all modern deep multimodal systems, because it learns *modality-specific* representations
and their *interaction* simultaneously, letting gradient signal from the task reshape
each encoder. Its cost is that it requires end-to-end differentiable training, large
data, and the engineering apparatus of deep learning; it is not, in general, something
one fits with a few lines of base R. The contrastive and transformer models of
@sec-mmf-clip and @sec-mmf-mllm are joint-fusion systems at scale.

A useful way to hold the three together: early fusion fuses *features*, late fusion
fuses *decisions*, and joint fusion fuses *representations*, learning those
representations as it goes. The runnable demonstration below contrasts the two that can
be built from off-the-shelf, maintained packages: early fusion with a penalized linear
model and late fusion with per-modality random forests combined by a stacking
meta-learner. Joint fusion is treated conceptually, because its honest demonstration
requires a deep-learning stack outside this book's runnable scope.

### A Runnable Early-versus-Late Demonstration {#sec-mmf-demo}

The demonstration is deliberately synthetic so that the data-generating process is
known and the comparison is interpretable. We simulate per-customer features standing
in for three modalities. A 20-dimensional **text-embedding block** stands in for, say,
a sentence-embedding of a customer's reviews and messages; a 15-dimensional
**image-feature block** stands in for a CNN or vision-transformer embedding of
customer-associated images; and an 8-dimensional **behavioral block** stands in for
clickstream and transaction summaries. Each block is generated from a single latent
driver plus independent noise, so that within a block the features are correlated
(as real embeddings are) and across blocks the signal is genuinely complementary: the
binary outcome (a conversion, say) depends on all three latents. This construction is
exactly the situation fusion is meant for, and it lets us see what each strategy
recovers.

```{r mmf-setup, message=FALSE, warning=FALSE}
set.seed(58)
suppressMessages({
  library(glmnet)        # penalized regression for early fusion
  library(randomForest)  # per-modality learners for late fusion
})

n      <- 1500            # customers
d_text <- 20             # text-embedding dimension
d_img  <- 15             # image-feature dimension
d_beh  <- 8              # behavioral-feature dimension

# One latent driver per modality. The outcome depends on all three,
# so no single modality is sufficient: this is the case for fusion.
z_text <- rnorm(n)
z_img  <- rnorm(n)
z_beh  <- rnorm(n)

# Each modality block: a latent loading plus independent noise, so that
# features within a block are correlated, as learned embeddings are.
make_block <- function(z, d, loading) {
  L <- rnorm(d, mean = 0, sd = loading)        # per-dimension loadings
  outer(z, L) + matrix(rnorm(n * d, sd = 1), n, d)
}

X_text <- make_block(z_text, d_text, loading = 0.9)
X_img  <- make_block(z_img,  d_img,  loading = 0.9)
X_beh  <- make_block(z_beh,  d_beh,  loading = 1.1)
colnames(X_text) <- paste0("txt", seq_len(d_text))
colnames(X_img)  <- paste0("img", seq_len(d_img))
colnames(X_beh)  <- paste0("beh", seq_len(d_beh))

# Binary outcome driven by all three latents (true complementarity).
eta <- 1.1 * z_text + 0.8 * z_img + 0.9 * z_beh - 0.5
y   <- rbinom(n, size = 1, prob = 1 / (1 + exp(-eta)))

# Train / test split.
train <- sample(seq_len(n), size = 1000)
test  <- setdiff(seq_len(n), train)

# Rank-based AUC (Mann-Whitney form); no extra package needed.
auc <- function(y_true, p_hat) {
  r  <- rank(p_hat)
  n1 <- sum(y_true == 1); n0 <- sum(y_true == 0)
  (sum(r[y_true == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
```

With the data in hand, **early fusion** concatenates the three blocks into one wide
matrix and fits a single elastic-net logistic model over it. The penalty matters here:
the concatenated input is 43-dimensional with correlated columns, and regularization is
what keeps a wide, partly redundant feature stack from overfitting. This is the typical
shape of a real fused-embedding regression, where the right-hand side is hundreds or
thousands of embedding dimensions wide.

```{r mmf-early, message=FALSE, warning=FALSE}
X_all <- cbind(X_text, X_img, X_beh)        # concatenation = early fusion

cv_early <- cv.glmnet(X_all[train, ], y[train],
                      family = "binomial", alpha = 0.5)
p_early  <- as.vector(predict(cv_early, X_all[test, ],
                              s = "lambda.min", type = "response"))

cat("Early fusion (elastic-net) AUC:", round(auc(y[test], p_early), 3), "\n")
```

**Late fusion** fits a separate random forest to each modality, then combines their
predictions. We show two combiners. The first is a simple average of the three
per-modality probabilities, the most common and most robust late-fusion rule. The
second is a **stacked** combiner: a logistic meta-learner trained on the base models'
predictions. Crucially, the meta-learner is trained on each forest's *out-of-bag*
predictions for the training customers, not on its in-sample fits, so that the stacking
model learns to weight modalities on honest, held-out signal rather than on the base
models' overfit training scores. This out-of-bag construction is what makes stacking
defensible rather than leaky.

```{r mmf-late, message=FALSE, warning=FALSE}
fit_rf <- function(X) {
  randomForest(x = X[train, ], y = factor(y[train]), ntree = 300)
}
prob_test <- function(model, X) {
  predict(model, X[test, ], type = "prob")[, 2]
}

rf_text <- fit_rf(X_text)
rf_img  <- fit_rf(X_img)
rf_beh  <- fit_rf(X_beh)

# Per-modality predictions on the test set.
p_text <- prob_test(rf_text, X_text)
p_img  <- prob_test(rf_img,  X_img)
p_beh  <- prob_test(rf_beh,  X_beh)

# Combiner 1: simple average of decisions.
p_late_avg <- (p_text + p_img + p_beh) / 3

# Combiner 2: stacked logistic meta-learner trained on OUT-OF-BAG
# predictions, so the stack does not see the base models' in-sample fits.
oob <- function(model) model$votes[, 2]
stack_train <- data.frame(
  y = factor(y[train]),
  a = oob(rf_text), b = oob(rf_img), c = oob(rf_beh)
)
meta <- glm(y ~ a + b + c, data = stack_train, family = "binomial")
p_late_stack <- as.vector(predict(
  meta, newdata = data.frame(a = p_text, b = p_img, c = p_beh),
  type = "response"
))

cat("Late fusion (average)  AUC:", round(auc(y[test], p_late_avg),   3), "\n")
cat("Late fusion (stacked)  AUC:", round(auc(y[test], p_late_stack), 3), "\n")
```

To make the value of fusion itself visible, we also report what a single modality
achieves on its own. The comparison is the point of the whole exercise: because the
outcome depends on all three latents, no unimodal model can reach what a fused model
reaches, and the gap between the best single modality and either fusion strategy is the
empirical case for fusing.

```{r mmf-compare, message=FALSE, warning=FALSE}
results <- data.frame(
  approach = c("Text only", "Image only", "Behavior only",
               "Early fusion (concat)",
               "Late fusion (average)", "Late fusion (stacked)"),
  AUC = round(c(
    auc(y[test], p_text), auc(y[test], p_img), auc(y[test], p_beh),
    auc(y[test], p_early), auc(y[test], p_late_avg), auc(y[test], p_late_stack)
  ), 3)
)
print(results[order(-results$AUC), ], row.names = FALSE)
```

The qualitative pattern is what the data-generating process guarantees and what the
literature reports on real data. Every unimodal model trails the fused models, because
each sees only one of the three drivers. Among the fusion strategies, early and late
fusion land close together on this clean synthetic problem; which one wins in practice
depends on the data. Early fusion tends to lead when cross-modal interactions are strong
and the sample is large enough to estimate a wide model, because only early (or joint)
fusion can see raw features from two modalities together. Late fusion tends to lead when
modalities are heterogeneous in scale and noise, when some modalities are frequently
missing, or when sample size is tight relative to the concatenated dimension, because
each base model is fit and regularized to its own modality and the combiner has few
parameters to estimate. The honest takeaway is not that one strategy dominates but that
the choice is an empirical, validated decision, and that fusing beats not fusing whenever
the modalities carry complementary signal.

## Shared Representations and Contrastive Models {#sec-mmf-clip}

Early and late fusion as demonstrated above take the per-modality representations as
*given*. The deeper question is where a shared, comparable representation across
modalities comes from in the first place. The answer that reorganized the field is
**contrastive image-text pretraining**, exemplified by CLIP (Contrastive
Language-Image Pre-training), introduced by Radford and colleagues at OpenAI in 2021
(ICML 2021; the foundational paper has no canonical Crossref DOI and is cited here by
name and venue). The idea is conceptually simple and is best understood as a learned,
*joint* embedding space.

CLIP trains two encoders together: an image encoder and a text encoder. It is fed a very
large corpus of image-caption pairs scraped from the web. For each batch, it computes an
image embedding for every image and a text embedding for every caption, and it optimizes
a contrastive objective: the embedding of an image should have high similarity (inner
product) to the embedding of *its own* caption and low similarity to the embeddings of
the *other* captions in the batch, and symmetrically for captions. After training on
hundreds of millions of pairs, the two encoders share a single embedding space in which
an image of a red sneaker and the text "a red sneaker" land near each other, while
unrelated images and texts land far apart.

Two properties make this transformative for marketing measurement. The first is **a
common space**: because images and text are embedded into the *same* geometry, the
cosine similarity between an image embedding and a text embedding is meaningful. One can
score how well a product photo matches the phrase "premium and minimalist," or rank
thousands of ad creatives by their alignment to a brand-attribute phrase, with no
labeled training data for that attribute. The second is **zero-shot transfer**: novel
categories can be scored by writing them as text prompts rather than by collecting
labeled examples, which collapses the cost of building an image classifier for a new
marketing construct from a labeling project to a sentence. Successor models in the same
family (for example SigLIP, which replaces the contrastive softmax with a sigmoid loss)
refine the training objective while preserving the shared-space property.

For the analyst, the practical consequence is that the per-modality feature blocks of
@sec-mmf-demo are increasingly drawn from a *single shared encoder family* rather than
from separate, incommensurable models. When the text embedding and the image embedding
already live in a common space, fusion becomes partly a matter of geometry rather than
of ad hoc concatenation, and the comparability that contrastive pretraining buys is
itself a form of joint fusion performed once, at scale, by the foundation model
provider. In marketing, vision-language models of this lineage have been shown to
classify marketing images effectively and to bridge the gap from image-only
convolutional pipelines to general-purpose foundation models [@witte2026language].

## Multimodal LLMs and the LLM-as-Instrument {#sec-mmf-mllm}

The contrastive models of the previous section produce *embeddings*. The next
development absorbs vision encoders into generative language models to produce
**multimodal large language models** (MLLMs): systems that ingest interleaved text and
images (and, increasingly, audio and video frames) and *emit text*, including
structured outputs. Architecturally, an MLLM typically couples a vision encoder of the
CLIP lineage to a transformer language model (the transformer architecture is due to
Vaswani and colleagues, NeurIPS 2017, and like CLIP is cited by name and venue rather
than by a canonical Crossref DOI), so that image patches are projected into the
language model's token space and processed alongside words. The GPT-4o-class, Claude,
and Gemini model families, together with open-weight vision-language models, are the
current instances. For the marketing researcher, what matters is less the architecture
than the new *interface* it offers: one can hand the model an image and a question
about it, or a product listing and an instruction to extract attributes, and receive a
structured answer.

This reframes the foundation model as a **measurement instrument**. Rather than training
a bespoke classifier for each marketing construct, the analyst prompts a general model
to read an artifact and emit a measure: the sentiment of a review, the attributes
visible in a product photo, the emotional arc of a video ad, the persuasion tactics in
ad copy. The text chapter introduced the LLM-as-annotator pattern; the multimodal
version extends it across modalities, and the integrative survey of machine learning and
artificial intelligence in marketing situates this shift within the broader arc of
connecting computational power to substantive marketing insight [@ma2020machine]. The
appeal is obvious: speed, breadth, and zero marginal labeling cost. The cautions are
equally real and are taken up in @sec-mmf-demand, because an LLM-derived measure is a
generated regressor par excellence, produced by an opaque model whose errors may be
systematic and outcome-correlated.

A more radical proposal pushes the instrument metaphor further: the **LLM as simulated
respondent**, or *silicon sampling*. Here the language model is prompted to role-play a
consumer with specified demographics and dispositions, and its responses are treated as
synthetic survey or experimental data, a "silicon sample" standing in for human
participants. The attraction is the prospect of cheap, fast pilot studies, pretesting of
stimuli, and exploration of segments that are expensive to recruit. The literature that
examines this practice in consumer and marketing research is explicit that it is a
frontier with serious hazards rather than a settled method, and it offers guidelines
accordingly [@sarstedt2024silicon]. The central concerns are that silicon samples can
homogenize away the heterogeneity that is the whole point of sampling, can reflect the
biases and the training-data demographics of the underlying model rather than any real
population, can be confidently wrong in ways that are hard to detect without the human
data they are meant to replace, and can drift as the underlying model is updated, so
that a "respondent" is not a stable object across time. The responsible posture treats
silicon samples as a hypothesis-generating and stimulus-pretesting tool whose outputs
require validation against human data before they bear any inferential weight, never as
a drop-in substitute for primary research.

## Demand and Response Estimation with Multimodal Data {#sec-mmf-demand}

The destination of all this representation work, for the empirical marketing scientist,
is usually a model of demand or response in which text and image features appear on the
right-hand side. A flagship example fuses the visual frames, the audio, and the spoken
and on-screen text of short-form influencer video advertisements to predict sales,
demonstrating both that the fused representation carries real predictive signal and that
the fusion of modalities outperforms any one of them [@yang2024engagement]. The pattern
generalizes: image features predict product return rates, visual brand portrayal
predicts brand perceptions, and review text predicts demand, each by inserting a learned
representation into an otherwise standard predictive or causal model.

Precisely because the representation is *learned*, the identification cautions that have
recurred throughout this part apply with full force, and fusion compounds them. Three
deserve restatement in the multimodal setting.

The first is the **generated-regressor problem**. Every fused feature is the output of
an upstream encoder estimated with error. Inserting it into a second-stage regression
as if it were measured without error understates standard errors and, more seriously,
can bias coefficients when the encoder's error is correlated with the outcome. With
several modalities, there are several such channels, and they need not be independent;
a shared foundation-model backbone can induce correlated errors *across* the text and
image features that the analyst treats as separate covariates. The honest analysis
acknowledges that the right-hand-side variables are estimates and propagates that
uncertainty, whether by sample splitting, by bootstrapping the full pipeline including
the encoding step, or by the two-stage corrections the methodology chapters develop.

The second is **endogeneity of the artifact**. Multimodal artifacts are choices.
A firm selects which photographs to show, an influencer designs a video, a consumer
decides whether and what to post. The features extracted from these artifacts are
therefore correlated with the unobserved strategies and types that also drive the
outcome. A measured association between a visual feature and sales may reflect that
better firms make better images *and* sell more, not that the image causes the sale.
Fusion does not solve this and can obscure it by burying the endogenous artifact inside
a high-dimensional embedding where the analyst loses sight of it. Credible causal claims
still require the identification apparatus of the rest of the book: experiments,
instruments, or design-based variation in the artifact.

The third is **construct validity of the learned measure**. A CLIP similarity to
"luxurious," or an MLLM's judgment that an ad is "authentic," is a model's operational
proxy for a human construct, and the proxy must be validated against human judgment on a
held-out sample before it is trusted, exactly as the sentiment-analysis benchmarking
logic of the text chapter requires. A fused, foundation-model-derived feature is more
opaque than a hand-built one, which raises rather than lowers the validation burden. The
convenience of zero-shot measurement is real, but it does not exempt the measure from
the requirement that it actually measure what it claims to.

## Industry and Production Practice {#sec-mmf-production}

Operationalizing multimodal fusion at scale is as much an engineering problem as a
modeling one, and the production patterns have converged on a recognizable stack.

At the center sits the **embedding pipeline**. Raw artifacts (images, text, audio,
video frames) are passed through encoders, usually hosted **foundation-model APIs** or
self-hosted open-weight models, to produce embeddings. Because encoding is the expensive
step and embeddings are reused across many downstream tasks, embeddings are computed once
and persisted rather than recomputed per model. This is the role of the **feature store**:
a system that holds precomputed per-entity features, including embeddings, keyed by entity
(customer, product, creative) and timestamp, and serves them consistently to both training
and inference. The feature store solves two problems that otherwise sink multimodal
projects. It enforces **train-serve consistency**, guaranteeing that the embedding a model
sees at inference time was produced by the same encoder version as the one it was trained
on, and it enables **point-in-time correctness**, returning the feature value as it stood
at the moment of a historical event rather than its current value, which is essential to
avoid leakage in any model trained on time-stamped marketing data.

**Versioning** is the discipline that holds the stack together. Foundation models are
updated, and an embedding produced by one model version is not interchangeable with one
produced by another; cosine similarities and learned downstream weights are only valid
within a fixed encoder version. Production systems therefore pin encoder versions, store
the version alongside every embedding, and treat a model upgrade as a re-embedding and
re-validation event, not a transparent swap. This is the engineering face of the silicon-
sampling drift concern from @sec-mmf-mllm: a measure built on a foundation model inherits
that model's mutability.

A few further practices recur. Embeddings are often **dimension-reduced** (by PCA or a
learned projection) before entering downstream models, to control the width of the fused
input. **Vector databases** index embeddings for similarity search and retrieval, which
is the substrate of retrieval-augmented pipelines. And **cost and latency** govern
architecture choices in production in a way they never do in a paper: batch precomputation
of embeddings, caching, and the choice between a hosted API and a self-hosted model are
driven by throughput and unit economics as much as by accuracy. None of this changes the
statistics, but all of it determines whether a multimodal model is something a firm can
actually run.

## The Representation Throughline {#sec-mmf-throughline}

It is now possible to state plainly the idea that has run beneath every chapter of this
part. **Every modality reduces to a learned representation, and once it is a vector, the
book's downstream methods take over.** Text became embeddings; images became CNN or
vision-transformer features; audio became prosodic and spectral features or wav2vec-style
embeddings; video became fused frame, audio, and text features; behavior, geography,
networks, and sensors each became a feature vector. The encoders differ, the modalities
differ, but the output type is the same, and that common output type is the seam along
which the unstructured-data part attaches to the rest of the book.

This is why the part can exist as a coherent unit rather than a list of unrelated
techniques. The regression, classification, choice, causal-inference, and Bayesian
machinery developed elsewhere in the book does not need to know whether a covariate
originated as a pixel, a phoneme, a click, or a word. It needs the covariate to be a
number, or a vector of numbers, with a defensible claim to measuring something. Fusion is
the operation that combines several such vectors; foundation models are the engines that
increasingly produce them; and the generated-regressor caution is the price of admission
that every one of them must pay. The representation is the universal interface, and the
discipline of the part is to remember, at every step, that the interface is *learned* and
therefore *uncertain*.

Stated as a principle: unstructured data of any modality becomes a learned, lossy feature
vector, and that vector is a generated regressor whose error may correlate with the
outcome. Every modality chapter specialized this caution; this capstone generalizes it.
The methods that consume the vector are powerful, but they inherit whatever the encoding
got wrong, and no amount of downstream sophistication repairs a representation that does
not measure what the analyst believes it measures.

## Frontier and Expansion {#sec-mmf-frontier}

Several directions are moving quickly enough to reshape the practice within the life of
this edition.

**Any-to-any and natively multimodal models.** The trajectory is toward single models
trained from the outset on text, images, audio, and video jointly, rather than a language
model with a vision encoder bolted on. As the modalities share more of the architecture
and the pretraining, the line between "fusion strategy" and "model" dissolves: fusion
becomes an internal property of a natively multimodal network rather than a choice the
analyst makes downstream.

**Agentic and tool-using measurement.** Foundation models that can call tools, retrieve
documents, and execute multistep procedures turn measurement from a single prompt into a
pipeline the model itself orchestrates. This raises the ceiling on what can be measured
from raw artifacts and simultaneously lowers the transparency of how the measure was
produced, sharpening the validation problem rather than relaxing it.

**Generative multimodal artifacts.** The same models that measure creative can generate
it. Synthetic ad images, video, and copy are now cheap to produce, which collapses the
cost of creative experimentation and at the same time pollutes the observational record
with machine-made artifacts whose provenance the analyst may not know. Modeling demand in
a world where some of the creative was generated by the same model class used to measure
it is an open and consequential problem.

**Causal multimodal inference.** The hardest and most valuable frontier is integrating
learned multimodal representations into credible causal designs: using fused features as
controls without inducing collider bias, isolating the causal effect of a manipulable
visual or textual attribute while holding the rest of a complex artifact fixed, and
propagating encoding uncertainty through a causal estimate. The methods exist in pieces
across the book; assembling them for high-dimensional, multimodal, foundation-model-
derived features is where the next decade of empirical marketing methodology will largely
be spent.

The arc of this part runs from a single modality measured in isolation to many modalities
fused through general-purpose foundation models. The constant across that arc is the
representation: learned, lossy, powerful, and uncertain. Holding all four of those
adjectives in mind at once is the whole discipline of doing marketing science with
unstructured and multimodal data.
