# Policy Evaluation and Cross-Market Spillovers {#sec-policy-spillovers}

A policy is evaluated on the market it names. A rule that restricts credit-card
marketing is judged by what happens to credit cards; an advertising ban is judged
by advertising; a privacy regulation is judged by tracking. This is the natural
way to audit an intervention, and it is very often the wrong way, because the
consumer whose behavior the rule changed did not stop having the need that drove
the behavior. She satisfied it somewhere else. The evaluation that looks only at
the targeted market records the intended effect at full strength and the
unintended effect not at all.

This chapter is about designing evaluations that see both. Its running case is
Section 304 of the Credit Card Accountability, Responsibility, and Disclosure
(CARD) Act of 2009, which effectively eliminated the marketing of credit cards on
and around U.S. college campuses. The provision worked: card holding among
students fell sharply. It also did something no one legislated. Students who
needed liquidity and no longer reached for a card reached instead for a student
loan, and student-loan balances rose by roughly eight percent on average and
fifteen percent among students from the least affluent neighborhoods
[@brown2026spillover]. Neither number is visible in a study of credit cards.

The pattern generalizes far past consumer finance, and it is the reason this
material belongs in a marketing-research book rather than only in a public-finance
one. Almost every consequential regulation of the last two decades restricts a
*marketing instrument*—what may be advertised, to whom a firm may target, which
data may be joined, which defaults may be pre-set, how a price may be displayed.
Restricting an instrument does not remove the demand it was serving; it reroutes
that demand into whichever substitute instrument, channel, or product the rule
left standing. A marketing scientist is unusually well placed to see the reroute,
because the substitution structure of consumer choice is our subject matter.

By the end of this chapter you should be able to state a spillover as a formal
estimand rather than a worry, distinguish the four mechanisms that generate one,
design an identification strategy in which the untargeted market is observed with
a credible control group, and recognize when a published policy evaluation has
measured only half of its own effect.

## Why Single-Market Evaluation Fails {#sec-psp-single-market}

Consider a consumer with a need $n$ that can be met by instruments
$j \in \{A, B\}$—two credit products, two channels, two information sources. A
policy $P$ raises the effective cost of instrument $A$: it bans the advertising
that made $A$ salient, deletes the tracking that made $A$ well-targeted, or
removes the campus table at which $A$ was sold. Write the consumer's use of each
instrument as $q_A$ and $q_B$. The evaluation that a regulator commissions
estimates

$$
\tau_A \;=\; \mathbb{E}\!\left[q_A(1) - q_A(0)\right],
$$ {#eq-psp-targeted}

the average effect of the policy on the targeted instrument, where $q_A(1)$ and
$q_A(0)$ are potential outcomes with and without the rule. What the regulator
*wants* is the effect on the underlying welfare-relevant outcome, which depends on
both instruments. The missing quantity is the **spillover estimand**

$$
\tau_B \;=\; \mathbb{E}\!\left[q_B(1) - q_B(0)\right],
$$ {#eq-psp-spillover}

and the ratio $\rho = \tau_B / |\tau_A|$—the **substitution rate**—is the single
number that determines whether the policy moved consumers out of a market or
merely moved them between markets. When $\rho \approx 0$ the policy suppressed the
behavior. When $\rho \approx 1$ it relabeled the behavior. Every intermediate
value is a mix, and nothing in the estimate of $\tau_A$ identifies which case
obtains.

The literature on choice architecture has been explicit that this is a systemic
gap rather than an occasional oversight. Surveys of the nudge evidence base note
that interventions are overwhelmingly evaluated on the dimension they target, so
that offsetting movements on non-targeted dimensions—which can shrink, erase, or
invert the measured benefit—remain largely unobserved [@beshears2020nudging]. The
behavioral-spillover literature makes the same point from the psychological side:
an intervention that changes one behavior systematically changes related ones,
sometimes reinforcing the intended direction and sometimes licensing its opposite
[@dolan2015ripples; @truelove2014spillover].

@fig-psp-leak shows the accounting. A policy evaluation is a claim about a
consumer's total behavior; measuring one market is a claim about one branch of it.

```{mermaid}
%%| label: fig-psp-leak
%%| fig-cap: "Why a single-market evaluation is not an evaluation. The rule suppresses the targeted instrument, but the underlying need persists and reroutes into whatever substitute the rule left standing. The commissioned study measures only the left branch."
flowchart TB
  N["Consumer need<br/>(liquidity, information,<br/>convenience) persists"] --> P["Policy raises the effective<br/>cost of instrument A"]
  P --> QA["Targeted market:<br/>use of A falls by tau_A"]
  P --> QB["Untargeted market:<br/>use of B rises by tau_B"]
  QA --> EV["Commissioned evaluation:<br/>reports tau_A only"]
  QB -.->|"unmeasured"| EV
  EV --> CL["Conclusion: policy worked"]
  QB --> TR["Truth depends on rho = tau_B / |tau_A|<br/>and on whether B is<br/>better or worse for the consumer"]
```

Two clarifications keep the framework honest. First, a spillover is not
automatically a failure. In the CARD Act case the substitution ran from a
high-rate revolving product to a low-rate, subsidized one; the entire welfare
question is whether the destination is better than the origin, and
@sec-welfare-calibration is devoted to answering it. Second, the sign of the
welfare effect and the sign of the behavioral effect are different objects. A
policy can raise borrowing (a behavioral increase that reads as a failure in the
press release) while raising welfare (because the borrowing is cheaper), which is
precisely what the anchor paper concludes.

## A Taxonomy of Spillovers {#sec-psp-taxonomy}

"Spillover" names at least four distinct phenomena, with different identification
requirements and different remedies. Confusing them is the most common conceptual
error in this literature.

**Cross-market substitution.** The same consumer meets the same need through a
different product. This is the CARD Act case: the student who would have revolved
on a card takes a larger loan. It is also the soda-tax case, where taxed beverages
give way to untaxed ones inside and outside the taxing jurisdiction
[@seiler2021], and the junk-food case, where banning advertising in one category
shifts demand across categories [@dubois2018banning]. Identification requires
observing the *same* consumers in the *untargeted* market.

**Behavioral spillover.** The intervention changes a psychological state—attention,
moral licensing, self-image, perceived permission—that then alters behavior in a
domain the policy never touched. A payment nudge that reduces revolving on the
targeted card can reduce or increase spending on a different card held by the same
person [@medina2021side]. Here the two behaviors need not be substitutes in any
budget sense; the link runs through the consumer's psychology
[@dolan2015ripples].

**Equilibrium and supply-side response.** Firms re-optimize. When tracking is
restricted, advertisers reallocate budget, ad prices move, and the composition of
surviving vendors changes; the measured effect on any one firm confounds the
consumer response with the industry's [@johnson2023privacy; @aridor2023privacy].
Recommender systems generate the same structure endogenously: an algorithm
designed to improve individual matches can concentrate aggregate sales, a
system-level consequence invisible in individual-level evaluation
[@fleder2009blockbuster].

The supply-side response can also relocate an effect across *margins* rather than
across markets, and this is the variant most likely to be missed, because the
harmed group is the group the rule was written to help. Brazil's disability
employment quota is the sharpest available case. A 2012 reform put real
enforcement behind a quota that had long been ignored, and it worked on the margin
the statute names: hiring of people with disabilities rose, most steeply at firms
that had been furthest out of compliance. But firms meeting a headcount target
re-optimized everything the target did not fix. Among workers with disabilities
already employed, job stability improved while wage gaps widened and promotion
rates slowed relative to non-disabled coworkers, and a 2015 nondiscrimination
statute did not undo the damage [@nardi2026diversity]. A representation metric and
a treatment-of-incumbents metric moved in opposite directions inside the same
firms. An evaluation that reports only the first records a success; the policy's
own beneficiaries would report something else. The lesson generalizes to any
mandate stated as a count—diverse-supplier targets, local-content rules,
minimum-shelf-space regulations, platform quotas for a favored seller class—and it
is the market-design analogue of the surrogate-metric problem in
@sec-metrics: the quantity that is easy to legislate is rarely the quantity that
carries the welfare.

**Interpersonal spillover (interference).** The treatment of one consumer changes
the outcomes of another—through networks, through congestion, through marketplace
prices. This is a violation of the stable-unit-treatment-value assumption rather
than a second market, and it is treated in @sec-causal-inference.

@tbl-psp-taxonomy summarizes what each type demands of a research design.

| Type | Mechanism | What must be observed | Typical design |
|:--|:--|:--|:--|
| Cross-market substitution | Same need, different product | Same units, both markets, pre/post | DiD or triple-diff on the untargeted outcome |
| Behavioral spillover | Changed attention, licensing, self-concept | An untreated domain plus a mechanism measure | Experiment with pre-registered secondary domain |
| Equilibrium / supply side | Firm re-optimization of prices, allocation, and unregulated margins | Market-level aggregates, entry and exit; incumbent outcomes on the margins the rule leaves free | Market-level DiD, structural supply model |
| Interpersonal (interference) | Treatment of $i$ affects $j$ | Network or geography linking units | Cluster or saturation randomization |

: Four spillover types and their design requirements. Only the first is a straightforward extension of a standard program evaluation; the others change the estimand or the randomization unit. {#tbl-psp-taxonomy}

## Identification: Finding a Control Group in the Untargeted Market {#sec-psp-identification}

The hard part of a spillover study is rarely the outcome. Student-loan balances are
well measured. The hard part is that the policy applied to everyone in the affected
population at the same moment, so there is no obvious untreated comparison.

The design in @brown2026spillover solves this with an exposure argument that is
worth studying as a template. Section 304 removed lenders *from campus*. Therefore
the intensity of a student's exposure to card marketing depends on how much time
she had already spent on campus when she made her borrowing decision. Incoming
freshmen make their student-loan choices before arriving; they were never exposed
to on-campus card marketing in either regime. Continuing sophomores and juniors
were exposed before the rule and not after. Freshmen are the control group,
continuing students the treated group, and the comparison is a
difference-in-differences in which the "untreated" units are inside the same
institution, the same semesters, and the same financial-aid system.

Three refinements make the design credible, and each is a transferable lesson.

*Drop the ambiguously treated.* Students enrolled in the two academic years
immediately after passage were partially exposed—they had seen campus marketing as
freshmen even if it was gone by their sophomore year—and financial-aid packages
are locked months before a semester begins, so even an instantly reacting aid
office could not have adjusted. Those cohorts are excluded rather than assigned to
either group. A design that forces every observation into treated or control buys
sample size with contamination.

*Exclude the population treated by a different provision of the same law.* Seniors
are mostly over 21, and the CARD Act's ability-to-pay and co-signer rules applied
differently to them. Including seniors would blend the on-campus marketing nudge
with a separate credit-supply restriction. When a statute has several sections,
the analysis sample must be the population for which only the section of interest
binds.

*Bound the calendar.* The post-period ends before the 150-percent rule of the
Moving Ahead for Progress in the 21st Century Act changed subsidized-loan
eligibility, because that rule altered both the sequencing of loan take-up inside
the aid office and the treatment of first-time borrowers—who are disproportionately
the control group. A later confounding policy that differentially hits the control
group is more dangerous than one that hits both groups.

The estimating equation is a non-parametric event study rather than a single
post-period dummy:

$$
y^{k}_{i,s} \;=\; \alpha^{k}_{s,q} \;+\; \alpha^{k}_{c} \;+\; \sum_{s \neq s_0} \beta^{k}_{s}\,\mathbb{1}(\text{continuing})_{i,s} \;+\; \gamma^{k} X_{i,s} \;+\; \epsilon^{k}_{i,s},
$$ {#eq-psp-eventstudy}

where $i$ indexes students, $s$ semesters, $k \in \{\text{take-up},\ \text{balance}\}$,
$\alpha^{k}_{s,q}$ is a semester-by-income-quartile fixed effect, $\alpha^{k}_{c}$
a class (freshman/sophomore/junior) fixed effect, and $s_0$ the omitted
pre-treatment semester. Letting the treatment coefficients vary semester by
semester does double duty: the pre-period coefficients test the parallel-trends
assumption, and the post-period coefficients show whether the effect is a
level shift, a trend break, or a transient.

The semester-by-quartile fixed effect deserves comment. Because the CARD Act
coincided with the financial crisis, a design with only semester effects would
attribute to the policy any recession shock that hit low-income students harder.
Interacting the time effect with income quartile absorbs exactly that, at the cost
of no longer identifying the average level effect of income. This is the general
move: **absorb the confound at the level at which it varies**, and accept the loss
of the parameters it collinearly consumes.

@fig-psp-design lays out the design.

```{mermaid}
%%| label: fig-psp-design
%%| fig-cap: "The exposure-based difference-in-differences used to identify a spillover from an economy-wide rule. Treatment status comes from how much on-campus marketing a student could have been exposed to before making the borrowing decision, not from any policy variation across places."
flowchart TB
  R["Section 304 removes card<br/>marketing from campus<br/>(national, one date)"] --> EX{"Was the student exposed to<br/>on-campus marketing before<br/>the borrowing decision?"}
  EX -->|"No: incoming freshmen<br/>decide before arriving"| C["Control group"]
  EX -->|"Yes pre-rule, no post-rule:<br/>sophomores and juniors"| T["Treated group"]
  EX -->|"Partially: cohorts in the<br/>two transition years"| D["Dropped as<br/>ambiguously treated"]
  EX -->|"Treated by other CARD Act<br/>sections: seniors over 21"| D2["Dropped to isolate<br/>the marketing provision"]
  C --> DD["Event-study DiD on the<br/>UNTARGETED outcome:<br/>student-loan balances"]
  T --> DD
  DD --> H["Triple difference by ZIP income:<br/>does the substitution concentrate<br/>where liquidity need is greatest?"]
```

### The heterogeneity test as a specification test {#sec-psp-triple}

A spillover estimate is more convincing when the model predicts *where* it should
be largest and the data agree. If the mechanism is liquidity substitution, then
students with the least outside liquidity should substitute most, and students with
enough family resources to have no shortfall should not substitute at all. The
triple-difference specification tests this:

$$
y^{b}_{i,s} \;=\; \alpha^{b}_{s,q} + \alpha^{b}_{c} + \beta^{b}_{1}\mathbb{1}(\text{cont.},\text{post})_{i,s} + \sum_{q=2}^{4}\beta^{b}_{q}\,\mathbb{1}(\text{cont.},\text{post})_{i,s}\times\mathbb{1}(Qtr = q)_{i} + \gamma^{b} X_{i,s} + \epsilon^{b}_{i,s}.
$$ {#eq-psp-triplediff}

In the CARD Act data the gradient is monotone and ends at zero: about \$227 more
borrowed per semester in the bottom ZIP-income quartile—roughly 15 percent of that
group's pre-policy mean, and about \$454 on an annual basis—then \$130 in the
second quartile, \$62 in the third, and a statistically indistinguishable-from-zero
\$-4 in the top [@brown2026spillover]. A null in the
group your theory says should be null is worth more than another significant
coefficient, because a confound like the recession has no reason to respect the
theory's ordering.

This is also where a spillover study earns its distributional claim. The average
effect (8 percent) understates what happened to the students the policy was
written to protect (15 percent), and the top quartile's zero rules out a broad
"everyone borrowed more after 2009" story.

## Simulating the Design {#sec-psp-simulation}

The following simulation builds a student-semester panel with the structure above:
a control class that is never exposed, a treated class that loses card access after
the rule, heterogeneous substitution by income quartile, and a recession shock that
hits low-income students in both groups. It then shows what a naive analysis of the
targeted market alone would report, what the event study recovers, and what happens
if the semester-by-quartile fixed effect is omitted.

```{r psp-sim-panel}
#| echo: true
#| message: false
#| warning: false

set.seed(66)

simulate_spillover_panel <- function(n_students = 12000,
                                     semesters  = 1:8,
                                     policy_at  = 5,
                                     card_drop  = 0.30,
                                     sub_by_q   = c(230, 130, 60, 0),
                                     recession  = c(-140, -90, -40, 0)) {
  # One row per student-semester. Class is fixed within a student.
  q      <- sample(1:4, n_students, replace = TRUE)
  cont   <- rbinom(n_students, 1, 0.65)          # 1 = sophomore/junior (treated)
  base   <- c(1442, 1397, 1318, 878)[q]          # pre-policy mean balance by quartile

  grid <- expand.grid(id = seq_len(n_students), sem = semesters)
  grid$q     <- q[grid$id]
  grid$cont  <- cont[grid$id]
  grid$post  <- as.integer(grid$sem >= policy_at)
  grid$base  <- base[grid$id]

  # Recession hits low-income students in BOTH groups: a pure confound.
  shock <- recession[grid$q] * grid$post
  # The spillover: only continuing students, only post-policy.
  spill <- sub_by_q[grid$q] * grid$cont * grid$post

  grid$balance <- grid$base + shock + spill +
    rnorm(nrow(grid), 0, 900)

  # Targeted market: card use falls for continuing students post-policy.
  p_card <- 0.80 - card_drop * grid$cont * grid$post
  grid$has_card <- rbinom(nrow(grid), 1, pmin(pmax(p_card, 0.01), 0.99))

  grid$sem_q <- interaction(grid$sem, grid$q, drop = TRUE)
  grid$semf  <- factor(grid$sem)
  grid
}

panel <- simulate_spillover_panel()
str(panel[, c("id", "sem", "q", "cont", "post", "balance", "has_card")])
```

The naive single-market evaluation looks decisive and says nothing about the loan
market:

```{r psp-naive}
#| echo: true
#| message: false
#| warning: false

naive_card <- lm(has_card ~ cont * post, data = panel)
round(coef(summary(naive_card))["cont:post", ], 4)
```

Now estimate the spillover on the untargeted outcome, first without and then with
the semester-by-quartile absorption. The difference is the recession confound.

```{r psp-did-compare}
#| echo: true
#| message: false
#| warning: false

# (a) Semester effects only: the recession contaminates the estimate because it
#     is correlated with income quartile, which is correlated with treatment mix.
fit_naive <- lm(balance ~ cont * post + semf + factor(q), data = panel)

# (b) Semester-by-quartile effects: absorbs any shock common to a quartile in a
#     semester, leaving only the treated-vs-control contrast within cell.
fit_absorb <- lm(balance ~ cont * post + sem_q, data = panel)

comparison <- rbind(
  `Semester FE only`        = coef(summary(fit_naive))["cont:post", 1:2],
  `Semester x quartile FE`  = coef(summary(fit_absorb))["cont:post", 1:2]
)
round(comparison, 1)
```

Both specifications recover a positive spillover here because the recession shock
hits treated and control students *within* a quartile equally; what the absorbed
specification buys is protection against the case where group composition shifts.
The event study makes the timing visible and provides the pre-trend test:

```{r psp-eventstudy}
#| echo: true
#| message: false
#| warning: false
#| label: fig-psp-event
#| fig-cap: "Event-study estimates of the spillover onto the untargeted market. Pre-policy coefficients are indistinguishable from zero (the parallel-trends check); post-policy coefficients jump and persist. The omitted period is the last pre-policy semester."

library(ggplot2)

panel$rel <- panel$sem - 4                       # semester 4 = last pre-policy
panel$rel <- relevel(factor(panel$rel), ref = "0")

es <- lm(balance ~ cont:rel + rel + cont + sem_q, data = panel)
cf <- coef(summary(es))
rows <- grep("^cont:rel", rownames(cf))

es_df <- data.frame(
  rel  = as.numeric(sub("^cont:rel", "", rownames(cf)[rows])),
  est  = cf[rows, 1],
  se   = cf[rows, 2]
)
es_df <- rbind(es_df, data.frame(rel = 0, est = 0, se = 0))
es_df <- es_df[order(es_df$rel), ]

ggplot(es_df, aes(rel, est)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_vline(xintercept = 0.5, linetype = "dashed", linewidth = 0.3) +
  geom_pointrange(aes(ymin = est - 1.96 * se, ymax = est + 1.96 * se)) +
  labs(x = "Semesters relative to the rule",
       y = "Effect on balance in the untargeted market ($)",
       title = "Spillover appears only after the policy") +
  theme_minimal(base_size = 11)
```

Finally, the heterogeneity test. The simulated substitution was set to decline
across quartiles and vanish at the top; a design that recovers that pattern has
passed a test a generic time shock would fail.

```{r psp-triple}
#| echo: true
#| message: false
#| warning: false

trip <- lm(balance ~ cont:post:factor(q) + cont * post + sem_q, data = panel)
cf3  <- coef(summary(trip))

# The bottom quartile is absorbed into cont:post, so the level effect for
# quartile k is the base coefficient plus the quartile-k interaction.
base_q1 <- cf3["cont:post", 1]
keep    <- grep("^cont:post:factor\\(q\\)", rownames(cf3))

est <- c(base_q1, base_q1 + cf3[keep, 1])
se  <- c(cf3["cont:post", 2], cf3[keep, 2])   # interaction SEs, ignoring covariance

data.frame(
  quartile   = c(1, as.numeric(sub(".*\\)", "", rownames(cf3)[keep]))),
  level_est  = round(est, 0),
  approx_se  = round(se, 0),
  true_value = c(230, 130, 60, 0),
  row.names  = NULL
)
```

The estimated gradient declines monotonically and reaches zero at the top
quartile, recovering the pattern that was built in. Individual cells are noisy
relative to their true values—which is the realistic case, since a triple
difference splits the sample four ways—so the claim a paper should make is about
the *ordering and the terminal zero*, not about any single quartile's point
estimate.

::: {.callout-note}
## What the simulation is and is not

These numbers are generated from a data-generating process the analyst wrote, so
recovering them is a check that the estimator is coded correctly, not evidence
about the CARD Act. The value of the exercise is diagnostic: it shows what a
correct event study looks like when the design holds, so that a real estimate
that does *not* look like this—drifting pre-trends, an effect that fades within
two periods, heterogeneity that contradicts the theory—can be recognized as a
problem rather than reported as a result.
:::

## The Alternative-Explanation Gauntlet {#sec-psp-alternatives}

Because a spillover study attributes movement in market $B$ to a rule about market
$A$, the reviewer's default hypothesis is that something else moved market $B$.
The CARD Act paper is a useful model of how to run this gauntlet systematically:
it enumerates the channels through which the contemporaneous financial crisis
could have raised student borrowing and tests each one, rather than asserting that
the design handles them [@brown2026spillover].

| Alternative explanation | Test used | Logic of the test |
|:--|:--|:--|
| Labor-market expectations worsened | Re-estimate on sophomores only, then juniors only | If expectations drive it, students nearer graduation (better-informed beliefs) should show larger effects; they do not |
| Home equity collapsed, cutting family support | Design comparison | The channel hits freshmen and continuing students equally, so it differences out |
| Composition of enrolled students changed | Covariates plus student fixed effects | Identify off the same student observed as freshman and as returning student |
| Credit supply tightened (ability-to-pay rules) | Adjusted-gross-income outcomes | If supply-side denial drove it, treated-control gaps in income should move; they are stable |
| Loan limits changed | Institutional check of limits by class and year | Subsidized limits unchanged; the one unsubsidized change applied to all classes, pre-period |
| Transition years contaminated | Re-estimate including dropped years | Interim coefficients are imprecise and trendless, as an ambiguous-treatment story predicts |

: The alternative-explanation gauntlet for a spillover claim. Each row names a rival story, the test that speaks to it, and why the test is informative rather than decorative. {#tbl-psp-gauntlet}

The transferable structure is that every row pairs a *specific* rival mechanism
with a test whose result would differ under the rival and under the maintained
hypothesis. A robustness table full of specifications that all give the same
answer for the same reason is not a gauntlet; it is a demonstration that the
estimate is numerically stable.

## Marketing Settings Where the Template Applies {#sec-psp-marketing}

The design generalizes to any setting where a rule, a platform change, or a firm
policy raises the cost of one marketing instrument while leaving substitutes
available. Several are already documented, and they show the range of what counts
as the "untargeted market."

**Targeting and privacy regulation.** Restrictions on tracking reduce the
effectiveness of behaviorally targeted display advertising [@goldfarb2011privacy]
and reshape which vendors survive in the data supply chain
[@johnson2023privacy; @aridor2023privacy]. The untargeted markets are contextual
advertising, first-party data collection, retail media, and offline channels; the
substitution question is whether restricting one form of targeting reduces total
persuasion or merely reallocates it to less measurable forms. Consumer-side privacy
controls raise the same issue in the other direction: giving users control over
personal-data use can *raise* the effectiveness of the ads that remain
[@tucker2014social].

**Advertising bans.** Banning advertising in a product category changes demand
inside the category and, through cross-category substitution, outside it
[@dubois2018banning]. Category-level evaluation of an advertising restriction is
the canonical single-market error.

**Excise taxes and other price interventions.** Beverage taxes generate
substitution across container sizes, across categories, and across the
jurisdictional border; the nutritional consequence of the policy is a function of
where the demand goes, not of how much taxed volume fell [@seiler2021].

**Contact regulation.** Do-not-call and anti-spam regimes restrict one outbound
channel; firms reallocate to channels the rule does not reach, and consumer
attention follows [@goh2011newspaper].

**Algorithmic and platform design.** A recommender optimized for individual match
quality can reduce aggregate variety, an effect that exists only at the level the
individual evaluation does not measure [@fleder2009blockbuster]. Platform policy
changes are the richest untapped setting for this template because the platform
observes both the targeted and the untargeted market natively.

**Firm-side marketing interventions.** The template does not require a government.
A retailer that suppresses a promotion, a bank that nudges cardholders toward
faster repayment [@medina2021side], or a subscription service that changes a
default is running the same experiment: the intended margin moves, and the
question is what the customer does with the need that the change did not
eliminate.

The cleanest firm-side demonstration is a platform feature rather than a rule.
@barach2026filtering randomized access to candidate filters in a large online
labor market—the search tool whose entire purpose is to make matching cheaper.
It did what it was built to do at the stage it was built for: employers with
filters interviewed 3.4 percent fewer applicants. But filters push every employer
toward the same easily observable signals, so the screening saving reappeared as a
bargaining problem one stage later, where offers were 8.9 percent more likely to be
rejected and contracted wages 2.5 percent higher among employers who actually used
them. A design metric owned by one team (interviews per hire) improved while a cost
owned by another (offer acceptance, wage bill) deteriorated. This is the same
structure as a policy spillover with the jurisdictional boundary replaced by an
organizational one, and it is why the template belongs in any evaluation of a
platform feature, not only in program evaluation.

::: {.callout-tip}
## Finding a spillover study in your own data

Three conditions make a setting viable. (1) A discrete change in the cost of one
instrument, with a date. (2) At least one substitute instrument observable for the
*same* units—this is the binding constraint, and it is why linked administrative
data, single-firm multi-product panels, and platform logs are so valuable. (3) A
source of variation in exposure that is not the policy date itself: who was on
campus, who was in the treated jurisdiction, who had already adopted, whose
contract renewed before the cutoff. Condition (3) is where most candidate projects
die, and it is worth solving before collecting anything.
:::

## From Effect to Evaluation {#sec-psp-toward-welfare}

Establishing that $\tau_B > 0$ completes the positive analysis and begins the
normative one. Knowing that students borrowed 8 percent more does not say whether
they were made better or worse off, because the answer depends on what the
alternative would have cost them, how likely they were to need the flexibility a
card provides, and whether their pre-policy behavior was a considered choice or a
mistake. Those questions cannot be answered by any regression on the two markets;
they require a model of what an optimizing consumer *should* do, which is the
subject of @sec-normative-benchmarks, and a calibration that turns the model into
numbers, which is the subject of @sec-welfare-calibration.

The ordering matters for how the research is written as well as how it is done.
The reduced-form spillover is the paper's empirical anchor and the reason a reader
believes anything follows; the model is what converts a behavioral magnitude into
a welfare claim. Papers that lead with the model and treat the estimate as an
illustration invert the credibility ordering and are read, correctly, as less
convincing.

## Key Takeaways {#sec-psp-takeaways}

- A policy evaluation restricted to the targeted market estimates @eq-psp-targeted
  and is silent on @eq-psp-spillover; the substitution rate
  $\rho = \tau_B / |\tau_A|$ decides whether a rule suppressed a behavior or
  relocated it.
- Four mechanisms generate spillovers—cross-market substitution, behavioral
  spillover, equilibrium response, and interpersonal interference
  (@tbl-psp-taxonomy)—and they impose different data and randomization
  requirements.
- Identification for an economy-wide rule usually comes from *differential
  exposure* rather than differential policy: find units whose decisions were made
  outside the reach of the intervention, as incoming freshmen were for on-campus
  card marketing.
- Design hygiene does most of the work: drop ambiguously treated cohorts, exclude
  populations bound by other provisions of the same statute, end the panel before
  the next confounding rule, and absorb confounds at the level at which they vary.
- Theory-predicted heterogeneity (@eq-psp-triplediff) is a specification test, not
  a supplementary finding: a monotone gradient that reaches zero exactly where the
  model says it should is evidence a generic time shock cannot easily mimic.
- Run an explicit alternative-explanation gauntlet (@tbl-psp-gauntlet) in which
  each rival mechanism is paired with a test whose outcome would differ under the
  rival.
- The untargeted outcome need not sit in another market. It can be another
  *margin* of the same transaction—wages and promotion rather than hiring, offer
  acceptance rather than screening cost—and when a mandate is written as a count,
  that margin is where the response goes
  [@nardi2026diversity; @barach2026filtering].
- Marketing is unusually rich in settings for this template because most modern
  regulation restricts a marketing instrument, and firms and platforms observe the
  substitute instruments natively.

## Further Reading {#sec-psp-further}

The anchor case is @brown2026spillover, which should be read alongside
@debbaut2016card on the CARD Act's direct effects and @agarwal2015regulating on
the Act's fee provisions, so that the marketing-restriction channel is separated
from the price-regulation channel. For the general argument that nudge evaluations
under-measure non-targeted outcomes, see @beshears2020nudging; for the
psychological version of the same claim, @dolan2015ripples and
@truelove2014spillover. On marketing-instrument regulation specifically,
@goldfarb2011privacy, @johnson2023privacy, and @aridor2023privacy cover privacy
and targeting; @dubois2018banning covers advertising bans; @seiler2021 covers
excise taxes and cross-market avoidance; @goh2011newspaper covers contact
regulation. @fleder2009blockbuster is the cleanest statement of a system-level
consequence invisible in individual-level evaluation. Two Organization Science
field studies extend the template past consumer markets and are worth reading
together: @nardi2026diversity on Brazil's disability quota, where a mandate
stated as a headcount raised hiring while widening wage and promotion gaps for
incumbent workers with disabilities, and @barach2026filtering, a randomized field
experiment showing that a matching feature which lowers screening cost raises
offer-stage cost instead. The identification
background for the designs used here is in @sec-causal-inference, and the
policy-and-well-being research tradition that motivates the questions is mapped in
@sec-public-policy-tcr-seminar.
