# Normative Benchmarks: Models of Optimal Behavior {#sec-normative-benchmarks}

You cannot call a choice a mistake without saying what the right choice was. This
is the sentence that separates a policy paper that survives review from one that
does not. An evaluation showing that a rule changed behavior establishes that
consumers responded; it establishes nothing about whether the response was good.
To get from "borrowing rose" to "students were made better off," someone has to
write down what an unconstrained, well-informed, correctly-optimizing consumer
would have done in the same situation—and then measure the distance between that
benchmark and the behavior actually observed.

This chapter is about constructing that benchmark. The object is a **normative
model**: a small analytical model of the consumer's problem whose solution
characterizes optimal behavior, whose comparative statics generate testable
predictions, and whose parameters are quantities that can be measured or bounded
rather than assumed. It is deliberately not a full structural model estimated on
the data. The distinction is central and is developed in @sec-nb-role: a normative
model earns its place by being *smaller* than a structural model, because it is
asked to do less.

The running construction is the student-borrowing model of @brown2026spillover,
which is unusually well suited to teaching because it is simple enough to solve on
paper and rich enough to generate a threshold rule, a welfare formula, and four
distinct forms of consumer error. We build it from primitives, derive its central
proposition, catalogue the departures from it, and then abstract the recipe so it
can be re-used on marketing problems that have nothing to do with credit.

## What a Normative Model Is For {#sec-nb-role}

Marketing research uses models for at least three incompatible purposes, and
conflating them produces papers that satisfy nobody.

A **descriptive** model reproduces observed behavior. Its success criterion is
fit, and its parameters are whatever makes the fit good.

A **structural** model recovers deep parameters—preferences, costs, beliefs—that
are invariant to the policy under study, so that counterfactual equilibria can be
simulated (@sec-structural-models). Its success criterion is that the recovered
parameters are policy-invariant and identified, which is demanding, and it usually
requires functional-form commitments the theory does not justify
[@nevo2010dogma; @rust2014limits].

A **normative** model characterizes what an agent *should* do given a stated
objective and constraint set. Its success criterion is neither fit nor
identification: it is that the optimality conditions are correct, that the
predictions distinguish optimizing from non-optimizing agents, and that the
comparative statics are sharp enough to be refuted. A normative model is allowed
to be wrong about average behavior. Indeed, if consumers systematically deviate
from it, that gap is the paper's finding.

@fig-nb-three-roles places the three side by side and shows what each licenses.

```{mermaid}
%%| label: fig-nb-three-roles
%%| fig-cap: "Three roles a model can play in a policy paper. The normative model is the one that licenses the word mistake; it is smaller than a structural model because it is not asked to reproduce the data."
flowchart TB
  Q["Policy question:<br/>did the rule help consumers?"] --> D["Descriptive model:<br/>reproduce observed choices"]
  Q --> S["Structural model:<br/>recover policy-invariant<br/>preferences and costs"]
  Q --> N["Normative model:<br/>characterize the optimal<br/>choice given objective<br/>and constraints"]
  D --> D2["Licenses: prediction<br/>within the observed regime"]
  S --> S2["Licenses: counterfactual<br/>equilibria and full welfare<br/>Costs: functional forms,<br/>identification burden"]
  N --> N2["Licenses: the words<br/>optimal, mistake, and<br/>welfare bound<br/>Costs: none if it stays small"]
  N2 --> W["Departures from N become<br/>the measurement target"]
```

The practical consequence is a sequencing rule. Write the normative model
*before* looking at the mechanism data, because its role is to tell you which
survey questions and which behavioral signatures are diagnostic. A model written
after the fact to rationalize a pattern is a description with equations.

## Building a Normative Model From Primitives {#sec-nb-build}

We now construct the benchmark. The economic content is a tension every marketing
scientist will recognize: a cheap instrument that must be committed to in advance
versus an expensive instrument that can be deployed after uncertainty resolves.
Call it the **price-versus-flexibility trade-off**. In the borrowing application,
the cheap-but-inflexible instrument is a student loan, taken at the start of the
year at a low posted rate on the full amount; the expensive-but-flexible one is a
credit card, a line of credit drawn only if and when the need materializes, at a
much higher rate.

### Primitives {#sec-nb-primitives}

At the start of period $t$ a consumer must finance consumption. She holds outside
funds $w$ and enters with debts $b^{c}_{0}$ on the flexible instrument and
$b^{s}_{0}$ on the committed one, writing $b_0 = b^{c}_{0} + b^{s}_{0}$. Period
utility is defined relative to a subsistence requirement $\underline{x}_t$:

$$
u_t(x) \;=\;
\begin{cases}
v_t\!\left(x - \underline{x}_t\right), & x \ge \underline{x}_t,\\[2pt]
-\infty, & x < \underline{x}_t,
\end{cases}
$$ {#eq-nb-utility}

with $v_t$ twice differentiable, increasing, and concave. The $-\infty$ branch is
doing real work: it says the requirement must be met, which is what makes the
financing decision non-trivial rather than a smooth trade-off against
consumption. Subsistence is stochastic and binary,

$$
\underline{x}_t \;=\;
\begin{cases}
h & \text{with probability } p,\\
\ell & \text{with probability } 1-p,
\end{cases}
\qquad h > \ell ,
$$ {#eq-nb-shock}

so $p$ is the probability of the high-need state—an unexpected expense, an
emergency, a demand shock. The consumer chooses borrowing $b^{s}_{t}$ and
$b^{c}_{t}$ to solve

$$
\max_{b^{s}_{t},\, b^{c}_{t}} \; u_t(x_t) \;+\; \beta V_t\!\left((1+r^{s})b^{s}_{t} + (1+r^{c})b^{c}_{t}\right),
$$ {#eq-nb-problem}

subject to an accounting constraint tying consumption to endowment and net new
borrowing, non-negativity, a card limit $\overline{b}^{c}$, and an availability
limit on the committed instrument. Here $0 \le r^{s} < r^{c} \le 1$ and
$\beta \in (0,1)$; the continuation value $V_t$ is decreasing in debt carried
forward, twice differentiable, and weakly concave.

Two structural features generate everything that follows. First, the committed
instrument is chosen *before* the state is realized and accrues interest on the
full amount drawn, whether or not it is spent. Second, the flexible instrument is
drawn *after* the state is known, at a higher price. This is insurance pricing:
paying $r^{s}$ with certainty to avoid paying $r^{c}$ in a state that occurs with
probability $p$.

### Reduction to a one-dimensional insurance problem {#sec-nb-reduction}

Because the committed instrument is strictly cheaper, any *known* expense—existing
debt, certain future outlays—is optimally financed with it. The only interesting
decision concerns the *uncertain* portion. Define the uncertain shortfall

$$
\Gamma \;\equiv\;
\begin{cases}
0, & w - b_0 > h,\\
h - (w - b_0), & \ell < w - b_0 < h,\\
h - \ell, & w - b_0 < \ell,
\end{cases}
$$ {#eq-nb-shortfall}

the amount the consumer would have to borrow in the high state but not in the low
state. The decision collapses to a single scalar: how much of $\Gamma$ to insure
in advance. Let $\gamma \in [0,\Gamma]$ be the insured amount. Writing
$A_h = (1+r^{s})(b_0 + h - w)$ and $A_\ell = (1+r^{s})(b_0 + \ell - w)$ for the
baseline debt carried into each state, the problem becomes

$$
\max_{0 \le \gamma \le \Gamma} \; J(\gamma) \;\equiv\; v_t(0) \;+\; p\,V_t\!\left(A_h + (r^{c}-r^{s})(\Gamma-\gamma)\right) \;+\; (1-p)\,V_t\!\left(A_\ell + r^{s}\gamma\right).
$$ {#eq-nb-insurance}

Read the two terms. In the high state the consumer needed all of $\Gamma$; the
insured part cost her the cheap rate, and the uninsured remainder $\Gamma - \gamma$
must be drawn on the flexible instrument, costing an extra $(r^{c}-r^{s})$ per
dollar. In the low state she did not need the money, and the $\gamma$ she borrowed
in advance costs her $r^{s}\gamma$ for nothing. Insurance has a premium and a
payout, exactly as it should.

### The optimality condition and the threshold rule {#sec-nb-threshold}

Differentiating @eq-nb-insurance,

$$
J'(\gamma) \;=\; -\,p\,(r^{c}-r^{s})\,V_t'\!\left(A_h + (r^{c}-r^{s})(\Gamma-\gamma)\right) \;+\; (1-p)\,r^{s}\,V_t'\!\left(A_\ell + r^{s}\gamma\right).
$$ {#eq-nb-foc}

Since $V_t' < 0$, the first term is positive (insuring lowers debt in the state
where debt is high) and the second is negative (insuring raises debt in the state
where it was unnecessary). Full insurance $\gamma^{*} = \Gamma$ is optimal when
$J'(\Gamma) \ge 0$, that is when

$$
p\,(r^{c}-r^{s})\,\bigl|V_t'(A_h)\bigr| \;\ge\; (1-p)\,r^{s}\,\bigl|V_t'(A_\ell + r^{s}\Gamma)\bigr| ,
$$ {#eq-nb-full}

and zero insurance $\gamma^{*}=0$—financing the whole shortfall on the expensive
flexible instrument—is optimal when $J'(0) \le 0$, that is when

$$
p\,(r^{c}-r^{s})\,\bigl|V_t'\!\left(A_h + (r^{c}-r^{s})\Gamma\right)\bigr| \;\le\; (1-p)\,r^{s}\,\bigl|V_t'(A_\ell)\bigr| .
$$ {#eq-nb-none}

Because $V_t$ is concave and decreasing, $|V_t'|$ is increasing in debt, so the
left side of @eq-nb-none is evaluated at a *higher* debt level than the left side
of @eq-nb-full. This yields the central result.

::: {.callout-important}
## The threshold characterization

There exist probabilities $p_1 \le p_2$ with $0 < p_1 \le p_2 \le r^{s}/r^{c}$
such that the optimal policy is: **fully insure** with the committed instrument
when $p \ge p_2$; **partially insure** when $p_1 < p < p_2$; and **finance the
entire shortfall on the flexible instrument** when $p \le p_1$. If the
continuation value is linear, the two thresholds coincide at exactly
$p_1 = p_2 = r^{s}/r^{c}$ [@brown2026spillover].
:::

The linear case is worth verifying because it delivers the formula that later
carries the entire welfare calculation. With $|V_t'|$ constant, @eq-nb-full
becomes $p(r^{c}-r^{s}) \ge (1-p)r^{s}$, hence $p\,r^{c} \ge r^{s}$, hence

$$
p \;\ge\; \frac{r^{s}}{r^{c}} .
$$ {#eq-nb-linear-threshold}

Everything hinges on two observable magnitudes: how likely the shock is, and how
much more the flexible instrument costs. That is a remarkably small informational
requirement for a normative statement, and it is the reason the model can be
calibrated from a survey and two posted interest rates rather than estimated.

Two comparative statics follow and are directly testable. Insuring becomes more
attractive when the shock is more likely and when the price gap is larger. And
under non-increasing absolute risk aversion, consumers who enter with more debt or
fewer resources are *less* willing to gamble on the expensive instrument—so the
poorer and more indebted consumer should, if optimizing, rely more on the cheap
committed instrument, not less.

@fig-nb-policy summarizes the decision.

```{mermaid}
%%| label: fig-nb-policy
%%| fig-cap: "The normative policy function. Known expenses always go on the cheap committed instrument; only the uncertain shortfall is a real decision, and it is resolved by comparing the shock probability to the price ratio of the two instruments."
flowchart TB
  E["Financing need this period"] --> K{"Is the expense known<br/>at the decision date?"}
  K -->|"Yes"| CH["Committed instrument:<br/>strictly dominates<br/>because r_s < r_c"]
  K -->|"No: uncertain shortfall Gamma"| PR{"Compare shock probability p<br/>to the price ratio r_s / r_c"}
  PR -->|"p >= p_2"| FI["Fully insure:<br/>borrow Gamma in advance<br/>on the committed instrument"]
  PR -->|"p_1 < p < p_2"| PI["Partially insure:<br/>interior gamma*"]
  PR -->|"p <= p_1"| NI["Do not insure:<br/>draw the flexible instrument<br/>only if the shock hits"]
  CH --> OUT["Normative benchmark:<br/>the behavior a well-informed<br/>optimizer would display"]
  FI --> OUT
  PI --> OUT
  NI --> OUT
```

## Solving and Exploring the Benchmark Numerically {#sec-nb-numeric}

Analytical thresholds are only useful if you can compute them for the parameters
of your setting. The code below specifies a concrete continuation value, solves
@eq-nb-insurance for the optimal insured amount, and locates $p_1$ and $p_2$. We
use $V(D) = -D^{1+\varphi}/(1+\varphi)$ with $\varphi \ge 0$, which is decreasing
and concave in debt, and which nests the linear benchmark at $\varphi = 0$.

```{r nb-solve}
#| echo: true
#| message: false
#| warning: false

# Continuation value over debt carried forward: decreasing and concave.
# phi = 0 is the linear (risk-neutral-in-debt) benchmark.
V     <- function(D, phi) if (phi == 0) -D else -(D^(1 + phi)) / (1 + phi)
Vprime <- function(D, phi) -(D^phi)          # negative, magnitude increasing in D

# Objective (4): choose gamma in [0, Gamma] to insure the uncertain shortfall.
objective <- function(gamma, p, rs, rc, Gamma, A_h, A_l, phi) {
  p * V(A_h + (rc - rs) * (Gamma - gamma), phi) +
    (1 - p) * V(A_l + rs * gamma, phi)
}

gamma_star <- function(p, rs, rc, Gamma, A_h, A_l, phi) {
  opt <- optimize(objective, interval = c(0, Gamma), maximum = TRUE,
                  p = p, rs = rs, rc = rc, Gamma = Gamma,
                  A_h = A_h, A_l = A_l, phi = phi)
  # The linear case (phi = 0) is bang-bang, so compare the interior candidate
  # against both corners rather than trusting a golden-section search on a
  # monotone objective.
  cand <- c(0, opt$maximum, Gamma)
  vals <- vapply(cand, objective, numeric(1),
                 p = p, rs = rs, rc = rc, Gamma = Gamma,
                 A_h = A_h, A_l = A_l, phi = phi)
  cand[which.max(vals)]
}

# Baseline calibration: posted subsidized/unsubsidized student loan rates versus
# a prime credit card rate, and a $1,000 uncertain shortfall.
rs <- 0.068; rc <- 0.195
Gamma <- 1000; A_h <- 2500; A_l <- 1500

# Linear case: the interior solution disappears and the cutoff is exactly rs/rc.
c(cutoff_linear = rs / rc,
  gamma_at_p_below = gamma_star(0.30, rs, rc, Gamma, A_h, A_l, phi = 0),
  gamma_at_p_above = gamma_star(0.45, rs, rc, Gamma, A_h, A_l, phi = 0))
```

With a linear continuation value the policy is bang-bang and switches exactly at
$r^{s}/r^{c} \approx 0.349$. Curvature opens an interior region between $p_1$ and
$p_2$:

```{r nb-policy-plot}
#| echo: true
#| message: false
#| warning: false
#| label: fig-nb-gamma
#| fig-cap: "Optimal insured share of the uncertain shortfall as a function of the shock probability, for three degrees of curvature in the continuation value. The dashed line marks the linear-case cutoff r_s / r_c. Curvature converts the knife-edge switch into an interval of partial insurance, but the qualitative rule is unchanged: insure more when shocks are likelier."

library(ggplot2)

grid <- expand.grid(p = seq(0.02, 0.95, by = 0.01), phi = c(0, 0.5, 1.5))
grid$gamma <- mapply(function(p, phi)
  gamma_star(p, rs, rc, Gamma, A_h, A_l, phi), grid$p, grid$phi)

ggplot(grid, aes(p, gamma / Gamma, colour = factor(phi))) +
  geom_vline(xintercept = rs / rc, linetype = "dashed", linewidth = 0.3) +
  geom_line(linewidth = 0.8) +
  scale_colour_discrete(name = "Curvature (phi)") +
  labs(x = "Probability of the high-need state (p)",
       y = "Optimal insured share  gamma* / Gamma",
       title = "The normative policy function") +
  theme_minimal(base_size = 11)
```

The comparative static in the price gap is what makes the model useful for a
marketing setting where the two instruments' prices are institutional facts:

```{r nb-comparative}
#| echo: true
#| message: false
#| warning: false

cutoffs <- data.frame(
  card_rate   = c(0.12, 0.155, 0.195, 0.215, 0.25),
  loan_rate   = 0.068
)
cutoffs$cutoff_p <- cutoffs$loan_rate / cutoffs$card_rate
cutoffs$reading  <- ifelse(cutoffs$cutoff_p > 0.5,
                           "flexible instrument optimal for most consumers",
                           "committed instrument optimal for most consumers")
cutoffs
```

At a 19.5 percent card rate against a 6.8 percent loan rate, any consumer facing
better than a roughly 35 percent chance of an emergency should be pre-borrowing on
the cheap instrument. That threshold is low enough that a large share of the
population sits above it—which is precisely why the empirical question of what
consumers actually do is interesting.

## Departures From the Benchmark {#sec-nb-departures}

A normative model becomes a research instrument only when you enumerate the ways
consumers can fail it, because each failure mode implies a different prediction and
a different welfare sign. The taxonomy that has become standard in consumer
financial regulation distinguishes ignorance of contract terms, ignorance of
financial concepts, ignorance of one's own financial history, and ignorance of
one's own future behavior [@campbell2016restoring; @campbell2011consumer].

**Ignorance of terms.** The consumer optimizes under wrong beliefs about
$r^{c}$ or $r^{s}$. If she believes the rates are equal, @eq-nb-linear-threshold
implies the flexible instrument is weakly better at *every* $p$, so she never
insures. Diagnostic: does she know the rate on her own account?

**Ignorance of concepts.** She misunderstands compounding, inflation, or risk, and
so misprices the future consequence of debt [@lusardi2014economic]. Diagnostic:
standard financial-literacy items. This is the departure most reliably tied to
carrying expensive debt, and it is also the one for which the marketing evidence on
remediation is least encouraging: financial education explains very little
variance in downstream behavior once endowed traits are accounted for
[@fernandes2014financial].

**Ignorance of own history.** She does not know what options she already has—for
example, believing she has exhausted her cheap borrowing capacity when unclaimed
subsidized capacity remains. She then behaves as if $r^{s}$ were prohibitive and
finances everything on the flexible instrument. Diagnostic: compare stated
availability against the administrative record. This departure is invisible in
survey-only and in records-only data; it requires the two to be linked, which is
the methodological point of @sec-complementary-evidence.

**Ignorance of own behavior.** She is present-biased and naive about it
[@odonoghue1999doing]. She selects the instrument correctly ex ante, then
overspends on the flexible one regardless of the state. Diagnostic: elicited
discounting, and the signature that card use is uncorrelated with the emergency
probability that is supposed to drive it.

@tbl-nb-departures maps each departure to its behavioral signature, its
measurement, and—critically—the sign of the welfare effect of a policy that
restricts the flexible instrument.

| Departure | Behavioral signature | How it is measured | Effect of restricting the flexible instrument |
|:--|:--|:--|:--|
| None (optimizer) | Insures iff $p \ge p_2$; known expenses on the cheap instrument | Consistency of choices with @eq-nb-linear-threshold | Weakly negative: choice restriction cannot help someone already optimizing |
| Terms unknown | Uses flexible instrument at all $p$; cannot state own rate | Direct question on own APR | Positive when $p > r^{s}/r^{c}$; negative when $p$ is very low |
| Concepts unknown | Revolves expensive debt; low literacy score | Standard literacy items | Positive on average, magnitude rising in $p$ |
| Own history unknown | Revolves while holding unclaimed cheap capacity | Survey response matched to records | Positive: the restriction pushes search toward the option she wrongly believed unavailable |
| Own behavior unknown (naive present bias) | Borrows the same total but on the expensive instrument; use uncorrelated with $p$ | Elicited discounting; correlation of use with stated shock risk | Weakly positive: same debt, lower price |

: Departures from the normative benchmark, their empirical signatures, and the welfare consequence of a policy that steers consumers away from the expensive flexible instrument. The welfare column is what converts a taxonomy into a research design. {#tbl-nb-departures}

The last column is the payoff of the whole exercise. Restricting choice can only
hurt an optimizer; every positive welfare entry in the table comes from a consumer
who was choosing badly. Therefore the empirical share of each type is not a
descriptive detail—it is the weight in the welfare aggregate, which is exactly how
@sec-welfare-calibration uses it.

::: {.callout-warning}
## The identification problem inside the taxonomy

Several departures predict the same *aggregate* behavior—more expensive borrowing—so
they cannot be separated by the borrowing data alone. Distinguishing them requires
measures that are diagnostic of the mechanism rather than the outcome: knowledge
of one's own rate separates terms-ignorance, literacy items separate
concept-ignorance, record-linked availability separates history-ignorance, and the
correlation between use and stated shock probability separates present bias.
Designing those measures is part of designing the model, not a downstream data
task.
:::

## Abstracting the Recipe {#sec-nb-recipe}

Strip the credit context away and the construction has four moving parts. Any
marketing problem with the same four parts admits the same treatment.

1. **Two or more instruments serving one need, ordered by price and flexibility.**
   Cheap-but-committed versus expensive-but-contingent. The ordering must be
   institutional, not estimated, so the benchmark does not depend on the data.
2. **Uncertainty resolved between the commitment and the contingent draw.** Without
   the timing wedge there is no insurance problem and the cheap instrument
   trivially dominates.
3. **A scalar reduction.** Push everything known into the dominant instrument so
   the decision collapses to a single choice variable. This is what keeps the model
   solvable in closed form and calibratable from a handful of moments.
4. **A cutoff expressed in observable quantities.** The threshold should be a
   function of prices and a probability, not of a preference parameter, because
   prices are posted and probabilities can be elicited.

@tbl-nb-translations applies the recipe to marketing settings.

| Setting | Cheap / committed instrument | Expensive / flexible instrument | Uncertainty | Normative cutoff governed by |
|:--|:--|:--|:--|:--|
| Subscription vs. à la carte | Annual plan or bundle | Pay-per-use | Usage volume next year | Usage probability vs. price ratio |
| Loyalty currency | Points banked toward a fixed redemption | Cash purchase at posted price | Whether the redemption occasion arises | Redemption probability vs. breakage-adjusted price gap |
| Insurance and warranties | Ex ante premium | Ex post out-of-pocket repair | Failure event | Failure probability vs. loading factor |
| Pre-commitment promotions | Prepaid or pre-ordered quantity | Spot purchase during the season | Demand realization | Stockout probability vs. price premium |
| Media and inventory buying | Upfront guaranteed buy | Spot / real-time auction | Audience availability and price | Scarcity probability vs. spot-price premium |
| Data and consent | Broad first-party consent gathered early | Costly per-instance targeting workarounds | Whether the targeting occasion arises | Occasion probability vs. workaround cost ratio |

: The price-versus-flexibility template translated into marketing settings. Each row supports the same normative analysis: characterize the cutoff, measure the departure, and evaluate a policy or platform change that pushes consumers across it. {#tbl-nb-translations}

The same structure appears in the choice-friction literature that produced the
best-known welfare estimates in applied microeconomics: consumers who stay in
dominated health-insurance plans [@handel2013adverse; @handel2015health], elderly
enrollees who choose dominated drug plans [@abaluck2011choice], homeowners who
fail to refinance at large certain gains [@keys2016failure], savers who split
contributions naively across offered funds [@benartzi2001naive], and buyers who
underweight future fuel costs [@busse2013myopic]. In each, the paper's contribution
is a credible normative benchmark plus a measured distance from it. Marketing has
the same opportunity in every category where a firm offers a menu whose optimal
selection is computable.

::: {.callout-tip}
## Keeping the model small on purpose

The temptation, once a normative model is written, is to enrich it: add
heterogeneity, add dynamics, estimate it. Resist until you know what the added
structure buys. In the anchor paper the authors explicitly decline full estimation
because it "requires substantial functional form restrictions on students'
preferences for which theory provides little guidance," and instead derive bounds
that hold for *any* utility function satisfying their assumptions
[@brown2026spillover]. A result that survives an entire class of preferences is
worth more to a referee than a point estimate that survives one.
:::

## From Benchmark to Welfare {#sec-nb-to-welfare}

The normative model has now produced three things a policy evaluation needs. It has
produced a *prediction*—who should use which instrument, and how that varies with
prices, shock risk, and resources—that the reduced-form heterogeneity analysis of
@sec-policy-spillovers can test. It has produced a *taxonomy* of departures, each
with a distinct measurable signature, which tells the survey what to ask. And it
has produced a *welfare expression*: because we know the optimal policy, we can
write the change in the consumer's continuation value when the flexible instrument
is taken away, separately by type.

That last object is the input to @sec-welfare-calibration, where the utility
function is normalized away, the expression is reduced to a formula in $p$ and
$r^{c}/r^{s}$, and the population shares from the survey turn it into a number.

## Key Takeaways {#sec-nb-takeaways}

- The word "mistake" requires a benchmark. A normative model supplies it by
  characterizing what an informed optimizer would do, and it is deliberately
  smaller than a structural model because it is not asked to fit the data.
- Write the normative model before the mechanism data are collected; its function
  is to determine which measurements are diagnostic.
- The workhorse structure is the price-versus-flexibility trade-off
  (@eq-nb-insurance): a cheap instrument committed to before uncertainty resolves
  versus an expensive one drawn after, which is formally an insurance problem.
- The solution is a threshold rule (@eq-nb-full, @eq-nb-none). Under a linear
  continuation value it collapses to the strikingly simple cutoff
  $p \ge r^{s}/r^{c}$ (@eq-nb-linear-threshold), a normative statement requiring
  only two posted prices and one elicited probability.
- Enumerate departures explicitly (@tbl-nb-departures). Each has a distinct
  behavioral signature, a distinct measurement, and—decisively—a distinct sign for
  the welfare effect of restricting the expensive instrument.
- Restricting choice can never help an agent who is already optimizing, so the
  entire welfare case for a paternalistic policy rests on the measured population
  share of each non-optimizing type.
- The recipe (@tbl-nb-translations) transfers to subscriptions, loyalty currencies,
  warranties, pre-commitment promotions, media buying, and consent regimes—any
  marketing menu whose optimal selection is computable from posted terms.

## Further Reading {#sec-nb-further}

The model developed here is that of @brown2026spillover; read it for the formal
propositions and the appendix proofs. The taxonomy of consumer financial mistakes
is due to @campbell2016restoring, with the policy framing in @campbell2011consumer;
@lusardi2014economic surveys the literacy evidence and @fernandes2014financial
provides the marketing field's most influential skeptical assessment of financial
education as a remedy. On naive present bias as a distinct departure, see
@odonoghue1999doing. For exemplary applications of the normative-benchmark strategy
in adjacent fields, @handel2013adverse and @handel2015health on insurance,
@abaluck2011choice on Medicare Part D, @keys2016failure on refinancing,
@benartzi2001naive on retirement allocation, and @busse2013myopic on durable-good
purchases. @gabaix2006shrouded supplies the supply-side counterpart—why competitive
firms may have no incentive to unshroud the attribute the consumer is getting
wrong—and @dellavigna2009psychology surveys the field evidence for departures from
the standard model. The relationship between normative, descriptive, and structural
modeling is developed further in @sec-model-building, @sec-analytical-models, and
@sec-structural-models.
