# Calibration and Welfare Bounds {#sec-welfare-calibration}

The normative model of @sec-normative-benchmarks tells us what an optimizing
consumer should do. The reduced-form design of @sec-policy-spillovers tells us
what consumers actually did when a policy moved them. Neither says whether they
were made better off. This chapter closes that gap, and it does so without
estimating the model.

That last clause is the methodological point. The instinctive route from a model
to a welfare number is full structural estimation: assume a utility function,
estimate its parameters, simulate the counterfactual. That route is often
unavailable, not because of computation but because theory does not pin down the
functional forms, and the answer turns out to depend on the arbitrary choices
[@rust2014limits; @nevo2010dogma]. The alternative developed here is to *normalize
the unknown utility away*, derive a bound that holds for every utility function in
a stated class, and calibrate the handful of quantities the bound actually
contains. In the anchor application the entire welfare conclusion rests on two
posted interest rates and one elicited probability.

## Three Routes From Model to Welfare Number {#sec-wc-routes}

Before building the bound it is worth seeing where it sits among the alternatives,
because the choice among them is a real design decision with real costs.

**Full structural estimation.** Specify preferences, estimate parameters from
choices, and compute compensating variation under the counterfactual. This gives
the most complete answer—distributional detail, general-equilibrium effects,
policies never observed—and demands the most: identification of every parameter
that enters the welfare integral, and defensible functional forms. It is the right
route when the counterfactual is far from anything in the data and when the
functional forms are disciplined by something other than convenience
(@sec-structural-models).

**Sufficient statistics.** Show that the welfare derivative depends on the deep
parameters only through a small number of estimable elasticities, then estimate
those [@chetty2009sufficient; @kleven2021sufficient]. This is the dominant
approach in modern public finance: it buys robustness by refusing to identify
anything it does not need. Its limits are that it is local—valid for marginal
policy changes—and that the sufficiency result must be derived anew for each
setting. The marginal-value-of-public-funds framework is the large-scale
application [@hendren2020unified].

**Bounds under a normalization.** Derive the worst possible welfare outcome for
the policy, express every other case as a fraction of it, and show that a linear
approximation to the unknown utility is conservative. This delivers a signed,
scaled answer valid for an entire class of preferences, at the cost of reporting a
range rather than a point and of saying nothing about magnitudes in dollars
without an extra step. It is the right route when the policy change is discrete,
the preference class is wide, and credibility matters more than precision.

@fig-wc-routes lays out the trade-off.

```{mermaid}
%%| label: fig-wc-routes
%%| fig-cap: "Three routes from a normative model to a welfare number, ordered by how much structure each requires. The bounds route buys robustness by answering a smaller question: not how large the effect is in dollars, but what fraction of a defined worst case it represents, and with what sign."
flowchart LR
  M["Normative model of<br/>optimal behavior"] --> A["Full structural estimation"]
  M --> B["Sufficient statistics"]
  M --> C["Bounds under a<br/>worst-case normalization"]
  A --> A2["Needs: functional forms,<br/>identification of every<br/>welfare-relevant parameter<br/>Gives: dollars, distribution,<br/>unobserved counterfactuals"]
  B --> B2["Needs: a sufficiency proof<br/>plus a few elasticities<br/>Gives: local welfare change,<br/>robust to deep parameters"]
  C --> C2["Needs: a defined worst case<br/>plus a concavity argument<br/>Gives: signed bound valid for<br/>a whole preference class"]
```

## Constructing the Bound {#sec-wc-bound}

We work with the model of @sec-normative-benchmarks. Recall its objects: an
uncertain shortfall $\Gamma$, a high-need state occurring with probability $p$,
rates $r^{s} < r^{c}$ on the committed and flexible instruments, baseline debt
levels $A_h$ and $A_\ell$ carried into each state, and a continuation value $V_t$
that is decreasing and weakly concave in debt.

### Step 1: define the worst case {#sec-wc-worstcase}

Ask what the most damaging possible version of the policy looks like. It is the
case in which every consumer chooses optimally and the shock probability is
vanishingly small: $p \to 0$. Because the requirement must still be met if the
shock occurs, the consumer must have liquidity available; the flexible instrument
is exactly the right tool, since it costs nothing unless drawn. Forcing her onto
the committed instrument makes her pay $r^{s}\Gamma$ in interest on money she
almost certainly will not need. Define the magnitude of that loss:

$$
L \;\equiv\; V_t(A_\ell) \;-\; V_t\!\left(A_\ell + r^{s}\Gamma\right) \;>\; 0 .
$$ {#eq-wc-worstcase}

$L$ is unknown—it depends on the utility function we refused to specify—but it is
a *fixed positive number* for any given consumer, and every welfare change in this
model can be expressed relative to it. That is the trick.

### Step 2: show the linear case is the conservative case {#sec-wc-linear}

Consider the linear function that passes through the two points defining $L$:

$$
\mathcal{L}_t(d) \;=\; \frac{V_t\!\left(A_\ell + r^{s}\Gamma\right) - V_t(A_\ell)}{r^{s}\Gamma}\; d \;=\; -\,\frac{L}{r^{s}\Gamma}\, d .
$$ {#eq-wc-linear}

Because $V_t$ is weakly concave and decreasing, replacing it with the chord
$\mathcal{L}_t$ understates the utility benefit of moving debt out of the
high-debt state and into the low-debt state. Any welfare *gain* computed with the
linear approximation is therefore a **lower bound** on the true gain
[@brown2026spillover]. This is what makes the exercise honest: if the linear
calculation says the policy helped, curvature can only help it more.

### Step 3: evaluate the two consumer types {#sec-wc-types}

For a consumer who chooses optimally, the welfare change from losing access to the
flexible instrument is

$$
\Delta W^{\text{opt}} \;=\; (1-p)\left[V_t\!\left(A_\ell + r^{s}\Gamma\right) - V_t\!\left(A_\ell + r^{s}\gamma^{*}\right)\right] \;+\; p\left[V_t(A_h) - V_t\!\left(A_h + (r^{c}-r^{s})(\Gamma-\gamma^{*})\right)\right],
$$ {#eq-wc-dwopt}

and for a consumer who—because she misunderstands the terms, the concepts, or her
own available options—uses the flexible instrument regardless of $p$,

$$
\Delta W^{\text{bd}} \;=\; (1-p)\left[V_t\!\left(A_\ell + r^{s}\Gamma\right) - V_t(A_\ell)\right] \;+\; p\left[V_t(A_h) - V_t\!\left(A_h + (r^{c}-r^{s})\Gamma\right)\right].
$$ {#eq-wc-dwbd}

Substituting the linear form @eq-wc-linear with slope magnitude
$m = L/(r^{s}\Gamma)$ and dividing through by $L$ collapses both expressions to
functions of two numbers. For the boundedly rational consumer,

$$
\frac{\Delta W^{\text{bd}}}{L} \;=\; \frac{m\Gamma\left(p\,r^{c} - r^{s}\right)}{m\,r^{s}\Gamma} \;=\; p\,\frac{r^{c}}{r^{s}} \;-\; 1 ,
$$ {#eq-wc-boundbd}

and for the optimizer, whose $\gamma^{*}$ is $\Gamma$ when $p \ge r^{s}/r^{c}$ and
$0$ otherwise,

$$
\frac{\Delta W^{\text{opt}}}{L} \;=\;
\begin{cases}
p\,\dfrac{r^{c}}{r^{s}} - 1 \;(<0), & p < \dfrac{r^{s}}{r^{c}},\\[10pt]
0, & p \ge \dfrac{r^{s}}{r^{c}} .
\end{cases}
$$ {#eq-wc-boundopt}

These two expressions carry the entire welfare analysis. Three properties deserve
emphasis.

*The preference function is gone.* @eq-wc-boundbd contains no utility parameter.
It holds for every $V_t$ in the assumed class, and it is a lower bound because of
Step 2.

*The scale is interpretable.* A value of $-1$ means the policy inflicted the full
worst-case harm; $0$ means no effect; a value above $0$ means net benefit, and
values above $1$ mean the benefit exceeds the worst-case harm in magnitude.

*There is a single break-even condition.* Both expressions are positive exactly
when

$$
p \;>\; \frac{r^{s}}{r^{c}} .
$$ {#eq-wc-breakeven}

The policy raises welfare, on average, if and only if the probability of the
high-need state exceeds the ratio of the two interest rates. With a 6.8 percent
loan rate and a 19.5 percent card rate, that threshold is roughly 35 percent
[@brown2026spillover]. This is the whole argument, and it fits on one line.

### Step 4: aggregate to a population {#sec-wc-aggregate}

An individual-level bound becomes a policy conclusion only after accounting for
who was actually affected. Three adjustments are required, and each corresponds to
a measurable share. Not everyone stopped using the flexible instrument
($\%\Delta CC$, the drop in adoption). Not everyone needs liquidity at all
($\%NL$, the share who could not simply ask family). And among those affected,
some optimize and some do not ($\hat{p}^{bd}$, the boundedly rational share).
Writing $\hat{p}$ for the elicited shock probability among those needing
liquidity,

$$
\overline{\Delta W}^{\text{opt}} \;=\; (\%\Delta CC)\,(\%NL)\cdot \min\!\left\{0,\; \hat{p}\,\frac{\overline{r}^{c}}{\overline{r}^{s}} - 1\right\},
$$ {#eq-wc-aggopt}

$$
\overline{\Delta W}^{\text{bd}} \;=\; (\%\Delta CC)\,(\%NL)\cdot\left[\hat{p}\,\frac{\overline{r}^{c}}{\overline{r}^{s}} - 1\right],
$$ {#eq-wc-aggbd}

$$
\overline{\Delta W} \;=\; \left(1-\hat{p}^{bd}\right)\overline{\Delta W}^{\text{opt}} \;+\; \hat{p}^{bd}\,\overline{\Delta W}^{\text{bd}} .
$$ {#eq-wc-aggtotal}

@fig-wc-pipeline shows the full pipeline from model to number.

```{mermaid}
%%| label: fig-wc-pipeline
%%| fig-cap: "The calibration pipeline. Each input enters at exactly one place and comes from a named source; the utility function never enters at all, having been normalized out by the worst-case scaling."
flowchart TB
  MOD["Normative model:<br/>threshold rule in p and r_s / r_c"] --> WC["Define worst case L:<br/>all optimize, p near zero"]
  WC --> LIN["Linear chord is conservative<br/>because V is concave"]
  LIN --> IND["Per-consumer bound:<br/>p (r_c / r_s) minus 1"]
  IND --> AGG["Aggregate over affected population"]
  R1["Posted loan rates<br/>(Dept. of Education)"] --> IND
  R2["Card rates by risk tier<br/>(regulator market report)"] --> IND
  S1["Elicited shock probability p<br/>(survey: emergencies last year)"] --> IND
  S2["Drop in adoption %dCC<br/>(pre vs post national surveys)"] --> AGG
  S3["Share needing liquidity %NL<br/>(survey: would you ask family?)"] --> AGG
  S4["Boundedly rational share<br/>(literacy items + record mismatch)"] --> AGG
  AGG --> OUT["Signed welfare range,<br/>overall and by subgroup"]
```

## Calibration Discipline {#sec-wc-discipline}

A bound is only as credible as its inputs, and the review process will interrogate
each one. The discipline is to state, for every quantity, where the number comes
from, what it would take to be wrong, and in which direction the error pushes the
conclusion.

| Quantity | Source | Direction if mismeasured | Defense |
|:--|:--|:--|:--|
| $r^{s}$, committed-instrument rate | Posted federal loan rates in the policy year | Understating raises the break-even $p$, making benefit harder to find | Posted, non-negotiable, identical across consumers |
| $r^{c}$, flexible-instrument rate | Regulator market report, by credit-risk tier | Overstating inflates the benefit | Report a range across tiers, not a point |
| $p$, shock probability | Survey: emergency expenditures experienced last year | Overstating inflates the benefit | Cross-check retrospective against prospective elicitation |
| $\%\Delta CC$, adoption drop | Pre-policy national survey vs. own post-policy survey | Overstating scales all effects up | Anchor to an independent industry survey |
| $\%NL$, share needing liquidity | Survey: would not ask family in an emergency | Overstating scales effects up | Conservative coding of ambiguous responses |
| $\hat{p}^{bd}$, boundedly rational share | Literacy items plus survey-versus-record mismatch on available credit | Overstating shifts weight to the positive branch | Report the optimizer branch separately so readers can reweight |

: Calibration ledger for the welfare bound. Every input is a measured share or a posted price; none is a preference parameter. The final column is what makes the ledger auditable rather than assumed. {#tbl-wc-ledger}

::: {.callout-important}
## Do not calibrate to the outcome you are explaining

The cardinal sin is to pick parameters so that the model reproduces the treatment
effect, then present the implied welfare number as evidence. The estimate and the
calibration must draw on non-overlapping information: here, the treatment effect
comes from administrative borrowing records, while the calibration comes from
posted rates, a regulator's market report, and a survey. When the two do
coincide—as when a moment used in calibration is also an estimated outcome—say so
explicitly and show the result with that moment excluded.
:::

## Reproducing the Welfare Table {#sec-wc-replication}

The following code implements @eq-wc-aggopt through @eq-wc-aggtotal and reproduces
the published welfare bounds of @brown2026spillover from its reported inputs. It
is worth running not because the arithmetic is hard but because a welfare table
whose every cell can be regenerated from six numbers is the strongest possible
demonstration that the analysis contains no hidden degrees of freedom.

```{r wc-welfare-table}
#| echo: true
#| message: false
#| warning: false

# --- Calibration inputs, as reported in the source study ---------------------
# Rate ranges: subsidized/unsubsidized federal loan rates in 2010, and prime/
# subprime average card rates from the regulator's CARD Act market report.
rs_range <- c(low = 0.068, high = 0.045)   # high benefit uses the LOWER loan rate
rc_range <- c(low = 0.195, high = 0.215)   # high benefit uses the HIGHER card rate

# Survey-calibrated population shares, overall and by ZIP income quartile.
cal <- data.frame(
  group   = c("All", "0-25 pctl", "25-50 pctl", "50-75 pctl", "75-100 pctl"),
  d_cc    = c(30.35, 33.25, 36.90, 27.06, 22.22) / 100,  # drop in card adoption
  need_l  = c(46.61, 52.50, 43.17, 45.80, 44.61) / 100,  # share needing liquidity
  p_hat   = c(47.12, 56.19, 49.72, 43.04, 35.57) / 100,  # prob. of high state
  p_bd    = c(62.06, 69.25, 62.93, 62.03, 52.40) / 100   # boundedly rational share
)

# --- The bound itself --------------------------------------------------------
per_consumer <- function(p, rc, rs) p * (rc / rs) - 1            # eq. 13 scaled

welfare_bounds <- function(d_cc, need_l, p_hat, p_bd) {
  scale <- d_cc * need_l
  ends <- c("low", "high")
  bd  <- sapply(ends, function(e) scale * per_consumer(p_hat, rc_range[[e]], rs_range[[e]]))
  opt <- sapply(ends, function(e) scale * min(0, per_consumer(p_hat, rc_range[[e]], rs_range[[e]])))
  tot <- (1 - p_bd) * opt + p_bd * bd
  c(opt_lo = opt[["low"]], opt_hi = opt[["high"]],
    bd_lo  = bd[["low"]],  bd_hi  = bd[["high"]],
    tot_lo = tot[["low"]], tot_hi = tot[["high"]])
}

res <- t(mapply(welfare_bounds, cal$d_cc, cal$need_l, cal$p_hat, cal$p_bd))
out <- data.frame(group = cal$group, round(100 * res, 2))
out
```

Read the table as the source study does. The optimizer branch is exactly zero
everywhere, because the calibrated shock probability exceeds the break-even ratio
$r^{s}/r^{c}$ in every subgroup: for a consumer who was already choosing well, the
policy is non-binding. Among the boundedly rational the gain is 5 to 18 percent of
the worst-case magnitude overall, and the population-weighted total is 3 to 11
percent. The gradient across quartiles is the substantive result: 7 to 20 percent
for the least affluent, 0.1 to 4 percent for the most affluent—the same ordering
the reduced-form spillover estimates produced, arrived at through an entirely
different chain of reasoning.

::: {.callout-note}
## Why the two branches move for different reasons

The optimizer branch is zero not because optimizers are unaffected in general but
because *at these calibrated values* they were already fully insuring. Lower the
shock probability below $r^{s}/r^{c}$ and the branch turns negative immediately:
the policy would then be destroying a genuinely useful option. The bound is
therefore not rigged toward a favorable answer; it is one calibrated input away
from delivering an unfavorable one.
:::

## Sensitivity Analysis Is the Result {#sec-wc-sensitivity}

In a bounds exercise the sensitivity analysis is not a robustness appendix; it is
the substance, because the reader's real question is how far the inputs would have
to move to flip the sign. Three presentations do most of the work.

**The break-even surface.** Plot the welfare bound over the two quantities that
actually enter it, and mark the zero contour.

```{r wc-heatmap}
#| echo: true
#| message: false
#| warning: false
#| label: fig-wc-surface
#| fig-cap: "The welfare bound for a boundedly rational consumer over the two quantities it depends on: the probability of the high-need state and the price ratio of the two instruments. The heavy contour is the break-even locus p = r_s / r_c; the point marks the study's central calibration. Everything above the contour is a welfare gain."

library(ggplot2)

surf <- expand.grid(p = seq(0.05, 0.90, by = 0.01),
                    ratio = seq(1.5, 6.0, by = 0.05))     # ratio = r_c / r_s
surf$bound <- surf$p * surf$ratio - 1

central <- data.frame(p = 0.4712, ratio = 0.195 / 0.068)

ggplot(surf, aes(ratio, p)) +
  geom_raster(aes(fill = pmax(pmin(bound, 2), -1))) +
  geom_contour(aes(z = bound), breaks = 0, colour = "white", linewidth = 0.9) +
  geom_point(data = central, size = 2.4, colour = "white") +
  scale_fill_gradient2(name = "Bound\n(share of\nworst case)",
                       low = "#8c1616", mid = "grey92", high = "#14507a",
                       midpoint = 0) +
  labs(x = "Price ratio  r_c / r_s",
       y = "Probability of high-need state (p)",
       title = "Where the policy helps and where it hurts") +
  theme_minimal(base_size = 11)
```

**The break-even input.** Report, for each subgroup, the value of the pivotal input
at which the conclusion flips. This is more useful to a policy audience than a
confidence interval, because it is stated in the units of the thing they might
dispute.

```{r wc-breakeven}
#| echo: true
#| message: false
#| warning: false

breakeven <- data.frame(
  group        = cal$group,
  p_calibrated = cal$p_hat,
  p_breakeven  = 0.068 / 0.195,               # conservative rate pair
  slack        = cal$p_hat - 0.068 / 0.195
)
breakeven$verdict <- ifelse(breakeven$slack > 0, "net benefit", "net harm")
round_cols <- c("p_calibrated", "p_breakeven", "slack")
breakeven[round_cols] <- lapply(breakeven[round_cols], round, 3)
breakeven
```

Even the most affluent quartile, with the lowest elicited emergency probability,
sits above the conservative break-even threshold—but only barely, which is the
honest way to describe a result whose sign is secure and whose magnitude is not.

**One-way sensitivity.** Vary each input across its plausible range holding the
rest at their central values, and show the resulting swing in the aggregate.

```{r wc-tornado}
#| echo: true
#| message: false
#| warning: false
#| label: fig-wc-tornado
#| fig-cap: "One-way sensitivity of the aggregate welfare bound. Each bar shows the range produced by moving a single calibrated input across its plausible interval with the others fixed. The price ratio and the shock probability dominate; the population shares scale the answer without threatening its sign."

base <- list(d_cc = 0.3035, need_l = 0.4661, p_hat = 0.4712,
             p_bd = 0.6206, rc = 0.195, rs = 0.068)

agg <- function(x) {
  bd  <- x$d_cc * x$need_l * (x$p_hat * x$rc / x$rs - 1)
  opt <- x$d_cc * x$need_l * min(0, x$p_hat * x$rc / x$rs - 1)
  100 * ((1 - x$p_bd) * opt + x$p_bd * bd)
}

ranges <- list(
  `Shock probability p`      = list(p_hat  = c(0.30, 0.60)),
  `Card rate r_c`            = list(rc     = c(0.155, 0.215)),
  `Loan rate r_s`            = list(rs     = c(0.045, 0.068)),
  `Adoption drop %dCC`       = list(d_cc   = c(0.20, 0.40)),
  `Needing liquidity %NL`    = list(need_l = c(0.35, 0.60)),
  `Bounded-rational share`   = list(p_bd   = c(0.45, 0.80))
)

rows <- do.call(rbind, lapply(names(ranges), function(nm) {
  par <- names(ranges[[nm]]); vals <- ranges[[nm]][[1]]
  lo <- base; lo[[par]] <- vals[1]
  hi <- base; hi[[par]] <- vals[2]
  data.frame(input = nm, low = agg(lo), high = agg(hi))
}))
rows$span  <- abs(rows$high - rows$low)
rows$input <- factor(rows$input, levels = rows$input[order(rows$span)])

ggplot(rows) +
  geom_hline(yintercept = agg(base), linetype = "dashed", linewidth = 0.3) +
  geom_segment(aes(x = input, xend = input, y = low, yend = high),
               linewidth = 3.2, colour = "grey55") +
  coord_flip() +
  labs(x = NULL, y = "Aggregate welfare bound (% of worst case)",
       title = "One input at a time",
       subtitle = "Dashed line: central calibration") +
  theme_minimal(base_size = 11)
```

## Failure Modes {#sec-wc-failures}

Welfare calibration goes wrong in recognizable ways, and reviewers know all of
them.

**Calling profit "welfare."** A firm-side calculation of incremental margin is not
a welfare analysis. If the paper's welfare object is consumer surplus, the
firm's response must either be shown to be irrelevant or be modeled; if the object
is total surplus, the producer side must be counted. State the object in the first
paragraph of the welfare section.

**Ignoring the supply side.** Restricting one marketing instrument changes firms'
incentives, and equilibrium price and product responses can undo or amplify the
consumer-side conclusion [@gabaix2006shrouded; @johnson2023privacy]. A partial
analysis is defensible when the treated market is small relative to the firms'
scale; it must be argued, not assumed.

**Unbounded or ill-defined worst cases.** The normalization in @eq-wc-worstcase
works because the worst case is finite and well-defined. If the worst case is
unbounded—if a consumer could in principle be driven to a $-\infty$ region of the
utility function—the scaling is meaningless, and the analysis must restrict the
domain first, as the anchor model does by assuming enough liquidity to meet
subsistence.

**Using revealed preference where the paper has argued preferences are unreliable.**
This is the deepest tension in behavioral welfare economics: the same evidence that
establishes consumers make mistakes disqualifies their choices as the welfare
metric [@bernheim2009beyond]. The bounds approach handles it by evaluating each
type against the *model's* optimum rather than against its own revealed choice,
and by reporting the optimizer branch separately so a reader who rejects the
paternalistic premise can read the number that assumes it away.

**Local answers presented as global.** Sufficient-statistic and marginal-elasticity
formulas are valid for small changes [@chetty2009sufficient]. A campus-wide
elimination of card marketing is not a small change, which is one reason a discrete
bounds argument suits it better.

::: {.callout-warning}
## The welfare claim is a modeling claim, and should be labeled as one

The strongest phrasing available to a bounds paper is that model-based evidence
*implies* the policy raised welfare, under a stated preference class and a stated
calibration. That is weaker than "we estimate that welfare rose by \$X" and much
more defensible. Papers that overclaim here invite a referee to attack the weakest
input, and there is always a weakest input.
:::

## Marketing Applications {#sec-wc-marketing}

The template applies wherever a firm or regulator moves consumers between two
instruments whose prices are observable and whose optimal use is characterizable.

*Plan and bundle migration.* When a carrier retires an à la carte option, the
migrated customer's welfare change is governed by the same ratio of committed to
contingent price and the probability of the usage event. The break-even usage rate
is computable before any estimation.

*Promotion restrictions and price-display rules.* Restricting a promotional
instrument moves purchases to other instruments; the welfare object is whether the
destination instrument is cheaper for the consumer at her usage frequency.

*Privacy and targeting regimes.* Here the flexible instrument is precise targeting
and the committed one is broad reach. Bounding the consumer-side effect requires
the analogue of $p$—the probability that the targeted occasion arises—and the
analogue of the price ratio, the cost premium of reaching the consumer without
targeting [@goldfarb2011privacy; @aridor2023privacy].

*Default architecture.* When a firm changes a default, the affected population is
those who would not have opted in, the "price ratio" is the cost difference between
the default and the alternative, and the boundedly rational share is the group for
whom the default is binding rather than preferred [@choi2003optimal;
@allcott2019welfare].

*Sin-good marketing.* Excise taxes and advertising bans on unhealthy categories
have an established optimal-policy literature that combines an internality
parameter with demand elasticities [@allcott2019regressive; @allcott2019sugar;
@odonoghue2006optimal]; a marketing paper can contribute the demand and
substitution side, which is where the field's comparative advantage lies
[@seiler2021; @dubois2018banning].

## Key Takeaways {#sec-wc-takeaways}

- There are three routes from a normative model to a welfare number—full
  structural estimation, sufficient statistics, and bounds under a normalization—and
  the bounds route trades precision for validity across an entire preference class.
- The construction is: define the worst case (@eq-wc-worstcase), show the linear
  chord is conservative because the value function is concave (@eq-wc-linear), and
  express every welfare change as a fraction of the worst case.
- The result is a formula with no preference parameters in it:
  $\Delta W / L = p\,(r^{c}/r^{s}) - 1$ (@eq-wc-boundbd), with a single break-even
  condition $p > r^{s}/r^{c}$ (@eq-wc-breakeven).
- Aggregation requires three measured shares—who stopped using the instrument, who
  needed liquidity, and who was choosing badly—and each enters at exactly one place
  (@eq-wc-aggtotal).
- Publish a calibration ledger (@tbl-wc-ledger) naming the source and the direction
  of bias for every input; a welfare table that can be regenerated from six numbers
  has no hidden degrees of freedom.
- In a bounds paper the sensitivity analysis *is* the result: report the break-even
  value of the pivotal input, the surface over the two quantities that matter, and
  a one-way sensitivity range.
- The most common failures are calling profit welfare, ignoring firm response,
  normalizing by an unbounded worst case, using revealed preference after arguing
  preferences are unreliable, and presenting local formulas as global answers.

## Further Reading {#sec-wc-further}

The bounds construction is developed in @brown2026spillover; the general case for
behavioral welfare analysis that does not treat choice as automatically revealing
preference is @bernheim2009beyond. For the sufficient-statistics alternative, start
with @chetty2009sufficient and the modern reassessment in @kleven2021sufficient,
then @hendren2020unified for large-scale application and @einav2010estimating for
an influential worked case in insurance. @chetty2015behavioral is the best short
statement of how behavioral parameters enter policy design; @allcott2019regressive
and @odonoghue2006optimal give the optimal-tax treatment of internalities, with
@allcott2019sugar as the accessible summary. @allcott2019welfare is the closest
methodological cousin to the exercise here—a welfare evaluation of a nudge that
takes the possibility of harm seriously. On partial identification as a general
strategy, @tamer2010partial and @manski2002. The structural alternative and its
critics are covered in @nevo2010dogma, @rust2014limits, and @dellavigna2018structural;
the estimation machinery itself is in @sec-structural-models, and the
identification background in @sec-causal-inference.
