# Complementary Evidence and Triangulation {#sec-complementary-evidence}

A welfare bound is an argument, and arguments can be wrong in ways that no amount
of internal rigor detects. The bound of @sec-welfare-calibration rests on a model
the authors wrote, a preference class they assumed, and survey moments they
elicited. Each link is defensible; none is verifiable from inside the chain. What
makes the conclusion believable is a second piece of evidence, produced by a
different method, whose failure modes have nothing to do with the first.

This is what the anchor paper does in its final move. Having concluded from a
calibrated model that restricting campus credit-card marketing raised student
welfare, it goes back to the administrative records and asks a blunter question:
did the students whose borrowing shifted actually do better in school? Final grade
point averages rose about 2 percent and on-time graduation rates by nearly 10
percent among students from lower-income communities, relative to the top-income
group whose borrowing was unaffected [@brown2026spillover]. That result is looser
and more suggestive than the model-based bound. It is also completely independent
of it: no utility function, no elicited probability, no assumption about the
preference class. If the model were wrong in some way the authors did not
anticipate, there is no obvious reason grades would move.

This chapter is about designing that second leg deliberately rather than adding it
opportunistically. It develops what makes evidence genuinely complementary,
how the anchor study built three legs that fail for different reasons, why the
third leg required a *new* identification strategy, and the inferential hazards—
multiple outcomes, forking paths, correlated biases—that turn triangulation into
an illusion of confirmation.

## What Makes Evidence Complementary {#sec-ce-definition}

Adding more evidence is not the same as adding independent evidence. A second
specification of the same regression on the same data confirms that the first was
computed correctly; it does not confirm that the design was valid. The useful
criterion is about *failure modes*:

> Two pieces of evidence are complementary for a claim when the assumptions that
> would have to fail for the first to be wrong are largely disjoint from the
> assumptions that would have to fail for the second.

Under that definition, robustness checks are usually not complementary, because
they share the identifying assumption. A different outcome measured on the same
design is partly complementary: it shares the design but not the measurement. A
different design on a different data source measuring a different construct is
fully complementary. @fig-ce-independence makes the logic explicit.

```{mermaid}
%%| label: fig-ce-independence
%%| fig-cap: "Why agreement between two analyses is informative only when their failure modes are disjoint. Two designs sharing an identifying assumption will agree whenever that assumption fails, so agreement carries almost no information about validity."
flowchart TB
  CL["Claim: the policy improved<br/>consumer welfare"] --> E1["Evidence A:<br/>model-based welfare bound"]
  CL --> E2["Evidence B:<br/>downstream real outcomes"]
  E1 --> F1["Fails if: preference class wrong,<br/>elicited p biased, worst case<br/>misdefined, shares mismeasured"]
  E2 --> F2["Fails if: control group invalid,<br/>outcome confounded, cohort<br/>composition changed"]
  F1 --> J{"Do the failure sets<br/>overlap?"}
  F2 --> J
  J -->|"Largely disjoint"| G["Agreement is informative:<br/>both would have to fail<br/>for different reasons"]
  J -->|"Shared assumption"| B["Agreement is nearly<br/>uninformative: one failure<br/>produces both results"]
```

There is a second, softer criterion that matters in practice: complementary
evidence should be *closer to the welfare object* than the primary evidence, or
else *further from it but more directly observed*. The model-based bound is close
to welfare but heavily assumed; graduation rates are further from the concept of
welfare but directly measured. Neither dominates, and the pairing is stronger than
either.

## The Three-Legged Design {#sec-ce-three-legs}

The anchor study is best understood as three studies sharing a research question.
@tbl-ce-legs sets them side by side.

| | Leg 1: administrative panel | Leg 2: matched survey | Leg 3: academic outcomes |
|:--|:--|:--|:--|
| Question | Did the untargeted market move? | Why did it move, and were choices optimal? | Did the shift show up in real outcomes? |
| Data | Student-semester borrowing records | Survey linked to individual records | Student-level GPA, graduation, major |
| Design | Event-study DiD, freshmen as control | Cross-sectional regressions with record-based covariates | Cohort DiD, top-income quartile as control |
| Identifying assumption | Parallel trends between entering and continuing classes | Survey responses measure beliefs and knowledge validly | Cohorts before and after the rule comparable within quartile |
| Main threat | Recession confound differentially by class | Survey run years after the policy | Compositional change in who enrolls |
| What it cannot do | Say whether the shift helped | Establish causality | Isolate the mechanism |

: The three legs of the anchor study. No leg is sufficient; each covers the others' blind spots, and each has a threat the others do not share. {#tbl-ce-legs}

The design lesson is in the last two rows. Every leg has a fatal-sounding weakness,
and in each case the weakness is answered by a different leg rather than by a
better version of the same leg. The survey cannot establish causality—so causality
comes from the panel. The panel cannot establish welfare—so welfare comes from the
model calibrated on the survey, and is corroborated by academic outcomes. The
academic outcomes cannot identify a mechanism—so the mechanism comes from the
survey.

### Leg 2: linking survey responses to administrative records {#sec-ce-matched}

The single most productive design choice in the study is unglamorous: the survey
was administered to students at the same institution and *matched to their
administrative records*. This converts a set of stated beliefs into a set of
belief-versus-reality comparisons, which is a categorically different kind of
measurement.

The payoff is visible in the departure that could not be detected any other way.
Roughly a fifth of card-holding students who had taken student loans were unaware
that additional loan capacity remained available to them, and students in that
position were markedly more likely to borrow on their card
[@brown2026spillover]. Neither data source alone can see this. The records show
unclaimed capacity but not the student's belief about it; the survey shows the
belief but not whether it is true. Only the join identifies *ignorance of one's own
financial history* as a distinct departure from the benchmark of
@sec-normative-benchmarks—and it is the departure with the clearest welfare
implication, since a consumer who wrongly believes the cheap option is exhausted
is not making a trade-off at all.

::: {.callout-tip}
## Designing a record-linked survey

Three rules make the join informative rather than decorative. (1) Ask at least one
question whose truth value the records determine, so belief accuracy becomes an
observable. (2) Ask about the *decision inputs* the normative model uses—here,
perceived emergency probability, perceived rates, perceived availability—not just
about the outcome, which the records already give you. (3) Report the selection
into the survey against the population on record-based characteristics, since you
can, and few survey studies can.
:::

The honest limitation, which the study states rather than buries, is temporal: the
survey was fielded years after the policy, so it describes the students of a later
cohort. The defense is not that this does not matter but that it is
*directionally* interpretable—the assumption needed is that suboptimal behavior
among card users had not substantially worsened in the interim, and comparisons to
contemporaneous industry surveys from the policy period support that. Naming the
assumption and testing it against outside data is the move; asserting that the
sample is "broadly representative" is not.

### Leg 3: when the complementary analysis needs its own identification {#sec-ce-newid}

Here is the craft point that most repays study. The event-study design of Leg 1
used incoming freshmen as controls, exploiting the fact that they made borrowing
decisions before arriving on campus. That design is unavailable for academic
outcomes, because GPA and on-time graduation are *student-level* outcomes
accumulated over four or more years: a student is not a freshman and a junior for
purposes of a graduation rate.

So the authors changed control groups. Since the borrowing spillover was
concentrated in the bottom income quartiles and was statistically zero in the top
quartile, the top quartile becomes a natural control for the academic analysis:
students whose financing did not change should show no achievement change. The
specification compares cohorts that began study just before the rule to those
beginning just after, differencing across income quartiles:

$$
y^{k}_{i} \;=\; \alpha_{q} \;+\; \alpha_{s} \;+\; \sum_{q=1}^{3}\beta^{k}_{q}\,\mathbb{1}(s \ge 2010)_{i}\times\mathbb{1}(Qtr = q)_{i} \;+\; \gamma^{k}X_{i} \;+\; \epsilon^{k}_{i},
$$ {#eq-ce-academic}

with $k \in \{\text{GPA},\ \text{on-time graduation},\ \text{major choice}\}$,
$\alpha_q$ income-quartile and $\alpha_s$ start-semester fixed effects.

Three features make this more than a mechanical re-run.

*The control group is chosen by a prior result, not by convenience.* The top
quartile is a valid control precisely because Leg 1 established that its borrowing
did not move. That is a genuine chain of inference across legs, and it is why the
legs must be presented in this order.

*A null outcome is included as a placebo.* Major choice (engineering or business)
shows no effect. If the estimated GPA and graduation gains were driven by a
compositional shift toward more able students in later cohorts, one would expect
lucrative-major enrollment to move too. It does not, which weakens the
compositional story [@brown2026spillover].

*The claim is downgraded to match the design.* The academic result is described as
suggestive and more direct, not as the paper's headline. A complementary analysis
that overclaims damages the primary result by association.

@fig-ce-chain shows how the legs connect.

```{mermaid}
%%| label: fig-ce-chain
%%| fig-cap: "How the three legs interlock. The heterogeneity result from the first leg supplies the control group for the third; the survey supplies the calibration inputs for the model; the model supplies the welfare claim that the third leg corroborates without sharing its assumptions."
flowchart TB
  L1["Leg 1: event-study DiD<br/>on borrowing records"] --> HET["Result: spillover concentrated<br/>in low-income quartiles,<br/>zero at the top"]
  HET --> CTRL["Top quartile becomes the<br/>control group for Leg 3"]
  L2["Leg 2: survey matched<br/>to records"] --> SHARES["Shares: needing liquidity,<br/>shock probability,<br/>boundedly rational"]
  SHARES --> MOD["Calibrated welfare bound"]
  L1 --> MOD
  CTRL --> L3["Leg 3: cohort DiD on GPA<br/>and on-time graduation"]
  MOD --> CONC["Conclusion: policy raised welfare,<br/>most for the least affluent"]
  L3 --> CONC
  L3 --> PL["Placebo: major choice<br/>shows no effect"]
```

## When Agreement Is Not Evidence {#sec-ce-correlated}

Triangulation has a failure mode of its own: two analyses can agree because they
share a bias rather than because both are right. The simulation below makes the
point quantitatively. Two designs estimate the same true effect; each is biased;
we vary how correlated the biases are and ask how much agreement between them
should update our belief.

```{r ce-correlated-bias}
#| echo: true
#| message: false
#| warning: false
#| label: fig-ce-agreement
#| fig-cap: "How informative is agreement between two studies? Each panel shows 1,500 pairs of estimates from two biased designs. When the biases are independent (left), agreement near the truth is genuinely diagnostic. When the biases are highly correlated (right), the two designs agree almost perfectly with each other while both being wrong."

library(ggplot2)
set.seed(69)

simulate_pair <- function(n = 1500, truth = 1.0, bias_sd = 0.8,
                          noise_sd = 0.25, rho = 0) {
  # Two designs, each with its own bias draw; rho is the correlation of biases.
  z  <- rnorm(n)
  b1 <- bias_sd * z
  b2 <- bias_sd * (rho * z + sqrt(1 - rho^2) * rnorm(n))
  data.frame(
    est_A = truth + b1 + rnorm(n, 0, noise_sd),
    est_B = truth + b2 + rnorm(n, 0, noise_sd),
    rho   = paste0("Bias correlation = ", rho)
  )
}

pairs_df <- do.call(rbind, lapply(c(0, 0.5, 0.95), function(r)
  simulate_pair(rho = r)))

ggplot(pairs_df, aes(est_A, est_B)) +
  geom_point(alpha = 0.18, size = 0.7) +
  geom_abline(slope = 1, intercept = 0, linewidth = 0.3) +
  geom_hline(yintercept = 1, linetype = "dashed", linewidth = 0.3) +
  geom_vline(xintercept = 1, linetype = "dashed", linewidth = 0.3) +
  facet_wrap(~ rho) +
  labs(x = "Estimate from design A", y = "Estimate from design B",
       title = "Agreement between designs is informative only when biases are independent",
       subtitle = "Dashed lines mark the true effect of 1.0") +
  theme_minimal(base_size = 11)
```

We can quantify the intuition: conditional on the two estimates agreeing closely,
how often are they both close to the truth?

```{r ce-agreement-value}
#| echo: true
#| message: false
#| warning: false

diagnostic_value <- function(rho, tol_agree = 0.25, tol_truth = 0.35) {
  d <- simulate_pair(n = 40000, rho = rho)
  agree <- abs(d$est_A - d$est_B) < tol_agree
  both_right <- abs(d$est_A - 1) < tol_truth & abs(d$est_B - 1) < tol_truth
  p_given <- mean(both_right[agree])
  p_uncond <- mean(both_right)
  c(rho = rho,
    P_agree = mean(agree),
    P_right_given_agree = p_given,
    P_right_unconditional = p_uncond,
    lift = p_given / p_uncond)
}

round(t(sapply(c(0, 0.5, 0.95), diagnostic_value)), 3)
```

The last column is the quantity of interest: how much observing agreement raises
the probability that both estimates are near the truth. With independent biases,
agreement is rare and carries a large lift—seeing it multiplies the odds that both
designs are close to right by a substantial factor. As the biases become
correlated, agreement becomes common and the lift shrinks toward one: the two
designs are converging on each other rather than on the truth, because they are
increasingly the same study run twice. The practical instruction is to ask, before
running the second analysis, *what would make both of these wrong at once*—and if
the answer is a short list, redesign rather than proceed.

## Inferential Hazards in Multi-Leg Designs {#sec-ce-hazards}

A paper with three legs and several outcomes per leg has many more chances to find
something. Three safeguards are now expected.

**Multiple hypotheses.** Reporting the effect on GPA, on-time graduation, major
choice, credits attempted, and retention invites a false positive somewhere.
Family-wise error control—stepdown procedures based on the bootstrap rather than a
crude Bonferroni correction—preserves power while controlling the error rate
[@list2019multiple]. The code below shows the difference on a set of correlated
outcomes.

```{r ce-mht}
#| echo: true
#| message: false
#| warning: false

set.seed(690)

# Five correlated academic outcomes; only the first two carry a true effect.
n <- 2500
treat  <- rbinom(n, 1, 0.5)
common <- rnorm(n)
true_effects <- c(GPA = 0.10, OnTime = 0.09, Major = 0, Credits = 0, Retention = 0)

Y <- sapply(true_effects, function(te)
  te * treat + 0.7 * common + rnorm(n, 0, 1))

pvals <- apply(Y, 2, function(y) coef(summary(lm(y ~ treat)))["treat", 4])

data.frame(
  outcome    = names(pvals),
  p_raw      = round(pvals, 4),
  p_holm     = round(p.adjust(pvals, "holm"), 4),
  p_bh_fdr   = round(p.adjust(pvals, "BH"), 4),
  truth      = ifelse(true_effects > 0, "real", "null")
)
```

The instructive pattern is not that correction is conservative—everyone knows
that—but *which* result it kills. The three null outcomes stay null under every
procedure, as they should. Among the two real effects, the stronger survives
family-wise control and the marginal one does not, which is exactly the situation
where a paper is tempted to report the raw p-value and describe the outcome as
"marginally significant." The defensible move is to declare in advance which
outcome is primary, control the error rate across the announced family, and label
the rest exploratory.

**Forking paths.** Every leg involves discretionary choices—which cohorts to drop,
how to bin income, which covariates to include. Reporting the full distribution of
estimates across defensible specifications, rather than one preferred
specification plus a robustness table, is now a recognized standard
[@simonsohn2020; @steegen2016multiverse]. A specification curve is especially
valuable in a spillover paper, because the sample-construction choices
(@sec-policy-spillovers) are numerous and consequential.

**Pre-specification.** Registering the primary outcome, the control group, and the
subgroup splits before analysis converts a suggestive complementary result into a
confirmatory one, at the cost of flexibility that observational work often needs
[@olken2015promises]. For retrospective administrative-data studies, the practical
compromise is to pre-specify the complementary analysis before running it, and to
say in the paper that this was done.

::: {.callout-warning}
## The complementary analysis should be pre-committed, not harvested

The strongest version of the third leg is one chosen *because the theory predicts
it*, announced before it is estimated. The weakest is one selected after searching
the outcome space for something significant. Since the reader cannot distinguish
these from the text, the burden is on the author to make the prediction explicit
before the result: the theory says a shift to cheaper financing should relieve
liquidity pressure during school, therefore we test achievement outcomes, and
therefore a null on major choice is a placebo rather than a disappointment.
:::

## Validating Models Against Held-Out Evidence {#sec-ce-validation}

A distinct and more demanding form of complementary analysis is to validate the
model itself out of sample. The canonical example estimates a behavioral model on
pre-experimental data and then predicts the effect of an experiment the model never
saw, comparing prediction to realization [@todd2006assessing]. This is the strongest
available evidence that a model's structure, not just its fit, is right.

Marketing has natural opportunities for this that are under-exploited. A firm that
calibrates a normative model of customer plan choice can predict the take-up
consequence of a plan change before rolling it out, then compare. A platform that
models optimal consumer search can predict the effect of a ranking change on an
untouched market. The general recipe is: calibrate on period one, predict period
two's policy change, publish the prediction, then report the comparison
[@dellavigna2018structural]. The relationship between such validation exercises
and the broader structural-versus-reduced-form debate is the subject of
@nevo2010dogma, @angrist2010credibility, and @rust2014limits; the practical
resolution most policy papers adopt is the one used here—reduced-form evidence for
the causal claim, a small model for the welfare claim, and an independent check
that the two are consistent.

## A Design Checklist {#sec-ce-checklist}

@tbl-ce-checklist converts the chapter into a planning tool. Fill it in before
collecting data, not after.

| Question | What to specify in advance |
|:--|:--|
| What is the primary claim? | One sentence, with the welfare object named |
| What is the primary evidence? | Design, estimand, identifying assumption |
| What is the complementary evidence? | A different data source or design measuring a different construct |
| Are the failure modes disjoint? | List what would have to fail for each; if the lists overlap, redesign |
| Does the complementary leg need its own identification? | Usually yes; name its control group and why it is valid |
| Is there a placebo outcome? | An outcome the theory says should not move |
| How many hypotheses are tested? | Pre-list them; state the multiplicity correction |
| What is the discretion in sample construction? | Enumerate; plan a specification curve |
| Which claims are confirmatory and which exploratory? | Label each in the text |
| What would change your conclusion? | The break-even value of the pivotal input |

: Pre-analysis checklist for a multi-leg policy paper. The rows that most often go unanswered—disjoint failure modes, a separate identification for the complementary leg, and an explicit placebo—are the ones referees ask about. {#tbl-ce-checklist}

## Key Takeaways {#sec-ce-takeaways}

- Evidence is complementary only when the assumptions whose failure would
  invalidate it are largely disjoint from those of the primary analysis; robustness
  checks that share the identifying assumption are not complementary.
- A strong policy paper has legs that fail for different reasons: a causal design
  that cannot speak to welfare, a mechanism study that cannot speak to causality,
  and a downstream-outcome test that cannot isolate mechanism (@tbl-ce-legs).
- Linking a survey to administrative records makes belief accuracy observable and
  is often the only way to detect ignorance-of-own-options as a distinct departure
  from the normative benchmark.
- The complementary leg usually needs a *new* identification strategy, and the best
  source for its control group is a heterogeneity result from the primary leg—a
  group the first analysis showed was unaffected.
- Include a placebo outcome the theory says should not move; a null there does more
  for credibility than another significant coefficient.
- Agreement between two analyses is informative only in proportion to how
  independent their biases are; ask what would make both wrong at once before
  running the second.
- Control the family-wise error rate across the outcomes a multi-leg design
  generates, report the specification distribution rather than one preferred model,
  and pre-commit the complementary test so it reads as confirmatory rather than
  harvested.
- Out-of-sample validation—calibrate, predict a change the model never saw, then
  compare—is the most demanding and most persuasive complementary analysis
  available.

## Further Reading {#sec-ce-further}

The three-leg design discussed throughout is @brown2026spillover. On out-of-sample
validation of behavioral models, @todd2006assessing remains the reference case, and
@dellavigna2018structural surveys the broader practice of combining structure with
experimental or quasi-experimental variation. The structural-versus-design debate
that frames these choices is set out in @angrist2010credibility, @nevo2010dogma,
and @rust2014limits. For the inferential hazards, @list2019multiple on multiple
hypothesis testing in experimental work, @simonsohn2020 and @steegen2016multiverse
on specification robustness, and @olken2015promises on pre-analysis plans; the
marketing field's own statement on experimental practice is in @nelson2020fieldexp.
The design and estimation background for the individual legs is in
@sec-causal-inference and @sec-surveys, and the reporting conventions that make a
multi-leg paper auditable are treated in @sec-reporting.
