## Industrial Organization

### Book Outline

#### Part I: Foundations of Industrial Organization

1.  Markets and Competition

    -   Perfect Competition

    -   Monopolistic Competition

    -   Oligopoly

    -   Monopoly

2.  Market Structures

    -   Homogeneous vs. Differentiated Products

    -   Entry and Exit Barriers

    -   Market Concentration

3.  Consumer Behavior

    -   Utility Theory

    -   Information Asymmetry

    -   Behavioral Economics

#### Part II: Marketing Strategies in Different Industrial Structures

1.  Price Strategies

    -   Price Discrimination

    -   Price Skimming

    -   Penetration Pricing

2.  Product Strategies

    -   Product Differentiation

    -   Branding

    -   Product Lifecycle

3.  Placement Strategies

    -   Channel Marketing

    -   Online vs Offline

    -   Supply Chain Management

4.  Promotion Strategies

    -   Advertising

    -   Sales Promotions

    -   Public Relations

#### Part III: Case Studies

1.  Fast-Moving Consumer Goods (FMCG)

2.  Technology Sector

3.  Automobile Industry

4.  Pharmaceuticals

5.  Services Industry

#### Part IV: Contemporary Issues

1.  Digital Transformation

2.  Sustainability in Industrial Marketing

3.  Globalization and Market Dynamics

# Key Questions in Industrial Organization

## Core Questions

1.  Why do firms behave the way that we observe?
2.  What are the effects of firm behavior (or business strategy) on consumers, other firms, and employees? What welfare standards should be used for evaluation?
3.  How can we design better business strategies for firms and formulate public policies for regulatory agencies?

## Methodological Approaches

-   Theoretical Approach: Modeling + Predictions
-   Empirical Approach: Data Analysis + Testing + Counterfactual & Welfare Analysis
-   Experimental and Behavioral Approaches

These approaches are complementary.

## Building an IO Model

### Starting Points

A. Ideas/Issues/Questions\
- From the Literature (IO and related fields)\
- From Other Sources (newspapers, magazines, TV news, conversations, consulting, etc.)

B. [Modeling Tools in IO]

### Modeling Tools in IO

1.  [Modeling Consumer Demands] for Products/Services

2.  Profit Maximization Subject to [Constraints]

    -   Constraints: Technology, Arbitrage, Incentives, Self-Selection, etc.

3.  [Game-Theoretic Models] of Competition

4.  [Comparative Static Analysis]

#### Modeling Consumer Demands

-   Single or Multiple Products?
    -   Classical Consumer Theory
-   Single or Multiple Units?
    -   Discrete Choice Theory
-   Are Products Differentiated?
    -   Complementary or Substitute?
    -   Vertically or Horizontally Differentiated?

Notes on Demand Complementarities

-   As the price of one product increases, the demand for that product decreases
    -   A monopolist charges lower prices than two independent firms would
    -   Complementarities can be a barrier to entry
-   Network Effects
-   Multi-Sided Markets
-   Ecosystems

#### Unconstrained or Constrained Profit Maximization

-   Choice Variables and Objective Function
-   Technology Constraints
-   Consumer Participation Constraints
    -   Bundling and Tied Selling
    -   Contract as a Barrier to Entry
    -   Arbitrage Constraints
-   Incentive Constraints (Principal-Agent Models)
    -   Second-Degree Price Discrimination
    -   Franchise Contracts

#### Game-Theoretic Models

##### A. Non-Cooperative Game-Theoretic Models

1.  Games in Normal Form (Simultaneous Moves)
    -   A Variety of Strategic Variables
    -   A Number of Players
    -   A Set of Strategies for Each Player
    -   A Payoff Function for Each Player
2.  Games in Extensive Form (Sequential Moves)
    -   First-Move Advantages
    -   Entry Deterrence
    -   Collusion through Repeated Interactions

##### B. Cooperative Game-Theoretic Models

-   Nash Bargaining
    -   Payoff Choice Set
    -   Bargaining Power
    -   Threat Point
-   Nash-in-Nash Bargaining

#### Comparative Static Analysis

-   A Model with Exogenous Parameters and Endogenous Variables
-   Behavioral Assumptions: Maximization or Equilibrium
-   How Do Optimal Solutions or Equilibrium Strategies Change with Respect to Exogenous Parameters or Policy Variables?

## Conclusions

-   Issues/Questions + Tools -\> Results/Findings
-   Time to Consider Whether the Results Make Intuitive Sense

## Presenting/Writing an IO Paper

1.  Describe how interesting and important the ideas/issues/questions are
    -   Examples, Statistical Numbers, Previous Studies, Policy Implementations, etc.
2.  Describe Your Approach and Summarize Major Findings
3.  Construct Your Model
4.  Conduct Your Analysis
    -   Determine Optimal Solutions or Equilibrium and Conduct Comparative Static Analysis
5.  Concluding Remarks
    -   Summary of the Analysis, Limitations, and Future Work

# Consumer Demands

## Questions

1.  How can we describe and formally model consumers' demands for certain products in response to firms' offers such as prices, quality, versions, advertising, warranty, and so on?

2.  What properties do demand systems (more than one product) possess?

3.  How can we estimate the demand system and compute own- and cross-price elasticities, diversion ratios?

## Approaches to Consumer Demand Systems

-   Business is a matching process between:

    -   A population of consumers with certain preferences (both observable and unobservable characteristics)

    -   A set of products and/or services (with specific observable and unobservable characteristics)

-   The microeconomics/IO literature offers three approaches:

    1.  Classical consumer theory with continuous choices

    2.  Discrete choice models with heterogeneity of consumer tastes

        -   Horizontal differentiation

        -   Vertical differentiation

        -   A combination of the above

    3.  A combination of discrete and continuous choices

## Part I: Classical Consumer Theory

-   A representative consumer with a utility function $U(q_0, q_1, \dots, q_n)$ and income $I$, facing a profile of prices $\mathbf{p} = (p_1, \dots, p_n)$, chooses a vector of continuous quantities $\mathbf{q} = (q_1, \dots, q_n)$ to maximize their utility subject to a budget constraint:

$$
\max_{q_0, \dots, q_n} U(q_0, q_1, \dots, q_n)
$$

subject to

$$
q_0 + \sum_{i=1}^n p_i q_i \le I
$$

where $q_0$ is a numeraire good with a normalized price $p_0 = 1$.

-   The solution to this constrained optimization problem yields the Walrasian (or Marshallian) demand system:

$$
q_i = D_i (\mathbf{p}, I), i = 1, 2, \dots, n
$$

Indirect utility function

$$
U(q_0^*, q^*) = V (p, I)
$$

Roy's Identity and Slutsky Equation

The Walrasian (or Marshallian) demand system satisfies the following properties:

-   Roy's Identity:

$$
-\frac{\partial V}{\partial p_i} / \frac{\partial V}{\partial I} = q_i (\mathbf{p}, I)
$$

where$V$is the maximum (indirect) utility.

-   Slutsky Equation:

$$
\frac{\partial h_i (\mathbf{p}, u)}{\partial p_j} = \frac{\partial q_i (\mathbf{p}, I)}{\partial p_j} + \frac{\partial q_i (\mathbf{p}, I)}{\partial I} q_j (\mathbf{p}, I)
$$

holds and forms the Slutsky substitution matrix, which is symmetric and negative semi-definite.

### A Special Case: CES Sub-utility and CES Demand System

-   Consider a special case:

$$
U = U(q_0, Q)
$$

For example, $U = q_0 Q$ where

$$
Q = (\sum_{i = 1}^n q_i^\rho)^{1/\rho}, \rho \le 1
$$

is a CES sub-utility function.

-   This leads to the CES demand system.

-   This functional form is handy when studying product diversity/variety.

#### CES Utility and Demand System

Consider a representative consumer with the following CES utility function over $n+1$ goods:

$$
U = q_0 [\sum_{i = 1}^n q_i^\rho]^{1/\rho}
$$

where $\rho \in (0,1)$, $q_0$ is the numeraire good.

To derive the demand systems, we will start with the utility maximization problem with the following budget constraint:

$$
P_0 q_0 + \sum_{i = 1}^n P_i q_i = M
$$

The Lagrangian for this problem is:

$$
\mathcal{L} = q_0 \left[ \sum_{i = 1}^n q_i^\rho \right]^{1/\rho} + \lambda \left( M - P_0 q_0 - \sum_{i = 1}^n P_i q_i \right)
$$

Taking the partial derivative with respect to $q_0$:

$$
\frac{\partial \mathcal{L}}{\partial q_0} = \left[ \sum_{i = 1}^n q_i^\rho \right]^{1/\rho} - P_0 \lambda
$$

From which we get:

$$
\left[ \sum_{i = 1}^n q_i^\rho \right]^{1/\rho} = P_0 \lambda
$$

For $q_i$:

$$
\frac{\partial \mathcal{L}}{\partial q_i} = q_0 \left[ \sum_{i = 1}^n q_i^\rho \right]^{1/\rho - 1} q_i^{\rho - 1} - P_i \lambda
$$

Which gives:

$$
q_0 \left[ \sum_{i = 1}^n q_i^\rho \right]^{1/\rho - 1} q_i^{\rho - 1} = P_i \lambda
$$

Combining the above expressions will give the direct demand function. Solving for $P_i$ gives the inverse demand function.

Lastly, for substitutability or complementarity of the goods: Given $\rho \in (0,1)$, the goods in the CES utility function are substitutes. This is because a value of $\rho$ between 0 and 1 indicates the elasticity of substitution between any two goods exceeds one, suggesting they are imperfect substitutes.

### Quasi-linear Utility and No Income Effect

-   Consider another special class: quasi-linear utility function

$$
U = q_0 + u(q_1 , \dots, q_n)
$$

-   The constrained maximization is equivalent to:

$$
\max_{q_1, \dots, q_n} u (q_1, \dots, q_n) - \sum_{i = 1}^n p_i q_i + I
$$

Hence,

$$
FOC: \frac{\partial u}{\partial q_i} = p_i, i = 1, 2, \dots, n
$$

-   The demands for the$n$products are independent of income $I$, as long as the income is sufficiently large.

Indirect utility function for this case is $U(q^*_1(\mathbf{p}) , \dots, q^*_n(\mathbf{p}) - \sum_{i=1}^n p_i q_i^* (\mathbf{p})$ = consumer surplus

-   The (indirect) maximum utility function is $V = v(p_1, \dots, p_n) + I$, which is useful for welfare analysis.

-   The relationship between demands and consumer surplus:

$$
- \frac{\partial v}{ \partial p_i} = q_i (\mathbf{p})
$$

which is a special case of Roy's Identity.

-   The Jacobian of the demand system:

$$
J = \left( \frac{\partial q_i (\mathbf{p})}{\partial p_j} \right) _{n \times n}
$$

is symmetric and negative semi-definite, which is a special case of the Slutsky equation.

Recover the utility function (Reverse engineer)

-   Suppose the Jacobian of the observed demand system

$$
J = \left( \frac{\partial q_i (\mathbf{p})}{\partial p_j} \right)_{n \times n}
$$

is symmetric and negative definite

-   Then the demand system can be derived from a representative consumer maximizing a (quasi-linear) concave utility function

-   Substitute the inverse demand system into the FOCs and solve a system of partial differential equations:

$$
\frac{\partial u}{\partial q_i} = p_i (q), i = 1, 2, \dots, n
$$

-   This can help with welfare analysis

Demand Substitutes vs. Complements

Consider $n = 2$

-   The two products are demand-substitutes if the demand for product $j$ increases with price $p_i$

-   The two products are demand-complements if the demand for product $j$ decreases with price $p_i$

-   Applying the Cramer's rule the first-order condition of the utility maximization leads to

$$
|J| \frac{\partial q_1}{\partial p_2} = |J| \frac{\partial q_2}{\partial p_1} = -u_{12} = -u_{21}
$$

where $|J|$ is the determinant of $J$ and is positive due to the negative definiteness of $J$ (or the strict concavity of the utility function)

-   Thus, under the quasi-linear utility, the products are

    -   demand-substitutes if $u_{12} = u_{21} <0$

    -   demand-complements if $u_{12} = u_{21} > 0$

Diversion Ratio

Consider $n = 2$ and imperfect substitutes

-   As price $p_1$ increases, how much demand reduction of product 1 is diverted to the consumption of product 2?

$$
d_{12} = \frac{\frac{\partial q_2}{\partial p_1}}{- \frac{\partial q_1}{\partial p_1}} > 0
$$

Negative definite substitution matrix (or the strict concavity of the utility function) implies that

$$
\frac{\partial q_1}{\partial p_1} \frac{\partial q_2}{\partial p_2} - \frac{\partial q_2}{\partial p_1} \frac{\partial q_1}{\partial p_2} >0
$$

or equivalently

$$
d_{12} d_{21} <1
$$

The Diversion Ratio shows how much of one product's demand shifts to another product if the first is unavailable. If you prefer coffee Brand A but switch to Brand B when A is out of stock, then sales have "diverted" from Brand A to Brand B.

Two products, $A$ and $B$, have quantities $Q_A$ and $Q_B$, and prices $P_A$ and $P_B$.

The Diversion Ratio $D_{AB}$ is:

$$
D_{AB} = \frac{\Delta Q_B / \Delta P_A}{\Delta Q_A / \Delta P_A}
$$

This measures the change in quantity of $B$ for a change in the price of $A$, normalized by the change in quantity of $A$ for the same price change.

```{r}
# Simulating data
set.seed(123)
n = 100
price_A = runif(n, 5, 10)
price_B = runif(n, 5, 10)
quantity_A = 100 - 5 * price_A + rnorm(n)
quantity_B = 80 - 4 * price_B + rnorm(n)

# Estimating Diversion Ratio
delta_QA = diff(quantity_A)
delta_PA = diff(price_A)
delta_QB = diff(quantity_B)

D_AB = mean(delta_QB / delta_PA) / mean(delta_QA / delta_PA)
D_AB
```

```{r}
library(ggplot2)
# Plotting
data = data.frame(delta_QA, delta_PA, delta_QB)
ggplot(data, aes(x = delta_PA)) +
  geom_point(aes(y = delta_QA), color = 'blue', alpha = 0.5) +
  geom_point(aes(y = delta_QB), color = 'red', alpha = 0.5) +
  geom_smooth(aes(y = delta_QA), method = 'lm', color = 'blue', se = FALSE) +
  geom_smooth(aes(y = delta_QB), method = 'lm', color = 'red', se = FALSE) +
  labs(x = "Change in Price of A",
       y = "Change in Quantity",
       title = "Change in Quantity vs Change in Price",
       subtitle = "Blue: Product A, Red: Product B") +
    causalverse::ama_theme()
```

### Representative Consumer Utility Maximization with Quasi-linear Quadratic Utility (and Linear Demand System)

Consider a representative consumer with the following quasi-linear utility function over three goods:

$$
U(q_0, q_1, q_2) = q_0 + \alpha_1 q_1 + \alpha_2 q_2 - [\beta_1 q_1^2 + 2 \gamma q_1 q_2 + \beta_2 q_2^2]/2
$$

with $\alpha_i >0, \beta_i >0, \beta_1 \beta_2 - \gamma^2 >0$ and $q_0$ is the numeraire good

Then the first-order conditions becomes

$$
p_1 = \alpha_1 - \beta_1 q_1 - \gamma q_2
$$

$$
p_2 = \alpha_2 - \beta_2 q_2 - \gamma q_1
$$

Solving the system of linear equations leads to a regular linear demand system and constant diversion ratios

#### Shubik and Levitan (1980) Form

Suppose $u(q_1, \dots, q_n)$ is a quadratic function as follow:

$$
u(q_1, \dots, q_n) = \alpha \sum_{i=1}^n q_i - \frac{n}{2(1 + \gamma)}[\sum_{i=1}^n q_i^2 + \frac{\gamma}{n} (\sum_{i = 1}^n q_i)^2]
$$

with $\alpha >0$ and $\gamma \in [0, \infty)$ is the degree of sustainability

This leads to the following linear demand system:

$$
q_i = \frac{1}{n}[ \alpha - p_i (1 + \gamma) + \frac{\gamma}{n} \sum_{j = 1}^n p_j]
$$

Two nice properties of this demand system:

1.  The aggregate demand, $Q = \sum_{i = 1}^nq_i = \alpha- \frac{1}{n}\sum_{j = 1}^n p_j$ does not depend on the degree of substitution among the products
2.  When the prices are symmetric ($p_i = p$), the aggregate demand, $Q = \alpha - p$ does not depend on the number of the products.

##### Application 1: Maximizing Buyer's Surplus subject to a Market-share Constraint

-   Consider a downstream buyer with a revenue function over two products $u(q_1, q_2)$ facing a pair of per-unit prices, $p_1, p_2$

-   However, the seller of product 1 sets a market-share based restriction: $q_1 / (q_1+ q_2) \ge s$ or $q_1 = 0$, where $s \in [0,1]$ is offered by firm 1.

-   A special case: $s = 1$ means an exclusive contract.

-   The buyer solves the following constrained surplus maximization:

$$
\max_{q_1, q_2} u(q_1, q_2) - p_1 q_1 - p_2 q_2
$$

such that $q_1 / (q_1 + q_2) \ge s$ or $q_1 = 0$

What properties does this demand system $q(p,s)$ have with respect to $p$ and $s$

Use the linear quadratic utility to solve this demand system.

##### Application 2: Strong Complements

Consider a quasi linear utility function $q_0 + u(q_1, \dots, q_n)$ and the demand system $q^*(p) = (q_1^* (p), \dots, q_n^* (p))$ for the $n$ products

The following two statements are equivalent:

1.  The demand system $q^*(p)$ satisfies the strong complement property: for $i = 2, \dots, n, q_i^*(p) /q_1^*(p)$ is independent of $p_1$
2.  For any constant marginal cost $c_i >0$ for $i = 1, 2, \dots, n$ the solution to a monopoly profit-maximization problem

$$
\max_p \sum_{i=1}^n [(p_i - c_i) q_i^*(p)]
$$

involves $p_i^* = c_i$ for $i = 2, \dots, n$

-   Let $v(p) = u (q^*(p)) - \sum_{i = 1}^n p_i q_i^*(p)$ be the indirect utility. The max profit conditional on offering the consumer at least $u$

$$
\pi(u) = \max_p \{ \sum_{i = 1}^n (p_i - c_i)q_i^*(p) 
$$

such that $v(p) \ge u$

is decreasing and concave

#### Consumers' Demands

To derive the consumer's demands for goods 1 and 2.

Differentiate the utility function with respect to $q_1$ and $q_2$:

1)  With respect to $q_1$:

$$
\frac{\partial U}{\partial q_1} = \alpha_1 - \beta_1 q_1 - \gamma q_2 = 0
$$

2)  With respect to $q_2$:

$$
\frac{\partial U}{\partial q_2} = \alpha_2 - \gamma q_1 - \beta_2 q_2 = 0
$$

From these equations, we can express $q_1$ and $q_2$ in terms of the parameters:

1)  $q_1$ in terms of $\alpha_1, \beta_1, \gamma, q_2$:

$$
q_1 = \frac{\alpha_1 - \gamma q_2}{\beta_1}
$$

2)  $q_2$ in terms of $\alpha_2, \beta_2, \gamma, q_1$:

$$
q_2 = \frac{\alpha_2 - \gamma q_1}{\beta_2}
$$

1)  Inverse demand for $q_1$:

$$
p_1 = \alpha_1 - \beta_1 q_1 - \gamma q_2
$$

2)  Inverse demand for $q_2$:

$$
p_2 = \alpha_2 - \gamma q_1 - \beta_2 q_2
$$

These equations represent how the prices of goods 1 and 2 (i.e., $p_1$ and $p_2$) change with the quantities consumed.

#### Degree of Substitutability/Complementarity

When considering how to parameterize the demand system to capture the degree of substitution or complementarity between goods $q_1$ and $q_2$, it's important to remember that:

-   A positive cross-partial derivative $\frac{\partial^2 U}{\partial q_1 \partial q_2}$ indicates that the two goods are complements.
-   A negative cross-partial derivative indicates that the two goods are substitutes.

Given the utility function, the cross-partial derivative is:

$$
\frac{\partial^2 U}{\partial q_1 \partial q_2} = \gamma 
$$

So the value and sign of $\gamma$ directly tells us about the substitution/complementarity relationship between $q_1$ and $q_2$:

-   Using $\gamma$ as an independent parameter: This approach directly captures the nature of the relationship between $q_1$ and $q_2$. A positive $\gamma$ would indicate complementarity, while a negative $\gamma$ would indicate substitutability.
-   Setting $\beta_1 = \beta_2 = 1 - \gamma$: This approach indirectly captures the relationship by tying the curvature of the utility function with respect to $q_1$ and $q_2$ to $\gamma$. If $\gamma$ increases while $\beta_1$ and $\beta_2$ decrease, the nature of the relationship between $q_1$ and $q_2$ can change. However, this choice might seem arbitrary and can limit the flexibility of the model, as it imposes restrictions on the curvature of the utility function in relation to $\gamma$.

#### Consumer's surplus

To derive the consumer's surplus for $q_1$ and $q_2$:

1.  Consumer's Surplus for $q_1$:

From the condition where marginal utility equals price, we get:

$$
\frac{\partial U}{\partial q_1} = p_1
$$

$$
\alpha_1 - \beta_1 q_1 - \gamma q_2 = p_1
$$

This gives the inverse demand function $p_1 = p_1(q_1, q_2)$.

The consumer's surplus for $q_1$ is:

$$
CS_1 = \int_0^{q_1} p_1(q_1', q_2) \, dq_1' - p_1 q_1
$$

2.  Consumer's Surplus for $q_2$:

Similarly, for $q_2$:

$$
\frac{\partial U}{\partial q_2} = p_2
$$

$$
\alpha_2 - \gamma q_1 - \beta_2 q_2 = p_2
$$

This gives the inverse demand function $p_2 = p_2(q_1, q_2)$.

The consumer's surplus for $q_2$ is:

$$
CS_2 = \int_0^{q_2} p_2(q_1, q_2') \, dq_2' - p_2 q_2
$$

The total consumer's surplus is:

$$
CS = CS_1 + CS_2
$$

#### Bertrand-Nash Equilibrium

Two firms produce 2 goods at constant marginal cost $c_1$ and $c_2$. We want to compute the Bertrand-Nash equilibrium when the firms compete by setting prices simultaneously. Additionally, we aim to determine their own and cross cost-pass-through (CPT) rates. Firm's Profit Maximization Problem

Firm 1's profit function is:

$$ \pi_1(p_1, p_2) = (p_1 - c_1) q_1(p_1, p_2) $$

The first order condition for firm 1's profit maximization is:

$$ q_1 + (p_1 - c_1) \frac{\partial q_1}{\partial p_1} = 0 $$

Similarly, for firm 2:

$$ q_2 + (p_2 - c_2) \frac{\partial q_2}{\partial p_2} = 0 $$

These equations will yield the Bertrand-Nash equilibrium prices $p_1^*$ and $p_2^*$.

Cost-Pass-Through (CPT) Rates

The own and cross CPT rates are:

$$ CPT_{11} = \frac{\partial p_1^*}{\partial c_1} $$ $$ CPT_{22} = \frac{\partial p_2^*}{\partial c_2} $$ $$ CPT_{12} = \frac{\partial p_1^*}{\partial c_2} $$ $$ CPT_{21} = \frac{\partial p_2^*}{\partial c_1} $$

#### Monopolist

When the two firms merge or collude to set prices, the firm becomes a monopolist over both goods. The monopolist's objective is to maximize the joint profits across both products, accounting for the cross-price effects.

The monopolist's profit function is given by: $$ \pi(p_1, p_2) = (p_1 - c_1) q_1(p_1, p_2) + (p_2 - c_2) q_2(p_1, p_2) $$

The first order conditions for maximizing this profit function are:

1\. $$ \frac{\partial \pi}{\partial p_1} = q_1 + (p_1 - c_1) \frac{\partial q_1}{\partial p_1} + (p_2 - c_2) \frac{\partial q_2}{\partial p_1} = 0 $$ 2. $$ \frac{\partial \pi}{\partial p_2} = q_2 + (p_1 - c_1) \frac{\partial q_1}{\partial p_2} + (p_2 - c_2) \frac{\partial q_2}{\partial p_2} = 0 $$

From these first order conditions, one can derive the joint-profit-maximizing prices, $p_1^M$ and $p_2^M$.

Cost-Pass-Through (CPT) Rates for the Monopolist

The CPT rates under monopoly are defined as:

$CPT_{11}^M = \frac{\partial p_1^M}{\partial c_1}$

$CPT_{22}^M = \frac{\partial p_2^M}{\partial c_2}$

$CPT_{12}^M = \frac{\partial p_1^M}{\partial c_2}$

$CPT_{21}^M = \frac{\partial p_2^M}{\partial c_1}$

Explanation

-   Monopoly vs. Competition: Typically, prices under monopoly (or collusion) are higher than under Bertrand competition due to the absence of competitive pressures.

-   CPT Rates: Under monopoly, the CPT rates may differ from the competitive scenario because the monopolist internalizes the cross-effects between the two goods.

-   Strategic Complements and Substitutes: Depending on the nature of the two products, the monopolist may treat them as strategic complements or substitutes when setting prices.

### Properties of Demand

1.  Single product $q= q(p), q = D(p)$
    1.  Downward-sloping: $q'(p) <0$
    2.  Price Elasticity: $\epsilon (p) = - q'(p) \frac{p}{q}$
2.  Multiple products

Given the linear demand system

$$
\begin{aligned}
q_1 &= q_1(p_1, p_2) =  \alpha_1 - \beta_{11} p_1 + \gamma_{12} p_2 + \dots \\
q_2 &= q_2(p_1, p_2) = \alpha_2 - \beta_{22} p_2 + \gamma_{21} p_1 + \dots
\end{aligned}
$$

$$
\begin{aligned}
q_1 &= q_1 (p_1, p_2) \\
q_2 &= q_2 (p_1, p_2)
\end{aligned}
$$

Own-price effect (similar to the single product case)

$$
\frac{\partial q_1}{\partial p_1} <0 
\frac{\partial q_2}{\partial q_2} <0
$$

Cross-price effects

$$
\begin{aligned}
\frac{\partial q_1}{\partial p_2}  \\
\frac{\partial q_2}{\partial q_1} 
\end{aligned}
$$

and

$$
\frac{\partial q_1}{\partial p_2}  = \frac{\partial q_2}{\partial q_1} 
$$

when

-   2 products are perfect substitute

-   2 products are perfect complement

-   2 products are independent

-   consumers have quasi-linear utility (and no income effect)

And the cross-price effects are positive if substitutes and negative if complements

Suppose we have quasi-linear utility maximization

$$
\max_{q_1, q_2} u(q_1, q_2) = p_1 q_1 - p_2 q_2
$$

FOCs

$$
u_1 = p_1 
u_2 = p2
$$

$$
\frac{q_1^*}{q_2^*}
$$

$$
\begin{aligned}
u_1 (q_1^*, q_2^*) = p_1 \\
u_2 (q_1^*, q_2^*) = p_2
\end{aligned}
$$

Differentiate with respect to $p_1$

$$
u_{11} \frac{\partial q_1^*}{\partial p_1} + u_{12} \frac{\partial q_2^*}{p_1} = 1 
u_{21} \frac{\partial q_1^*}{\partial p_1} + u_{22} \frac{\partial q_2^*}{\partial p_1} = 0
$$

To get

$$
\frac{\partial q_2^*}{\partial p_1} = - \frac{u_{21}}{u_{11} u_{22} - u_{12} u_{21} }
$$

Similarly,

$$
\frac{\partial q_1^*}{\partial p_2} = - \frac{u_{12}}{u_{11} u_{22} - u_{12} u_{21} }
$$

Whenever you have cross partial, you have $u_{12} = u_{21}$, then

$$
\frac{\partial q_1^*}{\partial p_2}  = \frac{\partial q_2^*}{\partial q_1} 
$$

Hence, whenever you assume that consumers have quasi-linear utility maximization function, the above equation will always hold (i.e.., cross-effects are equal).

From the Slutsky's equation, anytime we don't have the income effect, the cross-effects are equal.

However, when we have the income effects, we don't' know the size of $\frac{\partial q_i}{\partial I} q_j$ and other way around for product $j$. If they were to be equal then we have

$$
\frac{\partial h_i}{\partial p_j} = \frac{q_i}{p_j} + \frac{\partial p_i}{\partial I} q_j \equiv \frac{\partial h_j}{\partial p_i} = \frac{q_j}{p_i} + \frac{\partial p_j}{\partial I} q_i
$$

If this also holds,

$$
\frac{q_i}{\partial I} /q_i = \frac{q_j}{\partial I} /q_j
$$ then we have equal cross-price effects

In sum, if from our data, we have evidence for equal cross-price effects, we can say we have consistent evidence for either

-   the no income effect.

-   When income increases, then the consumption of two goods are the increasing similarly. As income increases (when you are richer), the additional products that you consume for the two products are similar.

Compare the cross-price effects with the own-price effects in 2 ways:

1.  Assume $- \frac{\partial q_i}{\partial p_i} > \frac{\partial q_j}{\partial p_i}$ so that the diversion ratio is less than 1
2.  Assume $- \frac{\partial q_i}{\partial p_i} > \frac{\partial q_i}{\partial p_j}$ so that the indirect diversion ratio is less than 1

And the own-price elasticities and cross-price elasticities are similarly defined

Remarks on the continuous approach

-   The classical approach has many advantages

    -   Nice properties of the demand system allow to derive intersecting pricing strategies.

    -   However, the approach does not pay much attention to the characteristics of products: quality, location, time, availability, consumers' info about its existence and quality, etc.

    -   Moreover, this approach does not formally model heterogeneity of consumer preferences

-   Discrete choice models are deigned to address some of the issues

#### Application 3: Properties of the Buyer's surplus Function

-   based on @armstrong2018multiproduct

-   Consumer optimization

    -   Representative consumer has same quasi-linear utility function: $u(q)$ where $u(0) = 0$, increasing and strictly concave

    -   Consumer optimization: $\max_q u(q) - p \times x$

    -   inverse demand function: $p_i(q) = \partial u(x) / \partial q_i$ or $p(q) \equiv \nabla u(q)$

    -   Total revenue from quantity $q$: $R(q) = q \times \nabla u(q)$

-   Consumer surplus from quantity $q$

$$
s(q) = u(q) - q \times \nabla u (q)
$$

### A Family of Utility and Homoerotic Consumer Surplus

Proposition [@armstrong2018multiproduct]

Consumer surplus $s(q)$ is homothetic in $q$ if and only if

$$
u(q) = h(q) + g(f(q))
$$

where $h$ and $f$ are homogeneous degree 1, and $g$ is concave with $g(0) = 0$

-   This is a much wider class than those where consumer surplus is homothetic in prices, which can be expressed simply as the homothetic $u(q)$ where $h = 0$

-   3 degrees of freedom in the family of tractable multi-product demand systems: $f(q), h(q), g(.)$

-   The advantages of homothetic $s$ and homogeneous $f, h$

Proof

-   "If" ($\leftarrow$): From consumer optimization, we have inverse demand

$$
p(q) = \nabla h(q) + g' (f(q)) \nabla f(q)
$$

since $q \times \nabla h(q) \equiv h(q)$ for a homogeneous degree 1 function

$$
R(q) = h(q) + g' (f(q))f(q)
$$

and hence consumer surplus is

$$
s(q) = g(f(q)) - g' (f(q)) f(q)
$$

and it is a monotonic function of the homogeneous function $f(q)$

Variable Decomposition Technique

Write quantities in "polar coordinates" form

$$
q = f(q) . \frac{q}{f(q)}
$$

-   $\frac{q}{f(q)}$ is homogeneous degree 0 and depends only on the ray from origin.

-   $f(q)$ measures how far along that ray $q$ lies

-   refer to $f(q)$ as "composite quantity" and $\frac{q}{f(q)}$ as "relative quantities"

Consumer Choice: A Two-stage Optimization

-   Consumer surplus with prices $p$ and quantities $q$ is

$$
u(q) - pq = g(f(q)) - f(q) \frac{pq - h(q)}{f(q)}
$$

-   in terms of two coordinates $f(q)$ and $\frac{q}{f(q)}$

-   for any given $f(q)$ surplus is maximized by minimizing $\frac{pq - h(q)}{f(q)}$

Let

$$
\phi(p) \equiv \min_{q \ge 0}: \frac{pq - h(q)}{f(q)}
$$

which is increasing and concave in $p$. Then, by the envelope theorem, the optimal relative quantities $q^r(p) \equiv \nabla \phi(p)$

The consumer facing price $p$ chooses relative quantities $q^r(p)$

given the relative quantities, $q^r(p)$, the consumer then chooses optimal composite quantity, $Q$, that maximizes

$$
g(Q) - Q \phi (p)
$$

which is concave in $Q$

$\hat{Q}(\phi(p))$ is the demand for composite quantity given $p$

So consumer demand function is

$$
q(p) = \hat{Q} (\phi (p)) \times q^r(p)
$$

$\phi(p)$ is the "composite price"

-   all prices with the same $\phi(p)$ induce the same composite quantity

-   inverse demand for composite quantity is $\phi(p) = g'(\hat{Q})$

Example 1: Linear Demand System

$$
u(q) = aq - \frac{1}{2} q^T B q
$$

$$
p(q) = a - Bq
$$

where $a >0$ and $B$ is a positive definite matrix

The utility unction can be decomposed as follows:

$$
h(q) = aq
$$

$$
f(q) = \sqrt{q^T B q}
$$

$$
g(f) = - \frac{1}{2} f^2
$$

$$
s(q) = \frac{f(q)^2}{2}
$$

Example 2: Logit Demand System

$$
q_i(p) = \frac{e^{(a_i - p_i)/\mu}}{k + \sum_j e^{(a_i - p_i)/\mu}}
$$

which corresponds to the consumer surplus function

$$
v(p) = \mu \log ( 1 + \frac{1}{k} \sum_j e^{(a_i - p_i)/\mu})
$$

Utility function representation:

$$
u(q) = aq + (\mu \log k)q + \mu \sum_i q_i \log \frac{f(q)}{q_i} + g(f(q))
$$

where

-   $h(q) = aq + (\mu \log k)q + \mu \sum_i q_i \log \frac{f(q)}{q_i}$

-   $f(q) \equiv \sum_j q_j$ (total quantity)

-   $g(f) = - \mu(f \log f + (1 - f) \log (1-f))$

-   $s(q) = - \mu\log(1 - f(q))$

Example 3: Complementary Products

-   Some consumption requires prior purchase of a base product ("access" or hardware)

-   Suppose the consumer enjoys utility $\hat{u}(y)$ from complementary services $y$

-   If $q_1$ is the number of accesses or hardware, and each requires complementary services, $y = q_2 /q_1$, where $q_2$ is the total supply of complementary services, then gross utility has the form

$$
\begin{aligned}
u(q_1, q_2) &= q_1 \hat{u}(y) + g(q_1) \\
&= q_1 \hat{u}(q_2 /q_1) + g(q_1)
\end{aligned}
$$

This is an example with $f(q) = q_1$ so consumer surplus is simply a function of the number of accesses or hardware.

Note that the ratio of the two demands, $q_2/q_1$ is indepdent of price $p_1$. This would have an important implication for monopoly profit maximization.

**Remarks on the Continuous Approach**

-   The classification approach has many advantages.

    -   Nice proprieties of the demand system allow to derive interesting pricing strategies

    -   But the approach does not pay much attention to the characteristic of products: quality, location time, availability, consumers' information about its existence and quality, etc.

    -   It does not formally model heterogeneity of consumer preferences

-   Discrete choice models are designed to address some of the issues.

## Part II: Vertical Differentiation: Quality

Differentiation is said to be **vertical**, if all consumers benefit when the level of the product's characteristics is augmented in the product space.

Consumers agree over the most preferred mix of characteristics such as quality

Example:

$$
u = 
\begin{cases}
\theta s - p & \text{if buys a good with quality s at price p}\\
0 & \text{does not buy}
\end{cases}
$$

1.  $u$ is the surplus derived from the consumption of the good
2.  $s > 0$ describes the quality of the good, which is common knowledge
3.  Utility is separable in quality and price
4.  Heterogeneity of consumers' tastes: $\theta >0$ is a taste parameter which may differ across consumers (unknown to firms). Firms only know that $\theta$ is distributed according to ta density function $f(\theta)$ and cumulative distribution $F(\theta), \theta \in [\underline{\theta}, \bar{\theta}]$

The key feature is that all consumers prefer higher quality to lower quality

Demand

1.  One Quality
    1.  When only one quality $s$ is offered at price $p$, the consumer with type $\theta$ purchases if and only if $\theta s - p \ge 0$ (i.e., $\theta \ge p/s$)

    2.  Let $N$ be the total number of identical consumers. The demand is\

        $$ D(p) = N [1 - F(\frac{p}{s})] $$

        which is downward sloping

    3.  In the case of uniform distribution,\
        $$D(p) = \begin{cases} N[1 - p/s] , \text{if } p \le s\\ 0, \text{if } p > s \end{cases}$$

    4.  Unlike the classical consumer theory, product quality is now explicitly incorpoated into the demand
2.  Two Qualities
    1.  Product line selection may help to extract more consumer surplus
    2.  Supposes two qualities are offered: $s_1 < s_2$. Without loss of generality assume $p_1 < p_2$
    3.  A consumer of type $\theta$ buys $s_1$ if and only if\
        $$\begin{cases} \theta s_1 - p_1 \ge 0 \\ \theta s_1 - p_1 \ge \theta s_2 - p_2 \end{cases}$$\
        buys $s_2$ if and only if\
        $$\begin{cases} \theta s_2 - p_ \ge 0 \\ \theta s_2 - p_2 \ge  \theta s_1 - p_1 \end{cases}$$
    4.  We can find out three critical points $p_1 / s_1, p_2 / s_2 , (p_2 - p_1)/(s_2 - s_1)$, which partition consumer types among 3 options.
    5.  There are two cases to be considered
        1.  Quality per dollar is higher for $s_2$, i.e.,\
            $$\frac{p_2}{s_2} \le \frac{p_1}{s_1}$$\
            $$\frac{s_2}{p_2} \ge \frac{s_1}{p_1}$$\
            $$\theta s_2 \ge \theta s_1 - p_1 \to \theta \ge \frac{p_2 - p_1}{s_2 - s_1} \equiv \theta^*$$\
            $$\begin{cases} D_2 (p_1, p_2) = N[1 - F(\frac{p_2}{s_2})]\\ D_1 (p_1, p_2) = 0 \end{cases}$$
        2.  Case 2:\
            $$\frac{p_2}{s_2} < \frac{p_1}{s_1}$$\
            $$\frac{s_2}{p_2} > \frac{s_1}{p_1}$$\
            $$\theta^* \equiv \frac{p_2 - p_1}{s_2 - s_1} >\frac{p_2}{s_2}$$\
            $$\begin{cases} D_2 (p_1, p_2) = N[F(\frac{p_2 - p_1}{s_2 - s_1}) - F(\frac{p_1}{s_1})]\\ D_1 (p_1, p_2) = N[1 - F(\frac{p_2-p_1}{s_2 - s_1})] \end{cases}$$
        3.  Put the above two cases together:

$$
D_1 (p_1, p_2) =
\begin{cases}
0 & \text{if } \frac{p_1}{p_2} \ge \frac{s_1}{s_2} \\
N[F(\frac{p_2 - p_1}{s_2 - s_1}) - F(\frac{p_1}{s_1})] & \text{if } \frac{p_1}{p_2} < \frac{s_1}{s_2}
\end{cases}
$$

$$
D_2 (p_1, p_2) = 
\begin{cases}
N[1 - F(\frac{p_2}{s_2})] & \text{if } \frac{p_1}{p_2} \ge \frac{s_1}{s_2} \\
N[1 - F(\frac{p_2 - p_1}{s_2 - s_1})] & \text{if } \frac{p_1}{p_2} < \frac{s_1}{s_2}
\end{cases}
$$

Example

$F(\theta) = \theta, \theta \in [0,1]$

$$ D_1 (p_1, p_2) = \begin{cases} 0 & \text{if } \frac{p_1}{p_2} \ge \frac{s_1}{s_2} \\ N[(\frac{p_2 - p_1}{s_2 - s_1}) - (\frac{p_1}{s_1})] & \text{if } \frac{p_1}{p_2} < \frac{s_1}{s_2} \end{cases} $$

$$ D_2 (p_1, p_2) =  \begin{cases} N[1 - (\frac{p_2}{s_2})] & \text{if } \frac{p_1}{p_2} \ge \frac{s_1}{s_2} \\ N[1 - (\frac{p_2 - p_1}{s_2 - s_1})] & \text{if } \frac{p_1}{p_2} < \frac{s_1}{s_2} \end{cases} $$

$$
\frac{\partial D_1}{\partial p_2} = 
\begin{cases}
0 \\
\frac{N}{s_2 - s_1} >0
\end{cases}
$$

$$
\frac{\partial D_2}{\partial p_1} =
\begin{cases}
0 \\ 
\frac{N}{s_2 - s_1} >0
\end{cases}
$$

They are substitutes. It depends on quality difference

A remarks on consumer heterogeneity

-   $\theta$ can be interpreted as consumer's willingness to pay for quality

-   Consumer's type $\theta$ can be related to income, which is a measurable characteristic to describe heterogeneity of consumers

-   Consumers differ by income, e.g.,\
    $U = u (I - p ) +s$\
    By Taylor's approximation

## Review

### Estimating Discrete-Choice Models of Product Differentiation @berry1994estimating

-   **Unobserved Demand Factors**:

    -   Structural interpretation suggests these represent unobserved demand influences.

    -   Correlation between price and these demand errors can lead to anomalies like upward sloping demand curves.

-   **Differentiated Products Markets**:

    -   Issues arise when product characteristics are unobserved.

    -   Characteristics like style, which are hard to quantify, influence demand.

    -   Products may be similar but differ in consumer perception of quality, durability, etc.

    -   Often, more characteristics are crucial to consumers than what the econometrician can observe.

-   **Endogeneity of Prices**:

    -   Not merely an econometric concern.

    -   Ignoring the correlation between price and unobserved product features can lead to misleading conclusions.

    -   Trajtenberg's study of CT scanners reveals anomalies in price's effect on demand due to unobserved product quality.

    -   Research on the automobile industry (Berry, Levinsohn, and Pakes) supports the significance of considering unobserved product characteristics.

-   **Estimation Methods**:

    -   For homogeneous goods, demand parameters can be estimated with traditional instrumental variables methods despite unobserved demand factors.

    -   In discrete-choice models, prices and unobserved product features influence demand non-linearly, complicating instrumental variables' application.

-   **Mean Utility Method**:

    -   Introduced as a solution to the nonlinear instrumental variables issue.

    -   Inverts the market shares function to find the mean utility levels defined by model primitives.

    -   This method relates mean utility levels to product features and prices through instrumental variables techniques.

    -   Applicable to various discrete-choice models, does not require a unique equilibrium, and is often less computationally intensive than other methods.

<!-- -->

-   **Simple Approach to Endogenous Prices**:

    -   Aggregate demand curves in differentiated products industry indicate quantity demanded:

        -   Decreases with a firm's own price.

        -   Increases with the price of rivals.

    -   Example provided is the constant elasticity framework:

        $\ln(q_j) = \alpha_j + \sum_k \eta_{jk} \ln(p_k) + \epsilon_j$

        -   where $\eta_{jk}$ is the elasticity of good $j$ with respect to the price of good $k$

    -   If $\epsilon$ and $p$ are correlated, the demand system can be estimated using traditional variable techniques.**Empirical Study Reference**:

        -   Bresnahan (1987) employs a discrete-choice model focusing on vertically differentiated products in the automobile market.

        -   Consumers value product quality based on observed product characteristics.

    -   **Features in Bresnahan's Model**:

        -   Consumer utility is modeled based on product characteristics.

        -   Consumers can opt for an "outside good".

        -   The model makes explicit use of first-order conditions of price-setting firms.

    -   **Shared Defect with Bresnahan's Model**:

        -   Like most empirical studies of differentiated products, product characteristics are treated as exogenous.

        -   However, product prices are determined within the model.

    -   Challenge: A system of $N$ goods results in $N^2$ elasticities, making estimation challenging for many real-world applications.

-   **Avoiding "Too Many Elasticities" Problem**:

    -   Place a priori restrictions on cross-price elasticities.

    -   Some might assume certain cross-price elasticities equal zero or some sets equal each other.

    -   This approach can be arbitrary, with economic theory often providing limited guidance.

-   **Structuring the Demand Problem**:

    -   To reduce the number of demand parameters, assumptions are made on consumer utility:

        -   Depends on product characteristics.

        -   Influenced by random consumer "tastes".

        -   A small set of parameters for estimation.

    -   Market demand results from aggregating individual consumer choices.

    -   Benefits:

        -   Derives all relevant demand elasticities from fewer utility parameters.

        -   Predicts demand for new and dissimilar products in different markets.

        -   Allows transitioning between aggregate demand and consumer utility statements.

-   **Discrete-choice Models**:

    -   Provide a method to structure demand effectively.

    -   Limitations:

        -   Rules out purchasing multiple items.

        -   Doesn't easily account for dynamic aspects of demand.

        -   Often places significant parametric restrictions on demand structure.

    -   Have a long history in econometrics, notably influenced by McFadden (1974).

    -   Increasing attention in theoretical literature on differentiated products oligopoly.

-   **Empirical Study Reference**:

    -   Bresnahan (1987) employs a discrete-choice model focusing on vertically differentiated products in the automobile market.

    -   Consumers value product quality based on observed product characteristics.

-   **Features in Bresnahan's Model that are also used in the paper**:

    -   Consumer utility is modeled based on product characteristics.

    -   Consumers can opt for an "outside good".

    -   The model makes explicit use of first-order conditions of price-setting firms.

-   **Shared Defect with Bresnahan's Model**:

    -   Product characteristics are treated as exogenous.

    -   However, product prices are determined within the model.

-   **Primitives of the Model**:

    -   Characteristics of products.

    -   Consumer preferences.

    -   Equilibrium notion.

-   **Assumptions for Participants**:

    -   All product characteristics and decisions are observable to all market participants.

-   **Limitations for the Econometrician**:

    -   Not all product characteristics are observable.

    -   Decisions of individual consumers may be unobserved.

-   **Econometrician's Observations**:

    -   Market outcomes including:

        -   Price.

        -   Quantities sold by each firm.

A large number $R$ independent markets.

$N_r$ firms in market $r$, with each firm producing 1 product

For product $j$ in market $r$, $z_{jr} \in R^{K_z}$ are **observed** characteristics, where this vector contains elements that can affect demand $x_j$ and marginal cost $w_j$

The characteristic of all firms in the market are $\mathbf{z} = (z_i, \dots, z_N)$

$(\xi_j, \omega_j)$ are **unobserved** characteristics of product $j$

-   $\xi_j$ is unobserved demand characteristic

-   $\omega_j$ is unobserved cost variable.

-   They are assumed to be mean independent of observed characteristics and independent across market (i.e., $\mathbf{z}, \mathbf{\xi}, \mathbf{\omega}$ are exogenous to the firm's pricing decisions)

The discrete choice model: The utility of consumer $i$ for product $j$ depends on the characteristics of the product ad the consumer $U(x_j, \xi_j, p_j, \nu_i, \theta_d)$ where

-   $x_j,\xi_j, p_j, \theta_d$ are observed product characteristics, unobserved product characteristics, price, and demand parameters

-   $\nu_i$ is consumer specific terms.

-   **Estimator Requirements**:

    -   Need for parametric assumptions on consumer-specific variables.

    -   These assumptions are similar to choosing a functional form for a homogeneous goods demand equation.

-   **Impact of Choices**:

    -   The selection of a utility function and the density of $\nu$ greatly influence the final model.

-   **Utility Specification**:

    -   The article focuses on a simple random coefficients specification for utility.

    -   Utility for consumer $i$ and product $j$ is defined by this specification:

$$
u_{ij} = x_j \tilde{\beta} - \alpha p_j + \xi_j + \epsilon_{ij}
$$

where

-   the consumer-specific taste parameters are $\tilde{\beta}, \epsilon_{ij}$

-   $\alpha$ is invariant across consumer.

-   $\xi_j$ is the mean of consumers' valuations of an unobserved product characteristic (e.g., product quality)

-   $\epsilon_{ij}$ is the distribution of consumer preferences about this mean

Consumer $i$'s taste parameter for characteristic $k$ is

$$
\tilde{\beta} = \beta_k + \sigma_k \zeta_{ik}
$$

where $\beta_k$ is the mean level of the taste parameter for product $k$ and and the mean-zero $\zeta_{ik}$ has i.i.d standard normal distribution across individuals and characteristics. Then

$$
u_{ij} = x_j \beta + \xi_j - \alpha p_j + \nu_{ij}
$$

with

$$
\nu_{ij} = \sum_k x_{jk} \sigma_k \zeta_{ik} + \epsilon_{ij}
$$

where $\nu_{ij}$ is mean-zero, heteroskedastic error that is the effects of the random taste parameters.

The mean utility level of product $j$ is

$$
\delta_j \equiv x_j \beta - \alpha p_j + \xi_j
$$

### Economic choices @mcfadden2001economic

[@cosguner2022dynamic] Dynamic Pricing for New Products Using a Utility-Based Generalization of the Bass Diffusion Model

-   **Objective**: Address the limitations of the Generalized Bass Model (GBM) in handling marketing variables for optimal dynamic pricing and advertising.

-   **Bass Model (BM)**:

    -   Strength: Renowned for new product sales forecasting.

    -   Shortcoming: GBM, an extension of BM, is limited in its prescriptive use. It only accounts for percentage changes in marketing variables, not their actual values. This limitation makes it hard to determine the ideal launch price for a product.

-   **Proposed Models**:

    1.  **Bass-Gumbel Diffusion Model (BGDM)**

    2.  **Bass-Logit Diffusion Model (BLDM)**

        -   Previously recognized and utilized.

        -   Utility-based extensions that offer prescriptions for both launch price and post-launch price trajectory.

-   **Performance**:

    -   Using empirical data across four product categories, both BGDM and BLDM notably surpass GBM in predicting new product sales.

-   **Estimation in Data Absence**: Discuss methodologies to estimate BGDM and BLDM when lacking prior sales data.

-   **Optimal Pricing Policy**:

    -   Contrast the BLDM's pricing policy with that of GBM.

    -   Extract optimal pricing strategies as suggested by BLDM across various model parameter ranges.

-   **Dynamic Pricing Approach**:

    -   Provides a computational-friendly method for managers to determine the best marketing strategies.

    -   Extended to cater to competitive, multiproduct scenarios.

[@cosguner2017behavioral] Behavioral Price Discrimination in Distribution Channels

-   **Objective**: Examine the strategic consequences of behavioral price discrimination (BPD) on manufacturers and retailers considering consumer switching costs.

-   **Differentiating Factor**:

    -   Unlike prior studies that consider only price elasticity differences among customers, this study also accounts for price elasticity variations over time due to switching costs.

-   **Data & Method**:

    -   Dynamic pricing model estimation using data from the cola category.

-   **Key Findings**:

    1.  **Retailer's Best Strategy**:

        -   It's more profitable for retailers to delegate data analytics and coupon customization to manufacturers rather than actively managing coupons themselves.

    2.  **Retailer as an Information Broker**:

        -   Retailers can profit significantly by selling their customer databases to manufacturers.

    3.  **Manufacturers' Disadvantage**:

        -   Despite having access to data, manufacturers face reduced profits, indicating the retailer's heightened channel power due to customer information.

    4.  **Simplicity in BPD**:

        -   Using just the most recent purchase data of customers can greatly affect firms' profitability. This method of BPD is easy and cost-effective for both manufacturers and retailers.

-   **Implications**:

    -   Customer information, especially recent purchase behavior, plays a pivotal role in channel dynamics.

    -   Retailers can harness this information not only for direct profit but also as a tool for channel leverage against manufacturers.

    -   Simplicity in data utilization can still yield significant benefits in a BPD strategy.

[@zhang2017dynamically] Email Marketing and Profitability

-   **Introduction**:

    -   Despite its profitability and ubiquity, email marketing is understudied in marketing literature.

    -   Previous research predominantly looks at either customer email responses or the mean effect of emails on purchases.

-   **Methodology**:

    -   Utilized data from a U.S. home improvement retailer.

    -   Employed a unified hidden Markov and copula framework to analyze customer email open behaviors alongside purchase patterns.

-   **Key Findings**:

    1.  **Discrepancy Between Email Activity and Purchases**:

        -   Active email readers don't necessarily translate to active buyers and vice versa.

    2.  **Non-linear Effect of Email Volume**:

        -   The number of emails dispatched by the retailer influences short- and long-term profitability in a non-linear manner.

-   **Counterfactual Insights**:

    -   The study introduces a decision support system aiding retailers in optimizing email contact strategies.

    -   Pinpointing the right volume of emails is crucial for sustaining profitability in the long run.

        -   Sending four emails (as opposed to the optimal seven) can lead to a 32% loss in lifetime profit per customer.

        -   Sending ten emails (versus the optimal seven) can result in a 16% dip in lifetime profit per customer.

-   **Implications**:

    -   Retailers must go beyond conventional wisdom and understand the nuanced relationship between email engagement and purchase behavior.

    -   It's vital to strike a balance in email frequency to maximize long-term profitability. Over or under-communicating can have significant profit implications.

[@chintagunta2010effects] Analyzing the Impact of Online User Reviews on Local Box Office Performance

-   **Objective**:

    -   Assess the effects (valence, volume, and variance) of nationwide online user reviews on the local box office sales performance of movies at the Designated Market Area (DMA) level.

-   **Complications Addressed**:

    1.  **Spatial Aggregation**: Aggregating data across diverse markets.

    2.  **Endogenous Rollout**: Serial correlation arising from the staged release of films.

    3.  **Other Serial Correlations**: Other unobserved factors potentially affecting the influence of user reviews.

-   **Data and Method**:

    -   Analyzed daily box office ticket sales for 148 movies out of 874 released in the U.S. over a 16-month span.

    -   Used user review data from Yahoo! Movies.

    -   Controlled for other potential factors influencing box office sales.

    -   Employed instruments for user ratings based on the staged release of movies across markets.

-   **Key Findings**:

    1.  **Valence Over Volume**: Unlike prior studies that highlighted volume as the primary determinant, this study found valence (sentiment) of reviews to be more impactful.

    2.  **Insignificance of Endogenous Rollout**: Overlooking the staged release pattern didn't significantly alter DMA-level results.

    3.  **Contrast with National Data**: Using aggregated national data yielded results similar to past studies, emphasizing volume over valence. The study further explored reasons for this discrepancy.

-   **Significance for Marketing Managers**:

    -   Allows assessment of a DMA's responsiveness to:

        -   Firm-initiated marketing (advertising).

        -   Consumer-generated marketing (online word-of-mouth).

    -   Provides actionable insights for studios and distributors in formulating product rollout strategies.

-   **Generalizability**:

    -   While the study primarily focuses on the motion picture industry, the approach to tackling the endogeneity of reviews can be applied to other sectors with sequential product releases.
