# Selected Topics in Marketing Strategy and Measurement {#sec-a1-others}

The chapters that precede this one each develop a single construct in depth. This
one is deliberately broader. It collects a set of problems that are central to
marketing science yet do not belong to any one substantive domain: how scholarly
knowledge itself diffuses and accrues influence; how measurement breaks down when
the unit of observation fractures across devices; how the marketing function earns
and loses power inside the firm; and how a firm's resources condition the returns
to the marketing actions it takes. What unites them is a methodological posture
rather than a topic. Each is, at bottom, an *identification* problem—an attempt to
recover a causal or latent quantity from data that were not generated by a clean
experiment—and each rewards the same discipline: state the estimand, write down
the model that maps it to observables, and be explicit about the assumption whose
failure would invalidate the answer.

The reader who finishes this chapter should be able to (i) reason about latent,
time-varying drivers of an outcome using a hidden-state formalism; (ii) diagnose
and sign the biases that cross-device identity fragmentation injects into
advertising-effect estimates; (iii) connect organizational structure—the power of
the marketing department, the mix of firm capabilities—to financial performance
through mediation and contingency arguments; and (iv) frame the
resource–deployment interaction that determines whether a given marketing action
pays off. Throughout, intuition leads and the formalism follows in full, as the
house style requires.

## The Citability of Research: Influence as a Latent, Dynamic State

Marketing scholars spend their careers producing a product—published research—and
then compete for its adoption by other scholars. Citations are the revealed-demand
data for that product, and the question of *why some papers are cited and others
are not* is a marketing question in the most literal sense. The overt drivers are
familiar: the prestige of the journal, the substance of the article, the standing
of the authors, and the self-reinforcing "Matthew effect" by which early advantage
compounds into later advantage. @li2015explaining argue that these observables miss
a latent driver, which they name **citability**: the time-varying degree to which
an article is relevant and relatable to its field, and hence the propensity with
which the field draws on it.

The key word is *time-varying*. A paper's influence is not a fixed attribute
stamped at publication; it evolves as the field's attention shifts, as
complementary work appears, and as the article's ideas move into or out of
fashion. Modeling this requires a construct that is (a) unobserved, (b) dynamic,
and (c) coupled to a count outcome—annual citations—that is frequently *zero*. The
zeros are not incidental. A large share of articles go uncited in any given year,
and a model that cannot represent uncitedness will misattribute those zeros to low
intrinsic quality rather than to a transient low-citability state.

### A Hidden-Markov Tobit Formalism

@li2015explaining meet these three requirements with a **discretized Tobit model
governed by a hidden Markov process**, fit to articles in seven premier marketing
journals published between 1996 and 2003. The construction is worth giving in full
because it is a template for any setting where a latent state drives a censored or
count outcome over time.

Let $S_{it} \in \{1,\dots,K\}$ denote the unobserved citability *state* of article
$i$ in year $t$. The state evolves as a first-order Markov chain with $K \times K$
transition matrix $\mathbf{Q}$, whose entries

$$
q_{jk} = \Pr\!\left(S_{it}=k \mid S_{i,t-1}=j\right)
$$ {#eq-citability-transition}

give the probability of moving from state $j$ to state $k$ between consecutive
years. @li2015explaining find that $K=2$ states suffice: a high-citability and a
low-citability regime. Conditional on the state, a latent citation propensity
$y_{it}^{*}$ is generated as a linear function of the overt drivers,

$$
y_{it}^{*} = \mathbf{x}_{it}^{\top}\boldsymbol{\beta}_{S_{it}} + \varepsilon_{it},
\qquad \varepsilon_{it}\sim\mathcal{N}(0,\sigma^{2}_{S_{it}}),
$$ {#eq-citability-latent}

where $\mathbf{x}_{it}$ collects journal, content, and author covariates and the
coefficient vector $\boldsymbol{\beta}_{S_{it}}$ is *state-dependent*—the same
covariate can matter in one regime and not the other. The observed citation count
is the Tobit-style censoring of the latent propensity at zero,

$$
y_{it} = \max\!\left(0,\; y_{it}^{*}\right),
$$ {#eq-citability-tobit}

so that an article in the low-citability state produces an observed zero whenever
its latent propensity falls below the threshold. This is precisely the mechanism
that lets the model represent uncitedness as a *state*, not as evidence of
permanent irrelevance.

> **Citability.** A latent, time-varying attribute of a published article
> measuring how relevant and relatable it is to its field, which governs—through a
> hidden Markov regime—the propensity of the field to cite it, and which can shift
> an article between high- and low-influence states over its life.

Estimation proceeds by maximum likelihood. The likelihood of an article's full
citation trajectory $\mathbf{y}_i = (y_{i1},\dots,y_{iT})$ marginalizes over the
unobserved state path using the **forward–backward (Baum–Welch) recursion**
standard for hidden Markov models: the forward variable
$\alpha_{it}(k)=\Pr(y_{i1},\dots,y_{it}, S_{it}=k)$ is propagated as

$$
\alpha_{it}(k) = \Big[\textstyle\sum_{j} \alpha_{i,t-1}(j)\,q_{jk}\Big]\,
f\!\left(y_{it}\mid S_{it}=k\right),
$$ {#eq-forward}

where $f(\cdot\mid S_{it}=k)$ is the censored-normal density implied by
@eq-citability-latent and @eq-citability-tobit, and the article-level likelihood
is $\sum_k \alpha_{iT}(k)$. Identification of the two regimes rests on three
assumptions that the analyst must defend: the Markov order (one lag of the state
carries all dynamics), the conditional independence of citation counts given the
state path, and sufficient separation of the state-specific parameters
$(\boldsymbol{\beta}_k,\sigma_k)$ for the states to be empirically distinguishable
rather than label-switched artifacts. The substantive payoff is that influence is
shown to be dynamic: the same paper can move between regimes, and the overt drivers
have *regime-specific* effects, a structure that static citation-count regressions
cannot represent. @fig-citability-hmm assembles these pieces into a single schematic,
from the latent two-state chain through the censoring that yields the observed count.

```{mermaid}
%%| label: fig-citability-hmm
%%| fig-cap: "The hidden-Markov citability model. An unobserved two-state chain (high vs. low citability) evolves via the transition matrix Q; conditional on the current state, overt drivers generate a latent citation propensity that is censored at zero to yield the observed annual citation count."
flowchart LR
    S1((High\ncitability)) -- q11 --> S1
    S1 -- q12 --> S2((Low\ncitability))
    S2 -- q21 --> S1
    S2 -- q22 --> S2
    S1 -. state-dependent .-> Y["Latent propensity\ny* = x'β_S + ε"]
    S2 -. state-dependent .-> Y
    Y -- "max(0, y*)" --> C["Observed\ncitations y"]
```

### From Influence to Relevance

Citability concerns influence *within* the academic field. A distinct and equally
strategic question is whether marketing scholarship is relevant *outside* it—to
practice. @jedidi2021 operationalize this with an index of **relevance to
marketing**, R2M, that scores publications on their relevance to practitioners
along two dimensions, the *topic* addressed and the *time* horizon over which it
matters. Read together, the two constructs partition the value of a marketing
paper into an internal coordinate (influence on other scholars, captured by
citability) and an external one (usefulness to managers, captured by R2M). The
pairing is a useful reminder that "impact" is not a scalar: a paper can score high
on one axis and low on the other, and the field's incentive structure determines
which axis researchers optimize.

## Measurement Under Identity Fragmentation

Causal measurement in marketing presumes that the analyst can match a treatment
(an ad exposure) to an outcome (a purchase) at the level of a single decision
maker. Modern digital reality violates that premise. A consumer browses on a
phone, researches on a laptop, and buys on a tablet, and absent perfect
cross-device linkage the analyst observes three *identities* where there is one
person. @lin2022 shows that this **identity fragmentation** is not a nuisance to be
waved away but a structured source of bias that contaminates the very
advertising-effect estimates marketing most cares about. The contribution is to
decompose the bias into three mechanically distinct channels and to state, for
each, the exact condition under which it vanishes.

### Three Channels of Bias

The first channel is **outcome fragmentation**, which produces classical
*attenuation*. When one person's two devices are counted as two identities, the
total purchase outcome is unchanged but is now spread across an artificially
inflated number of observations. The measured per-identity outcome is therefore a
shrunken version of the truth, and any regression of outcome on exposure inherits
the shrinkage. The disquieting implication is that this bias disappears only when
the true effect is itself zero: for any genuine advertising effect, outcome
fragmentation guarantees that the estimate is biased toward null.

The second channel is **exposure fragmentation**, which behaves like *omitted
variable bias*. The estimand of interest is the effect of a consumer's *total*
cross-device exposure, but the fragmented data condition only on within-device
exposure, omitting the consumer's exposures on her other devices. The omitted
cross-device exposure is correlated with both the included exposure and the
outcome, so the coefficient is biased. This bias vanishes only under the implausible
exclusion restriction that exposure on one device never influences purchase on
another—a condition that the entire premise of cross-channel advertising denies.

The third channel is **spurious covariance**, which arises from *device-level
activity bias*: consumers systematically use different devices for different stages
of the funnel—browsing on one, buying on another—so device usage is correlated with
purchase propensity for reasons unrelated to advertising. Differential exposure
levels and cross-device substitution then induce covariance between measured
exposure and outcome that is not causal. @lin2022 notes that estimating a
*separate* model for each device type can purge the device-level activity-bias
component of this channel, because within a device type the differential-usage
confound is held fixed. @tbl-fragmentation summarizes the three channels, the
statistical analogue of each, and the condition under which each bias vanishes.

```{r}
#| label: tbl-fragmentation
#| tbl-cap: "The three biases from identity fragmentation, their statistical analogue, and the condition under which each vanishes (after Lin and Misra 2022)."
#| echo: true
#| message: false
#| warning: false
frag <- data.frame(
  Channel = c("Outcome fragmentation",
              "Exposure fragmentation",
              "Spurious covariance"),
  `Statistical analogue` = c("Attenuation (measurement error)",
                             "Omitted-variable bias",
                             "Confounding via device-activity bias"),
  `Vanishes only if` = c("True effect equals zero",
                         "No cross-device exposure spillover",
                         "Per-device models (removes activity bias)"),
  check.names = FALSE
)
knitr::kable(frag)
```

### A Worked Illustration of Attenuation

The attenuation channel is sharp enough to demonstrate with a small simulation.
Suppose advertising has a genuine positive effect on purchase. We generate person-
level data, then *fragment* each person into two device-identities that split the
person's exposure and outcome, and compare the naive per-identity regression slope
to the truth.

```{r}
#| label: fig-fragmentation-attenuation
#| fig-cap: "Identity fragmentation attenuates the estimated advertising effect. The fragmented per-identity slope (red) lies below the true person-level slope (blue); the gap widens with the share of consumers using multiple devices."
#| echo: true
#| message: false
#| warning: false
set.seed(42)

simulate_attenuation <- function(frac_multi, n = 5000, beta_true = 1.0) {
  exposure <- rpois(n, lambda = 4)
  purchase <- beta_true * exposure + rnorm(n, sd = 2)
  multi    <- runif(n) < frac_multi          # which people use two devices

  # Each multi-device person is split into two identities sharing the totals.
  id_exposure <- c(exposure[!multi],
                   exposure[multi] / 2, exposure[multi] / 2)
  id_purchase <- c(purchase[!multi],
                   purchase[multi] / 2, purchase[multi] / 2)

  coef(lm(id_purchase ~ id_exposure))[["id_exposure"]]
}

fracs  <- seq(0, 0.9, by = 0.1)
slopes <- vapply(fracs, simulate_attenuation, numeric(1))

plot(fracs, slopes, type = "o", pch = 19, col = "red",
     ylim = c(0, 1.1), xlab = "Fraction of consumers using multiple devices",
     ylab = "Estimated advertising effect",
     main = "Attenuation under identity fragmentation")
abline(h = 1.0, col = "blue", lwd = 2, lty = 2)
legend("bottomleft", bty = "n",
       legend = c("Fragmented estimate", "True effect = 1.0"),
       col = c("red", "blue"), pch = c(19, NA), lty = c(1, 2))
```

As @fig-fragmentation-attenuation shows, the estimate degrades monotonically as
multi-device prevalence rises, exactly as the attenuation argument predicts.[^split] The lesson generalizes: any pipeline
that treats device-identities as exchangeable for persons will under-report
advertising effectiveness, and the under-reporting is worst in precisely the
mobile-heavy, multi-screen segments that advertisers most want to reach.

[^split]: The simulation splits each multi-device person's exposure and outcome
evenly across her two identities. Uneven splits, or splits correlated with the
outcome, additionally activate the spurious-covariance channel; the even split
isolates the pure attenuation effect.

### Remedies and Their Failure Modes

Three remedies are available, each buying identification at a price.
*Identity linking*—stitching device identities back into persons—is the
first-best in principle but treacherous in practice: *partially* linked data can be
worse than unlinked data, because a linkage that resolves some persons but not
others introduces a new, selection-driven heterogeneity that can dominate the
original bias. *Experiment-based adjustment* follows @coey2016, who recover the
true effect from an experiment under the **symmetric and independent exposures**
(SIE) assumption—that a consumer's device-level exposures are exchangeable and
independent. The estimator is clean when SIE holds, but @lin2022 shows it remains
biased whenever cross-device *activity* variation is present, because SIE is silent
about differential device usage. *Stratified aggregation* offers a more robust, if
coarser, route: first group consumers by geography, then bin each geographic group
by combinations of demographic characteristics (e.g., age, gender, education), and
estimate effects within bins where the fragmentation-inducing heterogeneity is
approximately constant. The choice among the three is a bias–variance trade-off,
and @lin2022 is best read as a map of where each method's identifying assumption
breaks.

## Marketing in the Firm: Power, Capabilities, and Performance

A second cluster of work asks how the marketing function fares *inside* the
organization and whether its standing translates into firm value. The constructs
are organizational—departmental power, firm-level capabilities, cross-functional
relationships—but the empirical strategy is the financial-economics one of linking
a marketing variable to long-run performance while ruling out confounds.

### The Power of the Marketing Department

@feng2015marketing track the **power of the marketing department**—its influence
relative to other functions—across a cross-industry panel of 612 public U.S. firms
from 1993 to 2008, using an objective metric rather than survey perceptions. Two
findings stand out. First, marketing-department power rose materially over the
period, against a prevailing narrative of marketing's organizational decline.
Second, and more consequentially, greater power predicts superior *long-run* total
shareholder return more strongly than it predicts short-run return on assets
(ROA)—the value of marketing influence accrues over horizons longer than a single
fiscal year.

The mechanism is a **mediation** argument built on market-based assets. Power
operates through two capability channels: long-run *market-based-asset building*
and short-run *market-based-asset leveraging*. These capabilities *partially*
mediate the effect of power on long-term shareholder value and *fully* mediate its
effect on short-term ROA. The distinction matters for inference. Under the standard
mediation logic, a fully mediated effect implies that, conditional on the
capability channel, the direct path from power to ROA is statistically
indistinguishable from zero—so the entire short-run benefit of marketing power runs
through asset leveraging. A partially mediated effect leaves a residual direct path,
indicating that long-run value creation has sources the measured capabilities do
not capture. Identifying mediation credibly requires the usual assumptions—no
unmeasured confounding of the power–mediator and mediator–outcome links, and
correct temporal ordering—which panel structure helps but does not guarantee.
@fig-mktpower-mediation traces these two capability channels and the differing
degrees to which each mediates the short-run and long-run performance paths.

```{mermaid}
%%| label: fig-mktpower-mediation
%%| fig-cap: "Marketing-department power and firm performance. Power operates through two market-based-asset capabilities: building (long-run) and leveraging (short-run). The capabilities fully mediate the effect on short-run ROA and partially mediate the effect on long-run shareholder value."
flowchart LR
    P["Marketing-department\npower"] --> B["Market-based-asset\nbuilding (long-run)"]
    P --> L["Market-based-asset\nleveraging (short-run)"]
    B -->|partial mediation| SV["Long-run\nshareholder value"]
    L -->|full mediation| ROA["Short-run\nROA"]
    P -. residual direct path .-> SV
```

### Capabilities and Growth as a Contingency

If power is one organizational asset, *capabilities* are another, and their returns
are not additive. @feng2017firm study how three firm capabilities—**marketing**,
**research and development (R&D)**, and **operations**—interact to drive revenue and
profit growth, using panel data on 612 public U.S. firms observed over 16 years
across 60 industries, and framed by **contingency theory**: the value of a
capability depends on its complements and on the environment.

The interaction structure is the substance. R&D capability *amplifies* the growth
effect of marketing capability—innovation gives marketing something worth
selling—whereas operations capability *attenuates* it, plausibly because an
operations orientation tuned for efficiency sits in tension with the
demand-expanding thrust of marketing. Crucially, these interaction signs are not
fixed constants but vary with two **boundary conditions**: market *munificence*
(the abundance of growth opportunities) and competitive *dynamism* (the volatility
of the competitive environment). A formalism makes the contingency explicit. Let
growth $g_i$ for firm $i$ depend on marketing ($M$), R&D ($R$), and operations
($O$) capabilities through

$$
g_i = \beta_M M_i + \beta_R R_i + \beta_O O_i
      + \gamma_{MR}\,(M_i R_i) + \gamma_{MO}\,(M_i O_i) + \mathbf{z}_i^{\top}\boldsymbol{\delta} + u_i,
$$ {#eq-capability-interaction}

where @feng2017firm find $\gamma_{MR}>0$ and $\gamma_{MO}<0$, and where the
interaction coefficients are themselves modeled as functions of munificence and
dynamism, $\gamma_{MR}=\gamma_{MR}(\text{munificence},\text{dynamism})$. The
managerial reading is direct: the optimal capability portfolio is
environment-specific, and a firm that pairs marketing with R&D in a munificent
market is exploiting a different complementarity than one defending share in a
dynamic one. Identification here rests on the exogeneity of the interacted
capabilities to the growth shock $u_i$—a strong assumption that the panel's
within-firm variation only partly relaxes.

### Cross-Functional Coopetition

The capability story treats functions as inputs to be combined; @luo2006cross asks
how the *relationships* among them should be configured. Marketing's interfaces
with other departments are conventionally cast as either cooperative or
competitive. @luo2006cross breaks that dichotomy with **coopetition**: the
simultaneous presence of cooperation *and* competition between functions. Drawing on
data from mid-level managers and senior executives, the study finds that
cross-functional coopetition *raises* both customer and financial outcomes—a result
that runs against the intuition that internal competition is purely destructive.

The effect is not direct. It is channeled through **market learning**: coopetition
generates the friction and information exchange that surface market intelligence,
and that learning is what improves outcomes. Mediation through learning is the
identifying claim, and it reframes internal competition as a knowledge-generating
mechanism rather than a coordination failure—provided the competitive element
remains coupled to a cooperative one rather than degenerating into pure rivalry.

## Market Share, Customers, and Shareholder Value

A third cluster connects classic strategic levers—market share, customer
portfolios—to profit and to the stock market's assessment of the firm.

### Why and When Market Share Drives Profit

Firms routinely set market share as an objective and treat it as a scorecard, on
the presumption that share causes profit. @bhattacharya2022examining interrogate
that presumption by estimating, *simultaneously*, the three theoretical mechanisms
that the literature offers for the share–profit link: **market power** (large share
confers pricing latitude), **quality signaling** (share is read by buyers as a
proxy for quality), and **operating efficiency** (scale lowers unit cost). The
simultaneity is the methodological point—prior work tends to examine one mechanism
in isolation, which cannot apportion the total effect.

The result is that market power and quality signaling jointly explain most of the
positive share-to-profit variance, while operating efficiency contributes little.
Strikingly, the same two mechanisms dominate even in the cases where share and
profit are *negatively* related—niche firms, and firms that "buy" share through
unprofitable promotion—indicating that the mechanisms are structural features of
the relationship rather than artifacts of the average case. The practical
corollary is that a share-building strategy pays off to the extent it strengthens
pricing power or quality perceptions, and not merely because it adds volume.

### When Firing Customers Destroys (or Preserves) Value

Not all customers are profitable, and firms sometimes manage the unprofitable ones
out. @feng2020impact ask what the stock market makes of **unprofitable customer
management** (UCM) and use an **event study** to find out: they measure the
abnormal stock return in a short window around announcements of UCM strategic
decisions. On average the market reacts *negatively*, with a short-run abnormal
return of about $-0.53\%$—investors do not, by default, applaud the firing of
customers.

The interpretation runs through **signaling theory**, decomposing the announcement
into properties of the *signal* (the UCM strategy itself), the *signaler* (the
firm), and the signaling *environment*. *Indirect* UCM strategies—raising fees,
reducing service, nudging unprofitable customers to leave—are received more
favorably than *direct* divestment, in which the firm overtly terminates customer
relationships. Moreover, specific indirect strategies, executed with clear strategic
intent and backed by strong marketing capabilities and positive media coverage, can
*offset* the average negative reaction entirely. The event-study logic is the same
one developed for brand-asset transactions in @sec-branding: the abnormal return
$\mathrm{AR}_{it}=R_{it}-\mathbb{E}[R_{it}\mid\text{market model}]$ is cumulated
over the event window and regressed on signal, signaler, and environment
characteristics, after screening the window for confounding announcements. Its
validity rests on market efficiency over the window and on the absence of
contemporaneous confounds.

## Resources, Deployment, and the Returns to Marketing Action

The final cluster takes up the resource-based view (RBV) of the firm and asks a
question the preceding clusters presuppose: *given* a firm's resources, what
determines whether its marketing actions actually pay off?

### Planning Capability and a Productive Paradox

Strategy scholars have long debated whether *formal planning* helps or hinders, and
the evidence is genuinely mixed. @slotegraaf2004paradox resolve part of the tension
by viewing marketing-planning capability through the RBV and uncovering a
**paradox**: a strong planning capability has two opposing consequences that can
*both* improve performance. On one hand, good planning *reduces* the need for
post-plan improvisation—execution follows the plan, and the firm avoids costly
mid-course scrambling. On the other hand, it introduces process *rigidity*—a
well-planned firm is less able to deviate when conditions change. The paradox is
that each effect can be performance-enhancing in its own right: reduced
improvisation lowers execution cost, while rigidity can protect a coherent strategy
from value-destroying ad hoc revision. The net effect is therefore not signed a
priori, which explains why the planning-performance literature reaches conflicting
conclusions—it has been estimating a composite of two channels with opposite
behavioral content.

### Resources Condition the Returns to Deployment

@slotegraaf2003role sharpen the central RBV question. Marketing research has tended
to study *either* a firm's resources *or* its strategic and tactical actions, but
neither alone explains competitive advantage. The authors introduce **market
deployment**—the specific marketing actions a firm takes (distribution pushes,
couponing, and the like)—and ask how a firm's *resource base* governs the *return*
to deployment. The estimand is therefore an interaction: the marginal effect of an
action on outcomes, as a function of the resources the firm brings to it.

Using a secondary-data design and a sequence of hierarchical regressions, they find
that resources directly shape deployment returns, and that the sign depends on the
*type* of resource. High levels of intangible *marketing* resources and
*technological* resources *enhance* the success of deployment activities such as
distribution and couponing. Higher levels of *financial* resources, by contrast,
*reduce* the efficiency of those same activities—plausibly because financial slack
dulls the discipline that makes deployment effective. A compact representation makes
the interaction explicit. Let $y_i$ be the outcome of a deployment action $D_i$ for
firm $i$ holding a resource bundle $\mathbf{r}_i$:

$$
y_i = \beta_0 + \beta_D D_i + \boldsymbol{\theta}^{\top}(\mathbf{r}_i \odot D_i)
      + \mathbf{r}_i^{\top}\boldsymbol{\phi} + \varepsilon_i,
$$ {#eq-deployment-interaction}

where $\odot$ denotes the element-wise product of the resource vector with the
deployment intensity, and the interaction vector $\boldsymbol{\theta}$ carries the
result: positive entries for intangible marketing and technological resources,
negative for financial resources. The managerial implication is that the
*same* marketing action is not equally productive across firms—its return is
contingent on the resource base—and that resource abundance is not uniformly good,
since financial slack can erode the efficiency of execution.

## Sustainability as a Brand-Level Action

A final, more focused result connects the resource-and-action theme to a salient
contemporary lever. @olsen2014green study the brand-level consequences of
introducing environmentally sustainable ("green") new products—an increasingly
common, resource-intensive marketing action—integrating **social identity** and
**framing** theories. Estimating a **three-stage least squares** (3SLS) system over
green product introductions from 75 brands during 2009–2012, they find that green
launches *can* improve brand attitude, but that the magnitude is conditional. The
volume of green-centric messaging, the product's typology, and the *credibility of
the information source* jointly determine how much attitude moves, and a brand's
strategic positioning together with its product category governs whether a green
launch is undertaken at all. The 3SLS specification is the methodologically
load-bearing choice: because the decision to *introduce* a green product and its
*effect* on brand attitude are jointly determined, single-equation estimation would
confound selection with effect, and the simultaneous-equations system is what
restores identification of the attitude effect net of the selection into green
launches.

## Key Takeaways

- **Latent dynamics need latent-state models.** Article influence is a time-varying
  hidden state; a hidden-Markov Tobit (@eq-citability-transition–@eq-citability-tobit)
  represents both the dynamics and the uncitedness that count regressions miss
  [@li2015explaining]. Impact is two-dimensional once external relevance (R2M) is
  added [@jedidi2021].
- **Identity fragmentation is structured bias, not noise.** Its three channels—
  attenuation, omitted cross-device exposure, and device-activity confounding—each
  vanish only under a specific, often implausible, condition, and partial identity
  linking can make matters worse [@lin2022; @coey2016].
- **Marketing's organizational standing is priced.** Department power predicts
  long-run shareholder value through market-based-asset capabilities [@feng2015marketing];
  capability returns are contingent and interactive [@feng2017firm]; and
  cross-functional coopetition pays off *through* market learning [@luo2006cross].
- **Share and customer decisions reach the stock market through mechanisms.** Market
  power and quality signaling—not efficiency—carry the share–profit link
  [@bhattacharya2022examining], and unprofitable-customer management is read by
  investors as a signal whose reception depends on its form [@feng2020impact].
- **Resources condition the returns to action.** Planning capability cuts both ways
  [@slotegraaf2004paradox]; intangible and technological resources lift deployment
  returns while financial slack erodes them [@slotegraaf2003role]; and green-product
  effects are identified only once selection into launching is modeled jointly
  [@olsen2014green].
