# Marketing Strategy {#sec-marketing-strategy}

Marketing strategy is the part of the field that asks how a firm wins, not how a
consumer chooses. Where most of the book studies decisions one customer at a time,
this chapter steps up a level of analysis to the firm, and asks how the firm
creates value for customers, captures part of that value for itself, and sustains
the arrangement against rivals who would rather it did not. Marketing strategy is
the integrated pattern of decisions through which a firm does this: which markets
to serve, which customers to win and keep, what to be known for, where to invest
the marketing budget, and how to organize the people who execute it. The unifying
test is performance. A marketing strategy is judged not by how clever it sounds
but by the customer, financial, stock-market, and societal outcomes it produces
[@varadarajan2010; @morgan2018].

This chapter is a teaching treatment of the field, complementing the doctoral
reading map in @sec-marketing-strategy-seminar. The seminar chapter is organized
as a fourteen-week syllabus that pairs each topic with its identification
challenge; this chapter develops the substantive theory in continuous prose, with
two small runnable demonstrations. Two commitments run through both. The first is
*integration*: a strategy is a coherent pattern, not a list of independent
marketing-mix moves, even though most empirical work has studied the parts rather
than the pattern [@morgan2018]. The second is *accountability*: the soft-sounding
constructs of strategy, orientation, capability, culture, relationships, are
repeatedly forced into contact with hard outcome data, because that contact is
what turns a plausible story into a finding.

## What Marketing Strategy Is {#sec-ms-domain}

It helps to fix vocabulary first, because the field uses two phrases that are
easily confused. *Strategic marketing* names the scholarly field: the study of how
marketing contributes to firm strategy and performance. *Marketing strategy* names
its object: the integrated pattern of choices a particular firm makes
[@varadarajan2010]. A doctoral student needs both senses, but the second is the
one with managerial bite.

### Levels of strategy

Strategy is layered, and marketing sits at more than one layer. *Corporate
strategy* decides which businesses the firm should be in. *Business strategy* (or
competitive strategy) decides how a given business unit competes in its chosen
market. *Functional strategy*, including marketing strategy in its narrow sense,
decides how each function supports the business-level choice. Marketing is unusual
among the functions in that it spans all three levels: it informs corporate
choices about which markets have attractive customers, it largely *is* the
business-level question of how to compete for those customers, and it supplies the
functional programs (the marketing mix) that execute the plan. This is why
marketing's claim to a seat at the strategy table is contested but persistent: the
function's natural object, the customer and the market, is exactly the object that
business-level strategy must get right.

### Strategy versus tactics

The cleanest way to separate strategy from tactics is by what each commits the
firm to and for how long. *Strategy* is the small set of choices that are costly to
reverse and that constrain everything downstream: which segment to target, what
position to own, which capabilities to build. *Tactics* are the many adjustable
moves, this quarter's price, this campaign's copy, this promotion's depth, that
execute a strategy and can be changed without changing the firm's direction. The
distinction is not a hierarchy of importance; a brilliant strategy with sloppy
tactics fails, and disciplined tactics cannot rescue an incoherent strategy. It is
a distinction of *reversibility and scope*. A useful diagnostic question is: if
this decision turns out wrong, how long and how expensive is the path back? Choices
with long, expensive paths back are strategic and deserve the deliberation the rest
of this chapter describes.

### The dependent variable

Because the field is defined by accountability, its recurring dependent variable is
*firm performance*, and that variable has conceptual layers worth separating
[@katsikeas2016; @morgan2012performance]. Closest to marketing actions are
*customer-mindset* outcomes (awareness, satisfaction, loyalty intentions). These
feed *marketplace* outcomes (market share, sales, customer retention), which feed
*accounting* outcomes (margins, cash flow, return on assets), which the capital
market prices into *financial-market* outcomes (stock returns, firm value, risk).
Much of the field's progress has come from tracing this chain link by link, and
much of its difficulty comes from the fact that causation can run backward along it
(profitable firms can afford more marketing) and that omitted firm-quality
variables can drive several links at once.

## Market Orientation {#sec-ms-market-orientation}

If marketing strategy has a single foundational construct on the implementation
side, it is *market orientation*: the degree to which a firm actually organizes
itself around its customers and competitors rather than merely professing to. Two
papers in 1990 defined the construct in complementary ways, and the field has spent
three decades measuring it and testing whether it pays.

### The behavioral view and the cultural view

@kohli1990 give the *behavioral* conceptualization. They define market orientation
as "the organizationwide generation of market intelligence pertaining to current
and future customer needs, dissemination of the intelligence across departments,
and responsiveness to it." The achievement of this definition is that it
decomposes a vague cultural posture into three *observable behaviors*, intelligence
generation, dissemination, and responsiveness, and thereby makes the construct
measurable (the MARKOR scale follows directly from it). Profitability is
deliberately placed as a *consequence* of orientation, not a part of it, which
keeps the construct clean enough to serve as a regressor against performance. The
seminar chapter develops this decomposition in detail in
@sec-strategy-market-orientation.

@narver1990 give the *cultural* conceptualization. They define market orientation
as the organizational culture that most effectively creates the behaviors needed
for superior value, and they decompose it into three components, customer
orientation, competitor orientation, and interfunctional coordination, plus two
decision criteria (long-term focus and profitability). Their MKTOR scale
operationalizes this view, and their paper supplied the first direct test linking
orientation to business profitability.

The two views are less rival than layered: culture (Narver and Slater) is the
disposition, and behavior (Kohli and Jaworski) is its enactment. A firm can hold
customer-oriented values yet fail to generate or act on intelligence; the
behavioral view measures whether the disposition actually shows up in routines.

### Does market orientation pay, and when?

The headline empirical question is whether more market-oriented firms perform
better, and the meta-analytic verdict is a qualified yes. @kirca2005 synthesize the
accumulated studies and find a positive, robust association between market
orientation and performance, but one whose size depends on the setting. The link is
stronger when performance is measured by managerial judgment than by objective
financials, stronger in studies using cost-based rather than revenue-based
outcomes, and moderated by national culture and industry. The honest reading is
that market orientation is *necessary but not sufficient*: it is a precondition for
converting market knowledge into value, but its payoff is contingent on the
capabilities (next section) that turn knowledge into action and on the competitive
context that determines how much customer focus is rewarded.

This sets up the field's signature identification worry. A cross-sectional
correlation between orientation and profit is consistent with orientation causing
profit, with profit funding orientation, or with some omitted quality (good
management, slack resources) causing both. The demonstration below uses simulated
data to show what an honest analyst can and cannot conclude from such a regression,
and how a moderator changes the picture.

### A simulated market-orientation to performance regression

The following chunk simulates a cross-section of firms, generates performance as a
function of market orientation *and* a moderator (competitive intensity), and then
recovers the relationship. The data-generating process is known, so we can see
exactly what the regression is estimating. The substantive point is that the
return to market orientation is larger where competition is fiercer: customer focus
matters most when customers have alternatives.

```{r ms-mo-regression, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({library(dplyr); library(ggplot2)})
set.seed(2024)

n <- 600

firms <- tibble(
  # market orientation (standardized index, e.g. MARKOR/MKTOR)
  mo        = rnorm(n, 0, 1),
  # competitive intensity moderator (0 = sheltered, 1 = cut-throat)
  intensity = runif(n, 0, 1),
  # an omitted "firm quality" that raises BOTH mo and performance
  quality   = rnorm(n, 0, 1)
)

# true model: MO helps, and helps MORE under high competitive intensity;
# firm quality confounds the naive association.
firms <- firms %>%
  mutate(
    mo_obs = mo + 0.5 * quality,                       # quality inflates MO
    perf   = 0.30 * mo_obs +                           # baseline MO effect
             0.60 * mo_obs * intensity +               # moderation
             0.80 * quality +                          # confound
             rnorm(n, 0, 1)
  )

# (1) naive regression: omits quality -> MO effect is biased upward
naive <- lm(perf ~ mo_obs, data = firms)

# (2) moderated regression controlling for quality -> recovers structure
moderated <- lm(perf ~ mo_obs * intensity + quality, data = firms)

round(rbind(
  naive     = c(coef(naive)["mo_obs"], NA, NA),
  moderated = coef(moderated)[c("mo_obs", "intensity", "mo_obs:intensity")]
), 3)
```

The naive slope on market orientation is inflated because it absorbs the omitted
firm-quality effect; the moderated specification, once quality is controlled,
recovers a positive main effect and a positive interaction, the return to market
orientation rises with competitive intensity. @fig-ms-mo-moderation plots the fitted
performance-orientation slope at low and high competitive intensity to make the
moderation visible.

```{r}
#| label: fig-ms-mo-moderation
#| fig-cap: "Simulated moderation of the market-orientation to performance link by competitive intensity. The fitted slope of performance on market orientation is steeper under high competitive intensity (customers have alternatives) than under low intensity, illustrating that market orientation is necessary but its payoff is contingent [@kirca2005]."
#| message: false
#| warning: false
grid <- expand.grid(
  mo_obs    = seq(-2.5, 2.5, length.out = 50),
  intensity = c(0.15, 0.85),
  quality   = 0
)
grid$perf <- predict(moderated, newdata = grid)
grid$Intensity <- factor(grid$intensity,
                         labels = c("Low competitive intensity",
                                    "High competitive intensity"))

ggplot(grid, aes(mo_obs, perf, colour = Intensity)) +
  geom_line(linewidth = 1.1) +
  labs(x = "Market orientation (standardized)",
       y = "Predicted firm performance",
       colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

The lesson generalizes well beyond this toy: a positive average effect can hide
substantial heterogeneity, and the moderators (here, competition; in the
literature, also market turbulence and national culture) are often the
managerially interesting part of the answer.

## Resources, Capabilities, and Sustainable Advantage {#sec-ms-rbv}

Market orientation tells the firm to listen to the market; it does not, by itself,
explain why some firms convert what they hear into durable advantage and others do
not. For that the field reaches to theories of competitive advantage, and the
dominant one is the resource-based view (RBV).

### The resource-based view and resource-advantage theory

The RBV holds that sustained advantage comes from resources that are valuable,
rare, imperfectly imitable, and non-substitutable (the VRIN tests), and that are
embedded in an organization able to exploit them [@wernerfelt1984]. The insight is
that resources easily bought or copied cannot support advantage for long, because
competitors acquire or replicate them; advantage must rest on resources that resist
the market for resources itself. Marketing's contribution to the RBV is the
observation that its most important assets, brand equity, customer relationships,
channel ties, market knowledge, are precisely the off-balance-sheet, relationship-
and knowledge-based resources that best pass the VRIN tests, a point developed as
the *market-based assets* framework [@Srivastava_1998] and treated at length in
@sec-marketing-finance.

Marketing has also produced its own theory of competition. @hunt1995 propose
*resource-advantage theory* (originally the comparative-advantage theory of
competition), an evolutionary, disequilibrium alternative to the equilibrium models
of industrial-organization economics. In their account, firms compete by seeking
*comparative advantages in resources* that yield marketplace positions of
competitive advantage and, in turn, superior financial performance; competition is
the ongoing, never-settled process of firms trying to neutralize or leapfrog one
another's resource advantages. The contrast with equilibrium IO is the heart of a
recurring seminar debate: are marketing resources truly inimitable sources of
lasting rents, or transient advantages constantly eroded by imitation? R-A theory's
answer is that the *process* of seeking advantage, not any particular static
advantage, is what marketing strategy is about.

### Marketing capabilities and dynamic capabilities

Resources are inert without the routines that deploy them, and those routines are
*capabilities*. @day1994 gives the canonical typology of the capabilities of
market-driven organizations: *outside-in* capabilities (market sensing, customer
linking, channel bonding) that connect the firm to its environment; *inside-out*
capabilities (manufacturing, logistics, financial management) that are activated by
market needs; and *spanning* capabilities (new-product development, pricing,
purchasing, service delivery) that integrate the two. A market-driven firm is one
whose outside-in capabilities are strong enough to guide the rest.

@vorhies2005 turn this typology into measurement, showing that distinct marketing
capabilities (in pricing, product development, channel management, marketing
communications, selling, market information management, and marketing planning and
implementation) can be benchmarked and that firms whose capabilities exceed
benchmarks perform better. @krasnikov2008 then run the horse race that matters for
resource allocation: in a meta-analysis comparing marketing, R&D, and operations
capabilities, marketing capabilities have the *largest* relative impact on firm
performance, a finding that pushes back on the engineering-centric view that R&D and
operations are where advantage is built. Crucially, market orientation and
marketing capabilities are *complements*, not substitutes: orientation supplies the
knowledge that capabilities convert into market position [@morgan2009].

The frontier worry is that static capabilities decay as markets change. *Dynamic
capabilities*, the firm's capacity to sense opportunities, seize them, and
reconfigure its resource base, address strategy under change [@teece1997]. @day2011
brings this into marketing with the *marketing capabilities gap*: the routines most
firms have were built for a stable, push-marketing world and lag the adaptive,
outside-in capabilities (vigilant market learning, adaptive experimentation, open
networking) that turbulent digital markets reward. The managerial implication is
that capability is a moving target; the resource-based test for *sustainability*
must be re-run as the environment shifts.

## Competitive Strategy and Positioning {#sec-ms-positioning}

Having decided to compete, the firm must decide *how*. The classic marketing answer
is the *segmentation, targeting, and positioning* (STP) sequence. *Segmentation*
partitions a heterogeneous market into groups of customers with similar needs and
responses. *Targeting* selects which segments the firm's resources and capabilities
let it serve profitably and defensibly. *Positioning* designs and communicates a
distinctive place in the minds of target customers, the position the firm intends
to own.

Underlying STP is *differentiation*: the firm must give target customers a reason to
prefer it that rivals cannot easily match. Differentiation can rest on the product,
the brand, the service, the channel, or the relationship, but the strategic test is
always the same VRIN test from the RBV, a basis of differentiation confers durable
advantage only if it is valuable to customers and costly for rivals to imitate. This
is why @daywensley1988 insist on separating the *sources* of advantage (superior
skills and resources) from the *positions* of advantage they produce (superior
customer value or lower relative cost) from the *performance outcomes* that follow
(satisfaction, loyalty, share, profitability). Confusing the three is the most
common analytic error in positioning research: a firm with high share (a position)
is not thereby shown to have a superior capability (a source), because the causal
arrow can run either way.

*Competitive dynamics* adds the recognition that positions are contested in real
time. Rivals respond to attacks, match price cuts, copy features, and counter-
position, so a position's value depends on how rivals will react, not only on how
customers will. The strategic question is therefore not merely "what position do we
want?" but "what position can we hold given how competitors will respond?", which is
where R-A theory's disequilibrium view [@hunt1995] and the dynamic-capabilities view
[@teece1997; @day2011] re-enter: a defensible position is one anchored to a
capability rivals cannot quickly reconfigure to match.

## The Marketing-Finance Interface: The Accountability Turn {#sec-ms-finance}

The most consequential development in marketing strategy over the last quarter
century is the *accountability turn*: the insistence that marketing demonstrate its
value in the language the boardroom and the capital market actually use, cash flow,
firm value, and risk. This is the subject of its own chapter
(@sec-marketing-finance); here we give the strategic logic and a panel
demonstration.

### Marketing and shareholder value

The hinge is the *market-based assets* framework [@Srivastava_1998], which reframes
marketing's outputs, brand equity, customer equity, channel relationships, as
intangible assets that drive the firm's future cash flows and therefore its value.
Under an efficient capital market, actions that build value with customers should
build value for shareholders, because a firm is worth the risk-adjusted present
value of its expected future cash flows and marketing assets shift those flows in
four ways: they can *accelerate* cash flows, *augment* their level, *reduce* their
volatility and vulnerability (lowering the discount rate), and raise their
*residual* (terminal) value [@Srivastava_1998].

@srinivasan2009 supply the canonical roadmap of how to measure these links: stock-
return response models, persistence and vector-autoregression models, and Tobin's
$q$ regressions, together with the identification toolkit (event studies, factor
models, panel methods) the field uses to defend causal claims. The meta-analytic
payoff is that marketing variables, brand, customer satisfaction, R&D, advertising,
have economically meaningful, positive elasticities on firm value
[@edeling2016; @edeling2021]. Brand equity (@sec-brand-equity) and customer equity
(@sec-clv) are the two strategic assets this literature prices most heavily:
strong-brand portfolios earn higher returns at lower risk, and the aggregate
lifetime value of the customer base tracks, and helps explain, market
capitalization [@kumar2009expanding].

### A simulated marketing-assets to firm-value panel

The following chunk illustrates, on simulated panel data, the workhorse Tobin's $q$
regression of the interface: firm value regressed on a marketing asset (here a brand-
equity index) with firm and year fixed effects and the standard financial controls
(leverage and cash flow). Fixed effects absorb time-invariant firm quality, so the
estimate comes from *within-firm* variation in the asset over time, the design that
makes the marketing-to-value claim credible rather than merely correlational. This
reuses the kind of machinery developed in @sec-marketing-finance.

```{r ms-firm-value-panel, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({library(dplyr)})
set.seed(7)

n_firm <- 120; n_year <- 8
firm_fe <- rnorm(n_firm, 0, 0.40)          # time-invariant firm quality
year_fe <- rnorm(n_year, 0, 0.10)

panel <- expand.grid(firm = 1:n_firm, year = 1:n_year) %>%
  mutate(
    brand_equity = 0.6 * firm_fe[firm] + rnorm(n(), 0, 1),  # quality raises BE
    leverage     = runif(n(), 0, 0.6),
    cash_flow    = rnorm(n(), 0, 1),
    # true within-firm effect of brand equity on Tobin's q is 0.25
    tobin_q = 1.2 + firm_fe[firm] + year_fe[year] +
              0.25 * brand_equity -        # the strategic effect of interest
              0.40 * leverage +
              0.15 * cash_flow +
              rnorm(n(), 0, 0.30)
  )

# Pooled OLS (no FE) overstates the effect: it confounds within- and between-firm
pooled <- lm(tobin_q ~ brand_equity + leverage + cash_flow, data = panel)

# Two-way fixed-effects: identify from WITHIN-firm changes in brand equity
fe <- lm(tobin_q ~ brand_equity + leverage + cash_flow +
           factor(firm) + factor(year), data = panel)

round(c(pooled_BE = coef(pooled)["brand_equity"],
        FE_BE      = coef(fe)["brand_equity"],
        truth      = 0.25), 3)
```

The pooled estimate is biased upward because it mixes the genuine within-firm effect
with the cross-firm correlation that firm quality induces between brand equity and
$q$; the two-way fixed-effects estimate, which uses only within-firm variation,
recovers the true 0.25 elasticity. This is the strategic content of the interface in
one line: marketing builds an asset, and a credibly identified panel shows that the
asset is priced into the value of the firm. The unresolved debate is whether
marketing *creates* that value or merely *signals* pre-existing quality, and whether
fixed effects and instruments fully purge the reverse-causality and omitted-quality
threats [@srinivasan2009; @edeling2021].

## Marketing Organization, the CMO, and Accountability {#sec-ms-organization}

Strategy is executed by an organization, and how marketing is organized, and how
much influence it wields, is itself a strategic variable. The marketing function's
influence inside the firm rose with the marketing concept and has been contested
ever since, with periodic claims that it is declining as analytics, finance, and
product functions absorb customer responsibility.

@verhoef2009influence ask what actually drives the marketing department's clout and
find that it rests on *accountability* (the ability to demonstrate marketing's
financial contribution), *innovativeness*, and a credible connection to the
customer, not on budget size or headcount. The implication is pointed: the
accountability turn is not only an intellectual movement but the practical price of
marketing's seat at the strategy table. Departments that cannot speak the language
of firm value lose influence to those that can.

@moorman2016 give the modern integrative framework, defining *marketing excellence*
as "a superior ability to perform essential customer-facing activities that improve
customer, financial, stock market, and societal outcomes" and resting it on four
organizational elements (capabilities, configuration, human capital, and culture)
mobilized through seven activities. This MARORG framework is developed as a worked
capability theory in @sec-strategy-excellence.

The sharpest test of marketing's organizational value is the *chief marketing
officer*. Does having a CMO in the top management team improve performance? The
natural design compares firms with and without a CMO [@nath2008], and the cleanest
estimate finds firms with a CMO exhibit roughly 15% higher Tobin's $q$
[@germann2015]. But CMO presence is not randomly assigned, so the credible design
combines firm fixed effects (identifying from within-firm changes in CMO presence)
with instrumental variables (whose exclusion restriction is the untestable
assumption on which the causal claim rests). The seminar chapter develops this as a
worked identification example in @sec-strategy-cmo, and it is the template for the
whole field: a strategic construct becomes a *finding* only when its estimator and
identifying assumptions are stated in full.

Underpinning all of this is *measurement*. @katsikeas2016 map the space of marketing
performance outcomes (customer-mindset, marketplace, accounting, financial-market),
and the recurring empirical result is that the *ability to measure marketing* is
itself associated with performance and with executives' confidence in the function.
Whether measurement causes performance or better firms simply measure more is the
same endogeneity question in a new guise.

## Innovation and Growth Strategy {#sec-ms-innovation}

Growth is the strategic objective that most directly tests a firm's resource base,
and *innovation* is its primary engine. The strategic questions are who innovates
radically, whether radical innovation pays, and how a firm can keep exploiting today's
business while exploring tomorrow's.

The intuition that large incumbents cannot innovate radically turns out to be
overstated. @chandy2000 show, against the "incumbent's curse" assumption, that
incumbents and large firms have introduced a substantial share of radical product
innovations, especially in some eras and industries; the curse is contingent, not a
law. Innovation's payoff is also two-dimensional: it affects both the *level* of firm
value and its *risk*, and radical and incremental innovations differ in how they move
each, so a complete strategic assessment of an innovation portfolio must weigh return
against the volatility it introduces.

The deeper organizational problem is *ambidexterity*: the firm must simultaneously
*exploit* existing competencies (refining today's products, serving today's
customers) and *explore* new ones (radical innovation, new markets), even though the
two demand opposed structures, cultures, and metrics. Exploitation rewards
efficiency, control, and incrementalism; exploration rewards slack, autonomy, and
tolerance of failure. The dynamic-capabilities view [@teece1997; @day2011] frames
ambidexterity as the higher-order capability of reconfiguring the resource base, and
it connects growth strategy back to the chapter's spine: the firms that sustain
advantage are those whose capabilities let them keep changing what they are good at.

## The Frontier and a Research Agenda {#sec-ms-frontier}

The field's own assessments of where it stands converge on a few themes. @varadarajan2010
defines the domain and its foundational premises and, in doing so, exposes how
fragmented the field's empirical base is. @morgan2019research survey the state of
marketing-strategy research and argue that the field has over-studied isolated
marketing-mix decisions and under-studied *integrated strategy*, the pattern that the
opening of this chapter insists is the real object; they call for more work on
strategy *content* (the substance of what firms decide), strategy *process* (how
decisions get made and implemented), and strategy *implementation*, and for designs
that take causal identification seriously.

Three further fronts are active. First, *empirical generalizations*: meta-analyses
now pin down replication-grade facts, advertising elasticities [@sethuraman2011],
the market-orientation to performance link [@kirca2005], marketing's elasticity on
firm value [@edeling2016], so that strategy debates can be settled with accumulated
evidence rather than single studies. Second, *integrating logics*: service-dominant
logic reframes value as co-created with customers rather than embedded in products
[@vargo2004], a lens that, depending on one's view, either adds explanatory power or
relabels existing constructs. Third, the *emerging modules* the field is racing to
absorb: digital and platform strategy, privacy as strategy, AI as both a marketing
capability and a competitive disruptor, ESG and firm value, and causal machine
learning for heterogeneous strategic effects. Each follows the same template this
chapter has modeled, a substantive construct, an estimator, and an explicit set of
identifying assumptions.

The through-line for a doctoral reader is that marketing strategy is where the field
is held accountable. Its constructs are interesting, but they earn their place only
when forced into contact with outcome data under a credible design. The two
simulations in this chapter, market orientation moderated by competition, and a
marketing asset priced into firm value, are deliberately small instances of that
discipline: each states a data-generating process, then shows what an honest
estimator can and cannot recover from it.

## Key Takeaways

- Marketing strategy is the firm-level, integrated pattern of choices through which
  a firm creates, captures, and sustains customer value, judged by customer,
  financial, stock-market, and societal performance [@varadarajan2010; @morgan2018].
  Strategy is distinguished from tactics by reversibility and scope, not by
  importance.
- *Market orientation* is the foundational implementation construct, conceived
  behaviorally as intelligence generation, dissemination, and responsiveness
  [@kohli1990] and culturally as customer, competitor, and interfunctional
  orientation [@narver1990]. It pays on average but is necessary, not sufficient,
  and its payoff is moderated by competition and context [@kirca2005].
- Sustainable advantage rests on VRIN resources [@wernerfelt1984] and the
  evolutionary, disequilibrium competition of resource-advantage theory [@hunt1995];
  marketing capabilities [@day1994; @vorhies2005] convert orientation into position
  and outweigh R&D and operations capabilities in their performance impact
  [@krasnikov2008], while dynamic capabilities [@teece1997; @day2011] keep advantage
  alive under change.
- Positioning (STP and differentiation) must separate the *sources*, *positions*,
  and *performance outcomes* of advantage [@daywensley1988], and a defensible
  position is anchored to a capability rivals cannot quickly match.
- The marketing-finance interface is the accountability turn: marketing assets are
  off-balance-sheet drivers of firm value [@Srivastava_1998; @srinivasan2009], and
  credibly identified panels price brand and customer equity into Tobin's $q$
  [@edeling2016; @kumar2009expanding]. Cross-references: @sec-marketing-finance,
  @sec-brand-equity, @sec-clv.
- Marketing's organizational influence depends on accountability, innovativeness, and
  customer connection [@verhoef2009influence]; marketing excellence rests on four
  elements and seven activities [@moorman2016]; and the CMO's value, a roughly 15%
  Tobin's $q$ premium, is a finding only because its design confronts endogeneity
  [@germann2015].
- The agenda [@morgan2019research; @morgan2012performance] is to study *integrated*
  strategy rather than isolated tactics, to accumulate empirical generalizations
  [@sethuraman2011], and to absorb digital, platform, privacy, and AI strategy under
  the same discipline of construct, estimator, and identifying assumption.
