# Marketing Strategy {#sec-marketing-strategy}

Marketing strategy is the part of the field that asks how a firm wins, not how a
consumer chooses. Where most of the book studies decisions one customer at a time,
this chapter steps up a level of analysis to the firm, and asks how the firm
creates value for customers, captures part of that value for itself, and sustains
the arrangement against rivals who would rather it did not. Marketing strategy is
the integrated pattern of decisions through which a firm does this: which markets
to serve, which customers to win and keep, what to be known for, where to invest
the marketing budget, and how to organize the people who execute it. The unifying
test is performance. A marketing strategy is judged not by how clever it sounds
but by the customer, financial, stock-market, and societal outcomes it produces
[@varadarajan2010; @morgan2018].

This chapter is a teaching treatment of the field, complementing the doctoral
reading map in @sec-marketing-strategy-seminar. The seminar chapter is organized
as a fourteen-week syllabus that pairs each topic with its identification
challenge; this chapter develops the substantive theory in continuous prose, with
two small runnable demonstrations. Two commitments run through both. The first is
*integration*: a strategy is a coherent pattern, not a list of independent
marketing-mix moves, even though most empirical work has studied the parts rather
than the pattern [@morgan2018]. The second is *accountability*: the soft-sounding
constructs of strategy, orientation, capability, culture, relationships, are
repeatedly forced into contact with hard outcome data, because that contact is
what turns a plausible story into a finding.

## What Marketing Strategy Is {#sec-ms-domain}

It helps to fix vocabulary first, because the field uses two phrases that are
easily confused. *Strategic marketing* names the scholarly field: the study of how
marketing contributes to firm strategy and performance. *Marketing strategy* names
its object: the integrated pattern of choices a particular firm makes
[@varadarajan2010]. A doctoral student needs both senses, but the second is the
one with managerial bite.

### Levels of strategy

Strategy is layered, and marketing sits at more than one layer. *Corporate
strategy* decides which businesses the firm should be in. *Business strategy* (or
competitive strategy) decides how a given business unit competes in its chosen
market. *Functional strategy*, including marketing strategy in its narrow sense,
decides how each function supports the business-level choice. Marketing is unusual
among the functions in that it spans all three levels: it informs corporate
choices about which markets have attractive customers, it largely *is* the
business-level question of how to compete for those customers, and it supplies the
functional programs (the marketing mix) that execute the plan. This is why
marketing's claim to a seat at the strategy table is contested but persistent: the
function's natural object, the customer and the market, is exactly the object that
business-level strategy must get right. @fig-strat-levels sketches the three levels
and marketing's span across them.

```{mermaid}
%%| label: fig-strat-levels
%%| fig-cap: "The levels of strategy and marketing's span. Corporate strategy decides which businesses the firm should be in, business strategy decides how a unit competes in its chosen market, and functional strategies support the business-level choice; marketing is unusual among the functions in operating at all three levels."
flowchart TB
  CS["Corporate strategy<br/>which businesses<br/>to be in"]
  BS["Business strategy<br/>how a business unit competes<br/>in its chosen market"]
  FS["Functional strategy<br/>how each function supports<br/>the business-level choice"]
  CS --> BS
  BS --> FS
  MK["Marketing"]
  MK -.->|"informs which markets have<br/>attractive customers"| CS
  MK -.->|"largely is the question of<br/>how to compete for customers"| BS
  MK -.->|"supplies the marketing mix<br/>that executes the plan"| FS
```

### Strategy versus tactics

The cleanest way to separate strategy from tactics is by what each commits the
firm to and for how long. *Strategy* is the small set of choices that are costly to
reverse and that constrain everything downstream: which segment to target, what
position to own, which capabilities to build. *Tactics* are the many adjustable
moves, this quarter's price, this campaign's copy, this promotion's depth, that
execute a strategy and can be changed without changing the firm's direction. The
distinction is not a hierarchy of importance; a brilliant strategy with sloppy
tactics fails, and disciplined tactics cannot rescue an incoherent strategy. It is
a distinction of *reversibility and scope*. A useful diagnostic question is: if
this decision turns out wrong, how long and how expensive is the path back? Choices
with long, expensive paths back are strategic and deserve the deliberation the rest
of this chapter describes.

### The dependent variable

Because the field is defined by accountability, its recurring dependent variable is
*firm performance*, and that variable has conceptual layers worth separating
[@katsikeas2016; @morgan2012performance]. Closest to marketing actions are
*customer-mindset* outcomes (awareness, satisfaction, loyalty intentions). These
feed *marketplace* outcomes (market share, sales, customer retention), which feed
*accounting* outcomes (margins, cash flow, return on assets), which the capital
market prices into *financial-market* outcomes (stock returns, firm value, risk).
Much of the field's progress has come from tracing this chain link by link, and
much of its difficulty comes from the fact that causation can run backward along it
(profitable firms can afford more marketing) and that omitted firm-quality
variables can drive several links at once. @fig-strat-performance-chain draws the
chain alongside both threats to reading it causally.

```{mermaid}
%%| label: fig-strat-performance-chain
%%| fig-cap: "The layered dependent variable of marketing strategy. Customer-mindset outcomes feed marketplace outcomes, which feed accounting outcomes, which the capital market prices into financial-market outcomes; causation can also run backward along the chain, and omitted firm quality can drive several links at once."
flowchart LR
  CM["Customer-mindset outcomes<br/>awareness, satisfaction,<br/>loyalty intentions"]
  MP["Marketplace outcomes<br/>market share, sales,<br/>customer retention"]
  AC["Accounting outcomes<br/>margins, cash flow,<br/>return on assets"]
  FN["Financial-market outcomes<br/>stock returns, firm value,<br/>risk"]
  CM --> MP
  MP --> AC
  AC --> FN
  AC -.->|"profitable firms can<br/>afford more marketing"| CM
  FQ["Omitted firm quality"]
  FQ -.-> CM
  FQ -.-> AC
  FQ -.-> FN
```

## Market Orientation {#sec-ms-market-orientation}

If marketing strategy has a single foundational construct on the implementation
side, it is *market orientation*: the degree to which a firm actually organizes
itself around its customers and competitors rather than merely professing to. Two
papers in 1990 defined the construct in complementary ways, and the field has spent
three decades measuring it and testing whether it pays.

### The behavioral view and the cultural view

@kohli1990 give the *behavioral* conceptualization. They define market orientation
as "the organizationwide generation of market intelligence pertaining to current
and future customer needs, dissemination of the intelligence across departments,
and responsiveness to it." The achievement of this definition is that it
decomposes a vague cultural posture into three *observable behaviors*, intelligence
generation, dissemination, and responsiveness, and thereby makes the construct
measurable (the MARKOR scale follows directly from it). Profitability is
deliberately placed as a *consequence* of orientation, not a part of it, which
keeps the construct clean enough to serve as a regressor against performance. The
seminar chapter develops this decomposition in detail in
@sec-strategy-market-orientation.

@narver1990 give the *cultural* conceptualization. They define market orientation
as the organizational culture that most effectively creates the behaviors needed
for superior value, and they decompose it into three components, customer
orientation, competitor orientation, and interfunctional coordination, plus two
decision criteria (long-term focus and profitability). Their MKTOR scale
operationalizes this view, and their paper supplied the first direct test linking
orientation to business profitability.

The two views are less rival than layered: culture (Narver and Slater) is the
disposition, and behavior (Kohli and Jaworski) is its enactment. A firm can hold
customer-oriented values yet fail to generate or act on intelligence; the
behavioral view measures whether the disposition actually shows up in routines.

### Does market orientation pay, and when?

The headline empirical question is whether more market-oriented firms perform
better, and the meta-analytic verdict is a qualified yes. @kirca2005 synthesize the
accumulated studies and find a positive, robust association between market
orientation and performance, but one whose size depends on the setting. The link is
stronger when performance is measured by managerial judgment than by objective
financials, stronger in studies using cost-based rather than revenue-based
outcomes, and moderated by national culture and industry. The honest reading is
that market orientation is *necessary but not sufficient*: it is a precondition for
converting market knowledge into value, but its payoff is contingent on the
capabilities (next section) that turn knowledge into action and on the competitive
context that determines how much customer focus is rewarded. A second meta-analysis
[@jaramillo2004metarelations] reaches a compatible verdict on the salesperson-level
version of the same relationship, which matters because it locates part of the
firm-level effect in individual behaviour rather than in structure.

Two qualifications are worth carrying forward. The first is that market orientation
is one strategic orientation among several — technology, entrepreneurial, and
learning orientations compete for the same organizational attention — and
@noble2002marketorientati find that the alternatives sometimes dominate depending on
the environment, so a firm maximizing market orientation alone may be optimizing the
wrong variable. The second is sharper: @jones2011marketorientati question whether
market orientation can be a source of *sustainable* competitive advantage at all,
given that it is observable, imitable, and widely taught. The construct may be a
threshold condition — costly to lack, insufficient to win on.

This sets up the field's signature identification worry. A cross-sectional
correlation between orientation and profit is consistent with orientation causing
profit, with profit funding orientation, or with some omitted quality (good
management, slack resources) causing both. Two methodological papers are worth
reading alongside any such regression: @ganesan2008crosssectional set out what a
cross-section can and cannot establish relative to a panel in exactly this
literature, and @ghosh2018endogeneisurvey treat endogeneity in survey research
directly, including the instrument-free approaches available when no credible
instrument exists. The demonstration below uses simulated data to show what an
honest analyst can and cannot conclude from such a regression, and how a moderator
changes the picture.

### A simulated market-orientation to performance regression

The following chunk simulates a cross-section of firms, generates performance as a
function of market orientation *and* a moderator (competitive intensity), and then
recovers the relationship. The data-generating process is known, so we can see
exactly what the regression is estimating. The substantive point is that the
return to market orientation is larger where competition is fiercer: customer focus
matters most when customers have alternatives.

```{r ms-mo-regression, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({library(dplyr); library(ggplot2)})
set.seed(2024)

n <- 600

firms <- tibble(
  # market orientation (standardized index, e.g. MARKOR/MKTOR)
  mo        = rnorm(n, 0, 1),
  # competitive intensity moderator (0 = sheltered, 1 = cut-throat)
  intensity = runif(n, 0, 1),
  # an omitted "firm quality" that raises BOTH mo and performance
  quality   = rnorm(n, 0, 1)
)

# true model: MO helps, and helps MORE under high competitive intensity;
# firm quality confounds the naive association.
firms <- firms %>%
  mutate(
    mo_obs = mo + 0.5 * quality,                       # quality inflates MO
    perf   = 0.30 * mo_obs +                           # baseline MO effect
             0.60 * mo_obs * intensity +               # moderation
             0.80 * quality +                          # confound
             rnorm(n, 0, 1)
  )

# (1) naive regression: omits quality -> MO effect is biased upward
naive <- lm(perf ~ mo_obs, data = firms)

# (2) moderated regression controlling for quality -> recovers structure
moderated <- lm(perf ~ mo_obs * intensity + quality, data = firms)

round(rbind(
  naive     = c(coef(naive)["mo_obs"], NA, NA),
  moderated = coef(moderated)[c("mo_obs", "intensity", "mo_obs:intensity")]
), 3)
```

The naive slope on market orientation is inflated because it absorbs the omitted
firm-quality effect; the moderated specification, once quality is controlled,
recovers a positive main effect and a positive interaction, the return to market
orientation rises with competitive intensity. @fig-ms-mo-moderation plots the fitted
performance-orientation slope at low and high competitive intensity to make the
moderation visible.

```{r}
#| label: fig-ms-mo-moderation
#| fig-cap: "Simulated moderation of the market-orientation to performance link by competitive intensity. The fitted slope of performance on market orientation is steeper under high competitive intensity (customers have alternatives) than under low intensity, illustrating that market orientation is necessary but its payoff is contingent [@kirca2005]."
#| message: false
#| warning: false
grid <- expand.grid(
  mo_obs    = seq(-2.5, 2.5, length.out = 50),
  intensity = c(0.15, 0.85),
  quality   = 0
)
grid$perf <- predict(moderated, newdata = grid)
grid$Intensity <- factor(grid$intensity,
                         labels = c("Low competitive intensity",
                                    "High competitive intensity"))

ggplot(grid, aes(mo_obs, perf, colour = Intensity)) +
  geom_line(linewidth = 1.1) +
  labs(x = "Market orientation (standardized)",
       y = "Predicted firm performance",
       colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

The lesson generalizes well beyond this toy: a positive average effect can hide
substantial heterogeneity, and the moderators (here, competition; in the
literature, also market turbulence and national culture) are often the
managerially interesting part of the answer.

## Resources, Capabilities, and Sustainable Advantage {#sec-ms-rbv}

Market orientation tells the firm to listen to the market; it does not, by itself,
explain why some firms convert what they hear into durable advantage and others do
not. For that the field reaches to theories of competitive advantage, and the
dominant one is the resource-based view (RBV).

### The resource-based view and resource-advantage theory

The RBV holds that sustained advantage comes from resources that are valuable,
rare, imperfectly imitable, and non-substitutable (the VRIN tests), and that are
embedded in an organization able to exploit them [@wernerfelt1984]. The insight is
that resources easily bought or copied cannot support advantage for long, because
competitors acquire or replicate them; advantage must rest on resources that resist
the market for resources itself. Marketing's contribution to the RBV is the
observation that its most important assets, brand equity, customer relationships,
channel ties, market knowledge, are precisely the off-balance-sheet, relationship-
and knowledge-based resources that best pass the VRIN tests, a point developed as
the *market-based assets* framework [@Srivastava_1998] and treated at length in
@sec-marketing-finance.

Marketing has also produced its own theory of competition. @hunt1995 propose
*resource-advantage theory* (originally the comparative-advantage theory of
competition), an evolutionary, disequilibrium alternative to the equilibrium models
of industrial-organization economics. In their account, firms compete by seeking
*comparative advantages in resources* that yield marketplace positions of
competitive advantage and, in turn, superior financial performance; competition is
the ongoing, never-settled process of firms trying to neutralize or leapfrog one
another's resource advantages. The contrast with equilibrium IO is the heart of a
recurring seminar debate: are marketing resources truly inimitable sources of
lasting rents, or transient advantages constantly eroded by imitation? R-A theory's
answer is that the *process* of seeking advantage, not any particular static
advantage, is what marketing strategy is about.

### Marketing capabilities and dynamic capabilities

Resources are inert without the routines that deploy them, and those routines are
*capabilities*. @day1994 gives the canonical typology of the capabilities of
market-driven organizations: *outside-in* capabilities (market sensing, customer
linking, channel bonding) that connect the firm to its environment; *inside-out*
capabilities (manufacturing, logistics, financial management) that are activated by
market needs; and *spanning* capabilities (new-product development, pricing,
purchasing, service delivery) that integrate the two. A market-driven firm is one
whose outside-in capabilities are strong enough to guide the rest.

@vorhies2005 turn this typology into measurement, showing that distinct marketing
capabilities (in pricing, product development, channel management, marketing
communications, selling, market information management, and marketing planning and
implementation) can be benchmarked and that firms whose capabilities exceed
benchmarks perform better. @krasnikov2008 then run the horse race that matters for
resource allocation: in a meta-analysis comparing marketing, R&D, and operations
capabilities, marketing capabilities have the *largest* relative impact on firm
performance, a finding that pushes back on the engineering-centric view that R&D and
operations are where advantage is built. Crucially, market orientation and
marketing capabilities are *complements*, not substitutes: orientation supplies the
knowledge that capabilities convert into market position [@morgan2009].

The frontier worry is that static capabilities decay as markets change. *Dynamic
capabilities*, the firm's capacity to sense opportunities, seize them, and
reconfigure its resource base, address strategy under change [@teece1997]. @day2011
brings this into marketing with the *marketing capabilities gap*: the routines most
firms have were built for a stable, push-marketing world and lag the adaptive,
outside-in capabilities (vigilant market learning, adaptive experimentation, open
networking) that turbulent digital markets reward. The managerial implication is
that capability is a moving target; the resource-based test for *sustainability*
must be re-run as the environment shifts.

## Competitive Strategy and Positioning {#sec-ms-positioning}

Having decided to compete, the firm must decide *how*. The classic marketing answer
is the *segmentation, targeting, and positioning* (STP) sequence. *Segmentation*
partitions a heterogeneous market into groups of customers with similar needs and
responses. *Targeting* selects which segments the firm's resources and capabilities
let it serve profitably and defensibly. *Positioning* designs and communicates a
distinctive place in the minds of target customers, the position the firm intends
to own.

Underlying STP is *differentiation*: the firm must give target customers a reason to
prefer it that rivals cannot easily match. Differentiation can rest on the product,
the brand, the service, the channel, or the relationship, but the strategic test is
always the same VRIN test from the RBV, a basis of differentiation confers durable
advantage only if it is valuable to customers and costly for rivals to imitate. This
is why @daywensley1988 insist on separating the *sources* of advantage (superior
skills and resources) from the *positions* of advantage they produce (superior
customer value or lower relative cost) from the *performance outcomes* that follow
(satisfaction, loyalty, share, profitability). Confusing the three is the most
common analytic error in positioning research: a firm with high share (a position)
is not thereby shown to have a superior capability (a source), because the causal
arrow can run either way. @fig-strat-advantage separates the three layers.

```{mermaid}
%%| label: fig-strat-advantage
%%| fig-cap: "Sources, positions, and performance outcomes of competitive advantage. Superior skills and resources produce positions of superior customer value or lower relative cost, which produce performance outcomes; the three layers must be kept distinct because the causal arrow between position and source can run either way."
flowchart LR
  SO["Sources of advantage<br/>superior skills<br/>and resources"]
  PO["Positions of advantage<br/>superior customer value<br/>or lower relative cost"]
  PF["Performance outcomes<br/>satisfaction, loyalty,<br/>share, profitability"]
  SO --> PO
  PO --> PF
  PO -.->|"the causal arrow can<br/>run either way"| SO
```

*Competitive dynamics* adds the recognition that positions are contested in real
time. Rivals respond to attacks, match price cuts, copy features, and counter-
position, so a position's value depends on how rivals will react, not only on how
customers will. The strategic question is therefore not merely "what position do we
want?" but "what position can we hold given how competitors will respond?", which is
where R-A theory's disequilibrium view [@hunt1995] and the dynamic-capabilities view
[@teece1997; @day2011] re-enter: a defensible position is one anchored to a
capability rivals cannot quickly reconfigure to match.

## The Marketing-Finance Interface: The Accountability Turn {#sec-ms-finance}

The most consequential development in marketing strategy over the last quarter
century is the *accountability turn*: the insistence that marketing demonstrate its
value in the language the boardroom and the capital market actually use, cash flow,
firm value, and risk. This is the subject of its own chapter
(@sec-marketing-finance); here we give the strategic logic and a panel
demonstration.

The turn was partly a response to an internal criticism worth reading in the
original. @reibstein2009guesteditorial argued that marketing academia had drifted
toward problems that were tractable rather than important, and that the field's
declining influence on practice was a consequence rather than an accident. It is a
useful counterweight to this chapter's methodological emphasis: identification
discipline is necessary, but a well-identified answer to a question no one is asking
is not a contribution.

### Marketing and shareholder value

The hinge is the *market-based assets* framework [@Srivastava_1998], which reframes
marketing's outputs, brand equity, customer equity, channel relationships, as
intangible assets that drive the firm's future cash flows and therefore its value.
Under an efficient capital market, actions that build value with customers should
build value for shareholders, because a firm is worth the risk-adjusted present
value of its expected future cash flows and marketing assets shift those flows in
four ways: they can *accelerate* cash flows, *augment* their level, *reduce* their
volatility and vulnerability (lowering the discount rate), and raise their
*residual* (terminal) value [@Srivastava_1998].

@srinivasan2009 supply the canonical roadmap of how to measure these links: stock-
return response models, persistence and vector-autoregression models, and Tobin's
$q$ regressions, together with the identification toolkit (event studies, factor
models, panel methods) the field uses to defend causal claims. The meta-analytic
payoff is that marketing variables, brand, customer satisfaction, R&D, advertising,
have economically meaningful, positive elasticities on firm value
[@edeling2016; @edeling2021]. Brand equity (@sec-brand-equity) and customer equity
(@sec-clv) are the two strategic assets this literature prices most heavily:
strong-brand portfolios earn higher returns at lower risk, and the aggregate
lifetime value of the customer base tracks, and helps explain, market
capitalization [@kumar2009expanding].

### A simulated marketing-assets to firm-value panel

The following chunk illustrates, on simulated panel data, the workhorse Tobin's $q$
regression of the interface: firm value regressed on a marketing asset (here a brand-
equity index) with firm and year fixed effects and the standard financial controls
(leverage and cash flow). Fixed effects absorb time-invariant firm quality, so the
estimate comes from *within-firm* variation in the asset over time, the design that
makes the marketing-to-value claim credible rather than merely correlational. This
reuses the kind of machinery developed in @sec-marketing-finance.

```{r ms-firm-value-panel, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({library(dplyr)})
set.seed(7)

n_firm <- 120; n_year <- 8
firm_fe <- rnorm(n_firm, 0, 0.40)          # time-invariant firm quality
year_fe <- rnorm(n_year, 0, 0.10)

panel <- expand.grid(firm = 1:n_firm, year = 1:n_year) %>%
  mutate(
    brand_equity = 0.6 * firm_fe[firm] + rnorm(n(), 0, 1),  # quality raises BE
    leverage     = runif(n(), 0, 0.6),
    cash_flow    = rnorm(n(), 0, 1),
    # true within-firm effect of brand equity on Tobin's q is 0.25
    tobin_q = 1.2 + firm_fe[firm] + year_fe[year] +
              0.25 * brand_equity -        # the strategic effect of interest
              0.40 * leverage +
              0.15 * cash_flow +
              rnorm(n(), 0, 0.30)
  )

# Pooled OLS (no FE) overstates the effect: it confounds within- and between-firm
pooled <- lm(tobin_q ~ brand_equity + leverage + cash_flow, data = panel)

# Two-way fixed-effects: identify from WITHIN-firm changes in brand equity
fe <- lm(tobin_q ~ brand_equity + leverage + cash_flow +
           factor(firm) + factor(year), data = panel)

round(c(pooled_BE = coef(pooled)["brand_equity"],
        FE_BE      = coef(fe)["brand_equity"],
        truth      = 0.25), 3)
```

The pooled estimate is biased upward because it mixes the genuine within-firm effect
with the cross-firm correlation that firm quality induces between brand equity and
$q$; the two-way fixed-effects estimate, which uses only within-firm variation,
recovers the true 0.25 elasticity. This is the strategic content of the interface in
one line: marketing builds an asset, and a credibly identified panel shows that the
asset is priced into the value of the firm. The unresolved debate is whether
marketing *creates* that value or merely *signals* pre-existing quality, and whether
fixed effects and instruments fully purge the reverse-causality and omitted-quality
threats [@srinivasan2009; @edeling2021].

## Marketing Organization, the CMO, and Accountability {#sec-ms-organization}

Strategy is executed by an organization, and how marketing is organized, and how
much influence it wields, is itself a strategic variable. The marketing function's
influence inside the firm rose with the marketing concept and has been contested
ever since, with periodic claims that it is declining as analytics, finance, and
product functions absorb customer responsibility.

@verhoef2009influence ask what actually drives the marketing department's clout and
find that it rests on *accountability* (the ability to demonstrate marketing's
financial contribution), *innovativeness*, and a credible connection to the
customer, not on budget size or headcount. The implication is pointed: the
accountability turn is not only an intellectual movement but the practical price of
marketing's seat at the strategy table. Departments that cannot speak the language
of firm value lose influence to those that can.

@moorman2016 give the modern integrative framework, defining *marketing excellence*
as "a superior ability to perform essential customer-facing activities that improve
customer, financial, stock market, and societal outcomes" and resting it on four
organizational elements (capabilities, configuration, human capital, and culture)
mobilized through seven activities. This MARORG framework is developed as a worked
capability theory in @sec-strategy-excellence.

The sharpest test of marketing's organizational value is the *chief marketing
officer*. Does having a CMO in the top management team improve performance? The
natural design compares firms with and without a CMO [@nath2008], and the cleanest
estimate finds firms with a CMO exhibit roughly 15% higher Tobin's $q$
[@germann2015]. But CMO presence is not randomly assigned, so the credible design
combines firm fixed effects (identifying from within-firm changes in CMO presence)
with instrumental variables (whose exclusion restriction is the untestable
assumption on which the causal claim rests). The seminar chapter develops this as a
worked identification example in @sec-strategy-cmo, and it is the template for the
whole field: a strategic construct becomes a *finding* only when its estimator and
identifying assumptions are stated in full.

Underpinning all of this is *measurement*. @katsikeas2016 map the space of marketing
performance outcomes (customer-mindset, marketplace, accounting, financial-market),
and the recurring empirical result is that the *ability to measure marketing* is
itself associated with performance and with executives' confidence in the function.
Whether measurement causes performance or better firms simply measure more is the
same endogeneity question in a new guise.

## Innovation and Growth Strategy {#sec-ms-innovation}

Growth is the strategic objective that most directly tests a firm's resource base,
and *innovation* is its primary engine. The strategic questions are who innovates
radically, whether radical innovation pays, and how a firm can keep exploiting today's
business while exploring tomorrow's.

The intuition that large incumbents cannot innovate radically turns out to be
overstated. @chandy2000 show, against the "incumbent's curse" assumption, that
incumbents and large firms have introduced a substantial share of radical product
innovations, especially in some eras and industries; the curse is contingent, not a
law. Innovation's payoff is also two-dimensional: it affects both the *level* of firm
value and its *risk*, and radical and incremental innovations differ in how they move
each, so a complete strategic assessment of an innovation portfolio must weigh return
against the volatility it introduces.

The deeper organizational problem is *ambidexterity*: the firm must simultaneously
*exploit* existing competencies (refining today's products, serving today's
customers) and *explore* new ones (radical innovation, new markets), even though the
two demand opposed structures, cultures, and metrics. Exploitation rewards
efficiency, control, and incrementalism; exploration rewards slack, autonomy, and
tolerance of failure. The dynamic-capabilities view [@teece1997; @day2011] frames
ambidexterity as the higher-order capability of reconfiguring the resource base, and
it connects growth strategy back to the chapter's spine: the firms that sustain
advantage are those whose capabilities let them keep changing what they are good at.

## Corporate Development: Acquisitions, Alliances, and Divestitures {#sec-ms-corpdev}

The chapter has so far treated the resource base as something the firm *builds*:
market orientation generates knowledge, capabilities convert it, dynamic
capabilities reconfigure it. But firms also *buy* and *sell* resources outright.
Corporate development, the acquisition, alliance, and divestiture decisions through
which a firm changes what it owns, is where the resource-based view of
@sec-ms-rbv confronts an actual market for resources, and it is where marketing
strategy's claims are tested at their most expensive. A single acquisition can
commit more capital than a decade of advertising, and by the reversibility
criterion of @sec-ms-domain it is the most strategic decision a firm makes. This
section develops the strategy-side logic of that decision; @sec-marketing-finance
treats the same events through the valuation machinery of the event study, and the
two should be read together.

The organizing claim is that corporate development is best understood as a *chain*
rather than a decision. A firm first chooses a mode of resource combination, then
forms a motive for the particular deal, then designs an integration that either
realizes or destroys the motive's premise, all under an ownership and stakeholder
structure that determines who captures whatever value results. Most of the
literature's apparent contradictions dissolve once studies are located on this
chain, because a variable that matters at one link is noise at another.
@fig-ms-cd-chain lays out the sequence and the moderators that attach to each link.

```{mermaid}
%%| label: fig-ms-cd-chain
%%| fig-cap: "Corporate development as a chain. The firm chooses a mode of resource combination, forms a motive for the particular deal, and designs an integration that realizes or destroys the motive's premise; ownership structure and stakeholder relationships determine who captures the resulting value, and the observable outcome is a distribution of performance rather than a single average effect."
flowchart TB
  MODE["Mode choice<br/>build, ally, buy,<br/>or sell"]
  MOTIVE["Deal motive<br/>consolidation, growth,<br/>capability, or option"]
  INTEG["Integration design<br/>degree, speed,<br/>autonomy"]
  DIST["Distribution of<br/>performance<br/>mean and spread"]
  MODE --> MOTIVE
  MOTIVE --> INTEG
  INTEG --> DIST
  REL["Resource relatedness<br/>and prior experience"]
  ATT["Managerial attention<br/>to candidate synergies"]
  CULT["Cultural fit and<br/>employee resistance"]
  OWN["Ownership structure<br/>and stakeholder ties"]
  REL -.-> MODE
  ATT -.-> MOTIVE
  CULT -.-> INTEG
  OWN -.-> DIST
```

### The mode choice: build, ally, buy, or sell

The first link is the most often assumed away. Empirical work typically studies
acquisitions alone, or alliances alone, and estimates what predicts the observed
mode; but a firm that acquires has by construction declined to ally, to build
internally, and to divest, so any determinant of acquisition estimated in isolation
is contaminated by the unmodeled availability of the alternatives.
@villalonga2005choice make this the object of study rather than a nuisance,
modeling the choice among acquisitions, alliances, and divestitures jointly. Their
finding is that the modes are governed by a common logic of resource fit and
experience: firms select the mode whose governance costs are lowest given the
relatedness of the resources involved and their own accumulated experience with
that mode. The methodological lesson generalizes far beyond corporate development,
and is the same lesson the CMO literature learned in @sec-ms-organization: a
strategic choice is a *selection*, and estimating its consequences requires
modeling why it was selected.

@wang2007alliance sharpen this by insisting the choice is a property of the *pair*,
not of the acquirer. In their dyadic account, alliance and acquisition are
alternative solutions to the same problem of combining two firms' resources, and
which solution appears depends on both parties' resource profiles and on the
relative bargaining position each brings. A resource combination that looks like an
acquisition from the acquirer's balance sheet may have been an alliance from the
target's perspective until the target's outside options thinned. This is the
combination-mode analogue of the sources-versus-positions distinction drawn by
@daywensley1988, in that the observed mode is a *position*, and inferring the
underlying resource logic from it requires care.

@capron2007private add the informational dimension. Acquirers of *private* targets
face far greater information asymmetry than acquirers of public ones, and the
asymmetry cuts both ways: it raises the risk of overpaying for an unobservable
lemon, but it also means that a bidder with privileged private information can buy
an asset the rest of the market cannot price. Their finding, that acquirers with
prior experience and relevant knowledge select private targets and earn higher
returns from them, is the mode-choice link's version of the resource-based
prediction, an advantage in *evaluating* a resource is itself a resource.

### Does acquisition pay? From the average to the distribution

The second question is the one the field has argued about longest, and the answer
has changed shape. The early accumulation of event studies produced a rough
consensus, catalogued in @tbl-mf-mna of @sec-marketing-finance, that targets gain
and acquirers roughly break even, with relatedness usually but not always
associated with larger gains. @datta1992 formalize that accumulation in a
meta-analysis of the factors influencing wealth creation, finding systematic
effects for deal characteristics but far more unexplained variance than a
well-behaved literature should have.

@barney1988relatedness supplies the theoretical reason the variance is there, and
it is the single most important idea in this section. Relatedness *cannot* by
itself produce bidder gains, because if a related combination generates synergy
that any related bidder could generate, then competition among bidders in the
market for corporate control transfers the entire synergy to the target's
shareholders through the acquisition premium. Bidders earn returns only when their
synergy with a particular target is *unique*, not merely related, and privately
known. This is the VRIN logic of @sec-ms-rbv applied to the market for firms, and
it reframes the empirical question: the interesting variable is not whether the
firms are similar but whether the acquirer's synergy is inimitable and unpriced.
It also explains why the similarity-versus-complementarity debate has never
resolved, since both similarity and complementarity can be either common knowledge
or private, and only the private version pays.

@king2004meta bring the accumulated evidence to its uncomfortable conclusion.
Meta-analyzing post-acquisition performance across the standard moderators,
relatedness, method of payment, prior acquisition experience, they find no
significant positive average effect on acquirer performance, and, more tellingly,
residual variance too large for the identified moderators to explain. Their title
names the diagnosis: the moderators that matter have not been identified. The
review literature has since organized what remains unexplained rather than
resolving it, with @haleblian2009 mapping the antecedents and consequences of
acquisition activity, and @devers2020behavioral charting the behavioral turn toward
the perceptions, emotions, and cognitive limits of the executives who actually make
these decisions.

@rabier2017 offers the most productive response, and it is a response about the
shape of the outcome rather than its mean. Classifying deals by motive, she shows
that acquisition motives shift not only the *average* of acquisition performance
but its entire *distribution*: deals motivated by synergy seeking exhibit both
higher upside and heavier downside than deals motivated by consolidation or
efficiency. A literature that reports means will therefore find nothing while a
genuine and managerially crucial effect sits in the second and higher moments. The
simulation below makes the mechanism concrete, because it is easy to state and easy
to miss.

```{r ms-cd-motive-sim, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({library(dplyr); library(ggplot2)})
set.seed(1988)

# A realistically sized acquisition sample (event studies rarely exceed
# a few hundred usable deals).
n <- 600

# Two motive populations, pooled in the data an analyst actually observes.
# Consolidation deals: modest but reliable cost synergies (small mean, tight).
# Synergy-seeking deals: uncertain growth synergies (no better on average,
# far wider spread) -- the Rabier (2017) distributional claim.
deals <- tibble(
  motive = rep(c("Consolidation", "Synergy seeking"), each = n / 2),
  car    = c(rnorm(n / 2, mean = 0.5, sd = 2.5),
             rnorm(n / 2, mean = 0.0, sd = 9.0))
)

# (1) The average-effect test a mean-based literature would run
pooled_t <- t.test(car ~ motive, data = deals)

# (2) The same data described by moment and by tail
by_motive <- deals %>%
  group_by(motive) %>%
  summarise(mean = mean(car), sd = sd(car),
            p10  = unname(quantile(car, 0.10)),
            p90  = unname(quantile(car, 0.90)),
            .groups = "drop")

# (3) Who actually occupies the extremes of the POOLED distribution?
#     (composition of each tail, not each motive's own tail rate)
cut_lo <- quantile(deals$car, 0.05); cut_hi <- quantile(deals$car, 0.95)
tails <- c(
  synergy_share_of_worst_5pct = mean(deals$motive[deals$car <= cut_lo] == "Synergy seeking"),
  synergy_share_of_best_5pct  = mean(deals$motive[deals$car >= cut_hi] == "Synergy seeking")
)

list(
  mean_test    = round(c(difference = unname(-diff(pooled_t$estimate)),
                         t_stat     = unname(pooled_t$statistic),
                         p_value    = pooled_t$p.value), 3),
  by_motive    = as.data.frame(by_motive) %>% mutate(across(where(is.numeric), ~round(.x, 2))),
  tail_make_up = round(tails, 3)
)
```

The mean difference is negligible and nowhere near significance; a meta-analysis
pooling studies like this would report exactly the null that @king2004meta report.
Yet synergy-seeking deals carry more than three times the standard deviation, and
they account for *every* deal in both the worst and the best five percent of the
pooled sample. An entire literature could conclude that acquisition motive does not
matter while motive in fact determines who ends up in the tails.
@fig-ms-cd-motive-dist draws the two distributions with the pooled mean marked,
which is the picture a mean-based literature never produces.

```{r}
#| label: fig-ms-cd-motive-dist
#| fig-cap: "Simulated distributions of acquirer abnormal returns by deal motive. Consolidation deals cluster tightly around a small positive mean; synergy-seeking deals have no higher a mean but a far wider spread, supplying most of both tails. A test of average effects finds nothing, while the distributions differ profoundly, which is the mechanism behind the unidentified moderators of @king2004meta and the distributional result of @rabier2017."
#| message: false
#| warning: false
ggplot(deals, aes(car, fill = motive, colour = motive)) +
  geom_density(alpha = 0.25, linewidth = 0.9) +
  geom_vline(xintercept = mean(deals$car), linetype = "dashed") +
  coord_cartesian(xlim = c(-25, 25)) +
  labs(x = "Acquirer abnormal return (%)",
       y = "Density", fill = NULL, colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

The lesson is the distributional counterpart of the moderation lesson from
@sec-ms-market-orientation. There, a positive average hid heterogeneity in the
slope; here, a null average hides heterogeneity in the *spread*. A field that asks
only "does it pay on average?" will declare a phenomenon dead while its most
important variation is still unexamined, and the corrective is to report quantiles,
variances, and tail shares alongside means.

### Motive as the master variable

If motive governs the distribution, the natural next question is where motives come
from and whether they can be observed. Three recent contributions attack this from
different directions.

@bauer2024synergy treat pre-deal synergy evaluation as a cognitive act rather than
a valuation exercise. In their attention-based account, managers do not evaluate all
candidate synergies and select the best; they evaluate the synergies their attention
structures make salient, and the firm's existing routines, recent experiences, and
organizational identity determine what those are. The consequence is that synergy
estimates are systematically incomplete in patterned rather than random ways, which
explains both why cost synergies are chronically over-represented in deal rationales
(they are legible, quantifiable, and attributable to identifiable line items) and
why revenue and capability synergies are chronically underestimated. For a marketing
audience this is a pointed finding: the synergies marketing is best placed to
deliver, cross-selling into a combined customer base, brand-portfolio consolidation,
channel access, are precisely the ones that attention structures dominated by
finance and operations are least likely to surface.

@feldman2022synergy give the construct the theoretical treatment it had been
missing. They distinguish synergy types by *source* and trace each type's *life
cycle*, arguing that synergies differ in how quickly they can be realized, how long
they persist, and how much of their value accrues to the acquirer rather than
leaking to the target's shareholders or to customers. The life-cycle framing matters
for empirical design: a study whose window is three days after announcement measures
investor beliefs about fast-realizing synergies, while a study whose window is three
years measures realized slow synergies net of integration costs, and the two need
not agree even when both are correct.

@piezunka2026 add the observation that a firm's motives are read by third parties
who then act on their reading. Studying what happens to a target's external
collaborators after an acquisition, they show that collaboration post-acquisition
depends on the acquirer's motives: collaborators who infer that the acquirer intends
to absorb and redirect the target withdraw, while collaborators who infer a
capability-preserving motive stay. The target's relational assets, in other words,
are not transferred by the purchase agreement; they are re-negotiated with every
partner, and the acquirer's perceived intent is the term of that negotiation. This
is the corporate-development instance of a general marketing truth, that
relationships are held by counterparties, not by owners.

### Integration: degree, speed, and the marketing content of the deal

Motive is a premise; integration either delivers it or does not. The integration
literature is the largest in corporate development, and marketing has contributed
to it directly.

@homburg2005marketing bring the marketing function into the picture explicitly,
studying how *marketing integration*, the consolidation of brands, sales forces,
product lines, and customer-facing processes, affects postmerger performance. Their
result is that marketing integration matters through two conflicting channels: it
generates cost savings and market-power gains, but it also disrupts customer
relationships and internal marketing capability during the transition. The net
effect is therefore contingent rather than uniformly positive, and the contingency
is a marketing variable.

@homburg2006speed then take on the folk wisdom that faster integration is better.
The received managerial view holds that a long integration is a long period of
uncertainty, so speed limits damage. They show this is contingent on two kinds of
relatedness. Where *external* relatedness is high, the merging firms serve similar
markets and customers, so speed pays because the market-facing benefits of
combination arrive sooner and the window of customer confusion is short. Where
*internal* relatedness is high, the firms have similar internal structures and
processes, and speed *hurts*, because rapid consolidation of overlapping internal
systems destroys the tacit knowledge and routines that made each firm work. Speed is
therefore not a success factor but a decision variable whose optimum depends on
where the relatedness sits, which is why the literature that treated it as a main
effect produced conflicting results.

@bauer2014antecedents provide the integrative test, modeling strategic
complementarity, cultural fit, and the *degree* and *speed* of integration together
rather than one at a time. Their central finding is that these antecedents are not
additive: complementarity and cultural fit set the ceiling on what integration can
achieve, and degree and speed determine how much of that ceiling is reached, with
the wrong combination of high degree and poor cultural fit actively destroying
value. The cultural channel has a long lineage. @chatterjee1992culture showed that
perceived cultural differences between merging firms predict lower shareholder
value in related mergers, linking equity effects to human capital, and
@larsson1999 established that synergy realization depends jointly on combination
potential, the *extent* of integration, and the absence of employee resistance,
which is the organizational precondition that complementarity alone cannot supply.

The learning literature explains how firms get better at this. @zollo2004deliberate
show that acquisition performance improves not with raw experience but with
*deliberate learning*: firms that explicitly codify what they learned from past
integrations, into manuals, checklists, and playbooks, build an integration
capability, while firms that merely accumulate deals do not. This is a clean
instance of the dynamic-capabilities argument of @teece1997, and it carries a sharp
managerial implication: experience is only a resource once it has been articulated.

Two papers resolve what integration should actually be integrated. @puranam2006organizing
identify the coordination-autonomy dilemma in technology acquisitions: structural
integration improves coordination between acquirer and target but destroys the
target's autonomy and thereby its innovative output, and the resolution depends on
whether the target's innovation trajectory was already established at the time of
acquisition. @puranam2007leverage refine this into a distinction between what
acquirers *know* and what they *do*, showing that acquirers can leverage a target's
technology without full structural integration when the relevant knowledge can be
transferred through other channels. @devarakonda2024 extend this line with a
revealing measurement strategy, tracking post-acquisition patent reassignments to
observe where resource control actually lands inside the corporate structure, which
turns an organizational-design question into something empirically visible.

Finally, @cording2008ambiguity address the causal ambiguity that makes this whole
literature hard. Integration decisions are far from performance in the causal
chain, and the intervening steps are usually unmeasured, so estimated
integration-to-performance links are noisy and unstable. Their remedy is to model
*intermediate goals*, internal reorganization and market expansion, as mediators, so
that the causal path is broken into steps each of which is measurable. This is the
same discipline the performance chain of @fig-strat-performance-chain imposes on
marketing strategy generally, and @bauer2020marketing carry it back into marketing
by showing that marketing fit moderates how marketing-integration decisions
translate into intermediate goals such as market expansion. @graebner2017pmi review
the postmerger integration process as a whole and reach a compatible verdict: the
field has plenty of variance-model evidence about integration attributes and far
too little process evidence about how integration unfolds over time.

### Stakeholders, and whose relationships come with the deal

The most consequential recent move is to stop treating the target as a bundle of
assets and start treating it as a node in a web of relationships.
@odziemkowska2026 develop this into the concept of *stakeholder synergies*: value
that arises from combining the acquirer's and target's relationships with external
stakeholders, distinct from the operational and financial synergies the literature
has always counted. Because stakeholder relationships are relational, non-tradeable,
and slow to build, they pass the VRIN tests of @sec-ms-rbv better than most physical
assets, which makes them a plausible source of exactly the unique, inimitable
synergy that @barney1988relatedness argued is the only kind bidders can profit from.
The concept also cuts the other way: a target's stakeholder relationships can be
*damaged* by the combination, so stakeholder considerations belong on both sides of
the synergy ledger.

@bettinazzi2017 supply the acquirer-side complement, showing that firms with a
stronger stakeholder orientation realize better acquisition performance, plausibly
because the routines that sustain attention to employees, customers, and suppliers
are precisely the routines that keep those constituencies engaged through the
disruption of integration. And @umashankar2021 deliver the finding that should
concern marketing most: acquisitions tend to *reduce* customer satisfaction, because
executive attention shifts from customers to financial integration, and the
resulting dissatisfaction erodes the very synergies the deal was meant to capture,
though marketing expertise in the upper echelons mitigates the damage. Read together
with @piezunka2026, the message is consistent. The relationships that make a target
valuable are held by counterparties who can walk, and the deal itself is the event
most likely to make them consider it.

### Ownership, governance, and who sits on each side

Whether a deal creates value and whether a given shareholder captures it are
different questions, and the difference is governance.

@goranova2010 make the sharpest version of this point. Institutional investors
frequently hold stakes in *both* the acquirer and the target, and such an investor's
interest is in the combined value of the pair rather than in the acquirer's price.
Overlapping institutional ownership therefore predicts higher premiums paid, because
owners on both sides are, in effect, moving money from one pocket to another and are
indifferent to the transfer that acquirer-only shareholders resist. The finding
generalizes into a warning that runs through corporate governance: "the
shareholders" is not a single interest, and any study treating shareholder wealth as
a scalar is aggregating over parties whose incentives point in opposite directions.

The broader ownership-structure literature supplies the mechanism.
@demiguel2004 show, on Spanish data, that firm value is a non-monotonic function of
insider ownership, rising through a convergence-of-interest region where more
ownership better aligns managers with shareholders and falling through an
entrenchment region where controlling insiders are insulated from discipline. The
existence of these regions is what makes ownership a *strategic* variable rather
than an accounting detail: the same governance change has opposite effects depending
on where the firm sits. @kabir1997 add the takeover-defense dimension, showing on
Dutch data that anti-takeover devices and ownership structure jointly shape stock
returns; the market for corporate control that @barney1988relatedness treats as the
force competing away synergies can itself be blunted by defenses, and the returns
consequences depend on which parties the defenses protect.

@feldman2019family bring family control into the same frame. Studying the stock
market's reaction to acquisitions and divestitures by family firms, they show that
the market prices family control differently across these two transaction types,
because family owners' distinctive objectives, control retention, long horizons,
socioemotional wealth, cut differently for buying than for selling. Their earlier
work on corporate divestitures and family control [@feldman2016divestitures]
establishes the divestiture half of the pattern. The strategic content is that the
identity of the controlling owner is not a control variable but a determinant of
which deals are done and how they are received.

@arikan2010 study a governance channel specific to young firms. Newly public
acquirers carry affiliations from their IPO, with underwriters and with venture
capitalists, and these affiliations can either certify the acquirer's quality to the
market or create conflicts of interest, since advisors have their own stake in deal
flow. Their finding, that the effect depends on the nature and quality of the
affiliation rather than its mere presence, is the corporate-development version of
the signaling-versus-substance debate that runs through
@sec-ms-finance. Finally, @tang2025 turns to the people who actually run these
processes, showing that dedicated corporate development executives, the in-house
professionals who source and execute deals, are associated with better M&A
performance. This parallels the CMO evidence of @germann2015 exactly: a functional
specialist in the top team is a capability, and demonstrating so requires confronting
the same selection problem, since firms that hire such specialists differ from those
that do not.

### Divestiture, and the buyer of divested assets

Divestiture is the half of corporate development that receives a fraction of the
attention, though @villalonga2005choice already established that it belongs in the
same choice set. The seller-side logic is reasonably well understood: firms divest
to correct prior over-diversification, to fund higher-value uses, or to exit
businesses whose fit has decayed, and the market generally rewards focus-increasing
divestitures.

The buyer side is thinner, and @laamanen2014 is one of the few direct studies.
Examining acquirers of *divested* assets in the U.S. software industry, they show
that buying a unit another firm has decided to sell is a distinct strategic act with
its own performance signature, not merely a small acquisition. The asymmetry is
informational, and reminiscent of @capron2007private. The seller knows why it is
selling, so the buyer faces adverse selection, but the seller is also motivated and
the asset is often available below the price a comparable standalone firm would
command. Whether a buyer profits depends on whether it has a use for the asset that
the seller demonstrably did not, which is once again @barney1988relatedness in a new
setting.

### The entrepreneurial mirror

Acquisition is one way to assemble a resource bundle; founding a firm is the other,
and founding, allying, acquiring, and divesting are four answers to a single
question—which resources should sit inside this firm's boundary. The entrepreneurship
literature is developed on its own terms in @sec-entrepreneurship; two of its results
belong here because they mirror the corporate-development findings above exactly.

@agarwal2026startup examine how founders' *pre-entry knowledge* and *post-entry
learning* shape the distribution of startup performance. Their result mirrors
@rabier2017 with striking precision: pre-entry knowledge does not simply raise
average performance but reshapes the whole distribution, and the effects on the
upper and lower tails differ from each other and from the effect on the mean. A
literature reporting only average returns to founder experience would therefore
understate what founder knowledge does, in exactly the way the acquisition
literature understated what motive does. The theoretical unification is worth
stating plainly: whenever performance depends on combining resources whose fit is
uncertain, the interesting action is in the second moment, because good fit and bad
fit are both amplified by the act of combination.

@howell2026solo show that solo founding is viable under identifiable conditions
rather than uniformly inferior to team founding. For corporate development this
matters directly, because startups are the targets: a solo-founded target
concentrates its critical knowledge in one person whose retention becomes the
integration problem, which is the coordination-autonomy dilemma of
@puranam2006organizing reduced to a single employee.

### Holes in the literature worth filling {#sec-ms-corpdev-gaps}

The seams in this literature are unusually visible, and several of them sit exactly
where marketing has data and constructs that strategy does not.

*The dependent variable is almost never a customer variable.* The performance chain
of @fig-strat-performance-chain has four layers, and corporate development research
lives almost entirely in the top two, announcement returns and accounting profit.
@umashankar2021 is close to alone in putting a customer-mindset outcome on the left
side. Nobody has estimated the *customer-side mediation* of acquisition performance:
if deals fail on average, do they fail because customers defect, and what fraction
of the variance in acquirer returns is attributable to post-deal churn? The data now
exist to answer this, and @cording2008ambiguity have already supplied the mediation
design.

*Motives are inferred, never measured.* @rabier2017 classifies motives from filings
and @bauer2024synergy show that stated synergies are the product of managerial
attention rather than of complete evaluation. The obvious next step, measuring
stated synergy beliefs at scale from deal documents and earnings calls using the
text methods of @sec-text-as-data, and then comparing stated with realized synergies
at the customer level, has not been taken. This would convert motive from a
classification into a measured construct with a prediction error, which is what the
attention-based view actually implies.

*Distributional thinking has not reached the integration literature.* @rabier2017
and @agarwal2026startup have moved from means to distributions, but every major
integration finding, speed [@homburg2006speed], degree and cultural fit
[@bauer2014antecedents], deliberate learning [@zollo2004deliberate], is estimated on
conditional means. The unanswered question is whether fast integration raises the
mean while fattening the left tail, which would reconcile the managerial enthusiasm
for speed with the evidence that it often disappoints. Quantile and distributional
regression on integration attributes is a straightforward and under-exploited design.

*Stakeholder synergy has not been extended to customers.* @odziemkowska2026 develop
stakeholder synergies with non-market stakeholders in view, and @bettinazzi2017
measure stakeholder orientation as a firm attribute. Neither treats the *customer
relationship portfolio* as the stakeholder asset being combined, even though it is
the stakeholder asset marketing can measure best. The customer-equity machinery of
@sec-clv makes this tractable: two firms' customer bases can be overlapping,
complementary, or antagonistic, and the resulting relational synergy should be
estimable from individual-level data rather than assumed.

*Overlapping owners have been studied; overlapping customers have not.*
@goranova2010 showed that owners on both sides of the deal change the price. The
direct analogue, *customers* on both sides of the deal, is unstudied, though it is
plainly consequential: when a large share of the target's customers are already the
acquirer's customers, the deal buys less incremental revenue than the top lines
suggest and raises the risk that a single dissatisfying integration damages two
relationships at once. Customer-base overlap is measurable and belongs in the
synergy calculation.

*Integration speed is identified almost entirely from retrospective surveys.* Both
@homburg2006speed and @bauer2014antecedents rely on managers recalling how fast
integration went and how well it turned out, a design in which speed is endogenous
to expected difficulty and both measures share method variance. Sources of exogenous
variation exist, regulatory review periods that delay integration, staggered
antitrust remedies, mandated hold-separate orders, and the causal-inference toolkit
of @sec-causal-inference is built for exactly this. An identified estimate of the
speed effect would be a genuine contribution to a thirty-year debate.

*The buyer side of divestiture remains nearly empty.* @laamanen2014 stands close to
alone, in one industry. What happens to the *customers* of a divested unit, whether
relationships survive the change of parent, and whether divested-asset buyers select
on customer quality are all open, and all measurable.

*Founding-team structure and target quality have never been connected.* @howell2026solo
establish when solo founding is viable and @agarwal2026startup establish how founder
knowledge shapes the performance distribution, while @puranam2006organizing establish
that autonomy preserves acquired innovation. Whether solo-founded ventures make
systematically better or worse acquisition targets, given that their critical
knowledge is concentrated in a single retainable person, follows immediately from
these three and has not been asked.

*Data assets are the new acquisition rationale, and nobody has priced the constraint.*
An increasing share of deals are justified by the target's customer data and the
models trained on it, yet the privacy regime of @sec-privacy can make the combined
use of two customer databases legally impossible, so the synergy is unrealizable no
matter how well integration is executed. Studying deals whose stated rationale is
data combination, against variation in privacy regulation, would identify a
first-order modern constraint on synergy realization that the existing frameworks
[@feldman2022synergy] do not contain.

## The Frontier and a Research Agenda {#sec-ms-frontier}

The field's own assessments of where it stands converge on a few themes. @varadarajan2010
defines the domain and its foundational premises and, in doing so, exposes how
fragmented the field's empirical base is. @morgan2019research survey the state of
marketing-strategy research and argue that the field has over-studied isolated
marketing-mix decisions and under-studied *integrated strategy*, the pattern that the
opening of this chapter insists is the real object; they call for more work on
strategy *content* (the substance of what firms decide), strategy *process* (how
decisions get made and implemented), and strategy *implementation*, and for designs
that take causal identification seriously.

Three further fronts are active. First, *empirical generalizations*: meta-analyses
now pin down replication-grade facts, advertising elasticities [@sethuraman2011],
the market-orientation to performance link [@kirca2005], marketing's elasticity on
firm value [@edeling2016], so that strategy debates can be settled with accumulated
evidence rather than single studies. Second, *integrating logics*: service-dominant
logic reframes value as co-created with customers rather than embedded in products
[@vargo2004], a lens that, depending on one's view, either adds explanatory power or
relabels existing constructs. Third, the *emerging modules* the field is racing to
absorb: digital and platform strategy, privacy as strategy, AI as both a marketing
capability and a competitive disruptor, ESG and firm value, and causal machine
learning for heterogeneous strategic effects. Each follows the same template this
chapter has modeled, a substantive construct, an estimator, and an explicit set of
identifying assumptions.

The through-line for a doctoral reader is that marketing strategy is where the field
is held accountable. Its constructs are interesting, but they earn their place only
when forced into contact with outcome data under a credible design. The two
simulations in this chapter, market orientation moderated by competition, and a
marketing asset priced into firm value, are deliberately small instances of that
discipline: each states a data-generating process, then shows what an honest
estimator can and cannot recover from it.

## Key Takeaways

- Marketing strategy is the firm-level, integrated pattern of choices through which
  a firm creates, captures, and sustains customer value, judged by customer,
  financial, stock-market, and societal performance [@varadarajan2010; @morgan2018].
  Strategy is distinguished from tactics by reversibility and scope, not by
  importance.
- *Market orientation* is the foundational implementation construct, conceived
  behaviorally as intelligence generation, dissemination, and responsiveness
  [@kohli1990] and culturally as customer, competitor, and interfunctional
  orientation [@narver1990]. It pays on average but is necessary, not sufficient,
  and its payoff is moderated by competition and context [@kirca2005].
- Sustainable advantage rests on VRIN resources [@wernerfelt1984] and the
  evolutionary, disequilibrium competition of resource-advantage theory [@hunt1995];
  marketing capabilities [@day1994; @vorhies2005] convert orientation into position
  and outweigh R&D and operations capabilities in their performance impact
  [@krasnikov2008], while dynamic capabilities [@teece1997; @day2011] keep advantage
  alive under change.
- Positioning (STP and differentiation) must separate the *sources*, *positions*,
  and *performance outcomes* of advantage [@daywensley1988], and a defensible
  position is anchored to a capability rivals cannot quickly match.
- The marketing-finance interface is the accountability turn: marketing assets are
  off-balance-sheet drivers of firm value [@Srivastava_1998; @srinivasan2009], and
  credibly identified panels price brand and customer equity into Tobin's $q$
  [@edeling2016; @kumar2009expanding]. Cross-references: @sec-marketing-finance,
  @sec-brand-equity, @sec-clv.
- Marketing's organizational influence depends on accountability, innovativeness, and
  customer connection [@verhoef2009influence]; marketing excellence rests on four
  elements and seven activities [@moorman2016]; and the CMO's value, a roughly 15%
  Tobin's $q$ premium, is a finding only because its design confronts endogeneity
  [@germann2015].
- Corporate development (@sec-ms-corpdev) is a chain, not a decision: firms choose a
  mode [@villalonga2005choice; @wang2007alliance; @capron2007private], form a motive,
  design an integration, and distribute the proceeds under a governance structure.
  Bidders profit only from synergies that are *unique* and privately known, because
  competition transfers common synergies to the target through the premium
  [@barney1988relatedness]. Averages hide the action: post-acquisition performance has
  no reliable mean effect [@king2004meta], while motive reshapes the whole
  *distribution* [@rabier2017], the same distributional logic that governs new-venture
  performance [@agarwal2026startup].
- Integration is a decision variable, not a virtue. Speed pays under high external
  relatedness and hurts under high internal relatedness [@homburg2006speed];
  complementarity and cultural fit set the ceiling that degree and speed then reach or
  miss [@bauer2014antecedents; @chatterjee1992culture]; codified experience, not raw
  experience, builds integration capability [@zollo2004deliberate]; and autonomy must
  be traded against coordination when the acquired asset is innovation
  [@puranam2006organizing]. Marketing integration is the channel through which much of
  this reaches customers [@homburg2005marketing], who tend to lose out
  [@umashankar2021].
- The relationships that make a target valuable are held by counterparties who can
  leave: stakeholder synergies are a distinct and VRIN-passing value source
  [@odziemkowska2026; @bettinazzi2017], and a target's collaborators act on their
  reading of the acquirer's motive [@piezunka2026]. Who captures the value depends on
  ownership: overlapping institutional owners raise premiums [@goranova2010], insider
  ownership affects value non-monotonically [@demiguel2004], takeover defenses blunt
  the market for corporate control [@kabir1997], and family control prices buying and
  selling differently [@feldman2019family].
- The agenda [@morgan2019research; @morgan2012performance] is to study *integrated*
  strategy rather than isolated tactics, to accumulate empirical generalizations
  [@sethuraman2011], and to absorb digital, platform, privacy, and AI strategy under
  the same discipline of construct, estimator, and identifying assumption.
